### 2014-04-30 doc/publications/incob2014/paper.tex
start

%% BioMed_Central_Tex_Template_v1.06
%%                                      %
%  bmc_article.tex            ver: 1.06 %
%                                       %

%%IMPORTANT: do not delete the first line of this template
%%It must be present to enable the BMC Submission system to
%%recognise this template!!

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                     %%
%%  LaTeX template for BioMed Central  %%
%%     journal article submissions     %%
%%                                     %%
%%          <8 June 2012>              %%
%%                                     %%
%%                                     %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                                                 %%
%% For instructions on how to fill out this Tex template           %%
%% document please refer to Readme.html and the instructions for   %%
%% authors page on the biomed central website                      %%
%% http://www.biomedcentral.com/info/authors/                      %%
%%                                                                 %%
%% Please do not use \input{...} to include other tex files.       %%
%% Submit your LaTeX manuscript as one .tex document.              %%
%%                                                                 %%
%% All additional figures and files should be attached             %%
%% separately and not embedded in the \TeX\ document itself.       %%
%%                                                                 %%
%% BioMed Central currently use the MikTex distribution of         %%
%% TeX for Windows) of TeX and LaTeX.  This is available from      %%
%% http://www.miktex.org                                           %%
%%                                                                 %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%% additional documentclass options:
%  [doublespacing]
%  [linenumbers]   - put the line numbers on margins

%%% loading packages, author definitions

%\documentclass[twocolumn]{bmcart}% uncomment this for twocolumn layout and comment line below
\documentclass{bmcart}

%%% Load packages
%\usepackage{amsthm,amsmath}
%\RequirePackage{natbib}
%\RequirePackage{hyperref}
\usepackage[utf8]{inputenc} %unicode support
%\usepackage[applemac]{inputenc} %applemac support if unicode package fails
%\usepackage[latin1]{inputenc} %UNIX support if unicode package fails

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                             %%
%%  If you wish to display your graphics for   %%
%%  your own use using includegraphic or       %%
%%  includegraphics, then comment out the      %%
%%  following two lines of code.               %%
%%  NB: These line *must* be included when     %%
%%  submitting to BMC.                         %%
%%  All figure files must be submitted as      %%
%%  separate graphics through the BMC          %%
%%  submission process, not included in the    %%
%%  submitted article.                         %%
%%                                             %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\def\includegraphic{}
\def\includegraphics{}

%%% Put your definitions there:
\startlocaldefs
\endlocaldefs

%%% Begin ...
\begin{document}

%%% Start of article front matter
\begin{frontmatter}

\begin{fmbox}
\dochead{Research}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% Enter the title of your article here     %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\title{A sample article title}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% Enter the authors here                   %%
%%                                          %%
%% Specify information, if available,       %%
%% in the form:                             %%
%%   <key>={<id1>,<id2>}                    %%
%%   <key>=                                 %%
%% Comment or delete the keys which are     %%
%% not used. Repeat \author command as much %%
%% as required.                             %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\author[
   addressref={aff1},                   % id's of addresses, e.g. {aff1,aff2}
   corref={aff1},                       % id of corresponding address, if any
   noteref={n1},                        % id's of article notes, if any
   email={jane.e.doe@cambridge.co.uk}   % email address
]{\inits{JE}\fnm{Jane E} \snm{Doe}}
\author[
   addressref={aff1,aff2},
   email={john.RS.Smith@cambridge.co.uk}
]{\inits{JRS}\fnm{John RS} \snm{Smith}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% Enter the authors' addresses here        %%
%%                                          %%
%% Repeat \address commands as much as      %%
%% required.                                %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\address[id=aff1]{%                           % unique id
  \orgname{Department of Zoology, Cambridge}, % university, etc
  \street{Waterloo Road},                     %
  %\postcode{}                                % post or zip code
  \city{London},                              % city
  \cny{UK}                                    % country
}
\address[id=aff2]{%
  \orgname{Marine Ecology Department, Institute of Marine Sciences Kiel},
  \street{D\"{u}sternbrooker Weg 20},
  \postcode{24105}
  \city{Kiel},
  \cny{Germany}
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% Enter short notes here                   %%
%%                                          %%
%% Short notes will be after addresses      %%
%% on first page.                           %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{artnotes}
%\note{Sample of title note}     % note to the article
\note[id=n1]{Equal contributor} % note, connected to author
\end{artnotes}

\end{fmbox}% comment this for two column layout

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% The Abstract begins here                 %%
%%                                          %%
%% Please refer to the Instructions for     %%
%% authors on http://www.biomedcentral.com  %%
%% and include the section headings         %%
%% accordingly for your article type.       %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{abstractbox}

\begin{abstract} % abstract
\parttitle{First part title} %if any
Text for this section.

\parttitle{Second part title} %if any
Text for this section.
\end{abstract}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% The keywords begin here                  %%
%%                                          %%
%% Put each keyword in separate \kwd{}.     %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{keyword}
\kwd{sample}
\kwd{article}
\kwd{author}
\end{keyword}

% MSC classifications codes, if any
%\begin{keyword}[class=AMS]
%\kwd[Primary ]{}
%\kwd{}
%\kwd[; secondary ]{}
%\end{keyword}

\end{abstractbox}
%
%\end{fmbox}% uncomment this for twcolumn layout

\end{frontmatter}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% The Main Body begins here                %%
%%                                          %%
%% Please refer to the instructions for     %%
%% authors on:                              %%
%% http://www.biomedcentral.com/info/authors%%
%% and include the section headings         %%
%% accordingly for your article type.       %%
%%                                          %%
%% See the Results and Discussion section   %%
%% for details on how to create sub-sections%%
%%                                          %%
%% use \cite{...} to cite references        %%
%%  \cite{koon} and                         %%
%%  \cite{oreg,khar,zvai,xjon,schn,pond}    %%
%%  \nocite{smith,marg,hunn,advi,koha,mouse}%%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%% start of article main body
% <put your article body there>

%%%%%%%%%%%%%%%%
%% Background %%
%%
\section*{Content}
Text and results for this section, as per the individual journal's instructions for authors. %\cite{koon,oreg,khar,zvai,xjon,schn,pond,smith,marg,hunn,advi,koha,mouse}

\section*{Background}
Text for this section.

\section*{Runtime}
Text for this section.

\section*{Accuracy}
Text for this section.

\section*{Real World Example}
Text for this section.

K-means clustering 
We Cluster samples from a VCF file containing the variants from chromosome one. The content of the VCF file represents
Mahout provides libraries for clustering data, 'sequences', although the data must first be converted from the VCF format to
sequence files. The sequence files should consist of key-value pairs, where
Using a custom Java program, we transpose the

\subsection*{Sub-heading for section}
Text for this sub-heading \ldots
\subsubsection*{Sub-sub heading for section}
Text for this sub-sub-heading \ldots
\paragraph*{Sub-sub-sub heading for section}
Text for this sub-sub-sub-headi	ng \ldots
In this section we examine the growth rate of the mean of $Z_0$, $Z_1$ and $Z_2$. In
addition, we examine a common modeling assumption and note the
importance of considering the tails of the extinction time $T_x$ in
studies of escape dynamics.
We will first consider the expected resistant population at $vT_x$ for
some $v>0$, (and temporarily assume $\alpha=0$)
%

%
If we assume that sensitive cells follow a deterministic decay
$Z_0(t)=xe^{\lambda_0 t}$ and approximate their extinction time as
$T_x\approx-\frac{1}{\lambda_0}\log x$, then we can heuristically
estimate the expected value as
%

%
Thus we observe that this expected value is finite for all $v>0$ (also see \cite{koon,khar,zvai,xjon,marg}).

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% Backmatter begins here                   %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{backmatter}

\section*{Competing interests}
  The authors declare that they have no competing interests.

\section*{Author's contributions}
    Text for this section \ldots

\section*{Acknowledgements}
  Text for this section \ldots
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                  The Bibliography                       %%
%%                                                         %%
%%  Bmc_mathpys.bst  will be used to                       %%
%%  create a .BBL file for submission.                     %%
%%  After submission of the .TEX file,                     %%
%%  you will be prompted to submit your .BBL file.         %%
%%                                                         %%
%%                                                         %%
%%  Note that the displayed Bibliography will not          %%
%%  necessarily be rendered by Latex exactly as specified  %%
%%  in the online Instructions for Authors.                %%
%%                                                         %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% if your bibliography is in bibtex format, use those commands:
\bibliographystyle{bmc-mathphys} % Style BST file
\bibliography{bmc_article}      % Bibliography file (usually '*.bib' )

% or include bibliography directly:
% \begin{thebibliography}
% \bibitem{b1}
% \end{thebibliography}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                               %%
%% Figures                       %%
%%                               %%
%% NB: this is for captions and  %%
%% Titles. All graphics must be  %%
%% submitted separately and NOT  %%
%% included in the Tex document  %%
%%                               %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%
%% Do not use \listoffigures as most will included as separate files

\section*{Figures}
  \begin{figure}[h!]
  \caption{\csentence{Sample figure title.}
      A short description of the figure content
      should go here.}
      \end{figure}

\begin{figure}[h!]
  \caption{\csentence{Sample figure title.}
      Figure legend text.}
      \end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                               %%
%% Tables                        %%
%%                               %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%% Use of \listoftables is discouraged.
%%
\section*{Tables}
\begin{table}[h!]
\caption{Sample table title. This is where the description of the table should go.}
      \begin{tabular}{cccc}
        \hline
           & B1  &B2   & B3\\ \hline
        A1 & 0.1 & 0.2 & 0.3\\
        A2 & ... & ..  & .\\
        A3 & ..  & .   & .\\ \hline
      \end{tabular}
\end{table}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                               %%
%% Additional Files              %%
%%                               %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section*{Additional Files}
  \subsection*{Additional file 1 --- Sample additional file title}
    Additional file descriptions text (including details of how to
    view the file, if it is in a non-standard format or the file extension).  This might
    refer to a multi-page table or a figure.

  \subsection*{Additional file 2 --- Sample additional file title}
    Additional file descriptions text.

\end{backmatter}
\end{document}

### 2014-05-01 doc/publications/incob2014/paper.tex
initial setup for a methodology paper in BMC Genomics

@@
 
-%\documentclass[twocolumn]{bmcart}% uncomment this for twocolumn layout and comment line below
-\documentclass{bmcart}
+\documentclass[twocolumn]{bmcart}% uncomment this for twocolumn layout and comment line below
+%\documentclass{bmcart}
 
@@
 \begin{fmbox}
-\dochead{Research}
+\dochead{Methodology}
 
@@
 
-\title{A sample article title}
+\title{Scalable Clustering of Genotype Information using MapReduce}
 
@@
 \author[
-   addressref={aff1},                   % id's of addresses, e.g. {aff1,aff2}
+   addressref={aff1,aff2},                   % id's of addresses, e.g. {aff1,aff2}
+%   noteref={n1},                        % id's of article notes, if any
+   email={Aidan.O'Brien@csiro.au}   % email address
+]{\inits{AR}\fnm{Aidan R} \snm{O'Brien}}
+\author[
+   addressref={aff2},
+   email={f.buske@garvan.org.au}
+]{\inits{FA}\fnm{Fabian A} \snm{Buske}}
+\author[
+   addressref={aff1},
    corref={aff1},                       % id of corresponding address, if any
-   noteref={n1},                        % id's of article notes, if any
-   email={jane.e.doe@cambridge.co.uk}   % email address
-]{\inits{JE}\fnm{Jane E} \snm{Doe}}
-\author[
-   addressref={aff1,aff2},
-   email={john.RS.Smith@cambridge.co.uk}
-]{\inits{JRS}\fnm{John RS} \snm{Smith}}
+   email={Denis.Bauer@CSIRO.au}
+]{\inits{DC}\fnm{Denis C} \snm{Bauer}}
 
@@
 \address[id=aff1]{%                           % unique id
-  \orgname{Department of Zoology, Cambridge}, % university, etc
-  \street{Waterloo Road},                     %
-  %\postcode{}                                % post or zip code
-  \city{London},                              % city
-  \cny{UK}                                    % country
+  \orgname{Division of Computational Informatics, CSIRO}, % university, etc
+  \street{11 Julius Av},                     %
+  \postcode{2113}                                % post or zip code
+  \city{Sydney},                              % city
+  \cny{Australia}                                    % country
 }
 \address[id=aff2]{%
-  \orgname{Marine Ecology Department, Institute of Marine Sciences Kiel},
-  \street{D\"{u}sternbrooker Weg 20},
-  \postcode{24105}
-  \city{Kiel},
-  \cny{Germany}
+  \orgname{School of Biomedical Sciences and Pharmacy, Faculty of Health},
+%  \street{},
+  \postcode{2308}
+  \city{Newcastle},
+  \cny{Australia}
+}
+\address[id=aff3]{%
+  \orgname{Cancer Epigenetics Program, Cancer Research Division, Kinghorn Cancer Centre, Garvan Institute of Medical Research},
+  \street{384 Victoria St},
+  \postcode{2010}
+  \city{Sydney},
+  \cny{Australia}
 }
@@
 
-\begin{artnotes}
+%\begin{artnotes}
 %\note{Sample of title note}     % note to the article
-\note[id=n1]{Equal contributor} % note, connected to author
-\end{artnotes}
+%\note[id=n1]{Equal contributor} % note, connected to author
+%\end{artnotes}
 
-\end{fmbox}% comment this for two column layout
+%\end{fmbox}% comment this for two column layout
 
@@
 \begin{abstract} % abstract
-\parttitle{First part title} %if any
-Text for this section.
-
-\parttitle{Second part title} %if any
-Text for this section.
+\parttitle{Background}
+\parttitle{Results}
+\parttitle{Conclusions}
 \end{abstract}
@@
 \begin{keyword}
-\kwd{sample}
-\kwd{article}
-\kwd{author}
+\kwd{Hadoop}
+\kwd{Mahout}
+\kwd{Genotype}
+\kwd{Clustering}
+\kwd{MapReduce}
 \end{keyword}
@@
 %
-%\end{fmbox}% uncomment this for twcolumn layout
+\end{fmbox}% uncomment this for twcolumn layout
 
@@
 %%
-\section*{Content}
-Text and results for this section, as per the individual journal's instructions for authors. %\cite{koon,oreg,khar,zvai,xjon,schn,pond,smith,marg,hunn,advi,koha,mouse}
-
 \section*{Background}
@@
 
-\section*{Runtime}
+\section*{Results and discussion}
+\subsection*{Runtime}
 Text for this section.
 
-\section*{Accuracy}
+\subsection*{Accuracy}
 Text for this section.
 
-\section*{Real World Example}
+\subsection*{Real World Example}
 Text for this section.
 
-K-means clustering 
+\section*{Conclusion}
+
+\section*{Methods}
+\subsection*{K-means clustering}
 We Cluster samples from a VCF file containing the variants from chromosome one. The content of the VCF file represents
@@
 Using a custom Java program, we transpose the
-
-\subsection*{Sub-heading for section}
-Text for this sub-heading \ldots
-\subsubsection*{Sub-sub heading for section}
-Text for this sub-sub-heading \ldots
-\paragraph*{Sub-sub-sub heading for section}
-Text for this sub-sub-sub-headi	ng \ldots
-In this section we examine the growth rate of the mean of $Z_0$, $Z_1$ and $Z_2$. In
-addition, we examine a common modeling assumption and note the
-importance of considering the tails of the extinction time $T_x$ in
-studies of escape dynamics.
-We will first consider the expected resistant population at $vT_x$ for
-some $v>0$, (and temporarily assume $\alpha=0$)
-%
-
-%
-If we assume that sensitive cells follow a deterministic decay
-$Z_0(t)=xe^{\lambda_0 t}$ and approximate their extinction time as
-$T_x\approx-\frac{1}{\lambda_0}\log x$, then we can heuristically
-estimate the expected value as
-%
-
-%
-Thus we observe that this expected value is finite for all $v>0$ (also see \cite{koon,khar,zvai,xjon,marg}).
 
@@
 \begin{backmatter}
+
+\section*{Availability of supporting data}
+\section*{List of abbreviations}
 
@@
 \bibliographystyle{bmc-mathphys} % Style BST file
-\bibliography{bmc_article}      % Bibliography file (usually '*.bib' )
+\bibliography{genotypeClustering}      % Bibliography file (usually '*.bib' )
 

### 2014-05-01 doc/publications/incob2014/paper.tex
a bit more content

@@
 %\RequirePackage{natbib}
-%\RequirePackage{hyperref}
+\RequirePackage{hyperref}
 \usepackage[utf8]{inputenc} %unicode support
@@
 \section*{Background}
-Text for this section.
+Grouping individuals based on the genomic profile is a commonly performed tasks to identify population structure~\cite{Gao2007} or elucidate different haplotype involvement in diseases suceptibility~\cite{Laitman2013}.
+Traditionally both the number of individuals and included genotypes were relatively small and libraries in Bioconductor sufficient.
+However, recent technological advances in whole genome sequencing instruments have made population-scale sequencing feasible.
+It is hence economical to generate studies with sample sizes currently reserved for larger consortia such as the 1000 genomes project~\cite{1KG2012} or TCGA~\cite{TCGA2013}.
+At the same time, whole genome sequencing will enable the inclusion of rare or even somatic mutations in the analysis.
+This drastic increase in both sample numbers and feature per sample requires a massively parallele approach to data processing.
+
+As a result of these big data challenges, MapReduce approaches are increasingly being used in bioinformatics (for reviews see~cite{Almeida2013,Zou2013, Qiu2010,Taylor2010}).
+Specifically, for sequence analysis tools for read mapping~\cite{Schatz2009}, duplicate removal~\cite{Jourdren2012}, and variant calling~\cite{Langmead2009, McKenna2010} were released over recent years.
+
+%SeqWare~{OConnor2010} Libraries \cite{Doering2008,Schumacher2014,Nordberg2013}
+
+GWAS\cite{Huang2013, Guo2014}
+
+\url{https://mahout.apache.org/}
 
@@
 \section*{Conclusion}
+Note,  Domain-specific language for linear algebraic operations on top of Apache Spark \url{http://spark.apache.org/}
+ADAM~\cite{Massie2013}
 
 \section*{Methods}
+\subsection*{Datasets}
+Overview see Table~\ref{datasets}
+
 \subsection*{K-means clustering}
@@
 \begin{table}[h!]
-\caption{Sample table title. This is where the description of the table should go.}
-      \begin{tabular}{cccc}
+\caption{Datasets used in this study.}
+      \begin{tabular}{lcc}
         \hline
-           & B1  &B2   & B3\\ \hline
-        A1 & 0.1 & 0.2 & 0.3\\
-        A2 & ... & ..  & .\\
-        A3 & ..  & .   & .\\ \hline
+           & individuals  & variants  \\ \hline
+        Chr1 subset & & \\
+        1000genomes trios & & \\
+        hapmap & ... & ..  \\ \hline
       \end{tabular}
+      \label{datasets}
 \end{table}

### 2014-05-01 doc/publications/incob2014/paper.tex
intro

@@
 As a result of these big data challenges, MapReduce approaches are increasingly being used in bioinformatics (for reviews see~cite{Almeida2013,Zou2013, Qiu2010,Taylor2010}).
-Specifically, for sequence analysis tools for read mapping~\cite{Schatz2009}, duplicate removal~\cite{Jourdren2012}, and variant calling~\cite{Langmead2009, McKenna2010} were released over recent years.
+This is especially the case for sequence analysis tasks, such as read mapping~\cite{Schatz2009}, duplicate removal~\cite{Jourdren2012}, and variant calling~\cite{Langmead2009, McKenna2010} or Genome Wide Analysis Study based tasks~\cite{Huang2013, Guo2014}.
+
+At the same time, sophisticated machine learning methodologies are increasingly adapted to utilise MapReduce paradigms overcoming callings that arise due to their iterative nature~\cite{Chu2009}.
+Specifically, the Mahout project\url{https://mahout.apache.org/} has been developed extensively~\cite{Ranger2007, Owen2011} and was successfully applied in the clinical informatics space~\cite{Dong2013}.
+
+We therefore link the two areas by providing an interface between Mahout and the standard variant data format, VCF~\cite{1KG2012}, which opens up the application of Mahout's different machine learning algorithms to be applied to genotype-based tasks.
+To demonstrate the capability we cluster variant datasets from the 1000 genomes project to determine population structure as well as ancestry using different algorithms available in Mahout.
+In the first two sections we benchmark the performance and accuracy of Mahout's implementations against standard R based implementations on a reduced version of the data and in the last section we investigate the performance on the full dataset, demonstrating how seamless Mahout can transit to population-scale tasks.
 
 %SeqWare~{OConnor2010} Libraries \cite{Doering2008,Schumacher2014,Nordberg2013}
-
-GWAS\cite{Huang2013, Guo2014}
-
-\url{https://mahout.apache.org/}
 
@@
 \subsection*{Datasets}
+Trio data from the 1000 genomes project -> determine in the clustering can identify ancestry 
+Sequence data from hapmap samples -> determine ethnic groups
+Scaled down version of the hap map data (Chr1) for R to run on
 Overview see Table~\ref{datasets}

### 2014-05-01 doc/publications/incob2014/paper.tex
Added a couple of tables templates

@@
            & individuals  & variants  \\ \hline
-        Chr1 subset & & \\
-        1000genomes trios & & \\
-        hapmap & ... & ..  \\ \hline
+        Chr1 subset & 1024 & 3,007,196 \\
+        1000genomes trios & 1024 & ...\\
+        hapmap & ... & ...  \\ \hline
+      \end{tabular}
+      \label{datasets}
+\end{table}
+
+\begin{table}[h!]
+\caption{Adjusted Rand index (accuracy).}
+      \begin{tabular}{lccc}
+        \hline
+           & Chr1 subset & 1000genomes trios  & hapmap \\ \hline
+        Mahout, k=4 & 0.664 & ..&..\\
+        Mahout, k=10 & 0.714 & ..&..\\
+        Mahout, k=20 & 0.769 & .. & ..\\ 
+        Mahout, k=30 & 0.728 & .. & ..\\ 
+        R, k=4 & .. & - & -\\
+        R, k=10 & .. & - & -\\
+        R, k=20 & .. & - & -\\ 
+        R, k=30 & .. & - & -\\ \hline
+      \end{tabular}
+      \label{datasets}
+\end{table}
+
+\begin{table}[h!]
+\caption{Time taken to cluster samples (k=20).}
+      \begin{tabular}{lccc}
+        \hline
+           & Chr1 subset & 1000genomes trios  & hapmap \\ \hline
+        Mahout, 1 node & .. & ..&..\\
+        Mahout, 2 nodes & .. & ..&..\\
+        Mahout, 3 nodes & .. & .. & ..\\ 
+        Mahout, 4 nodes & .. & .. & ..\\ 
+        R & .. & - & -\\ \hline
       \end{tabular}

### 2014-05-02 doc/publications/incob2014/paper.tex
abstract

@@
 \author[
-   addressref={aff2},
+   addressref={aff3, aff4},
    email={f.buske@garvan.org.au}
@@
 }
+\address[id=aff4]{%
+  \orgname{UNSW Medicine, University of New South Wales},
+%  \street{},
+  \postcode{2052}
+  \city{Sydney},
+  \cny{Australia}
+}
 
@@
 \begin{abstract} % abstract
-\parttitle{Background}
-\parttitle{Results}
-\parttitle{Conclusions}
+\parttitle{Background} Processing genomic information from whole genome sequence studies pose computational challenges due to the unprecedented data volume generated, which render transitional approaches insufficient.
+However, by utilising advancements in modern hardware accelerators and data processing we can provide the means for scalable solutions.
+We therefore aim to provide the interface between standard genomic data formats and advanced and scalable analysis libraries like Mahout.
+\parttitle{Results} We achieve an XX-fold speedup by using the scalable k-means MapReduce implementation over the equivalent analysis performed in R and YY-fold speedup for PCA, respectively, by comparable accuracy.
+Due to memory in the R-Implementation the comparison had to be limited to variant data from chromosome 1 only, however we ran the full dataset (aa GB, bb individuals {\`a} cc variants ) through Mahout demonstrating the scalability.
+\parttitle{Conclusions} Using modern compute paradigms is essential to scale to modern genomic research in an efficient sustainable way. 
 \end{abstract}
@@
 \section*{Background}
-Grouping individuals based on the genomic profile is a commonly performed tasks to identify population structure~\cite{Gao2007} or elucidate different haplotype involvement in diseases suceptibility~\cite{Laitman2013}.
-Traditionally both the number of individuals and included genotypes were relatively small and libraries in Bioconductor sufficient.
-However, recent technological advances in whole genome sequencing instruments have made population-scale sequencing feasible.
+Grouping individuals based on the genomic profile is a commonly performed tasks to identify population structure~\cite{Gao2007} or elucidate different haplotype involvement in diseases susceptibility~\cite{Laitman2013}.
+Traditionally both the number of individuals and included genotypes, typically from SNP arrays, were relatively small and libraries in Bioconductor sufficient.
+However, recent technological advances in whole genome sequencing have made population-scale sequencing feasible.
 It is hence economical to generate studies with sample sizes currently reserved for larger consortia such as the 1000 genomes project~\cite{1KG2012} or TCGA~\cite{TCGA2013}.
-At the same time, whole genome sequencing will enable the inclusion of rare or even somatic mutations in the analysis.
-This drastic increase in both sample numbers and feature per sample requires a massively parallele approach to data processing.
+At the same time, whole genome sequencing enables the inclusion of rare or even somatic mutations in the analysis, increasing the feature space by orders of magnitude.
+This drastic increase in both sample numbers and features per sample requires a massively parallele approach to data processing.
 
-As a result of these big data challenges, MapReduce approaches are increasingly being used in bioinformatics (for reviews see~cite{Almeida2013,Zou2013, Qiu2010,Taylor2010}).
+As a result of these big data challenges, MapReduce approaches are increasingly being used in bioinformatics (for reviews see~cite{Zou2013, Qiu2010,Taylor2010}).
 This is especially the case for sequence analysis tasks, such as read mapping~\cite{Schatz2009}, duplicate removal~\cite{Jourdren2012}, and variant calling~\cite{Langmead2009, McKenna2010} or Genome Wide Analysis Study based tasks~\cite{Huang2013, Guo2014}.

### 2014-05-02 vcf-clusterer/src/net/aidan/kmeansClusterer.java
Add java source

@@ -1,0 +1,301 @@
+/**
+ * 
+ */
+/**
+ * @author obr17q
+ *
+ */
+package net.aidan;
+
+import java.io.BufferedReader;
+import java.io.BufferedWriter;
+import java.io.File;
+import java.io.FileReader;
+import java.io.FileWriter;
+import java.io.IOException;
+import java.util.Arrays;
+
+import org.apache.hadoop.conf.Configuration;
+import org.apache.hadoop.conf.Configured;
+import org.apache.hadoop.fs.FileSystem;
+import org.apache.hadoop.fs.Path;
+import org.apache.hadoop.io.SequenceFile;
+import org.apache.hadoop.io.Text;
+import org.apache.hadoop.util.Tool;
+import org.apache.hadoop.util.ToolRunner;
+import org.apache.mahout.math.DenseVector;
+import org.apache.mahout.math.NamedVector;
+import org.apache.mahout.math.SequentialAccessSparseVector;
+import org.apache.mahout.math.VectorWritable;
+import org.apache.mahout.utils.vectors.RowIdJob;
+import org.apache.mahout.clustering.canopy.CanopyDriver;
+import org.apache.mahout.clustering.kmeans.Kluster;
+import org.apache.mahout.clustering.kmeans.KMeansDriver;
+import org.apache.mahout.clustering.kmeans.RandomSeedGenerator;
+import org.apache.mahout.clustering.streaming.mapreduce.StreamingKMeansDriver;
+import org.apache.mahout.common.distance.EuclideanDistanceMeasure;
+import org.apache.mahout.common.distance.ManhattanDistanceMeasure;
+
+public class kmeansClusterer extends Configured implements Tool {
+	public kmeansClusterer() {}
+	
+	// Some variables
+	public static final String INPUT_FILE =  "file.vcf";           
+	public static final String OUTPUT_DIRECTORY = "clustering";
+	public static final boolean SINGLE_MACHINE = false;
+	public static final int k = 10;
+	public static long end = 0;
+	public int run(String[] args) throws Exception {
+		long start = System.currentTimeMillis();
+		
+		
+		Configuration conf = super.getConf();
+		
+		/*
+		 * This adds the configuration files to 'conf'. Shouldn't have to do this.
+		 * Stopped loading them automatically when I switched to Hadoop2.2
[…]

### 2014-05-02 vcf-clusterer/src/net/aidan/outputFormatter.java

@@ -1,0 +1,77 @@
+package net.aidan;
+
+import java.io.BufferedWriter;
+import java.io.File;
+import java.io.FileWriter;
+import java.io.PrintWriter;
+import java.io.StringWriter;
+import java.util.LinkedHashMap;
+import java.util.Map;
+
+import org.apache.hadoop.conf.Configuration;
+import org.apache.hadoop.fs.FileStatus;
+import org.apache.hadoop.fs.FileSystem;
+import org.apache.hadoop.fs.Path;
+import org.apache.hadoop.io.IntWritable;
+import org.apache.hadoop.io.SequenceFile;
+import org.apache.mahout.clustering.Cluster;
+import org.apache.mahout.clustering.classify.WeightedPropertyVectorWritable;
+import org.apache.mahout.math.NamedVector;
+
+public class outputFormatter {
+	String outputFile = "resultFileCluster.txt";
+	LinkedHashMap<String,String> mpIdData = new LinkedHashMap<String, String>();
+	
+	public void fileRead(Configuration conf) {
+		System.out.println("Writing summary to "+ outputFile);
+		try {
+			FileSystem fs = FileSystem.get(conf);
+			Path outPath = new Path("clustering/output/" + Cluster.CLUSTERED_POINTS_DIR + "/");
+			FileStatus[] status = fs.listStatus(outPath);
+			
+			IntWritable key = new IntWritable();
+			WeightedPropertyVectorWritable value = new WeightedPropertyVectorWritable();
+			
+			for (FileStatus s: status) {
+				Path newPath = s.getPath();
+				if (!newPath.toString().contains("part-m-")){
+					System.out.println(newPath);
+					continue;
+				}
+				SequenceFile.Reader reader = new SequenceFile.Reader(fs, newPath, conf);
+				while (reader.next(key, value)) {
+					NamedVector vect = (NamedVector) value.getVector();
+					String vecName = vect.getName();
+					String clusterName = key.toString();
+
+					//Create a new element or add to existing element.
+					if (mpIdData.containsKey(clusterName)) {
+						String val = mpIdData.get(clusterName) + "," + vecName;
+						mpIdData.put(clusterName, val);
+					} else {
+						mpIdData.put(clusterName, vecName);
+					}				
+				}
+				reader.close();
+			}
+			FileWriter f1 = new FileWriter(outputFile,true);
+			BufferedWriter out = new BufferedWriter(f1);
[…]

### 2014-05-02 vcf-clusterer/src/net/aidan/pedigreeMatcher.java

@@ -1,0 +1,105 @@
+package net.aidan;
+
+import java.io.BufferedReader;
+import java.io.BufferedWriter;
+import java.io.FileNotFoundException;
+import java.io.FileReader;
+import java.io.FileWriter;
+import java.io.IOException;
+import java.util.HashMap;
+import java.util.Iterator;
+import java.util.Map;
+import java.util.Set;
+
+public class pedigreeMatcher {
+	HashMap<String, String[]> hm = new HashMap<String, String[]>();
+	public pedigreeMatcher(String pedFile) throws IOException {
+		BufferedReader ins =
+				new BufferedReader(new FileReader(pedFile));
+		String line = null;
+		while ((line = ins.readLine()) != null) {
+			String[] splitted = line.split("	");
+			hm.put(splitted[1],splitted);
+		}
+		ins.close();
+	}
+	
+	//Why??
+	public void printPedigree() {
+		Iterator i = hm.entrySet().iterator();
+		while(i.hasNext()) {
+	         Map.Entry me = (Map.Entry)i.next();
+	         System.out.print(me.getKey() + ": ");
+	         for (String s:(String[])me.getValue()) {
+	        	 System.out.print(s+ " ");
+	         }
+	         System.out.print("\n");
+	      }
+	}
+	
+	public void findMatches(String clusteredPoints) throws IOException {
+		BufferedWriter w = new BufferedWriter(new FileWriter("populations.txt"));
+		BufferedReader ins =
+				new BufferedReader(new FileReader(clusteredPoints));
+		String line = null;
+		while ((line = ins.readLine()) != null) {
+			String[] samples = line.split(": ");
+			w.write(line.split(":")[1] + ": ");
+			String[] sampleArray = samples[2].split(",");
+			for (String s:sampleArray){
+				w.write(hm.get(s)[6]+",");				
+			}
+			w.write("\n");
+		}
+		w.close();
+		ins.close();
+	}
+	
+	public void adjRandIndex(String clusteredPoints) throws IOException {
+		Iterator i = hm.entrySet().iterator();
+		String truth = "truth=[";
+		String clust = "clust=[";
+		
+		
+		BufferedReader ins =
+				new BufferedReader(new FileReader("populations.txt"));
+		String line = null;
+		int clusterId = 0;
+		while ((line = ins.readLine()) != null) {
+			
[…]

### 2014-05-02 vcf-clusterer/vcf-clusterer.sh

@@ -1,0 +1,185 @@
+#!/bin/bash
+#PBS -N hadoop_job
+#PBS -l nodes=2:ppn=12,vmem=48GB
+#PBS -l walltime=60:00
+#PBS -j oe
+
+### load modules to set the appropriate variables
+# especially HADOOP_DATA_DIR and HADOOP_LOG_DIR
+module load jdk/1.7.0_11 hadoop/2.2.0 hpchadoop openmpi
+# avoid some warnings... HADOOP_ROOT is used by ASC hpcHadoop instead
+test -n "$HADOOP_HOME" && unset HADOOP_HOME
+
+
+
+###
+# Classpaths for jars that are distributed to nodes
+export LIBJARS=/home/cci/obr17q/mahout-distribution-0.9/mahout-core-0.9.jar,/home/cci/obr17q/mahout-distribution-0.9/mahout-math-0.9.jar,/home/cci/obr17q/mahout-distribution-0.9/mahout-integration-0.9.jar,/home/cci/obr17q/mahout-distribution-0.9/lib/commons-cli-2.0-mahout.jar,/home/cci/obr17q/mahout-distribution-0.9/lib/commons-collections-3.2.1.jar,/home/cci/obr17q/mahout-distribution-0.9/lib/guava-16.0.jar,/home/cci/obr17q/mahout-distribution-0.9/lib/commons-math3-3.2.jar
+export HADOOP_CLASSPATH=/home/cci/obr17q/mahout-distribution-0.9/mahout-core-0.9.jar:/home/cci/obr17q/mahout-distribution-0.9/mahout-math-0.9.jar:/home/cci/obr17q/mahout-distribution-0.9/mahout-integration-0.9.jar:/home/cci/obr17q/mahout-distribution-0.9/lib/commons-cli-2.0-mahout.jar:/home/cci/obr17q/mahout-distribution-0.9/lib/commons-collections-3.2.1.jar:/home/cci/obr17q/mahout-distribution-0.9/lib/guava-16.0.jar:/home/cci/obr17q/mahout-distribution-0.9/lib/commons-math3-3.2.jar
+
+
+
+### These may not actually be required..
+# I added them during some trial and error
+export HADOOP_HOME=/apps/hadoop/2.2.0
+export HADOOP_MAPRED_HOME=/apps/hadoop/2.2.0
+export HADOOP_COMMON_HOME=/apps/hadoop/2.2.0
[…]

### 2014-05-14 doc/publications/incob2014/paper.tex
added image

@@
   \begin{figure}[h!]
-  \caption{\csentence{Sample figure title.}
-      A short description of the figure content
-      should go here.}
+  \caption{\csentence{Illustration of Mahout-based clustering of genotypes.}
+      The image shows how the here introduced interface converts the vcf file to a mahout usable vector-based data type.
+Based on k-means it illustrates how the Map-step procedure groups of vectors by their nearest cluster center and the Reduce-step averages the vectors in each group to find the new value of the updated cluster center.
+The mahout-produced output is then converted into a visualisation.}
       \end{figure}

### 2014-05-14 doc/publications/incob2014/paper.tex
Updates to methods

@@
 \subsection*{Datasets}
-Trio data from the 1000 genomes project -> determine in the clustering can identify ancestry 
+For k-means clustering of individuals, we used variant call format (VCF) files from the 1000 Genomes Project.
+These VCF files contain information about the genetic variants between each of the 1092 individuals and a reference genome
+derived from GRCh37. Each of the VCF files are partitioned by chromosome.
+We used the VCF file containing chromosome 1 variants (3,007,196 variants) as a subset of the entire human variant dataset.
+For the entire dataset, we used each of the chromosomal VCF files. Metadata available from the 1000 Genomes Consortium
+specifies additional attributes for each individual, including population and familial data.
+Table~\ref{datasets} contains an overview of the datasets.
+
+> determine in the clustering can identify ancestry 
 Sequence data from hapmap samples -> determine ethnic groups
-Scaled down version of the hap map data (Chr1) for R to run on
 Overview see Table~\ref{datasets}
 
-\subsection*{K-means clustering}
+\subsection*{Converting to arrays}
+Mahout requires sequence files as input to its clustering algorithms and provides libraries for writing vectors to these files.
+These libraries require a feature vector and optionally a sample name. The vectors must contain an integer for each element.
+Therefore, we need to preprocess the data from the VCF files into a suitable format. The variant data values, which are tab separated,
+must be converted to an integer. For this, we simply determine the Hamming distance between each tab separated value, and the reference genome.
+Therefore, a value identical to the reference will therefore be 0, a heterozygous variant will be 1 and a homozygous variant will be 2.
+
+\subsection*{Mahout K-means clustering}
 We Cluster samples from a VCF file containing the variants from chromosome one. The content of the VCF file represents
+
 Mahout provides libraries for clustering data, 'sequences', although the data must first be converted from the VCF format to
@@
 Using a custom Java program, we transpose the
+
+\subsection*{Clustering quality}
+We scored the clusters with the Adjusted Rand index. This index compares two different clusterings and assigns a value based
+on their similarity. The value is zero for independent clusterings and one for identical clusterings. We compared the mahout k-means
+clusters and the R k-means clusters to the known clusters based on meta data from the 1000 genome project. For the known clusters,
+individuals were partitioned by their super population code (ASN, EUR, AFR, AMR). We also compared the mahout clusters to the
+R clusters for the Chr1 subset data.
+
+\subsection*{Speed}
 

### 2014-05-15 doc/publications/incob2014/paper.tex
small changes to the method section

@@
 
-\def\includegraphic{}
-\def\includegraphics{}
+%\def\includegraphic{}
+%\def\includegraphics{}
+\usepackage{graphicx}
 
@@
 \endlocaldefs
+
+% Our variable names
+\newcommand{\chrOnePop}{Chr1\_Population}
+\newcommand{\FullPop}{Full\_Population}
+\newcommand{\FullFam}{Chr1\_Family}
 
@@
 \subsection*{Datasets}
-For k-means clustering of individuals, we used variant call format (VCF) files from the 1000 Genomes Project.
-These VCF files contain information about the genetic variants between each of the 1092 individuals and a reference genome
-derived from GRCh37. Each of the VCF files are partitioned by chromosome.
-We used the VCF file containing chromosome 1 variants (3,007,196 variants) as a subset of the entire human variant dataset.
-For the entire dataset, we used each of the chromosomal VCF files. Metadata available from the 1000 Genomes Consortium
-specifies additional attributes for each individual, including population and familial data.
+We used variant call format (VCF) files from the 1000 Genomes Project.
+These VCF files contain information about the genetic variants between each of the 1092 individuals and a reference genome derived from GRCh37. 
+Each of the VCF files are partitioned by chromosome.
+Metadata available from the 1000 Genomes Consortium specifies additional attributes for each individual, including population and familial data.
+To be able to compare with R, we created a subset of the entire human variant dataset by using only variants from chromosome 1 (\chrOnePop) where we investigate the correct grouping of the population structure.
+The population structure is also investigated using the full dataset, i.e. all chromosomes, using Mahout libraries (\FullPop). 
+In addition, we also investigate whether the full dataset allows the accurate prediction of the more detailed family association (\FullFam).
 Table~\ref{datasets} contains an overview of the datasets.
-
-> determine in the clustering can identify ancestry 
-Sequence data from hapmap samples -> determine ethnic groups
-Overview see Table~\ref{datasets}
 
@@
 \subsection*{Clustering quality}
-We scored the clusters with the Adjusted Rand index. This index compares two different clusterings and assigns a value based
-on their similarity. The value is zero for independent clusterings and one for identical clusterings. We compared the mahout k-means
+We scored the clusters with the Adjusted Rand index. 
+%TODO: can you provide the formular
+This index compares two different clusterings and assigns a value based on their similarity.
+The value is zero for independent clusterings and one for identical clusterings.
+We compared the Mahout k-means
 clusters and the R k-means clusters to the known clusters based on meta data from the 1000 genome project. For the known clusters,
-individuals were partitioned by their super population code (ASN, EUR, AFR, AMR). We also compared the mahout clusters to the
+individuals were partitioned by their super population code (ASN, EUR, AFR, AMR). 
+%TODO: "also on \chrOnePop" on what dataset did you compare the R performance otherwise ?
+We also compared the Mahout clusters to the
 R clusters for the Chr1 subset data.
@@
   \begin{figure}[h!]
+  \includegraphics[type=pdf,ext=.pdf,read=.pdf, scale=0.40]{images/signature}
   \caption{\csentence{Illustration of Mahout-based clustering of genotypes.}
@@
            & individuals  & variants  \\ \hline
-        Chr1 subset & 1024 & 3,007,196 \\
-        1000genomes trios & 1024 & ...\\
-        hapmap & ... & ...  \\ \hline
+        \chrOnePop & 1024 & 3,007,196 \\
+        \FullPop\ &  & \\
+        \FullFam\ & 1024 & ...\\
       \end{tabular}
@@
       \end{tabular}
-      \label{datasets}
+      \label{datasetsAcc}
 \end{table}
@@
       \end{tabular}
-      \label{datasets}
+      \label{datasetsTime}
 \end{table}

### 2014-05-15 doc/publications/incob2014/paper.tex
pre-processing section

--- 2014-05-14 doc/publications/incob2014/paper.tex
@@
 
-\subsection*{Converting to arrays}
-Mahout requires sequence files as input to its clustering algorithms and provides libraries for writing vectors to these files.
-These libraries require a feature vector and optionally a sample name. The vectors must contain an integer for each element.
-Therefore, we need to preprocess the data from the VCF files into a suitable format. The variant data values, which are tab separated,
-must be converted to an integer. For this, we simply determine the Hamming distance between each tab separated value, and the reference genome.
-Therefore, a value identical to the reference will therefore be 0, a heterozygous variant will be 1 and a homozygous variant will be 2.
+\subsection*{Pre-processing}
+Mahout requires sequence files as input to its clustering algorithms, and it provides libraries for writing vectors to this file format.
+Separate vectors for each sample, in this case individual, contain an entry for each of the sample's features, where each entry is
+a double. Although samples can subsequently be appended to sequence files, the entire feature-set for each sample must be
+appended in a single write call. Because the features for each sample are present in columns as tab-separated elements in VCF
+files, we recursively transpose the data into CSV files, where each comma-separated row contains the feature-set for a sample
+as well as the sample identity. Concurrently, we convert the values from the VCF file to doubles, as required for Mahout's sequence
+file format and the many algorithms it can serve as input for. For this process, we simply determine the Hamming distance between
+a variant and the reference genome. Therefore, a value identical to the reference will be 0, a heterozygous variant on either strand
+will be 1 and a homozygous variant will be 2.
+At this stage, the CSV files from more than one VCF file are concatenated into the one file. This resulting CSV file, or potential multiple
+CSV files for additional individuals, serve as the input for a custom Java function. This performs line by line, extracting the sample name
+and feature-set from the CSV file and calls a Mahout function to store this data as a named \texttt{DenseVector} object. The vectors are
+populated primarily by zeroes,  we convert this object to a \texttt{SequentialAccessSparseVector} which stores only non-zero doubles
+in one array, and maps their position in a second array. This object is serialized, and the resulting file, or files, serve as input
+to the majority of Mahout's algorithms.
 

### 2014-05-16 doc/publications/incob2014/paper.tex
Few changes

--- 2014-05-15 doc/publications/incob2014/paper.tex
@@
 \subsection*{Runtime}
-Text for this section.
+We measured the runtime seperately for pre-processing stage as well as the k-means clustering stage.
 
@@
 \subsection*{Mahout K-means clustering}
-We Cluster samples from a VCF file containing the variants from chromosome one. The content of the VCF file represents
-
-Mahout provides libraries for clustering data, 'sequences', although the data must first be converted from the VCF format to
-sequence files. The sequence files should consist of key-value pairs, where
-Using a custom Java program, we transpose the
+To cluster the samples from the sequence files, we use Mahout 0.9 in conjunction with Hadoop 2.2.0. 
+The clustering steps are the same for each sample, however, due to the increased feature size we designate more memory to the
+YARN containers. For the first step, we specify \texttt{k}, the number of clusters. The function \texttt{RandomSeedGenerator.buildRandom}
+creates a new sequence file containing \texttt{k} random centroids from the samples sequence file. The sequence file
+of samples and the sequence file of centroids serve as the input to \texttt{KMeansDriver.run}, a static Mahout API function that launches
+a k-means clustering job on the Hadoop service.
 
@@
 \subsection*{Speed}
+We measured the speed for various stages of the processing and clustering. We used
 

### 2014-05-16 doc/publications/incob2014/paper.tex
changes from other computer

@@
 These VCF files contain information about the genetic variants between each of the 1092 individuals and a reference genome derived from GRCh37. 
-Each of the VCF files are partitioned by chromosome.
-Metadata available from the 1000 Genomes Consortium specifies additional attributes for each individual, including population and familial data.
+Each of the VCF files are partitioned by chromosome.\\
+Metadata available from the 1000 Genomes Consortium specifies additional attributes for each individual, including population and familial data.\\
 To be able to compare with R, we created a subset of the entire human variant dataset by using only variants from chromosome 1 (\chrOnePop) where we investigate the correct grouping of the population structure.
@@
 Mahout requires sequence files as input to its clustering algorithms, and it provides libraries for writing vectors to this file format.
-Separate vectors for each sample, in this case individual, contain an entry for each of the sample's features, where each entry is
-a double. Although samples can subsequently be appended to sequence files, the entire feature-set for each sample must be
-appended in a single write call. Because the features for each sample are present in columns as tab-separated elements in VCF
-files, we recursively transpose the data into CSV files, where each comma-separated row contains the feature-set for a sample
+Separate vectors for each sample, in this case, an individual, contain entries for each of the sample's features, where each entry is
+a \texttt{double}. Although additional samples can subsequently be appended to sequence files, the entire feature-set for each sample must be
+appended in a single write operation. Because the features for each sample are present in columns, as tab-separated elements in VCF
+files, we recursively transpose the data into CSV files. Each comma-separated row contains the feature-set for a sample,
 as well as the sample identity. Concurrently, we convert the values from the VCF file to doubles, as required for Mahout's sequence
-file format and the many algorithms it can serve as input for. For this process, we simply determine the Hamming distance between
-a variant and the reference genome. Therefore, a value identical to the reference will be 0, a heterozygous variant on either strand
+file format and the many algorithms this format can serve as input for. We set the value of the double to the Hamming distance between
+each variant and the reference genome. Therefore, a value identical to the reference will be 0, a heterozygous variant on either strand
 will be 1 and a homozygous variant will be 2.
-At this stage, the CSV files from more than one VCF file are concatenated into the one file. This resulting CSV file, or potential multiple
-CSV files for additional individuals, serve as the input for a custom Java function. This performs line by line, extracting the sample name
-and feature-set from the CSV file and calls a Mahout function to store this data as a named \texttt{DenseVector} object. The vectors are
-populated primarily by zeroes,  we convert this object to a \texttt{SequentialAccessSparseVector} which stores only non-zero doubles
-in one array, and maps their position in a second array. This object is serialized, and the resulting file, or files, serve as input
-to the majority of Mahout's algorithms.
+At this stage, the CSV files from more than one VCF file are concatenated into the one file. This resulting CSV file serves as the input for
+a custom Java function. This iterates over the CSV file, extracting the sample name and feature-set from each line, and calls a Mahout
+function to store this data as a named \texttt{DenseVector} object. The vectors are populated primarily by zeroes, so we convert this object to 
+the more efficient \texttt{SequentialAccessSparseVector} object, which stores only non-zero doubles in an array, and maps their position in a second array.
+
+This object is serialized, and the resulting file can serve as input to the majority of Mahout's algorithms.
 
@@
 To cluster the samples from the sequence files, we use Mahout 0.9 in conjunction with Hadoop 2.2.0. 
-The clustering steps are the same for each sample, however, due to the increased feature size we designate more memory to the
-YARN containers. For the first step, we specify \texttt{k}, the number of clusters. The function \texttt{RandomSeedGenerator.buildRandom}
-creates a new sequence file containing \texttt{k} random centroids from the samples sequence file. The sequence file
+The clustering steps are the same for each dataset, however, due to the increased feature size we designate more memory to the
+YARN containers (a container being the resources allocated to each map or reduce iteration). For the first step, we specify \texttt{k},
+the number of clusters. The function \texttt{RandomSeedGenerator.buildRandom} creates a new sequence file containing \texttt{k}
+random centroids from the samples sequence file. The sequence file
 of samples and the sequence file of centroids serve as the input to \texttt{KMeansDriver.run}, a static Mahout API function that launches
@@
         \chrOnePop & 1024 & 3,007,196 \\
-        \FullPop\ &  & \\
+        \FullPop\ & 1024 & 38,219,238 \\
         \FullFam\ & 1024 & ...\\

### 2014-05-16 doc/publications/incob2014/paper.tex
Signed-off-by: Aidan O'Brien <ayduno@gmail.com>

--- 2014-05-16 doc/publications/incob2014/paper.tex
@@
 We measured the runtime seperately for pre-processing stage as well as the k-means clustering stage. 
+Pre-processing the data to Mahout's sequence files took approximately 21 minutes (1255s) and used approximately 1.8GB (1831160kb) physical memory (pmem) and 7.6GB (7646232kb) virtual memory (vmem).
+Conversly,  pre-processing
+the data to R's in memory matrix took approximately 56 minutes (3341s) and used approximately 17.7GB (17696940kb) pmem and 18GB (18000196kb) vmem.
 
@@
 Table~\ref{datasets} contains an overview of the datasets.
+
+\subsection*{Hardware}
+We processed the data using a Portable Batch System (PBS) cluster. We allocated one node and 64GB memory for the comparison between Mahout and R.
+Each node has eight CPU cores, however both applications run a single thread for pre-processing the data.
+For the (\FullPop) dataset, we ran the job on a PBS cluster of x nodes, with twelve CPU cores and 48GB of memory per node.
 
@@
 a k-means clustering job on the Hadoop service.
+
+\subsection*{R K-means clustering}
+We use 'readVcf' from the R library, 'VariantAnnotation', to iteratively read in the VCF file. For each iteration, we convert the data to a matrix and
+change the values for each variant to the Hamming distance, as mentioned previously. We then transpose the matrix, so individuals are represented
+by rows, and append the matrix to that from the previous iteration. Once the matrix contains all of the variants from the VCF file, we cluster the data
+using the 'kmeans' algorithm from the 'stats' library.
 

### 2014-05-17 vcf-clusterer/src/net/aidan/kmeansClusterer.java
Updated

@@ -50,7 +50,6 @@ import org.apache.mahout.common.distance.EuclideanDistanceMeasure;
 import org.apache.mahout.common.distance.ManhattanDistanceMeasure;
 import org.apache.mahout.math.hadoop.stochasticsvd.SSVDHelper;
 
-import au.com.bytecode.opencsv.CSVReader;
 public class kmeansClusterer extends Configured implements Tool {
 	public kmeansClusterer() {}
 	
@@ -88,27 +87,28 @@ public class kmeansClusterer extends Configured implements Tool {
 		
 		
 		//Transpose a VCF file to CSV file(s)
-		transposeVCF();
-		mergeCSV();
+		//transposeVCF();
+		//mergeCSV();
 		
 		//Submits a job to Hadoop to read in the CSV file(s) and convert them to a Mahout sequence file
-		csvToSequences(conf, fs);
+		//csvToSequences(conf, fs);
 		
 		//Probably redundant because of next line (Or could be used for non-random centroids)
 		//chooseInitialCentroids(conf, fs);
 		
 		//Create k centroids at random from the Mahout sequence file
 		Path centroids = RandomSeedGenerator.buildRandom(conf,new Path(OUTPUT_DIRECTORY + "/points/file1"),new Path(OUTPUT_DIRECTORY + "/clusters"),k,new EuclideanDistanceMeasure());
-		end = System.currentTimeMillis();
-		System.out.println("Time taken: " + (end - start) + " ms");
+		//end = System.currentTimeMillis();
+		//System.out.println("Time taken: " + (end - start) + " ms");
 		//Submit the k-means clustering job to the cluster
 		System.out.println("Launching KMeansDriver");
 		KMeansDriver.run(conf, new Path(OUTPUT_DIRECTORY + "/points"),
-				//new Path(OUTPUT_DIRECTORY + "/clusters"),
+		     	//new Path(OUTPUT_DIRECTORY + "/clusters"),
 				centroids,
 				new Path(OUTPUT_DIRECTORY + "/output"), 0.001, 20, true, 0.001, SINGLE_MACHINE);
 
-		
+		end = System.currentTimeMillis();
+		System.out.println("Time taken: " + (end - start) + " ms");
[…]

### 2014-05-17 vcf-clusterer/src/net/aidan/kmeansClusterer.java
changes

@@ -88,24 +88,24 @@ public class kmeansClusterer extends Configured implements Tool {
 		
 		
 		//Transpose a VCF file to CSV file(s)
-		transposeVCF();
-		mergeCSV();
+		//transposeVCF();
+		//mergeCSV();
 		
 		//Submits a job to Hadoop to read in the CSV file(s) and convert them to a Mahout sequence file
-		csvToSequences(conf, fs);
+		//csvToSequences(conf, fs);
 		
 		//Probably redundant because of next line (Or could be used for non-random centroids)
-		//chooseInitialCentroids(conf, fs);
+		chooseInitialCentroids(conf, fs);
 		
 		//Create k centroids at random from the Mahout sequence file
-		Path centroids = RandomSeedGenerator.buildRandom(conf,new Path(OUTPUT_DIRECTORY + "/points/file1"),new Path(OUTPUT_DIRECTORY + "/clusters"),k,new EuclideanDistanceMeasure());
-		end = System.currentTimeMillis();
-		System.out.println("Time taken: " + (end - start) + " ms");
+		//Path centroids = RandomSeedGenerator.buildRandom(conf,new Path(OUTPUT_DIRECTORY + "/points/file1"),new Path(OUTPUT_DIRECTORY + "/clusters"),k,new EuclideanDistanceMeasure());
+		//end = System.currentTimeMillis();
+		//System.out.println("Time taken: " + (end - start) + " ms");
 		//Submit the k-means clustering job to the cluster
-		System.out.println("Launching KMeansDriver");
+		//System.out.println("Launching KMeansDriver");
 		KMeansDriver.run(conf, new Path(OUTPUT_DIRECTORY + "/points"),
-				//new Path(OUTPUT_DIRECTORY + "/clusters"),
-				centroids,
+				new Path(OUTPUT_DIRECTORY + "/clusters"),
+		//		centroids,
 				new Path(OUTPUT_DIRECTORY + "/output"), 0.001, 20, true, 0.001, SINGLE_MACHINE);
 
 		

### 2014-05-17 doc/publications/incob2014/paper.tex
compare

@@
 We measured the runtime seperately for pre-processing stage as well as the k-means clustering stage. 
-Pre-processing the data to Mahout's sequence files took approximately 21 minutes (1255s) and used approximately 1.8GB (1831160kb) physical memory (pmem) and 7.6GB (7646232kb) virtual memory (vmem).
-Conversly,  pre-processing
-the data to R's in memory matrix took approximately 56 minutes (3341s) and used approximately 17.7GB (17696940kb) pmem and 18GB (18000196kb) vmem.
+Following are the results from the (\chrOnePop) dataset. Pre-processing the data to Mahout's sequence files took approximately 21 minutes (1255s) and used
+approximately 1.8GB (1831160kb) physical memory (pmem) and 7.6GB (7646232kb) virtual memory (vmem).
+Conversly,  pre-processing the data to R's in memory matrix took approximately 56 minutes (3341s) and used
+approximately 17.7GB (17696940kb) pmem and 18GB (18000196kb) vmem. The size of the Mahout sequence file,
+which is saved to disk, is 835MB. K-means clustering the data, however, was faster with R, possibly due to the data already being present in memory.
+R took 7s, whereas Mahout took 10 minutes.\\
+Although the Mahout pipeline was about 25 minutes faster than R's.
+the clustering step with R was much faster than with Mahout.
+This is likely due to the additional overhead involved in setting up the Hadoop nodes, spawning the map and reduce instances, and distributing the data between the instances.
+R on the other hand, already had the data in memory, and only had to process that data.
+In the case of this comparison, the data set is likely too small to truly reap the benefit of Hadoop's
+MapReduce algorithm. The main benefit of mahout in this scenario is the memory consumption. There is nearly a 10-fold increase for Mahout
+compared to R.
 

### 2014-05-17 doc/publications/incob2014/paper.tex
addition

@@
 % Our variable names
-\newcommand{\chrOnePop}{Chr1\_Population}
+\newcommand{\chrOnePop}{Chr19\_Population}
 \newcommand{\FullPop}{Full\_Population}
@@
 \subsection*{Accuracy}
-Text for this section.
+For this example, we knew the correct clusters from 1000 Genomes metadata. This metadata assigns a population and family value to each of the 1092 individuals.
+With the foreknowledge of the correct clusters, the Adjusted Rand Index enables us to assign a score to cluster accuracy~\cite{Hubert1985}.
 
@@
 \subsection*{R K-means clustering}
-We use 'readVcf' from the R library, 'VariantAnnotation', to iteratively read in the VCF file. For each iteration, we convert the data to a matrix and
+We use `readVcf' from the R library, `VariantAnnotation', to iteratively read in the VCF file. For each iteration, we convert the data to a matrix and
 change the values for each variant to the Hamming distance, as mentioned previously. We then transpose the matrix, so individuals are represented
 by rows, and append the matrix to that from the previous iteration. Once the matrix contains all of the variants from the VCF file, we cluster the data
-using the 'kmeans' algorithm from the 'stats' library.
+using the `kmeans' algorithm from the `stats' library.
 
@@
 %TODO: can you provide the formular
-This index compares two different clusterings and assigns a value based on their similarity.
+This index compares two different cluster sets and assigns a value based on their similarity.
 The value is zero for independent clusterings and one for identical clusterings.
@@
 
-\subsection*{Speed}
-We measured the speed for various stages of the processing and clustering. We used
+\subsection*{Performance}
+We recorded the memory usage and run-time from the PBS logs. For the Mahout jobs, we queued separate jobs
+for pre-processing and k-means clustering to get an accurate representation of the time and memory requirements for each stage.
+For the R job, where we could not easily separate the pre-processing and clustering components, we printed the time between
+stages using the function, \texttt{proc.time()}.
 
@@
            & individuals  & variants  \\ \hline
-        \chrOnePop & 1024 & 3,007,196 \\
-        \FullPop\ & 1024 & 38,219,238 \\
-        \FullFam\ & 1024 & ...\\
+        \chrOnePop & 1092 & 816,115 \\
+        \FullPop\ & 1092 & 38,219,238 \\
+        \FullFam\ & 1092 & ...\\
       \end{tabular}
@@
 \caption{Adjusted Rand index (accuracy).}
-      \begin{tabular}{lccc}
+      \begin{tabular}{lcccc}
         \hline
-           & Chr1 subset & 1000genomes trios  & hapmap \\ \hline
+           & \chrOnePop & \FullPop  & hapmap \\ \hline
         Mahout, k=4 & 0.664 & ..&..\\
@@
 \begin{table}[h!]
-\caption{Time taken to cluster samples (k=20).}
-      \begin{tabular}{lccc}
+\caption{Resources consumed.}
+      \begin{tabular}{lccccc}
       \hline
-           & Chr1 subset & 1000genomes trios  & hapmap \\ \hline
-        Mahout, 1 node & .. & ..&..\\
-        Mahout, 2 nodes & .. & ..&..\\
-        Mahout, 3 nodes & .. & .. & ..\\ 
-        Mahout, 4 nodes & .. & .. & ..\\ 
-        R & .. & - & -\\ \hline
+      &&\multicolumn{2}{ c }{Memory (kb)}&\multicolumn{2}{ c }{Time (s)} \\
+        \hline
+           & Nodes & pmem & vmem  & pre-processing & clustering \\ \hline
+        R, \chrOnePop & 1 & 17696940 & 18000196 &3341 & 7\\
+        Mahout,\chrOnePop & 1 & 1831160 & 7646232 & 1255 & 594\\
+        Mahout, \FullPop & .. &.. & .. & ..&..\\ 
+        Mahout, 4 nodes & .. & .. & .. & ..&..\\ 
+        \hline
       \end{tabular}

### 2014-05-17 doc/publications/incob2014/paper.tex
section and dataset rewritten

@@
 \newcommand{\FullFam}{Chr1\_Family}
+
+\newcommand{\OneReal}{OneChr\_1K}
+\newcommand{\SixReal}{SixChr\_1K}
+\newcommand{\FullReal}{Full\_1K}
+\newcommand{\SixArtiExact}{SixChr\_10K\_0}
+\newcommand{\SixArtiNoise}{SixChr\_100K\_10}
 
@@
 \section*{Results and discussion}
-\subsection*{Runtime}
-We measured the runtime seperately for pre-processing stage as well as the k-means clustering stage. 
-Following are the results from the (\chrOnePop) dataset. Pre-processing the data to Mahout's sequence files took approximately 21 minutes (1255s) and used
-approximately 1.8GB (1831160kb) physical memory (pmem) and 7.6GB (7646232kb) virtual memory (vmem).
-Conversly,  pre-processing the data to R's in memory matrix took approximately 56 minutes (3341s) and used
-approximately 17.7GB (17696940kb) pmem and 18GB (18000196kb) vmem. The size of the Mahout sequence file,
-which is saved to disk, is 835MB. K-means clustering the data, however, was faster with R, possibly due to the data already being present in memory.
-R took 7s, whereas Mahout took 10 minutes.\\
-Although the Mahout pipeline was about 25 minutes faster than R's.
-the clustering step with R was much faster than with Mahout.
+
+% COMPARISON TO R
+\subsection*{Clustering in Mahout uses less memory than in R}
+%TODO: I would drop the numbers in seconds and kb -- just have that in the table...
+In this section, we compare the runtime between the k-means implementation in R and Mahout for the small \OneReal\ dataset. 
+To do this we measured the runtime separately for pre-processing stage as well as the k-means clustering stage itself (see Methods~\ref{Sec:preprocessing}). 
+Pre-processing the data to Mahout's sequence files took 21 minutes (1255s), used 1.8GB (1831160kb) physical memory (pmem) and 7.6GB (7646232kb) virtual memory (vmem).
+Conversely,  pre-processing the data to R's in memory matrix took  56 minutes (3341s), used 17.7GB (17696940kb) pmem and 18GB (18000196kb) vmem. 
+The size of the Mahout sequence file, which is saved to disk, is 835MB. 
+K-means clustering the data, however, was faster with R, using 7s, whereas Mahout took 10 minutes. See Table~\ref{datasetsTime}. \\
+Due to the small number of classifying variants, cluster performance is very poor for both methods, with the best Adjusted Rand Index score for clustering individuals in one of 4 populations (see Methods) being 0.769 for 20 clusters in Mahout and XX with YY clusters in R (see Table~\ref{datasetsAcc}).
+
+Although the Mahout pipeline was about 25 minutes faster than R's, the actual clustering step took two orders of magnitude longer. 
 This is likely due to the additional overhead involved in setting up the Hadoop nodes, spawning the map and reduce instances, and distributing the data between the instances. 
 R on the other hand, already had the data in memory, and only had to process that data. 
-In the case of this comparison, the data set is likely too small to truly reap the benefit of Hadoop's
-MapReduce algorithm. The main benefit of mahout in this scenario is the memory consumption. There is nearly a 10-fold increase for Mahout
-compared to R.
+In the case of this comparison, the data set is likely too small to truly reap the benefit of Hadoop's MapReduce algorithm. 
+The main benefit of mahout in this scenario is the memory consumption with nearly 10-fold reduction for Mahout compared to R.
 
-\subsection*{Accuracy}
-For this example, we knew the correct clusters from 1000 Genomes metadata. This metadata assigns a population and family value to each of the 1092 individuals.
-With the foreknowledge of the correct clusters, the Adjusted Rand Index enables us to assign a score to cluster accuracy~\cite{Hubert1985}.
+% SCALING UP FEATURES IN MAHOUT
+\subsection*{Mahout scales to 15 Million genomic variants}
+In this section we demonstrate Mahout's scaling ability by increasing the numbers of features, here variants per sample. 
+Including more regions for the similarity calculation will increase the cluster performance, however also increases the memory consumption. 
+R therefore was not able to perform the clustering on even the medium sized dataset (\SixReal , 15 Million variants), we hence only report resource consumption and accuracy for Mahout.
+On this dataset Mahout's prepossessing took XX minutes, with XXGB of physical and XXGB of virtual memory, and the clustering took XX minutes, with XXGB of physical and XXGB of virtual memory. 
+The clustering accuracy has increased by XX-fold to YY with XX clusters compared to the performance on \OneReal\ for the 4-group-based population task (see Table~\ref{datasetsAcc}).
+Using additional variants now also allows the more detailed association to the 855 families.
+The best Adjusted Rand Index score here is XX for XX clusters (see Table~\ref{datasetsAcc}).
 
-\subsection*{Real World Example}
-Text for this section.
+Utilizing the whole dataset hence would have likely increased the performance further.
+However, on \FullReal\ (38 Million variants) Hadoop returns an 'out-of-memory' error on the name-node before the jobs can even be distributed to the mappers.
+%TODO: is this a correct statement?
+It is not possible to use Mahout's standard implementation for this dataset, as the supported Hadoop-setup only allows the communication between nodes with up to 8GB of memory and a scaling beyond that would have been required.
+
+While R does not scale to either of the two orders of magnitude incased datasets, Mahout copes will with up to 15 Million variants.
+Going beyond this would require an implementation that also paralizes the individual variants as memory directly scales with the size of this feature vector.
+
+% SCALING UP SAMPLES IN MAHOUT
+\subsection*{Mahout scales effortless to an order of magnitude more samples}
+In this section we investigate Mahout's ability to scale with more samples.
+The resource consumption remained virtually identical between \SixReal\ and the order of magnitude larger dataset \SixArtiExact, which took XX minutes, with XXGB of physical and XXGB of virtual memory, and the clustering took XX minutes, with XXGB of physical and XXGB of virtual memory.
+The clustering accuracy is YY with XX clusters.
+
+%TODO: can you think of a less general statement as conclusion
+Scaling effortless to massively more samples is the strength of Hadoop, hence compute problems need to be carefully designed to harness this ability.
 
 \section*{Conclusion}
-Note,  Domain-specific language for linear algebraic operations on top of Apache Spark \url{http://spark.apache.org/}
-ADAM~\cite{Massie2013}
+In this paper we have applied Mahout's k-means clustering to the task of genotype based population and family association. 
+We provide an interface from VCF format to Mahout and a toolset to visualise and interpret Mahout's clustering output. 
+We find that on small datasets actual clustering stage is two orders of magnitudes slower in Mahout compared to R (10m vs 7s), likely due to the overhead of setting up and co-ordinating the communication between the Hadoop nodes.
+However, increasing either the number of samples or the number of features per sample quickly exceeds the ability of R.
+Mahout on the other hand scales effortless to an order of magnitude more samples (10,000) and copes well with 15 Million features per samples.
+As memory consumption directly scales with the number of features per sample we observed that Mahout's standard set-up of Hadoop does not scale to a whole exome dataset (38 Million features per sample).
+However, this might be resolved once Mahout moves from MapReduce to a domain-specific language for linear algebraic operations on top of Apache Spark \url{http://spark.apache.org/}.
+
+In conclusion, Mahout offers a scalable out-of-the box solution that is directly applicable to genotype-based clustering for population or family association.
+
+%This efficient a high performance in-memory map-reduce system has already demonstrated to deliver a  50x speed-up on a whole-genome dataset compared to usual multithreaded approaches, see ADAM~\cite{Massie2013}
 
@@
 These VCF files contain information about the genetic variants between each of the 1092 individuals and a reference genome derived from GRCh37. 
-Each of the VCF files are partitioned by chromosome.\\
+Each of the VCF files are partitioned by chromosome.
 Metadata available from the 1000 Genomes Consortium specifies additional attributes for each individual, including population and familial data.\\
-To be able to compare with R, we created a subset of the entire human variant dataset by using only variants from chromosome 1 (\chrOnePop) where we investigate the correct grouping of the population structure.
-The population structure is also investigated using the full dataset, i.e. all chromosomes, using Mahout libraries (\FullPop). 
-In addition, we also investigate whether the full dataset allows the accurate prediction of the more detailed family association (\FullFam).
+To be able to compare with R, we created a subset of the entire human variant dataset by using only variants from chromosome 19 (\OneReal ) where we investigate the correct grouping of the population structure.
+Scaling up using Mahout, we investigate how the accuracy improves when using more variants from the genome with \SixReal\ and \FullReal, which uses chromosome 1-5 in addition to 19 and the full genome, respectively.
+We also investigate whether we can accurately predict the more detailed family association using these datasets. 
+In addition, we demonstrate Mahout's scalability by increasing the number of samples in the study to 10,920 by duplicating the original 1092 samples 10 times with zero and 10\% noise levels, i.e.
+randomly changing 10\% of the genotypes.
 Table~\ref{datasets} contains an overview of the datasets.
@@
 \subsection*{Pre-processing}
+\label{Sec:preprocessing}
 Mahout requires sequence files as input to its clustering algorithms, and it provides libraries for writing vectors to this file format.
@@
 R clusters for the Chr1 subset data.
+With the foreknowledge of the correct clusters, the Adjusted Rand Index enables us to assign a score to cluster accuracy~\cite{Hubert1985}.
 
@@
 
-\begin{figure}[h!]
-  \caption{\csentence{Sample figure title.}
-      Figure legend text.}
-      \end{figure}
+%\begin{figure}[h!]
+%  \caption{\csentence{Sample figure title.}
+ %     Figure legend text.}
+ %     \end{figure}
 
@@
 \caption{Datasets used in this study.}
-      \begin{tabular}{lcc}
+      \begin{tabular}{lrrrrr}
         \hline
-           & individuals  & variants  \\ \hline
-        \chrOnePop & 1092 & 816,115 \\
-        \FullPop\ & 1092 & 38,219,238 \\
-        \FullFam\ & 1092 & ...\\
+           & individuals  & variants  & population & families & noise\\ 
+           & & &groups&& \\ \hline
+        \OneReal & 1,092 & 816,115  & 4 & 855 & NA\\
+        \SixReal & 1,092 & 15,161,339 & 4 & 855 & NA\\
+        \FullReal\ & 1,092 & 38,219,238 & 4 & 855 & NA\\
+	\SixArtiExact\ & 10,920 & 15,161,339 & 4 & 855 & 0 \\ 
+	\SixArtiNoise\ & 10,920 & 15,161,339 & 4 & 855 & 10 \\ 
       \end{tabular}
@@
 
-\section*{Additional Files}
-  \subsection*{Additional file 1 --- Sample additional file title}
-    Additional file descriptions text (including details of how to
-    view the file, if it is in a non-standard format or the file extension).  This might
-    refer to a multi-page table or a figure.
+%\section*{Additional Files}
+%  \subsection*{Additional file 1 --- Sample additional file title}
+%    Additional file descriptions text (including details of how to
+%    view the file, if it is in a non-standard format or the file extension).  This might
+ %   refer to a multi-page table or a figure.
 
-  \subsection*{Additional file 2 --- Sample additional file title}
-    Additional file descriptions text.
+%  \subsection*{Additional file 2 --- Sample additional file title}
+ %   Additional file descriptions text.
 

### 2014-05-17 doc/publications/incob2014/paper.tex
funding and contribution

@@
 \section*{Author's contributions}
-    Text for this section \ldots
+A.R.O implemented the method and performed the results under the supervision of D.C.B. and F.A.B.
+All authors wrote the paper.
 
 \section*{Acknowledgements}
-  Text for this section \ldots
+A.R.O was funded by the NSW Cancer Institute Big Data Big Impact schema, F.A.B by the National Health and Medical Research Council [1051757] and D.C.B by Commonwealth Scientific and Industrial Research Organisation's Transformational Capability Platform, Science and Industry Endowment Fund and Information Management and Technology Services.
+The authors would like to thank Piotr Szul and Gareth Williams for their help with setting up Hadoop on the HPC system.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2014-05-17 doc/publications/incob2014/paper.tex
abbreviations

@@
 However, recent technological advances in whole genome sequencing have made population-scale sequencing feasible.
-It is hence economical to generate studies with sample sizes currently reserved for larger consortia such as the 1000 genomes project~\cite{1KG2012} or TCGA~\cite{TCGA2013}.
+It is hence economical to generate studies with sample sizes currently reserved for larger consortia such as the 1000 genomes project~\cite{1KG2012} or the cancer genome atlas (TCGA)~\cite{TCGA2013}.
 At the same time, whole genome sequencing enables the inclusion of rare or even somatic mutations in the analysis, increasing the feature space by orders of magnitude.
@@
 
-We therefore link the two areas by providing an interface between Mahout and the standard variant data format, VCF~\cite{1KG2012}, which opens up the application of Mahout's different machine learning algorithms to be applied to genotype-based tasks.
+We therefore link the two areas by providing an interface between Mahout and the standard variant data format, variant call format (VCF)~\cite{1KG2012}, which opens up the application of Mahout's different machine learning algorithms to be applied to genotype-based tasks.
 To demonstrate the capability we cluster variant datasets from the 1000 genomes project to determine population structure as well as ancestry using different algorithms available in Mahout.
@@
 \section*{List of abbreviations}
+TCGA -- the cancer genome atlas
+VCF -- variant call format
+PBS -- Portable Batch System
 
@@
   \caption{\csentence{Illustration of Mahout-based clustering of genotypes.}
-      The image shows how the here introduced interface converts the vcf file to a mahout usable vector-based data type.
+      The image shows how the here introduced interface converts the VCF file to a mahout usable vector-based data type.
 Based on k-means it illustrates how the Map-step procedure groups of vectors by their nearest cluster center and the Reduce-step averages the vectors in each group to find the new value of the updated cluster center.

### 2014-05-17 doc/publications/incob2014/paper.tex
small change to intro

@@
 We therefore link the two areas by providing an interface between Mahout and the standard variant data format, variant call format (VCF)~\cite{1KG2012}, which opens up the application of Mahout's different machine learning algorithms to be applied to genotype-based tasks.
-To demonstrate the capability we cluster variant datasets from the 1000 genomes project to determine population structure as well as ancestry using different algorithms available in Mahout.
+To demonstrate the capability we cluster variant datasets from the 1000 genomes project to determine population structure as well as ancestry using the k-means clustering algorithm available in Mahout.
 In the first two sections we benchmark the performance and accuracy of Mahout's implementations against standard R based implementations on a reduced version of the data and in the last section we investigate the performance on the full dataset, demonstrating how seamless Mahout can transit to population-scale tasks.

### 2014-05-17 doc/publications/incob2014/paper.tex
Some changes

@@
 \newcommand{\FullReal}{Full\_1K}
-\newcommand{\SixArtiExact}{SixChr\_10K\_0}
-\newcommand{\SixArtiNoise}{SixChr\_100K\_10}
+\newcommand{\SixArtiExact}{Chr19\_5K\_0}
+\newcommand{\SixArtiNoise}{Chr19\_5K\_10}
 
@@
         \FullReal\ & 1,092 & 38,219,238 & 4 & 855 & NA\\
-	\SixArtiExact\ & 10,920 & 15,161,339 & 4 & 855 & 0 \\ 
-	\SixArtiNoise\ & 10,920 & 15,161,339 & 4 & 855 & 10 \\ 
+	\SixArtiExact\ & 5,460 & 816,115 & 4 & 855 & 0 \\ 
+	\SixArtiNoise\ & 5,460 & 816,115 & 4 & 855 & 10 \\ 
       \end{tabular}
@@
 \caption{Adjusted Rand index (accuracy).}
-      \begin{tabular}{lcccc}
+      \begin{tabular}{lcc}
         \hline
-           & \chrOnePop & \FullPop  & hapmap \\ \hline
-        Mahout, k=4 & 0.664 & ..&..\\
-        Mahout, k=10 & 0.714 & ..&..\\
-        Mahout, k=20 & 0.769 & .. & ..\\ 
-        Mahout, k=30 & 0.728 & .. & ..\\ 
-        R, k=4 & .. & - & -\\
-        R, k=10 & .. & - & -\\
-        R, k=20 & .. & - & -\\ 
-        R, k=30 & .. & - & -\\ \hline
+           & \chrOnePop & \FullPop   \\ \hline
+        Mahout, k=4 & 0.664\\
+        Mahout, k=10 & 0.714 \\
+        Mahout, k=20 & 0.769 \\ 
+        Mahout, k=30 & 0.728  \\ 
+        R, k=4 & .. & - \\
+        R, k=10 & .. & - \\
+        R, k=20 & .. & - \\ 
+        R, k=30 & .. & - \\ \hline
       \end{tabular}
@@
            & Nodes & pmem & vmem  & pre-processing & clustering \\ \hline
-        R, \chrOnePop & 1 & 17696940 & 18000196 &3341 & 7\\
-        Mahout,\chrOnePop & 1 & 1831160 & 7646232 & 1255 & 594\\
-        Mahout, \FullPop & .. &.. & .. & ..&..\\ 
-        Mahout, 4 nodes & .. & .. & .. & ..&..\\ 
+        R, \chrOnePop & 1 & 17,696,940 & 18,000,196 &3,341 & 7\\
+        Mahout,\chrOnePop & 1 & 1,831,160 & 7646232 & 1,255 & 594\\
+        Mahout, \SixReal & 1 & 13,036,836 & 36,542,380 & 19,500 &..\\ 
+        Mahout, \SixArtiExact & 1 & 5,804,660 & 39,075,548 & 2,024 & ..\\ 
         \hline

### 2014-05-17 doc/publications/incob2014/paper.tex
text updates

@@
 % Our variable names
-\newcommand{\chrOnePop}{Chr19\_Population}
-\newcommand{\FullPop}{Full\_Population}
-\newcommand{\FullFam}{Chr1\_Family}
+%\newcommand{\chrOnePop}{Chr19\_Population}
+%\newcommand{\FullPop}{Full\_Population}
+%\newcommand{\FullFam}{Chr1\_Family}
 
@@
 
-As a result of these big data challenges, MapReduce approaches are increasingly being used in bioinformatics (for reviews see~cite{Zou2013, Qiu2010,Taylor2010}).
-This is especially the case for sequence analysis tasks, such as read mapping~\cite{Schatz2009}, duplicate removal~\cite{Jourdren2012}, and variant calling~\cite{Langmead2009, McKenna2010} or Genome Wide Analysis Study based tasks~\cite{Huang2013, Guo2014}.
+As a result of these big data challenges, MapReduce approaches are increasingly being used in bioinformatics (for reviews see~\cite{Zou2013, Qiu2010,Taylor2010}).
+This is especially the case for sequence analysis tasks, such as read mapping~\cite{Schatz2009}, duplicate removal~\cite{Jourdren2012}, and variant calling~\cite{Langmead2009, McKenna2010} as well as Genome Wide Analysis Study based tasks~\cite{Huang2013, Guo2014}.
 
-At the same time, sophisticated machine learning methodologies are increasingly adapted to utilise MapReduce paradigms overcoming callings that arise due to their iterative nature~\cite{Chu2009}.
-Specifically, the Mahout project\url{https://mahout.apache.org/} has been developed extensively~\cite{Ranger2007, Owen2011} and was successfully applied in the clinical informatics space~\cite{Dong2013}.
+At the same time, theoretical computer scientists adapt MapReduce paradigms to cope with the iterative nature of machine learning tasks~\cite{Chu2009}.
+Specifically, the Mahout project \url{https://mahout.apache.org/} has been developed extensively~\cite{Ranger2007, Owen2011} and was successfully applied in the clinical informatics space~\cite{Dong2013}.
 
-We therefore link the two areas by providing an interface between Mahout and the standard variant data format, variant call format (VCF)~\cite{1KG2012}, which opens up the application of Mahout's different machine learning algorithms to be applied to genotype-based tasks.
+In this paper we link the two areas of parallel machine learning and bioinformatics by providing an interface between Mahout and the standard variant data format, variant call format (VCF)~\cite{1KG2012}, which opens up the application of Mahout's different machine learning algorithms to be applied to genotype-based tasks.
+
 To demonstrate the capability we cluster variant datasets from the 1000 genomes project to determine population structure as well as ancestry using the k-means clustering algorithm available in Mahout.
-In the first two sections we benchmark the performance and accuracy of Mahout's implementations against standard R based implementations on a reduced version of the data and in the last section we investigate the performance on the full dataset, demonstrating how seamless Mahout can transit to population-scale tasks.
+In the first section we benchmark the performance and accuracy of Mahout's implementations against a standard R based implementation on a reduced version of the data and in the last two sections we investigate how Mahout scales to either more samples or more features per samples.
 
@@
 \subsection*{Clustering in Mahout uses less memory than in R}
-%TODO: I would drop the numbers in seconds and kb -- just have that in the table...
 In this section, we compare the runtime between the k-means implementation in R and Mahout for the small \OneReal\ dataset. 
-To do this we measured the runtime separately for pre-processing stage as well as the k-means clustering stage itself (see Methods~\ref{Sec:preprocessing}). 
-Pre-processing the data to Mahout's sequence files took 21 minutes (1255s), used 1.8GB (1831160kb) physical memory (pmem) and 7.6GB (7646232kb) virtual memory (vmem).
-Conversely,  pre-processing the data to R's in memory matrix took  56 minutes (3341s), used 17.7GB (17696940kb) pmem and 18GB (18000196kb) vmem. 
+To do this we measured the runtime separately for pre-processing stage as well as the k-means clustering stage itself (see Methods). 
+Pre-processing the data to Mahout's sequence files takes 21 minutes, uses 1.8GB physical memory (pmem) and 7.6GB virtual memory (vmem).
+Conversely,  pre-processing the data to R's in memory matrix takes 56 minutes, uses 17.7GB pmem and 18GB vmem. 
 The size of the Mahout sequence file, which is saved to disk, is 835MB. 
-K-means clustering the data, however, was faster with R, using 7s, whereas Mahout took 10 minutes. See Table~\ref{datasetsTime}. \\
+K-means clustering the data, however, is faster with R, using 7s, whereas Mahout takes 10 minutes. See Table~\ref{datasetsTime}. \\
 Due to the small number of classifying variants, cluster performance is very poor for both methods, with the best Adjusted Rand Index score for clustering individuals in one of 4 populations (see Methods) being 0.769 for 20 clusters in Mahout and XX with YY clusters in R (see Table~\ref{datasetsAcc}).
 
-Although the Mahout pipeline was about 25 minutes faster than R's, the actual clustering step took two orders of magnitude longer. 
+Although the Mahout pipeline is about 25 minutes faster than R's, the actual clustering step takes two orders of magnitude longer. 
 This is likely due to the additional overhead involved in setting up the Hadoop nodes, spawning the map and reduce instances, and distributing the data between the instances. 
@@
 In the case of this comparison, the data set is likely too small to truly reap the benefit of Hadoop's MapReduce algorithm. 
-The main benefit of mahout in this scenario is the memory consumption with nearly 10-fold reduction for Mahout compared to R.
+The main benefit of Mahout in this scenario is the memory consumption with nearly 10-fold reduction for Mahout compared to R.
 
@@
 R therefore was not able to perform the clustering on even the medium sized dataset (\SixReal , 15 Million variants), we hence only report resource consumption and accuracy for Mahout.
-On this dataset Mahout's prepossessing took XX minutes, with XXGB of physical and XXGB of virtual memory, and the clustering took XX minutes, with XXGB of physical and XXGB of virtual memory. 
+On this dataset Mahout's prepossessing takes XX minutes, with XXGB of physical and XXGB of virtual memory, and the clustering takes XX minutes, with XXGB of physical and XXGB of virtual memory. 
 The clustering accuracy has increased by XX-fold to YY with XX clusters compared to the performance on \OneReal\ for the 4-group-based population task (see Table~\ref{datasetsAcc}).
@@
 
-Utilizing the whole dataset hence would have likely increased the performance further.
+Utilizing the whole dataset hence would likely increased the performance further.
 However, on \FullReal\ (38 Million variants) Hadoop returns an 'out-of-memory' error on the name-node before the jobs can even be distributed to the mappers.
@@
 In this section we investigate Mahout's ability to scale with more samples.
-The resource consumption remained virtually identical between \SixReal\ and the order of magnitude larger dataset \SixArtiExact, which took XX minutes, with XXGB of physical and XXGB of virtual memory, and the clustering took XX minutes, with XXGB of physical and XXGB of virtual memory.
+The resource consumption remains virtually identical between \SixReal\ and the order of magnitude larger dataset \SixArtiExact, which takes XX minutes, with XXGB of physical and XXGB of virtual memory, and the clustering takes XX minutes, with XXGB of physical and XXGB of virtual memory.
 The clustering accuracy is YY with XX clusters.
@@
 In this paper we have applied Mahout's k-means clustering to the task of genotype based population and family association. 
-We provide an interface from VCF format to Mahout and a toolset to visualise and interpret Mahout's clustering output. 
-We find that on small datasets actual clustering stage is two orders of magnitudes slower in Mahout compared to R (10m vs 7s), likely due to the overhead of setting up and co-ordinating the communication between the Hadoop nodes.
+We provide an interface from VCF format to Mahout and a toolset to interpret Mahout's clustering output. 
+We find that on small datasets the actual clustering stage is two orders of magnitudes slower in Mahout compared to R (10 minutes vs 7 seconds), likely due to the overhead of setting up and co-ordinating the communication between the Hadoop nodes.
 However, increasing either the number of samples or the number of features per sample quickly exceeds the ability of R.
-Mahout on the other hand scales effortless to an order of magnitude more samples (10,000) and copes well with 15 Million features per samples.
-As memory consumption directly scales with the number of features per sample we observed that Mahout's standard set-up of Hadoop does not scale to a whole exome dataset (38 Million features per sample).
+Mahout on the other hand scales effortlessly to five times more samples (5,460) and copes well with 15 Million features per samples.
+As memory consumption directly scales with the number of features per sample, we observe that Mahout's standard set-up of Hadoop does not scale to a whole exome dataset (38 Million features per sample).
 However, this might be resolved once Mahout moves from MapReduce to a domain-specific language for linear algebraic operations on top of Apache Spark \url{http://spark.apache.org/}.
@@
 We also investigate whether we can accurately predict the more detailed family association using these datasets. 
-In addition, we demonstrate Mahout's scalability by increasing the number of samples in the study to 10,920 by duplicating the original 1092 samples 10 times with zero and 10\% noise levels, i.e.
+In addition, we demonstrate Mahout's scalability by increasing the number of samples in the study to 5,460 by duplicating the original 1092 samples 5 times with zero and 10\% noise levels, i.e.
 randomly changing 10\% of the genotypes.
@@
 Each node has eight CPU cores, however both applications run a single thread for pre-processing the data.
-For the (\FullPop) dataset, we ran the job on a PBS cluster of x nodes, with twelve CPU cores and 48GB of memory per node.
+%TODO update these numbers
+For the \SixReal\ and \FullReal\ dataset, we ran the job on a PBS cluster of x nodes, with twelve CPU cores and 48GB of memory per node.
 
@@
 \subsection*{Clustering quality}
-We scored the clusters with the Adjusted Rand index. 
-%TODO: can you provide the formular
-This index compares two different cluster sets and assigns a value based on their similarity.
+We scored the clusters with the Adjusted Rand index~\cite{Hubert1985}. 
+This index compares two different cluster sets, here the prediction $X = \{ X_1, X_2, \ldots , X_r \}$ and the annotated data $Y = \{ Y_1, Y_2, \ldots , Y_s \}$, and assigns a value based the overlap between $X$ and $Y$ as captured in a contingency table $\left[n_{ij}\right]$ where each entry $n_{ij}$ denotes the number of objects in common between $X_i$ and $Y_j : n_{ij}=|X_i \cap Y_j|$.
+The Adjusted Rand index is then 
+\begin{eqnarray}\label{eqexpmuts}
+ARI=\frac{\sum_{ij}{{n_{ij}\choose 2}} - \left[ \sum_{i}{{a_i\choose2}} \right]}{bla} 
+\end{eqnarray}
+
+%
+
 The value is zero for independent clusterings and one for identical clusterings. 
@@
 R clusters for the Chr1 subset data.
-With the foreknowledge of the correct clusters, the Adjusted Rand Index enables us to assign a score to cluster accuracy~\cite{Hubert1985}.
 
@@
         \hline
-           & \chrOnePop & \FullPop   \\ \hline
+%           & \chrOnePop & \FullPop   \\ \hline
         Mahout, k=4 & 0.664\\
@@
            & Nodes & pmem & vmem  & pre-processing & clustering \\ \hline
-        R, \chrOnePop & 1 & 17,696,940 & 18,000,196 &3,341 & 7\\
-        Mahout,\chrOnePop & 1 & 1,831,160 & 7646232 & 1,255 & 594\\
+        R, \OneReal & 1 & 17,696,940 & 18,000,196 &3,341 & 7\\
+        Mahout,\OneReal & 1 & 1,831,160 & 7646232 & 1,255 & 594\\
         Mahout, \SixReal & 1 & 13,036,836 & 36,542,380 & 19,500 &..\\ 

### 2014-05-17 doc/publications/incob2014/paper.tex
changes

@@
 K-means clustering the data, however, is faster with R, using 7s, whereas Mahout takes 10 minutes. See Table~\ref{datasetsTime}. \\
-Due to the small number of classifying variants, cluster performance is very poor for both methods, with the best Adjusted Rand Index score for clustering individuals in one of 4 populations (see Methods) being 0.769 for 20 clusters in Mahout and XX with YY clusters in R (see Table~\ref{datasetsAcc}).
+Due to the small number of classifying variants, cluster performance is very poor for both methods, with the best Adjusted Rand Index score for clustering individuals in one of 4 populations (see Methods) being 0.769 for 20 clusters in Mahout and 0.614 with 10 clusters in R (see Table~\ref{datasetsAcc}).
 
@@
 R therefore was not able to perform the clustering on even the medium sized dataset (\SixReal , 15 Million variants), we hence only report resource consumption and accuracy for Mahout.
-On this dataset Mahout's prepossessing takes XX minutes, with XXGB of physical and XXGB of virtual memory, and the clustering takes XX minutes, with XXGB of physical and XXGB of virtual memory. 
-The clustering accuracy has increased by XX-fold to YY with XX clusters compared to the performance on \OneReal\ for the 4-group-based population task (see Table~\ref{datasetsAcc}).
-Using additional variants now also allows the more detailed association to the 855 families.
-The best Adjusted Rand Index score here is XX for XX clusters (see Table~\ref{datasetsAcc}).
+On this dataset Mahout's pre-processing takes 325 minutes, with 13GB of physical and 37GB of virtual memory, and the clustering takes 80 minutes, with XXGB of physical and XXGB of virtual memory. 
+The clustering accuracy has increased by 35\% to 0.812 with 10 clusters compared to the performance on \OneReal\ for the 4-group-based population task (see Table~\ref{datasetsAcc}).
 
@@
 %           & \chrOnePop & \FullPop   \\ \hline
-        Mahout, k=4 & 0.664\\
-        Mahout, k=10 & 0.714 \\
-        Mahout, k=20 & 0.769 \\ 
-        Mahout, k=30 & 0.728  \\ 
-        R, k=4 & .. & - \\
-        R, k=10 & .. & - \\
-        R, k=20 & .. & - \\ 
-        R, k=30 & .. & - \\ \hline
+        R, \OneReal & 0.614 & - \\ 
+        Mahout, \OneReal & 0.605\\
+        Mahout, \SixReal & 0.812 \\
+        Mahout, \SixArtiExact & 0.600  \\
+        Mahout, \SixArtiNoise & 0.140 \\ \hline
       \end{tabular}
@@
         Mahout,\OneReal & 1 & 1,831,160 & 7646232 & 1,255 & 594\\
-        Mahout, \SixReal & 1 & 13,036,836 & 36,542,380 & 19,500 &..\\ 
-        Mahout, \SixArtiExact & 1 & 5,804,660 & 39,075,548 & 2,024 & ..\\ 
+        Mahout, \SixReal & 5 & 13,036,836 & 36,542,380 & 19,500 & 4820 \\ 
+        Mahout, \SixArtiExact & 3 & 5,804,660 & 39,075,548 & 2,024 & 1395\\ 
         \hline

### 2014-05-17 doc/publications/incob2014/paper.tex
eq

--- 2014-05-17 doc/publications/incob2014/paper.tex
@@
 We therefore aim to provide the interface between standard genomic data formats and advanced and scalable analysis libraries like Mahout.
-\parttitle{Results} We achieve an XX-fold speedup by using the scalable k-means MapReduce implementation over the equivalent analysis performed in R and YY-fold speedup for PCA, respectively, by comparable accuracy.
-Due to memory in the R-Implementation the comparison had to be limited to variant data from chromosome 1 only, however we ran the full dataset (aa GB, bb individuals {\`a} cc variants ) through Mahout demonstrating the scalability.
+\parttitle{Results} We achieve an 2-fold speedup by using the scalable k-means MapReduce implementation over the equivalent analysis performed in R, by comparable accuracy.
+However, the real benefit lies in scaling beyond R's capability to a population-size analysis.
+We successfully clustered more than 5000 individuals each having more than 15 Million variants.
 \parttitle{Conclusions} Using modern compute paradigms is essential to scale to modern genomic research in an efficient sustainable way. 
@@
 Each node has eight CPU cores, however both applications run a single thread for pre-processing the data.
-%TODO update these numbers
-For the \SixReal\ and \FullReal\ dataset, we ran the job on a PBS cluster of x nodes, with twelve CPU cores and 48GB of memory per node.
+%TODO check numbers are correct ?
+For the \SixReal\ and \FullReal\ dataset, we ran the job on a PBS cluster of 5 nodes, with twelve CPU cores and 48GB of memory per node.
 
@@
 The Adjusted Rand index is then 
-\begin{eqnarray}\label{eqexpmuts}
-ARI=\frac{\sum_{ij}{{n_{ij}\choose 2}} - \left[ \sum_{i}{{a_i\choose2}} \right]}{bla} 
-\end{eqnarray}
-
-%
+{\tiny
+\begin{eqnarray*}
+ARI=\frac{\sum_{ij}{{n_{ij}\choose 2}} - \left[ \sum_{i}{{a_i\choose2}} \sum_{j}{{b_i\choose2}} \right] / {n\choose2}}{\frac{1}{2} \left[ \sum_{i}{{a_{i}\choose 2} + \sum_{j}{{b_{j}\choose 2}}} \right] - \left[ \sum_{i}{{a_{i}\choose 2} + \sum_{j}{{b_{j}\choose 2}}} \right] / {n\choose2}}
+\end{eqnarray*}
+}
+where $a$ and $b$ are the sums over the rows and columns of the contingency table.
 
@@
 clusters and the R k-means clusters to the known clusters based on meta data from the 1000 genome project. For the known clusters,
-individuals were partitioned by their super population code (ASN, EUR, AFR, AMR). 
+individuals were partitioned by their super population code (ASN, EUR, AFR, AMR) as well as the 855 families.
 %TODO: "also on \chrOnePop" on what dataset did you compare the R performance otherwise ?
-We also compared the Mahout clusters to the
-R clusters for the Chr1 subset data.
+%We also compared the Mahout clusters to the
+%R clusters for the Chr1 subset data.
 

### 2014-05-17 doc/publications/incob2014/paper.tex
table2 heading
text

@@
 In this paper we have applied Mahout's k-means clustering to the task of genotype based population and family association. 
-We provide an interface from VCF format to Mahout and a toolset to interpret Mahout's clustering output. 
+We provide an interface from VCF format to Mahout and a toolset to interpret Mahout's clustering output (see Figure~\ref{fig:sign})
 We find that on small datasets the actual clustering stage is two orders of magnitudes slower in Mahout compared to R (10 minutes vs 7 seconds), likely due to the overhead of setting up and co-ordinating the communication between the Hadoop nodes.
@@
   \begin{figure}[h!]
+  \label{fig:sign}
   \includegraphics[type=pdf,ext=.pdf,read=.pdf, scale=0.40]{images/signature}
@@
         \hline
-%           & \chrOnePop & \FullPop   \\ \hline
-        R, \OneReal & 0.614 & - \\ 
+         Tool, Dataset & accuracy  \\ \hline
+        R, \OneReal & 0.614  \\ 
         Mahout, \OneReal & 0.605\\

### 2014-05-19 doc/publications/incob2014/paper.tex
Updated

@@
 R therefore was not able to perform the clustering on even the medium sized dataset (\SixReal , 15 Million variants), we hence only report resource consumption and accuracy for Mahout.
-On this dataset Mahout's pre-processing takes 325 minutes, with 13GB of physical and 37GB of virtual memory, and the clustering takes 80 minutes, with XXGB of physical and XXGB of virtual memory. 
+On this dataset Mahout's pre-processing takes 325 minutes, with 13GB of physical and 37GB of virtual memory, and the clustering takes 80 minutes.
 The clustering accuracy has increased by 35\% to 0.812 with 10 clusters compared to the performance on \OneReal\ for the 4-group-based population task (see Table~\ref{datasetsAcc}).
@@
 In this section we investigate Mahout's ability to scale with more samples.
-The resource consumption remains virtually identical between \SixReal\ and the order of magnitude larger dataset \SixArtiExact, which takes XX minutes, with XXGB of physical and XXGB of virtual memory, and the clustering takes XX minutes, with XXGB of physical and XXGB of virtual memory.
-The clustering accuracy is YY with XX clusters.
+The resource consumption remains virtually identical between \SixReal\ and the order of magnitude larger dataset \SixArtiExact, which takes 33 minutes, with 6GB of physical and 39GB of virtual memory, and the clustering takes 23 minutes
+The clustering accuracy is 0.812 with 10 clusters.
 
@@
 In this paper we have applied Mahout's k-means clustering to the task of genotype based population and family association. 
-We provide an interface from VCF format to Mahout and a toolset to interpret Mahout's clustering output (see Figure~\ref{fig:sign})
+We provide an interface from VCF format to Mahout and a toolset to interpret Mahout's clustering output (see Figure 1)
 We find that on small datasets the actual clustering stage is two orders of magnitudes slower in Mahout compared to R (10 minutes vs 7 seconds), likely due to the overhead of setting up and co-ordinating the communication between the Hadoop nodes.

### 2014-10-24 doc/publications/journal2014/paper.tex
converted paper to briefings in bioinf

--- 2014-05-19 doc/publications/incob2014/paper.tex
@@
-%% BioMed_Central_Tex_Template_v1.06
-%%                                      %
-%  bmc_article.tex            ver: 1.06 %
-%                                       %
+\documentclass{bioinfo}
+\copyrightyear{2005}
+\pubyear{2005}
 
-%%IMPORTANT: do not delete the first line of this template
-%%It must be present to enable the BMC Submission system to
-%%recognise this template!!
+\usepackage[hyphens]{url}
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                                     %%
-%%  LaTeX template for BioMed Central  %%
-%%     journal article submissions     %%
-%%                                     %%
-%%          <8 June 2012>              %%
-%%                                     %%
-%%                                     %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                                                                 %%
-%% For instructions on how to fill out this Tex template           %%
-%% document please refer to Readme.html and the instructions for   %%
-%% authors page on the biomed central website                      %%
-%% http://www.biomedcentral.com/info/authors/                      %%
-%%                                                                 %%
-%% Please do not use \input{...} to include other tex files.       %%
-%% Submit your LaTeX manuscript as one .tex document.              %%
-%%                                                                 %%
-%% All additional figures and files should be attached             %%
-%% separately and not embedded in the \TeX\ document itself.       %%
-%%                                                                 %%
-%% BioMed Central currently use the MikTex distribution of         %%
-%% TeX for Windows) of TeX and LaTeX.  This is available from      %%
-%% http://www.miktex.org                                           %%
-%%                                                                 %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%%% additional documentclass options:
-%  [doublespacing]
-%  [linenumbers]   - put the line numbers on margins
-
-%%% loading packages, author definitions
-
-\documentclass[twocolumn]{bmcart}% uncomment this for twocolumn layout and comment line below
-%\documentclass{bmcart}
-
-%%% Load packages
-%\usepackage{amsthm,amsmath}
-%\RequirePackage{natbib}
-\RequirePackage{hyperref}
-\usepackage[utf8]{inputenc} %unicode support
-%\usepackage[applemac]{inputenc} %applemac support if unicode package fails
-%\usepackage[latin1]{inputenc} %UNIX support if unicode package fails
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                                             %%
-%%  If you wish to display your graphics for   %%
-%%  your own use using includegraphic or       %%
-%%  includegraphics, then comment out the      %%
-%%  following two lines of code.               %%
-%%  NB: These line *must* be included when     %%
-%%  submitting to BMC.                         %%
-%%  All figure files must be submitted as      %%
-%%  separate graphics through the BMC          %%
-%%  submission process, not included in the    %%
-%%  submitted article.                         %%
-%%                                             %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%\def\includegraphic{}
-%\def\includegraphics{}
-\usepackage{graphicx}
-
-%%% Put your definitions there:
-\startlocaldefs
-\endlocaldefs
-
-% Our variable names
-%\newcommand{\chrOnePop}{Chr19\_Population}
-%\newcommand{\FullPop}{Full\_Population}
-%\newcommand{\FullFam}{Chr1\_Family}
+\begin{document}
+\firstpage{1}
 
@@
 
-%%% Begin ...
-\begin{document}
+\title[MapReduce Clustering]{Scalable Clustering of Genotype Information using MapReduce}
+\author[O'Brien \textit{et~al}]{Aidan R O'Brien\,$^{1}$, Fabian A Buske\,$^{2}$ and Denis C. Bauer\,$^1$\footnote{to whom correspondence should be addressed}}
+\address{$^{1}$CSIRO, Digital Productivity Flagship, 11 Julius Av, 2113, Sydney, Australia\\
+$^{2}$Cancer Epigenetics Program, Cancer Research Division, Kinghorn Cancer Centre, Garvan Institute of Medical Research, 384 Victoria St, 2010, Sydney, Australia}
 
-%%% Start of article front matter
-\begin{frontmatter}
+\history{Received on XXXXX; revised on XXXXX; accepted on XXXXX}
 
-\begin{fmbox}
-\dochead{Methodology}
+\editor{Associate Editor: XXXXXXX}
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                                          %%
-%% Enter the title of your article here     %%
-%%                                          %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\maketitle
 
-\title{Scalable Clustering of Genotype Information using MapReduce}
+\begin{abstract}
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                                          %%
-%% Enter the authors here                   %%
-%%                                          %%
-%% Specify information, if available,       %%
-%% in the form:                             %%
-%%   <key>={<id1>,<id2>}                    %%
-%%   <key>=                                 %%
-%% Comment or delete the keys which are     %%
-%% not used. Repeat \author command as much %%
-%% as required.                             %%
-%%                                          %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\author[
-   addressref={aff1,aff2},                   % id's of addresses, e.g. {aff1,aff2}
-%   noteref={n1},                        % id's of article notes, if any
-   email={Aidan.O'Brien@csiro.au}   % email address
-]{\inits{AR}\fnm{Aidan R} \snm{O'Brien}}
-\author[
-   addressref={aff3, aff4},
-   email={f.buske@garvan.org.au}
-]{\inits{FA}\fnm{Fabian A} \snm{Buske}}
-\author[
-   addressref={aff1},
-   corref={aff1},                       % id of corresponding address, if any
-   email={Denis.Bauer@CSIRO.au}
-]{\inits{DC}\fnm{Denis C} \snm{Bauer}}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                                          %%
-%% Enter the authors' addresses here        %%
-%%                                          %%
-%% Repeat \address commands as much as      %%
-%% required.                                %%
-%%                                          %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\address[id=aff1]{%                           % unique id
-  \orgname{Division of Computational Informatics, CSIRO}, % university, etc
-  \street{11 Julius Av},                     %
-  \postcode{2113}                                % post or zip code
-  \city{Sydney},                              % city
-  \cny{Australia}                                    % country
-}
-\address[id=aff2]{%
-  \orgname{School of Biomedical Sciences and Pharmacy, Faculty of Health},
-%  \street{},
-  \postcode{2308}
-  \city{Newcastle},
-  \cny{Australia}
-}
-\address[id=aff3]{%
-  \orgname{Cancer Epigenetics Program, Cancer Research Division, Kinghorn Cancer Centre, Garvan Institute of Medical Research},
-  \street{384 Victoria St},
-  \postcode{2010}
-  \city{Sydney},
-  \cny{Australia}
-}
-\address[id=aff4]{%
-  \orgname{UNSW Medicine, University of New South Wales},
-%  \street{},
-  \postcode{2052}
-  \city{Sydney},
-  \cny{Australia}
-}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                                          %%
-%% Enter short notes here                   %%
-%%                                          %%
-%% Short notes will be after addresses      %%
-%% on first page.                           %%
-%%                                          %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%\begin{artnotes}
-%\note{Sample of title note}     % note to the article
-%\note[id=n1]{Equal contributor} % note, connected to author
-%\end{artnotes}
-
-%\end{fmbox}% comment this for two column layout
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                                          %%
-%% The Abstract begins here                 %%
-%%                                          %%
-%% Please refer to the Instructions for     %%
-%% authors on http://www.biomedcentral.com  %%
-%% and include the section headings         %%
-%% accordingly for your article type.       %%
-%%                                          %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\begin{abstractbox}
-
-\begin{abstract} % abstract
-\parttitle{Background} Processing genomic information from whole genome sequence studies pose computational challenges due to the unprecedented data volume generated, which render transitional approaches insufficient.
+\section{Motivation:}
+Processing genomic information from whole genome sequence studies pose computational challenges due to the unprecedented data volume generated, which render transitional approaches insufficient.
 However, by utilising advancements in modern hardware accelerators and data processing we can provide the means for scalable solutions.
 We therefore aim to provide the interface between standard genomic data formats and advanced and scalable analysis libraries like Mahout.
-\parttitle{Results} We achieve an 2-fold speedup by using the scalable k-means MapReduce implementation over the equivalent analysis performed in R, by comparable accuracy.
+\section{Results:}
+We achieve an 2-fold speedup by using the scalable k-means MapReduce implementation over the equivalent analysis performed in R, by comparable accuracy.
 However, the real benefit lies in scaling beyond R's capability to a population-size analysis.
 We successfully clustered more than 5000 individuals each having more than 15 Million variants.
-\parttitle{Conclusions} Using modern compute paradigms is essential to scale to modern genomic research in an efficient sustainable way. 
+
+\section{Availability:}
+Using modern compute paradigms is essential to scale to modern genomic research in an efficient sustainable way.
+
+\section{Contact:} \href{Denis.Bauer@CSIRO.au}{Denis.Bauer@CSIRO.au}
 \end{abstract}
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                                          %%
-%% The keywords begin here                  %%
-%%                                          %%
-%% Put each keyword in separate \kwd{}.     %%
-%%                                          %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Introduction}
 
-\begin{keyword}
-\kwd{Hadoop}
-\kwd{Mahout}
-\kwd{Genotype}
-\kwd{Clustering}
-\kwd{MapReduce}
-\end{keyword}
-
-% MSC classifications codes, if any
-%\begin{keyword}[class=AMS]
-%\kwd[Primary ]{}
-%\kwd{}
-%\kwd[; secondary ]{}
-%\end{keyword}
-
-\end{abstractbox}
-%
-\end{fmbox}% uncomment this for twcolumn layout
-
-\end{frontmatter}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                                          %%
-%% The Main Body begins here                %%
-%%                                          %%
-%% Please refer to the instructions for     %%
-%% authors on:                              %%
-%% http://www.biomedcentral.com/info/authors%%
-%% and include the section headings         %%
-%% accordingly for your article type.       %%
-%%                                          %%
-%% See the Results and Discussion section   %%
-%% for details on how to create sub-sections%%
-%%                                          %%
-%% use \cite{...} to cite references        %%
-%%  \cite{koon} and                         %%
-%%  \cite{oreg,khar,zvai,xjon,schn,pond}    %%
-%%  \nocite{smith,marg,hunn,advi,koha,mouse}%%
-%%                                          %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%%%%%%%%%%%%%%%%%%%%%%%%% start of article main body
-% <put your article body there>
-
-%%%%%%%%%%%%%%%%
-%% Background %%
-%%
-\section*{Background}
 Grouping individuals based on the genomic profile is a commonly performed tasks to identify population structure~\cite{Gao2007} or elucidate different haplotype involvement in diseases susceptibility~\cite{Laitman2013}.
@@
 This is especially the case for sequence analysis tasks, such as read mapping~\cite{Schatz2009}, duplicate removal~\cite{Jourdren2012}, and variant calling~\cite{Langmead2009, McKenna2010} as well as Genome Wide Analysis Study based tasks~\cite{Huang2013, Guo2014}.
+
+%Mention Goolge's population clustering in the intro
 
@@
 
-\section*{Methods}
+\begin{methods}
+\section{Methods}
 \subsection*{Datasets}
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                                          %%
-%% Backmatter begins here                   %%
-%%                                          %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\end{methods}
 
-\begin{backmatter}
-
-\section*{Availability of supporting data}
-\section*{List of abbreviations}
-TCGA -- the cancer genome atlas
-VCF -- variant call format
-PBS -- Portable Batch System
-
-\section*{Competing interests}
-  The authors declare that they have no competing interests.
-
-\section*{Author's contributions}
-A.R.O implemented the method and performed the results under the supervision of D.C.B. and F.A.B.
-All authors wrote the paper.
-
-\section*{Acknowledgements}
-A.R.O was funded by the NSW Cancer Institute Big Data Big Impact schema, F.A.B by the National Health and Medical Research Council [1051757] and D.C.B by Commonwealth Scientific and Industrial Research Organisation's Transformational Capability Platform, Science and Industry Endowment Fund and Information Management and Technology Services.
-The authors would like to thank Piotr Szul and Gareth Williams for their help with setting up Hadoop on the HPC system.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                  The Bibliography                       %%
-%%                                                         %%
-%%  Bmc_mathpys.bst  will be used to                       %%
-%%  create a .BBL file for submission.                     %%
-%%  After submission of the .TEX file,                     %%
-%%  you will be prompted to submit your .BBL file.         %%
-%%                                                         %%
-%%                                                         %%
-%%  Note that the displayed Bibliography will not          %%
-%%  necessarily be rendered by Latex exactly as specified  %%
-%%  in the online Instructions for Authors.                %%
-%%                                                         %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-% if your bibliography is in bibtex format, use those commands:
-\bibliographystyle{bmc-mathphys} % Style BST file
-\bibliography{genotypeClustering}      % Bibliography file (usually '*.bib' )
-
-% or include bibliography directly:
-% \begin{thebibliography}
-% \bibitem{b1}
-% \end{thebibliography}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                               %%
-%% Figures                       %%
-%%                               %%
-%% NB: this is for captions and  %%
-%% Titles. All graphics must be  %%
-%% submitted separately and NOT  %%
-%% included in the Tex document  %%
-%%                               %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%%
-%% Do not use \listoffigures as most will included as separate files
-
-\section*{Figures}
-  \begin{figure}[h!]
-  \label{fig:sign}
-  \includegraphics[type=pdf,ext=.pdf,read=.pdf, scale=0.40]{images/signature}
-  \caption{\csentence{Illustration of Mahout-based clustering of genotypes.}
-      The image shows how the here introduced interface converts the VCF file to a mahout usable vector-based data type.
-Based on k-means it illustrates how the Map-step procedure groups of vectors by their nearest cluster center and the Reduce-step averages the vectors in each group to find the new value of the updated cluster center.
-The mahout-produced output is then converted into a visualisation.}
-      \end{figure}
-
-%\begin{figure}[h!]
-%  \caption{\csentence{Sample figure title.}
- %     Figure legend text.}
- %     \end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                               %%
-%% Tables                        %%
-%%                               %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%% Use of \listoftables is discouraged.
-%%
-\section*{Tables}
-\begin{table}[h!]
-\caption{Datasets used in this study.}
-      \begin{tabular}{lrrrrr}
-        \hline
+\begin{table}[!t]
+\processtable{Datasets used in this study.\label{Tab:01}}
+{\begin{tabular}{lrrrrr}\toprule
 & individuals  & variants  & population & families & noise\\
-           & & &groups&& \\ \hline
+& & &groups&& \\\midrule
         \OneReal & 1,092 & 816,115  & 4 & 855 & NA\\
@@
 	\SixArtiExact\ & 5,460 & 816,115 & 4 & 855 & 0 \\ 
-	\SixArtiNoise\ & 5,460 & 816,115 & 4 & 855 & 10 \\ 
-      \end{tabular}
-      \label{datasets}
+	\SixArtiNoise\ & 5,460 & 816,115 & 4 & 855 & 10 \\\botrule
+\end{tabular}}{Datasets used in this study}
 \end{table}
 
-\begin{table}[h!]
-\caption{Adjusted Rand index (accuracy).}
-      \begin{tabular}{lcc}
-        \hline
-         Tool, Dataset & accuracy  \\ \hline
+\begin{table}[!t]
+\processtable{Adjusted Rand index (accuracy).\label{Tab:01}}
+{\begin{tabular}{lcc}\toprule
+Tool, Dataset & accuracy \\\midrule
       R, \OneReal & 0.614  \\ 
@@
         Mahout, \SixArtiExact & 0.600  \\
-        Mahout, \SixArtiNoise & 0.140 \\ \hline
-      \end{tabular}
-      \label{datasetsAcc}
+        Mahout, \SixArtiNoise & 0.140 \\\botrule
+\end{tabular}}{Datasets used in this study}
 \end{table}
 
-\begin{table}[h!]
-\caption{Resources consumed.}
-      \begin{tabular}{lccccc}
-      \hline
+\begin{table*}[!t]
+\processtable{Resources consumed.\label{Tab:01}}
+{\begin{tabular}{lccccc}\toprule
 &&\multicolumn{2}{ c }{Memory (kb)}&\multicolumn{2}{ c }{Time (s)} \\
-        \hline
-           & Nodes & pmem & vmem  & pre-processing & clustering \\ \hline
+& Nodes & pmem & vmem  & pre-processing & clustering\\\midrule
          R, \OneReal & 1 & 17,696,940 & 18,000,196 &3,341 & 7\\
@@
         Mahout, \SixReal & 5 & 13,036,836 & 36,542,380 & 19,500 & 4820 \\ 
-        Mahout, \SixArtiExact & 3 & 5,804,660 & 39,075,548 & 2,024 & 1395\\ 
-        \hline
-      \end{tabular}
-      \label{datasetsTime}
-\end{table}
+        Mahout, \SixArtiExact & 3 & 5,804,660 & 39,075,548 & 2,024 & 1395 \\\botrule
+\end{tabular}}{Datasets used in this study}
+\end{table*}
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                               %%
-%% Additional Files              %%
-%%                               %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\begin{figure}[!tpb]%figure1
+\centerline{\includegraphics[type=pdf,ext=.pdf,read=.pdf, scale=0.40]{signature}}
+        \label{fig:sign}
+        \caption{{\bf Illustration of Mahout-based clustering of genotypes.}
+      The image shows how the here introduced interface converts the VCF file to a mahout usable vector-based data type.
+Based on k-means it illustrates how the Map-step procedure groups of vectors by their nearest cluster center and the Reduce-step averages the vectors in each group to find the new value of the updated cluster center.
+The mahout-produced output is then converted into a visualisation.}
 
-%\section*{Additional Files}
-%  \subsection*{Additional file 1 --- Sample additional file title}
-%    Additional file descriptions text (including details of how to
-%    view the file, if it is in a non-standard format or the file extension).  This might
- %   refer to a multi-page table or a figure.
+\end{figure}
 
-%  \subsection*{Additional file 2 --- Sample additional file title}
- %   Additional file descriptions text.
+\section*{Acknowledgement}
+A.R.O was funded by the NSW Cancer Institute Big Data Big Impact schema, F.A.B by the National Health and Medical Research Council [1051757] and D.C.B by Commonwealth Scientific and Industrial Research Organisation's Transformational Capability Platform, Science and Industry Endowment Fund and Information Management and Technology Services.
+The authors would like to thank Piotr Szul and Gareth Williams for their help with setting up Hadoop on the HPC system.
 
-\end{backmatter}
+\bibliographystyle{natbib}
+%\bibliographystyle{achemnat}
+%\bibliographystyle{plainnat}
+%\bibliographystyle{abbrv}
+%\bibliographystyle{bioinformatics}
+\bibliography{genotypeClustering}  
 \end{document}

### 2014-11-07 vcf-clusterer/src/net/aidan/CompositeKey.java
MapReduce class to convert VCF to Mahout's sequence file.

@@ -1,0 +1,88 @@
+package net.aidan;
+
+import java.io.DataInput;
+import java.io.DataOutput;
+import java.io.IOException;
+
+import org.apache.hadoop.io.IntWritable;
+import org.apache.hadoop.io.WritableComparable;
+
+/**
+* This key is a composite key. The "actual"
+* key is the UDID. The secondary sort will be performed against the datetime.
+*/
+public class CompositeKey implements WritableComparable<CompositeKey> {
+
+	private IntWritable individualid;
+	private IntWritable variantlocation;
+	
+	
+	public CompositeKey() {
+    	set(new IntWritable(), new IntWritable());
+	}
+
+	public CompositeKey(IntWritable individualid, IntWritable variantlocation) {
+		this.individualid = individualid;
+		this.variantlocation = variantlocation;
+	}
+
+    public void set(IntWritable individualid, IntWritable variantlocation) {
+        this.individualid = individualid;
+        this.variantlocation = variantlocation;
+    }
+    public void set(int individualid, int variantlocation) {
+        this.individualid.set(individualid);
+        this.variantlocation.set(variantlocation);
+    }
+	
+	@Override
+	public String toString() {
+		return "[" + individualid.toString() + ',' + variantlocation + "]";
+	}
+	
+	@Override
+	public void readFields(DataInput in) throws IOException {
+		this.individualid.readFields(in);
+		this.variantlocation.readFields(in);
+	}
+	
+	@Override
+	public void write(DataOutput out) throws IOException {
+		individualid.write(out);
+		variantlocation.write(out);
+	}
+	
+	@Override
+	public int compareTo(CompositeKey o) {
+	
+		int result = individualid.compareTo(o.individualid);
+		if (0 == result) {
+			result = variantlocation.compareTo(o.variantlocation);
+		}
+		return result;
+	}
+	
+	/**
+	* Gets the udid.
+	*
+	* @return UDID.
+	*/
+	public IntWritable getIndividualId() {
+		return individualid;
+	}
+	
+	public void setIndividualId(IntWritable individualid) {
+		this.individualid = individualid;
+	}
+	
+
+	public IntWritable getVariantLocation() {
+		return variantlocation;
+	}
+	
+	public void setDatetime(IntWritable variantlocation) {
+		this.variantlocation = variantlocation;
+	}
+	
+
+}

### 2014-11-07 vcf-clusterer/src/net/aidan/GenotypeWritable.java

@@ -1,0 +1,93 @@
+package net.aidan;
+
+import java.io.DataInput;
+import java.io.DataOutput;
+import java.io.IOException;
+
+import org.apache.hadoop.io.DoubleWritable;
+import org.apache.hadoop.io.IntWritable;
+import org.apache.hadoop.io.LongWritable;
+import org.apache.hadoop.io.WritableComparable;
+
+/**
+ * A custom Writable implementation for Request information.
+ *
+ * This is simple Custom Writable, and does not implement Comparable or RawComparator
+ */
+public class GenotypeWritable extends Object implements WritableComparable<GenotypeWritable> {
+
+    private IntWritable location;
+    private DoubleWritable variant;
+   
+
+    public GenotypeWritable(IntWritable location, DoubleWritable variant) {
+    	set(location, variant);
+    }
+
+    public GenotypeWritable() {
+    	set(new IntWritable(), new DoubleWritable());
+    }
+    
+    public GenotypeWritable(int location, Double variant) {
+    	set(new IntWritable(location), new DoubleWritable(variant));
+    }
+    
+    public IntWritable getLocation() {
+    	return location;
+    }
+    
+    public DoubleWritable getVariant() {
+    	return variant;
+    }
+        
+    public void set(IntWritable location, DoubleWritable variant) {
+        this.location = location;
+        this.variant = variant;
+    }
+    
+
+    @Override
+    public void readFields(DataInput in) throws IOException {
+        location.readFields(in);
+        variant.readFields(in);
+    }
+ 
+    @Override
+    public void write(DataOutput out) throws IOException {
+        location.write(out);
+        variant.write(out);
+    }
+ 
+    @Override
+    public String toString() {
+        return "[" + location + ", " + variant + "]";
+    }
+ 
+    @Override
+    public int compareTo(GenotypeWritable tp) {
+        int cmp = location.compareTo(tp.location);
+ 
+        if (cmp != 0) {
+            return cmp;
+        }
+ 
+        return variant.compareTo(tp.variant);
+    }
+ 
+    @Override
+    public int hashCode(){
+        return location.hashCode()*163 + variant.hashCode();
+    }
+ 
+    @Override
+    public boolean equals(Object o)
+    {
+        if(o instanceof GenotypeWritable)
+        {
+        	GenotypeWritable tp = (GenotypeWritable) o;
+            return location.equals(tp.location) && variant.equals(tp.variant);
+        }
+        return false;
+    }
+ 
+}

### 2014-11-07 vcf-clusterer/src/net/aidan/IndividualKeyGroupingComparator.java

@@ -1,0 +1,19 @@
+package net.aidan;
+
+import org.apache.hadoop.io.WritableComparable;
+import org.apache.hadoop.io.WritableComparator;
+
+public class IndividualKeyGroupingComparator extends WritableComparator {
+
+	protected IndividualKeyGroupingComparator() {
+	super(CompositeKey.class, true);
+	}
+	
+	@Override
+	public int compare(WritableComparable w1, WritableComparable w2) {
+		CompositeKey key1 = (CompositeKey) w1;
+		CompositeKey key2 = (CompositeKey) w2;
+		// (check on IndividualId)
+		return key1.getIndividualId().compareTo(key2.getIndividualId());
+	}
+}

### 2014-11-07 vcf-clusterer/src/net/aidan/IndividualKeyPartitioner.java

@@ -1,0 +1,27 @@
+package net.aidan;
+
+import org.apache.hadoop.io.DoubleWritable;
+import org.apache.hadoop.io.IntWritable;
+import org.apache.hadoop.io.Text;
+import org.apache.hadoop.mapreduce.Partitioner;
+import org.apache.hadoop.mapreduce.lib.partition.HashPartitioner;
+ 
+public class IndividualKeyPartitioner extends Partitioner<CompositeKey, DoubleWritable> {
+ 
+	HashPartitioner<IntWritable, DoubleWritable> hashPartitioner = new HashPartitioner<IntWritable, DoubleWritable>();
+	IntWritable newKey = new IntWritable();
+	 
+	@Override
+	public int getPartition(CompositeKey key, DoubleWritable value, int numReduceTasks) {
+		 
+		try {
+			// Execute the default partitioner over the first part of the key
+			newKey =key.getIndividualId();
+			return hashPartitioner.getPartition(newKey, value, numReduceTasks);
+		} catch (Exception e) {
+			e.printStackTrace();
+			return (int) (Math.random() * numReduceTasks); // this would return a random value in the range
+			// [0,numReduceTasks)
+		}
+	}
+}

### 2014-11-07 vcf-clusterer/src/net/aidan/LocationKeyComparator.java

@@ -1,0 +1,26 @@
+package net.aidan;
+
+import org.apache.hadoop.io.WritableComparable;
+import org.apache.hadoop.io.WritableComparator;
+
+public class LocationKeyComparator extends WritableComparator {
+protected LocationKeyComparator() {
+super(CompositeKey.class, true);
+}
+@SuppressWarnings("rawtypes")
+@Override
+public int compare(WritableComparable w1, WritableComparable w2) {
+	
+	CompositeKey key1 = (CompositeKey) w1;
+	CompositeKey key2 = (CompositeKey) w2;
+	
+	// (first check on udid)
+	int compare = key1.getIndividualId().compareTo(key2.getIndividualId());
+	if (compare == 0) {
+		// only if we are in the same input group should we try and sort by value (datetime)
+		return key1.getVariantLocation().compareTo(key2.getVariantLocation());
+	}
+	
+	return compare;
+	}
+}

### 2014-11-07 vcf-clusterer/src/net/aidan/csvClusterer.java

@@ -1,0 +1,170 @@
+/**
+ * 
+ */
+/**
+ * @author obr17q
+ *
+ */
+package net.aidan;
+
+import java.io.IOException;
+
+import org.apache.commons.logging.Log;
+import org.apache.commons.logging.LogFactory;
+import org.apache.hadoop.conf.Configuration;
+import org.apache.hadoop.conf.Configured;
+import org.apache.hadoop.fs.FileSystem;
+import org.apache.hadoop.fs.Path;
+import org.apache.hadoop.io.DoubleWritable;
+import org.apache.hadoop.io.LongWritable;
+import org.apache.hadoop.io.Text;
+import org.apache.hadoop.mapreduce.Job;
+import org.apache.hadoop.mapreduce.Mapper;
+import org.apache.hadoop.mapreduce.Reducer;
+import org.apache.hadoop.mapreduce.lib.input.FileInputFormat;
+import org.apache.hadoop.mapreduce.lib.input.TextInputFormat;
+import org.apache.hadoop.mapreduce.lib.output.FileOutputFormat;
+import org.apache.hadoop.mapreduce.lib.output.SequenceFileOutputFormat;
+import org.apache.hadoop.util.GenericOptionsParser;
+import org.apache.hadoop.util.Tool;
+import org.apache.hadoop.util.ToolRunner;
+import org.apache.mahout.math.SequentialAccessSparseVector;
+import org.apache.mahout.math.VectorWritable;
+
+
+
+public class csvClusterer extends Configured implements Tool {
+	//public csvClusterer() {}
+	
+	// Some variables
+	public static String[] idArray;
+	public static final String SEQUENCE_OUT_DIRECTORY = "clustering";
+	public static String INPUT_DIRECTORY;
+	public static final boolean SINGLE_MACHINE = false;
+	public static final int k = 50;
+	public static final int FEATURE_SIZE = 15161339;
+	public static long end = 0;
+	public static long start = System.currentTimeMillis();
+	private static final Log log = LogFactory.getLog(csvClusterer.class);
+
+		
+	public static class TokenizerMapper extends Mapper<LongWritable, Text, Text, VectorWritable>{
+		
+		private Text individual_id = new Text();
+		private VectorWritable genotype = new VectorWritable();
+		
+		public void map(LongWritable key, Text value, Context context) throws IOException, InterruptedException {
[…]

### 2014-11-07 vcf-clusterer/src/net/aidan/kmeansClusterer.java

@@ -57,6 +57,7 @@ public class kmeansClusterer extends Configured implements Tool {
 	public static final String OUTPUT_DIRECTORY = "clustering";
 	public static final boolean SINGLE_MACHINE = false;
 	public static final int k = 10;
+	public static final int FEATURE_SIZE = 15161339;
 	public static long end = 0;
 	public static long start = System.currentTimeMillis();
 	public int run(String[] args) throws Exception {
@@ -71,14 +72,21 @@ public class kmeansClusterer extends Configured implements Tool {
 		 * It may work whithout these lines now, with other changes I made,
 		 * but I haven't tested it yet. 
 		 */
-		final String HADOOP_CONF_DIR = System.getenv("HADOOP_CONF_DIR");
+		/*final String HADOOP_CONF_DIR = System.getenv("HADOOP_CONF_DIR");
 		conf.addResource(new Path(HADOOP_CONF_DIR, "core-site.xml"));
 		conf.addResource(new Path(HADOOP_CONF_DIR, "hdfs-site.xml"));
 		conf.addResource(new Path(HADOOP_CONF_DIR, "mapred-default.xml"));
 		conf.addResource(new Path(HADOOP_CONF_DIR, "mapred-site.xml"));
 		conf.addResource(new Path(HADOOP_CONF_DIR, "yarn-default.xml"));		
-		conf.addResource(new Path(HADOOP_CONF_DIR, "yarn-site.xml"));
+		conf.addResource(new Path(HADOOP_CONF_DIR, "yarn-site.xml"));*/
 		//conf.set("dfs.blocksize","134217728");
+		conf.set("mapreduce.job.reduces", "40");
+		
+		conf.set("mapreduce.map.memory.mb","6144");
+		conf.set("mapreduce.map.java.opts","-Xmx4915m");
+
+		
+		
 		conf.set("io.compression.codecs",
 				"org.apache.hadoop.io.compress.GzipCodec,org.apache.hadoop.io.compress.DefaultCodec,org.apache.hadoop.io.compress.BZip2Codec");
 			
@@ -94,18 +102,18 @@ public class kmeansClusterer extends Configured implements Tool {
 		//csvToSequences(conf, fs);
 		
 		//Probably redundant because of next line (Or could be used for non-random centroids)
-		chooseInitialCentroids(conf, fs);
+		//chooseInitialCentroids(conf, fs);
 		
[…]

### 2014-11-07 doc/publications/journal2014/paper.tex
pre-processing section

@@
 \label{Sec:preprocessing}
-Mahout requires sequence files as input to its clustering algorithms, and it provides libraries for writing vectors to this file format.
-Separate vectors for each sample, in this case, an individual, contain entries for each of the sample's features, where each entry is
-a \texttt{double}. Although additional samples can subsequently be appended to sequence files, the entire feature-set for each sample must be
-appended in a single write operation. Because the features for each sample are present in columns, as tab-separated elements in VCF
-files, we recursively transpose the data into CSV files. Each comma-separated row contains the feature-set for a sample,
-as well as the sample identity. Concurrently, we convert the values from the VCF file to doubles, as required for Mahout's sequence
-file format and the many algorithms this format can serve as input for. We set the value of the double to the Hamming distance between
-each variant and the reference genome. Therefore, a value identical to the reference will be 0, a heterozygous variant on either strand
-will be 1 and a homozygous variant will be 2.
-At this stage, the CSV files from more than one VCF file are concatenated into the one file. This resulting CSV file serves as the input for
-a custom Java function. This iterates over the CSV file, extracting the sample name and feature-set from each line, and calls a Mahout
-function to store this data as a named \texttt{DenseVector} object. The vectors are populated primarily by zeroes, so we convert this object to 
-the more efficient \texttt{SequentialAccessSparseVector} object, which stores only non-zero doubles in an array, and maps their position in a second array.
+Mahout requires Hadoop's `SequenceFiles' as input to its machine learning algorithms. SequenceFiles are binary flat files consisting of key/value pairs. In this case, each of the keys is a
+label for an individual. The associated value for each key is a `sparse vector' representing the individual's genomic variants. We use sparse vectors, rather than
+dense vectors, to significantly reduce processing time and required disk space. Each sparse vector consists of two separate vectors. An item in the first vector is
+effectively a key for an item at the same position in the send vector. The key represents a genomic location, where its value represents the variant. Zero values (positions where there is no variant)
+are omitted. Therefore, the fewer variants in an individual's genome, the smaller the sparse vector size. In comparison, `dense vectors' are single arrays that
+include every value. Therefore, the size of a dense vector isn't reduced with a large number of non-variants.
 
-This object is serialized, and the resulting file can serve as input to the majority of Mahout's algorithms.
+The values of the sparse vectors, which, as previously mentioned, represent an individuals genomic variants, are a \texttt{double} java type.
+We set the value of each double to the Hamming distance between each variant and the reference genome. Therefore, a value identical to the reference will be 0 (and omitted from the sparse vector),
+a heterozygous variant will be 1 and a homozygous variant will be 2.
+
+We wrote a custom `MapReduce' class to convert VCF files to the aforementioned sequence files. This should work on any number of files,
+of any file size; limited only by disk space and time.
+The `map' step creates key value pairs where the primary key is an individual, the secondary key is a variant location, and the value is the variant.
+We used a secondary key as this allowed us to implement custom functions to collect and sort the output of each mapper, based on this secondary key. This significantly reduced
+the overall processing time.
+The `reduce' step collects these sorted key/value pairs by primary key (individual ID) and adds the secondary key (location) and value (variant) to a sparse array.
+The algorithm
+saves these sparse arrays and keys to disk in the aforementioned SequenceFile format.
 

### 2014-11-07 doc/publications/journal2014/paper.tex
Hardware section

@@
 \subsection*{Hardware}
-We processed the data using a Portable Batch System (PBS) cluster. We allocated one node and 64GB memory for the comparison between Mahout and R.
-Each node has eight CPU cores, however both applications run a single thread for pre-processing the data.
-%TODO check numbers are correct ?
+We processed the data using a cluster of VMs on the Microsoft Azure cloud service. Each of the VMs has 7GB memory and and 4 CPU cores.
+Six of the VMs have the YARN NodeManager installed, and therefore these 6 machines process the data. The seventh machine dedicates the majority of its memory to
+the HDFS NameNode, ResourceManager and monitoring services. An eighth, single core machine, hosts the Cloudera manager service.
+With the 6 quad-core machines, we can run a total of 24 YARN containers simultaneously.
+
+We used a single VM for the comparison between Mahout and R.
 For the \SixReal\ and \FullReal\ dataset, we ran the job on a PBS cluster of 5 nodes, with twelve CPU cores and 48GB of memory per node.

### 2014-12-17 doc/publications/journal2014/paper.tex
updates

--- 2014-11-07 doc/publications/journal2014/paper.tex
@@
 
-\newcommand{\OneReal}{OneChr\_1K}
-\newcommand{\SixReal}{SixChr\_1K}
-\newcommand{\FullReal}{Full\_1K}
-\newcommand{\SixArtiExact}{Chr19\_5K\_0}
-\newcommand{\SixArtiNoise}{Chr19\_5K\_10}
+\newcommand{\OnePhaseone}{Phase1\_Chr1}
+\newcommand{\NinteenPhaseone}{Phase1\_Chr19}
+\newcommand{\SevenPhaseone}{Phase1\_Chr1-7}
+\newcommand{\OnePhasethree}{Phase3\_Chr1}
+\newcommand{\ThreePhasethree}{Phase3\_Chr1-3}
+\newcommand{\FullPhasethree}{Phase3\_Chr1-22}
 
@@
 \subsection*{Clustering in Mahout uses less memory than in R}
-In this section, we compare the runtime between the k-means implementation in R and Mahout for the small \OneReal\ dataset. 
+In this section, we compare the runtime between the k-means implementation in R and Mahout for the small \NinteenPhaseone\ dataset. 
 To do this we measured the runtime separately for pre-processing stage as well as the k-means clustering stage itself (see Methods). 
@@
 Including more regions for the similarity calculation will increase the cluster performance, however also increases the memory consumption. 
-R therefore was not able to perform the clustering on even the medium sized dataset (\SixReal , 15 Million variants), we hence only report resource consumption and accuracy for Mahout.
+R therefore was not able to perform the clustering on even the medium sized dataset (\SevenPhaseone , 15 Million variants), we hence only report resource consumption and accuracy for Mahout.
 On this dataset Mahout's pre-processing takes 325 minutes, with 13GB of physical and 37GB of virtual memory, and the clustering takes 80 minutes.
-The clustering accuracy has increased by 35\% to 0.812 with 10 clusters compared to the performance on \OneReal\ for the 4-group-based population task (see Table~\ref{datasetsAcc}).
+The clustering accuracy has increased by 35\% to 0.812 with 10 clusters compared to the performance on \NinteenPhaseone\ for the 4-group-based population task (see Table~\ref{datasetsAcc}).
 
 Utilizing the whole dataset hence would likely increased the performance further.
-However, on \FullReal\ (38 Million variants) Hadoop returns an 'out-of-memory' error on the name-node before the jobs can even be distributed to the mappers.
+However, on \FullPhasethree\ (38 Million variants) Hadoop returns an 'out-of-memory' error on the name-node before the jobs can even be distributed to the mappers.
 %TODO: is this a correct statement?
@@
 In this section we investigate Mahout's ability to scale with more samples.
-The resource consumption remains virtually identical between \SixReal\ and the order of magnitude larger dataset \SixArtiExact, which takes 33 minutes, with 6GB of physical and 39GB of virtual memory, and the clustering takes 23 minutes
+The resource consumption remains virtually identical between \SevenPhaseone\ and the order of magnitude larger dataset \OnePhasethree, which takes 33 minutes, with 6GB of physical and 39GB of virtual memory, and the clustering takes 23 minutes
 The clustering accuracy is 0.812 with 10 clusters.
@@
 Metadata available from the 1000 Genomes Consortium specifies additional attributes for each individual, including population and familial data.\\
-To be able to compare with R, we created a subset of the entire human variant dataset by using only variants from chromosome 19 (\OneReal ) where we investigate the correct grouping of the population structure.
-Scaling up using Mahout, we investigate how the accuracy improves when using more variants from the genome with \SixReal\ and \FullReal, which uses chromosome 1-5 in addition to 19 and the full genome, respectively.
+To be able to compare with R, we created a subset of the entire human variant dataset by using only variants from chromosome 19 (\NinteenPhaseone ) where we investigate the correct grouping of the population structure.
+Scaling up using Mahout, we investigate how the accuracy improves when using more variants from the genome with \SevenPhaseone\ and \FullPhasethree, which uses chromosome 1-5 in addition to 19 and the full genome, respectively.
 We also investigate whether we can accurately predict the more detailed family association using these datasets. 
@@
 
-\subsection*{Hardware}
-We processed the data using a Portable Batch System (PBS) cluster. We allocated one node and 64GB memory for the comparison between Mahout and R.
-Each node has eight CPU cores, however both applications run a single thread for pre-processing the data.
-%TODO check numbers are correct ?
-For the \SixReal\ and \FullReal\ dataset, we ran the job on a PBS cluster of 5 nodes, with twelve CPU cores and 48GB of memory per node.
+\subsection*{Hardware and Software}
+We processed the data using a cluster of five VMs on the Microsoft Azure cloud service.
+Three of the VMs, the 'worker nodes', each have 14GB memory and and 8 CPU cores.
+These three VMs run the Hadoop2 NodeManager.
+The NodeManager manages the Yet Another Resource Negotiator (YARN) 'containers', where YARN allows the user
+to specify the resources allocated to each container.
+The fourth VM runs the HDFS NameNode, ResourceManager and monitoring services, and the fifth VM, a small, single core machine, runs the Cloudera manager service.
+With this hardware setup, we can run 24 YARN containers with 1.5GB memory per container, or 12 YARN containers with 3GB memory per container.
+Storage consists of a 3TB Hadoop Distributed File System (HDFS) partition distributed across the three worker nodes. Each of these nodes runs a DataNode with a 1TB disk.
+
+We used a single VM for the comparison between Mahout and R.
 
@@
 \subsection*{Mahout K-means clustering}
-To cluster the samples from the sequence files, we use Mahout 0.9 in conjunction with Hadoop 2.2.0. 
+To cluster the samples from the sequence files, we use Mahout 0.9 in conjunction with Hadoop 2.5.0. 
 The clustering steps are the same for each dataset, however, due to the increased feature size we designate more memory to the
-YARN containers (a container being the resources allocated to each map or reduce iteration). For the first step, we specify \texttt{k},
+YARN containers. For the first step, we specify \texttt{k},
 the number of clusters. The function \texttt{RandomSeedGenerator.buildRandom} creates a new sequence file containing \texttt{k}
@@
 {\begin{tabular}{lrrrrr}\toprule
-& individuals  & variants  & population & families & noise\\
+& individuals  & variants  & population & families\\
 & & &groups&& \\\midrule
-        \OneReal & 1,092 & 816,115  & 4 & 855 & NA\\
-        \SixReal & 1,092 & 15,161,339 & 4 & 855 & NA\\
-        \FullReal\ & 1,092 & 38,219,238 & 4 & 855 & NA\\
-	\SixArtiExact\ & 5,460 & 816,115 & 4 & 855 & 0 \\ 
-	\SixArtiNoise\ & 5,460 & 816,115 & 4 & 855 & 10 \\\botrule
+        \NinteenPhaseone & 1,092 & 816,115  & 4 & 855\\
+        \OnePhaseone & 1,092 & 3,007,196  & 4 & 855\\
+        \SevenPhaseone & 1,092 & 18,984,880 & 4 & 855\\
+	\OnePhasethree\ & 2,535 & 6,468,094 & 5 & 855\\
+	\ThreePhasethree\ & 2,535 & 19,381,970 & 5 & 855\\
+	\FullPhasethree\ & 2,535 & 79,000,000 & 5 & 855\\\botrule
 \end{tabular}}{Datasets used in this study}
@@
 Tool, Dataset & accuracy \\\midrule
-      R, \OneReal & 0.614  \\ 
-        Mahout, \OneReal & 0.605\\
-        Mahout, \SixReal & 0.812 \\
-        Mahout, \SixArtiExact & 0.600  \\
-        Mahout, \SixArtiNoise & 0.140 \\\botrule
+      R, \NinteenPhaseone & 0.614  \\ 
+        Mahout, \NinteenPhaseone & 0.730\\
+        Mahout, \OnePhaseone & 0.891\\
+        Mahout, \SevenPhaseone & 0.812 \\
+        Mahout, \OnePhasethree & 0.877  \\\botrule
 \end{tabular}}{Datasets used in this study}
@@
 {\begin{tabular}{lccccc}\toprule
-&&\multicolumn{2}{ c }{Memory (kb)}&\multicolumn{2}{ c }{Time (s)} \\
-& Nodes & pmem & vmem  & pre-processing & clustering\\\midrule
-         R, \OneReal & 1 & 17,696,940 & 18,000,196 &3,341 & 7\\
-        Mahout,\OneReal & 1 & 1,831,160 & 7646232 & 1,255 & 594\\
-        Mahout, \SixReal & 5 & 13,036,836 & 36,542,380 & 19,500 & 4820 \\ 
-        Mahout, \SixArtiExact & 3 & 5,804,660 & 39,075,548 & 2,024 & 1395 \\\botrule
+&&\multicolumn{2}{ c }{Memory (GB)}&\multicolumn{2}{ c }{Time} \\
+& - & map & reduce  & pre-processing & clustering\\\midrule
+         R, \NinteenPhaseone & - & 1.5 & 2 &3,341 & 7\\
+        Mahout,\NinteenPhaseone & - & 1.5 & 2 & 2m 59s & 9m 12s\\
+        Mahout,\OnePhaseone & - & 1.5 & 2 & 7m 24s & 23m 9s\\
+        Mahout, \SevenPhaseone & - & 1.5 & 2 & 47m 43s & - \\ 
+        Mahout, \SevenPhaseone & - & 2 & 4 & - & 4820 \\ 
+        Mahout, \OnePhasethree & - & 1.5 & 2 & 27m 48s & 89m 42s \\
+        Mahout, \ThreePhasethree & - & 1.5 & 2 & 82m 14s & - \\
+        Mahout, \FullPhasethree & - & 1.5 & 2 & 336m 8s & - \\\botrule
 \end{tabular}}{Datasets used in this study}
@@
 \begin{figure}[!tpb]%figure1
-\centerline{\includegraphics[type=pdf,ext=.pdf,read=.pdf, scale=0.40]{signature}}
+\centerline{\includegraphics[type=pdf,ext=.pdf,read=.pdf, scale=0.40]{images/signature}}
         \label{fig:sign}

### 2014-12-17 vcf-clusterer/src/net/aidan/KmeansClusterer.java
latest code

@@ -30,6 +30,7 @@ import org.apache.hadoop.fs.FileSystem;
 import org.apache.hadoop.fs.Path;
 import org.apache.hadoop.io.SequenceFile;
 import org.apache.hadoop.io.Text;
+import org.apache.hadoop.util.GenericOptionsParser;
 import org.apache.hadoop.util.Tool;
 import org.apache.hadoop.util.ToolRunner;
 import org.apache.mahout.math.DenseVector;
@@ -46,300 +47,75 @@ import org.apache.mahout.clustering.kmeans.Kluster;
 import org.apache.mahout.clustering.kmeans.KMeansDriver;
 import org.apache.mahout.clustering.kmeans.RandomSeedGenerator;
 import org.apache.mahout.clustering.streaming.mapreduce.StreamingKMeansDriver;
+import org.apache.mahout.common.distance.DistanceMeasure;
 import org.apache.mahout.common.distance.EuclideanDistanceMeasure;
 import org.apache.mahout.common.distance.ManhattanDistanceMeasure;
 import org.apache.mahout.math.hadoop.stochasticsvd.SSVDHelper;
 
-public class kmeansClusterer extends Configured implements Tool {
-	public kmeansClusterer() {}
+public class KmeansClusterer extends Configured implements Tool {
+	public KmeansClusterer() {}
 	
 	// Some variables     
-	public static final String OUTPUT_DIRECTORY = "clustering";
-	public static final boolean SINGLE_MACHINE = false;
-	public static final int k = 10;
+	public static String INPUT_DIRECTORY;
+	public static String INTERMEDIATE_DIRECTORY;
+	public static String OUTPUT_DIRECTORY;
+	public static int k;
 	public static final int FEATURE_SIZE = 15161339;
 	public static long end = 0;
 	public static long start = System.currentTimeMillis();
+	
+	
 	public int run(String[] args) throws Exception {
-
-		
-		
-		Configuration conf = super.getConf();
-		
-		/*
-		 * This adds the configuration files to 'conf'. Shouldn't have to do this.
-		 * Stopped loading them automatically when I switched to Hadoop2.2
-		 * It may work whithout these lines now, with other changes I made,
-		 * but I haven't tested it yet. 
-		 */
[…]

### 2014-12-17 vcf-clusterer/src/net/aidan/outputFormatter.java

@@ -7,6 +7,7 @@ import java.io.PrintWriter;
 import java.io.StringWriter;
 import java.util.LinkedHashMap;
 import java.util.Map;
+import java.util.Scanner;
 
 import org.apache.hadoop.conf.Configuration;
 import org.apache.hadoop.fs.FileStatus;
@@ -16,32 +17,44 @@ import org.apache.hadoop.io.IntWritable;
 import org.apache.hadoop.io.SequenceFile;
 import org.apache.mahout.clustering.Cluster;
 import org.apache.mahout.clustering.classify.WeightedPropertyVectorWritable;
+import org.apache.mahout.clustering.classify.WeightedVectorWritable;
 import org.apache.mahout.math.NamedVector;
 
 public class outputFormatter {
 	String outputFile = "resultFileCluster.txt";
 	LinkedHashMap<String,String> mpIdData = new LinkedHashMap<String, String>();
 	
-	public void fileRead(Configuration conf) {
-		System.out.println("Writing summary to "+ outputFile);
+	public void fileRead(Configuration conf, String path, String namesFile) {
+		System.out.println("Writing summary to "+ outputFile + "...");
 		try {
 			FileSystem fs = FileSystem.get(conf);
-			Path outPath = new Path("clustering/output/" + Cluster.CLUSTERED_POINTS_DIR + "/");
-			FileStatus[] status = fs.listStatus(outPath);
+			Path inPath = new Path(path);
+			FileStatus[] status = fs.listStatus(inPath);
 			
 			IntWritable key = new IntWritable();
 			WeightedPropertyVectorWritable value = new WeightedPropertyVectorWritable();
 			
+			String content = new Scanner(new File(namesFile)).useDelimiter("\\Z").next();
+			String[] sampleNames = content.split(",");
+			
+			System.out.println("Number of individuals: " + sampleNames.length);
+			
+			
+			
 			for (FileStatus s: status) {
 				Path newPath = s.getPath();
 				if (!newPath.toString().contains("part-m-")){
-					System.out.println(newPath);
 					continue;
 				}
+				System.out.println(newPath);
 				SequenceFile.Reader reader = new SequenceFile.Reader(fs, newPath, conf);
 				while (reader.next(key, value)) {
 					NamedVector vect = (NamedVector) value.getVector();
-					String vecName = vect.getName();
+					
+
+					//String vecName = vect.getName();
+					String vecName= sampleNames[Integer.parseInt(vect.getName())];
+
[…]

### 2014-12-17 vcf-clusterer/src/net/aidan/pedigreeMatcher.java

@@ -13,6 +13,7 @@ import java.util.Set;
 
 public class pedigreeMatcher {
 	HashMap<String, String[]> hm = new HashMap<String, String[]>();
+	
 	public pedigreeMatcher(String pedFile) throws IOException {
 		BufferedReader ins =
 				new BufferedReader(new FileReader(pedFile));
@@ -23,6 +24,13 @@ public class pedigreeMatcher {
 		}
 		ins.close();
 	}
+	
+	
+	
+	
+	
+	
+	
 	
 	//Why??
 	public void printPedigree() {
@@ -71,21 +79,28 @@ public class pedigreeMatcher {
 			String[] sampleArray = samples[1].split(",");
 			
 			for (String s:sampleArray){
-				truth+=clusterId+",";
-				if (s.contains("CHB")|s.contains("JPT")|s.contains("CHS")|s.contains("ASN")) {
-					clust += 0+",";
+				if (s.contains("CHB")|s.contains("JPT")|s.contains("CHS")|s.contains("CDX")|s.contains("KHV")|s.contains("EAS")) {
+					truth += 0+",";
+					clust += clusterId+",";
 					
 				} else if (s.contains("CEU")|s.contains("TSI")|s.contains("FIN")|s.contains("GBR")|s.contains("IBS")|s.contains("EUR")) {
-					clust += 1+",";
+					truth += 1+",";
+					clust += clusterId+",";
 					
-				} else if (s.contains("YRI")|s.contains("LWK")|s.contains("ASW")|s.contains("ACB")|s.contains("AFR")) {
-					clust += 2+",";
+				} else if (s.contains("YRI")|s.contains("LWK")|s.contains("GWD")|s.contains("MSL")|s.contains("ESN")|s.contains("ASW")|s.contains("ACB")|s.contains("AFR")) {
+					truth += 2+",";
+					clust += clusterId+",";
 					
-				} else if (s.contains("MXL")|s.contains("PUR")|s.contains("CLM")|s.contains("AMR")) {
-					clust += 3+",";
+				} else if (s.contains("MXL")|s.contains("PUR")|s.contains("CLM")|s.contains("PEL")|s.contains("AMR")) {
+					truth += 3+",";
+					clust += clusterId+",";
 					
-				}else {
-					System.out.println(s);
+				} else if (s.contains("GIH")|s.contains("PJL")|s.contains("BEB")|s.contains("STU")|s.contains("ITU")|s.contains("SAS")) {
+					truth += 4+",";
[…]

### 2014-12-17 vcf-clusterer/src/net/aidan/VcfToSeq.java
latest

@@ -33,19 +33,19 @@ import org.apache.mahout.math.VectorWritable;
 
 
 
-public class csvClusterer extends Configured implements Tool {
+public class VcfToSeq extends Configured implements Tool {
 	//public csvClusterer() {}
 	
 	// Some variables
 	public static String[] idArray;
-	public static final String SEQUENCE_OUT_DIRECTORY = "clustering";
+	public static String SEQUENCE_OUT_DIRECTORY = "clustering";
 	public static String INPUT_DIRECTORY;
 	public static final boolean SINGLE_MACHINE = false;
 	public static final int k = 50;
 	public static final int FEATURE_SIZE = 15161339;
 	public static long end = 0;
 	public static long start = System.currentTimeMillis();
-	private static final Log log = LogFactory.getLog(csvClusterer.class);
+	private static final Log log = LogFactory.getLog(VcfToSeq.class);
 
 		
 	public static class TokenizerMapper extends Mapper<LongWritable, Text, Text, VectorWritable>{
@@ -81,19 +81,25 @@ public class csvClusterer extends Configured implements Tool {
 		public int individualId;
 
 		public void map(LongWritable key, Text value, Context context) throws IOException, InterruptedException {
-			
+			String variantString;
+			String[] alleles;
+			double variantDouble;
 			String line = value.toString();
 			if (String.valueOf(line.charAt(0)).equals("#")){
 				return;
 			}
 
 			String[] itr = line.split("\\s+");
-			//System.out.println((int) (key.get()/8000) + " " + itr[1]);
-			location = Integer.parseInt(itr[0]+"0000000")+(int) key.get()/8000;
+			
+			int chrId = Integer.parseInt(itr[0]+"0000000")-10000000;
+			location = chrId+ (int) (key.get()/10000);
+		
 			int l = itr.length;
 			for (int i = 9  ; i < l; i++) {
-				String variantString = itr[i];
-				double variantDouble = Double.parseDouble(variantString.split("\\|")[0]);
+				variantString = itr[i];
+				alleles = variantString.split("\\|");
+				variantDouble = Math.sqrt(Double.parseDouble(alleles[0])+Double.parseDouble(alleles[1]));
+				
 				if (variantDouble != 0) {
 					variant.set(variantDouble);
 					individualId = i-9;
[…]

### 2014-12-17 vcf-clusterer/src/net/aidan/GetResults.java
latest

@@ -1,5 +1,87 @@
 package net.aidan;
 
-public class createCenters {
+import java.io.IOException;
 
+import org.apache.hadoop.conf.Configuration;
+import org.apache.hadoop.conf.Configured;
+import org.apache.hadoop.util.GenericOptionsParser;
+import org.apache.hadoop.util.Tool;
+import org.apache.hadoop.util.ToolRunner;
+
+public class GetResults extends Configured implements Tool {
+	public int run(String[] args) throws Exception {
+		Configuration conf = this.getConf();
+		GenericOptionsParser optionParser = new GenericOptionsParser(conf, args);
+		String[] remainingArgs = optionParser.getRemainingArgs();
+		int mode = 0;
+		if (remainingArgs.length > 1) {
+			mode = Integer.parseInt(remainingArgs[1]);
+		}
+		
+		String sample_names = "";
+		if (remainingArgs.length > 2) {
+			sample_names = "data/" + remainingArgs[2];
+		}	
+		
+		//Writes clusters with their associated samples to "resultFileCluster.txt"
+		switch (mode) {
+		case 1:
+			sequenceToText(conf, remainingArgs[0], sample_names);
+			break;
+		case 2:
+			textToRandArray();
+			break;
+		default:
+			sequenceToText(conf, remainingArgs[0], sample_names);
+			textToRandArray();
+			break;		
+		}
+
+		//Other unrelated stuff
+		//String[] args2 = new String[] {
+		//		"-i",OUTPUT_DIRECTORY + "/points",
+		//		"-o",OUTPUT_DIRECTORY + "/matrix"};
+		//RowIdJob.main(args2);
+		
+		//DistributedRowMatrix m = new DistributedRowMatrix(new Path(OUTPUT_DIRECTORY + "/matrix"), new Path(OUTPUT_DIRECTORY + "/tmp"), 1092, 49970);
+		//m.setConf(conf);
+		//m.columnMeans();
+		
+		
+		//Path[] path = new Path[1];
+		//path[0] = new Path(OUTPUT_DIRECTORY + "/points");
+		//int k2 = 2;
+		//SSVDSolver ssvdsolver = new SSVDSolver(conf, path, new Path(OUTPUT_DIRECTORY + "/SSVTOUT"), 4, k2, 2, 3);
+		//ssvdsolver.setOverwrite(true);
+		//ssvdsolver.run();
+		//Vector svalues = ssvdsolver.getSingularValues().viewPart(0, k2);
[…]

### 2014-12-18 doc/publications/journal2014/paper.tex
Additions

@@
 However, the real benefit lies in scaling beyond R's capability to a population-size analysis.
-We successfully clustered more than 5000 individuals each having more than 15 Million variants.
+We successfully clustered more than 2,500 individuals each having more than 19 Million variants.
 
@@
 At the same time, whole genome sequencing enables the inclusion of rare or even somatic mutations in the analysis, increasing the feature space by orders of magnitude.
-This drastic increase in both sample numbers and features per sample requires a massively parallele approach to data processing.
+This drastic increase in both sample numbers and features per sample requires a massively parallel approach to data processing.
 
@@
 
-To demonstrate the capability we cluster variant datasets from the 1000 genomes project to determine population structure as well as ancestry using the k-means clustering algorithm available in Mahout.
+To demonstrate the capability we cluster variant datasets from the 1000 genomes project to determine population structure using the k-means clustering algorithm available in Mahout.
 In the first section we benchmark the performance and accuracy of Mahout's implementations against a standard R based implementation on a reduced version of the data and in the last two sections we investigate how Mahout scales to either more samples or more features per samples.
@@
 \subsection*{Clustering in Mahout uses less memory than in R}
-In this section, we compare the runtime between the k-means implementation in R and Mahout for the small \NinteenPhaseone\ dataset. 
-To do this we measured the runtime separately for pre-processing stage as well as the k-means clustering stage itself (see Methods). 
-Pre-processing the data to Mahout's sequence files takes 21 minutes, uses 1.8GB physical memory (pmem) and 7.6GB virtual memory (vmem).
-Conversely,  pre-processing the data to R's in memory matrix takes 56 minutes, uses 17.7GB pmem and 18GB vmem. 
-The size of the Mahout sequence file, which is saved to disk, is 835MB. 
-K-means clustering the data, however, is faster with R, using 7s, whereas Mahout takes 10 minutes. See Table~\ref{datasetsTime}. \\
-Due to the small number of classifying variants, cluster performance is very poor for both methods, with the best Adjusted Rand Index score for clustering individuals in one of 4 populations (see Methods) being 0.769 for 20 clusters in Mahout and 0.614 with 10 clusters in R (see Table~\ref{datasetsAcc}).
+In this section, we compare the runtime between the k-means clustering implementation in R and Mahout for the small \NinteenPhaseone\ dataset (see Methods). 
+We measured the runtime separately for pre-processing stage as well as the clustering stage. 
+With Hadoop, pre-processing the data takes approximately 3 minutes, with an allocation of up to 2GB memory per container.
+The resulting data is stored as a Hadoop 'SequenceFile' of about 835MB in size.
+Conversely,  pre-processing the data to R's in-memory matrix takes 56 minutes and uses 17.7GB memory. 
+K-means clustering the data, however, is faster with R, using 7s, whereas Mahout takes just over 9 minutes. See Table~\ref{datasetsTime}. \\
+Due to the relatively small number of variants, cluster performance is very poor for both methods, with the best Adjusted Rand Index score for clustering individuals in one of 4 populations (see Methods) being 0.730 in Mahout and 0.614 in R (see Table~\ref{datasetsAcc}).
 
-Although the Mahout pipeline is about 25 minutes faster than R's, the actual clustering step takes two orders of magnitude longer. 
-This is likely due to the additional overhead involved in setting up the Hadoop nodes, spawning the map and reduce instances, and distributing the data between the instances. 
-R on the other hand, already had the data in memory, and only had to process that data. 
-In the case of this comparison, the data set is likely too small to truly reap the benefit of Hadoop's MapReduce algorithm. 
-The main benefit of Mahout in this scenario is the memory consumption with nearly 10-fold reduction for Mahout compared to R.
+Although the overall Mahout pipeline is about 43 minutes faster than R's, Mahout's k-means clustering takes two orders of magnitude longer. 
+This is likely due to the additional overhead involved in spawning the map and reduce processes, and distributing the data between nodes.
+R, on the other hand, already has the entire matrix in-memory, on a single machine, this speeds up the clustering significantly as no disk-access or networking is required.
+However, the amount of memory available on a single machine can quickly becomes a limiting factor.
+As an example, the entire set of 1000 Genome Project Phase 3 variants would require approximately 1.6TB to be completely loaded into memory as a double-precision matrix (see Table~\ref{datasetsAcc}).
+
+Also, although Hadoop, in this case, has access to a total of 36GB of memory distributed over 3 machines, the container size is only 2GB. Therefore, we could
+easily run the same job on a single 12GB node, at the expense of time.
+
+This example demonstrates the potential for scalability in Hadoop, compared to single machine solutions, in respect to it's potential applications to genomics.
+We will proceed to demonstrate that Mahout can expand with relative ease to encompass larger data, although this depends on the task and how it can be parallelized.
 
 % SCALING UP FEATURES IN MAHOUT
-\subsection*{Mahout scales to 15 Million genomic variants}
-In this section we demonstrate Mahout's scaling ability by increasing the numbers of features, here variants per sample. 
-Including more regions for the similarity calculation will increase the cluster performance, however also increases the memory consumption. 
-R therefore was not able to perform the clustering on even the medium sized dataset (\SevenPhaseone , 15 Million variants), we hence only report resource consumption and accuracy for Mahout.
-On this dataset Mahout's pre-processing takes 325 minutes, with 13GB of physical and 37GB of virtual memory, and the clustering takes 80 minutes.
-The clustering accuracy has increased by 35\% to 0.812 with 10 clusters compared to the performance on \NinteenPhaseone\ for the 4-group-based population task (see Table~\ref{datasetsAcc}).
+\subsection*{Hadoop scales to 19 Million genomic variants}
+In this section we demonstrate Mahout's scaling ability by increasing the numbers of features (variants per sample).
 
-Utilizing the whole dataset hence would likely increased the performance further.
-However, on \FullPhasethree\ (38 Million variants) Hadoop returns an 'out-of-memory' error on the name-node before the jobs can even be distributed to the mappers.
-%TODO: is this a correct statement?
-It is not possible to use Mahout's standard implementation for this dataset, as the supported Hadoop-setup only allows the communication between nodes with up to 8GB of memory and a scaling beyond that would have been required.
+R therefore was not able to perform the clustering on even the second smallest dataset (\OnePhaseone , 3 Million variants).
+This dataset, stored as a double-precision matrix would consume approximately 26GB memory, hence why the in-memory implementation of R fails.
+Therefore we will only
+report the results for Hadoop.
+On this dataset Hadoop's pre-processing takes 7 minutes, using the same container size as the smaller dataset. Clustering takes 23 minutes.
+The clustering accuracy increased by 35\% to 0.891, compared to the performance on \NinteenPhaseone\ (see Table~\ref{datasetsAcc}).
 
-While R does not scale to either of the two orders of magnitude incased datasets, Mahout copes will with up to 15 Million variants.
-Going beyond this would require an implementation that also paralizes the individual variants as memory directly scales with the size of this feature vector.
+To perform k-means clustering on the entire dataset in a reasonable amount of time would require a larger cluster with more memory per container.
+This is because each Map task loads the entire feature vector for the clusters into memory.
+Although we could increase the container size to the full 12GB per node, this would mean only 3 containers (1 per node) could run simultaneously, rather than the maximum of 18, and would greatly increase the processing time.
+Pre-processing the data, however, does scale to the entire feature-set, without an increase in container memory. Having the data in the distributed
+SequenceFile format opens the door to other distributed analysis techniques.
+
+While R does not scale to either of the two orders of magnitude incased datasets, We have demonstrated Hadoop to cope well with up to 70 Million variants.
+Although k-means clustering with a feature-set of this size is not feasible on a compute cluster of this size, clustering a smaller subset of the genome is a trivial task.
 
 % SCALING UP SAMPLES IN MAHOUT
-\subsection*{Mahout scales effortless to an order of magnitude more samples}
-In this section we investigate Mahout's ability to scale with more samples.
-The resource consumption remains virtually identical between \SevenPhaseone\ and the order of magnitude larger dataset \OnePhasethree, which takes 33 minutes, with 6GB of physical and 39GB of virtual memory, and the clustering takes 23 minutes
-The clustering accuracy is 0.812 with 10 clusters.
+\subsection*{Hadoop scales effortless to an order of magnitude more samples}
+In this section we investigate Mahout's ability to scale with more samples, i.e. the 1092 individuals in the Phase 1 dataset and the 2535 individuals in the Phase 3 dataset.
+We used the \SevenPhaseone\ and the \ThreePhasethree\ data-sets specifically because they both contain a similar number of variants (19 million).
+The container size required for pre-processing the data is identical for both datasets, with the Phase 3 dataset requiring less than double the time of the smaller Phase 1 dataset.
+In this case, due to the number of Feature Vectors increasing, rather than the feature size, the memory requirements for k-means clustering should remain similar, as the cluster centres loaded into the containers' memory are approximately the same size.
+This proved to be the case, as 4GB allocated per container was the minimum for both datasets. The time to complete clustering was much higher though,
+especially due to the fact that only 9 4GB containers can run simultaneously, rather than the 24 1.5GB containers. 
+The clustering accuracy is 0.816 for the Phase 1 dataset and 0.xxx for the Phase 3 dataset.
 
@@
 \section*{Conclusion}
-In this paper we have applied Mahout's k-means clustering to the task of genotype based population and family association. 
-We provide an interface from VCF format to Mahout and a toolset to interpret Mahout's clustering output (see Figure 1)
-We find that on small datasets the actual clustering stage is two orders of magnitudes slower in Mahout compared to R (10 minutes vs 7 seconds), likely due to the overhead of setting up and co-ordinating the communication between the Hadoop nodes.
+In this paper we have applied Mahout's k-means clustering to the task of genotype based population. 
+We provide an interface from VCF format to Hadoop and Mahout, and a toolset to interpret Mahout's clustering output (see Figure 1)
+We find that on small datasets the actual clustering stage is two orders of magnitudes slower in Mahout compared to R (9 minutes vs 7 seconds), likely due to the overhead of setting up and co-ordinating the communication between the Hadoop nodes.
 However, increasing either the number of samples or the number of features per sample quickly exceeds the ability of R.
-Mahout on the other hand scales effortlessly to five times more samples (5,460) and copes well with 15 Million features per samples.
-As memory consumption directly scales with the number of features per sample, we observe that Mahout's standard set-up of Hadoop does not scale to a whole exome dataset (38 Million features per sample).
+Mahout on the other hand scales effortlessly to more than twice the number samples (2,535) and can process data with over 79 Million features per samples.
+As the memory required for k-means clustering directly scales with the number of features per sample, we observe that Mahout's implementation
+of k-means does not scale to the entire exome dataset, unless memory per container were increased beyond the limits of our cluster.
 However, this might be resolved once Mahout moves from MapReduce to a domain-specific language for linear algebraic operations on top of Apache Spark \url{http://spark.apache.org/}.
@@
 \subsection*{Datasets}
-We used variant call format (VCF) files from the 1000 Genomes Project.
-These VCF files contain information about the genetic variants between each of the 1092 individuals and a reference genome derived from GRCh37. 
-Each of the VCF files are partitioned by chromosome.
+We used variant call format (VCF) files from phase 1 and 3 of the 1000 Genomes Project.
+The VCF files contain information about the genetic variants between each of the 1092 and 2,535 individuals respectively and a reference genome derived from GRCh37. 
+Each of the VCF files contain the variants in one chromosome, and we only use the 22 autosomes.
 Metadata available from the 1000 Genomes Consortium specifies additional attributes for each individual, including population and familial data.\\
-To be able to compare with R, we created a subset of the entire human variant dataset by using only variants from chromosome 19 (\NinteenPhaseone ) where we investigate the correct grouping of the population structure.
-Scaling up using Mahout, we investigate how the accuracy improves when using more variants from the genome with \SevenPhaseone\ and \FullPhasethree, which uses chromosome 1-5 in addition to 19 and the full genome, respectively.
-We also investigate whether we can accurately predict the more detailed family association using these datasets. 
-In addition, we demonstrate Mahout's scalability by increasing the number of samples in the study to 5,460 by duplicating the original 1092 samples 5 times with zero and 10\% noise levels, i.e.
-randomly changing 10\% of the genotypes.
+We used different subsets of the data for different comparisons.
+For example, to be able to compare Hadoop with R, we used a subset of the entire human variant dataset containing only chromosome 19 variants (\NinteenPhaseone ).
+Scaling up using Hadoop, we investigate how the accuracy improves when using more variants from the genome with \SevenPhaseone\ and \FullPhasethree, which include chromosome 1-7 and the full genome, respectively.
 Table~\ref{datasets} contains an overview of the datasets.
@@
 Mahout requires Hadoop's `SequenceFiles' as input to its machine learning algorithms. SequenceFiles are binary flat files consisting of key/value pairs. In this case, each of the keys is a
-label for an individual. The associated value for each key is a `sparse vector' representing the individual's genomic variants. We use sparse vectors, rather than
-dense vectors, to significantly reduce processing time and required disk space. Each sparse vector consists of two separate vectors. An item in the first vector is
-effectively a key for an item at the same position in the send vector. The key represents a genomic location, where its value represents the variant. Zero values (positions where there is no variant)
+label for an individual, where the associated value for each key is a `sparse vector' representing the individual's genomic variants. We use sparse vectors, rather than
+dense vectors, to significantly reduce required memory and disk space. Each sparse vector consists of two separate vectors. An item in the first vector is
+effectively a key for an item at the same position in the second vector.
+The key represents a genomic location, where its value represents the variant.
+Zero values (positions where there is no variant)
 are omitted. Therefore, the fewer variants in an individual's genome, the smaller the sparse vector size. In comparison, `dense vectors' are single arrays that
-include every value. Therefore, the size of a dense vector isn't reduced with a large number of non-variants.
+include every position where a variant may fall, whether or not there actually is a variant. Therefore, the size of a dense vector isn't reduced with a large number of non-variants.
 
 The values of the sparse vectors, which, as previously mentioned, represent an individuals genomic variants, are a \texttt{double} java type.
-We set the value of each double to the Hamming distance between each variant and the reference genome. Therefore, a value identical to the reference will be 0 (and omitted from the sparse vector),
-a heterozygous variant will be 1 and a homozygous variant will be 2.
+We set the value of each double to the sum of the square root of the Hamming distance between each variant allele and the reference genome.
+Therefore, a value identical to the reference, 0|0, will be 0 (and omitted from the sparse vector),
+a heterozygous variant, 0|1 or 1|0, will be 1 (sqrt1) and a homozygous variant, 1|1, will be 2 (sqrt1 + sqrt1).
+The square root of the each allele prevents variants such as 1|1 colliding with the variant 2|0.
 
-We wrote a custom `MapReduce' class to convert VCF files to the aforementioned sequence files. This should work on any number of files,
+We wrote a custom `MapReduce' class in Java to convert VCF files to the aforementioned sequence files. This is scalable and should work on any number of files,
 of any file size; limited only by disk space and time.
 The `map' step creates key value pairs where the primary key is an individual, the secondary key is a variant location, and the value is the variant.
-We used a secondary key as this allowed us to implement custom functions to collect and sort the output of each mapper, based on this secondary key. This significantly reduced
-the overall processing time.
-The `reduce' step collects these sorted key/value pairs by primary key (individual ID) and adds the secondary key (location) and value (variant) to a sparse array.
-The algorithm
-saves these sparse arrays and keys to disk in the aforementioned SequenceFile format.
+We used a secondary key as this allowed us to implement custom 'collect' and 'sort' classes to process the output of each mapper, based on this secondary key.
+This significantly reduced the overall processing time.
+The `reduce' step processes the key/value pairs, which were collected by their primary key (individual ID), and sorted by their secondary key (location), and  adds the values (variants) to a sparse array.
+The algorithm saves these sparse arrays and keys to disk in the aforementioned SequenceFile format.
 
 \subsection*{Mahout K-means clustering}
-To cluster the samples from the sequence files, we use Mahout 0.9 in conjunction with Hadoop 2.5.0. 
+To cluster the samples from the sequence files, we use Mahout 0.9 with Hadoop 2.5.0. 
 The clustering steps are the same for each dataset, however, due to the increased feature size we designate more memory to the
@@
 We use `readVcf' from the R library, `VariantAnnotation', to iteratively read in the VCF file. For each iteration, we convert the data to a matrix and
-change the values for each variant to the Hamming distance, as mentioned previously. We then transpose the matrix, so individuals are represented
+convert the values for each variant to a double, as mentioned previously. We then transpose the matrix, so individuals are represented
 by rows, and append the matrix to that from the previous iteration. Once the matrix contains all of the variants from the VCF file, we cluster the data
@@
 \subsection*{Performance}
-We recorded the memory usage and run-time from the PBS logs. For the Mahout jobs, we queued separate jobs
-for pre-processing and k-means clustering to get an accurate representation of the time and memory requirements for each stage.
+We collected the job time from the 'Elapsed' time as reported by the Hadoop JobHistory server. This is measured from the job's start time to finish time.
+Because k-means clustering performs multiple iterations, and therefore multiple MapReduce jobs, the time we reported is the sum of the time each iteration consumed.
 For the R job, where we could not easily separate the pre-processing and clustering components, we printed the time between
@@
 {\begin{tabular}{lrrrrr}\toprule
-& individuals  & variants  & population & families\\
-& & &groups&& \\\midrule
-        \NinteenPhaseone & 1,092 & 816,115  & 4 & 855\\
-        \OnePhaseone & 1,092 & 3,007,196  & 4 & 855\\
-        \SevenPhaseone & 1,092 & 18,984,880 & 4 & 855\\
-	\OnePhasethree\ & 2,535 & 6,468,094 & 5 & 855\\
-	\ThreePhasethree\ & 2,535 & 19,381,970 & 5 & 855\\
-	\FullPhasethree\ & 2,535 & 79,000,000 & 5 & 855\\\botrule
-\end{tabular}}{Datasets used in this study}
+& individuals & variants & size (GB) & population\\
+& & & &groups& \\\midrule
+        \NinteenPhaseone & 1,092 & 816,115 & 7.13  & 4\\
+        \OnePhaseone & 1,092 & 3,007,196 & 26.27  & 4\\
+        \SevenPhaseone & 1,092 & 18,984,880 & 165.85 & 4\\
+	\OnePhasethree\ & 2,535 & 6,468,094 & 131.17 & 5\\
+	\ThreePhasethree\ & 2,535 & 19,381,970 & 393.06 & 5\\
+	\FullPhasethree\ & 2,535 & 79,000,000 & 1602 & 5\\\botrule
+\end{tabular}}{Datasets used in this study; based on portions of the Phase 1 and Phase 3 data from the 1000 genomes project.
+Size is the approximate space required for storage of a double-precision matrix of size n,m using the formula: n*m*8bytes/10\textsuperscript{9}bytes.
+}
 \end{table}
@@
         Mahout, \OnePhaseone & 0.891\\
-        Mahout, \SevenPhaseone & 0.812 \\
+        Mahout, \SevenPhaseone & 0.816 \\
         Mahout, \OnePhasethree & 0.877  \\\botrule
-\end{tabular}}{Datasets used in this study}
+\end{tabular}}{The adjusted Rand index for each dataset.
+The 'truth' index is the hypothetical cluster-set where there is a cluster for each super-population, with individuals belonging to the same super-population grouped together.}
 \end{table}
@@
 \begin{table*}[!t]
-\processtable{Resources consumed.\label{Tab:01}}
+\processtable{Resources allocated and time taken.\label{Tab:01}}
 {\begin{tabular}{lccccc}\toprule
@@
 & - & map & reduce  & pre-processing & clustering\\\midrule
-         R, \NinteenPhaseone & - & 1.5 & 2 &3,341 & 7\\
+         R, \NinteenPhaseone & - & 1.5 & 2 & 55m 41s & 7s\\
         Mahout,\NinteenPhaseone & - & 1.5 & 2 & 2m 59s & 9m 12s\\
@@
         Mahout, \SevenPhaseone & - & 1.5 & 2 & 47m 43s & - \\ 
-        Mahout, \SevenPhaseone & - & 2 & 4 & - & 4820 \\ 
+        Mahout, \SevenPhaseone & - & 4 & 4 & - & 10h 43m 50s \\ 
         Mahout, \OnePhasethree & - & 1.5 & 2 & 27m 48s & 89m 42s \\
         Mahout, \ThreePhasethree & - & 1.5 & 2 & 82m 14s & - \\
+        Mahout, \ThreePhasethree & - & 4 & 4 & - & (in progress) \\
         Mahout, \FullPhasethree & - & 1.5 & 2 & 336m 8s & - \\\botrule
-\end{tabular}}{Datasets used in this study}
+\end{tabular}}{The resources allocated to each MapReduce yarn-container and time taken for the job(s) to complete.
+Where clustering required bigger containers than pre-processing, an additional line is included in the table for clustering, following pre-processing.}
 \end{table*}

### 2014-12-18 doc/publications/journal2014/paper.tex
comments for paper

@@
 R, on the other hand, already has the entire matrix in-memory, on a single machine, this speeds up the clustering significantly as no disk-access or networking is required.
-However, the amount of memory available on a single machine can quickly becomes a limiting factor.
+However, the amount of memory available on a single machine quickly becomes a limiting factor.
+% DB: very nice but how do you get to the 1.6TB ? 
 As an example, the entire set of 1000 Genome Project Phase 3 variants would require approximately 1.6TB to be completely loaded into memory as a double-precision matrix (see Table~\ref{datasetsAcc}).
 
+% DB: not clear you need to spell out your argument a bit more...
 Also, although Hadoop, in this case, has access to a total of 36GB of memory distributed over 3 machines, the container size is only 2GB. Therefore, we could
@@
 This example demonstrates the potential for scalability in Hadoop, compared to single machine solutions, in respect to it's potential applications to genomics. 
+% DB: is this sentence adding information?
 We will proceed to demonstrate that Mahout can expand with relative ease to encompass larger data, although this depends on the task and how it can be parallelized.
@@
 \subsection*{Hadoop scales to 19 Million genomic variants}
-In this section we demonstrate Mahout's scaling ability by increasing the numbers of features (variants per sample).
+%DB: phase1 full genome ?
+In this section we demonstrate Mahout's scaling ability by increasing the numbers of features (variants per sample) iteratively to 3 Million variants (\OnePhaseone) and 18 Million ( \SevenPhaseone\, respectively.
 
-R therefore was not able to perform the clustering on even the second smallest dataset (\OnePhaseone , 3 Million variants).
-This dataset, stored as a double-precision matrix would consume approximately 26GB memory, hence why the in-memory implementation of R fails.
-Therefore we will only
-report the results for Hadoop.
+%DB: Where is the result from \SevenPhaseone\
 On this dataset Hadoop's pre-processing takes 7 minutes, using the same container size as the smaller dataset. Clustering takes 23 minutes.
 The clustering accuracy increased by 35\% to 0.891, compared to the performance on \NinteenPhaseone\ (see Table~\ref{datasetsAcc}).
+R's in-memory implementation fails to store this dataset as a double-precision matrix would consume approximately 26GB memory, we hence only
+report the results for Hadoop.
 
+% ??
 To perform k-means clustering on the entire dataset in a reasonable amount of time would require a larger cluster with more memory per container.
@@
 In this section we investigate Mahout's ability to scale with more samples, i.e. the 1092 individuals in the Phase 1 dataset and the 2535 individuals in the Phase 3 dataset.
+%DB: ok this makes sense from a technical point of view (how fast, memory) but you should not really discuss the accuracy ...
 We used the \SevenPhaseone\ and the \ThreePhasethree\ data-sets specifically because they both contain a similar number of variants (19 million).
@@
 individuals were partitioned by their super population code (ASN, EUR, AFR, AMR) as well as the 855 families.
-%TODO: "also on \chrOnePop" on what dataset did you compare the R performance otherwise ?
-%We also compared the Mahout clusters to the
-%R clusters for the Chr1 subset data.
 
@@
 
-\begin{table}[!t]
-\processtable{Adjusted Rand index (accuracy).\label{Tab:01}}
-{\begin{tabular}{lcc}\toprule
-Tool, Dataset & accuracy \\\midrule
-        R, \NinteenPhaseone & 0.614  \\ 
-        Mahout, \NinteenPhaseone & 0.730\\
-        Mahout, \OnePhaseone & 0.891\\
-        Mahout, \SevenPhaseone & 0.816 \\
-        Mahout, \OnePhasethree & 0.877  \\\botrule
-\end{tabular}}{The adjusted Rand index for each dataset.
+%DB: let's join 
+%\begin{table}[!t]
+%\processtable{Adjusted Rand index (accuracy).\label{Tab:01}}
+%{\begin{tabular}{lcc}\toprule
+%Tool, Dataset & accuracy \\\midrule
+%        R, \NinteenPhaseone & 0.614  \\ 
+%        Mahout, \NinteenPhaseone & 0.730\\
+%        Mahout, \OnePhaseone & 0.891\\
+%        Mahout, \SevenPhaseone & 0.816 \\
+%        Mahout, \OnePhasethree & 0.877  \\\botrule
+%\end{tabular}}{The adjusted Rand index for each dataset.
 The 'truth' index is the hypothetical cluster-set where there is a cluster for each super-population, with individuals belonging to the same super-population grouped together.}
-\end{table}
+%\end{table}
 
+%DB: what is the dash doing in the table ?
 \begin{table*}[!t]
 \processtable{Resources allocated and time taken.\label{Tab:01}}
-{\begin{tabular}{lccccc}\toprule
-&&\multicolumn{2}{ c }{Memory (GB)}&\multicolumn{2}{ c }{Time} \\
-& - & map & reduce  & pre-processing & clustering\\\midrule
-         R, \NinteenPhaseone & - & 1.5 & 2 & 55m 41s & 7s\\
-        Mahout,\NinteenPhaseone & - & 1.5 & 2 & 2m 59s & 9m 12s\\
-        Mahout,\OnePhaseone & - & 1.5 & 2 & 7m 24s & 23m 9s\\
-        Mahout, \SevenPhaseone & - & 1.5 & 2 & 47m 43s & - \\ 
-        Mahout, \SevenPhaseone & - & 4 & 4 & - & 10h 43m 50s \\ 
-        Mahout, \OnePhasethree & - & 1.5 & 2 & 27m 48s & 89m 42s \\
-        Mahout, \ThreePhasethree & - & 1.5 & 2 & 82m 14s & - \\
-        Mahout, \ThreePhasethree & - & 4 & 4 & - & (in progress) \\
-        Mahout, \FullPhasethree & - & 1.5 & 2 & 336m 8s & - \\\botrule
+{\begin{tabular}{ll|ccc|rr|c}\toprule
+Tool& Dataset&&\multicolumn{2}{ c |}{Memory (GB)}&\multicolumn{2}{ c |}{Time} & Accuracy\\
+&& - & map & reduce  & pre-processing & clustering & \\\midrule
+         R &\NinteenPhaseone & - & 1.5 & 2 & 55m 41s & 7s &0.614\\
+        Mahout & \NinteenPhaseone & - & 1.5 & 2 & 2m 59s & 9m 12s & 0.730\\
+         & \OnePhaseone & - & 1.5 & 2 & 7m 24s & 23m 9s & 0.891\\
+        & \SevenPhaseone & - & 1.5 & 2 & 47m 43s & - & 0.816\\  %DB: why are there two lines for \SevenPhaseone
+        & \SevenPhaseone & - & 4.0 & 4 & - & 10h 43m 50s \\ 
+        & \OnePhasethree & - & 1.5 & 2 & 27m 48s & 89m 42s & 0.877\\
+        & \ThreePhasethree & - & 1.5 & 2 & 82m 14s & - \\
+        & \ThreePhasethree & - & 4.0 & 4 & - & (in progress) \\
+        & \FullPhasethree & - & 1.5 & 2 & 336m 8s & - \\\botrule
 \end{tabular}}{The resources allocated to each MapReduce yarn-container and time taken for the job(s) to complete.
@@
 \begin{figure}[!tpb]%figure1
-\centerline{\includegraphics[type=pdf,ext=.pdf,read=.pdf, scale=0.40]{images/signature}}
+\centerline{\includegraphics[type=pdf,ext=.pdf,read=.pdf, scale=0.40]{signature}}
         \label{fig:sign}

### 2015-02-09 doc/publications/journal2014/paper.tex
Fixed 'fixme's as well as added latest data and rewirtten some sections

@@
 \newcommand{\SevenPhaseone}{Phase1\_Chr1-7}
+\newcommand{\FullPhaseone}{Phase1\_Chr1-22}
 \newcommand{\OnePhasethree}{Phase3\_Chr1}
@@
 % DB: very nice but how do you get to the 1.6TB ?
+% AO: Fixed
 As an example, the entire set of 1000 Genome Project Phase 3 variants would require approximately 1.6TB to be completely loaded into memory as a double-precision matrix (see Table~\ref{datasetsAcc}).
+This is based on the equation: n*m*8bytes/1GB (where 1GB is 10\^9 bytes).
+n*m are the matrix dimensions and 8bytes is the memory a double occupies.
 
 % DB: not clear you need to spell out your argument a bit more...
-Also, although Hadoop, in this case, has access to a total of 36GB of memory distributed over 3 machines, the container size is only 2GB. Therefore, we could
-easily run the same job on a single 12GB node, at the expense of time.
-
-This example demonstrates the potential for scalability in Hadoop, compared to single machine solutions, in respect to it's potential applications to genomics. 
-% DB: is this sentence adding information?
-We will proceed to demonstrate that Mahout can expand with relative ease to encompass larger data, although this depends on the task and how it can be parallelized.
+% AO: Fixed
+By dividing jobs up into separate 'containers', Hadoop provides great versatility in scaling options.
+Although in this case, Hadoop has access to 36GB of memory distributed over 3 machines, the size of each container is only 2GB.
+Each containers is effectively a self-contained job that is independent from any other container.
+Therefore the more resources that the user has available, the more containers that can be run in parallel, decreasing the run-time of the overall job.
+This means that our Hadoop job can scale from a single machine with minimal resources (albeit with a longer run-time), to a huge compute cluster with terabytes of RAM.
+This benefit of the MapReduce paradigm allows much greater scalability over linear jobs designed in R, for example.
 
 % SCALING UP FEATURES IN MAHOUT
-\subsection*{Hadoop scales to 19 Million genomic variants}
+\subsection*{Hadoop scales to genome-wide clustering}
 %DB: phase1 full genome ?
-In this section we demonstrate Mahout's scaling ability by increasing the numbers of features (variants per sample) iteratively to 3 Million variants (\OnePhaseone) and 18 Million ( \SevenPhaseone\, respectively.
+%AO: Fixed
+In this section we demonstrate Mahout's scaling ability by increasing the numbers of features (variants per sample) input to our algorithm to include the full 79 Million variants from the Phase 3 dataset ( \FullPhasethree\ ).
 
 %DB: Where is the result from \SevenPhaseone\
-On this dataset Hadoop's pre-processing takes 7 minutes, using the same container size as the smaller dataset. Clustering takes 23 minutes.
-The clustering accuracy increased by 35\% to 0.891, compared to the performance on \NinteenPhaseone\ (see Table~\ref{datasetsAcc}).
-R's in-memory implementation fails to store this dataset as a double-precision matrix would consume approximately 26GB memory, we hence only
-report the results for Hadoop.
+%AO: Fixed
+We tested on various arbitrarily sized datasets to demonstrate the different container sizes required by the pre-processing and clustering, as well as the time required.
+On the OnePhaseone dataset, pre-processing takes just under 10 minutes. Clustering took approximately 12 minutes and the container size was the same as with clustering the OnePhasenineteen dataset.
+We then increased the input to include the first seven chromosomes. Pre-processing increased to approximately 1 hour and clustering took about 5 hours and 45 minutes.
+Our pre-processing algorithm was successful with the same container size as the previous examples, however, we increased both map and reduce container size for the clustering step to 4GB to accommodate the increased vector size.
+Finally we used the entire phase one dataset as input. Although pre-processing now takes TIME, it was again successful with the same sized containers as previously.
 
-% ??
-To perform k-means clustering on the entire dataset in a reasonable amount of time would require a larger cluster with more memory per container.
-This is because each Map task loads the entire feature vector for the clusters into memory.
-Although we could increase the container size to the full 12GB per node, this would mean only 3 containers (1 per node) could run simultaneously, rather than the maximum of 18, and would greatly increase the processing time.
-Pre-processing the data, however, does scale to the entire feature-set, without an increase in container memory. Having the data in the distributed
-SequenceFile format opens the door to other distributed analysis techniques.
+The significantly increased time required for this larger job is due to the larger container size.
+With a container size of 6GB, only up to six containers can run simultaneously, as opposed to 24 1.5GB containers.
 
-While R does not scale to either of the two orders of magnitude incased datasets, We have demonstrated Hadoop to cope well with up to 70 Million variants.
-Although k-means clustering with a feature-set of this size is not feasible on a compute cluster of this size, clustering a smaller subset of the genome is a trivial task.
+While R does not scale to the larger datasets datasets, we have demonstrated Hadoop to cope well with pre-processing and clustering variants from 22 phase1 VCF files.
 
 % SCALING UP SAMPLES IN MAHOUT
-\subsection*{Hadoop scales effortless to an order of magnitude more samples}
-In this section we investigate Mahout's ability to scale with more samples, i.e. the 1092 individuals in the Phase 1 dataset and the 2535 individuals in the Phase 3 dataset.
+\subsection*{Hadoop scales effortless to more than twice the number of samples}
+In this section we investigate Mahout's ability to scale with more samples, i.e. the 2504 individuals from the Phase 3 dataset compared to the 1092 individuals in the Phase 1 dataset.
 %DB: ok this makes sense from a technical point of view (how fast, memory) but you should not really discuss the accuracy ...
@@
 \section*{Conclusion}
-In this paper we have applied Mahout's k-means clustering to the task of genotype based population. 
+In this paper we have applied Mahout's k-means clustering to the task of clustering individuals, based on their genomic variants. 
 We provide an interface from VCF format to Hadoop and Mahout, and a toolset to interpret Mahout's clustering output (see Figure 1)
 We find that on small datasets the actual clustering stage is two orders of magnitudes slower in Mahout compared to R (9 minutes vs 7 seconds), likely due to the overhead of setting up and co-ordinating the communication between the Hadoop nodes.
-However, increasing either the number of samples or the number of features per sample quickly exceeds the ability of R.
-Mahout on the other hand scales effortlessly to more than twice the number samples (2,535) and can process data with over 79 Million features per samples.
+However, increasing either the number of samples or the number of features per sample quickly exceeds the capability of R.
+Mahout on the other hand scales effortlessly to more than twice the number samples (2,504) and can process VCF files containing over 79 Million variants per samples.
 As the memory required for k-means clustering directly scales with the number of features per sample, we observe that Mahout's implementation
 of k-means does not scale to the entire exome dataset, unless memory per container were increased beyond the limits of our cluster.
-However, this might be resolved once Mahout moves from MapReduce to a domain-specific language for linear algebraic operations on top of Apache Spark \url{http://spark.apache.org/}.
 
@@
 \subsection*{Datasets}
-We used variant call format (VCF) files from phase 1 and 3 of the 1000 Genomes Project.
-The VCF files contain information about the genetic variants between each of the 1092 and 2,535 individuals respectively and a reference genome derived from GRCh37. 
-Each of the VCF files contain the variants in one chromosome, and we only use the 22 autosomes.
+We used variant call format (VCF) files from phases 1 and 3 of the 1000 Genomes Project.
+Each VCF file details the genetic variation (SNPs and indels) for one chromosome, between each individual and a reference genome derived from GRCh37. 
+The phase 1 and phase 3 datasets contain 1,092 and 2,504 individuals, respectively.
 Metadata available from the 1000 Genomes Consortium specifies additional attributes for each individual, including population and familial data.\\
-We used different subsets of the data for different comparisons.
-For example, to be able to compare Hadoop with R, we used a subset of the entire human variant dataset containing only chromosome 19 variants (\NinteenPhaseone ).
-Scaling up using Hadoop, we investigate how the accuracy improves when using more variants from the genome with \SevenPhaseone\ and \FullPhasethree, which include chromosome 1-7 and the full genome, respectively.
+We used different subsets of the data for different comparisons. For example, to be able to compare Hadoop with R, we used the VCF file containing chromosome 19 variants (\NinteenPhaseone ).
+Then to demonstrate scaling with Hadoop, we use subsets that contain more variants, i.e. \SevenPhaseone\ and \FullPhasethree, which include chromosome 1-7 and the full genome, respectively.
 Table~\ref{datasets} contains an overview of the datasets.
@@
 These three VMs run the Hadoop2 NodeManager.
-The NodeManager manages the Yet Another Resource Negotiator (YARN) 'containers', where YARN allows the user
-to specify the resources allocated to each container.
+The NodeManager manages the Yet Another Resource Negotiator (YARN) 'containers'.
+The user specifies the memory size of these containers, where map containers can be a different size to reduce containers.
 The fourth VM runs the HDFS NameNode, ResourceManager and monitoring services, and the fifth VM, a small, single core machine, runs the Cloudera manager service.
-With this hardware setup, we can run 24 YARN containers with 1.5GB memory per container, or 12 YARN containers with 3GB memory per container.
+With a total of 42GB memory and 24 CPU cores available between the node managers, we can run 24 YARN containers with 1.5GB memory per container, or 12 YARN containers with 3GB memory per container, in parallel.
 Storage consists of a 3TB Hadoop Distributed File System (HDFS) partition distributed across the three worker nodes. Each of these nodes runs a DataNode with a 1TB disk.
@@
 We collected the job time from the 'Elapsed' time as reported by the Hadoop JobHistory server. This is measured from the job's start time to finish time.
-Because k-means clustering performs multiple iterations, and therefore multiple MapReduce jobs, the time we reported is the sum of the time each iteration consumed.
+Because k-means clustering performs multiple iterations, and therefore multiple MapReduce jobs, the time we reported is the sum of the elapsed time of each iteration.
 For the R job, where we could not easily separate the pre-processing and clustering components, we printed the time between
@@
 {\begin{tabular}{lrrrrr}\toprule
-& individuals & variants & size (GB) & population\\
+& individuals & variants & in-memory size (GB) & population\\
 & & & &groups& \\\midrule
@@
         \SevenPhaseone & 1,092 & 18,984,880 & 165.85 & 4\\
-	\OnePhasethree\ & 2,535 & 6,468,094 & 131.17 & 5\\
-	\ThreePhasethree\ & 2,535 & 19,381,970 & 393.06 & 5\\
-	\FullPhasethree\ & 2,535 & 79,000,000 & 1602 & 5\\\botrule
+	\OnePhasethree\ & 2,504 & 6,468,094 & 131.17 & 5\\
+	\ThreePhasethree\ & 2,504 & 19,381,970 & 393.06 & 5\\
+	\FullPhasethree\ & 2,504 & 79,000,000 & 1602 & 5\\\botrule
 \end{tabular}}{Datasets used in this study; based on portions of the Phase 1 and Phase 3 data from the 1000 genomes project.
-Size is the approximate space required for storage of a double-precision matrix of size n,m using the formula: n*m*8bytes/10\textsuperscript{9}bytes.
+In-memory size is the approximate memory required to store the data as a double-precision matrix of size n,m using the formula: n*m*8bytes/10\textsuperscript{9}bytes.
 }
@@
 
-%DB: let's join 
-%\begin{table}[!t]
-%\processtable{Adjusted Rand index (accuracy).\label{Tab:01}}
-%{\begin{tabular}{lcc}\toprule
-%Tool, Dataset & accuracy \\\midrule
-%        R, \NinteenPhaseone & 0.614  \\ 
-%        Mahout, \NinteenPhaseone & 0.730\\
-%        Mahout, \OnePhaseone & 0.891\\
-%        Mahout, \SevenPhaseone & 0.816 \\
-%        Mahout, \OnePhasethree & 0.877  \\\botrule
-%\end{tabular}}{The adjusted Rand index for each dataset.
-The 'truth' index is the hypothetical cluster-set where there is a cluster for each super-population, with individuals belonging to the same super-population grouped together.}
-%\end{table}
-
-%DB: what is the dash doing in the table ?
 \begin{table*}[!t]
 \processtable{Resources allocated and time taken.\label{Tab:01}}
-{\begin{tabular}{ll|ccc|rr|c}\toprule
-Tool& Dataset&&\multicolumn{2}{ c |}{Memory (GB)}&\multicolumn{2}{ c |}{Time} & Accuracy\\
-&& - & map & reduce  & pre-processing & clustering & \\\midrule
-         R &\NinteenPhaseone & - & 1.5 & 2 & 55m 41s & 7s &0.614\\
-        Mahout & \NinteenPhaseone & - & 1.5 & 2 & 2m 59s & 9m 12s & 0.730\\
-         & \OnePhaseone & - & 1.5 & 2 & 7m 24s & 23m 9s & 0.891\\
-        & \SevenPhaseone & - & 1.5 & 2 & 47m 43s & - & 0.816\\  %DB: why are there two lines for \SevenPhaseone
-        & \SevenPhaseone & - & 4.0 & 4 & - & 10h 43m 50s \\ 
-        & \OnePhasethree & - & 1.5 & 2 & 27m 48s & 89m 42s & 0.877\\
-        & \ThreePhasethree & - & 1.5 & 2 & 82m 14s & - \\
-        & \ThreePhasethree & - & 4.0 & 4 & - & (in progress) \\
-        & \FullPhasethree & - & 1.5 & 2 & 336m 8s & - \\\botrule
-\end{tabular}}{The resources allocated to each MapReduce yarn-container and time taken for the job(s) to complete.
-Where clustering required bigger containers than pre-processing, an additional line is included in the table for clustering, following pre-processing.}
+{\begin{tabular}{ll|cc|cc|rr|c}\toprule
+Tool & Dataset & \multicolumn{2}{ c |}{Preprocesing (GB)} & \multicolumn{2}{ c |}{Clustering (GB)} & \multicolumn{2}{ c |}{Time} & Accuracy\\
+& & map & reduce  & map & reduce  & pre-processing & clustering & \\\midrule
+        R &\NinteenPhaseone & - & - & - & - & 55m 41s & 7s &0.614\\
+        Mahout & \NinteenPhaseone & 1536 & 3072 & 1536 & 3072 & 6m 10s & 8m 32s & 0.705\\
+        & \OnePhaseone & 1536 & 3072 & 1536 & 3072 & 9m 52s & 12m 08s & 0.771\\
+        & \SevenPhaseone & 1536 & 3072 & 4096 & 4096  & 60m 35s & 5h 44m & 0.716\\
+        & \FullPhaseone & 1536 & 3072 & 6144 & 6144  & 108m 47s &  & \\
+        & \OnePhasethree & 1536 & 3072 & 1536 & 3072 & 23m 27s & 21m 14s & 0.873\\
+        & \ThreePhasethree & 1536 & 3072 & 2048 & 4096 & 63m 49s & 93m 18s & 0.871\\\
+        & \FullPhasethree & 1536 & 3072 & 6144 & 6144 & 5h 36m & 21h 5m &0.623\\\botrule
+\end{tabular}}{The resources allocated to each MapReduce yarn-container and time taken for the job(s) to complete.}
 \end{table*}

### 2015-02-10 doc/publications/journal2014/paper.tex
latest changes

@@
 
-\newcommand{\OnePhaseone}{Phase1\_Chr1}
-\newcommand{\NinteenPhaseone}{Phase1\_Chr19}
-\newcommand{\SevenPhaseone}{Phase1\_Chr1-7}
-\newcommand{\FullPhaseone}{Phase1\_Chr1-22}
-\newcommand{\OnePhasethree}{Phase3\_Chr1}
-\newcommand{\ThreePhasethree}{Phase3\_Chr1-3}
-\newcommand{\FullPhasethree}{Phase3\_Chr1-22}
+\newcommand{\OnePhaseone}{phase1\_chr1}
+\newcommand{\NinteenPhaseone}{phase1\_chr19}
+\newcommand{\SevenPhaseone}{phase1\_chr1-7}
+\newcommand{\FullPhaseone}{phase3\_chr1-22}
+\newcommand{\OnePhasethree}{phase3\_chr1}
+\newcommand{\ThreePhasethree}{phase3\_chr1-3}
+\newcommand{\FullPhasethree}{phase3\_chr1-22}
 
@@
 % COMPARISON TO R
-\subsection*{Clustering in Mahout uses less memory than in R}
-In this section, we compare the runtime between the k-means clustering implementation in R and Mahout for the small \NinteenPhaseone\ dataset (see Methods). 
+\subsection*{Clustering in Mahout requires less memory than in R}
+In this section, we compare the runtime and resource requirements between the k-means clustering implementation in R and Mahout for the small \NinteenPhaseone\ dataset (see Methods). 
 We measured the runtime separately for pre-processing stage as well as the clustering stage. 
-With Hadoop, pre-processing the data takes approximately 3 minutes, with an allocation of up to 2GB memory per container.
+With Hadoop, pre-processing the data takes approximately 6 minutes, with an allocation of up to 3GB memory per 'container'.
 The resulting data is stored as a Hadoop 'SequenceFile' of about 835MB in size.
-Conversely,  pre-processing the data to R's in-memory matrix takes 56 minutes and uses 17.7GB memory. 
-K-means clustering the data, however, is faster with R, using 7s, whereas Mahout takes just over 9 minutes. See Table~\ref{datasetsTime}. \\
-Due to the relatively small number of variants, cluster performance is very poor for both methods, with the best Adjusted Rand Index score for clustering individuals in one of 4 populations (see Methods) being 0.730 in Mahout and 0.614 in R (see Table~\ref{datasetsAcc}).
+Conversely,  pre-processing the data to R's in-memory matrix takes 56 minutes and required 17.7GB memory. 
+K-means clustering the data, however, is faster with R, using 7s, whereas Mahout takes just under 9 minutes. See Table~\ref{datasetsTime}. \\
+Due to the relatively small number of variants, cluster performance is relatively poor for both methods, with the best Adjusted Rand Index score for clustering individuals in one of 4 populations (see Methods) being 0.705 in Mahout and 0.614 in R (see Table~\ref{datasetsAcc}).
 
-Although the overall Mahout pipeline is about 43 minutes faster than R's, Mahout's k-means clustering takes two orders of magnitude longer. 
+Although the overall Mahout pipeline is about 40 minutes faster than R's, Mahout's k-means clustering takes two orders of magnitude longer. 
 This is likely due to the additional overhead involved in spawning the map and reduce processes, and distributing the data between nodes.
-R, on the other hand, already has the entire matrix in-memory, on a single machine, this speeds up the clustering significantly as no disk-access or networking is required.
-However, the amount of memory available on a single machine quickly becomes a limiting factor.
+R, on the other hand, already has the entire matrix in-memory, on a single machine.
+This should speed up the clustering significantly as no disk-access or networking is required.
+However, the amount of memory available on a single machine will quickly become a limiting factor.
 % DB: very nice but how do you get to the 1.6TB ?
 % AO: Fixed
-As an example, the entire set of 1000 Genome Project Phase 3 variants would require approximately 1.6TB to be completely loaded into memory as a double-precision matrix (see Table~\ref{datasetsAcc}).
-This is based on the equation: n*m*8bytes/1GB (where 1GB is 10\^9 bytes).
-n*m are the matrix dimensions and 8bytes is the memory a double occupies.
+As an example, the entire set of 1000 Genome Project phase3 variants could require approximately 1.6TB to be completely loaded into memory as a double-precision matrix (see Table~\ref{datasetsAcc}).
+This is based on the equation:
+$^{n \times m \times 8bytes}/_{1GB}$, where $n \times m$ are the matrix dimensions, 8bytes is the memory a double occupies (i.e. each matrix element), and  $1GB$ is $10^9$bytes.
 
 % DB: not clear you need to spell out your argument a bit more...
-% AO: Fixed
-By dividing jobs up into separate 'containers', Hadoop provides great versatility in scaling options.
-Although in this case, Hadoop has access to 36GB of memory distributed over 3 machines, the size of each container is only 2GB.
-Each containers is effectively a self-contained job that is independent from any other container.
-Therefore the more resources that the user has available, the more containers that can be run in parallel, decreasing the run-time of the overall job.
+% AO: Fixed maybe
+The primary benefit of Hadoop is its scalability. This is achieved through MapReduce jobs which allows the task at hand to be separated into 'containers'.
+Although in the previous example, Hadoop had access to 36GB of memory distributed over 3 machines, the size of each container was only 3GB.
+Each of these containers is effectively a self-contained job that is independent from any other container.
+Therefore, the more resources that the user has available, the more containers that can be run in parallel, either on the same machine or separate, and in turn decreasing the run-time of the overall job.
 This means that our Hadoop job can scale from a single machine with minimal resources (albeit with a longer run-time), to a huge compute cluster with terabytes of RAM.
-This benefit of the MapReduce paradigm allows much greater scalability over linear jobs designed in R, for example.
+This benefit allows Hadoop jobs to not only scale vertically, i.e. to a larger single machine as R can, but also horizontally, to a cluster of many machines.
 
@@
 %DB: phase1 full genome ?
-%AO: Fixed
-In this section we demonstrate Mahout's scaling ability by increasing the numbers of features (variants per sample) input to our algorithm to include the full 79 Million variants from the Phase 3 dataset ( \FullPhasethree\ ).
+%AO: Fixed nearly
+In this section we demonstrate Mahout's scaling ability by increasing the numbers of features (variants per sample) input to our algorithm to include the 79 Million variants from the entire phase3 dataset (\FullPhasethree{}).
 
@@
 %AO: Fixed
-We tested on various arbitrarily sized datasets to demonstrate the different container sizes required by the pre-processing and clustering, as well as the time required.
-On the OnePhaseone dataset, pre-processing takes just under 10 minutes. Clustering took approximately 12 minutes and the container size was the same as with clustering the OnePhasenineteen dataset.
-We then increased the input to include the first seven chromosomes. Pre-processing increased to approximately 1 hour and clustering took about 5 hours and 45 minutes.
-Our pre-processing algorithm was successful with the same container size as the previous examples, however, we increased both map and reduce container size for the clustering step to 4GB to accommodate the increased vector size.
-Finally we used the entire phase one dataset as input. Although pre-processing now takes TIME, it was again successful with the same sized containers as previously.
-
-The significantly increased time required for this larger job is due to the larger container size.
-With a container size of 6GB, only up to six containers can run simultaneously, as opposed to 24 1.5GB containers.
+We tested on various arbitrarily sized datasets to demonstrate the different container sizes required by pre-processing and clustering, as well as the time required.
+On the \OnePhaseone{} dataset, pre-processing takes just under 10 minutes and clustering takes approximately 12 minutes with a container size the same as with the \NinteenPhaseone{} dataset.
+We then increased the input to include the first seven chromosomes (\SevenPhaseone{}).
+In this case, pre-processing time increased to approximately 1 hour, and clustering to about 5 hours and 45 minutes.
+Our pre-processing algorithm was successful with the same container size as the previous examples.
+However, we increased both the map and reduce container size for the clustering step to 4GB to accommodate the increased vector size.
+Finally we used the entire phase one dataset as input. Although pre-processing now takes 108 minutes, it was again successful with the same sized containers as previously.
+Clustering, however, not only required larger containers, but also took over 20 hours to complete.
+This significant increase in time required is in part due to the larger container size.
+With each container consuming 6GB of memory, only up to six containers can run simultaneously, as opposed to 24 1.5GB containers from the smaller examples.
 
@@
 \subsection*{Hadoop scales effortless to more than twice the number of samples}
-In this section we investigate Mahout's ability to scale with more samples, i.e. the 2504 individuals from the Phase 3 dataset compared to the 1092 individuals in the Phase 1 dataset.
+In this section we investigate Mahout's ability to scale with more samples, i.e. the 2504 individuals from the phase3 dataset compared to the 1092 individuals in the phase1 dataset.
 %DB: ok this makes sense from a technical point of view (how fast, memory) but you should not really discuss the accuracy ...
-We used the \SevenPhaseone\ and the \ThreePhasethree\ data-sets specifically because they both contain a similar number of variants (19 million).
-The container size required for pre-processing the data is identical for both datasets, with the Phase 3 dataset requiring less than double the time of the smaller Phase 1 dataset.
+
+The container size required for pre-processing the data is identical for both datasets, with the phase3 dataset requiring less than double the time of the smaller phase1 dataset.
 In this case, due to the number of Feature Vectors increasing, rather than the feature size, the memory requirements for k-means clustering should remain similar, as the cluster centres loaded into the containers' memory are approximately the same size.
@@
 especially due to the fact that only 9 4GB containers can run simultaneously, rather than the 24 1.5GB containers. 
-The clustering accuracy is 0.816 for the Phase 1 dataset and 0.xxx for the Phase 3 dataset.
+The clustering accuracy is 0.816 for the phase1 dataset and 0.xxx for the phase3 dataset.
 
@@
 We provide an interface from VCF format to Hadoop and Mahout, and a toolset to interpret Mahout's clustering output (see Figure 1)
-We find that on small datasets the actual clustering stage is two orders of magnitudes slower in Mahout compared to R (9 minutes vs 7 seconds), likely due to the overhead of setting up and co-ordinating the communication between the Hadoop nodes.
-However, increasing either the number of samples or the number of features per sample quickly exceeds the capability of R.
-Mahout on the other hand scales effortlessly to more than twice the number samples (2,504) and can process VCF files containing over 79 Million variants per samples.
-As the memory required for k-means clustering directly scales with the number of features per sample, we observe that Mahout's implementation
-of k-means does not scale to the entire exome dataset, unless memory per container were increased beyond the limits of our cluster.
+We find that on small datasets the clustering stage is two orders of magnitudes slower in Mahout compared to R (9 minutes vs 7 seconds), likely due to the overhead of setting up and co-ordinating the communication between the Hadoop nodes.
+However, increasing either the number of samples or the number of features per sample quickly exceeds the capability of R on this machine due to memory limitations.
+Hadoop on the other hand scales effortlessly to more than twice the number samples (2,504) and can process VCF files that contain over 79 Million variants per samples.
 
@@
 Each VCF file details the genetic variation (SNPs and indels) for one chromosome, between each individual and a reference genome derived from GRCh37. 
-The phase 1 and phase 3 datasets contain 1,092 and 2,504 individuals, respectively.
+The phase1 and phase3 datasets contain 1,092 and 2,504 individuals, respectively.
 Metadata available from the 1000 Genomes Consortium specifies additional attributes for each individual, including population and familial data.\\
-We used different subsets of the data for different comparisons. For example, to be able to compare Hadoop with R, we used the VCF file containing chromosome 19 variants (\NinteenPhaseone ).
-Then to demonstrate scaling with Hadoop, we use subsets that contain more variants, i.e. \SevenPhaseone\ and \FullPhasethree, which include chromosome 1-7 and the full genome, respectively.
+We used different subsets of the data for different comparisons. For example, to be able to compare Hadoop with R, we used the VCF file containing chromosome 19 variants (\NinteenPhaseone{}).
+Then to demonstrate scaling with Hadoop, we use subsets that contain more variants, i.e. \SevenPhaseone{} and \FullPhasethree{}, which include chromosome 1-7 and the full genome, respectively.
 Table~\ref{datasets} contains an overview of the datasets.
@@
 We processed the data using a cluster of five VMs on the Microsoft Azure cloud service.
-Three of the VMs, the 'worker nodes', each have 14GB memory and and 8 CPU cores.
-These three VMs run the Hadoop2 NodeManager.
-The NodeManager manages the Yet Another Resource Negotiator (YARN) 'containers'.
-The user specifies the memory size of these containers, where map containers can be a different size to reduce containers.
+Three of these VMs, the 'worker nodes', each have 14GB memory and and 8 CPU cores.
+These three VMs run the Hadoop2 NodeManager, which manages the Yet Another Resource Negotiator (YARN) 'containers'.
+The user is able to specify the resources allocated to each container, where map containers can have a different allocation to reduce containers.
 The fourth VM runs the HDFS NameNode, ResourceManager and monitoring services, and the fifth VM, a small, single core machine, runs the Cloudera manager service.
-With a total of 42GB memory and 24 CPU cores available between the node managers, we can run 24 YARN containers with 1.5GB memory per container, or 12 YARN containers with 3GB memory per container, in parallel.
+With a total of 42GB memory and 24 CPU cores available between the node managers, we can simultaneously run up to 24 YARN containers with 1.5GB memory per container, or 12 YARN containers with 3GB memory per container.
 Storage consists of a 3TB Hadoop Distributed File System (HDFS) partition distributed across the three worker nodes. Each of these nodes runs a DataNode with a 1TB disk.
@@
 
-We wrote a custom `MapReduce' class in Java to convert VCF files to the aforementioned sequence files. This is scalable and should work on any number of files,
+We wrote two `MapReduce' classes in Java to convert VCF files to the aforementioned sequence files. This is scalable and should work on any number of files,
 of any file size; limited only by disk space and time. 
-The `map' step creates key value pairs where the primary key is an individual, the secondary key is a variant location, and the value is the variant.
-We used a secondary key as this allowed us to implement custom 'collect' and 'sort' classes to process the output of each mapper, based on this secondary key.
-This significantly reduced the overall processing time.
-The `reduce' step processes the key/value pairs, which were collected by their primary key (individual ID), and sorted by their secondary key (location), and  adds the values (variants) to a sparse array.
-The algorithm saves these sparse arrays and keys to disk in the aforementioned SequenceFile format.
+The `map' stage of the first job parses each line of the VCF file to key value pairs.
+Each of the keys are an identical arbitrary number, where the value for each key is a sparse vector of the variants from each line.
+The variants in each vector are in the same order as they appear in the VCF file.
+In this stage, we can define parameters to exclude certain variants that may be deemed low quality, or only appear in a certain number of individuals.
+We use identical keys so that each key value pair is sent to the same reducer, and this reducer assigns a unique key to each value from the previous step.
+We require one reducer in this step to make each variant key a consecutive number, rather than an arbitrary feature, such as its genomic location.
+Because keys are represented by integers, this helps to avoid integer overflows by keeping the keys relatively small.
+
+The second job effectively transposes the output from the first job. The mappers create key value pairs where the key is composite of the individual id and variant id, and the value is the variant.
+This composite key, which is essentially a concatenation of two keys, allowed us to implement custom 'collect' and 'sort' classes to speed up the process.
+These classes divide up the key value pairs by their individual ID and sort them by their location ID, so each sorted collection of key/value pairs can be sent to a separate reducer.
+ The reducers then write the variants for each individual to a sparse vector. These are saved to disk as the value for each individual ID key.
 
@@
 The clustering steps are the same for each dataset, however, due to the increased feature size we designate more memory to the
-YARN containers. For the first step, we specify \texttt{k},
-the number of clusters. The function \texttt{RandomSeedGenerator.buildRandom} creates a new sequence file containing \texttt{k}
+YARN containers. For the first step, we specify \texttt{k}, the number of clusters. The function \texttt{RandomSeedGenerator.buildRandom} creates a new sequence file containing \texttt{k}
 random centroids from the samples sequence file. The sequence file
@@
 The value is zero for independent clusterings and one for identical clusterings. 
-We compared the Mahout k-means
-clusters and the R k-means clusters to the known clusters based on meta data from the 1000 genome project. For the known clusters,
-individuals were partitioned by their super population code (ASN, EUR, AFR, AMR) as well as the 855 families.
+We compared the Mahout k-means clusters and the R k-means clusters to the known clusters based on meta data from the 1000 genome project.
+For the known clusters, individuals from the both phase1 and phase3 datasets were partitioned by their super population code, ASN, EUR, AFR, and AMR, where phase3 included EAS and SAS instead of ASN.
 
 \subsection*{Performance}
-We collected the job time from the 'Elapsed' time as reported by the Hadoop JobHistory server. This is measured from the job's start time to finish time.
-Because k-means clustering performs multiple iterations, and therefore multiple MapReduce jobs, the time we reported is the sum of the elapsed time of each iteration.
+We recorded the job time from the 'Elapsed' time as reported by the Hadoop JobHistory server. This is measured from the job's start time to finish time.
+Because k-means clustering performs multiple iterations, and therefore spawns multiple MapReduce jobs, the time we reported is the sum of the elapsed time of each job.
 For the R job, where we could not easily separate the pre-processing and clustering components, we printed the time between
@@
 & & & &groups& \\\midrule
-        \NinteenPhaseone & 1,092 & 816,115 & 7.13  & 4\\
-        \OnePhaseone & 1,092 & 3,007,196 & 26.27  & 4\\
-        \SevenPhaseone & 1,092 & 18,984,880 & 165.85 & 4\\
-	\OnePhasethree\ & 2,504 & 6,468,094 & 131.17 & 5\\
-	\ThreePhasethree\ & 2,504 & 19,381,970 & 393.06 & 5\\
-	\FullPhasethree\ & 2,504 & 79,000,000 & 1602 & 5\\\botrule
-\end{tabular}}{Datasets used in this study; based on portions of the Phase 1 and Phase 3 data from the 1000 genomes project.
+        \NinteenPhaseone{} & 1,092 & 816,115 & 7.13  & 4\\
+        \OnePhaseone{} & 1,092 & 3,007,196 & 26.27  & 4\\
+        \SevenPhaseone{} & 1,092 & 18,984,880 & 165.85 & 4\\
+	\OnePhasethree{} & 2,504 & 6,468,094 & 131.17 & 5\\
+	\ThreePhasethree{} & 2,504 & 19,381,970 & 393.06 & 5\\
+	\FullPhasethree{} & 2,504 & 79,000,000 & 1602 & 5\\\botrule
+\end{tabular}}{Datasets used in this study; based on portions of the phase1 and phase3 data from the 1000 genomes project.
 In-memory size is the approximate memory required to store the data as a double-precision matrix of size n,m using the formula: n*m*8bytes/10\textsuperscript{9}bytes.
@@
 & & map & reduce  & map & reduce  & pre-processing & clustering & \\\midrule
-        R &\NinteenPhaseone & - & - & - & - & 55m 41s & 7s &0.614\\
-        Mahout & \NinteenPhaseone & 1536 & 3072 & 1536 & 3072 & 6m 10s & 8m 32s & 0.705\\
-        & \OnePhaseone & 1536 & 3072 & 1536 & 3072 & 9m 52s & 12m 08s & 0.771\\
-        & \SevenPhaseone & 1536 & 3072 & 4096 & 4096  & 60m 35s & 5h 44m & 0.716\\
-        & \FullPhaseone & 1536 & 3072 & 6144 & 6144  & 108m 47s &  & \\
-        & \OnePhasethree & 1536 & 3072 & 1536 & 3072 & 23m 27s & 21m 14s & 0.873\\
-        & \ThreePhasethree & 1536 & 3072 & 2048 & 4096 & 63m 49s & 93m 18s & 0.871\\\
-        & \FullPhasethree & 1536 & 3072 & 6144 & 6144 & 5h 36m & 21h 5m &0.623\\\botrule
+        R &\NinteenPhaseone{} & - & - & - & - & 55m 41s & 7s &0.614\\
+        Mahout & \NinteenPhaseone{} & 1536 & 3072 & 1536 & 3072 & 6m 10s & 8m 32s & 0.705\\
+        & \OnePhaseone{} & 1536 & 3072 & 1536 & 3072 & 9m 52s & 12m 08s & 0.771\\
+        & \SevenPhaseone{} & 1536 & 3072 & 4096 & 4096  & 60m 35s & 5h 44m & 0.716\\
+        & \FullPhaseone{} & 1536 & 3072 & 6144 & 6144  & 108m 47s &  & \\
+        & \OnePhasethree{} & 1536 & 3072 & 1536 & 3072 & 23m 27s & 21m 14s & 0.873\\
+        & \ThreePhasethree{} & 1536 & 3072 & 2048 & 4096 & 63m 49s & 93m 18s & 0.871\\\
+        & \FullPhasethree{} & 1536 & 3072 & 6144 & 6144 & 5h 36m & 21h 5m &0.623\\\botrule
 \end{tabular}}{The resources allocated to each MapReduce yarn-container and time taken for the job(s) to complete.}

### 2015-02-11 doc/publications/journal2014/paper.tex
Updates

@@
         & \OnePhaseone{} & 1536 & 3072 & 1536 & 3072 & 9m 52s & 12m 08s & 0.771\\
-        & \SevenPhaseone{} & 1536 & 3072 & 4096 & 4096  & 60m 35s & 5h 44m & 0.716\\
-        & \FullPhaseone{} & 1536 & 3072 & 6144 & 6144  & 108m 47s &  & \\
+        & \SevenPhaseone{} & 1536 & 3072 & 4096 & 4096  & 43m 56s & 5h 44m & 0.716\\
+        & \FullPhaseone{} & 1536 & 3072 & 6144 & 6144  & 85m 29s & 7h 43m & 0.723\\
         & \OnePhasethree{} & 1536 & 3072 & 1536 & 3072 & 23m 27s & 21m 14s & 0.873\\

### 2015-02-12 doc/publications/journal2014/paper.tex
missing comments

--- 2015-02-09 doc/publications/journal2014/paper.tex
@@
 \newcommand{\SevenPhaseone}{Phase1\_Chr1-7}
-\newcommand{\FullPhaseone}{Phase1\_Chr1-22}
 \newcommand{\OnePhasethree}{Phase3\_Chr1}
@@
 % DB: very nice but how do you get to the 1.6TB ? 
-% AO: Fixed
 As an example, the entire set of 1000 Genome Project Phase 3 variants would require approximately 1.6TB to be completely loaded into memory as a double-precision matrix (see Table~\ref{datasetsAcc}).
-This is based on the equation: n*m*8bytes/1GB (where 1GB is 10\^9 bytes).
-n*m are the matrix dimensions and 8bytes is the memory a double occupies.
 
 % DB: not clear you need to spell out your argument a bit more...
-% AO: Fixed
-By dividing jobs up into separate 'containers', Hadoop provides great versatility in scaling options.
-Although in this case, Hadoop has access to 36GB of memory distributed over 3 machines, the size of each container is only 2GB.
-Each containers is effectively a self-contained job that is independent from any other container.
-Therefore the more resources that the user has available, the more containers that can be run in parallel, decreasing the run-time of the overall job.
-This means that our Hadoop job can scale from a single machine with minimal resources (albeit with a longer run-time), to a huge compute cluster with terabytes of RAM.
-This benefit of the MapReduce paradigm allows much greater scalability over linear jobs designed in R, for example.
+Also, although Hadoop, in this case, has access to a total of 36GB of memory distributed over 3 machines, the container size is only 2GB. Therefore, we could
+easily run the same job on a single 12GB node, at the expense of time.
+
+This example demonstrates the potential for scalability in Hadoop, compared to single machine solutions, in respect to it's potential applications to genomics. 
+% DB: is this sentence adding information?
+We will proceed to demonstrate that Mahout can expand with relative ease to encompass larger data, although this depends on the task and how it can be parallelized.
 
 % SCALING UP FEATURES IN MAHOUT
-\subsection*{Hadoop scales to genome-wide clustering}
+\subsection*{Hadoop scales to 19 Million genomic variants}
 %DB: phase1 full genome ?
-%AO: Fixed
-In this section we demonstrate Mahout's scaling ability by increasing the numbers of features (variants per sample) input to our algorithm to include the full 79 Million variants from the Phase 3 dataset ( \FullPhasethree\ ).
+In this section we demonstrate Mahout's scaling ability by increasing the numbers of features (variants per sample) iteratively to 3 Million variants (\OnePhaseone) and 18 Million ( \SevenPhaseone\, respectively.
 
 %DB: Where is the result from \SevenPhaseone\
-%AO: Fixed
-We tested on various arbitrarily sized datasets to demonstrate the different container sizes required by the pre-processing and clustering, as well as the time required.
-On the OnePhaseone dataset, pre-processing takes just under 10 minutes. Clustering took approximately 12 minutes and the container size was the same as with clustering the OnePhasenineteen dataset.
-We then increased the input to include the first seven chromosomes. Pre-processing increased to approximately 1 hour and clustering took about 5 hours and 45 minutes.
-Our pre-processing algorithm was successful with the same container size as the previous examples, however, we increased both map and reduce container size for the clustering step to 4GB to accommodate the increased vector size.
-Finally we used the entire phase one dataset as input. Although pre-processing now takes TIME, it was again successful with the same sized containers as previously.
+On this dataset Hadoop's pre-processing takes 7 minutes, using the same container size as the smaller dataset. Clustering takes 23 minutes.
+The clustering accuracy increased by 35\% to 0.891, compared to the performance on \NinteenPhaseone\ (see Table~\ref{datasetsAcc}).
+R's in-memory implementation fails to store this dataset as a double-precision matrix would consume approximately 26GB memory, we hence only
+report the results for Hadoop.
 
-The significantly increased time required for this larger job is due to the larger container size.
-With a container size of 6GB, only up to six containers can run simultaneously, as opposed to 24 1.5GB containers.
+% ??
+To perform k-means clustering on the entire dataset in a reasonable amount of time would require a larger cluster with more memory per container.
+This is because each Map task loads the entire feature vector for the clusters into memory.
+Although we could increase the container size to the full 12GB per node, this would mean only 3 containers (1 per node) could run simultaneously, rather than the maximum of 18, and would greatly increase the processing time.
+Pre-processing the data, however, does scale to the entire feature-set, without an increase in container memory. Having the data in the distributed
+SequenceFile format opens the door to other distributed analysis techniques.
 
-While R does not scale to the larger datasets datasets, we have demonstrated Hadoop to cope well with pre-processing and clustering variants from 22 phase1 VCF files.
+While R does not scale to either of the two orders of magnitude incased datasets, We have demonstrated Hadoop to cope well with up to 70 Million variants.
+Although k-means clustering with a feature-set of this size is not feasible on a compute cluster of this size, clustering a smaller subset of the genome is a trivial task.
 
 % SCALING UP SAMPLES IN MAHOUT
-\subsection*{Hadoop scales effortless to more than twice the number of samples}
-In this section we investigate Mahout's ability to scale with more samples, i.e. the 2504 individuals from the Phase 3 dataset compared to the 1092 individuals in the Phase 1 dataset.
+\subsection*{Hadoop scales effortless to an order of magnitude more samples}
+In this section we investigate Mahout's ability to scale with more samples, i.e. the 1092 individuals in the Phase 1 dataset and the 2535 individuals in the Phase 3 dataset.
 %DB: ok this makes sense from a technical point of view (how fast, memory) but you should not really discuss the accuracy ...
@@
 \section*{Conclusion}
-In this paper we have applied Mahout's k-means clustering to the task of clustering individuals, based on their genomic variants. 
+In this paper we have applied Mahout's k-means clustering to the task of genotype based population. 
 We provide an interface from VCF format to Hadoop and Mahout, and a toolset to interpret Mahout's clustering output (see Figure 1)
 We find that on small datasets the actual clustering stage is two orders of magnitudes slower in Mahout compared to R (9 minutes vs 7 seconds), likely due to the overhead of setting up and co-ordinating the communication between the Hadoop nodes.
-However, increasing either the number of samples or the number of features per sample quickly exceeds the capability of R.
-Mahout on the other hand scales effortlessly to more than twice the number samples (2,504) and can process VCF files containing over 79 Million variants per samples.
+However, increasing either the number of samples or the number of features per sample quickly exceeds the ability of R.
+Mahout on the other hand scales effortlessly to more than twice the number samples (2,535) and can process data with over 79 Million features per samples.
 As the memory required for k-means clustering directly scales with the number of features per sample, we observe that Mahout's implementation
 of k-means does not scale to the entire exome dataset, unless memory per container were increased beyond the limits of our cluster.
+However, this might be resolved once Mahout moves from MapReduce to a domain-specific language for linear algebraic operations on top of Apache Spark \url{http://spark.apache.org/}.
 
@@
 \subsection*{Datasets}
-We used variant call format (VCF) files from phases 1 and 3 of the 1000 Genomes Project.
-Each VCF file details the genetic variation (SNPs and indels) for one chromosome, between each individual and a reference genome derived from GRCh37. 
-The phase 1 and phase 3 datasets contain 1,092 and 2,504 individuals, respectively.
+We used variant call format (VCF) files from phase 1 and 3 of the 1000 Genomes Project.
+The VCF files contain information about the genetic variants between each of the 1092 and 2,535 individuals respectively and a reference genome derived from GRCh37. 
+Each of the VCF files contain the variants in one chromosome, and we only use the 22 autosomes.
 Metadata available from the 1000 Genomes Consortium specifies additional attributes for each individual, including population and familial data.\\
-We used different subsets of the data for different comparisons. For example, to be able to compare Hadoop with R, we used the VCF file containing chromosome 19 variants (\NinteenPhaseone ).
-Then to demonstrate scaling with Hadoop, we use subsets that contain more variants, i.e. \SevenPhaseone\ and \FullPhasethree, which include chromosome 1-7 and the full genome, respectively.
+We used different subsets of the data for different comparisons.
+For example, to be able to compare Hadoop with R, we used a subset of the entire human variant dataset containing only chromosome 19 variants (\NinteenPhaseone ).
+Scaling up using Hadoop, we investigate how the accuracy improves when using more variants from the genome with \SevenPhaseone\ and \FullPhasethree, which include chromosome 1-7 and the full genome, respectively.
 Table~\ref{datasets} contains an overview of the datasets.
@@
 These three VMs run the Hadoop2 NodeManager.
-The NodeManager manages the Yet Another Resource Negotiator (YARN) 'containers'.
-The user specifies the memory size of these containers, where map containers can be a different size to reduce containers.
+The NodeManager manages the Yet Another Resource Negotiator (YARN) 'containers', where YARN allows the user
+to specify the resources allocated to each container.
 The fourth VM runs the HDFS NameNode, ResourceManager and monitoring services, and the fifth VM, a small, single core machine, runs the Cloudera manager service.
-With a total of 42GB memory and 24 CPU cores available between the node managers, we can run 24 YARN containers with 1.5GB memory per container, or 12 YARN containers with 3GB memory per container, in parallel.
+With this hardware setup, we can run 24 YARN containers with 1.5GB memory per container, or 12 YARN containers with 3GB memory per container.
 Storage consists of a 3TB Hadoop Distributed File System (HDFS) partition distributed across the three worker nodes. Each of these nodes runs a DataNode with a 1TB disk.
@@
 We collected the job time from the 'Elapsed' time as reported by the Hadoop JobHistory server. This is measured from the job's start time to finish time.
-Because k-means clustering performs multiple iterations, and therefore multiple MapReduce jobs, the time we reported is the sum of the elapsed time of each iteration.
+Because k-means clustering performs multiple iterations, and therefore multiple MapReduce jobs, the time we reported is the sum of the time each iteration consumed.
 For the R job, where we could not easily separate the pre-processing and clustering components, we printed the time between
@@
 {\begin{tabular}{lrrrrr}\toprule
-& individuals & variants & in-memory size (GB) & population\\
+& individuals & variants & size (GB) & population\\
 & & & &groups& \\\midrule
@@
         \SevenPhaseone & 1,092 & 18,984,880 & 165.85 & 4\\
-	\OnePhasethree\ & 2,504 & 6,468,094 & 131.17 & 5\\
-	\ThreePhasethree\ & 2,504 & 19,381,970 & 393.06 & 5\\
-	\FullPhasethree\ & 2,504 & 79,000,000 & 1602 & 5\\\botrule
+	\OnePhasethree\ & 2,535 & 6,468,094 & 131.17 & 5\\
+	\ThreePhasethree\ & 2,535 & 19,381,970 & 393.06 & 5\\
+	\FullPhasethree\ & 2,535 & 79,000,000 & 1602 & 5\\\botrule
 \end{tabular}}{Datasets used in this study; based on portions of the Phase 1 and Phase 3 data from the 1000 genomes project.
-In-memory size is the approximate memory required to store the data as a double-precision matrix of size n,m using the formula: n*m*8bytes/10\textsuperscript{9}bytes.
+Size is the approximate space required for storage of a double-precision matrix of size n,m using the formula: n*m*8bytes/10\textsuperscript{9}bytes.
 }
@@
 
+%DB: let's join 
+%\begin{table}[!t]
+%\processtable{Adjusted Rand index (accuracy).\label{Tab:01}}
+%{\begin{tabular}{lcc}\toprule
+%Tool, Dataset & accuracy \\\midrule
+%        R, \NinteenPhaseone & 0.614  \\ 
+%        Mahout, \NinteenPhaseone & 0.730\\
+%        Mahout, \OnePhaseone & 0.891\\
+%        Mahout, \SevenPhaseone & 0.816 \\
+%        Mahout, \OnePhasethree & 0.877  \\\botrule
+%\end{tabular}}{The adjusted Rand index for each dataset.
+The 'truth' index is the hypothetical cluster-set where there is a cluster for each super-population, with individuals belonging to the same super-population grouped together.}
+%\end{table}
+
+%DB: what is the dash doing in the table ?
+%DB: additional column for preprocessing and kmeans just like for time and have the map/reduce in the same and same number of after comma digits. 
 \begin{table*}[!t]
 \processtable{Resources allocated and time taken.\label{Tab:01}}
-{\begin{tabular}{ll|cc|cc|rr|c}\toprule
-Tool & Dataset & \multicolumn{2}{ c |}{Preprocesing (GB)} & \multicolumn{2}{ c |}{Clustering (GB)} & \multicolumn{2}{ c |}{Time} & Accuracy\\
-& & map & reduce  & map & reduce  & pre-processing & clustering & \\\midrule
-        R &\NinteenPhaseone & - & - & - & - & 55m 41s & 7s &0.614\\
-        Mahout & \NinteenPhaseone & 1536 & 3072 & 1536 & 3072 & 6m 10s & 8m 32s & 0.705\\
-        & \OnePhaseone & 1536 & 3072 & 1536 & 3072 & 9m 52s & 12m 08s & 0.771\\
-        & \SevenPhaseone & 1536 & 3072 & 4096 & 4096  & 60m 35s & 5h 44m & 0.716\\
-        & \FullPhaseone & 1536 & 3072 & 6144 & 6144  & 108m 47s &  & \\
-        & \OnePhasethree & 1536 & 3072 & 1536 & 3072 & 23m 27s & 21m 14s & 0.873\\
-        & \ThreePhasethree & 1536 & 3072 & 2048 & 4096 & 63m 49s & 93m 18s & 0.871\\\
-        & \FullPhasethree & 1536 & 3072 & 6144 & 6144 & 5h 36m & 21h 5m &0.623\\\botrule
-\end{tabular}}{The resources allocated to each MapReduce yarn-container and time taken for the job(s) to complete.}
+{\begin{tabular}{ll|ccc|rr|c}\toprule
+Tool& Dataset&&\multicolumn{2}{ c |}{Memory (GB)}&\multicolumn{2}{ c |}{Time} & Accuracy\\
+&& - & map & reduce  & pre-processing & clustering & \\\midrule
+         R &\NinteenPhaseone & - & 1.5 & 2 & 55m 41s & 7s &0.614\\
+        Mahout & \NinteenPhaseone & - & 1.5 & 2 & 2m 59s & 9m 12s & 0.730\\
+         & \OnePhaseone & - & 1.5 & 2 & 7m 24s & 23m 9s & 0.891\\
+        & \SevenPhaseone & - & 1.5 & 2 & 47m 43s & - & 0.816\\  %DB: why are there two lines for \SevenPhaseone
+        & \SevenPhaseone & - & 4.0 & 4 & - & 10h 43m 50s \\ 
+        & \OnePhasethree & - & 1.5 & 2 & 27m 48s & 89m 42s & 0.877\\
+        & \ThreePhasethree & - & 1.5 & 2 & 82m 14s & - \\
+        & \ThreePhasethree & - & 4.0 & 4 & - & (in progress) \\
+        & \FullPhasethree & - & 1.5 & 2 & 336m 8s & - \\\botrule
+\end{tabular}}{The resources allocated to each MapReduce yarn-container and time taken for the job(s) to complete.
+Where clustering required bigger containers than pre-processing, an additional line is included in the table for clustering, following pre-processing.}
 \end{table*}

### 2015-02-17 doc/publications/journal2014/paper.tex

--- 2015-02-11 doc/publications/journal2014/paper.tex
@@
 \newcommand{\SevenPhaseone}{phase1\_chr1-7}
-\newcommand{\FullPhaseone}{phase3\_chr1-22}
+\newcommand{\FullPhaseone}{phase3\_chr1-22} % DB: phase 1 ?
 \newcommand{\OnePhasethree}{phase3\_chr1}
@@
 We measured the runtime separately for pre-processing stage as well as the clustering stage. 
+%DB: can you explain 'container' or refer to methods where you have more room for discussing
 With Hadoop, pre-processing the data takes approximately 6 minutes, with an allocation of up to 3GB memory per 'container'.
@@
 
-Although the overall Mahout pipeline is about 40 minutes faster than R's, Mahout's k-means clustering takes two orders of magnitude longer. 
-This is likely due to the additional overhead involved in spawning the map and reduce processes, and distributing the data between nodes.
+Although the overall Mahout pipeline is about 40 minutes faster than R's, the actual k-means clustering takes two orders of magnitude longer in Mahout. 
+This is due to the additional overhead involved in spawning the map and reduce processes, and distributing the data between nodes.
 R, on the other hand, already has the entire matrix in-memory, on a single machine. 
-This should speed up the clustering significantly as no disk-access or networking is required.
-However, the amount of memory available on a single machine will quickly become a limiting factor.
-% DB: very nice but how do you get to the 1.6TB ?
-% AO: Fixed
+Since no disk-access or networking is required the clustering can be performed faster. However, the amount of memory available on a single machine will quickly become a limiting factor. 
 As an example, the entire set of 1000 Genome Project phase3 variants could require approximately 1.6TB to be completely loaded into memory as a double-precision matrix (see Table~\ref{datasetsAcc}).
 This is based on the equation:
-$^{n \times m \times 8bytes}/_{1GB}$, where $n \times m$ are the matrix dimensions, 8bytes is the memory a double occupies (i.e. each matrix element), and  $1GB$ is $10^9$bytes.
+$\frac{n \times m \times 8\text{bytes}}{1\text{GB}}$, where $n \times m$ are the matrix dimensions, 8bytes is the memory a double occupies (i.e. each matrix element), and  1GB is $10^9$bytes.
 
@@
 % AO: Fixed maybe
-The primary benefit of Hadoop is its scalability. This is achieved through MapReduce jobs which allows the task at hand to be separated into 'containers'.
-Although in the previous example, Hadoop had access to 36GB of memory distributed over 3 machines, the size of each container was only 3GB.
-Each of these containers is effectively a self-contained job that is independent from any other container.
-Therefore, the more resources that the user has available, the more containers that can be run in parallel, either on the same machine or separate, and in turn decreasing the run-time of the overall job.
-This means that our Hadoop job can scale from a single machine with minimal resources (albeit with a longer run-time), to a huge compute cluster with terabytes of RAM.
-This benefit allows Hadoop jobs to not only scale vertically, i.e. to a larger single machine as R can, but also horizontally, to a cluster of many machines.
+% DB: good explanation but I'm not sure what it does here (maybe reuse above or in methods)
+
+%The primary benefit of Hadoop is its scalability. This is achieved through MapReduce jobs, which allow the task at hand to be separated into 'containers'.
+%Although in the previous example, Hadoop had access to 36GB of memory distributed over 3 machines, the size of each container was only 3GB.
+%Each of these containers is effectively a self-contained job that is independent from any other container.
+%Therefore, the more resources that the user has available, the more containers that can be run in parallel, either on the same machine or separate, and in turn decreasing the run-time of the overall job.
+%This means that our Hadoop job can scale from a single machine with minimal resources (albeit with a longer run-time), to a huge compute cluster with terabytes of RAM.
+%This benefit allows Hadoop jobs to not only scale vertically, i.e. to a larger single machine as R can, but also horizontally, to a cluster of many machines.
 
@@
 \subsection*{Hadoop scales to genome-wide clustering}
-%DB: phase1 full genome ?
-%AO: Fixed nearly
-In this section we demonstrate Mahout's scaling ability by increasing the numbers of features (variants per sample) input to our algorithm to include the 79 Million variants from the entire phase3 dataset (\FullPhasethree{}).
+%DB phase 1 or phase 3 ?
+In this section we demonstrate Mahout's scaling ability by increasing the numbers of features (variants per sample) to include the 79 Million variants from the entire phase3 dataset (\FullPhasethree{}).
 
-%DB: Where is the result from \SevenPhaseone\
-%AO: Fixed
 We tested on various arbitrarily sized datasets to demonstrate the different container sizes required by pre-processing and clustering, as well as the time required.
@@
 However, we increased both the map and reduce container size for the clustering step to 4GB to accommodate the increased vector size.
+%DB phase 1 or 3 ?
 Finally we used the entire phase one dataset as input. Although pre-processing now takes 108 minutes, it was again successful with the same sized containers as previously.
 Clustering, however, not only required larger containers, but also took over 20 hours to complete.
-This significant increase in time required is in part due to the larger container size.
+%DB: you cannot use 'significant' without having performed some sort of statistical test to put a p-value to that statement.
+So use 'substantial' instead.
+Also this is a limitation from the cluster not a general thing so write it that way.
+This substantial time increase is in part due to the larger container size.
 With each container consuming 6GB of memory, only up to six containers can run simultaneously, as opposed to 24 1.5GB containers from the smaller examples.
 
-While R does not scale to the larger datasets datasets, we have demonstrated Hadoop to cope well with pre-processing and clustering variants from 22 phase1 VCF files.
+% DB: we need to discuss why the accuracy is actually dropping when you include more chromosomes. Its counterintuitive ... do you have a suggestion?
+
+%While R does not scale to the larger datasets datasets, we have demonstrated Hadoop to cope well with pre-processing and clustering variants from 22 phase1 VCF files.
+This demonstrates Hadoop's ability to cope with genome wide information, which merely inures a runtime increase rather than memory increase which quickly become the limiting factor as demonstrated by R's inability to process the data.
 
@@
 \subsection*{Hadoop scales effortless to more than twice the number of samples}
+% Can you rewrite this section to lead to the below concluding statement
 In this section we investigate Mahout's ability to scale with more samples, i.e. the 2504 individuals from the phase3 dataset compared to the 1092 individuals in the phase1 dataset.
-%DB: ok this makes sense from a technical point of view (how fast, memory) but you should not really discuss the accuracy ...
 
+% less than double ? Isn't it 85m 29s vs 5h 36m
 The container size required for pre-processing the data is identical for both datasets, with the phase3 dataset requiring less than double the time of the smaller phase1 dataset.
-In this case, due to the number of Feature Vectors increasing, rather than the feature size, the memory requirements for k-means clustering should remain similar, as the cluster centres loaded into the containers' memory are approximately the same size.
+%In this case, due to the number of Feature Vectors increasing, rather than the feature size, we expect the memory requirements for k-means clustering should remain similar, as the cluster centres loaded into the containers' memory are approximately the same size.
+We expect the memory requirement for the k-means clustering to stay the same as the number of cluster centres loaded into the containers' memory remain constant despite the increase of feature vectors oppose to feature size.
 This proved to be the case, as 4GB allocated per container was the minimum for both datasets. The time to complete clustering was much higher though,
 especially due to the fact that only 9 4GB containers can run simultaneously, rather than the 24 1.5GB containers. 
+%DB double-check numbers
 The clustering accuracy is 0.816 for the phase1 dataset and 0.xxx for the phase3 dataset.
 
-%TODO: can you think of a less general statement as conclusion
-Scaling effortless to massively more samples is the strength of Hadoop, hence compute problems need to be carefully designed to harness this ability.
+%DB: do you agree with this concluding statement?
+%Scaling effortless to massively more samples is the strength of Hadoop, hence compute problems need to be carefully designed to harness this ability.  
+Hadoop's strength is billed to be its ability to scale to massively more samples, however on a finite-resource cluster we observe a limitation to this scalability, which results in a dramatic increase in runtime.
+
+\subsection*{Result visualisation }
+% DB: in the conclusion we discussed that we provide a "toolset to interpret Mahout's clustering output".
+We hence need to show the R-image for that, i.e.
+spit out distance matrix and read in with R to plot, e.g.
+doing Multidimensional Scaling to squash the matrix in 2D http://www.statmethods.net/advstats/mds.html
 
@@
 However, increasing either the number of samples or the number of features per sample quickly exceeds the capability of R on this machine due to memory limitations.
+%DB ok effortlessly might be overstated ?
 Hadoop on the other hand scales effortlessly to more than twice the number samples (2,504) and can process VCF files that contain over 79 Million variants per samples.
@@
 We used different subsets of the data for different comparisons. For example, to be able to compare Hadoop with R, we used the VCF file containing chromosome 19 variants (\NinteenPhaseone{}).
-Then to demonstrate scaling with Hadoop, we use subsets that contain more variants, i.e. \SevenPhaseone{} and \FullPhasethree{}, which include chromosome 1-7 and the full genome, respectively.
+To demonstrate scaling with Hadoop, we use subsets that contain more variants, i.e. \SevenPhaseone{} and \FullPhaseone{}, which include chromosome 1-7 and the full genome, respectively.
 Table~\ref{datasets} contains an overview of the datasets.
@@
 \subsection*{Hardware and Software}
+%DB what about Piotr's cluster I thought you used that one for the full genome data of phase 1 and phase 3 ?
 We processed the data using a cluster of five VMs on the Microsoft Azure cloud service.
@@
 {\begin{tabular}{lrrrrr}\toprule
-& individuals & variants & in-memory size (GB) & population\\
-& & & &groups& \\\midrule
+name& individuals & variants & in-memory  & population\\
+& & &size (GB) &groups& \\\midrule
         \NinteenPhaseone{} & 1,092 & 816,115 & 7.13  & 4\\
@@
 	\ThreePhasethree{} & 2,504 & 19,381,970 & 393.06 & 5\\
-	\FullPhasethree{} & 2,504 & 79,000,000 & 1602 & 5\\\botrule
+	\FullPhasethree{} & 2,504 & 79,000,000 & 1602.00 & 5\\\botrule
 \end{tabular}}{Datasets used in this study; based on portions of the phase1 and phase3 data from the 1000 genomes project.

### 2015-02-17 doc/publications/journal2014/paper.tex
something

--- 2015-02-11 doc/publications/journal2014/paper.tex
@@
         & \OnePhaseone{} & 1536 & 3072 & 1536 & 3072 & 9m 52s & 12m 08s & 0.771\\
-        & \SevenPhaseone{} & 1536 & 3072 & 4096 & 4096  & 43m 56s & 5h 44m & 0.716\\
+        & \SevenPhaseone{} & 1536 & 3072 & 4096 & 4096  & 43m 56s & 3h 28m & 0.813\\
         & \FullPhaseone{} & 1536 & 3072 & 6144 & 6144  & 85m 29s & 7h 43m & 0.723\\

### 2015-03-02 doc/publications/journal2014/paper.tex
Latest changes

@@
 \newcommand{\SevenPhaseone}{phase1\_chr1-7}
-\newcommand{\FullPhaseone}{phase3\_chr1-22} % DB: phase 1 ?
+\newcommand{\FullPhaseone}{phase1\_chr1-22}
 \newcommand{\OnePhasethree}{phase3\_chr1}
@@
 \subsection*{Clustering in Mahout requires less memory than in R}
-In this section, we compare the runtime and resource requirements between the k-means clustering implementation in R and Mahout for the small \NinteenPhaseone\ dataset (see Methods). 
+In this section, we compare the runtime and resource requirements between the k-means clustering implementation in R and Mahout for the small \NinteenPhaseone{} dataset (see Methods). 
 We measured the runtime separately for pre-processing stage as well as the clustering stage. 
@@
 Conversely,  pre-processing the data to R's in-memory matrix takes 56 minutes and required 17.7GB memory. 
-K-means clustering the data, however, is faster with R, using 7s, whereas Mahout takes just under 9 minutes. See Table~\ref{datasetsTime}. \\
-Due to the relatively small number of variants, cluster performance is relatively poor for both methods, with the best Adjusted Rand Index score for clustering individuals in one of 4 populations (see Methods) being 0.705 in Mahout and 0.614 in R (see Table~\ref{datasetsAcc}).
+K-means clustering the data, however, is faster with R, using 7s, whereas Mahout takes just under 9 minutes. See Table~\ref{Tab:02}. \\
+Due to the relatively small number of variants, cluster performance is relatively poor for both methods, with the best Adjusted Rand Index score for clustering individuals in one of 4 populations (see Methods) being 0.705 in Mahout and 0.614 in R (see Table~\ref{Tab:02}).
 
@@
 Since no disk-access or networking is required the clustering can be performed faster. However, the amount of memory available on a single machine will quickly become a limiting factor. 
-As an example, the entire set of 1000 Genome Project phase3 variants could require approximately 1.6TB to be completely loaded into memory as a double-precision matrix (see Table~\ref{datasetsAcc}).
+As an example, the entire set of 1000 Genome Project phase3 variants could require approximately 1.6TB to be completely loaded into memory as a double-precision matrix (see Table~\ref{Tab:01}).
 This is based on the equation:
@@
 %DB phase 1 or phase 3 ?
-In this section we demonstrate Mahout's scaling ability by increasing the numbers of features (variants per sample) to include the 79 Million variants from the entire phase3 dataset (\FullPhasethree{}).
 
-We tested on various arbitrarily sized datasets to demonstrate the different container sizes required by pre-processing and clustering, as well as the time required.
-On the \OnePhaseone{} dataset, pre-processing takes just under 10 minutes and clustering takes approximately 12 minutes with a container size the same as with the \NinteenPhaseone{} dataset.
+In this section we demonstrate Mahout's scaling ability by increasing the number of features (variants per sample) to include the 38 million variants from the phase1 dataset (\FullPhaseone{}) and the 79 Million variants from the phase3 dataset (\FullPhasethree{}).
+
+We also ran jobs using intermediately sized phase1 datasets (see Table~\ref{Tab:02}) to demonstrate the different container sizes required by pre-processing and k-means clustering, and time required.
+On the \OnePhaseone{} dataset, pre-processing takes just under 10 minutes and clustering takes approximately 12 minutes with the same container size as with the \NinteenPhaseone{} dataset.
 We then increased the input to include the first seven chromosomes (\SevenPhaseone{}).
-In this case, pre-processing time increased to approximately 1 hour, and clustering to about 5 hours and 45 minutes.
+In this case, pre-processing time increased to approximately 45 minutes, and clustering to 3 hours and 28 minutes.
 Our pre-processing algorithm was successful with the same container size as the previous examples.
 However, we increased both the map and reduce container size for the clustering step to 4GB to accommodate the increased vector size.
-%DB phase 1 or 3 ?
-Finally we used the entire phase one dataset as input. Although pre-processing now takes 108 minutes, it was again successful with the same sized containers as previously.
-Clustering, however, not only required larger containers, but also took over 20 hours to complete.
+
+Finally we used the entire phase one dataset as input. Although pre-processing now takes 85 minutes, it was again successful with the same sized containers as previously.
+Clustering, however, not only required larger containers, but also took over 7 hours to complete.
 %DB: you cannot use 'significant' without having performed some sort of statistical test to put a p-value to that statement.
@@
 Also this is a limitation from the cluster not a general thing so write it that way.
+
 This substantial time increase is in part due to the larger container size.
-With each container consuming 6GB of memory, only up to six containers can run simultaneously, as opposed to 24 1.5GB containers from the smaller examples.
+With each container allocated 6GB of memory, our cluster can only run up to six containers simultaneously, as opposed to 24 1.5GB containers (from the smaller examples).
+This restriction in simultaneous containers is due to the limited memory available on our cluster.
 
+Although we compare phase3 datasets in the next section because they have a larger number of samples, the \FullPhasethree{} dataset also has the largest feature size, so we mention that in this section to demonstrate Hadoops ability to scale with feature-size.
+The most notable difference is the time required to complete clustering, at just over 21 hours.
+Although this is a substantial increase in time, the job still completes successfully.
+As we mention in a later section, this time can be reduced dramatically on a larger cluster.
 % DB: we need to discuss why the accuracy is actually dropping when you include more chromosomes. Its counterintuitive ... do you have a suggestion?
@@
 %While R does not scale to the larger datasets datasets, we have demonstrated Hadoop to cope well with pre-processing and clustering variants from 22 phase1 VCF files.
-This demonstrates Hadoop's ability to cope with genome wide information, which merely inures a runtime increase rather than memory increase which quickly become the limiting factor as demonstrated by R's inability to process the data.
+This section demonstrates Hadoop's ability to cope with genome wide information, which predominately inures a runtime increase.
+Although a memory increase is also present, the memory increases in proportion to vector length, rather than the size of the dataset, meaning we could still successfully run a job
+on the variants from an entire genome without encountering limitations due to the memory our cluster has available.
+%rather than memory increase which quickly become the limiting factor as demonstrated by R's inability to process the data.
 
 % SCALING UP SAMPLES IN MAHOUT
-\subsection*{Hadoop scales effortless to more than twice the number of samples}
+\subsection*{Hadoop scales effortlessly to more than twice the number of samples}
 % Can you rewrite this section to lead to the below concluding statement
-In this section we investigate Mahout's ability to scale with more samples, i.e. the 2504 individuals from the phase3 dataset compared to the 1092 individuals in the phase1 dataset.
+In this section we investigate Mahout's ability to scale with more samples, i.e.
+the 2504 individuals from the phase3 dataset compared to the 1092 individuals in the phase1 dataset.
+We focus on the \SevenPhaseone{} and \ThreePhasethree{} datasets due to their relatively similar feature size.
 
-% less than double ? Isn't it 85m 29s vs 5h 36m
-The container size required for pre-processing the data is identical for both datasets, with the phase3 dataset requiring less than double the time of the smaller phase1 dataset.
+The container size required for pre-processing the data is identical for both datasets, however the \ThreePhasethree{} dataset requiring about twenty minutes more time than the smaller \SevenPhaseone{} dataset.
+The clustering accuracy is 0.813 for the phase1 dataset and 0.825 for the phase3 dataset.
 %In this case, due to the number of Feature Vectors increasing, rather than the feature size, we expect the memory requirements for k-means clustering should remain similar, as the cluster centres loaded into the containers' memory are approximately the same size.
-We expect the memory requirement for the k-means clustering to stay the same as the number of cluster centres loaded into the containers' memory remain constant despite the increase of feature vectors oppose to feature size.
-This proved to be the case, as 4GB allocated per container was the minimum for both datasets. The time to complete clustering was much higher though,
-especially due to the fact that only 9 4GB containers can run simultaneously, rather than the 24 1.5GB containers. 
-%DB double-check numbers
-The clustering accuracy is 0.816 for the phase1 dataset and 0.xxx for the phase3 dataset.
+%We expect the memory requirement for the k-means clustering to stay the same as the number of cluster centres loaded into the containers' memory remain constant despite the increase of feature vectors oppose to feature size.
+We expect the memory requirement for the k-means clustering algorithm to be the similar for both datasets, because the larger size of \SevenPhaseone{} is due of more feature vectors (more individuals), rather than larger feature vectors (more variants).
+Mahouts iterative k-means algorithm loads a set number of these vectors into memory, based on the number of k-means clusters.
+And therefore, if the feature size doesn't increase, this memory requirement also won't increase.
+This proved to be the case, as 4GB allocated per container was the minimum for both datasets.
 
@@
 %Scaling effortless to massively more samples is the strength of Hadoop, hence compute problems need to be carefully designed to harness this ability.  
-Hadoop's strength is billed to be its ability to scale to massively more samples, however on a finite-resource cluster we observe a limitation to this scalability, which results in a dramatic increase in runtime.
+%Hadoop's strength is billed to be its ability to scale to massively more samples, however on a finite-resource cluster we observe a limitation to this scalability, which results in a dramatic increase in runtime.
+Hadoop's strength is billed to be its ability to scale to massively more samples.
+We observe this to be true, as increasing the samples by more than twice, which is the case from \SevenPhaseone{} to \ThreePhasethree{}, does not result in greater memory requirements, unlike increasing the number of variants per sample.
 
@@
 
+% AO: new section
+\subsection*{Cluster size and containers}
+An important consideration when designing a MapReduce job is the container size.
+A container is essentially an abstraction of resources, i.e.
+memory and CPU cores.
+Tasks are run in parallel in these containers, and each container is independent of other containers.
+The resources Hadoop allocates to each container is a predefined value, specified in its configuration.
+According to the Cloudera documentation, the general rule of thumb is that one CPU core is allocated to each container, and the available memory per node should be divided equally between each of those containers.
+For example, on a node with 8 CPU cores and 32GB memory, each container should be allocated 1 CPU core and 4GB Memory.
+Effectively allowing a maximum of eight 4GB containers to run in parallel on that node.
+
+When a MapReduce job is launched, Hadoop spawns a number of map tasks and reduce tasks. These tasks are placed in a queue until a node has resources available to run a container.
+The more containers that can run simultaneously, the faster the job will complete.
+Generally, the number of containers is limited by the available CPU cores, as each container requires an entire CPU core.
+However for jobs that deal
+with large datasets, the memory allocated per container may not be enough.
+We could modify the above example to allocate each container 1 CPU core and 8GB memory, however now only four containers can fit within the node's 32GB memory, resulting in the other 4 CPU cores being disused.
+It is therefore optimal to
+allocate the appropriate resources to containers, and only increase container memory when necessary, as overallocation of memory will simply limit the number of simultaneous containers each node can run.
+
+To demonstrate the importance of specifying these parameters correctly, we ran the same job using different configurations.
+The relatively small job, \OnePhaseone{}, took an additional 5 minutes with a map container size of 4GB, as opposed to 1.5GB.
+Yet this larger container size is still too small for the \FullPhasethree{} dataset, which requires at least 6GB allocated to each map container.
+
+To demonstrate the scaling ability of jobs to a larger cluster, we ran the same clustering job on a development cluster we had access to.
+As presented in Table \ref{Tab:03}, the \OnePhaseone{} job saw an improvement of approximately 70\%.
+The \FullPhasethree{} example, on the other hand, was 78\% faster.
+A limitation in both these examples is that the number of reduces in k-means clustering is limited to the number of k-means clusters.
+A task with five clusters, for example, will only be able to spawn five reduce containers, no matter how many resources are available.
+
 \section*{Conclusion}
@@
 We provide an interface from VCF format to Hadoop and Mahout, and a toolset to interpret Mahout's clustering output (see Figure 1)
-We find that on small datasets the clustering stage is two orders of magnitudes slower in Mahout compared to R (9 minutes vs 7 seconds), likely due to the overhead of setting up and co-ordinating the communication between the Hadoop nodes.
+We find that on small datasets the clustering stage is two orders of magnitudes slower in Mahout compared to R (9 minutes vs 7 seconds), due to the overhead of setting up and co-ordinating the communication between the Hadoop nodes.
 However, increasing either the number of samples or the number of features per sample quickly exceeds the capability of R on this machine due to memory limitations.
@@
 Hadoop on the other hand scales effortlessly to more than twice the number samples (2,504) and can process VCF files that contain over 79 Million variants per samples.
-
+The MapReduce functions can scale from a relatively small and accessible cluster with 16 cores, to a much larger cluster with close to 500 CPU cores. 
 In conclusion, Mahout offers a scalable out-of-the box solution that is directly applicable to genotype-based clustering for population or family association.
@@
 We set the value of each double to the sum of the square root of the Hamming distance between each variant allele and the reference genome.
-Therefore, a value identical to the reference, 0|0, will be 0 (and omitted from the sparse vector),
-a heterozygous variant, 0|1 or 1|0, will be 1 (sqrt1) and a homozygous variant, 1|1, will be 2 (sqrt1 + sqrt1).
-The square root of the each allele prevents variants such as 1|1 colliding with the variant 2|0.
+Therefore, a value identical to the reference, 0\textbar{}0, will be 0 (and omitted from the sparse vector),
+a heterozygous variant, 0\textbar{}1 or 1\textbar{}0, will be 1 (sqrt1) and a homozygous variant, 1\textbar{}1, will be 2 (sqrt1 + sqrt1).
+The square root of the each allele prevents variants such as 1\textbar{}1 colliding with the variant 2\textbar{}0.
 
@@
 The variants in each vector are in the same order as they appear in the VCF file.
-In this stage, we can define parameters to exclude certain variants that may be deemed low quality, or only appear in a certain number of individuals.
-We use identical keys so that each key value pair is sent to the same reducer, and this reducer assigns a unique key to each value from the previous step.
-We require one reducer in this step to make each variant key a consecutive number, rather than an arbitrary feature, such as its genomic location.
+
+Additionally, in this stage, we can define parameters to exclude certain variants that may be deemed low quality, or only appear in a certain number of individuals.
+We use identical keys so that each key value pair is sent to the same reducer, and this reducer assigns a unique key to each value.
+We require one reducer in this step to make each the key for each variant a consecutive number, rather than an arbitrary feature, such as its genomic location.
 Because keys are represented by integers, this helps to avoid integer overflows by keeping the keys relatively small.
 
-The second job effectively transposes the output from the first job. The mappers create key value pairs where the key is composite of the individual id and variant id, and the value is the variant.
+The second job effectively transposes the output from the first job. The mappers create key value pairs where the key is a composite of the individual id and variant id, and the value is the variant.
 This composite key, which is essentially a concatenation of two keys, allowed us to implement custom 'collect' and 'sort' classes to speed up the process.
 These classes divide up the key value pairs by their individual ID and sort them by their location ID, so each sorted collection of key/value pairs can be sent to a separate reducer.
- The reducers then write the variants for each individual to a sparse vector. These are saved to disk as the value for each individual ID key.
+ The reducers then write the variants for each individual to a sparse vector and stores each vector as a value, where the key is the individual ID.
 
@@
         \SevenPhaseone{} & 1,092 & 18,984,880 & 165.85 & 4\\
+        \FullPhaseone{} & 1,092 & 38,219,238 & 333.88 & 4\\
 	\OnePhasethree{} & 2,504 & 6,468,094 & 131.17 & 5\\
@@
 \begin{table*}[!t]
-\processtable{Resources allocated and time taken.\label{Tab:01}}
+\processtable{Resources allocated and time taken.\label{Tab:02}}
 {\begin{tabular}{ll|cc|cc|rr|c}\toprule
@@
         Mahout & \NinteenPhaseone{} & 1536 & 3072 & 1536 & 3072 & 6m 10s & 8m 32s & 0.705\\
-        & \OnePhaseone{} & 1536 & 3072 & 1536 & 3072 & 9m 52s & 12m 08s & 0.771\\
+        & \OnePhaseone{} & 1536 & 3072 & 1536 & 3072 & 9m 52s & 12m 49s & 0.798\\
         & \SevenPhaseone{} & 1536 & 3072 & 4096 & 4096  & 43m 56s & 3h 28m & 0.813\\
         & \FullPhaseone{} & 1536 & 3072 & 6144 & 6144  & 85m 29s & 7h 43m & 0.723\\
-        & \OnePhasethree{} & 1536 & 3072 & 1536 & 3072 & 23m 27s & 21m 14s & 0.873\\
-        & \ThreePhasethree{} & 1536 & 3072 & 2048 & 4096 & 63m 49s & 93m 18s & 0.871\\\
-        & \FullPhasethree{} & 1536 & 3072 & 6144 & 6144 & 5h 36m & 21h 5m &0.623\\\botrule
+        & \OnePhasethree{} & 1536 & 3072 & 1536 & 3072 & 23m 27s & 21m 14s & 0.622\\
+        & \ThreePhasethree{} & 1536 & 3072 & 4096 & 4096 & 63m 49s & 4h 21m & 0.825\\
+        & \FullPhasethree{} & 1536 & 3072 & 6144 & 6144 & 5h 36m & 21h 5m &0.840\\\botrule
+\end{tabular}}{The resources allocated to each MapReduce yarn-container and time taken for the job(s) to complete.}
+\end{table*}
+
+\begin{table*}[!t]
+\processtable{Resources allocated and time taken.\label{Tab:03}}
+{\begin{tabular}{l|cc|cc|r}\toprule
+Dataset & \multicolumn{2}{ c |}{Available Resources} & \multicolumn{2}{ c |}{Container Size (GB)} & Time\\
+& CPU cores & Memory (GB)  & map & reduce & \\\midrule
+        \OnePhaseone{} & 24 & 36 & 1536 & 3072 & 12m 49s\\
+        \OnePhaseone{} & 24 & 36 & 4096 & 4096  & 17m 47s\\
+        \OnePhaseone{} & 448 & 1310 & 1536 & 4096  & 5m 22s\\
+        \FullPhasethree{} & 24 & 36 & 6144 & 6144 & 21h 4m\\
+        \FullPhasethree{} & 448 & 1310 & 6144 & 6144 & 4h 42m\\\botrule
 \end{tabular}}{The resources allocated to each MapReduce yarn-container and time taken for the job(s) to complete.}

### 2015-03-02 doc/publications/journal2014/paper.tex
More changes

@@
 %While R does not scale to the larger datasets datasets, we have demonstrated Hadoop to cope well with pre-processing and clustering variants from 22 phase1 VCF files.
-This section demonstrates Hadoop's ability to cope with genome wide information, which predominately inures a runtime increase.
-Although a memory increase is also present, the memory increases in proportion to vector length, rather than the size of the dataset, meaning we could still successfully run a job
-on the variants from an entire genome without encountering limitations due to the memory our cluster has available.
+This section demonstrates Hadoop's ability to cope with genome wide information, which predominantly results in a runtime increase.
+The memory requirements are also increased, however the memory required increases in proportion to vector length,
+rather than the entire size of the dataset.
+This means we can successfully run a job on the variants from an entire genome on our cluster, or potentially an even smaller cluster, as long as 6GB is available for containers.
 %rather than memory increase which quickly become the limiting factor as demonstrated by R's inability to process the data.
@@
 the 2504 individuals from the phase3 dataset compared to the 1092 individuals in the phase1 dataset.
-We focus on the \SevenPhaseone{} and \ThreePhasethree{} datasets due to their relatively similar feature size.
+We focus on the \SevenPhaseone{} and \ThreePhasethree{} datasets due to their similar feature size.
 
-The container size required for pre-processing the data is identical for both datasets, however the \ThreePhasethree{} dataset requiring about twenty minutes more time than the smaller \SevenPhaseone{} dataset.
+The container size required for pre-processing the data is identical for both datasets, however the \ThreePhasethree{} dataset required about twenty minutes more time than the smaller \SevenPhaseone{} dataset.
 The clustering accuracy is 0.813 for the phase1 dataset and 0.825 for the phase3 dataset.
@@
 %We expect the memory requirement for the k-means clustering to stay the same as the number of cluster centres loaded into the containers' memory remain constant despite the increase of feature vectors oppose to feature size.
-We expect the memory requirement for the k-means clustering algorithm to be the similar for both datasets, because the larger size of \SevenPhaseone{} is due of more feature vectors (more individuals), rather than larger feature vectors (more variants).
-Mahouts iterative k-means algorithm loads a set number of these vectors into memory, based on the number of k-means clusters.
+We expect the memory requirement for the k-means clustering algorithm to be the similar for both datasets, because the larger size of \SevenPhaseone{} is a result of more feature vectors (more individuals), rather than larger feature vectors (more variants).
+Mahout's iterative k-means algorithm loads a set number of these vectors into memory, based on the number of k-means clusters.
 And therefore, if the feature size doesn't increase, this memory requirement also won't increase.
@@
 Hadoop's strength is billed to be its ability to scale to massively more samples.
-We observe this to be true, as increasing the samples by more than twice, which is the case from \SevenPhaseone{} to \ThreePhasethree{}, does not result in greater memory requirements, unlike increasing the number of variants per sample.
+We observe this to be true, as increasing the samples by two and a half, which is the case from \SevenPhaseone{} to \ThreePhasethree{}, does not result in greater memory requirements,
+unlike increasing the number of variants per sample.
 
-\subsection*{Result visualisation }
+\subsection*{Result visualisation}
 % DB: in the conclusion we discussed that we provide a "toolset to interpret Mahout's clustering output".
@@
 According to the Cloudera documentation, the general rule of thumb is that one CPU core is allocated to each container, and the available memory per node should be divided equally between each of those containers.
-For example, on a node with 8 CPU cores and 32GB memory, each container should be allocated 1 CPU core and 4GB Memory.
+For example, on a node with 8 CPU cores and 32GB memory, each container should be allocated 1 CPU core and 4GB memory.
 Effectively allowing a maximum of eight 4GB containers to run in parallel on that node.
 
-When a MapReduce job is launched, Hadoop spawns a number of map tasks and reduce tasks. These tasks are placed in a queue until a node has resources available to run a container.
-The more containers that can run simultaneously, the faster the job will complete.
-Generally, the number of containers is limited by the available CPU cores, as each container requires an entire CPU core.
-However for jobs that deal
-with large datasets, the memory allocated per container may not be enough.
+When a MapReduce job is launched, Hadoop spawns a number of map tasks and reduce tasks.
+These tasks are placed in a queue until a node has enough resources available to run a container, and generally, the more containers that can run simultaneously, the faster the job will complete.
+The number of containers for smaller jobs are generally limited by the available CPU cores, as each container requires an entire CPU core.
+However for jobs that deal with large datasets, the memory allocated per container (following Cloudera's guidelines) may not be enough.
 We could modify the above example to allocate each container 1 CPU core and 8GB memory, however now only four containers can fit within the node's 32GB memory, resulting in the other 4 CPU cores being disused.
@@
 
-To demonstrate the importance of specifying these parameters correctly, we ran the same job using different configurations.
-The relatively small job, \OnePhaseone{}, took an additional 5 minutes with a map container size of 4GB, as opposed to 1.5GB.
-Yet this larger container size is still too small for the \FullPhasethree{} dataset, which requires at least 6GB allocated to each map container.
+To demonstrate the importance of specifying these parameters correctly, we ran the identical jobs, using different Hadoop configurations.
+The \OnePhaseone{} dataset only requires 1.5GB memory per container, as we demonstrated previously.
+Increasing the map container size to 4GB results in the job taking an additional 5 minutes (see Table \ref{Tab:03}).
+However, this larger container size is still too small for the \FullPhasethree{} dataset, which requires at least 6GB allocated to each map container.
+Running a job using this dataset and 4GB memory per container results in the job failing.
 

### 2015-03-02 doc/publications/journal2014/paper.tex
More stuff

@@
 
-To demonstrate the scaling ability of jobs to a larger cluster, we ran the same clustering job on a development cluster we had access to.
+One of the main selling points of Hadoop, is that it will run on commodity hardware.
+For this reason we used a relatively small cluster for our benchmarks.
+Although our cluster uses VMs on Microsoft Azure, it would be relatively simple for a lab to
+create a similar Hadoop setup in-house.
+However, to demonstrate the scaling ability of Hadoop and our MapReduce algorithms, we also ran the clustering jobs on a much larger development Hadoop cluster.
 As presented in Table \ref{Tab:03}, the \OnePhaseone{} job saw an improvement of approximately 70\%.
 The \FullPhasethree{} example, on the other hand, was 78\% faster.
-A limitation in both these examples is that the number of reduces in k-means clustering is limited to the number of k-means clusters.
-A task with five clusters, for example, will only be able to spawn five reduce containers, no matter how many resources are available.
+Although both jobs were more than twice as fast on the larger cluster, the speed doesn't scale relative to the size of the cluster.
+This is due to a limitation in these examples, where the number of reduces in k-means clustering is limited to the number of k-means clusters (k).
+A task where k is five, for example, will only be able to spawn five reduce containers, no matter how many resources are available.
+Although sometimes inevitable, depending on the task, this bottleneck of maximum simultaneous reduce tasks is an important consideration to take into account when designing MapReduce jobs.
 
@@
 %DB ok effortlessly might be overstated ?
-Hadoop on the other hand scales effortlessly to more than twice the number samples (2,504) and can process VCF files that contain over 79 Million variants per samples.
+Hadoop on the other hand scales effortlessly to more than twice the number samples (2,504).
+It can also process VCF files that contain over 79 Million variants per samples, albeit with higher memory allocation.
 The MapReduce functions can scale from a relatively small and accessible cluster with 16 cores, to a much larger cluster with close to 500 CPU cores. 
@@
 Three of these VMs, the 'worker nodes', each have 14GB memory and and 8 CPU cores.
-These three VMs run the Hadoop2 NodeManager, which manages the Yet Another Resource Negotiator (YARN) 'containers'.
+These three VMs run the Hadoop2 NodeManager, which manages the Yet Another Resource Negotiator (YARN) 'containers' (see discussion).
 The user is able to specify the resources allocated to each container, where map containers can have a different allocation to reduce containers.
@@
 Storage consists of a 3TB Hadoop Distributed File System (HDFS) partition distributed across the three worker nodes. Each of these nodes runs a DataNode with a 1TB disk.
+
+The larger cluster mainly differs in regards to hardware. It has a total of 448 CPU cores and 1.31TB memory available.
 

### 2015-03-03 doc/publications/journal2014/paper.tex
comments

@@
 %DB: can you explain 'container' or refer to methods where you have more room for discussing
+%DB: is this addressed?
 With Hadoop, pre-processing the data takes approximately 6 minutes, with an allocation of up to 3GB memory per 'container'.
@@
 % DB: good explanation but I'm not sure what it does here (maybe reuse above or in methods)
-
+% DB: Delete this ?
 %The primary benefit of Hadoop is its scalability. This is achieved through MapReduce jobs, which allow the task at hand to be separated into 'containers'.
@@
 
+%DB: lets switch section 
+% 1) increase samples \SevenPhaseone{} and \ThreePhasethree{} -> Hadoop scales: does not increase memory requirement, and runtime would have stayed the same with more resource
+% 2) increase features \OnePhasethree, \ThreePhasethree, \FullPhasethree -> 
+% that way we get around explaining why FullPhaseOne is worse than \SevenPhaseone
+
 % SCALING UP FEATURES IN MAHOUT
 \subsection*{Hadoop scales to genome-wide clustering}
-%DB phase 1 or phase 3 ?
-
 In this section we demonstrate Mahout's scaling ability by increasing the number of features (variants per sample) to include the 38 million variants from the phase1 dataset (\FullPhaseone{}) and the 79 Million variants from the phase3 dataset (\FullPhasethree{}).
 
+% This section feels like a listing numbers rather than a "story".
+Maybe having a plot we can make a more general statement.
+So can you make a plot #features vs memory and in a separate on vs time with different colours showing pre-post processing.
+Or one plot memory vs time with colours as features and shape as pre-post ?
 We also ran jobs using intermediately sized phase1 datasets (see Table~\ref{Tab:02}) to demonstrate the different container sizes required by pre-processing and k-means clustering, and time required.
@@
 Clustering, however, not only required larger containers, but also took over 7 hours to complete.
-%DB: you cannot use 'significant' without having performed some sort of statistical test to put a p-value to that statement.
-So use 'substantial' instead.
-Also this is a limitation from the cluster not a general thing so write it that way.
 
@@
 
+% mentioning phase 3 here makes it difficult to understand.
 Although we compare phase3 datasets in the next section because they have a larger number of samples, the \FullPhasethree{} dataset also has the largest feature size, so we mention that in this section to demonstrate Hadoops ability to scale with feature-size.
@@
 As we mention in a later section, this time can be reduced dramatically on a larger cluster.
-% DB: we need to discuss why the accuracy is actually dropping when you include more chromosomes. Its counterintuitive ... do you have a suggestion?
 
-%While R does not scale to the larger datasets datasets, we have demonstrated Hadoop to cope well with pre-processing and clustering variants from 22 phase1 VCF files.
+%DB how else would you expect memory to grow 
 This section demonstrates Hadoop's ability to cope with genome wide information, which predominantly results in a runtime increase.
@@
 \subsection*{Hadoop scales effortlessly to more than twice the number of samples}
-% Can you rewrite this section to lead to the below concluding statement
 In this section we investigate Mahout's ability to scale with more samples, i.e.

### 2015-03-05 doc/publications/journal2014/paper.tex
start of new sections

@@
 
-%DB: lets switch section 
-% 1) increase samples \SevenPhaseone{} and \ThreePhasethree{} -> Hadoop scales: does not increase memory requirement, and runtime would have stayed the same with more resource
-% 2) increase features \OnePhasethree, \ThreePhasethree, \FullPhasethree -> 
-% that way we get around explaining why FullPhaseOne is worse than \SevenPhaseone
+% (1) SCALING UP FEATURES IN MAHOUT
+\subsection*{Hadoop scales to genome-wide clustering}
+In this section we demonstrate Mahout's scaling ability by increasing the number of features (variants per sample) to include the entire phase3 autosome dataset (\FullPhasethree{}).
+We focus on the \FullPhasethree{} dataset in this section because it has 79 million variants compared to \FullPhaseone{}'s 38 million.
 
-% SCALING UP FEATURES IN MAHOUT
-\subsection*{Hadoop scales to genome-wide clustering}
-In this section we demonstrate Mahout's scaling ability by increasing the number of features (variants per sample) to include the 38 million variants from the phase1 dataset (\FullPhaseone{}) and the 79 Million variants from the phase3 dataset (\FullPhasethree{}).
+We start by clustering one chromosome. Our algorithm completes successfully, however, it only achieves an adjusted Rand index of 0.622.
+Increasing the dataset to include the first three chromosomes (\ThreePhasethree{}) results in an adjusted Rand index on 0.825
+Finally we include all 22 autosomes as input to our algorithm (\FullPhasethree{}).
+Although the adjusted rand index isn't much higher than the previous example, at 0.84, the pre-processing and k-means complete successfully.
 
@@
 Or one plot memory vs time with colours as features and shape as pre-post ?
-We also ran jobs using intermediately sized phase1 datasets (see Table~\ref{Tab:02}) to demonstrate the different container sizes required by pre-processing and k-means clustering, and time required.
-On the \OnePhaseone{} dataset, pre-processing takes just under 10 minutes and clustering takes approximately 12 minutes with the same container size as with the \NinteenPhaseone{} dataset.
-We then increased the input to include the first seven chromosomes (\SevenPhaseone{}).
+
+%% OLD
+%%We also ran jobs using intermediately sized phase1 datasets (see Table~\ref{Tab:02}) to demonstrate the different container sizes required by pre-processing and k-means clustering, and time required.
+%On the \OnePhaseone{} dataset, pre-processing takes just under 10 minutes and clustering takes approximately 12 minutes with the same container size as with the \NinteenPhaseone{} dataset.
+%We then increased the input to include the first seven chromosomes (\SevenPhaseone{}).
 In this case, pre-processing time increased to approximately 45 minutes, and clustering to 3 hours and 28 minutes.
-Our pre-processing algorithm was successful with the same container size as the previous examples.
+%Our pre-processing algorithm was successful with the same container size as the previous examples.
 However, we increased both the map and reduce container size for the clustering step to 4GB to accommodate the increased vector size.
-
-Finally we used the entire phase one dataset as input. Although pre-processing now takes 85 minutes, it was again successful with the same sized containers as previously.
-Clustering, however, not only required larger containers, but also took over 7 hours to complete.
-
-This substantial time increase is in part due to the larger container size.
+%
+%Finally we used the entire phase one dataset as input. Although pre-processing now takes 85 minutes, it was again successful with the same sized containers as previously.
+%Clustering, however, not only required larger containers, but also took over 7 hours to complete.
+%
+%This substantial time increase is in part due to the larger container size.
 With each container allocated 6GB of memory, our cluster can only run up to six containers simultaneously, as opposed to 24 1.5GB containers (from the smaller examples).
-This restriction in simultaneous containers is due to the limited memory available on our cluster.
-
-% mentioning phase 3 here makes it difficult to understand.
-Although we compare phase3 datasets in the next section because they have a larger number of samples, the \FullPhasethree{} dataset also has the largest feature size, so we mention that in this section to demonstrate Hadoops ability to scale with feature-size.
-The most notable difference is the time required to complete clustering, at just over 21 hours.
+%This restriction in simultaneous containers is due to the limited memory available on our cluster.
+%
+%% mentioning phase 3 here makes it difficult to understand.
+%Although we compare phase3 datasets in the next section because they have a larger number of samples, the \FullPhasethree{} dataset also has the largest feature size, so we mention that in this section to demonstrate Hadoops ability to scale with feature-size.
+%The most notable difference is the time required to complete clustering, at just over 21 hours.
 Although this is a substantial increase in time, the job still completes successfully.
 As we mention in a later section, this time can be reduced dramatically on a larger cluster.
-
-%DB how else would you expect memory to grow 
-This section demonstrates Hadoop's ability to cope with genome wide information, which predominantly results in a runtime increase.
+%
+%%DB how else would you expect memory to grow 
+%This section demonstrates Hadoop's ability to cope with genome wide information, which predominantly results in a runtime increase.
 The memory requirements are also increased, however the memory required increases in proportion to vector length,
-rather than the entire size of the dataset.
+%rather than the entire size of the dataset.
 This means we can successfully run a job on the variants from an entire genome on our cluster, or potentially an even smaller cluster, as long as 6GB is available for containers.
 %rather than memory increase which quickly become the limiting factor as demonstrated by R's inability to process the data. 
+% /OLD
 
-% SCALING UP SAMPLES IN MAHOUT
+% (2) SCALING UP SAMPLES IN MAHOUT
 \subsection*{Hadoop scales effortlessly to more than twice the number of samples}
-In this section we investigate Mahout's ability to scale with more samples, i.e.
-the 2504 individuals from the phase3 dataset compared to the 1092 individuals in the phase1 dataset.
-We focus on the \SevenPhaseone{} and \ThreePhasethree{} datasets due to their similar feature size.
 
-The container size required for pre-processing the data is identical for both datasets, however the \ThreePhasethree{} dataset required about twenty minutes more time than the smaller \SevenPhaseone{} dataset.
-The clustering accuracy is 0.813 for the phase1 dataset and 0.825 for the phase3 dataset.
-%In this case, due to the number of Feature Vectors increasing, rather than the feature size, we expect the memory requirements for k-means clustering should remain similar, as the cluster centres loaded into the containers' memory are approximately the same size.
-%We expect the memory requirement for the k-means clustering to stay the same as the number of cluster centres loaded into the containers' memory remain constant despite the increase of feature vectors oppose to feature size.
-We expect the memory requirement for the k-means clustering algorithm to be the similar for both datasets, because the larger size of \SevenPhaseone{} is a result of more feature vectors (more individuals), rather than larger feature vectors (more variants).
-Mahout's iterative k-means algorithm loads a set number of these vectors into memory, based on the number of k-means clusters.
+In this section we investigate Mahout's ability to scale with more samples.
+Mahout's k-means algorithm performs by splitting the clustering job up into separate tasks, where each task calculates the distance between a subset of samples and the k-means centroids.
+Mahout performs each of these calculations iteratively until they are complete for every sample.
+In the previous section we increased the sample size. Larger samples have and increased dimensionality, and hence require more memory and also more time to calculate the distances.
+In this case however, we increase the number of samples.
+By simply increasing the number of samples, the k-means algorithm will perform more iterations, however the resource cost of each iteration should remain the same.
+To demonstrate this, we compare pre-processing and clustering of the 2504 individuals from the phase3 dataset compared to the 1092 individuals in the phase1 dataset.
+We specifically use the \SevenPhaseone{} and \ThreePhasethree{} datasets because of their similar feature size.
+Even with twice the number of samples in the larger dataset, the algorithm completed successfully for both datasets with the same container size.
+However the \ThreePhasethree{} dataset did require about twenty minutes more time than the smaller \SevenPhaseone{} dataset.
+This demonstrates Hadoop's strength of scaling with sample size.
+Rather than the larger comparison requiring more resources, it simple required more time.
+The container size however, cannot be an arbitrary value, as we discuss in the next section.
+
+%% OLD
+%The container size required for pre-processing the data is identical for both datasets, however the \ThreePhasethree{} dataset required about twenty minutes more time than the smaller \SevenPhaseone{} dataset.
+%The clustering accuracy is 0.813 for the phase1 dataset and 0.825 for the phase3 dataset.
+%%In this case, due to the number of Feature Vectors increasing, rather than the feature size, we expect the memory requirements for k-means clustering should remain similar, as the cluster centres loaded into the containers' memory are approximately the same size.
+%%We expect the memory requirement for the k-means clustering to stay the same as the number of cluster centres loaded into the containers' memory remain constant despite the increase of feature vectors oppose to feature size.
+%We expect the memory requirement for the k-means clustering algorithm to be the similar for both datasets, because the larger size of \SevenPhaseone{} is a result of more feature vectors (more individuals), rather than larger feature vectors (more variants).
+%Mahout's iterative k-means algorithm loads a set number of these vectors into memory, based on the number of k-means clusters.
 And therefore, if the feature size doesn't increase, this memory requirement also won't increase.
-This proved to be the case, as 4GB allocated per container was the minimum for both datasets.
+%This proved to be the case, as 4GB allocated per container was the minimum for both datasets.
 
-%DB: do you agree with this concluding statement?
-%Scaling effortless to massively more samples is the strength of Hadoop, hence compute problems need to be carefully designed to harness this ability.  
-%Hadoop's strength is billed to be its ability to scale to massively more samples, however on a finite-resource cluster we observe a limitation to this scalability, which results in a dramatic increase in runtime.
-Hadoop's strength is billed to be its ability to scale to massively more samples.
+%%DB: do you agree with this concluding statement?
+%%Scaling effortless to massively more samples is the strength of Hadoop, hence compute problems need to be carefully designed to harness this ability.  
+%%Hadoop's strength is billed to be its ability to scale to massively more samples, however on a finite-resource cluster we observe a limitation to this scalability, which results in a dramatic increase in runtime.
+%Hadoop's strength is billed to be its ability to scale to massively more samples.
 We observe this to be true, as increasing the samples by two and a half, which is the case from \SevenPhaseone{} to \ThreePhasethree{}, does not result in greater memory requirements,
-unlike increasing the number of variants per sample.
+%unlike increasing the number of variants per sample.
 
-\subsection*{Result visualisation}
-% DB: in the conclusion we discussed that we provide a "toolset to interpret Mahout's clustering output".
-We hence need to show the R-image for that, i.e.
-spit out distance matrix and read in with R to plot, e.g.
-doing Multidimensional Scaling to squash the matrix in 2D http://www.statmethods.net/advstats/mds.html
-
-% AO: new section
+% (3) CLUSTER AND CONTAINER SIZE
 \subsection*{Cluster size and containers}
@@
 Although sometimes inevitable, depending on the task, this bottleneck of maximum simultaneous reduce tasks is an important consideration to take into account when designing MapReduce jobs.
+
+\subsection*{Result visualisation}
+% DB: in the conclusion we discussed that we provide a "toolset to interpret Mahout's clustering output".
+We hence need to show the R-image for that, i.e.
+spit out distance matrix and read in with R to plot, e.g.
+doing Multidimensional Scaling to squash the matrix in 2D http://www.statmethods.net/advstats/mds.html
 
@@
         & \SevenPhaseone{} & 1536 & 3072 & 4096 & 4096  & 43m 56s & 3h 28m & 0.813\\
-        & \FullPhaseone{} & 1536 & 3072 & 6144 & 6144  & 85m 29s & 7h 43m & 0.723\\
+       % & \FullPhaseone{} & 1536 & 3072 & 6144 & 6144  & 85m 29s & 7h 43m & 0.723\\
         & \OnePhasethree{} & 1536 & 3072 & 1536 & 3072 & 23m 27s & 21m 14s & 0.622\\

### 2015-03-05 doc/publications/journal2014/paper.tex
More writing

@@
 \subsection*{Hadoop scales to genome-wide clustering}
-In this section we demonstrate Mahout's scaling ability by increasing the number of features (variants per sample) to include the entire phase3 autosome dataset (\FullPhasethree{}).
-We focus on the \FullPhasethree{} dataset in this section because it has 79 million variants compared to \FullPhaseone{}'s 38 million.
+In this section we demonstrate Mahout's ability to pre-process and cluster genome-wide variants and its effect on clustering quality.
+We focus on the larger \FullPhasethree{} dataset in this section which has 79 million variants.
+However, we begin with a subset of this dataset, \OnePhasethree{}. Although our algorithm completes successfully, it achieves a relatively low adjusted Rand index of 0.622.
 
-We start by clustering one chromosome. Our algorithm completes successfully, however, it only achieves an adjusted Rand index of 0.622.
-Increasing the dataset to include the first three chromosomes (\ThreePhasethree{}) results in an adjusted Rand index on 0.825
-Finally we include all 22 autosomes as input to our algorithm (\FullPhasethree{}).
+Next we increase the input to include the first three chromosomes (\ThreePhasethree{}), which has just under 20 million variants.
+This completes successfully and results in an adjusted Rand index of 0.825.
+Finally we include all 22 autosomes (\FullPhasethree{}) as input to our algorithm.
 Although the adjusted rand index isn't much higher than the previous example, at 0.84, the pre-processing and k-means complete successfully.
+-Conclusionary statement-
 
@@
 
-In this section we investigate Mahout's ability to scale with more samples.
-Mahout's k-means algorithm performs by splitting the clustering job up into separate tasks, where each task calculates the distance between a subset of samples and the k-means centroids.
-Mahout performs each of these calculations iteratively until they are complete for every sample.
-In the previous section we increased the sample size. Larger samples have and increased dimensionality, and hence require more memory and also more time to calculate the distances.
-In this case however, we increase the number of samples.
-By simply increasing the number of samples, the k-means algorithm will perform more iterations, however the resource cost of each iteration should remain the same.
-To demonstrate this, we compare pre-processing and clustering of the 2504 individuals from the phase3 dataset compared to the 1092 individuals in the phase1 dataset.
+In this section we investigate Hadoop's ability to scale with more samples.
+Mahout's k-means algorithm performs by splitting clustering jobs up into separate, smaller tasks, where each task calculates the distance between a subset of samples and the k-means centroids.
+Hadoop creates a 'container' for each task, and processes them iteratively until every task is complete.
+We discuss containers in greater detail in the next section, however a container is essentially a pre-defined allocation of memory and CPU cores.
+In the previous section we increased the sample size. Larger samples have an increased dimensionality, and hence require more memory per container, and also more time to complete.
+In this case, however, we increase the number of samples.
+With more samples, the k-means algorithm will divide the job up into more tasks, however the memory requirement of each task, and hence, container, should remain the same because the dimensionality is the same.
+Instead the job will take more time because there are more tasks for Mahout to complete.
+To demonstrate this, we compare k-means clustering of the 2504 individuals from the phase3 dataset compared to the 1092 individuals in the phase1 dataset.
 We specifically use the \SevenPhaseone{} and \ThreePhasethree{} datasets because of their similar feature size.
-Even with twice the number of samples in the larger dataset, the algorithm completed successfully for both datasets with the same container size.
-However the \ThreePhasethree{} dataset did require about twenty minutes more time than the smaller \SevenPhaseone{} dataset.
+The algorithm completed successfully with the same sized containers for both datasets, even with more than twice the number of samples in the larger dataset.
+However, the \ThreePhasethree{} dataset required about twenty minutes more time than the smaller \SevenPhaseone{} dataset.
 This demonstrates Hadoop's strength of scaling with sample size.
-Rather than the larger comparison requiring more resources, it simple required more time.
-The container size however, cannot be an arbitrary value, as we discuss in the next section.
+Rather than the larger comparison requiring us to allocate more resources allocated to containers, it simply required more time.
+Therefore, increasing the sample size of a job should not require reconfiguration of Hadoop or a larger cluster.
 
@@
 \subsection*{Cluster size and containers}
-An important consideration when designing a MapReduce job is the container size.
-A container is essentially an abstraction of resources, i.e.
+When designing a MapReduce job, it is important that the user specify container size correctly.
+Overallocation of resources can result in wasted resources, whereas under allocation can result in failed jobs.
+To elaborate, a container is essentially an abstraction of resources, i.e.
 memory and CPU cores.
-Tasks are run in parallel in these containers, and each container is independent of other containers.
+After Hadoop divides a job up into tasks, each task runs in parallel in a container.
+In addition, each container is independent of other containers.
 The resources Hadoop allocates to each container is a predefined value, specified in its configuration.
@@
 These tasks are placed in a queue until a node has enough resources available to run a container, and generally, the more containers that can run simultaneously, the faster the job will complete.
-The number of containers for smaller jobs are generally limited by the available CPU cores, as each container requires an entire CPU core.
-However for jobs that deal with large datasets, the memory allocated per container (following Cloudera's guidelines) may not be enough.
-We could modify the above example to allocate each container 1 CPU core and 8GB memory, however now only four containers can fit within the node's 32GB memory, resulting in the other 4 CPU cores being disused.
+For smaller jobs, the maximum number of containers are generally limited by the available CPU cores because each container requires an entire CPU core.
+The tasks running in each container may not be able to utilise all of the memory allocated to that container.
+However, for jobs that deal with large datasets, the memory allocated per container (following Cloudera's guidelines) may not be enough.
+We could modify the above example to allocate each container 1 CPU core and 8GB memory, however now only four containers can fit within the node's 32GB memory, resulting in the other 4 CPU cores being disused (because only four containers can run simultaneously).
 It is therefore optimal to
@@
 Increasing the map container size to 4GB results in the job taking an additional 5 minutes (see Table \ref{Tab:03}).
-However, this larger container size is still too small for the \FullPhasethree{} dataset, which requires at least 6GB allocated to each map container.
-Running a job using this dataset and 4GB memory per container results in the job failing.
+This larger container size is still too small for the \FullPhasethree{} dataset, which we demonstrated to require 6GB memory per container.
+Running a job on this dataset with 4GB memory per container results in the job failing.
 
@@
 	\ThreePhasethree{} & 2,504 & 19,381,970 & 393.06 & 5\\
-	\FullPhasethree{} & 2,504 & 79,000,000 & 1602.00 & 5\\\botrule
+	\FullPhasethree{} & 2,504 & 81,271,745 & 1602.00 & 5\\\botrule
 \end{tabular}}{Datasets used in this study; based on portions of the phase1 and phase3 data from the 1000 genomes project.

### 2015-03-06 doc/publications/journal2014/paper.tex
updated sections

@@
 \newcommand{\FullPhasethree}{phase3\_chr1-22}
+\newcommand{\ARI}{Adjusted Rand Index}
 
@@
 We measured the runtime separately for pre-processing stage as well as the clustering stage.
-%DB: can you explain 'container' or refer to methods where you have more room for discussing
-%DB: is this addressed?
-With Hadoop, pre-processing the data takes approximately 6 minutes, with an allocation of up to 3GB memory per 'container'.
+
+%With Hadoop, pre-processing the data takes approximately 6 minutes, with an allocation of up to 3GB memory per 'container'.
+The resulting data is stored as a Hadoop 'SequenceFile' of about 835MB in size.
+With Hadoop, pre-processing the data takes approximately 6 minutes, with an allocation of up to 3GB memory. 
 The resulting data is stored as a Hadoop 'SequenceFile' of about 835MB in size.
 Conversely,  pre-processing the data to R's in-memory matrix takes 56 minutes and required 17.7GB memory. 
-K-means clustering the data, however, is faster with R, using 7s, whereas Mahout takes just under 9 minutes. See Table~\ref{Tab:02}. \\
-Due to the relatively small number of variants, cluster performance is relatively poor for both methods, with the best Adjusted Rand Index score for clustering individuals in one of 4 populations (see Methods) being 0.705 in Mahout and 0.614 in R (see Table~\ref{Tab:02}).
+K-means clustering the data, however, is faster with R, using 7s, whereas Mahout takes just under 9 minutes (see Table~\ref{Tab:02}). 
+This is due to the additional overhead involved in spawning the map and reduce processes, and distributing the data between nodes.
+R, on the other hand, already has the entire matrix in-memory, on a single machine and since no disk-access or networking is required the clustering can be performed faster. 
+Due to the relatively small number of variants, cluster accuracy is relatively poor for both methods, with the best \ARI{} score for clustering individuals in one of 4 populations (see Methods) being 0.705 in Mahout and 0.614 in R (see Table~\ref{Tab:02}).
 
-Although the overall Mahout pipeline is about 40 minutes faster than R's, the actual k-means clustering takes two orders of magnitude longer in Mahout. 
-This is due to the additional overhead involved in spawning the map and reduce processes, and distributing the data between nodes.
-R, on the other hand, already has the entire matrix in-memory, on a single machine. 
-Since no disk-access or networking is required the clustering can be performed faster. However, the amount of memory available on a single machine will quickly become a limiting factor. 
+%Although the overall Mahout pipeline is about 40 minutes faster than R's, the actual k-means clustering takes two orders of magnitude longer in Mahout. 
+%This is due to the additional overhead involved in spawning the map and reduce processes, and distributing the data between nodes.
+%R, on the other hand, already has the entire matrix in-memory, on a single machine. 
+%Since no disk-access or networking is required the clustering can be performed faster.
+
+While the overall processing time is already 40 minutes faster in Mahout than R, the real bottleneck for larger dataset will be the amount of memory available on a single machine. 
 As an example, the entire set of 1000 Genome Project phase3 variants could require approximately 1.6TB to be completely loaded into memory as a double-precision matrix (see Table~\ref{Tab:01}).
@@
 
-% DB: not clear you need to spell out your argument a bit more...
-% AO: Fixed maybe
-% DB: good explanation but I'm not sure what it does here (maybe reuse above or in methods)
-% DB: Delete this ?
-%The primary benefit of Hadoop is its scalability. This is achieved through MapReduce jobs, which allow the task at hand to be separated into 'containers'.
-%Although in the previous example, Hadoop had access to 36GB of memory distributed over 3 machines, the size of each container was only 3GB.
-%Each of these containers is effectively a self-contained job that is independent from any other container.
-%Therefore, the more resources that the user has available, the more containers that can be run in parallel, either on the same machine or separate, and in turn decreasing the run-time of the overall job.
-%This means that our Hadoop job can scale from a single machine with minimal resources (albeit with a longer run-time), to a huge compute cluster with terabytes of RAM.
-%This benefit allows Hadoop jobs to not only scale vertically, i.e. to a larger single machine as R can, but also horizontally, to a cluster of many machines.
+Therefore, the data processed in the subsequent sections are too large to be processed with standard R, we hence we focus on evaluating different settings of the Mahout implementation only.
 
@@
 \subsection*{Hadoop scales to genome-wide clustering}
-In this section we demonstrate Mahout's ability to pre-process and cluster genome-wide variants and its effect on clustering quality.
-We focus on the larger \FullPhasethree{} dataset in this section which has 79 million variants.
-However, we begin with a subset of this dataset, \OnePhasethree{}. Although our algorithm completes successfully, it achieves a relatively low adjusted Rand index of 0.622.
+%In this section we demonstrate Mahout's ability to pre-process and cluster genome-wide variants and its effect on clustering quality.
+%We focus on the larger \FullPhasethree{} dataset in this section which has 79 million variants.
+%However, we begin with a subset of this dataset, \OnePhasethree{}. Although our algorithm completes successfully, it achieves a relatively low adjusted Rand index of 0.622.
+%Next we increase the input to include the first three chromosomes (\ThreePhasethree{}), which has just under 20 million variants.
+This completes successfully and results in an adjusted Rand index of 0.825.
+%Finally we include all 22 autosomes (\FullPhasethree{}) as input to our algorithm.
+Although the adjusted rand index isn't much higher than the previous example, at 0.84, the pre-processing and k-means complete successfully.
+%-Conclusionary statement-
 
-Next we increase the input to include the first three chromosomes (\ThreePhasethree{}), which has just under 20 million variants.
-This completes successfully and results in an adjusted Rand index of 0.825.
-Finally we include all 22 autosomes (\FullPhasethree{}) as input to our algorithm.
-Although the adjusted rand index isn't much higher than the previous example, at 0.84, the pre-processing and k-means complete successfully.
--Conclusionary statement-
+In this section we demonstrate Mahout's ability to cluster genome-wide variants. 
+To demonstrate accuracy-gain and resource-consumption we evaluate whole-genome data (\FullPhasethree{}), which contains 79 million variants, as well as two subsets with 19 (\ThreePhasethree{}) and 6 million (\OnePhasethree{}) variants, respectively.
+As expected, providing variants from only chromosome one (\OnePhasethree{}) achieves a relatively low \ARI{} of 0.622 (see Table~\ref{Tab:02})
+Note, phase3 data has one more population group, which explains the decrease in accuracy compared to the previous section.
 
-% This section feels like a listing numbers rather than a "story".
-Maybe having a plot we can make a more general statement.
-So can you make a plot #features vs memory and in a separate on vs time with different colours showing pre-post processing.
-Or one plot memory vs time with colours as features and shape as pre-post ?
-
-%% OLD
-%%We also ran jobs using intermediately sized phase1 datasets (see Table~\ref{Tab:02}) to demonstrate the different container sizes required by pre-processing and k-means clustering, and time required.
-%On the \OnePhaseone{} dataset, pre-processing takes just under 10 minutes and clustering takes approximately 12 minutes with the same container size as with the \NinteenPhaseone{} dataset.
-%We then increased the input to include the first seven chromosomes (\SevenPhaseone{}).
-In this case, pre-processing time increased to approximately 45 minutes, and clustering to 3 hours and 28 minutes.
-%Our pre-processing algorithm was successful with the same container size as the previous examples.
-However, we increased both the map and reduce container size for the clustering step to 4GB to accommodate the increased vector size.
-%
-%Finally we used the entire phase one dataset as input. Although pre-processing now takes 85 minutes, it was again successful with the same sized containers as previously.
-%Clustering, however, not only required larger containers, but also took over 7 hours to complete.
-%
-%This substantial time increase is in part due to the larger container size.
-With each container allocated 6GB of memory, our cluster can only run up to six containers simultaneously, as opposed to 24 1.5GB containers (from the smaller examples).
-%This restriction in simultaneous containers is due to the limited memory available on our cluster.
-%
-%% mentioning phase 3 here makes it difficult to understand.
-%Although we compare phase3 datasets in the next section because they have a larger number of samples, the \FullPhasethree{} dataset also has the largest feature size, so we mention that in this section to demonstrate Hadoops ability to scale with feature-size.
-%The most notable difference is the time required to complete clustering, at just over 21 hours.
-Although this is a substantial increase in time, the job still completes successfully.
-As we mention in a later section, this time can be reduced dramatically on a larger cluster.
-%
-%%DB how else would you expect memory to grow 
-%This section demonstrates Hadoop's ability to cope with genome wide information, which predominantly results in a runtime increase.
-The memory requirements are also increased, however the memory required increases in proportion to vector length,
-%rather than the entire size of the dataset.
-This means we can successfully run a job on the variants from an entire genome on our cluster, or potentially an even smaller cluster, as long as 6GB is available for containers.
-%rather than memory increase which quickly become the limiting factor as demonstrated by R's inability to process the data. 
-% /OLD
+Increasing the input to include the first three chromosomes (\ThreePhasethree{}) improves the accuracy with an \ARI{} of 0.825 (Table~\ref{Tab:02}).
+We are reaching a saturation point for k-means clustering here, as including all 22 autosomes (\FullPhasethree{}) improves the \ARI{} only marginally to 0.84 (Table~\ref{Tab:02}).
+%DB is that really GB or should that be 6.1GB?
+However, this demonstrates that performing clustering using whole-genome variants on research infrastructure is feasible using Mahout as the resource consumption lies within realistic boundaries.
+The time consumption increases to 25 hours and memory to 6144 GB (Table~\ref{Tab:02}).
 
@@
 \subsection*{Hadoop scales effortlessly to more than twice the number of samples}
+%In this section we investigate Hadoop's ability to scale with more samples. 
+%Mahout's k-means algorithm performs by splitting clustering jobs up into separate, smaller tasks, where each task calculates the distance between a subset of samples and the k-means centroids.
+%Hadoop creates a 'container' for each task, and processes them iteratively until every task is complete.
+%In the previous section we increased the sample size. 
+%Larger samples have an increased dimensionality, and hence require more memory per container, and also more time to complete.
+%In this case, however, we increase the number of samples. 
+%With more samples, the k-means algorithm will divide the job up into more tasks, however the memory requirement of each task, and hence, container, should remain the same because the dimensionality is the same.
+%Instead the job will take more time because there are more tasks for Mahout to complete.
+%In the previous section we increased the sample size. 
+%Larger samples have an increased dimensionality, and hence require more memory per container, and also more time to complete.
+%In this case, however, we increase the number of samples. 
+%With more samples, the k-means algorithm will divide the job up into more tasks, however the memory requirement of each task, and hence, container, should remain the same because the dimensionality is the same.
+%Instead the job will take more time because there are more tasks for Mahout to complete.
 
-In this section we investigate Hadoop's ability to scale with more samples.
-Mahout's k-means algorithm performs by splitting clustering jobs up into separate, smaller tasks, where each task calculates the distance between a subset of samples and the k-means centroids.
-Hadoop creates a 'container' for each task, and processes them iteratively until every task is complete.
-We discuss containers in greater detail in the next section, however a container is essentially a pre-defined allocation of memory and CPU cores.
-In the previous section we increased the sample size. Larger samples have an increased dimensionality, and hence require more memory per container, and also more time to complete.
-In this case, however, we increase the number of samples.
-With more samples, the k-means algorithm will divide the job up into more tasks, however the memory requirement of each task, and hence, container, should remain the same because the dimensionality is the same.
-Instead the job will take more time because there are more tasks for Mahout to complete.
+Hadoop's main advantage over other parallelisation techniques is its ability to scale to more samples without altering the code. 
+Mahout's k-means algorithm is hence structured such that it splits the clustering problem up into separate, smaller tasks, where each task calculates the distance between a subset of samples and the k-means centroids.
+The automatic scaling is then achieved by Hadoop creating 'containers' for each task, and processes them iteratively until every task is complete.
+We discuss containers in greater detail in the next section, however, a container is essentially a pre-defined allocation of memory and CPU cores.
+By increasing the number of samples, the k-means algorithm will simply divide the job up into more tasks, which enables the memory requirement of each task to stay the same. 
+This is because the dimensionality is constant, compared to the previous section where we increased the number of variants and observe a subsequent increase in memory.
+
 To demonstrate this, we compare k-means clustering of the 2504 individuals from the phase3 dataset compared to the 1092 individuals in the phase1 dataset.
 We specifically use the \SevenPhaseone{} and \ThreePhasethree{} datasets because of their similar feature size.
-The algorithm completed successfully with the same sized containers for both datasets, even with more than twice the number of samples in the larger dataset.
-However, the \ThreePhasethree{} dataset required about twenty minutes more time than the smaller \SevenPhaseone{} dataset.
+%The algorithm completed successfully with the same sized containers for both datasets, even with more than twice the number of samples in the larger dataset.
+%However, the \ThreePhasethree{} dataset required about twenty minutes more time than the smaller \SevenPhaseone{} dataset. 
+The algorithm completed successfully with the same sized containers for both datasets, however, the \ThreePhasethree{} dataset required about twenty minutes more time than the smaller \SevenPhaseone{} dataset (Table~\ref{Tab:02}).
 This demonstrates Hadoop's strength of scaling with sample size.
-Rather than the larger comparison requiring us to allocate more resources allocated to containers, it simply required more time.
+
+%Rather than the larger comparison requiring us to allocate more resources allocated to containers, it simply required more time.
 Therefore, increasing the sample size of a job should not require reconfiguration of Hadoop or a larger cluster.
@@
 \subsection*{Cluster size and containers}
-When designing a MapReduce job, it is important that the user specify container size correctly.
-Overallocation of resources can result in wasted resources, whereas under allocation can result in failed jobs.
-To elaborate, a container is essentially an abstraction of resources, i.e.
-memory and CPU cores.
-After Hadoop divides a job up into tasks, each task runs in parallel in a container.
-In addition, each container is independent of other containers.
+In the previous section we observed that while the memory remained constant the runtime increased for larger sample sizes. 
+On a larger cluster the runtime would also have remained constant. 
+%When designing a MapReduce job, it is important that the user specify container size correctly.
+However, as research is carried out on resource-finite infrastructure, we discuss in this section how to specify the optimal container size when designing MapReduce jobs to incur only a minimal increase in runtime.
+%Overallocation of resources can result in wasted resources, whereas under allocation can result in failed jobs.
+
+To elaborate, a container is essentially an abstraction of resources, i.e. memory and CPU cores. 
+After Hadoop divides a job up into tasks, each task runs in parallel in a container. In addition, each container is independent of other containers.
 The resources Hadoop allocates to each container is a predefined value, specified in its configuration. 
+% DB reference or version
 According to the Cloudera documentation, the general rule of thumb is that one CPU core is allocated to each container, and the available memory per node should be divided equally between each of those containers.
@@
 The \OnePhaseone{} dataset only requires 1.5GB memory per container, as we demonstrated previously.
-Increasing the map container size to 4GB results in the job taking an additional 5 minutes (see Table \ref{Tab:03}).
-This larger container size is still too small for the \FullPhasethree{} dataset, which we demonstrated to require 6GB memory per container.
-Running a job on this dataset with 4GB memory per container results in the job failing.
+%Increasing the map container size to 4GB results in the job taking an additional 5 minutes (see Table \ref{Tab:03}).
+Increasing the map container size to 4GB results in the job taking an additional 5 minutes (see Table \ref{Tab:03}), because fewer jobs can run in parallel. 
+%This larger container size is still too small for the \FullPhasethree{} dataset, which we demonstrated to require 6GB memory per container. 
+%Running a job on this dataset with 4GB memory per container results in the job failing.
+Attempting to run the larger data set (\FullPhasethree{}) with either 1.5GB or 4GB will result in the job failing, because the feature vector cannot be loaded completely, this requires at least 6GB memory per container and 21h runtime.
 
 One of the main selling points of Hadoop, is that it will run on commodity hardware. 
-For this reason we used a relatively small cluster for our benchmarks.
-Although our cluster uses VMs on Microsoft Azure, it would be relatively simple for a lab to
-create a similar Hadoop setup in-house.
+%For this reason we used a relatively small cluster for our benchmarks. Although our cluster uses VMs on Microsoft Azure, it would be relatively simple for a lab to
+%create a similar Hadoop setup in-house. 
+To mimic the available resource in a standard resource setting we used only a small VM cluster on Microsoft Azure for the benchmarking in this paper so far.
 However, to demonstrate the scaling ability of Hadoop and our MapReduce algorithms, we also ran the clustering jobs on a much larger development Hadoop cluster. 
-As presented in Table \ref{Tab:03}, the \OnePhaseone{} job saw an improvement of approximately 70\%.
-The \FullPhasethree{} example, on the other hand, was 78\% faster.
+%As presented in Table \ref{Tab:03}, the \OnePhaseone{} job saw an improvement of approximately 70\%.
+%The \FullPhasethree{} example, on the other hand, was 78\% faster.
+As presented in Table \ref{Tab:03}, the \OnePhaseone{} job saw an improvement of approximately 70\% when providing 448 cores of 13GB. 
+With these resource, the \FullPhasethree{} example, was 78\% faster.
+
 Although both jobs were more than twice as fast on the larger cluster, the speed doesn't scale relative to the size of the cluster.
@@
 \subsection*{Clustering quality}
-We scored the clusters with the Adjusted Rand index~\cite{Hubert1985}. 
+We scored the clusters with the \ARI{}~\cite{Hubert1985}. 
 This index compares two different cluster sets, here the prediction $X = \{ X_1, X_2, \ldots , X_r \}$ and the annotated data $Y = \{ Y_1, Y_2, \ldots , Y_s \}$, and assigns a value based the overlap between $X$ and $Y$ as captured in a contingency table $\left[n_{ij}\right]$ where each entry $n_{ij}$ denotes the number of objects in common between $X_i$ and $Y_j : n_{ij}=|X_i \cap Y_j|$.
-The Adjusted Rand index is then 
+The \ARI{} is then 
 {\tiny
@@
         & \FullPhasethree{} & 1536 & 3072 & 6144 & 6144 & 5h 36m & 21h 5m &0.840\\\botrule
-\end{tabular}}{The resources allocated to each MapReduce yarn-container and time taken for the job(s) to complete.}
+\end{tabular}}{The resources allocated to each MapReduce yarn-container and time taken for the job(s) to complete as well as the accuracy measured as \ARI{}.}
 \end{table*}
@@
 \processtable{Resources allocated and time taken.\label{Tab:03}}
-{\begin{tabular}{l|cc|cc|r}\toprule
+{\begin{tabular}{cl|cc|cc|r}\toprule
 Dataset & \multicolumn{2}{ c |}{Available Resources} & \multicolumn{2}{ c |}{Container Size (GB)} & Time\\

### 2015-03-06 doc/publications/journal2014/paper.tex
updated sections

@@
 \title[MapReduce Clustering]{Scalable Clustering of Genotype Information using MapReduce}
-\author[O'Brien \textit{et~al}]{Aidan R O'Brien\,$^{1}$, Fabian A Buske\,$^{2}$ and Denis C. Bauer\,$^1$\footnote{to whom correspondence should be addressed}}
+\author[O'Brien \textit{et~al}]{Aidan R O'Brien\,$^{1}$, Neil Saunders\,$^1$,Fabian A Buske\,$^{2}$ and Denis C. Bauer\,$^1$\footnote{to whom correspondence should be addressed}}
 \address{$^{1}$CSIRO, Digital Productivity Flagship, 11 Julius Av, 2113, Sydney, Australia\\
@@
 However, this demonstrates that performing clustering using whole-genome variants on research infrastructure is feasible using Mahout as the resource consumption lies within realistic boundaries.
-The time consumption increases to 25 hours and memory to 6144 GB (Table~\ref{Tab:02}).
+The time consumption increases to 25 hours and memory to 6 GB (Table~\ref{Tab:02}).
 
@@
 % DB: in the conclusion we discussed that we provide a "toolset to interpret Mahout's clustering output". 
-We hence need to show the R-image for that, i.e.
-spit out distance matrix and read in with R to plot, e.g.
-doing Multidimensional Scaling to squash the matrix in 2D http://www.statmethods.net/advstats/mds.html
+% DB: can we discuss here how we get a visual representation from the clustering (ClusterDumper+Gephi) and what the image tells us..
 

### 2015-03-09 doc/publications/journal2014/paper.tex
Fixed table

@@
 % DB reference or version
-According to the Cloudera documentation, the general rule of thumb is that one CPU core is allocated to each container, and the available memory per node should be divided equally between each of those containers.
+According to the Hortonworks documentation \cite{Horton2013}, the general rule of thumb is that one CPU core is allocated to each container, and the available memory per node should be divided equally between each of those containers.
 For example, on a node with 8 CPU cores and 32GB memory, each container should be allocated 1 CPU core and 4GB memory.
@@
 %Increasing the map container size to 4GB results in the job taking an additional 5 minutes (see Table \ref{Tab:03}).
-Increasing the map container size to 4GB results in the job taking an additional 5 minutes (see Table \ref{Tab:03}), because fewer jobs can run in parallel. 
+Increasing the map container size to 4GB results in the job taking an additional 5 minutes (see Table~\ref{Tab:03}), because fewer jobs can run in parallel. 
 %This larger container size is still too small for the \FullPhasethree{} dataset, which we demonstrated to require 6GB memory per container. 
@@
 %The \FullPhasethree{} example, on the other hand, was 78\% faster.
-As presented in Table \ref{Tab:03}, the \OnePhaseone{} job saw an improvement of approximately 70\% when providing 448 cores of 13GB. 
+As presented in Table~\ref{Tab:03}, the \OnePhaseone{} job saw an improvement of approximately 70\% when providing 448 cores of 13GB. 
 With these resource, the \FullPhasethree{} example, was 78\% faster.
@@
 To demonstrate scaling with Hadoop, we use subsets that contain more variants, i.e. \SevenPhaseone{} and \FullPhaseone{}, which include chromosome 1-7 and the full genome, respectively.
-Table~\ref{datasets} contains an overview of the datasets.
+Table~\ref{Tab:01} contains an overview of the datasets.
 
@@
 {\begin{tabular}{ll|cc|cc|rr|c}\toprule
-Tool & Dataset & \multicolumn{2}{ c |}{Preprocesing (GB)} & \multicolumn{2}{ c |}{Clustering (GB)} & \multicolumn{2}{ c |}{Time} & Accuracy\\
+Tool & Dataset & \multicolumn{2}{ c |}{Preprocesing (MB)} & \multicolumn{2}{ c |}{Clustering (MB)} & \multicolumn{2}{ c |}{Time} & Accuracy\\
 & & map & reduce  & map & reduce  & pre-processing & clustering & \\\midrule
@@
 \processtable{Resources allocated and time taken.\label{Tab:03}}
-{\begin{tabular}{cl|cc|cc|r}\toprule
-Dataset & \multicolumn{2}{ c |}{Available Resources} & \multicolumn{2}{ c |}{Container Size (GB)} & Time\\
+{\begin{tabular}{c|cc|cc|c}\toprule
+Dataset & \multicolumn{2}{ c |}{Available Resources} & \multicolumn{2}{ c |}{Container Size (MB)} & Time\\
 & CPU cores & Memory (GB)  & map & reduce & \\\midrule

### 2015-03-11 doc/publications/journal2014/paper.tex
Added methods, results and figure relevant to Gephi visualisation

@@
 
+\begin{figure}[!tpb]%figure1
+\centerline{\includegraphics[type=pdf,ext=.pdf,read=.pdf, scale=0.40]{gephi-p3}}
+        \label{fig:sign}
+        \caption{{\bf Visualisation of k-means clusters derived from the phase 3 dataset.}
+        Nodes are coloured by super population: AFR (green), AMR (blue), EAS (pink), EUR (red) and SAS (cyan). 
+}
+\end{figure}
+
 \subsection*{Result visualisation}
@@
 % DB: can we discuss here how we get a visual representation from the clustering (ClusterDumper+Gephi) and what the image tells us..
+GraphML files generated using Mahout Cluster Dumper can be visualised using popular freely-available tools such as Gephi~\cite{ICWSM09154}.
+An example visualisation using the phase 3 dataset, with nodes coloured by super population, is shown in Figure XX.
+
+The clusters generated by k-means clustering and their population structure are readily apparent.
+Three clusters are homogeneous, containing only EAS (1 cluster) or SAS individuals (2 clusters, the second small and unconnected).
+Two clusters contain predominantly one super population plus a minor component of another: AMR + AFR or AFR + AMR.
+The sixth cluster is predominantly a mix of EUR + AMR, with 3 AFR individuals and 1 SAS.
 
@@
 
+\subsection*{Graph visualisation}
+GraphML files generated using Mahout Cluster Dumper were visualised using Gephi.
+Layouts were applied in the following order: Random, ForceAtlas 2, Yifan Hu Proportional.
+For ForceAtlas 2 behaviour alternatives ``dissuade hubs'', ``linlog mode'' and ``prevent overlap'' were set to true, ``approximate repulsion'' was set to false.
+
 \end{methods}
@@
 
-\begin{figure}[!tpb]%figure1
+\begin{figure}[!tpb]%figure2
 \centerline{\includegraphics[type=pdf,ext=.pdf,read=.pdf, scale=0.40]{signature}}

### 2015-03-11 doc/publications/journal2014/paper.tex
Corrected figure numbers

@@
 GraphML files generated using Mahout Cluster Dumper can be visualised using popular freely-available tools such as Gephi~\cite{ICWSM09154}.
-An example visualisation using the phase 3 dataset, with nodes coloured by super population, is shown in Figure XX.
+An example visualisation using the phase 3 dataset, with nodes coloured by super population, is shown in Figure 1.
 
@@
 In this paper we have applied Mahout's k-means clustering to the task of clustering individuals, based on their genomic variants. 
-We provide an interface from VCF format to Hadoop and Mahout, and a toolset to interpret Mahout's clustering output (see Figure 1)
+We provide an interface from VCF format to Hadoop and Mahout, and a toolset to interpret Mahout's clustering output (see Figure 2)
 We find that on small datasets the clustering stage is two orders of magnitudes slower in Mahout compared to R (9 minutes vs 7 seconds), due to the overhead of setting up and co-ordinating the communication between the Hadoop nodes.

### 2015-03-11 doc/publications/journal2014/paper.tex
Fixing things here and there

@@
 
-To demonstrate this, we compare k-means clustering of the 2504 individuals from the phase3 dataset compared to the 1092 individuals in the phase1 dataset.
+To demonstrate this, we compare k-means clustering of the 2504 individuals from the phase3 dataset compared to the 1092 individuals from the phase1 dataset.
 We specifically use the \SevenPhaseone{} and \ThreePhasethree{} datasets because of their similar feature size.
@@
 To elaborate, a container is essentially an abstraction of resources, i.e. memory and CPU cores. 
-After Hadoop divides a job up into tasks, each task runs in parallel in a container. In addition, each container is independent of other containers.
-The resources Hadoop allocates to each container is a predefined value, specified in its configuration. 
+After Hadoop divides a job up into tasks, each task runs in parallel in a container. Also, each container is independent of other containers.
+The resources Hadoop allocates to each container is a predefined value, specified by the user in Hadoop's configuration. 
 % DB reference or version
-According to the Hortonworks documentation \cite{Horton2013}, the general rule of thumb is that one CPU core is allocated to each container, and the available memory per node should be divided equally between each of those containers.
+According to the Hortonworks documentation (\cite{Horton2013}), the general rule of thumb is that one CPU core is allocated to each container, and the available memory per node should be divided equally between each of those containers.
 For example, on a node with 8 CPU cores and 32GB memory, each container should be allocated 1 CPU core and 4GB memory.
@@
 
-To demonstrate the importance of specifying these parameters correctly, we ran the identical jobs, using different Hadoop configurations. 
+To demonstrate the importance of specifying these parameters correctly, we ran identical jobs, using different Hadoop configurations. 
 The \OnePhaseone{} dataset only requires 1.5GB memory per container, as we demonstrated previously.
 %Increasing the map container size to 4GB results in the job taking an additional 5 minutes (see Table \ref{Tab:03}).
-Increasing the map container size to 4GB results in the job taking an additional 5 minutes (see Table~\ref{Tab:03}), because fewer jobs can run in parallel. 
+Increasing the map container size to 4GB results in the job taking an additional 5 minutes (see Table~\ref{Tab:03}), because fewer tasks can run in parallel. 
 %This larger container size is still too small for the \FullPhasethree{} dataset, which we demonstrated to require 6GB memory per container. 
 %Running a job on this dataset with 4GB memory per container results in the job failing.
-Attempting to run the larger data set (\FullPhasethree{}) with either 1.5GB or 4GB will result in the job failing, because the feature vector cannot be loaded completely, this requires at least 6GB memory per container and 21h runtime.
+Attempting to run the larger data set (\FullPhasethree{}) with either 1.5GB or 4GB will result in the job failing, because the feature vectors cannot be loaded completely, this requires at least 6GB memory per container and 21h runtime.
 
@@
 As presented in Table~\ref{Tab:03}, the \OnePhaseone{} job saw an improvement of approximately 70\% when providing 448 cores of 13GB. 
-With these resource, the \FullPhasethree{} example, was 78\% faster.
+With these resources, the \FullPhasethree{} example, was 78\% faster.
 
 Although both jobs were more than twice as fast on the larger cluster, the speed doesn't scale relative to the size of the cluster.
-This is due to a limitation in these examples, where the number of reduces in k-means clustering is limited to the number of k-means clusters (k).
-A task where k is five, for example, will only be able to spawn five reduce containers, no matter how many resources are available.
-Although sometimes inevitable, depending on the task, this bottleneck of maximum simultaneous reduce tasks is an important consideration to take into account when designing MapReduce jobs.
+This is due to a limitation in these examples, where the number of reduce tasks in k-means clustering is limited to the number of k-means clusters (k).
+A task where k is five, for example, will only be able to spawn five reduce tasks, no matter how many resources are available.
+Although sometimes inevitable, this bottleneck of maximum simultaneous reduce tasks is an important consideration to take into account when designing MapReduce jobs.
 

### 2015-03-11 doc/publications/journal2014/paper.tex
changes to abstract

--- 2015-03-11 doc/publications/journal2014/paper.tex
@@
 \section{Motivation:}
-Processing genomic information from whole genome sequence studies pose computational challenges due to the unprecedented data volume generated, which render transitional approaches insufficient.
-However, by utilising advancements in modern hardware accelerators and data processing we can provide the means for scalable solutions.
-We therefore aim to provide the interface between standard genomic data formats and advanced and scalable analysis libraries like Mahout.
+Processing genomic information from whole genome sequence studies pose computational challenges due to the unprecedented data volume, which render transitional approaches insufficient. 
+However, advancements in modern hardware accelerators and data processing provides the means for overcoming these limitations. 
+Here we introduce the interface we developed between the standard variant format (VCF) and the scalable analysis library Mahout, which opens up the application of advanced, efferent machine learning algorithms to genomic data.
 \section{Results:}

### 2015-03-11 doc/publications/journal2014/paper.tex
More small changes

--- 2015-03-11 doc/publications/journal2014/paper.tex
@@
 However, recent technological advances in whole genome sequencing have made population-scale sequencing feasible.
-It is hence economical to generate studies with sample sizes currently reserved for larger consortia such as the 1000 genomes project~\cite{1KG2012} or the cancer genome atlas (TCGA)~\cite{TCGA2013}.
+It is hence economical to generate studies with sample sizes currently reserved for larger consortia such as the 1000 Genomes Project~\cite{1KG2012} or the cancer genome atlas (TCGA)~\cite{TCGA2013}.
 At the same time, whole genome sequencing enables the inclusion of rare or even somatic mutations in the analysis, increasing the feature space by orders of magnitude.
@@
 
-To demonstrate the capability we cluster variant datasets from the 1000 genomes project to determine population structure using the k-means clustering algorithm available in Mahout.
+To demonstrate the capability we cluster variant datasets from the 1000 Genomes Project to determine population structure using the k-means clustering algorithm available in Mahout.
 In the first section we benchmark the performance and accuracy of Mahout's implementations against a standard R based implementation on a reduced version of the data and in the last two sections we investigate how Mahout scales to either more samples or more features per samples.
@@
 While the overall processing time is already 40 minutes faster in Mahout than R, the real bottleneck for larger dataset will be the amount of memory available on a single machine. 
-As an example, the entire set of 1000 Genome Project phase3 variants could require approximately 1.6TB to be completely loaded into memory as a double-precision matrix (see Table~\ref{Tab:01}).
+As an example, the entire set of 1000 Genomes Project phase3 variants could require approximately 1.6TB to be completely loaded into memory as a double-precision matrix (see Table~\ref{Tab:01}).
 This is based on the equation:
@@
 We scored the clusters with the \ARI{}~\cite{Hubert1985}. 
-This index compares two different cluster sets, here the prediction $X = \{ X_1, X_2, \ldots , X_r \}$ and the annotated data $Y = \{ Y_1, Y_2, \ldots , Y_s \}$, and assigns a value based the overlap between $X$ and $Y$ as captured in a contingency table $\left[n_{ij}\right]$ where each entry $n_{ij}$ denotes the number of objects in common between $X_i$ and $Y_j : n_{ij}=|X_i \cap Y_j|$.
+This index compares two different cluster sets, here the prediction $X = \{ X_1, X_2, \ldots , X_r \}$ and the annotated data $Y = \{ Y_1, Y_2, \ldots , Y_s \}$, and assigns a value based on the overlap between $X$ and $Y$ as captured in a contingency table $\left[n_{ij}\right]$ where each entry $n_{ij}$ denotes the number of objects in common between $X_i$ and $Y_j : n_{ij}=|X_i \cap Y_j|$.
 The \ARI{} is then 
@@
 The value is zero for independent clusterings and one for identical clusterings. 
-We compared the Mahout k-means clusters and the R k-means clusters to the known clusters based on meta data from the 1000 genome project.
+We compared the Mahout k-means clusters and the R k-means clusters to the known clusters based on meta data from the 1000 Genomes Project.
 For the known clusters, individuals from the both phase1 and phase3 datasets were partitioned by their super population code, ASN, EUR, AFR, and AMR, where phase3 included EAS and SAS instead of ASN.
@@
 \subsection*{Graph visualisation}
-GraphML files generated using Mahout Cluster Dumper were visualised using Gephi.
-Layouts were applied in the following order: Random, ForceAtlas 2, Yifan Hu Proportional.
+GraphML files generated using Mahout Cluster Dumper were visualised using Gephi. Layouts were applied in the following order: Random, ForceAtlas 2, Yifan Hu Proportional.
 For ForceAtlas 2 behaviour alternatives ``dissuade hubs'', ``linlog mode'' and ``prevent overlap'' were set to true, ``approximate repulsion'' was set to false.
@@
 	\FullPhasethree{} & 2,504 & 81,271,745 & 1602.00 & 5\\\botrule
-\end{tabular}}{Datasets used in this study; based on portions of the phase1 and phase3 data from the 1000 genomes project.
+\end{tabular}}{Datasets used in this study; based on portions of the phase1 and phase3 data from the 1000 Genomes Project.
 In-memory size is the approximate memory required to store the data as a double-precision matrix of size n,m using the formula: n*m*8bytes/10\textsuperscript{9}bytes.

### 2015-03-11 doc/publications/journal2014/paper.tex
changes to text

--- 2015-03-11 doc/publications/journal2014/paper.tex
@@
 \title[MapReduce Clustering]{Scalable Clustering of Genotype Information using MapReduce}
-\author[O'Brien \textit{et~al}]{Aidan R O'Brien\,$^{1}$, Neil Saunders\,$^1$,Fabian A Buske\,$^{2}$ and Denis C. Bauer\,$^1$\footnote{to whom correspondence should be addressed}}
+\author[O'Brien \textit{et~al}]{Aidan R. O'Brien\,$^{1}$, Neil Saunders\,$^1$,Fabian A. Buske\,$^{2}$ and Denis C. Bauer\,$^1$\footnote{to whom correspondence should be addressed}}
 \address{$^{1}$CSIRO, Digital Productivity Flagship, 11 Julius Av, 2113, Sydney, Australia\\
@@
 \section{Motivation:}
-Processing genomic information from whole genome sequence studies pose computational challenges due to the unprecedented data volume, which render transitional approaches insufficient. 
+Processing genomic information from whole genome sequence studies poses computational challenges due to the unprecedented data volume, which render transitional approaches insufficient. 
 However, advancements in modern hardware accelerators and data processing provides the means for overcoming these limitations. 
-Here we introduce the interface we developed between the standard variant format (VCF) and the scalable analysis library Mahout, which opens up the application of advanced, efferent machine learning algorithms to genomic data.
+%Here we introduce the interface we developed between the standard variant format (VCF) and the scalable analysis library Mahout, which opens up the application of advanced, efferent machine learning algorithms to genomic data.
+Here we introduce an interface between the standard variant format (VCF) and the scalable analysis library Mahout, which opens up the application of advanced, efficient machine learning algorithms to genomic data.
+
 \section{Results:}
-We achieve an 2-fold speedup by using the scalable k-means MapReduce implementation over the equivalent analysis performed in R, by comparable accuracy.
-However, the real benefit lies in scaling beyond R's capability to a population-size analysis.
-We successfully clustered more than 2,500 individuals each having more than 19 Million variants.
-
+We achieve more than 2-fold speedup by using the scalable k-means MapReduce implementation over the equivalent analysis performed in R, by comparable accuracy.
+However, the real advantage lies in the scalability of MapReduce beyond R's capability to a genome-wide analysis.
+We successfully clustered more than 2,500 individuals each having more than 81 Million variants.
 \section{Availability:}
-Using modern compute paradigms is essential to scale to modern genomic research in an efficient sustainable way.
+%DB this probably is not the final URL...
+The package is written in Java and available at \url{https://stash.csiro.au/users/bau04c/repos/bioinfbiglearning/browse}.
 
@@
 
-Grouping individuals based on the genomic profile is a commonly performed tasks to identify population structure~\cite{Gao2007} or elucidate different haplotype involvement in diseases susceptibility~\cite{Laitman2013}.
-Traditionally both the number of individuals and included genotypes, typically from SNP arrays, were relatively small and libraries in Bioconductor sufficient.
+Grouping individuals based on their genomic profile is a commonly performed tasks to identify population structure~\citep{Gao2007} or elucidate different haplotype involvement in diseases susceptibility~\citep{Laitman2013}.
+Traditionally both, the number of individuals and included genotypes (typically from SNP arrays), were relatively small and libraries in Bioconductor, which commonly perform all calculations in memory, sufficient.
 However, recent technological advances in whole genome sequencing have made population-scale sequencing feasible. 
-It is hence economical to generate studies with sample sizes currently reserved for larger consortia such as the 1000 genomes project~\cite{1KG2012} or the cancer genome atlas (TCGA)~\cite{TCGA2013}.
+It is hence economical to generate studies with sample sizes previously reserved for larger consortia such as the 1000 genomes project~\citep{1KG2012} or The Cancer Genome Atlas, TCGA~\citep{TCGA2013}.
 At the same time, whole genome sequencing enables the inclusion of rare or even somatic mutations in the analysis, increasing the feature space by orders of magnitude.
@@
 
-As a result of these big data challenges, MapReduce approaches are increasingly being used in bioinformatics (for reviews see~\cite{Zou2013, Qiu2010,Taylor2010}).
-This is especially the case for sequence analysis tasks, such as read mapping~\cite{Schatz2009}, duplicate removal~\cite{Jourdren2012}, and variant calling~\cite{Langmead2009, McKenna2010} as well as Genome Wide Analysis Study based tasks~\cite{Huang2013, Guo2014}.
+As a result of these big data challenges, MapReduce approaches are increasingly being used in bioinformatics (for reviews see~\citep{Zou2013, Qiu2010,Taylor2010}).
+This is especially the case for sequence analysis tasks, such as read mapping~\citep{Schatz2009}, duplicate removal~\citep{Jourdren2012}, and variant calling~\citep{Langmead2009, McKenna2010} as well as Genome Wide Analysis Study based tasks~\citep{Huang2013, Guo2014}.
 
 %Mention Goolge's population clustering in the intro
+At the same time, computer scientists extent MapReduce paradigms to cope with the iterative nature of machine learning tasks~\citep{Chu2009}.
+Specifically, the Mahout project \url{https://mahout.apache.org/} has been developed extensively~\citep{Ranger2007, Owen2011} and was successfully applied in the clinical informatics space~\citep{Dong2013}.
 
-At the same time, theoretical computer scientists adapt MapReduce paradigms to cope with the iterative nature of machine learning tasks~\cite{Chu2009}.
-Specifically, the Mahout project \url{https://mahout.apache.org/} has been developed extensively~\cite{Ranger2007, Owen2011} and was successfully applied in the clinical informatics space~\cite{Dong2013}.
-
-In this paper we link the two areas of parallel machine learning and bioinformatics by providing an interface between Mahout and the standard variant data format, variant call format (VCF)~\cite{1KG2012}, which opens up the application of Mahout's different machine learning algorithms to be applied to genotype-based tasks.
-
-To demonstrate the capability we cluster variant datasets from the 1000 genomes project to determine population structure using the k-means clustering algorithm available in Mahout.
-In the first section we benchmark the performance and accuracy of Mahout's implementations against a standard R based implementation on a reduced version of the data and in the last two sections we investigate how Mahout scales to either more samples or more features per samples.
-
-%SeqWare~{OConnor2010} Libraries \cite{Doering2008,Schumacher2014,Nordberg2013}
+In this paper, we link the two areas of parallel machine learning and bioinformatics by providing an interface between Mahout and the standard variant data format, variant call format (VCF)~\citep{1KG2012}, which opens up the application of Mahout's different machine learning algorithms to be applied to genotype-based tasks.
+To demonstrate the capability, we cluster variant datasets from the 1000 genomes project to determine population structure using the k-means clustering algorithm available in Mahout.
+In the first section we benchmark the performance and accuracy of Mahout's implementations against a standard R based implementation on a subset of the data and in following two sections we investigate Mahout's scaleability with respect to sample size and features per sample.
+We then discuss design considerations for the resource allocation and give a visualisation example in the last two sections.
 
@@
 \subsection*{Clustering in Mahout requires less memory than in R}
-In this section, we compare the runtime and resource requirements between the k-means clustering implementation in R and Mahout for the small \NinteenPhaseone{} dataset (see Methods). 
+In this section, we compare the runtime and resource requirements between the k-means clustering implementation in R and Mahout for the small \NinteenPhaseone{} dataset (see Methods, Table~\ref{Tab:01}).
 We measured the runtime separately for pre-processing stage as well as the clustering stage.
 
-%With Hadoop, pre-processing the data takes approximately 6 minutes, with an allocation of up to 3GB memory per 'container'.
-The resulting data is stored as a Hadoop 'SequenceFile' of about 835MB in size.
 With Hadoop, pre-processing the data takes approximately 6 minutes, with an allocation of up to 3GB memory. 
@@
 R, on the other hand, already has the entire matrix in-memory, on a single machine and since no disk-access or networking is required the clustering can be performed faster. 
-Due to the relatively small number of variants, cluster accuracy is relatively poor for both methods, with the best \ARI{} score for clustering individuals in one of 4 populations (see Methods) being 0.705 in Mahout and 0.614 in R (see Table~\ref{Tab:02}).
+Due to the relatively small number of variants, cluster accuracy is relatively poor for both methods, with the best \ARI{}~\citep{Hubert1985} score for clustering individuals in one of 4 populations (see Methods) being 0.705 in Mahout and 0.614 in R (see Table~\ref{Tab:02}, an \ARI{} of 1 means perfect clustering).
 
-%Although the overall Mahout pipeline is about 40 minutes faster than R's, the actual k-means clustering takes two orders of magnitude longer in Mahout. 
-%This is due to the additional overhead involved in spawning the map and reduce processes, and distributing the data between nodes.
-%R, on the other hand, already has the entire matrix in-memory, on a single machine. 
-%Since no disk-access or networking is required the clustering can be performed faster.
-
-While the overall processing time is already 40 minutes faster in Mahout than R, the real bottleneck for larger dataset will be the amount of memory available on a single machine. 
+While the overall processing time is 40 minutes faster in Mahout compared to R, the real bottleneck for large datasets is the amount of memory available on a single machine. 
 As an example, the entire set of 1000 Genome Project phase3 variants could require approximately 1.6TB to be completely loaded into memory as a double-precision matrix (see Table~\ref{Tab:01}).
 This is based on the equation:
-$\frac{n \times m \times 8\text{bytes}}{1\text{GB}}$, where $n \times m$ are the matrix dimensions, 8bytes is the memory a double occupies (i.e. each matrix element), and  1GB is $10^9$bytes.
+$\frac{n \times m \times 8\text{bytes}}{1\text{GB}}$, where $n \times m$ are the matrix dimensions, 8bytes is the memory a \texttt{double} occupies (i.e.
+each matrix element), and  1GB is $10^9$ bytes.
 
-Therefore, the data processed in the subsequent sections are too large to be processed with standard R, we hence we focus on evaluating different settings of the Mahout implementation only.
+Therefore, the data processed in the subsequent sections are too large to be processed with standard R, hence we focus on evaluating different settings of the Mahout implementation only.
 
@@
 \subsection*{Hadoop scales to genome-wide clustering}
-%In this section we demonstrate Mahout's ability to pre-process and cluster genome-wide variants and its effect on clustering quality.
-%We focus on the larger \FullPhasethree{} dataset in this section which has 79 million variants.
-%However, we begin with a subset of this dataset, \OnePhasethree{}. Although our algorithm completes successfully, it achieves a relatively low adjusted Rand index of 0.622.
-%Next we increase the input to include the first three chromosomes (\ThreePhasethree{}), which has just under 20 million variants.
-This completes successfully and results in an adjusted Rand index of 0.825.
-%Finally we include all 22 autosomes (\FullPhasethree{}) as input to our algorithm.
-Although the adjusted rand index isn't much higher than the previous example, at 0.84, the pre-processing and k-means complete successfully.
-%-Conclusionary statement-
-
 In this section we demonstrate Mahout's ability to cluster genome-wide variants. 
-To demonstrate accuracy-gain and resource-consumption we evaluate whole-genome data (\FullPhasethree{}), which contains 79 million variants, as well as two subsets with 19 (\ThreePhasethree{}) and 6 million (\OnePhasethree{}) variants, respectively.
-As expected, providing variants from only chromosome one (\OnePhasethree{}) achieves a relatively low \ARI{} of 0.622 (see Table~\ref{Tab:02})
+To demonstrate accuracy-gain and resource-consumption we evaluate whole-genome data (\FullPhasethree{}), which contains 79 million variants, as well as two subsets with 19 (\ThreePhasethree{}) and 6 million (\OnePhasethree{}) variants, respectively  (see Table~\ref{Tab:01}).
+As expected, providing variants from only chromosome one (\OnePhasethree{}) holds to little information and hence achieves a relatively low \ARI{} of 0.622 (see Table~\ref{Tab:02}).
 Note, phase3 data has one more population group, which explains the decrease in accuracy compared to the previous section.
@@
 We are reaching a saturation point for k-means clustering here, as including all 22 autosomes (\FullPhasethree{}) improves the \ARI{} only marginally to 0.84 (Table~\ref{Tab:02}).
-%DB is that really GB or should that be 6.1GB?
-However, this demonstrates that performing clustering using whole-genome variants on research infrastructure is feasible using Mahout as the resource consumption lies within realistic boundaries.
+However, this demonstrates that performing clustering using whole-genome variants on research infrastructure using Mahout is feasible as the resource consumption lies within realistic boundaries. 
 The time consumption increases to 25 hours and memory to 6 GB (Table~\ref{Tab:02}).
@@
 \subsection*{Hadoop scales effortlessly to more than twice the number of samples}
-%In this section we investigate Hadoop's ability to scale with more samples. 
-%Mahout's k-means algorithm performs by splitting clustering jobs up into separate, smaller tasks, where each task calculates the distance between a subset of samples and the k-means centroids.
-%Hadoop creates a 'container' for each task, and processes them iteratively until every task is complete.
-%In the previous section we increased the sample size. 
-%Larger samples have an increased dimensionality, and hence require more memory per container, and also more time to complete.
-%In this case, however, we increase the number of samples. 
-%With more samples, the k-means algorithm will divide the job up into more tasks, however the memory requirement of each task, and hence, container, should remain the same because the dimensionality is the same.
-%Instead the job will take more time because there are more tasks for Mahout to complete.
-%In the previous section we increased the sample size. 
-%Larger samples have an increased dimensionality, and hence require more memory per container, and also more time to complete.
-%In this case, however, we increase the number of samples. 
-%With more samples, the k-means algorithm will divide the job up into more tasks, however the memory requirement of each task, and hence, container, should remain the same because the dimensionality is the same.
-%Instead the job will take more time because there are more tasks for Mahout to complete.
-
 Hadoop's main advantage over other parallelisation techniques is its ability to scale to more samples without altering the code. 
-Mahout's k-means algorithm is hence structured such that it splits the clustering problem up into separate, smaller tasks, where each task calculates the distance between a subset of samples and the k-means centroids.
-The automatic scaling is then achieved by Hadoop creating 'containers' for each task, and processes them iteratively until every task is complete.
+Mahout's k-means algorithm is structured such that it splits the clustering problem into separate, smaller tasks, where each task calculates the distance between a subset of samples and the k-means centroids.
+The scalability is achieved by Hadoop creating 'containers' for each task and subsequent iterative processing until every task is completed.
 We discuss containers in greater detail in the next section, however, a container is essentially a pre-defined allocation of memory and CPU cores.
-By increasing the number of samples, the k-means algorithm will simply divide the job up into more tasks, which enables the memory requirement of each task to stay the same. 
+By increasing the number of samples, the k-means algorithm simply divides the job into more tasks, which enables the memory requirement of each task to stay the same.
 This is because the dimensionality is constant, compared to the previous section where we increased the number of variants and observe a subsequent increase in memory.
@@
 We specifically use the \SevenPhaseone{} and \ThreePhasethree{} datasets because of their similar feature size.
-%The algorithm completed successfully with the same sized containers for both datasets, even with more than twice the number of samples in the larger dataset.
-%However, the \ThreePhasethree{} dataset required about twenty minutes more time than the smaller \SevenPhaseone{} dataset. 
 The algorithm completed successfully with the same sized containers for both datasets, however, the \ThreePhasethree{} dataset required about twenty minutes more time than the smaller \SevenPhaseone{} dataset (Table~\ref{Tab:02}).
@@
 
-%Rather than the larger comparison requiring us to allocate more resources allocated to containers, it simply required more time.
-Therefore, increasing the sample size of a job should not require reconfiguration of Hadoop or a larger cluster.
-
-%% OLD
-%The container size required for pre-processing the data is identical for both datasets, however the \ThreePhasethree{} dataset required about twenty minutes more time than the smaller \SevenPhaseone{} dataset.
-%The clustering accuracy is 0.813 for the phase1 dataset and 0.825 for the phase3 dataset.
-%%In this case, due to the number of Feature Vectors increasing, rather than the feature size, we expect the memory requirements for k-means clustering should remain similar, as the cluster centres loaded into the containers' memory are approximately the same size.
-%%We expect the memory requirement for the k-means clustering to stay the same as the number of cluster centres loaded into the containers' memory remain constant despite the increase of feature vectors oppose to feature size.
-%We expect the memory requirement for the k-means clustering algorithm to be the similar for both datasets, because the larger size of \SevenPhaseone{} is a result of more feature vectors (more individuals), rather than larger feature vectors (more variants).
-%Mahout's iterative k-means algorithm loads a set number of these vectors into memory, based on the number of k-means clusters.
-And therefore, if the feature size doesn't increase, this memory requirement also won't increase.
-%This proved to be the case, as 4GB allocated per container was the minimum for both datasets.
-
-%%DB: do you agree with this concluding statement?
-%%Scaling effortless to massively more samples is the strength of Hadoop, hence compute problems need to be carefully designed to harness this ability.  
-%%Hadoop's strength is billed to be its ability to scale to massively more samples, however on a finite-resource cluster we observe a limitation to this scalability, which results in a dramatic increase in runtime.
-%Hadoop's strength is billed to be its ability to scale to massively more samples.
-We observe this to be true, as increasing the samples by two and a half, which is the case from \SevenPhaseone{} to \ThreePhasethree{}, does not result in greater memory requirements,
-%unlike increasing the number of variants per sample.
-
 % (3) CLUSTER AND CONTAINER SIZE
 \subsection*{Cluster size and containers}
-In the previous section we observed that while the memory remained constant the runtime increased for larger sample sizes. 
-On a larger cluster the runtime would also have remained constant. 
-%When designing a MapReduce job, it is important that the user specify container size correctly.
+In the previous section we observed that while the memory remained constant the runtime increases with sample size. 
+On a larger Hadoop cluster the runtime could also stay constant if the number of additional tasks can be accommodated by additional compute nodes. 
 However, as research is carried out on resource-finite infrastructure, we discuss in this section how to specify the optimal container size when designing MapReduce jobs to incur only a minimal increase in runtime.
-%Overallocation of resources can result in wasted resources, whereas under allocation can result in failed jobs.
 
 To elaborate, a container is essentially an abstraction of resources, i.e. memory and CPU cores. 
-After Hadoop divides a job up into tasks, each task runs in parallel in a container. Also, each container is independent of other containers.
+After Hadoop divides a job into tasks, each task runs in parallel in a container. Also, each container is independent of other containers.
 The resources Hadoop allocates to each container is a predefined value, specified by the user in Hadoop's configuration. 
-% DB reference or version
-According to the Hortonworks documentation (\cite{Horton2013}), the general rule of thumb is that one CPU core is allocated to each container, and the available memory per node should be divided equally between each of those containers.
+According to the Hortonworks documentation~\citep{Horton2013}, the general rule of thumb is that one CPU core is allocated to each container, and the available memory per node should be divided equally between each of those containers.
 For example, on a node with 8 CPU cores and 32GB memory, each container should be allocated 1 CPU core and 4GB memory.
@@
 However, for jobs that deal with large datasets, the memory allocated per container (following Cloudera's guidelines) may not be enough.
-We could modify the above example to allocate each container 1 CPU core and 8GB memory, however now only four containers can fit within the node's 32GB memory, resulting in the other 4 CPU cores being disused (because only four containers can run simultaneously).
-It is therefore optimal to
-allocate the appropriate resources to containers, and only increase container memory when necessary, as overallocation of memory will simply limit the number of simultaneous containers each node can run.
+We could modify the above example to allocate each container 1 CPU core and 8GB memory, however now only four containers can fit within the node's 32GB memory, resulting in the other 4 CPU cores being idle (as only four containers can run simultaneously).
+It is therefore optimal to allocate the appropriate resources to containers, and only increase container memory when necessary, as overallocation of memory will simply limit the number of simultaneous containers each node can run.
 
@@
 The \OnePhaseone{} dataset only requires 1.5GB memory per container, as we demonstrated previously.
-%Increasing the map container size to 4GB results in the job taking an additional 5 minutes (see Table \ref{Tab:03}).
 Increasing the map container size to 4GB results in the job taking an additional 5 minutes (see Table~\ref{Tab:03}), because fewer tasks can run in parallel. 
-%This larger container size is still too small for the \FullPhasethree{} dataset, which we demonstrated to require 6GB memory per container. 
-%Running a job on this dataset with 4GB memory per container results in the job failing.
-Attempting to run the larger data set (\FullPhasethree{}) with either 1.5GB or 4GB will result in the job failing, because the feature vectors cannot be loaded completely, this requires at least 6GB memory per container and 21h runtime.
+Attempting to run the larger data set (\FullPhasethree{}) with either 1.5GB or 4GB will result in the job failing, because the feature vectors do not fit into the available memory. 
+At least 6GB memory per container and 21h runtime are required to finish the job.
 
 One of the main selling points of Hadoop, is that it will run on commodity hardware. 
-%For this reason we used a relatively small cluster for our benchmarks. Although our cluster uses VMs on Microsoft Azure, it would be relatively simple for a lab to
-%create a similar Hadoop setup in-house. 
-To mimic the available resource in a standard resource setting we used only a small VM cluster on Microsoft Azure for the benchmarking in this paper so far.
-However, to demonstrate the scaling ability of Hadoop and our MapReduce algorithms, we also ran the clustering jobs on a much larger development Hadoop cluster. 
-%As presented in Table \ref{Tab:03}, the \OnePhaseone{} job saw an improvement of approximately 70\%.
-%The \FullPhasethree{} example, on the other hand, was 78\% faster.
+To mimic the available resource in a standard resource setting we used only a small Virtual Machine cluster on Microsoft Azure for the benchmarking in this paper so far.
+However, to demonstrate the scalability of Hadoop and our MapReduce algorithms, we also ran the clustering jobs on a much larger development Hadoop cluster. 
 As presented in Table~\ref{Tab:03}, the \OnePhaseone{} job saw an improvement of approximately 70\% when providing 448 cores of 13GB. 
@@
 
-\subsection*{Result visualisation}
-% DB: in the conclusion we discussed that we provide a "toolset to interpret Mahout's clustering output". 
-% DB: can we discuss here how we get a visual representation from the clustering (ClusterDumper+Gephi) and what the image tells us..
-GraphML files generated using Mahout Cluster Dumper can be visualised using popular freely-available tools such as Gephi~\cite{ICWSM09154}.
-An example visualisation using the phase 3 dataset, with nodes coloured by super population, is shown in Figure 1.
+\subsection*{Cluster visualisation}
+GraphML files generated using Mahout Cluster Dumper can be visualised using popular freely-available tools such as Gephi~\citep{ICWSM09154}.
+An example visualisation using the \FullPhasethree{} dataset, with nodes coloured by super population, is shown in Figure 1.
 
 The clusters generated by k-means clustering and their population structure are readily apparent. 
-Three clusters are homogeneous, containing only EAS (1 cluster) or SAS individuals (2 clusters, the second small and unconnected).
-Two clusters contain predominantly one super population plus a minor component of another: AMR + AFR or AFR + AMR.
-The sixth cluster is predominantly a mix of EUR + AMR, with 3 AFR individuals and 1 SAS.
+Three clusters are homogeneous, containing only East Asian (EAS, pink cluster) or South Asian (SAS) individuals (2 clusters, cyan cluster and the second small and unconnected).
+Two clusters contain predominantly one super population plus a minor component of another: Ad Mixed American (AMR) + African (AFR) or AFR + AMR. 
+The sixth cluster is predominantly a mix of European (EUR) + AMR, with 3 AFR individuals and 1 SAS.
 
@@
 In this paper we have applied Mahout's k-means clustering to the task of clustering individuals, based on their genomic variants. 
-We provide an interface from VCF format to Hadoop and Mahout, and a toolset to interpret Mahout's clustering output (see Figure 2)
-We find that on small datasets the clustering stage is two orders of magnitudes slower in Mahout compared to R (9 minutes vs 7 seconds), due to the overhead of setting up and co-ordinating the communication between the Hadoop nodes.
-However, increasing either the number of samples or the number of features per sample quickly exceeds the capability of R on this machine due to memory limitations.
-%DB ok effortlessly might be overstated ?
+We provide an interface from VCF format to Hadoop and Mahout, and a toolset to interpret Mahout's clustering output (see Figure 2).
+Disregarding the pre-processing time, we find that on small datasets the clustering stage is two orders of magnitudes slower in Mahout compared to R (9 minutes vs 7 seconds), due to the overhead of setting up and co-ordinating the communication between the Hadoop nodes.
+However, increasing either the number of samples or the number of features per sample quickly exceeds the capability of R on the used machine due to memory limitations.
 Hadoop on the other hand scales effortlessly to more than twice the number samples (2,504).
-It can also process VCF files that contain over 79 Million variants per samples, albeit with higher memory allocation.
-The MapReduce functions can scale from a relatively small and accessible cluster with 16 cores, to a much larger cluster with close to 500 CPU cores. 
+It can also process VCF files that contain over 81 Million variants per samples, albeit with higher but still reasonable memory allocation (6G).
+The MapReduce functions can scale from a relatively small and accessible cluster with 16 cores, to a much larger cluster with close to 500 CPU cores, although designing the container size for optimal resource and runtime consumption remains an expert task.
 In conclusion, Mahout offers a scalable out-of-the box solution that is directly applicable to genotype-based clustering for population or family association.
-
-%This efficient a high performance in-memory map-reduce system has already demonstrated to deliver a  50x speed-up on a whole-genome dataset compared to usual multithreaded approaches, see ADAM~\cite{Massie2013}
 
@@
 \subsection*{Hardware and Software}
-%DB what about Piotr's cluster I thought you used that one for the full genome data of phase 1 and phase 3 ?
 We processed the data using a cluster of five VMs on the Microsoft Azure cloud service.
@@
 \subsection*{Clustering quality}
-We scored the clusters with the \ARI{}~\cite{Hubert1985}. 
+We scored the clusters with the \ARI{}~\citep{Hubert1985}. 
 This index compares two different cluster sets, here the prediction $X = \{ X_1, X_2, \ldots , X_r \}$ and the annotated data $Y = \{ Y_1, Y_2, \ldots , Y_s \}$, and assigns a value based the overlap between $X$ and $Y$ as captured in a contingency table $\left[n_{ij}\right]$ where each entry $n_{ij}$ denotes the number of objects in common between $X_i$ and $Y_j : n_{ij}=|X_i \cap Y_j|$.
@@
 \begin{table*}[!t]
-\processtable{Resources allocated and time taken.\label{Tab:02}}
+\processtable{Resource allocation and time consumption.\label{Tab:02}}
 {\begin{tabular}{ll|cc|cc|rr|c}\toprule
@@
 \begin{table*}[!t]
-\processtable{Resources allocated and time taken.\label{Tab:03}}
+\processtable{Resource allocation and time consumption.\label{Tab:03}}
 {\begin{tabular}{c|cc|cc|c}\toprule
@@
 \section*{Acknowledgement}
-A.R.O was funded by the NSW Cancer Institute Big Data Big Impact schema, F.A.B by the National Health and Medical Research Council [1051757] and D.C.B by Commonwealth Scientific and Industrial Research Organisation's Transformational Capability Platform, Science and Industry Endowment Fund and Information Management and Technology Services.
+A.R.O was funded by the NSW Cancer Institute Big Data Big Impact schema, F.A.B by the National Health and Medical Research Council [1051757] and both D.C.B and N.S by Commonwealth Scientific and Industrial Research Organisation's Transformational Capability Platform, Science and Industry Endowment Fund and Information Management and Technology Services.
+The computation on Azure was funded by Microsoft Azure Research Award.
 The authors would like to thank Piotr Szul and Gareth Williams for their help with setting up Hadoop on the HPC system.

### 2015-03-11 doc/publications/journal2014/paper.tex

@@
 While the overall processing time is 40 minutes faster in Mahout compared to R, the real bottleneck for large datasets is the amount of memory available on a single machine. 
-As an example, the entire set of 1000 Genome Project phase3 variants could require approximately 1.6TB to be completely loaded into memory as a double-precision matrix (see Table~\ref{Tab:01}).
+As an example, the entire set of 1000 Genomes Project phase3 variants could require approximately 1.6TB to be completely loaded into memory as a double-precision matrix (see Table~\ref{Tab:01}).
 This is based on the equation:
@@
 We scored the clusters with the \ARI{}~\citep{Hubert1985}. 
-This index compares two different cluster sets, here the prediction $X = \{ X_1, X_2, \ldots , X_r \}$ and the annotated data $Y = \{ Y_1, Y_2, \ldots , Y_s \}$, and assigns a value based the overlap between $X$ and $Y$ as captured in a contingency table $\left[n_{ij}\right]$ where each entry $n_{ij}$ denotes the number of objects in common between $X_i$ and $Y_j : n_{ij}=|X_i \cap Y_j|$.
+This index compares two different cluster sets, here the prediction $X = \{ X_1, X_2, \ldots , X_r \}$ and the annotated data $Y = \{ Y_1, Y_2, \ldots , Y_s \}$, and assigns a value based on the overlap between $X$ and $Y$ as captured in a contingency table $\left[n_{ij}\right]$ where each entry $n_{ij}$ denotes the number of objects in common between $X_i$ and $Y_j : n_{ij}=|X_i \cap Y_j|$.
 The \ARI{} is then 

### 2015-04-02 doc/publications/journal2014/paper.tex
coverletter
Affiliation added

@@
 \title[MapReduce Clustering]{Scalable Clustering of Genotype Information using MapReduce}
-\author[O'Brien \textit{et~al}]{Aidan R. O'Brien\,$^{1}$, Neil Saunders\,$^1$,Fabian A. Buske\,$^{2}$ and Denis C. Bauer\,$^1$\footnote{to whom correspondence should be addressed}}
+\author[O'Brien \textit{et~al}]{Aidan R. O'Brien\,$^{1,2}$, Neil Saunders\,$^1$,Fabian A. Buske\,$^{3,4}$ and Denis C. Bauer\,$^1$\footnote{to whom correspondence should be addressed}}
 \address{$^{1}$CSIRO, Digital Productivity Flagship, 11 Julius Av, 2113, Sydney, Australia\\
-$^{2}$Cancer Epigenetics Program, Cancer Research Division, Kinghorn Cancer Centre, Garvan Institute of Medical Research, 384 Victoria St, 2010, Sydney, Australia}
+$^{2}$School of Biomedical Sciences and Pharmacy, Faculty of Health, 2308 Newcastle, Australia\\
+$^{3}$Cancer Epigenetics Program, Cancer Research Division, Kinghorn Cancer Centre, Garvan Institute of Medical Research, 384 Victoria St, 2010, Sydney, Australia\\
+$^{4}$UNSW Medicine, University of New South Wales, 2052 Sydney, Australia.}
 
@@
 However, advancements in modern hardware accelerators and data processing provides the means for overcoming these limitations. 
-%Here we introduce the interface we developed between the standard variant format (VCF) and the scalable analysis library Mahout, which opens up the application of advanced, efferent machine learning algorithms to genomic data.
 Here we introduce an interface between the standard variant format (VCF) and the scalable analysis library Mahout, which opens up the application of advanced, efficient machine learning algorithms to genomic data.
@@
 \section{Availability:}
-%DB this probably is not the final URL...
-The package is written in Java and available at \url{https://stash.csiro.au/users/bau04c/repos/bioinfbiglearning/browse}.
+The package is written in Java and available at \url{https://github.com/BauerLab/GeMaIn}.
 
@@
 
-%Mention Goolge's population clustering in the intro
 At the same time, computer scientists extent MapReduce paradigms to cope with the iterative nature of machine learning tasks~\citep{Chu2009}.

### 2015-04-08 doc/publications/journal2014/paper.tex
updated format

@@
 \usepackage[hyphens]{url}
+\usepackage[square,numbers]{natbib}
 
@@
 Processing genomic information from whole genome sequence studies poses computational challenges due to the unprecedented data volume, which render transitional approaches insufficient. 
-However, advancements in modern hardware accelerators and data processing provides the means for overcoming these limitations. 
-Here we introduce an interface between the standard variant format (VCF) and the scalable analysis library Mahout, which opens up the application of advanced, efficient machine learning algorithms to genomic data.
+However, advancements in modern hardware accelerators and data processing provides the means for overcoming these limitations, with MapReduce on Hadoop-hardware the most popular approach.  
+Here we compare implementation and resource consumption when clustering genome-wide variants using R as well as the MapReduce-based library, Mahout.
+To do this, we developed an interface between Mahout and the standard variant format (VCF), which opens up the use of advanced, efficient machine learning algorithms to genomic data.
 
@@
 \end{abstract}
+
+\section{Author CV}
+\paragraph{Aidan R O'Brien, BSc(Hons), is a joined researcher at Newcastle University and CSIRO and interested in BigData approaches for pan-cancer research.}
+
+\paragraph{Neil Saunders, PhD, is a research scientist at CSIRO interested in systems biology approaches for human health.}
+
+\paragraph{Fabian A. Buske, PhD, is a computer science expert focusing on no-SQL database systems.}
+
+\paragraph{Denis C Bauer, PhD, is team leader of the transformational bioinformatics team at CSIRO active in the personalised genomics space.}
 
@@
 
+\section{Key Points}
+\begin{itemize}
+\item May bioinformatics tasks can be converted into a MapReduce-paradigm making Hadoop-based approaches attractive.
+\item Scaling seamlessly to larger data sets has been celebrated as Hadoop's superior functionality over conventional parallelisation methodologies, however this only applies on resource-uncapped machines.
+\item On machines in resource settings, we demonstrate that achieving optimal performance still requires careful resource adjustment taking machine architecture and dataset features into account. 
+\item Unlike R, Hadoop-based approaches can handle clustering of genome wide sequencing variants of cohorts with 2500 individuals
+\end{itemize}
+
 \section*{Acknowledgement}
@@
 
-\bibliographystyle{natbib}
+%\bibliographystyle{natbib}
 %\bibliographystyle{achemnat}
 %\bibliographystyle{plainnat}
-%\bibliographystyle{abbrv}
+\bibliographystyle{abbrv}
 %\bibliographystyle{bioinformatics}

### 2015-04-08 doc/publications/journal2014/paper.tex
fix bio

@@
 \section{Author CV}
-\paragraph{Aidan R O'Brien, BSc(Hons), is a joined researcher at Newcastle University and CSIRO and interested in BigData approaches for pan-cancer research.}
+\paragraph{Aidan R O'Brien, BBiot(Hons), is a joined researcher at Newcastle University and CSIRO and interested in BigData approaches for pan-cancer research.}
 

### 2015-04-09 doc/publications/journal2014/paper.tex
format changes

@@
 Processing genomic information from whole genome sequence studies poses computational challenges due to the unprecedented data volume, which render transitional approaches insufficient. 
-However, advancements in modern hardware accelerators and data processing provides the means for overcoming these limitations, with MapReduce on Hadoop-hardware the most popular approach.  
-Here we compare implementation and resource consumption when clustering genome-wide variants using R as well as the MapReduce-based library, Mahout.
+However, advancements in modern hardware accelerators and data processing provides the means for overcoming these limitations, with Hadoop the most popular approach.  
+Here we compare implementation and resource consumption when clustering genome-wide variants using R as well as the Hadoop-based library, Mahout.
 To do this, we developed an interface between Mahout and the standard variant format (VCF), which opens up the use of advanced, efficient machine learning algorithms to genomic data.
-
 \section{Results:}
@@
 \begin{itemize}
-\item May bioinformatics tasks can be converted into a MapReduce-paradigm making Hadoop-based approaches attractive.
+\item Many bioinformatics tasks can be converted into the MapReduce-paradigm making Hadoop-based approaches attractive.
 \item Scaling seamlessly to larger data sets has been celebrated as Hadoop's superior functionality over conventional parallelisation methodologies, however this only applies on resource-uncapped machines.
-\item On machines in resource settings, we demonstrate that achieving optimal performance still requires careful resource adjustment taking machine architecture and dataset features into account. 
-\item Unlike R, Hadoop-based approaches can handle clustering of genome wide sequencing variants of cohorts with 2500 individuals
+\item On machines available in research settings, we demonstrate that achieving optimal performance still requires careful resource adjustment taking machine architecture and dataset features into account.
+\item Unlike R, Hadoop-based approaches make clustering of genome wide sequencing variants from cohorts with 2500 individuals feasible on machines available in research settings.
 \end{itemize}

### 2015-07-27 doc/publications/journal2014/paper_old.tex
structure for GB

--- 2015-04-09 doc/publications/journal2014/paper.tex
@@
 \section{Motivation:}
-Processing genomic information from whole genome sequence studies poses computational challenges due to the unprecedented data volume, which render transitional approaches insufficient. 
-However, advancements in modern hardware accelerators and data processing provides the means for overcoming these limitations, with Hadoop the most popular approach.  
-Here we compare implementation and resource consumption when clustering genome-wide variants using R as well as the Hadoop-based library, Mahout.
-To do this, we developed an interface between Mahout and the standard variant format (VCF), which opens up the use of advanced, efficient machine learning algorithms to genomic data.
+Genomic information is increasingly being used for medical research, giving rise to the need for efficient analysis methodology able to cope with thousands of individuals and millions of variants.  
+The widely used Hadoop MapReduce architecture and associated machine learning library, Mahout, provide the means for tackling computationally challenging tasks. 
+However, many genomic analyses do not fit the Map-Reduce paradigm. 
+We therefore survey the recently developed Spark engine, which is less rigid in the parallelisation architecture and hence more suitable for bioinformatics tasks, along with its associated machine learning library, MLlib.
+To do this, we developed an interface from Mahout and MLlib to the standard variant format (VCF), which opens up the usage of advanced, efficient machine learning algorithms to genomic data. 
 \section{Results:}
-We achieve more than 2-fold speedup by using the scalable k-means MapReduce implementation over the equivalent analysis performed in R, by comparable accuracy.
-However, the real advantage lies in the scalability of MapReduce beyond R's capability to a genome-wide analysis.
 We successfully clustered more than 2,500 individuals each having more than 81 Million variants.
+Both Hadoop and Spark have superior performance over traditional approaches.
+We observe a 50 fold speedup when using the efficient in-memory Spark compute engine compared to Hadoop.
+Furthermore, our implementation achieves a better performance compared to a previously published Spark-based genome clustering approach, ADAM. 
 \section{Availability:}

### 2015-07-28 doc/publications/journal2014/paper.tex
added GB styles

--- 2015-04-08 doc/publications/journal2014/paper.tex
@@
-\documentclass{bioinfo}
-\copyrightyear{2005}
-\pubyear{2005}
+%% BioMed_Central_Tex_Template_v1.06
+%%                                      %
+%  bmc_article.tex            ver: 1.06 %
+%                                       %
 
-\usepackage[hyphens]{url}
-\usepackage[square,numbers]{natbib}
+%%IMPORTANT: do not delete the first line of this template
+%%It must be present to enable the BMC Submission system to
+%%recognise this template!!
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%                                     %%
+%%  LaTeX template for BioMed Central  %%
+%%     journal article submissions     %%
+%%                                     %%
+%%          <8 June 2012>              %%
+%%                                     %%
+%%                                     %%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%                                                                 %%
+%% For instructions on how to fill out this Tex template           %%
+%% document please refer to Readme.html and the instructions for   %%
+%% authors page on the biomed central website                      %%
+%% http://www.biomedcentral.com/info/authors/                      %%
+%%                                                                 %%
+%% Please do not use \input{...} to include other tex files.       %%
+%% Submit your LaTeX manuscript as one .tex document.              %%
+%%                                                                 %%
+%% All additional figures and files should be attached             %%
+%% separately and not embedded in the \TeX\ document itself.       %%
+%%                                                                 %%
+%% BioMed Central currently use the MikTex distribution of         %%
+%% TeX for Windows) of TeX and LaTeX.  This is available from      %%
+%% http://www.miktex.org                                           %%
+%%                                                                 %%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%%% additional documentclass options:
+%  [doublespacing]
+%  [linenumbers]   - put the line numbers on margins
+
+%%% loading packages, author definitions
+
+%\documentclass[twocolumn]{bmcart}% uncomment this for twocolumn layout and comment line below
+\documentclass{bmcart}
+
+%%% Load packages
+%\usepackage{amsthm,amsmath}
+%\RequirePackage{natbib}
+%\RequirePackage[authoryear]{natbib}% uncomment this for author-year bibliography
+%\RequirePackage{hyperref}
+\usepackage[utf8]{inputenc} %unicode support
+%\usepackage[applemac]{inputenc} %applemac support if unicode package fails
+%\usepackage[latin1]{inputenc} %UNIX support if unicode package fails
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%                                             %%
+%%  If you wish to display your graphics for   %%
+%%  your own use using includegraphic or       %%
+%%  includegraphics, then comment out the      %%
+%%  following two lines of code.               %%
+%%  NB: These line *must* be included when     %%
+%%  submitting to BMC.                         %%
+%%  All figure files must be submitted as      %%
+%%  separate graphics through the BMC          %%
+%%  submission process, not included in the    %%
+%%  submitted article.                         %%
+%%                                             %%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\def\includegraphic{}
+\def\includegraphics{}
+
+%%% Put your definitions there:
+\startlocaldefs
+\endlocaldefs
+
+%%% Begin ...
 \begin{document}
-\firstpage{1}
 
-\newcommand{\OnePhaseone}{phase1\_chr1}
-\newcommand{\NinteenPhaseone}{phase1\_chr19}
-\newcommand{\SevenPhaseone}{phase1\_chr1-7}
-\newcommand{\FullPhaseone}{phase1\_chr1-22}
-\newcommand{\OnePhasethree}{phase3\_chr1}
-\newcommand{\ThreePhasethree}{phase3\_chr1-3}
-\newcommand{\FullPhasethree}{phase3\_chr1-22}
-\newcommand{\ARI}{Adjusted Rand Index}
+%%% Start of article front matter
+\begin{frontmatter}
 
-\title[MapReduce Clustering]{Scalable Clustering of Genotype Information using MapReduce}
-\author[O'Brien \textit{et~al}]{Aidan R. O'Brien\,$^{1,2}$, Neil Saunders\,$^1$,Fabian A. Buske\,$^{3,4}$ and Denis C. Bauer\,$^1$\footnote{to whom correspondence should be addressed}}
-\address{$^{1}$CSIRO, Digital Productivity Flagship, 11 Julius Av, 2113, Sydney, Australia\\
-$^{2}$School of Biomedical Sciences and Pharmacy, Faculty of Health, 2308 Newcastle, Australia\\
-$^{3}$Cancer Epigenetics Program, Cancer Research Division, Kinghorn Cancer Centre, Garvan Institute of Medical Research, 384 Victoria St, 2010, Sydney, Australia\\
-$^{4}$UNSW Medicine, University of New South Wales, 2052 Sydney, Australia.}
+\begin{fmbox}
+\dochead{Research}
 
-\history{Received on XXXXX; revised on XXXXX; accepted on XXXXX}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%                                          %%
+%% Enter the title of your article here     %%
+%%                                          %%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-\editor{Associate Editor: XXXXXXX}
+\title{VariantSpark: Population Scale Clustering of Genotype Information}
 
-\maketitle
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%                                          %%
+%% Enter the authors here                   %%
+%%                                          %%
+%% Specify information, if available,       %%
+%% in the form:                             %%
+%%   <key>={<id1>,<id2>}                    %%
+%%   <key>=                                 %%
+%% Comment or delete the keys which are     %%
+%% not used. Repeat \author command as much %%
+%% as required.                             %%
+%%                                          %%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-\begin{abstract}
+\author[
+   addressref={aff1,aff2},                   % id's of addresses, e.g. {aff1,aff2}
+%   corref={aff1},                       % id of corresponding address, if any
+%   noteref={n1},                        % id's of article notes, if any
+%   email={jane.e.doe@cambridge.co.uk}   % email address
+]{\inits{}\fnm{Aidan R.} \snm{O'Brien}}
+\author[
+   addressref={aff1},
+]{\inits{}\fnm{Neil} \snm{Saunders}}
+\author[
+   addressref={aff3,aff4},
+%   email={john.RS.Smith@cambridge.co.uk}
+]{\inits{}\fnm{Fabian A.} \snm{Buske}}
+\author[
+   addressref={aff1},
+   email={Denis.Bauer@CSIRO.au}
+%   email={john.RS.Smith@cambridge.co.uk}
+]{\inits{}\fnm{Denis C.} \snm{Bauer}}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%                                          %%
+%% Enter the authors' addresses here        %%
+%%                                          %%
+%% Repeat \address commands as much as      %%
+%% required.                                %%
+%%                                          %%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-\section{Motivation:}
-Processing genomic information from whole genome sequence studies poses computational challenges due to the unprecedented data volume, which render transitional approaches insufficient. 
-However, advancements in modern hardware accelerators and data processing provides the means for overcoming these limitations, with MapReduce on Hadoop-hardware the most popular approach.  
-Here we compare implementation and resource consumption when clustering genome-wide variants using R as well as the MapReduce-based library, Mahout.
-To do this, we developed an interface between Mahout and the standard variant format (VCF), which opens up the use of advanced, efficient machine learning algorithms to genomic data.
+\address[id=aff1]{%                           % unique id
+  \orgname{CSIRO, Health \& Biosecurity Flagship}, % university, etc
+  \street{11 Julius Av},                     %
+  \postcode{2113},                                % post or zip code
+  \city{Sydney},                              % city
+  \cny{Australia}                                    % country
+}
+\address[id=aff2]{%
+  \orgname{School of Biomedical Sciences and Pharmacy, Faculty of Health},
+  \postcode{2308},
+  \city{Newcastle},
+  \cny{Australia}
+}
+\address[id=aff3]{%
+  \orgname{Cancer Epigenetics Program, Cancer Research Division, Kinghorn Cancer Centre, Garvan Institute of Medical Research},
+  \street{384 Victoria St},
+  \postcode{2010},
+  \city{Sydney},
+  \cny{Australia}
+}
+\address[id=aff4]{%
+  \orgname{UNSW Medicine, University of New South Wales},
+%  \street{384 Victoria St},
+  \postcode{2052},
+  \city{Sydney},
+  \cny{Australia}
+}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%                                          %%
+%% Enter short notes here                   %%
+%%                                          %%
+%% Short notes will be after addresses      %%
+%% on first page.                           %%
+%%                                          %%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-\section{Results:}
-We achieve more than 2-fold speedup by using the scalable k-means MapReduce implementation over the equivalent analysis performed in R, by comparable accuracy.
-However, the real advantage lies in the scalability of MapReduce beyond R's capability to a genome-wide analysis.
-We successfully clustered more than 2,500 individuals each having more than 81 Million variants.
-\section{Availability:}
-The package is written in Java and available at \url{https://github.com/BauerLab/GeMaIn}.
+\begin{artnotes}
+%\note{Sample of title note}     % note to the article
+\note[id=n1]{Equal contributor} % note, connected to author
+\end{artnotes}
 
-\section{Contact:} \href{Denis.Bauer@CSIRO.au}{Denis.Bauer@CSIRO.au}
+\end{fmbox}% comment this for two column layout
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%                                          %%
+%% The Abstract begins here                 %%
+%%                                          %%
+%% Please refer to the Instructions for     %%
+%% authors on http://www.biomedcentral.com  %%
+%% and include the section headings         %%
+%% accordingly for your article type.       %%
+%%                                          %%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{abstractbox}
+
+\begin{abstract} % abstract
+\parttitle{Motivation}: Genomic information is increasingly used in medical research, giving rise to the need for efficient analysis methodology able to cope with thousands of individuals and millions of variants.
+The widely used Hadoop MapReduce architecture and associated machine-learning library, Mahout, provide the means for tackling computationally challenging tasks.
+However, many genomic analyses do not fit the Map-Reduce paradigm.
+We therefore survey the recently developed Spark engine, along with its associated machine learning library, MLlib, which are more flexible in the parallelisation architecture and hence more suitable for population-scale bioinformatics tasks.
+To do this, we developed an interface from Mahout and MLlib to the standard variant format (VCF), which opens up the usage of advanced, efficient machine learning algorithms to genomic data.
+
+\parttitle{Results}: We successfully clustered more than 2,500 individuals each having more than 81 Million variants.
+Both Hadoop and Spark have superior performance over traditional approaches.
+We observe a 50 fold speedup when using the efficient in-memory Spark compute engine compared to Hadoop.
+Furthermore, our implementation achieves a better performance compared to a previously published Spark-based genome clustering approach, ADAM.
+
+\parttitle{Availability}: The package is written in Java and available at https: //github.com/BauerLab/GeMaIn.
 \end{abstract}
 
-\section{Author CV}
-\paragraph{Aidan R O'Brien, BSc(Hons), is a joined researcher at Newcastle University and CSIRO and interested in BigData approaches for pan-cancer research.}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%                                          %%
+%% The keywords begin here                  %%
+%%                                          %%
+%% Put each keyword in separate \kwd{}.     %%
+%%                                          %%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-\paragraph{Neil Saunders, PhD, is a research scientist at CSIRO interested in systems biology approaches for human health.}
+\begin{keyword}
+\kwd{sample}
+\kwd{article}
+\kwd{author}
+\end{keyword}
 
-\paragraph{Fabian A. Buske, PhD, is a computer science expert focusing on no-SQL database systems.}
+% MSC classifications codes, if any
+%\begin{keyword}[class=AMS]
+%\kwd[Primary ]{}
+%\kwd{}
+%\kwd[; secondary ]{}
+%\end{keyword}
 
-\paragraph{Denis C Bauer, PhD, is team leader of the transformational bioinformatics team at CSIRO active in the personalised genomics space.}
+\end{abstractbox}
+%
+%\end{fmbox}% uncomment this for twcolumn layout
 
-\section{Introduction}
+\end{frontmatter}
 
-Grouping individuals based on their genomic profile is a commonly performed tasks to identify population structure~\citep{Gao2007} or elucidate different haplotype involvement in diseases susceptibility~\citep{Laitman2013}.
-Traditionally both, the number of individuals and included genotypes (typically from SNP arrays), were relatively small and libraries in Bioconductor, which commonly perform all calculations in memory, sufficient.
-However, recent technological advances in whole genome sequencing have made population-scale sequencing feasible. 
-It is hence economical to generate studies with sample sizes previously reserved for larger consortia such as the 1000 genomes project~\citep{1KG2012} or The Cancer Genome Atlas, TCGA~\citep{TCGA2013}.
-At the same time, whole genome sequencing enables the inclusion of rare or even somatic mutations in the analysis, increasing the feature space by orders of magnitude.
-This drastic increase in both sample numbers and features per sample requires a massively parallel approach to data processing.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%                                          %%
+%% The Main Body begins here                %%
+%%                                          %%
+%% Please refer to the instructions for     %%
+%% authors on:                              %%
+%% http://www.biomedcentral.com/info/authors%%
+%% and include the section headings         %%
+%% accordingly for your article type.       %%
+%%                                          %%
+%% See the Results and Discussion section   %%
+%% for details on how to create sub-sections%%
+%%                                          %%
+%% use \cite{...} to cite references        %%
+%%  \cite{koon} and                         %%
+%%  \cite{oreg,khar,zvai,xjon,schn,pond}    %%
+%%  \nocite{smith,marg,hunn,advi,koha,mouse}%%
+%%                                          %%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-As a result of these big data challenges, MapReduce approaches are increasingly being used in bioinformatics (for reviews see~\citep{Zou2013, Qiu2010,Taylor2010}).
-This is especially the case for sequence analysis tasks, such as read mapping~\citep{Schatz2009}, duplicate removal~\citep{Jourdren2012}, and variant calling~\citep{Langmead2009, McKenna2010} as well as Genome Wide Analysis Study based tasks~\citep{Huang2013, Guo2014}.
+%%%%%%%%%%%%%%%%%%%%%%%%% start of article main body
+% <put your article body there>
 
-At the same time, computer scientists extent MapReduce paradigms to cope with the iterative nature of machine learning tasks~\citep{Chu2009}.
-Specifically, the Mahout project \url{https://mahout.apache.org/} has been developed extensively~\citep{Ranger2007, Owen2011} and was successfully applied in the clinical informatics space~\citep{Dong2013}.
+%%%%%%%%%%%%%%%%
+%% Background %%
+%%
+\section*{Background}
 
-In this paper, we link the two areas of parallel machine learning and bioinformatics by providing an interface between Mahout and the standard variant data format, variant call format (VCF)~\citep{1KG2012}, which opens up the application of Mahout's different machine learning algorithms to be applied to genotype-based tasks.
-To demonstrate the capability, we cluster variant datasets from the 1000 genomes project to determine population structure using the k-means clustering algorithm available in Mahout.
-In the first section we benchmark the performance and accuracy of Mahout's implementations against a standard R based implementation on a subset of the data and in following two sections we investigate Mahout's scaleability with respect to sample size and features per sample.
-We then discuss design considerations for the resource allocation and give a visualisation example in the last two sections.
+\section*{Results}
+\subsection*{Spark enables faster clustering of individuals compared to traditional methods}
+Add the results for the comparison of python, R, spark and hadoop here \ldots
 
-\section*{Results and discussion}
+\subsection*{VariantSpark allows genome wide sampling of variants to improve clustering quality}
+Add the results for the scalling from 1\% to X\% \ldots
 
-% COMPARISON TO R
-\subsection*{Clustering in Mahout requires less memory than in R}
-In this section, we compare the runtime and resource requirements between the k-means clustering implementation in R and Mahout for the small \NinteenPhaseone{} dataset (see Methods, Table~\ref{Tab:01}).
-We measured the runtime separately for pre-processing stage as well as the clustering stage.
+\subsection*{Graph visualization}
+Add the image here \ldots
 
-With Hadoop, pre-processing the data takes approximately 6 minutes, with an allocation of up to 3GB memory. 
-The resulting data is stored as a Hadoop 'SequenceFile' of about 835MB in size.
-Conversely,  pre-processing the data to R's in-memory matrix takes 56 minutes and required 17.7GB memory. 
-K-means clustering the data, however, is faster with R, using 7s, whereas Mahout takes just under 9 minutes (see Table~\ref{Tab:02}). 
-This is due to the additional overhead involved in spawning the map and reduce processes, and distributing the data between nodes.
-R, on the other hand, already has the entire matrix in-memory, on a single machine and since no disk-access or networking is required the clustering can be performed faster. 
-Due to the relatively small number of variants, cluster accuracy is relatively poor for both methods, with the best \ARI{}~\citep{Hubert1985} score for clustering individuals in one of 4 populations (see Methods) being 0.705 in Mahout and 0.614 in R (see Table~\ref{Tab:02}, an \ARI{} of 1 means perfect clustering).
+\section*{Methods}
+\subsection*{Implementation}
 
-While the overall processing time is 40 minutes faster in Mahout compared to R, the real bottleneck for large datasets is the amount of memory available on a single machine. 
-As an example, the entire set of 1000 Genomes Project phase3 variants could require approximately 1.6TB to be completely loaded into memory as a double-precision matrix (see Table~\ref{Tab:01}).
-This is based on the equation:
-$\frac{n \times m \times 8\text{bytes}}{1\text{GB}}$, where $n \times m$ are the matrix dimensions, 8bytes is the memory a \texttt{double} occupies (i.e.
-each matrix element), and  1GB is $10^9$ bytes.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%                                          %%
+%% Backmatter begins here                   %%
+%%                                          %%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-Therefore, the data processed in the subsequent sections are too large to be processed with standard R, hence we focus on evaluating different settings of the Mahout implementation only.
+\begin{backmatter}
 
-% (1) SCALING UP FEATURES IN MAHOUT
-\subsection*{Hadoop scales to genome-wide clustering}
-In this section we demonstrate Mahout's ability to cluster genome-wide variants. 
-To demonstrate accuracy-gain and resource-consumption we evaluate whole-genome data (\FullPhasethree{}), which contains 79 million variants, as well as two subsets with 19 (\ThreePhasethree{}) and 6 million (\OnePhasethree{}) variants, respectively  (see Table~\ref{Tab:01}).
-As expected, providing variants from only chromosome one (\OnePhasethree{}) holds to little information and hence achieves a relatively low \ARI{} of 0.622 (see Table~\ref{Tab:02}).
-Note, phase3 data has one more population group, which explains the decrease in accuracy compared to the previous section.
+\section*{Competing interests}
+  The authors declare that they have no competing interests.
 
-Increasing the input to include the first three chromosomes (\ThreePhasethree{}) improves the accuracy with an \ARI{} of 0.825 (Table~\ref{Tab:02}).
-We are reaching a saturation point for k-means clustering here, as including all 22 autosomes (\FullPhasethree{}) improves the \ARI{} only marginally to 0.84 (Table~\ref{Tab:02}).
-However, this demonstrates that performing clustering using whole-genome variants on research infrastructure using Mahout is feasible as the resource consumption lies within realistic boundaries. 
-The time consumption increases to 25 hours and memory to 6 GB (Table~\ref{Tab:02}).
+\section*{Author's contributions}
+    Text for this section \ldots
 
-% (2) SCALING UP SAMPLES IN MAHOUT
-\subsection*{Hadoop scales effortlessly to more than twice the number of samples}
-Hadoop's main advantage over other parallelisation techniques is its ability to scale to more samples without altering the code. 
-Mahout's k-means algorithm is structured such that it splits the clustering problem into separate, smaller tasks, where each task calculates the distance between a subset of samples and the k-means centroids.
-The scalability is achieved by Hadoop creating 'containers' for each task and subsequent iterative processing until every task is completed.
-We discuss containers in greater detail in the next section, however, a container is essentially a pre-defined allocation of memory and CPU cores.
-By increasing the number of samples, the k-means algorithm simply divides the job into more tasks, which enables the memory requirement of each task to stay the same.
-This is because the dimensionality is constant, compared to the previous section where we increased the number of variants and observe a subsequent increase in memory.
-
-To demonstrate this, we compare k-means clustering of the 2504 individuals from the phase3 dataset compared to the 1092 individuals from the phase1 dataset.
-We specifically use the \SevenPhaseone{} and \ThreePhasethree{} datasets because of their similar feature size.
-The algorithm completed successfully with the same sized containers for both datasets, however, the \ThreePhasethree{} dataset required about twenty minutes more time than the smaller \SevenPhaseone{} dataset (Table~\ref{Tab:02}).
-This demonstrates Hadoop's strength of scaling with sample size.
-
-% (3) CLUSTER AND CONTAINER SIZE
-\subsection*{Cluster size and containers}
-In the previous section we observed that while the memory remained constant the runtime increases with sample size. 
-On a larger Hadoop cluster the runtime could also stay constant if the number of additional tasks can be accommodated by additional compute nodes. 
-However, as research is carried out on resource-finite infrastructure, we discuss in this section how to specify the optimal container size when designing MapReduce jobs to incur only a minimal increase in runtime.
-
-To elaborate, a container is essentially an abstraction of resources, i.e. memory and CPU cores. 
-After Hadoop divides a job into tasks, each task runs in parallel in a container. Also, each container is independent of other containers.
-The resources Hadoop allocates to each container is a predefined value, specified by the user in Hadoop's configuration. 
-According to the Hortonworks documentation~\citep{Horton2013}, the general rule of thumb is that one CPU core is allocated to each container, and the available memory per node should be divided equally between each of those containers.
-For example, on a node with 8 CPU cores and 32GB memory, each container should be allocated 1 CPU core and 4GB memory.
-Effectively allowing a maximum of eight 4GB containers to run in parallel on that node.
-
-When a MapReduce job is launched, Hadoop spawns a number of map tasks and reduce tasks. 
-These tasks are placed in a queue until a node has enough resources available to run a container, and generally, the more containers that can run simultaneously, the faster the job will complete.
-For smaller jobs, the maximum number of containers are generally limited by the available CPU cores because each container requires an entire CPU core.
-The tasks running in each container may not be able to utilise all of the memory allocated to that container.
-However, for jobs that deal with large datasets, the memory allocated per container (following Cloudera's guidelines) may not be enough.
-We could modify the above example to allocate each container 1 CPU core and 8GB memory, however now only four containers can fit within the node's 32GB memory, resulting in the other 4 CPU cores being idle (as only four containers can run simultaneously).
-It is therefore optimal to allocate the appropriate resources to containers, and only increase container memory when necessary, as overallocation of memory will simply limit the number of simultaneous containers each node can run.
-
-To demonstrate the importance of specifying these parameters correctly, we ran identical jobs, using different Hadoop configurations. 
-The \OnePhaseone{} dataset only requires 1.5GB memory per container, as we demonstrated previously.
-Increasing the map container size to 4GB results in the job taking an additional 5 minutes (see Table~\ref{Tab:03}), because fewer tasks can run in parallel. 
-Attempting to run the larger data set (\FullPhasethree{}) with either 1.5GB or 4GB will result in the job failing, because the feature vectors do not fit into the available memory. 
-At least 6GB memory per container and 21h runtime are required to finish the job.
-
-One of the main selling points of Hadoop, is that it will run on commodity hardware. 
-To mimic the available resource in a standard resource setting we used only a small Virtual Machine cluster on Microsoft Azure for the benchmarking in this paper so far.
-However, to demonstrate the scalability of Hadoop and our MapReduce algorithms, we also ran the clustering jobs on a much larger development Hadoop cluster. 
-As presented in Table~\ref{Tab:03}, the \OnePhaseone{} job saw an improvement of approximately 70\% when providing 448 cores of 13GB. 
-With these resources, the \FullPhasethree{} example, was 78\% faster.
-
-Although both jobs were more than twice as fast on the larger cluster, the speed doesn't scale relative to the size of the cluster.
-This is due to a limitation in these examples, where the number of reduce tasks in k-means clustering is limited to the number of k-means clusters (k).
-A task where k is five, for example, will only be able to spawn five reduce tasks, no matter how many resources are available.
-Although sometimes inevitable, this bottleneck of maximum simultaneous reduce tasks is an important consideration to take into account when designing MapReduce jobs.
-
-\begin{figure}[!tpb]%figure1
-\centerline{\includegraphics[type=pdf,ext=.pdf,read=.pdf, scale=0.40]{gephi-p3}}
-        \label{fig:sign}
-        \caption{{\bf Visualisation of k-means clusters derived from the phase 3 dataset.}
-        Nodes are coloured by super population: AFR (green), AMR (blue), EAS (pink), EUR (red) and SAS (cyan). 
-}
-\end{figure}
-
-\subsection*{Cluster visualisation}
-GraphML files generated using Mahout Cluster Dumper can be visualised using popular freely-available tools such as Gephi~\citep{ICWSM09154}.
-An example visualisation using the \FullPhasethree{} dataset, with nodes coloured by super population, is shown in Figure 1.
-
-The clusters generated by k-means clustering and their population structure are readily apparent. 
-Three clusters are homogeneous, containing only East Asian (EAS, pink cluster) or South Asian (SAS) individuals (2 clusters, cyan cluster and the second small and unconnected).
-Two clusters contain predominantly one super population plus a minor component of another: Ad Mixed American (AMR) + African (AFR) or AFR + AMR. 
-The sixth cluster is predominantly a mix of European (EUR) + AMR, with 3 AFR individuals and 1 SAS.
-
-\section*{Conclusion}
-In this paper we have applied Mahout's k-means clustering to the task of clustering individuals, based on their genomic variants. 
-We provide an interface from VCF format to Hadoop and Mahout, and a toolset to interpret Mahout's clustering output (see Figure 2).
-Disregarding the pre-processing time, we find that on small datasets the clustering stage is two orders of magnitudes slower in Mahout compared to R (9 minutes vs 7 seconds), due to the overhead of setting up and co-ordinating the communication between the Hadoop nodes.
-However, increasing either the number of samples or the number of features per sample quickly exceeds the capability of R on the used machine due to memory limitations.
-Hadoop on the other hand scales effortlessly to more than twice the number samples (2,504).
-It can also process VCF files that contain over 81 Million variants per samples, albeit with higher but still reasonable memory allocation (6G).
-The MapReduce functions can scale from a relatively small and accessible cluster with 16 cores, to a much larger cluster with close to 500 CPU cores, although designing the container size for optimal resource and runtime consumption remains an expert task.
-In conclusion, Mahout offers a scalable out-of-the box solution that is directly applicable to genotype-based clustering for population or family association.
-
-\begin{methods}
-\section{Methods}
-\subsection*{Datasets}
-We used variant call format (VCF) files from phases 1 and 3 of the 1000 Genomes Project.
-Each VCF file details the genetic variation (SNPs and indels) for one chromosome, between each individual and a reference genome derived from GRCh37. 
-The phase1 and phase3 datasets contain 1,092 and 2,504 individuals, respectively.
-Metadata available from the 1000 Genomes Consortium specifies additional attributes for each individual, including population and familial data.\\
-We used different subsets of the data for different comparisons. For example, to be able to compare Hadoop with R, we used the VCF file containing chromosome 19 variants (\NinteenPhaseone{}).
-To demonstrate scaling with Hadoop, we use subsets that contain more variants, i.e. \SevenPhaseone{} and \FullPhaseone{}, which include chromosome 1-7 and the full genome, respectively.
-Table~\ref{Tab:01} contains an overview of the datasets.
-
-\subsection*{Hardware and Software}
-We processed the data using a cluster of five VMs on the Microsoft Azure cloud service.
-Three of these VMs, the 'worker nodes', each have 14GB memory and and 8 CPU cores.
-These three VMs run the Hadoop2 NodeManager, which manages the Yet Another Resource Negotiator (YARN) 'containers' (see discussion).
-The user is able to specify the resources allocated to each container, where map containers can have a different allocation to reduce containers.
-The fourth VM runs the HDFS NameNode, ResourceManager and monitoring services, and the fifth VM, a small, single core machine, runs the Cloudera manager service.
-With a total of 42GB memory and 24 CPU cores available between the node managers, we can simultaneously run up to 24 YARN containers with 1.5GB memory per container, or 12 YARN containers with 3GB memory per container.
-Storage consists of a 3TB Hadoop Distributed File System (HDFS) partition distributed across the three worker nodes. Each of these nodes runs a DataNode with a 1TB disk.
-
-The larger cluster mainly differs in regards to hardware. It has a total of 448 CPU cores and 1.31TB memory available.
-
-We used a single VM for the comparison between Mahout and R.
-
-\subsection*{Pre-processing}
-\label{Sec:preprocessing}
-Mahout requires Hadoop's `SequenceFiles' as input to its machine learning algorithms. SequenceFiles are binary flat files consisting of key/value pairs. In this case, each of the keys is a
-label for an individual, where the associated value for each key is a `sparse vector' representing the individual's genomic variants. We use sparse vectors, rather than
-dense vectors, to significantly reduce required memory and disk space. Each sparse vector consists of two separate vectors. An item in the first vector is
-effectively a key for an item at the same position in the second vector.
-The key represents a genomic location, where its value represents the variant.
-Zero values (positions where there is no variant)
-are omitted. Therefore, the fewer variants in an individual's genome, the smaller the sparse vector size. In comparison, `dense vectors' are single arrays that
-include every position where a variant may fall, whether or not there actually is a variant. Therefore, the size of a dense vector isn't reduced with a large number of non-variants.
-
-The values of the sparse vectors, which, as previously mentioned, represent an individuals genomic variants, are a \texttt{double} java type.
-We set the value of each double to the sum of the square root of the Hamming distance between each variant allele and the reference genome.
-Therefore, a value identical to the reference, 0\textbar{}0, will be 0 (and omitted from the sparse vector),
-a heterozygous variant, 0\textbar{}1 or 1\textbar{}0, will be 1 (sqrt1) and a homozygous variant, 1\textbar{}1, will be 2 (sqrt1 + sqrt1).
-The square root of the each allele prevents variants such as 1\textbar{}1 colliding with the variant 2\textbar{}0.
-
-We wrote two `MapReduce' classes in Java to convert VCF files to the aforementioned sequence files. This is scalable and should work on any number of files,
-of any file size; limited only by disk space and time. 
-The `map' stage of the first job parses each line of the VCF file to key value pairs.
-Each of the keys are an identical arbitrary number, where the value for each key is a sparse vector of the variants from each line.
-The variants in each vector are in the same order as they appear in the VCF file.
-
-Additionally, in this stage, we can define parameters to exclude certain variants that may be deemed low quality, or only appear in a certain number of individuals.
-We use identical keys so that each key value pair is sent to the same reducer, and this reducer assigns a unique key to each value.
-We require one reducer in this step to make each the key for each variant a consecutive number, rather than an arbitrary feature, such as its genomic location.
-Because keys are represented by integers, this helps to avoid integer overflows by keeping the keys relatively small.
-
-The second job effectively transposes the output from the first job. The mappers create key value pairs where the key is a composite of the individual id and variant id, and the value is the variant.
-This composite key, which is essentially a concatenation of two keys, allowed us to implement custom 'collect' and 'sort' classes to speed up the process.
-These classes divide up the key value pairs by their individual ID and sort them by their location ID, so each sorted collection of key/value pairs can be sent to a separate reducer.
- The reducers then write the variants for each individual to a sparse vector and stores each vector as a value, where the key is the individual ID.
-
-\subsection*{Mahout K-means clustering}
-To cluster the samples from the sequence files, we use Mahout 0.9 with Hadoop 2.5.0. 
-The clustering steps are the same for each dataset, however, due to the increased feature size we designate more memory to the
-YARN containers. For the first step, we specify \texttt{k}, the number of clusters. The function \texttt{RandomSeedGenerator.buildRandom} creates a new sequence file containing \texttt{k}
-random centroids from the samples sequence file. The sequence file
-of samples and the sequence file of centroids serve as the input to \texttt{KMeansDriver.run}, a static Mahout API function that launches
-a k-means clustering job on the Hadoop service.
-
-\subsection*{R K-means clustering}
-We use `readVcf' from the R library, `VariantAnnotation', to iteratively read in the VCF file. For each iteration, we convert the data to a matrix and
-convert the values for each variant to a double, as mentioned previously. We then transpose the matrix, so individuals are represented
-by rows, and append the matrix to that from the previous iteration. Once the matrix contains all of the variants from the VCF file, we cluster the data
-using the `kmeans' algorithm from the `stats' library.
-
-\subsection*{Clustering quality}
-We scored the clusters with the \ARI{}~\citep{Hubert1985}. 
-This index compares two different cluster sets, here the prediction $X = \{ X_1, X_2, \ldots , X_r \}$ and the annotated data $Y = \{ Y_1, Y_2, \ldots , Y_s \}$, and assigns a value based on the overlap between $X$ and $Y$ as captured in a contingency table $\left[n_{ij}\right]$ where each entry $n_{ij}$ denotes the number of objects in common between $X_i$ and $Y_j : n_{ij}=|X_i \cap Y_j|$.
-The \ARI{} is then 
-{\tiny
-\begin{eqnarray*}
-ARI=\frac{\sum_{ij}{{n_{ij}\choose 2}} - \left[ \sum_{i}{{a_i\choose2}} \sum_{j}{{b_i\choose2}} \right] / {n\choose2}}{\frac{1}{2} \left[ \sum_{i}{{a_{i}\choose 2} + \sum_{j}{{b_{j}\choose 2}}} \right] - \left[ \sum_{i}{{a_{i}\choose 2} + \sum_{j}{{b_{j}\choose 2}}} \right] / {n\choose2}}
-\end{eqnarray*}
-}
-where $a$ and $b$ are the sums over the rows and columns of the contingency table.
-
-The value is zero for independent clusterings and one for identical clusterings. 
-We compared the Mahout k-means clusters and the R k-means clusters to the known clusters based on meta data from the 1000 Genomes Project.
-For the known clusters, individuals from the both phase1 and phase3 datasets were partitioned by their super population code, ASN, EUR, AFR, and AMR, where phase3 included EAS and SAS instead of ASN.
-
-\subsection*{Performance}
-We recorded the job time from the 'Elapsed' time as reported by the Hadoop JobHistory server. This is measured from the job's start time to finish time.
-Because k-means clustering performs multiple iterations, and therefore spawns multiple MapReduce jobs, the time we reported is the sum of the elapsed time of each job.
-For the R job, where we could not easily separate the pre-processing and clustering components, we printed the time between
-stages using the function, \texttt{proc.time()}.
-
-\subsection*{Graph visualisation}
-GraphML files generated using Mahout Cluster Dumper were visualised using Gephi. Layouts were applied in the following order: Random, ForceAtlas 2, Yifan Hu Proportional.
-For ForceAtlas 2 behaviour alternatives ``dissuade hubs'', ``linlog mode'' and ``prevent overlap'' were set to true, ``approximate repulsion'' was set to false.
-
-\end{methods}
-
-\begin{table}[!t]
-\processtable{Datasets used in this study.\label{Tab:01}}
-{\begin{tabular}{lrrrrr}\toprule
-name& individuals & variants & in-memory  & population\\
-& & &size (GB) &groups& \\\midrule
-        \NinteenPhaseone{} & 1,092 & 816,115 & 7.13  & 4\\
-        \OnePhaseone{} & 1,092 & 3,007,196 & 26.27  & 4\\
-        \SevenPhaseone{} & 1,092 & 18,984,880 & 165.85 & 4\\
-        \FullPhaseone{} & 1,092 & 38,219,238 & 333.88 & 4\\
-	\OnePhasethree{} & 2,504 & 6,468,094 & 131.17 & 5\\
-	\ThreePhasethree{} & 2,504 & 19,381,970 & 393.06 & 5\\
-	\FullPhasethree{} & 2,504 & 81,271,745 & 1602.00 & 5\\\botrule
-\end{tabular}}{Datasets used in this study; based on portions of the phase1 and phase3 data from the 1000 Genomes Project.
-In-memory size is the approximate memory required to store the data as a double-precision matrix of size n,m using the formula: n*m*8bytes/10\textsuperscript{9}bytes.
-}
-\end{table}
-
-\begin{table*}[!t]
-\processtable{Resource allocation and time consumption.\label{Tab:02}}
-{\begin{tabular}{ll|cc|cc|rr|c}\toprule
-Tool & Dataset & \multicolumn{2}{ c |}{Preprocesing (MB)} & \multicolumn{2}{ c |}{Clustering (MB)} & \multicolumn{2}{ c |}{Time} & Accuracy\\
-& & map & reduce  & map & reduce  & pre-processing & clustering & \\\midrule
-        R &\NinteenPhaseone{} & - & - & - & - & 55m 41s & 7s &0.614\\
-        Mahout & \NinteenPhaseone{} & 1536 & 3072 & 1536 & 3072 & 6m 10s & 8m 32s & 0.705\\
-        & \OnePhaseone{} & 1536 & 3072 & 1536 & 3072 & 9m 52s & 12m 49s & 0.798\\
-        & \SevenPhaseone{} & 1536 & 3072 & 4096 & 4096  & 43m 56s & 3h 28m & 0.813\\
-       % & \FullPhaseone{} & 1536 & 3072 & 6144 & 6144  & 85m 29s & 7h 43m & 0.723\\
-        & \OnePhasethree{} & 1536 & 3072 & 1536 & 3072 & 23m 27s & 21m 14s & 0.622\\
-        & \ThreePhasethree{} & 1536 & 3072 & 4096 & 4096 & 63m 49s & 4h 21m & 0.825\\
-        & \FullPhasethree{} & 1536 & 3072 & 6144 & 6144 & 5h 36m & 21h 5m &0.840\\\botrule
-\end{tabular}}{The resources allocated to each MapReduce yarn-container and time taken for the job(s) to complete as well as the accuracy measured as \ARI{}.}
-\end{table*}
-
-\begin{table*}[!t]
-\processtable{Resource allocation and time consumption.\label{Tab:03}}
-{\begin{tabular}{c|cc|cc|c}\toprule
-Dataset & \multicolumn{2}{ c |}{Available Resources} & \multicolumn{2}{ c |}{Container Size (MB)} & Time\\
-& CPU cores & Memory (GB)  & map & reduce & \\\midrule
-        \OnePhaseone{} & 24 & 36 & 1536 & 3072 & 12m 49s\\
-        \OnePhaseone{} & 24 & 36 & 4096 & 4096  & 17m 47s\\
-        \OnePhaseone{} & 448 & 1310 & 1536 & 4096  & 5m 22s\\
-        \FullPhasethree{} & 24 & 36 & 6144 & 6144 & 21h 4m\\
-        \FullPhasethree{} & 448 & 1310 & 6144 & 6144 & 4h 42m\\\botrule
-\end{tabular}}{The resources allocated to each MapReduce yarn-container and time taken for the job(s) to complete.}
-\end{table*}
-
-\begin{figure}[!tpb]%figure2
-\centerline{\includegraphics[type=pdf,ext=.pdf,read=.pdf, scale=0.40]{signature}}
-        \label{fig:sign}
-        \caption{{\bf Illustration of Mahout-based clustering of genotypes.}
-      The image shows how the here introduced interface converts the VCF file to a mahout usable vector-based data type.
-Based on k-means it illustrates how the Map-step procedure groups of vectors by their nearest cluster center and the Reduce-step averages the vectors in each group to find the new value of the updated cluster center.
-The mahout-produced output is then converted into a visualisation.}
-
-\end{figure}
-
-\section{Key Points}
-\begin{itemize}
-\item May bioinformatics tasks can be converted into a MapReduce-paradigm making Hadoop-based approaches attractive.
-\item Scaling seamlessly to larger data sets has been celebrated as Hadoop's superior functionality over conventional parallelisation methodologies, however this only applies on resource-uncapped machines.
-\item On machines in resource settings, we demonstrate that achieving optimal performance still requires careful resource adjustment taking machine architecture and dataset features into account. 
-\item Unlike R, Hadoop-based approaches can handle clustering of genome wide sequencing variants of cohorts with 2500 individuals
-\end{itemize}
-
-\section*{Acknowledgement}
+\section*{Acknowledgements}
   A.R.O was funded by the NSW Cancer Institute Big Data Big Impact schema, F.A.B by the National Health and Medical Research Council [1051757] and both D.C.B and N.S by Commonwealth Scientific and Industrial Research Organisation's Transformational Capability Platform, Science and Industry Endowment Fund and Information Management and Technology Services.
@@
 
-%\bibliographystyle{natbib}
-%\bibliographystyle{achemnat}
-%\bibliographystyle{plainnat}
-\bibliographystyle{abbrv}
-%\bibliographystyle{bioinformatics}
-\bibliography{genotypeClustering}  
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%                  The Bibliography                       %%
+%%                                                         %%
+%%  Bmc_mathpys.bst  will be used to                       %%
+%%  create a .BBL file for submission.                     %%
+%%  After submission of the .TEX file,                     %%
+%%  you will be prompted to submit your .BBL file.         %%
+%%                                                         %%
+%%                                                         %%
+%%  Note that the displayed Bibliography will not          %%
+%%  necessarily be rendered by Latex exactly as specified  %%
+%%  in the online Instructions for Authors.                %%
+%%                                                         %%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+% if your bibliography is in bibtex format, use those commands:
+\bibliographystyle{bmc-mathphys} % Style BST file (bmc-mathphys, vancouver, spbasic).
+\bibliography{genotypeClustering}      % Bibliography file (usually '*.bib' )
+% for author-year bibliography (bmc-mathphys or spbasic)
+% a) write to bib file (bmc-mathphys only)
+% @settings{label, options="nameyear"}
+% b) uncomment next line
+%\nocite{label}
+
+% or include bibliography directly:
+% \begin{thebibliography}
+% \bibitem{b1}
+% \end{thebibliography}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%                               %%
+%% Figures                       %%
+%%                               %%
+%% NB: this is for captions and  %%
+%% Titles. All graphics must be  %%
+%% submitted separately and NOT  %%
+%% included in the Tex document  %%
+%%                               %%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%%
+%% Do not use \listoffigures as most will included as separate files
+
+\section*{Figures}
+  \begin{figure}[h!]
+  \caption{\csentence{Sample figure title.}
+      A short description of the figure content
+      should go here.}
+      \end{figure}
+
+\begin{figure}[h!]
+  \caption{\csentence{Sample figure title.}
+      Figure legend text.}
+      \end{figure}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%                               %%
+%% Tables                        %%
+%%                               %%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%% Use of \listoftables is discouraged.
+%%
+\section*{Tables}
+\begin{table}[h!]
+\caption{Sample table title. This is where the description of the table should go.}
+      \begin{tabular}{cccc}
+        \hline
+           & B1  &B2   & B3\\ \hline
+        A1 & 0.1 & 0.2 & 0.3\\
+        A2 & ... & ..  & .\\
+        A3 & ..  & .   & .\\ \hline
+      \end{tabular}
+\end{table}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%                               %%
+%% Additional Files              %%
+%%                               %%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section*{Additional Files}
+  \subsection*{Additional file 1 --- Sample additional file title}
+    Additional file descriptions text (including details of how to
+    view the file, if it is in a non-standard format or the file extension).  This might
+    refer to a multi-page table or a figure.
+
+  \subsection*{Additional file 2 --- Sample additional file title}
+    Additional file descriptions text.
+
+\end{backmatter}
 \end{document}

### 2015-07-28 doc/publications/journal2014/paper.tex
comparison results

@@
 \subsection*{Spark enables faster clustering of individuals compared to traditional methods}
-Add the results for the comparison of python, R, spark and hadoop here \ldots
+
+We compare the time required to cluster individuals based on their chromosome 22 variants using various implementations.
+We compare our VariantSpark implementation with ADAM and our Hadoop implementation, as well as a Python and R implementation.
+To keep the comparison fair for the Python and R implementations, we perform the comparisons on a single virtual-machine, rather than a cluster.
+Therefore, for k-means clustering, each implementation has access to the same resources.
+Each of the k-means clustering implementations that we compare require the input to be row vectors.
+We need to pre-process the VCF input files to this format, so we compare the time required for the pre-processing step, as well as the time required for k-means clustering.
+The python and R pre-processing implementation only runs on a single thread, so we don't include this in the comparison.
+
+Pre-processing the data is fastest in VariantSpark, requiring 2mins and 58secs. Next fastest is the ADAM implementation, requiring 12min and 48secs. Our Hadoop implementation took 14min and 22secs.
+Unlike VariantSpark and our Hadoop implementation, which pre-process VCF files directly, ADAM requires the input to be in their ADAM file format.
+VCF files must be converted to this binary format before they can be pre-processed into the row vector format.
+Although this additional pre-processing step is only required as a one-off for each input file, it requires an additional 13mins and 13secs.
+
+Clustering the samples is also fastest in VariantSpark, albeit by a smaller margin.
+The time required is 1min and 20secs compared to 1min and 52secs, in VariantSpark and ADAM, respectively.
+Hadoop however, required 14mins and 23secs.
+The similarity between VariantSpark and ADAM is expected, as they both use Spark's MLlib k-means implementation.
+The 30 second speed increase in VariantSpark is likely due to VariantSpark's pre-processing implementation converting VCF files to sparse vectors, whereas ADAM creates dense vectors.
+Due to the sparse nature of these row-vectors, the sparse vector format is likely to reduce some of the initial overhead.
+
+We also compared the k-means clustering times to a multi-threaded approach in Python and R.
+Although slower than the Spark implementations in VariantSpark and ADAM, k-means clustering in both Python and R were faster than our Hadoop implementation.
+Python required 11min and 29sec, whereas R required 7min and 25sec.
+This demonstrates one of Hadoop's disadvantages (one which Spark dramatically improves on).
+Hadoop can scale to datasets that exceed memory size as wall as distributing work across nodes in a cluster.
+However, in this example, the entire dataset can fit in the memory on a single machine.
+The k-means implemtations
+we use in R and Python cluster the in-memory row vectors.
+Hadoop, however, writes the output of each MapReduce iteration to disk.
+This unnecessary disk IO forms a bottleneck that results in the longer run-times.
+Spark, on the other hand, can cache the intermediate results in memory, virtually eliminating the disk IO bottleneck on smaller datasets.
+
+To summarise the total times for each big-data platform, VariantSpark was 4mins and 18secs, ADAM was 14mins and 14secs (or 27mins and 53secs including the one-off conversion of VCF to ADAM) and our Hadoop implementation was 28mins and 45secs.
 
 \subsection*{VariantSpark allows genome wide sampling of variants to improve clustering quality}
-Add the results for the scalling from 1\% to X\% \ldots
+Add the results for the scalling from 1\% to X\%
 

### 2015-08-14 doc/publications/journal2014/paper.tex
updated introduction

@@
 
-\title{VariantSpark: Population Scale Clustering of Genotype Information}
+\title{Variant{\sc Spark}: Population Scale Clustering of Genotype Information}
 
@@
 \begin{abstract} % abstract
-\parttitle{Motivation}: Genomic information is increasingly used in medical research, giving rise to the need for efficient analysis methodology able to cope with thousands of individuals and millions of variants.
+\parttitle{Motivation}: Genomic information is increasingly used in medical practice, giving rise to the need for efficient analysis methodology able to cope with thousands of individuals and millions of variants.
 The widely used Hadoop MapReduce architecture and associated machine-learning library, Mahout, provide the means for tackling computationally challenging tasks.
 However, many genomic analyses do not fit the Map-Reduce paradigm.
-We therefore survey the recently developed Spark engine, along with its associated machine learning library, MLlib, which are more flexible in the parallelisation architecture and hence more suitable for population-scale bioinformatics tasks.
+We therefore survey the recently developed {\sc Spark} engine, along with its associated machine learning library, MLlib, which are more flexible in the parallelisation architecture and hence more suitable for population-scale bioinformatics tasks.
 To do this, we developed an interface from Mahout and MLlib to the standard variant format (VCF), which opens up the usage of advanced, efficient machine learning algorithms to genomic data.
@@
 \parttitle{Results}: We successfully clustered more than 2,500 individuals each having more than 81 Million variants.
-Both Hadoop and Spark have superior performance over traditional approaches.
-We observe a 50 fold speedup when using the efficient in-memory Spark compute engine compared to Hadoop.
-Furthermore, our implementation achieves a better performance compared to a previously published Spark-based genome clustering approach, ADAM.
+Both Hadoop and {\sc Spark} have superior performance over traditional approaches.
+We observe a 50 fold speedup when using the efficient in-memory {\sc Spark} compute engine compared to Hadoop.
+Furthermore, our implementation achieves a better performance compared to a previously published {\sc Spark}-based genome clustering approach, ADAM.
 
-\parttitle{Availability}: The package is written in Java and available at https: //github.com/BauerLab/GeMaIn.
+\parttitle{Availability}: The package is written in Java and available at https: //github.com/BauerLab/VariantSpark.
 \end{abstract}
@@
 
+Genomic information is increasingly used in medical practice.
+A commonly performed task in such applications is grouping individuals based on their genomic profile to identify population association~\cite{Gao2007} or elucidate different haplotype involvement in diseases susceptibility~\cite{Laitman2013}.
+Due to the decreasing sequencing cost it is economical to generate studies with sample sizes previously reserved for larger consortia such as the 1000 genomes project~\cite{1KG2012} or The Cancer Genome Atlas, TCGA~\cite{TCGA2013}.
+At the same time, whole genome sequencing enables the inclusion of rare or even somatic mutations in the analysis, increasing the feature space by orders of magnitude.
+This drastic increase in both sample numbers and features per sample requires a massively parallel approach to data processing~\cite{Stein2010}.
+Traditional parallelisation strategies like OpenMPI or hardware accelerators (GPGPU) cannot scale to variable data sizes or require purpose-build hardware.
+
+Addressing this issue, {\sc Apache Hadoop MapReduce}~\cite{Borthakur2007} transforms data into 'key-value pairs' that can then be distributed between multiple nodes across a commodity compute-cluster depending on the size of the problem.
+MapReduce approaches are increasingly being used in bioinformatics (for reviews see~\cite{Zou2013, Qiu2010,Taylor2010}). 
+This is especially the case for sequence analysis tasks, such as read mapping~\cite{Schatz2009}, duplicate removal~\cite{Jourdren2012}, and variant calling~\cite{Langmead2009, McKenna2010} as well as Genome Wide Analysis Study based tasks~\cite{Huang2013, Guo2014}.
+Apache also developed a machine learning library, Mahout~\cite{Owen2011}, which allows efficient out-of-the-box analysis to be applied to clinical applications, such as medical health records~\cite{Ko2014}.
+Unfortunately, the MapReduce paradigm is not always the optimal solution, specifically for bioinformatics or machine learning applications that require iterative in-memory computation.
+In addition, Hadoop is disk IO intensive, and this can prove to be a bottleneck in processing-speed.
+
+{\sc Apache Spark}~\cite{Zaharia2011} is a more recent compute engine, which overcomes many of Hadoop's limitations. 
+One of the main benefits is that it allows programs to cache data in memory; potentially eliminating, or at least reducing, the bottleneck of disk IO. 
+When utilising caching, Apache claim {\sc Spark} to be 100x faster than Hadoop. 
+Although {\sc Spark} allows MapReduce-like programs, it does not require programs to exactly model the MapReduce paradigm, which in turn allows more flexibility when designing programs. 
+Recognising the capability, the Big Data Genomics (BDG) group has recently demonstrated the strength of {\sc Spark} in a genomic clustering application using ADAM, a set of formats and APIs as well as processing stage implementations for genomic data~\cite{Massie2013}.
+While the speedup over traditional methods was impressive, being limited by constraints within this general genomics framework hampered performance.
+
+We hence developed a more focused purpose-built application in {\sc Spark} to perform genomic clustering of individuals. 
+We utilise {\sc Spark}'s machine-learning library, MLlib, and provide an interface from the standard variant data format, variant call format (VCF)~\cite{1KG2012}, which opens up the application of MLlib's different machine learning algorithms a wide range of genotype-based analysis tasks.
+To demonstrate its capability, we cluster variant datasets from the 1000 genomes project~\cite{1KG2012} to determine population structure using the k-means clustering algorithm available in MLlib's. 
+In the first section we benchmark the performance and accuracy of MLlib's implementations against Hadoop's Mahout as well as more traditional methods (R, Python) limiting the used data to chromosome 22.
+In section two we demonstrate {\sc Spark}'s full capacity by seamlessly scaling from 1\% to 100\% of the human genome.
+In section three we replicate the analysis by using XX genomes from the Personal Genome Project~\cite{Lunshof2013}.
+In the last section we discuss the pipeline for visualising the resulting cluster.
+
 \section*{Results}
-\subsection*{Spark enables faster clustering of individuals compared to traditional methods}
+\subsection*{{\sc Spark} enables faster clustering of individuals compared to traditional methods}
 
 We compare the time required to cluster individuals based on their chromosome 22 variants using various implementations.
-We compare our VariantSpark implementation with ADAM and our Hadoop implementation, as well as a Python and R implementation.
+We compare our Variant{\sc Spark} implementation with ADAM and our Hadoop implementation, as well as a Python and R implementation.
 To keep the comparison fair for the Python and R implementations, we perform the comparisons on a single virtual-machine, rather than a cluster.
@@
 
-Pre-processing the data is fastest in VariantSpark, requiring 2mins and 58secs. Next fastest is the ADAM implementation, requiring 12min and 48secs. Our Hadoop implementation took 14min and 22secs.
-Unlike VariantSpark and our Hadoop implementation, which pre-process VCF files directly, ADAM requires the input to be in their ADAM file format.
+Pre-processing the data is fastest in Variant{\sc Spark}, requiring 2mins and 58secs.
+Next fastest is the ADAM implementation, requiring 12min and 48secs.
+Our Hadoop implementation took 14min and 22secs.
+Unlike Variant{\sc Spark} and our Hadoop implementation, which pre-process VCF files directly, ADAM requires the input to be in their ADAM file format.
 VCF files must be converted to this binary format before they can be pre-processed into the row vector format.
@@
 
-Clustering the samples is also fastest in VariantSpark, albeit by a smaller margin.
-The time required is 1min and 20secs compared to 1min and 52secs, in VariantSpark and ADAM, respectively.
+Clustering the samples is also fastest in Variant{\sc Spark}, albeit by a smaller margin.
+The time required is 1min and 20secs compared to 1min and 52secs, in Variant{\sc Spark} and ADAM, respectively.
 Hadoop however, required 14mins and 23secs.
-The similarity between VariantSpark and ADAM is expected, as they both use Spark's MLlib k-means implementation.
-The 30 second speed increase in VariantSpark is likely due to VariantSpark's pre-processing implementation converting VCF files to sparse vectors, whereas ADAM creates dense vectors.
+The similarity between Variant{\sc Spark} and ADAM is expected, as they both use {\sc Spark}'s MLlib k-means implementation.
+The 30 second speed increase in Variant{\sc Spark} is likely due to Variant{\sc Spark}'s pre-processing implementation converting VCF files to sparse vectors, whereas ADAM creates dense vectors.
 Due to the sparse nature of these row-vectors, the sparse vector format is likely to reduce some of the initial overhead.
@@
 We also compared the k-means clustering times to a multi-threaded approach in Python and R.
-Although slower than the Spark implementations in VariantSpark and ADAM, k-means clustering in both Python and R were faster than our Hadoop implementation.
+Although slower than the {\sc Spark} implementations in Variant{\sc Spark} and ADAM, k-means clustering in both Python and R were faster than our Hadoop implementation.
 Python required 11min and 29sec, whereas R required 7min and 25sec.
-This demonstrates one of Hadoop's disadvantages (one which Spark dramatically improves on).
+This demonstrates one of Hadoop's disadvantages (one which {\sc Spark} dramatically improves on).
 Hadoop can scale to datasets that exceed memory size as wall as distributing work across nodes in a cluster.
@@
 This unnecessary disk IO forms a bottleneck that results in the longer run-times.
-Spark, on the other hand, can cache the intermediate results in memory, virtually eliminating the disk IO bottleneck on smaller datasets.
+{\sc Spark}, on the other hand, can cache the intermediate results in memory, virtually eliminating the disk IO bottleneck on smaller datasets.
 
-To summarise the total times for each big-data platform, VariantSpark was 4mins and 18secs, ADAM was 14mins and 14secs (or 27mins and 53secs including the one-off conversion of VCF to ADAM) and our Hadoop implementation was 28mins and 45secs.
+To summarise the total times for each big-data platform, Variant{\sc Spark} was 4mins and 18secs, ADAM was 14mins and 14secs (or 27mins and 53secs including the one-off conversion of VCF to ADAM) and our Hadoop implementation was 28mins and 45secs.
 
-\subsection*{VariantSpark allows genome wide sampling of variants to improve clustering quality}
+\subsection*{Variant{\sc Spark} allows genome wide sampling of variants to improve clustering quality}
 Add the results for the scalling from 1\% to X\%
+
+\subsection*{Personal genomics project}
 

### 2015-08-14 doc/publications/journal2014/paper.tex
Results sections

--- 2015-07-28 doc/publications/journal2014/paper.tex
@@
 However, in this example, the entire dataset can fit in the memory on a single machine.
-The k-means implemtations
+The k-means implementation
 we use in R and Python cluster the in-memory row vectors.
@@
 
-To summarise the total times for each big-data platform, VariantSpark was 4mins and 18secs, ADAM was 14mins and 14secs (or 27mins and 53secs including the one-off conversion of VCF to ADAM) and our Hadoop implementation was 28mins and 45secs.
+To summarise the total times for each big-data platform, VariantSpark was 4mins and 18secs, ADAM was 14mins and 14secs (or 27mins and 53secs including the one-off conversion of VCF to ADAM), and our Hadoop implementation was 28mins and 45secs.
 
 \subsection*{VariantSpark allows genome wide sampling of variants to improve clustering quality}
-Add the results for the scalling from 1\% to X\%
+To demonstrate the scalability of Spark when applied to clustering individuals on their variants, we scale the size up from chromosome 22 to all 22 autosomes.
+Although this pre-processing and clustering is successful, the memory required for clustering increases dramatically.
+
+Where our previous clustering of chromosome 22 required no more than 4GB memory per executor, clustering the entire genome increases this requirement to over 20GB.
+On the other hand, the memory requirements for our pre-processing implementation remain unchanged, at 2 gigabytes.
+One factor which results in these differing memory requirements is the sparsity of the vectors used to store the variants.
+Due to their sparse nature (the abundance of 0s) we can store them efficiently as Sparse Vectors, rather than Dense Vectors.
+However, to cluster these vectors, the k-means algorithm converts them to Dense Vectors.
+With Dense Vectors, all the 0 values (which are excluded from Sparse Vectors so don't require any memory) are explicitly defined, considerably increasing the memory required.
 

### 2015-08-14 doc/publications/journal2014/paper.tex

--- 2015-08-14 doc/publications/journal2014/paper.tex
@@
 \subsection*{Variant{\sc Spark} allows genome wide sampling of variants to improve clustering quality}
-Add the results for the scalling from 1\% to X\%
+
+To demonstrate the scalability of Spark when applied to clustering individuals on their variants, we scale the size up from chromosome 22 to all 22 autosomes.
+Although this pre-processing and clustering is successful, the memory required for clustering increases dramatically.
+
+Where our previous clustering of chromosome 22 required no more than 4GB memory per executor, clustering the entire genome increases this requirement to over 20GB.
+On the other hand, the memory requirements for our pre-processing implementation remain unchanged, at 2 gigabytes.
+One factor which results in these differing memory requirements is the sparsity of the vectors used to store the variants.
+Due to their sparse nature (the abundance of 0s) we can store them efficiently as Sparse Vectors, rather than Dense Vectors.
+However, to cluster these vectors, the k-means algorithm converts them to Dense Vectors.
+With Dense Vectors, all the 0 values (which are excluded from Sparse Vectors so don't require any memory) are explicitly defined, considerably increasing the memory required.
 

### 2015-08-17 doc/publications/journal2014/paper.tex
some changes in result section  1

--- 2015-08-14 doc/publications/journal2014/paper.tex
@@
 \endlocaldefs
+
+\newcommand{\variantSpark}{{\sc Variant Spark}}
+\newcommand{\ARI}{Adjusted Rand Index}
 
@@
 
-We compare the time required to cluster individuals based on their chromosome 22 variants using various implementations.
+In this section we compare the time required to cluster individuals based on genomic variants using \variantSpark\ against ADAM as well as the more traditional approaches using Hadoop Mahout, Python and R.
+We limit the genomic variants to only chromosome 22 as the traditional approaches have substantially larger memory consumption rendering a whole-genome input infeasible.  
+Furthermore, we perform the comparisons on a single virtual-machine to ensure the five different approaches have access to the same resources.
+We use k-means clustering algorithm in the respective implementations, which all require the VCF input files to be pre-processed (see methods). 
+We therefore compare the time required for the pre-processing step, as well as the time required for k-means clustering.
+% DENIS: include this in the table too
+%The python and R pre-processing implementation only runs on a single thread, so we don't include this in the comparison.
+
+%We compare the time required to cluster individuals based on their chromosome 22 variants using various implementations.
 We compare our Variant{\sc Spark} implementation with ADAM and our Hadoop implementation, as well as a Python and R implementation.
-To keep the comparison fair for the Python and R implementations, we perform the comparisons on a single virtual-machine, rather than a cluster.
+%To keep the comparison fair for the Python and R implementations, we perform the comparisons on a single virtual-machine, rather than a cluster.
 Therefore, for k-means clustering, each implementation has access to the same resources.
-Each of the k-means clustering implementations that we compare require the input to be row vectors.
+%Each of the k-means clustering implementations that we compare require the input to be row vectors.
 We need to pre-process the VCF input files to this format, so we compare the time required for the pre-processing step, as well as the time required for k-means clustering.
-The python and R pre-processing implementation only runs on a single thread, so we don't include this in the comparison.
+%The python and R pre-processing implementation only runs on a single thread, so we don't include this in the comparison.
 
-Pre-processing the data is fastest in Variant{\sc Spark}, requiring 2mins and 58secs.
-Next fastest is the ADAM implementation, requiring 12min and 48secs.
-Our Hadoop implementation took 14min and 22secs.
-Unlike Variant{\sc Spark} and our Hadoop implementation, which pre-process VCF files directly, ADAM requires the input to be in their ADAM file format.
+As shown in Table~\ref{fivewaycomparison}, pre-processing the data is fastest in \variantSpark\, requiring 2mins and 58secs. %178s 
+This is almost 80\% faster than the ADAM implementation (12min and 48secs) %768s
+or our Hadoop implementation(14min and 22secs). %862s
+Unlike \variantSpark\ or our Hadoop implementation, which pre-process VCF files directly, ADAM requires the input to be in their ADAM file format. 
+VCF files must hence be converted to this binary format before they can be pre-processed into the row vector format. 
+Although this additional pre-processing step is only required as a one-off for each input file, it requires an additional 13mins and 13secs %793s
+rendering our approach an order of magnitude faster. 
+[Results for Python and R here].
+This substantially different runtime for the pre-processing in Python and R is due to these implementations does not natively supporting multithreading while \variantSpark, ADAM and Hadoop using XX CPUs.
+
+%Pre-processing the data is fastest in \variantSpark\, requiring 2mins and 58secs. Next fastest is the ADAM implementation, requiring 12min and 48secs. 
+%Our Hadoop implementation took 14min and 22secs.
+%Unlike Variant{\sc Spark} and our Hadoop implementation, which pre-process VCF files directly, ADAM requires the input to be in their ADAM file format.
 VCF files must be converted to this binary format before they can be pre-processed into the row vector format.
@@
 
-Clustering the samples is also fastest in Variant{\sc Spark}, albeit by a smaller margin.
-The time required is 1min and 20secs compared to 1min and 52secs, in Variant{\sc Spark} and ADAM, respectively.
-Hadoop however, required 14mins and 23secs.
-The similarity between Variant{\sc Spark} and ADAM is expected, as they both use {\sc Spark}'s MLlib k-means implementation.
-The 30 second speed increase in Variant{\sc Spark} is likely due to Variant{\sc Spark}'s pre-processing implementation converting VCF files to sparse vectors, whereas ADAM creates dense vectors.
-Due to the sparse nature of these row-vectors, the sparse vector format is likely to reduce some of the initial overhead.
+Clustering the samples 30\% fastest in \variantSpark\ compared to ADAM (1min and 20secs compared to 1min and 52secs) and  %80 vs 112
+90\% faster than Hadoop (14mins and 23secs), see Table~\ref{fivewaycomparison}.
+The similarity between \variantSpark\ and ADAM is expected, as they both use {\sc Spark}'s MLlib k-means implementation. 
+The 30 second speed increase in \variantSpark\ is likely due us converting the VCF files to sparse vectors, whereas ADAM creates dense vectors.
+%Due to the sparse nature of these row-vectors, the sparse vector format is likely to reduce some of the initial overhead.
+We also compared the k-means clustering times to a multi-threaded approach in Python and R. 
+Although an order of magnitude slower than \variantSpark\ and ADAM, k-means clustering in both Python and R were faster than our Hadoop implementation.
+Python required 11min and 29sec, whereas R required 7min and 25sec.
+This demonstrates one of Hadoop's disadvantages
 
-We also compared the k-means clustering times to a multi-threaded approach in Python and R.
-Although slower than the {\sc Spark} implementations in Variant{\sc Spark} and ADAM, k-means clustering in both Python and R were faster than our Hadoop implementation.
-Python required 11min and 29sec, whereas R required 7min and 25sec.
-This demonstrates one of Hadoop's disadvantages (one which {\sc Spark} dramatically improves on).
+ (one which {\sc Spark} dramatically improves on).
 Hadoop can scale to datasets that exceed memory size as wall as distributing work across nodes in a cluster.
@@
 
-To summarise the total times for each big-data platform, Variant{\sc Spark} was 4mins and 18secs, ADAM was 14mins and 14secs (or 27mins and 53secs including the one-off conversion of VCF to ADAM) and our Hadoop implementation was 28mins and 45secs.
+%Clustering the samples is also fastest in \variantSpark\, albeit by a smaller margin.
+The time required is 1min and 20secs compared to 1min and 52secs, in Variant{\sc Spark} and ADAM, respectively.
+Hadoop however, required 14mins and 23secs.
+%The similarity between Variant{\sc Spark} and ADAM is expected, as they both use {\sc Spark}'s MLlib k-means implementation.
+The 30 second speed increase in Variant{\sc Spark} is likely due to Variant{\sc Spark}'s pre-processing implementation converting VCF files to sparse vectors, whereas ADAM creates dense vectors.
+%Due to the sparse nature of these row-vectors, the sparse vector format is likely to reduce some of the initial overhead.
+
+%We also compared the k-means clustering times to a multi-threaded approach in Python and R.
+Although slower than the {\sc Spark} implementations in Variant{\sc Spark} and ADAM, k-means clustering in both Python and R were faster than our Hadoop implementation.
+Python required 11min and 29sec, whereas R required 7min and 25sec.
+This demonstrates one of Hadoop's disadvantages (one which {\sc Spark} dramatically improves on).
+Hadoop can scale to datasets that exceed memory size as wall as distributing work across nodes in a cluster.
+However, in this example, the entire dataset can fit in the memory on a single machine.
+The k-means implemtations
+%we use in R and Python cluster the in-memory row vectors.
+Hadoop, however, writes the output of each MapReduce iteration to disk.
+This unnecessary disk IO forms a bottleneck that results in the longer run-times.
+{\sc Spark}, on the other hand, can cache the intermediate results in memory, virtually eliminating the disk IO bottleneck on smaller datasets.
+
+%To summarise the total times for each big-data platform, Variant{\sc Spark} was 4mins and 18secs, ADAM was 14mins and 14secs (or 27mins and 53secs including the one-off conversion of VCF to ADAM) and our Hadoop implementation was 28mins and 45secs.
 
@@
 \section*{Tables}
+\label{fivewaycomparison}
 \begin{table}[h!]
-\caption{Sample table title. This is where the description of the table should go.}
-      \begin{tabular}{cccc}
+\caption{The resources consumption of the five compared methods as well as the accuracy measured as \ARI{} on chromosome 22.}
+      \begin{tabular}{|c|ccc|ccc|c|c|}
         \hline
-           & B1  &B2   & B3\\ \hline
-        A1 & 0.1 & 0.2 & 0.3\\
-        A2 & ... & ..  & .\\
-        A3 & ..  & .   & .\\ \hline
+           Tool &  \multicolumn{3}{ c |}{Pre-processing} & \multicolumn{3}{ c |}{Clustering} & Accuracy \\
+& threads & memory & time  &threads & memory & time  & \\
+  \hline
+\variantSpark\ &  & \\ 
+ADAM & & \\
+Hadoop & & \\
+R &  1&  \\
+Python & 1&  \\
+  \hline
       \end{tabular}

### 2015-08-17 doc/publications/journal2014/paper.tex
updated results

--- 2015-08-14 doc/publications/journal2014/paper.tex
@@
 
-\title{Variant{\sc Spark}: Population Scale Clustering of Genotype Information}
+\title{VariantSpark: Population Scale Clustering of Genotype Information}
 
@@
 
-\parttitle{Results}: We successfully clustered more than 2,500 individuals each having more than 81 Million variants.
+\parttitle{Results}: We successfully clustered more than 2,500 individuals each having more than 38 Million variants.
 Both Hadoop and {\sc Spark} have superior performance over traditional approaches.
@@
 
-\parttitle{Availability}: The package is written in Java and available at https: //github.com/BauerLab/VariantSpark.
+\parttitle{Availability}: The package is written in Scala and available at https: //github.com/BauerLab/VariantSpark.
 \end{abstract}
@@
 We compare the time required to cluster individuals based on their chromosome 22 variants using various implementations.
-We compare our Variant{\sc Spark} implementation with ADAM and our Hadoop implementation, as well as a Python and R implementation.
+We compare our {\sc VariantSpark} implementation with ADAM and our Hadoop implementation, as well as a Python and R implementation.
 To keep the comparison fair for the Python and R implementations, we perform the comparisons on a single virtual-machine, rather than a cluster.
@@
 
-Pre-processing the data is fastest in Variant{\sc Spark}, requiring 2mins and 58secs.
+Pre-processing the data is fastest in {\sc VariantSpark}, requiring 2mins and 58secs.
 Next fastest is the ADAM implementation, requiring 12min and 48secs.
 Our Hadoop implementation took 14min and 22secs.
-Unlike Variant{\sc Spark} and our Hadoop implementation, which pre-process VCF files directly, ADAM requires the input to be in their ADAM file format.
+Unlike {\sc VariantSpark} and our Hadoop implementation, which pre-process VCF files directly, ADAM requires the input to be in their ADAM file format.
 VCF files must be converted to this binary format before they can be pre-processed into the row vector format.
@@
 
-Clustering the samples is also fastest in Variant{\sc Spark}, albeit by a smaller margin.
-The time required is 1min and 20secs compared to 1min and 52secs, in Variant{\sc Spark} and ADAM, respectively.
-Hadoop however, required 14mins and 23secs.
-The similarity between Variant{\sc Spark} and ADAM is expected, as they both use {\sc Spark}'s MLlib k-means implementation.
-The 30 second speed increase in Variant{\sc Spark} is likely due to Variant{\sc Spark}'s pre-processing implementation converting VCF files to sparse vectors, whereas ADAM creates dense vectors.
+Clustering the samples is also fastest in {\sc VariantSpark}, albeit by a smaller margin.
+The time required is 1min and 20secs compared to 1min and 52secs, in {\sc VariantSpark} and ADAM, respectively.
+Hadoop, however, required 14mins and 23secs.
+The similarity between {\sc VariantSpark} and ADAM is expected, as they both use {\sc Spark}'s MLlib k-means implementation.
+The 30 second speed increase in {\sc VariantSpark} is likely due to {\sc VariantSpark}'s pre-processing implementation converting VCF files to sparse vectors, whereas ADAM creates dense vectors.
 Due to the sparse nature of these row-vectors, the sparse vector format is likely to reduce some of the initial overhead.
 
-We also compared the k-means clustering times to a multi-threaded approach in Python and R.
-Although slower than the {\sc Spark} implementations in Variant{\sc Spark} and ADAM, k-means clustering in both Python and R were faster than our Hadoop implementation.
-Python required 11min and 29sec, whereas R required 7min and 25sec.
-This demonstrates one of Hadoop's disadvantages (one which {\sc Spark} dramatically improves on).
-Hadoop can scale to datasets that exceed memory size as wall as distributing work across nodes in a cluster.
-However, in this example, the entire dataset can fit in the memory on a single machine.
-The k-means implemtations
-we use in R and Python cluster the in-memory row vectors.
-Hadoop, however, writes the output of each MapReduce iteration to disk.
-This unnecessary disk IO forms a bottleneck that results in the longer run-times.
-{\sc Spark}, on the other hand, can cache the intermediate results in memory, virtually eliminating the disk IO bottleneck on smaller datasets.
+We also compared k-means clustering times to a multi-threaded approach in Python and R.
+Although slower than the {\sc Spark} implementations in {\sc VariantSpark} and ADAM, k-means clustering in both Python and R were faster than our Hadoop implementation.
+Python required 11min and 29sec, whereas R required 7min and 25sec. This demonstrates one of Hadoop's disadvantages (one which {\sc Spark} dramatically improves on): the cost of disk IO.
+Where Hadoop's advantage is that it can scale to datasets that exceed memory size, in this example, the entire dataset can fit in memory on a single machine.
+So, where the k-means implementations we use in R and Python store the output from each k-means iteration in-memory, Hadoop, writes the output of each iteration to disk, whether or not the amount of available memory is sufficient to store this data.
+This unnecessary disk IO forms a bottleneck that results in the longer run-times for Mahout's k-means implementation.
+{\sc Spark}, on the other hand, will cache the intermediate results in memory, virtually eliminating the disk IO bottleneck on smaller datasets that fit entirely in memory, or at least reducing the disk IO by caching a portion of the data.
 
-To summarise the total times for each big-data platform, Variant{\sc Spark} was 4mins and 18secs, ADAM was 14mins and 14secs (or 27mins and 53secs including the one-off conversion of VCF to ADAM) and our Hadoop implementation was 28mins and 45secs.
+To summarise the total times for each big-data platform, {\sc VariantSpark} was 4mins and 18secs, ADAM was 14mins and 14secs (or 27mins and 53secs including the one-off conversion of VCF to ADAM) and our Hadoop implementation was 28mins and 45secs.
 
-\subsection*{Variant{\sc Spark} allows genome wide sampling of variants to improve clustering quality}
+\subsection*{{\sc VariantSpark} allows genome wide sampling of variants to improve clustering quality}
 
-To demonstrate the scalability of Spark when applied to clustering individuals on their variants, we scale the size up from chromosome 22 to all 22 autosomes.
+To demonstrate the scalability of {\sc Spark}, we continue to cluster individuals from the 1000 Genome Project, however, rather than clustering on a single chromosome, we cluster the individuals on their variants from all 22 autosomes (.
 Although this pre-processing and clustering is successful, the memory required for clustering increases dramatically.

### 2015-08-17 doc/publications/journal2014/paper.tex

--- 2015-08-17 doc/publications/journal2014/paper.tex
@@
 We also compared the k-means clustering times to a multi-threaded approach in Python and R.
+
 Although an order of magnitude slower than \variantSpark\ and ADAM, k-means clustering in both Python and R were faster than our Hadoop implementation.
 Python required 11min and 29sec, whereas R required 7min and 25sec.
-This demonstrates one of Hadoop's disadvantages
+This demonstrates one of Hadoop's disadvantages (one which {\sc Spark} dramatically improves on): the cost of disk IO.
+Where Hadoop's advantage is that it can scale to datasets that exceed memory size, in this example, the entire dataset can fit in memory on a single machine.
+So, where the k-means implemtations we use in R and Python store the output from each k-means iteration in memory, Hadoop writes the output of each iteration to disk, whether or not the amount of available memory is sufficient to store this data.
 
- (one which {\sc Spark} dramatically improves on).
-Hadoop can scale to datasets that exceed memory size as wall as distributing work across nodes in a cluster.
-However, in this example, the entire dataset can fit in the memory on a single machine.
-The k-means implemtations
-we use in R and Python cluster the in-memory row vectors.
-Hadoop, however, writes the output of each MapReduce iteration to disk.
 This unnecessary disk IO forms a bottleneck that results in the longer run-times.
@@
 
-\subsection*{Variant{\sc Spark} allows genome wide sampling of variants to improve clustering quality}
-Add the results for the scalling from 1\% to X\%
+\subsection*{{\sc VariantSpark} allows genome wide sampling of variants to improve clustering quality}
+
+To demonstrate the scalability of {\sc Spark}, we continue to cluster individuals from the 1000 Genome Project, however, rather than clustering on a single chromosome, we cluster the individuals on their variants from all 22 autosomes (.
+Although this pre-processing and clustering is successful, the memory required for clustering increases dramatically.
+
+Where our previous clustering of chromosome 22 required no more than 4GB memory per executor, clustering the entire genome increases this requirement to over 20GB.
+On the other hand, the memory requirements for our pre-processing implementation remain unchanged, at 2 gigabytes.
+One factor which results in these differing memory requirements is the sparsity of the vectors used to store the variants.
+Due to their sparse nature (the abundance of 0s) we can store them efficiently as Sparse Vectors, rather than Dense Vectors.
+However, to cluster these vectors, the k-means algorithm converts them to Dense Vectors.
+With Dense Vectors, all the 0 values (which are excluded from Sparse Vectors so don't require any memory) are explicitly defined, considerably increasing the memory required.
 

### 2015-08-18 doc/publications/journal2014/paper.tex
methods and more results

@@
 
-\newcommand{\variantSpark}{{\sc Variant Spark}}
+\newcommand{\variantSpark}{{\sc VariantSpark}}
 \newcommand{\ARI}{Adjusted Rand Index}
@@
 
-To demonstrate the scalability of {\sc Spark}, we continue to cluster individuals from the 1000 Genome Project, however, rather than clustering on a single chromosome, we cluster the individuals on their variants from all 22 autosomes (.
-Although this pre-processing and clustering is successful, the memory required for clustering increases dramatically.
+To demonstrate the scalability of {\sc Spark}, we continue to cluster the same individuals from the 1000 Genome Project, however, rather than clustering on a single chromosome, we cluster the individuals on different subsets of variants from their 22 autosomes, and finally every variant.
+Pre-processing the data takes approximately 12 minutes for a 20\% sample.
+For 40\% of the genome, the time required is about 19 minutes, and finally, to preprocess 100\% of the genome, the time required is about 40 minutes.
+For each of these sample sizes, the memory required doesn't exceed a modest 2GB per executor, even for the entire autosome. Memory requirements for clustering, however, increase by a
 
 Where our previous clustering of chromosome 22 required no more than 4GB memory per executor, clustering the entire genome increases this requirement to over 20GB.
-On the other hand, the memory requirements for our pre-processing implementation remain unchanged, at 2 gigabytes.
 One factor which results in these differing memory requirements is the sparsity of the vectors used to store the variants.
@@
 \subsection*{Implementation}
+For each of the implementations, we pre-process the VCF files to arrays of features.
+In /VariantSpark/ we read in the VCF files as text files to a Resilient Data Set (RDD).
+RDDs allow us to process the files, line-by-line, in parallel.
+We parse each line as tab-separated values, and "zip" each value with it's heading from the VCF file.
+The key-value pairs (KVPs) from each line in the VCF file are stored in separate arrays. We then "zip" each array with a unique index, where these indexes form identifiers for the alleles.
+At this point, we have one array for each line from the VCF file.
+The values in the arrays are KVPs of the parsed tab-separated values zipped with their heading (individual ID), and each array has a unique index.
+We now use a "flatMap" to convert these indexed arrays of KVPs into a less convoluted data-structure, as well as to filter out KVPs that aren't alleles (i.e. KVPs that are VCF metadata).
+This results in KVPs where the key is the individual ID, and the value is a tuple of the allele and it's index.
+We effectively have one KVP for each allele from the VCF file.
+At this point, each allele is still a string straight from the VCF file (i.e.
+"0\textbar1:13"), however, we require doubles.
+To convert this to a double, we take the Hamming distance of the allele, where a 1 represents a heterozygous variant, 2 a homozygous variant, and 0, no variant.
+Also, for each individual, we filter out any alleles that have no variant, leaving us with KVPs of variants (homozygous and heterozygous) only.
+We can now group these KVPs by their key (the individual ID) and map them to sparse vectors.
+This results in a sparse vector for each individual, the format required by MLLibs machine learning algorithms.
+
+For our ADAM comparison, we followed the ADAM implementation from (http://bdgenomics.org/blog/2015/02/02/scalable-genomes-clustering-with-adam-and-spark/).
+
+ For our R implementation, we read in the
 
@@
   \hline
-\variantSpark\ &  & \\ 
-ADAM & & \\
-Hadoop & & \\
-R &  1&  \\
-Python & 1&  \\
+\variantSpark\	& 8	& 32	& 2min 58sec	& 8	& 32	& 1min 20sec	& x	\\ 
+ADAM		& 8	& 32	& 12min 48sec	& 8	& 32	& 1min 52sec	& x	\\
+Hadoop		& 8	& 32	& 14min 22sec	& 8	& 32	& 14min 23sec	& x	\\
+R			& 1	& 32	& 67min		& 8	& 32	& 7min 25sec	& x	\\
+Python		& 1	& 32	& 92min		& 8	& 32	& 11min 29sec	& x	\\
   \hline

### 2015-08-18 doc/publications/journal2014/paper.tex
final section 1

@@
 We therefore compare the time required for the pre-processing step, as well as the time required for k-means clustering.
-% DENIS: include this in the table too
-%The python and R pre-processing implementation only runs on a single thread, so we don't include this in the comparison.
-
-%We compare the time required to cluster individuals based on their chromosome 22 variants using various implementations.
-We compare our Variant{\sc Spark} implementation with ADAM and our Hadoop implementation, as well as a Python and R implementation.
-%To keep the comparison fair for the Python and R implementations, we perform the comparisons on a single virtual-machine, rather than a cluster.
-Therefore, for k-means clustering, each implementation has access to the same resources.
-%Each of the k-means clustering implementations that we compare require the input to be row vectors.
-We need to pre-process the VCF input files to this format, so we compare the time required for the pre-processing step, as well as the time required for k-means clustering.
-%The python and R pre-processing implementation only runs on a single thread, so we don't include this in the comparison.
 
@@
 Although this additional pre-processing step is only required as a one-off for each input file, it requires an additional 13mins and 13secs %793s
-rendering our approach an order of magnitude faster. 
-[Results for Python and R here].
-This substantially different runtime for the pre-processing in Python and R is due to these implementations does not natively supporting multithreading while \variantSpark, ADAM and Hadoop using XX CPUs.
-
-%Pre-processing the data is fastest in \variantSpark\, requiring 2mins and 58secs. Next fastest is the ADAM implementation, requiring 12min and 48secs. 
-%Our Hadoop implementation took 14min and 22secs.
-%Unlike Variant{\sc Spark} and our Hadoop implementation, which pre-process VCF files directly, ADAM requires the input to be in their ADAM file format.
-VCF files must be converted to this binary format before they can be pre-processed into the row vector format.
-Although this additional pre-processing step is only required as a one-off for each input file, it requires an additional 13mins and 13secs.
+rendering our approach almost an order of magnitude faster. 
+R on the other hand takes 67 minutes and Python 92 minutes.
+This substantially different runtime for the pre-processing is due to the Python and R implementations not natively supporting multithreading while \variantSpark, ADAM and Hadoop using 8 CPUs.
 
@@
 The similarity between \variantSpark\ and ADAM is expected, as they both use {\sc Spark}'s MLlib k-means implementation. 
-The 30 second speed increase in \variantSpark\ is likely due us converting the VCF files to sparse vectors, whereas ADAM creates dense vectors.
-%Due to the sparse nature of these row-vectors, the sparse vector format is likely to reduce some of the initial overhead.
+The 30 second speed increase in \variantSpark\ is likely due us converting the VCF files to sparse vectors, whereas ADAM creates dense vectors making operations less efficient.
+
 We also compared the k-means clustering times to a multi-threaded approach in Python and R. 
+Although an order of magnitude slower than \variantSpark\ and ADAM, k-means clustering in both Python and R were faster than our Hadoop implementation; Python required 11min and 29sec, whereas R required 7min and 25sec.
+This is due the dataset (chromosome 22) being small enough to fit into memory on a single machine, which enables R and Python to store the output from each k-means iteration directly in memory whereas Hadoop is designed to write the output of each iteration to disk.
+This demonstrates the improvements of {\sc Spark} over Hadoop where datasets can be processed in memory, virtually eliminating the disk IO bottleneck on smaller datasets.
 
-Although an order of magnitude slower than \variantSpark\ and ADAM, k-means clustering in both Python and R were faster than our Hadoop implementation.
-Python required 11min and 29sec, whereas R required 7min and 25sec.
-This demonstrates one of Hadoop's disadvantages (one which {\sc Spark} dramatically improves on): the cost of disk IO.
-Where Hadoop's advantage is that it can scale to datasets that exceed memory size, in this example, the entire dataset can fit in memory on a single machine.
-So, where the k-means implemtations we use in R and Python store the output from each k-means iteration in memory, Hadoop writes the output of each iteration to disk, whether or not the amount of available memory is sufficient to store this data.
-
-This unnecessary disk IO forms a bottleneck that results in the longer run-times.
-{\sc Spark}, on the other hand, can cache the intermediate results in memory, virtually eliminating the disk IO bottleneck on smaller datasets.
-
-%Clustering the samples is also fastest in \variantSpark\, albeit by a smaller margin.
-The time required is 1min and 20secs compared to 1min and 52secs, in Variant{\sc Spark} and ADAM, respectively.
-Hadoop however, required 14mins and 23secs.
-%The similarity between Variant{\sc Spark} and ADAM is expected, as they both use {\sc Spark}'s MLlib k-means implementation.
-The 30 second speed increase in Variant{\sc Spark} is likely due to Variant{\sc Spark}'s pre-processing implementation converting VCF files to sparse vectors, whereas ADAM creates dense vectors.
-%Due to the sparse nature of these row-vectors, the sparse vector format is likely to reduce some of the initial overhead.
-
-%We also compared the k-means clustering times to a multi-threaded approach in Python and R.
-Although slower than the {\sc Spark} implementations in Variant{\sc Spark} and ADAM, k-means clustering in both Python and R were faster than our Hadoop implementation.
-Python required 11min and 29sec, whereas R required 7min and 25sec.
-This demonstrates one of Hadoop's disadvantages (one which {\sc Spark} dramatically improves on).
-Hadoop can scale to datasets that exceed memory size as wall as distributing work across nodes in a cluster.
-However, in this example, the entire dataset can fit in the memory on a single machine.
-The k-means implemtations
-%we use in R and Python cluster the in-memory row vectors.
-Hadoop, however, writes the output of each MapReduce iteration to disk.
-This unnecessary disk IO forms a bottleneck that results in the longer run-times.
-{\sc Spark}, on the other hand, can cache the intermediate results in memory, virtually eliminating the disk IO bottleneck on smaller datasets.
-
-%To summarise the total times for each big-data platform, Variant{\sc Spark} was 4mins and 18secs, ADAM was 14mins and 14secs (or 27mins and 53secs including the one-off conversion of VCF to ADAM) and our Hadoop implementation was 28mins and 45secs.
+To summarise, \variantSpark\ completes the whole variant clustering task in just over 4 minutes, which is 80\% faster than ADAM (28 mins), 85\% faster than Hadoop (29 min), 94\% faster than R (75 min), and 95\% faster than Python (103 min).
 
@@
 To demonstrate the scalability of {\sc Spark}, we continue to cluster the same individuals from the 1000 Genome Project, however, rather than clustering on a single chromosome, we cluster the individuals on different subsets of variants from their 22 autosomes, and finally every variant.
-Pre-processing the data takes approximately 12 minutes for a 20\% sample.
-For 40\% of the genome, the time required is about 19 minutes, and finally, to preprocess 100\% of the genome, the time required is about 40 minutes.
+Pre-processing the data takes approximately 12 minutes for 20\% of the genome, 19 minutes for 40\% and 40 minutes for 100\% of the genome, respectively.
 For each of these sample sizes, the memory required doesn't exceed a modest 2GB per executor, even for the entire autosome. Memory requirements for clustering, however, increase by a

### 2015-08-18 doc/publications/journal2014/paper.tex
more methods
more updates

@@
 Pre-processing the data takes approximately 12 minutes for 20\% of the genome, 19 minutes for 40\% and 40 minutes for 100\% of the genome, respectively.
-For each of these sample sizes, the memory required doesn't exceed a modest 2GB per executor, even for the entire autosome. Memory requirements for clustering, however, increase by a
+For each of these sample sizes, the memory required doesn't exceed a modest 2GB per executor, even for the entire autosome.
+Memory requirements for clustering, however, increase in correlation with the dataset.
 
 Where our previous clustering of chromosome 22 required no more than 4GB memory per executor, clustering the entire genome increases this requirement to over 20GB.
+This scales in a linear trend, where 20\% of the genome requires 6GB and 40\% of the genome requires 12GB.
+To cluster 100\% of the genome, we need to allocate
+at least 24GB to each executor. 
 One factor which results in these differing memory requirements is the sparsity of the vectors used to store the variants.
@@
 For each of the implementations, we pre-process the VCF files to arrays of features.
-In /VariantSpark/ we read in the VCF files as text files to a Resilient Data Set (RDD).
+In \variantSpark\ we read in the VCF files as text files to a Resilient Data Set (RDD).
 RDDs allow us to process the files, line-by-line, in parallel.
@@
 
- For our R implementation, we read in the
+For our R implementation, we read in the VCF file as a table and convert it to a matrix, dropping the columns that don't represent alleles.
+As with our \variantSpark\ pre-processing, we convert the strings that represent each allele to a numeric value.
+We then transpose the matrix, resulting in a data-structure where each row represents an individual.
+
+Our Python implementation reads in a VCF file as tab-separated values, and stores it as a pandas dataframe. We convert the allele strings to numeric values and then transpose the dataframe.
 

### 2015-08-19 doc/publications/journal2014/paper.tex
added table2

@@
 
+\label{scalingcomparison}
+\begin{table}[h!]
+\caption{The resources consumption on different subsets of the entire autosome (chromosomes 1-22). Memory specified is memory allocated to each executor.}
+      \begin{tabular}{|c|ccc|ccc|c|c|}
+        \hline
+           Portion &  \multicolumn{3}{ c |}{Pre-processing} & \multicolumn{3}{ c |}{Clustering}  \\
+& executors & memory & time  & executors & memory & time \\
+  \hline
+20\%		& 64	& 2	& 11min 53sec	& 64	& 6	& 1hr 10min	\\
+40\%		& 64	& 2	& 19min 9sec	& 64	& 12	& 2hr 19min	\\
+60\%		& 64	& 2	& 67min		& x	& 32	& x	\\
+100\%	& 64	& 2	& 40min 48sec	& x	& 24	& 14hr 44min	\\
+  \hline
+      \end{tabular}
+\end{table}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2015-08-19 doc/publications/journal2014/paper.tex
some more methods

@@
 However, to cluster these vectors, the k-means algorithm converts them to Dense Vectors.
-With Dense Vectors, all the 0 values (which are excluded from Sparse Vectors so don't require any memory) are explicitly defined, considerably increasing the memory required.
+With Dense Vectors, the zero values (which are excluded from Sparse Vectors so don't require any memory) are explicitly defined, considerably increasing the memory required.
 
@@
 For our ADAM comparison, we followed the ADAM implementation from (http://bdgenomics.org/blog/2015/02/02/scalable-genomes-clustering-with-adam-and-spark/).
+
+We used our previous Hadoop implementation based on the MapReduce model. This makes use of KVPs, similar to \variantSpark\ however, it is explicitly based on the MapReduce model.
+The Map stage creates keys from the VCF file, where the key is a tuple of a primary and secondary key.
+The primary key is the individual ID and the secondary key is the allele ID.
+The value is the variant.
+The primary key ensures that KVPs for each individual are distributed to the same node during the MapReduce shuffle stage. The KVPs for each individual are then sorted by their secondary key.
+Finally, the Reduce stage creates a sparse vector for each individual from the individual's KVPs.
 
@@
 \begin{table}[h!]
-\caption{The resources consumption on different subsets of the entire autosome (chromosomes 1-22). Memory specified is memory allocated to each executor.}
+\caption{The resources consumption on different subsets of the entire autosome (chromosomes 1-22). Memory specified is the memory allocated to each executor.}
       \begin{tabular}{|c|ccc|ccc|c|c|}
@@
 40\%		& 64	& 2	& 19min 9sec	& 64	& 12	& 2hr 19min	\\
-60\%		& 64	& 2	& 67min		& x	& 32	& x	\\
-100\%	& 64	& 2	& 40min 48sec	& x	& 24	& 14hr 44min	\\
+60\%		& 64	& 2	& 26min 34sec	& 64	& 17	& 3hr 33min	\\
+100\%	& 64	& 2	& 40min 48sec	& 40	& 24	& 14hr 44min	\\
   \hline

### 2015-09-02 doc/publications/journal2014/paper.tex
updated methods

@@
 %\RequirePackage{hyperref}
+\PassOptionsToPackage{hyphens}{url}\usepackage{hyperref}
 \usepackage[utf8]{inputenc} %unicode support
@@
 
-\section*{Results}
+\section*{Results and discussions}
 \subsection*{{\sc Spark} enables faster clustering of individuals compared to traditional methods}
 
-In this section we compare the time required to cluster individuals based on genomic variants using \variantSpark\ against ADAM as well as the more traditional approaches using Hadoop Mahout, Python and R.
+In this section we compare the time required to cluster individuals based on genomic variants using \variantSpark{} against ADAM as well as the more traditional approaches using Hadoop Mahout, Python and R.
 We limit the genomic variants to only chromosome 22 as the traditional approaches have substantially larger memory consumption rendering a whole-genome input infeasible.  
@@
 
-As shown in Table~\ref{fivewaycomparison}, pre-processing the data is fastest in \variantSpark\, requiring 2mins and 58secs. %178s 
+As shown in Table~\ref{fivewaycomparison}, pre-processing the data is fastest in \variantSpark{}, requiring 2mins and 58secs. %178s 
 This is almost 80\% faster than the ADAM implementation (12min and 48secs) %768s
 or our Hadoop implementation(14min and 22secs). %862s
-Unlike \variantSpark\ or our Hadoop implementation, which pre-process VCF files directly, ADAM requires the input to be in their ADAM file format. 
+Unlike \variantSpark{} or our Hadoop implementation, which pre-process VCF files directly, ADAM requires the input to be in their ADAM file format. 
 VCF files must hence be converted to this binary format before they can be pre-processed into the row vector format. 
@@
 R on the other hand takes 67 minutes and Python 92 minutes.
-This substantially different runtime for the pre-processing is due to the Python and R implementations not natively supporting multithreading while \variantSpark, ADAM and Hadoop using 8 CPUs.
+This substantially different runtime for the pre-processing is due to the Python and R implementations not natively supporting multithreading while \variantSpark{}, ADAM and Hadoop using 8 CPUs.
 
-Clustering the samples 30\% fastest in \variantSpark\ compared to ADAM (1min and 20secs compared to 1min and 52secs) and  %80 vs 112
+Clustering the samples 30\% fastest in \variantSpark{} compared to ADAM (1min and 20secs compared to 1min and 52secs) and  %80 vs 112
 90\% faster than Hadoop (14mins and 23secs), see Table~\ref{fivewaycomparison}.
-The similarity between \variantSpark\ and ADAM is expected, as they both use {\sc Spark}'s MLlib k-means implementation. 
-The 30 second speed increase in \variantSpark\ is likely due us converting the VCF files to sparse vectors, whereas ADAM creates dense vectors making operations less efficient.
+The similarity between \variantSpark{} and ADAM is expected, as they both use {\sc Spark}'s MLlib k-means implementation. 
+The 30 second speed increase in \variantSpark{} is likely due us converting the VCF files to sparse vectors, whereas ADAM creates dense vectors making operations less efficient.
 
 We also compared the k-means clustering times to a multi-threaded approach in Python and R. 
-Although an order of magnitude slower than \variantSpark\ and ADAM, k-means clustering in both Python and R were faster than our Hadoop implementation; Python required 11min and 29sec, whereas R required 7min and 25sec.
+Although an order of magnitude slower than \variantSpark{} and ADAM, k-means clustering in both Python and R were faster than our Hadoop implementation; Python required 11min and 29sec, whereas R required 7min and 25sec.
 This is due the dataset (chromosome 22) being small enough to fit into memory on a single machine, which enables R and Python to store the output from each k-means iteration directly in memory whereas Hadoop is designed to write the output of each iteration to disk.
@@
 
-To summarise, \variantSpark\ completes the whole variant clustering task in just over 4 minutes, which is 80\% faster than ADAM (28 mins), 85\% faster than Hadoop (29 min), 94\% faster than R (75 min), and 95\% faster than Python (103 min).
+To summarise, \variantSpark{} completes the whole variant clustering task in just over 4 minutes, which is 80\% faster than ADAM (28 mins), 85\% faster than Hadoop (29 min), 94\% faster than R (75 min), and 95\% faster than Python (103 min).
 
-\subsection*{{\sc VariantSpark} allows genome wide sampling of variants to improve clustering quality}
+\subsection*{\variantSpark{} allows genome wide sampling of variants to improve clustering quality}
 
 To demonstrate the scalability of {\sc Spark}, we continue to cluster the same individuals from the 1000 Genome Project, however, rather than clustering on a single chromosome, we cluster the individuals on different subsets of variants from their 22 autosomes, and finally every variant.
-Pre-processing the data takes approximately 12 minutes for 20\% of the genome, 19 minutes for 40\% and 40 minutes for 100\% of the genome, respectively.
-For each of these sample sizes, the memory required doesn't exceed a modest 2GB per executor, even for the entire autosome.
+Pre-processing the data takes approximately 12 minutes, 19 minutes and 40 minutes for 20\%, 40\% and 100\% of the genome, respectively.
+For each of these sample sizes, the memory required doesn't exceed a modest 2GB per executor, even for all 22 autosomes.
 Memory requirements for clustering, however, increase in correlation with the dataset.
 
-Where our previous clustering of chromosome 22 required no more than 4GB memory per executor, clustering the entire genome increases this requirement to over 20GB.
+Where our previous clustering of chromosome 22 required no more than 4GB memory per executor, when clustering the entire genome, this increases this requirement to over 20GB.
 This scales in a linear trend, where 20\% of the genome requires 6GB and 40\% of the genome requires 12GB.
@@
 
-\subsection*{Personal genomics project}
+\subsection*{Personal Genome Project}
+To demonstrate the versatility of \variantSpark{} we also process and cluster individuals from the Personal Genomics Project. Unlike the 1000 genomes project, this data is open
 
@@
 
-\section*{Methods}
-\subsection*{Implementation}
+\section*{Conclusions}
+
+\section*{Materials and methods}
+\subsection*{Computational resources}
+We completed the Chromosome 22 comparisons on a virtual machine (VM) hosted on Microsoft Azure. This VM is an A7 Linux instance with 8 cores, 56GB RAM and Ubuntu installed as the OS. 
+For the whole-genome clustering, we used our in-house Hadoop cluster. This uses Hadoop 2.5.0, managed by CDH 5. We use Spark 1.3.1. This 13 node cluster has a total of 416 cores and 1.22TB memory.
+
+\subsection*{VariantSpark Implementation}
 For each of the implementations, we pre-process the VCF files to arrays of features.
-In \variantSpark\ we read in the VCF files as text files to a Resilient Data Set (RDD).
+In \variantSpark{} we read in the VCF files as text files to a Resilient Data Set (RDD).
 RDDs allow us to process the files, line-by-line, in parallel.
-We parse each line as tab-separated values, and "zip" each value with it's heading from the VCF file.
-The key-value pairs (KVPs) from each line in the VCF file are stored in separate arrays. We then "zip" each array with a unique index, where these indexes form identifiers for the alleles.
+We parse each line as tab-separated values, and `zip' each value with it's heading from the VCF file.
+The key-value pairs (KVPs) from each line in the VCF file are stored in separate arrays. We then `zip' each array with a unique index, where these indexes form identifiers for the alleles.
 At this point, we have one array for each line from the VCF file.
 The values in the arrays are KVPs of the parsed tab-separated values zipped with their heading (individual ID), and each array has a unique index.
-We now use a "flatMap" to convert these indexed arrays of KVPs into a less convoluted data-structure, as well as to filter out KVPs that aren't alleles (i.e. KVPs that are VCF metadata).
+We now use a `flatMap' to convert these indexed arrays of KVPs into a less convoluted data-structure, as well as to filter out KVPs that aren't alleles (i.e. KVPs that are VCF metadata).
 This results in KVPs where the key is the individual ID, and the value is a tuple of the allele and it's index.
@@
 At this point, each allele is still a string straight from the VCF file (i.e.
-"0\textbar1:13"), however, we require doubles.
+`0\textbar1'), however, we require doubles.
 To convert this to a double, we take the Hamming distance of the allele, where a 1 represents a heterozygous variant, 2 a homozygous variant, and 0, no variant.
@@
 
-For our ADAM comparison, we followed the ADAM implementation from (http://bdgenomics.org/blog/2015/02/02/scalable-genomes-clustering-with-adam-and-spark/).
+\subsection*{ADAM Implementation}
+For our ADAM comparison, we followed the ADAM implementation from (\url{http://bdgenomics.org/blog/2015/02/02/scalable-genomes-clustering-with-adam-and-spark/}).
 
-We used our previous Hadoop implementation based on the MapReduce model. This makes use of KVPs, similar to \variantSpark\ however, it is explicitly based on the MapReduce model.
-The Map stage creates keys from the VCF file, where the key is a tuple of a primary and secondary key.
-The primary key is the individual ID and the secondary key is the allele ID.
-The value is the variant.
-The primary key ensures that KVPs for each individual are distributed to the same node during the MapReduce shuffle stage. The KVPs for each individual are then sorted by their secondary key.
-Finally, the Reduce stage creates a sparse vector for each individual from the individual's KVPs.
+\subsection*{Hadoop Implementation}
+We use our previous Hadoop implementation, which is based on the MapReduce model. This makes use of KVPs, similarly to \variantSpark{}, however, it is explicitly based on the MapReduce model.
+Because of this, and because we need a unique range of identifiers for alleles (in the smallest range possible), we need to run an initial MapReduce task to index the lines.
+This is comparable (however, more cumbersome) to the `zip' operation in \variantSpark{}.
+The second MapReduce task does the bulk of the work.
+The Map stage begins by creating KVPs from the VCF file.
+For each KVP, the key is a tuple of a primary and secondary key, where the primary key is the individual ID and the secondary key is the allele ID.
+The value for each KVP is the variant.
+The primary key ensures that KVPs for each individual are distributed to the same node during the MapReduce shuffle stage.
+After being distributed, the KVPs for each individual are sorted by their secondary key.
+Now that the KVPs for each individual are physically located on the same hardware, the Reduce stage can efficiently create a sparse vector for each individual from these KVPs.
+The sparse vectors (one for each individual) are saved to disk, ready to be used as the input to one of MLLibs algorithms.
 
+\subsection*{R Implementation}
 For our R implementation, we read in the VCF file as a table and convert it to a matrix, dropping the columns that don't represent alleles.
-As with our \variantSpark\ pre-processing, we convert the strings that represent each allele to a numeric value.
+As with our \variantSpark{} pre-processing, we convert the strings that represent each allele to a numeric value.
 We then transpose the matrix, resulting in a data-structure where each row represents an individual.
 
+\subsection*{Python Implementation}
 Our Python implementation reads in a VCF file as tab-separated values, and stores it as a pandas dataframe. We convert the allele strings to numeric values and then transpose the dataframe.
+
+%% TODO %%
+\subsection*{Personal Genome Project VCF making}
+The PGP data.....
 
@@
   \hline
-\variantSpark\	& 8	& 32	& 2min 58sec	& 8	& 32	& 1min 20sec	& x	\\ 
+\variantSpark{}	& 8	& 32	& 2min 58sec	& 8	& 32	& 1min 20sec	& 0.84	\\ 
 ADAM		& 8	& 32	& 12min 48sec	& 8	& 32	& 1min 52sec	& x	\\
@@
 R			& 1	& 32	& 67min		& 8	& 32	& 7min 25sec	& x	\\
-Python		& 1	& 32	& 92min		& 8	& 32	& 11min 29sec	& x	\\
+Python		& 1	& 32	& 92min		& 8	& 32	& 11min 29sec	& 0.84	\\
   \hline

### 2015-09-03 doc/publications/journal2014/paper.tex
comparison

@@
 The similarity between \variantSpark{} and ADAM is expected, as they both use {\sc Spark}'s MLlib k-means implementation. 
-The 30 second speed increase in \variantSpark{} is likely due us converting the VCF files to sparse vectors, whereas ADAM creates dense vectors making operations less efficient.
+The 30 second speed increase in \variantSpark{} is likely due to us converting the VCF files to sparse vectors, whereas ADAM creates dense vectors which are less memory efficient.
 
-We also compared the k-means clustering times to a multi-threaded approach in Python and R. 
-Although an order of magnitude slower than \variantSpark{} and ADAM, k-means clustering in both Python and R were faster than our Hadoop implementation; Python required 11min and 29sec, whereas R required 7min and 25sec.
-This is due the dataset (chromosome 22) being small enough to fit into memory on a single machine, which enables R and Python to store the output from each k-means iteration directly in memory whereas Hadoop is designed to write the output of each iteration to disk.
-This demonstrates the improvements of {\sc Spark} over Hadoop where datasets can be processed in memory, virtually eliminating the disk IO bottleneck on smaller datasets.
+To demonstrate {\sc Spark}'s advantages over parallel approaches to k-means clustering in existing libraries, we compare the clustering libraries used in {\variantSpark}, to those in Python and R. 
+In regards to speed, clustering using Python and R were both an order of magnitude slower than \variantSpark{} and ADAM, however both were faster than our Hadoop implementation, where Python required 11min and 29sec and R required 7min and 25sec.
+This is due the dataset (chromosome 22) being small enough to fit into memory on a single machine.
+This enables Python and R to store the output from each k-means iteration directly in memory.
+Hadoop, however, is designed to write the entire output of each iteration to disk.
+Although this feature allows for greater scalability, as memory availability is less of an issue, the bottleneck introduced from disk IO increases processing time notably.
+This demonstrates one of the major improvements of {\sc Spark} over Hadoop, where {\sc Spark} can cache datasets in memory.
+This virtually eliminates the disk IO bottleneck on smaller datasets, or at least reduces it on larger datasets where only a portion of the data can be cached in memory.
+
+For smaller datasets, we've demonstrated {\sc Spark} to be faster than it's single-machine counterparts, however it's also important that this speedup isn't at the cost of quality.
+The 1000 Genome data is ideal for investigating the clustering quality, as the consortium provides metadata which lists the super-population that each individual belongs to.
+This knowledge allows us to create a `truth' clustering, where individuals are clustered by their super-population.
+With the four super-populations (AMR, AFR, EAS, SAS), we can divide the individuals up into four clusters based on these labels. We can then compare this truth clustering with the k-means clusters.
+For this comparison, we use the Adjusted Rand Index (ARI) metric. To generate the ARI, an algorithm takes two separate clusterings and returns a value based on their similarity.
+Two identical clusterings will result in a 1, whereas 0 represents two clusterings that do not agree on any points (individuals).
+On Chromosome 22, each or the algorithms resulted in an ARI of 0.84.
+Although not perfect, this demonstrates that the speed-up in spark isn't at the cost of quality, when compared to R and Python.
+A result of 1 is probably unlikely because, as mentioned by 10.15252/embr.201439469, although differences in the distribution of alleles exist between populations, the boundaries are rather diffuse.
+%TODO - proper citation
 
@@
 \section*{Conclusions}
+By using the k-means to cluster the data, we can compare the results with
 
@@
 \subsection*{Computational resources}
-We completed the Chromosome 22 comparisons on a virtual machine (VM) hosted on Microsoft Azure. This VM is an A7 Linux instance with 8 cores, 56GB RAM and Ubuntu installed as the OS. 
+We completed the Chromosome 22 comparisons on a virtual machine (VM) hosted on Microsoft Azure. This VM is an A7 Linux instance with 8 cores, 56GB memory and Ubuntu as the OS. 
 For the whole-genome clustering, we used our in-house Hadoop cluster. This uses Hadoop 2.5.0, managed by CDH 5. We use Spark 1.3.1. This 13 node cluster has a total of 416 cores and 1.22TB memory.
@@
 At this point, each allele is still a string straight from the VCF file (i.e.
-`0\textbar1'), however, we require doubles.
+`0\textbar1'), however, we require doubles for MLLib.
 To convert this to a double, we take the Hamming distance of the allele, where a 1 represents a heterozygous variant, 2 a homozygous variant, and 0, no variant.
@@
 
-%% TODO %%
+%% TODO (and better heading) %%
 \subsection*{Personal Genome Project VCF making}

### 2015-09-03 doc/publications/journal2014/paper.tex
more results

@@
 \begin{abstract} % abstract
-\parttitle{Motivation}: Genomic information is increasingly used in medical practice, giving rise to the need for efficient analysis methodology able to cope with thousands of individuals and millions of variants.
+\parttitle{Motivation} Genomic information is increasingly used in medical practice, giving rise to the need for efficient analysis methodology able to cope with thousands of individuals and millions of variants.
 The widely used Hadoop MapReduce architecture and associated machine-learning library, Mahout, provide the means for tackling computationally challenging tasks.
@@
 
-\parttitle{Results}: We successfully clustered more than 2,500 individuals each having more than 38 Million variants.
+\parttitle{Results} We successfully clustered more than 2,500 individuals each having more than 38 Million variants.
 Both Hadoop and {\sc Spark} have superior performance over traditional approaches.
@@
 
-\parttitle{Availability}: The package is written in Scala and available at https: //github.com/BauerLab/VariantSpark.
+\parttitle{Availability} The package is written in Scala and available at \url{https://github.com/BauerLab/VariantSpark}.
 \end{abstract}
@@
 \section*{Results and discussions}
+
 \subsection*{{\sc Spark} enables faster clustering of individuals compared to traditional methods}
@@
 
-To summarise, \variantSpark{} completes the whole variant clustering task in just over 4 minutes, which is 80\% faster than ADAM (28 mins), 85\% faster than Hadoop (29 min), 94\% faster than R (75 min), and 95\% faster than Python (103 min).
-
 \subsection*{\variantSpark{} allows genome wide sampling of variants to improve clustering quality}
 
-To demonstrate the scalability of {\sc Spark}, we continue to cluster the same individuals from the 1000 Genome Project, however, rather than clustering on a single chromosome, we cluster the individuals on different subsets of variants from their 22 autosomes, and finally every variant.
-Pre-processing the data takes approximately 12 minutes, 19 minutes and 40 minutes for 20\%, 40\% and 100\% of the genome, respectively.
-For each of these sample sizes, the memory required doesn't exceed a modest 2GB per executor, even for all 22 autosomes.
-Memory requirements for clustering, however, increase in correlation with the dataset.
+Although we've demonstrated the advantages of \variantSpark{} with smaller datasets, the main advantage of {\sc Spark} (and hense \variantSpark) is it's scalability and ability to process datasets that exceed memory limits.
+To demonstrate the scalability, we continue to cluster the same individuals from the 1000 Genome Project, however, rather than clustering on a single chromosome, we cluster the individuals on different subsets of variants from their 22 autosomes, and finally every variant.
+We run this comparison on our in-house Hadoop cluster. Pre-processing the VCF files takes approximately 12 minutes, 19 minutes and 40 minutes for 20\%, 40\% and 100\% of the genome, respectively.
+In each case, the memory required doesn't exceed a modest 2GB per executor, even for 100\% of the genome. Memory requirements for clustering, however, increase in correlation with the dataset.
 
-Where our previous clustering of chromosome 22 required no more than 4GB memory per executor, when clustering the entire genome, this increases this requirement to over 20GB.
-This scales in a linear trend, where 20\% of the genome requires 6GB and 40\% of the genome requires 12GB.
-To cluster 100\% of the genome, we need to allocate
-at least 24GB to each executor. 
+Where our previous clustering of chromosome 22 required no more than 4GB memory per executor, scaling to 100\% of the genome increases the requirement to more than 24GB.
+This scales in an approximate linear trend, where 20\% of the genome requires 6GB and 40\% of the genome requires 12GB.
+24GB per executor for k-means clustering makes it much more resource intensive than pre-processing's 2GB per executor. 
 One factor which results in these differing memory requirements is the sparsity of the vectors used to store the variants.
-Due to their sparse nature (the abundance of 0s) we can store them efficiently as Sparse Vectors, rather than Dense Vectors.
-However, to cluster these vectors, the k-means algorithm converts them to Dense Vectors.
-With Dense Vectors, the zero values (which are excluded from Sparse Vectors so don't require any memory) are explicitly defined, considerably increasing the memory required.
+Due to their sparse nature (the abundance of 0s) we can store them efficiently as Sparse Vectors in our pre-processing stage.
+However, to cluster these vectors, the k-means algorithm converts these sparse vectors to dense vectors.
+This is much less memory-efficient, as dense vectors explicitly define zero values (which are excluded from Sparse Vectors so don't require any memory), considerably increasing the memory required.
+So although we demonstrate the potential of k-means clustering to scale to 100\% of the genome, machine learning algorithms that can deal with sparse vectors would potentially be able to scale much further, or for example, easily deal with more alleles.
 
@@
 \section*{Conclusions}
-By using the k-means to cluster the data, we can compare the results with
+\variantSpark{} 
+To summarise, \variantSpark{} completes the whole variant clustering task in just over 4 minutes, which is 80\% faster than ADAM (28 mins), 85\% faster than Hadoop (29 min), 94\% faster than R (75 min), and 95\% faster than Python (103 min).
 

### 2015-09-03 doc/publications/journal2014/paper.tex
more stuff

@@
 
+\subsection*{Datasets}
+We used  phase 1 variants from the 1000 Genomes Project. On the 22 autosomes, this dataset contains variants for 1092 individuals, spread over a total 38,219,238 alleles.
+We use chromosome 22 for some comparisons, and this contains variants over 494,328 alleles.
+These individuals from these datasets are distributed across four super populations, African (AFR), Mixed American (AMR), East Asian (EAS) and European (EUR).
+
 \subsection*{VariantSpark Implementation}

### 2015-09-04 doc/publications/journal2014/paper.tex
section for neil

@@
 \subsection*{R Implementation}
-For our R implementation, we read in the VCF file as a table and convert it to a matrix, dropping the columns that don't represent alleles.
+For our R implementation, we read in a VCF file as a table and convert it to a matrix, dropping the columns that don't represent alleles.
 As with our \variantSpark{} pre-processing, we convert the strings that represent each allele to a numeric value.
@@
 \subsection*{Python Implementation}
-Our Python implementation reads in a VCF file as tab-separated values, and stores it as a pandas dataframe. We convert the allele strings to numeric values and then transpose the dataframe.
+Our Python implementation reads in lines from a VCF file as tab-separated values and stores the lines in a pandas DataFrame.
+The column headings are the individual IDs and the row headings are the allele locations.
+We drop the first columns that aren't alleles, and then convert the remaining allele strings to numeric values.
+We can then transpose the DataFrame.
+Although to cluster this DataFrame (using sci-kit learn), we simply need to convert it to a matrix with {\sc .as\textunderscore{}matrix()}
 
-%% TODO (and better heading) %%
+%% TODO - NEIL - (and better heading) %%
 \subsection*{Personal Genome Project VCF making}

### 2015-09-07 doc/publications/journal2014/paper.tex
cleanup and consistency

@@
 \newcommand{\variantSpark}{{\sc VariantSpark}}
+\newcommand{\kMeans}{\textit{k}-means}
 \newcommand{\ARI}{Adjusted Rand Index}
@@
 \parttitle{Motivation} Genomic information is increasingly used in medical practice, giving rise to the need for efficient analysis methodology able to cope with thousands of individuals and millions of variants.
-The widely used Hadoop MapReduce architecture and associated machine-learning library, Mahout, provide the means for tackling computationally challenging tasks.
+The widely used Hadoop MapReduce architecture and associated machine learning library, Mahout, provide the means for tackling computationally challenging tasks.
 However, many genomic analyses do not fit the Map-Reduce paradigm.
@@
 A commonly performed task in such applications is grouping individuals based on their genomic profile to identify population association~\cite{Gao2007} or elucidate different haplotype involvement in diseases susceptibility~\cite{Laitman2013}.
-Due to the decreasing sequencing cost it is economical to generate studies with sample sizes previously reserved for larger consortia such as the 1000 genomes project~\cite{1KG2012} or The Cancer Genome Atlas, TCGA~\cite{TCGA2013}.
+Due to the decreasing sequencing cost it is economical to generate studies with sample sizes previously reserved for larger consortia such as the 1000 Genomes Project~\cite{1KG2012} or The Cancer Genome Atlas, TCGA~\cite{TCGA2013}.
 At the same time, whole genome sequencing enables the inclusion of rare or even somatic mutations in the analysis, increasing the feature space by orders of magnitude.
@@
 
-Addressing this issue, {\sc Apache Hadoop MapReduce}~\cite{Borthakur2007} transforms data into 'key-value pairs' that can then be distributed between multiple nodes across a commodity compute-cluster depending on the size of the problem.
+Addressing this issue, {\sc Apache Hadoop MapReduce}~\cite{Borthakur2007} transforms data into `key-value pairs' that can then be distributed between multiple nodes across a commodity compute-cluster depending on the size of the problem.
 MapReduce approaches are increasingly being used in bioinformatics (for reviews see~\cite{Zou2013, Qiu2010,Taylor2010}). 
@@
 Unfortunately, the MapReduce paradigm is not always the optimal solution, specifically for bioinformatics or machine learning applications that require iterative in-memory computation.
-In addition, Hadoop is disk IO intensive, and this can prove to be a bottleneck in processing-speed.
+Specifically, Hadoop is disk IO intensive, and this can prove to be a bottleneck in processing-speed.
 
@@
 Although {\sc Spark} allows MapReduce-like programs, it does not require programs to exactly model the MapReduce paradigm, which in turn allows more flexibility when designing programs. 
-Recognising the capability, the Big Data Genomics (BDG) group has recently demonstrated the strength of {\sc Spark} in a genomic clustering application using ADAM, a set of formats and APIs as well as processing stage implementations for genomic data~\cite{Massie2013}.
+Recognising the capability, the Big Data Genomics (BDG) group recently demonstrated the strength of {\sc Spark} in a genomic clustering application using ADAM, a set of formats and APIs as well as processing stage implementations for genomic data~\cite{Massie2013}.
 While the speedup over traditional methods was impressive, being limited by constraints within this general genomics framework hampered performance.
@@
 We hence developed a more focused purpose-built application in {\sc Spark} to perform genomic clustering of individuals. 
-We utilise {\sc Spark}'s machine-learning library, MLlib, and provide an interface from the standard variant data format, variant call format (VCF)~\cite{1KG2012}, which opens up the application of MLlib's different machine learning algorithms a wide range of genotype-based analysis tasks.
-To demonstrate its capability, we cluster variant datasets from the 1000 genomes project~\cite{1KG2012} to determine population structure using the k-means clustering algorithm available in MLlib's. 
+We utilise {\sc Spark}'s machine learning library, \mbox{MLlib}, and provide an interface from the standard variant data format, Variant Call Format (VCF)~\cite{1KG2012}, which opens up the application of MLlib's different machine learning algorithms a wide range of genotype-based analysis tasks.
+To demonstrate its capability, we cluster variant datasets from the 1000 Genomes Project~\cite{1KG2012} to determine population structure using the \kMeans{} clustering algorithm available in MLlib's.
 In the first section we benchmark the performance and accuracy of MLlib's implementations against Hadoop's Mahout as well as more traditional methods (R, Python) limiting the used data to chromosome 22.
@@
 
-In this section we compare the time required to cluster individuals based on genomic variants using \variantSpark{} against ADAM as well as the more traditional approaches using Hadoop Mahout, Python and R.
+In this section we compare the time required to cluster individuals based on genomic variants using \variantSpark{} against ADAM as well as the more traditional approaches using Hadoop (and Mahout), Python and R.
 We limit the genomic variants to only chromosome 22 as the traditional approaches have substantially larger memory consumption rendering a whole-genome input infeasible.  
-Furthermore, we perform the comparisons on a single virtual-machine to ensure the five different approaches have access to the same resources.
-We use k-means clustering algorithm in the respective implementations, which all require the VCF input files to be pre-processed (see methods). 
-We therefore compare the time required for the pre-processing step, as well as the time required for k-means clustering.
+Furthermore, we perform the comparisons on a single virtual machine to ensure the five different approaches have access to the same resources.
+We use \kMeans{} clustering algorithm in the respective implementations, which all require the VCF input files to be pre-processed (see methods). 
+We therefore compare the time required for the pre-processing step, as well as the time required for \kMeans{} clustering.
 
@@
 This is almost 80\% faster than the ADAM implementation (12min and 48secs) %768s
-or our Hadoop implementation(14min and 22secs). %862s
-Unlike \variantSpark{} or our Hadoop implementation, which pre-process VCF files directly, ADAM requires the input to be in their ADAM file format. 
-VCF files must hence be converted to this binary format before they can be pre-processed into the row vector format. 
-Although this additional pre-processing step is only required as a one-off for each input file, it requires an additional 13mins and 13secs %793s
+or our Hadoop implementation (14min and 22secs). %862s
+Unlike \variantSpark{} or our Hadoop implementation, which pre-process VCF files directly, ADAM requires the input to be in their ADAM file format, and VCF files must be converted to this binary format.
+Although this additional pre-processing step is only required as a one-off for each input file, it requires an additional 13mins and 13secs, %793s
 rendering our approach almost an order of magnitude faster. 
-R on the other hand takes 67 minutes and Python 92 minutes.
-This substantially different runtime for the pre-processing is due to the Python and R implementations not natively supporting multithreading while \variantSpark{}, ADAM and Hadoop using 8 CPUs.
+R on the other hand takes 67 minutes, and Python 92 minutes.
+This substantially different runtime for the pre-processing is due to the Python and R implementations not natively supporting multithreading while \variantSpark{}, ADAM and Hadoop are using 8 cores.
 
-Clustering the samples 30\% fastest in \variantSpark{} compared to ADAM (1min and 20secs compared to 1min and 52secs) and  %80 vs 112
+Clustering the samples is fastest in \variantSpark{} (1min and 20secs), being 30\% faster than ADAM (1min and 52secs) and  %80 vs 112
 90\% faster than Hadoop (14mins and 23secs), see Table~\ref{fivewaycomparison}.
-The similarity between \variantSpark{} and ADAM is expected, as they both use {\sc Spark}'s MLlib k-means implementation. 
+The similarity between \variantSpark{} and ADAM is expected, as they both use {\sc Spark}'s MLlib \kMeans{} implementation. 
 The 30 second speed increase in \variantSpark{} is likely due to us converting the VCF files to sparse vectors, whereas ADAM creates dense vectors which are less memory efficient.
 
-To demonstrate {\sc Spark}'s advantages over parallel approaches to k-means clustering in existing libraries, we compare the clustering libraries used in {\variantSpark}, to those in Python and R. 
-In regards to speed, clustering using Python and R were both an order of magnitude slower than \variantSpark{} and ADAM, however both were faster than our Hadoop implementation, where Python required 11min and 29sec and R required 7min and 25sec.
-This is due the dataset (chromosome 22) being small enough to fit into memory on a single machine.
-This enables Python and R to store the output from each k-means iteration directly in memory.
+To demonstrate {\sc Spark}'s advantages over parallel approaches to \kMeans{} clustering in existing libraries, we compare the clustering libraries used in {\variantSpark}, to those in Python and R.
+For this comparison, each implementation ran on 8 cores.
+In regards to speed, clustering using Python and R were both an order of magnitude slower than \variantSpark{} and ADAM, however both were faster than our Hadoop implementation.
+Python required 11mins and 29secs and R required 7mins and 25secs.
+The increased speed compared to Hadoop is due the dataset (chromosome 22) being small enough to fit into memory.
+This enables Python and R to store the output from each \kMeans{} iteration directly in memory.
 Hadoop, however, is designed to write the entire output of each iteration to disk.
-Although this feature allows for greater scalability, as memory availability is less of an issue, the bottleneck introduced from disk IO increases processing time notably.
-This demonstrates one of the major improvements of {\sc Spark} over Hadoop, where {\sc Spark} can cache datasets in memory.
-This virtually eliminates the disk IO bottleneck on smaller datasets, or at least reduces it on larger datasets where only a portion of the data can be cached in memory.
+Although this feature allows for greater scalability, as memory availability is less of an issue with huge datasets, the bottleneck introduced from disk IO increases processing time notably.
+This demonstrates one of the major improvements of {\sc Spark} over Hadoop, {\sc Spark}'s in-memory caching of datasets.
+Caching virtually eliminates the disk IO bottleneck on smaller datasets, and minimises the bottleneck on larger datasets where only a portion of the data can be cached in memory.
 
-For smaller datasets, we've demonstrated {\sc Spark} to be faster than it's single-machine counterparts, however it's also important that this speedup isn't at the cost of quality.
-The 1000 Genome data is ideal for investigating the clustering quality, as the consortium provides metadata which lists the super-population that each individual belongs to.
+For smaller datasets, we've demonstrated {\sc Spark} to be faster than its single-machine counterparts, however it's also important that this speedup isn't at the cost of quality.
+The 1000 Genomes data is ideal for investigating the clustering quality, as the consortium provides metadata which lists the super-population that each individual belongs to.
 This knowledge allows us to create a `truth' clustering, where individuals are clustered by their super-population.
-With the four super-populations (AMR, AFR, EAS, SAS), we can divide the individuals up into four clusters based on these labels. We can then compare this truth clustering with the k-means clusters.
+With the four super-populations (AMR, AFR, EAS, SAS), we can divide the individuals up into four clusters based on these labels. We can then compare this truth clustering with the \kMeans{} clusters.
 For this comparison, we use the Adjusted Rand Index (ARI) metric. To generate the ARI, an algorithm takes two separate clusterings and returns a value based on their similarity.
@@
 Although we've demonstrated the advantages of \variantSpark{} with smaller datasets, the main advantage of {\sc Spark} (and hense \variantSpark) is it's scalability and ability to process datasets that exceed memory limits.
-To demonstrate the scalability, we continue to cluster the same individuals from the 1000 Genome Project, however, rather than clustering on a single chromosome, we cluster the individuals on different subsets of variants from their 22 autosomes, and finally every variant.
+To demonstrate the scalability, we continue to cluster the same individuals from the 1000 Genomes Project, however, rather than clustering on a single chromosome, we cluster the individuals on different subsets of variants from their 22 autosomes, and finally every variant.
 We run this comparison on our in-house Hadoop cluster. Pre-processing the VCF files takes approximately 12 minutes, 19 minutes and 40 minutes for 20\%, 40\% and 100\% of the genome, respectively.
@@
 This scales in an approximate linear trend, where 20\% of the genome requires 6GB and 40\% of the genome requires 12GB.
-24GB per executor for k-means clustering makes it much more resource intensive than pre-processing's 2GB per executor. 
+24GB per executor for \kMeans{} clustering makes it much more resource intensive than pre-processing's 2GB per executor. 
 One factor which results in these differing memory requirements is the sparsity of the vectors used to store the variants.
 Due to their sparse nature (the abundance of 0s) we can store them efficiently as Sparse Vectors in our pre-processing stage.
-However, to cluster these vectors, the k-means algorithm converts these sparse vectors to dense vectors.
+However, to cluster these vectors, the \kMeans{} algorithm converts these sparse vectors to dense vectors.
 This is much less memory-efficient, as dense vectors explicitly define zero values (which are excluded from Sparse Vectors so don't require any memory), considerably increasing the memory required.
-So although we demonstrate the potential of k-means clustering to scale to 100\% of the genome, machine learning algorithms that can deal with sparse vectors would potentially be able to scale much further, or for example, easily deal with more alleles.
+So although we demonstrate the potential of \kMeans{} clustering to scale to 100\% of the genome, machine learning algorithms that can deal with sparse vectors would potentially be able to scale much further, or for example, easily deal with more alleles.
 
 \subsection*{Personal Genome Project}
-To demonstrate the versatility of \variantSpark{} we also process and cluster individuals from the Personal Genomics Project. Unlike the 1000 genomes project, this data is open
+To demonstrate the versatility of \variantSpark{} we also process and cluster individuals from the Personal Genomics Project. Unlike the 1000 Genomes Project, this data is open
 
@@
 We used  phase 1 variants from the 1000 Genomes Project. On the 22 autosomes, this dataset contains variants for 1092 individuals, spread over a total 38,219,238 alleles.
-We use chromosome 22 for some comparisons, and this contains variants over 494,328 alleles.
+We also used a subset of this dataset, the variants from chromosome 22, and this contains 494,328 alleles.
 These individuals from these datasets are distributed across four super populations, African (AFR), Mixed American (AMR), East Asian (EAS) and European (EUR).
@@
 At this point, each allele is still a string straight from the VCF file (i.e.
-`0\textbar1'), however, we require doubles for MLLib.
+`0\textbar{}1'), however, we require doubles for MLLib.
 To convert this to a double, we take the Hamming distance of the allele, where a 1 represents a heterozygous variant, 2 a homozygous variant, and 0, no variant.

### 2015-09-08 doc/publications/journal2014/paper.tex
more changes

@@
 \newcommand{\kMeans}{\textit{k}-means}
-\newcommand{\ARI}{Adjusted Rand Index}
+\newcommand{\ARI}{adjusted Rand index}
 
@@
 or our Hadoop implementation (14min and 22secs). %862s
-Unlike \variantSpark{} or our Hadoop implementation, which pre-process VCF files directly, ADAM requires the input to be in their ADAM file format, and VCF files must be converted to this binary format.
+Unlike \variantSpark{} or our Hadoop implementation, which pre-process VCF files directly, the ADAM framework requires the input to be in the ADAM file format, and VCF files must therefore be converted to this binary format.
 Although this additional pre-processing step is only required as a one-off for each input file, it requires an additional 13mins and 13secs, %793s
 rendering our approach almost an order of magnitude faster. 
-R on the other hand takes 67 minutes, and Python 92 minutes.
-This substantially different runtime for the pre-processing is due to the Python and R implementations not natively supporting multithreading while \variantSpark{}, ADAM and Hadoop are using 8 cores.
+R and Python are slower at pre-processing, taking 67 minutes and 92 minutes, respectively.
+This increased pre-processing time is due to the Python and R implementations not natively supporting multithreading, while \variantSpark{}, ADAM and Hadoop can use all eight available cores.
 
@@
 The increased speed compared to Hadoop is due the dataset (chromosome 22) being small enough to fit into memory.
-This enables Python and R to store the output from each \kMeans{} iteration directly in memory.
-Hadoop, however, is designed to write the entire output of each iteration to disk.
+This enables Python and R to store the output from each \kMeans{} iteration in memory.
+Hadoop, however, writes the entire output of each iteration to disk.
 Although this feature allows for greater scalability, as memory availability is less of an issue with huge datasets, the bottleneck introduced from disk IO increases processing time notably.
@@
 With the four super-populations (AMR, AFR, EAS, SAS), we can divide the individuals up into four clusters based on these labels. We can then compare this truth clustering with the \kMeans{} clusters.
-For this comparison, we use the Adjusted Rand Index (ARI) metric. To generate the ARI, an algorithm takes two separate clusterings and returns a value based on their similarity.
+For this comparison, we use the \ARI{} (ARI) metric. To generate the ARI, an algorithm takes two separate clusterings and returns a value based on their similarity.
 Two identical clusterings will result in a 1, whereas 0 represents two clusterings that do not agree on any points (individuals).
 On Chromosome 22, each or the algorithms resulted in an ARI of 0.84.
-Although not perfect, this demonstrates that the speed-up in spark isn't at the cost of quality, when compared to R and Python.
+Although not perfect, the identical results demonstrate that the speed-up in {\sc Spark} isn't at the cost of quality.
 A result of 1 is probably unlikely because, as mentioned by 10.15252/embr.201439469, although differences in the distribution of alleles exist between populations, the boundaries are rather diffuse.
@@
 \subsection*{Personal Genome Project}
-To demonstrate the versatility of \variantSpark{} we also process and cluster individuals from the Personal Genomics Project. Unlike the 1000 Genomes Project, this data is open
+To demonstrate the versatility of \variantSpark{} we also process and cluster individuals from the Personal Genomics Project (PGP).
+The PGP is open to the public for individuals to submit their genomic sequence along with any metadata, such as diseases or clinical features ~\cite{Lunshof2013}.
+The  VCF file we use contains 985,790 alleles.
+
+Clustering individuals leads to an ARI of 0.1. 
+Some observations are that the population groups to diverse in the PGP data.
+Each individual has metadata stating which country each of their grandparents are from, and only 40 individuals have all four grandparents from the same country.
+Also, being open to the public, the data may be more prone to error.
+For some individuals, the stated country of their grandparents changes between their updates.
+This could indicate some uncertainty about their origins, and hence, correct population.
+
+To try improve on these results, we only include the individuals with their grandparents were from the same country.
+We then grouped them into their approximate super-populations.
+Clustering these individuals results in an ARI of 0.29.
+Although still a far cry from the 1000 Genomes Project results, it's an improvement on the previous clustering, albeit with fewer individuals.
 
@@
 \section*{Conclusions}
-\variantSpark{} 
-To summarise, \variantSpark{} completes the whole variant clustering task in just over 4 minutes, which is 80\% faster than ADAM (28 mins), 85\% faster than Hadoop (29 min), 94\% faster than R (75 min), and 95\% faster than Python (103 min).
+\variantSpark{} excels at pre-processing VCF files into row vectors, being over an order of magnitude faster than ADAM.
+Memory consumption is also kept to a minimum.
+Where traditional approaches such as Python and R would fail when expanding to the whole genome,
+\variantSpark{} can pre-process VCF files with over 1000 individuals in under half an hour using minimal memory.
+
+Clustering these individuals is also much faster with \variantSpark{} than Python and R, although only marginally faster than ADAM.
+
+These benefits of speed, resource consumption and scalability demonstrate \variantSpark{} to be a useful tool in applying machine learning tools to genomic data.
+Although we only run \kMeans on the data, MLlib provides many other machine learning algorithms, such as Random Forests.
+Therefore it may be useful to investigate the novelty of applying these machine learning approaches to big genomic data.
+
+%completes the whole variant clustering task in just over 4 minutes, which is 80\% faster than ADAM (28 mins), 85\% faster than Hadoop (29 min), 94\% faster than R (75 min), and 95\% faster than Python (103 min).
 
@@
 At this point, each allele is still a string straight from the VCF file (i.e.
-`0\textbar{}1'), however, we require doubles for MLLib.
+`0\textbar{}1'), however, we require doubles for MLlib.
 To convert this to a double, we take the Hamming distance of the allele, where a 1 represents a heterozygous variant, 2 a homozygous variant, and 0, no variant.
@@
 We can now group these KVPs by their key (the individual ID) and map them to sparse vectors.
-This results in a sparse vector for each individual, the format required by MLLibs machine learning algorithms.
+This results in a sparse vector for each individual, the format required by MLlib's machine learning algorithms.
 
@@
 Now that the KVPs for each individual are physically located on the same hardware, the Reduce stage can efficiently create a sparse vector for each individual from these KVPs.
-The sparse vectors (one for each individual) are saved to disk, ready to be used as the input to one of MLLibs algorithms.
+The sparse vectors (one for each individual) are saved to disk, ready to be used as the input to one of MLlib's algorithms.
 

### 2015-09-10 doc/publications/journal2014/paper.tex
updated paper

@@
 or our Hadoop implementation (14min and 22secs). %862s
-Unlike \variantSpark{} or our Hadoop implementation, which pre-process VCF files directly, the ADAM framework requires the input to be in the ADAM file format, and VCF files must therefore be converted to this binary format.
-Although this additional pre-processing step is only required as a one-off for each input file, it requires an additional 13mins and 13secs, %793s
+Unlike \variantSpark{} and our Hadoop implementation, the ADAM framework cannot process the VCF files directly but requires them to be converted into a binary ADAM file format. 
+%Unlike \variantSpark{} or our Hadoop implementation, which pre-process VCF files directly, the ADAM framework requires the input to be in the ADAM file format, and VCF files must therefore be converted to this binary format.
+Although this additional pre-processing step is only required once for each input file, it requires an additional 13mins and 13secs, %793s
+rendering our approach almost an order of magnitude faster. 
+%Although this additional pre-processing step is only required as a one-off for each input file, it requires an additional 13mins and 13secs, %793s
 rendering our approach almost an order of magnitude faster. 
 R and Python are slower at pre-processing, taking 67 minutes and 92 minutes, respectively.
-This increased pre-processing time is due to the Python and R implementations not natively supporting multithreading, while \variantSpark{}, ADAM and Hadoop can use all eight available cores.
+This increased pre-processing time is due to the Python and R implementations not natively supporting multithreading, while \variantSpark{}, ADAM and Hadoop can use the eight available cores.
 
-Clustering the samples is fastest in \variantSpark{} (1min and 20secs), being 30\% faster than ADAM (1min and 52secs) and  %80 vs 112
-90\% faster than Hadoop (14mins and 23secs), see Table~\ref{fivewaycomparison}.
-The similarity between \variantSpark{} and ADAM is expected, as they both use {\sc Spark}'s MLlib \kMeans{} implementation. 
-The 30 second speed increase in \variantSpark{} is likely due to us converting the VCF files to sparse vectors, whereas ADAM creates dense vectors which are less memory efficient.
-
-To demonstrate {\sc Spark}'s advantages over parallel approaches to \kMeans{} clustering in existing libraries, we compare the clustering libraries used in {\variantSpark}, to those in Python and R.
-For this comparison, each implementation ran on 8 cores.
-In regards to speed, clustering using Python and R were both an order of magnitude slower than \variantSpark{} and ADAM, however both were faster than our Hadoop implementation.
-Python required 11mins and 29secs and R required 7mins and 25secs.
-The increased speed compared to Hadoop is due the dataset (chromosome 22) being small enough to fit into memory.
+Clustering the samples is also fastest in \variantSpark{} (1min and 20secs), despite \variantSpark{} and ADAM both using {\sc Spark}'s MLlib \kMeans{} implementation, see Table~\ref{fivewaycomparison}.
+We attribute the 30\% speedup over ADAM (1min and 52secs) to \variantSpark{} converting the VCF files to sparse vectors, whereas ADAM creates dense vectors which are less memory efficient.
+Multithreading \kMeans{} clustering is natively supported by both R and Python, we are therefore able to utilise 8 cores. 
+Despite this, both were an order of magnitude slower than \variantSpark{} and ADAM, with python requiring 11mins and 29secs and R requiring 7mins and 25secs.
+Out Hadoop implementation performed the worst with 14mins and 23secs.
+The substantially increased runtime is due the dataset (chromosome 22) being small enough to fit into memory.
 This enables Python and R to store the output from each \kMeans{} iteration in memory.
@@
 Although this feature allows for greater scalability, as memory availability is less of an issue with huge datasets, the bottleneck introduced from disk IO increases processing time notably.
-This demonstrates one of the major improvements of {\sc Spark} over Hadoop, {\sc Spark}'s in-memory caching of datasets.
+{\sc Spark} overcomes this limitation by performing in-memory caching of datasets.
+Caching virtually eliminates the disk IO bottleneck on smaller datasets, and minimises the bottleneck on larger datasets where at least a portion of the data can still be cached in memory.
+
+%Clustering the samples is fastest in \variantSpark{} (1min and 20secs), being 30\% faster than ADAM (1min and 52secs) and  %80 vs 112
+%90\% faster than Hadoop (14mins and 23secs), see Table~\ref{fivewaycomparison}.
+%The similarity between \variantSpark{} and ADAM is expected, as they both use {\sc Spark}'s MLlib \kMeans{} implementation. 
+%The 30 second speed increase in \variantSpark{} is likely due to us converting the VCF files to sparse vectors, whereas ADAM creates dense vectors which are less memory efficient.
+
+%To demonstrate {\sc Spark}'s advantages over parallel approaches to \kMeans{} clustering in existing libraries, we compare the clustering libraries used in {\variantSpark}, to those in Python and R. 
+%For this comparison, each implementation ran on 8 cores.
+%In regards to speed, clustering using Python and R were both an order of magnitude slower than \variantSpark{} and ADAM, however both were faster than our Hadoop implementation.
+Python required 11mins and 29secs and R required 7mins and 25secs.
+%The substantially increased speed compared to Hadoop is due the dataset (chromosome 22) being small enough to fit into memory.
+This enables Python and R to store the output from each \kMeans{} iteration in memory.
+Hadoop, however, writes the entire output of each iteration to disk.
+%Although this feature allows for greater scalability, as memory availability is less of an issue with huge datasets, the bottleneck introduced from disk IO increases processing time notably.
+%This demonstrates one of the major improvements of {\sc Spark} over Hadoop, {\sc Spark}'s in-memory caching of datasets.
 Caching virtually eliminates the disk IO bottleneck on smaller datasets, and minimises the bottleneck on larger datasets where only a portion of the data can be cached in memory.
 
-For smaller datasets, we've demonstrated {\sc Spark} to be faster than its single-machine counterparts, however it's also important that this speedup isn't at the cost of quality.
-The 1000 Genomes data is ideal for investigating the clustering quality, as the consortium provides metadata which lists the super-population that each individual belongs to.
+We also investigate the cluster quality for the five different methods by comparing the annotated super-population label (AMR, AFR, EAS, SAS) for each individual in the 1000 Genomes data to the assigned label from the clustering allowing four clusters.
+For this comparison, we use the \ARI{} (ARI) metric, which returns a value between 1 (perfect) and 0 (random clustering). 
+Using Chromosome 22, each of the algorithms resulted in an ARI of 0.84, confirming that the speed-up in {\sc Spark} is not at the cost of quality.
+We can substantially improve the accuracy to a perfect classification (ARI=1.0) by removing the fourth super-population, AMR (American). 
+Including AMR individuals, we observe that the majority of these individuals are placed in the same group as Europeans, likely accurately reflecting their migrational backgrounds. 
+Only a minority of AMR individuals form an independent group, likely comprising of genetic information otherwise not captured by the 26 sub-populations of the 1000 genomes project.
+This indicates that although allelic differences exist between populations, "genetic diversity is distributed on a continuum"~\cite{Hunter2014}.
+
+%For smaller datasets, we've demonstrated {\sc Spark} to be faster than its single-machine counterparts, however it's also important that this speedup isn't at the cost of quality.
+%The 1000 Genomes data is ideal for investigating the clustering quality, as the consortium provides metadata which lists the super-population that each individual belongs to.
 This knowledge allows us to create a `truth' clustering, where individuals are clustered by their super-population.
-With the four super-populations (AMR, AFR, EAS, SAS), we can divide the individuals up into four clusters based on these labels. We can then compare this truth clustering with the \kMeans{} clusters.
-For this comparison, we use the \ARI{} (ARI) metric. To generate the ARI, an algorithm takes two separate clusterings and returns a value based on their similarity.
-Two identical clusterings will result in a 1, whereas 0 represents two clusterings that do not agree on any points (individuals).
-On Chromosome 22, each or the algorithms resulted in an ARI of 0.84.
-Although not perfect, the identical results demonstrate that the speed-up in {\sc Spark} isn't at the cost of quality.
-A result of 1 is probably unlikely because, as mentioned by 10.15252/embr.201439469, although differences in the distribution of alleles exist between populations, the boundaries are rather diffuse.
-%TODO - proper citation
+%With the four super-populations (AMR, AFR, EAS, SAS), we can divide the individuals up into four clusters based on these labels. 
+%We can then compare this truth clustering with the \kMeans{} clusters.
+%For this comparison, we use the \ARI{} (ARI) metric. To generate the ARI, an algorithm takes two separate clusterings and returns a value based on their similarity.
+%Two identical clusterings will result in a 1, whereas 0 represents two clusterings that do not agree on any points (individuals). 
+%On Chromosome 22, each or the algorithms resulted in an ARI of 0.84. Although not perfect, the identical results demonstrate that the speed-up in {\sc Spark} isn't at the cost of quality.
+%A result of 1 is probably unlikely because, as mentioned by Hunter {\it et al.}~\cite{Hunter2014}, although differences in the distribution of alleles exist between populations, the boundaries are rather diffuse.
 
@@
 
-Although we've demonstrated the advantages of \variantSpark{} with smaller datasets, the main advantage of {\sc Spark} (and hense \variantSpark) is it's scalability and ability to process datasets that exceed memory limits.
-To demonstrate the scalability, we continue to cluster the same individuals from the 1000 Genomes Project, however, rather than clustering on a single chromosome, we cluster the individuals on different subsets of variants from their 22 autosomes, and finally every variant.
-We run this comparison on our in-house Hadoop cluster. Pre-processing the VCF files takes approximately 12 minutes, 19 minutes and 40 minutes for 20\%, 40\% and 100\% of the genome, respectively.
-In each case, the memory required doesn't exceed a modest 2GB per executor, even for 100\% of the genome. Memory requirements for clustering, however, increase in correlation with the dataset.
+Although we have demonstrated the advantages of \variantSpark{} over traditional methods on small datasets, the main advantage of {\sc Spark} is its ability to process datasets that exceed memory limits.
+% TODO add numbers
+To demonstrate this scalability, we increase the number of variants used in the clustering from the initial XX on chromosome 22 to YY including all variants in the genome.
 
-Where our previous clustering of chromosome 22 required no more than 4GB memory per executor, scaling to 100\% of the genome increases the requirement to more than 24GB.
-This scales in an approximate linear trend, where 20\% of the genome requires 6GB and 40\% of the genome requires 12GB.
-24GB per executor for \kMeans{} clustering makes it much more resource intensive than pre-processing's 2GB per executor. 
-One factor which results in these differing memory requirements is the sparsity of the vectors used to store the variants.
+%Although we have demonstrated the advantages of \variantSpark{} on smaller datasets, the main advantage of {\sc Spark} (and hense \variantSpark) is it's scalability and ability to process datasets that exceed memory limits.
+%To demonstrate the scalability, we continue to cluster the same individuals from the 1000 Genomes Project, however, rather than clustering on a single chromosome, we cluster the individuals on different subsets of variants from their 22 autosomes, and finally every variant.
+
+% TODO add numbers
+We run this comparison on our in-house Hadoop cluster using with XX cores.
+Pre-processing the VCF files takes approximately 12 minutes, 19 minutes and 40 minutes for 20\%, 40\% and 100\% of the genome, respectively.
+In each case, the memory required does not exceed a modest 2GB per executor, even for 100\% of the genome. 
+%TODO add numbers
+Clustering the data takes XX minutes, XX minutes and XX minutes, respectively.
+For clustering, the memory requirements increases approximate linearly with 6GB, 12GB and 24GB, respectively, per executor. 
+%Memory requirements for clustering, however, increase in correlation with the dataset. 
+%Where our previous clustering of chromosome 22 required no more than 4GB memory per executor, scaling to 100\% of the genome increases the requirement to more than 24GB. 
+%This scales in an approximate linear trend, where 20\% of the genome requires 6GB and 40\% of the genome requires 12GB.
+%24GB per executor for \kMeans{} clustering makes it much more resource intensive than pre-processing's 2GB per executor. 
+This substantial increase in memory requirement is due to \kMeans{} converting the provided sparse vectors into dense vectors during processing.  
+% DENIS: This conflicts with the 30 seconds gain over ADAM which you attribute to our clustering using sparse vectors, 
+% TODO: is it farther ?
+While \kMeans{} clustering can scale to 100\% of the genome, machine learning algorithms that can deal with sparse vectors would potentially be able to scale much further, for example process more alleles.
+
+%DENIS: do you see a change in ARI from chr22 to 20%, 40% ?
+
+%One factor which results in these differing memory requirements is the sparsity of the vectors used to store the variants.
 Due to their sparse nature (the abundance of 0s) we can store them efficiently as Sparse Vectors in our pre-processing stage.
@@
 This is much less memory-efficient, as dense vectors explicitly define zero values (which are excluded from Sparse Vectors so don't require any memory), considerably increasing the memory required.
-So although we demonstrate the potential of \kMeans{} clustering to scale to 100\% of the genome, machine learning algorithms that can deal with sparse vectors would potentially be able to scale much further, or for example, easily deal with more alleles.
+%So although we demonstrate the potential of \kMeans{} clustering to scale to 100\% of the genome, machine learning algorithms that can deal with sparse vectors would potentially be able to scale much further, or for example, easily deal with more alleles.
 
@@
 The PGP is open to the public for individuals to submit their genomic sequence along with any metadata, such as diseases or clinical features ~\cite{Lunshof2013}.
-The  VCF file we use contains 985,790 alleles.
+The VCF file we use contains 985,790 variants.
 
-Clustering individuals leads to an ARI of 0.1. 
-Some observations are that the population groups to diverse in the PGP data.
-Each individual has metadata stating which country each of their grandparents are from, and only 40 individuals have all four grandparents from the same country.
-Also, being open to the public, the data may be more prone to error.
+% TODO add number
+Again we cluster the individuals allowing X clusters and compare the assigned labels to the same number of self-reported population labels.
+The observed ARI is 0.1. 
+This poor result may be due to the population annotation being too detailed and diverse especially so for self-reported and hence potential uncertain ancestry.
+%Some observations are that the population groups to diverse in the PGP data. 
+%Each individual has metadata stating which country each of their grandparents are from, and only 40 individuals have all four grandparents from the same country.
+%Also, being open to the public, the data may be more prone to error.
 For some individuals, the stated country of their grandparents changes between their updates.
 This could indicate some uncertainty about their origins, and hence, correct population.
+% TODO provide mapping
+We therefore only include individuals where all grandparents were reported to be from the same country and group them into their approximate super-populations (see Supplemental Table X). 
+This new dataset comprises XX individuals.
+%To try improve on these results, we only include the individuals with their grandparents were from the same country.
+%We then grouped them into their approximate super-populations. 
+Clustering these individuals results in an ARI of 0.29.
 
-To try improve on these results, we only include the individuals with their grandparents were from the same country.
-We then grouped them into their approximate super-populations.
-Clustering these individuals results in an ARI of 0.29.
-Although still a far cry from the 1000 Genomes Project results, it's an improvement on the previous clustering, albeit with fewer individuals.
+%DENIS: If you see an increase in accuracy above we can hypothesise here that 985,790 variants are not sufficient...
+otherwise a sentence around the population mix in potential real world clinical cohorts....
 

### 2015-09-10 doc/publications/journal2014/paper.tex
updated paper
fixed most comments

@@
 Unlike \variantSpark{} and our Hadoop implementation, the ADAM framework cannot process the VCF files directly but requires them to be converted into a binary ADAM file format. 
-%Unlike \variantSpark{} or our Hadoop implementation, which pre-process VCF files directly, the ADAM framework requires the input to be in the ADAM file format, and VCF files must therefore be converted to this binary format.
 Although this additional pre-processing step is only required once for each input file, it requires an additional 13mins and 13secs, %793s
-rendering our approach almost an order of magnitude faster. 
-%Although this additional pre-processing step is only required as a one-off for each input file, it requires an additional 13mins and 13secs, %793s
 rendering our approach almost an order of magnitude faster.
@@
 Clustering the samples is also fastest in \variantSpark{} (1min and 20secs), despite \variantSpark{} and ADAM both using {\sc Spark}'s MLlib \kMeans{} implementation, see Table~\ref{fivewaycomparison}.
+% DENIS: how sure are we that this is not just a fluke, e.g. I typically repeat this 5 times and report the average and std/ste ...
 We attribute the 30\% speedup over ADAM (1min and 52secs) to \variantSpark{} converting the VCF files to sparse vectors, whereas ADAM creates dense vectors which are less memory efficient.
@@
 Despite this, both were an order of magnitude slower than \variantSpark{} and ADAM, with python requiring 11mins and 29secs and R requiring 7mins and 25secs.
-Out Hadoop implementation performed the worst with 14mins and 23secs.
+Our Hadoop implementation performed the worst with 14mins and 23secs.
 The substantially increased runtime is due the dataset (chromosome 22) being small enough to fit into memory.
@@
 
-%Clustering the samples is fastest in \variantSpark{} (1min and 20secs), being 30\% faster than ADAM (1min and 52secs) and  %80 vs 112
-%90\% faster than Hadoop (14mins and 23secs), see Table~\ref{fivewaycomparison}.
-%The similarity between \variantSpark{} and ADAM is expected, as they both use {\sc Spark}'s MLlib \kMeans{} implementation. 
-%The 30 second speed increase in \variantSpark{} is likely due to us converting the VCF files to sparse vectors, whereas ADAM creates dense vectors which are less memory efficient.
-
-%To demonstrate {\sc Spark}'s advantages over parallel approaches to \kMeans{} clustering in existing libraries, we compare the clustering libraries used in {\variantSpark}, to those in Python and R. 
-%For this comparison, each implementation ran on 8 cores.
-%In regards to speed, clustering using Python and R were both an order of magnitude slower than \variantSpark{} and ADAM, however both were faster than our Hadoop implementation.
-Python required 11mins and 29secs and R required 7mins and 25secs.
-%The substantially increased speed compared to Hadoop is due the dataset (chromosome 22) being small enough to fit into memory.
-This enables Python and R to store the output from each \kMeans{} iteration in memory.
-Hadoop, however, writes the entire output of each iteration to disk.
-%Although this feature allows for greater scalability, as memory availability is less of an issue with huge datasets, the bottleneck introduced from disk IO increases processing time notably.
-%This demonstrates one of the major improvements of {\sc Spark} over Hadoop, {\sc Spark}'s in-memory caching of datasets.
-Caching virtually eliminates the disk IO bottleneck on smaller datasets, and minimises the bottleneck on larger datasets where only a portion of the data can be cached in memory.
-
 We also investigate the cluster quality for the five different methods by comparing the annotated super-population label (AMR, AFR, EAS, SAS) for each individual in the 1000 Genomes data to the assigned label from the clustering allowing four clusters.
 For this comparison, we use the \ARI{} (ARI) metric, which returns a value between 1 (perfect) and 0 (random clustering). 
-Using Chromosome 22, each of the algorithms resulted in an ARI of 0.84, confirming that the speed-up in {\sc Spark} is not at the cost of quality.
+Using chromosome 22, each of the algorithms resulted in an ARI of 0.84, confirming that the speed-up in {\sc Spark} is not at the cost of quality.
 We can substantially improve the accuracy to a perfect classification (ARI=1.0) by removing the fourth super-population, AMR (American). 
@@
 
-%For smaller datasets, we've demonstrated {\sc Spark} to be faster than its single-machine counterparts, however it's also important that this speedup isn't at the cost of quality.
-%The 1000 Genomes data is ideal for investigating the clustering quality, as the consortium provides metadata which lists the super-population that each individual belongs to.
-This knowledge allows us to create a `truth' clustering, where individuals are clustered by their super-population.
-%With the four super-populations (AMR, AFR, EAS, SAS), we can divide the individuals up into four clusters based on these labels. 
-%We can then compare this truth clustering with the \kMeans{} clusters.
-%For this comparison, we use the \ARI{} (ARI) metric. To generate the ARI, an algorithm takes two separate clusterings and returns a value based on their similarity.
-%Two identical clusterings will result in a 1, whereas 0 represents two clusterings that do not agree on any points (individuals). 
-%On Chromosome 22, each or the algorithms resulted in an ARI of 0.84. Although not perfect, the identical results demonstrate that the speed-up in {\sc Spark} isn't at the cost of quality.
-%A result of 1 is probably unlikely because, as mentioned by Hunter {\it et al.}~\cite{Hunter2014}, although differences in the distribution of alleles exist between populations, the boundaries are rather diffuse.
-
 \subsection*{\variantSpark{} allows genome wide sampling of variants to improve clustering quality}
@@
 Although we have demonstrated the advantages of \variantSpark{} over traditional methods on small datasets, the main advantage of {\sc Spark} is its ability to process datasets that exceed memory limits.
-% TODO add numbers
-To demonstrate this scalability, we increase the number of variants used in the clustering from the initial XX on chromosome 22 to YY including all variants in the genome.
+To demonstrate this scalability, we increase the number of variants used in the clustering from the initial 494,328 on chromosome 22 to 38,219,238 including all variants in the genome.
 
-%Although we have demonstrated the advantages of \variantSpark{} on smaller datasets, the main advantage of {\sc Spark} (and hense \variantSpark) is it's scalability and ability to process datasets that exceed memory limits.
-%To demonstrate the scalability, we continue to cluster the same individuals from the 1000 Genomes Project, however, rather than clustering on a single chromosome, we cluster the individuals on different subsets of variants from their 22 autosomes, and finally every variant.
-
-% TODO add numbers
-We run this comparison on our in-house Hadoop cluster using with XX cores.
-Pre-processing the VCF files takes approximately 12 minutes, 19 minutes and 40 minutes for 20\%, 40\% and 100\% of the genome, respectively.
+We run this comparison on our in-house Hadoop cluster using 64 cores, although this was limited to 40 cores with 100\% of the genome due to memory availability.
+Pre-processing the VCF files takes approximately 12 minutes, 19 minutes, 27 minutes and 40 minutes for 20\%, 40\%, 60\% and 100\% of the genome, respectively.
 In each case, the memory required does not exceed a modest 2GB per executor, even for 100\% of the genome. 
-%TODO add numbers
-Clustering the data takes XX minutes, XX minutes and XX minutes, respectively.
-For clustering, the memory requirements increases approximate linearly with 6GB, 12GB and 24GB, respectively, per executor. 
-%Memory requirements for clustering, however, increase in correlation with the dataset. 
-%Where our previous clustering of chromosome 22 required no more than 4GB memory per executor, scaling to 100\% of the genome increases the requirement to more than 24GB. 
-%This scales in an approximate linear trend, where 20\% of the genome requires 6GB and 40\% of the genome requires 12GB.
-%24GB per executor for \kMeans{} clustering makes it much more resource intensive than pre-processing's 2GB per executor. 
-This substantial increase in memory requirement is due to \kMeans{} converting the provided sparse vectors into dense vectors during processing.  
+Clustering the data takes 70 minutes, 139 minutes, 213 minutes and 884 mins, respectively.
+For clustering, the memory requirements increase approximately linearly with 6GB, 12GB, 17GB and 24GB, respectively, per executor. 
+This substantial increase in memory required is a result of the increased dimensionality of the data causing the \kMeans{} centroids, which MLlib stores as dense vectors, requiring excessive memory.  
 % DENIS: This conflicts with the 30 seconds gain over ADAM which you attribute to our clustering using sparse vectors,
+% AIDAN: Only centroids are dense. Reworded to explain this
 % TODO: is it farther ?
+% AIDAN: I had to check (http://www.quickanddirtytips.com/education/grammar/further-versus-farther) but I believe it is further.
 While \kMeans{} clustering can scale to 100\% of the genome, machine learning algorithms that can deal with sparse vectors would potentially be able to scale much further, for example process more alleles.
@@
 %DENIS: do you see a change in ARI from chr22 to 20%, 40% ?
+%AIDAN: not really..
 
@@
 \section*{Conclusions}
+%DENIS: mention genome wide sampling ability
+
 \variantSpark{} excels at pre-processing VCF files into row vectors, being over an order of magnitude faster than ADAM.

### 2015-09-11 doc/publications/journal2014/paper.tex
fixed more comments

@@
 % DENIS: how sure are we that this is not just a fluke, e.g. I typically repeat this 5 times and report the average and std/ste ...
+% AIDAN: This result was consistent. But it was on the Azure cluster so I can't get std/ste..
 We attribute the 30\% speedup over ADAM (1min and 52secs) to \variantSpark{} converting the VCF files to sparse vectors, whereas ADAM creates dense vectors which are less memory efficient.
@@
 % TODO: is it farther ?
-% AIDAN: I had to check (http://www.quickanddirtytips.com/education/grammar/further-versus-farther) but I believe it is further.
+% AIDAN: I had to check, but I believe it is further? (http://www.quickanddirtytips.com/education/grammar/further-versus-farther)
 While \kMeans{} clustering can scale to 100\% of the genome, machine learning algorithms that can deal with sparse vectors would potentially be able to scale much further, for example process more alleles.
@@
 %AIDAN: not really..
-
-%One factor which results in these differing memory requirements is the sparsity of the vectors used to store the variants.
-Due to their sparse nature (the abundance of 0s) we can store them efficiently as Sparse Vectors in our pre-processing stage.
-However, to cluster these vectors, the \kMeans{} algorithm converts these sparse vectors to dense vectors.
-This is much less memory-efficient, as dense vectors explicitly define zero values (which are excluded from Sparse Vectors so don't require any memory), considerably increasing the memory required.
-%So although we demonstrate the potential of \kMeans{} clustering to scale to 100\% of the genome, machine learning algorithms that can deal with sparse vectors would potentially be able to scale much further, or for example, easily deal with more alleles.
 
@@
 
-% TODO add number
-Again we cluster the individuals allowing X clusters and compare the assigned labels to the same number of self-reported population labels.
-The observed ARI is 0.1. 
-This poor result may be due to the population annotation being too detailed and diverse especially so for self-reported and hence potential uncertain ancestry.
-%Some observations are that the population groups to diverse in the PGP data. 
-%Each individual has metadata stating which country each of their grandparents are from, and only 40 individuals have all four grandparents from the same country.
-%Also, being open to the public, the data may be more prone to error.
-For some individuals, the stated country of their grandparents changes between their updates.
-This could indicate some uncertainty about their origins, and hence, correct population.
+%AIDAN: Changed this section..
+Again we cluster the individuals allowing 5 clusters and compare the assigned labels to the same number of self-reported `Race/ethnicity' labels (`White', `Hispanic or Latino', `Asian', `Black or African American' and `American Indian').
+The observed ARI is negative, indicating the clustering is completely random. 
+This poor result may be due to the labels being too broad and being a blanket label for a diverse range of individuals.
+For example, the `White' label includes individuals with grandparents from United States to Syrian Arab Republic to Bulgaria.
+The dataset is also very one-sided with 433 out of the 478 (over 90\%) of the individuals being classified as white, even though geographical location of their grandparents are very diverse.
+%This poor result may be due to the population annotation being too detailed and diverse, especially so for self-reported, and hence potential uncertain ancestry.
 % TODO provide mapping
 We therefore only include individuals where all grandparents were reported to be from the same country and group them into their approximate super-populations (see Supplemental Table X). 
-This new dataset comprises XX individuals.
-%To try improve on these results, we only include the individuals with their grandparents were from the same country.
-%We then grouped them into their approximate super-populations. 
+This new dataset comprises 36 individuals.
 Clustering these individuals results in an ARI of 0.29.
@@
 otherwise a sentence around the population mix in potential real world clinical cohorts....
+Although we see an increase, the relatively small ARI is likely due to the diversity in this real world clinical cohort across the small sample size.
+Although our filter removes individuals with a mixed background, we are left with only 36 individuals from 23 different countries.
+Applying a super-population mask, based on geographical distance, to these individuals reduces the number of groups from 23 to 5, however this grouping doesn't take migrations that occurred prior to their grandparents into account.
 

### 2015-09-11 doc/publications/journal2014/paper.tex
comments

@@
 In the first section we benchmark the performance and accuracy of MLlib's implementations against Hadoop's Mahout as well as more traditional methods (R, Python) limiting the used data to chromosome 22.
+% TODO fill in
 In section two we demonstrate {\sc Spark}'s full capacity by seamlessly scaling from 1\% to 100\% of the human genome.
@@
 Clustering the samples is also fastest in \variantSpark{} (1min and 20secs), despite \variantSpark{} and ADAM both using {\sc Spark}'s MLlib \kMeans{} implementation, see Table~\ref{fivewaycomparison}.
+% 80
 % DENIS: how sure are we that this is not just a fluke, e.g. I typically repeat this 5 times and report the average and std/ste ...
 % AIDAN: This result was consistent. But it was on the Azure cluster so I can't get std/ste..
-We attribute the 30\% speedup over ADAM (1min and 52secs) to \variantSpark{} converting the VCF files to sparse vectors, whereas ADAM creates dense vectors which are less memory efficient.
+% DENIS: OK but I thought all of them were run on the same machine and you are still generating results for python ... in any case can you write something like (repeated X times)
+We attribute the 30\% speedup over ADAM (1min and 52secs) to \variantSpark{} converting the VCF files to sparse vectors, whereas ADAM creates dense vectors which are less memory efficient. %
 Multithreading \kMeans{} clustering is natively supported by both R and Python, we are therefore able to utilise 8 cores. 
 Despite this, both were an order of magnitude slower than \variantSpark{} and ADAM, with python requiring 11mins and 29secs and R requiring 7mins and 25secs.
-Our Hadoop implementation performed the worst with 14mins and 23secs.
+Our Hadoop implementation performed the worst with 14mins and 23secs. % 863
 The substantially increased runtime is due the dataset (chromosome 22) being small enough to fit into memory.
@@
 
-We run this comparison on our in-house Hadoop cluster using 64 cores, although this was limited to 40 cores with 100\% of the genome due to memory availability.
+We run this comparison on our in-house Hadoop cluster using 40 cores.
+%We run this comparison on our in-house Hadoop cluster using 64 cores, although this was limited to 40 cores with 100\% of the genome due to memory availability.
 Pre-processing the VCF files takes approximately 12 minutes, 19 minutes, 27 minutes and 40 minutes for 20\%, 40\%, 60\% and 100\% of the genome, respectively.
@@
 For clustering, the memory requirements increase approximately linearly with 6GB, 12GB, 17GB and 24GB, respectively, per executor. 
-This substantial increase in memory required is a result of the increased dimensionality of the data causing the \kMeans{} centroids, which MLlib stores as dense vectors, requiring excessive memory.  
+This substantial increase in memory is a result of more \kMeans{} centroids being created, which MLlib stores as dense vectors. 
+%DENIS: do you see a change in ARI from chr22 to 20%, 40% ?
+%AIDAN: not really.. 
+%DENIS: so it is still exactly 0.84? Can you provide the number of Variants used in the different sets (e.g. is 20\% less than chrom 22)
+%TODO A citation wold be good here
+We do not observe an increase in cluster accuracy when providing variants from other chromosomes indicating that current practice of clustering based on a subset of the genome is sufficient to define the population boundaries~\cite{Hunter2014}.
+
+%This substantial increase in memory required is a result of the increased dimensionality of the data causing the \kMeans{} centroids, which MLlib stores as dense vectors, requiring excessive memory.
 % DENIS: This conflicts with the 30 seconds gain over ADAM which you attribute to our clustering using sparse vectors,
 % AIDAN: Only centroids are dense. Reworded to explain this
-% TODO: is it farther ?
-% AIDAN: I had to check, but I believe it is further? (http://www.quickanddirtytips.com/education/grammar/further-versus-farther)
+% DENIS: That means you allow more clusters ?
 While \kMeans{} clustering can scale to 100\% of the genome, machine learning algorithms that can deal with sparse vectors would potentially be able to scale much further, for example process more alleles.
-
-%DENIS: do you see a change in ARI from chr22 to 20%, 40% ?
-%AIDAN: not really..
 
@@
 
-%AIDAN: Changed this section..
 Again we cluster the individuals allowing 5 clusters and compare the assigned labels to the same number of self-reported `Race/ethnicity' labels (`White', `Hispanic or Latino', `Asian', `Black or African American' and `American Indian').
 The observed ARI is negative, indicating the clustering is completely random. 
-This poor result may be due to the labels being too broad and being a blanket label for a diverse range of individuals.
+This poor result may be due to the labels being too broad and not accurately reflecting genetic diversity.
+For example, the `White' label includes individuals with grandparents from United States, Syria, Arab Republic, or Bulgaria, amongst others.
+%This poor result may be due to the labels being too broad and being a blanket label for a diverse range of individuals.
 For example, the `White' label includes individuals with grandparents from United States to Syrian Arab Republic to Bulgaria.
 The dataset is also very one-sided with 433 out of the 478 (over 90\%) of the individuals being classified as white, even though geographical location of their grandparents are very diverse.
-%This poor result may be due to the population annotation being too detailed and diverse, especially so for self-reported, and hence potential uncertain ancestry.
 % TODO provide mapping
-We therefore only include individuals where all grandparents were reported to be from the same country and group them into their approximate super-populations (see Supplemental Table X). 
-This new dataset comprises 36 individuals.
+We therefore only include individuals where all grandparents were reported to be from the same country leaving 36 individuals from 23 different countries, which we group into their approximate super-populations (see Supplemental Table X).
 Clustering these individuals results in an ARI of 0.29.
 
-%DENIS: If you see an increase in accuracy above we can hypothesise here that 985,790 variants are not sufficient...
-otherwise a sentence around the population mix in potential real world clinical cohorts....
-Although we see an increase, the relatively small ARI is likely due to the diversity in this real world clinical cohort across the small sample size.
+Although we see an increase in cluster-accuracy when removing individuals with a mixed background and operating at super-population level, the clustering is not as precise as on the carefully curated and characterised individuals from the 1000 Genomes Project.
+%TODO add numbers
+This is especially the case since XX\% are 'White' Americans, which formed the group most difficult to cluster in the 1000 Genomes Project data.
+
+This highlights the issue for clinical application where ancestries influences treatment (e.g.
+HLA allele genotyping from SNP information~\cite{Zheng2014}) and accurate population association may not be known for patients with diverse migrational background.
+Hence fast genome-wide applicable approaches may help in providing the necessary depth to provide insight into genetic diversity.
+
+%Although we see an increase, the relatively small ARI is likely due to the diversity in this real world clinical cohort across the small sample size.
 Although our filter removes individuals with a mixed background, we are left with only 36 individuals from 23 different countries.
-Applying a super-population mask, based on geographical distance, to these individuals reduces the number of groups from 23 to 5, however this grouping doesn't take migrations that occurred prior to their grandparents into account.
+%Applying a super-population mask, based on geographical distance, to these individuals reduces the number of groups from 23 to 5, however this grouping doesn't take migrations that occurred prior to their grandparents into account.
 
@@
 \section*{Conclusions}
-%DENIS: mention genome wide sampling ability
 
-\variantSpark{} excels at pre-processing VCF files into row vectors, being over an order of magnitude faster than ADAM.
+\variantSpark{} performs clustering on VCF files with over 1000 individuals and 40 Million variants in under 15 hours using minimal memory (24GB). 
+\variantSpark{} supports random genome-wide sampling of variants allowing faster clustering for well characterised cohorts where 20\% of the genome is sufficient. 
+On the benchmarking dataset, it outperforms ADAM by almost an order of magnitude (4 vs 28 minutes) by processing VCF files directly and storing the information in sparse vectors. % 258 vs 1673 
+\variantSpark{} utilizes {\sc Spark}, which allows for in-memory cashing and hence performs 85\% faster than Hadoop (29 min), 94\% faster than R (75 min), and 95\% faster than Python (103 min).
+
+%\variantSpark{} excels at pre-processing VCF files into row vectors, being over an order of magnitude faster than ADAM.
 Memory consumption is also kept to a minimum.
 Where traditional approaches such as Python and R would fail when expanding to the whole genome,
-\variantSpark{} can pre-process VCF files with over 1000 individuals in under half an hour using minimal memory.
+%\variantSpark{} can pre-process VCF files with over 1000 individuals in under half an hour using minimal memory.
 
-Clustering these individuals is also much faster with \variantSpark{} than Python and R, although only marginally faster than ADAM.
+%Clustering these individuals is also much faster with \variantSpark{} than Python and R, although only marginally faster than ADAM.
 
-These benefits of speed, resource consumption and scalability demonstrate \variantSpark{} to be a useful tool in applying machine learning tools to genomic data.
+%These benefits of speed, resource consumption and scalability demonstrate \variantSpark{} to be a useful tool in applying machine learning tools to genomic data.
 Although we only run \kMeans on the data, MLlib provides many other machine learning algorithms, such as Random Forests.
-Therefore it may be useful to investigate the novelty of applying these machine learning approaches to big genomic data.
+%Therefore it may be useful to investigate the novelty of applying these machine learning approaches to big genomic data.
 

### 2015-09-15 doc/publications/journal2014/paper.tex
update table

@@
 % DENIS: OK but I thought all of them were run on the same machine and you are still generating results for python ... in any case can you write something like (repeated X times)
+% AIDAN: They were all run on Azure. I was just generating ARI results with Python the other day on another machine
 We attribute the 30\% speedup over ADAM (1min and 52secs) to \variantSpark{} converting the VCF files to sparse vectors, whereas ADAM creates dense vectors which are less memory efficient. %
@@
 \variantSpark{}	& 8	& 32	& 2min 58sec	& 8	& 32	& 1min 20sec	& 0.84	\\ 
-ADAM		& 8	& 32	& 12min 48sec	& 8	& 32	& 1min 52sec	& x	\\
-Hadoop		& 8	& 32	& 14min 22sec	& 8	& 32	& 14min 23sec	& x	\\
-R			& 1	& 32	& 67min		& 8	& 32	& 7min 25sec	& x	\\
+ADAM		& 8	& 32	& 12min 48sec	& 8	& 32	& 1min 52sec	& 0.84	\\
+Hadoop		& 8	& 32	& 14min 22sec	& 8	& 32	& 14min 23sec	& 0.84	\\
+R			& 1	& 32	& 67min		& 8	& 32	& 7min 25sec	& 0.84	\\
 Python		& 1	& 32	& 92min		& 8	& 32	& 11min 29sec	& 0.84	\\
+ADMIXTURE	& 1	& 32	& 10min 8sec	& 8	& 32 & 8min 19sec	& 0.25	\\
   \hline

### 2015-09-15 doc/publications/journal2014/paper.tex

--- 2015-09-11 doc/publications/journal2014/paper.tex
@@
 %% TODO - NEIL - (and better heading) %%
-\subsection*{Personal Genome Project VCF making}
-The PGP data.....
+\subsection*{Personal Genome Project Data}
+%% The PGP data..... 
+PGP genome data was downloaded from the PGP website (\url{https://my.pgp-hms.org/public_genetic_data}). Genotype data from 23andMe microarray platforms was sorted by genome 
+build. Those files corresponding to NCBI build 36 were converted to BED format, updated to build 37 using UCSC liftOver and then converted back to 23andMe format. The 
+23andMe files were then converted to VCF format using code obtained from the Broad Institute (\url{http://apol1.blogspot.com.au/2013/08/impute-apoe-and-apol1-with-23andme.html}). 
+Where individuals had genotype data for both genome builds 36 and 37, only the latter was used. Finally, individual VCF files were combined for clustering into one 
+file using VCFtools \cite{Danecek2011Variant}.
+
+[if code to be made available, say so here and where it will be]
 

### 2015-09-16 doc/publications/journal2014/paper.tex
images

@@
 For clustering, the memory requirements increase approximately linearly with 6GB, 12GB, 17GB and 24GB, respectively, per executor. 
+%This substantial increase in memory required is a result of the increased dimensionality of the data causing the \kMeans{} centroids, which MLlib stores as dense vectors, requiring excessive memory.
+% DENIS: This conflicts with the 30 seconds gain over ADAM which you attribute to our clustering using sparse vectors,
+% AIDAN: Only centroids are dense. Reworded to explain this
+% DENIS: That means you allow more clusters ?
 This substantial increase in memory is a result of more \kMeans{} centroids being created, which MLlib stores as dense vectors. 
@@
 We do not observe an increase in cluster accuracy when providing variants from other chromosomes indicating that current practice of clustering based on a subset of the genome is sufficient to define the population boundaries~\cite{Hunter2014}.
-
-%This substantial increase in memory required is a result of the increased dimensionality of the data causing the \kMeans{} centroids, which MLlib stores as dense vectors, requiring excessive memory.
-% DENIS: This conflicts with the 30 seconds gain over ADAM which you attribute to our clustering using sparse vectors,
-% AIDAN: Only centroids are dense. Reworded to explain this
-% DENIS: That means you allow more clusters ?
 While \kMeans{} clustering can scale to 100\% of the genome, machine learning algorithms that can deal with sparse vectors would potentially be able to scale much further, for example process more alleles.
@@
 On the benchmarking dataset, it outperforms ADAM by almost an order of magnitude (4 vs 28 minutes) by processing VCF files directly and storing the information in sparse vectors. % 258 vs 1673 
-\variantSpark{} utilizes {\sc Spark}, which allows for in-memory cashing and hence performs 85\% faster than Hadoop (29 min), 94\% faster than R (75 min), and 95\% faster than Python (103 min).
-
-%\variantSpark{} excels at pre-processing VCF files into row vectors, being over an order of magnitude faster than ADAM.
-Memory consumption is also kept to a minimum.
-Where traditional approaches such as Python and R would fail when expanding to the whole genome,
-%\variantSpark{} can pre-process VCF files with over 1000 individuals in under half an hour using minimal memory.
-
-%Clustering these individuals is also much faster with \variantSpark{} than Python and R, although only marginally faster than ADAM.
-
-%These benefits of speed, resource consumption and scalability demonstrate \variantSpark{} to be a useful tool in applying machine learning tools to genomic data.
-Although we only run \kMeans on the data, MLlib provides many other machine learning algorithms, such as Random Forests.
-%Therefore it may be useful to investigate the novelty of applying these machine learning approaches to big genomic data.
-
-%completes the whole variant clustering task in just over 4 minutes, which is 80\% faster than ADAM (28 mins), 85\% faster than Hadoop (29 min), 94\% faster than R (75 min), and 95\% faster than Python (103 min).
+\variantSpark{} utilises {\sc Spark}, which allows for in-memory cashing and hence performs 85\% faster than Hadoop (29 min).
+This novel parallelization approach is also faster than traditional multithreading with 94\% speedup over R (75 min), and 95\% over Python (103 min).
+These benefits of speed, resource consumption and scalability demonstrate \variantSpark{} to be the interface for applying other machine learning tools from MlLib to genomic data.
 
@@
 
+% DENIS yes, it will be in the git repository mentioned above (https://github.com/BauerLab/VariantSpark)
+% can you please push your code into https://stash.csiro.au/users/bau04c/repos/variantspark/browse
 [if code to be made available, say so here and where it will be]
@@
   \begin{figure}[h!]
-  \caption{\csentence{Sample figure title.}
-      A short description of the figure content
-      should go here.}
+  \includegraphics[type=pdf,ext=.pdf,read=.pdf, scale=0.45]{images/Resources}
+  \includegraphics[type=pdf,ext=.pdf,read=.pdf, scale=0.45]{images/Scaling}
+  \caption{\csentence{Comparison of method and genome-wide scaling experiment.}
+      {\bf Left} Runtime is given in seconds with 32 GB of memory on 8 threads (except for the pre-processing in R and Python where this was not supported). 
+      {\bf Right} Scaling from 20\% to 100\% of variants in the genome with maximal number of executors and lowest possible memory assignment.}
+      \end{figure}
+
+%Hadoop
+%1. Map
+%The mapper converts each line of the VCF file into a preliminary key-value pair, where the Key is the byte-position of the first character thereby functioning as a proxy for the genomic location (variant ID), and the Value is a string of the line.
+From this, the key-value pairs for the individual genotypes are created by extracting the variant from each sample as the new Value.
+The new key is a tuple consisting of the primary key, which is the individual ID (position of the parsed variants), and the variant ID as the secondary key.
+%The MapReduce partitioner uses custom Collect/sort functions to collect the tuples by their primary key (individual ID) and distributes them between nodes.
+The data is sorted by secondary key (variant ID)
+%2. Reduce
+%This data is input to the Reducers, which store the variants for each individual to a sparse vector with the Key as the individual ID and the value as the variant vector.
+%
+%Spark
+%1. The lines from the VCF file(s) are parsed independently by worker nodes, where each line is zipped with the heading from the VCF file resulting in tuples of individual ID and the variant. 
+%2.
+These arrays are each zipped again with the unique, sequential index (the Variant ID) resulting in tuples (essentially key/value pairs) of Variant IDs (?key?) and the array of tuples from step 1 (?value?).
+%3. The flatmap function restructures the data into Individual ID (?key?) and tuples with Variant ID and the Variant (?value?).
+%4. These tuples are grouped by the Individual ID, collecting all Variant ID and Variant tuples, and non-zero variants are saved as a sparse vector.
+\begin{figure}[h!]
+\includegraphics[type=pdf,ext=.pdf,read=.pdf, scale=0.25]{signature}
+  \caption{\csentence{Schematic of the Hadoop and Spark implementation.} The image shows the conceptual differences in the steps necessary to convert variants in VCF file format to a data structure readable by Mahout and MLlib, respectively.}
       \end{figure}

### 2015-09-16 doc/publications/journal2014/paper.tex
polish 1

--- 2015-09-16 doc/publications/journal2014/paper.tex
@@
 This drastic increase in both sample numbers and features per sample requires a massively parallel approach to data processing~\cite{Stein2010}.
-Traditional parallelisation strategies like OpenMPI or hardware accelerators (GPGPU) cannot scale to variable data sizes or require purpose-build hardware.
+Traditional parallelisation strategies like OpenMPI or hardware accelerators (GPGPU) cannot scale with variable data sizes or require purpose-build hardware.
 
@@
 Recognising the capability, the Big Data Genomics (BDG) group recently demonstrated the strength of {\sc Spark} in a genomic clustering application using ADAM, a set of formats and APIs as well as processing stage implementations for genomic data~\cite{Massie2013}.
-While the speedup over traditional methods was impressive, being limited by constraints within this general genomics framework hampered performance.
+While the speedup over traditional methods was impressive (50 fold speedup), being limited by constraints within this general genomics framework hampered performance.
 
 We hence developed a more focused purpose-built application in {\sc Spark} to perform genomic clustering of individuals. 
-We utilise {\sc Spark}'s machine learning library, \mbox{MLlib}, and provide an interface from the standard variant data format, Variant Call Format (VCF)~\cite{1KG2012}, which opens up the application of MLlib's different machine learning algorithms a wide range of genotype-based analysis tasks.
-To demonstrate its capability, we cluster variant datasets from the 1000 Genomes Project~\cite{1KG2012} to determine population structure using the \kMeans{} clustering algorithm available in MLlib's.
+We utilise {\sc Spark}'s machine learning library, \mbox{MLlib}, and provide an interface to the standard variant data format, Variant Call Format (VCF)~\cite{1KG2012}, which opens up the application of MLlib's different machine learning algorithms to a wide range of genotype-based analysis tasks.
+
+To demonstrate \variantSpark's capability, we cluster variant datasets from the 1000 Genomes Project~\cite{1KG2012} to determine population structure using the \kMeans{} clustering algorithm available in MLlib's.
 In the first section we benchmark the performance and accuracy of MLlib's implementations against Hadoop's Mahout as well as more traditional methods (R, Python) limiting the used data to chromosome 22.
-% TODO fill in
-In section two we demonstrate {\sc Spark}'s full capacity by seamlessly scaling from 1\% to 100\% of the human genome.
-In section three we replicate the analysis by using XX genomes from the Personal Genome Project~\cite{Lunshof2013}.
+In section two we demonstrate \variantSpark's full capacity by seamlessly scaling from 20\% to 100\% of the human genome.
+In section three we replicate the analysis by using 478 genomes from the Personal Genome Project~\cite{Lunshof2013}.
 In the last section we discuss the pipeline for visualising the resulting cluster.
@@
 In this section we compare the time required to cluster individuals based on genomic variants using \variantSpark{} against ADAM as well as the more traditional approaches using Hadoop (and Mahout), Python and R.
-We limit the genomic variants to only chromosome 22 as the traditional approaches have substantially larger memory consumption rendering a whole-genome input infeasible.  
+We limit the genomic variants to only chromosome 22 as the traditional approaches have substantially larger memory consumption, rendering a whole-genome input infeasible.  
 Furthermore, we perform the comparisons on a single virtual machine to ensure the five different approaches have access to the same resources.
-We use \kMeans{} clustering algorithm in the respective implementations, which all require the VCF input files to be pre-processed (see methods). 
-We therefore compare the time required for the pre-processing step, as well as the time required for \kMeans{} clustering.
+We use \kMeans{} clustering algorithms in the respective implementations, which require the VCF input files to be pre-processed (see methods). 
+We therefore compare the time required for the pre-processing step, as well as for \kMeans{} clustering.
 
-As shown in Table~\ref{fivewaycomparison}, pre-processing the data is fastest in \variantSpark{}, requiring 2mins and 58secs. %178s 
+As shown in Table 1, pre-processing the data is fastest in \variantSpark{}, requiring 2mins and 58secs. %178s 
 This is almost 80\% faster than the ADAM implementation (12min and 48secs) %768s
-or our Hadoop implementation (14min and 22secs). %862s
+or our Hadoop implementation (14min and 22secs), respectively. %862s
 Unlike \variantSpark{} and our Hadoop implementation, the ADAM framework cannot process the VCF files directly but requires them to be converted into a binary ADAM file format. 
 Although this additional pre-processing step is only required once for each input file, it requires an additional 13mins and 13secs, %793s
-rendering our approach almost an order of magnitude faster.
-R and Python are slower at pre-processing, taking 67 minutes and 92 minutes, respectively.
+rendering our approach almost an order of magnitude faster (see Figure 1).
+R and Python are also slower at pre-processing, taking 67 minutes and 92 minutes, respectively. 
 This increased pre-processing time is due to the Python and R implementations not natively supporting multithreading, while \variantSpark{}, ADAM and Hadoop can use the eight available cores.
 
-Clustering the samples is also fastest in \variantSpark{} (1min and 20secs), despite \variantSpark{} and ADAM both using {\sc Spark}'s MLlib \kMeans{} implementation, see Table~\ref{fivewaycomparison}.
-% 80
+Clustering the samples is also fastest in \variantSpark{} (1min and 20secs, see Table 1), despite \variantSpark{} and ADAM both using {\sc Spark}'s MLlib \kMeans{} implementation. % 80
 % DENIS: how sure are we that this is not just a fluke, e.g. I typically repeat this 5 times and report the average and std/ste ...
 % AIDAN: This result was consistent. But it was on the Azure cluster so I can't get std/ste..
-% DENIS: OK but I thought all of them were run on the same machine and you are still generating results for python ... in any case can you write something like (repeated X times)
-We attribute the 30\% speedup over ADAM (1min and 52secs) to \variantSpark{} converting the VCF files to sparse vectors, whereas ADAM creates dense vectors which are less memory efficient. %
-Multithreading \kMeans{} clustering is natively supported by both R and Python, we are therefore able to utilise 8 cores. 
-Despite this, both were an order of magnitude slower than \variantSpark{} and ADAM, with python requiring 11mins and 29secs and R requiring 7mins and 25secs.
-Our Hadoop implementation performed the worst with 14mins and 23secs. % 863
-The substantially increased runtime is due the dataset (chromosome 22) being small enough to fit into memory.
-This enables Python and R to store the output from each \kMeans{} iteration in memory.
+% DENIS: OK but I thought all of them were run on the same machine and you are still generating results for python ... 
+% TODO:  write something like (repeated X times)
+We attribute the 30\% speedup over ADAM (1min and 52secs) to \variantSpark{} converting the VCF files to sparse vectors, whereas ADAM creates dense vectors, which are less memory efficient. %
+
+Multithreading \kMeans{} clustering is natively supported by both R and Python, we are therefore able to utilise eight cores. 
+Despite this, both are an order of magnitude slower than \variantSpark{} and ADAM, with Python requiring 11mins and 29secs and R requiring 7mins and 25secs.
+
+Our Hadoop implementation performs the worst with 14mins and 23secs, showcasing the limitations of the Hadoop engine. % 863
+The substantially increased runtime is due to the dataset (chromosome 22) being small enough to fit into memory, which enables Python and R to store the output from each \kMeans{} iteration in memory.
 Hadoop, however, writes the entire output of each iteration to disk.
-Although this feature allows for greater scalability, as memory availability is less of an issue with huge datasets, the bottleneck introduced from disk IO increases processing time notably.
-{\sc Spark} overcomes this limitation by performing in-memory caching of datasets.
-Caching virtually eliminates the disk IO bottleneck on smaller datasets, and minimises the bottleneck on larger datasets where at least a portion of the data can still be cached in memory.
+Although this Hadoop feature allows for greater scalability, as disk IO is not the run-time determining factor for larger datasets, it introduces a bottleneck for smaller datasets.
+{\sc Spark} overcomes this limitation by performing in-memory caching of datasets, which eliminates the disk IO bottleneck on smaller datasets, and minimises the bottleneck on larger datasets where at least a portion of the data can be cached in memory.
 
 We also investigate the cluster quality for the five different methods by comparing the annotated super-population label (AMR, AFR, EAS, SAS) for each individual in the 1000 Genomes data to the assigned label from the clustering allowing four clusters.
+%TODO: 0 seems not to be the lowest number as we get a negative score for PGP
 For this comparison, we use the \ARI{} (ARI) metric, which returns a value between 1 (perfect) and 0 (random clustering). 
@@
 We run this comparison on our in-house Hadoop cluster using 40 cores.
-%We run this comparison on our in-house Hadoop cluster using 64 cores, although this was limited to 40 cores with 100\% of the genome due to memory availability.
-Pre-processing the VCF files takes approximately 12 minutes, 19 minutes, 27 minutes and 40 minutes for 20\%, 40\%, 60\% and 100\% of the genome, respectively.
+Pre-processing the VCF files takes approximately 12 minutes, 19 minutes, 27 minutes and 40 minutes for 20\%, 40\%, 60\% and 100\% of the genome, respectively (see Table 2).
 In each case, the memory required does not exceed a modest 2GB per executor, even for 100\% of the genome. 
 Clustering the data takes 70 minutes, 139 minutes, 213 minutes and 884 mins, respectively.
-For clustering, the memory requirements increase approximately linearly with 6GB, 12GB, 17GB and 24GB, respectively, per executor. 
+For clustering, the memory requirements increase approximately linearly to 24GB per executor (see Figure 1). 
 %This substantial increase in memory required is a result of the increased dimensionality of the data causing the \kMeans{} centroids, which MLlib stores as dense vectors, requiring excessive memory.
@@
 % DENIS: That means you allow more clusters ?
-This substantial increase in memory is a result of more \kMeans{} centroids being created, which MLlib stores as dense vectors. 
+The increase in memory is a result of more \kMeans{} centroids being created, which MLlib stores as dense vectors. 
+While \kMeans{} clustering can scale to 100\% of the genome, machine learning algorithms that can deal with sparse vectors would potentially be able to scale much further, for example process more alleles.
+
 %DENIS: do you see a change in ARI from chr22 to 20%, 40% ?
@@
 %DENIS: so it is still exactly 0.84? Can you provide the number of Variants used in the different sets (e.g. is 20\% less than chrom 22)
-%TODO A citation wold be good here
-We do not observe an increase in cluster accuracy when providing variants from other chromosomes indicating that current practice of clustering based on a subset of the genome is sufficient to define the population boundaries~\cite{Hunter2014}.
-While \kMeans{} clustering can scale to 100\% of the genome, machine learning algorithms that can deal with sparse vectors would potentially be able to scale much further, for example process more alleles.
+%TODO A citation would be good here
+We do not observe an increase in cluster accuracy when providing variants from other chromosomes, indicating that current practice of clustering based on a subset of the genome is sufficient to define the population boundaries~\cite{Hunter2014}.
 
@@
 To demonstrate the versatility of \variantSpark{} we also process and cluster individuals from the Personal Genomics Project (PGP).
-The PGP is open to the public for individuals to submit their genomic sequence along with any metadata, such as diseases or clinical features ~\cite{Lunshof2013}.
-The VCF file we use contains 985,790 variants.
+The PGP is open to the public for individuals to submit their genomic sequence along with any metadata, such as diseases or clinical features~\cite{Lunshof2013}.
+We obtain and curate the data (see methods) resulting in 985,790 variants from 478 individuals.
 
-Again we cluster the individuals allowing 5 clusters and compare the assigned labels to the same number of self-reported `Race/ethnicity' labels (`White', `Hispanic or Latino', `Asian', `Black or African American' and `American Indian').
+We cluster the individuals allowing five clusters and compare the assigned labels to the same number of self-reported `Race/ethnicity' labels (`White', `Hispanic or Latino', `Asian', `Black or African American' and `American Indian').
 The observed ARI is negative, indicating the clustering is completely random. 
@@
 For example, the `White' label includes individuals with grandparents from United States, Syria, Arab Republic, or Bulgaria, amongst others.
-%This poor result may be due to the labels being too broad and being a blanket label for a diverse range of individuals.
-For example, the `White' label includes individuals with grandparents from United States to Syrian Arab Republic to Bulgaria.
 The dataset is also very one-sided with 433 out of the 478 (over 90\%) of the individuals being classified as white, even though geographical location of their grandparents are very diverse.
@@
 
-This highlights the issue for clinical application where ancestries influences treatment (e.g.
+This highlights the issue for clinical application where ancestry influences treatment (e.g.
 HLA allele genotyping from SNP information~\cite{Zheng2014}) and accurate population association may not be known for patients with diverse migrational background.
 Hence fast genome-wide applicable approaches may help in providing the necessary depth to provide insight into genetic diversity. 
-
-%Although we see an increase, the relatively small ARI is likely due to the diversity in this real world clinical cohort across the small sample size.
-Although our filter removes individuals with a mixed background, we are left with only 36 individuals from 23 different countries.
-%Applying a super-population mask, based on geographical distance, to these individuals reduces the number of groups from 23 to 5, however this grouping doesn't take migrations that occurred prior to their grandparents into account.
+% DENIS: statement about why we cannot do more with this data...
+Unfortunately, this hypothesis could no not be tested as whole genome information is not available for these individuals.
 

### 2015-09-16 doc/publications/journal2014/paper.tex
a few fixes

--- 2015-09-16 doc/publications/journal2014/paper.tex
@@
 This drastic increase in both sample numbers and features per sample requires a massively parallel approach to data processing~\cite{Stein2010}.
-Traditional parallelisation strategies like OpenMPI or hardware accelerators (GPGPU) cannot scale to variable data sizes or require purpose-build hardware.
+Traditional parallelisation strategies like OpenMPI or hardware accelerators (GPGPU) cannot scale to variable data sizes or require purpose-built hardware.
 
-Addressing this issue, {\sc Apache Hadoop MapReduce}~\cite{Borthakur2007} transforms data into `key-value pairs' that can then be distributed between multiple nodes across a commodity compute-cluster depending on the size of the problem.
+Addressing this issue, {\sc Apache Hadoop MapReduce}~\cite{Borthakur2007} transforms data into `key-value pairs' that can then be distributed between multiple nodes across a commodity computer cluster, depending on the size of the problem.
 MapReduce approaches are increasingly being used in bioinformatics (for reviews see~\cite{Zou2013, Qiu2010,Taylor2010}). 
@@
 We utilise {\sc Spark}'s machine learning library, \mbox{MLlib}, and provide an interface from the standard variant data format, Variant Call Format (VCF)~\cite{1KG2012}, which opens up the application of MLlib's different machine learning algorithms a wide range of genotype-based analysis tasks.
-To demonstrate its capability, we cluster variant datasets from the 1000 Genomes Project~\cite{1KG2012} to determine population structure using the \kMeans{} clustering algorithm available in MLlib's.
+To demonstrate its capability, we cluster variant datasets from the 1000 Genomes Project~\cite{1KG2012} to determine population structure using the \kMeans{} clustering algorithm available in MLlib. 
 In the first section we benchmark the performance and accuracy of MLlib's implementations against Hadoop's Mahout as well as more traditional methods (R, Python) limiting the used data to chromosome 22.
-% TODO fill in
 In section two we demonstrate {\sc Spark}'s full capacity by seamlessly scaling from 1\% to 100\% of the human genome.
-In section three we replicate the analysis by using XX genomes from the Personal Genome Project~\cite{Lunshof2013}.
+In section three we replicate the analysis by using 478 genomes from the Personal Genome Project~\cite{Lunshof2013}.
 In the last section we discuss the pipeline for visualising the resulting cluster.
@@
 This poor result may be due to the labels being too broad and not accurately reflecting genetic diversity.
-For example, the `White' label includes individuals with grandparents from United States, Syria, Arab Republic, or Bulgaria, amongst others.
-%This poor result may be due to the labels being too broad and being a blanket label for a diverse range of individuals.
-For example, the `White' label includes individuals with grandparents from United States to Syrian Arab Republic to Bulgaria.
+For example, the `White' label includes individuals with grandparents from United States, Syria, Arab Republic, and Bulgaria, amongst others.
 The dataset is also very one-sided with 433 out of the 478 (over 90\%) of the individuals being classified as white, even though geographical location of their grandparents are very diverse.
@@
 
-%Although we see an increase, the relatively small ARI is likely due to the diversity in this real world clinical cohort across the small sample size.
-Although our filter removes individuals with a mixed background, we are left with only 36 individuals from 23 different countries.
-%Applying a super-population mask, based on geographical distance, to these individuals reduces the number of groups from 23 to 5, however this grouping doesn't take migrations that occurred prior to their grandparents into account.
-
 \subsection*{Graph visualization}
@@
 On the benchmarking dataset, it outperforms ADAM by almost an order of magnitude (4 vs 28 minutes) by processing VCF files directly and storing the information in sparse vectors. % 258 vs 1673 
-\variantSpark{} utilises {\sc Spark}, which allows for in-memory cashing and hence performs 85\% faster than Hadoop (29 min).
+\variantSpark{} utilises {\sc Spark}, which allows for in-memory caching and hence performs 85\% faster than Hadoop (29 min).
 This novel parallelization approach is also faster than traditional multithreading with 94\% speedup over R (75 min), and 95\% over Python (103 min).

### 2015-09-16 doc/publications/journal2014/paper.tex
admixture implementation

@@
 We also investigate the cluster quality for the five different methods by comparing the annotated super-population label (AMR, AFR, EAS, SAS) for each individual in the 1000 Genomes data to the assigned label from the clustering allowing four clusters.
-For this comparison, we use the \ARI{} (ARI) metric, which returns a value between 1 (perfect) and 0 (random clustering). 
+For this comparison, we use the \ARI{} (ARI) metric, which returns a value between -1 (independent labelings) and 1 (perfect match)~\cite{Hubert1985}.
 Using chromosome 22, each of the algorithms resulted in an ARI of 0.84, confirming that the speed-up in {\sc Spark} is not at the cost of quality.
@@
 To demonstrate the versatility of \variantSpark{} we also process and cluster individuals from the Personal Genomics Project (PGP).
-The PGP is open to the public for individuals to submit their genomic sequence along with any metadata, such as diseases or clinical features ~\cite{Lunshof2013}.
+The PGP is open to the public for individuals to submit their genomic sequence along with any metadata, such as diseases or clinical features~\cite{Lunshof2013}.
 The VCF file we use contains 985,790 variants.
@@
 Again we cluster the individuals allowing 5 clusters and compare the assigned labels to the same number of self-reported `Race/ethnicity' labels (`White', `Hispanic or Latino', `Asian', `Black or African American' and `American Indian').
-The observed ARI is negative, indicating the clustering is completely random. 
+The observed ARI is negative, indicating the clustering is random. 
 This poor result may be due to the labels being too broad and not accurately reflecting genetic diversity.
@@
 %TODO add numbers
-This is especially the case since XX\% are 'White' Americans, which formed the group most difficult to cluster in the 1000 Genomes Project data.
+This is especially the case since XX\% are `White' Americans, which formed the group most difficult to cluster in the 1000 Genomes Project data.
 
@@
 Although to cluster this DataFrame (using sci-kit learn), we simply need to convert it to a matrix with {\sc .as\textunderscore{}matrix()}
+
+\subsection*{ADMIXTURE Implementation}
+We use Genome Analysis Toolkit (GATK) to convert the variants from the VCF file, and a 1000 Genomes Project supplied .ped file, to a binary PLINK (.bed), binary marker information file (.bim) and pedigree stub file (.fam).
+These three files are used as input to ADMIXTURE, with K (the number of ancestral populations) set to 4.
 

### 2015-09-16 doc/publications/journal2014/paper.tex
remove VCF format

@@
 \subsection*{ADMIXTURE Implementation}
-We use Genome Analysis Toolkit (GATK) to convert the variants from the VCF file, and a 1000 Genomes Project supplied .ped file, to a binary PLINK (.bed), binary marker information file (.bim) and pedigree stub file (.fam).
+We use Genome Analysis Toolkit (GATK) to convert the variants from VCF, and a 1000 Genomes Project supplied .ped file, to a binary PLINK (.bed), binary marker information file (.bim) and pedigree stub file (.fam).
 These three files are used as input to ADMIXTURE, with K (the number of ancestral populations) set to 4.
@@
 build. Those files corresponding to NCBI build 36 were converted to BED format, updated to build 37 using UCSC liftOver and then converted back to 23andMe format. The 
-23andMe files were then converted to VCF format using code obtained from the Broad Institute (\url{http://apol1.blogspot.com.au/2013/08/impute-apoe-and-apol1-with-23andme.html}). 
+23andMe files were then converted to VCF using code obtained from the Broad Institute (\url{http://apol1.blogspot.com.au/2013/08/impute-apoe-and-apol1-with-23andme.html}). 
 Where individuals had genotype data for both genome builds 36 and 37, only the latter was used. Finally, individual VCF files were combined for clustering into one 
@@
 \includegraphics[type=pdf,ext=.pdf,read=.pdf, scale=0.25]{signature}
-  \caption{\csentence{Schematic of the Hadoop and Spark implementation.} The image shows the conceptual differences in the steps necessary to convert variants in VCF file format to a data structure readable by Mahout and MLlib, respectively.}
+  \caption{\csentence{Schematic of the Hadoop and Spark implementation.} The image shows the conceptual differences in the steps necessary to convert variants in VCF to a data structure readable by Mahout and MLlib, respectively.}
       \end{figure}

### 2015-09-16 doc/publications/journal2014/paper.tex

--- 2015-09-16 doc/publications/journal2014/paper.tex
@@
 A commonly performed task in such applications is grouping individuals based on their genomic profile to identify population association~\cite{Gao2007} or elucidate different haplotype involvement in diseases susceptibility~\cite{Laitman2013}.
+The commonly used tool is {\sc Admixture}~\cite{Alexander2009}, which performs maximum likelihood estimation of individual ancestries from multilocus SNP genotype datasets.
 Due to the decreasing sequencing cost it is economical to generate studies with sample sizes previously reserved for larger consortia such as the 1000 Genomes Project~\cite{1KG2012} or The Cancer Genome Atlas, TCGA~\cite{TCGA2013}.
@@
 This drastic increase in both sample numbers and features per sample requires a massively parallel approach to data processing~\cite{Stein2010}.
-Traditional parallelisation strategies like OpenMPI or hardware accelerators (GPGPU) cannot scale to variable data sizes or require purpose-built hardware.
+Traditional parallelisation strategies like OpenMPI or hardware accelerators (GPGPU) cannot scale with variable data sizes or require purpose-build hardware.
 
-Addressing this issue, {\sc Apache Hadoop MapReduce}~\cite{Borthakur2007} transforms data into `key-value pairs' that can then be distributed between multiple nodes across a commodity computer cluster, depending on the size of the problem.
+Addressing this issue, {\sc Apache Hadoop MapReduce}~\cite{Borthakur2007} transforms data into `key-value pairs' that can then be distributed between multiple nodes across a commodity compute-cluster depending on the size of the problem.
 MapReduce approaches are increasingly being used in bioinformatics (for reviews see~\cite{Zou2013, Qiu2010,Taylor2010}). 
@@
 Recognising the capability, the Big Data Genomics (BDG) group recently demonstrated the strength of {\sc Spark} in a genomic clustering application using ADAM, a set of formats and APIs as well as processing stage implementations for genomic data~\cite{Massie2013}.
-While the speedup over traditional methods was impressive, being limited by constraints within this general genomics framework hampered performance.
+While the speedup over traditional methods was impressive (50 fold speedup), being limited by constraints within this general genomics framework hampered performance.
 
 We hence developed a more focused purpose-built application in {\sc Spark} to perform genomic clustering of individuals. 
-We utilise {\sc Spark}'s machine learning library, \mbox{MLlib}, and provide an interface from the standard variant data format, Variant Call Format (VCF)~\cite{1KG2012}, which opens up the application of MLlib's different machine learning algorithms a wide range of genotype-based analysis tasks.
-To demonstrate its capability, we cluster variant datasets from the 1000 Genomes Project~\cite{1KG2012} to determine population structure using the \kMeans{} clustering algorithm available in MLlib. 
-In the first section we benchmark the performance and accuracy of MLlib's implementations against Hadoop's Mahout as well as more traditional methods (R, Python) limiting the used data to chromosome 22.
-In section two we demonstrate {\sc Spark}'s full capacity by seamlessly scaling from 1\% to 100\% of the human genome.
+We utilise {\sc Spark}'s machine learning library, \mbox{MLlib}, and provide an interface to the standard variant data format, Variant Call Format (VCF)~\cite{1KG2012}, which opens up the application of MLlib's different machine learning algorithms to a wide range of genotype-based analysis tasks.
+
+To demonstrate \variantSpark's capability, we cluster variant datasets from the 1000 Genomes Project~\cite{1KG2012} to determine population structure using the \kMeans{} clustering algorithm available in MLlib's.
+In the first section we benchmark \variantSpark's performance and accuracy against ADAM and a Hadoop Mahout implementation as well as more traditional methods (R, Python) and the purpose-build tool {\sc Admixture}~\cite{Alexander2009} limiting the used data to chromosome 22.
+In section two we demonstrate \variantSpark's full capacity by seamlessly scaling from 20\% to 100\% of the human genome.
 In section three we replicate the analysis by using 478 genomes from the Personal Genome Project~\cite{Lunshof2013}.
@@
 In this section we compare the time required to cluster individuals based on genomic variants using \variantSpark{} against ADAM as well as the more traditional approaches using Hadoop (and Mahout), Python and R.
-We limit the genomic variants to only chromosome 22 as the traditional approaches have substantially larger memory consumption rendering a whole-genome input infeasible.  
-Furthermore, we perform the comparisons on a single virtual machine to ensure the five different approaches have access to the same resources.
-We use \kMeans{} clustering algorithm in the respective implementations, which all require the VCF input files to be pre-processed (see methods). 
-We therefore compare the time required for the pre-processing step, as well as the time required for \kMeans{} clustering.
+We limit the genomic variants to only chromosome 22 as the traditional approaches have substantially larger memory consumption, rendering a whole-genome input infeasible.  
+Furthermore, we perform the comparisons on a single virtual machine to ensure the six different approaches have access to the same resources.
+We use \kMeans{} clustering algorithms in the respective implementations, which require the VCF input files to be pre-processed (see methods). 
+The state-of-the-art tool, {\sc Admixture}, also requires a pre-processing step from VCF to PED format, which we performed using GAKT~\cite{McKenna2010}.
+We therefore compare the time required for the pre-processing step, as well as for \kMeans{} clustering.
 
-As shown in Table~\ref{fivewaycomparison}, pre-processing the data is fastest in \variantSpark{}, requiring 2mins and 58secs. %178s 
+As shown in Table 1, pre-processing the data is fastest in \variantSpark{}, requiring 2mins and 58secs. %178s 
 This is almost 80\% faster than the ADAM implementation (12min and 48secs) %768s
-or our Hadoop implementation (14min and 22secs). %862s
+or our Hadoop implementation (14min and 22secs), respectively. %862s
 Unlike \variantSpark{} and our Hadoop implementation, the ADAM framework cannot process the VCF files directly but requires them to be converted into a binary ADAM file format. 
 Although this additional pre-processing step is only required once for each input file, it requires an additional 13mins and 13secs, %793s
-rendering our approach almost an order of magnitude faster.
-R and Python are slower at pre-processing, taking 67 minutes and 92 minutes, respectively.
+rendering our approach almost an order of magnitude faster (see Figure 1).
+R and Python are also slower at pre-processing, taking 67 minutes and 92 minutes, respectively. 
 This increased pre-processing time is due to the Python and R implementations not natively supporting multithreading, while \variantSpark{}, ADAM and Hadoop can use the eight available cores.
+The GATK-based VCF to PED conversion is also slower with over 10 minutes.
 
-Clustering the samples is also fastest in \variantSpark{} (1min and 20secs), despite \variantSpark{} and ADAM both using {\sc Spark}'s MLlib \kMeans{} implementation, see Table~\ref{fivewaycomparison}.
-% 80
+Clustering the samples is also fastest in \variantSpark{} (1min and 20secs, see Table 1), despite \variantSpark{} and ADAM both using {\sc Spark}'s MLlib \kMeans{} implementation. % 80
 % DENIS: how sure are we that this is not just a fluke, e.g. I typically repeat this 5 times and report the average and std/ste ...
 % AIDAN: This result was consistent. But it was on the Azure cluster so I can't get std/ste..
-% DENIS: OK but I thought all of them were run on the same machine and you are still generating results for python ... in any case can you write something like (repeated X times)
-% AIDAN: They were all run on Azure. I was just generating ARI results with Python the other day on another machine
-We attribute the 30\% speedup over ADAM (1min and 52secs) to \variantSpark{} converting the VCF files to sparse vectors, whereas ADAM creates dense vectors which are less memory efficient. %
-Multithreading \kMeans{} clustering is natively supported by both R and Python, we are therefore able to utilise 8 cores. 
-Despite this, both were an order of magnitude slower than \variantSpark{} and ADAM, with python requiring 11mins and 29secs and R requiring 7mins and 25secs.
-Our Hadoop implementation performed the worst with 14mins and 23secs. % 863
-The substantially increased runtime is due the dataset (chromosome 22) being small enough to fit into memory.
-This enables Python and R to store the output from each \kMeans{} iteration in memory.
+% DENIS: OK but I thought all of them were run on the same machine and you are still generating results for python ... 
+% TODO:  write something like (repeated X times)
+We attribute the 30\% speedup over ADAM (1min and 52secs) to \variantSpark{} converting the VCF files to sparse vectors, whereas ADAM creates dense vectors, which are less memory efficient. %
+Multithreading \kMeans{} clustering is natively supported by both R and Python, we are therefore able to utilise eight cores. 
+Despite this, both are an order of magnitude slower than \variantSpark{} and ADAM, with Python requiring 11mins and 29secs and R requiring 7mins and 25secs.
+{\sc Admixture} is also slower with a runtime of just over 8 minutes. 
+Our Hadoop implementation performs the worst with 14mins and 23secs, showcasing the limitations of the Hadoop engine. % 863
+The substantially increased runtime is due to the dataset (chromosome 22) being small enough to fit into memory, which enables Python and R to store the output from each \kMeans{} iteration in memory.
 Hadoop, however, writes the entire output of each iteration to disk.
-Although this feature allows for greater scalability, as memory availability is less of an issue with huge datasets, the bottleneck introduced from disk IO increases processing time notably.
-{\sc Spark} overcomes this limitation by performing in-memory caching of datasets.
-Caching virtually eliminates the disk IO bottleneck on smaller datasets, and minimises the bottleneck on larger datasets where at least a portion of the data can still be cached in memory.
+Although this Hadoop feature allows for greater scalability, as disk IO is not the run-time determining factor for larger datasets, it introduces a bottleneck for smaller datasets.
+{\sc Spark} overcomes this limitation by performing in-memory caching of datasets, which eliminates the disk IO bottleneck on smaller datasets, and minimises the bottleneck on larger datasets where at least a portion of the data can be cached in memory.
 
-We also investigate the cluster quality for the five different methods by comparing the annotated super-population label (AMR, AFR, EAS, SAS) for each individual in the 1000 Genomes data to the assigned label from the clustering allowing four clusters.
+We also investigate the cluster quality for the six different methods by comparing the annotated super-population label (AMR, AFR, EAS, SAS) for each individual in the 1000 Genomes data to the assigned label from the clustering allowing four clusters.
+%TODO: 0 seems not to be the lowest number as we get a negative score for PGP
 For this comparison, we use the \ARI{} (ARI) metric, which returns a value between 1 (perfect) and 0 (random clustering). 
 Using chromosome 22, each of the algorithms resulted in an ARI of 0.84, confirming that the speed-up in {\sc Spark} is not at the cost of quality.
+Even the state-of-the-art tool, {\sc Admixture} does returns a low ARI=0.25.
 We can substantially improve the accuracy to a perfect classification (ARI=1.0) by removing the fourth super-population, AMR (American). 
@@
 To demonstrate the versatility of \variantSpark{} we also process and cluster individuals from the Personal Genomics Project (PGP).
-The PGP is open to the public for individuals to submit their genomic sequence along with any metadata, such as diseases or clinical features ~\cite{Lunshof2013}.
-The VCF file we use contains 985,790 variants.
+The PGP is open to the public for individuals to submit their genomic sequence along with any metadata, such as diseases or clinical features~\cite{Lunshof2013}.
+We obtain and curate the data (see methods) resulting in 985,790 variants from 478 individuals.
 
-Again we cluster the individuals allowing 5 clusters and compare the assigned labels to the same number of self-reported `Race/ethnicity' labels (`White', `Hispanic or Latino', `Asian', `Black or African American' and `American Indian').
+We cluster the individuals allowing five clusters and compare the assigned labels to the same number of self-reported `Race/ethnicity' labels (`White', `Hispanic or Latino', `Asian', `Black or African American' and `American Indian').
 The observed ARI is negative, indicating the clustering is completely random. 
 This poor result may be due to the labels being too broad and not accurately reflecting genetic diversity.
-For example, the `White' label includes individuals with grandparents from United States, Syria, Arab Republic, and Bulgaria, amongst others.
+For example, the `White' label includes individuals with grandparents from United States, Syria, Arab Republic, or Bulgaria, amongst others.
 The dataset is also very one-sided with 433 out of the 478 (over 90\%) of the individuals being classified as white, even though geographical location of their grandparents are very diverse.
@@
 
-This highlights the issue for clinical application where ancestries influences treatment (e.g.
+This highlights the issue for clinical application where ancestry influences treatment (e.g.
 HLA allele genotyping from SNP information~\cite{Zheng2014}) and accurate population association may not be known for patients with diverse migrational background.
 Hence fast genome-wide applicable approaches may help in providing the necessary depth to provide insight into genetic diversity. 
+% DENIS: statement about why we cannot do more with this data...
+Unfortunately, this hypothesis could no not be tested as whole genome information is not available for these individuals.
 
@@
 On the benchmarking dataset, it outperforms ADAM by almost an order of magnitude (4 vs 28 minutes) by processing VCF files directly and storing the information in sparse vectors. % 258 vs 1673 
-\variantSpark{} utilises {\sc Spark}, which allows for in-memory caching and hence performs 85\% faster than Hadoop (29 min).
+\variantSpark{} utilises {\sc Spark}, which allows for in-memory cashing and hence performs 85\% faster than Hadoop (29 min).
 This novel parallelization approach is also faster than traditional multithreading with 94\% speedup over R (75 min), and 95\% over Python (103 min).
@@
 \begin{table}[h!]
-\caption{The resources consumption of the five compared methods as well as the accuracy measured as \ARI{} on chromosome 22.}
+\caption{The resources consumption of the six compared methods as well as the accuracy measured as \ARI{} on chromosome 22.}
       \begin{tabular}{|c|ccc|ccc|c|c|}

### 2015-09-16 doc/publications/journal2014/paper.tex
Admixture discussion

@@
 We observe a 50 fold speedup when using the efficient in-memory {\sc Spark} compute engine compared to Hadoop.
-Furthermore, our implementation achieves a better performance compared to a previously published {\sc Spark}-based genome clustering approach, ADAM.
+Furthermore, our implementation achieves a better performance compared to a previously published {\sc Spark}-based genome clustering approach, {\sc ADAM}.
 
@@
 Although {\sc Spark} allows MapReduce-like programs, it does not require programs to exactly model the MapReduce paradigm, which in turn allows more flexibility when designing programs. 
-Recognising the capability, the Big Data Genomics (BDG) group recently demonstrated the strength of {\sc Spark} in a genomic clustering application using ADAM, a set of formats and APIs as well as processing stage implementations for genomic data~\cite{Massie2013}.
+Recognising the capability, the Big Data Genomics (BDG) group recently demonstrated the strength of {\sc Spark} in a genomic clustering application using {\sc ADAM}, a set of formats and APIs as well as processing stage implementations for genomic data~\cite{Massie2013}.
 While the speedup over traditional methods was impressive (50 fold speedup), being limited by constraints within this general genomics framework hampered performance.
@@
 To demonstrate \variantSpark's capability, we cluster variant datasets from the 1000 Genomes Project~\cite{1KG2012} to determine population structure using the \kMeans{} clustering algorithm available in MLlib's.
-In the first section we benchmark \variantSpark's performance and accuracy against ADAM and a Hadoop Mahout implementation as well as more traditional methods (R, Python) and the purpose-build tool {\sc Admixture}~\cite{Alexander2009} limiting the used data to chromosome 22.
+In the first section we benchmark \variantSpark's performance and accuracy against {\sc ADAM} and a Hadoop Mahout implementation as well as more traditional methods (R, Python) and the purpose-build tool {\sc Admixture}~\cite{Alexander2009} limiting the used data to chromosome 22.
 In section two we demonstrate \variantSpark's full capacity by seamlessly scaling from 20\% to 100\% of the human genome.
@@
 
-In this section we compare the time required to cluster individuals based on genomic variants using \variantSpark{} against ADAM as well as the more traditional approaches using Hadoop (and Mahout), Python and R.
+In this section we compare the time required to cluster individuals based on genomic variants using \variantSpark{} against {\sc ADAM} as well as the more traditional approaches using Hadoop (and Mahout), Python and R.
 We limit the genomic variants to only chromosome 22 as the traditional approaches have substantially larger memory consumption, rendering a whole-genome input infeasible.  
@@
 As shown in Table 1, pre-processing the data is fastest in \variantSpark{}, requiring 2mins and 58secs. %178s 
-This is almost 80\% faster than the ADAM implementation (12min and 48secs) %768s
+This is almost 80\% faster than the {\sc ADAM} implementation (12min and 48secs) %768s
 or our Hadoop implementation (14min and 22secs), respectively. %862s
-Unlike \variantSpark{} and our Hadoop implementation, the ADAM framework cannot process the VCF files directly but requires them to be converted into a binary ADAM file format. 
+Unlike \variantSpark{} and our Hadoop implementation, the {\sc ADAM} framework cannot process the VCF files directly but requires them to be converted into a binary {\sc ADAM} file format. 
 Although this additional pre-processing step is only required once for each input file, it requires an additional 13mins and 13secs, %793s
@@
 R and Python are also slower at pre-processing, taking 67 minutes and 92 minutes, respectively. 
-This increased pre-processing time is due to the Python and R implementations not natively supporting multithreading, while \variantSpark{}, ADAM and Hadoop can use the eight available cores.
+This increased pre-processing time is due to the Python and R implementations not natively supporting multithreading, while \variantSpark{}, {\sc ADAM} and Hadoop can use the eight available cores.
 The GATK-based VCF to PED conversion is also slower with over 10 minutes.
 
-Clustering the samples is also fastest in \variantSpark{} (1min and 20secs, see Table 1), despite \variantSpark{} and ADAM both using {\sc Spark}'s MLlib \kMeans{} implementation. % 80
+Clustering the samples is also fastest in \variantSpark{} (1min and 20secs, see Table 1), despite \variantSpark{} and {\sc ADAM} both using {\sc Spark}'s MLlib \kMeans{} implementation. % 80
 % DENIS: how sure are we that this is not just a fluke, e.g. I typically repeat this 5 times and report the average and std/ste ...
@@
 % TODO:  write something like (repeated X times)
-We attribute the 30\% speedup over ADAM (1min and 52secs) to \variantSpark{} converting the VCF files to sparse vectors, whereas ADAM creates dense vectors, which are less memory efficient. %
+We attribute the 30\% speedup over {\sc ADAM} (1min and 52secs) to \variantSpark{} converting the VCF files to sparse vectors, whereas {\sc ADAM} creates dense vectors, which are less memory efficient.
+%
 Multithreading \kMeans{} clustering is natively supported by both R and Python, we are therefore able to utilise eight cores. 
-Despite this, both are an order of magnitude slower than \variantSpark{} and ADAM, with Python requiring 11mins and 29secs and R requiring 7mins and 25secs.
+Despite this, both are an order of magnitude slower than \variantSpark{} and {\sc ADAM}, with Python requiring 11mins and 29secs and R requiring 7mins and 25secs.
 {\sc Admixture} is also slower with a runtime of just over 8 minutes. 
@@
 %This substantial increase in memory required is a result of the increased dimensionality of the data causing the \kMeans{} centroids, which MLlib stores as dense vectors, requiring excessive memory.
-% DENIS: This conflicts with the 30 seconds gain over ADAM which you attribute to our clustering using sparse vectors,
+% DENIS: This conflicts with the 30 seconds gain over {\sc ADAM} which you attribute to our clustering using sparse vectors,
 % AIDAN: Only centroids are dense. Reworded to explain this
@@
 \variantSpark{} supports random genome-wide sampling of variants allowing faster clustering for well characterised cohorts where 20\% of the genome is sufficient. 
-On the benchmarking dataset, it outperforms ADAM by almost an order of magnitude (4 vs 28 minutes) by processing VCF files directly and storing the information in sparse vectors. % 258 vs 1673 
+On the benchmarking dataset, it outperforms {\sc ADAM} by almost an order of magnitude (4 vs 28 minutes) by processing VCF files directly and storing the information in sparse vectors. % 258 vs 1673 
 \variantSpark{} utilises {\sc Spark}, which allows for in-memory cashing and hence performs 85\% faster than Hadoop (29 min).
 This novel parallelization approach is also faster than traditional multithreading with 94\% speedup over R (75 min), and 95\% over Python (103 min).
+\variantSpark{} is also superior in performance and accuracy over the current state-of-the-art tool for individual ancestries determination, {\sc Admixture}.
 These benefits of speed, resource consumption and scalability demonstrate \variantSpark{} to be the interface for applying other machine learning tools from MlLib to genomic data.
@@
 
-\subsection*{ADAM Implementation}
-For our ADAM comparison, we followed the ADAM implementation from (\url{http://bdgenomics.org/blog/2015/02/02/scalable-genomes-clustering-with-adam-and-spark/}).
+\subsection*{{\sc ADAM} Implementation}
+For our {\sc ADAM} comparison, we followed the {\sc ADAM} implementation from (\url{http://bdgenomics.org/blog/2015/02/02/scalable-genomes-clustering-with-adam-and-spark/}).
 
@@
 \variantSpark{}	& 8	& 32	& 2min 58sec	& 8	& 32	& 1min 20sec	& 0.84	\\ 
-ADAM		& 8	& 32	& 12min 48sec	& 8	& 32	& 1min 52sec	& 0.84	\\
+{\sc ADAM}		& 8	& 32	& 12min 48sec	& 8	& 32	& 1min 52sec	& 0.84	\\
 Hadoop		& 8	& 32	& 14min 22sec	& 8	& 32	& 14min 23sec	& 0.84	\\
@@
 Python		& 1	& 32	& 92min		& 8	& 32	& 11min 29sec	& 0.84	\\
-ADMIXTURE	& 1	& 32	& 10min 8sec	& 8	& 32 & 8min 19sec	& 0.25	\\
+{\sc Admixture}	& 1	& 32	& 10min 8sec	& 8	& 32 & 8min 19sec	& 0.25	\\
   \hline

### 2015-09-16 doc/publications/journal2014/paper.tex

@@
 This drastic increase in both sample numbers and features per sample requires a massively parallel approach to data processing~\cite{Stein2010}.
-Traditional parallelisation strategies like OpenMPI or hardware accelerators (GPGPU) cannot scale with variable data sizes or require purpose-build hardware.
+Traditional parallelisation strategies like OpenMPI or hardware accelerators (GPGPU) cannot scale with variable data sizes or require purpose-built hardware.
 
-Addressing this issue, {\sc Apache Hadoop MapReduce}~\cite{Borthakur2007} transforms data into `key-value pairs' that can then be distributed between multiple nodes across a commodity compute-cluster depending on the size of the problem.
+Addressing this issue, {\sc Apache Hadoop MapReduce}~\cite{Borthakur2007} transforms data into `key-value pairs' that can then be distributed between multiple nodes across a commodity computer cluster, depending on the size of the problem.
 MapReduce approaches are increasingly being used in bioinformatics (for reviews see~\cite{Zou2013, Qiu2010,Taylor2010}). 
@@
 
-To demonstrate \variantSpark's capability, we cluster variant datasets from the 1000 Genomes Project~\cite{1KG2012} to determine population structure using the \kMeans{} clustering algorithm available in MLlib's.
+To demonstrate \variantSpark's capability, we cluster variant datasets from the 1000 Genomes Project~\cite{1KG2012} to determine population structure using the \kMeans{} clustering algorithm available in MLlib.
 In the first section we benchmark \variantSpark's performance and accuracy against {\sc ADAM} and a Hadoop Mahout implementation as well as more traditional methods (R, Python) and the purpose-build tool {\sc Admixture}~\cite{Alexander2009} limiting the used data to chromosome 22.
@@
 
-We also investigate the cluster quality for the six different methods by comparing the annotated super-population label (AMR, AFR, EAS, SAS) for each individual in the 1000 Genomes data to the assigned label from the clustering allowing four clusters.
-%TODO: 0 seems not to be the lowest number as we get a negative score for PGP
-For this comparison, we use the \ARI{} (ARI) metric, which returns a value between 1 (perfect) and 0 (random clustering). 
+We also investigate the cluster quality for the five different methods by comparing the annotated super-population label (AMR, AFR, EAS, SAS) for each individual in the 1000 Genomes data to the assigned label from the clustering allowing four clusters.
+For this comparison, we use the \ARI{} (ARI) metric, which returns a value between -1 (independent labelings) and 1 (perfect match)~\cite{Hubert1985}.
 Using chromosome 22, each of the algorithms resulted in an ARI of 0.84, confirming that the speed-up in {\sc Spark} is not at the cost of quality.
@@
 We cluster the individuals allowing five clusters and compare the assigned labels to the same number of self-reported `Race/ethnicity' labels (`White', `Hispanic or Latino', `Asian', `Black or African American' and `American Indian').
-The observed ARI is negative, indicating the clustering is completely random. 
+The observed ARI is negative, indicating the clustering is random. 
 This poor result may be due to the labels being too broad and not accurately reflecting genetic diversity.
@@
 On the benchmarking dataset, it outperforms {\sc ADAM} by almost an order of magnitude (4 vs 28 minutes) by processing VCF files directly and storing the information in sparse vectors. % 258 vs 1673 
-\variantSpark{} utilises {\sc Spark}, which allows for in-memory cashing and hence performs 85\% faster than Hadoop (29 min).
+\variantSpark{} utilises {\sc Spark}, which allows for in-memory caching and hence performs 85\% faster than Hadoop (29 min).
 This novel parallelization approach is also faster than traditional multithreading with 94\% speedup over R (75 min), and 95\% over Python (103 min).
@@
 
+\subsection*{ADMIXTURE Implementation}
+We use Genome Analysis Toolkit (GATK) to convert the variants from VCF, and a 1000 Genomes Project supplied .ped file, to a binary PLINK (.bed), binary marker information file (.bim) and pedigree stub file (.fam).
+These three files are used as input to ADMIXTURE, with K (the number of ancestral populations) set to 4.
+
 %% TODO - NEIL - (and better heading) %%

### 2015-09-17 doc/publications/journal2014/paper.tex
Edited PGP data methods section

--- 2015-09-16 doc/publications/journal2014/paper.tex
@@
 
-%\def\includegraphic{}
-%\def\includegraphics{}
-\usepackage{graphicx}
+\def\includegraphic{}
+\def\includegraphics{}
 
@@
 A commonly performed task in such applications is grouping individuals based on their genomic profile to identify population association~\cite{Gao2007} or elucidate different haplotype involvement in diseases susceptibility~\cite{Laitman2013}.
-The commonly used tool is {\sc Admixture}~\cite{Alexander2009}, which performs maximum likelihood estimation of individual ancestries from multilocus SNP genotype datasets.
 Due to the decreasing sequencing cost it is economical to generate studies with sample sizes previously reserved for larger consortia such as the 1000 Genomes Project~\cite{1KG2012} or The Cancer Genome Atlas, TCGA~\cite{TCGA2013}.
@@
 This drastic increase in both sample numbers and features per sample requires a massively parallel approach to data processing~\cite{Stein2010}.
-Traditional parallelisation strategies like OpenMPI or hardware accelerators (GPGPU) cannot scale with variable data sizes or require purpose-build hardware.
+Traditional parallelisation strategies like OpenMPI or hardware accelerators (GPGPU) cannot scale to variable data sizes or require purpose-build hardware.
 
@@
 Recognising the capability, the Big Data Genomics (BDG) group recently demonstrated the strength of {\sc Spark} in a genomic clustering application using ADAM, a set of formats and APIs as well as processing stage implementations for genomic data~\cite{Massie2013}.
-While the speedup over traditional methods was impressive (50 fold speedup), being limited by constraints within this general genomics framework hampered performance.
+While the speedup over traditional methods was impressive, being limited by constraints within this general genomics framework hampered performance.
 
 We hence developed a more focused purpose-built application in {\sc Spark} to perform genomic clustering of individuals. 
-We utilise {\sc Spark}'s machine learning library, \mbox{MLlib}, and provide an interface to the standard variant data format, Variant Call Format (VCF)~\cite{1KG2012}, which opens up the application of MLlib's different machine learning algorithms to a wide range of genotype-based analysis tasks.
-
-To demonstrate \variantSpark's capability, we cluster variant datasets from the 1000 Genomes Project~\cite{1KG2012} to determine population structure using the \kMeans{} clustering algorithm available in MLlib's.
-In the first section we benchmark \variantSpark's performance and accuracy against ADAM and a Hadoop Mahout implementation as well as more traditional methods (R, Python) and the purpose-build tool {\sc Admixture}~\cite{Alexander2009} limiting the used data to chromosome 22.
-In section two we demonstrate \variantSpark's full capacity by seamlessly scaling from 20\% to 100\% of the human genome.
-In section three we replicate the analysis by using 478 genomes from the Personal Genome Project~\cite{Lunshof2013}.
+We utilise {\sc Spark}'s machine learning library, \mbox{MLlib}, and provide an interface from the standard variant data format, Variant Call Format (VCF)~\cite{1KG2012}, which opens up the application of MLlib's different machine learning algorithms a wide range of genotype-based analysis tasks.
+To demonstrate its capability, we cluster variant datasets from the 1000 Genomes Project~\cite{1KG2012} to determine population structure using the \kMeans{} clustering algorithm available in MLlib's.
+In the first section we benchmark the performance and accuracy of MLlib's implementations against Hadoop's Mahout as well as more traditional methods (R, Python) limiting the used data to chromosome 22.
+% TODO fill in
+In section two we demonstrate {\sc Spark}'s full capacity by seamlessly scaling from 1\% to 100\% of the human genome.
+In section three we replicate the analysis by using XX genomes from the Personal Genome Project~\cite{Lunshof2013}.
 In the last section we discuss the pipeline for visualising the resulting cluster.
@@
 In this section we compare the time required to cluster individuals based on genomic variants using \variantSpark{} against ADAM as well as the more traditional approaches using Hadoop (and Mahout), Python and R.
-We limit the genomic variants to only chromosome 22 as the traditional approaches have substantially larger memory consumption, rendering a whole-genome input infeasible.  
-Furthermore, we perform the comparisons on a single virtual machine to ensure the six different approaches have access to the same resources.
-We use \kMeans{} clustering algorithms in the respective implementations, which require the VCF input files to be pre-processed (see methods). 
-The state-of-the-art tool, {\sc Admixture}, also requires a pre-processing step from VCF to PED format, which we performed using GAKT~\cite{McKenna2010}.
-We therefore compare the time required for the pre-processing step, as well as for \kMeans{} clustering.
+We limit the genomic variants to only chromosome 22 as the traditional approaches have substantially larger memory consumption rendering a whole-genome input infeasible.  
+Furthermore, we perform the comparisons on a single virtual machine to ensure the five different approaches have access to the same resources.
+We use \kMeans{} clustering algorithm in the respective implementations, which all require the VCF input files to be pre-processed (see methods). 
+We therefore compare the time required for the pre-processing step, as well as the time required for \kMeans{} clustering.
 
-As shown in Table 1, pre-processing the data is fastest in \variantSpark{}, requiring 2mins and 58secs. %178s 
+As shown in Table~\ref{fivewaycomparison}, pre-processing the data is fastest in \variantSpark{}, requiring 2mins and 58secs. %178s 
 This is almost 80\% faster than the ADAM implementation (12min and 48secs) %768s
-or our Hadoop implementation (14min and 22secs), respectively. %862s
+or our Hadoop implementation (14min and 22secs). %862s
 Unlike \variantSpark{} and our Hadoop implementation, the ADAM framework cannot process the VCF files directly but requires them to be converted into a binary ADAM file format. 
 Although this additional pre-processing step is only required once for each input file, it requires an additional 13mins and 13secs, %793s
-rendering our approach almost an order of magnitude faster (see Figure 1).
-R and Python are also slower at pre-processing, taking 67 minutes and 92 minutes, respectively. 
+rendering our approach almost an order of magnitude faster.
+R and Python are slower at pre-processing, taking 67 minutes and 92 minutes, respectively.
 This increased pre-processing time is due to the Python and R implementations not natively supporting multithreading, while \variantSpark{}, ADAM and Hadoop can use the eight available cores.
-The GATK-based VCF to PED conversion is also slower with over 10 minutes.
 
-Clustering the samples is also fastest in \variantSpark{} (1min and 20secs, see Table 1), despite \variantSpark{} and ADAM both using {\sc Spark}'s MLlib \kMeans{} implementation. % 80
+Clustering the samples is also fastest in \variantSpark{} (1min and 20secs), despite \variantSpark{} and ADAM both using {\sc Spark}'s MLlib \kMeans{} implementation, see Table~\ref{fivewaycomparison}.
+% 80
 % DENIS: how sure are we that this is not just a fluke, e.g. I typically repeat this 5 times and report the average and std/ste ...
 % AIDAN: This result was consistent. But it was on the Azure cluster so I can't get std/ste..
-% DENIS: OK but I thought all of them were run on the same machine and you are still generating results for python ... 
-% TODO:  write something like (repeated X times)
-We attribute the 30\% speedup over ADAM (1min and 52secs) to \variantSpark{} converting the VCF files to sparse vectors, whereas ADAM creates dense vectors, which are less memory efficient. %
-Multithreading \kMeans{} clustering is natively supported by both R and Python, we are therefore able to utilise eight cores. 
-Despite this, both are an order of magnitude slower than \variantSpark{} and ADAM, with Python requiring 11mins and 29secs and R requiring 7mins and 25secs.
-{\sc Admixture} is also slower with a runtime of just over 8 minutes. 
-Our Hadoop implementation performs the worst with 14mins and 23secs, showcasing the limitations of the Hadoop engine. % 863
-The substantially increased runtime is due to the dataset (chromosome 22) being small enough to fit into memory, which enables Python and R to store the output from each \kMeans{} iteration in memory.
+% DENIS: OK but I thought all of them were run on the same machine and you are still generating results for python ... in any case can you write something like (repeated X times)
+We attribute the 30\% speedup over ADAM (1min and 52secs) to \variantSpark{} converting the VCF files to sparse vectors, whereas ADAM creates dense vectors which are less memory efficient. %
+Multithreading \kMeans{} clustering is natively supported by both R and Python, we are therefore able to utilise 8 cores. 
+Despite this, both were an order of magnitude slower than \variantSpark{} and ADAM, with python requiring 11mins and 29secs and R requiring 7mins and 25secs.
+Our Hadoop implementation performed the worst with 14mins and 23secs. % 863
+The substantially increased runtime is due the dataset (chromosome 22) being small enough to fit into memory.
+This enables Python and R to store the output from each \kMeans{} iteration in memory.
 Hadoop, however, writes the entire output of each iteration to disk.
-Although this Hadoop feature allows for greater scalability, as disk IO is not the run-time determining factor for larger datasets, it introduces a bottleneck for smaller datasets.
-{\sc Spark} overcomes this limitation by performing in-memory caching of datasets, which eliminates the disk IO bottleneck on smaller datasets, and minimises the bottleneck on larger datasets where at least a portion of the data can be cached in memory.
+Although this feature allows for greater scalability, as memory availability is less of an issue with huge datasets, the bottleneck introduced from disk IO increases processing time notably.
+{\sc Spark} overcomes this limitation by performing in-memory caching of datasets.
+Caching virtually eliminates the disk IO bottleneck on smaller datasets, and minimises the bottleneck on larger datasets where at least a portion of the data can still be cached in memory.
 
-We also investigate the cluster quality for the six different methods by comparing the annotated super-population label (AMR, AFR, EAS, SAS) for each individual in the 1000 Genomes data to the assigned label from the clustering allowing four clusters.
-%TODO: 0 seems not to be the lowest number as we get a negative score for PGP
+We also investigate the cluster quality for the five different methods by comparing the annotated super-population label (AMR, AFR, EAS, SAS) for each individual in the 1000 Genomes data to the assigned label from the clustering allowing four clusters.
 For this comparison, we use the \ARI{} (ARI) metric, which returns a value between 1 (perfect) and 0 (random clustering). 
 Using chromosome 22, each of the algorithms resulted in an ARI of 0.84, confirming that the speed-up in {\sc Spark} is not at the cost of quality.
-Even the state-of-the-art tool, {\sc Admixture} does returns a low ARI=0.25.
 We can substantially improve the accuracy to a perfect classification (ARI=1.0) by removing the fourth super-population, AMR (American). 
@@
 We run this comparison on our in-house Hadoop cluster using 40 cores.
-Pre-processing the VCF files takes approximately 12 minutes, 19 minutes, 27 minutes and 40 minutes for 20\%, 40\%, 60\% and 100\% of the genome, respectively (see Table 2).
+%We run this comparison on our in-house Hadoop cluster using 64 cores, although this was limited to 40 cores with 100\% of the genome due to memory availability.
+Pre-processing the VCF files takes approximately 12 minutes, 19 minutes, 27 minutes and 40 minutes for 20\%, 40\%, 60\% and 100\% of the genome, respectively.
 In each case, the memory required does not exceed a modest 2GB per executor, even for 100\% of the genome. 
 Clustering the data takes 70 minutes, 139 minutes, 213 minutes and 884 mins, respectively.
-For clustering, the memory requirements increase approximately linearly to 24GB per executor (see Figure 1). 
+For clustering, the memory requirements increase approximately linearly with 6GB, 12GB, 17GB and 24GB, respectively, per executor. 
+This substantial increase in memory is a result of more \kMeans{} centroids being created, which MLlib stores as dense vectors. 
+%DENIS: do you see a change in ARI from chr22 to 20%, 40% ?
+%AIDAN: not really.. 
+%DENIS: so it is still exactly 0.84? Can you provide the number of Variants used in the different sets (e.g. is 20\% less than chrom 22)
+%TODO A citation wold be good here
+We do not observe an increase in cluster accuracy when providing variants from other chromosomes indicating that current practice of clustering based on a subset of the genome is sufficient to define the population boundaries~\cite{Hunter2014}.
+
 %This substantial increase in memory required is a result of the increased dimensionality of the data causing the \kMeans{} centroids, which MLlib stores as dense vectors, requiring excessive memory.
@@
 % DENIS: That means you allow more clusters ?
-The increase in memory is a result of more \kMeans{} centroids being created, which MLlib stores as dense vectors. 
 While \kMeans{} clustering can scale to 100\% of the genome, machine learning algorithms that can deal with sparse vectors would potentially be able to scale much further, for example process more alleles.
-
-%DENIS: do you see a change in ARI from chr22 to 20%, 40% ?
-%AIDAN: not really.. 
-%DENIS: so it is still exactly 0.84? Can you provide the number of Variants used in the different sets (e.g. is 20\% less than chrom 22)
-%TODO A citation would be good here
-We do not observe an increase in cluster accuracy when providing variants from other chromosomes, indicating that current practice of clustering based on a subset of the genome is sufficient to define the population boundaries~\cite{Hunter2014}.
 
@@
 To demonstrate the versatility of \variantSpark{} we also process and cluster individuals from the Personal Genomics Project (PGP).
-The PGP is open to the public for individuals to submit their genomic sequence along with any metadata, such as diseases or clinical features~\cite{Lunshof2013}.
-We obtain and curate the data (see methods) resulting in 985,790 variants from 478 individuals.
+The PGP is open to the public for individuals to submit their genomic sequence along with any metadata, such as diseases or clinical features ~\cite{Lunshof2013}.
+The VCF file we use contains 985,790 variants.
 
-We cluster the individuals allowing five clusters and compare the assigned labels to the same number of self-reported `Race/ethnicity' labels (`White', `Hispanic or Latino', `Asian', `Black or African American' and `American Indian').
+Again we cluster the individuals allowing 5 clusters and compare the assigned labels to the same number of self-reported `Race/ethnicity' labels (`White', `Hispanic or Latino', `Asian', `Black or African American' and `American Indian').
 The observed ARI is negative, indicating the clustering is completely random. 
@@
 For example, the `White' label includes individuals with grandparents from United States, Syria, Arab Republic, or Bulgaria, amongst others.
+%This poor result may be due to the labels being too broad and being a blanket label for a diverse range of individuals.
+For example, the `White' label includes individuals with grandparents from United States to Syrian Arab Republic to Bulgaria.
 The dataset is also very one-sided with 433 out of the 478 (over 90\%) of the individuals being classified as white, even though geographical location of their grandparents are very diverse.
@@
 
-This highlights the issue for clinical application where ancestry influences treatment (e.g.
+This highlights the issue for clinical application where ancestries influences treatment (e.g.
 HLA allele genotyping from SNP information~\cite{Zheng2014}) and accurate population association may not be known for patients with diverse migrational background.
 Hence fast genome-wide applicable approaches may help in providing the necessary depth to provide insight into genetic diversity.
-% DENIS: statement about why we cannot do more with this data...
-Unfortunately, this hypothesis could no not be tested as whole genome information is not available for these individuals.
+
+%Although we see an increase, the relatively small ARI is likely due to the diversity in this real world clinical cohort across the small sample size.
+Although our filter removes individuals with a mixed background, we are left with only 36 individuals from 23 different countries.
+%Applying a super-population mask, based on geographical distance, to these individuals reduces the number of groups from 23 to 5, however this grouping doesn't take migrations that occurred prior to their grandparents into account.
 
@@
 On the benchmarking dataset, it outperforms ADAM by almost an order of magnitude (4 vs 28 minutes) by processing VCF files directly and storing the information in sparse vectors. % 258 vs 1673 
-\variantSpark{} utilises {\sc Spark}, which allows for in-memory cashing and hence performs 85\% faster than Hadoop (29 min).
-This novel parallelization approach is also faster than traditional multithreading with 94\% speedup over R (75 min), and 95\% over Python (103 min).
-These benefits of speed, resource consumption and scalability demonstrate \variantSpark{} to be the interface for applying other machine learning tools from MlLib to genomic data.
+\variantSpark{} utilizes {\sc Spark}, which allows for in-memory cashing and hence performs 85\% faster than Hadoop (29 min), 94\% faster than R (75 min), and 95\% faster than Python (103 min).
+
+%\variantSpark{} excels at pre-processing VCF files into row vectors, being over an order of magnitude faster than ADAM.
+Memory consumption is also kept to a minimum.
+Where traditional approaches such as Python and R would fail when expanding to the whole genome,
+%\variantSpark{} can pre-process VCF files with over 1000 individuals in under half an hour using minimal memory.
+
+%Clustering these individuals is also much faster with \variantSpark{} than Python and R, although only marginally faster than ADAM.
+
+%These benefits of speed, resource consumption and scalability demonstrate \variantSpark{} to be a useful tool in applying machine learning tools to genomic data.
+Although we only run \kMeans on the data, MLlib provides many other machine learning algorithms, such as Random Forests.
+%Therefore it may be useful to investigate the novelty of applying these machine learning approaches to big genomic data.
+
+%completes the whole variant clustering task in just over 4 minutes, which is 80\% faster than ADAM (28 mins), 85\% faster than Hadoop (29 min), 94\% faster than R (75 min), and 95\% faster than Python (103 min).
 
@@
 PGP genome data was downloaded from the PGP website (\url{https://my.pgp-hms.org/public_genetic_data}). Genotype data from 23andMe microarray platforms was sorted by genome 
-build. Those files corresponding to NCBI build 36 were converted to BED format, updated to build 37 using UCSC liftOver and then converted back to 23andMe format. The 
-23andMe files were then converted to VCF format using code obtained from the Broad Institute (\url{http://apol1.blogspot.com.au/2013/08/impute-apoe-and-apol1-with-23andme.html}). 
+build. Shell scripts were written to convert NCBI build 36 files to BED format, update to build 37 using UCSC liftOver (\url{https://genome.ucsc.edu/cgi-bin/hgLiftOver}) and 
+then convert back to 23andMe format.
+The 23andMe files were then converted to VCF format using code obtained from the Broad Institute (\url{http://apol1.blogspot.com.au/2013/08/impute-apoe-and-apol1-with-23andme.html}).
 Where individuals had genotype data for both genome builds 36 and 37, only the latter was used. Finally, individual VCF files were combined for clustering into one 
 file using VCFtools \cite{Danecek2011Variant}.
-
-% DENIS yes, it will be in the git repository mentioned above (https://github.com/BauerLab/VariantSpark)
-% can you please push your code into https://stash.csiro.au/users/bau04c/repos/variantspark/browse
-[if code to be made available, say so here and where it will be]
 
@@
   \begin{figure}[h!]
-  \includegraphics[type=pdf,ext=.pdf,read=.pdf, scale=0.45]{images/Resources}
-  \includegraphics[type=pdf,ext=.pdf,read=.pdf, scale=0.45]{images/Scaling}
-  \caption{\csentence{Comparison of method and genome-wide scaling experiment.}
-      {\bf Left} Runtime is given in seconds with 32 GB of memory on 8 threads (except for the pre-processing in R and Python where this was not supported). 
-      {\bf Right} Scaling from 20\% to 100\% of variants in the genome with maximal number of executors and lowest possible memory assignment.}
-      \end{figure}
-
-%Hadoop
-%1. Map
-%The mapper converts each line of the VCF file into a preliminary key-value pair, where the Key is the byte-position of the first character thereby functioning as a proxy for the genomic location (variant ID), and the Value is a string of the line.
-From this, the key-value pairs for the individual genotypes are created by extracting the variant from each sample as the new Value.
-The new key is a tuple consisting of the primary key, which is the individual ID (position of the parsed variants), and the variant ID as the secondary key.
-%The MapReduce partitioner uses custom Collect/sort functions to collect the tuples by their primary key (individual ID) and distributes them between nodes.
-The data is sorted by secondary key (variant ID)
-%2. Reduce
-%This data is input to the Reducers, which store the variants for each individual to a sparse vector with the Key as the individual ID and the value as the variant vector.
-%
-%Spark
-%1. The lines from the VCF file(s) are parsed independently by worker nodes, where each line is zipped with the heading from the VCF file resulting in tuples of individual ID and the variant. 
-%2.
-These arrays are each zipped again with the unique, sequential index (the Variant ID) resulting in tuples (essentially key/value pairs) of Variant IDs (?key?) and the array of tuples from step 1 (?value?).
-%3. The flatmap function restructures the data into Individual ID (?key?) and tuples with Variant ID and the Variant (?value?).
-%4. These tuples are grouped by the Individual ID, collecting all Variant ID and Variant tuples, and non-zero variants are saved as a sparse vector.
-\begin{figure}[h!]
-\includegraphics[type=pdf,ext=.pdf,read=.pdf, scale=0.25]{signature}
-  \caption{\csentence{Schematic of the Hadoop and Spark implementation.} The image shows the conceptual differences in the steps necessary to convert variants in VCF file format to a data structure readable by Mahout and MLlib, respectively.}
+  \caption{\csentence{Sample figure title.}
+      A short description of the figure content
+      should go here.}
       \end{figure}
@@
 \begin{table}[h!]
-\caption{The resources consumption of the six compared methods as well as the accuracy measured as \ARI{} on chromosome 22.}
+\caption{The resources consumption of the five compared methods as well as the accuracy measured as \ARI{} on chromosome 22.}
       \begin{tabular}{|c|ccc|ccc|c|c|}
@@
 \variantSpark{}	& 8	& 32	& 2min 58sec	& 8	& 32	& 1min 20sec	& 0.84	\\ 
-ADAM		& 8	& 32	& 12min 48sec	& 8	& 32	& 1min 52sec	& 0.84	\\
-Hadoop		& 8	& 32	& 14min 22sec	& 8	& 32	& 14min 23sec	& 0.84	\\
-R			& 1	& 32	& 67min		& 8	& 32	& 7min 25sec	& 0.84	\\
+ADAM		& 8	& 32	& 12min 48sec	& 8	& 32	& 1min 52sec	& x	\\
+Hadoop		& 8	& 32	& 14min 22sec	& 8	& 32	& 14min 23sec	& x	\\
+R			& 1	& 32	& 67min		& 8	& 32	& 7min 25sec	& x	\\
 Python		& 1	& 32	& 92min		& 8	& 32	& 11min 29sec	& 0.84	\\
-ADMIXTURE	& 1	& 32	& 10min 8sec	& 8	& 32 & 8min 19sec	& 0.25	\\
   \hline

### 2015-09-17 doc/publications/journal2014/paper.tex
Re-edited PGP methods section

--- 2015-09-16 doc/publications/journal2014/paper.tex
@@
 PGP genome data was downloaded from the PGP website (\url{https://my.pgp-hms.org/public_genetic_data}). Genotype data from 23andMe microarray platforms was sorted by genome 
-build. Those files corresponding to NCBI build 36 were converted to BED format, updated to build 37 using UCSC liftOver and then converted back to 23andMe format. The 
-23andMe files were then converted to VCF using code obtained from the Broad Institute (\url{http://apol1.blogspot.com.au/2013/08/impute-apoe-and-apol1-with-23andme.html}). 
+build. Shell scripts were written to convert NCBI build 36 files to BED format, update to build 37 using UCSC liftOver (\url{https://genome.ucsc.edu/cgi-bin/hgLiftOver}) and 
+then convert back to 23andMe format.
+The 23andMe files were then converted to VCF format using code obtained from the Broad Institute (\url{http://apol1.blogspot.com.au/2013/08/impute-apoe-and-apol1-with-23andme.html}).
 Where individuals had genotype data for both genome builds 36 and 37, only the latter was used. Finally, individual VCF files were combined for clustering into one 
@@
 % can you please push your code into https://stash.csiro.au/users/bau04c/repos/variantspark/browse
-[if code to be made available, say so here and where it will be]
 

### 2015-09-17 doc/publications/journal2014/paper.tex
Added Gephi citation; cluster visualization = figure 3; some text for graph visualization

@@
 \subsection*{Graph visualization}
-Add the image here \ldots
+%% Add the image here \ldots
+A visualization of the clusters generated by \variantSpark{} from 1000 Genomes Project data, generated using Gephi \cite{ICWSM09154}, is 
+shown in Figure 3. Three of the four clusters (super-populations AFR, EAS and AMR) are relatively-homogeneous; the fourth is more mixed and 
+consists predominantly of individuals labelled EUR and AMR.
 
@@
 \begin{figure}[h!]
-  \caption{\csentence{Sample figure title.}
-      Figure legend text.}
+\includegraphics[type=pdf,ext=.pdf,read=.pdf, scale=0.25]{chr22-k4-gephi}
+  \caption{\csentence{Visualization of \variantSpark{} clusters.} Individuals from the 1000 Genomes project are clustered effectively by super-populations AFR, AMR and EAS; a fourth cluster is predominantly EUR + AMR.}
       \end{figure}

### 2015-09-17 doc/publications/journal2014/paper.tex
Figure 3 uses PDF with legend
added citation

@@
 Although {\sc Spark} allows MapReduce-like programs, it does not require programs to exactly model the MapReduce paradigm, which in turn allows more flexibility when designing programs. 
-Recognising the capability, the Big Data Genomics (BDG) group recently demonstrated the strength of {\sc Spark} in a genomic clustering application using {\sc ADAM}, a set of formats and APIs as well as processing stage implementations for genomic data~\cite{Massie2013}.
-While the speedup over traditional methods was impressive (50 fold speedup), being limited by constraints within this general genomics framework hampered performance.
+Recognising the capability, Wiewi{\'o}rka {\it et al.}~\cite{Wiewiorka2014} developed SparkSeq for high-throughput sequence data analysis and the Big Data Genomics (BDG) group recently demonstrated the strength of {\sc Spark} in a genomic clustering application using {\sc ADAM}, a set of formats and APIs as well as processing stage implementations for genomic data~\cite{Massie2013}.
+While the speedup of {\sc ADAM} over traditional methods was impressive (50 fold speedup), being limited by constraints within this general genomics framework hampered performance.
 

### 2015-09-17 doc/publications/journal2014/paper.tex
updates

@@
 In the first section we benchmark \variantSpark's performance and accuracy against {\sc ADAM} and a Hadoop Mahout implementation as well as more traditional methods (R, Python) and the purpose-build tool {\sc Admixture}~\cite{Alexander2009} limiting the used data to chromosome 22.
-In section two we demonstrate \variantSpark's full capacity by seamlessly scaling from 20\% to 100\% of the human genome.
-In section three we replicate the analysis by using 478 genomes from the Personal Genome Project~\cite{Lunshof2013}.
-In the last section we discuss the pipeline for visualising the resulting cluster.
+In the second section we discuss the pipeline for visualising the resulting cluster.
+In section three we demonstrate \variantSpark's full capacity by seamlessly scaling from 20\% to 100\% of the human genome.
+In the last section we replicate the analysis by using 478 genomes from the Personal Genome Project~\cite{Lunshof2013}.
 
@@
 Using chromosome 22, each of the algorithms resulted in an ARI of 0.84, confirming that the speed-up in {\sc Spark} is not at the cost of quality.
-Even the state-of-the-art tool, {\sc Admixture} does returns a low ARI=0.25.
+Even the state-of-the-art tool, {\sc Admixture}, does return a low ARI=0.25.
+It should be noted that {\sc Admixture}'s underlying statistical model does not take linkage disequilibrium (LD) into account.
+Therefore removing variants with high LD may result in higher accuracy.   
 We can substantially improve the accuracy to a perfect classification (ARI=1.0) by removing the fourth super-population, AMR (American). 
-Including AMR individuals, we observe that the majority of these individuals are placed in the same group as Europeans, likely accurately reflecting their migrational backgrounds. 
+This is due to the majority of AMR individuals being placed in the same group as Europeans, likely accurately reflecting their migrational backgrounds. 
 Only a minority of AMR individuals form an independent group, likely comprising of genetic information otherwise not captured by the 26 sub-populations of the 1000 genomes project.
-This indicates that although allelic differences exist between populations, "genetic diversity is distributed on a continuum"~\cite{Hunter2014}.
+This indicates that although allelic differences exist between populations, "genetic diversity is distributed on a continuum"~\cite{Pugach2015}.
+
+\subsection*{Graph visualisation}
+\variantSpark{} also supports visualisation of the resulting population clusters by generating a file that can be loaded in Gephi \cite{ICWSM09154}.
+Figure 3 visualises the clusters generated by \variantSpark{} from the 1000 Genomes Project data. 
+As discussed in the previous section, three of the four clusters (super-populations AFR, EAS and AMR) are relatively-homogeneous; the fourth is more mixed and 
+consists predominantly of individuals labelled EUR and AMR providing a visual representation of the ARI of 0.84.
 
 \subsection*{\variantSpark{} allows genome wide sampling of variants to improve clustering quality}
-
 Although we have demonstrated the advantages of \variantSpark{} over traditional methods on small datasets, the main advantage of {\sc Spark} is its ability to process datasets that exceed memory limits.
@@
 For clustering, the memory requirements increase approximately linearly to 24GB per executor (see Figure 1). 
-%This substantial increase in memory required is a result of the increased dimensionality of the data causing the \kMeans{} centroids, which MLlib stores as dense vectors, requiring excessive memory.
-% DENIS: This conflicts with the 30 seconds gain over {\sc ADAM} which you attribute to our clustering using sparse vectors,
-% AIDAN: Only centroids are dense. Reworded to explain this
-% DENIS: That means you allow more clusters ?
-The increase in memory is a result of more \kMeans{} centroids being created, which MLlib stores as dense vectors. 
+%TODO: doubleckeck
+The increase in memory is due to more distance measurements between variants and \kMeans{} centroids being created, which MLlib stores as dense vectors. 
 While \kMeans{} clustering can scale to 100\% of the genome, machine learning algorithms that can deal with sparse vectors would potentially be able to scale much further, for example process more alleles.
 
-%DENIS: do you see a change in ARI from chr22 to 20%, 40% ?
-%AIDAN: not really.. 
-%DENIS: so it is still exactly 0.84? Can you provide the number of Variants used in the different sets (e.g. is 20\% less than chrom 22)
-%TODO A citation would be good here
-We do not observe an increase in cluster accuracy when providing variants from other chromosomes, indicating that current practice of clustering based on a subset of the genome is sufficient to define the population boundaries~\cite{Hunter2014}.
+We do not observe an increase in cluster accuracy when providing variants from other chromosomes, indicating that current practice of clustering based on a subset of the genome is sufficient to define the existing population boundaries~\cite{Pugach2015}.
 
@@
 HLA allele genotyping from SNP information~\cite{Zheng2014}) and accurate population association may not be known for patients with diverse migrational background.
-Hence fast genome-wide applicable approaches may help in providing the necessary depth to provide insight into genetic diversity. 
-% DENIS: statement about why we cannot do more with this data...
-Unfortunately, this hypothesis could no not be tested as whole genome information is not available for these individuals.
-
-\subsection*{Graph visualization}
-%% Add the image here \ldots
-A visualization of the clusters generated by \variantSpark{} from 1000 Genomes Project data, generated using Gephi \cite{ICWSM09154}, is 
-shown in Figure 3. Three of the four clusters (super-populations AFR, EAS and AMR) are relatively-homogeneous; the fourth is more mixed and 
-consists predominantly of individuals labelled EUR and AMR.
+As noted by Patterson et al ~\cite{Patterson2006}, more markers are needed for populations with low genetic divergence. 
+It is hence likely that higher density genotyping (e.g.
+from genome sequencing) would help in elucidating population structure in this dataset, which demonstrates the need for fast whole-genome approaches.
+Unfortunately, this hypothesis can no not be tested as whole genome information is not available for these individuals.
 
@@
 \variantSpark{} utilises {\sc Spark}, which allows for in-memory caching and hence performs 85\% faster than Hadoop (29 min).
-This novel parallelization approach is also faster than traditional multithreading with 94\% speedup over R (75 min), and 95\% over Python (103 min).
+This novel parallelisation approach is also faster than traditional multithreading with 94\% speedup over R (75 min), and 95\% over Python (103 min).
 \variantSpark{} is also superior in performance and accuracy over the current state-of-the-art tool for individual ancestries determination, {\sc Admixture}.
@@
 We also used a subset of this dataset, the variants from chromosome 22, and this contains 494,328 alleles.
-These individuals from these datasets are distributed across four super populations, African (AFR), Mixed American (AMR), East Asian (EAS) and European (EUR).
+The individuals from these datasets are distributed across four super populations, African (AFR), Mixed American (AMR), East Asian (EAS) and European (EUR).
 
@@
 
-\subsection*{ADMIXTURE Implementation}
+\subsection*{{\sc Admixture} Implementation}
 We use Genome Analysis Toolkit (GATK) to convert the variants from VCF, and a 1000 Genomes Project supplied .ped file, to a binary PLINK (.bed), binary marker information file (.bim) and pedigree stub file (.fam).
-These three files are used as input to ADMIXTURE, with K (the number of ancestral populations) set to 4.
+These three files are used as input to {\sc Admixture}, with $K$ (the number of ancestral populations) set to 4.
 
@@
 \section*{Author's contributions}
-    Text for this section \ldots
+    DCB and FAB designed the overall study. ARO implemented \variantSpark\ and conducted the comparison. NS obtained the PGP data and generated the graph.
+    All authors contributed to the writing of the document.
 
@@
 \begin{figure}[h!]
-\includegraphics[type=pdf,ext=.pdf,read=.pdf, scale=0.25]{chr22-k4-gephi}
+\includegraphics[type=pdf,ext=.pdf,read=.pdf, scale=0.40]{chr22-k4-gephi-legend}
   \caption{\csentence{Visualization of \variantSpark{} clusters.} Individuals from the 1000 Genomes project are clustered effectively by super-populations AFR, AMR and EAS; a fourth cluster is predominantly EUR + AMR.}

### 2015-09-17 doc/publications/journal2014/paper.tex
abstract

@@
 \begin{abstract} % abstract
-\parttitle{Motivation} Genomic information is increasingly used in medical practice, giving rise to the need for efficient analysis methodology able to cope with thousands of individuals and millions of variants.
+\parttitle{Motivation} Genomic information is increasingly used in medical practice giving rise to the need for efficient analysis methodology able to cope with thousands of individuals and millions of variants.
 The widely used Hadoop MapReduce architecture and associated machine learning library, Mahout, provide the means for tackling computationally challenging tasks.
 However, many genomic analyses do not fit the Map-Reduce paradigm.
-We therefore survey the recently developed {\sc Spark} engine, along with its associated machine learning library, MLlib, which are more flexible in the parallelisation architecture and hence more suitable for population-scale bioinformatics tasks.
-To do this, we developed an interface from Mahout and MLlib to the standard variant format (VCF), which opens up the usage of advanced, efficient machine learning algorithms to genomic data.
+We therefore utilise the recently developed {\sc Spark} engine, along with its associated machine learning library, MLlib, which offers more flexibility in the parallelisation of population-scale bioinformatics tasks.
+\variantSpark{} provides an interface from MLlib to the standard variant format (VCF), offers seamless genome-wide sampling of variants and provides a pipeline for visualising results.
 
-\parttitle{Results} We successfully clustered more than 2,500 individuals each having more than 38 Million variants.
-Both Hadoop and {\sc Spark} have superior performance over traditional approaches.
-We observe a 50 fold speedup when using the efficient in-memory {\sc Spark} compute engine compared to Hadoop.
-Furthermore, our implementation achieves a better performance compared to a previously published {\sc Spark}-based genome clustering approach, {\sc ADAM}.
+\parttitle{Results} To demonstrate the capabilities of \variantSpark, we clustered more than 1,000 individuals with 38 Million variants each to determine the population structure in the dataset.
+\variantSpark{} is 80\% faster than the {\sc Spark}-based genome clustering approach, {\sc ADAM}, the comparable implementation using Hadoop/Mahout, as well as a {\sc Admixture}, a commonly used tool for determining individual ancestries.
+It is over 90\% faster than traditional implementations using R and Python.
+These benefits of speed, resource consumption and scalability enables \variantSpark{} to open up the usage of advanced, efficient machine learning algorithms to genomic data.
 
@@
 \variantSpark{} is also superior in performance and accuracy over the current state-of-the-art tool for individual ancestries determination, {\sc Admixture}.
-These benefits of speed, resource consumption and scalability demonstrate \variantSpark{} to be the interface for applying other machine learning tools from MlLib to genomic data.
+These benefits of speed, resource consumption and scalability allows \variantSpark{} to be the interface for applying other machine learning tools from MlLib to genomic data.
 

### 2015-09-17 doc/publications/journal2014/paper.tex
Update methods

@@
 Although we see an increase in cluster-accuracy when removing individuals with a mixed background and operating at super-population level, the clustering is not as precise as on the carefully curated and characterised individuals from the 1000 Genomes Project.
-%TODO add numbers
-This is especially the case since XX\% are `White' Americans, which formed the group most difficult to cluster in the 1000 Genomes Project data.
+This is especially the case since 65\% are `White' Americans, which formed the group most difficult to cluster in the 1000 Genomes Project data.
 
@@
 \subsection*{VariantSpark Implementation}
-For each of the implementations, we pre-process the VCF files to arrays of features.
+To cluster the individuals in a VCF file, we initially need to pre-process the variants to feature vectors.
 In \variantSpark{} we read in the VCF files as text files to a Resilient Data Set (RDD).
 RDDs allow us to process the files, line-by-line, in parallel.
-We parse each line as tab-separated values, and `zip' each value with it's heading from the VCF file.
-The key-value pairs (KVPs) from each line in the VCF file are stored in separate arrays. We then `zip' each array with a unique index, where these indexes form identifiers for the alleles.
-At this point, we have one array for each line from the VCF file.
-The values in the arrays are KVPs of the parsed tab-separated values zipped with their heading (individual ID), and each array has a unique index.
-We now use a `flatMap' to convert these indexed arrays of KVPs into a less convoluted data-structure, as well as to filter out KVPs that aren't alleles (i.e. KVPs that are VCF metadata).
-This results in KVPs where the key is the individual ID, and the value is a tuple of the allele and it's index.
-We effectively have one KVP for each allele from the VCF file.
-At this point, each allele is still a string straight from the VCF file (i.e.
-`0\textbar{}1'), however, we require doubles for MLlib.
-To convert this to a double, we take the Hamming distance of the allele, where a 1 represents a heterozygous variant, 2 a homozygous variant, and 0, no variant.
+We parse each line as tab-separated values, and `zip' each value with it's heading from the VCF file, as shown in Figure 2.
+The result is an array of key-value pairs (KVPs) for each line, where the column heading is the key and the variant is the value.
+We then `zip' each KVP array with a unique sequential index, where these indexes will serve as identifiers for the alleles.
+Now each KVP array is the value in a parent KVP, where the key is the unique index.
+We use a `flatMap' (step 3 of Figure 2) to convert these indexed arrays of KVPs into a less convoluted data-structure, as well as to filter out any extra metadata.
+This results in KVPs (number of variants x number of alleles) where the key is the individual ID, and the value is a tuple of an allele and its index.
+Effectively, we have a KVP for each allele from the VCF file.
+At this point, each allele is still a string straight from the VCF file (i.e. `0\textbar{}1'), however, we require doubles, for MLlib.
+To convert these strings to doubles, we use the Hamming distance of the allele, where a 1 represents a heterozygous variant, 2 a homozygous variant, and 0, no variant.
 Also, for each individual, we filter out any alleles that have no variant, leaving us with KVPs of variants (homozygous and heterozygous) only.
-We can now group these KVPs by their key (the individual ID) and map them to sparse vectors.
+We perform this `map' operation inside the previous `flatMap' (step 3).
+We can now group these KVPs by their key (the individual ID) and save them to sparse vectors (step 4).
 This results in a sparse vector for each individual, the format required by MLlib's machine learning algorithms.
@@
 Because of this, and because we need a unique range of identifiers for alleles (in the smallest range possible), we need to run an initial MapReduce task to index the lines.
-This is comparable (however, more cumbersome) to the `zip' operation in \variantSpark{}.
-The second MapReduce task does the bulk of the work.
-The Map stage begins by creating KVPs from the VCF file.
+This is comparable (however, more verbose) to the `zip' operation in \variantSpark{}.
+The second MapReduce task does the bulk of the work and is visualised in Figure 2.
+The Map stage (step 1) begins by creating KVPs from the VCF file.
 For each KVP, the key is a tuple of a primary and secondary key, where the primary key is the individual ID and the secondary key is the allele ID.
@@
 After being distributed, the KVPs for each individual are sorted by their secondary key.
-Now that the KVPs for each individual are physically located on the same hardware, the Reduce stage can efficiently create a sparse vector for each individual from these KVPs.
+Now that the KVPs for each individual are physically located on the same hardware, the Reduce stage (step 2) can efficiently create a sparse vector for each individual from these KVPs.
 The sparse vectors (one for each individual) are saved to disk, ready to be used as the input to one of MLlib's algorithms.
 
+%% Need to cite R package??
 \subsection*{R Implementation}
-For our R implementation, we read in a VCF file as a table and convert it to a matrix, dropping the columns that don't represent alleles.
+For our R implementation, we read in a VCF file as a table and convert it to a matrix, dropping columns that don't represent alleles.
 As with our \variantSpark{} pre-processing, we convert the strings that represent each allele to a numeric value.
 We then transpose the matrix, resulting in a data-structure where each row represents an individual.
+We convert the matrix to a `big.matrix', for the \kMeans{} algorithm from the `biganalytics' package.
 

### 2015-09-22 doc/publications/journal2014/paper.tex
supp table

@@
 \usepackage{graphicx}
+\usepackage{multirow}
 
@@
 % TODO provide mapping
-We therefore only include individuals where all grandparents were reported to be from the same country leaving 36 individuals from 23 different countries, which we group into their approximate super-populations (see Supplemental Table X).
-Clustering these individuals results in an ARI of 0.29.
+We therefore only include individuals where all grandparents were reported to be from the same country leaving 35 individuals from 22 different countries, which we group into their approximate super-populations (see Supplemental Table X).
+Clustering these individuals results in an ARI of 0.45.
 
@@
 We use Genome Analysis Toolkit (GATK) to convert the variants from VCF, and a 1000 Genomes Project supplied .ped file, to a binary PLINK (.bed), binary marker information file (.bim) and pedigree stub file (.fam).
-These three files are used as input to {\sc Admixture}, with $K$ (the number of ancestral populations) set to 4.
+These three files are used as input to {\sc Admixture}, with default options, and $K$ (the number of ancestral populations) set to 4.
 
@@
 
+\label{pgppops}
+\begin{table}[h!]
+\caption{The resources consumption on different subsets of the entire autosome (chromosomes 1-22). Memory specified is the memory allocated to each executor.}
+      \begin{tabular}{|c|c|c|c|}
+
+Super Population		&	Super Population Size &	Populations	&	Population size \\
+\multirow{4}{*}{AMR} 	&	\multirow{4}{*}{5}	&	Canada	&	2	\\
+					&					&	Mexico	&	1	\\
+					&					&	Puerto Rico               & 1               \\
+					&					&	US Minor Outlying Islands & 1               \\
+\multirow{2}{*}{EAS}		&	\multirow{2}{*}{6}	&	China                     & 5               \\
+					&					&	Taiwan, Province of China & 1	\\
+\multirow{14}{*}{EUR}	&	\multirow{14}{*}{20}	& Belgium				& 1	\\
+					&					& Estonia				& 1	\\
+					&					& Finland				& 1	\\
+					&					& France				& 1	\\
+					&					& Germany			& 1	\\
+					&					& Greece				& 1	\\
+					&					& Ireland				& 3	\\
+					&					& Poland				& 2	\\
+					&					& Russian Federation	& 1	\\
+					&					& Slovenia			& 1	\\
+					&					& Spain				& 2	\\
+					&					& Sweden				& 1	\\
+					&					& Ukraine				& 2	\\
+					&					& United Kingdom		& 2	\\
+\multirow{2}{*}{SAS}		&	\multirow{2}{*}{4}	& Bangladesh			& 1	\\
+					&					& India				& 3              
+      \end{tabular}
+\end{table}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2015-09-22 doc/publications/journal2014/paper.tex
polished methods

--- 2015-09-17 doc/publications/journal2014/paper.tex
@@
 \begin{keyword}
-\kwd{sample}
-\kwd{article}
-\kwd{author}
+\kwd{Genotype clustering}
+\kwd{SPARK}
+\kwd{BigData}
+\kwd{1000 Genomes Project}
+\kwd{Personal Genome Project}
+\kwd{Population structure}
 \end{keyword}
@@
 Genomic information is increasingly used in medical practice.
-A commonly performed task in such applications is grouping individuals based on their genomic profile to identify population association~\cite{Gao2007} or elucidate different haplotype involvement in diseases susceptibility~\cite{Laitman2013}.
+A commonly performed task in such applications is grouping individuals based on their genomic profile to identify population association~\cite{Gao2007} or elucidate haplotype involvement in diseases susceptibility~\cite{Laitman2013}.
 The commonly used tool is {\sc Admixture}~\cite{Alexander2009}, which performs maximum likelihood estimation of individual ancestries from multilocus SNP genotype datasets.
-Due to the decreasing sequencing cost it is economical to generate studies with sample sizes previously reserved for larger consortia such as the 1000 Genomes Project~\cite{1KG2012} or The Cancer Genome Atlas, TCGA~\cite{TCGA2013}.
+Due to the decreasing sequencing cost it is now economical to generate studies with sample sizes previously reserved for larger consortia such as the 1000 Genomes Project~\cite{1KG2012} or The Cancer Genome Atlas, TCGA~\cite{TCGA2013}.
 At the same time, whole genome sequencing enables the inclusion of rare or even somatic mutations in the analysis, increasing the feature space by orders of magnitude.
 This drastic increase in both sample numbers and features per sample requires a massively parallel approach to data processing~\cite{Stein2010}.
-Traditional parallelisation strategies like OpenMPI or hardware accelerators (GPGPU) cannot scale with variable data sizes or require purpose-built hardware.
+Traditional parallelisation strategies like MPI/OpenMP or hardware accelerators (GPGPU) cannot scale with variable data sizes at runtime~\cite{ReyesOrtiz2015121} or require purpose-built hardware.
 
@@
 Unfortunately, the MapReduce paradigm is not always the optimal solution, specifically for bioinformatics or machine learning applications that require iterative in-memory computation.
-Specifically, Hadoop is disk IO intensive, and this can prove to be a bottleneck in processing-speed.
+Specifically, Hadoop is relying extensively on hard disk input-output operations (disk IO), and this can prove to be a bottleneck in processing-speed.
 
@@
 One of the main benefits is that it allows programs to cache data in memory; potentially eliminating, or at least reducing, the bottleneck of disk IO. 
-When utilising caching, Apache claim {\sc Spark} to be 100x faster than Hadoop. 
-Although {\sc Spark} allows MapReduce-like programs, it does not require programs to exactly model the MapReduce paradigm, which in turn allows more flexibility when designing programs. 
+When utilising caching, Apache claim {\sc Spark} to be up to 100x faster than Hadoop. 
+Although {\sc Spark} allows MapReduce-like programs, it does not require programs to exactly model the MapReduce paradigm, which in turn allows more flexible software design. 
 Recognising the capability, Wiewi{\'o}rka {\it et al.}~\cite{Wiewiorka2014} developed SparkSeq for high-throughput sequence data analysis and the Big Data Genomics (BDG) group recently demonstrated the strength of {\sc Spark} in a genomic clustering application using {\sc ADAM}, a set of formats and APIs as well as processing stage implementations for genomic data~\cite{Massie2013}.
@@
 
-We hence developed a more focused purpose-built application in {\sc Spark} to perform genomic clustering of individuals. 
+We hence developed a purpose-built application in {\sc Spark} to perform genomic clustering of individuals. 
 We utilise {\sc Spark}'s machine learning library, \mbox{MLlib}, and provide an interface to the standard variant data format, Variant Call Format (VCF)~\cite{1KG2012}, which opens up the application of MLlib's different machine learning algorithms to a wide range of genotype-based analysis tasks.
@@
 To demonstrate \variantSpark's capability, we cluster variant datasets from the 1000 Genomes Project~\cite{1KG2012} to determine population structure using the \kMeans{} clustering algorithm available in MLlib.
-In the first section we benchmark \variantSpark's performance and accuracy against {\sc ADAM} and a Hadoop Mahout implementation as well as more traditional methods (R, Python) and the purpose-build tool {\sc Admixture}~\cite{Alexander2009} limiting the used data to chromosome 22.
+In the first section we benchmark \variantSpark's performance and accuracy against {\sc ADAM} and a Hadoop Mahout implementation as well as more traditional methods (R, Python) and the purpose-build tool {\sc Admixture}~\cite{Alexander2009}.
+% limiting the used data to chromosome 22.
 In the second section we discuss the pipeline for visualising the resulting cluster.
@@
 As shown in Table 1, pre-processing the data is fastest in \variantSpark{}, requiring 2mins and 58secs. %178s 
-This is almost 80\% faster than the {\sc ADAM} implementation (12min and 48secs) %768s
-or our Hadoop implementation (14min and 22secs), respectively. %862s
+This is almost 80\% faster than both the {\sc ADAM} implementation (12min and 48secs) %768s
+and our Hadoop implementation (14min and 22secs). %862s
 Unlike \variantSpark{} and our Hadoop implementation, the {\sc ADAM} framework cannot process the VCF files directly but requires them to be converted into a binary {\sc ADAM} file format. 
@@
 R and Python are also slower at pre-processing, taking 67 minutes and 92 minutes, respectively. 
-This increased pre-processing time is due to the Python and R implementations not natively supporting multithreading, while \variantSpark{}, {\sc ADAM} and Hadoop can use the eight available cores.
+This increased pre-processing time is due to the standard Python and R implementations not natively supporting multithreading, while \variantSpark{}, {\sc ADAM} and Hadoop can use the eight available cores.
 The GATK-based VCF to PED conversion is also slower with over 10 minutes.
@@
 Clustering the samples is also fastest in \variantSpark{} (1min and 20secs, see Table 1), despite \variantSpark{} and {\sc ADAM} both using {\sc Spark}'s MLlib \kMeans{} implementation. % 80
-% DENIS: how sure are we that this is not just a fluke, e.g. I typically repeat this 5 times and report the average and std/ste ...
-% AIDAN: This result was consistent. But it was on the Azure cluster so I can't get std/ste..
-% DENIS: OK but I thought all of them were run on the same machine and you are still generating results for python ... 
-% TODO:  write something like (repeated X times)
 We attribute the 30\% speedup over {\sc ADAM} (1min and 52secs) to \variantSpark{} converting the VCF files to sparse vectors, whereas {\sc ADAM} creates dense vectors, which are less memory efficient.
@@
 {\sc Admixture} is also slower with a runtime of just over 8 minutes. 
-Our Hadoop implementation performs the worst with 14mins and 23secs, showcasing the limitations of the Hadoop engine. % 863
-The substantially increased runtime is due to the dataset (chromosome 22) being small enough to fit into memory, which enables Python and R to store the output from each \kMeans{} iteration in memory.
-Hadoop, however, writes the entire output of each iteration to disk.
-Although this Hadoop feature allows for greater scalability, as disk IO is not the run-time determining factor for larger datasets, it introduces a bottleneck for smaller datasets.
-{\sc Spark} overcomes this limitation by performing in-memory caching of datasets, which eliminates the disk IO bottleneck on smaller datasets, and minimises the bottleneck on larger datasets where at least a portion of the data can be cached in memory.
+Our Hadoop implementation performs the worst with 14mins and 23secs, showcasing the limitations of the Hadoop engine for iterative algorithms such as \kMeans{}. % 863
+Hadoop writes the entire output of each \kMeans{} iteration to disk.
+This feature was designed to ensure scalability but here substantially increases the runtime as disk IO becomes the limiting factor for this small dataset (chromosome 22).
+Python and R store the output from each iteration in memory thereby eliminating the IO bottleneck but without being able to utilise the disk they are limited to datasets that fit into memory.  
+{\sc Spark} overcomes both limitations by performing in-memory caching for each executor, which eliminates the disk IO bottleneck on smaller datasets, while maintaining scalability to larger datasets where the data actively processed on executors is cashed in memory.
 
@@
 Using chromosome 22, each of the algorithms resulted in an ARI of 0.84, confirming that the speed-up in {\sc Spark} is not at the cost of quality.
-Even the state-of-the-art tool, {\sc Admixture}, does return a low ARI=0.25.
-It should be noted that {\sc Admixture}'s underlying statistical model does not take linkage disequilibrium (LD) into account.
+The state-of-the-art tool, {\sc Admixture}, returns a low ARI of 0.25. It should be noted that {\sc Admixture}'s underlying statistical model does not take linkage disequilibrium (LD) into account. 
 Therefore removing variants with high LD may result in higher accuracy.   
 We can substantially improve the accuracy to a perfect classification (ARI=1.0) by removing the fourth super-population, AMR (American). 
-This is due to the majority of AMR individuals being placed in the same group as Europeans, likely accurately reflecting their migrational backgrounds. 
+This is due to the majority of AMR individuals being placed in the same group as Europeans, likely reflecting their migrational backgrounds. 
 Only a minority of AMR individuals form an independent group, likely comprising of genetic information otherwise not captured by the 26 sub-populations of the 1000 genomes project.
@@
 \subsection*{Graph visualisation}
-\variantSpark{} also supports visualisation of the resulting population clusters by generating a file that can be loaded in Gephi \cite{ICWSM09154}.
+\variantSpark{} supports visualisation of the resulting population clusters by generating a file that can be loaded into {\sc Gephi}~\cite{ICWSM09154}.
 Figure 3 visualises the clusters generated by \variantSpark{} from the 1000 Genomes Project data. 
@@
 
-\subsection*{\variantSpark{} allows genome wide sampling of variants to improve clustering quality}
-Although we have demonstrated the advantages of \variantSpark{} over traditional methods on small datasets, the main advantage of {\sc Spark} is its ability to process datasets that exceed memory limits.
+\subsection*{\variantSpark{} enables genome wide sampling of variants to improve clustering quality}
+Although we have demonstrated the advantages of \variantSpark{} over traditional methods on small datasets, such as individual chromosomes, the main advantage of {\sc Spark} is its ability to process datasets that exceed memory limits.
 To demonstrate this scalability, we increase the number of variants used in the clustering from the initial 494,328 on chromosome 22 to 38,219,238 including all variants in the genome.
@@
 Pre-processing the VCF files takes approximately 12 minutes, 19 minutes, 27 minutes and 40 minutes for 20\%, 40\%, 60\% and 100\% of the genome, respectively (see Table 2).
-In each case, the memory required does not exceed a modest 2GB per executor, even for 100\% of the genome. 
+In each case, the memory required does not exceed a modest 2GB per executor, even for the complete genome. 
 Clustering the data takes 70 minutes, 139 minutes, 213 minutes and 884 mins, respectively.
 For clustering, the memory requirements increase approximately linearly to 24GB per executor (see Figure 1). 
-%TODO: doubleckeck
 The increase in memory is due to more distance measurements between variants and \kMeans{} centroids being created, which MLlib stores as dense vectors. 
@@
 
-We do not observe an increase in cluster accuracy when providing variants from other chromosomes, indicating that current practice of clustering based on a subset of the genome is sufficient to define the existing population boundaries~\cite{Pugach2015}.
+We do not observe an increase in cluster accuracy when providing variants from all chromosomes, indicating that current practice of clustering based on a subset of the genome is sufficient to define the existing population boundaries for populations with large genetic distance~\cite{Pugach2015}.
 
-\subsection*{Personal Genome Project}
+\subsection*{Determining the population structure of the Personal Genome Project}
 To demonstrate the versatility of \variantSpark{} we also process and cluster individuals from the Personal Genomics Project (PGP).
@@
 
-\variantSpark{} performs clustering on VCF files with over 1000 individuals and 40 Million variants in under 15 hours using minimal memory (24GB). 
+\variantSpark{} performs clustering on VCF files with over 1000 individuals and 40 million variants in under 15 hours using minimal memory (24GB). 
 \variantSpark{} supports random genome-wide sampling of variants allowing faster clustering for well characterised cohorts where 20\% of the genome is sufficient. 
@@
 \subsection*{Computational resources}
-We completed the Chromosome 22 comparisons on a virtual machine (VM) hosted on Microsoft Azure. This VM is an A7 Linux instance with 8 cores, 56GB memory and Ubuntu as the OS. 
-For the whole-genome clustering, we used our in-house Hadoop cluster. This uses Hadoop 2.5.0, managed by CDH 5. We use Spark 1.3.1. This 13 node cluster has a total of 416 cores and 1.22TB memory.
+We completed the Chromosome 22 comparisons on a virtual machine (VM) hosted on Microsoft Azure. This VM is an A7 Linux instance with 8 cores, 56GB memory running Ubuntu. 
+For the whole-genome clustering, we used our in-house Hadoop cluster with Hadoop 2.5.0, managed by cloudera's CDH 5.
+We use Spark 1.3.1.
+This 13 node cluster has a total of 416 cores and 1.22TB memory.
 
@@
 We used  phase 1 variants from the 1000 Genomes Project. On the 22 autosomes, this dataset contains variants for 1092 individuals, spread over a total 38,219,238 alleles.
-We also used a subset of this dataset, the variants from chromosome 22, and this contains 494,328 alleles.
+We also used a subset of this dataset, the variants from chromosome 22 containing 494,328 alleles.
 The individuals from these datasets are distributed across four super populations, African (AFR), Mixed American (AMR), East Asian (EAS) and European (EUR).
@@
 RDDs allow us to process the files, line-by-line, in parallel.
-We parse each line as tab-separated values, and `zip' each value with it's heading from the VCF file, as shown in Figure 2.
-The result is an array of key-value pairs (KVPs) for each line, where the column heading is the key and the variant is the value.
-We then `zip' each KVP array with a unique sequential index, where these indexes will serve as identifiers for the alleles.
-Now each KVP array is the value in a parent KVP, where the key is the unique index.
-We use a `flatMap' (step 3 of Figure 2) to convert these indexed arrays of KVPs into a less convoluted data-structure, as well as to filter out any extra metadata.
-This results in KVPs (number of variants x number of alleles) where the key is the individual ID, and the value is a tuple of an allele and its index.
-Effectively, we have a KVP for each allele from the VCF file.
-At this point, each allele is still a string straight from the VCF file (i.e. `0\textbar{}1'), however, we require doubles, for MLlib.
-To convert these strings to doubles, we use the Hamming distance of the allele, where a 1 represents a heterozygous variant, 2 a homozygous variant, and 0, no variant.
-Also, for each individual, we filter out any alleles that have no variant, leaving us with KVPs of variants (homozygous and heterozygous) only.
+We parse each line as tab-separated values, and Spark's `zip' each value with it's heading (Sample name) from the VCF file, resulting in key-value pairs (KVPs) for each line see Figure 2 (Spark step 1).
+We then `zip' each pair with a unique sequential index, which later-on will serve as identifier for each allele (and keep the genome loci order), resulting in a KVP array (step 2).
+We use `flatMap' to transpose the data structure to facilitate the grouping of variants by individuals.
+For the \kMeans{} clustering, we convert the VCF genotypes (i.e.
+`0\textbar{}1') to doubles using the Hamming distance to the reference, where a 1 represents a heterozygous variant, 2 a homozygous variant, and 0, no variant.
+To remove uninformative KVPs,  we filter out any alleles that have no variant. 
 We perform this `map' operation inside the previous `flatMap' (step 3).
 We can now group these KVPs by their key (the individual ID) and save them to sparse vectors (step 4).
-This results in a sparse vector for each individual, the format required by MLlib's machine learning algorithms.
 
@@
 \subsection*{Hadoop Implementation}
-We use our previous Hadoop implementation, which is based on the MapReduce model. This makes use of KVPs, similarly to \variantSpark{}, however, it is explicitly based on the MapReduce model.
-Because of this, and because we need a unique range of identifiers for alleles (in the smallest range possible), we need to run an initial MapReduce task to index the lines.
+Our Hadoop implementation is based on the MapReduce model, which utilises KVPs, similarly to \variantSpark{}.
+As we need a unique range of identifiers for alleles (in the smallest range possible), we need to run an initial MapReduce task to index the lines.
 This is comparable (however, more verbose) to the `zip' operation in \variantSpark{}.
@@
 Now that the KVPs for each individual are physically located on the same hardware, the Reduce stage (step 2) can efficiently create a sparse vector for each individual from these KVPs.
-The sparse vectors (one for each individual) are saved to disk, ready to be used as the input to one of MLlib's algorithms.
 
-%% Need to cite R package??
 \subsection*{R Implementation}
-For our R implementation, we read in a VCF file as a table and convert it to a matrix, dropping columns that don't represent alleles.
+For our R implementation, we read in a VCF file as a table and convert it to a matrix, filtering columns where all variant as conform to the reference.
 As with our \variantSpark{} pre-processing, we convert the strings that represent each allele to a numeric value.
 We then transpose the matrix, resulting in a data-structure where each row represents an individual.
-We convert the matrix to a `big.matrix', for the \kMeans{} algorithm from the `biganalytics' package.
+We convert the matrix to a `big.matrix', for the \kMeans{} algorithm from the `biganalytics' package (\url{https://cran.r-project.org/web/packages/biganalytics/index.html}).
 
 \subsection*{Python Implementation}
-Our Python implementation reads in lines from a VCF file as tab-separated values and stores the lines in a pandas DataFrame.
+Our Python implementation reads in lines from a VCF file as tab-separated values and stores the lines in a pandas DataFrame (\url{http://pandas.pydata.org}).
 The column headings are the individual IDs and the row headings are the allele locations.
-We drop the first columns that aren't alleles, and then convert the remaining allele strings to numeric values.
-We can then transpose the DataFrame.
-Although to cluster this DataFrame (using sci-kit learn), we simply need to convert it to a matrix with {\sc .as\textunderscore{}matrix()}
+We filter out reference-conform data and convert the remaining allele strings to numeric values. 
+We cluster this DataFrame using sci-kit learn (\url{http://scikit-learn.org/stable/}), upon converting it to a matrix with {\sc .as\textunderscore{}matrix()}.
 
 \subsection*{{\sc Admixture} Implementation}
-We use Genome Analysis Toolkit (GATK) to convert the variants from VCF, and a 1000 Genomes Project supplied .ped file, to a binary PLINK (.bed), binary marker information file (.bim) and pedigree stub file (.fam).
+We use Genome Analysis Toolkit ({\sc -T VariantsToBinaryPed})~\cite{McKenna2010} to convert the variants from VCF, and a 1000 Genomes Project supplied .ped file, to a binary PLINK (.bed), binary marker information file (.bim) and pedigree stub file (.fam).
 These three files are used as input to {\sc Admixture}, with $K$ (the number of ancestral populations) set to 4.
 
-%% TODO - NEIL - (and better heading) %%
 \subsection*{Personal Genome Project Data}
-%% The PGP data..... 
-PGP genome data was downloaded from the PGP website (\url{https://my.pgp-hms.org/public_genetic_data}). Genotype data from 23andMe microarray platforms was sorted by genome 
+Genome data was downloaded from the Personal Genome Project Data (PGP) website (\url{https://my.pgp-hms.org/public_genetic_data}). Genotype data from 23andMe microarray platforms was sorted by genome 
 build. Shell scripts were written to convert NCBI build 36 files to BED format, update to build 37 using UCSC liftOver (\url{https://genome.ucsc.edu/cgi-bin/hgLiftOver}) and 
@@
 file using VCFtools \cite{Danecek2011Variant}.
-
-% DENIS yes, it will be in the git repository mentioned above (https://github.com/BauerLab/VariantSpark)
-% can you please push your code into https://stash.csiro.au/users/bau04c/repos/variantspark/browse
 
@@
   \caption{\csentence{Comparison of method and genome-wide scaling experiment.}
-      {\bf Left} Runtime is given in seconds with 32 GB of memory on 8 threads (except for the pre-processing in R and Python where this was not supported). 
+      {\bf Left} Runtime for clustering variants from chromosome 22 is given in seconds with 32 GB of memory on 8 threads (except for the pre-processing in R and Python where this was not supported). 
       {\bf Right} Scaling from 20\% to 100\% of variants in the genome with maximal number of executors and lowest possible memory assignment.}
@@
 
-%Hadoop
-%1. Map
-%The mapper converts each line of the VCF file into a preliminary key-value pair, where the Key is the byte-position of the first character thereby functioning as a proxy for the genomic location (variant ID), and the Value is a string of the line.
-From this, the key-value pairs for the individual genotypes are created by extracting the variant from each sample as the new Value.
-The new key is a tuple consisting of the primary key, which is the individual ID (position of the parsed variants), and the variant ID as the secondary key.
-%The MapReduce partitioner uses custom Collect/sort functions to collect the tuples by their primary key (individual ID) and distributes them between nodes.
-The data is sorted by secondary key (variant ID)
-%2. Reduce
-%This data is input to the Reducers, which store the variants for each individual to a sparse vector with the Key as the individual ID and the value as the variant vector.
-%
-%Spark
-%1. The lines from the VCF file(s) are parsed independently by worker nodes, where each line is zipped with the heading from the VCF file resulting in tuples of individual ID and the variant. 
-%2.
-These arrays are each zipped again with the unique, sequential index (the Variant ID) resulting in tuples (essentially key/value pairs) of Variant IDs (?key?) and the array of tuples from step 1 (?value?).
-%3. The flatmap function restructures the data into Individual ID (?key?) and tuples with Variant ID and the Variant (?value?).
-%4. These tuples are grouped by the Individual ID, collecting all Variant ID and Variant tuples, and non-zero variants are saved as a sparse vector.
 \begin{figure}[h!]
@@
 \includegraphics[type=pdf,ext=.pdf,read=.pdf, scale=0.40]{chr22-k4-gephi-legend}
-  \caption{\csentence{Visualization of \variantSpark{} clusters.} Individuals from the 1000 Genomes project are clustered effectively by super-populations AFR, AMR and EAS; a fourth cluster is predominantly EUR + AMR.}
+  \caption{\csentence{Visualisation of \variantSpark{} predicted clusters.} Individuals from the 1000 Genomes project are clustered effectively by super-populations AFR, AMR and EAS; a fourth cluster is predominantly EUR + AMR.}
       \end{figure}
@@
 \begin{table}[h!]
-\caption{The resources consumption of the six compared methods as well as the accuracy measured as \ARI{} on chromosome 22.}
+\caption{The resource consumption of the six compared methods as well as the accuracy measured as \ARI{} on chromosome 22.}
       \begin{tabular}{|c|ccc|ccc|c|c|}
@@
 \section*{Additional Files}
-  \subsection*{Additional file 1 --- Sample additional file title}
-    Additional file descriptions text (including details of how to
-    view the file, if it is in a non-standard format or the file extension).  This might
-    refer to a multi-page table or a figure.
-
-  \subsection*{Additional file 2 --- Sample additional file title}
-    Additional file descriptions text.
+  \subsection*{Additional file 1 --- PGP population labels}
+    Conversion map self-reported descriptive population labels to super-population labels.
 

### 2015-09-22 doc/publications/journal2014/paper.tex

@@
 We use Genome Analysis Toolkit ({\sc -T VariantsToBinaryPed})~\cite{McKenna2010} to convert the variants from VCF, and a 1000 Genomes Project supplied .ped file, to a binary PLINK (.bed), binary marker information file (.bim) and pedigree stub file (.fam).
-These three files are used as input to {\sc Admixture}, with $K$ (the number of ancestral populations) set to 4.
+These three files are used as input to {\sc Admixture}, with default options and $K$ (the number of ancestral populations) set to 4.
 

### 2015-09-22 doc/publications/journal2014/paper.tex
couple of changes

@@
 This feature was designed to ensure scalability but here substantially increases the runtime as disk IO becomes the limiting factor for this small dataset (chromosome 22).
-Python and R store the output from each iteration in memory thereby eliminating the IO bottleneck but without being able to utilise the disk they are limited to datasets that fit into memory.  
-{\sc Spark} overcomes both limitations by performing in-memory caching for each executor, which eliminates the disk IO bottleneck on smaller datasets, while maintaining scalability to larger datasets where the data actively processed on executors is cashed in memory.
+Python and R store the output from each iteration in memory, thereby eliminating the IO bottleneck, but without being able to utilise the disk they are limited to datasets that fit into memory.
+{\sc Spark} overcomes both limitations by performing in-memory caching for each executor, which eliminates the disk IO bottleneck on smaller datasets, while maintaining scalability to larger datasets by spilling data that exceeds memory availability to disk.
 
@@
 To cluster the individuals in a VCF file, we initially need to pre-process the variants to feature vectors.
-In \variantSpark{} we read in the VCF files as text files to a Resilient Data Set (RDD).
+In \variantSpark{} we read in the VCF files as text files to a Resilient Distributed Dataset (RDD).
 RDDs allow us to process the files, line-by-line, in parallel.
-We parse each line as tab-separated values, and Spark's `zip' each value with it's heading (Sample name) from the VCF file, resulting in key-value pairs (KVPs) for each line see Figure 2 (Spark step 1).
+We parse each line as tab-separated values, and with Spark's `zip' function, `zip' each value with it's heading (Sample name) from the VCF file, resulting in key-value pairs (KVPs) for each line see Figure 2 (Spark step 1).
 We then `zip' each pair with a unique sequential index, which later-on will serve as identifier for each allele (and keep the genome loci order), resulting in a KVP array (step 2).

### 2015-09-22 doc/publications/journal2014/paper.tex
couple of changes

@@
 Figure 3 visualises the clusters generated by \variantSpark{} from the 1000 Genomes Project data. 
-As discussed in the previous section, three of the four clusters (super-populations AFR, EAS and AMR) are relatively-homogeneous; the fourth is more mixed and 
+As discussed in the previous section, three of the four clusters (super-populations AFR, EAS and AMR) are relatively homogeneous; the fourth is more mixed and 
 consists predominantly of individuals labelled EUR and AMR providing a visual representation of the ARI of 0.84.

### 2015-09-22 doc/publications/journal2014/paper.tex
submission

@@
 \usepackage{graphicx}
-\usepackage{multirow}
 
@@
 Although {\sc Spark} allows MapReduce-like programs, it does not require programs to exactly model the MapReduce paradigm, which in turn allows more flexible software design. 
-Recognising the capability, Wiewi{\'o}rka {\it et al.}~\cite{Wiewiorka2014} developed SparkSeq for high-throughput sequence data analysis and the Big Data Genomics (BDG) group recently demonstrated the strength of {\sc Spark} in a genomic clustering application using {\sc ADAM}, a set of formats and APIs as well as processing stage implementations for genomic data~\cite{Massie2013}.
+Recognising the capability, Wiewi{\'o}rka {\it et al.}~\cite{Wiewiorka2014} developed {\sc SparkSeq} for high-throughput sequence data analysis and the Big Data Genomics (BDG) group recently demonstrated the strength of {\sc Spark} in a genomic clustering application using {\sc ADAM}, a set of formats and APIs as well as processing stage implementations for genomic data~\cite{Massie2013}.
 While the speedup of {\sc ADAM} over traditional methods was impressive (50 fold speedup), being limited by constraints within this general genomics framework hampered performance.
@@
 We use \kMeans{} clustering algorithms in the respective implementations, which require the VCF input files to be pre-processed (see methods). 
-The state-of-the-art tool, {\sc Admixture}, also requires a pre-processing step from VCF to PED format, which we performed using GAKT~\cite{McKenna2010}.
+The state-of-the-art tool, {\sc Admixture}, also requires a pre-processing step from VCF to PED format, which we performed using GATK~\cite{McKenna2010}.
 We therefore compare the time required for the pre-processing step, as well as for \kMeans{} clustering.
 
-As shown in Table 1, pre-processing the data is fastest in \variantSpark{}, requiring 2mins and 58secs. %178s 
-This is almost 80\% faster than both the {\sc ADAM} implementation (12min and 48secs) %768s
-and our Hadoop implementation (14min and 22secs). %862s
+As shown in Table 1, pre-processing the data is fastest in \variantSpark{}, requiring 3 minutes. %2mins and 58secs. %178s 
+This is almost 80\% faster than both the {\sc ADAM} implementation (13 minutes) %(12min and 48secs) %768s
+and our Hadoop implementation (14 minutes). %(14min and 22secs). %862s
 Unlike \variantSpark{} and our Hadoop implementation, the {\sc ADAM} framework cannot process the VCF files directly but requires them to be converted into a binary {\sc ADAM} file format. 
-Although this additional pre-processing step is only required once for each input file, it requires an additional 13mins and 13secs, %793s
+Although this additional pre-processing step is only required once for each input file, it uses an additional 13 minutes, %13mins and 13secs, %793s
 rendering our approach almost an order of magnitude faster (see Figure 1).
@@
 This increased pre-processing time is due to the standard Python and R implementations not natively supporting multithreading, while \variantSpark{}, {\sc ADAM} and Hadoop can use the eight available cores.
-The GATK-based VCF to PED conversion is also slower with over 10 minutes.
+The GATK-based VCF to PED conversion for {\sc Admixture} is also slower with over 10 minutes.
 
@@
 Multithreading \kMeans{} clustering is natively supported by both R and Python, we are therefore able to utilise eight cores. 
-Despite this, both are an order of magnitude slower than \variantSpark{} and {\sc ADAM}, with Python requiring 11mins and 29secs and R requiring 7mins and 25secs.
+Despite this, both are an order of magnitude slower than \variantSpark{} and {\sc ADAM}, with Python requiring 11 minutes  and R requiring 8 minutes. %11and 29secs 7mins and 25secs
 {\sc Admixture} is also slower with a runtime of just over 8 minutes. 
-Our Hadoop implementation performs the worst with 14mins and 23secs, showcasing the limitations of the Hadoop engine for iterative algorithms such as \kMeans{}. % 863
+Our Hadoop implementation performs the worst with 14 minutes, showcasing the limitations of the Hadoop engine for iterative algorithms such as \kMeans{}. % 863 14mins and 23secs
 Hadoop writes the entire output of each \kMeans{} iteration to disk.
@@
 For this comparison, we use the \ARI{} (ARI) metric, which returns a value between -1 (independent labelings) and 1 (perfect match)~\cite{Hubert1985}.
+
 Using chromosome 22, each of the algorithms resulted in an ARI of 0.84, confirming that the speed-up in {\sc Spark} is not at the cost of quality.
@@
 The dataset is also very one-sided with 433 out of the 478 (over 90\%) of the individuals being classified as white, even though geographical location of their grandparents are very diverse.
-% TODO provide mapping
-We therefore only include individuals where all grandparents were reported to be from the same country leaving 35 individuals from 22 different countries, which we group into their approximate super-populations (see Supplemental Table X).
+We therefore only include individuals where all grandparents were reported to be from the same country leaving 35 individuals from 22 different countries, which we group into their approximate super-populations (see Supplemental Table 1).
 Clustering these individuals results in an ARI of 0.45.
@@
 \variantSpark{} utilises {\sc Spark}, which allows for in-memory caching and hence performs 85\% faster than Hadoop (29 min).
-This novel parallelisation approach is also faster than traditional multithreading with 94\% speedup over R (75 min), and 95\% over Python (103 min).
+\variantSpark{} scales to data sizes that are not feasible to process using R or Python due their requirement to load the whole dataset into memory. 
+But even on the small benchmarking dataset, \variantSpark{}'s novel parallelisation approach is faster than traditional multithreading with 94\% speedup over R (75 min), and 95\% over Python (103 min).
 \variantSpark{} is also superior in performance and accuracy over the current state-of-the-art tool for individual ancestries determination, {\sc Admixture}.
@@
 
-\label{pgppops}
-\begin{table}[h!]
-\caption{The resources consumption on different subsets of the entire autosome (chromosomes 1-22). Memory specified is the memory allocated to each executor.}
-      \begin{tabular}{|c|c|c|c|}
-
-Super Population		&	Super Population Size &	Populations	&	Population size \\
-\multirow{4}{*}{AMR} 	&	\multirow{4}{*}{5}	&	Canada	&	2	\\
-					&					&	Mexico	&	1	\\
-					&					&	Puerto Rico               & 1               \\
-					&					&	US Minor Outlying Islands & 1               \\
-\multirow{2}{*}{EAS}		&	\multirow{2}{*}{6}	&	China                     & 5               \\
-					&					&	Taiwan, Province of China & 1	\\
-\multirow{14}{*}{EUR}	&	\multirow{14}{*}{20}	& Belgium				& 1	\\
-					&					& Estonia				& 1	\\
-					&					& Finland				& 1	\\
-					&					& France				& 1	\\
-					&					& Germany			& 1	\\
-					&					& Greece				& 1	\\
-					&					& Ireland				& 3	\\
-					&					& Poland				& 2	\\
-					&					& Russian Federation	& 1	\\
-					&					& Slovenia			& 1	\\
-					&					& Spain				& 2	\\
-					&					& Sweden				& 1	\\
-					&					& Ukraine				& 2	\\
-					&					& United Kingdom		& 2	\\
-\multirow{2}{*}{SAS}		&	\multirow{2}{*}{4}	& Bangladesh			& 1	\\
-					&					& India				& 3              
-      \end{tabular}
-\end{table}
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2015-09-22 variantspark/src/main/scala/au/csiro/obr17q/variantspark/App.scala
added variantspark source

@@ -1,0 +1,15 @@
+package au.csiro.obr17q.variantspark
+
+/**
+ * @author ${user.name}
+ */
+object App {
+  
+  def foo(x : Array[String]) = x.foldLeft("")((a,b) => a + b)
+  
+  def main(args : Array[String]) {
+    println( "Hello World!" )
+    println("concat arguments = " + foo(args))
+  }
+
+}

### 2015-09-22 variantspark/src/main/scala/au/csiro/obr17q/variantspark/CommonFunctions.scala

@@ -1,0 +1,34 @@
+package au.csiro.obr17q.variantspark
+
+import org.apache.spark.mllib.linalg.Vector
+import org.apache.spark.mllib.linalg.{Vector=>MLVector, Vectors}
+import org.apache.log4j.Logger
+import org.apache.log4j.Level
+import java.lang.Math.sqrt
+object CommonFunctions {
+   //val logger = Logger.getLogger("au.csiro.obr17q.variantspark");
+   //logger.setLevel(Warn)
+
+  
+  // Takes a VCF string and returns the distance between the variant and wildtype.
+  // Returns m if variant is missing (i.e. .|.)
+  // Useful to either impute 0 for missign variants, or -1 to filter out missing variants.
+  def variantDist(n: String, m: Int): Double = {
+    val alleles = n.split(":")(0)
+    val distance = if(!alleles.matches("""^[0-9][\||\/][0-9]$""")) m else alleles.split("""[\|]|[/]""").map(_.toDouble).reduce(_+_)
+    return distance
+  }
+  
+  
+  def numNonzeros(n: Array[Double]): Int = {
+    // same as values.count(_ != 0.0) but faster
+    var nnz = 0
+    n.foreach { v =>
+      if (v != 0.0) {
+        nnz += 1
+      }
+    }
+    return nnz
+  }
+    
+}

### 2015-09-22 variantspark/src/main/scala/au/csiro/obr17q/variantspark/CustomSettings.scala

@@ -1,0 +1,41 @@
+package au.csiro.obr17q.variantspark
+
+/**
+ * @author obr17q
+ */
+
+class BaseMap(val superPopId: String, val popArray: Array[String])
+class SuperPopulationMap(val pops: Map[String, String])
+
+object BaseMap {
+  def apply(superPopId: String, popArray: Array[String]) = new BaseMap(superPopId, popArray)
+}
+
+object CustomSettings {
+
+  val PGPpopulationArray =
+            BaseMap("EUR", Array("Sweden", "Finland", "Ireland", "France", "Germany", "Greece", "United Kingdom", "Spain", "Poland", "Belgium")) ::
+            BaseMap("SAS", Array("India","Bangladesh")) ::
+            BaseMap("AMR", Array("United States Minor Outlying Islands",  "Canada")) ::
+            BaseMap("EUE", Array("Bulgeria", "Ukraine", "Estonia", "Slovenia", "Russian Federation")) ::
+            BaseMap("MXL", Array("Puerto Rico", "Mexico")) ::
+            BaseMap("AFR", Array("South Africa")) ::
+            BaseMap("ASN", Array("China", "Taiwan, Province of China")) ::
+            Nil
+
+
+        val TGPpopulationArray =
+            BaseMap("MEX", Array()) ::
+            BaseMap("EUR", Array("GBR", "FIN", "IBS", "CEU", "TSI")) ::
+            BaseMap("EAS", Array("CHS", "CHB", "CDX", "JPT", "KHV" )) ::
+            BaseMap("AMR", Array("CLM", "MXL", "PEL", "PUR" )) ::
+            BaseMap("SAS", Array("BEB", "GIH", "STU", "ITU", "PJL")) ::
+            BaseMap("AFR", Array("ACB", "ASW", "GWD", "ESN", "LWK", "MSL", "YRI")) ::
+            Nil
+        
+        implicit val popmap = new SuperPopulationMap(
+            TGPpopulationArray.map( BaseMap => {
+              BaseMap.popArray.map(PopId => (PopId, BaseMap.superPopId))
+            }).flatMap(popTuple => popTuple.map(p => (p._1, p._2))).toMap
+        )
+}

### 2015-09-22 variantspark/src/main/scala/au/csiro/obr17q/variantspark/PopulationMap.scala

@@ -1,0 +1,45 @@
+package au.csiro.obr17q.variantspark
+
+import au.com.bytecode.opencsv.CSVParser
+import au.csiro.obr17q.variantspark.CustomSettings._
+
+/**
+ * @author obr17q
+ */
+class IndividualMap(val IndividualId: String, val PopulationId: String, val SuperPopulationId: String = "")
+      (val AdditionalOne: String = "", val AdditionalTwo: String = "", val AdditionalThree: String = "", val AdditionalFour: String = "") extends java.io.Serializable {
+  def toTup = (PopulationId, (IndividualId, AdditionalOne, AdditionalTwo, AdditionalThree, AdditionalFour))
+  def toMap = Map(PopulationId -> (IndividualId, AdditionalOne, AdditionalTwo, AdditionalThree, AdditionalFour))
+  def toIndo = (IndividualId, (PopulationId,SuperPopulationId))
+  def toIndoAll = (IndividualId, (PopulationId, SuperPopulationId, AdditionalOne, AdditionalTwo, AdditionalThree, AdditionalFour))
+
+  def filterGroups(include:Array[String], exclude:Array[String]):Boolean = {
+    val to_include = if (include.size > 0 && include(0) != "") include.contains(PopulationId) else true
+    val to_exclude = if (exclude.size > 0 && exclude(0) != "") !exclude.contains(PopulationId) else true
+    if (to_include == false || to_exclude == false) false else true
+  }
+}
+
+class PopulationMap(val PopFiles:Iterator[String], val header:String,
+                    val FileDeliminator:Char, val IndividualIdCol:Int, val PopulationCol:Int)
+                    (implicit val e1: Int = PopulationCol, val e2: Int = PopulationCol, val e3: Int = PopulationCol, val e4: Int = PopulationCol) {
+  val too = PopFiles.toList
+  val Populations = (too
+    .filter( !_.startsWith(header) )
+    .map ( lines => {
+      val parser = new CSVParser(FileDeliminator, '"')
+      val l = parser.parseLine(lines)
+      val individualId = l(IndividualIdCol)
+      val Population = l(PopulationCol)
+      //val SuperPopulation = popmap.pops(Population)
+      //( l(IndividualIdCol), l(PopulationCol) ) // (IndividualID, Population)
+      new IndividualMap(individualId, Population, Population)( l(e1), l(e2), l(e3), l(e4))
[…]

### 2015-09-22 variantspark/src/main/scala/au/csiro/obr17q/variantspark/SparkApp.scala

@@ -1,0 +1,23 @@
+package au.csiro.obr17q.variantspark
+
+import org.apache.spark.SparkContext
+import org.apache.spark.SparkConf
+import org.apache.log4j.Logger
+import org.apache.log4j.Level
+//import org.bdgenomics.adam.rdd.ADAMContext
+
+trait SparkApp {
+  def defaultMasterUrl = "local"
+  Logger.getLogger("org").setLevel(Level.OFF)
+  Logger.getLogger("akka").setLevel(Level.OFF)
+  
+  lazy val conf = new SparkConf()
+  //.set("spark.shuffle.memoryFraction", "0.2")
+  //.set("spark.storage.memoryFraction", "0.4")
+  .set("spark.yarn.executor.memoryOverhead", "2048")
+  .set("spark.driver.maxResultSize","2048")
+  .set("spark.default.parallelism", "256")
+  lazy val masterUrl =  if (System.getenv("MASTER") != null) System.getenv("MASTER") else defaultMasterUrl
+  lazy val sc = { println("MASTER: " + masterUrl) ; new SparkContext(masterUrl,this.getClass().getName(),conf)}
+  //lazy val ac = new ADAMContext(sc)
+}

### 2015-09-22 variantspark/src/main/scala/au/csiro/obr17q/variantspark/TcgaClustering.scala

@@ -1,0 +1,44 @@
+package au.csiro.obr17q.variantspark
+
+/**
+ * @author obr17q
+ */
+object TcgaClustering extends SparkApp{
+  
+  def main(args:Array[String]) {
+    val args1 = {
+      if (masterUrl == "local") {
+        /*
+         * Define offline variables here!!!
+         */
+        Array(
+            "data/data.vcf", //Input VCF file
+            "0", "2811814",  //Filter for variants between these two values
+            "3",             //Number of clusters (k)
+            "GBR,ASW,CHB",   //Groups for inclusion
+            "0.1"            //Sample size (0 - 1)
+        )
+      } else {
+      args
+      }
+    }
+    if (args1.length < 1) {
+        println("Usage: CsvClusterer <input-path>")
+    }
+    
+    
+    
+    val VcfFiles = sc.textFile(args1(0))
+    
+    
+    
+    
+    
+    
+    
+  }
+  
+  
+  
+  
+}

### 2015-09-22 variantspark/src/main/scala/au/csiro/obr17q/variantspark/VcfClustering.scala

@@ -1,0 +1,262 @@
+package au.csiro.obr17q.variantspark
+
+import au.csiro.obr17q.variantspark.CommonFunctions._
+import au.csiro.obr17q.variantspark.CustomSettings._
+
+import au.com.bytecode.opencsv.CSVParser
+import org.apache.spark.mllib.linalg.Vector
+import org.apache.spark.mllib.linalg.{Vector=>MLVector, Vectors}
+import org.apache.spark.mllib.clustering.KMeans
+import org.apache.spark.rdd.RDD
+import sys.process._
+import java.util.Arrays;
+import java.io.Writer;
+import java.io.BufferedWriter;
+import java.io.OutputStreamWriter;
+import java.io.PrintWriter;
+import org.json.JSONObject;
+import org.json.JSONArray;
+import scala.io.Source
+
+
+
+
+
+object VcfClustering extends SparkApp {
+  conf.setAppName("VCF cluster")
+  def main(args:Array[String]) {
+
+    val HOME = if (masterUrl == "local") "/Users/obr17q/" else "/OSM/HOME-CDC/obr17q/"
+    val args1 = {
+      if (masterUrl == "local") {
+        /*
+         * Define offline variables here!!!
+         */
+        Array(
+            "data/merged.vcf", //Input VCF file
+            "0", "2811814",  //Filter for variants between these two values
+            "3",             //Number of clusters (k)
+            "",   //Groups for inclusion
+            "",   //Groups for exclusion
+            "0.1"            //Sample size (0 - 1)
+        )
+      } else {
+      args
+      }
+    }
+    if (args1.length < 1) {
+        println("Usage: CsvClusterer <input-path>")
+    }
+
+
+    val StartPos = args1(1).toLong
+    val EndPos = args1(2).toLong
+    val k = args1(3).toInt
+    val IncludeGroups = args1(4).split('|')
+    val ExcludeGroups = args1(5).split('|')
+    val SampleSize = args1(6).toFloat
+    val VariantRe = """^[A-Za-z]{2}[A-Fa-f0-9]{4,6}$""".r // More generic individual ID matcher
+    
+    // The VCF file(s)
+    val VcfFiles = sc.textFile(args1(0), 10)
+        
+    // Returns the heading row from the FASTA file
+    val Headings = VcfFiles
+    .filter( _.startsWith("#CHROM") )
+    .map(
+      line => {
[…]

### 2015-09-22 variantspark/src/main/scala/au/csiro/obr17q/variantspark/VcfForest.scala

@@ -1,0 +1,67 @@
+package au.csiro.obr17q.variantspark
+
+import au.com.bytecode.opencsv.CSVParser
+import org.apache.spark.rdd.RDD
+import org.apache.spark.mllib.linalg.Vector
+import org.apache.spark.mllib.linalg.{Vector=>MLVector, Vectors}
+import org.apache.spark.mllib.tree.RandomForest
+import org.apache.spark.mllib.regression.LabeledPoint
+
+
+/**
+ * @author obr17q
+ */
+object VcfForest extends SparkApp {
+  conf.setAppName("VCF foresting")
+    
+  def main(args:Array[String]) {
+    val numClasses = 2 // 0 - healthy, 1 - obese
+    val categoricalFeaturesInfo = Map[Int, Int]((0, 2))
+    val numTrees = 5
+    val featureSubsetStrategy = "auto"
+    val impurity = "gini"
+    val maxDepth = 6
+    val maxBins = 100
+    val seed = 4363465
+  
+
+    val genoFiles = sc.textFile("data/tcga_csv.csv");
+
+    // Population information
+    val Populations = genoFiles
+    .filter( !_.startsWith("bcr_sample_barcode") )
+    .mapPartitions ( lines => {
+      val parser = new CSVParser(',')
+      lines.map ( line => {
+        val l = parser.parseLine(line)
+        val name = l(0)
+        val obese = l(1).toDouble
+        val genotype = l.slice(2,l.length).map(_.toDouble)
+        val genotypeVector = Vectors.dense(genotype)
+        val n = l.length
+        LabeledPoint(obese, genotypeVector)
+        //( name, obese, v ) // IndividualID, ObesityStatus, Relationship
+      } )
+    } )
+
+
+    val splits = Populations.randomSplit(Array(0.6, 0.4))
+    val (trainingData, testData) = (splits(0), splits(1))     
+    //trainingData.foreach(x => println(x.label))
+
+    val model = RandomForest.trainClassifier(trainingData, numClasses, categoricalFeaturesInfo,
+        numTrees, featureSubsetStrategy, impurity, maxDepth, maxBins)
+
+          
+    // Evaluate model on test instances and compute test error
+    val labelAndPreds = testData.map { point =>
+      val prediction = model.predict(point.features)
+      (point.label, prediction)
+    }
+    val testErr = labelAndPreds.filter(r => r._1 != r._2).count.toDouble / testData.count()
[…]

### 2015-09-22 variantspark/src/main/scala/au/csiro/obr17q/variantspark/VcfSelection.scala

@@ -1,0 +1,79 @@
+package au.csiro.obr17q.variantspark
+
+import au.com.bytecode.opencsv.CSVParser
+import org.apache.spark.rdd.RDD
+import org.apache.spark.mllib.linalg.Vector
+import org.apache.spark.mllib.linalg.{Vector=>MLVector, Vectors}
+import org.apache.spark.mllib.tree.RandomForest
+import org.apache.spark.mllib.regression.LabeledPoint
+import org.apache.spark.mllib.feature.ChiSqSelector
+
+
+/**
+ * @author obr17q
+ */
+object VcfSelection extends SparkApp {
+  conf.setAppName("VCF foresting")
+    
+  def main(args:Array[String]) {
+    val numClasses = 2 // 0 - healthy, 1 - obese
+    val categoricalFeaturesInfo = Map[Int, Int]()
+    val numTrees = 500
+    val featureSubsetStrategy = "auto"
+    val impurity = "entropy"
+    val maxDepth = 20
+    val maxBins = 300
+    val seed = 4363465
+  
+
+    val genoFiles = sc.textFile("data/tcga_csv_1k.csv");
+
+    
+    // Population information
+    val featureLabels = genoFiles
+    .first().split(',').splitAt(2)._2
+
+    // Population information
+    val Populations = genoFiles
+    .filter( !_.startsWith("bcr_sample_barcode") )
+    .mapPartitions ( lines => {
+      val parser = new CSVParser(',')
+      lines.map ( line => {
+        val l = parser.parseLine(line)
+        val name = l(0)
+        val obese = l(1).toDouble
+        val genotype = l.slice(2,l.length).map(_.toDouble)
+        val genotypeVector = Vectors.dense(genotype)
+        val n = l.length
+        LabeledPoint(obese, genotypeVector)
+        //( name, obese, v ) // IndividualID, ObesityStatus, Relationship
+      } )
+    } )
+
+    
+    val selector = new ChiSqSelector(10)
+    val transformer = selector.fit(Populations)
+    val filteredData = Populations.map { lp =>
+      LabeledPoint(lp.label, transformer.transform(lp.features)) 
+    }
+    val splits = filteredData.randomSplit(Array(0.6, 0.4))
+    val (trainingData, testData) = (splits(0), splits(1)) 
+    
+    val model = RandomForest.trainClassifier(trainingData, numClasses, categoricalFeaturesInfo,
+        numTrees, featureSubsetStrategy, impurity, maxDepth, maxBins)
+        
[…]

### 2015-09-22 variantspark/src/test/scala/au/csiro/obr17q/variantspark/PopulationMapTest.scala

@@ -1,0 +1,35 @@
+package au.csiro.obr17q.variantspark;
+
+import org.junit.Assert._;
+import org.junit.Test;
+
+class PopulationMapTest {
+
+	@Test	def test1000GenomeLikeData() {
+    val CSV = Iterator(
+        "sample\tpop\tsuper_pop\tgender",
+        "HG00096\ta1\tEUR\tmale",
+        "HG00097\ta1\tEUR\tfemale",
+        "HG00098\ta2\tEUR\tfemale",
+        "HG00099\ta2\tEUR\tfemale",
+        "HG00100\tb1\tEUR\tfemale",
+        "HG00101\tb2\tEUR\tmale",
+        "HG00102\tb2\tEUR\tfemale",
+        "HG00102\tc1\tEUR\tfemale",
+        "HG00102\tc2\tEUR\tfemale"
+    )
+
+		val popMap = new PopulationMap(CSV, "sample", '\t', 0, 1)
+
+		assertEquals("The output should be a count of individuals.", 9, popMap.returnMap().length)
+    assertEquals("The output should be a count of individuals who belong to a1.", 2, popMap.returnMap(IncludeGroups = Array("a1")).length)
+    assertEquals("The output should be a count of individuals who belong to a1 or a2.", 4, popMap.returnMap(IncludeGroups = Array("a1", "a2")).length)
+    assertEquals("The output should be a count of individuals who belong to a1 or a2.", 4, popMap.returnMap(IncludeGroups = Array("a1", "a2"), ExcludeGroups = Array("")).length)
+    assertEquals("The output should be a count of individuals who belong to a1 or a2.", 4, popMap.returnMap(IncludeGroups = Array("a1", "a2"), ExcludeGroups = Array("b1", "b2")).length)
+    assertEquals("The output should be a count of individuals who belong to a1 or a2.", 4, popMap.returnMap(IncludeGroups = Array("a1", "a2", "c1"), ExcludeGroups = Array("c1")).length)
+    assertEquals("The output should be a count of individuals who belong to a1 or a2 or c1.", 5, popMap.returnMap(IncludeGroups = Array("a1", "a2", "c1")).length)
+    assertEquals("The output should be a count of individuals who do not belong to b1.", 8, popMap.returnMap(ExcludeGroups = Array("b1")).length)
+    assertEquals("The output should be a count of individuals who do not belong to b1 or b2.", 6, popMap.returnMap(ExcludeGroups = Array("b1", "b2")).length)
+	}
+  
+}

### 2015-09-22 variantspark.sh
Added files to root

@@ -1,0 +1,56 @@
+# Default configuration
+StartPosition=0
+EndPosition=2147483647
+IncludeGroups=""
+ExcludeGroups=""
+SampleSize=1
+
+# Parse options
+while getopts ":hc:" opt; do
+  case $opt in
+    c)
+      config_file=$OPTARG >&2
+      if [ ! -f $config_file ]; then
+        echo "Config file not found"
+        exit 1
+      fi
+      source $config_file
+      ;;
+    h)
+      echo "Usage: vcf-clusterer [-pc] [-i input] [-o output] [-k clusters]
+  -p    pre-process files in input directory
+  -c    run k-means clustering job on pre-processed files
+  -i    directory containing VCF files, relative path on HDFS
+  -o    output directory, relative path on HDFS
+  -l    when pre-processing, ignore variants occuring in less than this many people
+  -m    when pre-processing, ignore variants occuring in more than this many people
+  -k    number of clusters for k-means clustering" >&2
+      exit 1
+      ;;
+    \?)
+      echo "Invalid option: -$OPTARG
+Try vcf-clusterer -h for more information." >&2
+      exit 1
+      ;;
+    :)
+      echo "Option -$OPTARG requires an argument.
+Try vcf-clusterer -h for more information." >&2
+      exit 1
+      ;;
+  esac
+done
+
+spark-submit --class au.csiro.obr17q.variantspark.VcfClustering \
+      --master yarn-client \
+      --num-executors $num_executors \
+      --executor-memory $executor_memory \
+      --executor-cores $executor_cores \
+      --driver-memory $driver_memory \
+      workspace/variantspark/target/variantspark-0.0.1-SNAPSHOT.jar \
+      $input_file \
+      $StartPosition \
+      $EndPosition \
+      $clusters \
+      "$IncludeGroups" \
+      "$ExcludeGroups" \
+      $SampleSize

### 2015-10-07 doc/publications/Recomb15/recomb15.tex
recomb first draft

% This is LLNCS.DOC the documentation file of
% the LaTeX2e class from Springer-Verlag
% for Lecture Notes in Computer Science, version 2.4
\documentclass{llncs}
\usepackage{llncsdoc}
%

% Variable definitions
\newcommand{\variantSpark}{{\sc VariantSpark}}
\newcommand{\kMeans}{\textit{k}-means}
\newcommand{\ARI}{adjusted Rand index}

\begin{document}

\newcounter{save}\setcounter{save}{\value{section}}
{\def\addtocontents#1#2{}%
\def\addcontentsline#1#2#3{}%
\def\markboth#1#2{}%
%
\title{Using VariantSpark to uncover the link between obesity and cancer}

\author{Aidan R. O'Brien\inst{1,2}, Jason Ross\inst{1}, Robert Dunne\inst{1}, Firoz Anwar\inst{1}, \and Denis C. Bauer\inst{1}}

\institute{
CSIRO, 11 Julius Av, 2113, Sydney, Australia \\
\and
School of Biomedical Sciences and Pharmacy, Faculty of Health, University of Newcastle, 2308, Newcastle, Australia \\
}

\maketitle
%
\begin{abstract}
This paragraph shall summarize the contents of the paper
in short terms.
\end{abstract}
%
\section{Introduction}
%

Obesity is the second leading risk factor for cancer~\cite{DePergola2013}. 
For colorectal cancer (CRC) there is a 2\% increase in risk per 1 kg/m$^2$, which is similar to the pancreatic, ovarian and prostate cancer risk (about RR 1.06 - 1.10 per 5 kg/m2), while other rarer cancers such as Kidney, gallbladder and endometrial cancers have an even higher risk of up to RR 1.5 per 5 kg/m2. 
The currently observed shift in the global population from a median BMI of 25 to 30 means we must expect a 10\% to 50\% increase in cancer cases in the future, which is akin to the temporal relationship observed with cigarette smoking and lung cancer. 
While the associations between obesity and cancer are robust, almost nothing is known about the mechanism.

BMI itself has a genetic component as demonstrated in a recent meta study, which finds 97 BMI-associated loci using linear regression against BMI adjusting for study-specific covariates~\cite{Locke2015}. 
It is therefore critical to study the impact of obesity on cancer on a large well characterised cohort such as that of the cancer genome atlas (TCGA)~\cite{1KG2012}. 
Hakimi {\it et al.} investigated implication of obesity in the renal cell carcinoma mortality using TCGA data of somatic mutation, copy number, global methylation, and gene/mRNA expression~\cite{DePergola2013}. 
They used a logistic regression models to infer the associations between BMI and disease and a multivariable competing risks regression models to estimate associations between BMI and cancer specific mortality. 
They concluded that BMI is not an independent prognostic factor for cancer specific mortality after controlling for stage and grade. 
But as Locke {\it et al.}~\cite{Locke2015} pointed out in their study, only five of the 97 loci have an independent association signal, the rest seem jointly predict the metabolic phenotype.

In fact, several studies have shown that single gene or single SNP approaches often fail to identify strong feature association or find marginal effect of the feature on the target phenotype~\cite{Bureau2005,Yoo2012,Qi2012}. 
%Such approaches undermine the inter-relationship between multiple trait loci and often specify one locus a time (Pociot, Karlsen et al. 2004). In addition to that, 
The phenotypic relationship between obesity and cancer may hence manifest itself in small additive multi-loci effects, known as epistasis~\cite{Mackay2014}, which is not interrogated with logistic regression models.
Using conventional machine learning analysis to elucidate this additive effects is challenging as there are not enough samples to robustly fit the model over such a high-dimensional feature space~\cite{Chen2012}. 
As a result machine learning methods such as random forest (RF) have become popular as the underlying decision trees are build from a subsets of the features, which make the approach robust against overfitting issues~\cite{Breiman2001}. 
However, training a RF on such an expansive dataset as the TCGA is computationally challenging and traditional massively parallel approaches to data processing are not easily applicable to machine learning tasks, as they iteratively refine models based on information from the full dataset.

We therefore utilise the {\sc Spark} engine in our recently developed framework, \variantSpark~\cite{OBrien}, for applying machine learning methods to standard variant files in Variant Call Format (VCF). 
Unlike {\sc Hadoop}, {\sc Spark} allows a more flexible software design to utilise node-parallelisation and offers in memory cashing, both critically components for efficient "big learning" tasks.

To demonstrate \variantSpark's capability, we train a RF on the genomic profiles and BMI values of cancer patients from the TCGA dataset to detect differences in the mutational signature of tumours arising in a lean and obese background. 
In the first section we determine the cross-validation accuracy of the model and discuss overfitting potential. 
In section two record the runtime and memory consumption of the approach. 
In the last section we compare the genomic profiles of samples that can be fitted by a model very well as oppose to those that do not and discuss the influence of cancer type and tissue of origin.

%The major outcomes for us are:
%1) Can we detect a difference in the mutational signature of tumours arising in a lean and obese background?
%2) Can we characterise the observed differences and highlight new possible modes of therapy?
%3) Can we develop a biomarker(s) that identifies this group of tumours with an obese-specific mutational signature?

%
\section{Results}
%
\subsubsection{Random forest accurately predicts BMI pan-cancer}

\begin{table}
\caption{Cross-validated results, mean accuracy (and standard error).}
\begin{center}
\renewcommand{\arraystretch}{1.4}
\setlength\tabcolsep{3pt}
\begin{tabular}{lcccc}
\hline\noalign{\smallskip}
method  & test folds & training folds & hold out \\
5-fold  & X (Y) & X (Y) & X (Y) \\
\noalign{\smallskip}
\hline
\end{tabular}
\end{center}
\end{table}

\subsubsection{Runtime}

\begin{table}
\caption{Runtime and memory consuption}
\begin{center}
\renewcommand{\arraystretch}{1.4}
\setlength\tabcolsep{3pt}
\begin{tabular}{lcccc}
\hline\noalign{\smallskip}
method  & nodes & runtime & memory \\
\noalign{\smallskip}
\hline
\end{tabular}
\end{center}
\end{table}

\subsubsection{Biological insights}

% 
\begin{figure}
\vspace{2.5cm}
\caption{This is the caption of the figure displaying a white eagle and
a white horse on a snow field}
\end{figure}

\bibliographystyle{plain}
\bibliography{recomb}

\end{document}

### 2015-10-07 doc/publications/Recomb15/recomb15.tex
grammar

@@
 We therefore utilise the {\sc Spark} engine in our recently developed framework, \variantSpark~\cite{OBrien}, for applying machine learning methods to standard variant files in Variant Call Format (VCF).
-Unlike {\sc Hadoop}, {\sc Spark} allows a more flexible software design to utilise node-parallelisation and offers in memory cashing, both critically components for efficient "big learning" tasks.
+Unlike {\sc Hadoop}, {\sc Spark} allows a more flexible software design to utilise node-parallelisation and offers in-memory cashing, both critical components for efficient "big learning" tasks.
 

### 2015-10-22 doc/publications/Recomb15/recomb15.tex
recomb change

@@
 %
-\title{Using VariantSpark to uncover the link between obesity and cancer}
+\title{Using VariantSpark to learn population structure from genomic profiles}
 
@@
 
-Obesity is the second leading risk factor for cancer~\cite{DePergola2013}. 
-For colorectal cancer (CRC) there is a 2\% increase in risk per 1 kg/m$^2$, which is similar to the pancreatic, ovarian and prostate cancer risk (about RR 1.06 - 1.10 per 5 kg/m2), while other rarer cancers such as Kidney, gallbladder and endometrial cancers have an even higher risk of up to RR 1.5 per 5 kg/m2.
-The currently observed shift in the global population from a median BMI of 25 to 30 means we must expect a 10\% to 50\% increase in cancer cases in the future, which is akin to the temporal relationship observed with cigarette smoking and lung cancer.
-While the associations between obesity and cancer are robust, almost nothing is known about the mechanism.
+%TODO add a paper where ethnicity (i.e. allelefrequency) is important for drug dosage. 
+Determining a persons ethnicity has become important for medical applications (EXAMPLES).
+This highlights the issue for clinical application where ancestry influences treatment (e.g.
+HLA allele genotyping from SNP information~\cite{Zheng2014}) and accurate population association may not be known for patients with diverse migrational background.
 
-BMI itself has a genetic component as demonstrated in a recent meta study, which finds 97 BMI-associated loci using linear regression against BMI adjusting for study-specific covariates~\cite{Locke2015}.
-It is therefore critical to study the impact of obesity on cancer on a large well characterised cohort such as that of the cancer genome atlas (TCGA)~\cite{1KG2012}. 
-Hakimi {\it et al.} investigated implication of obesity in the renal cell carcinoma mortality using TCGA data of somatic mutation, copy number, global methylation, and gene/mRNA expression~\cite{DePergola2013}.
-They used a logistic regression models to infer the associations between BMI and disease and a multivariable competing risks regression models to estimate associations between BMI and cancer specific mortality.
-They concluded that BMI is not an independent prognostic factor for cancer specific mortality after controlling for stage and grade. 
-But as Locke {\it et al.}~\cite{Locke2015} pointed out in their study, only five of the 97 loci have an independent association signal, the rest seem jointly predict the metabolic phenotype.
+In previous work we have developed \variantSpark~\cite{OBrien}, a \kMeans\ approach using {\sc Spark} machine library, {\sc MLlib}, to cluster samples from the 1000 genome project~\cite{1KG2012}.
+%TODO add value of the kmeans method on the full dataset (the equivalent of the ARI=0.84)
+We observed a perfect separation for the super populations African (AFR), East Asian (EAS) and European (EUR), however when adding the fourth super-population group, American (AMR), we observe that only a subset of the AMR individual form an independent cluster while the majority falls into the EUR cluster, resulting in an accuracy of XX, with 0 being random and 1 a perfect clustering.
+As Spark enabled our approach to utilises whole genome information (40 Million variants), we argue that the analysis already includes a sufficient number of markers to separate populations with low genetic divergence~\cite{Patterson2006}.
 
+As \kMeans, does not take higher order interactions between features into account it is hence possible that the chosen methodology is not capable to accurately capture the underlying population structure.
 In fact, several studies have shown that single gene or single SNP approaches often fail to identify strong feature association or find marginal effect of the feature on the target phenotype~\cite{Bureau2005,Yoo2012,Qi2012}.
-%Such approaches undermine the inter-relationship between multiple trait loci and often specify one locus a time (Pociot, Karlsen et al. 2004). In addition to that, 
-The phenotypic relationship between obesity and cancer may hence manifest itself in small additive multi-loci effects, known as epistasis~\cite{Mackay2014}, which is not interrogated with logistic regression models.
+As recently shown by Locke {\it et al.}~\cite{Locke2015} in a meta-study around genetic susceptibility of obesity only five of the 97 genomic loci have an independent association signal, the rest seem to jointly predict the metabolic phenotype.
+Ethnicity may hence manifest itself in small additive multi-loci effects, known as epistasis~\cite{Mackay2014}, and hence requires to investigate the effect of one variant against all other variants in the feature vector.
+
 Using conventional machine learning analysis to elucidate this additive effects is challenging as there are not enough samples to robustly fit the model over such a high-dimensional feature space~\cite{Chen2012}.
 As a result machine learning methods such as random forest (RF) have become popular as the underlying decision trees are build from a subsets of the features, which make the approach robust against overfitting issues~\cite{Breiman2001}.
-However, training a RF on such an expansive dataset as the TCGA is computationally challenging and traditional massively parallel approaches to data processing are not easily applicable to machine learning tasks, as they iteratively refine models based on information from the full dataset.
+However, training a RF on such an expansive dataset as the 1000 genomes is computationally challenging and traditional massively parallel approaches to data processing are not easily applicable to machine learning tasks, as they iteratively refine models based on information from the full dataset.
 
-We therefore utilise the {\sc Spark} engine in our recently developed framework, \variantSpark~\cite{OBrien}, for applying machine learning methods to standard variant files in Variant Call Format (VCF).
-Unlike {\sc Hadoop}, {\sc Spark} allows a more flexible software design to utilise node-parallelisation and offers in-memory cashing, both critical components for efficient "big learning" tasks.
+We therefore use {\sc Spark} through or recently developed \variantSpark~\cite{OBrien} framework, which allows us to apply RF in the MLlib library to the variant file in the standard Variant Call Format (VCF) format.
+Please note that unlike {\sc Hadoop}, {\sc Spark} allows a more flexible software design to utilise node-parallelisation and offers in-memory cashing, both critical components for efficient "big learning" tasks.
 
-To demonstrate \variantSpark's capability, we train a RF on the genomic profiles and BMI values of cancer patients from the TCGA dataset to detect differences in the mutational signature of tumours arising in a lean and obese background.
-In the first section we determine the cross-validation accuracy of the model and discuss overfitting potential. 
-In section two record the runtime and memory consumption of the approach. 
-In the last section we compare the genomic profiles of samples that can be fitted by a model very well as oppose to those that do not and discuss the influence of cancer type and tissue of origin.
+To demonstrate \variantSpark's capability, we train a RF on the genomic profiles of the 1000 genome data using the annotated super-population labels. 
+In the first section we determine the cross-validation accuracy of RF on three populations and compare against the unsupervised \kMeans\ clustering and the reported accuracy of a deep-learning approach using {\sc ADAM}~\cite{Massie2013}.
+In section two we record the performance of RF and \kMeans\ on the full set of populations. 
+In the last section we discuss the memory consumptions and runtime of RF compared to \kMeans.
 
-%The major outcomes for us are:
-%1) Can we detect a difference in the mutational signature of tumours arising in a lean and obese background?
-%2) Can we characterise the observed differences and highlight new possible modes of therapy?
-%3) Can we develop a biomarker(s) that identifies this group of tumours with an obese-specific mutational signature?
-
-%
+%TODO check accuracy of titles
 \section{Results}
 %
-\subsubsection{Random forest accurately predicts BMI pan-cancer}
+\subsubsection{Random Forests and \kMeans\ capture the population better than deep learning in a three population test}
+We perform 10-fold cross-validation and report the mean accuracy over all training and testing folds along with the standard error.
+We compare against the reported accuracy of the deep learning approach using the ADAM framework as reported on their webpage~\cite{Ferguson}. 
+The analysis was performed on only three of the 14 populations, we hence restrict our comparison also to ASW (Americans of African Ancestry in SW USA), CHB (Han Chinese in Bejing, China), and GBR (British in England and Scotland).
+
+%TODO[say that this may be a too easy dataset to benchmark the performance differences between kmeans and RF (hence the full dataset needed)]
 
 \begin{table}
-\caption{Cross-validated results, mean accuracy (and standard error).}
+\caption{Cross-validated results, mean accuracy (and standard error) on 3 populations.}
 \begin{center}
@@
 method  & test folds & training folds & hold out \\
-5-fold  & X (Y) & X (Y) & X (Y) \\
+RF  & X (Y) & X (Y) & X (Y) \\
+\kMeans & X (Y) & X (Y) & X (Y) \\
+deep learning & X (Y) & X (Y) & X (Y) \\
 \noalign{\smallskip}
@@
 
-\subsubsection{Runtime}
+\subsubsection{Random Forests is more accurate than \kMeans\ in capturing the full population structure}
+In this section we train and test our two methods on the full dataset of 14 populations and assess the accuracy with respect to the accurate classification into the 4 super-populations.
+
+%TODO [Discuss supervised vs unsupervised]
+%TODO can you drill down to the population clusters and see if and where the best performing method still makes mistakes - this will help in the conclusion to argue that actually the labels are wrong rather than the method not being able to capture the structure (e.g.
+say if AMR are mixed in with GBR TSI)
 
 \begin{table}
-\caption{Runtime and memory consuption}
+\caption{Cross-validated results, mean accuracy (and standard error) on the 4 super-populations.}
+\begin{center}
+\renewcommand{\arraystretch}{1.4}
+\setlength\tabcolsep{3pt}
+\begin{tabular}{lcccc}
+\hline\noalign{\smallskip}
+method  & test folds & training folds & hold out \\
+RF  & X (Y) & X (Y) & X (Y) \\
+\kMeans & X (Y) & X (Y) & X (Y) \\
+\noalign{\smallskip}
+\hline
+\end{tabular}
+\end{center}
+\end{table}
+
+\subsubsection{Kmeans clustering is faster and consumes less memory than Random Forest}
+
+%TODO [write up the observations and argue why one is less resource hungry than the other one]
+
+\begin{table}
+\caption{Runtime and memory consumption on the 4 super-populations}
 \begin{center}
@@
 method  & nodes & runtime & memory \\
+RF  & X (Y) & X (Y) & X (Y) \\
+\kMeans & X (Y) & X (Y) & X (Y) \\
 \noalign{\smallskip}
@@
 
-\subsubsection{Biological insights}
-
-% 
-\begin{figure}
-\vspace{2.5cm}
-\caption{This is the caption of the figure displaying a white eagle and
-a white horse on a snow field}
-\end{figure}
+\section{Conclusion}
+In this paper we investigated the benefit of using a supervised method, Random Forest, compared to an unsupervised method,\kMeans, to determine the ethnicity from whole genome variant data. 
+%TODO fill in the next sections
+We find that X is better than Y because Z.
+We also compare to the supervised deep learning approach as reported by Big Data Genomics using the ADAM framework and find X.
+While we can improve the performance by X\% using the RF, even a supervised method is unable to accurately predict the assigned population labels. 
+We therefore ague that the remaining mixture in X is reflective of the migrational background and hence the underlying genomic profile of Y is accurately similar to X.
+%TODO add
+This may be an important observation for EXAMPLES ABOVE.
 
 \bibliographystyle{plain}
+%TODO you need to find a way of shortening the Locket paper author list
 \bibliography{recomb}

### 2015-10-22 doc/publications/journal2014/images/plotting.R

@@ -3,24 +3,29 @@ library("reshape2")
 library(plyr)
 
 
-table1 <- read.table("~/Documents/datahome/GeMaIn/doc/publications/journal2014/images/Table1.txt", header=TRUE, quote="\"")
+table1 <- read.table("~/Documents/datahome/VariantSpark/doc/publications/journal2014/images/Table1.txt", header=TRUE, quote="\"")
+table1$name<-factor(table1$name,levels=c("Python","R","Hadoop","Adam","ADMIXTURE","VariantSpark"))
 
-pdf("~/Documents/datahome/GeMaIn/doc/publications/journal2014/images/Resources.pdf", width=5, height=4)
+pdf("~/Documents/datahome/VariantSpark/doc/publications/journal2014/images/Resources.pdf", width=5, height=4)
 ggplot(table1, aes(x=factor(name), y=time, fill=task)) + geom_bar(stat="identity") +
   ylab("time in seconds") + xlab("method") +
   theme(axis.text.x = element_text(angle = 45, hjust = 1))
 dev.off()
 
-table2 <- read.table("~/Documents/datahome/GeMaIn/doc/publications/journal2014/images/Table2.txt", header=TRUE, quote="\"")
+table2 <- read.table("~/Documents/datahome/VariantSpark/doc/publications/journal2014/images/Table2.txt", header=TRUE, quote="\"")
 %table2$totalMem=table2$memory*table2$executors
 %result2 <- cbind(table2[1],melt(table2[,c(2,5,6)]))
 result2 <- cbind(table2[1],melt(table2[,c(2,3,4,5)]))
-levels(result2$task)=c("pre-processing","clustering")
+result2$task<-factor(result2$task,levels=c("pre-processing","clustering"))
 
-pdf("~/Documents/datahome/GeMaIn/doc/publications/journal2014/images/Scaling.pdf", width=5, height=4)
+pdf("~/Documents/datahome/VariantSpark/doc/publications/journal2014/images/Scaling.pdf", width=5, height=4)
 ggplot(result2, aes(x=variants, y=value)) + geom_line(aes(colour = variable)) +
   facet_grid(variable ~ task, scales = "free") +
   ylab("value") + xlab("number of variants (%)") +
   theme(axis.text.x = element_text(angle = 45, hjust = 1))
 dev.off()
 
+
+data=data.frame(samples=c(1,270, 1097,11000,70000), dates=c(2001,2005,2012,2015,2017))
+ggplot(data, aes(x=dates, y=samples)) + geom_line()
+  

### 2015-10-24 doc/publications/Recomb15/recomb15.tex
recomb change
updates

@@
 
-In previous work we have developed \variantSpark~\cite{OBrien}, a \kMeans\ approach using {\sc Spark} machine library, {\sc MLlib}, to cluster samples from the 1000 genome project~\cite{1KG2012}.
+In previous work we have developed \variantSpark~\cite{OBrien}, a \kMeans{} approach using {\sc Spark} machine library, {\sc MLlib}, to cluster samples from the 1000 Genomes Project~\cite{1KG2012}.
 %TODO add value of the kmeans method on the full dataset (the equivalent of the ARI=0.84)
-We observed a perfect separation for the super populations African (AFR), East Asian (EAS) and European (EUR), however when adding the fourth super-population group, American (AMR), we observe that only a subset of the AMR individual form an independent cluster while the majority falls into the EUR cluster, resulting in an accuracy of XX, with 0 being random and 1 a perfect clustering.
+We observed a perfect separation for the super populations African (AFR), East Asian (EAS) and European (EUR), however when adding the fourth super-population group, American (AMR), we observe that only a subset of the AMR individual form an independent cluster while the majority falls into the EUR cluster, resulting in an accuracy of 0.84, with -1 being random and 1 a perfect clustering.
 As Spark enabled our approach to utilises whole genome information (40 Million variants), we argue that the analysis already includes a sufficient number of markers to separate populations with low genetic divergence~\cite{Patterson2006}.
 
-As \kMeans, does not take higher order interactions between features into account it is hence possible that the chosen methodology is not capable to accurately capture the underlying population structure.
+As \kMeans{}, does not take higher order interactions between features into account it is hence possible that the chosen methodology is not capable to accurately capture the underlying population structure.
 In fact, several studies have shown that single gene or single SNP approaches often fail to identify strong feature association or find marginal effect of the feature on the target phenotype~\cite{Bureau2005,Yoo2012,Qi2012}.
@@
 Using conventional machine learning analysis to elucidate this additive effects is challenging as there are not enough samples to robustly fit the model over such a high-dimensional feature space~\cite{Chen2012}.
-As a result machine learning methods such as random forest (RF) have become popular as the underlying decision trees are build from a subsets of the features, which make the approach robust against overfitting issues~\cite{Breiman2001}.
-However, training a RF on such an expansive dataset as the 1000 genomes is computationally challenging and traditional massively parallel approaches to data processing are not easily applicable to machine learning tasks, as they iteratively refine models based on information from the full dataset.
+As a result machine learning methods such as random forests (RF) have become popular as the underlying decision trees are build from a subsets of the features, which make the approach robust against overfitting issues~\cite{Breiman2001}.
+However, training an RF on such an expansive dataset as the 1000 genomes is computationally challenging and traditional massively parallel approaches to data processing are not easily applicable to machine learning tasks, as they iteratively refine models based on information from the full dataset.
+
+%TODO add paragraph about deep learning
 
 We therefore use {\sc Spark} through or recently developed \variantSpark~\cite{OBrien} framework, which allows us to apply RF in the MLlib library to the variant file in the standard Variant Call Format (VCF) format.
-Please note that unlike {\sc Hadoop}, {\sc Spark} allows a more flexible software design to utilise node-parallelisation and offers in-memory cashing, both critical components for efficient "big learning" tasks.
+Please note that unlike {\sc Hadoop}, {\sc Spark} allows a more flexible software design to utilise node-parallelisation and offers in-memory cashing, both critical components for efficient ``big learning" tasks.
 
-To demonstrate \variantSpark's capability, we train a RF on the genomic profiles of the 1000 genome data using the annotated super-population labels. 
-In the first section we determine the cross-validation accuracy of RF on three populations and compare against the unsupervised \kMeans\ clustering and the reported accuracy of a deep-learning approach using {\sc ADAM}~\cite{Massie2013}.
-In section two we record the performance of RF and \kMeans\ on the full set of populations. 
-In the last section we discuss the memory consumptions and runtime of RF compared to \kMeans.
+To demonstrate \variantSpark's capability, we train an RF on the genomic profiles of the 1000 Genomes data using the annotated super-population labels. 
+In the first section we determine the cross-validation accuracy of RF on three populations and compare against the unsupervised \kMeans{} clustering and the reported accuracy of a deep-learning approach using {\sc ADAM}~\cite{Massie2013}.
+In section two we record the performance of RF and \kMeans{} on the full set of populations. 
+In the last section we discuss the memory consumptions and runtime of RF compared to \kMeans{}.
 
@@
 %
-\subsubsection{Random Forests and \kMeans\ capture the population better than deep learning in a three population test}
-We perform 10-fold cross-validation and report the mean accuracy over all training and testing folds along with the standard error.
+\subsubsection{Logistic regression and \kMeans{} capture the population better than deep learning in a three population test}
+We perform 5-fold cross-validation and report the mean accuracy over all training and testing folds along with the standard error.
+We also perform 5 runs of \kMeans and report the mean accuracy of the resulting models along with the standard error. 
 We compare against the reported accuracy of the deep learning approach using the ADAM framework as reported on their webpage~\cite{Ferguson}. 
 The analysis was performed on only three of the 14 populations, we hence restrict our comparison also to ASW (Americans of African Ancestry in SW USA), CHB (Han Chinese in Bejing, China), and GBR (British in England and Scotland).
+
+The mean error for our \kMeans clustering models is approximately XXX.
+This is similar to the deep-learning approach where they achieved an accuracy of 99\% (two misclassified individuals).
+Our supervised logistic regression result, however, improves upon this result.
+With logistic regression, we achieve an accuracy of 100\%.
+This is the average across five folds, with a standard error of 0.
+This is on the above three populations, ASW, CHB and GBR, and these three populations belong to the super-populations, AFR, EAS and EUR, respectively.
+As we mention in the introduction, we observed a perfect clustering result when we cluster the entirety of these three super-populations, with an ARI of 1.
+However, when we introduce the fourth super-population into the mix, AMR, the ARI drops to 0.84.
+Given the perfect result when building the logistic regression model on this small subset of data, we expand our dataset to include every population from the 1000 Genomes Project phase1 dataset.
 
@@
 
-\subsubsection{Random Forests is more accurate than \kMeans\ in capturing the full population structure}
+\subsubsection{Logistic regression is more accurate than \kMeans{} in capturing the full population structure}
 In this section we train and test our two methods on the full dataset of 14 populations and assess the accuracy with respect to the accurate classification into the 4 super-populations.
+Using 5-fold cross-validation to assess the logistic regression classification results, we observe an average ARI of \(0.93 \pm{} XX\).
+This is an improvement on our previous result from \kMeans{} clustering the same dataset
+where we observed an ARI of \(0.83\).
+We also record the precision and recall for each of the groups, and observe that for each classification, the precision and recall for EAS were always 1.0.
+Therefore each logistic regression model we built was able to classify every EAS individual correctly while not including any false positives in this group.
+We consistently observed a relatively low precision for the AMR group (\(0.85\pm{} XX \)).
 
@@
 
-\subsubsection{Kmeans clustering is faster and consumes less memory than Random Forest}
+\subsubsection{\kMeans{} clustering is faster and consumes less memory than random forests}
+
+Although memory requirements scale linearly with
 
@@
 \section{Conclusion}
-In this paper we investigated the benefit of using a supervised method, Random Forest, compared to an unsupervised method,\kMeans, to determine the ethnicity from whole genome variant data. 
+In this paper we investigated the benefit of using a supervised method, random forests, compared to an unsupervised method,\kMeans, to determine the ethnicity from whole genome variant data. 
 %TODO fill in the next sections

### 2015-10-24 doc/publications/Recomb15/recomb15.tex
updates

@@
 This is similar to the deep-learning approach where they achieved an accuracy of 99\% (two misclassified individuals).
-Our supervised logistic regression result, however, improves upon this result.
-With logistic regression, we achieve an accuracy of 100\%.
+Our supervised logistic regression approach, however, improves upon this result.
+With this approach, we achieve an accuracy of 100\%.
 This is the average across five folds, with a standard error of 0.
-This is on the above three populations, ASW, CHB and GBR, and these three populations belong to the super-populations, AFR, EAS and EUR, respectively.
-As we mention in the introduction, we observed a perfect clustering result when we cluster the entirety of these three super-populations, with an ARI of 1.
+This is on the above three populations, ASW, CHB and GBR, that belong to the super-populations, AFR, EAS and EUR, respectively.
+As we mention in the introduction, we observed a perfect clustering result when we cluster the entirety of these three super-populations, with an ARI of 1.0.
 However, when we introduce the fourth super-population into the mix, AMR, the ARI drops to 0.84.

### 2015-10-25 doc/publications/Recomb15/recomb15.tex
first part of updating to logReg

@@
 %TODO add value of the kmeans method on the full dataset (the equivalent of the ARI=0.84)
+% Denis: umm correct/total happend to be 0.84 too ? or are you quoting here the ARI ? I want to have the same accuracy measure as reported by bdgenomics for all comparisons in this paper
 We observed a perfect separation for the super populations African (AFR), East Asian (EAS) and European (EUR), however when adding the fourth super-population group, American (AMR), we observe that only a subset of the AMR individual form an independent cluster while the majority falls into the EUR cluster, resulting in an accuracy of 0.84, with -1 being random and 1 a perfect clustering.
-As Spark enabled our approach to utilises whole genome information (40 Million variants), we argue that the analysis already includes a sufficient number of markers to separate populations with low genetic divergence~\cite{Patterson2006}.
+As Spark enabled our approach to utilises whole genome information (40 Million variants), we argue that the analysis already includes a sufficient number of markers to separate populations with low genetic divergence~\cite{Patterson2006}, and hence investigate here whether a supervised method yields better results.
 
-As \kMeans{}, does not take higher order interactions between features into account it is hence possible that the chosen methodology is not capable to accurately capture the underlying population structure.
-In fact, several studies have shown that single gene or single SNP approaches often fail to identify strong feature association or find marginal effect of the feature on the target phenotype~\cite{Bureau2005,Yoo2012,Qi2012}.
-As recently shown by Locke {\it et al.}~\cite{Locke2015} in a meta-study around genetic susceptibility of obesity only five of the 97 genomic loci have an independent association signal, the rest seem to jointly predict the metabolic phenotype.
-Ethnicity may hence manifest itself in small additive multi-loci effects, known as epistasis~\cite{Mackay2014}, and hence requires to investigate the effect of one variant against all other variants in the feature vector.
+%TODO Rob can you think of a better description of what it does
+One of the simplest supervised machine learning methods is logistic regression, which creates a decision boundary by fitting a linear combination of the feature variables. 
+It is therefore able to adjust the weight of the individual genotypes to increase the accuracy of the multi-class prediction.
 
-Using conventional machine learning analysis to elucidate this additive effects is challenging as there are not enough samples to robustly fit the model over such a high-dimensional feature space~\cite{Chen2012}.
-As a result machine learning methods such as random forests (RF) have become popular as the underlying decision trees are build from a subsets of the features, which make the approach robust against overfitting issues~\cite{Breiman2001}.
-However, training an RF on such an expansive dataset as the 1000 genomes is computationally challenging and traditional massively parallel approaches to data processing are not easily applicable to machine learning tasks, as they iteratively refine models based on information from the full dataset.
+%TODO do we need to cite ~\cite{Bureau2005,Yoo2012,Qi2012}. 
+We also investigate more complex supervised approaches, namely Random Forest (RF) and deep learning (DL), as for example Locke {\it et al.}~\cite{Locke2015} showed in a recent meta-study around genetic susceptibility of obesity that only five of the 97 genomic loci have an independent association signal the rest seem to jointly predict the complex metabolic phenotype.
+We therefore hypothesise that ethnicity is a similarly complex phenotype that requires higher order dependencies between features to be taken into account, as done by RF and DL.
+%TODO Rob do you want to add two sentences about RF and DL and the respective benefits (ideally with respect to the multi variable interactions and RF benefit of being hard to overfit).
+%Using conventional machine learning analysis to elucidate this additive effects is challenging as there are not enough samples to robustly fit the model over such a high-dimensional feature space~\cite{Chen2012}.
+%As a result machine learning methods such as random forests (RF) have become popular as the underlying decision trees are build from a subsets of the features, which make the approach robust against overfitting issues~\cite{Breiman2001}.
+However, training a supervised learning method on such an expansive dataset as the 1000 genomes is computationally challenging and traditional massively parallel approaches to data processing are not easily applicable to machine learning tasks, as they iteratively refine models based on information from the full dataset.
 
-%TODO add paragraph about deep learning
-
-We therefore use {\sc Spark} through or recently developed \variantSpark~\cite{OBrien} framework, which allows us to apply RF in the MLlib library to the variant file in the standard Variant Call Format (VCF) format.
+We therefore use {\sc Spark} through our recently developed \variantSpark~\cite{OBrien} framework, which allows us to apply RF in the MLlib library to the variant file in the standard Variant Call Format (VCF) format.
 Please note that unlike {\sc Hadoop}, {\sc Spark} allows a more flexible software design to utilise node-parallelisation and offers in-memory cashing, both critical components for efficient ``big learning" tasks.
@@
 \section{Results}
-%
+
+\subsubsection{Random Forest does not scale to whole genome analysis and is outperformed by logistic regression}
+In this section we compare RF against \kMeans\ and the logistic regression.
+%TODO elaborate on the error
+Despite using {\sc Spark} and {\s MLlib}, RF is not able to scale to the full dataset of 1000 individuals and 40 Million variables, due to XXX. 
+We therefore use only chromosome 22 for the comparison.
+%TODO elaborate on accuracy measures
+As reported in our earlier work, \kMeans\ is not able to fit the dataset accurately (XX\%, ARI=Y). 
+RF achieves a higher accuracy (XX\%, ARI=Y) but is outperformed by the simple logistic regression (XX\%, ARI=Y).
+This indicates that a multivariate approach may not be needed to accurately fit ethnicity.
+
+%TODO report bd's accuracy measure as the main measure (add ARI as the additional information linking to our paper)
 \subsubsection{Logistic regression and \kMeans{} capture the population better than deep learning in a three population test}
+In this section we investigate whether the multivariate approach, deep learning, is able to acheive a better result than RF by again comparing it against \kMeans\ and logistic regression.  
 We perform 5-fold cross-validation and report the mean accuracy over all training and testing folds along with the standard error.

### 2015-10-25 doc/publications/Recomb15/recomb15.tex
last part of updating to logReg

@@
 
-%TODO add a paper where ethnicity (i.e. allelefrequency) is important for drug dosage. 
-Determining a persons ethnicity has become important for medical applications (EXAMPLES).
-This highlights the issue for clinical application where ancestry influences treatment (e.g.
-HLA allele genotyping from SNP information~\cite{Zheng2014}) and accurate population association may not be known for patients with diverse migrational background.
+%TODO other examples?
+Determining a person's ethnicity has become important for medical applications, such as HLA allele genotyping from SNP information~\cite{Zheng2014}.
+Such clinical application are hence severely hampered for patients where accurate population association is unknown due to a diverse migrational background.
+Determining the population association from genomic information may provide vital insights for treatment and diagnostic in such cases.
 
-In previous work we have developed \variantSpark~\cite{OBrien}, a \kMeans{} approach using {\sc Spark} machine library, {\sc MLlib}, to cluster samples from the 1000 Genomes Project~\cite{1KG2012}.
+Aiding this task, we developed \variantSpark~\cite{OBrien}, a \kMeans{} approach using {\sc Spark}'s machine learning library, {\sc MLlib}, which is capable of clustering variants from whole genome sequencing data.
+We demonstrate its use by clustering samples from the 1000 Genomes Project~\cite{1KG2012}.
 %TODO add value of the kmeans method on the full dataset (the equivalent of the ARI=0.84)
-% Denis: umm correct/total happend to be 0.84 too ? or are you quoting here the ARI ? I want to have the same accuracy measure as reported by bdgenomics for all comparisons in this paper
-We observed a perfect separation for the super populations African (AFR), East Asian (EAS) and European (EUR), however when adding the fourth super-population group, American (AMR), we observe that only a subset of the AMR individual form an independent cluster while the majority falls into the EUR cluster, resulting in an accuracy of 0.84, with -1 being random and 1 a perfect clustering.
-As Spark enabled our approach to utilises whole genome information (40 Million variants), we argue that the analysis already includes a sufficient number of markers to separate populations with low genetic divergence~\cite{Patterson2006}, and hence investigate here whether a supervised method yields better results.
+% Denis: correct/total happened to be 0.84 too ?
+or are you quoting here the ARI ?
+I want to have the same accuracy measure as reported by bdgenomics for all comparisons in this paper, also it goes between 0 and 100 right ?
 
+Using \kMeans{} we observed a perfect separation for the super populations African (AFR), East Asian (EAS) and European (EUR), however when adding the fourth super-population group, American (AMR), we observe that only a subset of the AMR individual form an independent cluster while the majority falls into the EUR cluster, resulting in an accuracy of 0.84, with -1 being random and 1 a perfect clustering.
+We argue that the suboptimal clustering is a result of using an unsupervised method rather than an insufficient number of genomic markers~\cite{Patterson2006} and hence hypothesise that a supervised machine learning method may be capable of separating these subgroups of low genetic divergence.
+Indeed, bigdata genomics recently investigated the supervised method, deep learning, for clustering the 1000 genomes data by ethnicity~\cite{Ferguson}.
+
+In this paper we therefore systematically evaluate supervised and unsupervised methods for determining population association and compare our results to the ones obtained using deep learning. 
 %TODO Rob can you think of a better description of what it does
-One of the simplest supervised machine learning methods is logistic regression, which creates a decision boundary by fitting a linear combination of the feature variables. 
+One of the simplest supervised machine learning methods is logistic regression (LR), which creates a decision boundary by fitting a linear combination of the feature variables. 
 It is therefore able to adjust the weight of the individual genotypes to increase the accuracy of the multi-class prediction.
@@
 %TODO do we need to cite ~\cite{Bureau2005,Yoo2012,Qi2012}. 
-We also investigate more complex supervised approaches, namely Random Forest (RF) and deep learning (DL), as for example Locke {\it et al.}~\cite{Locke2015} showed in a recent meta-study around genetic susceptibility of obesity that only five of the 97 genomic loci have an independent association signal the rest seem to jointly predict the complex metabolic phenotype.
-We therefore hypothesise that ethnicity is a similarly complex phenotype that requires higher order dependencies between features to be taken into account, as done by RF and DL.
+We also investigate more complex supervised approaches, namely Random Forest (RF), as for example Locke {\it et al.}~\cite{Locke2015} showed in a recent meta-study around genetic susceptibility of obesity that only five of the 97 genomic loci have an independent association signal the rest seem to jointly predict the complex metabolic phenotype.
+We therefore investigate whether ethnicity is a similarly complex phenotype that requires higher order dependencies between features to be taken into account. %, as done by RF and DL.
 %TODO Rob do you want to add two sentences about RF and DL and the respective benefits (ideally with respect to the multi variable interactions and RF benefit of being hard to overfit).
@@
 
-We therefore use {\sc Spark} through our recently developed \variantSpark~\cite{OBrien} framework, which allows us to apply RF in the MLlib library to the variant file in the standard Variant Call Format (VCF) format.
+We therefore use {\sc Spark} through our recently developed \variantSpark~\cite{OBrien} framework, which allows us to apply RF in the {\sc MLlib} library to the variant file in the standard Variant Call Format (VCF) format.
 Please note that unlike {\sc Hadoop}, {\sc Spark} allows a more flexible software design to utilise node-parallelisation and offers in-memory cashing, both critical components for efficient ``big learning" tasks.
 
-To demonstrate \variantSpark's capability, we train an RF on the genomic profiles of the 1000 Genomes data using the annotated super-population labels. 
-In the first section we determine the cross-validation accuracy of RF on three populations and compare against the unsupervised \kMeans{} clustering and the reported accuracy of a deep-learning approach using {\sc ADAM}~\cite{Massie2013}.
-In section two we record the performance of RF and \kMeans{} on the full set of populations. 
-In the last section we discuss the memory consumptions and runtime of RF compared to \kMeans{}.
+To demonstrate \variantSpark's capability, we determine the cross-validation accuracy of the supervised multivariate RF method and compare it against the simple supervised method LR as well as the unsupervised \kMeans{} in the first section.
+In the second section we compare LR and \kMeans{} against the reported accuracy of the supervised multivariate DL method.
+In section three we record the accuracy of LR and \kMeans{} on the full dataset set using all populations. 
+In the last section we discuss the memory consumptions and runtime of LR compared to \kMeans{}.
 
@@
 \subsubsection{Random Forest does not scale to whole genome analysis and is outperformed by logistic regression}
-In this section we compare RF against \kMeans\ and the logistic regression.
+In this section we compare RF against \kMeans\ and the LR.
 %TODO elaborate on the error
-Despite using {\sc Spark} and {\s MLlib}, RF is not able to scale to the full dataset of 1000 individuals and 40 Million variables, due to XXX. 
-We therefore use only chromosome 22 for the comparison.
+Despite using {\sc Spark} and {\sc MLlib}, RF is not able to scale to the full dataset of 1000 individuals and 40 Million variables, due to XXX. 
+We therefore use only chromosome 22 for the comparison in this section.
+We train a RF and LR on this subset of genomic profiles from the 1000 Genomes data using the annotated labels of all super-populations.
+We perform 5-fold cross-validation and report the mean accuracy over all training and testing folds along with the standard error.
+We also perform 5 runs of \kMeans{} with 4 clusters and report the mean accuracy of the resulting models along with the standard error.
+
 %TODO elaborate on accuracy measures
-As reported in our earlier work, \kMeans\ is not able to fit the dataset accurately (XX\%, ARI=Y). 
+As reported in our earlier work, \kMeans\ is not able to fit the dataset accurately (XX\%, ARI=Y), see table~\ref{RF}.
 RF achieves a higher accuracy (XX\%, ARI=Y) but is outperformed by the simple logistic regression (XX\%, ARI=Y).
@@
 
+\begin{table}
+\caption{Cross-validated results, mean accuracy (and standard error) on variants from chromosome 22 of 4 super-populations.}
+\begin{center}
+\renewcommand{\arraystretch}{1.4}
+\setlength\tabcolsep{3pt}
+\begin{tabular}{lcccc}
+\hline\noalign{\smallskip}
+method  & test folds & training folds & hold out \\
+RF  & X (Y) & X (Y) & X (Y) \\
+\kMeans & X (Y) & X (Y) & X (Y) \\
+LR & X (Y) & X (Y) & X (Y) \\
+\noalign{\smallskip}
+\hline
+\label{RF}
+\end{tabular}
+\end{center}
+\end{table}
+
 %TODO report bd's accuracy measure as the main measure (add ARI as the additional information linking to our paper)
 \subsubsection{Logistic regression and \kMeans{} capture the population better than deep learning in a three population test}
-In this section we investigate whether the multivariate approach, deep learning, is able to acheive a better result than RF by again comparing it against \kMeans\ and logistic regression.  
-We perform 5-fold cross-validation and report the mean accuracy over all training and testing folds along with the standard error.
-We also perform 5 runs of \kMeans and report the mean accuracy of the resulting models along with the standard error. 
-We compare against the reported accuracy of the deep learning approach using the ADAM framework as reported on their webpage~\cite{Ferguson}. 
-The analysis was performed on only three of the 14 populations, we hence restrict our comparison also to ASW (Americans of African Ancestry in SW USA), CHB (Han Chinese in Bejing, China), and GBR (British in England and Scotland).
+In this section we investigate whether the multivariate approach, deep learning, is able to achieve a better result than RF by again comparing it against \kMeans{} and logistic regression.  
+We compare against the reported accuracy of the deep learning approach using the {\sc ADAM} framework~\cite{Massie2013} as reported on their webpage~\cite{Ferguson}. 
+As the reported results were performed on only three of the 14 populations, we hence restrict our comparison also to ASW (Americans of African Ancestry in SW USA), CHB (Han Chinese in Bejing, China), and GBR (British in England and Scotland).
 
-The mean error for our \kMeans clustering models is approximately XXX.
+The mean error for our \kMeans{} clustering models is approximately XXX.
 This is similar to the deep-learning approach where they achieved an accuracy of 99\% (two misclassified individuals).
 Our supervised logistic regression approach, however, improves upon this result.
-With this approach, we achieve an accuracy of 100\%.
-This is the average across five folds, with a standard error of 0.
-This is on the above three populations, ASW, CHB and GBR, that belong to the super-populations, AFR, EAS and EUR, respectively.
-As we mention in the introduction, we observed a perfect clustering result when we cluster the entirety of these three super-populations, with an ARI of 1.0.
+With this approach, we achieve a mean accuracy of 100\% across five folds, with a standard error of 0.
+%This is on the above three populations, ASW, CHB and GBR, that belong to the super-populations, AFR, EAS and EUR, respectively.
+%As we mention in the introduction, we observed a perfect clustering result when we cluster the entirety of these three super-populations, with an ARI of 1.0.
 However, when we introduce the fourth super-population into the mix, AMR, the ARI drops to 0.84.
+Again we observe the complex multivariate analysis to be inferior to the simple logistic regression.
+
 Given the perfect result when building the logistic regression model on this small subset of data, we expand our dataset to include every population from the 1000 Genomes Project phase1 dataset.
@@
 \section{Conclusion}
-In this paper we investigated the benefit of using a supervised method, random forests, compared to an unsupervised method,\kMeans, to determine the ethnicity from whole genome variant data. 
-%TODO fill in the next sections
-We find that X is better than Y because Z.
-We also compare to the supervised deep learning approach as reported by Big Data Genomics using the ADAM framework and find X.
-While we can improve the performance by X\% using the RF, even a supervised method is unable to accurately predict the assigned population labels. 
+In this paper we systematically investigated the performance of supervised and unsupervised methods for determining the ethnicity from whole genome variant data. 
+We found that a supervised method, logistic regression, outperforms the unsupervised approach, \kMeans{} (XX vs YY). 
+We also found that ethnicity can be better predicted by the simple logistic regression method compared to the multivariate approaches such as random forest or deep learning as reported by Big Data Genomics using the ADAM framework (XX vs Y1 and Y2, respectively).
+Furthermore, we observe a limitation within Java, which limits the random forest implementation of {\sc Spark}'s {\sc MLlib} to XX features for the 1069 individuals of the 1000 genome project.
+This property would limit the 100,000 genome project to using only YY features for their 70,000 individuals, when using the same random forest implementation.  
+We conclude that while logistic regression outperforms all other methods it remains unable to accurately classify individuals from [Sub-populations] from the AMR super-population. 
 We therefore ague that the remaining mixture in X is reflective of the migrational background and hence the underlying genomic profile of Y is accurately similar to X.
-%TODO add
-This may be an important observation for EXAMPLES ABOVE.
+This may be an important observation for medical applications that need accurate population association.
 

### 2015-10-25 doc/publications/Recomb15/recomb15.tex
updates

--- 2015-10-24 doc/publications/Recomb15/recomb15.tex
@@
 %
-\subsubsection{Logistic regression and \kMeans{} capture the population better than deep learning in a three population test}
-We perform 5-fold cross-validation and report the mean accuracy over all training and testing folds along with the standard error.
-We also perform 5 runs of \kMeans and report the mean accuracy of the resulting models along with the standard error. 
+\subsubsection{Logistic regression and captures the population better than deep learning in a three population test}
+To test the quality We perform 5-fold cross-validation and report the mean accuracy over all training and testing folds along with the standard error.
+We also perform 5 runs of \kMeans and report the mean Adjusted Rand Index (ARI) of the resulting models along with the standard error. 
 We compare against the reported accuracy of the deep learning approach using the ADAM framework as reported on their webpage~\cite{Ferguson}. 
@@
 
-The mean error for our \kMeans clustering models is approximately XXX.
-This is similar to the deep-learning approach where they achieved an accuracy of 99\% (two misclassified individuals).
-Our supervised logistic regression approach, however, improves upon this result.
-With this approach, we achieve an accuracy of 100\%.
-This is the average across five folds, with a standard error of 0.
-This is on the above three populations, ASW, CHB and GBR, that belong to the super-populations, AFR, EAS and EUR, respectively.
-As we mention in the introduction, we observed a perfect clustering result when we cluster the entirety of these three super-populations, with an ARI of 1.0.
+We use the Adjusted Rand Index to test our \kMeans clustering quality as it allows us to compare the result to the expected clustering (based on populations) and takes into account the structure, rather than just the error-rate.
+We see an average ARI across five runs of \(0.772 \pm 0.086 \).
+Although the average ARI is relatively low, it varies greatly between runs.
+The two best models clustered all but one and all but two individuals correctly, with an ARI of 0.988 and 0.978, respectively.
+However, for the three less-accurate models, a common trait is that ASW and GBR individuals are grouped into one cluster with CHB individuals in a second cluster.
+The third cluster would contain between 1 and 10 seemingly random individuals.
+One possible reason for this varied result is the cluster initialization step.
+Some individuals may have enough variation from the majority of their population, that if the algorithm uses one of their coordinates as an initial centroid, they will cluster separately from the rest of their population.
+
+Although we see two of our five runs meet the accuracy of the deep-learning approach, where they achieved an accuracy of 99\% (or ARI of 0.975, with two misclassified individuals), our logistic regression approach demonstrates a consistent improvement.
+With this approach, we achieve an accuracy of \(99.997 \pm 0.003\) across five folds.
+
+To reiterate, these results are from the three populations, ASW, CHB and GBR, that belong to the super-populations, AFR, EAS and EUR, respectively.
+As we mention in the introduction, we observed a perfect clustering result when we cluster the entirety of these three super-populations (rather than one population from each), with an ARI of 1.0.
 However, when we introduce the fourth super-population into the mix, AMR, the ARI drops to 0.84.
-Given the perfect result when building the logistic regression model on this small subset of data, we expand our dataset to include every population from the 1000 Genomes Project phase1 dataset.
-
-%TODO[say that this may be a too easy dataset to benchmark the performance differences between kmeans and RF (hence the full dataset needed)]
+Given the perfect result from the logistic regression models on the three populations, and the improvement over \kMeans{}, we proceed with logistic regression on the four super-populations.
 
@@
 
-\subsubsection{Logistic regression is more accurate than \kMeans{} in capturing the full population structure}
+\subsubsection{Logistic regression is more accurate than \kMeans{} in capturing the full super-population structure}
 In this section we train and test our two methods on the full dataset of 14 populations and assess the accuracy with respect to the accurate classification into the 4 super-populations.
-Using 5-fold cross-validation to assess the logistic regression classification results, we observe an average ARI of \(0.93 \pm{} XX\).
+Using 10-fold cross-validation to assess the logistic regression classification results, we observe an average ARI of \(0.93 \pm{} XX\).
 This is an improvement on our previous result from \kMeans{} clustering the same dataset
 where we observed an ARI of \(0.83\).
-We also record the precision and recall for each of the groups, and observe that for each classification, the precision and recall for EAS were always 1.0.
-Therefore each logistic regression model we built was able to classify every EAS individual correctly while not including any false positives in this group.
-We consistently observed a relatively low precision for the AMR group (\(0.85\pm{} XX \)).
+We also record the precision and recall for each of the groups, and observe that for each classification, and both values are always 1.0 for the EAS super-population.
+This demonstrates that each logistic regression model was able to classify every EAS individual correctly while not including any false positives in this group.
+For the AMR group, however, we consistantly observed a relatively low precision (\(0.85\pm{} XX \)).
+This was generally due to individuals from the EUR super-population being misclassified as AMR.
+As a result, this led to EUR having a relatively low recall (\(0.9 \pm XX \)).
+Although not present in the result of every fold, another observation is that AFR individuals would sometimes be misclassified as AMR.
+In all ten folds, we didn't observe AFR being classified as EUR, and vice-verse, nor AMR individuals being misclassified as another super-population.
+These results demonstrate the diversity, or relative difference of some of the groups.
 

### 2015-10-25 doc/publications/Recomb15/recomb15.tex

--- 2015-10-25 doc/publications/Recomb15/recomb15.tex
@@
 
-The mean error for our \kMeans{} clustering models is approximately XXX.
+The mean error for our \kMeans clustering models is approximately XXX.
 This is similar to the deep-learning approach where they achieved an accuracy of 99\% (two misclassified individuals).
 Our supervised logistic regression approach, however, improves upon this result.
-With this approach, we achieve a mean accuracy of 100\% across five folds, with a standard error of 0.
-%This is on the above three populations, ASW, CHB and GBR, that belong to the super-populations, AFR, EAS and EUR, respectively.
-%As we mention in the introduction, we observed a perfect clustering result when we cluster the entirety of these three super-populations, with an ARI of 1.0.
+With this approach, we achieve an accuracy of 100\%.
+This is the average across five folds, with a standard error of 0.
+This is on the above three populations, ASW, CHB and GBR, that belong to the super-populations, AFR, EAS and EUR, respectively.
+As we mention in the introduction, we observed a perfect clustering result when we cluster the entirety of these three super-populations, with an ARI of 1.0.
 However, when we introduce the fourth super-population into the mix, AMR, the ARI drops to 0.84.
-Again we observe the complex multivariate analysis to be inferior to the simple logistic regression.
 
 Given the perfect result when building the logistic regression model on this small subset of data, we expand our dataset to include every population from the 1000 Genomes Project phase1 dataset.
+
+Although we see two of our five runs meet the accuracy of the deep-learning approach, where they achieved an accuracy of 99\% (or ARI of 0.975, with two misclassified individuals), our logistic regression approach demonstrates a consistent improvement.
+With this approach, we achieve an accuracy of \(99.997 \pm 0.003\) across five folds.
+
+To reiterate, these results are from the three populations, ASW, CHB and GBR, that belong to the super-populations, AFR, EAS and EUR, respectively.
+As we mention in the introduction, we observed a perfect clustering result when we cluster the entirety of these three super-populations (rather than one population from each), with an ARI of 1.0.
+However, when we introduce the fourth super-population into the mix, AMR, the ARI drops to 0.84.
+Given the perfect result from the logistic regression models on the three populations, and the improvement over \kMeans{}, we proceed with logistic regression on the four super-populations.
 
@@
 In this section we train and test our two methods on the full dataset of 14 populations and assess the accuracy with respect to the accurate classification into the 4 super-populations.
-Using 5-fold cross-validation to assess the logistic regression classification results, we observe an average ARI of \(0.93 \pm{} XX\).
+Using 10-fold cross-validation to assess the logistic regression classification results, we observe an average ARI of \(0.93 \pm{} XX\).
 This is an improvement on our previous result from \kMeans{} clustering the same dataset
 where we observed an ARI of \(0.83\).
-We also record the precision and recall for each of the groups, and observe that for each classification, the precision and recall for EAS were always 1.0.
-Therefore each logistic regression model we built was able to classify every EAS individual correctly while not including any false positives in this group.
-We consistently observed a relatively low precision for the AMR group (\(0.85\pm{} XX \)).
+We also record the precision and recall for each of the groups, and observe that for each classification, and both values are always 1.0 for the EAS super-population.
+This demonstrates that each logistic regression model was able to classify every EAS individual correctly while not including any false positives in this group.
+For the AMR group, however, we consistently observed a relatively low precision (\(0.85\pm{} XX \)).
+This was generally due to individuals from the EUR super-population being misclassified as AMR.
+As a result, this led to EUR having a relatively low recall (\(0.9 \pm XX \)).
+Although not present in the result of every fold, another observation is that AFR individuals would sometimes be misclassified as AMR.
+In all ten folds, we didn't observe AFR being classified as EUR, and vice-verse, nor AMR individuals being misclassified as another super-population.
+These results demonstrate the diversity, or relative difference of some of the groups.
 

### 2015-10-26 doc/publications/Recomb15/recomb15.tex
make two small changes -- why are the lines so long? hundreds of characters

@@
 %TODO other examples?
-Determining a person's ethnicity has become important for medical applications, such as HLA allele genotyping from SNP information~\cite{Zheng2014}.
-Such clinical application are hence severely hampered for patients where accurate population association is unknown due to a diverse migrational background.
-Determining the population association from genomic information may provide vital insights for treatment and diagnostic in such cases.
+
+Determining a person's ethnicity has become important for medical applications, such as HLA allele genotyping from SNP
+information~\cite{Zheng2014}. Such clinical application are severely hampered for patients where accurate population
+association is unknown due to a diverse migrational background. Determining the population association from genomic
+information may provide vital insights for treatment and diagnostic in such cases.
 

### 2015-10-26 doc/publications/Recomb15/recomb15.tex
made all the lines shorter (don't you guys ahve a carriage return?). make all the TODOs into marginpars

@@
 
-Aiding this task, we developed \variantSpark~\cite{OBrien}, a \kMeans{} approach using {\sc Spark}'s machine learning library, {\sc MLlib}, which is capable of clustering variants from whole genome sequencing data.
-We demonstrate its use by clustering samples from the 1000 Genomes Project~\cite{1KG2012}.
-%TODO add value of the kmeans method on the full dataset (the equivalent of the ARI=0.84)
-% Denis: correct/total happened to be 0.84 too ?
-or are you quoting here the ARI ?
-I want to have the same accuracy measure as reported by bdgenomics for all comparisons in this paper, also it goes between 0 and 100 right ?
+Aiding this task, we developed \variantSpark~\cite{OBrien}, a \kMeans{} approach using {\sc Spark}'s machine learning
+library, {\sc MLlib}, which is capable of clustering variants from whole genome sequencing data.  We demonstrate its use
+by clustering samples from the 1000 Genomes Project~\cite{1KG2012}.  \marginpar{TODO add value of the kmeans method on
+  the full dataset (the equivalent of the ARI=0.84) Denis: correct/total happened to be 0.84 too ? or are you quoting
+  here the ARI ? I want to have the same accuracy measure as reported by bdgenomics for all comparisons in this paper,
+  also it goes between 0 and 100 right ?}
 
-Using \kMeans{} we observed a perfect separation for the super populations African (AFR), East Asian (EAS) and European (EUR), however when adding the fourth super-population group, American (AMR), we observe that only a subset of the AMR individual form an independent cluster while the majority falls into the EUR cluster, resulting in an accuracy of 0.84, with -1 being random and 1 a perfect clustering.
-We argue that the suboptimal clustering is a result of using an unsupervised method rather than an insufficient number of genomic markers~\cite{Patterson2006} and hence hypothesise that a supervised machine learning method may be capable of separating these subgroups of low genetic divergence.
-Indeed, bigdata genomics recently investigated the supervised method, deep learning, for clustering the 1000 genomes data by ethnicity~\cite{Ferguson}.
+Using \kMeans{} we observed a perfect separation for the super populations African (AFR), East Asian (EAS) and European
+(EUR), however when adding the fourth super-population group, American (AMR), we observe that only a subset of the AMR
+individual form an independent cluster while the majority falls into the EUR cluster, resulting in an accuracy of 0.84,
+with -1 being random and 1 a perfect clustering.  We argue that the suboptimal clustering is a result of using an
+unsupervised method rather than an insufficient number of genomic markers~\cite{Patterson2006} and hence hypothesise
+that a supervised machine learning method may be capable of separating these subgroups of low genetic divergence.
+Indeed, bigdata genomics recently investigated the supervised method, deep learning, for clustering the 1000 genomes
+data by ethnicity~\cite{Ferguson}.
 
-In this paper we therefore systematically evaluate supervised and unsupervised methods for determining population association and compare our results to the ones obtained using deep learning. 
-%TODO Rob can you think of a better description of what it does
-One of the simplest supervised machine learning methods is logistic regression (LR), which creates a decision boundary by fitting a linear combination of the feature variables. 
-It is therefore able to adjust the weight of the individual genotypes to increase the accuracy of the multi-class prediction.
+In this paper we therefore systematically evaluate supervised and unsupervised methods for determining population
+association and compare our results to the ones obtained using deep learning.
+\marginpar{TODO Rob can you think of a better description of what it does}
+One of the simplest supervised machine learning methods is logistic regression (LR), which creates a decision boundary
+by fitting a linear combination of the feature variables.  It is therefore able to adjust the weight of the individual
+genotypes to increase the accuracy of the multi-class prediction.
 
-%TODO do we need to cite ~\cite{Bureau2005,Yoo2012,Qi2012}. 
-We also investigate more complex supervised approaches, namely Random Forest (RF), as for example Locke {\it et al.}~\cite{Locke2015} showed in a recent meta-study around genetic susceptibility of obesity that only five of the 97 genomic loci have an independent association signal the rest seem to jointly predict the complex metabolic phenotype.
-We therefore investigate whether ethnicity is a similarly complex phenotype that requires higher order dependencies between features to be taken into account. %, as done by RF and DL.
-%TODO Rob do you want to add two sentences about RF and DL and the respective benefits (ideally with respect to the multi variable interactions and RF benefit of being hard to overfit).
-%Using conventional machine learning analysis to elucidate this additive effects is challenging as there are not enough samples to robustly fit the model over such a high-dimensional feature space~\cite{Chen2012}.
-%As a result machine learning methods such as random forests (RF) have become popular as the underlying decision trees are build from a subsets of the features, which make the approach robust against overfitting issues~\cite{Breiman2001}.
-However, training a supervised learning method on such an expansive dataset as the 1000 genomes is computationally challenging and traditional massively parallel approaches to data processing are not easily applicable to machine learning tasks, as they iteratively refine models based on information from the full dataset.
+\marginpar{TODO do we need to cite ~\cite{Bureau2005,Yoo2012,Qi2012}.}  We also investigate more complex supervised
+approaches, namely Random Forest (RF), as for example Locke {\it et al.}~\cite{Locke2015} showed in a recent meta-study
+around genetic susceptibility of obesity that only five of the 97 genomic loci have an independent association signal
+the rest seem to jointly predict the complex metabolic phenotype.  We therefore investigate whether ethnicity is a
+similarly complex phenotype that requires higher order dependencies between features to be taken into
+account. %, as done by RF and DL.
+\marginpar{TODO Rob do you want to add two sentences about RF and DL and the respective benefits (ideally with respect
+  to the multi variable interactions and RF benefit of being hard to overfit).  Using conventional machine learning
+  analysis to elucidate this additive effects is challenging as there are not enough samples to robustly fit the model
+  over such a high-dimensional feature space~\cite{Chen2012}.  As a result machine learning methods such as random
+  forests (RF) have become popular as the underlying decision trees are build from a subsets of the features, which make
+  the approach robust against overfitting issues~\cite{Breiman2001}.  However, training a supervised learning method on
+  such an expansive dataset as the 1000 genomes is computationally challenging and traditional massively parallel
+  approaches to data processing are not easily applicable to machine learning tasks, as they iteratively refine models
+  based on information from the full dataset.}
 
-We therefore use {\sc Spark} through our recently developed \variantSpark~\cite{OBrien} framework, which allows us to apply RF in the {\sc MLlib} library to the variant file in the standard Variant Call Format (VCF) format.
-Please note that unlike {\sc Hadoop}, {\sc Spark} allows a more flexible software design to utilise node-parallelisation and offers in-memory cashing, both critical components for efficient ``big learning" tasks.
+We therefore use {\sc Spark} through our recently developed \variantSpark~\cite{OBrien} framework, which allows us to
+apply RF in the {\sc MLlib} library to the variant file in the standard Variant Call Format (VCF) format.  Please note
+that unlike {\sc Hadoop}, {\sc Spark} allows a more flexible software design to utilise node-parallelisation and offers
+in-memory cashing, both critical components for efficient ``big learning" tasks.
 
-To demonstrate \variantSpark's capability, we determine the cross-validation accuracy of the supervised multivariate RF method and compare it against the simple supervised method LR as well as the unsupervised \kMeans{} in the first section.
-In the second section we compare LR and \kMeans{} against the reported accuracy of the supervised multivariate DL method.
-In section three we record the accuracy of LR and \kMeans{} on the full dataset set using all populations. 
+To demonstrate \variantSpark's capability, we determine the cross-validation accuracy of the supervised multivariate RF
+method and compare it against the simple supervised method LR as well as the unsupervised \kMeans{} in the first
+section.  In the second section we compare LR and \kMeans{} against the reported accuracy of the supervised multivariate
+DL method.  In section three we record the accuracy of LR and \kMeans{} on the full dataset set using all populations.
 In the last section we discuss the memory consumptions and runtime of LR compared to \kMeans{}.
 
-%TODO check accuracy of titles
+\marginpar{TODO check accuracy of titles}
 \section{Results}
@@
 In this section we compare RF against \kMeans\ and the LR.  
-%TODO elaborate on the error
-Despite using {\sc Spark} and {\sc MLlib}, RF is not able to scale to the full dataset of 1000 individuals and 40 Million variables, due to XXX. 
-We therefore use only chromosome 22 for the comparison in this section.
-We train a RF and LR on this subset of genomic profiles from the 1000 Genomes data using the annotated labels of all super-populations.
-We perform 5-fold cross-validation and report the mean accuracy over all training and testing folds along with the standard error.
-We also perform 5 runs of \kMeans{} with 4 clusters and report the mean accuracy of the resulting models along with the standard error.
+\marginpar{TODO elaborate on the error} 
+Despite using {\sc Spark} and {\sc MLlib}, RF is not able to scale to the full dataset of 1000 individuals and 40
+Million variables, due to XXX.  We therefore use only chromosome 22 for the comparison in this section.  We train a RF
+and LR on this subset of genomic profiles from the 1000 Genomes data using the annotated labels of all
+super-populations.  We perform 5-fold cross-validation and report the mean accuracy over all training and testing folds
+along with the standard error.  We also perform 5 runs of \kMeans{} with 4 clusters and report the mean accuracy of the
+resulting models along with the standard error.
 
-%TODO elaborate on accuracy measures
+\marginpar{TODO elaborate on accuracy measures}
 As reported in our earlier work, \kMeans\ is not able to fit the dataset accurately (XX\%, ARI=Y), see table~\ref{RF}.
@@
 
-%TODO report bd's accuracy measure as the main measure (add ARI as the additional information linking to our paper)
+\marginpar{TODO report bd's accuracy measure as the main measure (add ARI as the additional information linking to our
+  paper)}
 \subsubsection{Logistic regression and \kMeans{} capture the population better than deep learning in a three population test}
-In this section we investigate whether the multivariate approach, deep learning, is able to achieve a better result than RF by again comparing it against \kMeans{} and logistic regression.  
-We compare against the reported accuracy of the deep learning approach using the {\sc ADAM} framework~\cite{Massie2013} as reported on their webpage~\cite{Ferguson}. 
-As the reported results were performed on only three of the 14 populations, we hence restrict our comparison also to ASW (Americans of African Ancestry in SW USA), CHB (Han Chinese in Bejing, China), and GBR (British in England and Scotland).
+In this section we investigate whether the multivariate approach, deep learning, is able to achieve a better result than
+RF by again comparing it against \kMeans{} and logistic regression.  We compare against the reported accuracy of the
+deep learning approach using the {\sc ADAM} framework~\cite{Massie2013} as reported on their webpage~\cite{Ferguson}.
+As the reported results were performed on only three of the 14 populations, we hence restrict our comparison also to ASW
+(Americans of African Ancestry in SW USA), CHB (Han Chinese in Bejing, China), and GBR (British in England and
+Scotland).
 
-The mean error for our \kMeans clustering models is approximately XXX.
-This is similar to the deep-learning approach where they achieved an accuracy of 99\% (two misclassified individuals).
-Our supervised logistic regression approach, however, improves upon this result.
-With this approach, we achieve an accuracy of 100\%.
-This is the average across five folds, with a standard error of 0.
-This is on the above three populations, ASW, CHB and GBR, that belong to the super-populations, AFR, EAS and EUR, respectively.
-As we mention in the introduction, we observed a perfect clustering result when we cluster the entirety of these three super-populations, with an ARI of 1.0.
-However, when we introduce the fourth super-population into the mix, AMR, the ARI drops to 0.84.
+The mean error for our \kMeans clustering models is approximately XXX. This is similar to the deep-learning approach
+where they achieved an accuracy of 99\% (two misclassified individuals). Our supervised logistic regression approach,
+however, improves upon this result.  With this approach, we achieve an accuracy of 100\%. This is the average across
+five folds, with a standard error of 0. This is on the above three populations, ASW, CHB and GBR, that belong to the
+super-populations, AFR, EAS and EUR, respectively.  As we mention in the introduction, we observed a perfect clustering
+result when we cluster the entirety of these three super-populations, with an ARI of 1.0. However, when we introduce the
+fourth super-population into the mix, AMR, the ARI drops to 0.84.
 
-Given the perfect result when building the logistic regression model on this small subset of data, we expand our dataset to include every population from the 1000 Genomes Project phase1 dataset.
+Given the perfect result when building the logistic regression model on this small subset of data, we expand our dataset
+to include every population from the 1000 Genomes Project phase1 dataset.
 
-Although we see two of our five runs meet the accuracy of the deep-learning approach, where they achieved an accuracy of 99\% (or ARI of 0.975, with two misclassified individuals), our logistic regression approach demonstrates a consistent improvement.
-With this approach, we achieve an accuracy of \(99.997 \pm 0.003\) across five folds.
+Although we see two of our five runs meet the accuracy of the deep-learning approach, where they achieved an accuracy of
+99\% (or ARI of 0.975, with two misclassified individuals), our logistic regression approach demonstrates a consistent
+improvement.  With this approach, we achieve an accuracy of \(99.997 \pm 0.003\) across five folds.
 
-To reiterate, these results are from the three populations, ASW, CHB and GBR, that belong to the super-populations, AFR, EAS and EUR, respectively.
-As we mention in the introduction, we observed a perfect clustering result when we cluster the entirety of these three super-populations (rather than one population from each), with an ARI of 1.0.
-However, when we introduce the fourth super-population into the mix, AMR, the ARI drops to 0.84.
-Given the perfect result from the logistic regression models on the three populations, and the improvement over \kMeans{}, we proceed with logistic regression on the four super-populations.
+To reiterate, these results are from the three populations, ASW, CHB and GBR, that belong to the super-populations, AFR,
+EAS and EUR, respectively.  As we mention in the introduction, we observed a perfect clustering result when we cluster
+the entirety of these three super-populations (rather than one population from each), with an ARI of 1.0. However, when
+we introduce the fourth super-population into the mix, AMR, the ARI drops to 0.84.  Given the perfect result from the
+logistic regression models on the three populations, and the improvement over \kMeans{}, we proceed with logistic
+regression on the four super-populations.
 
-%TODO[say that this may be a too easy dataset to benchmark the performance differences between kmeans and RF (hence the full dataset needed)]
+\marginpar{TODO[say that this may be a too easy dataset to benchmark the performance differences between kmeans and RF
+  (hence the full dataset needed)]}
 
@@
 \subsubsection{Logistic regression is more accurate than \kMeans{} in capturing the full population structure}
-In this section we train and test our two methods on the full dataset of 14 populations and assess the accuracy with respect to the accurate classification into the 4 super-populations.
-Using 10-fold cross-validation to assess the logistic regression classification results, we observe an average ARI of \(0.93 \pm{} XX\).
-This is an improvement on our previous result from \kMeans{} clustering the same dataset
-where we observed an ARI of \(0.83\).
-We also record the precision and recall for each of the groups, and observe that for each classification, and both values are always 1.0 for the EAS super-population.
-This demonstrates that each logistic regression model was able to classify every EAS individual correctly while not including any false positives in this group.
-For the AMR group, however, we consistently observed a relatively low precision (\(0.85\pm{} XX \)).
-This was generally due to individuals from the EUR super-population being misclassified as AMR.
-As a result, this led to EUR having a relatively low recall (\(0.9 \pm XX \)).
-Although not present in the result of every fold, another observation is that AFR individuals would sometimes be misclassified as AMR.
-In all ten folds, we didn't observe AFR being classified as EUR, and vice-verse, nor AMR individuals being misclassified as another super-population.
-These results demonstrate the diversity, or relative difference of some of the groups.
+In this section we train and test our two methods on the full dataset of 14 populations and assess the accuracy with
+respect to the accurate classification into the 4 super-populations.  Using 10-fold cross-validation to assess the
+logistic regression classification results, we observe an average ARI of \(0.93 \pm{} XX\). This is an improvement on
+our previous result from \kMeans{} clustering the same dataset where we observed an ARI of \(0.83\).  We also record the
+precision and recall for each of the groups, and observe that for each classification, and both values are always 1.0
+for the EAS super-population. This demonstrates that each logistic regression model was able to classify every EAS
+individual correctly while not including any false positives in this group.  For the AMR group, however, we consistently
+observed a relatively low precision (\(0.85\pm{} XX \)). This was generally due to individuals from the EUR
+super-population being misclassified as AMR. As a result, this led to EUR having a relatively low recall (\(0.9 \pm XX
+\)).  Although not present in the result of every fold, another observation is that AFR individuals would sometimes be
+misclassified as AMR. In all ten folds, we didn't observe AFR being classified as EUR, and vice-verse, nor AMR
+individuals being misclassified as another super-population.  These results demonstrate the diversity, or relative
+difference of some of the groups.
 
-%TODO [Discuss supervised vs unsupervised]
-%TODO can you drill down to the population clusters and see if and where the best performing method still makes mistakes - this will help in the conclusion to argue that actually the labels are wrong rather than the method not being able to capture the structure (e.g.
-say if AMR are mixed in with GBR TSI)
+\marginpar{TODO [Discuss supervised vs unsupervised]} \marginpar{TODO can you drill down to the population clusters and
+  see if and where the best performing method still makes mistakes - this will help in the conclusion to argue that
+  actually the labels are wrong rather than the method not being able to capture the structure (e.g. say if AMR are
+  mixed in with GBR TSI)}
 
@@
 
-%TODO [write up the observations and argue why one is less resource hungry than the other one]
+\marginpar{TODO [write up the observations and argue why one is less resource hungry than the other one]}
 
@@
 \section{Conclusion}
-In this paper we systematically investigated the performance of supervised and unsupervised methods for determining the ethnicity from whole genome variant data. 
-We found that a supervised method, logistic regression, outperforms the unsupervised approach, \kMeans{} (XX vs YY). 
-We also found that ethnicity can be better predicted by the simple logistic regression method compared to the multivariate approaches such as random forest or deep learning as reported by Big Data Genomics using the ADAM framework (XX vs Y1 and Y2, respectively).
-Furthermore, we observe a limitation within Java, which limits the random forest implementation of {\sc Spark}'s {\sc MLlib} to XX features for the 1069 individuals of the 1000 genome project.
-This property would limit the 100,000 genome project to using only YY features for their 70,000 individuals, when using the same random forest implementation.  
-We conclude that while logistic regression outperforms all other methods it remains unable to accurately classify individuals from [Sub-populations] from the AMR super-population. 
-We therefore ague that the remaining mixture in X is reflective of the migrational background and hence the underlying genomic profile of Y is accurately similar to X.
-This may be an important observation for medical applications that need accurate population association.
+In this paper we systematically investigated the performance of supervised and unsupervised methods for determining the
+ethnicity from whole genome variant data.  We found that a supervised method, logistic regression, outperforms the
+unsupervised approach, \kMeans{} (XX vs YY).  We also found that ethnicity can be better predicted by the simple
+logistic regression method compared to the multivariate approaches such as random forest or deep learning as reported by
+Big Data Genomics using the ADAM framework (XX vs Y1 and Y2, respectively).  Furthermore, we observe a limitation within
+Java, which limits the random forest implementation of {\sc Spark}'s {\sc MLlib} to XX features for the 1069 individuals
+of the 1000 genome project.  This property would limit the 100,000 genome project to using only YY features for their
+70,000 individuals, when using the same random forest implementation.  We conclude that while logistic regression
+outperforms all other methods it remains unable to accurately classify individuals from [Sub-populations] from the AMR
+super-population.  We therefore ague that the remaining mixture in X is reflective of the migrational background and
+hence the underlying genomic profile of Y is accurately similar to X.  This may be an important observation for medical
+applications that need accurate population association.
 
 \bibliographystyle{plain}
-%TODO you need to find a way of shortening the Locket paper author list
+\marginpar{TODO you need to find a way of shortening the Locket paper author list}
 \bibliography{recomb}

### 2015-10-26 doc/publications/Recomb15/recomb15.tex
logreg results

--- 2015-10-25 doc/publications/Recomb15/recomb15.tex
@@
 In this section we train and test our two methods on the full dataset of 14 populations and assess the accuracy with respect to the accurate classification into the 4 super-populations.
-Using 10-fold cross-validation to assess the logistic regression classification results, we observe an average ARI of \(0.93 \pm{} XX\).
+Using 10-fold cross-validation to assess the logistic regression classification results, we observe an average ARI of \(0.936 \pm{} 0.014\).
 This is an improvement on our previous result from \kMeans{} clustering the same dataset
-where we observed an ARI of \(0.83\).
-We also record the precision and recall for each of the groups, and observe that for each classification, and both values are always 1.0 for the EAS super-population.
-This demonstrates that each logistic regression model was able to classify every EAS individual correctly while not including any false positives in this group.
-For the AMR group, however, we consistently observed a relatively low precision (\(0.85\pm{} XX \)).
-This was generally due to individuals from the EUR super-population being misclassified as AMR.
-As a result, this led to EUR having a relatively low recall (\(0.9 \pm XX \)).
-Although not present in the result of every fold, another observation is that AFR individuals would sometimes be misclassified as AMR.
-In all ten folds, we didn't observe AFR being classified as EUR, and vice-verse, nor AMR individuals being misclassified as another super-population.
+in our previous work, where we observed an ARI of \(0.83\).
+
+To determine where the model is underperforming, we also record the precision and recall for each of the groups, and observe that for each model.
+For each model, the precision and recall are always 1.0 for the EAS super-population.
+This demonstrates that each logistic regression model was able to classify every EAS individual correctly while not misclassifying any individuals as EAS.
+For the AMR group, however, we consistently observed a relatively low precision (\(0.829 \pm{} 0.02 \)).
+This is due to individuals from the AFR and EUR super-populations being misclassified as AMR.
+Consequently, this led to these two super-populations having a relatively low recall of \(0.956 \pm 0.01 \) and \(0.950 \pm 0.02 \), respectively.
+Although these misclassifications as AMR are relatively frequent, we didn't observe any AFR individuals being classified as EUR, or vice-versa, in any of the ten folds.
+Also, only in one fold did we observe one AMR individual misclassified as another super-population (AFR).
 These results demonstrate the diversity, or relative difference of some of the groups.

### 2015-10-26 doc/publications/Recomb15/recomb15.tex
logreg results

--- 2015-10-26 doc/publications/Recomb15/recomb15.tex
@@
 \subsubsection{Logistic regression is more accurate than \kMeans{} in capturing the full population structure}
-In this section we train and test our two methods on the full dataset of 14 populations and assess the accuracy with
-respect to the accurate classification into the 4 super-populations.  Using 10-fold cross-validation to assess the
-logistic regression classification results, we observe an average ARI of \(0.93 \pm{} XX\). This is an improvement on
-our previous result from \kMeans{} clustering the same dataset where we observed an ARI of \(0.83\).  We also record the
-precision and recall for each of the groups, and observe that for each classification, and both values are always 1.0
-for the EAS super-population. This demonstrates that each logistic regression model was able to classify every EAS
-individual correctly while not including any false positives in this group.  For the AMR group, however, we consistently
-observed a relatively low precision (\(0.85\pm{} XX \)). This was generally due to individuals from the EUR
-super-population being misclassified as AMR. As a result, this led to EUR having a relatively low recall (\(0.9 \pm XX
-\)).  Although not present in the result of every fold, another observation is that AFR individuals would sometimes be
-misclassified as AMR. In all ten folds, we didn't observe AFR being classified as EUR, and vice-verse, nor AMR
-individuals being misclassified as another super-population.  These results demonstrate the diversity, or relative
-difference of some of the groups.
+In this section we train and test our two methods on the full dataset of 14 populations and assess the accuracy with respect to the accurate
+classification into the 4 super-populations.
+Using 10-fold cross-validation to assess the logistic regression classification results, we observe an average ARI of \(0.936 \pm{} 0.014\).
+This is an improvement on our previous result from \kMeans{} clustering the same dataset in our previous work, where we observed an ARI of \(0.83\).
+
+To determine where the model is underperforming, we also record the precision and recall for each of the groups, and observe that for each model.
+For each model, the precision and recall are always 1.0 for the EAS super-population.
+This demonstrates that each logistic regression model was able to classify every EAS individual correctly while not misclassifying any individuals as EAS.
+For the AMR group, however, we consistently observed a relatively low precision (\(0.829 \pm{} 0.02 \)).
+This is due to individuals from the AFR and EUR super-populations being misclassified as AMR.
+Consequently, this led to these two super-populations having a relatively low recall of \(0.956 \pm 0.01 \) and \(0.950 \pm 0.02 \), respectively.
+Although these misclassifications as AMR are relatively frequent, we didn't observe any AFR individuals being classified as EUR, or vice-versa, in any of the ten folds.
+Also, only in one fold did we observe one AMR individual misclassified as another super-population (AFR).
+These results demonstrate the diversity, or relative difference of some of the groups.
 

### 2015-10-26 doc/publications/Recomb15/recomb15.tex
logreg results

@@
 
-To determine where the model is underperforming, we also record the precision and recall for each of the groups, and observe that for each model.
-For each model, the precision and recall are always 1.0 for the EAS super-population.
-This demonstrates that each logistic regression model was able to classify every EAS individual correctly while not misclassifying any individuals as EAS.
+To determine where the model is underperforming, we also record the precision and recall for each of the groups.
+We observe that for each model, the precision and recall are always 1.0 for the EAS super-population.
+This demonstrates that each logistic regression model was able to classify every EAS individual correctly, while not misclassifying any other individuals as EAS.
 For the AMR group, however, we consistently observed a relatively low precision (\(0.829 \pm{} 0.02 \)).
 This is due to individuals from the AFR and EUR super-populations being misclassified as AMR.
-Consequently, this led to these two super-populations having a relatively low recall of \(0.956 \pm 0.01 \) and \(0.950 \pm 0.02 \), respectively.
+Consequently, this led to these two super-populations having a lower recall of \(0.956 \pm 0.01 \) and \(0.950 \pm 0.02 \), respectively.
 Although these misclassifications as AMR are relatively frequent, we didn't observe any AFR individuals being classified as EUR, or vice-versa, in any of the ten folds.

### 2015-10-26 doc/publications/Recomb15/recomb15.tex
logreg results
chr22 results

@@
 Despite using {\sc Spark} and {\sc MLlib}, RF is not able to scale to the full dataset of 1000 individuals and 40
-Million variables, due to XXX.  We therefore use only chromosome 22 for the comparison in this section.  We train a RF
+Million variables, due to a limitation of the Java virtual machine (maximum array size) that causes {\sc Mllib}'s
+random forest implementation to fail. This error is due to the number of features present in each sample,
+with a potential solution being to limit the features through feature selection, or simply using a subset of the dataset.
+We therefore use only chromosome 22 (494,328 alleles) for the comparison in this section.  We train a RF
 and LR on this subset of genomic profiles from the 1000 Genomes data using the annotated labels of all
-super-populations.  We perform 5-fold cross-validation and report the mean accuracy over all training and testing folds
-along with the standard error.  We also perform 5 runs of \kMeans{} with 4 clusters and report the mean accuracy of the
+super-populations.  We perform 10-fold cross-validation and report the mean accuracy over all training and testing folds
+along with the standard error.  We also perform 10 runs of \kMeans{} with 4 clusters (\(k=4\)) and report the mean accuracy of the
 resulting models along with the standard error.
@@
 \marginpar{TODO elaborate on accuracy measures}
-As reported in our earlier work, \kMeans\ is not able to fit the dataset accurately (XX\%, ARI=Y), see table~\ref{RF}.
-RF achieves a higher accuracy (XX\%, ARI=Y) but is outperformed by the simple logistic regression (XX\%, ARI=Y).
+As reported in our earlier work, \kMeans\ is not able to fit the dataset accurately.
+Although capable of fitting a relatively accurate model (ARI=0.832), the average ARI over ten models is  \(0.553 \pm 0.07\), see table~\ref{RF}.
+RF achieves a better fitting model (XX\%, ARI=Y) but is outperformed by the simple logistic regression, with an ARI of \(0.923 \pm 0.02\).
 This indicates that a multivariate approach may not be needed to accurately fit ethnicity.
@@
 As the reported results were performed on only three of the 14 populations, we hence restrict our comparison also to ASW
-(Americans of African Ancestry in SW USA), CHB (Han Chinese in Bejing, China), and GBR (British in England and
+(Americans of African Ancestry in SW USA), CHB (Han Chinese in Beijing, China), and GBR (British in England and
 Scotland).
@@
 
-Given the perfect result when building the logistic regression model on this small subset of data, we expand our dataset
-to include every population from the 1000 Genomes Project phase1 dataset.
-
 Although we see two of our five runs meet the accuracy of the deep-learning approach, where they achieved an accuracy of
@@
 the entirety of these three super-populations (rather than one population from each), with an ARI of 1.0. However, when
-we introduce the fourth super-population into the mix, AMR, the ARI drops to 0.84.  Given the perfect result from the
-logistic regression models on the three populations, and the improvement over \kMeans{}, we proceed with logistic
-regression on the four super-populations.
+we introduce the fourth super-population into the mix, AMR, the ARI drops to 0.84.  Given the perfect result when building
+the logistic regression model on this small subset of data, we expand our dataset to include every population from the
+1000 Genomes Project phase1 dataset.
 

### 2015-10-26 doc/publications/Recomb15/recomb15.tex
action on my TODO list

--- 2015-10-26 doc/publications/Recomb15/recomb15.tex
@@
 \usepackage{llncsdoc}
+\usepackage{placeins}
 %
@@
 
-\marginpar{TODO do we need to cite ~\cite{Bureau2005,Yoo2012,Qi2012}.}  We also investigate more complex supervised
+\marginpar{TODO do we need to cite ~\cite{Bureau2005,Yoo2012,Qi2012}.}  
+We also investigate more complex supervised
 approaches, namely Random Forest (RF), as for example Locke {\it et al.}~\cite{Locke2015} showed in a recent meta-study
@@
 account. %, as done by RF and DL.
+
+RF inherit many of the benefits of decision trees. It handles large numbers of variables transparently and
+provides an unbiased estimate of the generalization error (in addition, it handles missing values and unbalanced
+population classes sizes). They provide a measure of variable importance and allow for interactions between
+variables to be investigated. However, in this problem, with many millions of variables, inspecting plots of interactions
+is not feasible.  We note that decision trees (and random forests) model local variable interactions -- there are 
+no main effects or global interaction effects.
+
+For shallow neural networks there are a number of methods for controlling overfitting including regularization, max norm
+constraints, and dropout. Whether the problem of overfitting is exacerbated or not in deep networks is an open problem.
+With a large number of variables (like this problem) some regularization is going to be necesssary. 
+%http://stats.stackexchange.com/questions/21152/obtaining-knowledge-from-a-random-forest
+
 \marginpar{TODO Rob do you want to add two sentences about RF and DL and the respective benefits (ideally with respect
@@
 
+\FloatBarrier
+\clearpage
 \marginpar{TODO report bd's accuracy measure as the main measure (add ARI as the additional information linking to our
@@
 \subsubsection{Logistic regression is more accurate than \kMeans{} in capturing the full population structure}
-In this section we train and test our two methods on the full dataset of 14 populations and assess the accuracy with respect to the accurate
-classification into the 4 super-populations.
-Using 10-fold cross-validation to assess the logistic regression classification results, we observe an average ARI of \(0.936 \pm{} 0.014\).
-This is an improvement on our previous result from \kMeans{} clustering the same dataset in our previous work, where we observed an ARI of \(0.83\).
-
-To determine where the model is underperforming, we also record the precision and recall for each of the groups.
-We observe that for each model, the precision and recall are always 1.0 for the EAS super-population.
-This demonstrates that each logistic regression model was able to classify every EAS individual correctly, while not misclassifying any other individuals as EAS.
-For the AMR group, however, we consistently observed a relatively low precision (\(0.829 \pm{} 0.02 \)).
-This is due to individuals from the AFR and EUR super-populations being misclassified as AMR.
-Consequently, this led to these two super-populations having a lower recall of \(0.956 \pm 0.01 \) and \(0.950 \pm 0.02 \), respectively.
-Although these misclassifications as AMR are relatively frequent, we didn't observe any AFR individuals being classified as EUR, or vice-versa, in any of the ten folds.
-Also, only in one fold did we observe one AMR individual misclassified as another super-population (AFR).
+In this section we train and test our two methods on the full dataset of 14 populations and assess the accuracy with respect to the accurate classification into the 4 super-populations.
+Using 10-fold cross-validation to assess the logistic regression classification results, we observe an average ARI of \(0.93 \pm{} XX\).
+This is an improvement on our previous result from \kMeans{} clustering the same dataset
+where we observed an ARI of \(0.83\).
+We also record the precision and recall for each of the groups, and observe that for each classification, and both values are always 1.0 for the EAS super-population.
+This demonstrates that each logistic regression model was able to classify every EAS individual correctly while not including any false positives in this group.
+For the AMR group, however, we consistently observed a relatively low precision (\(0.85\pm{} XX \)).
+This was generally due to individuals from the EUR super-population being misclassified as AMR.
+As a result, this led to EUR having a relatively low recall (\(0.9 \pm XX \)).
+Although not present in the result of every fold, another observation is that AFR individuals would sometimes be misclassified as AMR.
+In all ten folds, we didn't observe AFR being classified as EUR, and vice-verse, nor AMR individuals being misclassified as another super-population.
 These results demonstrate the diversity, or relative difference of some of the groups.

### 2015-10-26 doc/publications/Recomb15/recomb15.tex
chr22 results

--- 2015-10-26 doc/publications/Recomb15/recomb15.tex
@@
 and LR on this subset of genomic profiles from the 1000 Genomes data using the annotated labels of all
-super-populations.  We perform 10-fold cross-validation and report the mean accuracy over all training and testing folds
-along with the standard error.  We also perform 10 runs of \kMeans{} with 4 clusters (\(k=4\)) and report the mean accuracy of the
+super-populations.  We perform 10-fold cross-validation and report the mean error over all training and testing folds
+along with the standard error of the mean (SEM). We also use the Adjusted Rand Index as a metric, as it allows us to
+compare the classification results with the \kMeans{} clustering result. We also perform 10 runs of \kMeans{} with 4 clusters (\(k=4\)) and report the mean accuracy of the
 resulting models along with the standard error.
@@
 As reported in our earlier work, \kMeans\ is not able to fit the dataset accurately.
-Although capable of fitting a relatively accurate model (ARI=0.832), the average ARI over ten models is  \(0.553 \pm 0.07\), see table~\ref{RF}.
-RF achieves a better fitting model (XX\%, ARI=Y) but is outperformed by the simple logistic regression, with an ARI of \(0.923 \pm 0.02\).
+Although capable of fitting a relatively accurate model \(ARI=0.837\), the average ARI over ten models is  \(0.596 \pm 0.067\), see table~\ref{RF}.
+RF achieves a better fitting model \(ARI = 0.757 \pm 0.011\) but is outperformed by the simple logistic regression, with an ARI of \(0.923 \pm 0.015\).
 This indicates that a multivariate approach may not be needed to accurately fit ethnicity.
@@
 \hline\noalign{\smallskip}
-method  & test folds & training folds & hold out \\
-RF  & X (Y) & X (Y) & X (Y) \\
-\kMeans & X (Y) & X (Y) & X (Y) \\
-LR & X (Y) & X (Y) & X (Y) \\
+method  & ARI & error (\%) & hold out \\
+RF  & 0.757 (0.011) &0.13 (0.007) & X (Y) \\
+\kMeans & 0.596 (0.066) & -- & X (Y) \\
+LR & 0.923 (0.015) & 0.034 (0.006) & X (Y) \\
 \noalign{\smallskip}
@@
 \hline\noalign{\smallskip}
-method  & test folds & training folds & hold out \\
+method  & ARI & training folds & hold out \\
 RF  & X (Y) & X (Y) & X (Y) \\
 \kMeans & X (Y) & X (Y) & X (Y) \\
-deep learning & X (Y) & X (Y) & X (Y) \\
+deep learning & 0.976 & X (Y) & X (Y) \\
 \noalign{\smallskip}

### 2015-10-26 doc/publications/Recomb15/recomb15.tex
precision/recall table

@@
 \hline\noalign{\smallskip}
-method  & ARI & error (\%) & hold out \\
-RF  & 0.757 (0.011) &0.13 (0.007) & X (Y) \\
-\kMeans & 0.596 (0.066) & -- & X (Y) \\
-LR & 0.923 (0.015) & 0.034 (0.006) & X (Y) \\
+method  & ARI & error (\%) \\
+RF  & 0.757 (0.011) &0.13 (0.007) \\
+\kMeans & 0.596 (0.066) & -- \\
+LR & 0.923 (0.015) & 0.034 (0.006) \\
 \noalign{\smallskip}
@@
 
+\begin{table}
+\caption{Cross-validated results, mean accuracy (and standard error) on the 4 super-populations.}
+\begin{center}
+\renewcommand{\arraystretch}{1.4}
+\setlength\tabcolsep{3pt}
+\begin{tabular}{lcccc}
+\hline\noalign{\smallskip}
+method  & test folds & training folds & hold out \\
+RF  & X (Y) & X (Y) & X (Y) \\
+\kMeans & X (Y) & X (Y) & X (Y) \\
+\noalign{\smallskip}
+\hline
+\end{tabular}
+\end{center}
+\end{table}
+
 To determine where the model is underperforming, we also record the precision and recall for each of the groups.
@@
 \begin{table}
-\caption{Cross-validated results, mean accuracy (and standard error) on the 4 super-populations.}
+\caption{Mean precision and recall (and standard error of mean) for each super-population across the ten logistic regression folds.}
 \begin{center}
@@
 \hline\noalign{\smallskip}
-method  & test folds & training folds & hold out \\
-RF  & X (Y) & X (Y) & X (Y) \\
-\kMeans & X (Y) & X (Y) & X (Y) \\
+super-population  & precision & recall \\
+AFR  & 0.997 (0.003) & 0.956 (0.013) \\
+AMR & 0.829 (0.020) & 0.994 (0.005) \\
+EAS  & 1 (0) & 1 (0) \\
+EUR  & 1 (0) & 0.950 (0.015) \\
 \noalign{\smallskip}

### 2015-10-26 doc/publications/Recomb15/recomb15.tex
Remove duplicate para and add stuff

@@
 five folds, with a standard error of 0. This is on the above three populations, ASW, CHB and GBR, that belong to the
-super-populations, AFR, EAS and EUR, respectively.  As we mention in the introduction, we observed a perfect clustering
-result when we cluster the entirety of these three super-populations, with an ARI of 1.0. However, when we introduce the
-fourth super-population into the mix, AMR, the ARI drops to 0.84.
+super-populations, AFR, EAS and EUR, respectively.
 
@@
 
-Although memory requirements scale linearly with
+Time and resources vary greatly between the three algorithms, with the most resource intensive being random forests.
+This is likely due to RF being an ensemble method. Where logistic regression and \kMeans{} clustering build a single model,
+RF is building a forest of trees. So not only is training longer, but then each sample in the test fold must be classified by each and every tree.
+
+LR is
 

### 2015-10-26 doc/publications/Recomb15/recomb15.tex
compare sup/unsup

@@
 In this section we compare RF against \kMeans\ and the LR.  
-\marginpar{TODO elaborate on the error} 
+\marginpar{TODO elaborate on the error Aidan: See near end of paragraph} 
 Despite using {\sc Spark} and {\sc MLlib}, RF is not able to scale to the full dataset of 1000 individuals and 40
@@
 super-populations.  We perform 10-fold cross-validation and report the mean error over all training and testing folds
-along with the standard error of the mean (SEM). We also use the Adjusted Rand Index as a metric, as it allows us to
-compare the classification results with the \kMeans{} clustering result. We also perform 10 runs of \kMeans{} with 4 clusters (\(k=4\)) and report the mean accuracy of the
+along with the standard error of the mean (SEM). We also use the Adjusted Rand Index as a metric for model quality
+as it allows us to compare the classification results with the \kMeans{} clustering model. We also calculate the error
+for the two classification methods, which is simply the number of misclassified individuals divided by the number of samples.
+We perform 10 runs of \kMeans{} with 4 clusters (\(k=4\)) and report the mean ARI of the
 resulting models along with the standard error.
 
-\marginpar{TODO elaborate on accuracy measures}
+\marginpar{TODO elaborate on accuracy measures Aidan: See above}
 As reported in our earlier work, \kMeans\ is not able to fit the dataset accurately.
-Although capable of fitting a relatively accurate model \(ARI=0.837\), the average ARI over ten models is  \(0.596 \pm 0.067\), see table~\ref{RF}.
-RF achieves a better fitting model \(ARI = 0.757 \pm 0.011\) but is outperformed by the simple logistic regression, with an ARI of \(0.923 \pm 0.015\).
+Although capable of fitting a relatively accurate model (\(ARI=0.837\)), the average ARI over ten models is  \(0.596 \pm 0.067\), see table~\ref{RF}.
+RF achieves a better fitting model with a mean ARI of \(0.769 \pm 0.021\) but is outperformed by the simple logistic regression, with an ARI of \(0.923 \pm 0.015\).
 This indicates that a multivariate approach may not be needed to accurately fit ethnicity.
@@
 \begin{table}
-\caption{Cross-validated results, mean accuracy (and standard error) on variants from chromosome 22 of 4 super-populations.}
+\caption{Cross-validated results, ARI and error (and standard error of mean) on variants from chromosome 22 of 4 super-populations.}
 \begin{center}
@@
 \hline\noalign{\smallskip}
-method  & ARI & error (\%) \\
-RF  & 0.757 (0.011) &0.13 (0.007) \\
-\kMeans & 0.596 (0.066) & -- \\
+method  & ARI & error \\
+RF  & 0.769 (0.021) &0.118 (0.011) \\
+\kMeans & 0.596 (0.067) & -- \\
 LR & 0.923 (0.015) & 0.034 (0.006) \\
@@
 \end{table}
+
+These results also demonstrate how much the quality of the models can vary between runs (or folds) for supervised and unsupervised learning.
+For both of the supervised methods, RF and LR, the ARI was fairly consistent when testing on each holdout set, with a standard deviation of
+\( 0.064 \) and 0.048, respectively. For the \kMeans{} models however, the standard deviation was 0.211.
+One pitfall of \kMeans{} clustering is that depending on the starting points, the resulting model may represent a local optimum of the dataset ~\cite{Steinley2003}.
+{\sc MLlib} uses {\sc \kMeans\textbar\textbar}~\cite{Bahmani2012} (a parallel implementation of \kMeans{}++) for the cluster initialisation.
+Although this produces close to the optimal solution for the initial cluster centres, these initial cluster centres are different for each model.
+Due to the amount of variance in the 1000 Genomes dataset, it's possible that multiple local optima are close to the global optimum, and the algorithm
+is choosing one that doesn't represent population structure.
 
@@
 
-The mean error for our \kMeans clustering models is approximately XXX. This is similar to the deep-learning approach
-where they achieved an accuracy of 99\% (two misclassified individuals). Our supervised logistic regression approach,
+The mean ARI for our \kMeans clustering models is \(0.960 \pm 0.004 \). Although slightly lower, this value is similar to the deep-learning approach
+where they achieved an ARI of 0.976 (where one ASW and one GBR individual is misclassified as CHB). Our supervised logistic regression approach,
 however, improves upon this result.  With this approach, we achieve an accuracy of 100\%. This is the average across
-five folds, with a standard error of 0. This is on the above three populations, ASW, CHB and GBR, that belong to the
-super-populations, AFR, EAS and EUR, respectively.
+ten folds, with a standard error of 0.
 
-Although we see two of our five runs meet the accuracy of the deep-learning approach, where they achieved an accuracy of
+Although we see two of our ten runs meet the accuracy of the deep-learning approach, where they achieved an accuracy of
 99\% (or ARI of 0.975, with two misclassified individuals), our logistic regression approach demonstrates a consistent
-improvement.  With this approach, we achieve an accuracy of \(99.997 \pm 0.003\) across five folds.
+improvement.  With this approach, we achieve an accuracy of \(99.997 \pm 0.003\) across ten folds.
 
 To reiterate, these results are from the three populations, ASW, CHB and GBR, that belong to the super-populations, AFR,
-EAS and EUR, respectively.  As we mention in the introduction, we observed a perfect clustering result when we cluster
-the entirety of these three super-populations (rather than one population from each), with an ARI of 1.0. However, when
+EAS and EUR, respectively.  As we mention in the introduction, we observed a perfect clustering result when we previously clustered
+the these three super-populations (rather than one population from each), with an ARI of 1.0. However, when
 we introduce the fourth super-population into the mix, AMR, the ARI drops to 0.84.  Given the perfect result when building
@@
 \hline\noalign{\smallskip}
-method  & ARI & training folds & hold out \\
-RF  & X (Y) & X (Y) & X (Y) \\
-\kMeans & X (Y) & X (Y) & X (Y) \\
-deep learning & 0.976 & X (Y) & X (Y) \\
+method  & ARI & error \\
+LR  & X (Y) & -- \\
+\kMeans & 0.960 (0.004) & -- \\
+deep learning & 0.976 & 0.008 \\
 \noalign{\smallskip}
@@
 \hline\noalign{\smallskip}
-method  & test folds & training folds & hold out \\
-RF  & X (Y) & X (Y) & X (Y) \\
-\kMeans & X (Y) & X (Y) & X (Y) \\
+method  & ARI & error \\
+LR  & 0.936 (0.014) & 0.029 (0.005) \\
+\kMeans & 0.84 & -- \\
 \noalign{\smallskip}

### 2015-10-26 doc/publications/Recomb15/recomb15.tex
compare sup/unsup

@@
 \( 0.064 \) and 0.048, respectively. For the \kMeans{} models however, the standard deviation was 0.211.
-One pitfall of \kMeans{} clustering is that depending on the starting points, the resulting model may represent a local optimum of the dataset ~\cite{Steinley2003}.
+One pitfall of \kMeans{} clustering is that depending on the starting points, the resulting model may represent a local optimum of the dataset~\cite{Steinley2003}.
 {\sc MLlib} uses {\sc \kMeans\textbar\textbar}~\cite{Bahmani2012} (a parallel implementation of \kMeans{}++) for the cluster initialisation.
-Although this produces close to the optimal solution for the initial cluster centres, these initial cluster centres are different for each model.
-Due to the amount of variance in the 1000 Genomes dataset, it's possible that multiple local optima are close to the global optimum, and the algorithm
-is choosing one that doesn't represent population structure.
+Although this produces close to the optimal solution for the initial cluster centres, due to the heuristical nature of the algorithm,
+these initial cluster centres will be different for each model. Due to the amount of variance in the 1000 Genomes dataset, it's possible that multiple local
+optima are close to the global optimum, and the algorithm is choosing one that doesn't represent population structure.
 

### 2015-10-26 doc/publications/Recomb15/recomb15.tex
more touchups

@@
 super-populations.  We perform 10-fold cross-validation and report the mean error over all training and testing folds
-along with the standard error of the mean (SEM). We also use the Adjusted Rand Index as a metric for model quality
-as it allows us to compare the classification results with the \kMeans{} clustering model. We also calculate the error
+along with the standard error of the mean (SEM). We also use the Adjusted Rand Index (ARI) as a metric for model quality
+as it allows us to compare the classification results with \kMeans{} clustering models. We also calculate the error
 for the two classification methods, which is simply the number of misclassified individuals divided by the number of samples.
@@
 
-\marginpar{TODO elaborate on accuracy measures Aidan: See above}
+\marginpar{TODO elaborate on accuracy measures Aidan: See above paragraph}
 As reported in our earlier work, \kMeans\ is not able to fit the dataset accurately.
 Although capable of fitting a relatively accurate model (\(ARI=0.837\)), the average ARI over ten models is  \(0.596 \pm 0.067\), see table~\ref{RF}.
-RF achieves a better fitting model with a mean ARI of \(0.769 \pm 0.021\) but is outperformed by the simple logistic regression, with an ARI of \(0.923 \pm 0.015\).
+RF achieves a better fitting model with a mean ARI of \(0.769 \pm 0.021\), however it's outperformed by the simple logistic regression, with an ARI of \(0.923 \pm 0.015\).
 This indicates that a multivariate approach may not be needed to accurately fit ethnicity.
@@
 however, improves upon this result.  With this approach, we achieve an accuracy of 100\%. This is the average across
-ten folds, with a standard error of 0.
-
-Although we see two of our ten runs meet the accuracy of the deep-learning approach, where they achieved an accuracy of
-99\% (or ARI of 0.975, with two misclassified individuals), our logistic regression approach demonstrates a consistent
-improvement.  With this approach, we achieve an accuracy of \(99.997 \pm 0.003\) across ten folds.
+ten folds.
 
@@
 ethnicity from whole genome variant data.  We found that a supervised method, logistic regression, outperforms the
-unsupervised approach, \kMeans{} (XX vs YY).  We also found that ethnicity can be better predicted by the simple
-logistic regression method compared to the multivariate approaches such as random forest or deep learning as reported by
-Big Data Genomics using the ADAM framework (XX vs Y1 and Y2, respectively).  Furthermore, we observe a limitation within
+unsupervised approach, \kMeans{} (0.936 vs 0.84).  We also found that ethnicity can be better predicted by the simple
+logistic regression method compared to the unsupervised \kMeans{} approach or the multivariate deep learning approach as reported by
+Big Data Genomics using the ADAM framework (1 vs 0.96 and 0.976, respectively).  Furthermore, we observe a limitation within
 Java, which limits the random forest implementation of {\sc Spark}'s {\sc MLlib} to XX features for the 1069 individuals
-of the 1000 genome project.  This property would limit the 100,000 genome project to using only YY features for their
+of the 1000 genome project.  This property would make feature selection mandatory for the 100,000 genome project when classifying their
 70,000 individuals, when using the same random forest implementation.  We conclude that while logistic regression

### 2015-10-26 doc/publications/Recomb15/recomb15.tex
more touchups

@@
 \marginpar{TODO report bd's accuracy measure as the main measure (add ARI as the additional information linking to our
-  paper)}
+  paper) Aidan:See email}
 \subsubsection{Logistic regression and \kMeans{} capture the population better than deep learning in a three population test}
@@
 where they achieved an ARI of 0.976 (where one ASW and one GBR individual is misclassified as CHB). Our supervised logistic regression approach,
-however, improves upon this result.  With this approach, we achieve an accuracy of 100\%. This is the average across
-ten folds.
+however, improves upon this result.  With this approach, we achieve an ARI of 1 across the ten folds.
 
@@
 we introduce the fourth super-population into the mix, AMR, the ARI drops to 0.84.  Given the perfect result when building
-the logistic regression model on this small subset of data, we expand our dataset to include every population from the
-1000 Genomes Project phase1 dataset.
+the logistic regression model on this small subset of data, and the similarity with the other models,
+we expand our dataset to include every population from the 1000 Genomes Project phase1 dataset.
 
 \marginpar{TODO[say that this may be a too easy dataset to benchmark the performance differences between kmeans and RF
-  (hence the full dataset needed)]}
+  (hence the full dataset needed) Aidan: done(probably)]}
 
@@
 method  & ARI & error \\
-LR  & X (Y) & -- \\
+LR  & 1.000 (0.000) & 0.000 \\
 \kMeans & 0.960 (0.004) & -- \\
@@
 
-\marginpar{TODO [Discuss supervised vs unsupervised]} \marginpar{TODO can you drill down to the population clusters and
+\marginpar{TODO [Discuss supervised vs unsupervised] Aidan: wrote in previous section (where I mention initialisation)} \marginpar{TODO can you drill down to the population clusters and
   see if and where the best performing method still makes mistakes - this will help in the conclusion to argue that
   actually the labels are wrong rather than the method not being able to capture the structure (e.g. say if AMR are
-  mixed in with GBR TSI)}
+  mixed in with GBR TSI) Aidan: not in time :( }
 

### 2015-10-26 doc/publications/Recomb15/recomb15.tex
final comments

@@
 library, {\sc MLlib}, which is capable of clustering variants from whole genome sequencing data.  We demonstrate its use
-by clustering samples from the 1000 Genomes Project~\cite{1KG2012}.  \marginpar{TODO add value of the kmeans method on
-  the full dataset (the equivalent of the ARI=0.84) Denis: correct/total happened to be 0.84 too ? or are you quoting
-  here the ARI ? I want to have the same accuracy measure as reported by bdgenomics for all comparisons in this paper,
-  also it goes between 0 and 100 right ?}
+by clustering samples from the 1000 Genomes Project~\cite{1KG2012}.
 
@@
 (EUR), however when adding the fourth super-population group, American (AMR), we observe that only a subset of the AMR
-individual form an independent cluster while the majority falls into the EUR cluster, resulting in an accuracy of 0.84,
+individual form an independent cluster while the majority falls into the EUR cluster. 
+We quantified this by calculating the \ARI\ (ARI), which compares the predicted and annotated labels. \kMeans achieved an ARI=0.84,
 with -1 being random and 1 a perfect clustering.  We argue that the suboptimal clustering is a result of using an
@@
 association and compare our results to the ones obtained using deep learning.
-\marginpar{TODO Rob can you think of a better description of what it does}
+%\marginpar{TODO Rob can you think of a better description of what it does}
 One of the simplest supervised machine learning methods is logistic regression (LR), which creates a decision boundary
@@
 
-\marginpar{TODO do we need to cite ~\cite{Bureau2005,Yoo2012,Qi2012}.}  
+%\marginpar{TODO do we need to cite ~\cite{Bureau2005,Yoo2012,Qi2012}.}  
 We also investigate more complex supervised
@@
 
-\marginpar{TODO Rob do you want to add two sentences about RF and DL and the respective benefits (ideally with respect
-  to the multi variable interactions and RF benefit of being hard to overfit).  Using conventional machine learning
-  analysis to elucidate this additive effects is challenging as there are not enough samples to robustly fit the model
-  over such a high-dimensional feature space~\cite{Chen2012}.  As a result machine learning methods such as random
-  forests (RF) have become popular as the underlying decision trees are build from a subsets of the features, which make
-  the approach robust against overfitting issues~\cite{Breiman2001}.  However, training a supervised learning method on
-  such an expansive dataset as the 1000 genomes is computationally challenging and traditional massively parallel
-  approaches to data processing are not easily applicable to machine learning tasks, as they iteratively refine models
-  based on information from the full dataset.}
+However, training a supervised learning method on such an expansive dataset as the 1000 genomes is computationally 
+challenging and traditional massively parallel approaches to data processing are not easily applicable to machine learning 
+tasks, as they iteratively refine models based on information from the full dataset.
 
 We therefore use {\sc Spark} through our recently developed \variantSpark~\cite{OBrien} framework, which allows us to
-apply RF in the {\sc MLlib} library to the variant file in the standard Variant Call Format (VCF) format.  Please note
+apply machine learning algorithms in the {\sc MLlib} library to the variant file in the standard Variant Call Format (VCF) format.  Please note
 that unlike {\sc Hadoop}, {\sc Spark} allows a more flexible software design to utilise node-parallelisation and offers
@@
 
-\marginpar{TODO check accuracy of titles}
 \section{Results}
@@
 \marginpar{TODO elaborate on accuracy measures Aidan: See above paragraph}
-As reported in our earlier work, \kMeans\ is not able to fit the dataset accurately.
-Although capable of fitting a relatively accurate model (\(ARI=0.837\)), the average ARI over ten models is  \(0.596 \pm 0.067\), see table~\ref{RF}.
-RF achieves a better fitting model with a mean ARI of \(0.769 \pm 0.021\), however it's outperformed by the simple logistic regression, with an ARI of \(0.923 \pm 0.015\).
-This indicates that a multivariate approach may not be needed to accurately fit ethnicity.
+As reported in our earlier work, \kMeans\ is not able to fit the dataset accurately, with an average ARI over ten models of \(0.596 \pm 0.067\), see table~\ref{RF}.
+%Although capable of fitting a relatively accurate model (\(ARI=0.837\)), The average ARI over ten models is  \(0.596 \pm 0.067\), see table~\ref{RF}.
+RF achieves a better fitting model with a mean ARI of \(0.769 \pm 0.021\), however it is outperformed by the LR, with an ARI of \(0.923 \pm 0.015\).
+This indicates that while applying a supervised learning method results in higher accuracy, a multivariate approach may not be needed to accurately fit ethnicity.
 
@@
 Although this produces close to the optimal solution for the initial cluster centres, due to the heuristical nature of the algorithm,
-these initial cluster centres will be different for each model. Due to the amount of variance in the 1000 Genomes dataset, it's possible that multiple local
-optima are close to the global optimum, and the algorithm is choosing one that doesn't represent population structure.
+these initial cluster centres will be different for each model. Due to the amount of variance in the 1000 Genomes dataset, it is possible that multiple local
+optima are close to the global optimum, and the algorithm is choosing one that does not represent population structure.
 
@@
 \subsubsection{Logistic regression and \kMeans{} capture the population better than deep learning in a three population test}
-In this section we investigate whether the multivariate approach, deep learning, is able to achieve a better result than
+In this section we investigate whether another multivariate approach, DL, is able to achieve a better result than
 RF by again comparing it against \kMeans{} and logistic regression.  We compare against the reported accuracy of the
@@
 
-The mean ARI for our \kMeans clustering models is \(0.960 \pm 0.004 \). Although slightly lower, this value is similar to the deep-learning approach
-where they achieved an ARI of 0.976 (where one ASW and one GBR individual is misclassified as CHB). Our supervised logistic regression approach,
-however, improves upon this result.  With this approach, we achieve an ARI of 1 across the ten folds.
+The mean ARI for our \kMeans clustering models on this dataset is \(0.960 \pm 0.004 \). 
+%DENIS: lower compared to what ?
+Although slightly lower, this value is similar to the one reported for the deep-learning approach, which achieved an ARI of 0.976 (where one ASW and one GBR individual is misclassified as CHB). 
+Our supervised logistic regression approach, however, improves upon this result and achieves an ARI of 1 across the ten folds.
 
+%DENIS: I don't understand this 
 To reiterate, these results are from the three populations, ASW, CHB and GBR, that belong to the super-populations, AFR,
 EAS and EUR, respectively.  As we mention in the introduction, we observed a perfect clustering result when we previously clustered
-the these three super-populations (rather than one population from each), with an ARI of 1.0. However, when
-we introduce the fourth super-population into the mix, AMR, the ARI drops to 0.84.  Given the perfect result when building
+ these three super-populations (rather than one population from each), with an ARI of 1.0. However, when
+we introduce the fourth super-population into the mix, AMR, the ARI drops to 0.84.
+
+Given the perfect result when building
 the logistic regression model on this small subset of data, and the similarity with the other models,
 we expand our dataset to include every population from the 1000 Genomes Project phase1 dataset.
-
-\marginpar{TODO[say that this may be a too easy dataset to benchmark the performance differences between kmeans and RF
-  (hence the full dataset needed) Aidan: done(probably)]}
 
@@
 \subsubsection{Logistic regression is more accurate than \kMeans{} in capturing the full population structure}
-In this section we train and test our two methods on the full dataset of 14 populations and assess the accuracy with respect to the accurate
-classification into the 4 super-populations.
+In this section we apply both the supervised LR and unsupervised \kMeans\ method on the full dataset of 14 populations and assess the 
+classification accuracy into the 4 super-populations.
 Using 10-fold cross-validation to assess the logistic regression classification results, we observe an average ARI of \(0.936 \pm{} 0.014\).
+%DENIS: above you argue that kMeans has local optima and results might hence be too optimistic but here you only train one model. So either remove the above section or add some repeats to this one
 This is an improvement on our previous result from \kMeans{} clustering the same dataset in our previous work, where we observed an ARI of \(0.83\).
@@
 This is due to individuals from the AFR and EUR super-populations being misclassified as AMR.
-Consequently, this led to these two super-populations having a lower recall of \(0.956 \pm 0.01 \) and \(0.950 \pm 0.02 \), respectively.
-Although these misclassifications as AMR are relatively frequent, we didn't observe any AFR individuals being classified as EUR, or vice-versa, in any of the ten folds.
+Consequently, this leds to these two super-populations having a lower recall of \(0.956 \pm 0.01 \) and \(0.950 \pm 0.02 \), respectively.
+Although these misclassifications as AMR are relatively frequent, we did not observe any AFR individuals being classified as EUR, or {\it vice versa}, in any of the ten folds.
 Also, only in one fold did we observe one AMR individual misclassified as another super-population (AFR).
@@
 unsupervised approach, \kMeans{} (0.936 vs 0.84).  We also found that ethnicity can be better predicted by the simple
-logistic regression method compared to the unsupervised \kMeans{} approach or the multivariate deep learning approach as reported by
+logistic regression method compared to the 
+%unsupervised \kMeans{} approach or the 
+multivariate random forest or deep learning approach as reported by
 Big Data Genomics using the ADAM framework (1 vs 0.96 and 0.976, respectively).  Furthermore, we observe a limitation within
@@
 70,000 individuals, when using the same random forest implementation.  We conclude that while logistic regression
-outperforms all other methods it remains unable to accurately classify individuals from [Sub-populations] from the AMR
-super-population.  We therefore ague that the remaining mixture in X is reflective of the migrational background and
-hence the underlying genomic profile of Y is accurately similar to X.  This may be an important observation for medical
+outperforms all other methods it remains unable to accurately identify all individuals from EUR (lowest precision) or misclassify AMR (lowest recall) individuals. 
+We therefore ague that the remaining mixture is potentially reflective of the migrational background, which may be an important observation for medical
 applications that need accurate population association.
+
+%DENIS - does recomb paper need a method section ??
 

### 2015-10-26 doc/publications/Recomb15/recomb15.tex
added my_plain.bst -- modified to reduce bib authors
small changes

@@
 \usepackage{placeins}
+\usepackage{hyperref}
 %
@@
 \newcommand{\variantSpark}{{\sc VariantSpark}}
-\newcommand{\kMeans}{\textit{k}-means}
+\newcommand{\kMeans}{\textit{k}-means }
 \newcommand{\ARI}{adjusted Rand index}
@@
 
+\marginpar{I don't understand this. Can't you just sak a patient about their ancestry?}
 Determining a person's ethnicity has become important for medical applications, such as HLA allele genotyping from SNP
@@
 (EUR), however when adding the fourth super-population group, American (AMR), we observe that only a subset of the AMR
-individual form an independent cluster while the majority falls into the EUR cluster. 
-We quantified this by calculating the \ARI\ (ARI), which compares the predicted and annotated labels. \kMeans achieved an ARI=0.84,
-with -1 being random and 1 a perfect clustering.  We argue that the suboptimal clustering is a result of using an
-unsupervised method rather than an insufficient number of genomic markers~\cite{Patterson2006} and hence hypothesise
-that a supervised machine learning method may be capable of separating these subgroups of low genetic divergence.
-Indeed, bigdata genomics recently investigated the supervised method, deep learning, for clustering the 1000 genomes
-data by ethnicity~\cite{Ferguson}.
+individual form an independent cluster while the majority falls into the EUR cluster.  We quantified this by calculating
+the \ARI\ ( ARI), which compares the predicted and annotated labels. \kMeans achieved an ARI of 0.84, with -1 being
+random and 1 a perfect clustering.  We argue that the suboptimal clustering is a result of using an unsupervised method
+rather than an insufficient number of genomic markers~\cite{Patterson2006} and hence hypothesise that a supervised
+machine learning method may be capable of separating these subgroups of low genetic divergence.  Indeed, bigdata
+genomics recently investigated the supervised method, deep learning, for clustering the 1000 genomes data by
+ethnicity~\cite{Ferguson}.  \marginpar{``bigdata genomics'' seems to be a domain name \url{http://bdgenomics.org/} --
+  how should we cite this}
 
@@
 approaches, namely Random Forest (RF), as for example Locke {\it et al.}~\cite{Locke2015} showed in a recent meta-study
-around genetic susceptibility of obesity that only five of the 97 genomic loci have an independent association signal
+around genetic susceptibility to obesity that only five of the 97 genomic loci have an independent association signal
 the rest seem to jointly predict the complex metabolic phenotype.  We therefore investigate whether ethnicity is a
@@
 This indicates that while applying a supervised learning method results in higher accuracy, a multivariate approach may not be needed to accurately fit ethnicity.
+
+\marginpar{not sure what is meant here. RF, LR etc are not multivariate -- there is only one $y$ variable, obesity. They
+are multi-variable.}
 
@@
 
-\bibliographystyle{plain}
-\marginpar{TODO you need to find a way of shortening the Locket paper author list}
+\bibliographystyle{my_plain}
+\marginpar{FIXED:  TODO you need to find a way of shortening the Locket paper author list}
 \bibliography{recomb}

### 2015-10-26 doc/publications/Recomb15/recomb15.tex
broken thing

@@
 \usepackage{hyperref}
+\usepackage{amsmath}
 %
@@
 \begin{abstract}
-This paragraph shall summarize the contents of the paper
-in short terms.
+Using our 
+We apply various machine learning algorithms to the 1092 individuals in the 1000 Genomes Project dataset.
+Although each algorithm we investigate can cluster individuals from three populations with near perfect accuracy,
+we find logistic regression to be superior. Comparing  With the entire dataset,
+
 \end{abstract}
@@
 
-\marginpar{I don't understand this. Can't you just sak a patient about their ancestry?}
+\marginpar{I don't understand this. Can't you just ask a patient about their ancestry?}
 Determining a person's ethnicity has become important for medical applications, such as HLA allele genotyping from SNP
@@
 One of the simplest supervised machine learning methods is logistic regression (LR), which creates a decision boundary
-by fitting a linear combination of the feature variables.  It is therefore able to adjust the weight of the individual
-genotypes to increase the accuracy of the multi-class prediction.
+by fitting a linear combination of the feature variables.  In this appliction it is not fesible to model interactions,
+so the model is retricted to main effects.
 
@@
 section.  In the second section we compare LR and \kMeans{} against the reported accuracy of the supervised multivariate
-DL method.  In section three we record the accuracy of LR and \kMeans{} on the full dataset set using all populations.
+deep learning (DL) method.  In section three we record the accuracy of LR and \kMeans{} on the full dataset set using all populations.
 In the last section we discuss the memory consumptions and runtime of LR compared to \kMeans{}.
+
+\clearpage
+\hrule
+I think the use of the term ``multivariate'' is a bit loose. We have only one $y$ variable -- obesity. By analogy with linear regression:
+\begin{itemize}
+\item simple linear regression $y=\beta_0 + \beta_1x_1$
+\item multi-variable regression $y=\beta_0 + \beta_1x_1 + \beta_2 x_2$
+\item multivariate regression $\underset{n\times m}{Y}= \underset{(n\times p+1)}{X} \underset{(p+1 \times m)}{B}$
+\end{itemize}
+So RF, DL and LR are all multi-variable (in this application).
+
+A multivarite technique we could use is PCA, $T = X W$ where $W$ is a $p \times p$ matrix whose columns are the eigenvectors of $X^tX$. 
+\hrule
 
@@
 In this section we compare RF against \kMeans\ and the LR.  
-\marginpar{TODO elaborate on the error Aidan: See near end of paragraph} 
 Despite using {\sc Spark} and {\sc MLlib}, RF is not able to scale to the full dataset of 1000 individuals and 40
@@
 
-\marginpar{TODO elaborate on accuracy measures Aidan: See above paragraph}
 As reported in our earlier work, \kMeans\ is not able to fit the dataset accurately, with an average ARI over ten models of \(0.596 \pm 0.067\), see table~\ref{RF}.
@@
 This indicates that while applying a supervised learning method results in higher accuracy, a multivariate approach may not be needed to accurately fit ethnicity.
-
-\marginpar{not sure what is meant here. RF, LR etc are not multivariate -- there is only one $y$ variable, obesity. They
-are multi-variable.}
 
@@
 
-These results also demonstrate how much the quality of the models can vary between runs (or folds) for supervised and unsupervised learning.
-For both of the supervised methods, RF and LR, the ARI was fairly consistent when testing on each holdout set, with a standard deviation of
-\( 0.064 \) and 0.048, respectively. For the \kMeans{} models however, the standard deviation was 0.211.
-One pitfall of \kMeans{} clustering is that depending on the starting points, the resulting model may represent a local optimum of the dataset~\cite{Steinley2003}.
-{\sc MLlib} uses {\sc \kMeans\textbar\textbar}~\cite{Bahmani2012} (a parallel implementation of \kMeans{}++) for the cluster initialisation.
-Although this produces close to the optimal solution for the initial cluster centres, due to the heuristical nature of the algorithm,
-these initial cluster centres will be different for each model. Due to the amount of variance in the 1000 Genomes dataset, it is possible that multiple local
-optima are close to the global optimum, and the algorithm is choosing one that does not represent population structure.
+%These results also demonstrate how much the quality of the models can vary between runs (or folds) for supervised and unsupervised learning.
+%For both of the supervised methods, RF and LR, the ARI was fairly consistent when testing on each holdout set, with a standard deviation of
+%\( 0.064 \) and 0.048, respectively. For the \kMeans{} models however, the standard deviation was 0.211.
+%One pitfall of \kMeans{} clustering is that depending on the starting points, the resulting model may represent a local optimum of the dataset~\cite{Steinley2003}.
+%{\sc MLlib} uses {\sc \kMeans\textbar\textbar}~\cite{Bahmani2012} (a parallel implementation of \kMeans{}++) for the cluster initialisation.
+%Although this produces close to the optimal solution for the initial cluster centres, due to the heuristical nature of the algorithm,
+%these initial cluster centres will be different for each model. Due to the amount of variance in the 1000 Genomes dataset, it is possible that multiple local
+%optima are close to the global optimum, and the algorithm is choosing one that does not represent population structure.
 
-\FloatBarrier
-\clearpage
-\marginpar{TODO report bd's accuracy measure as the main measure (add ARI as the additional information linking to our
-  paper) Aidan:See email}
 \subsubsection{Logistic regression and \kMeans{} capture the population better than deep learning in a three population test}
@@
 
-The mean ARI for our \kMeans clustering models on this dataset is \(0.960 \pm 0.004 \). 
-%DENIS: lower compared to what ?
-Although slightly lower, this value is similar to the one reported for the deep-learning approach, which achieved an ARI of 0.976 (where one ASW and one GBR individual is misclassified as CHB). 
+The mean ARI for our \kMeans{} clustering models on this dataset is \(0.960 \pm 0.004 \). 
+This value is slightly less than the one reported for the deep-learning approach, which achieved an ARI of 0.976 (where one ASW and one GBR individual are misclassified as CHB). 
 Our supervised logistic regression approach, however, improves upon this result and achieves an ARI of 1 across the ten folds.
 
-%DENIS: I don't understand this 
-To reiterate, these results are from the three populations, ASW, CHB and GBR, that belong to the super-populations, AFR,
-EAS and EUR, respectively.  As we mention in the introduction, we observed a perfect clustering result when we previously clustered
- these three super-populations (rather than one population from each), with an ARI of 1.0. However, when
-we introduce the fourth super-population into the mix, AMR, the ARI drops to 0.84.
-
-Given the perfect result when building
-the logistic regression model on this small subset of data, and the similarity with the other models,
-we expand our dataset to include every population from the 1000 Genomes Project phase1 dataset.
+Although we see a slight improvement with logistic regression in regards to other algorithms, their results are still very similar.
+However, there is not much room in the first place for logistic regression to demonstrate any potential increases in model accuracy.
+We therefore proceed to apply our method to the entire 1000 Genomes Project phase 1 dataset, which we previously clustered with an ARI of 0.84.
 
@@
 This is due to individuals from the AFR and EUR super-populations being misclassified as AMR.
-Consequently, this leds to these two super-populations having a lower recall of \(0.956 \pm 0.01 \) and \(0.950 \pm 0.02 \), respectively.
+Consequently, this leads to these two super-populations having a lower recall of \(0.956 \pm 0.01 \) and \(0.950 \pm 0.02 \), respectively.
 Although these misclassifications as AMR are relatively frequent, we did not observe any AFR individuals being classified as EUR, or {\it vice versa}, in any of the ten folds.
@@
 These results demonstrate the diversity, or relative difference of some of the groups.
-
-\marginpar{TODO [Discuss supervised vs unsupervised] Aidan: wrote in previous section (where I mention initialisation)} \marginpar{TODO can you drill down to the population clusters and
-  see if and where the best performing method still makes mistakes - this will help in the conclusion to argue that
-  actually the labels are wrong rather than the method not being able to capture the structure (e.g. say if AMR are
-  mixed in with GBR TSI) Aidan: not in time :( }
 
@@
 This is likely due to RF being an ensemble method. Where logistic regression and \kMeans{} clustering build a single model,
-RF is building a forest of trees. So not only is training longer, but then each sample in the test fold must be classified by each and every tree.
+RF is building an forest of decision trees. So not only is training longer, but then each sample in the test fold must be classified by each and every tree.
 
@@
 \bibliographystyle{my_plain}
-\marginpar{FIXED:  TODO you need to find a way of shortening the Locket paper author list}
 \bibliography{recomb}

### 2015-10-26 doc/publications/Recomb15/recomb15.tex

@@
 In this paper we therefore systematically evaluate supervised and unsupervised methods for determining population
-association and compare our results to the ones obtained using deep learning.
+association and compare our results to those obtained using deep learning.
 %\marginpar{TODO Rob can you think of a better description of what it does}
 One of the simplest supervised machine learning methods is logistic regression (LR), which creates a decision boundary
-by fitting a linear combination of the feature variables.  In this appliction it is not fesible to model interactions,
-so the model is retricted to main effects.
+by fitting a linear combination of the feature variables.  In this appliction, it is not feasible to model interactions and
+so the model is restricted to main effects.
 
@@
 We also investigate more complex supervised
-approaches, namely Random Forest (RF), as for example Locke {\it et al.}~\cite{Locke2015} showed in a recent meta-study
-around genetic susceptibility to obesity that only five of the 97 genomic loci have an independent association signal
-the rest seem to jointly predict the complex metabolic phenotype.  We therefore investigate whether ethnicity is a
-similarly complex phenotype that requires higher order dependencies between features to be taken into
-account. %, as done by RF and DL.
+approaches, namely Random Forest (RF), with the rationale being that interaction effects may dominate. A recent meta-study by
+Locke {\it et al.}~\cite{Locke2015} found across 97 genomic loci associated with genetic susceptibility to obesity, only five
+had an independent association signal and the rest seem to jointly predict this complex metabolic phenotype.  
+We therefore investigate whether ethnicity is a similarly complex phenotype that requires higher order dependencies between
+features to be taken into account. %, as done by RF and DL.
 
@@
 provides an unbiased estimate of the generalization error (in addition, it handles missing values and unbalanced
-population classes sizes). They provide a measure of variable importance and allow for interactions between
+population classes sizes). RF provide a measure of variable importance and allow for interactions between
 variables to be investigated. However, in this problem, with many millions of variables, inspecting plots of interactions

### 2015-10-26 doc/publications/Recomb15/recomb15.tex
updates

--- 2015-10-26 doc/publications/Recomb15/recomb15.tex
@@
 
-\clearpage
-\hrule
-I think the use of the term ``multivariate'' is a bit loose. We have only one $y$ variable -- obesity. By analogy with linear regression:
-\begin{itemize}
-\item simple linear regression $y=\beta_0 + \beta_1x_1$
-\item multi-variable regression $y=\beta_0 + \beta_1x_1 + \beta_2 x_2$
-\item multivariate regression $\underset{n\times m}{Y}= \underset{(n\times p+1)}{X} \underset{(p+1 \times m)}{B}$
-\end{itemize}
-So RF, DL and LR are all multi-variable (in this application).
+%\clearpage
+%\hrule
+%I think the use of the term ``multivariate'' is a bit loose. We have only one $y$ variable -- obesity. By analogy with linear regression:
+%\begin{itemize}
+%\item simple linear regression $y=\beta_0 + \beta_1x_1$
+%\item multi-variable regression $y=\beta_0 + \beta_1x_1 + \beta_2 x_2$
+%\item multivariate regression $\underset{n\times m}{Y}= \underset{(n\times p+1)}{X} \underset{(p+1 \times m)}{B}$
+%\end{itemize}
+%So RF, DL and LR are all multi-variable (in this application).
 
-A multivarite technique we could use is PCA, $T = X W$ where $W$ is a $p \times p$ matrix whose columns are the eigenvectors of $X^tX$. 
-\hrule
+%A multivarite technique we could use is PCA, $T = X W$ where $W$ is a $p \times p$ matrix whose columns are the eigenvectors of $X^tX$. 
+%\hrule
 
@@
 random forest implementation to fail. This error is due to the number of features present in each sample,
-with a potential solution being to limit the features through feature selection, or simply using a subset of the dataset.
-We therefore use only chromosome 22 (494,328 alleles) for the comparison in this section.  We train a RF
-and LR on this subset of genomic profiles from the 1000 Genomes data using the annotated labels of all
-super-populations.  We perform 10-fold cross-validation and report the mean error over all training and testing folds
-along with the standard error of the mean (SEM). We also use the Adjusted Rand Index (ARI) as a metric for model quality
-as it allows us to compare the classification results with \kMeans{} clustering models. We also calculate the error
+rather than the number of samples, and a potential solution is to either limit the features through feature selection, or to simply use a subset of the dataset.
+We therefore use only chromosome 22 (494,328 variants) for the comparison in this section.  We train a RF
+and LR on this subset of genomic profiles from the 1000 Genomes Project dataset using the annotated labels of all
+super-populations.  We perform 10-fold cross-validation and calculate the Adjusted Rand Index (ARI) for use as a metric for model quality.
+The ARI allows us to compare the classification results with \kMeans{} clustering results, as it takes the partitioning of the dataset into account, rather than just labels.
+report the mean error over all training and testing folds
+along with the standard error of the mean (SEM)
+We also calculate the error
 for the two classification methods, which is simply the number of misclassified individuals divided by the number of samples.
@@
 Using 10-fold cross-validation to assess the logistic regression classification results, we observe an average ARI of \(0.936 \pm{} 0.014\).
-%DENIS: above you argue that kMeans has local optima and results might hence be too optimistic but here you only train one model. So either remove the above section or add some repeats to this one
 This is an improvement on our previous result from \kMeans{} clustering the same dataset in our previous work, where we observed an ARI of \(0.83\).
@@
 
-Time and resources vary greatly between the three algorithms, with the most resource intensive being random forests.
-This is likely due to RF being an ensemble method. Where logistic regression and \kMeans{} clustering build a single model,
-RF is building an forest of decision trees. So not only is training longer, but then each sample in the test fold must be classified by each and every tree.
-
-LR is
-
-\marginpar{TODO [write up the observations and argue why one is less resource hungry than the other one]}
+Time and resources vary greatly between the three algorithms, with the most resource intensive being random forests (see table~\ref{resources}).
+This is likely due to RF being an ensemble method. For example, where logistic regression and \kMeans{} clustering build a single model,
+RF is building an forest of decision trees.
+When training on chromosome 22 (494,328 variants), resource requirements are kept to a minimum, with no more than 3GB memory required per executor (node)
+(6GB for RF). Although we specify 80 executors, the same algorithm could successfully complete on a much smaller cluster, at the expense of time. 
+However, when we train the algorithms on the entire genome, memory requirements increase to 24GB. This limited us to using 40 executors, due to memory availability.
+The memory requirements also restricts a job of that size to clusters with at least 24GB available. This demonstrates that pre-processing the data to reduce its dimensionality will
+be an essential step for future work, especially as datasets grow outwards.
 
@@
 \hline\noalign{\smallskip}
-method  & nodes & runtime & memory \\
-RF  & X (Y) & X (Y) & X (Y) \\
-\kMeans & X (Y) & X (Y) & X (Y) \\
+method  & chromosomes & executors & runtime & memory (GB) \\
+\kMeans & 22 & 80 & 2mins 3secs & 3 \\
+LR  & 22 & 80 & 5mins 1sec & 3 \\
+RF & 22 & 80 & 17mins 20secs & 6 \\
+\kMeans & 1-22 & 40 & 14hrs 44mins & 24 \\
+LR  & 1-22 & 40 & 30hrs 12mins & 24 \\
 \noalign{\smallskip}
 \hline
+\label{resources}
 \end{tabular}
@@
 Big Data Genomics using the ADAM framework (1 vs 0.96 and 0.976, respectively).  Furthermore, we observe a limitation within
-Java, which limits the random forest implementation of {\sc Spark}'s {\sc MLlib} to XX features for the 1069 individuals
-of the 1000 genome project.  This property would make feature selection mandatory for the 100,000 genome project when classifying their
+Java, which limits the random forest implementation of {\sc Spark}'s {\sc MLlib} to XX features for the 1096 individuals
+of the 1000 Genomes Project.  This property would make feature selection mandatory for the 100,000 genome project when classifying their
 70,000 individuals, when using the same random forest implementation.  We conclude that while logistic regression

### 2015-10-27 doc/publications/Recomb15/recomb15.tex
updates

@@
 \begin{abstract}
-Using our 
-We apply various machine learning algorithms to the 1092 individuals in the 1000 Genomes Project dataset.
-Although each algorithm we investigate can cluster individuals from three populations with near perfect accuracy,
-we find logistic regression to be superior. Comparing  With the entire dataset,
+Using VariantSpark, we compare various machine learning algorithms on the 1092 individuals in the 1000 Genomes Project dataset.
+Although each algorithm that we investigate can classify or cluster individuals from three populations with near perfect accuracy,
+we find logistic regression to outperform the others. We observe a similar outcome on the entire dataset, with logistic regression
+classifying the majority of the individuals correctly.
 
@@
 individual form an independent cluster while the majority falls into the EUR cluster.  We quantified this by calculating
-the \ARI\ ( ARI), which compares the predicted and annotated labels. \kMeans achieved an ARI of 0.84, with -1 being
+the \ARI\ (ARI), which compares the predicted and annotated labels. \kMeans achieved an ARI of 0.84, with -1 being
 random and 1 a perfect clustering.  We argue that the suboptimal clustering is a result of using an unsupervised method
@@
 machine learning method may be capable of separating these subgroups of low genetic divergence.  Indeed, bigdata
-genomics recently investigated the supervised method, deep learning, for clustering the 1000 genomes data by
+genomics recently investigated the supervised method, deep learning, for clustering the 1000 Genomes data by
 ethnicity~\cite{Ferguson}.  \marginpar{``bigdata genomics'' seems to be a domain name \url{http://bdgenomics.org/} --
@@
 We also investigate more complex supervised
-approaches, namely Random Forest (RF), as for example Locke {\it et al.}~\cite{Locke2015} showed in a recent meta-study
-around genetic susceptibility to obesity that only five of the 97 genomic loci have an independent association signal
-the rest seem to jointly predict the complex metabolic phenotype.  We therefore investigate whether ethnicity is a
-similarly complex phenotype that requires higher order dependencies between features to be taken into
-account. %, as done by RF and DL.
+approaches, namely Random Forests (RF), with the rationale being that interaction effects may dominate. A recent meta-study by
+Locke {\it et al.}~\cite{Locke2015} found across 97 genomic loci associated with genetic susceptibility to obesity, only five
+had an independent association signal and the rest seem to jointly predict this complex metabolic phenotype.  
+We therefore investigate whether ethnicity is a similarly complex phenotype that requires higher order dependencies between
+features to be taken into account. %, as done by RF and DL.
 
@@
 provides an unbiased estimate of the generalization error (in addition, it handles missing values and unbalanced
-population classes sizes). They provide a measure of variable importance and allow for interactions between
+population class sizes). RF provide a measure of variable importance and allow for interactions between
 variables to be investigated. However, in this problem, with many millions of variables, inspecting plots of interactions
@@
 
-However, training a supervised learning method on such an expansive dataset as the 1000 genomes is computationally 
+However, training a supervised learning method on such an expansive dataset as the 1000 Genomes Project phase 1 dataset is computationally 
 challenging and traditional massively parallel approaches to data processing are not easily applicable to machine learning 
@@
 super-populations.  We perform 10-fold cross-validation and calculate the Adjusted Rand Index (ARI) for use as a metric for model quality.
-The ARI allows us to compare the classification results with \kMeans{} clustering results, as it takes the partitioning of the dataset into account, rather than just labels.
-report the mean error over all training and testing folds
-along with the standard error of the mean (SEM)
-We also calculate the error
-for the two classification methods, which is simply the number of misclassified individuals divided by the number of samples.
-We perform 10 runs of \kMeans{} with 4 clusters (\(k=4\)) and report the mean ARI of the
-resulting models along with the standard error.
+This metric allows us to compare the classification results with \kMeans{} clustering results.
+We also report the mean error for the two classification methods, which is simply the number of misclassified individuals divided by the number of samples.
 
-As reported in our earlier work, \kMeans\ is not able to fit the dataset accurately, with an average ARI over ten models of \(0.596 \pm 0.067\), see table~\ref{RF}.
+We perform 10 runs of \kMeans{} with 4 clusters (\(k=4\)). As reported in our earlier work, \kMeans\ is not able to fit the dataset accurately,
+with an average ARI over ten models of \(0.596 \pm 0.067\) (see table~\ref{RF}).
 %Although capable of fitting a relatively accurate model (\(ARI=0.837\)), The average ARI over ten models is  \(0.596 \pm 0.067\), see table~\ref{RF}.
@@
 deep learning approach using the {\sc ADAM} framework~\cite{Massie2013} as reported on their webpage~\cite{Ferguson}.
-As the reported results were performed on only three of the 14 populations, we hence restrict our comparison also to ASW
+As the reported results were performed on only three of the 14 populations, we similarly restrict our comparison to ASW
 (Americans of African Ancestry in SW USA), CHB (Han Chinese in Beijing, China), and GBR (British in England and

### 2015-10-27 doc/publications/Recomb15/recomb15.tex
updates

@@
 \begin{abstract}
-Using VariantSpark, we compare various machine learning algorithms on the 1092 individuals in the 1000 Genomes Project dataset.
-Although each algorithm that we investigate can classify or cluster individuals from three populations with near perfect accuracy,
-we find logistic regression to outperform the others. We observe a similar outcome on the entire dataset, with logistic regression
-classifying the majority of the individuals correctly.
+Using VariantSpark, we compare various machine learning algorithms on the 1000 Genomes Project dataset.
+The algorithms we investigate classify or cluster the individuals from three populations with near perfect accuracy,
+with logistic regression outperforming the others.
+We proceed to investigate the algorithms on all four super-populations, where we also observe
+logistic regression to outperform the rest.
+However, even with the parallel nature of {\sc Spark}, resource requirements increase dramatically with the feature size,
+indicating that feature selection will be a requirement for larger datasets.
 
@@
 
-\marginpar{I don't understand this. Can't you just ask a patient about their ancestry?}
+%\marginpar{I don't understand this. Can't you just ask a patient about their ancestry?}
 Determining a person's ethnicity has become important for medical applications, such as HLA allele genotyping from SNP
@@
 genomics recently investigated the supervised method, deep learning, for clustering the 1000 Genomes data by
-ethnicity~\cite{Ferguson}.  \marginpar{``bigdata genomics'' seems to be a domain name \url{http://bdgenomics.org/} --
-  how should we cite this}
+ethnicity~\cite{Ferguson}.
+%\marginpar{``bigdata genomics'' seems to be a domain name \url{http://bdgenomics.org/} --  how should we cite this}
 
@@
 deep learning (DL) method.  In section three we record the accuracy of LR and \kMeans{} on the full dataset set using all populations.
-In the last section we discuss the memory consumptions and runtime of LR compared to \kMeans{}.
+In the last section we discuss the memory consumptions and runtime of LR compared to \kMeans{} clustering.
 
@@
 \subsubsection{Random Forest does not scale to whole genome analysis and is outperformed by logistic regression}
-In this section we compare RF against \kMeans\ and the LR.  
+In this section we compare RF against \kMeans\ and LR.  
 Despite using {\sc Spark} and {\sc MLlib}, RF is not able to scale to the full dataset of 1000 individuals and 40
@@
 %Although capable of fitting a relatively accurate model (\(ARI=0.837\)), The average ARI over ten models is  \(0.596 \pm 0.067\), see table~\ref{RF}.
-RF achieves a better fitting model with a mean ARI of \(0.769 \pm 0.021\), however it is outperformed by the LR, with an ARI of \(0.923 \pm 0.015\).
+RF achieves a better fitting model with a mean ARI of \(0.769 \pm 0.021\), however it is outperformed by LR, with an ARI of \(0.923 \pm 0.015\).
 This indicates that while applying a supervised learning method results in higher accuracy, a multivariate approach may not be needed to accurately fit ethnicity.
@@
 
-Although we see a slight improvement with logistic regression in regards to other algorithms, their results are still very similar.
+Although we see this slight improvement with logistic regression in regards to other algorithms, their results are still very similar.
 However, there is not much room in the first place for logistic regression to demonstrate any potential increases in model accuracy.
@@
 Using 10-fold cross-validation to assess the logistic regression classification results, we observe an average ARI of \(0.936 \pm{} 0.014\).
-This is an improvement on our previous result from \kMeans{} clustering the same dataset in our previous work, where we observed an ARI of \(0.83\).
+This is an improvement on our result from \kMeans{} clustering the same dataset in our previous work, where we observed an ARI of \(0.84\) (see table~\ref{ALLCHR}).
 
 \begin{table}
-\caption{Cross-validated results, mean accuracy (and standard error) on the 4 super-populations.}
+\caption{Cross-validated results, mean accuracy (and standard error of the mean) on the 4 super-populations.}
 \begin{center}
@@
 \hline
+\label{ALLCHR}
 \end{tabular}
@@
 
-To determine where the model is underperforming, we also record the precision and recall for each of the groups.
+To determine where the model is underperforming, we also record the precision and recall for each of the groups (see table~\ref{PR}).
 We observe that for each model, the precision and recall are always 1.0 for the EAS super-population.
@@
 \hline
+\label{PR}
 \end{tabular}
@@
 
-\subsubsection{\kMeans{} clustering is faster and consumes less memory than random forests}
+\subsubsection{\kMeans{} clustering is faster and consumes less memory than random forests and logistic regression}
 
-Time and resources vary greatly between the three algorithms, with the most resource intensive being random forests (see table~\ref{resources}).
+Time and resource consumption vary greatly between the three algorithms, with the most resource intensive being random forests (see table~\ref{resources}).
 This is likely due to RF being an ensemble method. For example, where logistic regression and \kMeans{} clustering build a single model,
-RF is building an forest of decision trees.
+RF is building an forest of decision trees. Although memory requirements for LR are identical to \kMeans{} clustering, LR requires more than twice as much time.
+
 When training on chromosome 22 (494,328 variants), resource requirements are kept to a minimum, with no more than 3GB memory required per executor (node)
-(6GB for RF). Although we specify 80 executors, the same algorithm could successfully complete on a much smaller cluster, at the expense of time. 
-However, when we train the algorithms on the entire genome, memory requirements increase to 24GB. This limited us to using 40 executors, due to memory availability.
-The memory requirements also restricts a job of that size to clusters with at least 24GB available. This demonstrates that pre-processing the data to reduce its dimensionality will
+for \kMeans and LR, with 6GB required for RF. Although we specify 80 executors, the same algorithm could successfully complete on a much smaller cluster, albeit at the expense of time. 
+However, when we train the algorithms on the entire genome, memory requirements increase to 24GB. This limited us to using 40 executors, due to memory availability, and would also restrict
+a job of this size to clusters with at least 24GB available. This demonstrates that pre-processing the data to reduce its dimensionality will
 be an essential step for future work, especially as datasets grow outwards.
@@
 Big Data Genomics using the ADAM framework (1 vs 0.96 and 0.976, respectively).  Furthermore, we observe a limitation within
-Java, which limits the random forest implementation of {\sc Spark}'s {\sc MLlib} to XX features for the 1096 individuals
+Java, which limits the random forest implementation of {\sc Spark}'s {\sc MLlib} to approximately 10 million features for the 1096 individuals
 of the 1000 Genomes Project.  This property would make feature selection mandatory for the 100,000 genome project when classifying their

### 2015-10-27 doc/publications/Recomb15/recomb15.tex
removed dones

--- 2015-10-26 doc/publications/Recomb15/recomb15.tex
@@
 \usepackage{placeins}
-\usepackage{hyperref}
-\usepackage{amsmath}
 %
@@
 \newcommand{\variantSpark}{{\sc VariantSpark}}
-\newcommand{\kMeans}{\textit{k}-means }
+\newcommand{\kMeans}{\textit{k}-means}
 \newcommand{\ARI}{adjusted Rand index}
@@
 \begin{abstract}
-Using our 
-We apply various machine learning algorithms to the 1092 individuals in the 1000 Genomes Project dataset.
-Although each algorithm we investigate can cluster individuals from three populations with near perfect accuracy,
-we find logistic regression to be superior. Comparing  With the entire dataset,
-
+This paragraph shall summarize the contents of the paper
+in short terms.
 \end{abstract}
@@
 
-\marginpar{I don't understand this. Can't you just ask a patient about their ancestry?}
 Determining a person's ethnicity has become important for medical applications, such as HLA allele genotyping from SNP
@@
 (EUR), however when adding the fourth super-population group, American (AMR), we observe that only a subset of the AMR
-individual form an independent cluster while the majority falls into the EUR cluster.  We quantified this by calculating
-the \ARI\ ( ARI), which compares the predicted and annotated labels. \kMeans achieved an ARI of 0.84, with -1 being
-random and 1 a perfect clustering.  We argue that the suboptimal clustering is a result of using an unsupervised method
-rather than an insufficient number of genomic markers~\cite{Patterson2006} and hence hypothesise that a supervised
-machine learning method may be capable of separating these subgroups of low genetic divergence.  Indeed, bigdata
-genomics recently investigated the supervised method, deep learning, for clustering the 1000 genomes data by
-ethnicity~\cite{Ferguson}.  \marginpar{``bigdata genomics'' seems to be a domain name \url{http://bdgenomics.org/} --
-  how should we cite this}
+individual form an independent cluster while the majority falls into the EUR cluster. 
+We quantified this by calculating the \ARI\ (ARI), which compares the predicted and annotated labels. \kMeans achieved an ARI=0.84,
+with -1 being random and 1 a perfect clustering.  We argue that the suboptimal clustering is a result of using an
+unsupervised method rather than an insufficient number of genomic markers~\cite{Patterson2006} and hence hypothesise
+that a supervised machine learning method may be capable of separating these subgroups of low genetic divergence.
+Indeed, bigdata genomics recently investigated the supervised method, deep learning, for clustering the 1000 genomes
+data by ethnicity~\cite{Ferguson}.
 
@@
 One of the simplest supervised machine learning methods is logistic regression (LR), which creates a decision boundary
-by fitting a linear combination of the feature variables.  In this appliction it is not fesible to model interactions,
-so the model is retricted to main effects.
+by fitting a linear combination of the feature variables.  It is therefore able to adjust the weight of the individual
+genotypes to increase the accuracy of the multi-class prediction.
 
@@
 approaches, namely Random Forest (RF), as for example Locke {\it et al.}~\cite{Locke2015} showed in a recent meta-study
-around genetic susceptibility to obesity that only five of the 97 genomic loci have an independent association signal
+around genetic susceptibility of obesity that only five of the 97 genomic loci have an independent association signal
 the rest seem to jointly predict the complex metabolic phenotype.  We therefore investigate whether ethnicity is a
@@
 section.  In the second section we compare LR and \kMeans{} against the reported accuracy of the supervised multivariate
-deep learning (DL) method.  In section three we record the accuracy of LR and \kMeans{} on the full dataset set using all populations.
+DL method.  In section three we record the accuracy of LR and \kMeans{} on the full dataset set using all populations.
 In the last section we discuss the memory consumptions and runtime of LR compared to \kMeans{}.
-
-%\clearpage
-%\hrule
-%I think the use of the term ``multivariate'' is a bit loose. We have only one $y$ variable -- obesity. By analogy with linear regression:
-%\begin{itemize}
-%\item simple linear regression $y=\beta_0 + \beta_1x_1$
-%\item multi-variable regression $y=\beta_0 + \beta_1x_1 + \beta_2 x_2$
-%\item multivariate regression $\underset{n\times m}{Y}= \underset{(n\times p+1)}{X} \underset{(p+1 \times m)}{B}$
-%\end{itemize}
-%So RF, DL and LR are all multi-variable (in this application).
-
-%A multivarite technique we could use is PCA, $T = X W$ where $W$ is a $p \times p$ matrix whose columns are the eigenvectors of $X^tX$. 
-%\hrule
 
@@
 random forest implementation to fail. This error is due to the number of features present in each sample,
-rather than the number of samples, and a potential solution is to either limit the features through feature selection, or to simply use a subset of the dataset.
-We therefore use only chromosome 22 (494,328 variants) for the comparison in this section.  We train a RF
-and LR on this subset of genomic profiles from the 1000 Genomes Project dataset using the annotated labels of all
-super-populations.  We perform 10-fold cross-validation and calculate the Adjusted Rand Index (ARI) for use as a metric for model quality.
-The ARI allows us to compare the classification results with \kMeans{} clustering results, as it takes the partitioning of the dataset into account, rather than just labels.
-report the mean error over all training and testing folds
-along with the standard error of the mean (SEM)
-We also calculate the error
+with a potential solution being to limit the features through feature selection, or simply using a subset of the dataset.
+We therefore use only chromosome 22 (494,328 alleles) for the comparison in this section.  We train a RF
+and LR on this subset of genomic profiles from the 1000 Genomes data using the annotated labels of all
+super-populations.  We perform 10-fold cross-validation and report the mean error over all training and testing folds
+along with the standard error of the mean (SEM). We also use the Adjusted Rand Index (ARI) as a metric for model quality
+as it allows us to compare the classification results with \kMeans{} clustering models. We also calculate the error
 for the two classification methods, which is simply the number of misclassified individuals divided by the number of samples.
@@
 
-%These results also demonstrate how much the quality of the models can vary between runs (or folds) for supervised and unsupervised learning.
-%For both of the supervised methods, RF and LR, the ARI was fairly consistent when testing on each holdout set, with a standard deviation of
-%\( 0.064 \) and 0.048, respectively. For the \kMeans{} models however, the standard deviation was 0.211.
-%One pitfall of \kMeans{} clustering is that depending on the starting points, the resulting model may represent a local optimum of the dataset~\cite{Steinley2003}.
-%{\sc MLlib} uses {\sc \kMeans\textbar\textbar}~\cite{Bahmani2012} (a parallel implementation of \kMeans{}++) for the cluster initialisation.
-%Although this produces close to the optimal solution for the initial cluster centres, due to the heuristical nature of the algorithm,
-%these initial cluster centres will be different for each model. Due to the amount of variance in the 1000 Genomes dataset, it is possible that multiple local
-%optima are close to the global optimum, and the algorithm is choosing one that does not represent population structure.
+These results also demonstrate how much the quality of the models can vary between runs (or folds) for supervised and unsupervised learning.
+For both of the supervised methods, RF and LR, the ARI was fairly consistent when testing on each holdout set, with a standard deviation of
+\( 0.064 \) and 0.048, respectively. For the \kMeans{} models however, the standard deviation was 0.211.
+One pitfall of \kMeans{} clustering is that depending on the starting points, the resulting model may represent a local optimum of the dataset~\cite{Steinley2003}.
+{\sc MLlib} uses {\sc \kMeans\textbar\textbar}~\cite{Bahmani2012} (a parallel implementation of \kMeans{}++) for the cluster initialisation.
+Although this produces close to the optimal solution for the initial cluster centres, due to the heuristical nature of the algorithm,
+these initial cluster centres will be different for each model. Due to the amount of variance in the 1000 Genomes dataset, it is possible that multiple local
+optima are close to the global optimum, and the algorithm is choosing one that does not represent population structure.
 
+\FloatBarrier
+\clearpage
 \subsubsection{Logistic regression and \kMeans{} capture the population better than deep learning in a three population test}
@@
 
-The mean ARI for our \kMeans{} clustering models on this dataset is \(0.960 \pm 0.004 \). 
-This value is slightly less than the one reported for the deep-learning approach, which achieved an ARI of 0.976 (where one ASW and one GBR individual are misclassified as CHB). 
+The mean ARI for our \kMeans clustering models on this dataset is \(0.960 \pm 0.004 \). 
+%DENIS: lower compared to what ?
+Although slightly lower, this value is similar to the one reported for the deep-learning approach, which achieved an ARI of 0.976 (where one ASW and one GBR individual is misclassified as CHB). 
 Our supervised logistic regression approach, however, improves upon this result and achieves an ARI of 1 across the ten folds.
 
-Although we see a slight improvement with logistic regression in regards to other algorithms, their results are still very similar.
-However, there is not much room in the first place for logistic regression to demonstrate any potential increases in model accuracy.
-We therefore proceed to apply our method to the entire 1000 Genomes Project phase 1 dataset, which we previously clustered with an ARI of 0.84.
+%DENIS: I don't understand this 
+To reiterate, these results are from the three populations, ASW, CHB and GBR, that belong to the super-populations, AFR,
+EAS and EUR, respectively.  As we mention in the introduction, we observed a perfect clustering result when we previously clustered
+ these three super-populations (rather than one population from each), with an ARI of 1.0. However, when
+we introduce the fourth super-population into the mix, AMR, the ARI drops to 0.84.
+
+Given the perfect result when building
+the logistic regression model on this small subset of data, and the similarity with the other models,
+we expand our dataset to include every population from the 1000 Genomes Project phase1 dataset.
 
@@
 Using 10-fold cross-validation to assess the logistic regression classification results, we observe an average ARI of \(0.936 \pm{} 0.014\).
+%DENIS: above you argue that kMeans has local optima and results might hence be too optimistic but here you only train one model. So either remove the above section or add some repeats to this one
 This is an improvement on our previous result from \kMeans{} clustering the same dataset in our previous work, where we observed an ARI of \(0.83\).
@@
 This is due to individuals from the AFR and EUR super-populations being misclassified as AMR.
-Consequently, this leads to these two super-populations having a lower recall of \(0.956 \pm 0.01 \) and \(0.950 \pm 0.02 \), respectively.
+Consequently, this leds to these two super-populations having a lower recall of \(0.956 \pm 0.01 \) and \(0.950 \pm 0.02 \), respectively.
 Although these misclassifications as AMR are relatively frequent, we did not observe any AFR individuals being classified as EUR, or {\it vice versa}, in any of the ten folds.
@@
 
-Time and resources vary greatly between the three algorithms, with the most resource intensive being random forests (see table~\ref{resources}).
-This is likely due to RF being an ensemble method. For example, where logistic regression and \kMeans{} clustering build a single model,
-RF is building an forest of decision trees.
-When training on chromosome 22 (494,328 variants), resource requirements are kept to a minimum, with no more than 3GB memory required per executor (node)
-(6GB for RF). Although we specify 80 executors, the same algorithm could successfully complete on a much smaller cluster, at the expense of time. 
-However, when we train the algorithms on the entire genome, memory requirements increase to 24GB. This limited us to using 40 executors, due to memory availability.
-The memory requirements also restricts a job of that size to clusters with at least 24GB available. This demonstrates that pre-processing the data to reduce its dimensionality will
-be an essential step for future work, especially as datasets grow outwards.
+Time and resources vary greatly between the three algorithms, with the most resource intensive being random forests.
+This is likely due to RF being an ensemble method. Where logistic regression and \kMeans{} clustering build a single model,
+RF is building a forest of trees. So not only is training longer, but then each sample in the test fold must be classified by each and every tree.
+
+LR is
+
+\marginpar{TODO [write up the observations and argue why one is less resource hungry than the other one]}
 
@@
 \hline\noalign{\smallskip}
-method  & chromosomes & executors & runtime & memory (GB) \\
-\kMeans & 22 & 80 & 2mins 3secs & 3 \\
-LR  & 22 & 80 & 5mins 1sec & 3 \\
-RF & 22 & 80 & 17mins 20secs & 6 \\
-\kMeans & 1-22 & 40 & 14hrs 44mins & 24 \\
-LR  & 1-22 & 40 & 30hrs 12mins & 24 \\
+method  & nodes & runtime & memory \\
+RF  & X (Y) & X (Y) & X (Y) \\
+\kMeans & X (Y) & X (Y) & X (Y) \\
 \noalign{\smallskip}
 \hline
-\label{resources}
 \end{tabular}
@@
 Big Data Genomics using the ADAM framework (1 vs 0.96 and 0.976, respectively).  Furthermore, we observe a limitation within
-Java, which limits the random forest implementation of {\sc Spark}'s {\sc MLlib} to XX features for the 1096 individuals
-of the 1000 Genomes Project.  This property would make feature selection mandatory for the 100,000 genome project when classifying their
+Java, which limits the random forest implementation of {\sc Spark}'s {\sc MLlib} to XX features for the 1069 individuals
+of the 1000 genome project.  This property would make feature selection mandatory for the 100,000 genome project when classifying their
 70,000 individuals, when using the same random forest implementation.  We conclude that while logistic regression
@@
 
-\bibliographystyle{my_plain}
+\bibliographystyle{plain}
+\marginpar{TODO you need to find a way of shortening the Locket paper author list}
 \bibliography{recomb}

### 2015-10-27 doc/publications/Recomb15/recomb15.tex
neaten

@@
 section.  In the second section we compare LR and \kMeans{} against the reported accuracy of the supervised multivariate
-DL method.  In section three we record the accuracy of LR and \kMeans{} on the full dataset set using all populations.
+deep learning (DL) method.  In section three we record the accuracy of LR and \kMeans{} on the full dataset set using all populations.
 In the last section we discuss the memory consumptions and runtime of LR compared to \kMeans{}.
@@
 
-These results also demonstrate how much the quality of the models can vary between runs (or folds) for supervised and unsupervised learning.
-For both of the supervised methods, RF and LR, the ARI was fairly consistent when testing on each holdout set, with a standard deviation of
-\( 0.064 \) and 0.048, respectively. For the \kMeans{} models however, the standard deviation was 0.211.
-One pitfall of \kMeans{} clustering is that depending on the starting points, the resulting model may represent a local optimum of the dataset~\cite{Steinley2003}.
-{\sc MLlib} uses {\sc \kMeans\textbar\textbar}~\cite{Bahmani2012} (a parallel implementation of \kMeans{}++) for the cluster initialisation.
-Although this produces close to the optimal solution for the initial cluster centres, due to the heuristical nature of the algorithm,
-these initial cluster centres will be different for each model. Due to the amount of variance in the 1000 Genomes dataset, it is possible that multiple local
-optima are close to the global optimum, and the algorithm is choosing one that does not represent population structure.
+%These results also demonstrate how much the quality of the models can vary between runs (or folds) for supervised and unsupervised learning.
+%For both of the supervised methods, RF and LR, the ARI was fairly consistent when testing on each holdout set, with a standard deviation of
+%\( 0.064 \) and 0.048, respectively. For the \kMeans{} models however, the standard deviation was 0.211.
+%One pitfall of \kMeans{} clustering is that depending on the starting points, the resulting model may represent a local optimum of the dataset~\cite{Steinley2003}.
+%{\sc MLlib} uses {\sc \kMeans\textbar\textbar}~\cite{Bahmani2012} (a parallel implementation of \kMeans{}++) for the cluster initialisation.
+%Although this produces close to the optimal solution for the initial cluster centres, due to the heuristical nature of the algorithm,
+%these initial cluster centres will be different for each model. Due to the amount of variance in the 1000 Genomes dataset, it is possible that multiple local
+%optima are close to the global optimum, and the algorithm is choosing one that does not represent population structure.
 
-\FloatBarrier
-\clearpage
 \subsubsection{Logistic regression and \kMeans{} capture the population better than deep learning in a three population test}
@@
 
-\bibliographystyle{plain}
-\marginpar{TODO you need to find a way of shortening the Locket paper author list}
+\bibliographystyle{my_plain}
 \bibliography{recomb}

### 2015-11-20 doc/publications/journal2014/paper.tex
R and others

@@
 \subsection*{Computational resources}
-We completed the Chromosome 22 comparisons on a virtual machine (VM) hosted on Microsoft Azure. This VM is an A7 Linux instance with 8 cores, 56GB memory running Ubuntu. 
-For the whole-genome clustering, we used our in-house Hadoop cluster with Hadoop 2.5.0, managed by cloudera's CDH 5.
-We use Spark 1.3.1.
-This 13 node cluster has a total of 416 cores and 1.22TB memory.
+We completed the Chromosome 22 comparisons on a virtual machine (VM) hosted on Microsoft Azure. This VM is an A7 Linux instance with 8 cores, 56GB
+memory running Ubuntu. 
+For the whole-genome clustering, we used our in-house Hadoop cluster with Hadoop 2.5.0, managed by cloudera's CDH 5. We use Spark 1.3.1. This 13 node
+cluster has a total of 416 cores and 1.22TB memory.
 
 \subsection*{Datasets}
-We used phase 1 variants from the 1000 Genomes Project. On the 22 autosomes, this dataset contains variants for 1092 individuals, spread over a total 38,219,238 alleles.
-We also used a subset of this dataset, the variants from chromosome 22 containing 494,328 alleles.
+We use the 22 autosomes from the 1000 Genomes Project phase 1 dataset. This dataset contains variants from 1092 individuals, across 38,219,238 alleles.
+For the smaller dataset, we use the variants from chromosome 22, with 494,328 alleles.
 The individuals from these datasets are distributed across four super populations, African (AFR), Mixed American (AMR), East Asian (EAS) and European (EUR).
+We also cluster the Phase 3 dataset, which, contains variants from 2535 individuals, across 81,271,745 variants. As well as the above for super populations, this
+dataset also includes South Asian (SAS).
 
 \subsection*{VariantSpark Implementation}
-To cluster the individuals in a VCF file, we initially need to pre-process the variants to feature vectors.
-In \variantSpark{} we read in the VCF files as text files to a Resilient Distributed Dataset (RDD).
-RDDs allow us to process the files, line-by-line, in parallel.
-We parse each line as tab-separated values, and with Spark's `zip' function, `zip' each value with it's heading (Sample name) from the VCF file, resulting in key-value pairs (KVPs) for each line see Figure 2 (Spark step 1).
-We then `zip' each pair with a unique sequential index, which later-on will serve as identifier for each allele (and keep the genome loci order), resulting in a KVP array (step 2).
-We use `flatMap' to transpose the data structure to facilitate the grouping of variants by individuals.
-For the \kMeans{} clustering, we convert the VCF genotypes (i.e.
-`0\textbar{}1') to doubles using the Hamming distance to the reference, where a 1 represents a heterozygous variant, 2 a homozygous variant, and 0, no variant.
-To remove uninformative KVPs, we filter out any alleles that have no variant. 
-We perform this `map' operation inside the previous `flatMap' (step 3).
-We can now group these KVPs by their key (the individual ID) and save them to sparse vectors (step 4).
+To cluster individuals from a VCF file, we initially need to pre-process the variants to feature vectors.
+In \variantSpark{} we read in VCF files as text files, to a Resilient Distributed Dataset (RDD). RDDs allow us to process the files, line-by-line, in parallel.
+
+We parse each line as tab-separated values, and store these the values from each line in an array.
+
+For each array, we use Spark's `zip' function to `zip' each value with it's heading (Sample name) from the VCF file. Now each array element is a
+key-value pair (KVP) of the previous value and its heading. For the KVPs that contain an allele as the value, the key (derived from the heading) is
+the Individual ID (see Figure 2, Spark step 1).
+
+For each array, we remove any KVPs that represent data other than alleles (such as the genomic location). With the remaining KVPs, which all
+represent alleles, we apply a function to convert these from the strings found in VCF files (i.e. `0\textbar{}1'), to doubles, where the double is the
+Hamming distance to the reference. I.e. a 1 represents a heterozygous variant, 2, a homozygous variant, and 0, no variant. Because we will
+eventually be converting this data to sparse vectors, we remove any KVPs with a value of 0 (i.e. no variant).
+
+At this stage, we can optionally filter variants that don't match a specific criteria. For the Phase 3 dataset, we exclude rare variants that are present in
+only one individual. Because each array now only contains variants (rather than every allele), we can simply drop arrays with a length of one, as the length
+of each array is equal to the number of individuals who have a variant at that allele.
+
+Now that the dataset only contains alleles we are interested in, we `zip' each array with a unique sequential index (allele ID). This index will serve
+as identifier for each allele (and keep the genome loci order) (step 2).
+
+Although we now have the data we require, the variants are stored in arrays of alleles, whereas we require the data to be transposed into arrays of individuals.
+To facilitate the grouping of variants by individuals, we use `flatMap' to flatten out the arrays into a collection of KVPs, where the key is the individual ID, and the
+value is a tuple of the variant ID and the variant.
+
+We can now group these KVPs by their key (individual ID) and save them to sparse vectors (step 4).
 
@@
 \subsection*{R Implementation}
-For our R implementation, we read in a VCF file as a table and convert it to a matrix, filtering columns where all variant as conform to the reference.
-As with our \variantSpark{} pre-processing, we convert the strings that represent each allele to a numeric value.
-We then transpose the matrix, resulting in a data-structure where each row represents an individual.
-We convert the matrix to a `big.matrix', for the \kMeans{} algorithm from the `biganalytics' package (\url{https://cran.r-project.org/web/packages/biganalytics/index.html}).
+For our R implementation, we read in a VCF file using {\sc read.table}, and then convert the resulting table to a matrix ({\sc vcfMatrix} with {\sc as.data.frame.matrix}.
+We remove the first 9 columns which leaves us with just the variants. As with our \variantSpark{} pre-processing, we convert the strings that represent each allele to a numeric value.
+We then transpose the matrix with {\sc t(vcfMatrix)}, which results in a data-structure where each row represents an individual. We convert the matrix to a {\sc big.matrix} object,
+as required by the \kMeans{} algorithm from the `biganalytics' package (\url{https://cran.r-project.org/web/packages/biganalytics/index.html}), and then call {\sc bigkmeans} with
+the {\sc big.matrix} object and the required number of clusters as arguments.
 

### 2015-11-20 doc/publications/journal2014/paper.tex
python

@@
 \subsection*{Python Implementation}
-Our Python implementation reads in lines from a VCF file as tab-separated values and stores the lines in a pandas DataFrame (\url{http://pandas.pydata.org}).
+Our Python implementation reads in lines from a VCF file as tab-separated values using {\sc DataFrame.from\textunderscore{}csv}, and stores the data in a pandas DataFrame (\url{http://pandas.pydata.org}).
 The column headings are the individual IDs and the row headings are the allele locations.
-We filter out reference-conform data and convert the remaining allele strings to numeric values. 
-We cluster this DataFrame using sci-kit learn (\url{http://scikit-learn.org/stable/}), upon converting it to a matrix with {\sc .as\textunderscore{}matrix()}.
+We the first 9 columns and convert the remaining allele strings to numeric values.
+We convert the DataFrame to a matrix with {\sc .as\textunderscore{}matrix()} and cluster the matrix using sci-kit learn (\url{http://scikit-learn.org/stable/}).
 

### 2015-11-25 variantspark/src/main/scala/au/csiro/obr17q/variantspark/FlatVariantToSparse.scala
Added flat to sparse

@@ -1,0 +1,42 @@
+package au.csiro.obr17q.variantspark
+
+import au.csiro.obr17q.variantspark.CommonFunctions._
+import org.apache.spark.mllib.clustering.KMeans
+import org.apache.spark.mllib.linalg.Vectors
+import org.apache.spark
+import scala.io.Source
+import org.apache.spark.sql.functions.max
+import org.apache.spark.rdd.RDD
+import au.csiro.obr17q.variantspark.model.SubjectSparseVariant
+
+
+object FlatVariantToSparse extends SparkApp {
+  conf.setAppName("VCF cluster")
+  def main(args:Array[String]) {
+
+    if (args.length < 1) {
+        println("Usage: CsvClusterer <input-path>")
+    }
+
+    val inputFiles = args(0)
+    val output = args(1) 
+    val chunks = args(2).toInt
+    val sqlContext = new org.apache.spark.sql.SQLContext(sc)
+    import sqlContext.implicits._
+    val flatVariandDF = sqlContext.parquetFile(inputFiles)    
+    println(flatVariandDF.schema)
+    
+    
+    val maxVariantIndex = flatVariandDF.agg(max(flatVariandDF("variantIndex")))
+      .collectAsList().get(0).getInt(0) //flatVariandDF.rdd.map(r => r.getInt(1)).max();
+    
+    println(maxVariantIndex)
+    val parirRDD:RDD[(String,(Int,Double))] = flatVariandDF.rdd.map(r => (r.getString(0), (r.getInt(1), r.getDouble(2))))
+    parirRDD
+      .groupByKey(chunks)
+      .mapValues(i => Vectors.sparse(maxVariantIndex + 1, i.to[Seq]).toSparse)
+      .map {case (subjectId, sparseVector) =>
+        SubjectSparseVariant(subjectId, sparseVector.size, sparseVector.indices, sparseVector.values)}
+      .toDF().saveAsParquetFile(output)  
+  } 
+}

### 2015-11-25 variantspark/src/main/scala/au/csiro/obr17q/variantspark/VcfParser.scala

@@ -68,7 +68,7 @@ class VcfParser (val VcfFileNames: String, val VariantCutoff: Int, val sc: Spark
   /**
     * Number of variants in the file
     */
-  val variantCount : Int =
+  lazy val variantCount : Int =
     VcfLineRdd.count().toInt
 
 

### 2015-11-25 variantspark/src/main/scala/au/csiro/obr17q/variantspark/model/SubjectSparseVariant.scala

@@ -1,0 +1,5 @@
+package au.csiro.obr17q.variantspark.model
+
+case class SubjectSparseVariant(val subjectId:String, val size:Int, val indices:Array[Int], val values:Array[Double]) {
+
+}

### 2015-11-25 doc/publications/journal2014/paper.tex
updates for reviewers

@@
 We run this comparison on our in-house Hadoop cluster using 40 cores.
-Pre-processing the VCF files takes approximately 12 minutes, 19 minutes, 27 minutes and 40 minutes for 20\%, 40\%, 60\% and 100\% of the genome, respectively (see Table 2).
+Pre-processing the VCF files takes approximately 12, 19, 27 and 40 minutes for 20, 40, 60 and 100 percent of the genome, respectively (see Table 2).
 In each case, the memory required does not exceed a modest 2GB per executor, even for the complete genome. 
@@
 We do not observe an increase in cluster accuracy when providing variants from all chromosomes, indicating that current practice of clustering based on a subset of the genome is sufficient to define the existing population boundaries for populations with large genetic distance~\cite{Pugach2015}.
+
+To further demonstrate the scalability of \variantSpark{}, we also cluster the larger 1000 Genomes Project phase 3 VCF files. Although \kMeans{} clustering completes successfully on this dataset,
+it takes over 27 hours (see Table 2) due to more than twice the number of individuals, and hence more variants.
+For comparison, the uncompressed size of the phase 3 files is 770GB vs the 161GB of the phase 1 dataset.
 
@@
 the {\sc big.matrix} object and the required number of clusters as arguments.
+
+%Compare with our old implementation still? Or just have the VariantAnnotation one.
+%Do we acknowledge that this is a reviewer suggestion in any way?
+We also compare this approach to one using the `VariantAnnotation' package.
+Reading the VCF file to a matrix with `readGT' is faster than reading a VCF file to a table using R's `read.table', at 116 seconds compared to 207 seconds,
+however, because `readGT' only includes genotype information in the matrix, we can simplify our function that converts this information into numeric values.
+This speeds up processing time greatly.
+More implementation details are available
+in the supplementary material.
 
@@
 100\%	& 64	& 2	& 40min 48sec	& 40	& 24	& 14hr 44min	\\
+???\%	& 64 & 2	& ??min ??sec	& 40 & 24	& 27hr 46min	\\
   \hline

### 2015-11-25 variantspark/src/main/scala/au/csiro/obr17q/variantspark/FlatVariantToSparse.scala
Changed to sparse to use fastutil

@@ -3,14 +3,25 @@ package au.csiro.obr17q.variantspark
 import au.csiro.obr17q.variantspark.CommonFunctions._
 import org.apache.spark.mllib.clustering.KMeans
 import org.apache.spark.mllib.linalg.Vectors
+import org.apache.spark.mllib.linalg.Vector
 import org.apache.spark
 import scala.io.Source
 import org.apache.spark.sql.functions.max
 import org.apache.spark.rdd.RDD
 import au.csiro.obr17q.variantspark.model.SubjectSparseVariant
+import it.unimi.dsi.fastutil.ints.Int2DoubleOpenHashMap
+import it.unimi.dsi.fastutil.ints.Int2DoubleMap
+import it.unimi.dsi.fastutil.ints.Int2DoubleRBTreeMap
+import org.apache.spark.mllib.linalg.SparseVector
 
 
 object FlatVariantToSparse extends SparkApp {
+  
+  def map2Sparse(size:Int, map: Int2DoubleMap):SparseVector = {
+     val sorted = new Int2DoubleRBTreeMap(map);
+     Vectors.sparse(size, sorted.keySet().toIntArray(), sorted.values().toDoubleArray()).toSparse
+  }
+  
   conf.setAppName("VCF cluster")
   def main(args:Array[String]) {
 
@@ -33,8 +44,11 @@ object FlatVariantToSparse extends SparkApp {
     println(maxVariantIndex)
     val parirRDD:RDD[(String,(Int,Double))] = flatVariandDF.rdd.map(r => (r.getString(0), (r.getInt(1), r.getDouble(2))))
     parirRDD
-      .groupByKey(chunks)
-      .mapValues(i => Vectors.sparse(maxVariantIndex + 1, i.to[Seq]).toSparse)
+      .aggregateByKey(new Int2DoubleOpenHashMap(),chunks)(
+          (m, t) =>  {m.put(t._1,t._2); m},
+          (m1,m2) => {m1.putAll(m2); m1}
+       )
+      .mapValues(i => map2Sparse(maxVariantIndex,i))
       .map {case (subjectId, sparseVector) =>
         SubjectSparseVariant(subjectId, sparseVector.size, sparseVector.indices, sparseVector.values)}
       .toDF().saveAsParquetFile(output)  

### 2015-11-25 variantspark/src/main/scala/au/csiro/obr17q/variantspark/SparseClustering.scala

@@ -1,0 +1,196 @@
+package au.csiro.obr17q.variantspark
+
+import au.csiro.obr17q.variantspark.CommonFunctions._
+import org.apache.spark.mllib.clustering.KMeans
+import org.apache.spark.mllib.linalg.Vectors
+
+import scala.io.Source
+
+object SparseClustering extends SparkApp {
+  conf.setAppName("VCF cluster")
+  def main(args:Array[String]) {
+
+   
+    if (args.length < 1) {
+        println("Usage: CsvClusterer <input-path>")
+    }
+
+    val inputFiles = args(0)
+    val k = args(1).toInt
+    val IncludeGroups = args(2).split('|')
+    val ExcludeGroups = args(3).split('|')
+    val VariantCutoff = args(4).toInt
+
+    
+    
+    
+    
+    //val PopFiles = Source.fromFile("data/PGPParticipantSurvey-20150831064509.csv").getLines()
+    //val Populations = sc.parallelize(new PopulationMap(PopFiles, 1, ',', 0, 16 ).returnMap(IncludeGroups, ExcludeGroups))
+
+    val PopFiles = Source.fromFile("../data/ALL.panel").getLines()
+    val Populations = sc.parallelize(new MetaDataParser(PopFiles, 1, '\t', "NA", 0, 1 ).returnMap(IncludeGroups, ExcludeGroups))
+
+    
+    val pops = Populations.map(_.toPops).collectAsMap();
+    println(pops)
+    val pops_br = sc.broadcast(pops);
+    val sqlContext = new org.apache.spark.sql.SQLContext(sc)
+    import sqlContext.implicits._
+    val sparseVariat = sqlContext.parquetFile(inputFiles)    
+    println(sparseVariat.schema)
+
+    
+    val IndividualVariants = 
+      sparseVariat.rdd
+        .filter(r => pops_br.value.contains(r.getString(0)))
+        .map(r=> (pops_br.value(r.getString(0)), r.getString(0), Vectors.sparse(r.getInt(1),
+          r.getSeq[Int](2).toArray, r.getSeq[Double](3).toArray)))
+      
+      
+      
+
+
+    /** Print populations included and number of individuals (slow) **/
+    //IndividualVariants.map(p => (p._1, 1)).join(Populations.map(_.toPops)).map(p => (p._2._2, 1)).reduceByKey(_ + _).collect().foreach(println)
+
[…]

### 2015-11-25 variantspark/src/main/scala/au/csiro/obr17q/variantspark/FlatVariantToSparse.scala
Fixed issues with sparse size

@@ -48,7 +48,7 @@ object FlatVariantToSparse extends SparkApp {
           (m, t) =>  {m.put(t._1,t._2); m},
           (m1,m2) => {m1.putAll(m2); m1}
        )
-      .mapValues(i => map2Sparse(maxVariantIndex,i))
+      .mapValues(i => map2Sparse(maxVariantIndex+1,i))
       .map {case (subjectId, sparseVector) =>
         SubjectSparseVariant(subjectId, sparseVector.size, sparseVector.indices, sparseVector.values)}
       .toDF().saveAsParquetFile(output)  

### 2015-11-25 variantspark/src/main/scala/au/csiro/obr17q/variantspark/ValidateSparse.scala

@@ -1,0 +1,59 @@
+package au.csiro.obr17q.variantspark
+
+import au.csiro.obr17q.variantspark.CommonFunctions._
+import org.apache.spark.mllib.clustering.KMeans
+import org.apache.spark.mllib.linalg.Vectors
+
+import scala.io.Source
+
+object ValidateSparse extends SparkApp {
+  conf.setAppName("VCF cluster")
+  
+  
+  def assert(test:Boolean, msg: String) {
+    if (!test) {
+      throw new RuntimeException("Failed "+ msg)
+    }   
+  }
+  def main(args:Array[String]) {
+
+   
+    if (args.length < 1) {
+        println("Usage: CsvClusterer <input-path>")
+    }
+
+    val inputFiles = args(0)
+
+    
+    
+    
+    
+    //val PopFiles = Source.fromFile("data/PGPParticipantSurvey-20150831064509.csv").getLines()
+    //val Populations = sc.parallelize(new PopulationMap(PopFiles, 1, ',', 0, 16 ).returnMap(IncludeGroups, ExcludeGroups))
+
+    val sqlContext = new org.apache.spark.sql.SQLContext(sc)
+    import sqlContext.implicits._
+    val sparseVariat = sqlContext.parquetFile(inputFiles)    
+    println(sparseVariat.schema)
+
+    
+    val IndividualVariants = 
+      sparseVariat.rdd
+        .map(r=> (r.getString(0), r.getInt(1),
+          r.getSeq[Int](2).toArray, r.getSeq[Double](3).toArray))
+        .foreach{ case (subjectId, size, indices, values) => 
+        
+
+          
+          println(s"${subjectId}: size: ${size}, indices: ${indices.length},values: ${values.length}" + 
+              s", ${indices.toSeq.min}, ${indices.toSeq.max},${values.toSeq.min}, ${values.toSeq.max}")
+          assert(size > 0, "Size > 0 ")
+          assert(indices.length == values.length, "indices.length == values.length")
+          assert(indices.toSeq.max < size, "indices.toSeq.max < size")
+          assert(indices.toSeq.min >=0 , "indices.toSeq.min >=0")
+          assert(values.toSeq.min >=0 , "values.toSeq.min >=0")
+    }
+        
+    
+  } 
+}

### 2015-11-26 variantspark/src/main/scala/au/csiro/obr17q/variantspark/FlatVariantToSparseByPosition.scala
Changed locus sparse to group by

@@ -49,11 +49,8 @@ object FlatVariantToSparseByPosition extends SparkApp {
     
     val parirRDD:RDD[(Int,(Int,Double))] = flatVariandDF.rdd.map(r => (r.getInt(1), (br_subjectToIndexMap.value(r.getString(0)), r.getDouble(2))))
     parirRDD
-      .aggregateByKey(new Int2DoubleOpenHashMap(),chunks)(
-          (m, t) =>  {m.put(t._1,t._2); m},
-          (m1,m2) => {m1.putAll(m2); m1}
-       )
-      .mapValues(i => map2Sparse(subjectsSize,i))
+      .groupByKey(chunks)
+      .mapValues(i => Vectors.sparse(subjectsSize,i.to[Seq]).toSparse)
       .map {case (locusId, sparseVector) =>
         LocusSparseVariant(locusId, sparseVector.size, sparseVector.indices, sparseVector.values)}
       .toDF().saveAsParquetFile(output)  

### 2015-11-26 doc/publications/journal2014/paper.tex
updates to letter and paper

--- 2015-11-20 doc/publications/journal2014/paper.tex
@@
 We do not observe an increase in cluster accuracy when providing variants from all chromosomes, indicating that current practice of clustering based on a subset of the genome is sufficient to define the existing population boundaries for populations with large genetic distance~\cite{Pugach2015}.
+
+%To further demonstrate the scalability of \variantSpark{}, we also cluster the larger 1000 Genomes Project phase 3 VCF files. Although \kMeans{} clustering completes successfully on this dataset,
+%it takes over 27 hours (see Table 2) due to more than twice the number of individuals, and hence more variants.
+For comparison, the uncompressed size of the phase 3 files is 770GB vs the 161GB of the phase 1 dataset.
+%TODO add value
+To further demonstrate the scalability of \variantSpark{}, we also cluster the 1000 Genomes Project phase 3 data, which contains 3000 individuals from 5 superpopulations and as a result has over 80 Million variants.
+The uncompressed size of the phase 3 files is 770GB compared to the 161GB of the phase 1 dataset.
+\variantSpark{} successfully completes the clustering in just 27 hours (see Table 2) with an ARI of XX. % if XX is smaller than 0.84 we need to explain more.
 
@@
 
-\variantSpark{} performs clustering on VCF files with over 1000 individuals and 40 million variants in under 15 hours using minimal memory (24GB). 
+%\variantSpark{} performs clustering on VCF files with over 1000 individuals and 40 million variants in under 15 hours using minimal memory (24GB). 
+\variantSpark{} performs clustering on VCF files with over 3000 individuals and 80 million variants in 27 hours using minimal memory (24GB). 
 \variantSpark{} supports random genome-wide sampling of variants allowing faster clustering for well characterised cohorts where 20\% of the genome is sufficient. 
@@
 These benefits of speed, resource consumption and scalability allows \variantSpark{} to be the interface for applying other machine learning tools from MlLib to genomic data. 
+%TODO added sentence
+Utilising MLlib as well as Spark.ML will enable supervised machine learning applications to e.g.
+identify variants that jointly interact with phenotypes as well as include electronic health record in addition to the genomic feature vector to e.g.
+capture medical history as well as predispositions for diagnosis and treatment decisions.
 
@@
 The individuals from these datasets are distributed across four super populations, African (AFR), Mixed American (AMR), East Asian (EAS) and European (EUR).
-We also cluster the Phase 3 dataset, which, contains variants from 2535 individuals, across 81,271,745 variants. As well as the above for super populations, this
+We also cluster the Phase 3 dataset, which contains variants from 2535 individuals, across 81,271,745 variants. As well as the above for super populations, this
 dataset also includes South Asian (SAS).
@@
 For our R implementation, we read in a VCF file using {\sc read.table}, and then convert the resulting table to a matrix ({\sc vcfMatrix} with {\sc as.data.frame.matrix}.
+%TODO VariantAnnotation reads in the file quicker but causes issue down the track so overall read.table is faster, right ? - can you add this explanation to the supplementary material.
+In the supplementary material we benchmark this approach against reading the VCF file in using the `VariantAnnotation' package.
 We remove the first 9 columns which leaves us with just the variants. As with our \variantSpark{} pre-processing, we convert the strings that represent each allele to a numeric value.
@@
 \includegraphics[type=pdf,ext=.pdf,read=.pdf, scale=0.25]{signature}
-  \caption{\csentence{Schematic of the Hadoop and Spark implementation.} The image shows the conceptual differences in the steps necessary to convert variants in VCF to a data structure readable by Mahout and MLlib, respectively.}
+\caption{\csentence{Schematic overview of VariantSpark}.
+The image shows the flow from the input VCF file to the machine learning library and onto the visualization.
+It highlights the differences between the Hadoop and Spark implementations for converting data in VCF format to a data structure readable by Mahout and MLlib, respectively.}
+%  \caption{\csentence{Schematic of the Hadoop and Spark implementation.} The image shows the conceptual differences in the steps necessary to convert variants in VCF to a data structure readable by Mahout and MLlib, respectively.}
       \end{figure}
@@
 \includegraphics[type=pdf,ext=.pdf,read=.pdf, scale=0.40]{chr22-k4-gephi-legend}
-  \caption{\csentence{Visualisation of \variantSpark{} predicted clusters.} Individuals from the 1000 Genomes project are clustered effectively by super-populations AFR, AMR and EAS; a fourth cluster is predominantly EUR + AMR.}
+  \caption{\csentence{Visualisation of \variantSpark{} predicted clusters.} The figure shows the four clusters predicted for the 1000 Genomes data.
+Individuals from the super-populations AFR, AMR and EAS are accurately grouped into distinct clusters.
+The fourth cluster contains predominantly EUR + AMR individuals potentially accurately reflecting migrational backgrounds.}
+%  \caption{\csentence{Visualisation of \variantSpark{} predicted clusters.} Individuals from the 1000 Genomes project are clustered effectively by super-populations AFR, AMR and EAS; a fourth cluster is predominantly EUR + AMR.}
       \end{figure}

### 2015-11-26 variantspark/src/main/scala/au/csiro/obr17q/variantspark/algo/TestWideKMeans.scala
Fixex spars clusring

@@ -2,6 +2,8 @@ package au.csiro.obr17q.variantspark.algo
 
 import au.csiro.obr17q.variantspark.SparkApp
 import org.apache.spark.mllib.linalg.Vectors
+import org.apache.spark.mllib.linalg.Vector
+import org.apache.spark.rdd.RDD
 
 object TestWideKMeans extends SparkApp {
   conf.setAppName("VCF cluster")
@@ -9,6 +11,7 @@ object TestWideKMeans extends SparkApp {
   
   def main(args:Array[String]) {
     println("Testign WideKMeans")
+    
     
     val dims = 1000
     val centersNo = 5
@@ -19,7 +22,7 @@ object TestWideKMeans extends SparkApp {
     val clusterAssignment = Range(0,samples).map(i => Math.floor(Math.random()*centersNo).toInt).toList
     println(clusterAssignment)
     
-    val data = centers.map(v =>
+    val data:RDD[Vector] = centers.map(v =>
       Vectors.dense(clusterAssignment.map(c => v(c) + (Math.random() * clusterVariance - clusterVariance/2)).toArray)
     )
     

### 2015-11-26 variantspark/src/main/scala/au/csiro/obr17q/variantspark/algo/WideKMeans.scala

@@ -5,6 +5,11 @@ import org.apache.spark.mllib.linalg.Vector
 import org.apache.spark.mllib.linalg.Vectors
 import java.util.Arrays
 import org.apache.spark.SparkContext
+import org.apache.spark.mllib.linalg.SparseVector
+
+object WideKMeans {
+  def sqr(d:Double) = d*d
+}
 
 class WideKMeans(kk:Int, kiter:Int) {
   
@@ -75,7 +80,8 @@ class WideKMeans(kk:Int, kiter:Int) {
      val clusterAssignment = data.zip(clusterCentres)
         .aggregate(Array.fill(dims)(Array.fill(k)(0.0)))(
             (dists, vac) => {
-              for(i <- Range(0,dims); j<- Range(0,k)) {dists(i)(j) += (vac._1(i) - vac._2(j))*(vac._1(i) - vac._2(j)) }
+              val dv = vac._1.toArray
+              for(i <- Range(0,dims); j<- Range(0,k)) {dists(i)(j) += WideKMeans.sqr(dv(i) - vac._2(j)) }                
               dists
             } ,  // add sqr(dists - vectorsAndCenters) for all (i, j)
             (dist1, dist2) => { 

### 2015-11-26 doc/publications/journal2014/paper.tex
updated paper and letter

@@
 For comparison, the uncompressed size of the phase 3 files is 770GB vs the 161GB of the phase 1 dataset.
-%TODO add value
-To further demonstrate the scalability of \variantSpark{}, we also cluster the 1000 Genomes Project phase 3 data, which contains 3000 individuals from 5 superpopulations and as a result has over 80 Million variants.
+To further demonstrate the scalability of \variantSpark{}, we also cluster the 1000 Genomes Project phase 3 data, which contains 3000 individuals from 5 super-populations and as a result has over 80 Million variants.
 The uncompressed size of the phase 3 files is 770GB compared to the 161GB of the phase 1 dataset.
-\variantSpark{} successfully completes the clustering in just 27 hours (see Table 2) with an ARI of XX. % if XX is smaller than 0.84 we need to explain more.
+\variantSpark{} successfully completes the clustering in just 27 hours (see Table 2) with an ARI of 0.82.
 
@@
 These benefits of speed, resource consumption and scalability allows \variantSpark{} to be the interface for applying other machine learning tools from MlLib to genomic data. 
-%TODO added sentence
 Utilising MLlib as well as Spark.ML will enable supervised machine learning applications to e.g.
@@
 %TODO VariantAnnotation reads in the file quicker but causes issue down the track so overall read.table is faster, right ? - can you add this explanation to the supplementary material.
-In the supplementary material we benchmark this approach against reading the VCF file in using the `VariantAnnotation' package.
+In the supplementary material we benchmark this faster approach against reading the VCF file in using the `VariantAnnotation' package.
 We remove the first 9 columns which leaves us with just the variants. As with our \variantSpark{} pre-processing, we convert the strings that represent each allele to a numeric value.

### 2015-11-26 variantspark/src/main/scala/au/csiro/obr17q/variantspark/algo/TestWideKMeans.scala
Fixed sparse imple

@@ -23,7 +23,7 @@ object TestWideKMeans extends SparkApp {
     println(clusterAssignment)
     
     val data:RDD[Vector] = centers.map(v =>
-      Vectors.dense(clusterAssignment.map(c => v(c) + (Math.random() * clusterVariance - clusterVariance/2)).toArray)
+      Vectors.dense(clusterAssignment.map(c => v(c) + (Math.random() * clusterVariance - clusterVariance/2)).toArray).toSparse
     )
     
     val test = data.cache().count()

### 2015-11-26 variantspark/src/main/scala/au/csiro/obr17q/variantspark/algo/WideKMeans.scala

@@ -41,7 +41,8 @@ class WideKMeans(kk:Int, kiter:Int) {
       val clusterAssignment = data.zip(clusterCentres)
         .aggregate(Array.fill(dims)(Array.fill(k)(0.0)))(
             (dists, vac) => {
-              for(i <- Range(0,dims); j<- Range(0,k)) {dists(i)(j) += (vac._1(i) - vac._2(j))*(vac._1(i) - vac._2(j)) }
+              val dv = vac._1.toArray
+              for(i <- Range(0,dims); j<- Range(0,k)) {dists(i)(j) += WideKMeans.sqr(dv(i) - vac._2(j)) }
               dists
             } ,  // add sqr(dists - vectorsAndCenters) for all (i, j)
             (dist1, dist2) => { 

### 2015-11-26 variantspark/src/main/scala/au/csiro/obr17q/variantspark/FlatVariantToSparseByPosition.scala
Added writting index to sparse by locus

@@ -15,6 +15,10 @@ import it.unimi.dsi.fastutil.ints.Int2DoubleRBTreeMap
 import org.apache.spark.mllib.linalg.SparseVector
 import scala.collection.mutable.HashMap
 import au.csiro.obr17q.variantspark.model.LocusSparseVariant
+import au.csiro.pbdava.sparkle.LoanUtils
+import com.github.tototoshi.csv.CSVWriter
+import java.io.FileWriter
+import java.io.File
 
 
 object FlatVariantToSparseByPosition extends SparkApp {
@@ -54,5 +58,10 @@ object FlatVariantToSparseByPosition extends SparkApp {
       .map {case (locusId, sparseVector) =>
         LocusSparseVariant(locusId, sparseVector.size, sparseVector.indices, sparseVector.values)}
       .toDF().saveAsParquetFile(output)  
+      
+    // and also commit the dictionary
+    LoanUtils.withCloseable(new CSVWriter(new FileWriter(new File(output, "_index.csv")))) { cvsWriter =>
+        subjectToIndexMap.foreach(t => cvsWriter.writeRow(t.productIterator.toSeq))
+    }
   } 
 }

### 2015-11-26 variantspark/src/main/scala/au/csiro/pbdava/sparkle/LoanUtils.scala

@@ -1,0 +1,14 @@
+package au.csiro.pbdava.sparkle
+
+import java.io.Closeable
+
+object LoanUtils {
+
+  def withCloseable[C <: Closeable,R](cl:C)(func:C => R):R = {
+    try {
+      func(cl)
+    } finally {
+      cl.close()
+    }
+  }  
+}

### 2015-11-26 run-clustering.sh
Added running scripts

@@ -1,0 +1,44 @@
+#!/bin/bash
+
+
+input_file="\
+/flush/obr17q/phase3/ALL.chr1.phase3_shapeit2_mvncall_integrated_v5a.20130502.genotypes.vcf,\
+/flush/obr17q/phase3/ALL.chr2.phase3_shapeit2_mvncall_integrated_v5a.20130502.genotypes.vcf,\
+/flush/obr17q/phase3/ALL.chr3.phase3_shapeit2_mvncall_integrated_v5a.20130502.genotypes.vcf,\
+/flush/obr17q/phase3/ALL.chr4.phase3_shapeit2_mvncall_integrated_v5a.20130502.genotypes.vcf,\
+/flush/obr17q/phase3/ALL.chr5.phase3_shapeit2_mvncall_integrated_v5a.20130502.genotypes.vcf,\
+/flush/obr17q/phase3/ALL.chr6.phase3_shapeit2_mvncall_integrated_v5a.20130502.genotypes.vcf,\
+/flush/obr17q/phase3/ALL.chr7.phase3_shapeit2_mvncall_integrated_v5a.20130502.genotypes.vcf,\
+/flush/obr17q/phase3/ALL.chr8.phase3_shapeit2_mvncall_integrated_v5a.20130502.genotypes.vcf,\
+/flush/obr17q/phase3/ALL.chr9.phase3_shapeit2_mvncall_integrated_v5a.20130502.genotypes.vcf,\
+/flush/obr17q/phase3/ALL.chr10.phase3_shapeit2_mvncall_integrated_v5a.20130502.genotypes.vcf,\
+/flush/obr17q/phase3/ALL.chr11.phase3_shapeit2_mvncall_integrated_v5a.20130502.genotypes.vcf,\
+/flush/obr17q/phase3/ALL.chr12.phase3_shapeit2_mvncall_integrated_v5a.20130502.genotypes.vcf,\
+/flush/obr17q/phase3/ALL.chr13.phase3_shapeit2_mvncall_integrated_v5a.20130502.genotypes.vcf,\
+/flush/obr17q/phase3/ALL.chr14.phase3_shapeit2_mvncall_integrated_v5a.20130502.genotypes.vcf,\
+/flush/obr17q/phase3/ALL.chr15.phase3_shapeit2_mvncall_integrated_v5a.20130502.genotypes.vcf,\
+/flush/obr17q/phase3/ALL.chr16.phase3_shapeit2_mvncall_integrated_v5a.20130502.genotypes.vcf,\
[…]

### 2015-11-26 run-flat2sparse-bylocus.sh

@@ -1,0 +1,14 @@
+#!/bin/bash
+
+rm -rf /flush2/szu004/variant/phase3.sparse.bylocus.parquet
+
+spark-submit --class au.csiro.obr17q.variantspark.FlatVariantToSparseByPosition \
+      --master yarn-client \
+      --num-executors 100 \
+      --executor-memory 8G \
+      --executor-cores 1 \
+      --driver-memory 8G \
+      variantspark/target/variantspark-0.0.1-SNAPSHOT-all.jar \
+      /flush2/szu004/variant/phase3.flat.parquet \
+      /flush2/szu004/variant/phase3.sparse.bylocus.parquet \
+      1000

### 2015-11-26 run-sparse-clustering.sh

@@ -1,0 +1,20 @@
+#!/bin/bash
+
+input_file="/flush2/szu004/variant/phase3.sparse.parquet"
+clusters=5
+IncludeGroups=""
+ExcludeGroups=""
+VariantCutoff="0"
+
+spark-submit --class au.csiro.obr17q.variantspark.SparseClustering \
+      --master yarn-client \
+      --num-executors 50 \
+      --executor-memory 16G \
+      --executor-cores 1 \
+      --driver-memory 24G \
+      variantspark/target/variantspark-0.0.1-SNAPSHOT.jar \
+      $input_file \
+      $clusters \
+      "$IncludeGroups" \
+      "$ExcludeGroups" \
+      $VariantCutoff

### 2015-11-26 run-sparse-wide-clusterin.sh

@@ -1,0 +1,16 @@
+#!/bin/bash
+
+input_file="/flush2/szu004/variant/phase3.sparse.bylocus.parquet"
+clusters=5
+iters=20
+
+spark-submit --class au.csiro.obr17q.variantspark.SparseWideClustering \
+      --master yarn-client \
+      --num-executors 100 \
+      --executor-memory 4G \
+      --executor-cores 1 \
+      --driver-memory 8G \
+      variantspark/target/variantspark-0.0.1-SNAPSHOT-all.jar \
+      $input_file \
+      $clusters \
+      $iters

### 2015-11-26 variantspark/src/main/scala/au/csiro/obr17q/variantspark/SparseWideClustering.scala
Added ouput generation for wide sparse clustering

@@ -5,6 +5,11 @@ import org.apache.spark.mllib.clustering.KMeans
 import org.apache.spark.mllib.linalg.Vectors
 import scala.io.Source
 import au.csiro.obr17q.variantspark.algo.WideKMeans
+import au.csiro.pbdava.sparkle.LoanUtils
+import com.github.tototoshi.csv.CSVReader
+import java.io.File
+import java.io.FileReader
+import com.github.tototoshi.csv.CSVWriter
 
 object SparseWideClustering extends SparkApp {
   conf.setAppName("VCF cluster")
@@ -18,6 +23,7 @@ object SparseWideClustering extends SparkApp {
     val inputFiles = args(0)
     val k = args(1).toInt
     val iter = args(2).toInt
+    val output = args(3)
     
     val sqlContext = new org.apache.spark.sql.SQLContext(sc)
     import sqlContext.implicits._
@@ -35,7 +41,18 @@ object SparseWideClustering extends SparkApp {
     val kmeans = new WideKMeans(k, iter)
     val result  = kmeans.run(data)  
     println("Result:")
-    println(kmeans.assignClusters(data, result).toList)
+    val clusterAssignment = kmeans.assignClusters(data, result).toList
+    println(clusterAssignment)
     
+    val indexSubjectMap = LoanUtils.withCloseable(CSVReader.open(new FileReader(new File(inputFiles, "_index.csv")))){
+      csvReader =>
+        csvReader.iterator.map { x => (x(1).toInt,x(0))}.toMap   
+    }
+    
+    LoanUtils.withCloseable(CSVWriter.open(output)) { cswWriter =>
+      clusterAssignment.zipWithIndex.map{ case (cluster,index) => (indexSubjectMap(index), cluster)}
+        .foreach(t => cswWriter.writeRow(t.productIterator.toSeq))
+    }
+  
   } 
 }

### 2015-11-26 r-vcf-clusterer.R
Added R file

@@ -1,0 +1,34 @@
+library(biganalytics)
+
+f1 ="/Users/obr17q/Documents/workspace/genespark/data/data2.vcf"
+
+### Converts a variant to a numeric, i.e. "1|1:12" to 2 (1+1)
+hamming <- function(a){
+  a <- substr(a, 0, 3)
+  a <- strsplit(a, "|", fixed = TRUE)
+  a <- vapply(a, function(x) sum(as.numeric(x)), as.numeric(1))
+}
+
+### With VariantAnnotation
+vcf1 <- as.data.frame(readGT(f1), stringsAsFactors=FALSE) #Read VCF file
+vcf1 <- sapply(vcf1, hamming) #Apply Hamming function
+vcf1 <- t(vcf1) #Transpose matrix
+
+### Without VariantAnnotation
+vcf2 <- read.table(f1, sep="\t",as.is=TRUE) #Read VCF file
+vcf2[1:9] <- list() #Remove the first 9 columns (these aren't variants)
+vcf2 <- sapply(vcf2, hamming) #Apply Hamming function
+vcf2 <- t(vcf2) #Transpose matrix
+
+### Do the two methods return the same matrix?
+all(vcf1[1] == vcf2[1])
+
+### Create big.matrix as required by bigkmeans
+x <- as.big.matrix(vcf1)
+
+### Build a kmeans model
+ans <- bigkmeans(x, 2, nstart=20)
+
+### Calculate within set sum of squared error
+wss[i] <- sum(bigkmeans(x, centers=3)$withinss)
+rand

### 2015-11-26 doc/publications/journal2014/paper.tex
add r time

@@
 Hadoop		& 8	& 32	& 14min 22sec	& 8	& 32	& 14min 23sec	& 0.84	\\
-R			& 1	& 32	& 67min		& 8	& 32	& 7min 25sec	& 0.84	\\
+R			& 1	& 32	& 34min 30sec	& 8	& 32	& 7min 25sec	& 0.84	\\
 Python		& 1	& 32	& 92min		& 8	& 32	& 11min 29sec	& 0.84	\\

### 2015-11-26 doc/publications/journal2014/paper.tex
update R methods

@@
 \subsection*{R Implementation}
-For our R implementation, we read in a VCF file using {\sc read.table}, and then convert the resulting table to a matrix ({\sc vcfMatrix} with {\sc as.data.frame.matrix}.
+For our R implementation, we utilize {\sc read.table} to read in a VCF file as a tab-delimited document.
+As this also captures the first nine columns that don't contain variants, we subsequently remove those columns with {\sc vcf[1:9] <- list()}.
 %TODO VariantAnnotation reads in the file quicker but causes issue down the track so overall read.table is faster, right ? - can you add this explanation to the supplementary material.
-In the supplementary material we benchmark this faster approach against reading the VCF file in using the `VariantAnnotation' package.
-We remove the first 9 columns which leaves us with just the variants. As with our \variantSpark{} pre-processing, we convert the strings that represent each allele to a numeric value.
+In the supplementary material we benchmark a slightly faster (by less than one minute) approach of reading the VCF file in using the `VariantAnnotation' package.
+
+As with our \variantSpark{} pre-processing, we convert the strings that represent each allele to a numeric value. This process is identical for both of our R approaches, and essentially consists
+of applying our {\sc Hamming} function to the dataframe with {\sc sapply}.
 We then transpose the matrix with {\sc t(vcfMatrix)}, which results in a data-structure where each row represents an individual. We convert the matrix to a {\sc big.matrix} object,

### 2015-11-27 variantspark/src/main/scala/au/csiro/obr17q/variantspark/EvalClustering.scala
Added adjusted rand index calculation and clustring evaluation:

@@ -1,0 +1,37 @@
+package au.csiro.obr17q.variantspark
+
+import au.csiro.obr17q.variantspark.CommonFunctions._
+import org.apache.spark.mllib.clustering.KMeans
+import org.apache.spark.mllib.linalg.Vectors
+import scala.io.Source
+import au.csiro.pbdava.sparkle.LoanUtils
+import com.github.tototoshi.csv.CSVReader
+import au.csiro.obr17q.variantspark.metrics.Metrics
+
+object EvalClustering extends SparkApp {
+  conf.setAppName("VCF cluster")
+  def main(args:Array[String]) {
+    if (args.length < 1) {
+        println("Usage: CsvClusterer <input-path>")
+    }
+
+    val inputFile = args(0)
+
+
+    val PopFiles = Source.fromFile("data/ALL.panel").getLines()
+    val Populations = sc.parallelize(new MetaDataParser(PopFiles, 1, '\t', "NA", 0, 1 ).returnMap(Array(), Array()))
+    
+    val SuperPopulationUniqueId = Populations.map(_.SuperPopulationId).distinct().zipWithIndex() //For ARI
+    val superPopToId = SuperPopulationUniqueId.collectAsMap()
+    val subjectToSuperPopId = Populations.map(ind => (ind.IndividualId, superPopToId(ind.SuperPopulationId))).collectAsMap()
+    val zippedClustring = LoanUtils.withCloseable(CSVReader.open(inputFile)) { csvReader =>
+      csvReader.iterator
+      .filter(subjectAndCluster => subjectToSuperPopId.contains(subjectAndCluster(0)))
+      .map(subjectAndCluster => (subjectToSuperPopId(subjectAndCluster(0)).toInt, subjectAndCluster(1).toInt)).toList
+    }
+    println(zippedClustring)
+    val clustering = zippedClustring.unzip
+    println("RandIndex:")
+    println(Metrics.adjustedRandIndex(clustering._1, clustering._2))
+  } 
+}

### 2015-11-27 variantspark/src/main/scala/au/csiro/obr17q/variantspark/IndividualMap.scala

@@ -3,7 +3,7 @@ package au.csiro.obr17q.variantspark
 /**
  * @author obr17q
  */
-class IndividualMap (val IndividualId: String, val PopulationId: String, val SuperPopulationId: String = "")
+case class IndividualMap (val IndividualId: String, val PopulationId: String, val SuperPopulationId: String = "")
 (val AdditionalOne: String = "", val AdditionalTwo: String = "", val AdditionalThree: String = "", val AdditionalFour: String = "")
 (val Weight: Double = 0, val Height: Double = 0, val BMI: Double = 0)
 (val Sex: Int = 0)

### 2015-11-27 variantspark/src/main/scala/au/csiro/obr17q/variantspark/MetaDataParser.scala

@@ -20,7 +20,7 @@ class MetaDataParser(val PopFiles:Iterator[String], val HeaderLines:Int, val Fil
         val l = parser.parseLine(line)
         val individualId = l(IndividualIdCol)
         val Population = l(PopulationCol)
-        val SuperPopulation = l(extra1)
+        val SuperPopulation = l(extra1+1)
         //val SuperPopulation = popmap.pops(Population)
         //( l(IndividualIdCol), l(PopulationCol) ) // (IndividualID, Population)
         

### 2015-11-27 variantspark/src/main/scala/au/csiro/obr17q/variantspark/metrics/Metrics.scala

@@ -1,0 +1,21 @@
+package au.csiro.obr17q.variantspark.metrics
+
+object Metrics {
+
+  def over2(n:Int) = n*(n-1)/2
+  
+  def adjustedRandIndex(c1:List[Int], c2:List[Int]): Double = {
+    require(c1.size == c2.size)
+    val c1PartitionNo = c1.max + 1
+    val c2PartitionNo = c2.max + 1
+    val n = c1.size
+    val contingencyMatrix = Array.fill(c1PartitionNo)(Array.fill(c2PartitionNo)(0))
+    Range(0,c1.size).foreach { i  => contingencyMatrix(c1(i))(c2(i)) +=1 }    
+    val c1Bound = contingencyMatrix.map(_.sum)
+    val c2Bound = Range(0,c2PartitionNo).map(j=> contingencyMatrix.map(_(j)).sum)
+    val index:Int = contingencyMatrix.flatMap(_.map(over2)).sum
+    val sumAi = c1Bound.map(over2).sum
+    val sumBi = c2Bound.map(over2).sum
+    (index.toDouble - sumAi*sumBi/over2(n).toDouble)/((sumAi+sumBi)/2.0 - sumAi*sumBi/over2(n).toDouble)
+  }
+}

### 2015-11-27 doc/publications/journal2014/paper.tex
update to paper and letter

@@
    addressref={aff1},
-]{\inits{}\fnm{Neil} \snm{Saunders}}
+]{\inits{}\fnm{Neil F.W.} \snm{Saunders}}
 \author[
@@
 ]{\inits{}\fnm{Fabian A.} \snm{Buske}}
+\author[
+   addressref={aff2},
+   email={rodney.scott@newcastle.edu.au}
+]{\inits{}\fnm{Rodney J.} \snm{Scott}}
 \author[
@@
 rendering our approach almost an order of magnitude faster (see Figure 1).
-R and Python are also slower at pre-processing, taking 67 minutes and 92 minutes, respectively. 
+R and Python are also slower at pre-processing, taking 34 minutes and 92 minutes, respectively. %34min 30sec % 2070
 This increased pre-processing time is due to the standard Python and R implementations not natively supporting multithreading, while \variantSpark{}, {\sc ADAM} and Hadoop can use the eight available cores.
@@
 
-%To further demonstrate the scalability of \variantSpark{}, we also cluster the larger 1000 Genomes Project phase 3 VCF files. Although \kMeans{} clustering completes successfully on this dataset,
-%it takes over 27 hours (see Table 2) due to more than twice the number of individuals, and hence more variants.
-For comparison, the uncompressed size of the phase 3 files is 770GB vs the 161GB of the phase 1 dataset.
 To further demonstrate the scalability of \variantSpark{}, we also cluster the 1000 Genomes Project phase 3 data, which contains 3000 individuals from 5 super-populations and as a result has over 80 Million variants.
 The uncompressed size of the phase 3 files is 770GB compared to the 161GB of the phase 1 dataset.
-\variantSpark{} successfully completes the clustering in just 27 hours (see Table 2) with an ARI of 0.82.
+\variantSpark{} successfully completes the clustering in 27 hours (see Table 2) with an ARI of 0.82.
 
@@
 
-%\variantSpark{} performs clustering on VCF files with over 1000 individuals and 40 million variants in under 15 hours using minimal memory (24GB). 
-\variantSpark{} performs clustering on VCF files with over 3000 individuals and 80 million variants in 27 hours using minimal memory (24GB). 
+%TODO add value
+\variantSpark{} performs clustering on VCF files with over 3000 individuals and 80 million variants in XX hours using minimal memory (24GB). 
 \variantSpark{} supports random genome-wide sampling of variants allowing faster clustering for well characterised cohorts where 20\% of the genome is sufficient. 
 On the benchmarking dataset, it outperforms {\sc ADAM} by almost an order of magnitude (4 vs 28 minutes) by processing VCF files directly and storing the information in sparse vectors. % 258 vs 1673 
-\variantSpark{} utilises {\sc Spark}, which allows for in-memory caching and hence performs 85\% faster than Hadoop (29 min).
+\variantSpark{} utilises {\sc Spark}, which allows for in-memory caching and hence performs 86\% faster than Hadoop (29 min). %7-fold speedup
 \variantSpark{} scales to data sizes that are not feasible to process using R or Python due their requirement to load the whole dataset into memory. 
-But even on the small benchmarking dataset, \variantSpark{}'s novel parallelisation approach is faster than traditional multithreading with 94\% speedup over R (75 min), and 95\% over Python (103 min).
+But even on the small benchmarking dataset, \variantSpark{}'s novel parallelisation approach is faster than traditional multithreading with 90\% speedup over R (42 min), and 96\% over Python (103 min).
+%10 fold %25-fold
 \variantSpark{} is also superior in performance and accuracy over the current state-of-the-art tool for individual ancestries determination, {\sc Admixture}.
@@
 \subsection*{R Implementation}
-For our R implementation, we utilize {\sc read.table} to read in a VCF file as a tab-delimited document.
+%For our R implementation, we utilize {\sc read.table} to read in a VCF file as a tab-delimited document.
 As this also captures the first nine columns that don't contain variants, we subsequently remove those columns with {\sc vcf[1:9] <- list()}.
-%TODO VariantAnnotation reads in the file quicker but causes issue down the track so overall read.table is faster, right ? - can you add this explanation to the supplementary material.
-In the supplementary material we benchmark a slightly faster (by less than one minute) approach of reading the VCF file in using the `VariantAnnotation' package.
+For our R implementation, we utilize {\sc read.table} to read in a VCF file as a tab-delimited document and remove the first nine columns that do not contain variants. 
+%In the supplementary material we benchmark a slightly faster (by less than one minute) approach of reading the VCF file in using the `VariantAnnotation' package.
+In the supplementary material we benchmark this against an approach reading the VCF file using the `VariantAnnotation' package, which causes inefficiencies for the subsequent data manipulation tasks and is hence overall slower.
 
@@
 \section*{Author's contributions}
-    DCB and FAB designed the overall study. ARO implemented \variantSpark\ and conducted the comparison. NS obtained the PGP data and generated the graph.
+    DCB and FAB designed the overall study. ARO implemented \variantSpark\ and conducted the comparison. NFS obtained the PGP data and generated the graph. RJS guided the interpretation. 
     All authors contributed to the writing of the document.
@@
 \section*{Acknowledgements}
-  A.R.O was funded by the NSW Cancer Institute Big Data Big Impact schema, F.A.B by the National Health and Medical Research Council [1051757] and both D.C.B and N.S by Commonwealth Scientific and Industrial Research Organisation's Transformational Capability Platform, Science and Industry Endowment Fund and Information Management and Technology Services.
+  ARO was funded by the NSW Cancer Institute Big Data Big Impact schema, FAB by the National Health and Medical Research Council [1051757] and both DCB and NFS by Commonwealth Scientific and Industrial Research Organisation's Transformational Capability Platform, Science and Industry Endowment Fund and Information Management and Technology Services.
 The computation on Azure was funded by Microsoft Azure Research Award.
@@
 \section*{Additional Files}
-  \subsection*{Additional file 1 --- PGP population labels}
+  \subsection*{Additional file 1 --- PGP population labels and implementation details}
     Conversion map self-reported descriptive population labels to super-population labels. 
+    Details of the R implementation.
 

### 2015-11-27 doc/publications/journal2014/paper.tex
time for phase 3

--- 2015-11-26 doc/publications/journal2014/paper.tex
@@
 100\%	& 64	& 2	& 40min 48sec	& 40	& 24	& 14hr 44min	\\
-???\%	& 64 & 2	& ??min ??sec	& 40 & 24	& 27hr 46min	\\
+100\%	& 64 & 2	& 234min 24sec	& 40 & 24	& 27hr 46min	\\
   \hline

### 2015-11-27 doc/publications/journal2014/paper.tex
update to paper and letter

--- 2015-11-27 doc/publications/journal2014/paper.tex
@@
 \author[
+   addressref={aff5},
+   email={Yi.Guo@csiro.au}
+%   email={john.RS.Smith@cambridge.co.uk}
+]{\inits{}\fnm{Yi} \snm{Guo}}
+\author[
    addressref={aff3,aff4},
@@
    addressref={aff1},
+   corref={aff1},
    email={Denis.Bauer@CSIRO.au}
@@
 %  \street{384 Victoria St},
+  \postcode{2052},
+  \city{Sydney},
+  \cny{Australia}
+}
+\address[id=aff5]{%
+  \orgname{CSIRO, Data61},
   \postcode{2052},
@@
 The uncompressed size of the phase 3 files is 770GB compared to the 161GB of the phase 1 dataset.
-\variantSpark{} successfully completes the clustering in 27 hours (see Table 2) with an ARI of 0.82.
+\variantSpark{} successfully completes the clustering in 30 hours (see Table 2) with an ARI of 0.82.
 
@@
 
-%TODO add value
-\variantSpark{} performs clustering on VCF files with over 3000 individuals and 80 million variants in XX hours using minimal memory (24GB). 
+\variantSpark{} performs clustering on VCF files with over 3000 individuals and 80 million variants in 30 hours using minimal memory (24GB). 
 \variantSpark{} supports random genome-wide sampling of variants allowing faster clustering for well characterised cohorts where 20\% of the genome is sufficient. 
@@
 \subsection*{R Implementation}
-%For our R implementation, we utilize {\sc read.table} to read in a VCF file as a tab-delimited document.
-As this also captures the first nine columns that don't contain variants, we subsequently remove those columns with {\sc vcf[1:9] <- list()}.
-For our R implementation, we utilize {\sc read.table} to read in a VCF file as a tab-delimited document and remove the first nine columns that do not contain variants. 
-%In the supplementary material we benchmark a slightly faster (by less than one minute) approach of reading the VCF file in using the `VariantAnnotation' package.
-In the supplementary material we benchmark this against an approach reading the VCF file using the `VariantAnnotation' package, which causes inefficiencies for the subsequent data manipulation tasks and is hence overall slower.
+We utilise the {\sc readGT} from the {\sc VariantAnnotation} package for reading in the VCF file and extracting the the genotype matrix.  
+In the supplementary material we demonstrate that this approach is approximately one minute faster than using R's built-in {\sc read.table} function, which requires  removing the first nine columns that do not contain genotypes.
 
-As with our \variantSpark{} pre-processing, we convert the strings that represent each allele to a numeric value. This process is identical for both of our R approaches, and essentially consists
-of applying our {\sc Hamming} function to the dataframe with {\sc sapply}.
+As with our \variantSpark{} pre-processing, we convert the strings that represent each allele to a numeric value. 
+This process consists of applying our {\sc Hamming} function to the dataframe with {\sc sapply}.
 We then transpose the matrix with {\sc t(vcfMatrix)}, which results in a data-structure where each row represents an individual. We convert the matrix to a {\sc big.matrix} object,

### 2015-11-27 doc/publications/journal2014/paper.tex
change a thing

--- 2015-11-27 doc/publications/journal2014/paper.tex
@@
 \subsection*{Python Implementation}
-Our Python implementation reads in lines from a VCF file as tab-separated values using {\sc DataFrame.from\textunderscore{}csv}, and stores the data in a pandas DataFrame (\url{http://pandas.pydata.org}).
+Our Python implementation reads in lines from a VCF file as tab-separated values using {\sc DataFrame.read\textunderscore{}csv}, and stores the data in a pandas DataFrame (\url{http://pandas.pydata.org}).
 The column headings are the individual IDs and the row headings are the allele locations.

### 2015-11-27 doc/publications/journal2014/paper.tex
typos and remove a sentence

@@
 \subsection*{R Implementation}
-We utilise the {\sc readGT} from the {\sc VariantAnnotation} package for reading in the VCF file and extracting the the genotype matrix.  
-In the supplementary material we demonstrate that this approach is approximately one minute faster than using R's built-in {\sc read.table} function, which requires  removing the first nine columns that do not contain genotypes.
+We utilise {\sc readGT} from the {\sc VariantAnnotation} package for reading in the VCF file and extracting the genotype matrix.  
+In the supplementary material we demonstrate that this approach is approximately one minute faster than using R's built-in {\sc read.table} function.
 

### 2015-11-27 doc/publications/journal2014/paper.tex
fixed typos in r/python/adm/pgp

@@
 The column headings are the individual IDs and the row headings are the allele locations.
-We the first 9 columns and convert the remaining allele strings to numeric values.
-We convert the DataFrame to a matrix with {\sc .as\textunderscore{}matrix()} and cluster the matrix using sci-kit learn (\url{http://scikit-learn.org/stable/}).
+We remove the first 9 columns and convert the remaining allele strings to numeric values by applying our {\sc Hamming} function to the DataFrame.
+We then convert the DataFrame to a matrix with {\sc .as\textunderscore{}matrix()} and cluster the matrix using `scikit-learn' (\url{http://scikit-learn.org/stable/}).
 
@@
 \subsection*{Personal Genome Project Data}
-Genome data was downloaded from the Personal Genome Project Data (PGP) website (\url{https://my.pgp-hms.org/public_genetic_data}). Genotype data from 23andMe microarray platforms was sorted by genome 
-build. Shell scripts were written to convert NCBI build 36 files to BED format, update to build 37 using UCSC liftOver (\url{https://genome.ucsc.edu/cgi-bin/hgLiftOver}) and 
+We acquired genomic data from the Personal Genome Project (PGP) website (\url{https://my.pgp-hms.org/public_genetic_data}). We sorted the genotype data from 23andMe microarray platforms by genome 
+build. Using custom shell scripts, we convert NCBI build 36 files to BED format, update to build 37 using UCSC liftOver (\url{https://genome.ucsc.edu/cgi-bin/hgLiftOver}) and 
 then convert back to 23andMe format.
-The 23andMe files were then converted to VCF format using code obtained from the Broad Institute (\url{http://apol1.blogspot.com.au/2013/08/impute-apoe-and-apol1-with-23andme.html}).
-Where individuals had genotype data for both genome builds 36 and 37, only the latter was used. Finally, individual VCF files were combined for clustering into one 
-file using VCFtools \cite{Danecek2011Variant}.
+We then convert the 23andMe files to VCF format using code obtained from the Broad Institute (\url{http://apol1.blogspot.com.au/2013/08/impute-apoe-and-apol1-with-23andme.html}).
+Where individuals had genotype data for both genome builds 36 and 37, only the latter was used. Finally, we combined the individual VCF files into one 
+file for clustering, using VCFtools \cite{Danecek2011Variant}.
 
@@
 R			& 1	& 32	& 34min 30sec	& 8	& 32	& 7min 25sec	& 0.84	\\
-Python		& 1	& 32	& 92min		& 8	& 32	& 11min 29sec	& 0.84	\\
+Python		& 1	& 32	& 34min 15sec	& 8	& 32	& 11min 29sec	& 0.84	\\
 {\sc Admixture}	& 1	& 32	& 10min 8sec	& 8	& 32 & 8min 19sec	& 0.25	\\

### 2015-11-27 doc/publications/journal2014/paper.tex
a few typos

@@
 We also cluster the Phase 3 dataset, which contains variants from 2535 individuals, across 81,271,745 variants. As well as the above for super populations, this
-dataset also includes South Asian (SAS).
+dataset also includes the South Asian (SAS) super population.
 
@@
 To cluster individuals from a VCF file, we initially need to pre-process the variants to feature vectors.
-In \variantSpark{} we read in VCF files as text files, to a Resilient Distributed Dataset (RDD). RDDs allow us to process the files, line-by-line, in parallel.
+In \variantSpark{} we read in VCF files as text files to a Resilient Distributed Dataset (RDD). RDDs allow us to process the files, line-by-line, in parallel.
 
-We parse each line as tab-separated values, and store these the values from each line in an array.
+We parse each line as tab-separated values, and store the values from each line in an array.
 
-For each array, we use Spark's `zip' function to `zip' each value with it's heading (Sample name) from the VCF file. Now each array element is a
+For each array, we use Spark's `zip' function to create tuples of the values with their respective heading (sample name). Now each array element is a
 key-value pair (KVP) of the previous value and its heading. For the KVPs that contain an allele as the value, the key (derived from the heading) is
-the Individual ID (see Figure 2, Spark step 1).
+the individual ID (see Figure 2, Spark step 1).
 
@@
 
-Now that the dataset only contains alleles we are interested in, we `zip' each array with a unique sequential index (allele ID). This index will serve
+Now that the dataset only contains alleles we are interested in, we zip each array with a unique sequential index (allele ID). This index will serve
 as identifier for each allele (and keep the genome loci order) (step 2).

### 2015-11-27 doc/publications/journal2014/paper.tex
a few typos

@@
 This is especially the case for sequence analysis tasks, such as read mapping~\cite{Schatz2009}, duplicate removal~\cite{Jourdren2012}, and variant calling~\cite{Langmead2009, McKenna2010} as well as Genome Wide Analysis Study based tasks~\cite{Huang2013, Guo2014}.
-Apache also developed a machine learning library, Mahout~\cite{Owen2011}, which allows efficient out-of-the-box analysis to be applied to clinical applications, such as medical health records~\cite{Ko2014}.
+Apache has also developed a machine learning library, Mahout~\cite{Owen2011}, which allows efficient out-of-the-box analysis to be applied to clinical applications, such as medical health records~\cite{Ko2014}.
 Unfortunately, the MapReduce paradigm is not always the optimal solution, specifically for bioinformatics or machine learning applications that require iterative in-memory computation.
@@
 This increased pre-processing time is due to the standard Python and R implementations not natively supporting multithreading, while \variantSpark{}, {\sc ADAM} and Hadoop can use the eight available cores.
-The GATK-based VCF to PED conversion for {\sc Admixture} is also slower with over 10 minutes.
+The GATK-based VCF to PED conversion for {\sc Admixture} is also slower, with a runtime of over 10 minutes.
 
@@
 {\sc Admixture} is also slower with a runtime of just over 8 minutes. 
-Our Hadoop implementation performs the worst with 14 minutes, showcasing the limitations of the Hadoop engine for iterative algorithms such as \kMeans{}. % 863 14mins and 23secs
+Our Hadoop implementation performs the worst, at 14 minutes, showcasing the limitations of the Hadoop engine for iterative algorithms such as \kMeans{}. % 863 14mins and 23secs
 Hadoop writes the entire output of each \kMeans{} iteration to disk.
-This feature was designed to ensure scalability but here substantially increases the runtime as disk IO becomes the limiting factor for this small dataset (chromosome 22).
-Python and R store the output from each iteration in memory, thereby eliminating the IO bottleneck, but without being able to utilise the disk they are limited to datasets that fit into memory.
+This feature enables scalability but, in this case, substantially increases the runtime as disk IO becomes the limiting factor for this small dataset (chromosome 22).
+Python and R store the output from each iteration in memory, thereby eliminating the IO bottleneck, but without being able to utilise disk storage they are limited to datasets that fit into memory.
 {\sc Spark} overcomes both limitations by performing in-memory caching for each executor, which eliminates the disk IO bottleneck on smaller datasets, while maintaining scalability to larger datasets by spilling data that exceeds memory availability to disk.

### 2015-11-27 doc/publications/journal2014/paper.tex
a few typos

@@
 
-We also investigate the cluster quality for the five different methods by comparing the annotated super-population label (AMR, AFR, EAS, SAS) for each individual in the 1000 Genomes data to the assigned label from the clustering allowing four clusters.
+We also investigate the cluster quality for the five different methods by comparing the annotated super-population label (AMR, AFR, EAS, SAS) for each individual in the 1000 Genomes data to the label assigned by \kMeans{} clustering.
 For this comparison, we use the \ARI{} (ARI) metric, which returns a value between -1 (independent labelings) and 1 (perfect match)~\cite{Hubert1985}.
@@
 Only a minority of AMR individuals form an independent group, likely comprising of genetic information otherwise not captured by the 26 sub-populations of the 1000 genomes project.
-This indicates that although allelic differences exist between populations, "genetic diversity is distributed on a continuum"~\cite{Pugach2015}.
+This indicates that although allelic differences exist between populations, ``genetic diversity is distributed on a continuum"~\cite{Pugach2015}.
 
@@
 Although we have demonstrated the advantages of \variantSpark{} over traditional methods on small datasets, such as individual chromosomes, the main advantage of {\sc Spark} is its ability to process datasets that exceed memory limits.
-To demonstrate this scalability, we increase the number of variants used in the clustering from the initial 494,328 on chromosome 22 to 38,219,238 including all variants in the genome.
+To demonstrate this scalability, we increase the number of variants used in the clustering from the initial 494,328 on chromosome 22 to 38,219,238 variants across the 1000 Genomes Project dataset.
 
@@
 The increase in memory is due to more distance measurements between variants and \kMeans{} centroids being created, which MLlib stores as dense vectors. 
-While \kMeans{} clustering can scale to 100\% of the genome, machine learning algorithms that can deal with sparse vectors would potentially be able to scale much further, for example process more alleles.
+While \kMeans{} clustering can scale to 100\% of the genome, machine learning algorithms that can deal with sparse vectors would potentially be able to scale much further and potentially process more alleles.
 
@@
 To further demonstrate the scalability of \variantSpark{}, we also cluster the 1000 Genomes Project phase 3 data, which contains 3000 individuals from 5 super-populations and as a result has over 80 Million variants.
-The uncompressed size of the phase 3 files is 770GB compared to the 161GB of the phase 1 dataset.
+The uncompressed size of the phase 3 files is 770GB compared to 161GB for the phase 1 dataset.
 \variantSpark{} successfully completes the clustering in 30 hours (see Table 2) with an ARI of 0.82.
@@
 For example, the `White' label includes individuals with grandparents from United States, Syria, Arab Republic, or Bulgaria, amongst others.
-The dataset is also very one-sided with 433 out of the 478 (over 90\%) of the individuals being classified as white, even though geographical location of their grandparents are very diverse.
+The dataset is also very one-sided with 433 out of the 478 (over 90\%) of the individuals labeled as white, even though geographical location of their grandparents are very diverse.
 We therefore only include individuals where all grandparents were reported to be from the same country leaving 35 individuals from 22 different countries, which we group into their approximate super-populations (see Supplemental Table 1).
@@
 from genome sequencing) would help in elucidating population structure in this dataset, which demonstrates the need for fast whole-genome approaches.
-Unfortunately, this hypothesis can no not be tested as whole genome information is not available for these individuals.
+Unfortunately, this hypothesis can not be tested as whole genome information is not available for these individuals.
 

### 2015-11-27 doc/publications/journal2014/paper.tex
updated times in document

@@
 rendering our approach almost an order of magnitude faster (see Figure 1).
-R and Python are also slower at pre-processing, taking 34 minutes and 92 minutes, respectively. %34min 30sec % 2070
+R and Python are also slower at pre-processing, each taking approximately 34 minutes. %34min 30sec % 2070
 This increased pre-processing time is due to the standard Python and R implementations not natively supporting multithreading, while \variantSpark{}, {\sc ADAM} and Hadoop can use the eight available cores.
@@
 \variantSpark{} scales to data sizes that are not feasible to process using R or Python due their requirement to load the whole dataset into memory. 
-But even on the small benchmarking dataset, \variantSpark{}'s novel parallelisation approach is faster than traditional multithreading with 90\% speedup over R (42 min), and 96\% over Python (103 min).
+But even on the small benchmarking dataset, \variantSpark{}'s novel parallelisation approach is faster than traditional multithreading with 90\% speedup over R (42 min), and 91\% over Python (45 min).
 %10 fold %25-fold
@@
 memory running Ubuntu. 
-For the whole-genome clustering, we used our in-house Hadoop cluster with Hadoop 2.5.0, managed by cloudera's CDH 5. We use Spark 1.3.1. This 13 node
+For the whole-genome clustering, we used our in-house Hadoop cluster with Hadoop 2.5.0, managed by Cloudera's CDH 5. We use Spark 1.3.1. This 13 node
 cluster has a total of 416 cores and 1.22TB memory.

### 2015-11-27 doc/publications/journal2014/paper.tex
typos and passive

@@
 \subsection*{Personal Genome Project Data}
-We acquired genomic data from the Personal Genome Project (PGP) website (\url{https://my.pgp-hms.org/public_genetic_data}). We sorted the genotype data from 23andMe microarray platforms by genome 
-build. Using custom shell scripts, we convert NCBI build 36 files to BED format, update to build 37 using UCSC liftOver (\url{https://genome.ucsc.edu/cgi-bin/hgLiftOver}) and 
+We acquired genomic data from the Personal Genome Project (PGP) website (\url{https://my.pgp-hms.org/public_genetic_data}). We sort the genotype data from 23andMe microarray platforms by genome 
+build. Using a custom shell script, we convert NCBI build 36 files to BED format, update to build 37 using UCSC liftOver (\url{https://genome.ucsc.edu/cgi-bin/hgLiftOver}) and 
 then convert back to 23andMe format.
-We then convert the 23andMe files to VCF format using code obtained from the Broad Institute (\url{http://apol1.blogspot.com.au/2013/08/impute-apoe-and-apol1-with-23andme.html}).
-Where individuals had genotype data for both genome builds 36 and 37, only the latter was used. Finally, we combined the individual VCF files into one 
+We then convert the 23andMe files to VCF using code obtained from the Broad Institute (\url{http://apol1.blogspot.com.au/2013/08/impute-apoe-and-apol1-with-23andme.html}).
+Where individuals had genotype data for both genome builds 36 and 37, we only use the latter. Finally, we combined the individual VCF files into one 
 file for clustering, using VCFtools \cite{Danecek2011Variant}.

### 2015-11-27 py-vcf-clusterer.py
Added python clusterer

@@ -1,0 +1,25 @@
+import pandas as pd
+import io
+file_path = "/Users/obr17q/Documents/workspace/genespark/data/data2.vcf"
+
+### Convets strings from a serial to integers
+def hamming( variant ):
+    return variant\
+    .str.split(':')\
+    .str.get(0)\
+    .str.split('|')\
+    .apply(lambda x: sum(map(int, x)))
+
+### Read in the VCF file and drop first 9 columns
+with io.open(file_path, 'r') as f:
+    for line in f:
+        if line[0:2] == '##':
+            pass
+        else:
+            break
+    vcf = pd.read_csv(f, sep='\t', header=None)
+    vcf = vcf.drop(vcf.columns[[range(0,9)]], axis=1)
+
+### Apply Hamming to DataFrame and transpose
+vcf = vcf.apply(hamming).T
+

### 2015-11-27 py-vcf-clusterer.py
update kmeans

@@ -1,5 +1,7 @@
 import pandas as pd
 import io
+from sklearn.cluster import KMeans
+
 file_path = "/Users/obr17q/Documents/workspace/genespark/data/data2.vcf"
 
 ### Convets strings from a serial to integers
@@ -23,3 +25,12 @@ with io.open(file_path, 'r') as f:
 ### Apply Hamming to DataFrame and transpose
 vcf = vcf.apply(hamming).T
 
+### Convert DataFrame to matrix
+vcfMatrix = vcf.as_matrix()
+
+### Build kmeans model and fit data
+model = KMeans(n_clusters=4)
+model.fit(vcfMatrix)
+
+### Labels
+labels = model.labels_

### 2015-11-27 doc/publications/journal2014/paper.tex
final version for publication

@@
 
-\parttitle{Results} To demonstrate the capabilities of \variantSpark, we clustered more than 1,000 individuals with 38 Million variants each to determine the population structure in the dataset.
+\parttitle{Results} To demonstrate the capabilities of \variantSpark, we clustered more than 3,000 individuals with 80 Million variants each to determine the population structure in the dataset.
 \variantSpark{} is 80\% faster than the {\sc Spark}-based genome clustering approach, {\sc ADAM}, the comparable implementation using Hadoop/Mahout, as well as a {\sc Admixture}, a commonly used tool for determining individual ancestries.
@@
 Recognising the capability, Wiewi{\'o}rka {\it et al.}~\cite{Wiewiorka2014} developed {\sc SparkSeq} for high-throughput sequence data analysis and the Big Data Genomics (BDG) group recently demonstrated the strength of {\sc Spark} in a genomic clustering application using {\sc ADAM}, a set of formats and APIs as well as processing stage implementations for genomic data~\cite{Massie2013}.
-While the speedup of {\sc ADAM} over traditional methods was impressive (50 fold speedup), being limited by constraints within this general genomics framework hampered performance.
+ADAM is expected to be one of the cornerstones of the Precision Medicine Initiative and the Global Alliance for Genomics and Health~\cite{Paten2015}.
+While the speedup of {\sc ADAM} over traditional methods was impressive (50 fold speedup), being limited by constraints within this general genomics framework can hamper performance.
 
-We hence developed a purpose-built application in {\sc Spark} to perform genomic clustering of individuals. 
+We hence developed a purpose-built appraoch in {\sc Spark} to perform machine learning tasks on genomic data, such as clustering of individual genomes. 
 We utilise {\sc Spark}'s machine learning library, \mbox{MLlib}, and provide an interface to the standard variant data format, Variant Call Format (VCF)~\cite{1KG2012}, which opens up the application of MLlib's different machine learning algorithms to a wide range of genotype-based analysis tasks.
@@
 
-In this section we compare the time required to cluster individuals based on genomic variants using \variantSpark{} against {\sc ADAM} as well as the more traditional approaches using Hadoop (and Mahout), Python and R.
+In this section we compare the time required to cluster individuals based on genomic variants using \variantSpark{} against {\sc ADAM} and the more traditional approaches using Hadoop (and Mahout), Python and R, as well as the purpose-build tool {\sc Admixture}.
 We limit the genomic variants to only chromosome 22 as the traditional approaches have substantially larger memory consumption, rendering a whole-genome input infeasible.
@@
 Although we have demonstrated the advantages of \variantSpark{} over traditional methods on small datasets, such as individual chromosomes, the main advantage of {\sc Spark} is its ability to process datasets that exceed memory limits.
-To demonstrate this scalability, we increase the number of variants used in the clustering from the initial 494,328 on chromosome 22 to 38,219,238 variants across the 1000 Genomes Project dataset.
+To demonstrate this scalability, we increase the number of variants used in the clustering from the initial 494,328 on chromosome 22 to 38,219,238 variants across the 1000 Genomes Project dataset (phase 1).
 
-We run this comparison on our in-house Hadoop cluster using 40 cores.
+We run this comparison on our in-house {\sc Hadoop/Spark} cluster using 40 cores.
 Pre-processing the VCF files takes approximately 12, 19, 27 and 40 minutes for 20, 40, 60 and 100 percent of the genome, respectively (see Table 2).
@@
 HLA allele genotyping from SNP information~\cite{Zheng2014}) and accurate population association may not be known for patients with diverse migrational background.
-As noted by Patterson et al ~\cite{Patterson2006}, more markers are needed for populations with low genetic divergence. 
+As noted by Patterson {\it et al.}~\cite{Patterson2006}, more markers are needed for populations with low genetic divergence. 
 It is hence likely that higher density genotyping (e.g.
@@
 But even on the small benchmarking dataset, \variantSpark{}'s novel parallelisation approach is faster than traditional multithreading with 90\% speedup over R (42 min), and 91\% over Python (45 min).
-%10 fold %25-fold
+%10 fold %10-fold
 \variantSpark{} is also superior in performance and accuracy over the current state-of-the-art tool for individual ancestries determination, {\sc Admixture}.
 These benefits of speed, resource consumption and scalability allows \variantSpark{} to be the interface for applying other machine learning tools from MlLib to genomic data. 
-Utilising MLlib as well as Spark.ML will enable supervised machine learning applications to e.g.
+Utilising MLlib as well as the more recent addition, Spark.ML, will enable supervised machine learning applications to e.g.
 identify variants that jointly interact with phenotypes as well as include electronic health record in addition to the genomic feature vector to e.g.
@@
 We parse each line as tab-separated values, and store the values from each line in an array.
-
-For each array, we use Spark's `zip' function to create tuples of the values with their respective heading (sample name). Now each array element is a
+For each array, we use {\sc Spark}'s `zip' function to create tuples of the values with their respective heading (sample name). Now each array element is a
 key-value pair (KVP) of the previous value and its heading. For the KVPs that contain an allele as the value, the key (derived from the heading) is
@@
 
-At this stage, we can optionally filter variants that don't match a specific criteria. For the Phase 3 dataset, we exclude rare variants that are present in
+At this stage, we can optionally filter variants that do not match a specific criteria. For the Phase 3 dataset, we exclude rare variants that are present in
 only one individual. Because each array now only contains variants (rather than every allele), we can simply drop arrays with a length of one, as the length
@@
 the {\sc big.matrix} object and the required number of clusters as arguments.
+See git repository for more details.
 
@@
 We then convert the DataFrame to a matrix with {\sc .as\textunderscore{}matrix()} and cluster the matrix using `scikit-learn' (\url{http://scikit-learn.org/stable/}).
+See git repository for more details.
 
@@
 \section*{Author's contributions}
-    DCB and FAB designed the overall study. ARO implemented \variantSpark\ and conducted the comparison. NFS obtained the PGP data and generated the graph. RJS guided the interpretation. 
+    DCB and FAB designed the overall study. 
+    ARO implemented \variantSpark\ and conducted the comparison. 
+    YG contributed to the implementation comparison. 
+    NFS obtained the PGP data and generated the graphic. 
+    RJS guided the interpretation. 
     All authors contributed to the writing of the document.
@@
 \caption{The resource consumption of the six compared methods as well as the accuracy measured as \ARI{} on chromosome 22.}
-      \begin{tabular}{|c|ccc|ccc|c|c|}
+      \begin{tabular}{|c|ccr|ccr|c|c|}
         \hline
@@
 Python		& 1	& 32	& 34min 15sec	& 8	& 32	& 11min 29sec	& 0.84	\\
-{\sc Admixture}	& 1	& 32	& 10min 8sec	& 8	& 32 & 8min 19sec	& 0.25	\\
+{\sc Admixture}	& 1	& 32	& 10min 08sec	& 8	& 32 & 8min 19sec	& 0.25	\\
   \hline
@@
 \begin{table}[h!]
-\caption{The resources consumption on different subsets of the entire autosome (chromosomes 1-22). Memory specified is the memory allocated to each executor.}
-      \begin{tabular}{|c|ccc|ccc|c|c|}
+\caption{The resources consumption on different subsets of the entire autosome (chromosomes 1-22) of phase 1 as well as all of phase 3. Memory specified is the memory allocated to each executor.}
+      \begin{tabular}{|l|c|ccr|ccr|c|r|}
         \hline
-           Portion &  \multicolumn{3}{ c |}{Pre-processing} & \multicolumn{3}{ c |}{Clustering}  \\
-& executors & memory & time  & executors & memory & time \\
+           Data & Portion &  \multicolumn{3}{ c |}{Pre-processing} & \multicolumn{3}{ c |}{Clustering}  \\
+&& executors & memory & time  & executors & memory & time \\
   \hline
-20\%		& 64	& 2	& 11min 53sec	& 64	& 6	& 1hr 10min	\\
-40\%		& 64	& 2	& 19min 9sec	& 64	& 12	& 2hr 19min	\\
-60\%		& 64	& 2	& 26min 34sec	& 64	& 17	& 3hr 33min	\\
-100\%	& 64	& 2	& 40min 48sec	& 40	& 24	& 14hr 44min	\\
-100\%	& 64 & 2	& 234min 24sec	& 40 & 24	& 27hr 46min	\\
+phase 1 & 20\%		& 64	& 2	& 11min 53sec	& 64	& 6	& 1hr 10min	\\
+& 40\%		& 64	& 2	& 19min 09sec	& 64	& 12	& 2hr 19min	\\
+& 60\%		& 64	& 2	& 26min 34sec	& 64	& 17	& 3hr 33min	\\
+& 100\%	& 64	& 2	& 40min 48sec	& 40	& 24	& 14hr 44min	\\
+phase 3 &100\%	& 64 & 2	& 3hr 54min 24sec	& 40 & 24	& 27hr 46min \\
+%100\%	& 64 & 2	& 234min 24sec	& 40 & 24	& 27hr 46min	
   \hline

### 2015-11-30 doc/publications/journal2014/paper.tex

@@
 \parttitle{Results} To demonstrate the capabilities of \variantSpark, we clustered more than 3,000 individuals with 80 Million variants each to determine the population structure in the dataset.
-\variantSpark{} is 80\% faster than the {\sc Spark}-based genome clustering approach, {\sc ADAM}, the comparable implementation using Hadoop/Mahout, as well as a {\sc Admixture}, a commonly used tool for determining individual ancestries.
+\variantSpark{} is 80\% faster than the {\sc Spark}-based genome clustering approach, {\sc ADAM}, the comparable implementation using Hadoop/Mahout, as well as {\sc Admixture}, a commonly used tool for determining individual ancestries.
 It is over 90\% faster than traditional implementations using R and Python.

### 2015-12-08 doc/publications/journal2016/paper.tex
Changes

\documentclass[11pt]{amsart}
\usepackage{geometry}                % See geometry.pdf to learn the layout options. There are lots.
\geometry{letterpaper}                   % ... or a4paper or a5paper or ... 
%\geometry{landscape}                % Activate for for rotated page geometry
%\usepackage[parfill]{parskip}    % Activate to begin paragraphs with an empty line rather than an indent
\usepackage{graphicx}
\usepackage{amssymb}
\usepackage{epstopdf}
\DeclareGraphicsRule{.tif}{png}{.png}{`convert #1 `dirname #1`/`basename #1 .tif`.png}

\title{Brief Article}
\author{The Author}
%\date{}                                           % Activate to display a given date or no date

\begin{document}
\maketitle
\section{Introduction}
As of 2014, obesity effects more than 600 million individuals worldwide, a number which has doubled since 1980 (who2015). The WHO definition for obesity is a BMI of greater than or equal to 30.
In addition to these individuals, 1.9 million adults are overweight.
Although changes in lifestyle factors, such as diet, contribute to this rise in obesity (Binkley2000), as a multifactorial disease (Grundy1998), certain genetic variants may predispose individuals to obesity.

Recently, Wheeler2013 identified 29 loci with evidence of association with obesity from an analysis of approximately 2 million SNPs.
This is in addition to 52 other genetic loci identified by GWAS over the four years prior (Loos2012).
However, with the relatively low coverage of GWAS compared to whole genome sequencing approaches, it's possible that GWAS studies may miss rare variants.
With genome sequencing becoming more accessible, the number of SNPs and other variants in recent datasets is increasing. For example,
the 1000 Genomes Project phase 3 dataset (Abecasis2012) contains more than 80 million variant sites.

As the size of the data increases, and the dimensionality (number of variants) increases, so does the compute complexity.
For this reason, we implemented MLlib, an interface between Variant Call Format (VCF) files and Spark.
Unlike approaches in R and Python, MLlib can scale further and is faster on equivalent, smaller datasets.

% What we do
Using Spark's Machine learning library, `Spark ml', we build a Random Forest model based on The Cancer Genome Atlas (TCGA) data, with the aim to
elucidate genetic variants associated with obesity.

We choose Random Forests for reasons such as their robustness to noise ( Breiman2001)

%For example, individuals homozygous for risk alleles in in FTO,
%a transcription factor for IRX3, weigh on average 3 kg more than those homozygous for non-risk alleles.

%The decreasing cost of genome sequencing means we have more data

%\subsection{dfh}

\end{document}

### 2015-12-08 R/testRandomForest.R
Changed variable importance calculation

@@ -4,4 +4,4 @@ data <-read.table('wide-data.csv',header=FALSE, sep=",")
 X <- t(data[-1,])
 y <- as.factor(t(data[1,]))
 
-model <-randomForest(X,y,ntree=300, importance=TRUE)
+model <-randomForest(X,y,ntree=500, importance=TRUE)

### 2015-12-08 variantspark/src/main/scala/au/csiro/obr17q/variantspark/SparseWideForest.scala

@@ -56,9 +56,9 @@ object SparseWideForest extends SparkApp {
     
     
     val rf = new WideRandomForest()
-    val result  = rf.run(data,labels.toArray, 50)
+    val result  = rf.run(data,labels.toArray, 100)
     //println(result)
-    result.printout()
+    //result.printout()
     val variableImportnace = result.variableImportance
     
     variableImportnace.toSeq.sortBy(-_._2).take(20).foreach(println)

### 2015-12-08 variantspark/src/main/scala/au/csiro/obr17q/variantspark/algo/WideDecisionTree.scala

@@ -28,7 +28,7 @@ object WideDecisionTree {
     if (total == 0.0) 0.0 else 1 - counts.map(s => sqr(s / total)).sum
   }
 
-  def findSplit(currentSet: Array[Int], labels: Array[Int])(t: (Vector, Long)): (Double, Long, Int, Int, Array[Int], Array[Int]) = {
+  def findSplit(currentSet: Array[Int], labels: Array[Int])(t: (Vector, Long)): (Double, Double, Long, Int, Int, Array[Int], Array[Int]) = {
 
     val v = t._1.toArray
     val index: Long = t._2
@@ -58,19 +58,19 @@ object WideDecisionTree {
     }.min
     // calculat the best ginit split
     // actually also need to return new subsplits
-    (totalGini - splitGini, index, s, majorityLabel, currentSet.filter(i => v(i) <= s), currentSet.filter(i => v(i) > s))
+    (totalGini - splitGini, totalGini, index, s, majorityLabel, currentSet.filter(i => v(i) <= s), currentSet.filter(i => v(i) > s))
   }
 }
 
 case class DecisionTreeNode(variableIndex: Long, splitPoint: Int, majorityLabel: Int,
-    impurityReduction: Double, size: Int, left: DecisionTreeNode = null, right: DecisionTreeNode = null) {
+    impurityReduction: Double, nodeImpurity: Double, size: Int, left: DecisionTreeNode = null, right: DecisionTreeNode = null) {
 
   def isLeaf = (impurityReduction == 0)
 
   def printout(level: Int) {
     print(new String(Array.fill(level)(' ')))
     val nodeType = if (isLeaf) "leaf" else "split"
-    println(s"${nodeType}[${variableIndex}, ${splitPoint}, ${majorityLabel}, ${size}, ${impurityReduction}]")
+    println(s"${nodeType}[${variableIndex}, ${splitPoint}, ${majorityLabel}, ${size}, ${impurityReduction}, ${nodeImpurity}]")
     if (!isLeaf) {
       left.printout(level + 1)
       right.printout(level + 1)
@@ -79,7 +79,8 @@ case class DecisionTreeNode(variableIndex: Long, splitPoint: Int, majorityLabel:
 
   def countImportance(accumulations: Long2DoubleOpenHashMap, totalSize:Int) {
     if (!isLeaf) {
-      accumulations.addTo(variableIndex, impurityReduction)//*size/totalSize.toDouble)
+      
[…]

### 2015-12-09 R/testRandomForest.Rmd
Fixex sparse random forrest

Run rf on a small saple of genom
========================================================

Run rf on a small saple of genom

```{r}
library(randomForest)
data <-read.table('wide-data.csv',header=FALSE, sep=",")

X <- t(data[-1,])
y <- as.factor(t(data[1,]))

model <-randomForest(X,y,ntree=100, importance=TRUE)
```

Plot OOB error rates

```{r fig.width=7, fig.height=6}
plot(model)
```

```{r}
model$err.rate[90:100,]
```

Plot variable importance

```{r fig.width=10, fig.height=7}
varImpPlot(model)
```

### 2015-12-09 variantspark/src/main/scala/au/csiro/obr17q/variantspark/SparseDataExport.scala

@@ -34,12 +34,6 @@ object SparseDataExport extends SparkApp {
       csvReader =>
         csvReader.iterator.map { x => (x(1).toInt,x(0))}.toMap   
     }    
-    val data = 
-      sparseVariat.rdd
-        .map(r=> Vectors.sparse(r.getInt(1),
-          r.getSeq[Int](2).toArray, r.getSeq[Double](3).toArray))
-    val test = data.cache().count()
-    println(test)
     
     val PopFiles = Source.fromFile("data/ALL.panel").getLines()
     val Populations = sc.parallelize(new MetaDataParser(PopFiles, 1, '\t', "NA", 0, 1 ).returnMap(Array(), Array()))
@@ -48,12 +42,45 @@ object SparseDataExport extends SparkApp {
     val superPopToId = SuperPopulationUniqueId.collectAsMap()
     val subjectToSuperPopId= Populations.map(ind => (ind.IndividualId, superPopToId(ind.SuperPopulationId))).collectAsMap()
 
-    val labels = indexSubjectMap.toStream.sorted
+    val allLabels = indexSubjectMap.toStream.sorted
       .map({case (index,subject) => 
         subjectToSuperPopId.getOrElse(subject, superPopToId.size.toLong).toInt
       }).toArray
-     
-    LoanUtils.withCloseable(CSVWriter.open(output)) { cswWriter =>
+    
+   
+   val unknownLabel = superPopToId.size
+   val unknownLabelsCount = allLabels.count( _ == unknownLabel)
+   println(s"Unknown labels count: ${unknownLabelsCount}")
+   // need to get of unknow data   
+      
+   val indexCorrections = Array.fill(allLabels.length)(0)
+  
+   // I am sure there is a way to do this in a nice functional programming way but for now
+   var counts = 0;
+   allLabels.zipWithIndex.foreach { case(v,i) => 
+     if (v == unknownLabel) {
+       counts += 1
+     }
+     indexCorrections(i) = counts
+   }
+   
+   
+   val data = 
+      sparseVariat.rdd      
+        .map{r=> 
+          val (size, indexes,values) = (r.getInt(1),r.getSeq[Int](2).toArray, r.getSeq[Double](3))
+          Vectors.sparse(size - unknownLabelsCount,
+          indexes.filter(i => allLabels(i) < unknownLabel).map(i => i - indexCorrections(i)).toArray,
[…]

### 2015-12-09 variantspark/src/main/scala/au/csiro/obr17q/variantspark/SparseWideForest.scala

@@ -12,9 +12,13 @@ import java.io.FileReader
 import com.github.tototoshi.csv.CSVWriter
 import au.csiro.obr17q.variantspark.algo.WideDecisionTree
 import au.csiro.obr17q.variantspark.algo.WideRandomForest
+import au.csiro.obr17q.variantspark.algo.RandomForestParams
+import scala.collection.mutable.MutableList
 
 object SparseWideForest extends SparkApp {
   conf.setAppName("VCF cluster")
+  //conf.registerKryoClasses(Array(classOf[TreeSplitInfo]))
+  
   def main(args:Array[String]) {
 
    
@@ -24,6 +28,7 @@ object SparseWideForest extends SparkApp {
 
     val inputFiles = args(0)
     val output = args(1)
+    val ntree = args(2).toInt
     
     val sqlContext = new org.apache.spark.sql.SQLContext(sc)
     import sqlContext.implicits._
@@ -34,12 +39,6 @@ object SparseWideForest extends SparkApp {
       csvReader =>
         csvReader.iterator.map { x => (x(1).toInt,x(0))}.toMap   
     }    
-    val data = 
-      sparseVariat.rdd
-        .map(r=> Vectors.sparse(r.getInt(1),
-          r.getSeq[Int](2).toArray, r.getSeq[Double](3).toArray))
-    val test = data.cache().count()
-    println(test)
     
     val PopFiles = Source.fromFile("data/ALL.panel").getLines()
     val Populations = sc.parallelize(new MetaDataParser(PopFiles, 1, '\t', "NA", 0, 1 ).returnMap(Array(), Array()))
@@ -48,16 +47,49 @@ object SparseWideForest extends SparkApp {
     val superPopToId = SuperPopulationUniqueId.collectAsMap()
     val subjectToSuperPopId= Populations.map(ind => (ind.IndividualId, superPopToId(ind.SuperPopulationId))).collectAsMap()
 
-    val labels = indexSubjectMap.toStream.sorted
+    val allLabels = indexSubjectMap.toStream.sorted
       .map({case (index,subject) => 
         subjectToSuperPopId.getOrElse(subject, superPopToId.size.toLong).toInt
       }).toArray
-     
     
+   
+   val unknownLabel = superPopToId.size
+   val unknownLabelsCount = allLabels.count( _ == unknownLabel)
+   println(s"Unknown labels count: ${unknownLabelsCount}")
+   // need to get of unknow data   
+      
+   val indexCorrections = Array.fill(allLabels.length)(0)
+  
[…]

### 2015-12-09 variantspark/src/main/scala/au/csiro/obr17q/variantspark/algo/WideDecisionTree.scala

@@ -2,13 +2,18 @@ package au.csiro.obr17q.variantspark.algo
 
 import scala.Range
 import scala.collection.JavaConversions.asScalaSet
-
 import org.apache.spark.mllib.linalg.Vector
 import org.apache.spark.mllib.linalg.Vectors
 import org.apache.spark.rdd.RDD
 import org.apache.spark.rdd.RDD.rddToPairRDDFunctions
-
 import it.unimi.dsi.fastutil.longs.Long2DoubleOpenHashMap
+import org.apache.spark.mllib.linalg.DenseVector
+import org.apache.spark.mllib.linalg.SparseVector
+import scala.collection.mutable.MutableList
+import test.Utils
+
+
+
 
 object WideDecisionTree {
 
@@ -28,37 +33,85 @@ object WideDecisionTree {
     if (total == 0.0) 0.0 else 1 - counts.map(s => sqr(s / total)).sum
   }
 
-  def findSplit(currentSet: Array[Int], labels: Array[Int])(t: (Vector, Long)): (Double, Double, Long, Int, Int, Array[Int], Array[Int]) = {
-
-    val v = t._1.toArray
-    val index: Long = t._2
-    val labelsCount = labels.max + 1
-
-    val labelCounts = Array.fill(labelsCount)(0)
+  def giniImpurity(currentSet: Array[Int], labels: Array[Int], labelCount:Int):(Double,Int) = {
+    val labelCounts = Array.fill(labelCount)(0)
     currentSet.foreach(i => labelCounts(labels(i)) += 1)
-    val totalGini = giniImprity(labelCounts)
+    (giniImprity(labelCounts), labelCounts.zipWithIndex.max._2)
+  }
+  
+  class SparseVectorSequentialAccessor(val sv:SparseVector) {
+    var index = 0
+    def apply(i:Int):Double = {
+      while(index < sv.indices.length && sv.indices(index) < i) { 
+        index+=1
+      }
+      if (index < sv.indices.length && sv.indices(index) ==i ) sv.values(index) else 0.0  
+    }
+    
+  }
+  
+  def mergeForEeach(currentSet: Array[Int], v:Vector)(f:(Int,Double)=>Unit)  {
+    // assumes currentSet is sorted
+    // for dense vector
+    v match {
+      case dv:DenseVector => currentSet.foreach(i=> f(i,dv.values(i))) 
+      case sv:SparseVector => {
[…]

### 2015-12-09 variantspark/src/main/scala/au/csiro/obr17q/variantspark/workflow/TestApp.scala

@@ -3,6 +3,8 @@ package au.csiro.obr17q.variantspark.workflow
 object TestApp {
   def main(args:Array[String])  {
     println(Range(0,2).size)
-    println(0 until 2)
+
+    val result = List(1,2,3,4).toSeq.foldLeft(List[Int](0))((l,v) => (v + l.head) :: l ).reverse
+    println(result)
   }
 }

### 2015-12-09 run-sparse-wide-rf.sh
Fixed missing import and updated random forest running script

@@ -12,4 +12,5 @@ spark-submit --class au.csiro.obr17q.variantspark.SparseWideForest\
       --driver-memory 8G \
       variantspark/target/variantspark-0.0.1-SNAPSHOT-all.jar \
       $input_file \
-      /flush2/szu004/variant/forrest.csv
+      /flush2/szu004/variant/forrest.csv \
+      100

### 2015-12-09 variantspark/src/main/scala/au/csiro/obr17q/variantspark/algo/WideDecisionTree.scala

@@ -10,7 +10,6 @@ import it.unimi.dsi.fastutil.longs.Long2DoubleOpenHashMap
 import org.apache.spark.mllib.linalg.DenseVector
 import org.apache.spark.mllib.linalg.SparseVector
 import scala.collection.mutable.MutableList
-import test.Utils
 
 
 

### 2015-12-22 doc/publications/plos16/plos16.tex
Started intro

% Template for PLoS
% Version 3.1 February 2015
%
% To compile to pdf, run:
% latex plos.template
% bibtex plos.template
% latex plos.template
% latex plos.template
% dvipdf plos.template
%
% % % % % % % % % % % % % % % % % % % % % %
%
% -- IMPORTANT NOTE
%
% This template contains comments intended 
% to minimize problems and delays during our production 
% process. Please follow the template instructions
% whenever possible.
%
% % % % % % % % % % % % % % % % % % % % % % % 
%
% Once your paper is accepted for publication, 
% PLEASE REMOVE ALL TRACKED CHANGES in this file and leave only
% the final text of your manuscript.
%
% There are no restrictions on package use within the LaTeX files except that 
% no packages listed in the template may be deleted.
%
% Please do not include colors or graphics in the text.
%
% Please do not create a heading level below \subsection. For 3rd level headings, use \paragraph{}.
%
% % % % % % % % % % % % % % % % % % % % % % %
%
% -- FIGURES AND TABLES
%
% Please include tables/figure captions directly after the paragraph where they are first cited in the text.
%
% DO NOT INCLUDE GRAPHICS IN YOUR MANUSCRIPT
% - Figures should be uploaded separately from your manuscript file. 
% - Figures generated using LaTeX should be extracted and removed from the PDF before submission. 
% - Figures containing multiple panels/subfigures must be combined into one image file before submission.
% For figure citations, please use "Fig." instead of "Figure".
% See http://www.plosone.org/static/figureGuidelines for PLOS figure guidelines.
%
% Tables should be cell-based and may not contain:
% - tabs/spacing/line breaks within cells to alter layout or alignment
% - vertically-merged cells (no tabular environments within tabular environments, do not use \multirow)
% - colors, shading, or graphic objects
% See http://www.plosone.org/static/figureGuidelines#tables for table guidelines.
%
% For tables that exceed the width of the text column, use the adjustwidth environment as illustrated in the example table in text below.
%
% % % % % % % % % % % % % % % % % % % % % % % %
%
% -- EQUATIONS, MATH SYMBOLS, SUBSCRIPTS, AND SUPERSCRIPTS
%
% IMPORTANT
% Below are a few tips to help format your equations and other special characters according to our specifications. For more tips to help reduce the possibility of formatting errors during conversion, please see our LaTeX guidelines at http://www.plosone.org/static/latexGuidelines
%
% Please be sure to include all portions of an equation in the math environment.
%
% Do not include text that is not math in the math environment. For example, CO2 will be CO\textsubscript{2}.
%
% Please add line breaks to long display equations when possible in order to fit size of the column. 
%
% For inline equations, please do not include punctuation (commas, etc) within the math environment unless this is part of the equation.
%
% % % % % % % % % % % % % % % % % % % % % % % % 
%
% Please contact latex@plos.org with any questions.
%
% % % % % % % % % % % % % % % % % % % % % % % %

\documentclass[10pt,letterpaper]{article}
\usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}

% Use adjustwidth environment to exceed column width (see example table in text)
\usepackage{changepage}

% Use Unicode characters when possible
\usepackage[utf8]{inputenc}

% textcomp package and marvosym package for additional characters
\usepackage{textcomp,marvosym}

% fixltx2e package for \textsubscript
\usepackage{fixltx2e}

% amsmath and amssymb packages, useful for mathematical formulas and symbols
\usepackage{amsmath,amssymb}

% cite package, to clean up citations in the main text. Do not remove.
\usepackage{cite}

% Use nameref to cite supporting information files (see Supporting Information section for more info)
\usepackage{nameref,hyperref}

% line numbers
\usepackage[right]{lineno}

% ligatures disabled
\usepackage{microtype}
\DisableLigatures[f]{encoding = *, family = * }

% rotating package for sideways tables
\usepackage{rotating}

% Remove comment for double spacing
%\usepackage{setspace} 
%\doublespacing

% Text layout
\raggedright
\setlength{\parindent}{0.5cm}
\textwidth 5.25in 
\textheight 8.75in

% Bold the 'Figure #' in the caption and separate it from the title/caption with a period
% Captions will be left justified
\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off]{caption}

% Use the PLoS provided BiBTeX style
\bibliographystyle{plos2015}

% Remove brackets from numbering in List of References
\makeatletter
\renewcommand{\@biblabel}[1]{\quad#1.}
\makeatother

% Leave date blank
\date{}

% Header and Footer with logo
\usepackage{lastpage,fancyhdr,graphicx}
\usepackage{epstopdf}
\pagestyle{myheadings}
\pagestyle{fancy}
\fancyhf{}
\lhead{\includegraphics[width=2.0in]{PLOS-submission.eps}}
\rfoot{\thepage/\pageref{LastPage}}
\renewcommand{\footrule}{\hrule height 2pt \vspace{2mm}}
\fancyheadoffset[L]{2.25in}
\fancyfootoffset[L]{2.25in}
\lfoot{\sf PLOS}

%% Include all macros below

\newcommand{\lorem}{{\bf LOREM}}
\newcommand{\ipsum}{{\bf IPSUM}}

%% END MACROS SECTION

\begin{document}
\vspace*{0.35in}

% Title must be 250 characters or less.
% Please capitalize all terms in the title except conjunctions, prepositions, and articles.
\begin{flushleft}
{\Large
\textbf\newline{Article Title - Submission to PLOS Journals}
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Name1 Surname\textsuperscript{1,\Yinyang},
Name2 Surname\textsuperscript{2,\Yinyang},
Name3 Surname\textsuperscript{2,\textcurrency a},
Name4 Surname\textsuperscript{2,\ddag},
Name5 Surname\textsuperscript{2,\ddag},
Name6 Surname\textsuperscript{2},
Name7 Surname\textsuperscript{3,*},
with the Lorem Ipsum Consortium\textsuperscript{\textpilcrow}
\\
\bigskip
\bf{1} Affiliation Dept/Program/Center, Institution Name, City, State, Country
\\
\bf{2} Affiliation Dept/Program/Center, Institution Name, City, State, Country
\\
\bf{3} Affiliation Dept/Program/Center, Institution Name, City, State, Country
\\
\bigskip

% Insert additional author notes using the symbols described below. Insert symbol callouts after author names as necessary.
% 
% Remove or comment out the author notes below if they aren't used.
%
% Primary Equal Contribution Note
\Yinyang These authors contributed equally to this work.

% Additional Equal Contribution Note
% Also use this double-dagger symbol for special authorship notes, such as senior authorship.
\ddag These authors also contributed equally to this work.

% Current address notes
\textcurrency a Insert current address of first author with an address update
% \textcurrency b Insert current address of second author with an address update
% \textcurrency c Insert current address of third author with an address update

% Deceased author note
\dag Deceased

% Group/Consortium Author Note
\textpilcrow Membership list can be found in the Acknowledgments section.

% Use the asterisk to denote corresponding authorship and provide email address in note below.
* CorrespondingAuthor@institute.edu

\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}
Lorem ipsum dolor sit amet, consectetur adipiscing elit. Curabitur eget porta erat. Morbi consectetur est vel gravida pretium. Suspendisse ut dui eu ante cursus gravida non sed sem. Nullam sapien tellus, commodo id velit id, eleifend volutpat quam. Phasellus mauris velit, dapibus finibus elementum vel, pulvinar non tellus. Nunc pellentesque pretium diam, quis maximus dolor faucibus id. Nunc convallis sodales ante, ut ullamcorper est egestas vitae. Nam sit amet enim ultrices, ultrices elit pulvinar, volutpat risus.

% Please keep the Author Summary between 150 and 200 words
% Use first person. PLOS ONE authors please skip this step. 
% Author Summary not valid for PLOS ONE submissions.   
\section*{Author Summary}
Lorem ipsum dolor sit amet, consectetur adipiscing elit. Curabitur eget porta erat. Morbi consectetur est vel gravida pretium. Suspendisse ut dui eu ante cursus gravida non sed sem. Nullam sapien tellus, commodo id velit id, eleifend volutpat quam. Phasellus mauris velit, dapibus finibus elementum vel, pulvinar non tellus. Nunc pellentesque pretium diam, quis maximus dolor faucibus id. Nunc convallis sodales ante, ut ullamcorper est egestas vitae. Nam sit amet enim ultrices, ultrices elit pulvinar, volutpat risus.

\linenumbers

\section*{Introduction}

Important because:

Computational tools such as Apache Spark provide a framework for designing and implementing complex analyses.

In addition of Spark ML, provides popular machine learning libraries. It includes supervised approaches such as Decision trees, and the ensemble Random Forests, as well as unsupervised approaches such as k-means clustering.

We previously demonstrated the versatility and scalability of Spark by developing VariantSpark, a framework to allow users to apply Spark and Spark ML's algorithms to VCF files.

Although we successfully clustered the 2500 individuals from the 1000 Genomes Project phase 3 dataset based variants from the 22 autosomes, this task approached the limits of our cluster.

Although our VCF pre-processing phase succeeds with a minimal resource requirement (3GB per executor) the memory required for k-means clustering skyrockets, requiring 24GB per executor to cluster the entire genome.

The reason for this can be attributed to the "curse of dimensionality". Spark ML is designed to process large numbers of samples rather than a large number of dimensions. As a result, data is partitioned along sample boundaries, with entire samples distributed to worker nodes.
As the sample size grows, Spark can simple schedule more jobs to run to cope with the additional samples.

However with the 1000 Genomes Project data, the dimensionality far exceeds the number of samples. This limits the number of jobs that Spark can create, and instead forces it to create fewer, larger jobs.

Lorem ipsum dolor sit~\cite{bib1} amet, consectetur adipiscing elit. Curabitur eget porta erat. Morbi consectetur est vel gravida pretium. Suspendisse ut dui eu ante cursus gravida non sed sem. Nullam Eq.~(\ref{eq:schemeP}) sapien tellus, commodo id velit id, eleifend volutpat quam. Phasellus mauris velit, dapibus finibus elementum vel, pulvinar non tellus. Nunc pellentesque pretium diam, quis maximus dolor faucibus id.~\cite{bib2} Nunc convallis sodales ante, ut ullamcorper est egestas vitae. Nam sit amet enim ultrices, ultrices elit pulvinar, volutpat risus.

\begin{equation}\label{eq:schemeP} 
D_{coll} = \frac{D_f+\frac{[S]^2}{K_D S_T} D_S} {1+\frac{[S]^2}{K_D S_T}}, 
D_{sm} = \frac{D_f+ \frac{[S]}{K_D} D_S}{1+\frac{[S]}{K_D}},
\end{equation}

% You may title this section "Methods" or "Models". 
% "Models" is not a valid title for PLoS ONE authors. However, PLoS ONE
% authors may use "Analysis" 
\section*{Materials and Methods}
\subsection*{Etiam eget sapien nibh.}

% For figure citations, please use "Fig." instead of "Figure".
Nulla mi mi, Fig.~\ref{fig1} venenatis sed ipsum varius, volutpat euismod diam. Proin rutrum vel massa non gravida. Quisque tempor sem et dignissim rutrum. Lorem ipsum dolor sit amet, consectetur adipiscing elit. Morbi at justo vitae nulla elementum commodo eu id massa. In vitae diam ac augue semper tincidunt eu ut eros. Fusce fringilla erat porttitor lectus cursus, \nameref{S1_Video} vel sagittis arcu lobortis. Aliquam in enim semper, aliquam massa id, cursus neque. Praesent faucibus semper libero.

\begin{figure}[h]
\caption{{\bf Figure Title first bold sentence Nulla mi mi, venenatis sed ipsum varius, volutpat euismod diam.}
Figure Caption Proin rutrum vel massa non gravida. Quisque tempor sem et dignissim rutrum. A: Lorem ipsum dolor sit amet. B: Consectetur adipiscing elit.}
\label{fig1}
\end{figure}

\begin{enumerate}
\item{react}
\item{diffuse free particles}
\item{increment time by dt and go to 1}
\end{enumerate}

% Results and Discussion can be combined.
\section*{Results}
Nulla mi mi, venenatis sed ipsum varius, Table~\ref{table1} volutpat euismod diam. Proin rutrum vel massa non gravida. Quisque tempor sem et dignissim rutrum. Lorem ipsum dolor sit amet, consectetur adipiscing elit. Morbi at justo vitae nulla elementum commodo eu id massa. In vitae diam ac augue semper tincidunt eu ut eros. Fusce fringilla erat porttitor lectus cursus, vel sagittis arcu lobortis. Aliquam in enim semper, aliquam massa id, cursus neque. Praesent faucibus semper libero.

\begin{table}[!ht]
\begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
\caption{
{\bf Table caption Nulla mi mi, venenatis sed ipsum varius, volutpat euismod diam.}}
\begin{tabular}{|l|l|l|l|l|l|l|l|}
\hline
\multicolumn{4}{|l|}{\bf Heading1} & \multicolumn{4}{|l|}{\bf Heading2}\\ \hline
$cell1 row1$ & cell2 row 1 & cell3 row 1 & cell4 row 1 & cell5 row 1 & cell6 row 1 & cell7 row 1 & cell8 row 1\\ \hline
$cell1 row2$ & cell2 row 2 & cell3 row 2 & cell4 row 2 & cell5 row 2 & cell6 row 2 & cell7 row 2 & cell8 row 2\\ \hline
$cell1 row3$ & cell2 row 3 & cell3 row 3 & cell4 row 3 & cell5 row 3 & cell6 row 3 & cell7 row 3 & cell8 row 3\\ \hline
\end{tabular}
\begin{flushleft} Table notes Phasellus venenatis, tortor nec vestibulum mattis, massa tortor interdum felis, nec pellentesque metus tortor nec nisl. Ut ornare mauris tellus, vel dapibus arcu suscipit sed.
\end{flushleft}
\label{table1}
\end{adjustwidth}
\end{table}

\subsection*{\lorem\ and \ipsum\ Nunc blandit a tortor.}

Maecenas convallis mauris sit amet sem ultrices gravida. Etiam eget sapien nibh. Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula. Curabitur fringilla pulvinar lectus consectetur pellentesque. Quisque augue sem, tincidunt sit amet feugiat eget, ullamcorper sed velit. Sed non aliquet felis. Lorem ipsum dolor sit amet, consectetur adipiscing elit. Mauris commodo justo ac dui pretium imperdiet. Sed suscipit iaculis mi at feugiat.

\subsection*{Sed ac quam id nisi malesuada congue.}

Nulla mi mi, venenatis sed ipsum varius, volutpat euismod diam. Proin rutrum vel massa non gravida. Quisque tempor sem et dignissim rutrum. Lorem ipsum dolor sit amet, consectetur adipiscing elit. Morbi at justo vitae nulla elementum commodo eu id massa. In vitae diam ac augue semper tincidunt eu ut eros. Fusce fringilla erat porttitor lectus cursus, vel sagittis arcu lobortis. Aliquam in enim semper, aliquam massa id, cursus neque. Praesent faucibus semper libero.

% Please do not create a heading level below \subsection. For 3rd level headings, use \paragraph{}. 
\subsection*{Subsection 1}
Nulla mi mi, venenatis sed ipsum varius, volutpat euismod diam. Proin rutrum vel massa non gravida. Quisque tempor sem et dignissim rutrum. Lorem ipsum dolor sit amet, consectetur adipiscing elit. Morbi at justo vitae nulla elementum commodo eu id massa. In vitae diam ac augue semper tincidunt eu ut eros. Fusce fringilla erat porttitor lectus cursus, vel sagittis arcu lobortis. Aliquam in enim semper, aliquam massa id, cursus neque. Praesent faucibus semper libero.

\subsection*{Subsection 2}
\paragraph{3rd Level Heading.} Nulla mi mi, venenatis sed ipsum varius, volutpat euismod diam. Proin rutrum vel massa non gravida. Quisque tempor sem et dignissim rutrum. Lorem ipsum dolor sit amet, consectetur adipiscing elit. Morbi at justo vitae nulla elementum commodo eu id massa. In vitae diam ac augue semper tincidunt eu ut eros. Fusce fringilla erat porttitor lectus cursus, vel sagittis arcu lobortis. Aliquam in enim semper, aliquam massa id, cursus neque. Praesent faucibus semper libero.

\section*{Discussion}
Nulla mi mi, venenatis sed ipsum varius, Table~\ref{table1} volutpat euismod diam. Proin rutrum vel massa non gravida. Quisque tempor sem et dignissim rutrum. Lorem ipsum dolor sit amet, consectetur adipiscing elit. Morbi at justo vitae nulla elementum commodo eu id massa. In vitae diam ac augue semper tincidunt eu ut eros. Fusce fringilla erat porttitor lectus cursus, vel sagittis arcu lobortis. Aliquam in enim semper, aliquam massa id, cursus neque. Praesent faucibus semper libero.

\subsection*{\lorem\ and \ipsum\ Nunc blandit a tortor.}

CO\textsubscript{2} Maecenas convallis mauris sit amet sem ultrices gravida. Etiam eget sapien nibh. Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula. Curabitur fringilla pulvinar lectus consectetur pellentesque. Quisque augue sem, tincidunt sit amet feugiat eget, ullamcorper sed velit.

Sed non aliquet felis. Lorem ipsum dolor sit amet, consectetur adipiscing elit. Mauris commodo justo ac dui pretium imperdiet. Sed suscipit iaculis mi at feugiat. Ut neque ipsum, luctus id lacus ut, laoreet scelerisque urna. Phasellus venenatis, tortor nec vestibulum mattis, massa tortor interdum felis, nec pellentesque metus tortor nec nisl. Ut ornare mauris tellus, vel dapibus arcu suscipit sed. Nam condimentum sem eget mollis euismod. Nullam dui urna, gravida venenatis dui et, tincidunt sodales ex. Nunc est dui, sodales sed mauris nec, auctor sagittis leo. Aliquam tincidunt, ex in facilisis elementum, libero lectus luctus est, non vulputate nisl augue at dolor. For more information, see \nameref{S1_Text}.

\section*{Supporting Information}

% Include only the SI item label in the subsection heading. Use the \nameref{label} command to cite SI items in the text.
\subsection*{S1 Video}
\label{S1_Video}
{\bf Bold the first sentence.}  Maecenas convallis mauris sit amet sem ultrices gravida. Etiam eget sapien nibh. Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula. Curabitur fringilla pulvinar lectus consectetur pellentesque.

\subsection*{S1 Text}
\label{S1_Text}
{\bf Lorem Ipsum.} Maecenas convallis mauris sit amet sem ultrices gravida. Etiam eget sapien nibh. Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula. Curabitur fringilla pulvinar lectus consectetur pellentesque.

\subsection*{S1 Fig}
\label{S1_Fig}
{\bf Lorem Ipsum.} Maecenas convallis mauris sit amet sem ultrices gravida. Etiam eget sapien nibh. Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula. Curabitur fringilla pulvinar lectus consectetur pellentesque.

\subsection*{S2 Fig}
\label{S2_Fig}
{\bf Lorem Ipsum.} Maecenas convallis mauris sit amet sem ultrices gravida. Etiam eget sapien nibh. Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula. Curabitur fringilla pulvinar lectus consectetur pellentesque.

\subsection*{S1 Table}
\label{S1_Table}
{\bf Lorem Ipsum.} Maecenas convallis mauris sit amet sem ultrices gravida. Etiam eget sapien nibh. Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula. Curabitur fringilla pulvinar lectus consectetur pellentesque.

\section*{Acknowledgments}
Cras egestas velit mauris, eu mollis turpis pellentesque sit amet. Interdum et malesuada fames ac ante ipsum primis in faucibus. Nam id pretium nisi. Sed ac quam id nisi malesuada congue. Sed interdum aliquet augue, at pellentesque quam rhoncus vitae.

\nolinenumbers

%\section*{References}
% Either type in your references using
% \begin{thebibliography}{}
% \bibitem{}
% Text
% \end{thebibliography}
%
% OR
%
% Compile your BiBTeX database using our plos2015.bst
% style file and paste the contents of your .bbl file
% here.
% 
\begin{thebibliography}{10}
\bibitem{bib1}
Devaraju P, Gulati R, Antony PT, Mithun CB, Negi VS. Susceptibility to SLE in South Indian Tamils may be influenced by genetic selection pressure on TLR2 and TLR9 genes. Mol Immunol. 2014 Nov 22. pii: S0161-5890(14)00313-7. doi: 10.1016/j.molimm.2014.11.005

\bibitem{bib2}
Huynen MMTE, Martens P, Hilderlink HBM. The health impacts of globalisation: a conceptual framework. Global Health. 2005;1: 14. Available: http://www.globalizationandhealth.com/content/1/1/14.

\end{thebibliography}

\end{document}

### 2016-01-14 vcf-clusterer/src/au/csiro/GenotypeWritable.java
Getting up to date

@@ -6,7 +6,6 @@ import java.io.IOException;
 
 import org.apache.hadoop.io.DoubleWritable;
 import org.apache.hadoop.io.IntWritable;
-import org.apache.hadoop.io.LongWritable;
 import org.apache.hadoop.io.WritableComparable;
 
 /**

### 2016-01-14 vcf-clusterer/src/au/csiro/GetResults.java
Getting up to date

@@ -13,27 +13,35 @@ public class GetResults extends Configured implements Tool {
 		Configuration conf = this.getConf();
 		GenericOptionsParser optionParser = new GenericOptionsParser(conf, args);
 		String[] remainingArgs = optionParser.getRemainingArgs();
+		String phase = "phase1";
+		String sample_names = "data/phase3.txt";
+		String job_id = ".";
+		
 		int mode = 0;
+		
+		if (remainingArgs.length > 0) {
+			job_id = remainingArgs[0];
+		}
+		
 		if (remainingArgs.length > 1) {
 			mode = Integer.parseInt(remainingArgs[1]);
 		}
-		
-		String sample_names = "";
 		if (remainingArgs.length > 2) {
-			sample_names = "data/" + remainingArgs[2];
+			phase = remainingArgs[2];
+			sample_names = "data/" + phase;
 		}	
 		
 		//Writes clusters with their associated samples to "resultFileCluster.txt"
 		switch (mode) {
 		case 1:
-			sequenceToText(conf, remainingArgs[0], sample_names);
+			sequenceToText(conf, job_id, sample_names);
 			break;
 		case 2:
-			textToRandArray();
+			textToRandArray(conf, job_id, phase);
 			break;
 		default:
-			sequenceToText(conf, remainingArgs[0], sample_names);
-			textToRandArray();
+			sequenceToText(conf, job_id, sample_names);
+			textToRandArray(conf, job_id, phase);
 			break;		
 		}
 
@@ -64,13 +72,13 @@ public class GetResults extends Configured implements Tool {
 		of.fileRead(conf, input, sample_names);
 	}
 	
-	private static void textToRandArray() throws IOException {
+	private static void textToRandArray(Configuration conf, String jobId, String phase) throws IOException {
 		System.out.println("Printing arrays for ");
 		System.out.println("\n\n\n");
 		
 		pedigreeMatcher p = new pedigreeMatcher("data/integrated_call_samples.20130502.ALL.ped");
-		p.findMatches("resultFileCluster.txt");
-		p.adjRandIndex("resultFileCluster.txt");
+		p.findMatches(conf, jobId);
+		p.adjRandIndex(conf, jobId, phase);
 		
 		System.out.println("from sklearn import metrics");
 		System.out.println("metrics.adjusted_rand_score(truth, clust)");

### 2016-01-14 vcf-clusterer/src/au/csiro/IndividualKeyPartitioner.java
Getting up to date

@@ -2,7 +2,6 @@ package au.csiro;
 
 import org.apache.hadoop.io.DoubleWritable;
 import org.apache.hadoop.io.IntWritable;
-import org.apache.hadoop.io.Text;
 import org.apache.hadoop.mapreduce.Partitioner;
 import org.apache.hadoop.mapreduce.lib.partition.HashPartitioner;
  

### 2016-01-14 vcf-clusterer/src/au/csiro/KmeansClusterer.java
Getting up to date

@@ -23,29 +23,33 @@ public class KmeansClusterer extends Configured implements Tool {
 	public KmeansClusterer() {}
 	
 	// Some variables     
-	public static String INPUT_DIRECTORY;
-	public static String INTERMEDIATE_DIRECTORY;
-	public static String OUTPUT_DIRECTORY;
-	public static int k;
-	public static int iter;
+	public static String INPUT_DIRECTORY = "transposed";
+	public static String INTERMEDIATE_DIRECTORY ="clusters";
+	public static String OUTPUT_DIRECTORY ="output";
+	public static int k = 7;
+	public static int iter = 20;
 	
 	public int run(String[] args) throws Exception {
 		Configuration conf = this.getConf();
 		GenericOptionsParser optionParser = new GenericOptionsParser(conf, args);
 		String[] remainingArgs = optionParser.getRemainingArgs();
 		FileSystem fs = FileSystem.get(conf);
-		INPUT_DIRECTORY = remainingArgs[0];	
-		INTERMEDIATE_DIRECTORY = remainingArgs[1];		
-		OUTPUT_DIRECTORY = remainingArgs[2];	
-		k = Integer.parseInt(remainingArgs[3]);
-		iter = Integer.parseInt(remainingArgs[6]);
+		if(remainingArgs.length > 0){
+			INPUT_DIRECTORY = remainingArgs[0] + "/transposed";
+			INTERMEDIATE_DIRECTORY = remainingArgs[0] + "/clusters";
+			OUTPUT_DIRECTORY = remainingArgs[0] +"/output";	
+			k = Integer.parseInt(remainingArgs[1]);
+			iter = Integer.parseInt(remainingArgs[4]);
+		}
 		conf.setInt("mapreduce.job.reduces", k);
 		conf.set("yarn.app.mapreduce.am.resource.mb", "4096");
-		conf.set("yarn.app.mapreduce.am.command-opts", "-Xmx3303M");
-		conf.set("mapreduce.map.memory.mb", "6144");
-		conf.set("mapreduce.reduce.memory.mb", "6144");
-		conf.set("mapreduce.map.java.opts", "-Xmx4954M");	
-		conf.set("mapreduce.reduce.java.opts", "-Xmx4954M");
+		conf.set("yarn.app.mapreduce.am.command-opts", "-Xmx3277M");
+		
+		conf.set("mapreduce.map.memory.mb","4096");
[…]

### 2016-01-14 vcf-clusterer/src/au/csiro/VcfToSeq.java
Getting up to date

@@ -91,7 +91,7 @@ public class VcfToSeq extends Configured implements Tool {
 
 			String[] itr = line.split("\\s+");
 			
-			int chrId = Integer.parseInt(itr[0]+"0000000")-10000000;
+			//int chrId = Integer.parseInt(itr[0]+"0000000")-10000000;
 			//location = chrId+ (int) (key.get()/10000);
 			location = (int) (key.get()/10000);
 

### 2016-01-14 vcf-clusterer/src/au/csiro/outputFormatter.java
Getting up to date

@@ -2,7 +2,7 @@ package au.csiro;
 
 import java.io.BufferedWriter;
 import java.io.File;
-import java.io.FileWriter;
+import java.io.OutputStreamWriter;
 import java.io.PrintWriter;
 import java.io.StringWriter;
 import java.util.LinkedHashMap;
@@ -15,26 +15,37 @@ import org.apache.hadoop.fs.FileSystem;
 import org.apache.hadoop.fs.Path;
 import org.apache.hadoop.io.IntWritable;
 import org.apache.hadoop.io.SequenceFile;
-import org.apache.mahout.clustering.Cluster;
+import org.apache.hadoop.io.SequenceFile.Reader;
 import org.apache.mahout.clustering.classify.WeightedPropertyVectorWritable;
-import org.apache.mahout.clustering.classify.WeightedVectorWritable;
 import org.apache.mahout.math.NamedVector;
 
 public class outputFormatter {
-	String outputFile = "resultFileCluster.txt";
 	LinkedHashMap<String,String> mpIdData = new LinkedHashMap<String, String>();
 	
-	public void fileRead(Configuration conf, String path, String namesFile) {
-		System.out.println("Writing summary to "+ outputFile + "...");
+	public void fileRead(Configuration conf, String jobId, String namesFile) {
+		
+		
+		
+		
+		
+		System.out.println("Writing summary to "+ jobId + "/resultFileCluster.txt...");
 		try {
 			FileSystem fs = FileSystem.get(conf);
-			Path inPath = new Path(path);
+			
+			Path pt = new Path(jobId + "/resultFileCluster.txt");
+			BufferedWriter out =
+					new BufferedWriter(new OutputStreamWriter(fs.create(pt, true)));
+
+			Path inPath = new Path(jobId + "/output/clusteredPoints");
 			FileStatus[] status = fs.listStatus(inPath);
 			
 			IntWritable key = new IntWritable();
 			WeightedPropertyVectorWritable value = new WeightedPropertyVectorWritable();
 			
-			String content = new Scanner(new File(namesFile)).useDelimiter("\\Z").next();
+			Scanner scan = new Scanner(new File(namesFile));
+			String content = scan.useDelimiter("\\Z").next();
+			scan.close();
+			
 			String[] sampleNames = content.split(",");
 			
 			System.out.println("Number of individuals: " + sampleNames.length);
@@ -47,7 +58,7 @@ public class outputFormatter {
 					continue;
[…]

### 2016-01-14 vcf-clusterer/src/au/csiro/pedigreeMatcher.java
Getting up to date

@@ -3,11 +3,16 @@ package au.csiro;
 import java.io.BufferedReader;
 import java.io.BufferedWriter;
 import java.io.FileReader;
-import java.io.FileWriter;
 import java.io.IOException;
+import java.io.InputStreamReader;
+import java.io.OutputStreamWriter;
+import java.util.Arrays;
 import java.util.HashMap;
-import java.util.Iterator;
-import java.util.Map;
+import java.util.List;
+
+import org.apache.hadoop.conf.Configuration;
+import org.apache.hadoop.fs.FileSystem;
+import org.apache.hadoop.fs.Path;
 
 public class pedigreeMatcher {
 	HashMap<String, String[]> hm = new HashMap<String, String[]>();
@@ -22,33 +27,40 @@ public class pedigreeMatcher {
 		}
 		ins.close();
 	}
-	
-	
-	
-	
-	
-	
-	
-	
-	//Why??
-	public void printPedigree() {
-		Iterator i = hm.entrySet().iterator();
-		while(i.hasNext()) {
-	         Map.Entry me = (Map.Entry)i.next();
-	         System.out.print(me.getKey() + ": ");
-	         for (String s:(String[])me.getValue()) {
-	        	 System.out.print(s+ " ");
-	         }
-	         System.out.print("\n");
-	      }
+
+	public enum Phase1Pops {
+		EAS("JPT", "CHB", "CHS"), SAS(), EUR("CEU", "TSI", "FIN", "GBR", "IBS"), 
+		AFR("YRI", "LWK", "ASW"), AMR("MXL", "PUR", "CLM");
+		private final List<String> values;
+		Phase1Pops(String ...values) {
+			this.values = Arrays.asList(values);
+		}
+	    public List<String> getValues() {
+	        return values;
+	    }
 	}
-	
-	public void findMatches(String clusteredPoints) throws IOException {
-		BufferedWriter w = new BufferedWriter(new FileWriter("populations.txt"));
-		BufferedReader ins =
-				new BufferedReader(new FileReader(clusteredPoints));
+	public enum Phase3Pops {
+		EAS("CHB", "JPT", "CHS", "CDX", "KHV"),	SAS("GIH", "PJL", "BEB", "STU", "ITU"),	EUR("CEU", "TSI", "FIN", "GBR", "IBS"), 
+		AFR("YRI", "LWK", "GWD", "MSL", "ESN", "ASW", "ACB"), AMR("MXL", "PUR", "CLM", "PEL");
[…]

### 2016-01-14 variantspark/src/main/scala/au/csiro/obr17q/variantspark/VcfSelection.scala
Feature selection updates

@@ -1,5 +1,6 @@
 package au.csiro.obr17q.variantspark
 
+import au.csiro.obr17q.variantspark.model.VcfParser
 import org.apache.spark.mllib.feature.ChiSqSelector
 import org.apache.spark.mllib.linalg.{Vector => MLVector, Vectors}
 import org.apache.spark.mllib.regression.LabeledPoint
@@ -28,7 +29,7 @@ object VcfSelection extends SparkApp {
     val IndividualMeta = sc.parallelize(new MetaDataParser(PopFiles, HeaderLines = 1, '\t', "", 0, 1)(SexCol = 3, extra1 = 2).returnMap())
 
     
-    val vcfObject = new VcfParser(VcfFiles, VariantCutoff, sc)
+    val vcfObject = new VcfParser(VcfFiles, VariantCutoff, IndividualMeta, sc)
 
     val NoOfAlleles = vcfObject.variantCount
 

### 2016-01-14 variantspark/src/main/scala/au/csiro/obr17q/variantspark/SparkApp.scala
setting for running locally

@@ -6,15 +6,16 @@ import org.apache.spark.{SparkConf, SparkContext}
 
 trait SparkApp {
   def defaultMasterUrl = "yarn-client"
-  //Logger.getLogger("org").setLevel(Level.OFF)
-  //Logger.getLogger("akka").setLevel(Level.OFF)
+  Logger.getLogger("org").setLevel(Level.OFF)
+  Logger.getLogger("akka").setLevel(Level.OFF)
   
   lazy val conf = new SparkConf()
-  //.set("spark.shuffle.memoryFraction", "0.2")
-  //.set("spark.storage.memoryFraction", "0.4")
-  //.set("spark.yarn.executor.memoryOverhead", "2048")
-  .set("spark.driver.maxResultSize","2048")
-  //.set("spark.default.parallelism", "256")
+    //.setMaster("local")
+    //.set("spark.shuffle.memoryFraction", "0.2")
+    //.set("spark.storage.memoryFraction", "0.4")
+    //.set("spark.yarn.executor.memoryOverhead", "2048")
+    .set("spark.driver.maxResultSize","2048")
+    //.set("spark.default.parallelism", "256")
   lazy val masterUrl =  if (System.getenv("MASTER") != null) System.getenv("MASTER") else "yarn-client"
   lazy val sc = { println("MASTER: " + masterUrl) ; new SparkContext(conf)}
   lazy val sqlContext = new org.apache.spark.sql.SQLContext(sc)

### 2016-01-14 variantspark/src/main/scala/au/csiro/obr17q/variantspark/VcfLogRegression.scala
logistic regression

@@ -3,6 +3,7 @@ package au.csiro.obr17q.variantspark
 import java.util.Date
 
 import au.csiro.obr17q.variantspark.CommonFunctions._
+import au.csiro.obr17q.variantspark.model.VcfParser
 import org.apache.spark.mllib.classification.LogisticRegressionWithLBFGS
 import org.apache.spark.mllib.evaluation.MulticlassMetrics
 import org.apache.spark.mllib.linalg.{Vectors}
@@ -31,7 +32,7 @@ object VcfLogRegression extends SparkApp {
     //val IndividualMeta = sc.parallelize(new MetaDataParser(PopFiles, HeaderLines = 1, '\t', "", 0, 2 )(SexCol = 3).returnMap())
 
     
-    val vcfObject = new VcfParser(args(0), VariantCutoff, sc)
+    val vcfObject = new VcfParser(args(0), VariantCutoff, IndividualMeta, sc)
 
     val NoOfAlleles = vcfObject.variantCount
 

### 2016-01-14 variantspark/src/main/scala/au/csiro/obr17q/variantspark/VcfForest.scala
Refactoring the forest

@@ -1,121 +1,80 @@
 package au.csiro.obr17q.variantspark
 
-import org.apache.spark.mllib.linalg._
-import org.apache.spark.mllib.regression.LabeledPoint
+import au.csiro.obr17q.variantspark.model.CsvParser
 import org.apache.spark.ml.classification.RandomForestClassifier
 import org.apache.spark.ml.classification.RandomForestClassificationModel
+import org.apache.spark.ml.regression.RandomForestRegressionModel
 import org.apache.spark.ml.Pipeline
 import org.apache.spark.ml.PipelineModel
+import org.apache.spark.ml.regression.RandomForestRegressor
 import org.apache.spark.ml.tuning.{CrossValidator, ParamGridBuilder}
 import org.apache.spark.ml.evaluation.MulticlassClassificationEvaluator
-import scala.io.Source
 import org.apache.spark.sql.{DataFrame, Row}
 import org.apache.spark.ml.feature.{VectorIndexer, StringIndexer}
 import org.apache.spark.rdd.RDD
+import java.io._
 /**
  * @author obr17q
  */
 object VcfForest extends SparkApp {
   conf.setAppName("VCF foresting")
 
-  case class Record(individual: String, sampleType: String, bmi: Double, preLabel: String, features: Vector)
+  def main(args: Array[String]) {
 
-  
-  def main(args:Array[String]) {
-    
-    //val args = Array("data/smALL.vcf","5","5","100","auto","0.5")
-    
-    val seed = 3262362
-    val VcfFiles = args(0)
-    val NumTrees = args(1).toInt
+    val defaults = Array("/Users/obr17q/Desktop/coadread_mutations_bmi_filtered.csv","5","5","10","auto","0")
+
+
+    //val seed = 3262362
+    val VcfFiles = if (args.length > 0) args(0) else defaults(0)
+    val NumTrees = if (args.length > 1) args(1).toInt else defaults(1).toInt
     //val maxDepth = args(2).toInt
 
-    val FeatureSubsetStrategy = args(4)
-    val VariantCutoff = args(5).toInt
+    val FeatureSubsetStrategy = if (args.length > 4) args(4) else defaults(4)
+    val VariantCutoff = if (args.length > 5) args(5).toInt else defaults(5).toInt
 
 
-    val numFolds = 10
[…]

### 2016-01-14 variantspark/src/main/scala/au/csiro/obr17q/variantspark/VcfClustering.scala
refacting clustering

@@ -1,8 +1,13 @@
 package au.csiro.obr17q.variantspark
 
 import au.csiro.obr17q.variantspark.CommonFunctions._
-import org.apache.spark.mllib.clustering.KMeans
+import au.csiro.obr17q.variantspark.model.VcfParser
+import org.apache.spark.ml.Pipeline
+import org.apache.spark.ml.classification.RandomForestClassifier
+import org.apache.spark.ml.feature.StringIndexer
 import org.apache.spark.mllib.linalg.Vectors
+import org.apache.spark.rdd.RDD
+import org.apache.spark.ml.clustering.KMeans
 
 import scala.io.Source
 
@@ -45,12 +50,10 @@ object VcfClustering extends SparkApp {
     //val Populations = sc.parallelize(new PopulationMap(PopFiles, 1, ',', 0, 16 ).returnMap(IncludeGroups, ExcludeGroups))
 
     val PopFiles = Source.fromFile("data/ALL.panel").getLines()
-    val Populations = sc.parallelize(new MetaDataParser(PopFiles, 1, '\t', "NA", 0, 1 ).returnMap(IncludeGroups, ExcludeGroups))
-    val vcfParser = new VcfParser(VcfFiles, VariantCutoff, sc)
+    val IndividualMeta : RDD[IndividualMap] = sc.parallelize(new MetaDataParser(PopFiles, 1, '\t', "NA", 0, 1 ).returnMap(IncludeGroups, ExcludeGroups))
+    val vcfObject = new VcfParser(VcfFiles, VariantCutoff, IndividualMeta, sc)
 
-    val NoOfAlleles = vcfParser.variantCount
-
-    val FilteredAlleles = vcfParser.individualTuples
+    val FilteredAlleles = vcfObject.individualTuples
 
 
 
@@ -58,12 +61,12 @@ object VcfClustering extends SparkApp {
      * Vector of elements for each individual
      * Vector elements are zipped with the Individual ID
      */
-    val IndividualVariants = FilteredAlleles
-    .groupByKey //group by individual ID, i.e. get RDD of individuals
-    .join(Populations.map(_.toPops)) //filter out population groups you don't want
-    .map(h => (h._1, h._2._2, Vectors.sparse(NoOfAlleles, h._2._1.to[Seq] )))//.cache() //create sparse vectors
+    //val IndividualVariants = FilteredAlleles
+    //.groupByKey //group by individual ID, i.e. get RDD of individuals
[…]

### 2016-01-14 variantspark/src/main/scala/au/csiro/obr17q/variantspark/App.scala
What was this file doing in here?

@@ -1,15 +1,0 @@
-package au.csiro.obr17q.variantspark
-
-/**
- * @author ${user.name}
- */
-object App {
-  
-  def foo(x : Array[String]) = x.foldLeft("")((a,b) => a + b)
-  
-  def main(args : Array[String]) {
-    println( "Hello World!" )
-    println("concat arguments = " + foo(args))
-  }
-
-}

### 2016-01-14 variantspark/src/main/scala/au/csiro/obr17q/variantspark/VcfToFlatVariant.scala
remove spark sql import as is now in sparkapp

@@ -1,11 +1,8 @@
 package au.csiro.obr17q.variantspark
 
-import au.csiro.obr17q.variantspark.CommonFunctions._
-import org.apache.spark.mllib.clustering.KMeans
-import org.apache.spark.mllib.linalg.Vectors
-import scala.io.Source
 import org.apache.spark.rdd.RDD
 import au.csiro.obr17q.variantspark.model.FlatVariant
+import au.csiro.obr17q.variantspark.model.VcfParser
 
 object VcfToFlatVariant extends SparkApp {
   conf.setAppName("VCF cluster")
@@ -17,17 +14,16 @@ object VcfToFlatVariant extends SparkApp {
     val VcfFiles = args(0)
     val output = args(1)
     val VariantCutoff = args(2).toInt
-
+    val IndividualMeta: RDD[IndividualMap] = null
   
     //val PopFiles = Source.fromFile("data/PGPParticipantSurvey-20150831064509.csv").getLines()
     //val Populations = sc.parallelize(new PopulationMap(PopFiles, 1, ',', 0, 16 ).returnMap(IncludeGroups, ExcludeGroups))
-    val vcfParser = new VcfParser(VcfFiles, VariantCutoff, sc)
+    val vcfParser = new VcfParser(VcfFiles, VariantCutoff, IndividualMeta, sc)
     //val NoOfAlleles = vcfParser.variantCount
     val FilteredAlleles:RDD[FlatVariant] = vcfParser.individualVariants
 
-    val sqlContext = new org.apache.spark.sql.SQLContext(sc)
     import sqlContext.implicits._
 
-    FilteredAlleles.toDF().saveAsParquetFile(output)
+    FilteredAlleles.toDF().write.parquet(output)
   } 
 }

### 2016-01-14 variantspark/src/main/scala/au/csiro/obr17q/variantspark/FlatVariantToSparse.scala
remove sparksql imports

@@ -32,7 +32,6 @@ object FlatVariantToSparse extends SparkApp {
     val inputFiles = args(0)
     val output = args(1) 
     val chunks = args(2).toInt
-    val sqlContext = new org.apache.spark.sql.SQLContext(sc)
     import sqlContext.implicits._
     val flatVariandDF = sqlContext.parquetFile(inputFiles)    
     println(flatVariandDF.schema)

### 2016-01-14 variantspark/src/main/scala/au/csiro/obr17q/variantspark/FlatVariantToSparseByPosition.scala

@@ -38,9 +38,8 @@ object FlatVariantToSparseByPosition extends SparkApp {
     val inputFiles = args(0)
     val output = args(1) 
     val chunks = args(2).toInt
-    val sqlContext = new org.apache.spark.sql.SQLContext(sc)
     import sqlContext.implicits._
-    val flatVariandDF = sqlContext.parquetFile(inputFiles)    
+    val flatVariandDF = sqlContext.read.parquet(inputFiles)
     println(flatVariandDF.schema)
     val subjectIds = flatVariandDF.select("subjectId").distinct().collect()
  
@@ -57,7 +56,7 @@ object FlatVariantToSparseByPosition extends SparkApp {
       .mapValues(i => Vectors.sparse(subjectsSize,i.to[Seq]).toSparse)
       .map {case (locusId, sparseVector) =>
         LocusSparseVariant(locusId, sparseVector.size, sparseVector.indices, sparseVector.values)}
-      .toDF().saveAsParquetFile(output)  
+      .toDF().write.parquet(output)
       
     // and also commit the dictionary
     LoanUtils.withCloseable(new CSVWriter(new FileWriter(new File(output, "_index.csv")))) { cvsWriter =>

### 2016-01-14 variantspark/src/main/scala/au/csiro/obr17q/variantspark/SparkApp.scala

@@ -2,6 +2,7 @@ package au.csiro.obr17q.variantspark
 
 import org.apache.log4j.{Level, Logger}
 import org.apache.spark.{SparkConf, SparkContext}
+import org.apache.spark.sql.SQLContext
 //import org.bdgenomics.adam.rdd.ADAMContext
 
 trait SparkApp {
@@ -10,7 +11,7 @@ trait SparkApp {
   Logger.getLogger("akka").setLevel(Level.OFF)
   
   lazy val conf = new SparkConf()
-    //.setMaster("local")
+    .setMaster("local")
     //.set("spark.shuffle.memoryFraction", "0.2")
     //.set("spark.storage.memoryFraction", "0.4")
     //.set("spark.yarn.executor.memoryOverhead", "2048")
@@ -18,6 +19,6 @@ trait SparkApp {
     //.set("spark.default.parallelism", "256")
   lazy val masterUrl =  if (System.getenv("MASTER") != null) System.getenv("MASTER") else "yarn-client"
   lazy val sc = { println("MASTER: " + masterUrl) ; new SparkContext(conf)}
-  lazy val sqlContext = new org.apache.spark.sql.SQLContext(sc)
+  lazy val sqlContext = new SQLContext(sc)
   //lazy val ac = new ADAMContext(sc)
 }

### 2016-01-14 variantspark/src/main/scala/au/csiro/obr17q/variantspark/SparseClustering.scala

@@ -35,9 +35,7 @@ object SparseClustering extends SparkApp {
     val pops = Populations.map(_.toPops).collectAsMap();
     println(pops)
     val pops_br = sc.broadcast(pops);
-    val sqlContext = new org.apache.spark.sql.SQLContext(sc)
-    import sqlContext.implicits._
-    val sparseVariat = sqlContext.parquetFile(inputFiles)    
+    val sparseVariat = sqlContext.read.parquet(inputFiles)
     println(sparseVariat.schema)
 
     

### 2016-01-14 variantspark/src/main/scala/au/csiro/obr17q/variantspark/SparseDataExport.scala

@@ -25,9 +25,7 @@ object SparseDataExport extends SparkApp {
     val inputFiles = args(0)
     val output = args(1)
     
-    val sqlContext = new org.apache.spark.sql.SQLContext(sc)
-    import sqlContext.implicits._
-    val sparseVariat = sqlContext.parquetFile(inputFiles)    
+    val sparseVariat = sqlContext.read.parquet(inputFiles)
     println(sparseVariat.schema)
 
    val indexSubjectMap = LoanUtils.withCloseable(CSVReader.open(new FileReader(new File(inputFiles, "_index.csv")))){

### 2016-01-14 variantspark/src/main/scala/au/csiro/obr17q/variantspark/SparseWideAAT.scala

@@ -39,9 +39,7 @@ object SparseWideAAT extends SparkApp {
     val inputFiles = args(0)
     val output = args(1)
     
-    val sqlContext = new org.apache.spark.sql.SQLContext(sc)
-    import sqlContext.implicits._
-    val sparseVariat = sqlContext.parquetFile(inputFiles)    
+    val sparseVariat = sqlContext.read.parquet(inputFiles)
     println(sparseVariat.schema)
 
     

### 2016-01-14 variantspark/src/main/scala/au/csiro/obr17q/variantspark/SparseWideClustering.scala

@@ -25,9 +25,7 @@ object SparseWideClustering extends SparkApp {
     val iter = args(2).toInt
     val output = args(3)
     
-    val sqlContext = new org.apache.spark.sql.SQLContext(sc)
-    import sqlContext.implicits._
-    val sparseVariat = sqlContext.parquetFile(inputFiles)    
+    val sparseVariat = sqlContext.read.parquet(inputFiles)
     println(sparseVariat.schema)
 
     

### 2016-01-14 variantspark/src/main/scala/au/csiro/obr17q/variantspark/SparseWideForest.scala

@@ -31,10 +31,8 @@ object SparseWideForest extends SparkApp {
     val ntree = args(2).toInt
     val ntryfraction = if (args.length > 3) args(3).toDouble else Double.NaN
         
-    val sqlContext = new org.apache.spark.sql.SQLContext(sc)
-    import sqlContext.implicits._
-    val sparseVariat = sqlContext.parquetFile(inputFiles)    
-    println(sparseVariat.schema)
+    val sparseVariant = sqlContext.read.parquet(inputFiles)
+    println(sparseVariant.schema)
 
    val indexSubjectMap = LoanUtils.withCloseable(CSVReader.open(new FileReader(new File(inputFiles, "_index.csv")))){
       csvReader =>
@@ -72,7 +70,7 @@ object SparseWideForest extends SparkApp {
    
    
    val vectorData = 
-      sparseVariat.rdd      
+      sparseVariant.rdd
         .map{r=> 
           val (size, indexes,values) = (r.getInt(1),r.getSeq[Int](2).toArray, r.getSeq[Double](3))
           val filteredIndexs = MutableList[Int]()

### 2016-01-14 variantspark/src/main/scala/au/csiro/obr17q/variantspark/ValidateSparse.scala

@@ -31,9 +31,8 @@ object ValidateSparse extends SparkApp {
     //val PopFiles = Source.fromFile("data/PGPParticipantSurvey-20150831064509.csv").getLines()
     //val Populations = sc.parallelize(new PopulationMap(PopFiles, 1, ',', 0, 16 ).returnMap(IncludeGroups, ExcludeGroups))
 
-    val sqlContext = new org.apache.spark.sql.SQLContext(sc)
     import sqlContext.implicits._
-    val sparseVariat = sqlContext.parquetFile(inputFiles)    
+    val sparseVariat = sqlContext.read.parquet(inputFiles)
     println(sparseVariat.schema)
 
     

### 2016-01-14 variantspark/src/main/scala/au/csiro/obr17q/variantspark/VcfClustering.scala

@@ -1,11 +1,8 @@
 package au.csiro.obr17q.variantspark
 
-import au.csiro.obr17q.variantspark.CommonFunctions._
 import au.csiro.obr17q.variantspark.model.VcfParser
 import org.apache.spark.ml.Pipeline
-import org.apache.spark.ml.classification.RandomForestClassifier
 import org.apache.spark.ml.feature.StringIndexer
-import org.apache.spark.mllib.linalg.Vectors
 import org.apache.spark.rdd.RDD
 import org.apache.spark.ml.clustering.KMeans
 

### 2016-01-14 variantspark/src/main/scala/au/csiro/obr17q/variantspark/VcfForest.scala

@@ -39,7 +39,7 @@ object VcfForest extends SparkApp {
     val labelName = "bmi_cat"
     val maxDepth = Array(5, 10)
     //val maxDepth = Array(5)
-    val maxBins = Array(20, 40)
+    val maxBins = Array(20)
     //val maxBins = Array(10)
 
 
@@ -154,83 +154,33 @@ object VcfForest extends SparkApp {
 
 
 
-    //val Array(trainingData0, testData0) = data.randomSplit(Array(0.8, 0.2))
-    //val Array(trainingData1, testData1) = data.randomSplit(Array(0.8, 0.2))
-    //val Array(trainingData2, testData2) = data.randomSplit(Array(0.8, 0.2))
-    //val Array(trainingData3, testData3) = data.randomSplit(Array(0.8, 0.2))
-    //val Array(trainingData4, testData4) = data.randomSplit(Array(0.8, 0.2))
-    //val Array(trainingData5, testData5) = data.randomSplit(Array(0.8, 0.2))
-    //val Array(trainingData6, testData6) = data.randomSplit(Array(0.8, 0.2))
-    //val Array(trainingData7, testData7) = data.randomSplit(Array(0.8, 0.2))
-    //val Array(trainingData8, testData8) = data.randomSplit(Array(0.8, 0.2))
-    //val Array(trainingData9, testData9) = data.randomSplit(Array(0.8, 0.2))
+    //val Array(trainingData, testData) = data.randomSplit(Array(0.8, 0.2))
 
-    val filteredAlleleTuples0 = modelFit(data)
-    val filteredAlleleTuples1 = modelFit(data)
-    val filteredAlleleTuples2 = modelFit(data)
-    val filteredAlleleTuples3 = modelFit(data)
-    val filteredAlleleTuples4 = modelFit(data)
-    /*val filteredAlleleTuples5 = modelFit(data)
-    val filteredAlleleTuples6 = modelFit(data)
-    val filteredAlleleTuples7 = modelFit(data)
-    val filteredAlleleTuples8 = modelFit(data)
-    val filteredAlleleTuples9 = modelFit(data)
-    val filteredAlleleTuples10 = modelFit(data)
-    val filteredAlleleTuples11 = modelFit(data)
-    val filteredAlleleTuples12 = modelFit(data)
-    val filteredAlleleTuples13 = modelFit(data)
-    val filteredAlleleTuples14 = modelFit(data)
[…]

### 2016-01-14 variantspark/src/main/scala/au/csiro/obr17q/variantspark/VcfSparseForest.scala

@@ -60,9 +60,7 @@ object VcfSparseForest extends SparkApp {
       
     println(individualToSuperId)
       
-    val sqlContext = new org.apache.spark.sql.SQLContext(sc)
-    import sqlContext.implicits._
-    val sparseVariat = sqlContext.parquetFile(inputFiles)    
+    val sparseVariat = sqlContext.read.parquet(inputFiles)
     
     val br_individualToSuperId = sc.broadcast(individualToSuperId)
     

### 2016-01-14 variantspark/src/main/scala/au/csiro/obr17q/variantspark/model/CsvParser.scala

@@ -79,21 +79,20 @@ class CsvParser (val CsvFileNames: String, val VariantCutoff: Int, val sc: Spark
             features = Vectors.sparse(variantCount, p._2._1.to[Seq])))
     }.toDF
 
-
-    var printData = {
-      val variantCount = this.variantCount
-      val alleleTuples = this.alleleTuples
-      val IndividualMetaData = this.IndividualMetaData
-      CsvLineRDD
-        .map(line => ((line(10), line(7)), 1))
-        .reduceByKey(_ + _)
-        .map(p => (p._1._2, (p._1._1, p._2.toDouble)))
-        .join(alleleTuples)
-        .map(p => (p._2._1._1, (p._2._2, p._2._1._2)))
-        .groupByKey
-        .join(IndividualMetaData)
-        .map(p => ( p._1, p._2._2._1, p._2._2._2, Vectors.sparse(variantCount, p._2._1.to[Seq] ).toArray))
-    }
+  var printData = {
+    val variantCount = this.variantCount
+    val alleleTuples = this.alleleTuples
+    val IndividualMetaData = this.IndividualMetaData
+    CsvLineRDD
+      .map(line => ((line(10), line(7)), 1))
+      .reduceByKey(_ + _)
+      .map(p => (p._1._2, (p._1._1, p._2.toDouble)))
+      .join(alleleTuples)
+      .map(p => (p._2._1._1, (p._2._2, p._2._1._2)))
+      .groupByKey
+      .join(IndividualMetaData)
+      .map(p => ( p._1, p._2._2._1, p._2._2._2, Vectors.sparse(variantCount, p._2._1.to[Seq] ).toArray))
+  }
 
 
 

### 2016-01-14 variantspark/src/main/scala/au/csiro/obr17q/variantspark/model/VcfParser.scala

@@ -7,13 +7,18 @@ import au.csiro.obr17q.variantspark.VcfForest._
 import org.apache.spark.SparkContext
 import org.apache.spark.rdd.RDD
 import org.apache.spark.mllib.linalg._
-import model.FlatVariant
 
 /**
  * @author obr17q
  */
 
-private case class VcfRecord(individual: String, sampleType: String, bmi: Double, preLabel: String, features: Vector)
+private case class VcfRecord(
+                              individual: String,
+                              sampleType: String,
+                              bmi: Double,
+                              preLabel: String,
+                              features: Vector
+                            )
 
 class VcfParser (val VcfFileNames: String, val VariantCutoff: Int, val IndividualMeta: RDD[IndividualMap], val sc: SparkContext) extends scala.Serializable {
 
@@ -102,7 +107,7 @@ class VcfParser (val VcfFileNames: String, val VariantCutoff: Int, val Individua
     .map(h => (h._2, h._1.toInt))
   }
 
-/*
+
   def individualTuples : RDD[(String, (Int, Double))] = {
     VcfLineRdd
       .map(h => (h._1, h._3))
@@ -111,12 +116,15 @@ class VcfParser (val VcfFileNames: String, val VariantCutoff: Int, val Individua
           .map( i => ( i._1, (h._1.toInt, i._2) ) )
         })
     }
-*/
-  
+
+
+
+
+
   def individualVariants : RDD[FlatVariant] = {
     VcfLineRdd
     .flatMap( (h) => {
-      h._2
+      h._3
       .map( i => FlatVariant( i._1, h._1.toInt, i._2))
     })
   }

### 2016-03-08 doc/publications/plos16/plos16.tex
Intro

@@
 
-Important because:
+In the past few years, whole genome sequence (WGS) data has rapidly been becoming an essential part in the clinical space, where traditional methods are failing to produce results.
 
-Computational tools such as Apache Spark provide a framework for designing and implementing complex analyses.
+WGS has proven itself in discovering rare variants [http://www.sciencedirect.com/science/article/pii/S2352396414000498] and diagnosing unknown genetic disorders [MATT MIGHT],
 
-In addition of Spark ML, provides popular machine learning libraries.
-It includes supervised approaches such as Decision trees, and the ensemble Random Forests, as well as unsupervised approaches such as k-means clustering.
+One challenge with analysing WGS data over, for example, traditional genome-wide association (GWA) data, is developing software and algorithms to deal with the huge amounts of data.
+Although the sample size of WGS data-sets may be comparable to GWA, the dimensionality of the data may be many times larger. For example, the 1000 Genome data consists of
+approximately 2500 individuals with up to 80 million variants. The uncompressed file sizes are close to a terabyte. Of course, it's not just genomic data that is rapidly growing. Many other
+types of data
 
-We previously demonstrated the versatility and scalability of Spark by developing VariantSpark, a framework to allow users to apply Spark and Spark ML's algorithms to VCF files.
+To manage Developed by Apache, Hadoop is one platform for dealing with large data sets is 
+Platforms Fortunately computational tools such as Apache Spark are available as a framework for iteratively dealing with big data.
 
-Although we successfully clustered the 2500 individuals from the 1000 Genomes Project phase 3 dataset based variants from the 22 autosomes, this task approached the limits of our cluster.
+Spark also includes the machine learning (ML) library, Spark ML, providing popular machine learning libraries.
+Spark ML includes supervised approaches such as decision trees and the ensemble method, random forests, as well as
+unsupervised approaches such as k-means clustering.
+We previously demonstrated the versatility and scalability of Spark by developing VariantSpark, a framework allowing users to easily analyse Variant Call Format (VCF) files using ML algorithms on the Spark framework.
+Using VariantSpark,
+we successfully built a k-means model on the 2500 individuals from the 1000 Genomes Project, based on their genomic data.
+We also investigated classifying the samples using Random Forests.
+We chose random forests due to its
+ability to capture complex interactions between features.
+Also, due to it being an ensemble method where trees can be formed independently, which, is ideal for parallelisation.
+Furthermore, the ensemble of weak learners can help to minimise overfitting and variance.
 
-Although our VCF pre-processing phase succeeds with a minimal resource requirement (3GB per executor) the memory required for k-means clustering skyrockets, requiring 24GB per executor to cluster the entire genome.
+However, we observe that Spark ML's random forests implementation isn't able to handle the 1000 Genomes data, and is unable to build a model.
+The reason for the failure can be attributed to the "curse of dimensionality".
+Where Spark ML
+is designed to process a huge number of samples with a modest dimensionality, the 1000 Genomes data instead contains a huge number of dimensions (features).
+As a result, the default random forests implementation is unable to partition the data into small-enough tasks
+for nodes on the computer cluster to handle.
 
-The reason for this can be attributed to the "curse of dimensionality".
-Spark ML is designed to process large numbers of samples rather than a large number of dimensions.
-As a result, data is partitioned along sample boundaries, with entire samples distributed to worker nodes.
-As the sample size grows, Spark can simple schedule more jobs to run to cope with the additional samples.
-
-However with the 1000 Genomes Project data, the dimensionality far exceeds the number of samples.
-This limits the number of jobs that Spark can create, and instead forces it to create fewer, larger jobs.
-
+Therefore, we develop our own 'wide' version of these ML algorithms.
 Lorem ipsum dolor sit~\cite{bib1} amet, consectetur adipiscing elit.
@@
 Praesent faucibus semper libero.
+
+\subsection*{}
+To
+
+\subsection*{CursedForest copes with the curse of dimensionality}
+To
+
+\subsection*{CursedForest copes with the curse of dimensionality}
+To
 

### 2016-03-09 doc/publications/plos16/plos16.tex
some more changes

@@
 
-% Please keep the Author Summary between 150 and 200 words
-% Use first person. PLOS ONE authors please skip this step. 
-% Author Summary not valid for PLOS ONE submissions.   
-\section*{Author Summary}
-Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-Curabitur eget porta erat.
-Morbi consectetur est vel gravida pretium.
-Suspendisse ut dui eu ante cursus gravida non sed sem.
-Nullam sapien tellus, commodo id velit id, eleifend volutpat quam.
-Phasellus mauris velit, dapibus finibus elementum vel, pulvinar non tellus.
-Nunc pellentesque pretium diam, quis maximus dolor faucibus id.
-Nunc convallis sodales ante, ut ullamcorper est egestas vitae.
-Nam sit amet enim ultrices, ultrices elit pulvinar, volutpat risus.
-
 \linenumbers
@@
 
-To manage Developed by Apache, Hadoop is one platform for dealing with large data sets is 
+Developed by Apache, Hadoop is one platform for dealing with large data sets is 
 Platforms Fortunately computational tools such as Apache Spark are available as a framework for iteratively dealing with big data.
-
+Spark, based on Scala, allows for distributed applications that can scale to hundreds or thousands of nodes.
+A Spark application runs on a `driver' node, and this node divides and distributes work out to the many `worker' nodes.
 Spark also includes the machine learning (ML) library, Spark ML, providing popular machine learning libraries.
@@
 
-However, we observe that Spark ML's random forests implementation isn't able to handle the 1000 Genomes data, and is unable to build a model.
-The reason for the failure can be attributed to the "curse of dimensionality".
-Where Spark ML
-is designed to process a huge number of samples with a modest dimensionality, the 1000 Genomes data instead contains a huge number of dimensions (features).
-As a result, the default random forests implementation is unable to partition the data into small-enough tasks
+However, we observe that Spark ML's random forests implementation isn't able to handle the 1000 Genomes data, and cannot build a model on the entire dataset.
+Although Spark ML can build a RF model on a subset of the data (chromosome 1~and~2),
+the time taken is excessive due to the huge amount of data being aggregated and processed by the single driver node during intermediate stages of building the model.
+This means that the work doesn't scale as while the driver is processing data, the worker nodes are lying unused.
+
+The reason for the failure can be attributed to the "curse of dimensionality". Where Spark ML is designed to process a huge number of samples with a modest dimensionality, the 1000 Genomes data
+instead contains a huge number of dimensions (features). As a result, the default random forests implementation is unable to partition the data into small-enough tasks
 for nodes on the computer cluster to handle.
@@
 Therefore, we develop our own 'wide' version of these ML algorithms.
-Lorem ipsum dolor sit~\cite{bib1} amet, consectetur adipiscing elit.
-Curabitur eget porta erat.
-Morbi consectetur est vel gravida pretium.
-Suspendisse ut dui eu ante cursus gravida non sed sem.
-Nullam Eq.~(\ref{eq:schemeP}) sapien tellus, commodo id velit id, eleifend volutpat quam.
-Phasellus mauris velit, dapibus finibus elementum vel, pulvinar non tellus.
-Nunc pellentesque pretium diam, quis maximus dolor faucibus id.~\cite{bib2} Nunc convallis sodales ante, ut ullamcorper est egestas vitae.
-Nam sit amet enim ultrices, ultrices elit pulvinar, volutpat risus.
+Unlike the standard Apache implementation, these are able to easily process data with millions of dimensions without excessive resource requirements on the driver or worker nodes.
+Also, due to more efficient handling of high-dimensional data, equivalent jobs are likely to finish faster on wideML than Spark ML.
 
-\begin{equation}\label{eq:schemeP} 
-D_{coll} = \frac{D_f+\frac{[S]^2}{K_D S_T} D_S} {1+\frac{[S]^2}{K_D S_T}}, 
-D_{sm} = \frac{D_f+ \frac{[S]}{K_D} D_S}{1+\frac{[S]}{K_D}},
-\end{equation}
+%sit~\cite{bib1} Nullam Eq.~(\ref{eq:schemeP}) ~\cite{bib2}  Table~\ref{table1} CO\textsubscript{2}  see \nameref{S1_Text}.
+%\begin{equation}\label{eq:schemeP} 
+%D_{coll} = \frac{D_f+\frac{[S]^2}{K_D S_T} D_S} {1+\frac{[S]^2}{K_D S_T}}, 
+%D_{sm} = \frac{D_f+ \frac{[S]}{K_D} D_S}{1+\frac{[S]}{K_D}},
+%\end{equation}
 
@@
 % authors may use "Analysis" 
-\section*{Materials and Methods}
-\subsection*{Etiam eget sapien nibh.}
+\section*{Methods}
+\subsection*{VariantSpark.}
 
 % For figure citations, please use "Fig." instead of "Figure".
-Nulla mi mi, Fig.~\ref{fig1} venenatis sed ipsum varius, volutpat euismod diam.
-Proin rutrum vel massa non gravida.
-Quisque tempor sem et dignissim rutrum.
-Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-Morbi at justo vitae nulla elementum commodo eu id massa.
-In vitae diam ac augue semper tincidunt eu ut eros.
-Fusce fringilla erat porttitor lectus cursus, \nameref{S1_Video} vel sagittis arcu lobortis.
+%Nulla mi mi, Fig.~\ref{fig1}  Fusce fringilla erat porttitor lectus cursus, \nameref{S1_Video} vel sagittis arcu lobortis.
 Aliquam in enim semper, aliquam massa id, cursus neque.
 Praesent faucibus semper libero.
-
-\begin{figure}[h]
-\caption{{\bf Figure Title first bold sentence Nulla mi mi, venenatis sed ipsum varius, volutpat euismod diam.}
-Figure Caption Proin rutrum vel massa non gravida. Quisque tempor sem et dignissim rutrum. A: Lorem ipsum dolor sit amet. B: Consectetur adipiscing elit.}
-\label{fig1}
-\end{figure}
-
-\begin{enumerate}
-\item{react}
-\item{diffuse free particles}
-\item{increment time by dt and go to 1}
-\end{enumerate}
+%\begin{figure}[h]
+%\caption{{\bf Figure Title first bold sentence Nulla mi mi, venenatis sed ipsum varius, volutpat euismod diam.}
+%Figure Caption Proin rutrum vel massa non gravida. Quisque tempor sem et dignissim rutrum. A: Lorem ipsum dolor sit amet. B: Consectetur adipiscing elit.}
+%\label{fig1}
+%\end{figure}
+%\begin{enumerate}
+%\item{react}
+%\item{diffuse free particles}
+%\item{increment time by dt and go to 1}
+%\end{enumerate}
 
 % Results and Discussion can be combined.
-\section*{Results}
-Nulla mi mi, venenatis sed ipsum varius, Table~\ref{table1} volutpat euismod diam.
-Proin rutrum vel massa non gravida.
-Quisque tempor sem et dignissim rutrum.
-Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-Morbi at justo vitae nulla elementum commodo eu id massa.
-In vitae diam ac augue semper tincidunt eu ut eros.
-Fusce fringilla erat porttitor lectus cursus, vel sagittis arcu lobortis.
-Aliquam in enim semper, aliquam massa id, cursus neque.
-Praesent faucibus semper libero.
+% Please do not create a heading level below \subsection. For 3rd level headings, use \paragraph{}. 
+\section*{Results and Discussion}
 
-\subsection*{}
-To
+\subsection*{Machine learning with Spark ML}
+In this section we compare Apache's higher level machine learning library, Spark ML, with MLLib, the standard low-level library. From the results, we see that times are similar.
+
+There are many practical benefits for using Spark ML with VariantSpark, with a selection of these benefits derived from the `pipeline' system Spark ML uses.
+
+One benefit is that hyperparameter optimisation is built in. Although possible in
 
 \subsection*{CursedForest copes with the curse of dimensionality}
-To
+With Spark ML unable to cope with high-dimensional data, we develop a `wide' implementations of these ML algorithms. **technical stuff by Piotr here**
+These
 
-\subsection*{CursedForest copes with the curse of dimensionality}
-To
+\subsection*{Third section}
 
-\begin{table}[!ht]
-\begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
-\caption{
-{\bf Table caption Nulla mi mi, venenatis sed ipsum varius, volutpat euismod diam.}}
-\begin{tabular}{|l|l|l|l|l|l|l|l|}
-\hline
-\multicolumn{4}{|l|}{\bf Heading1} & \multicolumn{4}{|l|}{\bf Heading2}\\ \hline
-$cell1 row1$ & cell2 row 1 & cell3 row 1 & cell4 row 1 & cell5 row 1 & cell6 row 1 & cell7 row 1 & cell8 row 1\\ \hline
-$cell1 row2$ & cell2 row 2 & cell3 row 2 & cell4 row 2 & cell5 row 2 & cell6 row 2 & cell7 row 2 & cell8 row 2\\ \hline
-$cell1 row3$ & cell2 row 3 & cell3 row 3 & cell4 row 3 & cell5 row 3 & cell6 row 3 & cell7 row 3 & cell8 row 3\\ \hline
-\end{tabular}
-\begin{flushleft} Table notes Phasellus venenatis, tortor nec vestibulum mattis, massa tortor interdum felis, nec pellentesque metus tortor nec nisl.
+\subsection*{Conclusion}
+
+%\begin{table}[!ht]
+%\begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
+%\caption{
+%{\bf Table caption Nulla mi mi, venenatis sed ipsum varius, volutpat euismod diam.}}
+%\begin{tabular}{|l|l|l|l|l|l|l|l|}
+%\hline
+%\multicolumn{4}{|l|}{\bf Heading1} & \multicolumn{4}{|l|}{\bf Heading2}\\ \hline
+%$cell1 row1$ & cell2 row 1 & cell3 row 1 & cell4 row 1 & cell5 row 1 & cell6 row 1 & cell7 row 1 & cell8 row 1\\ \hline
+%$cell1 row2$ & cell2 row 2 & cell3 row 2 & cell4 row 2 & cell5 row 2 & cell6 row 2 & cell7 row 2 & cell8 row 2\\ \hline
+%$cell1 row3$ & cell2 row 3 & cell3 row 3 & cell4 row 3 & cell5 row 3 & cell6 row 3 & cell7 row 3 & cell8 row 3\\ \hline
+%\end{tabular}
+%\begin{flushleft} Table notes Phasellus venenatis, tortor nec vestibulum mattis, massa tortor interdum felis, nec pellentesque metus tortor nec nisl.
 Ut ornare mauris tellus, vel dapibus arcu suscipit sed.
-\end{flushleft}
-\label{table1}
-\end{adjustwidth}
-\end{table}
-
-\subsection*{\lorem\ and \ipsum\ Nunc blandit a tortor.}
-
-Maecenas convallis mauris sit amet sem ultrices gravida.
-Etiam eget sapien nibh.
-Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula.
-Curabitur fringilla pulvinar lectus consectetur pellentesque.
-Quisque augue sem, tincidunt sit amet feugiat eget, ullamcorper sed velit.
-Sed non aliquet felis.
-Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-Mauris commodo justo ac dui pretium imperdiet.
-Sed suscipit iaculis mi at feugiat.
-
-\subsection*{Sed ac quam id nisi malesuada congue.}
-
-Nulla mi mi, venenatis sed ipsum varius, volutpat euismod diam.
-Proin rutrum vel massa non gravida.
-Quisque tempor sem et dignissim rutrum.
-Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-Morbi at justo vitae nulla elementum commodo eu id massa.
-In vitae diam ac augue semper tincidunt eu ut eros.
-Fusce fringilla erat porttitor lectus cursus, vel sagittis arcu lobortis.
-Aliquam in enim semper, aliquam massa id, cursus neque.
-Praesent faucibus semper libero.
-
-% Please do not create a heading level below \subsection. For 3rd level headings, use \paragraph{}. 
-\subsection*{Subsection 1}
-Nulla mi mi, venenatis sed ipsum varius, volutpat euismod diam.
-Proin rutrum vel massa non gravida.
-Quisque tempor sem et dignissim rutrum.
-Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-Morbi at justo vitae nulla elementum commodo eu id massa.
-In vitae diam ac augue semper tincidunt eu ut eros.
-Fusce fringilla erat porttitor lectus cursus, vel sagittis arcu lobortis.
-Aliquam in enim semper, aliquam massa id, cursus neque.
-Praesent faucibus semper libero.
-
-\subsection*{Subsection 2}
-\paragraph{3rd Level Heading.} Nulla mi mi, venenatis sed ipsum varius, volutpat euismod diam.
-Proin rutrum vel massa non gravida.
-Quisque tempor sem et dignissim rutrum.
-Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-Morbi at justo vitae nulla elementum commodo eu id massa.
-In vitae diam ac augue semper tincidunt eu ut eros.
-Fusce fringilla erat porttitor lectus cursus, vel sagittis arcu lobortis.
-Aliquam in enim semper, aliquam massa id, cursus neque.
-Praesent faucibus semper libero.
-
-\section*{Discussion}
-Nulla mi mi, venenatis sed ipsum varius, Table~\ref{table1} volutpat euismod diam.
-Proin rutrum vel massa non gravida.
-Quisque tempor sem et dignissim rutrum.
-Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-Morbi at justo vitae nulla elementum commodo eu id massa.
-In vitae diam ac augue semper tincidunt eu ut eros.
-Fusce fringilla erat porttitor lectus cursus, vel sagittis arcu lobortis.
-Aliquam in enim semper, aliquam massa id, cursus neque.
-Praesent faucibus semper libero.
-
-\subsection*{\lorem\ and \ipsum\ Nunc blandit a tortor.}
-
-CO\textsubscript{2} Maecenas convallis mauris sit amet sem ultrices gravida.
-Etiam eget sapien nibh.
-Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula.
-Curabitur fringilla pulvinar lectus consectetur pellentesque.
-Quisque augue sem, tincidunt sit amet feugiat eget, ullamcorper sed velit.
-
-Sed non aliquet felis.
-Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-Mauris commodo justo ac dui pretium imperdiet.
-Sed suscipit iaculis mi at feugiat.
-Ut neque ipsum, luctus id lacus ut, laoreet scelerisque urna.
-Phasellus venenatis, tortor nec vestibulum mattis, massa tortor interdum felis, nec pellentesque metus tortor nec nisl.
-Ut ornare mauris tellus, vel dapibus arcu suscipit sed.
-Nam condimentum sem eget mollis euismod.
-Nullam dui urna, gravida venenatis dui et, tincidunt sodales ex.
-Nunc est dui, sodales sed mauris nec, auctor sagittis leo.
-Aliquam tincidunt, ex in facilisis elementum, libero lectus luctus est, non vulputate nisl augue at dolor.
-For more information, see \nameref{S1_Text}.
+%\end{flushleft}
+%\label{table1}
+%\end{adjustwidth}
+%\end{table}
 

### 2016-03-21 doc/publications/plos16/plos16.tex
plos paper comments

@@
 {\Large
-\textbf\newline{Article Title - Submission to PLOS Journals}
+\textbf\newline{Cursed Forest - A random forest implementation for 'big' and 'wide' data}
 }
@@
 \\
-Name1 Surname\textsuperscript{1,\Yinyang},
-Name2 Surname\textsuperscript{2,\Yinyang},
-Name3 Surname\textsuperscript{2,\textcurrency a},
-Name4 Surname\textsuperscript{2,\ddag},
-Name5 Surname\textsuperscript{2,\ddag},
-Name6 Surname\textsuperscript{2},
-Name7 Surname\textsuperscript{3,*},
-with the Lorem Ipsum Consortium\textsuperscript{\textpilcrow}
+Aidan O'Brien\textsuperscript{1},
+Piotr Szul\textsuperscript{2},
+Stephanie Li\textsuperscript{3},
+James Doecke\textsuperscript{3},
+Nick Ellis\textsuperscript{4},
+Robert Dunne\textsuperscript{5}, and
+Denis C. Bauer\textsuperscript{1,*}
+%, with the Lorem Ipsum Consortium\textsuperscript{\textpilcrow}
 \\
 \bigskip
-\bf{1} Affiliation Dept/Program/Center, Institution Name, City, State, Country
+\bf{1} Health \& Biosecurity, CSIRO, Sydney, NSW, Australia
 \\
-\bf{2} Affiliation Dept/Program/Center, Institution Name, City, State, Country
+\bf{2} Data61, CSIRO, Brisbane, QLD, Australia
 \\
-\bf{3} Affiliation Dept/Program/Center, Institution Name, City, State, Country
+\bf{3} Health \& Biosecurity, CSIRO, Brisbane, QLD, Australia
+\\
+\bf{4} Oceans \& Atmosphere, CSIRO, Brisbane, QLD, Australia
+\\
+\bf{5} Data61, CSIRO, Sydney, NSW, Australia
 \\
@@
 % Primary Equal Contribution Note
-\Yinyang These authors contributed equally to this work.
+%\Yinyang These authors contributed equally to this work.
 
@@
 % Also use this double-dagger symbol for special authorship notes, such as senior authorship.
-\ddag These authors also contributed equally to this work.
+%\ddag These authors also contributed equally to this work.
 
 % Current address notes
-\textcurrency a Insert current address of first author with an address update
+%\textcurrency a Insert current address of first author with an address update
 % \textcurrency b Insert current address of second author with an address update
@@
 % Deceased author note
-\dag Deceased
+%\dag Deceased
 
 % Group/Consortium Author Note
-\textpilcrow Membership list can be found in the Acknowledgments section.
+%\textpilcrow Membership list can be found in the Acknowledgments section.
 
 % Use the asterisk to denote corresponding authorship and provide email address in note below.
-* CorrespondingAuthor@institute.edu
+* Denis.Bauer@CSIRO.au
 
@@
 
-In the past few years, whole genome sequence (WGS) data has rapidly been becoming an essential part in the clinical space, where traditional methods are failing to produce results.
+The digital revolution or 'datafication' sees a dramatic increase in data collected about every aspect of life~\cite{Loebbecke2015}. 
+These dataset not only grow vertically by capturing more and more events but also horizontally by capturing more information about these events.
+The challenge of big and 'wide' data is especially pronounced in the health space where whole genome sequence (WGS) technology enabled researchers to interrogate all 3 billion basepairs of the human genome.
+While a large fraction of this data is identical between individuals, identifying the relevant and disease specific differences is the focus of complex trait analysis [citation].
 
-WGS has proven itself in discovering rare variants [http://www.sciencedirect.com/science/article/pii/S2352396414000498] and diagnosing unknown genetic disorders [MATT MIGHT],
+For example, the 1000 Genome data consists of approximately 2500 samples with up to 80 million variants~\cite{1KG2012}, denoting the differences between individuals. 
+This dataset consumes close to a terabyte of disk space.
+The richness of this data can be utilised in other datasets as well by filling in or "imputing" the information at genomic positions previously unobserved due to the less detailed resolution of older array technology [citation about imputation].
+This means that the large number of samples for which genome-wide association (GWA) data is available~\cite{Welter2013} can be extended horizontally by imputing the previously unobserved genotypes. 
+This provides the potential to generate datasets of hundreds of thousands of individuals with millions of variants, highlighting the need for incorporating modern compute paradigms to deal with these challenges.
 
-One challenge with analysing WGS data over, for example, traditional genome-wide association (GWA) data, is developing software and algorithms to deal with the huge amounts of data.
-Although the sample size of WGS data-sets may be comparable to GWA, the dimensionality of the data may be many times larger. For example, the 1000 Genome data consists of
-approximately 2500 individuals with up to 80 million variants. The uncompressed file sizes are close to a terabyte. Of course, it's not just genomic data that is rapidly growing. Many other
-types of data
+%WGS has proven itself in discovering rare variants [http://www.sciencedirect.com/science/article/pii/S2352396414000498] and diagnosing unknown genetic disorders [MATT MIGHT],
+%One challenge with analysing WGS data over, for example, traditional genome-wide association (GWA) data, is developing software and algorithms to deal with the huge amounts of data.
+%Although the sample size of WGS data-sets may be comparable to GWA, the dimensionality of the data may be many times larger. For example, the 1000 Genome data consists of
+%approximately 2500 individuals with up to 80 million variants. The uncompressed file sizes are close to a terabyte. Of course, it's not just genomic data that is rapidly growing. Many other
+%types of data
 
-Developed by Apache, Hadoop is one platform for dealing with large data sets is 
-Platforms Fortunately computational tools such as Apache Spark are available as a framework for iteratively dealing with big data.
-Spark, based on Scala, allows for distributed applications that can scale to hundreds or thousands of nodes.
+%Developed by Apache, Hadoop is one platform for dealing with large data sets is 
+%Platforms Fortunately computational tools such as Apache Spark are available as a framework for iteratively dealing with big data. 
+Distributing the compute tasks is one strategy to overcome this challenge and Apache Spark, based on Scala, allows for distributed applications to scale to hundreds or thousands of nodes.
 A Spark application runs on a `driver' node, and this node divides and distributes work out to the many `worker' nodes.
 Spark also includes the machine learning (ML) library, Spark ML, providing popular machine learning libraries. 
-Spark ML includes supervised approaches such as decision trees and the ensemble method, random forests, as well as
-unsupervised approaches such as k-means clustering.
-We previously demonstrated the versatility and scalability of Spark by developing VariantSpark, a framework allowing users to easily analyse Variant Call Format (VCF) files using ML algorithms on the Spark framework.
-Using VariantSpark,
-we successfully built a k-means model on the 2500 individuals from the 1000 Genomes Project, based on their genomic data.
-We also investigated classifying the samples using Random Forests.
-We chose random forests due to its
-ability to capture complex interactions between features.
+Spark ML includes supervised approaches such as decision trees and the ensemble method, random forests, as well as unsupervised approaches such as k-means clustering.
+
+We previously demonstrated the versatility and scalability of Spark by developing VariantSpark~\cite{OBrien2015}, a framework allowing users to easily analyse Variant Call Format (VCF) files using ML algorithms on the Spark framework.
+Using VariantSpark, we successfully built a k-means model on the full 2500x80Million matrix to cluster individuals by their ethnicity achieving an Adjusted Rand Index ARI=0.84, with 1 perfect and -1 random clustering.
+
+While unsupervised clustering is useful for selected clinical applications~\cite{Li2015}, a more common analysis task is supervised learning, i.e.
+given a phenotype of clinical label determine the most associated features.
+One of the simplest supervised machine learning methods is logistic regression (LR), which creates a decision boundary by fitting a linear combination of the feature variables.
+Processing the 'wide' genomic data poses difficulties for machine learning applications as having more features than samples causes the model to not generalise well and overfit the data. 
+This is known as the 'curse of dimensionality'~\cite{Bauer2014}. 
+Random Forests (RF) as an ensemble of weak learners overcomes this issue and is hence ideally suited to this application. 
+Furthermore, RF are able to capture interactions between features, which is of importance for modelling complex polygenic diseases.
+
+%We also investigated classifying the samples using Random Forests. 
+%We chose random forests due to its
+%ability to capture complex interactions between features.
 Also, due to it being an ensemble method where trees can be formed independently, which, is ideal for parallelisation.
@@
 
-However, we observe that Spark ML's random forests implementation isn't able to handle the 1000 Genomes data, and cannot build a model on the entire dataset.
-Although Spark ML can build a RF model on a subset of the data (chromosome 1~and~2),
-the time taken is excessive due to the huge amount of data being aggregated and processed by the single driver node during intermediate stages of building the model.
-This means that the work doesn't scale as while the driver is processing data, the worker nodes are lying unused.
+However, we observe that Spark's standard RF implementation is not able to handle the extremely 'wide' genomic data as it was developed for large number of samples with only modest dimensionality. 
+Although Spark ML can build a RF model on a subset of the data (chromosome 1~and~2), the time taken is excessive due to the huge amount of data being aggregated and processed by the single driver node during intermediate stages of building the model (see Result~\ref{comp}).
+This unbalanced work load where the driver node becomes the bottleneck and worker nodes being idle prevents a seamless scaling to larger datasets.
 
-The reason for the failure can be attributed to the "curse of dimensionality". Where Spark ML is designed to process a huge number of samples with a modest dimensionality, the 1000 Genomes data
-instead contains a huge number of dimensions (features). As a result, the default random forests implementation is unable to partition the data into small-enough tasks
-for nodes on the computer cluster to handle.
+%The reason for the failure can be attributed to the "curse of dimensionality". Where Spark ML is designed to process a huge number of samples with a modest dimensionality, the 1000 Genomes data
+%instead contains a huge number of dimensions (features). As a result, the default random forests implementation is unable to partition the data into small-enough tasks
+%for nodes on the computer cluster to handle.
 
-Therefore, we develop our own 'wide' version of these ML algorithms.
-Unlike the standard Apache implementation, these are able to easily process data with millions of dimensions without excessive resource requirements on the driver or worker nodes.
-Also, due to more efficient handling of high-dimensional data, equivalent jobs are likely to finish faster on wideML than Spark ML.
-
-%sit~\cite{bib1} Nullam Eq.~(\ref{eq:schemeP}) ~\cite{bib2}  Table~\ref{table1} CO\textsubscript{2}  see \nameref{S1_Text}.
-%\begin{equation}\label{eq:schemeP} 
-%D_{coll} = \frac{D_f+\frac{[S]^2}{K_D S_T} D_S} {1+\frac{[S]^2}{K_D S_T}}, 
-%D_{sm} = \frac{D_f+ \frac{[S]}{K_D} D_S}{1+\frac{[S]}{K_D}},
-%\end{equation}
+We therefore we developed CursedForest a Spark-based random forest algorithm able to process big and 'wide' data.
+Unlike the standard Apache implementation, CursedForest is able to easily process data with millions of dimensions without excessive resource requirements on the driver or worker nodes.
+%Also, due to more efficient handling of high-dimensional data, equivalent jobs are likely to finish faster on wideML than Spark ML.
+In the fist section we
 
@@
 
-\subsection*{Machine learning with Spark ML}
-In this section we compare Apache's higher level machine learning library, Spark ML, with MLLib, the standard low-level library. From the results, we see that times are similar.
+%OPTIONS
+% Spark ML vs MLlib
+% random forest, kmeans, logistic regression 
+% synthetic data demonstrating random forest can cope with n<<p
+% James dataset
+% GWAS dataset
 
-There are many practical benefits for using Spark ML with VariantSpark, with a selection of these benefits derived from the `pipeline' system Spark ML uses.
-
-One benefit is that hyperparameter optimisation is built in. Although possible in
+\subsection*{Synthetic data demonstrating that RF can deal with extreme dimensionality}
+Detailing random association between features and labels
+and whatever rob did
 
 \subsection*{CursedForest copes with the curse of dimensionality}
-With Spark ML unable to cope with high-dimensional data, we develop a `wide' implementations of these ML algorithms. **technical stuff by Piotr here**
-These
+\label{comp}
+Spark ML vs CursedForest vs Kmeans vs LR on small and full version (where applicable) of 1000G (performance and time/mem)
 
-\subsection*{Third section}
+%In this section we compare Apache's higher level machine learning library, Spark ML, with MLLib, the standard low-level library. From the results, we see that times are similar. 
+%There are many practical benefits for using Spark ML with VariantSpark, with a selection of these benefits derived from the `pipeline' system Spark ML uses.
+%One benefit is that hyperparameter optimisation is built in. Although possible in
+
+%\subsection*{CursedForest copes with the curse of dimensionality}
+%With Spark ML unable to cope with high-dimensional data, we develop a `wide' implementations of these ML algorithms. **technical stuff by Piotr here**
+%These
+
+\subsection*{James' dataset}
+Comparing CursedForest with feature selection + linear model (performance and time/mem)
 

### 2016-03-23 doc/publications/plos16/plos16.tex
fix backticks

@@
 {\Large
-\textbf\newline{Cursed Forest - A random forest implementation for 'big' and 'wide' data}
+\textbf\newline{Cursed Forest - A random forest implementation for `big' and `wide' data}
 }
@@
 
-The digital revolution or 'datafication' sees a dramatic increase in data collected about every aspect of life~\cite{Loebbecke2015}. 
+The digital revolution or `datafication' sees a dramatic increase in data collected about every aspect of life~\cite{Loebbecke2015}. 
 These dataset not only grow vertically by capturing more and more events but also horizontally by capturing more information about these events.
-The challenge of big and 'wide' data is especially pronounced in the health space where whole genome sequence (WGS) technology enabled researchers to interrogate all 3 billion basepairs of the human genome.
+The challenge of big and `wide' data is especially pronounced in the health space where whole genome sequence (WGS) technology enabled researchers to interrogate all 3 billion basepairs of the human genome.
 While a large fraction of this data is identical between individuals, identifying the relevant and disease specific differences is the focus of complex trait analysis [citation].
@@
 One of the simplest supervised machine learning methods is logistic regression (LR), which creates a decision boundary by fitting a linear combination of the feature variables.
-Processing the 'wide' genomic data poses difficulties for machine learning applications as having more features than samples causes the model to not generalise well and overfit the data. 
-This is known as the 'curse of dimensionality'~\cite{Bauer2014}. 
+Processing the `wide' genomic data poses difficulties for machine learning applications as having more features than samples causes the model to not generalise well and overfit the data. 
+This is known as the `curse of dimensionality'~\cite{Bauer2014}. 
 Random Forests (RF) as an ensemble of weak learners overcomes this issue and is hence ideally suited to this application. 
@@
 
-However, we observe that Spark's standard RF implementation is not able to handle the extremely 'wide' genomic data as it was developed for large number of samples with only modest dimensionality. 
+However, we observe that Spark's standard RF implementation is not able to handle the extremely `wide' genomic data as it was developed for large number of samples with only modest dimensionality. 
 Although Spark ML can build a RF model on a subset of the data (chromosome 1~and~2), the time taken is excessive due to the huge amount of data being aggregated and processed by the single driver node during intermediate stages of building the model (see Result~\ref{comp}).
@@
 
-We therefore we developed CursedForest a Spark-based random forest algorithm able to process big and 'wide' data.
+We therefore we developed CursedForest a Spark-based random forest algorithm able to process big and `wide' data.
 Unlike the standard Apache implementation, CursedForest is able to easily process data with millions of dimensions without excessive resource requirements on the driver or worker nodes.

### 2016-03-30 variantspark/src/main/scala/au/csiro/obr17q/variantspark/model/VcfParser.scala
Add missing header item and new print

@@ -23,7 +23,7 @@ case class VcfRecord(
 
 class VcfParser (val VcfFileNames: String, val VariantCutoff: Int, val IndividualMeta: RDD[IndividualMap], val sc: SparkContext, val sqlContext: SQLContext) extends scala.Serializable {
 
-  val NotAVariant = "#CHROM" :: "POS" :: "REF" :: "ALT" :: "QUAL" :: "FILTER" :: "INFO" :: "FORMAT" :: Nil
+  val NotAVariant = "#CHROM" :: "POS" :: "ID" :: "REF" :: "ALT" :: "QUAL" :: "FILTER" :: "INFO" :: "FORMAT" :: Nil
 
   val VcfFilesRDD = sc.textFile(VcfFileNames, 20)
 
@@ -53,6 +53,8 @@ class VcfParser (val VcfFileNames: String, val VariantCutoff: Int, val Individua
     val VariantCutoff = this.VariantCutoff
     val NotAVariant = this.NotAVariant
     val Headings = getHeadings(VcfFilesRDD)
+    println(s"Individuals: ${Headings.length - 9}")
+
     VcfFilesRDD
     .filter(!_.startsWith("#"))
     .mapPartitions( lines => {
@@ -79,7 +81,7 @@ class VcfParser (val VcfFileNames: String, val VariantCutoff: Int, val Individua
   def data = sqlContext
     .createDataFrame {
       val variantCount = this.variantCount
-      println(variantCount)
+      println(s"Variants: $variantCount")
       individualTuples
         .groupByKey //group by individual ID, i.e. get RDD of individuals
         .map(p => (p._1.split('_')(0).substring(0, 12), (p._1.split('_')(1), p._2))) // Split the TCGA key to get ID & type
@@ -132,7 +134,7 @@ class VcfParser (val VcfFileNames: String, val VariantCutoff: Int, val Individua
 
   def individualVariants : RDD[FlatVariant] = {
     VcfLineRdd
-    .flatMap( (h) => {
+    .flatMap( h => {
       h._3
       .map( i => FlatVariant( i._1, h._1.toInt, i._2))
     })

### 2016-03-30 variantspark/src/main/scala/au/csiro/obr17q/variantspark/model/FakeDataParser.scala
Added save function.. quick and easy way to enable use of same synthetic data on sparkML and wideRF

@@ -6,6 +6,8 @@ import org.apache.spark.SparkContext
 import org.apache.spark.mllib.linalg._
 import org.apache.spark.rdd.RDD
 import org.apache.spark.sql.SQLContext
+import org.apache.hadoop.conf.Configuration
+import org.apache.hadoop.fs._
 
 /**
   * Created by obr17q on 5/01/2016.
@@ -13,7 +15,7 @@ import org.apache.spark.sql.SQLContext
 
 private case class FakeDataRecord(individual: String, simulationOne: Double, simulationTwo: Double, features: Vector)
 
-class FakeDataParser (val n: Int, val p: Int, val sc: SparkContext, val sqlContext: SQLContext) extends scala.Serializable {
+class FakeDataParser (val samples: Int, val dims: Int, val save:Boolean = false, val sc: SparkContext, val sqlContext: SQLContext) extends scala.Serializable {
 
 
 
@@ -22,9 +24,9 @@ class FakeDataParser (val n: Int, val p: Int, val sc: SparkContext, val sqlConte
     * RDD of an line Arrays
     */
   private def CsvLineRDD: RDD[(Array[Double], Long)] = {
-    val p = this.p
-    sc.parallelize( List.fill(n)(Array))
-      .map(q => q.fill(p)(Random.nextDouble))
+    val dims = this.dims
+    sc.parallelize( List.fill(samples)(Array))
+      .map(q => q.fill(dims)(Random.nextDouble))
       .zipWithIndex
   }
 
@@ -35,7 +37,7 @@ class FakeDataParser (val n: Int, val p: Int, val sc: SparkContext, val sqlConte
     * count = noOfGenes
     */
   def featureTuples: List[(String, Int)] = {
-    (1 to p).toList
+    (1 to dims).toList
       .map(name => s"col$name")
       .zipWithIndex
   }
@@ -44,7 +46,7 @@ class FakeDataParser (val n: Int, val p: Int, val sc: SparkContext, val sqlConte
   val data = sqlContext
     .createDataFrame {
       val featureTuples = this.featureTuples
-      val p = this.p
+      val dims = this.dims
       val r = Random.nextDouble
       CsvLineRDD
         .map{ case(x, id) =>
@@ -53,12 +55,32 @@ class FakeDataParser (val n: Int, val p: Int, val sc: SparkContext, val sqlConte
             individual = s"row$id",
[…]

### 2016-03-30 variantspark/src/main/scala/au/csiro/obr17q/variantspark/VcfToFlatVariant.scala
Added line (commented) to alternately call parser for abeta csv files

@@ -2,7 +2,7 @@ package au.csiro.obr17q.variantspark
 
 import org.apache.spark.rdd.RDD
 import au.csiro.obr17q.variantspark.model.FlatVariant
-import au.csiro.obr17q.variantspark.model.VcfParser
+import au.csiro.obr17q.variantspark.model._
 import scala.io._
 
 object VcfToFlatVariant extends SparkApp {
@@ -18,6 +18,8 @@ object VcfToFlatVariant extends SparkApp {
     val IndividualMeta: RDD[IndividualMap] = null
 
     val vcfParser = new VcfParser(VcfFiles, VariantCutoff, IndividualMeta, sc, sqlContext)
+    //val vcfParser = new ABetaParser(VcfFiles, sc, sqlContext)
+
     //val NoOfAlleles = vcfParser.variantCount
     val FilteredAlleles:RDD[FlatVariant] = vcfParser.individualVariants
 

### 2016-03-30 variantspark/src/main/scala/au/csiro/obr17q/variantspark/model/ABetaParser.scala
Just updated

@@ -10,7 +10,7 @@ import org.apache.spark.sql.SQLContext
   * Created by obr17q on 5/01/2016.
   */
 
-private case class ABetaRecord(individual: String, aBeta: String, features: Vector)
+private case class ABetaRecord(individual: String, aBeta: Double, features: Vector)
 
 class ABetaParser (val CsvFileNames: String, val sc: SparkContext, val sqlContext: SQLContext) extends scala.Serializable {
 
@@ -37,7 +37,8 @@ class ABetaParser (val CsvFileNames: String, val sc: SparkContext, val sqlContex
 
 
   /**
-    * RDD of an line Arrays
+    * @return
+    * RDD[ Array(column1:String, column2:String, ..., columnN:String)]
     */
   private def CsvLineRDD: RDD[Array[String]] = {
     CsvFilesRDD
@@ -52,9 +53,9 @@ class ABetaParser (val CsvFileNames: String, val sc: SparkContext, val sqlContex
 
   /**
     * RDD of Variant IDs zipped with a unique index.
-    * RDD[(Variant, VariantIndex)]
+    * @return
+    * RDD[(Variant:String, VariantIndex:Int)]
     * RDD[(PGM1, 1221)]
-    * count = noOfGenes
     */
   def featureTuples: List[(String, Int)] = {
     //val NotAVariant = this.NotAVariant
@@ -64,15 +65,28 @@ class ABetaParser (val CsvFileNames: String, val sc: SparkContext, val sqlContex
   }
 
   /**
-    * RDD[(IndividualID, (BMI, BMI_CAT, MSI_STATUS))]
-    * RDD[(TCGA-CA-6717, (30.25, "obese", 2))]
+    * RDD[(IndividualID:String, abStatus:Int))]
     * count = noOfIndividuals
     */
-  private def IndividualMetaData: RDD[(String, (Double, String, String))] = {
+  def IndividualMetaData: RDD[(String, Int)] = {
     CsvLineRDD
-      .map(line => (line(10), (line(27).toDouble, line(28), line(23))))
+      .map(line => (line(0), line(1).toDouble.toInt))
       .distinct
   }
+
+  /**
+    * @return
+    * RDD[ FlatVariant(subjectId:String, variantIndex:Int, allele:Double)]
+    */
+
+  def individualVariants: RDD[FlatVariant] = {
+    val featureTuples = this.featureTuples
+    CsvLineRDD
+      .map(p => (p(0), p.drop(2).zip(featureTuples)))
[…]

### 2016-03-30 variantspark/src/main/scala/au/csiro/obr17q/variantspark/model/CsvParser.scala

@@ -1,10 +1,10 @@
 package au.csiro.obr17q.variantspark.model
 
 import au.com.bytecode.opencsv.CSVParser
-import au.csiro.obr17q.variantspark.VcfForest._
 import org.apache.spark.SparkContext
 import org.apache.spark.mllib.linalg._
 import org.apache.spark.rdd.RDD
+import org.apache.spark.sql.SQLContext
 
 /**
   * Created by obr17q on 5/01/2016.
@@ -12,7 +12,7 @@ import org.apache.spark.rdd.RDD
 
 private case class CsvRecord(individual: String, sampleType: String, bmi: Double, msi_status: String, bmi_cat: String, features: Vector)
 
-class CsvParser (val CsvFileNames: String, val VariantCutoff: Int, val sc: SparkContext) extends scala.Serializable {
+class CsvParser (val CsvFileNames: String, val VariantCutoff: Int, val sc: SparkContext, val sqlContext: SQLContext) extends scala.Serializable {
 
   private val CsvFilesRDD = sc.textFile(CsvFileNames, 5)
 
@@ -36,12 +36,12 @@ class CsvParser (val CsvFileNames: String, val VariantCutoff: Int, val sc: Spark
     * RDD[(PGM1, 1221)]
     * count = noOfGenes
     */
-  def alleleTuples: RDD[(String, Int)] = {
+  def featureTuples: List[(String, Int)] = {
     CsvLineRDD
       .map(line => line(7))
       .distinct
       .zipWithIndex
-      .map(p => (p._1, p._2.toInt))
+      .map(p => (p._1, p._2.toInt)).collect.toList
   }
 
   /**
@@ -58,13 +58,13 @@ class CsvParser (val CsvFileNames: String, val VariantCutoff: Int, val sc: Spark
   val data = sqlContext
     .createDataFrame {
       val variantCount = this.variantCount
-      val alleleTuples = this.alleleTuples
+      val featureTuples = this.featureTuples
       val IndividualMetaData = this.IndividualMetaData
       CsvLineRDD
         .map(line => ((line(10), line(7)), 1))
         .reduceByKey(_ + _)
         .map(p => (p._1._2, (p._1._1, p._2.toDouble)))
-        .join(alleleTuples)
+        .join(sc.parallelize(featureTuples))
         .map(p => (p._2._1._1, (p._2._2, p._2._1._2)))
         .groupByKey
         .join(IndividualMetaData)
[…]

### 2016-03-31 variantspark/src/main/scala/au/csiro/obr17q/variantspark/algo/WideRandomForest.scala
dims wrong?

@@ -21,8 +21,8 @@ case class WideRandomForestModel(trees: List[WideDecisionTreeModel], val labelCo
 
   def variableImportance: Map[Long, Double] = {
     // average by all trees
-    val accumulations = new Long2DoubleOpenHashMap();
-    val counts = new Long2IntOpenHashMap();
+    val accumulations = new Long2DoubleOpenHashMap()
+    val counts = new Long2IntOpenHashMap()
     trees.foreach { t =>
       val treeImportnace = t.variableImportanceAsFastMap
       //println(treeImportnace.toMap)
@@ -56,23 +56,28 @@ case class WideRandomForestModel(trees: List[WideDecisionTreeModel], val labelCo
 }
 
 case class RandomForestParams(
-    val oob:Boolean = true,
-    val nTryFraction:Double =  Double.NaN
+    oob:Boolean = true,
+    nTryFraction:Double =  Double.NaN
 )
 
 class WideRandomForest {
   def run(data: RDD[(Vector, Long)], labels: Array[Int], ntrees: Int, params:RandomForestParams = RandomForestParams()): WideRandomForestModel = {
-    // subsample 
+    // subsample
+    //dims seems to be the number of samples, not number of dimensions?
     val dims = labels.length
+    val features = data.count().toInt
+
     val labelCount = labels.max + 1
     val oobVotes = Array.fill(dims)(Array.fill(labelCount)(0))
-    
-    val ntryFraction = if (params.nTryFraction.isNaN() ) Math.sqrt(dims.toDouble)/dims.toDouble else params.nTryFraction
-    println(s"RF: Using ntryfraction: ${ntryFraction}")
-    
-    val trees = Range(0, ntrees).map { i =>
-      println(s"Building tree: ${i}")
-      val currentSample = for (i <- 0 until dims) yield (Math.random() * dims).toInt // sample with replacemnt
+    println(features.toDouble)
+    val ntryFraction = if (params.nTryFraction.isNaN ) Math.sqrt(features.toDouble)/features.toDouble else params.nTryFraction
+    println(s"RF: Using ntryfraction: $ntryFraction")
+    println(s"RF: i.e. trying ${(features * ntryFraction).toInt} features per split")
+
+    val trees = Range(0, ntrees).map { p =>
+      println(s"Building tree: $p")
+      //val currentSample = for (i <- 0 until dims) yield (Math.random * dims).toInt // sample with replacement
[…]

### 2016-04-02 doc/publications/plos16/plos16.tex
Just updated
added PLOS-submission.pdf as it was not being created automatically for me
added section on variable selection, figures, bibliography

@@
 \fancyhf{}
-\lhead{\includegraphics[width=2.0in]{PLOS-submission.eps}}
+\lhead{\includegraphics[width=2.0in]{PLOS-submission}}
 \rfoot{\thepage/\pageref{LastPage}}
@@
 %Also, due to more efficient handling of high-dimensional data, equivalent jobs are likely to finish faster on wideML than Spark ML.
-In the fist section we
+In the first section we
+
+\section*{Variable Selection Using Random Forests}
+
+\cite{Biau.2012} provide a proof that a random forest model will converge at a rate that depends on the cardinality of
+the set $S$ of  ``strong predictors'' rather than on the number of variables $p$. That is, given a function $y=f(S)$
+depending on a set of variables $S$, a random forest will converge to the true function $y$ with a rate that depends on
+the size of $S$ and not the number of noise variables in the data set. If the size of $S$ is small compared to the total
+number of variables, then the rate of convergence will be quicker.
+
+In a number of simulations (reproduced from \cite{Genuer.et.al.2010}) we see evidence that random forests can recover
+the variables defining a functional relationship in the presence of a substantial number of noise variables. See the
+supplementary information for full details. It is apparently that some of the model parameters, such
+as \texttt{mtry} -- the number of randomly selected variables considered as potential splits at each node, have to be altered
+from their default value when there are a large number of predictors.
+
+It is likely that in the genomic setting, variables will have a monotonic influence on the outcome. 
+We have taken \cite{Genuer.et.al.2010}'s tree function example and extended it to a much larger number of noise variables.
+The function, shown in figure  \ref{figure:tree_function.png}, is 
+defined on the first 5 predictor variables. We have added Gaussian noise to the function (figure \ref{figure:noisy_tree_function.png})
+and extended the size of the training data set from $n=100 \times p=1000$ to $n=100 \times  p=1000000$.
+The random forest model recovers variables 1 and 4 as the two
+most important variables (out of 1000000). It fails to uncover the other 3 variables, $\{2,3,5\}$.
+
+It would appear that even in setting with an extremely large number of noise variables, it may be possible to recover
+the significant varaiables if the functional relationship is simple.
+
+\begin{figure}[tbhp]
+ \includegraphics[totalheight=9cm]{./figs/tree_function.png}
+ \captionof{figure}{The tree function.}
+  \label{figure:tree_function.png}
+\end{figure}
+
+\begin{figure}[tbhp]
+ \includegraphics[totalheight=8cm]{./figs/noisy_tree_function.png}
+ \captionof{figure}{A plot of the ordered $y$ variable for the tree example, and the $y$ variable with added Gaussian
+   noise (in red).}
+  \label{figure:noisy_tree_function.png}
+\end{figure}
 
@@
 \section*{Methods}
+
 \subsection*{VariantSpark.}
@@
 %
-\begin{thebibliography}{10}
-\bibitem{bib1}
-Devaraju P, Gulati R, Antony PT, Mithun CB, Negi VS.
-Susceptibility to SLE in South Indian Tamils may be influenced by genetic selection pressure on TLR2 and TLR9 genes.
-Mol Immunol.
-2014 Nov 22.
-pii: S0161-5890(14)00313-7.
-doi: 10.1016/j.molimm.2014.11.005
 
-\bibitem{bib2}
-Huynen MMTE, Martens P, Hilderlink HBM. The health impacts of globalisation: a conceptual framework. Global Health. 2005;1: 14. Available: http://www.globalizationandhealth.com/content/1/1/14.
-
-\end{thebibliography}
+\bibliography{./CursedForest.bib}
 

### 2016-04-11 variantspark/src/main/scala/au/csiro/obr17q/variantspark/VcfForest.scala
Disable cross-validation, add regression metrics

@@ -7,7 +7,7 @@ import org.apache.spark.ml.{Pipeline, PipelineModel, PipelineStage}
 import org.apache.spark.ml.tuning.{CrossValidator, ParamGridBuilder}
 import org.apache.spark.ml.evaluation.{MulticlassClassificationEvaluator, RegressionEvaluator}
 import org.apache.spark.ml.feature.{IndexToString, StringIndexer, VectorIndexer}
-import org.apache.spark.mllib.evaluation.MulticlassMetrics
+import org.apache.spark.mllib.evaluation.{MulticlassMetrics, RegressionMetrics}
 import org.apache.spark.mllib.linalg._
 import org.apache.spark.sql.{DataFrame, Row}
 import org.apache.spark.rdd.RDD
@@ -105,11 +105,6 @@ object VcfForest extends SparkApp {
     //modObject.saveAs("100x1000.csv")
 
     val Array(trainingData, testData) = data.randomSplit(Array(0.8, 0.2))
-    val testSize: Int = testData.count.toInt
-    println(s"Samples: ${data.count}")
-    println(s"Testset: $testSize")
-
-
 
 
     //val gender = modObject.feature("col5")
@@ -198,11 +193,11 @@ object VcfForest extends SparkApp {
       .addGrid(rfClassifier.maxBins, maxBins)
       .build
 
-    val cv = new CrossValidator()
-      .setEstimator(pipeline)
-      .setEvaluator(regressionEvaluator)
-      .setEstimatorParamMaps(paramGrid)
-      .setNumFolds(numFolds)
+    //val cv = new CrossValidator()
+    //  .setEstimator(pipeline)
+    //  .setEvaluator(regressionEvaluator)
+    //  .setEstimatorParamMaps(paramGrid)
+    //  .setNumFolds(numFolds)
 
 
     /**
@@ -225,19 +220,25 @@ object VcfForest extends SparkApp {
       featureTuples
         .map(p => (p._2, (p._1, rf(p._2))))
         .sortBy(_._2._2, ascending = false)
-        .take(50)
+        .take(100)
         .foreach(println)
     }
 
     def modelFit(trainDF: DataFrame, testDF: DataFrame) = {
+
+      val testSize: Int = testDF.count.toInt
+      println(s"Total samples: ${data.count}")
+      println(s"Testset: $testSize")
+
+
       // Build the cross-validated model
-      val cvModel = cv.fit(trainDF)
+      val cvModel = pipeline.fit(trainDF)
 
 
[…]

### 2016-04-11 doc/publications/plos16/plos16.tex
paper sections

--- 2016-03-23 doc/publications/plos16/plos16.tex
@@
 One of the simplest supervised machine learning methods is logistic regression (LR), which creates a decision boundary by fitting a linear combination of the feature variables.
-Processing the `wide' genomic data poses difficulties for machine learning applications as having more features than samples causes the model to not generalise well and overfit the data. 
+Processing the `wide' genomic data poses difficulties for machine learning applications as having more features $p$ than samples $n$ causes the model to not generalise well and overfit the data. 
 This is known as the `curse of dimensionality'~\cite{Bauer2014}. 
@@
 %Also, due to more efficient handling of high-dimensional data, equivalent jobs are likely to finish faster on wideML than Spark ML.
-In the fist section we
+In the first section we
 
@@
 
-\subsection*{Synthetic data demonstrating that RF can deal with extreme dimensionality}
-Detailing random association between features and labels
-and whatever rob did
+\section*{Random Forests converges on wide data}
 
-\subsection*{CursedForest copes with the curse of dimensionality}
+As discussed in the Introduction, random forests are robust against overfitting and hence can generalise for datasets where $p > n$, however here we demonstrate that this holds true in extreme cases where $p \gg n$.
+
+Biau {\it et al.}~\cite{Biau.2012} provide a proof that a random forest model will converge at a rate that depends on the cardinality of
+the set $S$ of  ``strong predictors'' rather than on the number of variables $p$. That is, given a function $y=f(S)$
+depending on a set of variables $S$, a random forest will converge to the true function $y$ with a rate that depends on
+the size of $S$ and not the number of noise variables in the data set. If the size of $S$ is small compared to the total
+number of variables, then the rate of convergence will be quicker.
+
+To test this empirically we adapt the simulations developed by Genuer {\it et al.}~\cite{Genuer.et.al.2010} to include a much larger feature vector $p$ per sample.  
+The function, shown in Figure  \ref{figure:synth}a, is 
+defined on the first 5 predictor variables. We have added Gaussian noise to the function (Figure \ref{figure:synth}b)
+and extended the size of the training data set from $n=100 \times p=1000$ to $n=100 \times  p=1000000$.
+%TODO recovers 2 out of 5 ? I think you need to spell out why this is a good demonstrator that RF can deal with such a large number of features...
+%TODO also how is the added noise playing into this ?
+The random forest model recovers variables 1 and 4 as the two
+most important variables (out of 1000000). It fails to uncover the other 3 variables, $\{2,3,5\}$.
+
+% Important ?
+%In a number of simulations (reproduced from \cite{Genuer.et.al.2010}) we see evidence that random forests can recover
+%the variables defining a functional relationship in the presence of a substantial number of noise variables. See the
+%supplementary information for full details. It is apparently that some of the model parameters, such
+%as \texttt{mtry} -- the number of randomly selected variables considered as potential splits at each node, have to be altered
+%from their default value when there are a large number of predictors.
+
+%It is likely that in the genomic setting, variables will have a monotonic influence on the outcome. 
+%We have taken \cite{Genuer.et.al.2010}'s tree function example and extended it to a much larger number of noise variables.
+%The function, shown in figure  \ref{figure:tree_function.png}, is 
+%defined on the first 5 predictor variables. We have added Gaussian noise to the function (figure \ref{figure:noisy_tree_function.png})
+%and extended the size of the training data set from $n=100 \times p=1000$ to $n=100 \times  p=1000000$.
+%The random forest model recovers variables 1 and 4 as the two
+%most important variables (out of 1000000). It fails to uncover the other 3 variables, $\{2,3,5\}$.
+
+%It would appear that even in setting with an extremely large number of noise variables, it may be possible to recover
+%the significant varaiables if the functional relationship is simple.
+
+%TODO can you make the caption more descriptive, e.g.
+what are the splits in the three and what are the numbers; similarly for b) there are 100 samples (can you name the index axis samples) and why are there 9 targets (y) ?
+\begin{figure}[tbhp]
+\begin{tabular}{ll}
+a)& b)\\
+ \includegraphics[totalheight=6cm]{./figs/tree_function.png}&
+  \includegraphics[totalheight=6cm]{./figs/noisy_tree_function.png}
+ \end{tabular}
+ \captionof{figure}{{\bf Synthetic data demonstrating RF's capability to cope with $p \gg n$.} {\bf a)} The tree function, where each split means XXXX.
+{\bf b)} A plot of the ordered $y$ variable for the tree example, and the $y$ variable with added Gaussian
+   noise (in red).}
+  \label{figure:synth}
+\end{figure}
+
+\subsection*{CursedForest processes genome wide variant data}
 \label{comp}
@@
 
-%\subsection*{CursedForest copes with the curse of dimensionality}
-%With Spark ML unable to cope with high-dimensional data, we develop a `wide' implementations of these ML algorithms. **technical stuff by Piotr here**
-%These
+\begin{table}[!ht]
+%\begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
+\caption{
+{\bf Performance comparison between the different Machine Learning algorithms.}}
+\begin{tabular}{|l|l|l|l|l|l|l|l|}
+\hline
+%\multicolumn{4}{|l|}{\bf Heading1} & \multicolumn{4}{|l|}{\bf Heading2}\\ \hline
+\bf{Genome \%}  & \bf{method} & \bf{Accuracy} & \bf{Runtime (s)} & \bf{Memory} \\
+\hline
+X\% & CursedForest & &  &\\ \hline
+          & RF(Spark ML) & &  &\\ \hline
+          &  RF(R) & &  &\\ \hline
+          &  LR & &  &\\ \hline
+XX\% & CursedForest & &  &\\ \hline
+          & RF(Spark ML) & &  &\\ \hline
+          &  RF(R) & &  &\\ \hline
+          &  LR & &  &\\ \hline
+
+100\% & CursedForest & & & \\ \hline
+          & RF(Spark ML) & &  &\\ \hline
+          &  RF(R) & &  &\\ \hline
+          &  LR & &  &\\ \hline
+          & kmeans(Spark ML) & & & \\ \hline
+\end{tabular}
+\begin{flushleft} Table notes Phasellus venenatis, tortor nec vestibulum mattis, massa tortor interdum felis, nec pellentesque metus tortor nec nisl.
+Ut ornare mauris tellus, vel dapibus arcu suscipit sed.
+\end{flushleft}
+\label{table1}
+%\end{adjustwidth}
+\end{table}
+
+\subsection*{Random forest is sensitive to tree depht}
+
+\begin{table}[!ht]
+%\begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
+\caption{
+{\bf Sensitivity analysis on parameter choice.}}
+\begin{tabular}{|l|l|l|l|l|l|l|l|}
+\hline
+%\multicolumn{4}{|l|}{\bf Heading1} & \multicolumn{4}{|l|}{\bf Heading2}\\ \hline
+\bf{Mtry}  & \bf{Mtree} & \bf{depth} & \bf{Accuracy} & \bf{Runtime} & \bf{Memory} \\
+\hline
+&&&&&\\ \hline
+\end{tabular}
+\begin{flushleft} Table notes Phasellus venenatis, tortor nec vestibulum mattis, massa tortor interdum felis, nec pellentesque metus tortor nec nisl.
+Ut ornare mauris tellus, vel dapibus arcu suscipit sed.
+\end{flushleft}
+\label{table2}
+%\end{adjustwidth}
+\end{table}
 

### 2016-04-21 doc/publications/plos16/plos16.tex
Added RF numbers to table

@@
 This dataset consumes close to a terabyte of disk space.
-The richness of this data can be utilised in other datasets as well by filling in or "imputing" the information at genomic positions previously unobserved due to the less detailed resolution of older array technology [citation about imputation].
+The richness of this data can be utilised in other datasets as well by filling in or ``imputing" the information at genomic positions previously unobserved due to the less detailed resolution of older array technology [citation about imputation].
 This means that the large number of samples for which genome-wide association (GWA) data is available~\cite{Welter2013} can be extended horizontally by imputing the previously unobserved genotypes. 
@@
 %\multicolumn{4}{|l|}{\bf Heading1} & \multicolumn{4}{|l|}{\bf Heading2}\\ \hline
-\bf{Genome \%}  & \bf{method} & \bf{Accuracy} & \bf{Runtime (s)} & \bf{Memory} \\
+\bf{Genome \%}  & \bf{method} & \bf{Accuracy} & \bf{Runtime (s)} & \bf{Memory/Node} \\
 \hline
-X\% & CursedForest & &  &\\ \hline
-          & RF(Spark ML) & &  &\\ \hline
+8\% (chr1) & CursedForest & 0.0359 & 1hr 58min 23sec & 4GB \\ \hline
+          & RF(Spark ML) & 0.0395 & 8hr 29min 07sec & 8GB \\ \hline
           &  RF(R) & &  &\\ \hline
@@
 
-100\% & CursedForest & & & \\ \hline
-          & RF(Spark ML) & &  &\\ \hline
+100\% & CursedForest & 0.0231 & 7hr 14min 40sec & 8GB \\ \hline
+          & RF(Spark ML) & - & - & - \\ \hline
           &  RF(R) & &  &\\ \hline
-          &  LR & &  &\\ \hline
+          &  LR & - & - & - \\ \hline
           & kmeans(Spark ML) & & & \\ \hline

### 2016-04-22 doc/publications/plos16/plos16.tex
small edits -- added Notes section

@@
 
+\reversemarginpar
+
 % Text layout
@@
 % Please keep the abstract below 300 words
+
+\section*{Notes}
+\begin{itemize}
+\item what are you guys editing this with?  what about some carrige
+  returns or line feeds or something!
+\item Are we going to use the TCGA data in this paper?
+\item are we going to run sparkML or Curesed forest on the extended
+  simulation data \cite[]{Genuer.et.al.2010}? Has Aidan done this already?
+\end{itemize}
+
 \section*{Abstract}
-Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-Curabitur eget porta erat.
-Morbi consectetur est vel gravida pretium.
-Suspendisse ut dui eu ante cursus gravida non sed sem.
-Nullam sapien tellus, commodo id velit id, eleifend volutpat quam.
-Phasellus mauris velit, dapibus finibus elementum vel, pulvinar non tellus.
-Nunc pellentesque pretium diam, quis maximus dolor faucibus id.
-Nunc convallis sodales ante, ut ullamcorper est egestas vitae.
-Nam sit amet enim ultrices, ultrices elit pulvinar, volutpat risus.
+Lorem ipsum dolor sit amet, consectetur adipiscing elit. Curabitur eget porta erat. Morbi consectetur est vel gravida
+pretium. Suspendisse ut dui eu ante cursus gravida non sed sem. Nullam sapien tellus, commodo id velit id, eleifend
+volutpat quam. Phasellus mauris velit, dapibus finibus elementum vel, pulvinar non tellus. Nunc pellentesque pretium
+diam, quis maximus dolor faucibus id. Nunc convallis sodales ante, ut ullamcorper est egestas vitae. Nam sit amet enim
+ultrices, ultrices elit pulvinar, volutpat risus.
 
@@
 
-The digital revolution or `datafication' sees a dramatic increase in data collected about every aspect of life~\cite{Loebbecke2015}. 
-These dataset not only grow vertically by capturing more and more events but also horizontally by capturing more information about these events.
-The challenge of big and `wide' data is especially pronounced in the health space where whole genome sequence (WGS) technology enabled researchers to interrogate all 3 billion basepairs of the human genome.
-While a large fraction of this data is identical between individuals, identifying the relevant and disease specific differences is the focus of complex trait analysis [citation].
+The digital revolution or `datafication' sees a dramatic increase in data collected about every aspect of   \marginpar{is seeing, will see}
+life~\cite{Loebbecke2015}.  These dataset not only grow vertically by capturing more and more events but also
+horizontally by capturing more information about these events.  The challenge of big and `wide' data is especially
+pronounced in the health space where whole genome sequence (WGS) technology enabled researchers to interrogate all 3
+billion basepairs of the human genome.  While a large fraction of this data is identical between individuals, \marginpar{base-pairs}
+identifying the relevant and disease specific differences is the focus of complex trait analysis [citation].
 
-For example, the 1000 Genome data consists of approximately 2500 samples with up to 80 million variants~\cite{1KG2012}, denoting the differences between individuals. 
-This dataset consumes close to a terabyte of disk space.
-The richness of this data can be utilised in other datasets as well by filling in or ``imputing" the information at genomic positions previously unobserved due to the less detailed resolution of older array technology [citation about imputation].
-This means that the large number of samples for which genome-wide association (GWA) data is available~\cite{Welter2013} can be extended horizontally by imputing the previously unobserved genotypes. 
-This provides the potential to generate datasets of hundreds of thousands of individuals with millions of variants, highlighting the need for incorporating modern compute paradigms to deal with these challenges.
+For example, the 1000 Genome data consists of approximately 2500 samples with up to 80 million variants~\cite{1KG2012},
+denoting the differences between individuals.  This dataset consumes close to a terabyte of disk space.  The richness of
+this data can be utilised in other datasets as well by filling in or ``imputing" the information at genomic positions
+previously unobserved due to the less detailed resolution of older array technology [citation about imputation].  This
+means that the large number of samples for which genome-wide association (GWA) data is available~\cite{Welter2013} can
+be extended horizontally by imputing the previously unobserved genotypes.  This provides the potential to generate
+datasets of hundreds of thousands of individuals with millions of variants, highlighting the need for incorporating
+modern compute paradigms to deal with these challenges.
 
-%WGS has proven itself in discovering rare variants [http://www.sciencedirect.com/science/article/pii/S2352396414000498] and diagnosing unknown genetic disorders [MATT MIGHT],
-%One challenge with analysing WGS data over, for example, traditional genome-wide association (GWA) data, is developing software and algorithms to deal with the huge amounts of data.
-%Although the sample size of WGS data-sets may be comparable to GWA, the dimensionality of the data may be many times larger. For example, the 1000 Genome data consists of
-%approximately 2500 individuals with up to 80 million variants. The uncompressed file sizes are close to a terabyte. Of course, it's not just genomic data that is rapidly growing. Many other
-%types of data
+% WGS has proven itself in discovering rare variants
+% [http://www.sciencedirect.com/science/article/pii/S2352396414000498] and diagnosing unknown genetic disorders [MATT
+% MIGHT], One challenge with analysing WGS data over, for example, traditional genome-wide association (GWA) data, is
+% developing software and algorithms to deal with the huge amounts of data.  Although the sample size of WGS data-sets
+% may be comparable to GWA, the dimensionality of the data may be many times larger. For example, the 1000 Genome data
+% consists of approximately 2500 individuals with up to 80 million variants. The uncompressed file sizes are close to a
+% terabyte. Of course, it's not just genomic data that is rapidly growing. Many other types of data
 
-%Developed by Apache, Hadoop is one platform for dealing with large data sets is 
-%Platforms Fortunately computational tools such as Apache Spark are available as a framework for iteratively dealing with big data. 
-Distributing the compute tasks is one strategy to overcome this challenge and Apache Spark, based on Scala, allows for distributed applications to scale to hundreds or thousands of nodes.
-A Spark application runs on a `driver' node, and this node divides and distributes work out to the many `worker' nodes.
-Spark also includes the machine learning (ML) library, Spark ML, providing popular machine learning libraries. 
-Spark ML includes supervised approaches such as decision trees and the ensemble method, random forests, as well as unsupervised approaches such as k-means clustering.
+% Developed b Apache, Hadoop is one platform for dealing with large data sets is Platforms Fortunately computational
+% tools such as Apache Spark are available as a framework for iteratively dealing with big data.
+Distributing the compute tasks is one strategy to overcome this challenge. Apache Spark, based on Scala, allows for  
+\marginpar{is it based on scale or does it have a scale API?}
+distributed applications to scale to thousands of nodes.  A Spark application runs on a `driver' node, and
+this node divides and distributes work out to the many `worker' nodes.  Spark also includes the machine learning (ML)
+library, Spark ML, providing popular machine learning libraries.  Spark ML includes supervised approaches such as
+decision trees and the ensemble method, random forests, as well as unsupervised approaches such as k-means clustering.
 
-We previously demonstrated the versatility and scalability of Spark by developing VariantSpark~\cite{OBrien2015}, a framework allowing users to easily analyse Variant Call Format (VCF) files using ML algorithms on the Spark framework.
-Using VariantSpark, we successfully built a k-means model on the full 2500x80Million matrix to cluster individuals by their ethnicity achieving an Adjusted Rand Index ARI=0.84, with 1 perfect and -1 random clustering.
+We previously demonstrated the versatility and scalability of Spark by developing VariantSpark~\cite{OBrien2015}, a
+framework allowing users to easily analyse Variant Call Format (VCF) files using ML algorithms on the Spark framework.
+Using VariantSpark, we successfully built a k-means model on the full 2500x80Million matrix to cluster individuals by
+their ethnicity achieving an Adjusted Rand Index of 0.84 (with 1 being perfect clustering  and -1 random clustering).
 
-While unsupervised clustering is useful for selected clinical applications~\cite{Li2015}, a more common analysis task is supervised learning, i.e.
-given a phenotype of clinical label determine the most associated features.
-One of the simplest supervised machine learning methods is logistic regression (LR), which creates a decision boundary by fitting a linear combination of the feature variables.
-Processing the `wide' genomic data poses difficulties for machine learning applications as having more features $p$ than samples $n$ causes the model to not generalise well and overfit the data. 
-This is known as the `curse of dimensionality'~\cite{Bauer2014}. 
-Random Forests (RF) as an ensemble of weak learners overcomes this issue and is hence ideally suited to this application. 
-Furthermore, RF are able to capture interactions between features, which is of importance for modelling complex polygenic diseases.
+While unsupervised clustering is useful for selected clinical applications~\cite{Li2015}, a more common analysis task is
+supervised learning, i.e. given a phenotype of clinical label determine the most associated features.  One of the
+simplest supervised machine learning methods is logistic regression (LR), which creates a decision boundary by fitting a
+linear combination of the feature variables.  Processing the `wide' genomic data poses difficulties for machine learning
+applications as having more features $p$ than samples $n$ causes the model to overfit the data and generalize poorly.
+This is known as the `curse of dimensionality'~\cite{Bauer2014}.  Random Forests (RF) as an ensemble of weak learners 
+\marginpar{weak learners only in a certain sense. Originally ``weak learners'' was applied to decision stumps}
+overcomes this issue and is hence ideally suited to this application.  Furthermore, RF are able to capture interactions
+between features, which is of importance for modelling complex polygenic diseases.
 
-%We also investigated classifying the samples using Random Forests. 
-%We chose random forests due to its
-%ability to capture complex interactions between features.
-Also, due to it being an ensemble method where trees can be formed independently, which, is ideal for parallelisation.
-Furthermore, the ensemble of weak learners can help to minimise overfitting and variance.
+% We also investigated classifying the samples using Random Forests.  We chose random forests due to its ability to
+% capture complex interactions between features. Also, due to it being an ensemble method where trees can be formed
+% independently, which, is ideal for parallelisation. Furthermore, the ensemble of weak learners can help to minimise
+% overfitting and variance.
 
-However, we observe that Spark's standard RF implementation is not able to handle the extremely `wide' genomic data as it was developed for large number of samples with only modest dimensionality. 
-Although Spark ML can build a RF model on a subset of the data (chromosome 1~and~2), the time taken is excessive due to the huge amount of data being aggregated and processed by the single driver node during intermediate stages of building the model (see Result~\ref{comp}).
-This unbalanced work load where the driver node becomes the bottleneck and worker nodes being idle prevents a seamless scaling to larger datasets.
+However, we observe that Spark's standard RF implementation is not able to handle the extremely `wide' genomic data as
+it was developed for large number of samples with only modest dimensionality.  Although Spark ML can build a RF model on
+a subset of the data (chromosome 1~and~2), the time taken is excessive due to the huge amount of data being aggregated
+and processed by the single driver node during intermediate stages of building the model (see Result~\ref{comp}).  This
+unbalanced work load where the driver node becomes the bottleneck and worker nodes being idle prevents a seamless
+scaling to larger datasets.
 
-%The reason for the failure can be attributed to the "curse of dimensionality". Where Spark ML is designed to process a huge number of samples with a modest dimensionality, the 1000 Genomes data
-%instead contains a huge number of dimensions (features). As a result, the default random forests implementation is unable to partition the data into small-enough tasks
-%for nodes on the computer cluster to handle.
+% The reason for the failure can be attributed to the "curse of dimensionality". Where Spark ML is designed to process a
+% huge number of samples with a modest dimensionality, the 1000 Genomes data instead contains a huge number of
+% dimensions (features). As a result, the default random forests implementation is unable to partition the data into
+% small-enough tasks for nodes on the computer cluster to handle.
 
-We therefore we developed CursedForest a Spark-based random forest algorithm able to process big and `wide' data.
-Unlike the standard Apache implementation, CursedForest is able to easily process data with millions of dimensions without excessive resource requirements on the driver or worker nodes.
+We therefore we developed CursedForest, a Spark-based random forest algorithm able to process big and `wide' data. Unlike
+the standard Apache implementation, CursedForest is able to easily process data with millions of dimensions without
+excessive resource requirements on the driver or worker nodes.
 %Also, due to more efficient handling of high-dimensional data, equivalent jobs are likely to finish faster on wideML than Spark ML.
@@
 
-As discussed in the Introduction, random forests are robust against overfitting and hence can generalise for datasets where $p > n$, however here we demonstrate that this holds true in extreme cases where $p \gg n$.
+As discussed in the Introduction, random forests are robust against overfitting and hence can generalise for datasets
+where $p > n$, however here we demonstrate that this holds true in extreme cases where $p \gg n$.
 
-Biau {\it et al.}~\cite{Biau.2012} provide a proof that a random forest model will converge at a rate that depends on the cardinality of
-the set $S$ of  ``strong predictors'' rather than on the number of variables $p$. That is, given a function $y=f(S)$
-depending on a set of variables $S$, a random forest will converge to the true function $y$ with a rate that depends on
-the size of $S$ and not the number of noise variables in the data set. If the size of $S$ is small compared to the total
-number of variables, then the rate of convergence will be quicker.
+Biau {\it et al.}~\cite{Biau.2012} provide a proof that a random forest model will converge at a rate that depends on
+the cardinality of the set $S$ of ``strong predictors'' rather than on the number of variables $p$. That is, given a
+function $y=f(S)$ depending on a set of variables $S$, a random forest will converge to the true function $y$ with a
+rate that depends on the size of $S$ and not the number of noise variables in the data set. If the size of $S$ is small
+compared to the total number of variables, then the rate of convergence will be quicker.
+\marginpar{say something about the rate of convergence}
 
-To test this empirically we adapt the simulations developed by Genuer {\it et al.}~\cite{Genuer.et.al.2010} to include a much larger feature vector $p$ per sample.  
-The function, shown in Figure  \ref{figure:synth}a, is 
-defined on the first 5 predictor variables. We have added Gaussian noise to the function (Figure \ref{figure:synth}b)
-and extended the size of the training data set from $n=100 \times p=1000$ to $n=100 \times  p=1000000$.
+To test this empirically we adapt the simulations developed by Genuer {\it et al.}~\cite{Genuer.et.al.2010} to include a
+much larger feature vector $p$ per sample.  The function, shown in Figure \ref{figure:synth}a, is defined on the first 5
+predictor variables. We have added Gaussian noise to the function (Figure \ref{figure:synth}b) and extended the size of
+the training data set from $n=100 \times p=1000$ to $p=1000000$.
 %TODO recovers 2 out of 5 ? I think you need to spell out why this is a good demonstrator that RF can deal with such a large number of features...
@@
 
-% Important ?
-%In a number of simulations (reproduced from \cite{Genuer.et.al.2010}) we see evidence that random forests can recover
-%the variables defining a functional relationship in the presence of a substantial number of noise variables. See the
-%supplementary information for full details. It is apparently that some of the model parameters, such
-%as \texttt{mtry} -- the number of randomly selected variables considered as potential splits at each node, have to be altered
-%from their default value when there are a large number of predictors.
+% Important ?  In a number of simulations (reproduced from \cite{Genuer.et.al.2010}) we see evidence that random forests
+% can recover the variables defining a functional relationship in the presence of a substantial number of noise
+% variables. See the supplementary information for full details. It is apparently that some of the model parameters,
+% such as \texttt{mtry} -- the number of randomly selected variables considered as potential splits at each node, have
+% to be altered from their default value when there are a large number of predictors.
 
-%It is likely that in the genomic setting, variables will have a monotonic influence on the outcome. 
-%We have taken \cite{Genuer.et.al.2010}'s tree function example and extended it to a much larger number of noise variables.
-%The function, shown in figure  \ref{figure:tree_function.png}, is 
-%defined on the first 5 predictor variables. We have added Gaussian noise to the function (figure \ref{figure:noisy_tree_function.png})
-%and extended the size of the training data set from $n=100 \times p=1000$ to $n=100 \times  p=1000000$.
-%The random forest model recovers variables 1 and 4 as the two
-%most important variables (out of 1000000). It fails to uncover the other 3 variables, $\{2,3,5\}$.
+% It is likely that in the genomic setting, variables will have a monotonic influence on the outcome.  We have taken
+% \cite{Genuer.et.al.2010}'s tree function example and extended it to a much larger number of noise variables.  The
+% function, shown in figure \ref{figure:tree_function.png}, is defined on the first 5 predictor variables. We have added
+% Gaussian noise to the function (figure \ref{figure:noisy_tree_function.png}) and extended the size of the training
+% data set from $n=100 \times p=1000$ to $n=100 \times p=1000000$.  The random forest model recovers variables 1 and 4
+% as the two most important variables (out of 1000000). It fails to uncover the other 3 variables, $\{2,3,5\}$.
 
-%It would appear that even in setting with an extremely large number of noise variables, it may be possible to recover
-%the significant varaiables if the functional relationship is simple.
+% It would appear that even in setting with an extremely large number of noise variables, it may be possible to recover
+% the significant varaiables if the functional relationship is simple.
 
-%TODO can you make the caption more descriptive, e.g.
-what are the splits in the three and what are the numbers; similarly for b) there are 100 samples (can you name the index axis samples) and why are there 9 targets (y) ?
+% TODO can you make the caption more descriptive, e.g. what are the splits in the three and what are the numbers;
+% similarly for b) there are 100 samples (can you name the index axis samples) and why are there 9 targets (y) ?
+
 \begin{figure}[tbhp]
@@
  \end{tabular}
- \captionof{figure}{{\bf Synthetic data demonstrating RF's capability to cope with $p \gg n$.} {\bf a)} The tree function, where each split means XXXX.
-{\bf b)} A plot of the ordered $y$ variable for the tree example, and the $y$ variable with added Gaussian
-   noise (in red).}
+ \captionof{figure}{{\bf Synthetic data demonstrating RF's capability to cope with $p \gg n$.} {\bf a)} The tree
+   function, where each split means XXXX. {\bf b)} A plot of the ordered $y$ variable for the tree example, and the $y$
+   variable with added Gaussian noise (in red).}
   \label{figure:synth}
@@
 \label{comp}
-Spark ML vs CursedForest vs Kmeans vs LR on small and full version (where applicable) of 1000G (performance and time/mem)
+Spark ML vs CursedForest vs Kmeans vs LR on small and full version (where applicable) of 1000G (performance and
+time/mem)
 
-%In this section we compare Apache's higher level machine learning library, Spark ML, with MLLib, the standard low-level library. From the results, we see that times are similar. 
-%There are many practical benefits for using Spark ML with VariantSpark, with a selection of these benefits derived from the `pipeline' system Spark ML uses.
-%One benefit is that hyperparameter optimisation is built in. Although possible in
+% In this section we compare Apache's higher level machine learning library, Spark ML, with MLLib, the standard
+% low-level library. From the results, we see that times are similar.  There are many practical benefits for using Spark
+% ML with VariantSpark, with a selection of these benefits derived from the `pipeline' system Spark ML uses.  One
+% benefit is that hyperparameter optimisation is built in. Although possible in
 
@@
 \end{tabular}
-\begin{flushleft} Table notes Phasellus venenatis, tortor nec vestibulum mattis, massa tortor interdum felis, nec pellentesque metus tortor nec nisl.
-Ut ornare mauris tellus, vel dapibus arcu suscipit sed.
+\begin{flushleft} 
+  Table notes Phasellus venenatis, tortor nec vestibulum mattis, massa tortor interdum felis, nec pellentesque metus
+  tortor nec nisl. Ut ornare mauris tellus, vel dapibus arcu suscipit sed.
 \end{flushleft}
@@
 
-\subsection*{Random forest is sensitive to tree depht}
+\subsection*{Random forest is sensitive to tree depth}
 

### 2016-04-22 doc/publications/plos16/plos16.tex
2 sections -- parameter settings and decision trees

@@
 
-\subsection*{Random forest is sensitive to tree depth}
+\subsection*{Parameter settings}
+\marginpar{ was''Random forest is sensitive to tree depth''}
+In this setting, variable selection in very high dimensions, we nned to consider the parameter setting for the RF
+algorithm.
+
+I will use the R terms from the package RandomForests whcih incoporates the original Fortran code by Brieman and
+Cutler. We consider the parameters:
+\begin{itemize}
+\item \texttt{max_depth} the size (complexity) of the individual trees
+\item \texttt{ntree} the number of trees 
+\item \texttt{mtry} --  the number of variables considered at each split (if \texttt{mtry}=$p$, we have a boosted decison
+  tree model \marginpar{is this right?}
+\end{itemize}
+
+We note that it is possible for a RF model to have a very poor fit, and be useless for prediction, and still have a 
+useful ranking of variable importance.
+
+\subsection*{decision trees }
+\begin{itemize}
+\item benefits os a decision tree
+  \begin{itemize}
+  \item invariant to scale of the input data
+  \item handles categorical and real valued inputs 
+  \item can handle missing values 
+  \item insensitive to useless predictors
+  \end{itemize}
+\item benefits of random forests
+  \begin{itemize}
+  \item variable importance 
+  \item inherits the qualities of tree models
+  \end{itemize}
+\end{itemize}
 

### 2016-04-26 doc/publications/plos16/plos16.tex
small changes

@@
 \begin{itemize}
-\item what are you guys editing this with?  what about some carrige
+\item what are you guys editing this with?  what about some carriage
   returns or line feeds or something!
 \item Are we going to use the TCGA data in this paper?
-\item are we going to run sparkML or Curesed forest on the extended
+\item are we going to run sparkML or Cursed forest on the extended
   simulation data \cite[]{Genuer.et.al.2010}? Has Aidan done this already?
@@
 \begin{itemize}
-\item \texttt{max_depth} the size (complexity) of the individual trees
+\item the size and complexity of the individual trees is controlled in RandomForests by setting \texttt{nodesize}, the
+  minimum size of terminal nodes. It is controlled in Spark.ML by setting \texttt{maxDepth}, the maximum depth of each
+  tree in the forest. The SparkML documentation somewhat unhelpfully sets \texttt{maxDepth=4} in their classification
+  example\footnote{the documentation is available at \url{http://spark.apache.org/docs/latest/mllib-ensembles.html} }.
+  In the ATCG example the RF code produced trees with depths of 30+. We note that Random forests are a way of averaging
+  multiple deep decision trees, trained on different parts of the same training set, with the goal of reducing the
+  variance. \url{https://en.wikipedia.org/wiki/Random_forest} and refs therein.  
+\marginpar{are we including this  example?  Supplementary info?}
 \item \texttt{ntree} the number of trees 

### 2016-04-26 doc/publications/plos16/plos16.tex
iChanges

--- 2016-04-22 doc/publications/plos16/plos16.tex
@@
 %\multicolumn{4}{|l|}{\bf Heading1} & \multicolumn{4}{|l|}{\bf Heading2}\\ \hline
-\bf{Genome \%}  & \bf{method} & \bf{Accuracy} & \bf{Runtime (s)} & \bf{Memory/Node} \\
+\bf{Genome \%}  & \bf{method} & \bf{Accuracy} & \bf{Runtime} & \bf{Memory/Node} \\
 \hline
-8\% (chr1) & CursedForest & 0.0359 & 1hr 58min 23sec & 4GB \\ \hline
-          & RF(Spark ML) & 0.0395 & 8hr 29min 07sec & 8GB \\ \hline
-          &  RF(R) & &  &\\ \hline
+8\% (chr1) & CursedForest & XX & 1hr 58min 23sec & 4GB \\ \hline
+          & RF (Spark ML) & XX & 8hr 29min 07sec & 8GB \\ \hline
+          &  RF (R) & &  &\\ \hline
           &  LR & &  &\\ \hline
@@
 
-100\% & CursedForest & 0.0231 & 7hr 14min 40sec & 8GB \\ \hline
-          & RF(Spark ML) & - & - & - \\ \hline
-          &  RF(R) & &  &\\ \hline
+100\% & CursedForest & 0.96 & 7hr 14min 40sec & 8GB \\ \hline
+          & RF (Spark ML) & - & - & - \\ \hline
+          &  RF (R) & &  &\\ \hline
           &  LR & - & - & - \\ \hline
-          & kmeans(Spark ML) & & & \\ \hline
+          & kmeans (Spark ML) & 0.82 & 30hr 44min 00sec & 24GB \\ \hline
 \end{tabular}

### 2016-06-29 doc/publications/plos16/plos16.tex
added reference, some small changes

@@
 framework allowing users to easily analyse Variant Call Format (VCF) files using ML algorithms on the Spark framework.
-Using VariantSpark, we successfully built a k-means model on the full 2500x80Million matrix to cluster individuals by
+Using VariantSpark, we successfully built a k-means model on the full
+$2500 \times 80$ million matrix to cluster individuals by
 their ethnicity achieving an Adjusted Rand Index of 0.84 (with 1 being perfect clustering  and -1 random clustering).
@@
 applications as having more features $p$ than samples $n$ causes the model to overfit the data and generalize poorly.
-This is known as the `curse of dimensionality'~\cite{Bauer2014}.  Random Forests (RF) as an ensemble of weak learners 
-\marginpar{weak learners only in a certain sense. Originally ``weak learners'' was applied to decision stumps}
+This is known as the `curse of dimensionality'~\cite{Bauer2014}.  Random Forests~\cite{Breiman.2001} (RF) as an ensemble of weak learners 
+\marginpar{weak learners only in a misuse of the term. ``weak learners'' was applied to decision stumps}
+%http://stats.stackexchange.com/questions/77018/random-forest-is-it-a-boosting-algorithm
+%http://stats.stackexchange.com/questions/173390/gradient-boosting-tree-vs-random-forest
 overcomes this issue and is hence ideally suited to this application.  Furthermore, RF are able to capture interactions
@@
 
-We therefore we developed CursedForest, a Spark-based random forest algorithm able to process big and `wide' data. Unlike
+We therefore developed CursedForest, a Spark-based random forest algorithm able to process big and `wide' data. Unlike
 the standard Apache implementation, CursedForest is able to easily process data with millions of dimensions without
@@
 
-As discussed in the Introduction, random forests are robust against overfitting and hence can generalise for datasets
-where $p > n$, however here we demonstrate that this holds true in extreme cases where $p \gg n$.
+As discussed in the Introduction, random forests are robust against overfitting
+\marginpar{strong claim -- see disussion in Segal 2004, perhaps}
+and hence can generalise for datasets where $p > n$, however here we demonstrate that this holds true in extreme cases where $p \gg n$.
 

### 2016-06-29 doc/publications/plos16/plos16.tex
changes to plos16.tex

@@
 applications as having more features $p$ than samples $n$ causes the model to overfit the data and generalize poorly.
-This is known as the `curse of dimensionality'~\cite{Bauer2014}.  Random Forests~\cite{Breiman.2001} (RF) as an ensemble of weak learners 
+This is known as the `curse of dimensionality'~\cite{Bauer2014}.  Random Forests~\cite{Breiman.2001} (RF) 
+apply the technique of bootstrap aggregation or bagging to decision trees. The training set is randomly sampled $B$
+times with replacement. A tree is grown on each of these $B$ samples and the results are combined
+\begin{equation*}
+{\bar {h}}={\frac {1}{B}}\sum _{b=1}^{B} h(x;\theta_b)
+\end{equation*}
+
+Let $PE_f$ be the predicton error for the forest. Then
+\begin{equation*}
+PE_f \leq \hat{\rho} PR_b
+\end{equation*}
+where $\hat{\rho}$ is the weighted correlation of the residuals $Y-  h(X;\theta)$ and $Y-  h(X;\theta^\prime)$
+for independent $\theta,  \theta^\prime $
+
+So what is required for  accurate random forest regression is low correlation between residuals of differing tree members of
+the forest, and (ii) low prediction error for the individual trees.
+
+The strategy employed to achieve these ends is as follows:
+1. To keep individual error low, grow trees to maximum depth.
+To keep residual correlation low randomize via
+a) Grow each tree on a bootstrap sample from the training data.
+b) Specify the number of covariates. At each node of every tree
+select covariates and pick the best split of that node based on these
+covariates.\cite{Segal.2004}
+
 \marginpar{weak learners only in a misuse of the term. ``weak learners'' was applied to decision stumps}
@@
 
+\begin{table}
+\begin{minipage}{\textwidth}
+\begin{center}
+\begin{tabular}{|p{4.2cm}|l|l|l|l|l|l|l|}
+\hline
+\bf{Genome }  & \bf{method} & \bf{Accuracy} & \bf{Runtime} & \bf{Memory/Node} \\
+\hline
+ chr 22 - phase1 (2.1GB)  &  RF (R)\footnote{10 nodes, 500 trees, dpas03} 
+                                   &  0.456 & 2.89 & 18GB \\ \hline
+\end{tabular}
+\medskip
+\begin{flushleft} 
+\caption{some R results on smaller data sets.}
+\label{table:model_outputs}
+\end{flushleft}
+\end{center}
+%\end{sidewaystable}
+\end{minipage}
+\end{table}
+
+% \begin
+% \begin{table}[!ht]
+% \caption{
+% {\bf Performance comparison between the different Machine Learning algorithms.}}
+% \begin{tabular}{|p{4.2cm}|l|l|l|l|l|l|l|}
+% \hline
+% \bf{Genome }  & \bf{method} & \bf{Accuracy} & \bf{Runtime} & \bf{Memory/Node} \\
+% \hline
+%  chr 22 - phase1 (2.1GB)  &  RF (R) &  0.456 & 2.89 & 18GB \\ \hline
+% \end{tabular}
+% \begin{flushleft} 
+% \end{flushleft}
+% \label{table1}
+% %\end{adjustwidth}
+% \end{table}
+
 \subsection*{Parameter settings}

### 2016-07-04 doc/publications/plos16/plos16.tex
changes to plos.tex

@@
 applications as having more features $p$ than samples $n$ causes the model to overfit the data and generalize poorly.
-This is known as the `curse of dimensionality'~\cite{Bauer2014}.  Random Forests~\cite{Breiman.2001} (RF) 
+This is known as the `curse of dimensionality'~\cite{Bauer2014}.
+
+A random forest is an ensemble  of decision $h(x; \theta_b) b=1, \ldots, B$ where
+$x$ is the covariate vector of length $p$ with associated
+random vector $X$ and the $\theta_b$ are i.i.d. random vectors. 
+The training  data is independently drawn from the joint distribution of $(X,Y)
+and comprises $n$  $(p+1)$-tuples (x_1,y_1),\ldots, (x_n,y_n)$.
+For regression, the random forest prediction is the unweighted average over
+the collection: $\bar{h}(x)=\dfrac{1}{K}\Sigma_{k=1}^K h(x; \theta_k)$.
+
+The average prediction error for an individual tree $h(X; \theta)$ as
+\begin{equation}
+PE_t E_\theta E_{X,Y} (Y-h(X; \theta))^2
+\end{equation}
+Assume that for all  $\theta$ the tree is unbiased, i.e., $EY=E_Xh(X; \theta)$. Then
+\begin{equation}
+PE_f = \rho PE_t
+\end{equation}
+where $\rho$ is the weighted correlation between residuals
+$(Y-h(X;\theta)$ and  $(Y-h(X;\theta))$
+for independent $\theta, theta^\prime$.
+The inequality (3) pinpoints what is required for accurate random forest regression:
+(i) low correlation between residuals of differing tree members of
+the forest, and (ii) low prediction error for the individual trees. Further, the
+random forest will, in expectation, decrease the individual tree error, $PE_t$ , by
+the factor $\rho$. Accordingly, the randomization injected strives for low correlation.
+The strategy employed to achieve these ends is as follows:
+1. To keep individual error low, grow trees to maximum depth.
+bench.rf.
+
+Random Forests~\cite{Breiman.2001} (RF) 
 apply the technique of bootstrap aggregation or bagging to decision trees. The training set is randomly sampled $B$

### 2016-07-06 doc/publications/plos16/plos16.tex
changes to plos16.tex
added to .bib
changed layout of page
changes text

--- 2016-06-29 doc/publications/plos16/plos16.tex
@@
 
-\documentclass[10pt,letterpaper]{article}
+\documentclass[10pt,a4paper]{article}  %was \documentclass[10pt,letterpaper]{article}
+
 \usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}
@@
 
+\let\oldmarginpar\marginpar
+\renewcommand\marginpar[1]{\-\oldmarginpar[\raggedleft\footnotesize #1]%hv
+{\raggedright\footnotesize #1}}
 \reversemarginpar
+\setlength{\marginparwidth}{5cm}
 
@@
   simulation data \cite[]{Genuer.et.al.2010}? Has Aidan done this already?
+\item I dont think I can get R to handle chromosme 1 etc. May have to run CF on chromosme 22, see table
+  \ref{table:smaller_data}
+\item check Piotr's simulation
+\item run an argument as follows:
+  \begin{itemize}
+  \item variable  selection may succeed where accurate prediction does not
+  \item consider  \cite{Segal.2004} examples
+  \item  perhaps we argue that in the case where the relationship of $X$ and $y$ takes a particular form we can recover
+    the reltionship for very large $p$p
+  \end{itemize}
+\item \cite{Tuv.et.al.2009} feature selection via adding permuted variables. Do we want to consider this?
+\item chek out varSelRF
 \end{itemize}
@@
 applications as having more features $p$ than samples $n$ causes the model to overfit the data and generalize poorly.
-This is known as the `curse of dimensionality'~\cite{Bauer2014}.  Random Forests~\cite{Breiman.2001} (RF) as an ensemble of weak learners 
+This is known as the `curse of dimensionality'~\cite{Bauer2014}.  
+\marginpar{is this the reference we want for the `curse of dimensionality'}
+
+\subsection*{RF}
+
+Random Forests~\cite{Breiman.2001} (RF) apply the technique of bootstrap aggregation or bagging to decision trees.  The training
+data is independently drawn from the joint distribution of $(X,Y)$ and comprises $n$ $(p+1)$-tuples $(x_1,y_1),\ldots, (x_n,y_n)$.
+$X$ and $\theta_b$ for $b=1,\ldots,B$ are i.i.d. random vectors and a tree is grown on each of these $\theta_b$  samples 
+($X$ is randomly sampled $B$ times with replacement and a tree is grown on each of these $B$ samples).
+
+The results are combined in an appropriate way, i.e. for a regression problem,
+\begin{equation*}
+{\bar {h}}={\frac {1}{B}}\sum _{b=1}^{B} h(x;\theta_b).
+\end{equation*}
+%where $x$ is the covariate vector of length $p$
+%with associated random vector $X$ and the $\theta_b$ are i.i.d. random vectors.
+
+The average prediction error for an individual tree $h(X; \theta)$ is
+\begin{equation}
+PE_t = E_\theta E_{X,Y} (Y-h(X; \theta))^2.
+\end{equation}
+Assume that, for all  $\theta$, the tree is unbiased, i.e., $EY= E_X h(X; \theta)$. Then
+\begin{equation}
+PE_f \leq \rho PE_t
+\end{equation}
+where $\rho$ is the weighted correlation between residuals $(Y-h(X;\theta))$ and $(Y-h(X;\theta^\prime))$ for independent $\theta,
+\theta^\prime$.  
+So what is required for  accurate random forest regression is (i) low correlation between residuals of differing tree in
+the forest, and (ii) low prediction error for the individual trees \cite{Segal.2004}.
+
+\subsection*{benefits of RF}
+
 \marginpar{weak learners only in a misuse of the term. ``weak learners'' was applied to decision stumps}
@@
 %http://stats.stackexchange.com/questions/173390/gradient-boosting-tree-vs-random-forest
-overcomes this issue and is hence ideally suited to this application.  Furthermore, RF are able to capture interactions
-between features, which is of importance for modelling complex polygenic diseases.
+%overcomes this issue and is hence ideally suited to this application.  Furthermore, RF are able to capture interactions
+%between features, which is of importance for modelling complex polygenic diseases.
 
@@
 % overfitting and variance.
+Decision trees have a number of desirable features, they:
+\begin{itemize}
+  \item are scale invariant;
+  \item can handle categorical and real valued inputs;
+  \item can handle missing values;
+  \item are insensitive to useless predictors;
+  \item are able to capture interactions between features (which is of importance for modelling complex polygenic diseases).
+  \end{itemize}
+However, the tree fitting algorithm is greedy and may generate an unstable model. That is, a small change in the data may
+lead to a very different model. 
+\marginpar{a potential problem  with tree models is the tendencey to split on categorical variables with a high number of
+  levels, even when better splits are available}
+Random forests are a variance reduction technique. They take a tree models which have a low bias but a high variance and
+by combining them reduce the variance. The prediction error is the sum of the variance and the bias squared.
+
+The benefits of random forests models are
+  \begin{itemize}
+  \item they give an out-of-bag estimate of model accurace;
+  \item they provide a measure of variable importance;
+  \item they inherit many of the qualities of tree models.
+\marginpar{but not the handling of missing velues via surrogate variables. This would be difficult as the trees as split
+  on random subsets of variables}
+  \end{itemize}
 
@@
 
-As discussed in the Introduction, random forests are robust against overfitting
-\marginpar{strong claim -- see disussion in Segal 2004, perhaps}
-and hence can generalise for datasets where $p > n$, however here we demonstrate that this holds true in extreme cases where $p \gg n$.
+As discussed in the Introduction,
+\marginpar{I think this has been removed from the introduction}
+random forests are resistent against overfitting.
+However it is not true that they will not overfit; see \cite{Segal.2004} where examples of RF overfitting are given 
+\marginpar{discuss these examples}
+
+Subject to certain conditions, RF will not overfit even in the  $p > n$ case. 
+Here we demonstrate that this may hold true in extreme cases where $p \gg n$.
 
@@
 
+\begin{table}
+\begin{minipage}{\textwidth}
+\begin{center}
+\begin{tabular}{|p{4.2cm}|l|l|l|l|l|l|l|}
+\hline
+\bf{Genome }  & \bf{method} & \bf{Accuracy} & \bf{Runtime} & \bf{Memory/Node} \\
+\hline
+ chr 22 - phase1 (2.1GB)  &  RF (R)\footnote{10 nodes, 500 trees, dpas03} 
+                                   &  0.456 & 2.89 & 18GB \\ \hline
+\end{tabular}
+\medskip
+\begin{flushleft} 
+\caption{some R results on smaller data sets.}
+\label{table:smaller_data}
+\end{flushleft}
+\end{center}
+%\end{sidewaystable}
+\end{minipage}
+\end{table}
+
+% \begin
+% \begin{table}[!ht]
+% \caption{
+% {\bf Performance comparison between the different Machine Learning algorithms.}}
+% \begin{tabular}{|p{4.2cm}|l|l|l|l|l|l|l|}
+% \hline
+% \bf{Genome }  & \bf{method} & \bf{Accuracy} & \bf{Runtime} & \bf{Memory/Node} \\
+% \hline
+%  chr 22 - phase1 (2.1GB)  &  RF (R) &  0.456 & 2.89 & 18GB \\ \hline
+% \end{tabular}
+% \begin{flushleft} 
+% \end{flushleft}
+% \label{table1}
+% %\end{adjustwidth}
+% \end{table}
+
 \subsection*{Parameter settings}
@@
 useful ranking of variable importance.
-
-\subsection*{decision trees }
-\begin{itemize}
-\item benefits os a decision tree
-  \begin{itemize}
-  \item invariant to scale of the input data
-  \item handles categorical and real valued inputs 
-  \item can handle missing values 
-  \item insensitive to useless predictors
-  \end{itemize}
-\item benefits of random forests
-  \begin{itemize}
-  \item variable importance 
-  \item inherits the qualities of tree models
-  \end{itemize}
-\end{itemize}
 

### 2016-07-20 doc/publications/plos16/plos16.tex
Update discussion and table, remove some latin, update 20 2016 template

@@
 % Template for PLoS
-% Version 3.1 February 2015
-%
-% To compile to pdf, run:
-% latex plos.template
-% bibtex plos.template
-% latex plos.template
-% latex plos.template
-% dvipdf plos.template
+% Version 3.3 June 2016
 %
@@
 % Once your paper is accepted for publication, 
-% PLEASE REMOVE ALL TRACKED CHANGES in this file and leave only
-% the final text of your manuscript.
+% PLEASE REMOVE ALL TRACKED CHANGES in this file 
+% and leave only the final text of your manuscript. 
+% PLOS recommends the use of latexdiff to track changes during review, as this will help to maintain a clean tex file.
+% Visit https://www.ctan.org/pkg/latexdiff?lang=en for info or contact us at latex@plos.org.
+%
 %
@@
 %
-% Please do not create a heading level below \subsection. For 3rd level headings, use \paragraph{}.
+% The manuscript LaTeX source should be contained within a single file (do not use \input, \externaldocument, or similar commands).
 %
@@
 % - Figures containing multiple panels/subfigures must be combined into one image file before submission.
-% For figure citations, please use "Fig." instead of "Figure".
-% See http://www.plosone.org/static/figureGuidelines for PLOS figure guidelines.
+% For figure citations, please use "Fig" instead of "Figure".
+% See http://journals.plos.org/plosone/s/figures for PLOS figure guidelines.
 %
 % Tables should be cell-based and may not contain:
-% - tabs/spacing/line breaks within cells to alter layout or alignment
-% - vertically-merged cells (no tabular environments within tabular environments, do not use \multirow)
-% - colors, shading, or graphic objects
-% See http://www.plosone.org/static/figureGuidelines#tables for table guidelines.
+% - spacing/line breaks within cells to alter layout or alignment
+% - do not nest tabular environments (no tabular environments within tabular environments)
+% - no graphics or colored text (cell background color/shading OK)
+% See http://journals.plos.org/plosone/s/tables for table guidelines.
 %
@@
 % Below are a few tips to help format your equations and other special characters according to our specifications.
-For more tips to help reduce the possibility of formatting errors during conversion, please see our LaTeX guidelines at http://www.plosone.org/static/latexGuidelines
+For more tips to help reduce the possibility of formatting errors during conversion, please see our LaTeX guidelines at http://journals.plos.org/plosone/s/latex
 %
-% Please be sure to include all portions of an equation in the math environment.
+% For inline equations, please be sure to include all portions of an equation in the math environment.
+For example, x$^2$ is incorrect; this should be formatted as $x^2$ (or $\mathrm{x}^2$ if the romanized font is desired).
 %
-% Do not include text that is not math in the math environment. For example, CO2 will be CO\textsubscript{2}.
+% Do not include text that is not math in the math environment. For example, CO2 should be written as CO\textsubscript{2} instead of CO$_2$.
 %
@@
 % For inline equations, please do not include punctuation (commas, etc) within the math environment unless this is part of the equation.
+%
+% When adding superscript or subscripts outside of brackets/braces, please group using {}.  For example, change "[U(D,E,\gamma)]^2" to "{[U(D,E,\gamma)]}^2". 
+%
+% Do not use \cal for caligraphic font.  Instead, use \mathcal{}
 %
@@
 
-\documentclass[10pt,a4paper]{article}  %was \documentclass[10pt,letterpaper]{article}
+\documentclass[10pt,letterpaper]{article}
+\usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}
 
-\usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}
+% amsmath and amssymb packages, useful for mathematical formulas and symbols
+\usepackage{amsmath,amssymb}
 
@@
 % Use Unicode characters when possible
-\usepackage[utf8]{inputenc}
+\usepackage[utf8x]{inputenc}
 
@@
 \usepackage{textcomp,marvosym}
-
-% fixltx2e package for \textsubscript
-\usepackage{fixltx2e}
-
-% amsmath and amssymb packages, useful for mathematical formulas and symbols
-\usepackage{amsmath,amssymb}
 
@@
 
-% rotating package for sideways tables
-\usepackage{rotating}
+% color can be used to apply background shading to table cells only
+\usepackage[table]{xcolor}
+
+% array package and thick rules for tables
+\usepackage{array}
+
+% create "+" rule type for thick vertical lines
+\newcolumntype{+}{!{\vrule width 2pt}}
+
+% create \thickcline for thick horizontal lines of variable length
+\newlength\savedwidth
+\newcommand\thickcline[1]{%
+  \noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
+  \cline{#1}%
+  \noalign{\vskip\arrayrulewidth}%
+  \noalign{\global\arrayrulewidth\savedwidth}%
+}
+
+% \thickhline command for thick horizontal lines that span the table
+\newcommand\thickhline{\noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
+\hline
+\noalign{\global\arrayrulewidth\savedwidth}}
 
@@
 %\doublespacing
-
-\let\oldmarginpar\marginpar
-\renewcommand\marginpar[1]{\-\oldmarginpar[\raggedleft\footnotesize #1]%hv
-{\raggedright\footnotesize #1}}
-\reversemarginpar
-\setlength{\marginparwidth}{5cm}
 
@@
 \usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off]{caption}
+\renewcommand{\figurename}{Fig}
 
@@
 \fancyhf{}
-\lhead{\includegraphics[width=2.0in]{PLOS-submission}}
+\setlength{\headheight}{27.023pt}
+\lhead{\includegraphics[width=2.0in]{PLOS-submission.eps}}
 \rfoot{\thepage/\pageref{LastPage}}
@@
 
+\usepackage{multirow}
+
 \begin{document}
-\vspace*{0.35in}
+\vspace*{0.2in}
 
@@
 {\Large
-\textbf\newline{Cursed Forest - A random forest implementation for `big' and `wide' data}
+\textbf\newline{Cursed Forest - A random forest implementation for `big' and `wide' data} % Please use "title case" (capitalize all terms in the title except conjunctions, prepositions, and articles).
 }
@@
 \item I dont think I can get R to handle chromosme 1 etc. May have to run CF on chromosme 22, see table
-  \ref{table:smaller_data}
+  \ref{table:smaller_data} -- On it, Aidan
 \item check Piotr's simulation
@@
 \section*{Abstract}
-Lorem ipsum dolor sit amet, consectetur adipiscing elit. Curabitur eget porta erat. Morbi consectetur est vel gravida
-pretium. Suspendisse ut dui eu ante cursus gravida non sed sem. Nullam sapien tellus, commodo id velit id, eleifend
-volutpat quam. Phasellus mauris velit, dapibus finibus elementum vel, pulvinar non tellus. Nunc pellentesque pretium
-diam, quis maximus dolor faucibus id. Nunc convallis sodales ante, ut ullamcorper est egestas vitae. Nam sit amet enim
-ultrices, ultrices elit pulvinar, volutpat risus.
+The abstract.
 
@@
 \subsection*{VariantSpark.}
+Traditionally, VariantSpark made use of Spark's machine learning algrorithms `Spark ML'. These algorithms require data in a `DataFrame',
+similar to DataFrames in R.
+
+We use the same pre-processing algorithms in CursedForests. One difference is that we
 
@@
 
-\section*{Random Forests converges on wide data}
+\subsection*{Random Forests converges on wide data}
 
@@
 
-\subsection*{CursedForest processes genome wide variant data}
-\label{comp}
-Spark ML vs CursedForest vs Kmeans vs LR on small and full version (where applicable) of 1000G (performance and
-time/mem)
+\subsection*{CursedForest processes genome-wide variant data}
+%\label{comp}
+%Spark ML vs CursedForest vs Kmeans vs LR on small and full version (where applicable) of 1000G (performance and
+%time/mem)
 
-% In this section we compare Apache's higher level machine learning library, Spark ML, with MLLib, the standard
-% low-level library. From the results, we see that times are similar.  There are many practical benefits for using Spark
-% ML with VariantSpark, with a selection of these benefits derived from the `pipeline' system Spark ML uses.  One
-% benefit is that hyperparameter optimisation is built in. Although possible in
+Traditionally, VariantSpark used the random forest algorithm from SparkML. While SparkML algorithms demonstrate
+scalability when dealing with a large number of samples, this scalability quickly breaks down as we include more features.
+
+One of the restrictions preventing SparkML algorithms from scaling to high-dimensionality data is the fact that the feature-set
+for each individual is stored as a `Vector'. In Spark, Vectors are unable to be split and to process multiple vectors together,
+the entirety of each vector must be loaded into memory on the same node. Therefore, as dimensionality increases, the
+size of these Vectors must also increase, and subsequently this results in higher memory requirents.
+
+On the other hand, CursedForest (CF) is specifically designed to handle wide ``cursed" data. It avoids the correlation between
+memory and dimensionality by not using Vectors. Data is instead stored in a columnar form in a Parquet format.
 
 \begin{table}[!ht]
-%\begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
+\begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
 \caption{
-{\bf Performance comparison between the different Machine Learning algorithms.}}
-\begin{tabular}{|l|l|l|l|l|l|l|l|}
+{\bf Performance comparison between the different machine learning algorithms.}}
+\begin{tabular}{|l+l|l|l|l|l|l|l|}
 \hline
@@
 \hline
-8\% (chr1) & CursedForest & XX & 1hr 58min 23sec & 4GB \\ \hline
-          & RF (Spark ML) & XX & 8hr 29min 07sec & 8GB \\ \hline
-          &  RF (R) & &  &\\ \hline
-          &  LR & &  &\\ \hline
-XX\% & CursedForest & &  &\\ \hline
-          & RF(Spark ML) & &  &\\ \hline
-          &  RF(R) & &  &\\ \hline
+\multirow{3}{*}{8\% (chr1: 6,450,364)} & CursedForest & XX & 1hr 58min 23sec & 4GB \\
+& RF (Spark ML) & 0.94 & 8hr 29min 07sec & 8GB \\
+& LR &  & 10hr 30min & 8GB\\ \hline
+
+\multirow{3}{*}{24\% (chr1-3: 19,328,051)}  & CursedForest & &  &\\
+& RF(Spark ML) & &  &\\
 & LR & &  &\\ \hline
 
-100\% & CursedForest & 0.96 & 7hr 14min 40sec & 8GB \\ \hline
-          & RF (Spark ML) & - & - & - \\ \hline
-          &  RF (R) & &  &\\ \hline
-          &  LR & - & - & - \\ \hline
+\multirow{4}{*}{100\% (chr1-22: 81,047,467)} & CursedForest & 0.96 & 7hr 14min 40sec & 8GB \\ 
+& RF (Spark ML) & - & - & - \\
+& LR & - & - & - \\ 
 & kmeans (Spark ML) & 0.82 & 30hr 44min 00sec & 24GB \\ \hline
@@
 \begin{flushleft} 
-  Table notes Phasellus venenatis, tortor nec vestibulum mattis, massa tortor interdum felis, nec pellentesque metus
-  tortor nec nisl. Ut ornare mauris tellus, vel dapibus arcu suscipit sed.
+Comparison of different various machine learning libraries on different subsets of variant data 
+from the 1000 Genomes Project.
+
 \end{flushleft}
 \label{table1}
-%\end{adjustwidth}
+\end{adjustwidth}
 \end{table}
@@
 \hline
-\bf{Genome }  & \bf{method} & \bf{Accuracy} & \bf{Runtime} & \bf{Memory/Node} \\
+\bf{method} & \bf{Accuracy} & \bf{Runtime} & \bf{Memory/Node} \\
 \hline
- chr 22 - phase1 (2.1GB)  &  RF (R)\footnote{10 nodes, 500 trees, dpas03} 
+RF (R)\footnote{10 nodes, 500 trees, dpas03} 
                                    &  0.456 & 2.89 & 18GB \\ \hline
+CursedForest\footnote{something}
+                                   & X & X & X \\ \hline
 \end{tabular}
@@
 \begin{flushleft} 
-\caption{some R results on smaller data sets.}
+\caption{Comparion between Random Forests on R and CursedForests on a smaller
+dataset. Chromosome 22 - phase1 (2.1GB) }
 \label{table:smaller_data}
@@
 % \hline
-%  chr 22 - phase1 (2.1GB)  &  RF (R) &  0.456 & 2.89 & 18GB \\ \hline
+%  chr 	 - phase1 (2.1GB)  &  RF (R) &  0.456 & 2.89 & 18GB \\ \hline
 % \end{tabular}
@@
 
-\subsection*{James' dataset}
-Comparing CursedForest with feature selection + linear model (performance and time/mem)
+%\subsection*{James' dataset}
+%Comparing CursedForest with feature selection + linear model (performance and time/mem)
 
@@
 
-\subsection*{S1 Fig}
-\label{S1_Fig}
-{\bf Lorem Ipsum.} Maecenas convallis mauris sit amet sem ultrices gravida.
-Etiam eget sapien nibh.
-Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula.
-Curabitur fringilla pulvinar lectus consectetur pellentesque.
-
-\subsection*{S2 Fig}
-\label{S2_Fig}
-{\bf Lorem Ipsum.} Maecenas convallis mauris sit amet sem ultrices gravida.
-Etiam eget sapien nibh.
-Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula.
-Curabitur fringilla pulvinar lectus consectetur pellentesque.
-
 \subsection*{S1 Table}

### 2016-07-20 doc/publications/plos16/plos16.tex
merge the R table into the main table

@@
 
-On the other hand, CursedForest (CF) is specifically designed to handle wide ``cursed" data. It avoids the correlation between
-memory and dimensionality by not using Vectors. Data is instead stored in a columnar form in a Parquet format.
+On the other hand, CursedForest (CF) is specifically designed to handle wide ``cursed" data. It avoids the relation between
+memory and dimensionality, by not using Vectors. Instead, CursedForest stores data in the Parquet columnar format.
 
@@
 \hline
-\multirow{3}{*}{8\% (chr1: 6,450,364)} & CursedForest & XX & 1hr 58min 23sec & 4GB \\
+
+\multirow{3}{*}{phase1\_chr22: 1,092 x 490,036} & CursedForest &  &  &  \\
+& RF (Spark ML) &  &  &  \\
+& RF (R) & 0.456 & 2.89 & 18GB \\ \hline
+
+\multirow{3}{*}{phase3\_chr1: 2,504 x 6,450,364} & CursedForest & XX & 1hr 58min 23sec & 4GB \\
 & RF (Spark ML) & 0.94 & 8hr 29min 07sec & 8GB \\
@@
 
-\multirow{3}{*}{24\% (chr1-3: 19,328,051)}  & CursedForest & &  &\\
+\multirow{3}{*}{phase3\_chr1-3: 2,504 x 19,328,051}  & CursedForest & &  &\\
 & RF(Spark ML) & &  &\\
@@
 
-\multirow{4}{*}{100\% (chr1-22: 81,047,467)} & CursedForest & 0.96 & 7hr 14min 40sec & 8GB \\ 
+\multirow{4}{*}{phase3\_chr1-22: 2,504 x 81,047,467} & CursedForest & 0.96 & 7hr 14min 40sec & 8GB \\ 
 & RF (Spark ML) & - & - & - \\
@@
 Comparison of different various machine learning libraries on different subsets of variant data 
-from the 1000 Genomes Project.
+from the 1000 Genomes Project. Size of the data is individuals x variants
 
@@
 
-\begin{table}
-\begin{minipage}{\textwidth}
-\begin{center}
-\begin{tabular}{|p{4.2cm}|l|l|l|l|l|l|l|}
-\hline
-\bf{method} & \bf{Accuracy} & \bf{Runtime} & \bf{Memory/Node} \\
-\hline
-RF (R)\footnote{10 nodes, 500 trees, dpas03} 
-                                   &  0.456 & 2.89 & 18GB \\ \hline
-CursedForest\footnote{something}
-                                   & X & X & X \\ \hline
-\end{tabular}
-\medskip
-\begin{flushleft} 
-\caption{Comparion between Random Forests on R and CursedForests on a smaller
-dataset. Chromosome 22 - phase1 (2.1GB) }
-\label{table:smaller_data}
-\end{flushleft}
-\end{center}
+%\begin{table}
+%\begin{minipage}{\textwidth}
+%\begin{center}
+%\begin{tabular}{|p{4.2cm}|l|l|l|l|l|l|l|}
+%\hline
+%\bf{method} & \bf{Accuracy} & \bf{Runtime} & \bf{Memory/Node} \\
+%\hline
+%RF (R)\footnote{10 nodes, 500 trees, dpas03} 
+ %                                  &  0.456 & 2.89 & 18GB \\ \hline
+%CursedForest\footnote{something}
+ %                                  & X & X & X \\ \hline
+%\end{tabular}
+%\medskip
+%\begin{flushleft} 
+%\caption{Comparion between Random Forests on R and CursedForests on a smaller
+%dataset. Chromosome 22 - phase1 (2.1GB) }
+%\label{table:smaller_data}
+%\end{flushleft}
+%\end{center}
 %\end{sidewaystable}
-\end{minipage}
-\end{table}
+%\end{minipage}
+%\end{table}
 

### 2016-07-22 doc/publications/plos16/plos16.tex
edits to page dimensions and marginpar
move description of DT, aded discussion of Strobl

@@
 
-\documentclass[10pt,letterpaper]{article}
+\documentclass[10pt,a4paper]{article}  %\documentclass[10pt,letterpaper]{article}
 \usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}
@@
 \usepackage{multirow}
+
+%%local edits (rob)
+\let\oldmarginpar\marginpar
+\renewcommand\marginpar[1]{\-\oldmarginpar[\raggedleft\footnotesize #1]%hv
+{\raggedright\footnotesize #1}}
+\reversemarginpar
+\setlength{\marginparwidth}{5cm}
 
@@
 
+Decision trees have a number of desirable features, they:
+\begin{itemize}
+  \item are scale invariant;
+  \item can handle categorical and real valued inputs;
+  \item can handle missing values;
+  \item are insensitive to useless predictors;
+  \item are able to capture interactions between features (which is of importance for modelling complex polygenic diseases).
+  \end{itemize}
+However, the tree fitting algorithm is greedy and may generate an unstable model. That is, a small change in the data may
+lead to a very different model. 
+\marginpar{a potential problem  with tree models is the tendencey to split on categorical variables with a high number of
+  levels, even when better splits are available}
 \subsection*{RF}
@@
 \end{equation*}
+\marginpar{make it a classification example}
 %where $x$ is the covariate vector of length $p$
@@
 % overfitting and variance.
-Decision trees have a number of desirable features, they:
-\begin{itemize}
-  \item are scale invariant;
-  \item can handle categorical and real valued inputs;
-  \item can handle missing values;
-  \item are insensitive to useless predictors;
-  \item are able to capture interactions between features (which is of importance for modelling complex polygenic diseases).
-  \end{itemize}
-However, the tree fitting algorithm is greedy and may generate an unstable model. That is, a small change in the data may
-lead to a very different model. 
-\marginpar{a potential problem  with tree models is the tendencey to split on categorical variables with a high number of
-  levels, even when better splits are available}
+
 Random forests are a variance reduction technique. They take a tree models which have a low bias but a high variance and
@@
 Here we demonstrate that this may hold true in extreme cases where $p \gg n$.
+
+\cite{Strobl.et.al.2007} demosnstrate that a RF algorithm is subject to a bias in variable selection via two mechanisms:
+\begin{itemize}
+\item variable scale (for $X_I$ continuous) and number of levels (for $X_i$ discrete) inroduces a bias
+\item boosting introduces a bias.
+\end{itemize}
+These effects will be apparent in the CF algorithm as well. \marginpar{we can fix the boosting effect}
+However, there are many examples, (particularly in genomics) whre we have data sets that are
+\begin{itemize}
+\item very wide
+\item all variables have the same number of levels (measures at different points along the genome)
+\item the hypothesied functional relationship is of a form amenable to a DT.
+\end{itemize}
 

### 2016-07-22 doc/publications/plos16/plos16.tex
remove seconds

--- 2016-07-20 doc/publications/plos16/plos16.tex
@@
 %\multicolumn{4}{|l|}{\bf Heading1} & \multicolumn{4}{|l|}{\bf Heading2}\\ \hline
-\bf{Genome \%}  & \bf{method} & \bf{Accuracy} & \bf{Runtime} & \bf{Memory/Node} \\
+\bf{Genome \%}  & \bf{Method} & \bf{Accuracy} & \bf{Runtime} & \bf{Memory/Node} \\
 \hline
@@
 
-\multirow{3}{*}{phase3\_chr1: 2,504 x 6,450,364} & CursedForest & XX & 1hr 58min 23sec & 4GB \\
-& RF (Spark ML) & 0.94 & 8hr 29min 07sec & 8GB \\
-& LR &  & 10hr 30min & 8GB\\ \hline
+\multirow{3}{*}{phase3\_chr1: 2,504 x 6,450,364} & CursedForest & XX & 1hr 58min & 4GB \\
+& RF (Spark ML) & 0.94 & 8hr 29min & 8GB \\
+& LR &  & 2hr 6min & 8GB\\ \hline
 
-\multirow{3}{*}{phase3\_chr1-3: 2,504 x 19,328,051}  & CursedForest & &  &\\
+\multirow{3}{*}{phase3\_chr1-3: 2,504 x 19,328,051}  & CursedForest & 0.046 () & 4hr 0min & 4GB\\
 & RF(Spark ML) & &  &\\
-& LR & &  &\\ \hline
+& LR & & 14hr 48min & 8GB \\ \hline
 
-\multirow{4}{*}{phase3\_chr1-22: 2,504 x 81,047,467} & CursedForest & 0.96 & 7hr 14min 40sec & 8GB \\ 
+\multirow{4}{*}{phase3\_chr1-22: 2,504 x 81,047,467} & CursedForest & 0.96 & 7hr 14min & 8GB \\ 
 & RF (Spark ML) & - & - & - \\
 & LR & - & - & - \\ 
-& kmeans (Spark ML) & 0.82 & 30hr 44min 00sec & 24GB \\ \hline
+& kmeans (Spark ML) & 0.82 & 30hr 44min & 24GB \\ \hline
 \end{tabular}

### 2016-07-31 doc/publications/plos16/plos16.tex
edits to CursedForest.bib plos16.tex

--- 2016-07-22 doc/publications/plos16/plos16.tex
@@
 \item \cite{Tuv.et.al.2009} feature selection via adding permuted variables. Do we want to consider this?
-\item chek out varSelRF
+\item check out varSelRF
+I think the structure of the first section should be
+\begin{itemize}
+\item  convergence results from Biau 2012
+\item simulations reproduced from Genuer et al 2010
+\item we extend the simulations to many more variables
+\item discussion of the effect of the ntree and mtry parameters for variable selection with many noise variables
+\item  in the case of some functions (step, ramp) we can recover the significant variables in the presence of a very large number of noise variables
+\item in the case of genomic data, our functions of interest are more likely to be stop or ramp functions rather than harmonic functions. So we have some hope.
+\end{itemize}
+
+\item check out 
+  \begin{itemize}
+  \item D\'iaz-Uriarte, R., \& Alvarez de Andr\'es, S. (2006). Gene selection and classification of microarray data using random
+    forest. BMC Bioinformatics, 7, 3. http://doi.org/10.1186/1471-2105-7-3
+  \item Goldstein et al. (2011)
+  \item Chen (2012)
+  \item VSURF and  varSelRF
+  \end{itemize}
 \end{itemize}
@@
 
-The digital revolution or `datafication' sees a dramatic increase in data collected about every aspect of   \marginpar{is seeing, will see}
+The digital revolution is seeing a dramatic increase in data collected about every aspect of 
 life~\cite{Loebbecke2015}.  These dataset not only grow vertically by capturing more and more events but also
@@
 Distributing the compute tasks is one strategy to overcome this challenge. Apache Spark, based on Scala, allows for  
-\marginpar{is it based on scale or does it have a scale API?}
+\marginpar{is it based on Scala or does it have a Scala API?}
 distributed applications to scale to thousands of nodes.  A Spark application runs on a `driver' node, and
@@
 This is known as the `curse of dimensionality'~\cite{Bauer2014}.  
-\marginpar{is this the reference we want for the `curse of dimensionality'}
+\marginpar{is this the reference we want for the ``curse of dimensionality''  Bellman (1961) is a more ususl reference}
 
@@
 lead to a very different model.
-\marginpar{a potential problem  with tree models is the tendencey to split on categorical variables with a high number of
-  levels, even when better splits are available}
+
 \subsection*{RF}
@@
 
-The results are combined in an appropriate way, i.e. for a regression problem,
+The results are combined in an appropriate way. In a classification problem we take the majority vote for the trees over
+the $i=1,..,Q$ classes
 \begin{equation*}
-{\bar {h}}={\frac {1}{B}}\sum _{b=1}^{B} h(x;\theta_b).
+{\bar {h}}=  \arg \max_i \left(\sum_b I(h(x;\theta_b)=i)\right).
 \end{equation*}
-\marginpar{make it a classification example}
-%where $x$ is the covariate vector of length $p$
-%with associated random vector $X$ and the $\theta_b$ are i.i.d. random vectors.
+
+Random forests are a variance reduction technique. They take a tree models, which have a low bias but a high variance, and
+by combining them reduce the variance. The prediction error is the sum of the variance and the bias squared.
 
@@
 
-\subsection*{benefits of RF}
-
-\marginpar{weak learners only in a misuse of the term. ``weak learners'' was applied to decision stumps}
 %http://stats.stackexchange.com/questions/77018/random-forest-is-it-a-boosting-algorithm
@@
 
-Random forests are a variance reduction technique. They take a tree models which have a low bias but a high variance and
-by combining them reduce the variance. The prediction error is the sum of the variance and the bias squared.
-
 The benefits of random forests models are
@@
   \item they provide a measure of variable importance;
-  \item they inherit many of the qualities of tree models.
-\marginpar{but not the handling of missing velues via surrogate variables. This would be difficult as the trees as split
-  on random subsets of variables}
+  \item they inherit some of the qualities of tree models (but not, for example, the handling of missing values via surrogate
+    variables. This would be difficult as the trees as split on random subsets of variables).
   \end{itemize}
+
+Since the original implmentation of random forests there have been a number of developments leading to a better understanding of
+the algorithm. 
+\begin{itemize}
+\item random forests are resistent against overfitting. However it is not true that they will not overfit; see \cite{Segal.2004}
+  where an example of RF overfitting is  given, caused by the fact that successive trees were correlated;
+\item \cite{Strobl.et.al.2007} demonstrates that the RF algorithm is subject to a bias in variable selection via two mechanisms:
+  \begin{enumerate}
+  \item variable scale (for $X_I$ continuous) and number of levels (for $X_i$ discrete) inroduces a bias. A uninformative  variable with a large
+    number of levels may be selected over a more informative  variable with fewer levels;
+  \item bagging introduces a bias, but sampleing about $(1- 1/e) \approx 0.632$  of the data without 
+  \end{enumerate}
+\end{itemize}
+
+\cite{Strobl.et.al.2007} introoduce changes to the RF algorithm to fix 1) and 2). 
+However the level effect will be apparent in the CF algorithm as well. \marginpar{we can fix the boosting effect}
+However, there are many examples, (particularly in genomics) whre we have data sets that are
+\begin{itemize}
+\item very wide
+\item all variables have the same number of levels (measures at different points along the genome)
+\item the hypothesied functional relationship is of a form amenable to a DT.
+\end{itemize}
+
+Biau {\it et al.}~\cite{Biau.2012} provide a proof that a random forest model will converge at a rate that depends on
+the cardinality of the set $S$ of ``strong predictors'' rather than on the number of variables $p$. That is, given a
+function $y=f(S)$ depending on a set of variables $S$, a random forest will converge to the true function $y$ with a
+rate that depends on the size of $S$ and not the number of noise variables in the data set. If the size of $S$ is small
+compared to the total number of variables, then the rate of convergence will be quicker.
+\marginpar{say something about the rate of convergence}
+
+Subject to certain conditions, RF will not overfit even in the  $p > n$ case. 
+Here we demonstrate that this may hold true in extreme cases where $p \gg n$.
 
@@
 % GWAS dataset
-
-\subsection*{Random Forests converges on wide data}
-
-As discussed in the Introduction,
-\marginpar{I think this has been removed from the introduction}
-random forests are resistent against overfitting.
-However it is not true that they will not overfit; see \cite{Segal.2004} where examples of RF overfitting are given 
-\marginpar{discuss these examples}
-
-Subject to certain conditions, RF will not overfit even in the  $p > n$ case. 
-Here we demonstrate that this may hold true in extreme cases where $p \gg n$.
-
-\cite{Strobl.et.al.2007} demosnstrate that a RF algorithm is subject to a bias in variable selection via two mechanisms:
-\begin{itemize}
-\item variable scale (for $X_I$ continuous) and number of levels (for $X_i$ discrete) inroduces a bias
-\item boosting introduces a bias.
-\end{itemize}
-These effects will be apparent in the CF algorithm as well. \marginpar{we can fix the boosting effect}
-However, there are many examples, (particularly in genomics) whre we have data sets that are
-\begin{itemize}
-\item very wide
-\item all variables have the same number of levels (measures at different points along the genome)
-\item the hypothesied functional relationship is of a form amenable to a DT.
-\end{itemize}
-
-Biau {\it et al.}~\cite{Biau.2012} provide a proof that a random forest model will converge at a rate that depends on
-the cardinality of the set $S$ of ``strong predictors'' rather than on the number of variables $p$. That is, given a
-function $y=f(S)$ depending on a set of variables $S$, a random forest will converge to the true function $y$ with a
-rate that depends on the size of $S$ and not the number of noise variables in the data set. If the size of $S$ is small
-compared to the total number of variables, then the rate of convergence will be quicker.
-\marginpar{say something about the rate of convergence}
 
@@
 \subsection*{Parameter settings}
-\marginpar{ was''Random forest is sensitive to tree depth''}
-In this setting, variable selection in very high dimensions, we nned to consider the parameter setting for the RF
-algorithm.
+We consider the parameter setting for the RF algorithm. We use the R terms from the package RandomForests
+\cite{Liaw.and.Weiner.2002} which incoporates the original Fortran code by Brieman and Cutler. We incprporate the 
+advice of \cite{Liaw.and.Weiner.2002}, which we have found mirrors our own experience.
 
-I will use the R terms from the package RandomForests whcih incoporates the original Fortran code by Brieman and
-Cutler. We consider the parameters:
 \begin{itemize}
+\item \texttt{ntree} the number of trees.  The number of trees necessary for good performance grows with the number of predictors.
+  \marginpar{can we do proxmiity?}  \cite{Liaw.and.Weiner.2002} suggest that a high \texttt{ntree} is necessary to get stable
+  estimates of variable importance and proximity; however, even though the variable importance measures may vary from run to run,
+  the ranking of the importances is quite stable.  In addition, we note that it is possible for a RF model to have a very poor
+  fit, and be useless for prediction, and still have a useful ranking of variable importance.
+\item \texttt{mtry} --  the number of variables considered at each split (if \texttt{mtry}=$p$, we have a boosted decison
+  tree model).  If one has a very large number of variables but expects only very few to be ``important'', using larger \texttt{mtry} may give
+  better performance.
+\item  For classification problems where the class frequencies are extremely unbalanced it may be necessary to change the prediction
+  rule \marginpar{can we do this}
 \item the size and complexity of the individual trees is controlled in RandomForests by setting \texttt{nodesize}, the
@@
   tree in the forest. The SparkML documentation somewhat unhelpfully sets \texttt{maxDepth=4} in their classification
-  example\footnote{the documentation is available at \url{http://spark.apache.org/docs/latest/mllib-ensembles.html} }.
-  In the ATCG example the RF code produced trees with depths of 30+. We note that Random forests are a way of averaging
+  example\footnote{The documentation is available at \url{http://spark.apache.org/docs/latest/mllib-ensembles.html}.}.
+
+  In the ATCG example the RandomForests code produced trees with depths of 30+. We note that Random forests are a way of averaging
   multiple deep decision trees, trained on different parts of the same training set, with the goal of reducing the
@@
 \marginpar{are we including this  example?  Supplementary info?}
-\item \texttt{ntree} the number of trees 
-\item \texttt{mtry} --  the number of variables considered at each split (if \texttt{mtry}=$p$, we have a boosted decison
-  tree model \marginpar{is this right?}
 \end{itemize}
@@
 \end{document}
+
+  \item D\iaz-Uriarte, R., \& Alvarez de Andr\es, S. (2006). Gene selection and classification of microarray data using random
+    forest. BMC Bioinformatics, 7, 3. http://doi.org/10.1186/1471-2105-7-3
+  \item Goldstein, B. A., Polley, E. C., & Briggs, F. B. S. (2011). Random Forests for Genetic Association Studies. Statistical
+    Applications in Genetics and Molecular Biology, 10(1), 32. http://doi.org/10.2202/1544-6115.1691
+
+@article{Chen2012323,
+title = "Random forests for genomic data analysis ",
+journal = "Genomics ",
+volume = "99",
+number = "6",
+pages = "323 - 329",
+year = "2012",
+note = "",
+issn = "0888-7543",
+doi = "http://dx.doi.org/10.1016/j.ygeno.2012.04.003",
+url = "http://www.sciencedirect.com/science/article/pii/S0888754312000626",
+author = "Xi Chen and Hemant Ishwaran",
+keywords = "Random forests",
+keywords = "Random survival forests",
+keywords = "Classification",
+keywords = "Prediction",
+keywords = "Variable selection",
+keywords = "Genomic data analysis ",
+abstract = "Random forests (RF) is a popular tree-based ensemble machine learning tool that is highly data adaptive, applies to “large p, small n” problems, and is able to account for correlation as well as interactions among features.
+This makes \{RF\} particularly appealing for high-dimensional genomic data analysis.
+In this article, we systematically review the applications and recent progresses of \{RF\} for genomic data, including prediction and classification, variable selection, pathway analysis, genetic association and epistasis detection, and unsupervised learning.
+"
+}

### 2016-07-31 doc/publications/plos16/plos16.tex
plos16.tex

@@
 
+Cursed Forest uses:
+\begin{itemize}
+\item an ``information gained'' criteria for splitting.
+\item variabel importance is the sum of IG over all notes and all trees for that variable
+\item the model of parallelism is different to the ``foreach'' method -- we need to describe it
+\end{itemize}
+
 % For figure citations, please use "Fig." instead of "Figure".
@@
 % GWAS dataset
+
+\marginpar{These are regression examples, we nned some classification examples}
 

### 2016-08-01 doc/publications/plos16/plos16.tex
plos16.tex CursedForest.bib

@@
   \item Chen (2012)
-  \item VSURF and  varSelRF
+  \item VSURF and  varSelRF, 
+  \item ranger package
+  \item  r2VIM: A new variable selection method for random forests in
+    genome-wide association studies
+  \item CloudForest: A Scalable and Efficient Random Forest Implementation for Biological Data
   \end{itemize}
@@
 
-\cite{Strobl.et.al.2007} introoduce changes to the RF algorithm to fix 1) and 2). 
+\cite{Strobl.et.al.2007} introduce changes to the RF algorithm to fix 1) and 2). 
 However the level effect will be apparent in the CF algorithm as well. \marginpar{we can fix the boosting effect}
@@
 % authors may use "Analysis"
+
+\subsubsection*{variable selection}
+\cite{Genuer.et.al.2010} and \cite{Diaz.and.Alvarez.2006} describe iterative schemes for doing variable selection using the
+importance measure. See \cite{Chen.and.Ishwaran.2012} for a review of the area with particular reference to genomic data.
+\marginpar{ \cite{Chen.and.Ishwaran.2012} cite many papers on extensions of RF to handle GWAS data. Waht do we nned to do in this area}
+
 \section*{Methods}
@@
 \begin{itemize}
-\item an ``information gained'' criteria for splitting.
-\item variabel importance is the sum of IG over all notes and all trees for that variable
+\item an ``information gained'' criteria for splitting.  $f_q$ be the fraction of items labeled with value $q$ where $q= 1,
+  \ldots, Q$ in a set. The  entropy of the set is $H = - \sum^{m}_{i=1} f_i \log^{}_2 f_i$
+The expected information gain is the change in information entropy $H$ from a prior state $T$ to a state that takes some
+information $a$ as given: 
+\[ IG(T,a) = H(T) - H(T|a) \]
+In this case, the  the information gain is IG = Entropy(parent node) - Weighted Sum of Entropy(Children)
+\item variable importance is the sum of IG over all notes and all trees for that variable
 \item the model of parallelism is different to the ``foreach'' method -- we need to describe it

### 2016-08-01 doc/publications/plos16/plos16.tex
plos16.tex

@@
 \begin{itemize}
-\item what are you guys editing this with?  what about some carriage
-  returns or line feeds or something!
-\item Are we going to use the TCGA data in this paper?
-\item are we going to run sparkML or Cursed forest on the extended
-  simulation data \cite[]{Genuer.et.al.2010}? Has Aidan done this already?
+\item   I think the structure of the first section should be
+  \begin{itemize}
+  \item  convergence results from Biau 2012
+  \item consider  \cite{Segal.2004} examples
+  \item simulations reproduced from Genuer et al 2010 (no -- we need classification examples)
+  \item check Piotr's simulation
+  \item run sparkML or Cursed forest on the  simulation data 
+  \item we extend the simulations to many more variables 
   \item I dont think I can get R to handle chromosme 1 etc. May have to run CF on chromosme 22, see table
     \ref{table:smaller_data} -- On it, Aidan
-\item check Piotr's simulation
-\item run an argument as follows:
-  \begin{itemize}
   \item variable  selection may succeed where accurate prediction does not
-  \item consider  \cite{Segal.2004} examples
-  \item  perhaps we argue that in the case where the relationship of $X$ and $y$ takes a particular form we can recover
-    the reltionship for very large $p$p
+  \item discussion of the effect of the ntree and mtry parameters for variable selection with many noise variables
+  \item   perhaps we argue that in the case where the relationship of $X$ and $y$ takes a particular form (step, ramp) we can recover
+    the significant variables in the presence of a very large number of noise variables
+  \item in the case of genomic data, our functions of interest are more likely to be stop or ramp functions rather than harmonic
+    functions. So we have some hope.
   \end{itemize}
-\item \cite{Tuv.et.al.2009} feature selection via adding permuted variables. Do we want to consider this?
-\item check out varSelRF
-I think the structure of the first section should be
-\begin{itemize}
-\item  convergence results from Biau 2012
-\item simulations reproduced from Genuer et al 2010
-\item we extend the simulations to many more variables
-\item discussion of the effect of the ntree and mtry parameters for variable selection with many noise variables
-\item  in the case of some functions (step, ramp) we can recover the significant variables in the presence of a very large number of noise variables
-\item in the case of genomic data, our functions of interest are more likely to be stop or ramp functions rather than harmonic functions. So we have some hope.
-\end{itemize}
-
 \item check out 
   \begin{itemize}
-  \item D\'iaz-Uriarte, R., \& Alvarez de Andr\'es, S. (2006). Gene selection and classification of microarray data using random
-    forest. BMC Bioinformatics, 7, 3. http://doi.org/10.1186/1471-2105-7-3
+  \item D\'iaz-Uriarte, R., \& Alvarez de Andr\'es, S. (2006). (done)
   \item Goldstein et al. (2011)
-  \item Chen (2012)
-  \item VSURF and  varSelRF, 
+  \item Chen (2012)  (done)
+  \item VSURF and  varSelRF, (doee) recursive feature (forwards backwards) using variable importance
   \item ranger package
@@
   \item CloudForest: A Scalable and Efficient Random Forest Implementation for Biological Data
+  \item \cite{Tuv.et.al.2009} feature selection via adding permuted variables. Do we want to consider this? (no, leave it for now)
+  \end{itemize}
+\item other things to do
+  \begin{itemize}
+  \item what is the problem with bagging? Do we need to redo the thousand genome stuff?
   \end{itemize}

### 2016-08-01 doc/publications/plos16/plos16.tex
many changes. added tableofcontents to help organize the paper

@@
 % Please keep the abstract below 300 words
-
-\section*{Notes}
+\tableofcontents
+\clearpage
+\section{Notes}
 \begin{itemize}
@@
 
-\section*{Abstract}
+\clearpage
+
+\section{Abstract}
 The abstract.
@@
 
-\section*{Introduction}
+\section{Introduction}
 
 The digital revolution is seeing a dramatic increase in data collected about every aspect of 
-life~\cite{Loebbecke2015}.  These dataset not only grow vertically by capturing more and more events but also
+life~\cite{Loebbecke2015}.  These data sets are not only growing vertically by capturing more and more events but also
 horizontally by capturing more information about these events.  The challenge of big and `wide' data is especially
@@
 this data can be utilised in other datasets as well by filling in or ``imputing" the information at genomic positions
-previously unobserved due to the less detailed resolution of older array technology [citation about imputation].  This
-means that the large number of samples for which genome-wide association (GWA) data is available~\cite{Welter2013} can
-be extended horizontally by imputing the previously unobserved genotypes.  This provides the potential to generate
-datasets of hundreds of thousands of individuals with millions of variants, highlighting the need for incorporating
-modern compute paradigms to deal with these challenges.
+previously unobserved due to the less detailed resolution of older array technology [citation about imputation].
+\marginpar{citations missing} 
+This means that the large number of samples for which genome-wide association (GWA) data
+is available~\cite{Welter2013} can be extended horizontally by imputing the previously unobserved genotypes.  This
+provides the potential to generate datasets of hundreds of thousands of individuals with millions of variants,
+highlighting the need for incorporating modern compute paradigms to deal with these challenges.
 
@@
 While unsupervised clustering is useful for selected clinical applications~\cite{Li2015}, a more common analysis task is
-supervised learning, i.e. given a phenotype of clinical label determine the most associated features.  One of the
+supervised learning, i.e. given a phenotype or clinical label determine the most associated features.  One of the
 simplest supervised machine learning methods is logistic regression (LR), which creates a decision boundary by fitting a
@@
 This is known as the `curse of dimensionality'~\cite{Bauer2014}.  
-\marginpar{is this the reference we want for the ``curse of dimensionality''  Bellman (1961) is a more ususl reference}
+\marginpar{is this the reference we want for the ``curse of dimensionality''  Bellman (1961) is a more usual reference}
 
@@
 \begin{itemize}
-  \item are scale invariant;
+  \item are invariant to monotonic scaling of the data;
   \item can handle categorical and real valued inputs;
@@
 
-\subsection*{RF}
+\subsection{RF}
 
@@
 The results are combined in an appropriate way. In a classification problem we take the majority vote for the trees over
-the $i=1,..,Q$ classes
+the $q=1,..,Q$ classes,
 \begin{equation*}
-{\bar {h}}=  \arg \max_i \left(\sum_b I(h(x;\theta_b)=i)\right).
+{{h_f}}=  \arg \max_q \left(\sum_b I(h(x;\theta_b)=q)\right).
 \end{equation*}
 
-Random forests are a variance reduction technique. They take a tree models, which have a low bias but a high variance, and
-by combining them reduce the variance. The prediction error is the sum of the variance and the bias squared.
+Random forests are a variance reduction technique. They take unpruned tree models, which have a low bias but a high
+variance, and by combining them reduce the variance. The prediction error is the sum of the variance and the bias
+squared.
 
@@
 So what is required for  accurate random forest regression is (i) low correlation between residuals of differing tree in
-the forest, and (ii) low prediction error for the individual trees \cite{Segal.2004}.
-
-%http://stats.stackexchange.com/questions/77018/random-forest-is-it-a-boosting-algorithm
- %http://stats.stackexchange.com/questions/173390/gradient-boosting-tree-vs-random-forest
-%overcomes this issue and is hence ideally suited to this application.  Furthermore, RF are able to capture interactions
-%between features, which is of importance for modelling complex polygenic diseases.
-
-% We also investigated classifying the samples using Random Forests.  We chose random forests due to its ability to
-% capture complex interactions between features. Also, due to it being an ensemble method where trees can be formed
-% independently, which, is ideal for parallelisation. Furthermore, the ensemble of weak learners can help to minimise
-% overfitting and variance.
+the forest, and (ii) low prediction error for the individual trees \cite{Segal.2004}. In addition to growing the trees
+on  bootstrapped samples, each node in the tree is split on a randomly selected subset of the variables.
 
@@
   \begin{itemize}
-  \item they give an out-of-bag estimate of model accurace;
-  \item they provide a measure of variable importance;
+  \item they give an out-of-bag estimate of model accuracy, by using the samples not selected for the bootstrapped sample
+    as an independent test set for each tree;
+  \item they provide a measure of variable importance, for example by summing the number of times a variable is used to
+    split a node, over all nodes and all trees. More complex measures of variable importance are also used;
   \item they inherit some of the qualities of tree models (but not, for example, the handling of missing values via surrogate
@@
 
-Since the original implmentation of random forests there have been a number of developments leading to a better understanding of
+Since the original implementation of random forests there have been a number of developments leading to a better understanding of
 the algorithm. 
 \begin{itemize}
-\item random forests are resistent against overfitting. However it is not true that they will not overfit; see \cite{Segal.2004}
+\item random forests are resistant against overfitting. However it is not true that they will not overfit; see \cite{Segal.2004}
   where an example of RF overfitting is  given, caused by the fact that successive trees were correlated;
@@
   \begin{enumerate}
-  \item variable scale (for $X_I$ continuous) and number of levels (for $X_i$ discrete) inroduces a bias. A uninformative  variable with a large
+  \item differences in variable scale (for $X_i$ continuous) and number of levels (for $X_i$ discrete) introduces a bias. A uninformative  variable with a large
     number of levels may be selected over a more informative  variable with fewer levels;
-  \item bagging introduces a bias, but sampleing about $(1- 1/e) \approx 0.632$  of the data without 
+  \item bagging introduces a bias, but sampling $(1- 1/e) \approx 0.632$  of the data without replacement seems to be an
+    effective stragey for removing the problem.
   \end{enumerate}
@@
 However the level effect will be apparent in the CF algorithm as well. \marginpar{we can fix the boosting effect}
-However, there are many examples, (particularly in genomics) whre we have data sets that are
+However, there are many examples, (particularly in genomics) where we have data sets that are
 \begin{itemize}
@@
 \item all variables have the same number of levels (measures at different points along the genome)
-\item the hypothesied functional relationship is of a form amenable to a DT.
+\item the hypothesized functional relationship is of a form amenable to a decision tree.
 \end{itemize}
 
+\subsubsection{convergence}
 Biau {\it et al.}~\cite{Biau.2012} provide a proof that a random forest model will converge at a rate that depends on
@@
 
+\subsubsection{variable selection}
+\cite{Genuer.et.al.2010} and \cite{Diaz.and.Alvarez.2006} describe iterative schemes for doing variable selection using the
+importance measure. See \cite{Chen.and.Ishwaran.2012} for a review of the area with particular reference to genomic data.
+\marginpar{ \cite{Chen.and.Ishwaran.2012} cite many papers on extensions of RF to handle GWAS data. What do we need to do in this area}
+
+\section{Methods}
+
+\subsection{VariantSpark.}
+Traditionally, VariantSpark made use of Spark's machine learning algorithms `Spark ML'. These algorithms require data in a `DataFrame',
+similar to DataFrames in R.
+
 However, we observe that Spark's standard RF implementation is not able to handle the extremely `wide' genomic data as
@@
 
-% The reason for the failure can be attributed to the "curse of dimensionality". Where Spark ML is designed to process a
-% huge number of samples with a modest dimensionality, the 1000 Genomes data instead contains a huge number of
-% dimensions (features). As a result, the default random forests implementation is unable to partition the data into
-% small-enough tasks for nodes on the computer cluster to handle.
-
 We therefore developed CursedForest, a Spark-based random forest algorithm able to process big and `wide' data. Unlike
@@
 
-% You may title this section "Methods" or "Models". 
-% "Models" is not a valid title for PLoS ONE authors. However, PLoS ONE
-% authors may use "Analysis"
-
-\subsubsection*{variable selection}
-\cite{Genuer.et.al.2010} and \cite{Diaz.and.Alvarez.2006} describe iterative schemes for doing variable selection using the
-importance measure. See \cite{Chen.and.Ishwaran.2012} for a review of the area with particular reference to genomic data.
-\marginpar{ \cite{Chen.and.Ishwaran.2012} cite many papers on extensions of RF to handle GWAS data. Waht do we nned to do in this area}
-
-\section*{Methods}
-
-\subsection*{VariantSpark.}
-Traditionally, VariantSpark made use of Spark's machine learning algrorithms `Spark ML'. These algorithms require data in a `DataFrame',
-similar to DataFrames in R.
-
 We use the same pre-processing algorithms in CursedForests. One difference is that we
 
-Cursed Forest uses:
+Traditionally, VariantSpark used the random forest algorithm from SparkML. While SparkML algorithms demonstrate
+scalability when dealing with a large number of samples, this scalability quickly breaks down as we include more features.
+
+One of the restrictions preventing SparkML algorithms from scaling to high-dimensionality data is the fact that the feature-set
+for each individual is stored as a `Vector'. In Spark, Vectors are unable to be split and to process multiple vectors together,
+the entirety of each vector must be loaded into memory on the same node. Therefore, as dimensionality increases, the
+size of these Vectors must also increase, and subsequently this results in higher memory requirents.
+
+On the other hand, CursedForest (CF) is specifically designed to handle wide ``cursed" data. It avoids the relation between
+memory and dimensionality, by not using Vectors. Instead, CursedForest stores data in the Parquet columnar format.
+
+The current implementation of Cursed Forest uses:
 \begin{itemize}
-\item an ``information gained'' criteria for splitting.  $f_q$ be the fraction of items labeled with value $q$ where $q= 1,
-  \ldots, Q$ in a set. The  entropy of the set is $H = - \sum^{m}_{i=1} f_i \log^{}_2 f_i$
-The expected information gain is the change in information entropy $H$ from a prior state $T$ to a state that takes some
+\item an ``information gained'' criteria for splitting. Let $f_q$ be the fraction of items labeled with value $q$ where $q= 1,
+  \ldots, Q$ at a node. The  entropy of the node is $H = - \sum^{m}_{i=1} f_i \log^{}_2 f_i$
+The expected information gain is the change in entropy $H$ from a prior state $T$ to a state that takes some
 information $a$ as given: 
-\[ IG(T,a) = H(T) - H(T|a) \]
-In this case, the  the information gain is IG = Entropy(parent node) - Weighted Sum of Entropy(Children)
+\[ IG(T,a) = H(T) - H(T|a). \]
+In this case, the information gain is IG = Entropy(parent node) - Weighted Sum of Entropy(Children)
 \item variable importance is the sum of IG over all notes and all trees for that variable
-\item the model of parallelism is different to the ``foreach'' method -- we need to describe it
 \end{itemize}
 
-% For figure citations, please use "Fig." instead of "Figure".
-%Nulla mi mi, Fig.~\ref{fig1}  Fusce fringilla erat porttitor lectus cursus, \nameref{S1_Video} vel sagittis arcu lobortis.
-Aliquam in enim semper, aliquam massa id, cursus neque.
-Praesent faucibus semper libero.
-%\begin{figure}[h]
-%\caption{{\bf Figure Title first bold sentence Nulla mi mi, venenatis sed ipsum varius, volutpat euismod diam.}
-%Figure Caption Proin rutrum vel massa non gravida. Quisque tempor sem et dignissim rutrum. A: Lorem ipsum dolor sit amet. B: Consectetur adipiscing elit.}
-%\label{fig1}
-%\end{figure}
-%\begin{enumerate}
-%\item{react}
-%\item{diffuse free particles}
-%\item{increment time by dt and go to 1}
-%\end{enumerate}
-
-% Results and Discussion can be combined.
-% Please do not create a heading level below \subsection. For 3rd level headings, use \paragraph{}. 
-\section*{Results and Discussion}
+\section{Results and Discussion}
 
@@
 
-\marginpar{These are regression examples, we nned some classification examples}
+\marginpar{These are regression examples, we need some classification examples}
 
@@
 
-\subsection*{CursedForest processes genome-wide variant data}
+\subsection{CursedForest processes genome-wide variant data}
 %\label{comp}
@@
 %time/mem)
-
-Traditionally, VariantSpark used the random forest algorithm from SparkML. While SparkML algorithms demonstrate
-scalability when dealing with a large number of samples, this scalability quickly breaks down as we include more features.
-
-One of the restrictions preventing SparkML algorithms from scaling to high-dimensionality data is the fact that the feature-set
-for each individual is stored as a `Vector'. In Spark, Vectors are unable to be split and to process multiple vectors together,
-the entirety of each vector must be loaded into memory on the same node. Therefore, as dimensionality increases, the
-size of these Vectors must also increase, and subsequently this results in higher memory requirents.
-
-On the other hand, CursedForest (CF) is specifically designed to handle wide ``cursed" data. It avoids the relation between
-memory and dimensionality, by not using Vectors. Instead, CursedForest stores data in the Parquet columnar format.
 
@@
 
-\subsection*{Parameter settings}
+\subsection{Parameter settings}
 We consider the parameter setting for the RF algorithm. We use the R terms from the package RandomForests
-\cite{Liaw.and.Weiner.2002} which incoporates the original Fortran code by Brieman and Cutler. We incprporate the 
+\cite{Liaw.and.Weiner.2002} which incorporates the original Fortran code by Brieman and Cutler. We incorporate the 
 advice of \cite{Liaw.and.Weiner.2002}, which we have found mirrors our own experience.
@@
 \item \texttt{ntree} the number of trees.  The number of trees necessary for good performance grows with the number of predictors.
-  \marginpar{can we do proxmiity?}  \cite{Liaw.and.Weiner.2002} suggest that a high \texttt{ntree} is necessary to get stable
+  \marginpar{can we do proximity?}  \cite{Liaw.and.Weiner.2002} suggest that a high \texttt{ntree} is necessary to get stable
   estimates of variable importance and proximity; however, even though the variable importance measures may vary from run to run,
@@
   fit, and be useless for prediction, and still have a useful ranking of variable importance.
-\item \texttt{mtry} --  the number of variables considered at each split (if \texttt{mtry}=$p$, we have a boosted decison
+\item \texttt{mtry} --  the number of variables considered at each split (if \texttt{mtry}=$p$, we have a boosted decision
   tree model).  If one has a very large number of variables but expects only very few to be ``important'', using larger \texttt{mtry} may give
@@
 
-%\subsection*{James' dataset}
+%\subsection{James' dataset}
 %Comparing CursedForest with feature selection + linear model (performance and time/mem)
 
-\subsection*{Conclusion}
+\subsection{Conclusion}
 
-%\begin{table}[!ht]
-%\begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
-%\caption{
-%{\bf Table caption Nulla mi mi, venenatis sed ipsum varius, volutpat euismod diam.}}
-%\begin{tabular}{|l|l|l|l|l|l|l|l|}
-%\hline
-%\multicolumn{4}{|l|}{\bf Heading1} & \multicolumn{4}{|l|}{\bf Heading2}\\ \hline
-%$cell1 row1$ & cell2 row 1 & cell3 row 1 & cell4 row 1 & cell5 row 1 & cell6 row 1 & cell7 row 1 & cell8 row 1\\ \hline
-%$cell1 row2$ & cell2 row 2 & cell3 row 2 & cell4 row 2 & cell5 row 2 & cell6 row 2 & cell7 row 2 & cell8 row 2\\ \hline
-%$cell1 row3$ & cell2 row 3 & cell3 row 3 & cell4 row 3 & cell5 row 3 & cell6 row 3 & cell7 row 3 & cell8 row 3\\ \hline
-%\end{tabular}
-%\begin{flushleft} Table notes Phasellus venenatis, tortor nec vestibulum mattis, massa tortor interdum felis, nec pellentesque metus tortor nec nisl.
-Ut ornare mauris tellus, vel dapibus arcu suscipit sed.
-%\end{flushleft}
-%\label{table1}
-%\end{adjustwidth}
-%\end{table}
-
-\section*{Supporting Information}
+\section{Supporting Information}
 
@@
 }
+
+%http://stats.stackexchange.com/questions/77018/random-forest-is-it-a-boosting-algorithm
+ %http://stats.stackexchange.com/questions/173390/gradient-boosting-tree-vs-random-forest
+%overcomes this issue and is hence ideally suited to this application.  Furthermore, RF are able to capture interactions
+%between features, which is of importance for modelling complex polygenic diseases.
+
+% We also investigated classifying the samples using Random Forests.  We chose random forests due to its ability to
+% capture complex interactions between features. Also, due to it being an ensemble method where trees can be formed
+% independently, which, is ideal for parallelisation. Furthermore, the ensemble of weak learners can help to minimise
+% overfitting and variance.
+
+% The reason for the failure can be attributed to the "curse of dimensionality". Where Spark ML is designed to process a
+% huge number of samples with a modest dimensionality, the 1000 Genomes data instead contains a huge number of
+% dimensions (features). As a result, the default random forests implementation is unable to partition the data into
+% small-enough tasks for nodes on the computer cluster to handle.
+
+% For figure citations, please use "Fig." instead of "Figure".
+%Nulla mi mi, Fig.~\ref{fig1}  Fusce fringilla erat porttitor lectus cursus, \nameref{S1_Video} vel sagittis arcu lobortis.
+Aliquam in enim semper, aliquam massa id, cursus neque.
+Praesent faucibus semper libero.
+%\begin{figure}[h]
+%\caption{{\bf Figure Title first bold sentence Nulla mi mi, venenatis sed ipsum varius, volutpat euismod diam.}
+%Figure Caption Proin rutrum vel massa non gravida. Quisque tempor sem et dignissim rutrum. A: Lorem ipsum dolor sit amet. B: Consectetur adipiscing elit.}
+%\label{fig1}
+%\end{figure}
+%\begin{enumerate}
+%\item{react}
+%\item{diffuse free particles}
+%\item{increment time by dt and go to 1}
+%\end{enumerate}
+
+% Results and Discussion can be combined.
+% Please do not create a heading level below \subsection. For 3rd level headings, use \paragraph{}.
+
+%\begin{table}[!ht]
+%\begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
+%\caption{
+%{\bf Table caption Nulla mi mi, venenatis sed ipsum varius, volutpat euismod diam.}}
+%\begin{tabular}{|l|l|l|l|l|l|l|l|}
+%\hline
+%\multicolumn{4}{|l|}{\bf Heading1} & \multicolumn{4}{|l|}{\bf Heading2}\\ \hline
+%$cell1 row1$ & cell2 row 1 & cell3 row 1 & cell4 row 1 & cell5 row 1 & cell6 row 1 & cell7 row 1 & cell8 row 1\\ \hline
+%$cell1 row2$ & cell2 row 2 & cell3 row 2 & cell4 row 2 & cell5 row 2 & cell6 row 2 & cell7 row 2 & cell8 row 2\\ \hline
+%$cell1 row3$ & cell2 row 3 & cell3 row 3 & cell4 row 3 & cell5 row 3 & cell6 row 3 & cell7 row 3 & cell8 row 3\\ \hline
+%\end{tabular}
+%\begin{flushleft} Table notes Phasellus venenatis, tortor nec vestibulum mattis, massa tortor interdum felis, nec pellentesque metus tortor nec nisl.
+Ut ornare mauris tellus, vel dapibus arcu suscipit sed.
+%\end{flushleft}
+%\label{table1}
+%\end{adjustwidth}
+%\end{table}

### 2016-08-10 doc/publications/plos16/plos16.tex
Some changes

--- 2016-08-01 doc/publications/plos16/plos16.tex
@@
 {\Large
-\textbf\newline{Cursed Forest - A random forest implementation for `big' and `wide' data} % Please use "title case" (capitalize all terms in the title except conjunctions, prepositions, and articles).
+\textbf\newline{Cursed Forest - A random forest implementation for ``big'' and ``wide'' data} % Please use "title case" (capitalize all terms in the title except conjunctions, prepositions, and articles).
 }
@@
 \item Are we going to use the TCGA data in this paper?
-\item are we going to run sparkML or Cursed forest on the extended
+\item are we going to run spark ML or Cursed forest on the extended
   simulation data \cite[]{Genuer.et.al.2010}? Has Aidan done this already?
@@
 \item \cite{Tuv.et.al.2009} feature selection via adding permuted variables. Do we want to consider this?
-\item check out varSelRF
-I think the structure of the first section should be
-\begin{itemize}
-\item  convergence results from Biau 2012
-\item simulations reproduced from Genuer et al 2010
-\item we extend the simulations to many more variables
-\item discussion of the effect of the ntree and mtry parameters for variable selection with many noise variables
-\item  in the case of some functions (step, ramp) we can recover the significant variables in the presence of a very large number of noise variables
-\item in the case of genomic data, our functions of interest are more likely to be stop or ramp functions rather than harmonic functions. So we have some hope.
-\end{itemize}
-
-\item check out 
-  \begin{itemize}
-  \item D\'iaz-Uriarte, R., \& Alvarez de Andr\'es, S. (2006). Gene selection and classification of microarray data using random
-    forest. BMC Bioinformatics, 7, 3. http://doi.org/10.1186/1471-2105-7-3
-  \item Goldstein et al. (2011)
-  \item Chen (2012)
-  \item VSURF and  varSelRF, 
-  \item ranger package
-  \item  r2VIM: A new variable selection method for random forests in
-    genome-wide association studies
-  \item CloudForest: A Scalable and Efficient Random Forest Implementation for Biological Data
-  \end{itemize}
+\item chek out varSelRF
 \end{itemize}
@@
 
-The digital revolution is seeing a dramatic increase in data collected about every aspect of 
+The digital revolution or ``datafication'' sees a dramatic increase in data collected about every aspect of   \marginpar{is seeing, will see}
 life~\cite{Loebbecke2015}.  These dataset not only grow vertically by capturing more and more events but also
-horizontally by capturing more information about these events.  The challenge of big and `wide' data is especially
+horizontally by capturing more information about these events.  The challenge of big and ``wide'' data is especially
 pronounced in the health space where whole genome sequence (WGS) technology enabled researchers to interrogate all 3
@@
 denoting the differences between individuals.  This dataset consumes close to a terabyte of disk space.  The richness of
-this data can be utilised in other datasets as well by filling in or ``imputing" the information at genomic positions
+this data can be utilised in other datasets as well by filling in or ``imputing'' the information at genomic positions
 previously unobserved due to the less detailed resolution of older array technology [citation about imputation].  This
@@
 Distributing the compute tasks is one strategy to overcome this challenge. Apache Spark, based on Scala, allows for  
-\marginpar{is it based on Scala or does it have a Scala API?}
-distributed applications to scale to thousands of nodes.  A Spark application runs on a `driver' node, and
-this node divides and distributes work out to the many `worker' nodes.  Spark also includes the machine learning (ML)
-library, Spark ML, providing popular machine learning libraries.  Spark ML includes supervised approaches such as
-decision trees and the ensemble method, random forests, as well as unsupervised approaches such as k-means clustering.
+\marginpar{is it based on scale or does it have a scale API?}
+distributed applications to scale to thousands of nodes.  A Spark application runs on a ``driver'' node, and this node
+divides and distributes work out to the many ``worker'' nodes.  Spark also includes the machine learning (ML) library,
+Spark ML, providing support for a selection of machine learning algorithms.  Spark ML includes supervised approaches such
+as decision trees and the ensemble method, random forests, as well as unsupervised approaches such as k-means clustering.
 
@@
 simplest supervised machine learning methods is logistic regression (LR), which creates a decision boundary by fitting a
-linear combination of the feature variables.  Processing the `wide' genomic data poses difficulties for machine learning
+linear combination of the feature variables.  Processing the ``wide'' genomic data poses difficulties for machine learning
 applications as having more features $p$ than samples $n$ causes the model to overfit the data and generalize poorly.
-This is known as the `curse of dimensionality'~\cite{Bauer2014}.  
-\marginpar{is this the reference we want for the ``curse of dimensionality''  Bellman (1961) is a more ususl reference}
+This is known as the ``curse of dimensionality''~\cite{Bauer2014}.  
+\marginpar{is this the reference we want for the ``curse of dimensionality''}
 
@@
 lead to a very different model. 
-
-\subsection*{RF}
+\marginpar{a potential problem  with tree models is the tendencey to split on categorical variables with a high number of
+  levels, even when better splits are available}
+\subsection*{Random Forest}
 
@@
 
-The results are combined in an appropriate way. In a classification problem we take the majority vote for the trees over
-the $i=1,..,Q$ classes
+The results are combined in an appropriate way, i.e. for a regression problem,
 \begin{equation*}
-{\bar {h}}=  \arg \max_i \left(\sum_b I(h(x;\theta_b)=i)\right).
+{\bar {h}}={\frac {1}{B}}\sum _{b=1}^{B} h(x;\theta_b).
 \end{equation*}
-
-Random forests are a variance reduction technique. They take a tree models, which have a low bias but a high variance, and
-by combining them reduce the variance. The prediction error is the sum of the variance and the bias squared.
+\marginpar{make it a classification example}
+%where $x$ is the covariate vector of length $p$
+%with associated random vector $X$ and the $\theta_b$ are i.i.d. random vectors.
 
@@
 
+\subsection*{Benefits of RF}
+
+\marginpar{weak learners only in a misuse of the term. ``weak learners'' was applied to decision stumps}
 %http://stats.stackexchange.com/questions/77018/random-forest-is-it-a-boosting-algorithm
@@
 
-The benefits of random forests models are
+Random forests are a variance reduction technique. They take a tree models which have a low bias but a high variance and
+by combining them reduce the variance. The prediction error is the sum of the variance and the bias squared.
+
+The benefits of random forests models are:
   \begin{itemize}
@@
   \item they provide a measure of variable importance;
-  \item they inherit some of the qualities of tree models (but not, for example, the handling of missing values via surrogate
-    variables. This would be difficult as the trees as split on random subsets of variables).
+  \item they inherit many of the qualities of tree models.
+\marginpar{but not the handling of missing velues via surrogate variables. This would be difficult as the trees are split
+  on random subsets of variables}
   \end{itemize}
 
-Since the original implmentation of random forests there have been a number of developments leading to a better understanding of
-the algorithm. 
-\begin{itemize}
-\item random forests are resistent against overfitting. However it is not true that they will not overfit; see \cite{Segal.2004}
-  where an example of RF overfitting is  given, caused by the fact that successive trees were correlated;
-\item \cite{Strobl.et.al.2007} demonstrates that the RF algorithm is subject to a bias in variable selection via two mechanisms:
-  \begin{enumerate}
-  \item variable scale (for $X_I$ continuous) and number of levels (for $X_i$ discrete) inroduces a bias. A uninformative  variable with a large
-    number of levels may be selected over a more informative  variable with fewer levels;
-  \item bagging introduces a bias, but sampleing about $(1- 1/e) \approx 0.632$  of the data without 
-  \end{enumerate}
-\end{itemize}
-
-\cite{Strobl.et.al.2007} introduce changes to the RF algorithm to fix 1) and 2). 
-However the level effect will be apparent in the CF algorithm as well. \marginpar{we can fix the boosting effect}
-However, there are many examples, (particularly in genomics) whre we have data sets that are
-\begin{itemize}
-\item very wide
-\item all variables have the same number of levels (measures at different points along the genome)
-\item the hypothesied functional relationship is of a form amenable to a DT.
-\end{itemize}
-
-Biau {\it et al.}~\cite{Biau.2012} provide a proof that a random forest model will converge at a rate that depends on
-the cardinality of the set $S$ of ``strong predictors'' rather than on the number of variables $p$. That is, given a
-function $y=f(S)$ depending on a set of variables $S$, a random forest will converge to the true function $y$ with a
-rate that depends on the size of $S$ and not the number of noise variables in the data set. If the size of $S$ is small
-compared to the total number of variables, then the rate of convergence will be quicker.
-\marginpar{say something about the rate of convergence}
-
-Subject to certain conditions, RF will not overfit even in the  $p > n$ case. 
-Here we demonstrate that this may hold true in extreme cases where $p \gg n$.
-
-However, we observe that Spark's standard RF implementation is not able to handle the extremely `wide' genomic data as
+However, we observe that Spark ML's standard RF implementation is unable to handle the extremely ``wide'' genomic data as
 it was developed for large number of samples with only modest dimensionality.  Although Spark ML can build a RF model on
@@
 unbalanced work load where the driver node becomes the bottleneck and worker nodes being idle prevents a seamless
-scaling to larger datasets.
+scaling to larger datasets. The memory requirements per executor also increases with dimensionality due to the data types
+Spark ML uses, which we elaborate on in the discussion.
 
-% The reason for the failure can be attributed to the "curse of dimensionality". Where Spark ML is designed to process a
+% The reason for the failure can be attributed to the ``curse of dimensionality''. Where Spark ML is designed to process a
 % huge number of samples with a modest dimensionality, the 1000 Genomes data instead contains a huge number of
@@
 
-We therefore developed CursedForest, a Spark-based random forest algorithm able to process big and `wide' data. Unlike
+We therefore developed CursedForest, a Spark-based random forest algorithm able to process big and ``wide'' data. Unlike
 the standard Apache implementation, CursedForest is able to easily process data with millions of dimensions without
-excessive resource requirements on the driver or worker nodes.
+excessive resource requirements on the driver or worker nodes. It also greatly reduces memory requirements on worker nodes
+by partitioning the job differently than Spark ML.
 %Also, due to more efficient handling of high-dimensional data, equivalent jobs are likely to finish faster on wideML than Spark ML.
-In the first section we
+In the first section we explore some of the properties of random forests, demonstrating that any potential flaws in CursedForest
+are an inherent property of Random Forests. Here we provide demonstrations using synthetic data.
+We then demonstrate the scalability of CursedForest in respect to the dimensionality of the data, building an RF model on whole-genome
+data from the 1000 Genomes Project. Here we report on the accuracy (in terms of the ARI) and the error.
+Finally, given the role different parameter values can play in model construction, we explore the effect tuning these parameters
+can have on the prediction accuracy of a model.
 
@@
 % authors may use "Analysis" 
-
-\subsubsection*{variable selection}
-\cite{Genuer.et.al.2010} and \cite{Diaz.and.Alvarez.2006} describe iterative schemes for doing variable selection using the
-importance measure. See \cite{Chen.and.Ishwaran.2012} for a review of the area with particular reference to genomic data.
-\marginpar{ \cite{Chen.and.Ishwaran.2012} cite many papers on extensions of RF to handle GWAS data. Waht do we nned to do in this area}
-
 \section*{Methods}
@@
 \subsection*{VariantSpark.}
-Traditionally, VariantSpark made use of Spark's machine learning algrorithms `Spark ML'. These algorithms require data in a `DataFrame',
+Traditionally, VariantSpark made use of Spark's machine learning algrorithms, ``Spark ML''. These algorithms require data in a ``DataFrame'',
 similar to DataFrames in R.
@@
 We use the same pre-processing algorithms in CursedForests. One difference is that we
-
-Cursed Forest uses:
-\begin{itemize}
-\item an ``information gained'' criteria for splitting.  $f_q$ be the fraction of items labeled with value $q$ where $q= 1,
-  \ldots, Q$ in a set. The  entropy of the set is $H = - \sum^{m}_{i=1} f_i \log^{}_2 f_i$
-The expected information gain is the change in information entropy $H$ from a prior state $T$ to a state that takes some
-information $a$ as given: 
-\[ IG(T,a) = H(T) - H(T|a) \]
-In this case, the  the information gain is IG = Entropy(parent node) - Weighted Sum of Entropy(Children)
-\item variable importance is the sum of IG over all notes and all trees for that variable
-\item the model of parallelism is different to the ``foreach'' method -- we need to describe it
-\end{itemize}
 
@@
 
-\marginpar{These are regression examples, we nned some classification examples}
+\subsection*{Random Forests converges on wide data}
+
+As discussed in the Introduction,
+\marginpar{I think this has been removed from the introduction}
+random forests are resistent against overfitting.
+However it is not true that they will not overfit; see \cite{Segal.2004} where examples of RF overfitting are given 
+\marginpar{discuss these examples}
+
+Subject to certain conditions, RF will not overfit even in the  $p > n$ case. 
+Here we demonstrate that this may hold true in extreme cases where $p \gg n$.
+
+\cite{Strobl.et.al.2007} demosnstrate that a RF algorithm is subject to a bias in variable selection via two mechanisms:
+\begin{itemize}
+\item variable scale (for $X_I$ continuous) and number of levels (for $X_i$ discrete) inroduces a bias
+\item boosting introduces a bias.
+\end{itemize}
+These effects will be apparent in the CF algorithm as well. \marginpar{we can fix the boosting effect}
+However, there are many examples, (particularly in genomics) whre we have data sets that are
+\begin{itemize}
+\item very wide
+\item all variables have the same number of levels (measures at different points along the genome)
+\item the hypothesied functional relationship is of a form amenable to a DT.
+\end{itemize}
+
+Biau {\it et al.}~\cite{Biau.2012} provide a proof that a random forest model will converge at a rate that depends on
+the cardinality of the set $S$ of ``strong predictors'' rather than on the number of variables $p$. That is, given a
+function $y=f(S)$ depending on a set of variables $S$, a random forest will converge to the true function $y$ with a
+rate that depends on the size of $S$ and not the number of noise variables in the data set. If the size of $S$ is small
+compared to the total number of variables, then the rate of convergence will be quicker.
+\marginpar{say something about the rate of convergence}
 
@@
 
-Traditionally, VariantSpark used the random forest algorithm from SparkML. While SparkML algorithms demonstrate
+Traditionally, VariantSpark used the random forest algorithm from Spark ML. While Spark ML algorithms demonstrate
 scalability when dealing with a large number of samples, this scalability quickly breaks down as we include more features.
 
-One of the restrictions preventing SparkML algorithms from scaling to high-dimensionality data is the fact that the feature-set
-for each individual is stored as a `Vector'. In Spark, Vectors are unable to be split and to process multiple vectors together,
-the entirety of each vector must be loaded into memory on the same node. Therefore, as dimensionality increases, the
-size of these Vectors must also increase, and subsequently this results in higher memory requirents.
+One of the restrictions preventing Spark ML algorithms from scaling to high-dimensionality data is that the feature-set
+for each sample is stored as a vector data type. Spark vectors are stored in resilient distributed dataset (RDDs), and while RDDs
+can be split, vectors cannot. So as we add more samples (in our case individuals), Spark ML can simply split the RDD into more
+partitions and resource requirements will remain relatively stable. However, as the number of dimensions (variants) increases,
+the size of the vectors increases. This results in higher memory-requirements as increasingly large vectors must be loaded
+into memory.
 
-On the other hand, CursedForest (CF) is specifically designed to handle wide ``cursed" data. It avoids the relation between
+On the other hand, CursedForest (CF) is specifically designed to handle wide ``cursed'' data. It avoids the relation between
 memory and dimensionality, by not using Vectors. Instead, CursedForest stores data in the Parquet columnar format.
+To build a tree, CursedForest will distribute tasks across compute nodes to calculate the information gain for every feature
+in the feature-set. The result is then reduced to return the feature which gives the greatest information gain.
+This process is then repeated until CursedForest has calculated every node for said tree.
 
@@
 %\multicolumn{4}{|l|}{\bf Heading1} & \multicolumn{4}{|l|}{\bf Heading2}\\ \hline
-\bf{Genome \%}  & \bf{method} & \bf{Accuracy} & \bf{Runtime} & \bf{Memory/Node} \\
+\bf{Genome \%}  & \bf{Method} & \bf{Error (ARI)} & \bf{Runtime} & \bf{Memory/Node} \\
 \hline
 
-\multirow{3}{*}{phase1\_chr22: 1,092 x 490,036} & CursedForest &  &  &  \\
-& RF (Spark ML) &  &  &  \\
-& RF (R) & 0.456 & 2.89 & 18GB \\ \hline
+\multirow{3}{*}{phase1\_chr22: 1,092 x 490,036} & CursedForest & 0.077 (XX)  &  &  \\
+& RF - Spark ML &  &  &  \\
+& RF - R & 0.456 & 2.89 & 18GB \\ \hline
 
-\multirow{3}{*}{phase3\_chr1: 2,504 x 6,450,364} & CursedForest & XX & 1hr 58min 23sec & 4GB \\
-& RF (Spark ML) & 0.94 & 8hr 29min 07sec & 8GB \\
-& LR &  & 10hr 30min & 8GB\\ \hline
+\multirow{3}{*}{phase3\_chr1: 2,504 x 6,450,364} & CursedForest & 0.063 (XX) & 1hr 58min & 4GB \\
+& RF - Spark ML & 0.94 & 8hr 29min & 8GB \\
+& LR - Spark ML&  & 2hr 6min & 8GB\\ \hline
 
-\multirow{3}{*}{phase3\_chr1-3: 2,504 x 19,328,051}  & CursedForest & &  &\\
-& RF(Spark ML) & &  &\\
-& LR & &  &\\ \hline
+\multirow{3}{*}{phase3\_chr1-3: 2,504 x 19,328,051}  & CursedForest & 0.046 (XX) & 4hr 0min & 4GB\\
+& RF - Spark ML & &  &\\
+& LR - Spark ML & & 14hr 48min & 8GB \\ \hline
 
-\multirow{4}{*}{phase3\_chr1-22: 2,504 x 81,047,467} & CursedForest & 0.96 & 7hr 14min 40sec & 8GB \\ 
-& RF (Spark ML) & - & - & - \\
-& LR & - & - & - \\ 
-& kmeans (Spark ML) & 0.82 & 30hr 44min 00sec & 24GB \\ \hline
+\multirow{4}{*}{phase3\_chr1-22: 2,504 x 81,047,467} & CursedForest & XX (0.96) & 7hr 14min & 8GB \\ 
+& RF - Spark ML & - & - & - \\
+& LR - Spark ML & - & - & - \\ 
+& kmeans - Spark ML & - (0.82) & 30hr 44min & 24GB \\ \hline
 \end{tabular}
@@
 Comparison of different various machine learning libraries on different subsets of variant data 
-from the 1000 Genomes Project. Size of the data is individuals x variants
+from the 1000 Genomes Project. Size of the data is listed as individuals x variants.
 
@@
 \subsection*{Parameter settings}
-We consider the parameter setting for the RF algorithm. We use the R terms from the package RandomForests
-\cite{Liaw.and.Weiner.2002} which incoporates the original Fortran code by Brieman and Cutler. We incprporate the 
-advice of \cite{Liaw.and.Weiner.2002}, which we have found mirrors our own experience.
+\marginpar{ was``Random forest is sensitive to tree depth''}
+In this setting, variable selection in very high dimensions, we nned to consider the parameter setting for the RF
+algorithm.
 
+I will use the R terms from the package RandomForests whcih incoporates the original Fortran code by Brieman and
+Cutler. We consider the parameters:
 \begin{itemize}
-\item \texttt{ntree} the number of trees.  The number of trees necessary for good performance grows with the number of predictors.
-  \marginpar{can we do proxmiity?}  \cite{Liaw.and.Weiner.2002} suggest that a high \texttt{ntree} is necessary to get stable
-  estimates of variable importance and proximity; however, even though the variable importance measures may vary from run to run,
-  the ranking of the importances is quite stable.  In addition, we note that it is possible for a RF model to have a very poor
-  fit, and be useless for prediction, and still have a useful ranking of variable importance.
-\item \texttt{mtry} --  the number of variables considered at each split (if \texttt{mtry}=$p$, we have a boosted decison
-  tree model).  If one has a very large number of variables but expects only very few to be ``important'', using larger \texttt{mtry} may give
-  better performance.
-\item  For classification problems where the class frequencies are extremely unbalanced it may be necessary to change the prediction
-  rule \marginpar{can we do this}
 \item the size and complexity of the individual trees is controlled in RandomForests by setting \texttt{nodesize}, the
   minimum size of terminal nodes. It is controlled in Spark.ML by setting \texttt{maxDepth}, the maximum depth of each
-  tree in the forest. The SparkML documentation somewhat unhelpfully sets \texttt{maxDepth=4} in their classification
-  example\footnote{The documentation is available at \url{http://spark.apache.org/docs/latest/mllib-ensembles.html}.}.
-
-  In the ATCG example the RandomForests code produced trees with depths of 30+. We note that Random forests are a way of averaging
+  tree in the forest. The Spark ML documentation somewhat unhelpfully sets \texttt{maxDepth=4} in their classification
+  example\footnote{the documentation is available at \url{http://spark.apache.org/docs/latest/mllib-ensembles.html} }.
+  In the ATCG example the RF code produced trees with depths of 30+. We note that Random forests are a way of averaging
   multiple deep decision trees, trained on different parts of the same training set, with the goal of reducing the
@@
 \marginpar{are we including this  example?  Supplementary info?}
+\item \texttt{ntree} the number of trees 
+\item \texttt{mtry} --  the number of variables considered at each split (if \texttt{mtry}=$p$, we have a boosted decison
+  tree model \marginpar{is this right?}
 \end{itemize}
@@
 \begin{flushleft} 
-  Table notes Phasellus venenatis, tortor nec vestibulum mattis, massa tortor interdum felis, nec pellentesque metus
-  tortor nec nisl. Ut ornare mauris tellus, vel dapibus arcu suscipit sed.
+  Table notes here.
 \end{flushleft}
@@
 \subsection*{Conclusion}
+We find
 
@@
 % Include only the SI item label in the subsection heading. Use the \nameref{label} command to cite SI items in the text.
-\subsection*{S1 Video}
-\label{S1_Video}
-{\bf Bold the first sentence.}  Maecenas convallis mauris sit amet sem ultrices gravida.
+%\subsection*{S1 Video}
+%\label{S1_Video}
+%{\bf Bold the first sentence.}  Maecenas convallis mauris sit amet sem ultrices gravida.
 Etiam eget sapien nibh.
-Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula.
-Curabitur fringilla pulvinar lectus consectetur pellentesque.
-
-\subsection*{S1 Text}
-\label{S1_Text}
-{\bf Lorem Ipsum.} Maecenas convallis mauris sit amet sem ultrices gravida.
-Etiam eget sapien nibh.
-Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula.
-Curabitur fringilla pulvinar lectus consectetur pellentesque.
-
-\subsection*{S1 Table}
-\label{S1_Table}
-{\bf Lorem Ipsum.} Maecenas convallis mauris sit amet sem ultrices gravida.
-Etiam eget sapien nibh.
-Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula.
+Sed ac ipsum eget %enim egestas ullamcorper nec euismod ligula.
 Curabitur fringilla pulvinar lectus consectetur pellentesque.
@@
 \end{document}
-
-  \item D\iaz-Uriarte, R., \& Alvarez de Andr\es, S. (2006). Gene selection and classification of microarray data using random
-    forest. BMC Bioinformatics, 7, 3. http://doi.org/10.1186/1471-2105-7-3
-  \item Goldstein, B. A., Polley, E. C., & Briggs, F. B. S. (2011). Random Forests for Genetic Association Studies. Statistical
-    Applications in Genetics and Molecular Biology, 10(1), 32. http://doi.org/10.2202/1544-6115.1691
-
-@article{Chen2012323,
-title = "Random forests for genomic data analysis ",
-journal = "Genomics ",
-volume = "99",
-number = "6",
-pages = "323 - 329",
-year = "2012",
-note = "",
-issn = "0888-7543",
-doi = "http://dx.doi.org/10.1016/j.ygeno.2012.04.003",
-url = "http://www.sciencedirect.com/science/article/pii/S0888754312000626",
-author = "Xi Chen and Hemant Ishwaran",
-keywords = "Random forests",
-keywords = "Random survival forests",
-keywords = "Classification",
-keywords = "Prediction",
-keywords = "Variable selection",
-keywords = "Genomic data analysis ",
-abstract = "Random forests (RF) is a popular tree-based ensemble machine learning tool that is highly data adaptive, applies to “large p, small n” problems, and is able to account for correlation as well as interactions among features.
-This makes \{RF\} particularly appealing for high-dimensional genomic data analysis.
-In this article, we systematically review the applications and recent progresses of \{RF\} for genomic data, including prediction and classification, variable selection, pathway analysis, genetic association and epistasis detection, and unsupervised learning.
-"
-}

### 2016-08-10 doc/publications/plos16/plos16.tex

--- 2016-08-01 doc/publications/plos16/plos16.tex
@@
 denoting the differences between individuals.  This dataset consumes close to a terabyte of disk space.  The richness of
-this data can be utilised in other datasets as well by filling in or ``imputing" the information at genomic positions
+this data can be utilised in other datasets as well by filling in or ``imputing'' the information at genomic positions
 previously unobserved due to the less detailed resolution of older array technology [citation about imputation].
@@
 simplest supervised machine learning methods is logistic regression (LR), which creates a decision boundary by fitting a
-linear combination of the feature variables.  Processing the `wide' genomic data poses difficulties for machine learning
+linear combination of the feature variables.  Processing the ``wide'' genomic data poses difficulties for machine learning
 applications as having more features $p$ than samples $n$ causes the model to overfit the data and generalize poorly.
-This is known as the `curse of dimensionality'~\cite{Bauer2014}.  
+This is known as the ``curse of dimensionality''~\cite{Bauer2014}.  
 \marginpar{is this the reference we want for the ``curse of dimensionality''  Bellman (1961) is a more usual reference}
@@
 
-\subsection{RF}
+\subsection{Random Forest}
 
@@
 
-The benefits of random forests models are
+The benefits of random forests models are:
   \begin{itemize}
@@
 \subsection{VariantSpark.}
-Traditionally, VariantSpark made use of Spark's machine learning algorithms `Spark ML'. These algorithms require data in a `DataFrame',
+Traditionally, VariantSpark made use of Spark's machine learning algorithms ``Spark ML''. These algorithms require data in a ``DataFrame'',
 similar to DataFrames in R.
 
-However, we observe that Spark's standard RF implementation is not able to handle the extremely `wide' genomic data as
+However, we observe that Spark ML's RF implementation is not able to handle the extremely ``wide'' genomic data as
 it was developed for large number of samples with only modest dimensionality.  Although Spark ML can build a RF model on
@@
 unbalanced work load where the driver node becomes the bottleneck and worker nodes being idle prevents a seamless
-scaling to larger datasets.
+scaling to larger datasets. The memory requirements per executor also increases with dimensionality due to the data types
+Spark ML uses, which we elaborate on in the discussion.
 
-We therefore developed CursedForest, a Spark-based random forest algorithm able to process big and `wide' data. Unlike
+We therefore developed CursedForest, a Spark-based random forest algorithm able to process big and ``wide'' data. Unlike
 the standard Apache implementation, CursedForest is able to easily process data with millions of dimensions without
 excessive resource requirements on the driver or worker nodes.
-%Also, due to more efficient handling of high-dimensional data, equivalent jobs are likely to finish faster on wideML than Spark ML.
-In the first section we
+
+In the first section we explore some of the properties of random forests, demonstrating that any potential flaws in CursedForest
+are an inherent property of Random Forests. Here we provide demonstrations using synthetic data.
+We then demonstrate the scalability of CursedForest in respect to the dimensionality of the data, building an RF model on whole-genome
+data from the 1000 Genomes Project. Here we report on the accuracy (in terms of the ARI) and the error.
+Finally, given the role different parameter values can play in model construction, we explore the effect tuning these parameters
+can have on the prediction accuracy of a model.
 
@@
 \item the size and complexity of the individual trees is controlled in RandomForests by setting \texttt{nodesize}, the
-  minimum size of terminal nodes. It is controlled in Spark.ML by setting \texttt{maxDepth}, the maximum depth of each
+  minimum size of terminal nodes. It is controlled in Spark ML by setting \texttt{maxDepth}, the maximum depth of each
   tree in the forest. The SparkML documentation somewhat unhelpfully sets \texttt{maxDepth=4} in their classification

### 2016-08-10 doc/publications/plos16/plos16.tex
Fixed mention of Spark being based on scala
Made the description hopefully a bit clearer

@@
 % tools such as Apache Spark are available as a framework for iteratively dealing with big data.
-Distributing the compute tasks is one strategy to overcome this challenge. Apache Spark, based on Scala, allows for  
-\marginpar{is it based on scale or does it have a scale API?}
-distributed applications to scale to thousands of nodes.  A Spark application runs on a ``driver'' node, and this node
-divides and distributes work out to the many ``worker'' nodes.  Spark also includes the machine learning (ML) library,
+Distributing the compute tasks is one strategy to overcome this challenge. Apache Spark, with APIs available for
+Scala, Java, Python and R, allows for distributed applications to scale to thousands of nodes.
+A Spark application runs on a ``driver'' node, and this node
+divides and distributes work out to the many ``worker'' nodes, or ``executors''.  Spark also includes the machine learning (ML) library,
 Spark ML, providing support for a selection of machine learning algorithms.  Spark ML includes supervised approaches such
@@
 On the other hand, CursedForest (CF) is specifically designed to handle wide ``cursed'' data. It avoids the relation between
-memory and dimensionality, by not using Vectors. Instead, CursedForest stores data in the Parquet columnar format.
-To build a tree, CursedForest will distribute tasks across compute nodes to calculate the information gain for every feature
-in the feature-set. The result is then reduced to return the feature which gives the greatest information gain.
-This process is then repeated until CursedForest has calculated every node for said tree.
+memory and dimensionality by avoiding calculations that rely on entire feature vectors.
+
+Compared to Spark ML, CursedForest further parallelizes work down to individual features. 
+For example, for each node of a tree, CursedForest will distribute tasks that consist of single features (variants), for every individual.
+Each of these tasks will calculate the information gain for that specific feature.
+Once these tasks have completed, the results are reduced to return the feature which gives the greatest information gain.
+This process is then repeated until CursedForest has calculated the entire decicion tree.
 

### 2016-08-22 doc/publications/plos16/plos16.tex
small changes to paper

@@
   \begin{itemize}
-  \item  convergence results from Biau 2012
+  \item  convergence results from Biau 2012  (done)
   \item consider  \cite{Segal.2004} examples
   \item simulations reproduced from Genuer et al 2010 (no -- we need classification examples)
-  \item check Piotr's simulation
+  \item check Piotr's simulation  (done)
   \item run sparkML or Cursed forest on the  simulation data 
@@
   \item D\'iaz-Uriarte, R., \& Alvarez de Andr\'es, S. (2006). (done)
-  \item Goldstein et al. (2011)
+  \item \cite{Goldstein.et.al.2011}
+notes that the 0, 1, 2 encoding of alleles is an additive model i.e. each additional minor allele increases the effect 
+\begin{verbatim}
+Different genetic effects
+Type	Mechanism	Partition
+Additive	Each additional minor allele increases variation	0, 1, 2
+Dominant	Presence of at least 1 minor allele increases variation	0, 1/2
+Recessive	Two minor alleles needed for variation	0/1, 2
+Heterosis	Heterozygote leads to variation	0/2, 1
+\end{verbatim}
+
   \item Chen (2012)  (done)

### 2016-08-24 doc/publications/plos16/plos16.tex

--- 2016-08-10 doc/publications/plos16/plos16.tex
@@
 pronounced in the health space where whole genome sequence (WGS) technology enabled researchers to interrogate all 3
-billion basepairs of the human genome.  While a large fraction of this data is identical between individuals, \marginpar{base-pairs}
-identifying the relevant and disease specific differences is the focus of complex trait analysis [citation].
+billion base-pairs of the human genome.  
+%While a large fraction of this data is identical between individuals,
+
+Identifying the relevant and disease specific base-pair differences between individuals is the focus of genome wide association studies (GWAs)~\cite{CCC2007}, which typically ascertain the mutations prevalence between affected and unaffected individuals at every genomic position independently.
+% complex trait analysis~\cite{Yang2011}.
+These calculations were typically on a reduced-representation of the genome (SNP-arrays), however with WGS becoming more prevalent deep GWAs  sample numbers Crop ~\cite{Yano2016}
 
 For example, the 1000 Genome data consists of approximately 2500 samples with up to 80 million variants~\cite{1KG2012},
-denoting the differences between individuals.  This dataset consumes close to a terabyte of disk space.  The richness of
+denoting the differences between individuals.  This dataset consumes close to a terabyte of disk space.  
+The richness of
 this data can be utilised in other datasets as well by filling in or ``imputing'' the information at genomic positions

### 2016-08-24 doc/publications/plos16/plos16.tex
first attempt at the intro

@@
 billion base-pairs of the human genome.
-%While a large fraction of this data is identical between individuals,
 
-Identifying the relevant and disease specific base-pair differences between individuals is the focus of genome wide association studies (GWAs)~\cite{CCC2007}, which typically ascertain the mutations prevalence between affected and unaffected individuals at every genomic position independently.
-% complex trait analysis~\cite{Yang2011}.
-These calculations were typically on a reduced-representation of the genome (SNP-arrays), however with WGS becoming more prevalent deep GWAs  sample numbers Crop ~\cite{Yano2016}
+Identifying the relevant and disease specific base-pair difference between individuals is the focus of genome wide association studies (GWAs).
+These analysis are typically performed on only the base-pairs that differ frequently in the population (SNPs) with a linear or logistic regression analysis performed for each SNPs separately~\cite{CCC2007}.
+Not taking joint interactions between SNPs into account when filtering for potential disease mutations has since been demonstrated to to be inadequate, with variants identified from GWAS explaining only a small proportion of the overall heritability~\cite{Manolio2009}.
 
+Random Forests~\cite{Breiman.2001} (RF) are able to capture interactions between features and are hence better suited for modelling complex polygenic diseases. 
+Szymczak2016 {\it et al.} hence relied on random forests to perform variable selection in genome-wide association studies. 
+%DENIS Rob do you want to flesh this out more
+\marginpar{Rob do you want to flesh this out more}
+%(80000000 feat * 28800 s)/ 500000 feat = 4608000 s = 53 days
+%6000 samples
+On a simulated datasets with 500,000 features, r2VIM used 8 h (two threads).
+
+With WGS becoming more prevalent GWAs-type analyses need to be performed on datasets with a much larger number of features~\cite{Yano2016}. 
 For example, the 1000 Genome data consists of approximately 2500 samples with up to 80 million variants~\cite{1KG2012},
-denoting the differences between individuals.  This dataset consumes close to a terabyte of disk space.  
-The richness of
-this data can be utilised in other datasets as well by filling in or ``imputing'' the information at genomic positions
-previously unobserved due to the less detailed resolution of older array technology [citation about imputation].
-\marginpar{citations missing} 
-This means that the large number of samples for which genome-wide association (GWA) data
-is available~\cite{Welter2013} can be extended horizontally by imputing the previously unobserved genotypes.  This
-provides the potential to generate datasets of hundreds of thousands of individuals with millions of variants,
-highlighting the need for incorporating modern compute paradigms to deal with these challenges.
+This dataset consumes close to a terabyte of disk space.  
+The richness of this data can be utilised to ``fill in" or impute the information at previously unobserved genomic positions e.g. from older older array technology~\cite{Howie2012}.
+This provides the potential to impute samples in the GWAs catalogue and generate datasets of hundreds of thousands of individuals with millions of variants, highlighting the need for incorporating modern compute paradigms to deal with these challenges.
 
-% WGS has proven itself in discovering rare variants
-% [http://www.sciencedirect.com/science/article/pii/S2352396414000498] and diagnosing unknown genetic disorders [MATT
-% MIGHT], One challenge with analysing WGS data over, for example, traditional genome-wide association (GWA) data, is
-% developing software and algorithms to deal with the huge amounts of data.  Although the sample size of WGS data-sets
-% may be comparable to GWA, the dimensionality of the data may be many times larger. For example, the 1000 Genome data
-% consists of approximately 2500 individuals with up to 80 million variants. The uncompressed file sizes are close to a
-% terabyte. Of course, it's not just genomic data that is rapidly growing. Many other types of data
+%(80000000 feat * 26.423 s)/167698 feat =12605 s = 3.5 hours
+CloudForest~\cite{Bressler2015} written in Go, achieves fast running times by effective use of the CPU cache, optimising for different classes of features and efficiently multi-threading, it processes 167,698 features in 26.423s.
+%DENIS Rob do you want to flesh this out more
+\marginpar{Rob do you want to flesh this out more}
 
-% Developed b Apache, Hadoop is one platform for dealing with large data sets is Platforms Fortunately computational
-% tools such as Apache Spark are available as a framework for iteratively dealing with big data.
+However, expanding parallelisation beyond multi-threading and utilise CPUs from multiple compute nodes overcomes the limitation to hardware architecture hence allowing to scale to the large dimensional datasets and meet future demands.
+
+%DENIS Write clearer
 Distributing the compute tasks is one strategy to overcome this challenge. Apache Spark, with APIs available for

### 2016-08-30 doc/publications/plos16/plos16.tex
just added a space

--- 2016-08-24 doc/publications/plos16/plos16.tex
@@
 pronounced in the health space where whole genome sequence (WGS) technology enabled researchers to interrogate all 3
-billion base-pairs of the human genome.  
-%While a large fraction of this data is identical between individuals,
-
-Identifying the relevant and disease specific base-pair differences between individuals is the focus of genome wide association studies (GWAs)~\cite{CCC2007}, which typically ascertain the mutations prevalence between affected and unaffected individuals at every genomic position independently.
-% complex trait analysis~\cite{Yang2011}.
-These calculations were typically on a reduced-representation of the genome (SNP-arrays), however with WGS becoming more prevalent deep GWAs  sample numbers Crop ~\cite{Yano2016}
+billion base pairs of the human genome.  While a large fraction of this data is identical between individuals,
+identifying the relevant and disease specific differences is the focus of complex trait analysis [citation].
 
 For example, the 1000 Genome data consists of approximately 2500 samples with up to 80 million variants~\cite{1KG2012},
-denoting the differences between individuals.  This dataset consumes close to a terabyte of disk space.  
-The richness of
+denoting the differences between individuals.  This dataset consumes close to a terabyte of disk space.  The richness of
 this data can be utilised in other datasets as well by filling in or ``imputing'' the information at genomic positions

### 2016-08-31 doc/publications/plos16/plos16.tex
changes to plos16.tex

@@
   \item Chen (2012)  (done)
-  \item VSURF and  varSelRF, (doee) recursive feature (forwards backwards) using variable importance
-  \item ranger package
+  \item VSURF and  varSelRF, (done) recursive feature (forwards backwards) using variable importance
+  \item \cite{Wright.and.Ziegle.2016} ranger package (doing it!)
   \item  r2VIM: A new variable selection method for random forests in
     genome-wide association studies
-  \item CloudForest: A Scalable and Efficient Random Forest Implementation for Biological Data
+  \item CloudForest: A Scalable and Efficient Random Forest
+    Implementation for Biological Data
+   Ensembles of decision trees in go/golang   --  do we need to look
+   at this? 
   \item \cite{Tuv.et.al.2009} feature selection via adding permuted variables. Do we want to consider this? (no, leave it for now)
@@
 pronounced in the health space where whole genome sequence (WGS) technology enabled researchers to interrogate all 3
-billion base pairs of the human genome.  While a large fraction of this data is identical between individuals,
+billion basepairs of the human genome.  While a large fraction of this data is identical between individuals, \marginpar{base-pairs}
 identifying the relevant and disease specific differences is the focus of complex trait analysis [citation].

### 2016-08-31 doc/publications/plos16/plos16.tex
added timings to plos16.tex

@@
 \begin{table}[!ht]
+\begin{minipage}{\textwidth}
 \begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
@@
 & RF - Spark ML &  &  &  \\
-& RF - R & 0.456 & 2.89 & 18GB \\ \hline
-
+& RF - R &  0.05  & 24 min  & 18GB\footnote{using the \texttt{parallel} packages with 10 nodes}\\
+& ranger -R &         0.05 &        1 min 10 seconds  &          16GB  (one processor) \\
+ \hline
+\multirow{3}{*}{phase3\_chr22: 2,504 x 1,103,548} &  &  &  &  \\
+& RF - R & NA  & NA & NA\footnote{long vectors ($> 2^31-1$)  are not supported in the .Fortran interface}\\
+&  H2O   &        0.05 &     16 hours       &   600GB \footnote{java -Xmx600g -jar h2o.jar,   h2o.init(nthreads=20)} \\
+\hline
 \multirow{3}{*}{phase3\_chr1: 2,504 x 6,450,364} & CursedForest & 0.063 (XX) & 1hr 58min & 4GB \\
 & RF - Spark ML & 0.94 & 8hr 29min & 8GB \\
-& LR - Spark ML&  & 2hr 6min & 8GB\\ \hline
-
+& LR - Spark ML&  & 2hr 6min & 8GB\\ 
+& ranger  R   &  0.06   &     28 mins  &            num.threads=10\footnote{don't know how to get the memory used, This
+  was done on dpas-03} \\ 
+\hline
 \multirow{3}{*}{phase3\_chr1-3: 2,504 x 19,328,051}  & CursedForest & 0.046 (XX) & 4hr 0min & 4GB\\
 & RF - Spark ML & &  &\\
-& LR - Spark ML & & 14hr 48min & 8GB \\ \hline
+& LR - Spark ML & & 14hr 48min & 8GB \\ 
+& ranger  R   &  0.05   &     58 mins\footnote{ but took 6 hours to load the data}  &      909GB\footnote{I think this is total memory. The job was run with
+  ntasks-per-node=160 cores-per-socket=10 on ruby }  \\ 
+\hline
 
@@
 \end{adjustwidth}
+\end{minipage}
 \end{table}

### 2016-09-01 doc/publications/plos16/plos16.tex

@@
 
+Identifying the relevant and disease specific base-pair difference between individuals is the focus of genome wide association studies (GWAs).
+These analysis are typically performed on only the base-pairs that differ frequently in the population (SNPs) with a linear or logistic regression analysis performed for each SNPs separately~\cite{CCC2007}.
+Not taking joint interactions between SNPs into account when filtering for potential disease mutations has since been demonstrated to to be inadequate, with variants identified from GWAS explaining only a small proportion of the overall heritability~\cite{Manolio2009}.
+
+Random Forests~\cite{Breiman.2001} (RF) are able to capture interactions between features and are hence better suited for modelling complex polygenic diseases. 
+Szymczak2016 {\it et al.} hence relied on random forests to perform variable selection in genome-wide association studies. 
+%DENIS Rob do you want to flesh this out more
+\marginpar{Rob do you want to flesh this out more}
+%(80000000 feat * 28800 s)/ 500000 feat = 4608000 s = 53 days
+%6000 samples
+On a simulated datasets with 500,000 features, r2VIM used 8 h (two threads).
+
+With WGS becoming more prevalent GWAs-type analyses need to be performed on datasets with a much larger number of features~\cite{Yano2016}. 
 For example, the 1000 Genome data consists of approximately 2500 samples with up to 80 million variants~\cite{1KG2012},
-denoting the differences between individuals.  This dataset consumes close to a terabyte of disk space.  The richness of
-this data can be utilised in other datasets as well by filling in or ``imputing'' the information at genomic positions
-previously unobserved due to the less detailed resolution of older array technology [citation about imputation].
-\marginpar{citations missing} 
-This means that the large number of samples for which genome-wide association (GWA) data
-is available~\cite{Welter2013} can be extended horizontally by imputing the previously unobserved genotypes.  This
-provides the potential to generate datasets of hundreds of thousands of individuals with millions of variants,
-highlighting the need for incorporating modern compute paradigms to deal with these challenges.
+This dataset consumes close to a terabyte of disk space.  
+The richness of this data can be utilised to ``fill in" or impute the information at previously unobserved genomic positions e.g. from older older array technology~\cite{Howie2012}.
+This provides the potential to impute samples in the GWAs catalogue and generate datasets of hundreds of thousands of individuals with millions of variants, highlighting the need for incorporating modern compute paradigms to deal with these challenges.
 
-% WGS has proven itself in discovering rare variants
-% [http://www.sciencedirect.com/science/article/pii/S2352396414000498] and diagnosing unknown genetic disorders [MATT
-% MIGHT], One challenge with analysing WGS data over, for example, traditional genome-wide association (GWA) data, is
-% developing software and algorithms to deal with the huge amounts of data.  Although the sample size of WGS data-sets
-% may be comparable to GWA, the dimensionality of the data may be many times larger. For example, the 1000 Genome data
-% consists of approximately 2500 individuals with up to 80 million variants. The uncompressed file sizes are close to a
-% terabyte. Of course, it's not just genomic data that is rapidly growing. Many other types of data
+%(80000000 feat * 26.423 s)/167698 feat =12605 s = 3.5 hours
+CloudForest~\cite{Bressler2015} written in Go, achieves fast running times by effective use of the CPU cache, optimising for different classes of features and efficiently multi-threading, it processes 167,698 features in 26.423s.
+%DENIS Rob do you want to flesh this out more
+\marginpar{Rob do you want to flesh this out more}
 
-% Developed b Apache, Hadoop is one platform for dealing with large data sets is Platforms Fortunately computational
-% tools such as Apache Spark are available as a framework for iteratively dealing with big data.
+However, expanding parallelisation beyond multi-threading and utilise CPUs from multiple compute nodes overcomes the limitation to hardware architecture hence allowing to scale to the large dimensional datasets and meet future demands.
+
+%DENIS Write clearer
 Distributing the compute tasks is one strategy to overcome this challenge. Apache Spark, with APIs available for

### 2016-09-01 doc/publications/plos16/plos16.tex
updated intro

@@
 
+\newcommand{\cursedforest}{{\sc CursedForest}}
 \newcommand{\lorem}{{\bf LOREM}}
@@
 life~\cite{Loebbecke2015}.  These data sets are not only growing vertically by capturing more and more events but also
-horizontally by capturing more information about these events.  The challenge of big and `wide' data is especially
+horizontally by capturing more information about these events.  The challenge of big and ``wide'' data is especially
 pronounced in the health space where whole genome sequence (WGS) technology enabled researchers to interrogate all 3
-billion base pairs of the human genome.  While a large fraction of this data is identical between individuals,
-identifying the relevant and disease specific differences is the focus of complex trait analysis [citation].
+billion base pairs of the human genome.
 
 Identifying the relevant and disease specific base-pair difference between individuals is the focus of genome wide association studies (GWAs).
-These analysis are typically performed on only the base-pairs that differ frequently in the population (SNPs) with a linear or logistic regression analysis performed for each SNPs separately~\cite{CCC2007}.
-Not taking joint interactions between SNPs into account when filtering for potential disease mutations has since been demonstrated to to be inadequate, with variants identified from GWAS explaining only a small proportion of the overall heritability~\cite{Manolio2009}.
+These analysis are typically performed on only the most frequently differing base-pairs (SNPs) by applying linear or logistic regression analysis to each SNPs separately~\cite{CCC2007}.
+It has since been demonstrated that not taking joint interactions between SNPs into account when filtering for potential disease mutations is inadequate~\cite{Manolio2009}. 
+Complex Trait analysis has demonstrated that in order to explain the overall heritability a model needs to take the joint contribution of all variants into account~\cite{Yang2011}.
 
-Random Forests~\cite{Breiman.2001} (RF) are able to capture interactions between features and are hence better suited for modelling complex polygenic diseases. 
-Szymczak2016 {\it et al.} hence relied on random forests to perform variable selection in genome-wide association studies. 
-%DENIS Rob do you want to flesh this out more
-\marginpar{Rob do you want to flesh this out more}
-%(80000000 feat * 28800 s)/ 500000 feat = 4608000 s = 53 days
-%6000 samples
-On a simulated datasets with 500,000 features, r2VIM used 8 h (two threads).
+Random Forests~\cite{Breiman2001} (RF) are able to capture interactions between features while still allowing to prioritise SNPs for their contribution to disease. 
+The joint interaction is achieved by apply the technique of bootstrap aggregation or bagging to decision trees, while a measure of variable importance can be provided for example by summing the number of times a variable is used to split a node, over all nodes and all trees.
+Processing ``wide'' genomic data poses difficulties for machine learning applications as having more features $p$ than samples $n$ causes the model to overfit the data and generalise poorly. 
+This is known as the ``curse of dimensionality''~\cite{Bauer2014, bellman1961adaptive}. 
+RF are known to be more resistant to overfitting than other techniques~\cite{Segal.2004}. 
+%DENIS Rob, Piotr anything else to add or explain more clearly?
+\marginpar{Rob, Piotr anything else to add or explain more clearly?}
+RF are hence well suited for modelling complex polygenic diseases and identifying the strongest contributing features overall.
 
-With WGS becoming more prevalent GWAs-type analyses need to be performed on datasets with a much larger number of features~\cite{Yano2016}. 
+Several RF implementations exists, most notably {\sc party}~\cite{} and the {\sc random forest} implementation in R~\cite{}. 
+%DENIS Rob, Piotr can you extend this with some more detailed descriptions of party and RF in R
+\marginpar{Rob, Piotr can you extend this with some more detailed descriptions of party and RF in R}
+These traditional implementations are have been reported to not scale well with increasing feature size~\cite{Wright.and.Ziegle.2016}.
+
+Historically successful solutions are hence not catering for the big data demands that especially due to WGS are becoming more prevalent.
+Here, GWAs-type analyses need to be performed on datasets with a much larger number of features~\cite{Yano2016}.
 For example, the 1000 Genome data consists of approximately 2500 samples with up to 80 million variants~\cite{1KG2012},
@@
 %(80000000 feat * 26.423 s)/167698 feat =12605 s = 3.5 hours
-CloudForest~\cite{Bressler2015} written in Go, achieves fast running times by effective use of the CPU cache, optimising for different classes of features and efficiently multi-threading, it processes 167,698 features in 26.423s.
+{\sc CloudForest}~\cite{Bressler2015} written in Go, achieves fast running times by effective use of the CPU cache, optimising for different classes of features and efficiently multi-threading.%, it processes 167,698 features in 26.423s.
 %DENIS Rob do you want to flesh this out more
 \marginpar{Rob do you want to flesh this out more}
+Similarly, {\sc Ranger}~\cite{Wright.and.Ziegle.2016} written in C++ with an R front-end is specifically optimised for dealing with large data sets.
 
-However, expanding parallelisation beyond multi-threading and utilise CPUs from multiple compute nodes overcomes the limitation to hardware architecture hence allowing to scale to the large dimensional datasets and meet future demands.
-
-%DENIS Write clearer
-Distributing the compute tasks is one strategy to overcome this challenge. Apache Spark, with APIs available for
-Scala, Java, Python and R, allows for distributed applications to scale to thousands of nodes.
-A Spark application runs on a ``driver'' node, and this node
-divides and distributes work out to the many ``worker'' nodes, or ``executors''.  Spark also includes the machine learning (ML) library,
-Spark ML, providing support for a selection of machine learning algorithms.  Spark ML includes supervised approaches such
-as decision trees and the ensemble method, random forests, as well as unsupervised approaches such as k-means clustering.
+However, these parallelisation strategies are capable to only utilise the CPUs on one compute node (multi-threading), which we show provides faster compute time on smaller dataset but becomes the bottleneck for datasets comprising of millions of features.
+Hadoop/Spark overcomes this limitation by enabling programs to scale beyond compute-node boundaries and virtually utilise infinite number of CPUs.   
+A Spark application runs on a ``driver'' node, and this node divides and distributes work out to the many ``worker'' nodes, or ``executors''.
 
@@
 
-While unsupervised clustering is useful for selected clinical applications~\cite{Li2015}, a more common analysis task is
-supervised learning, i.e. given a phenotype or clinical label determine the most associated features.  One of the
-simplest supervised machine learning methods is logistic regression (LR), which creates a decision boundary by fitting a
-linear combination of the feature variables.  Processing the ``wide'' genomic data poses difficulties for machine learning
-applications as having more features $p$ than samples $n$ causes the model to overfit the data and generalize poorly.
-This is known as the ``curse of dimensionality''~\cite{Bauer2014}.  
-\marginpar{is this the reference we want for the ``curse of dimensionality''  Bellman (1961) is a more usual reference}
+Here we extend this toolkit to encompass supervised learning using random forest. 
+Spark ML's RF implementation is not able to handle the extremely ``wide'' genomic data as it was developed for large number of samples with only modest dimensionality.  
+Although Spark ML can build a RF model on
+a subset of the data (chromosome 1~and~2), the time taken is excessive due to the huge amount of data being aggregated
+and processed by the single driver node during intermediate stages of building the model (see Result~\ref{comp}).  This
+unbalanced work load where the driver node becomes the bottleneck and worker nodes being idle prevents a seamless
+scaling to larger datasets. The memory requirements per executor also increases with dimensionality due to the data types
+Spark ML uses, which we elaborate on in the discussion.
+
+We therefore developed \cursedforest\, a Spark-based random forest algorithm able to process big and ``wide'' data. 
+%Unlike
+%the standard Apache implementation, CursedForest is able to easily process data with millions of dimensions without
+%excessive resource requirements on the driver or worker nodes.
+%DENIS Piotr this is your 20 trees has the important feature in the top 5 and 100 trees has it always as the top element, do we need to spell this out better, it should be a result section.
+\marginpar{Piotr this is your 20 trees has the important feature in the top 5 and 100 trees has it always as the top element, do we need to spell this out better, it should be a result section.}
+Harnessing the virtually unlimited capability to parallelise tasks, \cursedforest\ can build a large number of trees, which boosts the accuracy for variable selection as demonstrated by {\sc r2VIM}~\cite{Szymczak2016} on smaller datasets.
+
+In the first section we explore some of the properties of random forests, demonstrating that any potential flaws in CursedForest
+are an inherent property of Random Forests. Here we provide demonstrations using synthetic data.
+We then demonstrate the scalability of CursedForest in respect to the dimensionality of the data, building an RF model on whole-genome
+data from the 1000 Genomes Project. Here we report on the accuracy (in terms of the ARI) and the error.
+Finally, given the role different parameter values can play in model construction, we explore the effect tuning these parameters
+can have on the prediction accuracy of a model.
+
+\section{Methods}
+
+% DENIS Rob can you condense this down into what is relevant for this paper and not explained elsewhere in the literature
+\marginpar{Rob can you condense this down into what is relevant for this paper and not explained elsewhere in the literature}
+\section{Overview of Random Forests}
 
@@
 
-\subsection{Random Forest}
+%\subsection{Random Forest}
 
-Random Forests~\cite{Breiman.2001} (RF) apply the technique of bootstrap aggregation or bagging to decision trees.  The training
+Random Forests~\cite{Breiman2001} (RF) apply the technique of bootstrap aggregation or bagging to decision trees.  The training
 data is independently drawn from the joint distribution of $(X,Y)$ and comprises $n$ $(p+1)$-tuples $(x_1,y_1),\ldots, (x_n,y_n)$.
@@
 
+% DENIS why is this relevant 
+\marginpar{why is this relevant, i.e. if it is something we show in the first result section it should be introduced there.}
 \subsubsection{convergence}
@@
 
+% DENIS why is this relevant 
+\marginpar{why is this relevant, i.e. if this is special to VariantSpark it should be explained better.}
 \subsubsection{variable selection}
@@
 
-\section{Methods}
-
 \subsection{VariantSpark.}
@@
 similar to DataFrames in R.
-
-However, we observe that Spark ML's RF implementation is not able to handle the extremely ``wide'' genomic data as
-it was developed for large number of samples with only modest dimensionality.  Although Spark ML can build a RF model on
-a subset of the data (chromosome 1~and~2), the time taken is excessive due to the huge amount of data being aggregated
-and processed by the single driver node during intermediate stages of building the model (see Result~\ref{comp}).  This
-unbalanced work load where the driver node becomes the bottleneck and worker nodes being idle prevents a seamless
-scaling to larger datasets. The memory requirements per executor also increases with dimensionality due to the data types
-Spark ML uses, which we elaborate on in the discussion.
-
-We therefore developed CursedForest, a Spark-based random forest algorithm able to process big and ``wide'' data. Unlike
-the standard Apache implementation, CursedForest is able to easily process data with millions of dimensions without
-excessive resource requirements on the driver or worker nodes.
-
-In the first section we explore some of the properties of random forests, demonstrating that any potential flaws in CursedForest
-are an inherent property of Random Forests. Here we provide demonstrations using synthetic data.
-We then demonstrate the scalability of CursedForest in respect to the dimensionality of the data, building an RF model on whole-genome
-data from the 1000 Genomes Project. Here we report on the accuracy (in terms of the ARI) and the error.
-Finally, given the role different parameter values can play in model construction, we explore the effect tuning these parameters
-can have on the prediction accuracy of a model.
-
-We use the same pre-processing algorithms in CursedForests. One difference is that we
 
@@
 \section{Results and Discussion}
-
-%OPTIONS
-% Spark ML vs MLlib
-% random forest, kmeans, logistic regression 
-% synthetic data demonstrating random forest can cope with n<<p
-% James dataset
-% GWAS dataset
 

### 2016-09-01 doc/publications/plos16/plos16.tex
small change to plos16.tex

@@
 
-Several RF implementations exists, most notably {\sc party}~\cite{} and the {\sc random forest} implementation in R~\cite{}. 
-%DENIS Rob, Piotr can you extend this with some more detailed descriptions of party and RF in R
+RF are very easy to parallize. As the forest is a sum of descision trees, it it possible to grow separate trees on 
+each processor and sum them. This mode of parallization is widely used for the {\sc random forest} implementation in
+R~\cite{}, and could also be used for the {\sc party}~\cite{} package.
 \marginpar{Rob, Piotr can you extend this with some more detailed descriptions of party and RF in R}
-These traditional implementations are have been reported to not scale well with increasing feature size~\cite{Wright.and.Ziegle.2016}.
+ However,
+this has several limitations. One is that these implementations reach a hard limit in R when they require vectors of a
+length greater than $2^{31}-1$. While R now has basic support for long vectors, not all fuctions can utilize it.
+These traditional implementations are have been reported to not scale well with increasing feature
+size~\cite{Wright.and.Ziegle.2016}.
+
+However a more basic problem may be that the parallization is at too course a grain. The Cursed Forest implemtation has
+a finer grained version of paraliization where the calculation for each split is handed off ot a separate processor.
 

### 2016-09-01 doc/publications/plos16/plos16.tex
spelling

@@
 
-RF are very easy to parallize. As the forest is a sum of descision trees, it it possible to grow separate trees on 
+RF are very easy to parallize. As the forest is a sum of decision trees, it it possible to grow separate trees on 
 each processor and sum them. This mode of parallization is widely used for the {\sc random forest} implementation in
@@
 this has several limitations. One is that these implementations reach a hard limit in R when they require vectors of a
-length greater than $2^{31}-1$. While R now has basic support for long vectors, not all fuctions can utilize it.
+length greater than $2^{31}-1$. While R now has basic support for long vectors, not all functions can utilize it.
 These traditional implementations are have been reported to not scale well with increasing feature
@@
 
-However a more basic problem may be that the parallization is at too course a grain. The Cursed Forest implemtation has
-a finer grained version of paraliization where the calculation for each split is handed off ot a separate processor.
+However a more basic problem may be that the parallization is at too course a grain. The Cursed Forest implementation has
+a finer grained version of parallization where the calculation for each split is handed off to a separate processor.
 

### 2016-09-02 doc/publications/plos16/plos16.tex
updated intro from robs comments

--- 2016-09-01 doc/publications/plos16/plos16.tex
@@
 Random Forests~\cite{Breiman2001} (RF) are able to capture interactions between features while still allowing to prioritise SNPs for their contribution to disease. 
-The joint interaction is achieved by apply the technique of bootstrap aggregation or bagging to decision trees, while a measure of variable importance can be provided for example by summing the number of times a variable is used to split a node, over all nodes and all trees.
-Processing ``wide'' genomic data poses difficulties for machine learning applications as having more features $p$ than samples $n$ causes the model to overfit the data and generalise poorly. 
-This is known as the ``curse of dimensionality''~\cite{Bauer2014, bellman1961adaptive}. 
-RF are known to be more resistant to overfitting than other techniques~\cite{Segal.2004}. 
-%DENIS Rob, Piotr anything else to add or explain more clearly?
-\marginpar{Rob, Piotr anything else to add or explain more clearly?}
-RF are hence well suited for modelling complex polygenic diseases and identifying the strongest contributing features overall.
+The joint interaction is achieved by appling the technique of bootstrap aggregation or bagging to decision trees, while a measure of variable importance can be provided, for example, by summing the number of times a variable is used to split a node, over all nodes and all trees.
+RF are well suited for processing ``wide'' genomic data for two reasons. 
+Firstly, while other machine learning applications have the propensity to overfit datasets with more features $p$ than samples $n$ (``curse of dimensionality''~\cite{Bauer2014, bellman1961adaptive}), RF are known to be more resistant through not immune to overfitting~\cite{Segal.2004}.
+Secondly, RF are also very easy to parallelise. As the forest is a sum of decision trees, it it possible to grow separate trees on each processor and combine the mature trees.
 
-Several RF implementations exists, most notably {\sc party}~\cite{} and the {\sc random forest} implementation in R~\cite{}. 
-%DENIS Rob, Piotr can you extend this with some more detailed descriptions of party and RF in R
-\marginpar{Rob, Piotr can you extend this with some more detailed descriptions of party and RF in R}
-These traditional implementations are have been reported to not scale well with increasing feature size~\cite{Wright.and.Ziegle.2016}.
+The most popular random forest software packages are R-based, although current implementations do not scale well with increasing feature size~\cite{Wright.and.Ziegle.2016}. 
+This is despite R now supporting long vectors (length greater than $2^{31}$), and the standard {\sc random forest} implementation in R using paralelization to grow multiple trees simultaneously~\cite{Liaw2002}.
+%While R has provisions to support long vectors (length greater than $2^{31}$), current R-based RF implementations, such as {\sc random forest} or {\sc party}~\cite{Hothorn2006}, do not scale well with increasing feature size~\cite{Wright.and.Ziegle.2016}.
+More suitable implementations for ``wide'' genomic data are hence developed in other programming languages.  
+{\sc CloudForest}~\cite{Bressler2015} written in Go, achieves fast running times by effective use of the CPU cache, optimising for different classes of features and efficiently multi-Similarly, {\sc Ranger}~\cite{Wright.and.Ziegle.2016} written in C++ with an R front-end is specifically optimised for dealing with large data sets.
 
-Historically successful solutions are hence not catering for the big data demands that especially due to WGS are becoming more prevalent.
-Here, GWAs-type analyses need to be performed on datasets with a much larger number of features~\cite{Yano2016}.
-For example, the 1000 Genome data consists of approximately 2500 samples with up to 80 million variants~\cite{1KG2012},
-This dataset consumes close to a terabyte of disk space.  
-The richness of this data can be utilised to ``fill in" or impute the information at previously unobserved genomic positions e.g. from older older array technology~\cite{Howie2012}.
+However, the use of traditional compute infrastructure limits the parallelisation strategies that can be employed by these methods. 
+The programs are limited to utilising only CPUs that are on the same computer node (multi-threading) as oppose to using virtually infinite numbers of CPUs across multiple nodes.
+This in turn limits the number of tasks that can be performed in parallel such that only splitting at whole tree-level is feasible as oppose to a more fine grained strategy of handing off the calculation for each tree node to a separate processor.
+Hadoop/Spark overcomes these limitation by enabling programs to scale beyond compute-node boundaries and hence enable more sophisticated parallelisation strategies.  
+%A Spark application runs on a ``driver'' node, and this node divides and distributes work out to the many ``worker'' nodes, or ``executors''.
+
+Here we introduce \cursedforest, a Hadoop/Spark based implementation of random forest specifically designed to cater for ``big" (many samples) and ``wide" (many features) data sets.
+\cursedforest\ is capable of parallelising the split for each node in a tree thereby being able to handle millions of features, such as in the 1000 Genome data, which consists of approximately 2500 samples with up to 80 million variants~\cite{1KG2012}.
+By also utilising Spark to read in and manipulate the standard genomic variant format (VCF) directly, \cursedforest\ outperforms existing tools on even small datasets where multi-threading is considered more performant.
+\marginpar{Piotr this is your 20 trees has the important feature in the top 5 and 100 trees has it always as the top element, do we need to spell this out better, it should be a result section.}
+%Harnessing the virtually unlimited capability to parallelise tasks, \cursedforest\ can build a large number of trees, which boosts the accuracy for variable selection as demonstrated by {\sc r2VIM}~\cite{Szymczak2016} on smaller datasets.
+Harnessing the virtually unlimited capability to parallelise tasks, \cursedforest\ can build a large number of trees, which boosts the accuracy for variable importance selection~\cite{Szymczak2016}.
+
+\cursedforest\ is included in our previously developed framework for Spark-based genomic data analysis, VariantSpark~\cite{OBrien2015}, which therefore now enables supervised and unsupervised machine learning.
+This offers a comprehensive analysis toolkit that can scale for the future data demands, where the richness from WGS data can be utilised to ``fill in" or impute the information at previously unobserved genomic positions in older array technology~\cite{Howie2012}.
 This provides the potential to impute samples in the GWAs catalogue and generate datasets of hundreds of thousands of individuals with millions of variants, highlighting the need for incorporating modern compute paradigms to deal with these challenges.
 
-%(80000000 feat * 26.423 s)/167698 feat =12605 s = 3.5 hours
-{\sc CloudForest}~\cite{Bressler2015} written in Go, achieves fast running times by effective use of the CPU cache, optimising for different classes of features and efficiently multi-threading.%, it processes 167,698 features in 26.423s.
-%DENIS Rob do you want to flesh this out more
-\marginpar{Rob do you want to flesh this out more}
-Similarly, {\sc Ranger}~\cite{Wright.and.Ziegle.2016} written in C++ with an R front-end is specifically optimised for dealing with large data sets.
+%Hadoop/Spark overcomes this limitation by enabling programs to scale beyond compute-node boundaries and virtually utilise infinite number of CPUs.   
+%A Spark application runs on a ``driver'' node, and this node divides and distributes work out to the many ``worker'' nodes, or ``executors''.  
+%Therefore, these historically successful solutions are not catering for the big data demands that especially due to WGS are becoming more prevalent.
+Here, classification or feature importance analysis needs to be performed on datasets with a much larger number of features~\cite{Yano2016}.
+%For example, the 1000 Genome data consists of approximately 2500 samples with up to 80 million variants~\cite{1KG2012},
+%This dataset consumes close to a terabyte of disk space.  
+%We previously demonstrated the versatility and scalability of Spark by developing VariantSpark~\cite{OBrien2015}, a
+%framework allowing users to easily analyse Variant Call Format (VCF) files using ML algorithms on the Spark framework.
+%Using VariantSpark, we successfully built a k-means model on the full
+%$2500 \times 80$ million matrix to cluster individuals by
+%their ethnicity achieving an Adjusted Rand Index of 0.84 (with 1 being perfect clustering  and -1 random clustering).
 
-However, these parallelisation strategies are capable to only utilise the CPUs on one compute node (multi-threading), which we show provides faster compute time on smaller dataset but becomes the bottleneck for datasets comprising of millions of features.
-Hadoop/Spark overcomes this limitation by enabling programs to scale beyond compute-node boundaries and virtually utilise infinite number of CPUs.   
-A Spark application runs on a ``driver'' node, and this node divides and distributes work out to the many ``worker'' nodes, or ``executors''.
-
-We previously demonstrated the versatility and scalability of Spark by developing VariantSpark~\cite{OBrien2015}, a
-framework allowing users to easily analyse Variant Call Format (VCF) files using ML algorithms on the Spark framework.
-Using VariantSpark, we successfully built a k-means model on the full
-$2500 \times 80$ million matrix to cluster individuals by
-their ethnicity achieving an Adjusted Rand Index of 0.84 (with 1 being perfect clustering  and -1 random clustering).
-
-Here we extend this toolkit to encompass supervised learning using random forest. 
-Spark ML's RF implementation is not able to handle the extremely ``wide'' genomic data as it was developed for large number of samples with only modest dimensionality.  
-Although Spark ML can build a RF model on
-a subset of the data (chromosome 1~and~2), the time taken is excessive due to the huge amount of data being aggregated
-and processed by the single driver node during intermediate stages of building the model (see Result~\ref{comp}).  This
-unbalanced work load where the driver node becomes the bottleneck and worker nodes being idle prevents a seamless
-scaling to larger datasets. The memory requirements per executor also increases with dimensionality due to the data types
-Spark ML uses, which we elaborate on in the discussion.
-
-We therefore developed \cursedforest\, a Spark-based random forest algorithm able to process big and ``wide'' data. 
+%Spark ML's RF implementation is not able to handle the extremely ``wide'' genomic data as it was developed for large number of samples with only modest dimensionality.  
+%Although Spark ML can build a RF model on
+%a subset of the data (chromosome 1~and~2), the time taken is excessive due to the huge amount of data being aggregated
+%and processed by the single driver node during intermediate stages of building the model (see Result~\ref{comp}).  This
+%unbalanced work load where the driver node becomes the bottleneck and worker nodes being idle prevents a seamless
+%scaling to larger datasets. The memory requirements per executor also increases with dimensionality due to the data types
+%Spark ML uses, which we elaborate on in the discussion.
+%
+%We therefore developed \cursedforest\, a Spark-based random forest algorithm able to process big and ``wide'' data. 
 %Unlike
@@
 %DENIS Piotr this is your 20 trees has the important feature in the top 5 and 100 trees has it always as the top element, do we need to spell this out better, it should be a result section.
-\marginpar{Piotr this is your 20 trees has the important feature in the top 5 and 100 trees has it always as the top element, do we need to spell this out better, it should be a result section.}
-Harnessing the virtually unlimited capability to parallelise tasks, \cursedforest\ can build a large number of trees, which boosts the accuracy for variable selection as demonstrated by {\sc r2VIM}~\cite{Szymczak2016} on smaller datasets.
 

### 2016-09-02 doc/publications/plos16/plos16.tex
methods + result

@@
 
+\marginpar{Needs to be rewritten once we actually have the sections} 
 In the first section we explore some of the properties of random forests, demonstrating that any potential flaws in CursedForest
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%
+% Method
+%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Methods}
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 % DENIS Rob can you condense this down into what is relevant for this paper and not explained elsewhere in the literature
+% >>CONDENSE
 \marginpar{Rob can you condense this down into what is relevant for this paper and not explained elsewhere in the literature}
-\section{Overview of Random Forests}
+\subsection{Overview of Random Forests}
 
@@
 
-\subsection{VariantSpark.}
+% <<CONDENSE
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% This section talks about the cursed forest implementation
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Cursed Forest.}
+% DENIS: Aidan and Piotr can you please explain here how Cursed forest is implemented
 Traditionally, VariantSpark made use of Spark's machine learning algorithms ``Spark ML''. These algorithms require data in a ``DataFrame'',
@@
 
-\section{Results and Discussion}
-
-\marginpar{These are regression examples, we need some classification examples}
-
-To test this empirically we adapt the simulations developed by Genuer {\it et al.}~\cite{Genuer.et.al.2010} to include a
-much larger feature vector $p$ per sample.  The function, shown in Figure \ref{figure:synth}a, is defined on the first 5
-predictor variables. We have added Gaussian noise to the function (Figure \ref{figure:synth}b) and extended the size of
-the training data set from $n=100 \times p=1000$ to $p=1000000$.
-%TODO recovers 2 out of 5 ? I think you need to spell out why this is a good demonstrator that RF can deal with such a large number of features...
-%TODO also how is the added noise playing into this ?
-The random forest model recovers variables 1 and 4 as the two
-most important variables (out of 1000000). It fails to uncover the other 3 variables, $\{2,3,5\}$.
-
-% Important ?  In a number of simulations (reproduced from \cite{Genuer.et.al.2010}) we see evidence that random forests
-% can recover the variables defining a functional relationship in the presence of a substantial number of noise
-% variables. See the supplementary information for full details. It is apparently that some of the model parameters,
-% such as \texttt{mtry} -- the number of randomly selected variables considered as potential splits at each node, have
-% to be altered from their default value when there are a large number of predictors.
-
-% It is likely that in the genomic setting, variables will have a monotonic influence on the outcome.  We have taken
-% \cite{Genuer.et.al.2010}'s tree function example and extended it to a much larger number of noise variables.  The
-% function, shown in figure \ref{figure:tree_function.png}, is defined on the first 5 predictor variables. We have added
-% Gaussian noise to the function (figure \ref{figure:noisy_tree_function.png}) and extended the size of the training
-% data set from $n=100 \times p=1000$ to $n=100 \times p=1000000$.  The random forest model recovers variables 1 and 4
-% as the two most important variables (out of 1000000). It fails to uncover the other 3 variables, $\{2,3,5\}$.
-
-% It would appear that even in setting with an extremely large number of noise variables, it may be possible to recover
-% the significant varaiables if the functional relationship is simple.
-
-% TODO can you make the caption more descriptive, e.g. what are the splits in the three and what are the numbers;
-% similarly for b) there are 100 samples (can you name the index axis samples) and why are there 9 targets (y) ?
-
-\begin{figure}[tbhp]
-\begin{tabular}{ll}
-a)& b)\\
- \includegraphics[totalheight=6cm]{./figs/tree_function.png}&
-  \includegraphics[totalheight=6cm]{./figs/noisy_tree_function.png}
- \end{tabular}
- \captionof{figure}{{\bf Synthetic data demonstrating RF's capability to cope with $p \gg n$.} {\bf a)} The tree
-   function, where each split means XXXX. {\bf b)} A plot of the ordered $y$ variable for the tree example, and the $y$
-   variable with added Gaussian noise (in red).}
-  \label{figure:synth}
-\end{figure}
-
-\subsection{CursedForest processes genome-wide variant data}
-%\label{comp}
-%Spark ML vs CursedForest vs Kmeans vs LR on small and full version (where applicable) of 1000G (performance and
-%time/mem)
-
-\begin{table}[!ht]
-\begin{minipage}{\textwidth}
-\begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
-\caption{
-{\bf Performance comparison between the different machine learning algorithms.}}
-\begin{tabular}{|l+l|l|l|l|l|l|l|}
-\hline
-%\multicolumn{4}{|l|}{\bf Heading1} & \multicolumn{4}{|l|}{\bf Heading2}\\ \hline
-\bf{Genome \%}  & \bf{Method} & \bf{Error (ARI)} & \bf{Runtime} & \bf{Memory/Node} \\
-\hline
-
-\multirow{3}{*}{phase1\_chr22: 1,092 x 490,036} & CursedForest & 0.077 (XX)  &  &  \\
-& RF - Spark ML &  &  &  \\
-& RF - R &  0.05  & 24 min  & 18GB\footnote{using the \texttt{parallel} packages with 10 nodes}\\
-& ranger -R &         0.05 &        1 min 10 seconds  &          16GB  (one processor) \\
- \hline
-\multirow{3}{*}{phase3\_chr22: 2,504 x 1,103,548} &  &  &  &  \\
-& RF - R & NA  & NA & NA\footnote{long vectors ($> 2^31-1$)  are not supported in the .Fortran interface}\\
-&  H2O   &        0.05 &     16 hours       &   600GB \footnote{java -Xmx600g -jar h2o.jar,   h2o.init(nthreads=20)} \\
-\hline
-\multirow{3}{*}{phase3\_chr1: 2,504 x 6,450,364} & CursedForest & 0.063 (XX) & 1hr 58min & 4GB \\
-& RF - Spark ML & 0.94 & 8hr 29min & 8GB \\
-& LR - Spark ML&  & 2hr 6min & 8GB\\ 
-& ranger  R   &  0.06   &     28 mins  &            num.threads=10\footnote{don't know how to get the memory used, This
-  was done on dpas-03} \\ 
-\hline
-\multirow{3}{*}{phase3\_chr1-3: 2,504 x 19,328,051}  & CursedForest & 0.046 (XX) & 4hr 0min & 4GB\\
-& RF - Spark ML & &  &\\
-& LR - Spark ML & & 14hr 48min & 8GB \\ 
-& ranger  R   &  0.05   &     58 mins\footnote{ but took 6 hours to load the data}  &      909GB\footnote{I think this is total memory. The job was run with
-  ntasks-per-node=160 cores-per-socket=10 on ruby }  \\ 
-\hline
-
-\multirow{4}{*}{phase3\_chr1-22: 2,504 x 81,047,467} & CursedForest & XX (0.96) & 7hr 14min & 8GB \\ 
-& RF - Spark ML & - & - & - \\
-& LR - Spark ML & - & - & - \\ 
-& kmeans - Spark ML & - (0.82) & 30hr 44min & 24GB \\ \hline
-\end{tabular}
-\begin{flushleft} 
-Comparison of different various machine learning libraries on different subsets of variant data 
-from the 1000 Genomes Project. Size of the data is listed as individuals x variants.
-
-\end{flushleft}
-\label{table1}
-\end{adjustwidth}
-\end{minipage}
-\end{table}
-
-%\begin{table}
-%\begin{minipage}{\textwidth}
-%\begin{center}
-%\begin{tabular}{|p{4.2cm}|l|l|l|l|l|l|l|}
-%\hline
-%\bf{method} & \bf{Accuracy} & \bf{Runtime} & \bf{Memory/Node} \\
-%\hline
-%RF (R)\footnote{10 nodes, 500 trees, dpas03} 
- %                                  &  0.456 & 2.89 & 18GB \\ \hline
-%CursedForest\footnote{something}
- %                                  & X & X & X \\ \hline
-%\end{tabular}
-%\medskip
-%\begin{flushleft} 
-%\caption{Comparion between Random Forests on R and CursedForests on a smaller
-%dataset. Chromosome 22 - phase1 (2.1GB) }
-%\label{table:smaller_data}
-%\end{flushleft}
-%\end{center}
-%\end{sidewaystable}
-%\end{minipage}
-%\end{table}
-
-% \begin
-% \begin{table}[!ht]
-% \caption{
-% {\bf Performance comparison between the different Machine Learning algorithms.}}
-% \begin{tabular}{|p{4.2cm}|l|l|l|l|l|l|l|}
-% \hline
-% \bf{Genome }  & \bf{method} & \bf{Accuracy} & \bf{Runtime} & \bf{Memory/Node} \\
-% \hline
-%  chr 	 - phase1 (2.1GB)  &  RF (R) &  0.456 & 2.89 & 18GB \\ \hline
-% \end{tabular}
-% \begin{flushleft} 
-% \end{flushleft}
-% \label{table1}
-% %\end{adjustwidth}
-% \end{table}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% This section talks about the datasets
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Datasets}
+%DENIS: Aidan can you please fill this in of where we got it from and how we process it for all the different tools to use
+\subsubsection{1000 Genomes}
+%DENIS: Rob how was the synthetic data generated
+\subsubsection{Synthetic data}
 
@@
 
-We note that it is possible for a RF model to have a very poor fit, and be useless for prediction, and still have a 
-useful ranking of variable importance.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%
+% Results
+%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Results and Discussion}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Section 1:  Feature selection 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{CursedForest grows denser forests to stabilise feature importance ranking}
+In this section we demonstrate that \cursedforest 's ability to rapidly grow large number of trees, \texttt{ntree}, in parallel results in a stable feature importance ranking.
+This concept was originally proposed by Liaw and Weiner~\cite{Liaw.and.Weiner.2002}, who suggested that though the variable importance measures may vary from run to run, the ranking of the variables will stabilise with growing \texttt{ntree}.
+
+To investigate this hypothesis for data sets with large number of feature variables we generated a synthetic dataset ...
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% DENIS: Rob I don't think this is relevant anymore?
+% >>CONDENSE
+%\marginpar{These are regression examples, we need some classification examples}
+%To test this empirically we adapt the simulations developed by Genuer {\it et al.}~\cite{Genuer.et.al.2010} to include a
+%much larger feature vector $p$ per sample.  The function, shown in Figure \ref{figure:synth}a, is defined on the first 5
+%predictor variables. We have added Gaussian noise to the function (Figure \ref{figure:synth}b) and extended the size of
+%the training data set from $n=100 \times p=1000$ to $p=1000000$.
+%%TODO recovers 2 out of 5 ? I think you need to spell out why this is a good demonstrator that RF can deal with such a large number of features...
+%%TODO also how is the added noise playing into this ?
+%The random forest model recovers variables 1 and 4 as the two
+%most important variables (out of 1000000). It fails to uncover the other 3 variables, $\{2,3,5\}$.
+
+%% Important ?  In a number of simulations (reproduced from \cite{Genuer.et.al.2010}) we see evidence that random forests
+%% can recover the variables defining a functional relationship in the presence of a substantial number of noise
+%% variables. See the supplementary information for full details. It is apparently that some of the model parameters,
+%% such as \texttt{mtry} -- the number of randomly selected variables considered as potential splits at each node, have
+%% to be altered from their default value when there are a large number of predictors.
+
+%% It is likely that in the genomic setting, variables will have a monotonic influence on the outcome.  We have taken
+%% \cite{Genuer.et.al.2010}'s tree function example and extended it to a much larger number of noise variables.  The
+%% function, shown in figure \ref{figure:tree_function.png}, is defined on the first 5 predictor variables. We have added
+%% Gaussian noise to the function (figure \ref{figure:noisy_tree_function.png}) and extended the size of the training
+%% data set from $n=100 \times p=1000$ to $n=100 \times p=1000000$.  The random forest model recovers variables 1 and 4
+%% as the two most important variables (out of 1000000). It fails to uncover the other 3 variables, $\{2,3,5\}$.
+
+%% It would appear that even in setting with an extremely large number of noise variables, it may be possible to recover
+%% the significant varaiables if the functional relationship is simple.
+
+%% TODO can you make the caption more descriptive, e.g. what are the splits in the three and what are the numbers;
+%% similarly for b) there are 100 samples (can you name the index axis samples) and why are there 9 targets (y) ?
+
+%\begin{figure}[tbhp]
+%\begin{tabular}{ll}
+%a)& b)\\
+%\includegraphics[totalheight=6cm]{./figs/tree_function.png}&
+% \includegraphics[totalheight=6cm]{./figs/noisy_tree_function.png}
+% \end{tabular}
+% \captionof{figure}{{\bf Synthetic data demonstrating RF's capability to cope with $p \gg n$.} {\bf a)} The tree
+%   function, where each split means XXXX. {\bf b)} A plot of the ordered $y$ variable for the tree example, and the $y$
+%   variable with added Gaussian noise (in red).}
+%  \label{figure:synth}
+%\end{figure}
+% <<CONDENSE
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Feature selection application example 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%\subsection{Feature selection application example}
+%Maybe we should have a feature selection on real data section. We could be using one of the smaller datasets used in the other papers?
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Section 2: Classification selection
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{CursedForest outperforms existing methods for classification problems}
+In this section we compare \cursedforest 's performance against other published methods for predicting ethnicity on the 1000 genomes dataset.
+
+\begin{table}[!ht]
+\begin{minipage}{\textwidth}
+\begin{adjustwidth}{-1.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
+\caption{
+{\bf Performance comparison between the different machine learning algorithms.}}
+\begin{tabular}{|p{1.2cm}|p{1cm}|p{1.5cm}+l|l|p{1.8cm}|p{1.5cm}|}
+\hline
+%\multicolumn{4}{|l|}{\bf Heading1} & \multicolumn{4}{|l|}{\bf Heading2}\\ \hline
+\bf{Data- set} & \bf{Samp- les} & \bf{Feat- ures}  & \bf{Method} & \bf{Error (ARI)} & \bf{Runtime} & \bf{Memory/ Node} \\
+\hline
+%\multirow{3}{*}{phase1\_chr22: 1,092 x 490,036} & CursedForest & 0.077 (XX)  &  &  \\
+phase1 chr22 &1,092 & 490,036 & CursedForest & 0.077 (XX)  &  &  \\
+&&& RF - Spark ML &  &  &  \\
+&&& RF - R &  0.05  & 24 min  & 18GB\footnote{using the \texttt{parallel} packages with 10 nodes}\\
+&&& ranger -R &         0.05 &        1 min 10 seconds  &          16GB  (one processor) \\
+ \hline
+phase3 chr22 & 2,504 & 1,103,548 &  &  &  &  \\
+&&& RF - R & NA  & NA & NA\footnote{long vectors ($> 2^31-1$)  are not supported in the .Fortran interface}\\
+&&&  H2O   &        0.05 &     16 hours       &   600GB \footnote{java -Xmx600g -jar h2o.jar,   h2o.init(nthreads=20)} \\
+\hline
+phase3 chr1 & 2,504 & 6,450,364 & CursedForest & 0.063 (XX) & 1hr 58min & 4GB \\
+&&& RF - Spark ML & 0.94 & 8hr 29min & 8GB \\
+&&& LR - Spark ML&  & 2hr 6min & 8GB\\ 
+&&& ranger  R   &  0.06   &     28 mins  &            num.threads = 10\footnote{don't know how to get the memory used, This
+  was done on dpas-03} \\ 
+\hline
+phase3 chr1-3 & 2,504 & 19,328,051  & CursedForest & 0.046 (XX) & 4hr 0min & 4GB\\
+&&& RF - Spark ML & &  &\\
+&&& LR - Spark ML & & 14hr 48min & 8GB \\ 
+&&& ranger  R   &  0.05   &     58 mins\footnote{ but took 6 hours to load the data}  &      909GB\footnote{I think this is total memory. The job was run with
+  ntasks-per-node=160 cores-per-socket=10 on ruby }  \\ 
+\hline
+phase3 chr1-22 & 2,504 & 81,047,467 & CursedForest & XX (0.96) & 7hr 14min & 8GB \\ 
+&&& RF - Spark ML & - & - & - \\
+&&& LR - Spark ML & - & - & - \\ 
+%&&& kmeans - Spark ML & - (0.82) & 30hr 44min & 24GB \\ 
+\hline
+\end{tabular}
+\begin{flushleft} 
+Comparison of different various machine learning libraries on different subsets of variant data 
+from the 1000 Genomes Project. 
+\end{flushleft}
+\label{table1}
+\end{adjustwidth}
+\end{minipage}
+\end{table}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Section 3: Sensitivity 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Sensitivity of parameter choice}
 
@@
 
-%\subsection{James' dataset}
-%Comparing CursedForest with feature selection + linear model (performance and time/mem)
-
-\subsection{Conclusion}
+\section{Conclusion}
 
@@
 
-%\section*{References}
-% Either type in your references using
-% \begin{thebibliography}{}
-% \bibitem{}
-% Text
-% \end{thebibliography}
-%
-% OR
-%
-% Compile your BiBTeX database using our plos2015.bst
-% style file and paste the contents of your .bbl file
-% here.
-%
-
 \bibliography{./CursedForest.bib}
@@
 
-  \item D\iaz-Uriarte, R., \& Alvarez de Andr\es, S. (2006). Gene selection and classification of microarray data using random
-    forest. BMC Bioinformatics, 7, 3. http://doi.org/10.1186/1471-2105-7-3
-  \item Goldstein, B. A., Polley, E. C., & Briggs, F. B. S. (2011). Random Forests for Genetic Association Studies. Statistical
-    Applications in Genetics and Molecular Biology, 10(1), 32. http://doi.org/10.2202/1544-6115.1691
+%  \item D\iaz-Uriarte, R., \& Alvarez de Andr\es, S. (2006). Gene selection and classification of microarray data using random
+%    forest. BMC Bioinformatics, 7, 3. http://doi.org/10.1186/1471-2105-7-3
+%  \item Goldstein, B. A., Polley, E. C., & Briggs, F. B. S. (2011). Random Forests for Genetic Association Studies. Statistical
+%    Applications in Genetics and Molecular Biology, 10(1), 32. http://doi.org/10.2202/1544-6115.1691
 
-@article{Chen2012323,
-title = "Random forests for genomic data analysis ",
-journal = "Genomics ",
-volume = "99",
-number = "6",
-pages = "323 - 329",
-year = "2012",
-note = "",
-issn = "0888-7543",
-doi = "http://dx.doi.org/10.1016/j.ygeno.2012.04.003",
-url = "http://www.sciencedirect.com/science/article/pii/S0888754312000626",
-author = "Xi Chen and Hemant Ishwaran",
-keywords = "Random forests",
-keywords = "Random survival forests",
-keywords = "Classification",
-keywords = "Prediction",
-keywords = "Variable selection",
-keywords = "Genomic data analysis ",
-abstract = "Random forests (RF) is a popular tree-based ensemble machine learning tool that is highly data adaptive, applies to “large p, small n” problems, and is able to account for correlation as well as interactions among features.
+%@article{Chen2012323,
+%title = "Random forests for genomic data analysis ",
+%journal = "Genomics ",
+%volume = "99",
+%number = "6",
+%pages = "323 - 329",
+%year = "2012",
+%note = "",
+%issn = "0888-7543",
+%doi = "http://dx.doi.org/10.1016/j.ygeno.2012.04.003",
+%url = "http://www.sciencedirect.com/science/article/pii/S0888754312000626",
+%author = "Xi Chen and Hemant Ishwaran",
+%keywords = "Random forests",
+%keywords = "Random survival forests",
+%keywords = "Classification",
+%keywords = "Prediction",
+%keywords = "Variable selection",
+%keywords = "Genomic data analysis ",
+%abstract = "Random forests (RF) is a popular tree-based ensemble machine learning tool that is highly data adaptive, applies to “large p, small n” problems, and is able to account for correlation as well as interactions among features.
 This makes \{RF\} particularly appealing for high-dimensional genomic data analysis.
@@
 "
-}
+%}
 

### 2016-09-09 doc/publications/plos16/plos16.tex
added values to table

@@
 \newcolumntype{+}{!{\vrule width 2pt}}
+
+\usepackage{footmisc}
 
@@
 %\multirow{3}{*}{phase1\_chr22: 1,092 x 490,036} & CursedForest & 0.077 (XX)  &  &  \\
-phase1 chr22 &1,092 & 490,036 & CursedForest & 0.077 (XX)  &  &  \\
-&&& RF - Spark ML &  &  &  \\
-&&& RF - R &  0.05  & 24 min  & 18GB\footnote{using the \texttt{parallel} packages with 10 nodes}\\
-&&& ranger -R &         0.05 &        1 min 10 seconds  &          16GB  (one processor) \\
+phase1 chr22 &1,092 & 490,036 & CursedForest & 0.077 (XX)  & 9min 16sec & 1GB\footnote{\label{note10}10 nodes} \\
+&&& RF - Spark ML & 0.069 (0.89) & 14min 13sec & 1GB\footref{note10}  \\
+&&& RF - R &  0.05  & 24min  & 18GB\footref{note1}\footnote{using the \texttt{parallel} packages}\\
+&&& ranger -R &         0.05 &        1min 10sec  &          16GB  (one processor) \\
  \hline
 phase3 chr22 & 2,504 & 1,103,548 &  &  &  &  \\
+&&& CursedForest & 0.053 & 9 min 8 sec & 2GB\footnote{\label{note20}20 nodes}\\
+&&& RF - Spark ML & 0.058 (0.92) & 1hr 1min & 2GB\footref{note20} \\
 &&& RF - R & NA  & NA & NA\footnote{long vectors ($> 2^31-1$)  are not supported in the .Fortran interface}\\
@@
 phase3 chr1 & 2,504 & 6,450,364 & CursedForest & 0.063 (XX) & 1hr 58min & 4GB \\
-&&& RF - Spark ML & 0.94 & 8hr 29min & 8GB \\
+&&& RF - Spark ML & (0.94) & 8hr 29min & 8GB \\
 &&& LR - Spark ML&  & 2hr 6min & 8GB\\ 
@@
 phase3 chr1-3 & 2,504 & 19,328,051  & CursedForest & 0.046 (XX) & 4hr 0min & 4GB\\
-&&& RF - Spark ML & &  &\\
+&&& RF - Spark ML & - & -  & -\\
 &&& LR - Spark ML & & 14hr 48min & 8GB \\ 
@@
 from the 1000 Genomes Project.
+Each random forest model consists of 50 trees.
 \end{flushleft}

### 2016-09-14 doc/publications/plos16/plos16.tex
brief start to feature importance

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Section 3: Feature Selection 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{CursedForest selects previously demonstrated cancer variants }
+Feature selection can be useful for determining variables the contribute to the label or for perhaps selecting the n most important
+variables for further analysis. Our Random Forest implementation allows the user to retrieve a subsection of features, ranked
+by their importance. We demonstrate this feature on The Cancer Genome Atlas (TCGA) BRCA dataset. We build this model on
+VCF files retrieved from the TCGA data portal. For the label, we use the cell type (i.e. tumor or normal).
+
+In Table~\ref{table3}, we display the top 10 features, ranked by their Gini index in the Random Forest model.
+
+\begin{table}[!ht]
+%\begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
+\centering
+\caption{
+{\bf Features from the TCGA BRCA dataset ranked by their importance.}}
+\begin{tabular}{|l|l|l|l|l|l|l|}
+\hline
+{\bf Variant} & {\bf Gene} & {\bf Importance}\\ \thickhline
+x:xxxx & cell2 row 1 & cell3 row 1\\ \hline
+x:xxxx & cell2 row 2 & cell3 row 2\\ \hline
+x:xxxx & cell2 row 3 & cell3 row 3\\ \hline
+\end{tabular}
+\begin{flushleft} Here the importance is the gini coefficient when building a model using the cell type as the label.
+\end{flushleft}
+\label{table3}
+%\end{adjustwidth}
+\end{table}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 % Section 3: Sensitivity 

### 2016-09-21 doc/publications/plos16/plos16.tex
changes to table in plos16.tex

--- 2016-09-02 doc/publications/plos16/plos16.tex
@@
 Random Forests~\cite{Breiman2001} (RF) are able to capture interactions between features while still allowing to prioritise SNPs for their contribution to disease. 
-The joint interaction is achieved by appling the technique of bootstrap aggregation or bagging to decision trees, while a measure of variable importance can be provided, for example, by summing the number of times a variable is used to split a node, over all nodes and all trees.
-RF are well suited for processing ``wide'' genomic data for two reasons. 
-Firstly, while other machine learning applications have the propensity to overfit datasets with more features $p$ than samples $n$ (``curse of dimensionality''~\cite{Bauer2014, bellman1961adaptive}), RF are known to be more resistant through not immune to overfitting~\cite{Segal.2004}.
-Secondly, RF are also very easy to parallelise. As the forest is a sum of decision trees, it it possible to grow separate trees on each processor and combine the mature trees.
+The joint interaction is achieved by apply the technique of bootstrap aggregation or bagging to decision trees, while a measure of variable importance can be provided for example by summing the number of times a variable is used to split a node, over all nodes and all trees.
+Processing ``wide'' genomic data poses difficulties for machine learning applications as having more features $p$ than samples $n$ causes the model to overfit the data and generalise poorly. 
+This is known as the ``curse of dimensionality''~\cite{Bauer2014, bellman1961adaptive}. 
+RF are known to be more resistant to overfitting than other techniques~\cite{Segal.2004}. 
+%DENIS Rob, Piotr anything else to add or explain more clearly?
+\marginpar{Rob, Piotr anything else to add or explain more clearly?}
+RF are hence well suited for modelling complex polygenic diseases and identifying the strongest contributing features overall.
 
-The most popular random forest software packages are R-based, although current implementations do not scale well with increasing feature size~\cite{Wright.and.Ziegle.2016}. 
-This is despite R now supporting long vectors (length greater than $2^{31}$), and the standard {\sc random forest} implementation in R using paralelization to grow multiple trees simultaneously~\cite{Liaw2002}.
-%While R has provisions to support long vectors (length greater than $2^{31}$), current R-based RF implementations, such as {\sc random forest} or {\sc party}~\cite{Hothorn2006}, do not scale well with increasing feature size~\cite{Wright.and.Ziegle.2016}.
-More suitable implementations for ``wide'' genomic data are hence developed in other programming languages.  
-{\sc CloudForest}~\cite{Bressler2015} written in Go, achieves fast running times by effective use of the CPU cache, optimising for different classes of features and efficiently multi-Similarly, {\sc Ranger}~\cite{Wright.and.Ziegle.2016} written in C++ with an R front-end is specifically optimised for dealing with large data sets.
+RF are very easy to parallize. As the forest is a sum of decision trees, it it possible to grow separate trees on 
+each processor and sum them. This mode of parallization is widely used for the {\sc random forest} implementation in
+R~\cite{}, and could also be used for the {\sc party}~\cite{} package.
+\marginpar{Rob, Piotr can you extend this with some more detailed descriptions of party and RF in R}
+ However,
+this has several limitations. One is that these implementations reach a hard limit in R when they require vectors of a
+length greater than $2^{31}-1$. While R now has basic support for long vectors, not all functions can utilize it.
+These traditional implementations are have been reported to not scale well with increasing feature
+size~\cite{Wright.and.Ziegle.2016}.
 
-However, the use of traditional compute infrastructure limits the parallelisation strategies that can be employed by these methods. 
-The programs are limited to utilising only CPUs that are on the same computer node (multi-threading) as oppose to using virtually infinite numbers of CPUs across multiple nodes.
-This in turn limits the number of tasks that can be performed in parallel such that only splitting at whole tree-level is feasible as oppose to a more fine grained strategy of handing off the calculation for each tree node to a separate processor.
-Hadoop/Spark overcomes these limitation by enabling programs to scale beyond compute-node boundaries and hence enable more sophisticated parallelisation strategies.  
-%A Spark application runs on a ``driver'' node, and this node divides and distributes work out to the many ``worker'' nodes, or ``executors''.
+However a more basic problem may be that the parallization is at too course a grain. The Cursed Forest implementation has
+a finer grained version of parallization where the calculation for each split is handed off to a separate processor.
 
-Here we introduce \cursedforest, a Hadoop/Spark based implementation of random forest specifically designed to cater for ``big" (many samples) and ``wide" (many features) data sets.
-\cursedforest\ is capable of parallelising the split for each node in a tree thereby being able to handle millions of features, such as in the 1000 Genome data, which consists of approximately 2500 samples with up to 80 million variants~\cite{1KG2012}.
-By also utilising Spark to read in and manipulate the standard genomic variant format (VCF) directly, \cursedforest\ outperforms existing tools on even small datasets where multi-threading is considered more performant.
-\marginpar{Piotr this is your 20 trees has the important feature in the top 5 and 100 trees has it always as the top element, do we need to spell this out better, it should be a result section.}
-%Harnessing the virtually unlimited capability to parallelise tasks, \cursedforest\ can build a large number of trees, which boosts the accuracy for variable selection as demonstrated by {\sc r2VIM}~\cite{Szymczak2016} on smaller datasets.
-Harnessing the virtually unlimited capability to parallelise tasks, \cursedforest\ can build a large number of trees, which boosts the accuracy for variable importance selection~\cite{Szymczak2016}.
-
-\cursedforest\ is included in our previously developed framework for Spark-based genomic data analysis, VariantSpark~\cite{OBrien2015}, which therefore now enables supervised and unsupervised machine learning.
-This offers a comprehensive analysis toolkit that can scale for the future data demands, where the richness from WGS data can be utilised to ``fill in" or impute the information at previously unobserved genomic positions in older array technology~\cite{Howie2012}.
+Historically successful solutions are hence not catering for the big data demands that especially due to WGS are becoming more prevalent.
+Here, GWAs-type analyses need to be performed on datasets with a much larger number of features~\cite{Yano2016}.
+For example, the 1000 Genome data consists of approximately 2500 samples with up to 80 million variants~\cite{1KG2012},
+This dataset consumes close to a terabyte of disk space.  
+The richness of this data can be utilised to ``fill in" or impute the information at previously unobserved genomic positions e.g. from older older array technology~\cite{Howie2012}.
 This provides the potential to impute samples in the GWAs catalogue and generate datasets of hundreds of thousands of individuals with millions of variants, highlighting the need for incorporating modern compute paradigms to deal with these challenges.
 
-%Hadoop/Spark overcomes this limitation by enabling programs to scale beyond compute-node boundaries and virtually utilise infinite number of CPUs.   
-%A Spark application runs on a ``driver'' node, and this node divides and distributes work out to the many ``worker'' nodes, or ``executors''.  
-%Therefore, these historically successful solutions are not catering for the big data demands that especially due to WGS are becoming more prevalent.
-Here, classification or feature importance analysis needs to be performed on datasets with a much larger number of features~\cite{Yano2016}.
-%For example, the 1000 Genome data consists of approximately 2500 samples with up to 80 million variants~\cite{1KG2012},
-%This dataset consumes close to a terabyte of disk space.  
-%We previously demonstrated the versatility and scalability of Spark by developing VariantSpark~\cite{OBrien2015}, a
-%framework allowing users to easily analyse Variant Call Format (VCF) files using ML algorithms on the Spark framework.
-%Using VariantSpark, we successfully built a k-means model on the full
-%$2500 \times 80$ million matrix to cluster individuals by
-%their ethnicity achieving an Adjusted Rand Index of 0.84 (with 1 being perfect clustering  and -1 random clustering).
+%(80000000 feat * 26.423 s)/167698 feat =12605 s = 3.5 hours
+{\sc CloudForest}~\cite{Bressler2015} written in Go, achieves fast running times by effective use of the CPU cache, optimising for different classes of features and efficiently multi-threading.%, it processes 167,698 features in 26.423s.
+%DENIS Rob do you want to flesh this out more
+\marginpar{Rob do you want to flesh this out more}
+Similarly, {\sc Ranger}~\cite{Wright.and.Ziegle.2016} written in C++ with an R front-end is specifically optimised for dealing with large data sets.
 
-%Spark ML's RF implementation is not able to handle the extremely ``wide'' genomic data as it was developed for large number of samples with only modest dimensionality.  
-%Although Spark ML can build a RF model on
-%a subset of the data (chromosome 1~and~2), the time taken is excessive due to the huge amount of data being aggregated
-%and processed by the single driver node during intermediate stages of building the model (see Result~\ref{comp}).  This
-%unbalanced work load where the driver node becomes the bottleneck and worker nodes being idle prevents a seamless
-%scaling to larger datasets. The memory requirements per executor also increases with dimensionality due to the data types
-%Spark ML uses, which we elaborate on in the discussion.
-%
-%We therefore developed \cursedforest\, a Spark-based random forest algorithm able to process big and ``wide'' data. 
-%Unlike
-%the standard Apache implementation, CursedForest is able to easily process data with millions of dimensions without
-%excessive resource requirements on the driver or worker nodes.
-%DENIS Piotr this is your 20 trees has the important feature in the top 5 and 100 trees has it always as the top element, do we need to spell this out better, it should be a result section.
+However, these parallelisation strategies are capable to only utilise the CPUs on one compute node (multi-threading), which we show provides faster compute time on smaller dataset but becomes the bottleneck for datasets comprising of millions of features.
+Hadoop/Spark overcomes this limitation by enabling programs to scale beyond compute-node boundaries and virtually utilise infinite number of CPUs.   
+A Spark application runs on a ``driver'' node, and this node divides and distributes work out to the many ``worker'' nodes, or ``executors''.
 
-\marginpar{Needs to be rewritten once we actually have the sections} 
+We previously demonstrated the versatility and scalability of Spark by developing VariantSpark~\cite{OBrien2015}, a
+framework allowing users to easily analyse Variant Call Format (VCF) files using ML algorithms on the Spark framework.
+Using VariantSpark, we successfully built a k-means model on the full
+$2500 \times 80$ million matrix to cluster individuals by
+their ethnicity achieving an Adjusted Rand Index of 0.84 (with 1 being perfect clustering  and -1 random clustering).
+
+Here we extend this toolkit to encompass supervised learning using random forest. 
+Spark ML's RF implementation is not able to handle the extremely ``wide'' genomic data as it was developed for large number of samples with only modest dimensionality.  
+Although Spark ML can build a RF model on
+a subset of the data (chromosome 1~and~2), the time taken is excessive due to the huge amount of data being aggregated
+and processed by the single driver node during intermediate stages of building the model (see Result~\ref{comp}).  This
+unbalanced work load where the driver node becomes the bottleneck and worker nodes being idle prevents a seamless
+scaling to larger datasets. The memory requirements per executor also increases with dimensionality due to the data types
+Spark ML uses, which we elaborate on in the discussion.
+
+We therefore developed \cursedforest\, a Spark-based random forest algorithm able to process big and ``wide'' data. 
+% Unlike the standard Apache implementation, CursedForest is able to easily process data with millions of dimensions without
+% excessive resource requirements on the driver or worker nodes.  DENIS Piotr this is your 20 trees has the important feature in
+% the top 5 and 100 trees has it always as the top element, do we need to spell this out better, it should be a result section.
+\marginpar{Piotr this is your 20 trees has the important feature in the top 5 and 100 trees has it always as the top element, do
+  we need to spell this out better, it should be a result section.}  
+Harnessing the virtually unlimited capability to parallelise
+tasks, \cursedforest\ can build a large number of trees, which boosts the accuracy for variable selection as demonstrated by {\sc
+  r2VIM}~\cite{Szymczak2016} on smaller datasets.
+
 In the first section we explore some of the properties of random forests, demonstrating that any potential flaws in CursedForest
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%
-% Method
-%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Methods}
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 % DENIS Rob can you condense this down into what is relevant for this paper and not explained elsewhere in the literature
-% >>CONDENSE
 \marginpar{Rob can you condense this down into what is relevant for this paper and not explained elsewhere in the literature}
-\subsection{Overview of Random Forests}
+\section{Overview of Random Forests}
 
@@
 
-% <<CONDENSE
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% This section talks about the cursed forest implementation
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Cursed Forest.}
-% DENIS: Aidan and Piotr can you please explain here how Cursed forest is implemented
+\subsection{VariantSpark.}
 Traditionally, VariantSpark made use of Spark's machine learning algorithms ``Spark ML''. These algorithms require data in a ``DataFrame'',
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% This section talks about the datasets
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Datasets}
-%DENIS: Aidan can you please fill this in of where we got it from and how we process it for all the different tools to use
-\subsubsection{1000 Genomes}
-%DENIS: Rob how was the synthetic data generated
-\subsubsection{Synthetic data}
+\section{Results and Discussion}
+
+\marginpar{These are regression examples, we need some classification examples}
+
+To test this empirically we adapt the simulations developed by Genuer {\it et al.}~\cite{Genuer.et.al.2010} to include a
+much larger feature vector $p$ per sample.  The function, shown in Figure \ref{figure:synth}a, is defined on the first 5
+predictor variables. We have added Gaussian noise to the function (Figure \ref{figure:synth}b) and extended the size of
+the training data set from $n=100 \times p=1000$ to $p=1000000$.
+%TODO recovers 2 out of 5 ? I think you need to spell out why this is a good demonstrator that RF can deal with such a large number of features...
+%TODO also how is the added noise playing into this ?
+The random forest model recovers variables 1 and 4 as the two
+most important variables (out of 1000000). It fails to uncover the other 3 variables, $\{2,3,5\}$.
+
+% Important ?  In a number of simulations (reproduced from \cite{Genuer.et.al.2010}) we see evidence that random forests
+% can recover the variables defining a functional relationship in the presence of a substantial number of noise
+% variables. See the supplementary information for full details. It is apparently that some of the model parameters,
+% such as \texttt{mtry} -- the number of randomly selected variables considered as potential splits at each node, have
+% to be altered from their default value when there are a large number of predictors.
+
+% It is likely that in the genomic setting, variables will have a monotonic influence on the outcome.  We have taken
+% \cite{Genuer.et.al.2010}'s tree function example and extended it to a much larger number of noise variables.  The
+% function, shown in figure \ref{figure:tree_function.png}, is defined on the first 5 predictor variables. We have added
+% Gaussian noise to the function (figure \ref{figure:noisy_tree_function.png}) and extended the size of the training
+% data set from $n=100 \times p=1000$ to $n=100 \times p=1000000$.  The random forest model recovers variables 1 and 4
+% as the two most important variables (out of 1000000). It fails to uncover the other 3 variables, $\{2,3,5\}$.
+
+% It would appear that even in setting with an extremely large number of noise variables, it may be possible to recover
+% the significant varaiables if the functional relationship is simple.
+
+% TODO can you make the caption more descriptive, e.g. what are the splits in the three and what are the numbers;
+% similarly for b) there are 100 samples (can you name the index axis samples) and why are there 9 targets (y) ?
+
+\begin{figure}[tbhp]
+\begin{tabular}{ll}
+a)& b)\\
+ \includegraphics[totalheight=6cm]{./figs/tree_function.png}&
+  \includegraphics[totalheight=6cm]{./figs/noisy_tree_function.png}
+ \end{tabular}
+ \captionof{figure}{{\bf Synthetic data demonstrating RF's capability to cope with $p \gg n$.} {\bf a)} The tree
+   function, where each split means XXXX. {\bf b)} A plot of the ordered $y$ variable for the tree example, and the $y$
+   variable with added Gaussian noise (in red).}
+  \label{figure:synth}
+\end{figure}
+
+\subsection{CursedForest processes genome-wide variant data}
+%\label{comp}
+%Spark ML vs CursedForest vs Kmeans vs LR on small and full version (where applicable) of 1000G (performance and
+%time/mem)
+
+\begin{table}[!ht]
+\begin{minipage}{\textwidth}
+\begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
+\caption{
+{\bf Performance comparison between the different machine learning algorithms.}}
+\begin{tabular}{|l+l|l|l|l|l|l|p{1cm}|}
+\hline
+%\multicolumn{4}{|l|}{\bf Heading1} & \multicolumn{4}{|l|}{\bf Heading2}                               \\ \hline
+\bf{Genome \%}                      & \bf{Method} & \bf{Error (ARI)} & \bf{Runtime} & \bf{Memory/Node} \\
+\hline
+
+\multirow{3}{*}{phase1\_chr22: 1,092 x 490,036} & CursedForest  & 0.077 (XX) &                  &                                                                   \\
+                                                & RF - Spark ML &            &                  &                                                                   \\
+                                                & RF - R        & 0.05       & 24 min           & 18GB\footnote{using the \texttt{parallel} packages with 10 nodes} \\
+                                                & ranger -R     & 0.94       & 1 min 10 seconds & 16GB  (one processor)                                             \\
+                                                & ranger c++    & 0.94       & 7 mins           & --nthreads 20                                                     \\
+
+ \hline
+\multirow{3}{*}{phase3\_chr22: 2,504 x 1,103,548}   &               &            &                  &                                                                                     \\
+                                                    & RF - R        & NA         & NA               & NA\footnote{long vectors ($> 2^31-1$)  are not supported in the .Fortran interface} \\
+                                                    & H2O           & 0.05       & 16 hours         & 600GB \footnote{java -Xmx600g -jar h2o.jar,   h2o.init(nthreads=20)}                \\
+                                                    & ranger  -R    & 0.94       & 4 1/2 mins       & 21GB (one processor)                                                                \\
+\hline
+\multirow{3}{*}{phase3\_chr1: 2,504 x 6,450,364}    & CursedForest  & 0.063 (XX) & 1hr 58min        & 4GB                                                                                 \\
+                                                    & RF - Spark ML & 0.94       & 8hr 29min        & 8GB                                                                                 \\
+                                                    & LR - Spark ML &            & 2hr 6min         & 8GB                                                                                 \\ 
+                                                    & ranger  R     & 0.95       & 28 mins          & num.threads=10\footnote{don't
+                                                                                                      know how to get the memory
+                                                                                                      used,   This was done on dpas-03}  \\ 
+\hline
+\multirow{3}{*}{phase3\_chr1-3: 2,504 x 19,328,051} & CursedForest  & 0.046 (XX) & 4hr 0min                                             & 4GB         \\
+                                                    & RF - Spark ML &            &                                                      &             \\
+                                                    & LR - Spark ML &            & 14hr 48min                                           & 8GB         \\ 
+                                                    & ranger  R     &  0.97      & 58 mins\footnote{ but took 6 hours to load the
+                                                                                   data}
+                                                                                                                                         &
+                                                                                                                                           909GB\footnote{I
+                                                                                                                                           think
+                                                                                                                                           this
+                                                                                                                                           is
+                                                                                                                                           total
+                                                                                                                                           memory. The
+                                                                                                                                           job
+                                                                                                                                           was
+                                                                                                                                           run
+                                                                                                                                           with
+                                                                                                                                                   ntasks-per-node=160
+                                                                                                                                           cores-per-socket=10
+                                                                                                                                           on
+                                                                                                                                           ruby
+                                                                                                                                           }  \\ 
+& ranger   -R        &             0.973     &      03:58           &       pearcey --cpus-per-task=15 --mem 1600gb   906GB \\
+\hline
+
+\multirow{4}{*}{phase3\_chr1-22: 2,504 x 81,047,467} & CursedForest & XX (0.96) & 7hr 14min & 8GB \\ 
+& RF - Spark ML & - & - & - \\
+& LR - Spark ML & - & - & - \\ 
+& kmeans - Spark ML & - (0.82) & 30hr 44min & 24GB \\ 
+& ranger (C++)       &        NA     &        NA     &             80 CPU -- runs out of memory (12.8GB per CPU) \\
+& ranger (C++)       &        NA     &        NA      &            160 CPU -- runs out of time at 20 hours \\
+\hline
+\end{tabular}
+\begin{flushleft} 
+Comparison of different various machine learning libraries on different subsets of variant data 
+from the 1000 Genomes Project. Size of the data is listed as individuals x variants.
+
+\end{flushleft}
+\label{table1}
+\end{adjustwidth}
+\end{minipage}
+\end{table}
+
+%\begin{table}
+%\begin{minipage}{\textwidth}
+%\begin{center}
+%\begin{tabular}{|p{4.2cm}|l|l|l|l|l|l|l|}
+%\hline
+%\bf{method} & \bf{Accuracy} & \bf{Runtime} & \bf{Memory/Node} \\
+%\hline
+%RF (R)\footnote{10 nodes, 500 trees, dpas03} 
+ %                                  &  0.456 & 2.89 & 18GB \\ \hline
+%CursedForest\footnote{something}
+ %                                  & X & X & X \\ \hline
+%\end{tabular}
+%\medskip
+%\begin{flushleft} 
+%\caption{Comparion between Random Forests on R and CursedForests on a smaller
+%dataset. Chromosome 22 - phase1 (2.1GB) }
+%\label{table:smaller_data}
+%\end{flushleft}
+%\end{center}
+%\end{sidewaystable}
+%\end{minipage}
+%\end{table}
+
+% \begin
+% \begin{table}[!ht]
+% \caption{
+% {\bf Performance comparison between the different Machine Learning algorithms.}}
+% \begin{tabular}{|p{4.2cm}|l|l|l|l|l|l|l|}
+% \hline
+% \bf{Genome }  & \bf{method} & \bf{Accuracy} & \bf{Runtime} & \bf{Memory/Node} \\
+% \hline
+%  chr 	 - phase1 (2.1GB)  &  RF (R) &  0.456 & 2.89 & 18GB \\ \hline
+% \end{tabular}
+% \begin{flushleft} 
+% \end{flushleft}
+% \label{table1}
+% %\end{adjustwidth}
+% \end{table}
 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%
-% Results
-%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Results and Discussion}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Section 1:  Feature selection 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{CursedForest grows denser forests to stabilise feature importance ranking}
-In this section we demonstrate that \cursedforest 's ability to rapidly grow large number of trees, \texttt{ntree}, in parallel results in a stable feature importance ranking.
-This concept was originally proposed by Liaw and Weiner~\cite{Liaw.and.Weiner.2002}, who suggested that though the variable importance measures may vary from run to run, the ranking of the variables will stabilise with growing \texttt{ntree}.
-
-To investigate this hypothesis for data sets with large number of feature variables we generated a synthetic dataset ...
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% DENIS: Rob I don't think this is relevant anymore?
-% >>CONDENSE
-%\marginpar{These are regression examples, we need some classification examples}
-%To test this empirically we adapt the simulations developed by Genuer {\it et al.}~\cite{Genuer.et.al.2010} to include a
-%much larger feature vector $p$ per sample.  The function, shown in Figure \ref{figure:synth}a, is defined on the first 5
-%predictor variables. We have added Gaussian noise to the function (Figure \ref{figure:synth}b) and extended the size of
-%the training data set from $n=100 \times p=1000$ to $p=1000000$.
-%%TODO recovers 2 out of 5 ? I think you need to spell out why this is a good demonstrator that RF can deal with such a large number of features...
-%%TODO also how is the added noise playing into this ?
-%The random forest model recovers variables 1 and 4 as the two
-%most important variables (out of 1000000). It fails to uncover the other 3 variables, $\{2,3,5\}$.
-
-%% Important ?  In a number of simulations (reproduced from \cite{Genuer.et.al.2010}) we see evidence that random forests
-%% can recover the variables defining a functional relationship in the presence of a substantial number of noise
-%% variables. See the supplementary information for full details. It is apparently that some of the model parameters,
-%% such as \texttt{mtry} -- the number of randomly selected variables considered as potential splits at each node, have
-%% to be altered from their default value when there are a large number of predictors.
-
-%% It is likely that in the genomic setting, variables will have a monotonic influence on the outcome.  We have taken
-%% \cite{Genuer.et.al.2010}'s tree function example and extended it to a much larger number of noise variables.  The
-%% function, shown in figure \ref{figure:tree_function.png}, is defined on the first 5 predictor variables. We have added
-%% Gaussian noise to the function (figure \ref{figure:noisy_tree_function.png}) and extended the size of the training
-%% data set from $n=100 \times p=1000$ to $n=100 \times p=1000000$.  The random forest model recovers variables 1 and 4
-%% as the two most important variables (out of 1000000). It fails to uncover the other 3 variables, $\{2,3,5\}$.
-
-%% It would appear that even in setting with an extremely large number of noise variables, it may be possible to recover
-%% the significant varaiables if the functional relationship is simple.
-
-%% TODO can you make the caption more descriptive, e.g. what are the splits in the three and what are the numbers;
-%% similarly for b) there are 100 samples (can you name the index axis samples) and why are there 9 targets (y) ?
-
-%\begin{figure}[tbhp]
-%\begin{tabular}{ll}
-%a)& b)\\
-%\includegraphics[totalheight=6cm]{./figs/tree_function.png}&
-% \includegraphics[totalheight=6cm]{./figs/noisy_tree_function.png}
-% \end{tabular}
-% \captionof{figure}{{\bf Synthetic data demonstrating RF's capability to cope with $p \gg n$.} {\bf a)} The tree
-%   function, where each split means XXXX. {\bf b)} A plot of the ordered $y$ variable for the tree example, and the $y$
-%   variable with added Gaussian noise (in red).}
-%  \label{figure:synth}
-%\end{figure}
-% <<CONDENSE
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Feature selection application example 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%\subsection{Feature selection application example}
-%Maybe we should have a feature selection on real data section. We could be using one of the smaller datasets used in the other papers?
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Section 2: Classification selection
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{CursedForest outperforms existing methods for classification problems}
-In this section we compare \cursedforest 's performance against other published methods for predicting ethnicity on the 1000 genomes dataset.
-
-\begin{table}[!ht]
-\begin{minipage}{\textwidth}
-\begin{adjustwidth}{-1.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
-\caption{
-{\bf Performance comparison between the different machine learning algorithms.}}
-\begin{tabular}{|p{1.2cm}|p{1cm}|p{1.5cm}+l|l|p{1.8cm}|p{1.5cm}|}
-\hline
-%\multicolumn{4}{|l|}{\bf Heading1} & \multicolumn{4}{|l|}{\bf Heading2}\\ \hline
-\bf{Data- set} & \bf{Samp- les} & \bf{Feat- ures}  & \bf{Method} & \bf{Error (ARI)} & \bf{Runtime} & \bf{Memory/ Node} \\
-\hline
-%\multirow{3}{*}{phase1\_chr22: 1,092 x 490,036} & CursedForest & 0.077 (XX)  &  &  \\
-phase1 chr22 &1,092 & 490,036 & CursedForest & 0.077 (XX)  &  &  \\
-&&& RF - Spark ML &  &  &  \\
-&&& RF - R &  0.05  & 24 min  & 18GB\footnote{using the \texttt{parallel} packages with 10 nodes}\\
-&&& ranger -R &         0.05 &        1 min 10 seconds  &          16GB  (one processor) \\
- \hline
-phase3 chr22 & 2,504 & 1,103,548 &  &  &  &  \\
-&&& RF - R & NA  & NA & NA\footnote{long vectors ($> 2^31-1$)  are not supported in the .Fortran interface}\\
-&&&  H2O   &        0.05 &     16 hours       &   600GB \footnote{java -Xmx600g -jar h2o.jar,   h2o.init(nthreads=20)} \\
-\hline
-phase3 chr1 & 2,504 & 6,450,364 & CursedForest & 0.063 (XX) & 1hr 58min & 4GB \\
-&&& RF - Spark ML & 0.94 & 8hr 29min & 8GB \\
-&&& LR - Spark ML&  & 2hr 6min & 8GB\\ 
-&&& ranger  R   &  0.06   &     28 mins  &            num.threads = 10\footnote{don't know how to get the memory used, This
-  was done on dpas-03} \\ 
-\hline
-phase3 chr1-3 & 2,504 & 19,328,051  & CursedForest & 0.046 (XX) & 4hr 0min & 4GB\\
-&&& RF - Spark ML & &  &\\
-&&& LR - Spark ML & & 14hr 48min & 8GB \\ 
-&&& ranger  R   &  0.05   &     58 mins\footnote{ but took 6 hours to load the data}  &      909GB\footnote{I think this is total memory. The job was run with
-  ntasks-per-node=160 cores-per-socket=10 on ruby }  \\ 
-\hline
-phase3 chr1-22 & 2,504 & 81,047,467 & CursedForest & XX (0.96) & 7hr 14min & 8GB \\ 
-&&& RF - Spark ML & - & - & - \\
-&&& LR - Spark ML & - & - & - \\ 
-%&&& kmeans - Spark ML & - (0.82) & 30hr 44min & 24GB \\ 
-\hline
-\end{tabular}
-\begin{flushleft} 
-Comparison of different various machine learning libraries on different subsets of variant data 
-from the 1000 Genomes Project. 
-\end{flushleft}
-\label{table1}
-\end{adjustwidth}
-\end{minipage}
-\end{table}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Section 3: Sensitivity 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Sensitivity of parameter choice}
+We note that it is possible for a RF model to have a very poor fit, and be useless for prediction, and still have a 
+useful ranking of variable importance.
 
@@
 
-\section{Conclusion}
+%\subsection{James' dataset}
+%Comparing CursedForest with feature selection + linear model (performance and time/mem)
+
+\subsection{Conclusion}
 
@@
 
+%\section*{References}
+% Either type in your references using
+% \begin{thebibliography}{}
+% \bibitem{}
+% Text
+% \end{thebibliography}
+%
+% OR
+%
+% Compile your BiBTeX database using our plos2015.bst
+% style file and paste the contents of your .bbl file
+% here.
+%
+
 \bibliography{./CursedForest.bib}
@@
 
-%  \item D\iaz-Uriarte, R., \& Alvarez de Andr\es, S. (2006). Gene selection and classification of microarray data using random
-%    forest. BMC Bioinformatics, 7, 3. http://doi.org/10.1186/1471-2105-7-3
-%  \item Goldstein, B. A., Polley, E. C., & Briggs, F. B. S. (2011). Random Forests for Genetic Association Studies. Statistical
-%    Applications in Genetics and Molecular Biology, 10(1), 32. http://doi.org/10.2202/1544-6115.1691
+  \item D\iaz-Uriarte, R., \& Alvarez de Andr\es, S. (2006). Gene selection and classification of microarray data using random
+    forest. BMC Bioinformatics, 7, 3. http://doi.org/10.1186/1471-2105-7-3
+  \item Goldstein, B. A., Polley, E. C., & Briggs, F. B. S. (2011). Random Forests for Genetic Association Studies. Statistical
+    Applications in Genetics and Molecular Biology, 10(1), 32. http://doi.org/10.2202/1544-6115.1691
 
-%@article{Chen2012323,
-%title = "Random forests for genomic data analysis ",
-%journal = "Genomics ",
-%volume = "99",
-%number = "6",
-%pages = "323 - 329",
-%year = "2012",
-%note = "",
-%issn = "0888-7543",
-%doi = "http://dx.doi.org/10.1016/j.ygeno.2012.04.003",
-%url = "http://www.sciencedirect.com/science/article/pii/S0888754312000626",
-%author = "Xi Chen and Hemant Ishwaran",
-%keywords = "Random forests",
-%keywords = "Random survival forests",
-%keywords = "Classification",
-%keywords = "Prediction",
-%keywords = "Variable selection",
-%keywords = "Genomic data analysis ",
-%abstract = "Random forests (RF) is a popular tree-based ensemble machine learning tool that is highly data adaptive, applies to “large p, small n” problems, and is able to account for correlation as well as interactions among features.
+@article{Chen2012323,
+title = "Random forests for genomic data analysis ",
+journal = "Genomics ",
+volume = "99",
+number = "6",
+pages = "323 - 329",
+year = "2012",
+note = "",
+issn = "0888-7543",
+doi = "http://dx.doi.org/10.1016/j.ygeno.2012.04.003",
+url = "http://www.sciencedirect.com/science/article/pii/S0888754312000626",
+author = "Xi Chen and Hemant Ishwaran",
+keywords = "Random forests",
+keywords = "Random survival forests",
+keywords = "Classification",
+keywords = "Prediction",
+keywords = "Variable selection",
+keywords = "Genomic data analysis ",
+abstract = "Random forests (RF) is a popular tree-based ensemble machine learning tool that is highly data adaptive, applies to “large p, small n” problems, and is able to account for correlation as well as interactions among features.
 This makes \{RF\} particularly appealing for high-dimensional genomic data analysis.
@@
 "
-%}
+}
 

### 2016-09-21 doc/publications/plos16/plos16.tex
added Piotr's table

--- 2016-09-14 doc/publications/plos16/plos16.tex
@@
 
+%(80000000 feat * 26.423 s)/167698 feat =12605 s = 3.5 hours
+{\sc CloudForest}~\cite{Bressler2015} written in Go, achieves fast running times by effective use of the CPU cache, optimising for different classes of features and efficiently multi-threading.%, it processes 167,698 features in 26.423s.
+%DENIS Rob do you want to flesh this out more
+\marginpar{Rob do you want to flesh this out more}
+Similarly, {\sc Ranger}~\cite{Wright.and.Ziegle.2016} written in C++ with an R front-end is specifically optimised for dealing with large data sets.
+
+However, these parallelisation strategies are capable to only utilise the CPUs on one compute node (multi-threading), which we show provides faster compute time on smaller dataset but becomes the bottleneck for datasets comprising of millions of features.
+Hadoop/Spark overcomes this limitation by enabling programs to scale beyond compute-node boundaries and virtually utilise infinite number of CPUs.   
+A Spark application runs on a ``driver'' node, and this node divides and distributes work out to the many ``worker'' nodes, or ``executors''.
+
+We previously demonstrated the versatility and scalability of Spark by developing VariantSpark~\cite{OBrien2015}, a
+framework allowing users to easily analyse Variant Call Format (VCF) files using ML algorithms on the Spark framework.
+Using VariantSpark, we successfully built a k-means model on the full
+$2500 \times 80$ million matrix to cluster individuals by
+their ethnicity achieving an Adjusted Rand Index of 0.84 (with 1 being perfect clustering  and -1 random clustering).
+
+Here we extend this toolkit to encompass supervised learning using random forest. 
+Spark ML's RF implementation is not able to handle the extremely ``wide'' genomic data as it was developed for large number of samples with only modest dimensionality.  
+Although Spark ML can build a RF model on
+a subset of the data (chromosome 1~and~2), the time taken is excessive due to the huge amount of data being aggregated
+and processed by the single driver node during intermediate stages of building the model (see Result~\ref{comp}).  This
+unbalanced work load where the driver node becomes the bottleneck and worker nodes being idle prevents a seamless
+scaling to larger datasets. The memory requirements per executor also increases with dimensionality due to the data types
+Spark ML uses, which we elaborate on in the discussion.
+
+We therefore developed \cursedforest\, a Spark-based random forest algorithm able to process big and ``wide'' data. 
+% Unlike the standard Apache implementation, CursedForest is able to easily process data with millions of dimensions without
+% excessive resource requirements on the driver or worker nodes.  DENIS Piotr this is your 20 trees has the important feature in
+% the top 5 and 100 trees has it always as the top element, do we need to spell this out better, it should be a result section.
+\marginpar{Piotr this is your 20 trees has the important feature in the top 5 and 100 trees has it always as the top element, do
+  we need to spell this out better, it should be a result section.}  
+Harnessing the virtually unlimited capability to parallelise
+tasks, \cursedforest\ can build a large number of trees, which boosts the accuracy for variable selection as demonstrated by {\sc
+  r2VIM}~\cite{Szymczak2016} on smaller datasets.
+
+=======
 %Hadoop/Spark overcomes this limitation by enabling programs to scale beyond compute-node boundaries and virtually utilise infinite number of CPUs.   
@@
 % DENIS Rob can you condense this down into what is relevant for this paper and not explained elsewhere in the literature
-% >>CONDENSE
+% 
 \marginpar{Rob can you condense this down into what is relevant for this paper and not explained elsewhere in the literature}
@@
 
-% <<CONDENSE
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 %\end{figure}
-% <<CONDENSE
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 {\bf Performance comparison between the different machine learning algorithms.}}
-\begin{tabular}{|p{1.2cm}|p{1cm}|p{1.5cm}+l|l|p{1.8cm}|p{1.5cm}|}
+\begin{tabular}{|l+l|l|l|l|l|l|p{1cm}|}
 \hline
-%\multicolumn{4}{|l|}{\bf Heading1} & \multicolumn{4}{|l|}{\bf Heading2}\\ \hline
-\bf{Data- set} & \bf{Samp- les} & \bf{Feat- ures}  & \bf{Method} & \bf{Error (ARI)} & \bf{Runtime} & \bf{Memory/ Node} \\
+%\multicolumn{4}{|l|}{\bf Heading1} & \multicolumn{4}{|l|}{\bf Heading2}                               \\ \hline
+\bf{Genome \%}                      & \bf{Method} & \bf{Error (ARI)} & \bf{Runtime} & \bf{Memory/Node} \\
 \hline
-%\multirow{3}{*}{phase1\_chr22: 1,092 x 490,036} & CursedForest & 0.077 (XX)  &  &  \\
-phase1 chr22 &1,092 & 490,036 & CursedForest & 0.077 (XX)  & 9min 16sec & 1GB\footnote{\label{note10}10 nodes} \\
-&&& RF - Spark ML & 0.069 (0.89) & 14min 13sec & 1GB\footref{note10}  \\
-&&& RF - R &  0.05  & 24min  & 18GB\footref{note1}\footnote{using the \texttt{parallel} packages}\\
-&&& ranger -R &         0.05 &        1min 10sec  &          16GB  (one processor) \\
+
+\multirow{3}{*}{phase1\_chr22: 1,092 x 490,036} & CursedForest  & 0.077 (XX) &                  &                                                                   \\
+                                                & RF - Spark ML &            &                  &                                                                   \\
+                                                & RF - R        & 0.05       & 24 min           & 18GB\footnote{using the \texttt{parallel} packages with 10 nodes} \\
+                                                & ranger -R     & 0.94       & 1 min 10 seconds & 16GB  (one processor)                                             \\
+                                                & ranger c++    & 0.94       & 7 mins           & --nthreads 20                                                     \\
+
  \hline
-phase3 chr22 & 2,504 & 1,103,548 &  &  &  &  \\
-&&& CursedForest & 0.053 & 9 min 8 sec & 2GB\footnote{\label{note20}20 nodes}\\
-&&& RF - Spark ML & 0.058 (0.92) & 1hr 1min & 2GB\footref{note20} \\
-&&& RF - R & NA  & NA & NA\footnote{long vectors ($> 2^31-1$)  are not supported in the .Fortran interface}\\
-&&&  H2O   &        0.05 &     16 hours       &   600GB \footnote{java -Xmx600g -jar h2o.jar,   h2o.init(nthreads=20)} \\
+\multirow{3}{*}{phase3\_chr22: 2,504 x 1,103,548}   &               &            &                  &                                                                                     \\
+                                                    & RF - R        & NA         & NA               & NA\footnote{long vectors ($> 2^31-1$)  are not supported in the .Fortran interface} \\
+                                                    & H2O           & 0.05       & 16 hours         & 600GB \footnote{java -Xmx600g -jar h2o.jar,   h2o.init(nthreads=20)}                \\
+                                                    & ranger  -R    & 0.94       & 4 1/2 mins       & 21GB (one processor)                                                                \\
 \hline
-phase3 chr1 & 2,504 & 6,450,364 & CursedForest & 0.063 (XX) & 1hr 58min & 4GB \\
-&&& RF - Spark ML & (0.94) & 8hr 29min & 8GB \\
-&&& LR - Spark ML&  & 2hr 6min & 8GB\\ 
-&&& ranger  R   &  0.06   &     28 mins  &            num.threads = 10\footnote{don't know how to get the memory used, This
-  was done on dpas-03} \\ 
+\multirow{3}{*}{phase3\_chr1: 2,504 x 6,450,364}    & CursedForest  & 0.063 (XX) & 1hr 58min        & 4GB                                                                                 \\
+                                                    & RF - Spark ML & 0.94       & 8hr 29min        & 8GB                                                                                 \\
+                                                    & LR - Spark ML &            & 2hr 6min         & 8GB                                                                                 \\ 
+                                                    & ranger  R     & 0.95       & 28 mins          & num.threads=10\footnote{don't
+                                                                                                      know how to get the memory
+                                                                                                      used,   This was done on dpas-03}  \\ 
 \hline
-phase3 chr1-3 & 2,504 & 19,328,051  & CursedForest & 0.046 (XX) & 4hr 0min & 4GB\\
-&&& RF - Spark ML & - & -  & -\\
-&&& LR - Spark ML & & 14hr 48min & 8GB \\ 
-&&& ranger  R   &  0.05   &     58 mins\footnote{ but took 6 hours to load the data}  &      909GB\footnote{I think this is total memory. The job was run with
-  ntasks-per-node=160 cores-per-socket=10 on ruby }  \\ 
+\multirow{3}{*}{phase3\_chr1-3: 2,504 x 19,328,051} & CursedForest  & 0.046 (XX) & 4hr 0min                                             & 4GB         \\
+                                                    & RF - Spark ML &            &                                                      &             \\
+                                                    & LR - Spark ML &            & 14hr 48min                                           & 8GB         \\ 
+                                                    & ranger  R     &  0.97      & 58 mins\footnote{ but took 6 hours to load the
+                                                                                   data}
+                                                                                                                                         &
+                                                                                                                                           909GB\footnote{I
+                                                                                                                                           think
+                                                                                                                                           this
+                                                                                                                                           is
+                                                                                                                                           total
+                                                                                                                                           memory. The
+                                                                                                                                           job
+                                                                                                                                           was
+                                                                                                                                           run
+                                                                                                                                           with
+                                                                                                                                                   ntasks-per-node=160
+                                                                                                                                           cores-per-socket=10
+                                                                                                                                           on
+                                                                                                                                           ruby
+                                                                                                                                           }  \\ 
+& ranger   -R        &             0.973     &      03:58           &       pearcey --cpus-per-task=15 --mem 1600gb   906GB \\
 \hline
-phase3 chr1-22 & 2,504 & 81,047,467 & CursedForest & XX (0.96) & 7hr 14min & 8GB \\ 
-&&& RF - Spark ML & - & - & - \\
-&&& LR - Spark ML & - & - & - \\ 
-%&&& kmeans - Spark ML & - (0.82) & 30hr 44min & 24GB \\ 
-\hline
+
+\multirow{4}{*}{phase3\_chr1-22: 2,504 x 81,047,467} & CursedForest & XX (0.96) & 7hr 14min & 8GB \\ 
+& RF - Spark ML & - & - & - \\
+& LR - Spark ML & - & - & - \\ 
+& kmeans - Spark ML & - (0.82) & 30hr 44min & 24GB \\ 
+& ranger (C++)       &        NA     &        NA     &             80 CPU -- runs out of memory (12.8GB per CPU) \\
+& ranger (C++)       &        NA     &        NA      &            160
+                                                        CPU -- runs
+                                                        out of time at
+                                                        20 hours \\
+
+% \begin{tabular}{|p{1.2cm}|p{1cm}|p{1.5cm}+l|l|p{1.8cm}|p{1.5cm}|}
+% \hline
+% %\multicolumn{4}{|l|}{\bf Heading1} & \multicolumn{4}{|l|}{\bf Heading2}\\ \hline
+% \bf{Data- set} & \bf{Samp- les} & \bf{Feat- ures}  & \bf{Method} & \bf{Error (ARI)} & \bf{Runtime} & \bf{Memory/ Node} \\
+% \hline
+% %\multirow{3}{*}{phase1\_chr22: 1,092 x 490,036} & CursedForest & 0.077 (XX)  &  &  \\
+% phase1 chr22 &1,092 & 490,036 & CursedForest & 0.077 (XX)  & 9min 16sec & 1GB\footnote{\label{note10}10 nodes} \\
+% &&& RF - Spark ML & 0.069 (0.89) & 14min 13sec & 1GB\footref{note10}  \\
+% &&& RF - R &  0.05  & 24min  & 18GB\footref{note1}\footnote{using the \texttt{parallel} packages}\\
+% &&& ranger -R &         0.05 &        1min 10sec  &          16GB  (one processor) \\
+%  \hline
+% phase3 chr22 & 2,504 & 1,103,548 &  &  &  &  \\
+% &&& CursedForest & 0.053 & 9 min 8 sec & 2GB\footnote{\label{note20}20 nodes}\\
+% &&& RF - Spark ML & 0.058 (0.92) & 1hr 1min & 2GB\footref{note20} \\
+% &&& RF - R & NA  & NA & NA\footnote{long vectors ($> 2^31-1$)  are not supported in the .Fortran interface}\\
+% &&&  H2O   &        0.05 &     16 hours       &   600GB \footnote{java -Xmx600g -jar h2o.jar,   h2o.init(nthreads=20)} \\
+% \hline
+% phase3 chr1 & 2,504 & 6,450,364 & CursedForest & 0.063 (XX) & 1hr 58min & 4GB \\
+% &&& RF - Spark ML & (0.94) & 8hr 29min & 8GB \\
+% &&& LR - Spark ML&  & 2hr 6min & 8GB\\ 
+% &&& ranger  R   &  0.06   &     28 mins  &            num.threads = 10\footnote{don't know how to get the memory used, This
+%   was done on dpas-03} \\ 
+% \hline
+% phase3 chr1-3 & 2,504 & 19,328,051  & CursedForest & 0.046 (XX) & 4hr 0min & 4GB\\
+% &&& RF - Spark ML & - & -  & -\\
+% &&& LR - Spark ML & & 14hr 48min & 8GB \\ 
+% &&& ranger  R   &  0.05   &     58 mins\footnote{ but took 6 hours to load the data}  &      909GB\footnote{I think this is total memory. The job was run with
+%   ntasks-per-node=160 cores-per-socket=10 on ruby }  \\ 
+% \hline
+% phase3 chr1-22 & 2,504 & 81,047,467 & CursedForest & XX (0.96) & 7hr 14min & 8GB \\ 
+% &&& RF - Spark ML & - & - & - \\
+% &&& LR - Spark ML & - & - & - \\ 
+% %&&& kmeans - Spark ML & - (0.82) & 30hr 44min & 24GB \\ 
+% \hline
 \end{tabular}
@@
 \end{minipage}
+\end{table}
+
+\begin{table}[!ht]
+\centering
+\caption{
+{\bf Time in seconds to build random forest with 100
+  trees with specified mtry fraction using given number of CPU cores.}}
+\begin{tabular}{|l|l|l|l|l|l|}
+\hline
+\multicolumn{2}{|l|}{\multirow{2}{*}{}}                   & \multicolumn{4}{|c|}{Number of CPU cores}  \\
+                                                            &  & 16   & 32      & 64      & 128  \\
+\multirow{3}{*}{mtry (fraction)}                     & 0.10 & 708.27  & 340.00  & 186.01 & 134.47 \\	
+                                                     & 0.25 & 1134.89 & 535.01  & 270.78 & 179.59 \\	
+                                                     & 0.80 & 2313.47 & 1125.70 & 569.10 & 346.58 \\	
+\hline
+\end{tabular}
+\begin{flushleft} 
+\end{flushleft}
+\label{table10}
 \end{table}

### 2016-09-21 doc/publications/plos16/plos16.tex
Added data for performnce and scalability section

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{CursedForest grows denser forests to stabilise feature importance ranking}
-In this section we demonstrate that \cursedforest 's ability to rapidly grow large number of trees, \texttt{ntree}, in parallel results in a stable feature importance ranking.
-This concept was originally proposed by Liaw and Weiner~\cite{Liaw.and.Weiner.2002}, who suggested that though the variable importance measures may vary from run to run, the ranking of the variables will stabilise with growing \texttt{ntree}.
+\subsection{CursedForest performance and scalability}
+%In this section we demonstrate that \cursedforest 's ability to rapidly grow large number of trees, \texttt{ntree}, in parallel results in a stable feature importance ranking.
+%This concept was originally proposed by Liaw and Weiner~\cite{Liaw.and.Weiner.2002}, who suggested that though the variable importance measures may vary from run to run, the ranking of the variables will stabilise with growing \texttt{ntree}.
 
-To investigate this hypothesis for data sets with large number of feature variables we generated a synthetic dataset ...
+%To investigate this hypothesis for data sets with large number of feature variables we generated a synthetic dataset ...
+
+\begin{table}[!ht]
+\centering
+\caption{
+{\bf Time in seconds to build random forest of 100
+  trees with specified mtry fraction using given number of CPU cores.}}
+\begin{tabular}{|l|l|l|l|l|l|}
+\hline
+\multicolumn{2}{|l|}{\multirow{2}{*}{}}                   & \multicolumn{4}{|c|}{Number of CPU cores}  \\
+                                                            &  & 16   & 32      & 64      & 128  \\
+\multirow{3}{*}{mtry (fraction)}                            & 0.10 & 708.3  & 340.0  & 186.0 & 134.5 \\
+                                                            & 0.25 & 1134.9 & 535.0  & 270.8 & 179.6 \\
+                                                            & 0.50 & 1724.1 & 856.6  & 428.7 & 243.8 \\
+                                                            & 0.80 & 2313.5 & 1125.7 & 569.1 & 364.6 \\
+
+\hline
+\end{tabular}
+\begin{flushleft} 
+\end{flushleft}
+\label{table10}
+\end{table}
+
+\begin{figure}[tbhp]
+\begin{tabular}{ll}
+a)& b)\\
+\includegraphics[totalheight=6cm]{./figs/mtry_cpu.png} & 
+\includegraphics[totalheight=6cm]{./figs/cpu_mtry_trees_per_hour.png} \\
+\end{tabular}
+\captionof{figure}{{\bf Scalablity of the Wide Random Forest on the synthetic dataset of 2.5M features and 5K samples.} 
+   {\bf a)} Time in seconds to build 100 trees for differnet mtry fractions. 
+   {\bf b)} Number of trees build per hour when using a growing number of CPU cores.}
+\label{figure:synth}
+\end{figure}
 
@@
 
-\begin{table}[!ht]
-\centering
-\caption{
-{\bf Time in seconds to build random forest with 100
-  trees with specified mtry fraction using given number of CPU cores.}}
-\begin{tabular}{|l|l|l|l|l|l|}
-\hline
-\multicolumn{2}{|l|}{\multirow{2}{*}{}}                   & \multicolumn{4}{|c|}{Number of CPU cores}  \\
-                                                            &  & 16   & 32      & 64      & 128  \\
-\multirow{3}{*}{mtry (fraction)}                     & 0.10 & 708.27  & 340.00  & 186.01 & 134.47 \\	
-                                                     & 0.25 & 1134.89 & 535.01  & 270.78 & 179.59 \\	
-                                                     & 0.80 & 2313.47 & 1125.70 & 569.10 & 346.58 \\	
-\hline
-\end{tabular}
-\begin{flushleft} 
-\end{flushleft}
-\label{table10}
-\end{table}
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2016-09-22 doc/publications/plos16/plos16.tex
Added table with data size scalability results
Preatiffied the table
fixed CursedForest.bib -- many duplicated entries

@@
 
-The digital revolution is seeing a dramatic increase in data collected about every aspect of 
-life~\cite{Loebbecke2015}.  These data sets are not only growing vertically by capturing more and more events but also
-horizontally by capturing more information about these events.  The challenge of big and ``wide'' data is especially
-pronounced in the health space where whole genome sequence (WGS) technology enabled researchers to interrogate all 3
-billion base pairs of the human genome.
+The digital revolution is seeing a dramatic increase in data collected about every aspect of life~\cite{Loebbecke2015}.
+These data sets are not only growing vertically, by capturing more and more events, but also horizontally by capturing
+more information about these events.  The challenge of big and ``wide'' data is especially pronounced in the health
+space where whole genome sequence (WGS) technology enabled researchers to interrogate all 3 billion base pairs of the
+human genome.
 
@@
 
+%nvars 150000  500000  2500000 10000000  50000000
+%nsamples          
+%1000  21.6  26.7  59.8  133.5 668.8
+%5000  45.7  71.9  186.2 729.3 2887.4
+%10000 70.2  120.2 397.3 2237.8  5747.8
+
+\begin{table}[!ht]
+\centering
+\caption{
+{\bf Time in seconds to build random forest of 100
+  trees on 128 CPU cores with varying numbers of variables and samples.}}
+\begin{tabular}{|l|l||l|l|l|l|l|}
+\hline
+\multicolumn{2}{|l||}{\multirow{2}{*}{}}               & \multicolumn{5}{c|}{Number of variables}  \\
+\cline{3-7}
+\multicolumn{2}{|l||}{}                                & \bf{150,000} & \bf{500,000} & \bf{2,500,000}  & \bf{10,000,000} & \bf{50,000,000} \\
+\hline                                                            
+\multirow{4}{*}{Number of samples}                          & \bf{1000} &  21.6 & 26.7 & 59.8 & 133.5 & 668.8 \\
+                                                            & \bf{5000} & 45.7 & 71.9 & 186.2 & 729.3 & 2887.4 \\
+                                                            & \bf{10000} & 70.2 & 120.2 & 397.3 & 2237.8 & 5747.8 \\
+\hline
+\end{tabular}
+\begin{flushleft} 
+\end{flushleft}
+\label{table10}
+\end{table}
+
 \begin{table}[!ht]
@@
   trees with specified mtry fraction using given number of CPU cores.}}
-\begin{tabular}{|l|l|l|l|l|l|}
+\begin{tabular}{|l|l||l|l|l|l|}
 \hline
-\multicolumn{2}{|l|}{\multirow{2}{*}{}}                   & \multicolumn{4}{|c|}{Number of CPU cores}  \\
-                                                            &  & 16   & 32      & 64      & 128  \\
-\multirow{3}{*}{mtry (fraction)}                            & 0.10 & 708.3  & 340.0  & 186.0 & 134.5 \\
-                                                            & 0.25 & 1134.9 & 535.0  & 270.8 & 179.6 \\
-                                                            & 0.50 & 1724.1 & 856.6  & 428.7 & 243.8 \\
-                                                            & 0.80 & 2313.5 & 1125.7 & 569.1 & 364.6 \\
+\multicolumn{2}{|l||}{\multirow{2}{*}{}}                   & \multicolumn{4}{|c|}{Number of CPU cores}  \\
+\cline{3-6}
+\multicolumn{2}{|l||}{}                                               & \bf{16} & \bf{32}      & \bf{64}      & \bf{128}  \\
+\hline
+\multirow{4}{*}{mtry (fraction)}                            & \bf{0.10} & 708.3  & 340.0  & 186.0 & 134.5 \\
+                                                            & \bf{0.25} & 1134.9 & 535.0  & 270.8 & 179.6 \\
+                                                            & \bf{0.50} & 1724.1 & 856.6  & 428.7 & 243.8 \\
+                                                            & \bf{0.80} & 2313.5 & 1125.7 & 569.1 & 364.6 \\
 

### 2016-09-22 doc/publications/plos16/plos16.tex
small changes to plos16.tex

@@
 
-Identifying the relevant and disease specific base-pair difference between individuals is the focus of genome wide association studies (GWAs).
-These analysis are typically performed on only the most frequently differing base-pairs (SNPs) by applying linear or logistic regression analysis to each SNPs separately~\cite{CCC2007}.
-It has since been demonstrated that not taking joint interactions between SNPs into account when filtering for potential disease mutations is inadequate~\cite{Manolio2009}. 
-Complex Trait analysis has demonstrated that in order to explain the overall heritability a model needs to take the joint contribution of all variants into account~\cite{Yang2011}.
-
-Random Forests~\cite{Breiman2001} (RF) are able to capture interactions between features while still allowing to prioritise SNPs for their contribution to disease. 
-The joint interaction is achieved by appling the technique of bootstrap aggregation or bagging to decision trees, while a measure of variable importance can be provided, for example, by summing the number of times a variable is used to split a node, over all nodes and all trees.
-RF are well suited for processing ``wide'' genomic data for two reasons. 
-Firstly, while other machine learning applications have the propensity to overfit datasets with more features $p$ than samples $n$ (``curse of dimensionality''~\cite{Bauer2014, bellman1961adaptive}), RF are known to be more resistant through not immune to overfitting~\cite{Segal.2004}.
-Secondly, RF are also very easy to parallelise. As the forest is a sum of decision trees, it it possible to grow separate trees on each processor and combine the mature trees.
-
-The most popular random forest software packages are R-based, although current implementations do not scale well with increasing feature size~\cite{Wright.and.Ziegle.2016}. 
-This is despite R now supporting long vectors (length greater than $2^{31}$), and the standard {\sc random forest} implementation in R using paralelization to grow multiple trees simultaneously~\cite{Liaw2002}.
-%While R has provisions to support long vectors (length greater than $2^{31}$), current R-based RF implementations, such as {\sc random forest} or {\sc party}~\cite{Hothorn2006}, do not scale well with increasing feature size~\cite{Wright.and.Ziegle.2016}.
-More suitable implementations for ``wide'' genomic data are hence developed in other programming languages.  
-{\sc CloudForest}~\cite{Bressler2015} written in Go, achieves fast running times by effective use of the CPU cache, optimising for different classes of features and efficiently multi-Similarly, {\sc Ranger}~\cite{Wright.and.Ziegle.2016} written in C++ with an R front-end is specifically optimised for dealing with large data sets.
-
-However, the use of traditional compute infrastructure limits the parallelisation strategies that can be employed by these methods. 
-The programs are limited to utilising only CPUs that are on the same computer node (multi-threading) as oppose to using virtually infinite numbers of CPUs across multiple nodes.
-This in turn limits the number of tasks that can be performed in parallel such that only splitting at whole tree-level is feasible as oppose to a more fine grained strategy of handing off the calculation for each tree node to a separate processor.
-Hadoop/Spark overcomes these limitation by enabling programs to scale beyond compute-node boundaries and hence enable more sophisticated parallelisation strategies.  
-%A Spark application runs on a ``driver'' node, and this node divides and distributes work out to the many ``worker'' nodes, or ``executors''.
-
-Here we introduce \cursedforest, a Hadoop/Spark based implementation of random forest specifically designed to cater for ``big" (many samples) and ``wide" (many features) data sets.
-\cursedforest\ is capable of parallelising the split for each node in a tree thereby being able to handle millions of features, such as in the 1000 Genome data, which consists of approximately 2500 samples with up to 80 million variants~\cite{1KG2012}.
-By also utilising Spark to read in and manipulate the standard genomic variant format (VCF) directly, \cursedforest\ outperforms existing tools on even small datasets where multi-threading is considered more performant.
-\marginpar{Piotr this is your 20 trees has the important feature in the top 5 and 100 trees has it always as the top element, do we need to spell this out better, it should be a result section.}
-%Harnessing the virtually unlimited capability to parallelise tasks, \cursedforest\ can build a large number of trees, which boosts the accuracy for variable selection as demonstrated by {\sc r2VIM}~\cite{Szymczak2016} on smaller datasets.
-Harnessing the virtually unlimited capability to parallelise tasks, \cursedforest\ can build a large number of trees, which boosts the accuracy for variable importance selection~\cite{Szymczak2016}.
-
-\cursedforest\ is included in our previously developed framework for Spark-based genomic data analysis, VariantSpark~\cite{OBrien2015}, which therefore now enables supervised and unsupervised machine learning.
-This offers a comprehensive analysis toolkit that can scale for the future data demands, where the richness from WGS data can be utilised to ``fill in" or impute the information at previously unobserved genomic positions in older array technology~\cite{Howie2012}.
-This provides the potential to impute samples in the GWAs catalogue and generate datasets of hundreds of thousands of individuals with millions of variants, highlighting the need for incorporating modern compute paradigms to deal with these challenges.
-
-%(80000000 feat * 26.423 s)/167698 feat =12605 s = 3.5 hours
-{\sc CloudForest}~\cite{Bressler2015} written in Go, achieves fast running times by effective use of the CPU cache, optimising for different classes of features and efficiently multi-threading.%, it processes 167,698 features in 26.423s.
-%DENIS Rob do you want to flesh this out more
-\marginpar{Rob do you want to flesh this out more}
-Similarly, {\sc Ranger}~\cite{Wright.and.Ziegle.2016} written in C++ with an R front-end is specifically optimised for dealing with large data sets.
-
-However, these parallelisation strategies are capable to only utilise the CPUs on one compute node (multi-threading), which we show provides faster compute time on smaller dataset but becomes the bottleneck for datasets comprising of millions of features.
-Hadoop/Spark overcomes this limitation by enabling programs to scale beyond compute-node boundaries and virtually utilise infinite number of CPUs.   
-A Spark application runs on a ``driver'' node, and this node divides and distributes work out to the many ``worker'' nodes, or ``executors''.
+Identifying the relevant and disease specific base-pair difference between individuals is the focus of genome wide
+association studies (GWAs).  These analysis are typically performed on only the most frequently differing base-pairs
+(SNPs) by applying linear or logistic regression analysis to each SNPs separately~\cite{CCC2007}.  It has since been
+demonstrated that not taking joint interactions between SNPs into account when filtering for potential disease mutations
+is inadequate~\cite{Manolio2009}.  Complex Trait analysis has demonstrated that in order to explain the overall
+heritability a model needs to take the joint contribution of all variants into account~\cite{Yang2011}.
+
+\marginpar{decison trees capture interaction -- there is no main effect term. RF inherit this property. I dont think it
+  has anything to do with boostrapping, the point of which is to ensure that successive trees are not correlated.}
+Random Forests~\cite{Breiman2001} (RF) are able to capture interactions between features while still allowing to
+prioritise SNPs for their contribution to disease.  The joint interaction is achieved by appling the technique of
+bootstrap aggregation or bagging to decision trees, while a measure of variable importance can be provided, for example,
+by summing the number of times a variable is used to split a node, over all nodes and all trees.  RF are well suited for
+processing ``wide'' genomic data for two reasons.  Firstly, while other machine learning applications have the
+propensity to overfit datasets with more features $p$ than samples $n$ (a consequence of the ``curse of dimensionality''~\cite{Bauer2014,
+  bellman1961adaptive}), decision trees are resistant to overfitting. Random Forests are not immune to
+overfitting as shown by ~\cite{Segal.2004} who gives  examples of the case where:
+\begin{itemize}
+\item a single tree is the most appropriate model; and
+\item the varioable are hightly correlated that the boostrap resampling does not give rise to uncorrelated trees.
+\end{itemize}
+However, keeping these difficult cases in mind, RF have a strong calim on being resistant to overfitting.
+Secondly, RF are also very easy to parallelise. As the forest is a sum of decision trees, it it possible to grow
+separate trees on each processor and combine the mature trees.
+
+The most popular random forest software packages are R-based, although current implementations do not scale well with
+increasing feature size~\cite{Wright.and.Ziegle.2016}.  This is despite R now supporting long vectors (length greater
+than $2^{31}$), and the {\sc randomForest} implementation in R lending itself using paralelization to grow multiple trees
+simultaneously~\cite{Liaw2002} and them combine them.
+More suitable implementations for ``wide'' genomic data are hence developed in other programming languages.  {\sc
+  CloudForest}~\cite{Bressler2015} written in Go, achieves fast running times by effective use of the CPU cache,
+optimising for different classes of features and efficiently multi-threading.
+Similarly, {\sc Ranger}~\cite{Wright.and.Ziegle.2016}
+written in C++ with an R front-end is specifically optimised for dealing with large data sets.
+
+However, the use of traditional compute infrastructure limits the parallelisation strategies that can be employed by
+these methods.  The programs are limited to utilising only CPUs that are on the same computer node (multi-threading) or
+else they run on CPUs distributed across nodes by virtue of the fact that there is no communication between the processes (a
+separate tree is grown on each node).  Hadoop/Spark overcomes these limitation by enabling programs to scale beyond
+compute-node boundaries and hence enable more sophisticated parallelisation strategies.
+In the case of Random forests this measn that the the comutations for each node of a tree can be handed off to separate processors.
+
+Here we introduce \cursedforest, a Hadoop/Spark based implementation of random forest specifically designed to cater for
+``big'' (many samples) and ``wide'' (many features) data sets. \cursedforest\ is capable of parallelising the split for
+each node in a tree thereby being able to handle millions of features, such as in the 1000 Genome data, which consists
+of approximately 2500 samples with up to 80 million variants~\cite{1KG2012}.  By also utilising Spark to read in and
+manipulate the standard genomic variant format (VCF) directly, \cursedforest\ outperforms existing tools on even small
+datasets where multi-threading is considered more performant.  \marginpar{Piotr this is your 20 trees has the important
+  feature in the top 5 and 100 trees has it always as the top element, do we need to spell this out better, it should be
+  a result section.}
+%Harnessing the virtually unlimited capability to parallelise tasks, \cursedforest\ can build a large number of trees,
+%which boosts the accuracy for variable selection as demonstrated by {\sc r2VIM}~\cite{Szymczak2016} on smaller
+%datasets. 
+Harnessing the virtually unlimited capability to parallelise tasks, \cursedforest\ can build a large number of trees,
+which boosts the accuracy for variable importance selection~\cite{Szymczak2016}.
+
+\cursedforest\ is included in our previously developed framework for Spark-based genomic data analysis,
+VariantSpark~\cite{OBrien2015}, which therefore now enables supervised and unsupervised machine learning.  This offers a
+comprehensive analysis toolkit that can scale for the future data demands, where the richness from WGS data can be
+utilised to ``fill in'' or impute the information at previously unobserved genomic positions in older array
+technology~\cite{Howie2012}.  This provides the potential to impute samples in the GWAs catalogue and generate datasets
+of hundreds of thousands of individuals with millions of variants, highlighting the need for incorporating modern
+compute paradigms to deal with these challenges.
+
+A Spark application runs on a ``driver'' node, and this node
+divides and distributes work out to the many ``worker'' nodes, or ``executors''.
 
@@
 framework allowing users to easily analyse Variant Call Format (VCF) files using ML algorithms on the Spark framework.
-Using VariantSpark, we successfully built a k-means model on the full
-$2500 \times 80$ million matrix to cluster individuals by
-their ethnicity achieving an Adjusted Rand Index of 0.84 (with 1 being perfect clustering  and -1 random clustering).
-
-Here we extend this toolkit to encompass supervised learning using random forest. 
-Spark ML's RF implementation is not able to handle the extremely ``wide'' genomic data as it was developed for large number of samples with only modest dimensionality.  
-Although Spark ML can build a RF model on
-a subset of the data (chromosome 1~and~2), the time taken is excessive due to the huge amount of data being aggregated
-and processed by the single driver node during intermediate stages of building the model (see Result~\ref{comp}).  This
-unbalanced work load where the driver node becomes the bottleneck and worker nodes being idle prevents a seamless
-scaling to larger datasets. The memory requirements per executor also increases with dimensionality due to the data types
-Spark ML uses, which we elaborate on in the discussion.
+Using VariantSpark, we successfully built a k-means model on the full $2500 \times 80$ million matrix to cluster
+individuals by their ethnicity achieving an Adjusted Rand Index of 0.84 (with 1 being perfect clustering and -1 random
+clustering).
+
+Here we extend this toolkit to encompass supervised learning using random forest.  Spark ML's RF implementation is not
+able to handle the extremely ``wide'' genomic data as it was developed for large number of samples with only modest
+dimensionality.  Although Spark ML can build a RF model on a subset of the data (chromosome 1~and~2), the time taken is
+excessive due to the huge amount of data being aggregated and processed by the single driver node during intermediate
+stages of building the model (see Result~\ref{comp}).  This unbalanced work load where the driver node becomes the
+bottleneck and worker nodes being idle prevents a seamless scaling to larger datasets. The memory requirements per
+executor also increases with dimensionality due to the data types Spark ML uses, which we elaborate on in the
+discussion.
 
@@
 % the top 5 and 100 trees has it always as the top element, do we need to spell this out better, it should be a result section.
-\marginpar{Piotr this is your 20 trees has the important feature in the top 5 and 100 trees has it always as the top element, do
-  we need to spell this out better, it should be a result section.}  
-Harnessing the virtually unlimited capability to parallelise
-tasks, \cursedforest\ can build a large number of trees, which boosts the accuracy for variable selection as demonstrated by {\sc
-  r2VIM}~\cite{Szymczak2016} on smaller datasets.
+\marginpar{Piotr this is your 20 trees has the important feature in the top 5 and 100 trees has it always as the top
+  element, do we need to spell this out better, it should be a result section.}  Harnessing the virtually unlimited
+capability to parallelise tasks, \cursedforest\ can build a large number of trees, which boosts the accuracy for
+variable selection as demonstrated by {\sc r2VIM}~\cite{Szymczak2016} on smaller datasets.
 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%
-% Method
-%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Methods}

### 2016-09-22 doc/publications/plos16/plos16.tex
some changes to plos16.tex

@@
 The digital revolution is seeing a dramatic increase in data collected about every aspect of life~\cite{Loebbecke2015}.
-These data sets are not only growing vertically, by capturing more and more events, but also horizontally by capturing
+These data sets are not only growing vertically, by capturing more events, but also horizontally by capturing
 more information about these events.  The challenge of big and ``wide'' data is especially pronounced in the health
@@
 processing ``wide'' genomic data for two reasons.  Firstly, while other machine learning applications have the
-propensity to overfit datasets with more features $p$ than samples $n$ (a consequence of the ``curse of dimensionality''~\cite{Bauer2014,
-  bellman1961adaptive}), decision trees are resistant to overfitting. Random Forests are not immune to
-overfitting as shown by ~\cite{Segal.2004} who gives  examples of the case where:
+propensity to overfit datasets with more features $p$ than samples $n$ (a consequence of the ``curse of
+dimensionality''~\cite{Bauer2014, bellman1961adaptive}), decision trees are resistant to overfitting. Random Forests are
+not immune to overfitting as shown by ~\cite{Segal.2004} who gives examples of overfitting in the case where:
 \begin{itemize}
 \item a single tree is the most appropriate model; and
-\item the varioable are hightly correlated that the boostrap resampling does not give rise to uncorrelated trees.
+\item the variables are hightly correlated so that the boostrap resampling does not give rise to uncorrelated trees.
 \end{itemize}
-However, keeping these difficult cases in mind, RF have a strong calim on being resistant to overfitting.
+However, keeping these difficult cases in mind, RF have a strong claim to being resistant to overfitting.
 Secondly, RF are also very easy to parallelise. As the forest is a sum of decision trees, it it possible to grow
-separate trees on each processor and combine the mature trees.
+separate trees on each processor and combine the trees.
 
@@
 increasing feature size~\cite{Wright.and.Ziegle.2016}.  This is despite R now supporting long vectors (length greater
-than $2^{31}$), and the {\sc randomForest} implementation in R lending itself using paralelization to grow multiple trees
-simultaneously~\cite{Liaw2002} and them combine them.
-More suitable implementations for ``wide'' genomic data are hence developed in other programming languages.  {\sc
-  CloudForest}~\cite{Bressler2015} written in Go, achieves fast running times by effective use of the CPU cache,
-optimising for different classes of features and efficiently multi-threading.
-Similarly, {\sc Ranger}~\cite{Wright.and.Ziegle.2016}
-written in C++ with an R front-end is specifically optimised for dealing with large data sets.
+than $2^{31}$), and the {\sc randomForest} implementation in R lending itself to paralelization to grow multiple trees
+simultaneously~\cite{Liaw2002} and them combine them.  More suitable implementations for ``wide'' genomic data have been
+developed in other programming languages.  {\sc CloudForest}~\cite{Bressler2015} written in Go, achieves fast running
+times by effective use of the CPU cache, optimising for different classes of features and efficiently multi-threading.
+Similarly, {\sc Ranger}~\cite{Wright.and.Ziegle.2016} written in C++ with an R front-end is specifically optimised for
+dealing with large data sets.
 
@@
 these methods.  The programs are limited to utilising only CPUs that are on the same computer node (multi-threading) or
-else they run on CPUs distributed across nodes by virtue of the fact that there is no communication between the processes (a
-separate tree is grown on each node).  Hadoop/Spark overcomes these limitation by enabling programs to scale beyond
-compute-node boundaries and hence enable more sophisticated parallelisation strategies.
-In the case of Random forests this measn that the the comutations for each node of a tree can be handed off to separate processors.
+else they run on CPUs distributed across nodes by virtue of the fact that there is no communication between the
+processes (a separate tree is grown on each node).  Hadoop/Spark overcomes these limitation by enabling programs to
+scale beyond compute-node boundaries and hence enable more sophisticated parallelisation strategies.  In the case of
+Random forests this means that the the comutations for each node of a tree can be handed off to separate processors.
 
@@
 manipulate the standard genomic variant format (VCF) directly, \cursedforest\ outperforms existing tools on even small
-datasets where multi-threading is considered more performant.  \marginpar{Piotr this is your 20 trees has the important
+datasets where multi-threading is considered more performant. \marginpar{is performant a word?}
+ \marginpar{Piotr this is your 20 trees has the important
   feature in the top 5 and 100 trees has it always as the top element, do we need to spell this out better, it should be
   a result section.}
-%Harnessing the virtually unlimited capability to parallelise tasks, \cursedforest\ can build a large number of trees,
-%which boosts the accuracy for variable selection as demonstrated by {\sc r2VIM}~\cite{Szymczak2016} on smaller
-%datasets. 
 Harnessing the virtually unlimited capability to parallelise tasks, \cursedforest\ can build a large number of trees,
@@
 
-We therefore developed \cursedforest\, a Spark-based random forest algorithm able to process big and ``wide'' data.
-% Unlike the standard Apache implementation, CursedForest is able to easily process data with millions of dimensions without
-% excessive resource requirements on the driver or worker nodes.  DENIS Piotr this is your 20 trees has the important feature in
-% the top 5 and 100 trees has it always as the top element, do we need to spell this out better, it should be a result section.
-\marginpar{Piotr this is your 20 trees has the important feature in the top 5 and 100 trees has it always as the top
-  element, do we need to spell this out better, it should be a result section.}  Harnessing the virtually unlimited
-capability to parallelise tasks, \cursedforest\ can build a large number of trees, which boosts the accuracy for
-variable selection as demonstrated by {\sc r2VIM}~\cite{Szymczak2016} on smaller datasets.
+%% We therefore developed \cursedforest\, a Spark-based random forest algorithm able to process big and ``wide'' data.
+%% % Unlike the standard Apache implementation, CursedForest is able to easily process data with millions of dimensions without
+%% % excessive resource requirements on the driver or worker nodes.  DENIS Piotr this is your 20 trees has the important feature in
+%% % the top 5 and 100 trees has it always as the top element, do we need to spell this out better, it should be a result section.
+%% \marginpar{Piotr this is your 20 trees has the important feature in the top 5 and 100 trees has it always as the top
+%%   element, do we need to spell this out better, it should be a result section.}  Harnessing the virtually unlimited
+%% capability to parallelise tasks, \cursedforest\ can build a large number of trees, which boosts the accuracy for
+%% variable selection as demonstrated by {\sc r2VIM}~\cite{Szymczak2016} on smaller datasets.
 
@@
 \marginpar{Needs to be rewritten once we actually have the sections} 
-In the first section we explore some of the properties of random forests, demonstrating that any potential flaws in CursedForest
-are an inherent property of Random Forests. Here we provide demonstrations using synthetic data.
-We then demonstrate the scalability of CursedForest in respect to the dimensionality of the data, building an RF model on whole-genome
-data from the 1000 Genomes Project. Here we report on the accuracy (in terms of the ARI) and the error.
-Finally, given the role different parameter values can play in model construction, we explore the effect tuning these parameters
-can have on the prediction accuracy of a model.
+In the first section we explore some of the
+properties of random forests, demonstrating that any potential flaws in CursedForest are an inherent property of Random
+Forests. Here we provide demonstrations using synthetic data.  We then demonstrate the scalability of CursedForest in
+respect to the dimensionality of the data, building an RF model on whole-genome data from the 1000 Genomes Project. Here
+we report on the accuracy (in terms of the ARI) and the error.  Finally, given the role different parameter values can
+play in model construction, we explore the effect tuning these parameters can have on the prediction accuracy of a
+model.
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% DENIS Rob can you condense this down into what is relevant for this paper and not explained elsewhere in the literature
-% 
 \marginpar{Rob can you condense this down into what is relevant for this paper and not explained elsewhere in the literature}
@@
 
-%\subsection{Random Forest}
-
 Random Forests~\cite{Breiman2001} (RF) apply the technique of bootstrap aggregation or bagging to decision trees.  The training
@@
 ($X$ is randomly sampled $B$ times with replacement and a tree is grown on each of these $B$ samples). 
-
 The results are combined in an appropriate way. In a classification problem we take the majority vote for the trees over
@@
 Random forests are a variance reduction technique. They take unpruned tree models, which have a low bias but a high
-variance, and by combining them reduce the variance. The prediction error is the sum of the variance and the bias
-squared.
+variance, and by combining them reduce the variance. As the prediction error is the sum of the variance and the bias
+squared, this strategy should lead to a low prediction error.
 
@@
 \begin{itemize}
-\item random forests are resistant against overfitting. However it is not true that they will not overfit; see \cite{Segal.2004}
-  where an example of RF overfitting is  given, caused by the fact that successive trees were correlated;
+\item random forests are resistant against overfitting. However it is not true that they will not overfit \cite[]{Segal.2004};
 \item \cite{Strobl.et.al.2007} demonstrates that the RF algorithm is subject to a bias in variable selection via two mechanisms:
   \begin{enumerate}
-  \item differences in variable scale (for $X_i$ continuous) and number of levels (for $X_i$ discrete) introduces a bias. A uninformative  variable with a large
-    number of levels may be selected over a more informative  variable with fewer levels;
+  \item differences in variable scale (for $X_i$ continuous) and number of levels (for $X_i$ discrete) introduces a
+    bias. A uninformative variable with a large number of levels may be selected over a more informative variable with
+    fewer levels;
   \item bagging introduces a bias, but sampling $(1- 1/e) \approx 0.632$  of the data without replacement seems to be an
@@
 
-\cite{Strobl.et.al.2007} introduce changes to the RF algorithm to fix 1) and 2). 
-However the level effect will be apparent in the CF algorithm as well. \marginpar{we can fix the boosting effect}
-However, there are many examples, (particularly in genomics) where we have data sets that are
+\cite{Strobl.et.al.2007} introduce changes to the RF algorithm to fix 1) and 2).  However the level effect will be
+apparent in the CF algorithm as well.  However, there are many examples, particularly in genomics, where we have data
+sets that are
 \begin{itemize}
-\item very wide
-\item all variables have the same number of levels (measures at different points along the genome)
+\item very wide;
+\item all variables have the same number of levels (for example, measuments at different points along the genome);
 \item the hypothesized functional relationship is of a form amenable to a decision tree.
@@
 
-% DENIS why is this relevant 
-\marginpar{why is this relevant, i.e. if it is something we show in the first result section it should be introduced there.}
 \subsubsection{convergence}
@@
 
-% DENIS why is this relevant 
 \marginpar{why is this relevant, i.e. if this is special to VariantSpark it should be explained better.}
 \subsubsection{variable selection}
-\cite{Genuer.et.al.2010} and \cite{Diaz.and.Alvarez.2006} describe iterative schemes for doing variable selection using the
-importance measure. See \cite{Chen.and.Ishwaran.2012} for a review of the area with particular reference to genomic data.
-\marginpar{ \cite{Chen.and.Ishwaran.2012} cite many papers on extensions of RF to handle GWAS data. What do we need to do in this area}
+\cite{Genuer.et.al.2010} and \cite{Diaz.and.Alvarez.2006} describe iterative schemes for doing variable selection using
+the importance measure. See \cite{Chen.and.Ishwaran.2012} for a review of the area with particular reference to genomic
+data.
+\marginpar{ \cite{Chen.and.Ishwaran.2012} cite many papers on extensions of RF to handle 
+GWAS data. What do we need to do in this area? What about linkage diseqelibrium?}
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% This section talks about the cursed forest implementation
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsection{Cursed Forest.}
 % DENIS: Aidan and Piotr can you please explain here how Cursed forest is implemented
-Traditionally, VariantSpark made use of Spark's machine learning algorithms ``Spark ML''. These algorithms require data in a ``DataFrame'',
-similar to DataFrames in R.
+Traditionally, VariantSpark made use of Spark's machine learning algorithms ``Spark ML''. These algorithms require data
+in a ``DataFrame'', similar to DataFrames in R.
 
 Traditionally, VariantSpark used the random forest algorithm from Spark ML. While Spark ML algorithms demonstrate
-scalability when dealing with a large number of samples, this scalability quickly breaks down as we include more features.
+scalability when dealing with a large number of samples, this scalability quickly breaks down as we include more
+features.
 
 One of the restrictions preventing Spark ML algorithms from scaling to high-dimensionality data is that the feature-set
-for each sample is stored as a vector data type. Spark vectors are stored in resilient distributed dataset (RDDs), and while RDDs
-can be split, vectors cannot. So as we add more samples (in our case individuals), Spark ML can simply split the RDD into more
-partitions and resource requirements will remain relatively stable. However, as the number of dimensions (variants) increases,
-the size of the vectors increases. This results in higher memory-requirements as increasingly large vectors must be loaded
-into memory.
+for each sample is stored as a vector data type. Spark vectors are stored in resilient distributed dataset (RDDs), and
+while RDDs can be split, vectors cannot. So as we add more samples (in our case individuals), Spark ML can simply split
+the RDD into more partitions and resource requirements will remain relatively stable. However, as the number of
+dimensions (variants) increases, the size of the vectors increases. This results in higher memory-requirements as
+increasingly large vectors must be loaded into memory.
 
-On the other hand, CursedForest (CF) is specifically designed to handle wide ``cursed'' data. It avoids the relation between
-memory and dimensionality by avoiding calculations that rely on entire feature vectors.
+On the other hand, CursedForest (CF) is specifically designed to handle wide ``cursed'' data. It avoids the relation
+between memory and dimensionality by avoiding calculations that rely on entire feature vectors.
 
-Compared to Spark ML, CursedForest further parallelizes work down to individual features. 
-For example, for each node of a tree, CursedForest will distribute tasks that consist of single features (variants), for every individual.
-Each of these tasks will calculate the information gain for that specific feature.
-Once these tasks have completed, the results are reduced to return the feature which gives the greatest information gain.
-This process is then repeated until CursedForest has calculated the entire decicion tree.
+Compared to Spark ML, CursedForest further parallelizes work down to individual features.  For example, for each node of
+a tree, CursedForest will distribute tasks that consist of single features (variants), for every individual.  Each of
+these tasks will calculate the information gain for that specific feature.  Once these tasks have completed, the results
+are reduced to return the feature which gives the greatest information gain.  This process is then repeated until
+CursedForest has calculated the entire decicion tree.
 
@@
 \subsection{CursedForest performance and scalability}
-%In this section we demonstrate that \cursedforest 's ability to rapidly grow large number of trees, \texttt{ntree}, in parallel results in a stable feature importance ranking.
-%This concept was originally proposed by Liaw and Weiner~\cite{Liaw.and.Weiner.2002}, who suggested that though the variable importance measures may vary from run to run, the ranking of the variables will stabilise with growing \texttt{ntree}.
+%In this section we demonstrate that \cursedforest 's ability to rapidly grow large number of trees, \texttt{ntree}, in
+%parallel results in a stable feature importance ranking.  This concept was originally proposed by Liaw and
+%Weiner~\cite{Liaw.and.Weiner.2002}, who suggested that though the variable importance measures may vary from run to
+%run, the ranking of the variables will stabilise with growing \texttt{ntree}.
 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% DENIS: Rob I don't think this is relevant anymore?
-% >>CONDENSE
-%\marginpar{These are regression examples, we need some classification examples}
-%To test this empirically we adapt the simulations developed by Genuer {\it et al.}~\cite{Genuer.et.al.2010} to include a
-%much larger feature vector $p$ per sample.  The function, shown in Figure \ref{figure:synth}a, is defined on the first 5
-%predictor variables. We have added Gaussian noise to the function (Figure \ref{figure:synth}b) and extended the size of
-%the training data set from $n=100 \times p=1000$ to $p=1000000$.
-%%TODO recovers 2 out of 5 ? I think you need to spell out why this is a good demonstrator that RF can deal with such a large number of features...
-%%TODO also how is the added noise playing into this ?
-%The random forest model recovers variables 1 and 4 as the two
-%most important variables (out of 1000000). It fails to uncover the other 3 variables, $\{2,3,5\}$.
-
-%% Important ?  In a number of simulations (reproduced from \cite{Genuer.et.al.2010}) we see evidence that random forests
-%% can recover the variables defining a functional relationship in the presence of a substantial number of noise
-%% variables. See the supplementary information for full details. It is apparently that some of the model parameters,
-%% such as \texttt{mtry} -- the number of randomly selected variables considered as potential splits at each node, have
-%% to be altered from their default value when there are a large number of predictors.
-
-%% It is likely that in the genomic setting, variables will have a monotonic influence on the outcome.  We have taken
-%% \cite{Genuer.et.al.2010}'s tree function example and extended it to a much larger number of noise variables.  The
-%% function, shown in figure \ref{figure:tree_function.png}, is defined on the first 5 predictor variables. We have added
-%% Gaussian noise to the function (figure \ref{figure:noisy_tree_function.png}) and extended the size of the training
-%% data set from $n=100 \times p=1000$ to $n=100 \times p=1000000$.  The random forest model recovers variables 1 and 4
-%% as the two most important variables (out of 1000000). It fails to uncover the other 3 variables, $\{2,3,5\}$.
-
-%% It would appear that even in setting with an extremely large number of noise variables, it may be possible to recover
-%% the significant varaiables if the functional relationship is simple.
-
-%% TODO can you make the caption more descriptive, e.g. what are the splits in the three and what are the numbers;
-%% similarly for b) there are 100 samples (can you name the index axis samples) and why are there 9 targets (y) ?
-
-%\begin{figure}[tbhp]
-%\begin{tabular}{ll}
-%a)& b)\\
-%\includegraphics[totalheight=6cm]{./figs/tree_function.png}&
-% \includegraphics[totalheight=6cm]{./figs/noisy_tree_function.png}
-% \end{tabular}
-% \captionof{figure}{{\bf Synthetic data demonstrating RF's capability to cope with $p \gg n$.} {\bf a)} The tree
-%   function, where each split means XXXX. {\bf b)} A plot of the ordered $y$ variable for the tree example, and the $y$
-%   variable with added Gaussian noise (in red).}
-%  \label{figure:synth}
-%\end{figure}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Feature selection application example 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%\subsection{Feature selection application example}
-%Maybe we should have a feature selection on real data section. We could be using one of the smaller datasets used in the other papers?
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \subsection{CursedForest outperforms existing methods for classification problems}
-In this section we compare \cursedforest 's performance against other published methods for predicting ethnicity on the 1000 genomes dataset.
+In this section we compare \cursedforest 's performance against other published methods for predicting ethnicity on the
+1000 genomes dataset.
 

### 2016-09-26 doc/publications/plos16/plos16.tex
Preatifgied tables

--- 2016-09-22 doc/publications/plos16/plos16.tex
@@
 
-The digital revolution is seeing a dramatic increase in data collected about every aspect of life~\cite{Loebbecke2015}.
-These data sets are not only growing vertically, by capturing more and more events, but also horizontally by capturing
-more information about these events.  The challenge of big and ``wide'' data is especially pronounced in the health
-space where whole genome sequence (WGS) technology enabled researchers to interrogate all 3 billion base pairs of the
-human genome.
+The digital revolution is seeing a dramatic increase in data collected about every aspect of 
+life~\cite{Loebbecke2015}.  These data sets are not only growing vertically by capturing more and more events but also
+horizontally by capturing more information about these events.  The challenge of big and ``wide'' data is especially
+pronounced in the health space where whole genome sequence (WGS) technology enabled researchers to interrogate all 3
+billion base pairs of the human genome.
 
@@
   trees on 128 CPU cores with varying numbers of variables and samples.}}
-\begin{tabular}{|l|l||l|l|l|l|l|}
-\hline
-\multicolumn{2}{|l||}{\multirow{2}{*}{}}               & \multicolumn{5}{c|}{Number of variables}  \\
+\begin{tabular}{|l|l|l|l|l|l|l|}
+\hline
+\multicolumn{2}{|l|}{\multirow{2}{*}{}}               & \multicolumn{5}{c|}{Number of variables}  \\
 \cline{3-7}
-\multicolumn{2}{|l||}{}                                & \bf{150,000} & \bf{500,000} & \bf{2,500,000}  & \bf{10,000,000} & \bf{50,000,000} \\
+\multicolumn{2}{|l|}{}                                & \bf{150,000} & \bf{500,000} & \bf{2,500,000}  & \bf{10,000,000} & \bf{50,000,000} \\
 \hline                                                            
@@
 
+\begin{figure}[tbhp]
+\begin{tabular}{ll}
+a)& b)\\
+\includegraphics[totalheight=6cm]{./figs/nsamples_nvars.png} & 
+\includegraphics[totalheight=6cm]{./figs/nvars_nsamples.png} \\
+\end{tabular}
+\captionof{figure}{{\bf Scalablity of the Wide Random Forest on synthetic datasets with varying number of samples and variables.} 
+   {\bf a)} Time in seconds to build 100 trees for differnet mtry fractions. 
+   {\bf b)} Number of trees build per hour when using a growing number of CPU cores.}
+\label{figure:synth}
+\end{figure}
+
 \begin{table}[!ht]
@@
   trees with specified mtry fraction using given number of CPU cores.}}
-\begin{tabular}{|l|l||l|l|l|l|}
-\hline
-\multicolumn{2}{|l||}{\multirow{2}{*}{}}                   & \multicolumn{4}{|c|}{Number of CPU cores}  \\
+\begin{tabular}{|l|l|l|l|l|l|}
+\hline
+\multicolumn{2}{|l|}{\multirow{2}{*}{}}                   & \multicolumn{4}{c|}{Number of CPU cores}  \\
 \cline{3-6}
-\multicolumn{2}{|l||}{}                                               & \bf{16} & \bf{32}      & \bf{64}      & \bf{128}  \\
+\multicolumn{2}{|l|}{}                                      & \bf{16}   & \bf{32}  & \bf{64}    & \bf{128} \\
 \hline

### 2016-09-26 doc/publications/plos16/plos16.tex

--- 2016-09-22 doc/publications/plos16/plos16.tex
@@
   trees on 128 CPU cores with varying numbers of variables and samples.}}
-\begin{tabular}{|l|l||l|l|l|l|l|}
+\begin{tabular}{|l|l|l|l|l|l|l|}
 \hline
-\multicolumn{2}{|l||}{\multirow{2}{*}{}}               & \multicolumn{5}{c|}{Number of variables}  \\
+\multicolumn{2}{|l|}{\multirow{2}{*}{}}               & \multicolumn{5}{c|}{Number of variables}  \\
 \cline{3-7}
-\multicolumn{2}{|l||}{}                                & \bf{150,000} & \bf{500,000} & \bf{2,500,000}  & \bf{10,000,000} & \bf{50,000,000} \\
+\multicolumn{2}{|l|}{}                                & \bf{150,000} & \bf{500,000} & \bf{2,500,000}  & \bf{10,000,000} & \bf{50,000,000} \\
 \hline                                                            
@@
 \hline
-\multicolumn{2}{|l||}{\multirow{2}{*}{}}                   & \multicolumn{4}{|c|}{Number of CPU cores}  \\
+\multicolumn{2}{|l|}{\multirow{2}{*}{}}                   & \multicolumn{4}{c|}{Number of CPU cores}  \\
 \cline{3-6}
-\multicolumn{2}{|l||}{}                                               & \bf{16} & \bf{32}      & \bf{64}      & \bf{128}  \\
+\multicolumn{2}{|l|}{}                                      & \bf{16}   & \bf{32}  & \bf{64}    & \bf{128} \\
 \hline
@@
                                                             & \bf{0.80} & 2313.5 & 1125.7 & 569.1 & 364.6 \\
-
 \hline

### 2016-09-28 doc/publications/plos16/plos16.tex
Added more pictures to the scalability seciton

@@
 
-%nvars 150000  500000  2500000 10000000  50000000
-%nsamples          
-%1000  21.6  26.7  59.8  133.5 668.8
-%5000  45.7  71.9  186.2 729.3 2887.4
-%10000 70.2  120.2 397.3 2237.8  5747.8
-
-\begin{table}[!ht]
-\centering
-\caption{
-{\bf Time in seconds to build random forest of 100
-  trees on 128 CPU cores with varying numbers of variables and samples.}}
-\begin{tabular}{|l|l|l|l|l|l|l|}
-\hline
-\multicolumn{2}{|l|}{\multirow{2}{*}{}}               & \multicolumn{5}{c|}{Number of variables}  \\
-\cline{3-7}
-\multicolumn{2}{|l|}{}                                & \bf{150,000} & \bf{500,000} & \bf{2,500,000}  & \bf{10,000,000} & \bf{50,000,000} \\
-\hline                                                            
-\multirow{4}{*}{Number of samples}                          & \bf{1000} &  21.6 & 26.7 & 59.8 & 133.5 & 668.8 \\
-                                                            & \bf{5000} & 45.7 & 71.9 & 186.2 & 729.3 & 2887.4 \\
-                                                            & \bf{10000} & 70.2 & 120.2 & 397.3 & 2237.8 & 5747.8 \\
-\hline
-\end{tabular}
-\begin{flushleft} 
-\end{flushleft}
-\label{table10}
-\end{table}
-
-\begin{figure}[tbhp]
-\begin{tabular}{ll}
-a)& b)\\
-\includegraphics[totalheight=6cm]{./figs/nsamples_nvars.png} & 
-\includegraphics[totalheight=6cm]{./figs/nvars_nsamples.png} \\
-\end{tabular}
-\captionof{figure}{{\bf Scalablity of the Wide Random Forest on synthetic datasets with varying number of samples and variables.} 
-   {\bf a)} Time in seconds to build 100 trees for differnet mtry fractions. 
-   {\bf b)} Number of trees build per hour when using a growing number of CPU cores.}
-\label{figure:synth}
-\end{figure}
-
 \begin{table}[!ht]
@@
    {\bf b)} Number of trees build per hour when using a growing number of CPU cores.}
+\label{figure:synth}
+\end{figure}
+
+%nvars 150000  500000  2500000 10000000  50000000
+%nsamples          
+%1000  21.6  26.7  59.8  133.5 668.8
+%5000  45.7  71.9  186.2 729.3 2887.4
+%10000 70.2  120.2 397.3 2237.8  5747.8
+
+\begin{table}[!ht]
+\centering
+\caption{
+{\bf Time in seconds to build random forest of 100
+  trees on 128 CPU cores with varying numbers of variables and samples.}}
+\begin{tabular}{|l|l|l|l|l|l|l|}
+\hline
+\multicolumn{2}{|l|}{\multirow{2}{*}{}}               & \multicolumn{5}{c|}{Number of variables}  \\
+\cline{3-7}
+\multicolumn{2}{|l|}{}                                & \bf{150,000} & \bf{500,000} & \bf{2,500,000}  & \bf{10,000,000} & \bf{50,000,000} \\
+\hline                                                            
+\multirow{4}{*}{Number of samples}                          & \bf{1000} &  21.6 & 26.7 & 59.8 & 133.5 & 668.8 \\
+                                                            & \bf{5000} & 45.7 & 71.9 & 186.2 & 729.3 & 2887.4 \\
+                                                            & \bf{10000} & 70.2 & 120.2 & 397.3 & 2237.8 & 5747.8 \\
+\hline
+\end{tabular}
+\begin{flushleft} 
+\end{flushleft}
+\label{table10}
+\end{table}
+
+\begin{figure}[tbhp]
+\begin{tabular}{ll}
+a)& b)\\
+\includegraphics[totalheight=6cm]{./figs/nvars_nsamples.png} & 
+\includegraphics[totalheight=6cm]{./figs/nsamples_nvars.png} \\
+\end{tabular}
+\captionof{figure}{{\bf Scalablity of the Wide Random Forest on synthetic datasets with varying number of samples and variables.} 
+   {\bf a)} Number of trees build per hour when growing number of variables. 
+   {\bf b)} Number of trees build per hour when growing number of samples.}
+\label{figure:synth}
+\end{figure}
+
+\begin{figure}[tbhp]
+\begin{tabular}{ll}
+a)& b)\\
+\includegraphics[totalheight=6cm]{./figs/var_oder.png} & 
+\includegraphics[totalheight=6cm]{./figs/var_oob.png} \\
+\end{tabular}
+\captionof{figure}{{\bf Variable importance (Mean Decrease Gini) and OOB (out of bag) classification error estimate for random forest with 100 trees trained on the synthetic dataset of 2.5M features and 5K samples.}
+   {\bf a)} Variable importance of top 20 variabels normalized to the sum of the linear model coefficients. {\bf Others} represents aggregated importance of variables not present in the linear model.
+   {\bf b)} OOB (out of bag) classification error estimate as funtion of mtry fraction.}
 \label{figure:synth}

### 2016-10-18 doc/publications/plos16/plos16.tex
added Piotr's results

@@
 \subsubsection{Synthetic data} 
+Each dataset consists of n samples and m variables where n << m, and values for each variable are ordinal factors with three levels represented as numbers 0, 1, 2 (which correspond to the ?typical?
+encoding of genomics variation) randomly generated from uniform distribution with equal probabilities.
+The dichotomous response is generated from a two level Bernoulli trail with p (probability of success) calculated by applying inverse logit transformation to the continuous linear response (y) of five variables:
+
+$$y = a_10* v_10 +  a_100*v_100 + a_1000*v_1000  +  a_10000*v_10000 + a_10000*v_100000  { where a_10 = 0.8, a_100 = 0.9, a_1000 = 1.1, a_10000 = 1.0, a_10000=1.15}$$
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Section 1:  Feature selection 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{CursedForest performance and scalability}
+% Section 1:  Feature Selection and classification on synthetic data
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{CursedForest can evaluate variable importance for millions of features  }\label{synthetic}
 %In this section we demonstrate that \cursedforest 's ability to rapidly grow large number of trees, \texttt{ntree}, in
@@
 %run, the ranking of the variables will stabilise with growing \texttt{ntree}.
-
-%To investigate this hypothesis for data sets with large number of feature variables we generated a synthetic dataset ...
-
-\begin{table}[!ht]
-\centering
-\caption{
-{\bf Time in seconds to build random forest of 100
-  trees with specified mtry fraction using given number of CPU cores.}}
-\begin{tabular}{|l|l||l|l|l|l|}
-\hline
-\multicolumn{2}{|l|}{\multirow{2}{*}{}}                   & \multicolumn{4}{c|}{Number of CPU cores}  \\
-\cline{3-6}
-\multicolumn{2}{|l|}{}                                      & \bf{16}   & \bf{32}  & \bf{64}    & \bf{128} \\
-\hline
-\multirow{4}{*}{mtry (fraction)}                            & \bf{0.10} & 708.3  & 340.0  & 186.0 & 134.5 \\
-                                                            & \bf{0.25} & 1134.9 & 535.0  & 270.8 & 179.6 \\
-                                                            & \bf{0.50} & 1724.1 & 856.6  & 428.7 & 243.8 \\
-                                                            & \bf{0.80} & 2313.5 & 1125.7 & 569.1 & 364.6 \\
-\hline
-\end{tabular}
-\begin{flushleft} 
-\end{flushleft}
-\label{table10}
-\end{table}
-
-\begin{figure}[tbhp]
-\begin{tabular}{ll}
-a)& b)\\
-\includegraphics[totalheight=6cm]{./figs/mtry_cpu.png} & 
-\includegraphics[totalheight=6cm]{./figs/cpu_mtry_trees_per_hour.png} \\
-\end{tabular}
-\captionof{figure}{{\bf Scalablity of the Wide Random Forest on the synthetic dataset of 2.5M features and 5K samples.} 
-   {\bf a)} Time in seconds to build 100 trees for differnet mtry fractions. 
-   {\bf b)} Number of trees build per hour when using a growing number of CPU cores.}
-\label{figure:synth}
-\end{figure}
-
-%nvars 150000  500000  2500000 10000000  50000000
-%nsamples          
-%1000  21.6  26.7  59.8  133.5 668.8
-%5000  45.7  71.9  186.2 729.3 2887.4
-%10000 70.2  120.2 397.3 2237.8  5747.8
-
-\begin{table}[!ht]
-\centering
-\caption{
-{\bf Time in seconds to build random forest of 100
-  trees on 128 CPU cores with varying numbers of variables and samples.}}
-\begin{tabular}{|l|l|l|l|l|l|l|}
-\hline
-\multicolumn{2}{|l|}{\multirow{2}{*}{}}               & \multicolumn{5}{c|}{Number of variables}  \\
-\cline{3-7}
-\multicolumn{2}{|l|}{}                                & \bf{150,000} & \bf{500,000} & \bf{2,500,000}  & \bf{10,000,000} & \bf{50,000,000} \\
-\hline                                                            
-\multirow{4}{*}{Number of samples}                          & \bf{1000} &  21.6 & 26.7 & 59.8 & 133.5 & 668.8 \\
-                                                            & \bf{5000} & 45.7 & 71.9 & 186.2 & 729.3 & 2887.4 \\
-                                                            & \bf{10000} & 70.2 & 120.2 & 397.3 & 2237.8 & 5747.8 \\
-\hline
-\end{tabular}
-\begin{flushleft} 
-\end{flushleft}
-\label{table10}
-\end{table}
-
-\begin{figure}[tbhp]
-\begin{tabular}{ll}
-a)& b)\\
-\includegraphics[totalheight=6cm]{./figs/nvars_nsamples.png} & 
-\includegraphics[totalheight=6cm]{./figs/nsamples_nvars.png} \\
-\end{tabular}
-\captionof{figure}{{\bf Scalablity of the Wide Random Forest on synthetic datasets with varying number of samples and variables.} 
-   {\bf a)} Number of trees build per hour when growing number of variables. 
-   {\bf b)} Number of trees build per hour when growing number of samples.}
-\label{figure:synth}
-\end{figure}
+In this section we demonstrate CursedForrest two main features, namely variable importance analysis to select the most associated set of features amongst millions of dimensions as well as its ability to accurately classify the labels of high-dimensional samples.
+%We therefore generate a synthetic dataset with varying numbers of variables (features) and samples, as discussed in the method section. 
+We therefore generate a synthetic dataset with 2.5 million variables (features) and 5000 samples, as discussed in the method section. 
+Briefly we generate features by randomly sampling from a three value vector (1,2,3) with equal probabilities, which mimics genomic mutation data and corresponds to homozygote reference, heterozygote or homozygote alternative.
+We then generate the response variable as a linear combination of 5 weighted features, modulating a complex trait of multiple interacting features.
+
+After fitting the random forest model and performing the variable importance analysis we hence expect to retrieve these 5 feature in order of the weight that was assigned to them. 
+We are running Apache Spark 1.6.1 on a YARN cluster with 12 worker nodes each with 16 Intel Xeon E5-2660@2.20GHz CPU cores and 128 GB or RAM. 
+We captured the OOB (out of bag) classification error and the 20 most important variables.
+
+Figure~\ref{figure:synth}a illustrates that increasing the number of predictors sampled for splitting at each node (mtry) increases the accuracy with which features are ranked. 
+From $mtry=0.5$ the sort order of the features reflects the weight originally assigned to them (dotted line).
+Further increasing $mtry$ improves the predicted importance measure, which starts to converge to the weight originally assigned to them (coefficients).
+It also increases the difference between the associated 5 variable and the remaining variables in the top 20 list. 
+By $mtry=0.8$ the variable with the smallest weight has a predicted importance measure that is 5 times larger than the aggregated importance of the 15 remaining variables from the list.
+
+Figure~\ref{figure:synth}b illustrates that increasing $mtry$ also improves the OOB (out of bag) classification error. 
+%Piotr can you please explain the remaining error on the synthetic data
+By $mtry=0.8$ the error is converging to X, which is due to Y.
+
+From the synthetic data we can conclude that CursedForest is able to perform accurate feature selection as well as accurate classification for large numbers of samples with large feature vectors.
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Section 2: Classification selection
+% Section 2: Feature Selection on real data 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{CursedForest selects previously demonstrated cancer variants }
+%Feature selection can be useful for determining variables the contribute to the label or for perhaps selecting the n most important
+%variables for further analysis. 
+As established in the pervious section, CursedForest allows the user to retrieve a subsection of features, ranked by their importance to identify the variables predictive of the label or selecting the n most important variables for further analysis.
+We demonstrate this feature on The Cancer Genome Atlas (TCGA) BRCA dataset. 
+We build this model on VCF files retrieved from the TCGA data portal. 
+For the label, we use the cell type (i.e. tumor or normal).
+
+In Table~\ref{table3}, we display the top 10 features, ranked by their Gini index in the Random Forest model.
+
+\begin{table}[!ht]
+%\begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
+\centering
+\caption{
+{\bf Features from the TCGA BRCA dataset ranked by their importance.}}
+\begin{tabular}{|l|l|l|l|l|l|l|}
+\hline
+{\bf Variant} & {\bf Gene} & {\bf Importance}\\ \thickhline
+x:xxxx & cell2 row 1 & cell3 row 1\\ \hline
+x:xxxx & cell2 row 2 & cell3 row 2\\ \hline
+x:xxxx & cell2 row 3 & cell3 row 3\\ \hline
+\end{tabular}
+\begin{flushleft} Here the importance is the gini coefficient when building a model using the cell type as the label.
+\end{flushleft}
+\label{table3}
+%\end{adjustwidth}
+\end{table}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Section 3: Classification selection
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsection{CursedForest outperforms existing methods for classification problems}
+As established in Section~\ref{synthetic}, CursedForest 
 In this section we compare \cursedforest 's performance against other published methods for predicting ethnicity on the
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Section 3: Feature Selection 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{CursedForest selects previously demonstrated cancer variants }
-Feature selection can be useful for determining variables the contribute to the label or for perhaps selecting the n most important
-variables for further analysis. Our Random Forest implementation allows the user to retrieve a subsection of features, ranked
-by their importance. We demonstrate this feature on The Cancer Genome Atlas (TCGA) BRCA dataset. We build this model on
-VCF files retrieved from the TCGA data portal. For the label, we use the cell type (i.e. tumor or normal).
-
-In Table~\ref{table3}, we display the top 10 features, ranked by their Gini index in the Random Forest model.
+% Section 4: Sensitivity 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Scalability analysis}
+
+In this section we explore the performance of CursedForest by testing its ability to scale to different sizes of data and computational resources.
+
+In order to asses these characteristics we ran CursedForst classification on synthetics datasets with varying numbers of variables (features) and samples, similar to the dataset used in to evaluate ranger [REF: ranger], allocating varying number of CPU cores to the CursedForest and also varying computational complexity of random forest by using range of mtry values.
+
+We investigate the different synthetic datasets generated for section 5.1.
+and measured the time of building a random forest of 100 trees and the results reported below are averages of 5 runs.
+To improve consistency of measurements all the cases were executed with the same random seed.
+
+Firstly we look at CurstedForest horizontal scalability for a medium size dataset of 2.5 million variables and 5000 samples, by varying mtry fraction and the number of CPU cores allocated to the execution.
+The raw results are presented in the table [REF table1] and visualized in figure [REF: Fig 1] below
+Regardless of the number of cores used, CursedForerst displays approximately  linear dependency between the execution time and the mtry parameter (presented here as a fraction of the number of variables) [Fig 1 a)
+
+CursedForest scales almost linearly with the number of CPU cores for medium values of mtry fraction but for both lower and higher values the performance degrades slightly (Fig 1 b)  In the latter case the likely cause is communication overhead (with lower mtry values the proportion of time for parallelizable computation to the time for internode communication is lower) while in the latter case its most likely caused by reaching the clusters computational capacity.
+
+Next we investigate CursedForest scalability with regards to the size of data, by varying the number of variables and sample for a fixed mtry fraction of 0.25 and execution of 128 CPU cores.
+The raw results are presented in the table [REF] and visualized in figure [ REF: Fig 2] below (please note log scale on the axes and the values on y axes are expresses as trees per hour).
+
+Generally, the number of trees per hour decreases with increased number of variables and samples sizes.
+Some irregularities in the graph can be attributed to computation vs communication tradeoff.
+It is also worth noting that keeping mtry fraction constants results with higher mtry values with growing number of variables, and this is what drives the performance down rather that the increase of dataset size itself.
+
+To conclude CursedForst is capable of processing datasets with 50M variables and 10K samples, which is the order of magnitude of the entire genome variants expected in such dataset, for which it can build approximately 60 trees per hour.
 
 \begin{table}[!ht]
-%\begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
 \centering
 \caption{
-{\bf Features from the TCGA BRCA dataset ranked by their importance.}}
+{\bf Time in seconds to build random forest of 100
+  trees with specified mtry fraction using given number of CPU cores.}}
+\begin{tabular}{|l|l||l|l|l|l|}
+\hline
+\multicolumn{2}{|l|}{\multirow{2}{*}{}}                   & \multicolumn{4}{c|}{Number of CPU cores}  \\
+\cline{3-6}
+\multicolumn{2}{|l|}{}                                      & \bf{16}   & \bf{32}  & \bf{64}    & \bf{128} \\
+\hline
+\multirow{4}{*}{mtry (fraction)}                            & \bf{0.10} & 708.3  & 340.0  & 186.0 & 134.5 \\
+                                                            & \bf{0.25} & 1134.9 & 535.0  & 270.8 & 179.6 \\
+                                                            & \bf{0.50} & 1724.1 & 856.6  & 428.7 & 243.8 \\
+                                                            & \bf{0.80} & 2313.5 & 1125.7 & 569.1 & 364.6 \\
+\hline
+\end{tabular}
+\begin{flushleft} 
+\end{flushleft}
+\label{table10}
+\end{table}
+
+\begin{figure}[tbhp]
+\begin{tabular}{ll}
+a)& b)\\
+\includegraphics[totalheight=6cm]{./figs/mtry_cpu.png} & 
+\includegraphics[totalheight=6cm]{./figs/cpu_mtry_trees_per_hour.png} \\
+\end{tabular}
+\captionof{figure}{{\bf Scalablity of the Wide Random Forest on the synthetic dataset of 2.5M features and 5K samples.} 
+   {\bf a)} Time in seconds to build 100 trees for differnet mtry fractions. 
+   {\bf b)} Number of trees build per hour when using a growing number of CPU cores.}
+\label{figure:synth}
+\end{figure}
+
+%nvars 150000  500000  2500000 10000000  50000000
+%nsamples          
+%1000  21.6  26.7  59.8  133.5 668.8
+%5000  45.7  71.9  186.2 729.3 2887.4
+%10000 70.2  120.2 397.3 2237.8  5747.8
+
+\begin{table}[!ht]
+\centering
+\caption{
+{\bf Time in seconds to build random forest of 100
+  trees on 128 CPU cores with varying numbers of variables and samples.}}
 \begin{tabular}{|l|l|l|l|l|l|l|}
 \hline
-{\bf Variant} & {\bf Gene} & {\bf Importance}\\ \thickhline
-x:xxxx & cell2 row 1 & cell3 row 1\\ \hline
-x:xxxx & cell2 row 2 & cell3 row 2\\ \hline
-x:xxxx & cell2 row 3 & cell3 row 3\\ \hline
+\multicolumn{2}{|l|}{\multirow{2}{*}{}}               & \multicolumn{5}{c|}{Number of variables}  \\
+\cline{3-7}
+\multicolumn{2}{|l|}{}                                & \bf{150,000} & \bf{500,000} & \bf{2,500,000}  & \bf{10,000,000} & \bf{50,000,000} \\
+\hline                                                            
+\multirow{4}{*}{Number of samples}                          & \bf{1000} &  21.6 & 26.7 & 59.8 & 133.5 & 668.8 \\
+                                                            & \bf{5000} & 45.7 & 71.9 & 186.2 & 729.3 & 2887.4 \\
+                                                            & \bf{10000} & 70.2 & 120.2 & 397.3 & 2237.8 & 5747.8 \\
+\hline
 \end{tabular}
-\begin{flushleft} Here the importance is the gini coefficient when building a model using the cell type as the label.
+\begin{flushleft} 
 \end{flushleft}
-\label{table3}
-%\end{adjustwidth}
+\label{table10}
 \end{table}
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Section 3: Sensitivity 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Sensitivity of parameter choice}
+\begin{figure}[tbhp]
+\begin{tabular}{ll}
+a)& b)\\
+\includegraphics[totalheight=6cm]{./figs/nvars_nsamples.png} & 
+\includegraphics[totalheight=6cm]{./figs/nsamples_nvars.png} \\
+\end{tabular}
+\captionof{figure}{{\bf Scalablity of the Wide Random Forest on synthetic datasets with varying number of samples and variables.} 
+   {\bf a)} Number of trees build per hour when growing number of variables. 
+   {\bf b)} Number of trees build per hour when growing number of samples.}
+\label{figure:synth}
+\end{figure}
 

### 2016-10-24 doc/publications/plos16/plos16.tex
added scaling content
some changes

@@
 than $2^{31}$), and the {\sc randomForest} implementation in R lending itself to paralelization to grow multiple trees
-simultaneously~\cite{Liaw2002} and them combine them.  More suitable implementations for ``wide'' genomic data have been
+simultaneously~\cite{Liaw.and.Weiner.2002} and them combine them.  More suitable implementations for ``wide'' genomic data have been
 developed in other programming languages.  {\sc CloudForest}~\cite{Bressler2015} written in Go, achieves fast running

### 2016-10-24 doc/publications/plos16/plos16.tex
some changes to plos16.tex

@@
 \section{Introduction}
+\marginpar{change parallelisation to parallelization}
 
@@
 more information about these events.  The challenge of big and ``wide'' data is especially pronounced in the health
-space where whole genome sequence (WGS) technology enabled researchers to interrogate all 3 billion base pairs of the
+space where whole genome sequence (WGS) technology enables researchers to interrogate all 3 billion base pairs of the
 human genome.
@@
 association studies (GWAs).  These analysis are typically performed on only the most frequently differing base-pairs
-(SNPs) by applying linear or logistic regression analysis to each SNPs separately~\cite{CCC2007}.  It has since been
+(SNPs) by applying linear or logistic regression analysis to each SNP separately~\cite{CCC2007}.  It has since been
 demonstrated that not taking joint interactions between SNPs into account when filtering for potential disease mutations
-is inadequate~\cite{Manolio2009}.  Complex Trait analysis has demonstrated that in order to explain the overall
+is inadequate~\cite{Manolio2009}.  
+\marginpar{is Complex Trait a thing? why capitals?}
+Complex Trait analysis has demonstrated that in order to explain the overall
 heritability a model needs to take the joint contribution of all variants into account~\cite{Yang2011}.
@@
 propensity to overfit datasets with more features $p$ than samples $n$ (a consequence of the ``curse of
-dimensionality''~\cite{Bauer2014, bellman1961adaptive}), decision trees are resistant to overfitting. Random Forests are
-not immune to overfitting as shown by ~\cite{Segal.2004} who gives examples of overfitting in the case where:
-\begin{itemize}
-\item a single tree is the most appropriate model; and
-\item the variables are hightly correlated so that the boostrap resampling does not give rise to uncorrelated trees.
-\end{itemize}
-However, keeping these difficult cases in mind, RF have a strong claim to being resistant to overfitting.
+dimensionality''~\cite{Bauer2014, bellman1961adaptive}), decision trees are resistant to overfitting. 
 Secondly, RF are also very easy to parallelise. As the forest is a sum of decision trees, it it possible to grow
-separate trees on each processor and combine the trees.
+separate trees on different processors and combine the trees.
 
@@
 scale beyond compute-node boundaries and hence enable more sophisticated parallelisation strategies.  In the case of
-Random forests this means that the the comutations for each node of a tree can be handed off to separate processors.
+random forests this means that the computations for each node of a tree can be handed off to separate processors.
 
@@
 ``big'' (many samples) and ``wide'' (many features) data sets. \cursedforest\ is capable of parallelising the split for
-each node in a tree thereby being able to handle millions of features, such as in the 1000 Genome data, which consists
+each node in a tree thereby handling millions of features, such as in the 1000 Genome data, which consists
 of approximately 2500 samples with up to 80 million variants~\cite{1KG2012}.  By also utilising Spark to read in and
 manipulate the standard genomic variant format (VCF) directly, \cursedforest\ outperforms existing tools on even small
-datasets where multi-threading is considered more performant. \marginpar{is performant a word?}
+datasets where multi-threading generally performs well. 
+\marginpar{``performant'' -- a person who performs, an actor or celebrant }
  \marginpar{Piotr this is your 20 trees has the important
@@
 
-=======
 %Hadoop/Spark overcomes this limitation by enabling programs to scale beyond compute-node boundaries and virtually utilise infinite number of CPUs.   
@@
 respect to the dimensionality of the data, building an RF model on whole-genome data from the 1000 Genomes Project. Here
-we report on the accuracy (in terms of the ARI) and the error.  Finally, given the role different parameter values can
-play in model construction, we explore the effect tuning these parameters can have on the prediction accuracy of a
+we report on the accuracy (in terms of the ARI)
+\marginpar{what is ARI?}
+ and the error.  Finally, given the role different parameter values can
+play in model construction, we explore the effect tuning these parameters can have on the prediction accuracy of the
 model.
@@
 \begin{itemize}
-\item random forests are resistant against overfitting. However it is not true that they will not overfit \cite[]{Segal.2004};
+\item random forests are resistant against overfitting. However it is not true that they will not overfit,
+  ~\cite{Segal.2004} gives examples of overfitting in the case where:
+  \begin{itemize}
+  \item a single tree is the most appropriate model; and
+  \item the variables are hightly correlated so that the boostrap resampling does not give rise to uncorrelated trees.
+  \end{itemize}
+However, keeping these difficult cases in mind, RF have a strong claim to being resistant to overfitting.
+
 \item \cite{Strobl.et.al.2007} demonstrates that the RF algorithm is subject to a bias in variable selection via two mechanisms:
@@
 \item an ``information gained'' criteria for splitting. Let $f_q$ be the fraction of items labeled with value $q$ where $q= 1,
-  \ldots, Q$ at a node. The  entropy of the node is $H = - \sum^{m}_{i=1} f_i \log^{}_2 f_i$
+  \ldots, Q$ at a node. The  entropy of the node is $H = - \sum^{m}_{i=1} f_i \log^{}_2 f_i.$
 The expected information gain is the change in entropy $H$ from a prior state $T$ to a state that takes some
@@
 \subsubsection{Synthetic data}
-Each dataset consists of n samples and m variables where n << m, and values for each variable are ordinal factors with three levels represented as numbers 0, 1, 2 (which correspond to the ?typical?
-encoding of genomics variation) randomly generated from uniform distribution with equal probabilities.
-The dichotomous response is generated from a two level Bernoulli trail with p (probability of success) calculated by applying inverse logit transformation to the continuous linear response (y) of five variables:
-
-$$y = a_10* v_10 +  a_100*v_100 + a_1000*v_1000  +  a_10000*v_10000 + a_10000*v_100000  { where a_10 = 0.8, a_100 = 0.9, a_1000 = 1.1, a_10000 = 1.0, a_10000=1.15}$$
+
+Each dataset consists of $n$ samples and $m$ variables where $n << m$, and values for each variable are ordinal factors with
+three levels represented as numbers $\{0, 1, 2\}$ (which correspond to an additive effect encoding of genomics variation) randomly
+generated from uniform distribution with equal probabilities.  The dichotomous response is generated from a two level
+Bernoulli trail with $p$ (probability of success) calculated by applying inverse logit transformation to the continuous
+linear response ($y$) of five variables:
+$$y = a_{10}* v_{10} + a_{100}*v_{100} + a_{1000}*v_{1000} + a_{10000}*v_{10000} + a_{10000}*v_{100000}$$
+where $a_{10} = 0.8, a_{100} = 0.9, a_{1000} = 1.1, a_{10000} = 1.0, a_{10000}=1.15$.
 
@@
 %run, the ranking of the variables will stabilise with growing \texttt{ntree}.
-In this section we demonstrate CursedForrest two main features, namely variable importance analysis to select the most associated set of features amongst millions of dimensions as well as its ability to accurately classify the labels of high-dimensional samples.
+In this section we demonstrate CursedForrest two main features, namely variable importance analysis to select the most
+associated set of features amongst millions of dimensions as well as its ability to accurately classify the labels of
+high-dimensional samples.
 %We therefore generate a synthetic dataset with varying numbers of variables (features) and samples, as discussed in the method section. 
-We therefore generate a synthetic dataset with 2.5 million variables (features) and 5000 samples, as discussed in the method section. 
-Briefly we generate features by randomly sampling from a three value vector (1,2,3) with equal probabilities, which mimics genomic mutation data and corresponds to homozygote reference, heterozygote or homozygote alternative.
-We then generate the response variable as a linear combination of 5 weighted features, modulating a complex trait of multiple interacting features.
-
-After fitting the random forest model and performing the variable importance analysis we hence expect to retrieve these 5 feature in order of the weight that was assigned to them. 
-We are running Apache Spark 1.6.1 on a YARN cluster with 12 worker nodes each with 16 Intel Xeon E5-2660@2.20GHz CPU cores and 128 GB or RAM. 
-We captured the OOB (out of bag) classification error and the 20 most important variables.
-
-Figure~\ref{figure:synth}a illustrates that increasing the number of predictors sampled for splitting at each node (mtry) increases the accuracy with which features are ranked. 
-From $mtry=0.5$ the sort order of the features reflects the weight originally assigned to them (dotted line).
-Further increasing $mtry$ improves the predicted importance measure, which starts to converge to the weight originally assigned to them (coefficients).
-It also increases the difference between the associated 5 variable and the remaining variables in the top 20 list. 
-By $mtry=0.8$ the variable with the smallest weight has a predicted importance measure that is 5 times larger than the aggregated importance of the 15 remaining variables from the list.
+We therefore generate a synthetic dataset with 2.5 million variables (features) and 5000 samples, as discussed in the
+method section.  Briefly we generate features by randomly sampling from a three value vector (1,2,3) with equal
+probabilities, which mimics genomic mutation data and corresponds to homozygote reference, heterozygote or homozygote
+alternative.  We then generate the response variable as a linear combination of 5 weighted features, modulating a
+complex trait of multiple interacting features.
+
+After fitting the random forest model and performing the variable importance analysis we hence expect to retrieve these
+5 feature in order of the weight that was assigned to them.  We are running Apache Spark 1.6.1 on a YARN cluster with 12
+worker nodes each with 16 Intel Xeon E5-2660@2.20GHz CPU cores and 128 GB or RAM.  We captured the OOB (out of bag)
+classification error and the 20 most important variables.
+
+Figure~\ref{figure:synth}a illustrates that increasing the number of predictors sampled for splitting at each node
+(mtry) increases the accuracy with which features are ranked.  From $mtry=0.5$ the sort order of the features reflects
+the weight originally assigned to them (dotted line). Further increasing $mtry$ improves the predicted importance
+measure, which starts to converge to the weight originally assigned to them (coefficients).  It also increases the
+difference between the associated 5 variable and the remaining variables in the top 20 list.  By $mtry=0.8$ the variable
+with the smallest weight has a predicted importance measure that is 5 times larger than the aggregated importance of the
+15 remaining variables from the list.
 
@@
 
-From the synthetic data we can conclude that CursedForest is able to perform accurate feature selection as well as accurate classification for large numbers of samples with large feature vectors.
+From the synthetic data we can conclude that CursedForest is able to perform accurate feature selection as well as
+accurate classification for large numbers of samples with large feature vectors.
 
@@
 \end{tabular}
-\captionof{figure}{{\bf Variable importance (Mean Decrease Gini) and OOB (out of bag) classification error estimate for random forest with 100 trees trained on the synthetic dataset of 2.5M features and 5K samples.}
-   {\bf a)} Variable importance of top 20 variabels normalized to the sum of the linear model coefficients. {\bf Others} represents aggregated importance of variables not present in the linear model.
-   {\bf b)} OOB (out of bag) classification error estimate as funtion of mtry fraction.}
+\captionof{figure}{{\bf Variable importance (Mean Decrease Gini) and OOB (out of bag) classification error estimate for
+    random forest with 100 trees trained on the synthetic dataset of 2.5M features and 5K samples.}  {\bf a)} Variable
+  importance of top 20 variabels normalized to the sum of the linear model coefficients. {\bf Others} represents
+  aggregated importance of variables not present in the linear model.  {\bf b)} OOB (out of bag) classification error
+  estimate as funtion of mtry fraction.}
 \label{figure:synth}
@@
 %variables for further analysis. 
-As established in the pervious section, CursedForest allows the user to retrieve a subsection of features, ranked by their importance to identify the variables predictive of the label or selecting the n most important variables for further analysis.
-We demonstrate this feature on The Cancer Genome Atlas (TCGA) BRCA dataset. 
-We build this model on VCF files retrieved from the TCGA data portal. 
-For the label, we use the cell type (i.e. tumor or normal).
+As established in the pervious section, CursedForest allows the user to retrieve a subsection of features, ranked by
+their importance to identify the variables predictive of the label or selecting the $n$ most important variables for
+further analysis.  We demonstrate this feature on The Cancer Genome Atlas (TCGA) BRCA dataset.  We build this model on
+VCF files retrieved from the TCGA data portal.  For the label, we use the cell type (i.e. tumor or normal).
 

### 2016-10-25 doc/publications/plos16/plos16.tex
General fixes

@@
 
-\documentclass[10pt,a4paper]{article}  %\documentclass[10pt,letterpaper]{article}
+\documentclass[10pt,letterpaper]{article}
 \usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}
@@
 \newcolumntype{+}{!{\vrule width 2pt}}
-
-\usepackage{footmisc}
 
@@
 \newcommand{\cursedforest}{{\sc CursedForest}}
-\newcommand{\lorem}{{\bf LOREM}}
-\newcommand{\ipsum}{{\bf IPSUM}}
 
@@
 
+%% Additional packages
+\usepackage{footmisc}
 \usepackage{multirow}
@@
 % Title must be 250 characters or less.
-% Please capitalize all terms in the title except conjunctions, prepositions, and articles.
 \begin{flushleft}
 {\Large
-\textbf\newline{Cursed Forest - A random forest implementation for ``big'' and ``wide'' data} % Please use "title case" (capitalize all terms in the title except conjunctions, prepositions, and articles).
+\textbf\newline{\cursedforest - A random forest implementation for ``big'' and ``wide'' data} % Please use "title case" (capitalize all terms in the title except conjunctions, prepositions, and articles).
 }
@@
 \bigskip
-\bf{1} Health \& Biosecurity, CSIRO, Sydney, NSW, Australia
+\textbf{1} Health \& Biosecurity, CSIRO, Sydney, NSW, Australia
 \\
-\bf{2} Data61, CSIRO, Brisbane, QLD, Australia
+\textbf{2} Data61, CSIRO, Brisbane, QLD, Australia
 \\
-\bf{3} Health \& Biosecurity, CSIRO, Brisbane, QLD, Australia
+\textbf{3} Health \& Biosecurity, CSIRO, Brisbane, QLD, Australia
 \\
-\bf{4} Oceans \& Atmosphere, CSIRO, Brisbane, QLD, Australia
+\textbf{4} Oceans \& Atmosphere, CSIRO, Brisbane, QLD, Australia
 \\
-\bf{5} Data61, CSIRO, Sydney, NSW, Australia
+\textbf{5} Data61, CSIRO, Sydney, NSW, Australia
 \\
@@
 \end{flushleft}
-% Please keep the abstract below 300 words
 \tableofcontents
@@
   \item check Piotr's simulation  (done)
-  \item run sparkML or Cursed forest on the  simulation data 
+  \item run sparkML or CursedForest on the  simulation data 
   \item we extend the simulations to many more variables 
@@
 
-\section{Abstract}
+% Please keep the abstract below 300 words
+\section*{Abstract}
 The abstract.
@@
 ``big'' (many samples) and ``wide'' (many features) data sets. \cursedforest\ is capable of parallelising the split for
-each node in a tree thereby handling millions of features, such as in the 1000 Genome data, which consists
+each node in a tree thereby handling millions of features, such as in the 1000 Genomes Project data, which consists
 of approximately 2500 samples with up to 80 million variants~\cite{1KG2012}.  By also utilising Spark to read in and
@@
 %% We therefore developed \cursedforest\, a Spark-based random forest algorithm able to process big and ``wide'' data.
-%% % Unlike the standard Apache implementation, CursedForest is able to easily process data with millions of dimensions without
+%% % Unlike the standard Apache implementation, \cursedforest\ is able to easily process data with millions of dimensions without
 %% % excessive resource requirements on the driver or worker nodes.  DENIS Piotr this is your 20 trees has the important feature in
@@
 %Unlike
-%the standard Apache implementation, CursedForest is able to easily process data with millions of dimensions without
+%the standard Apache implementation, \cursedforest\ is able to easily process data with millions of dimensions without
 %excessive resource requirements on the driver or worker nodes.
@@
 In the first section we explore some of the
-properties of random forests, demonstrating that any potential flaws in CursedForest are an inherent property of Random
-Forests. Here we provide demonstrations using synthetic data.  We then demonstrate the scalability of CursedForest in
+properties of random forests, demonstrating that any potential flaws in \cursedforest\ are an inherent property of Random
+Forests. Here we provide demonstrations using synthetic data.  We then demonstrate the scalability of \cursedforest\ in
 respect to the dimensionality of the data, building an RF model on whole-genome data from the 1000 Genomes Project. Here
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Cursed Forest.}
-% DENIS: Aidan and Piotr can you please explain here how Cursed forest is implemented
+\subsection{\cursedforest.}
+% DENIS: Aidan and Piotr can you please explain here how CursedForest is implemented
 Traditionally, VariantSpark made use of Spark's machine learning algorithms ``Spark ML''. These algorithms require data
@@
 
-On the other hand, CursedForest (CF) is specifically designed to handle wide ``cursed'' data. It avoids the relation
+On the other hand, \cursedforest\ (CF) is specifically designed to handle wide ``cursed'' data. It avoids the relation
 between memory and dimensionality by avoiding calculations that rely on entire feature vectors.
 
-Compared to Spark ML, CursedForest further parallelizes work down to individual features.  For example, for each node of
-a tree, CursedForest will distribute tasks that consist of single features (variants), for every individual.  Each of
+Compared to Spark ML, \cursedforest\ further parallelizes work down to individual features.  For example, for each node of
+a tree, \cursedforest\ will distribute tasks that consist of single features (variants), for every individual.  Each of
 these tasks will calculate the information gain for that specific feature.  Once these tasks have completed, the results
 are reduced to return the feature which gives the greatest information gain.  This process is then repeated until
-CursedForest has calculated the entire decicion tree.
-
-The current implementation of Cursed Forest uses:
+\cursedforest\ has calculated the entire decicion tree.
+
+The current implementation of \cursedforest\ uses:
 \begin{itemize}
@@
 %DENIS: Aidan can you please fill this in of where we got it from and how we process it for all the different tools to use
-\subsubsection{1000 Genomes}
+\subsubsection{1000 Genomes Project}
+
+We obtained the 1000 Genomes Project data as VCF files from their FTP site.
+
 %DENIS: Rob how was the synthetic data generated
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{CursedForest can evaluate variable importance for millions of features  }\label{synthetic}
-%In this section we demonstrate that \cursedforest 's ability to rapidly grow large number of trees, \texttt{ntree}, in
+\subsection{\cursedforest\ can evaluate variable importance for millions of features}\label{synthetic}
+%In this section we demonstrate that \cursedforest's ability to rapidly grow large number of trees, \texttt{ntree}, in
 %parallel results in a stable feature importance ranking.  This concept was originally proposed by Liaw and
@@
 %run, the ranking of the variables will stabilise with growing \texttt{ntree}.
-In this section we demonstrate CursedForrest two main features, namely variable importance analysis to select the most
+In this section we demonstrate \cursedforest\ two main features, namely variable importance analysis to select the most
 associated set of features amongst millions of dimensions as well as its ability to accurately classify the labels of
@@
 
-From the synthetic data we can conclude that CursedForest is able to perform accurate feature selection as well as
+From the synthetic data we can conclude that \cursedforest\ is able to perform accurate feature selection as well as
 accurate classification for large numbers of samples with large feature vectors.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{CursedForest selects previously demonstrated cancer variants }
+\subsection{\cursedforest\ selects previously demonstrated cancer variants }
 %Feature selection can be useful for determining variables the contribute to the label or for perhaps selecting the n most important
 %variables for further analysis. 
-As established in the pervious section, CursedForest allows the user to retrieve a subsection of features, ranked by
+As established in the pervious section, \cursedforest\ allows the user to retrieve a subsection of features, ranked by
 their importance to identify the variables predictive of the label or selecting the $n$ most important variables for
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{CursedForest outperforms existing methods for classification problems}
-As established in Section~\ref{synthetic}, CursedForest 
+\subsection{\cursedforest\ outperforms existing methods for classification problems}
+As established in Section~\ref{synthetic}, \cursedforest\ 
 In this section we compare \cursedforest 's performance against other published methods for predicting ethnicity on the
-1000 genomes dataset.
+1000 Genomes Project dataset.
 
@@
 \begin{minipage}{\textwidth}
-\begin{adjustwidth}{-1.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
+\begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
+\centering
 \caption{
@@
 \hline
-%\multicolumn{4}{|l|}{\bf Heading1} & \multicolumn{4}{|l|}{\bf Heading2}                               \\ \hline
 \bf{Genome \%}                      & \bf{Method} & \bf{Error (ARI)} & \bf{Runtime} & \bf{Memory/Node} \\
@@
 
-\multirow{3}{*}{phase1\_chr22: 1,092 x 490,036} & CursedForest  & 0.077 (XX) &                  &                                                                   \\
+\multirow{3}{*}{phase1\_chr22: 1,092 x 490,036} & \cursedforest\ & 0.077 (XX) &                  &                                                                   \\
                                                 & RF - Spark ML  &            &                  &                                                                   \\
                                                 & RF - R         & 0.05       & 24 min           & 18GB\footnote{using the \texttt{parallel} packages with 10 nodes} \\
-                                                & ranger -R     & 0.94       & 1 min 10 seconds & 16GB  (one processor)                                             \\
-                                                & ranger c++    & 0.94       & 7 mins           & --nthreads 20                                                     \\
+                                                & ranger - R      & 0.94       & 1 min 10 seconds & 16GB  (one processor)                                             \\
+                                                & ranger - c++     & 0.94       & 7 mins           & --nthreads 20                                                     \\
 
@@
                                                     & H2O           & 0.05       & 16 hours         & 600GB \footnote{java -Xmx600g -jar h2o.jar,   h2o.init(nthreads=20)} \\
-                                                    & ranger  -R    & 0.94       & 4 1/2 mins       & 21GB (one processor)                                                                \\
-\hline
-\multirow{3}{*}{phase3\_chr1: 2,504 x 6,450,364}    & CursedForest  & 0.063 (XX) & 1hr 58min        & 4GB                                                                                 \\
+                                                    & ranger - R    & 0.94       & 4 1/2 mins       & 21GB (one processor) \\
+\hline
+\multirow{3}{*}{phase3\_chr1: 2,504 x 6,450,364}    & \cursedforest\ & 0.063 (XX) & 1hr 58min        & 4GB \\
                                                     & RF - Spark ML  & 0.94       & 8hr 29min        & 8GB \\
                                                     & LR - Spark ML  &            & 2hr 6min         & 8GB \\
-                                                    & ranger  R     & 0.95       & 28 mins          & num.threads=10\footnote{don't
+                                                    & ranger - R      & 0.95       & 28 mins          & num.threads=10\footnote{don't
                                                                                                       know how to get the memory
@@
 \hline
-\multirow{3}{*}{phase3\_chr1-3: 2,504 x 19,328,051} & CursedForest  & 0.046 (XX) & 4hr 0min                                             & 4GB         \\
+\multirow{3}{*}{phase3\_chr1-3: 2,504 x 19,328,051} & \cursedforest\ & 0.046 (XX) & 4hr 0min                                             & 4GB \\
                                                     & RF - Spark ML  &            &                                                      & \\
                                                     & LR - Spark ML  &            & 14hr 48min                                           & 8GB \\
-                                                    & ranger  R     &  0.97      & 58 mins\footnote{ but took 6 hours to load the
+                                                    & ranger - R      &  0.97      & 58 mins\footnote{ but took 6 hours to load the
                                                                                    data}
@@
                                                                                                                                            }  \\
-& ranger   -R        &             0.973     &      03:58           &       pearcey --cpus-per-task=15 --mem 1600gb   906GB \\
-\hline
-
-\multirow{4}{*}{phase3\_chr1-22: 2,504 x 81,047,467} & CursedForest & XX (0.96) & 7hr 14min & 8GB \\ 
+& ranger - R        &             0.973     &      03:58           &       pearcey --cpus-per-task=15 --mem 1600gb   906GB \\
+\hline
+
+\multirow{4}{*}{phase3\_chr1-22: 2,504 x 81,047,467} & \cursedforest\ & XX (0.96) & 7hr 14min & 8GB \\
 & RF - Spark ML & - & - & - \\
@@
                                                         20 hours \\
-
+\hline
 % \begin{tabular}{|p{1.2cm}|p{1cm}|p{1.5cm}+l|l|p{1.8cm}|p{1.5cm}|}
@@
 
-In this section we explore the performance of CursedForest by testing its ability to scale to different sizes of data and computational resources.
-
-In order to asses these characteristics we ran CursedForst classification on synthetics datasets with varying numbers of variables (features) and samples, similar to the dataset used in to evaluate ranger [REF: ranger], allocating varying number of CPU cores to the CursedForest and also varying computational complexity of random forest by using range of mtry values.
+In this section we explore the performance of \cursedforest\ by testing its ability to scale to different sizes of data and computational resources.
+
+In order to asses these characteristics we ran \cursedforest\ classification on synthetics datasets with varying numbers of variables (features) and samples, similar to the dataset used in to evaluate ranger [REF: ranger], allocating varying number of CPU cores to the \cursedforest\ and also varying computational complexity of random forest by using range of mtry values.
 
@@
 The raw results are presented in the table [REF table1] and visualized in figure [REF: Fig 1] below
-Regardless of the number of cores used, CursedForerst displays approximately  linear dependency between the execution time and the mtry parameter (presented here as a fraction of the number of variables) [Fig 1 a)
-
-CursedForest scales almost linearly with the number of CPU cores for medium values of mtry fraction but for both lower and higher values the performance degrades slightly (Fig 1 b)  In the latter case the likely cause is communication overhead (with lower mtry values the proportion of time for parallelizable computation to the time for internode communication is lower) while in the latter case its most likely caused by reaching the clusters computational capacity.
-
-Next we investigate CursedForest scalability with regards to the size of data, by varying the number of variables and sample for a fixed mtry fraction of 0.25 and execution of 128 CPU cores.
+Regardless of the number of cores used, \cursedforest\ displays approximately  linear dependency between the execution time and the mtry parameter (presented here as a fraction of the number of variables) [Fig 1 a)
+
+\cursedforest\ scales almost linearly with the number of CPU cores for medium values of mtry fraction but for both lower and higher values the performance degrades slightly (Fig 1 b)  In the latter case the likely cause is communication overhead (with lower mtry values the proportion of time for parallelizable computation to the time for internode communication is lower) while in the latter case its most likely caused by reaching the clusters computational capacity.
+
+Next we investigate \cursedforest\ scalability with regards to the size of data, by varying the number of variables and sample for a fixed mtry fraction of 0.25 and execution of 128 CPU cores.
 The raw results are presented in the table [REF] and visualized in figure [ REF: Fig 2] below (please note log scale on the axes and the values on y axes are expresses as trees per hour).
@@
 
-To conclude CursedForst is capable of processing datasets with 50M variables and 10K samples, which is the order of magnitude of the entire genome variants expected in such dataset, for which it can build approximately 60 trees per hour.
+To conclude \cursedforest\ is capable of processing datasets with 50M variables and 10K samples, which is the order of magnitude of the entire genome variants expected in such dataset, for which it can build approximately 60 trees per hour.
 
@@
 %    Applications in Genetics and Molecular Biology, 10(1), 32. http://doi.org/10.2202/1544-6115.1691
-
-%@article{Chen2012323,
-%title = "Random forests for genomic data analysis ",
-%journal = "Genomics ",
-%volume = "99",
-%number = "6",
-%pages = "323 - 329",
-%year = "2012",
-%note = "",
-%issn = "0888-7543",
-%doi = "http://dx.doi.org/10.1016/j.ygeno.2012.04.003",
-%url = "http://www.sciencedirect.com/science/article/pii/S0888754312000626",
-%author = "Xi Chen and Hemant Ishwaran",
-%keywords = "Random forests",
-%keywords = "Random survival forests",
-%keywords = "Classification",
-%keywords = "Prediction",
-%keywords = "Variable selection",
-%keywords = "Genomic data analysis ",
-%abstract = "Random forests (RF) is a popular tree-based ensemble machine learning tool that is highly data adaptive, applies to “large p, small n” problems, and is able to account for correlation as well as interactions among features.
-This makes \{RF\} particularly appealing for high-dimensional genomic data analysis.
-In this article, we systematically review the applications and recent progresses of \{RF\} for genomic data, including prediction and classification, variable selection, pathway analysis, genetic association and epistasis detection, and unsupervised learning.
-"
-%}
 
@@
 % Results and Discussion can be combined.
-% Please do not create a heading level below \subsection. For 3rd level headings, use \paragraph{}.
 

### 2016-10-25 doc/publications/plos16/plos16.tex
Fix some semantics

@@
 \newcommand{\cursedforest}{{\sc CursedForest}}
+\newcommand{\mtry}{{\texttt mtry}}
+\newcommand{\ntree}{{\texttt ntree}}
 
@@
 {\Large
-\textbf\newline{\cursedforest - A random forest implementation for ``big'' and ``wide'' data} % Please use "title case" (capitalize all terms in the title except conjunctions, prepositions, and articles).
+\textbf\newline{\cursedforest\ - A Random Forest Implementation for ``Big'' and ``Wide'' Data} % Please use "title case" (capitalize all terms in the title except conjunctions, prepositions, and articles).
 }
@@
   \item variable  selection may succeed where accurate prediction does not
-  \item discussion of the effect of the ntree and mtry parameters for variable selection with many noise variables
+  \item discussion of the effect of the \ntree\ and \mtry\ parameters for variable selection with many noise variables
   \item   perhaps we argue that in the case where the relationship of $X$ and $y$ takes a particular form (step, ramp) we can recover
@@
 % Please keep the abstract below 300 words
-\section*{Abstract}
+\section{Abstract}
 The abstract.
@@
 \section{Introduction}
-\marginpar{change parallelisation to parallelization}
 
@@
 is inadequate~\cite{Manolio2009}.  
-\marginpar{is Complex Trait a thing? why capitals?}
-Complex Trait analysis has demonstrated that in order to explain the overall
+Complex trait analysis has demonstrated that in order to explain the overall
 heritability a model needs to take the joint contribution of all variants into account~\cite{Yang2011}.
 
-\marginpar{decison trees capture interaction -- there is no main effect term. RF inherit this property. I dont think it
+\marginpar{decison trees capture interaction -- there is no main effect term. Random forests inherit this property. I dont think it
   has anything to do with boostrapping, the point of which is to ensure that successive trees are not correlated.}
-Random Forests~\cite{Breiman2001} (RF) are able to capture interactions between features while still allowing to
+Random forests~\cite{Breiman2001} are able to capture interactions between features while still allowing to
 prioritise SNPs for their contribution to disease.  The joint interaction is achieved by appling the technique of
 bootstrap aggregation or bagging to decision trees, while a measure of variable importance can be provided, for example,
-by summing the number of times a variable is used to split a node, over all nodes and all trees.  RF are well suited for
+by summing the number of times a variable is used to split a node, over all nodes and all trees.  Random forests are well suited for
 processing ``wide'' genomic data for two reasons.  Firstly, while other machine learning applications have the
@@
 dimensionality''~\cite{Bauer2014, bellman1961adaptive}), decision trees are resistant to overfitting. 
-Secondly, RF are also very easy to parallelise. As the forest is a sum of decision trees, it it possible to grow
+Secondly, random forests are also very easy to parallelise. As the forest is a sum of decision trees, it it possible to grow
 separate trees on different processors and combine the trees.
@@
 
-However, the use of traditional compute infrastructure limits the parallelisation strategies that can be employed by
+However, the use of traditional compute infrastructure limits the parallelization strategies that can be employed by
 these methods.  The programs are limited to utilising only CPUs that are on the same computer node (multi-threading) or
@@
 processes (a separate tree is grown on each node).  Hadoop/Spark overcomes these limitation by enabling programs to
-scale beyond compute-node boundaries and hence enable more sophisticated parallelisation strategies.  In the case of
+scale beyond compute-node boundaries and hence enable more sophisticated parallelization strategies.  In the case of
 random forests this means that the computations for each node of a tree can be handed off to separate processors.
 
-Here we introduce \cursedforest, a Hadoop/Spark based implementation of random forest specifically designed to cater for
+Here we introduce \cursedforest, a Hadoop/Spark based implementation of random forests specifically designed to cater for
 ``big'' (many samples) and ``wide'' (many features) data sets. \cursedforest\ is capable of parallelising the split for
@@
 of approximately 2500 samples with up to 80 million variants~\cite{1KG2012}.  By also utilising Spark to read in and
-manipulate the standard genomic variant format (VCF) directly, \cursedforest\ outperforms existing tools on even small
+manipulate the standard genomic variant format (VCF) directly, \cursedforest\ outperforms existing tools even on small
 datasets where multi-threading generally performs well. 
@@
 
-A Spark application runs on a ``driver'' node, and this node
-divides and distributes work out to the many ``worker'' nodes, or ``executors''.
+A Spark application runs on a ``driver'' node and distributes tasks to the many ``worker'' nodes, or ``executors''.
 
@@
 
-Here we extend this toolkit to encompass supervised learning using random forest.  Spark ML's RF implementation is not
+Here we extend this toolkit to encompass supervised learning using random forests.  Spark ML's random forest implementation is not
 able to handle the extremely ``wide'' genomic data as it was developed for large number of samples with only modest
-dimensionality.  Although Spark ML can build a RF model on a subset of the data (chromosome 1~and~2), the time taken is
+dimensionality.  Although Spark ML can build a random forest model on a subset of the data (chromosome 1~and~2), the time taken is
 excessive due to the huge amount of data being aggregated and processed by the single driver node during intermediate
@@
 %Hadoop/Spark overcomes this limitation by enabling programs to scale beyond compute-node boundaries and virtually utilise infinite number of CPUs.   
-%A Spark application runs on a ``driver'' node, and this node divides and distributes work out to the many ``worker'' nodes, or ``executors''.  
 %Therefore, these historically successful solutions are not catering for the big data demands that especially due to WGS are becoming more prevalent.
@@
 
-%Spark ML's RF implementation is not able to handle the extremely ``wide'' genomic data as it was developed for large number of samples with only modest dimensionality.  
-%Although Spark ML can build a RF model on
+%Spark ML's random forest implementation is not able to handle the extremely ``wide'' genomic data as it was developed for large number of samples with only modest dimensionality.  
+%Although Spark ML can build a random forest model on
 %a subset of the data (chromosome 1~and~2), the time taken is excessive due to the huge amount of data being aggregated
@@
 In the first section we explore some of the
-properties of random forests, demonstrating that any potential flaws in \cursedforest\ are an inherent property of Random
-Forests. Here we provide demonstrations using synthetic data.  We then demonstrate the scalability of \cursedforest\ in
-respect to the dimensionality of the data, building an RF model on whole-genome data from the 1000 Genomes Project. Here
+properties of random forests, demonstrating that any potential flaws in \cursedforest\ are an inherent property of random
+forests. Here we provide demonstrations using synthetic data.  We then demonstrate the scalability of \cursedforest\ in
+respect to the dimensionality of the data, building a random forest model on whole-genome data from the 1000 Genomes Project. Here
 we report on the accuracy (in terms of the ARI)
@@
 
-Random Forests~\cite{Breiman2001} (RF) apply the technique of bootstrap aggregation or bagging to decision trees.  The training
+Random forests~\cite{Breiman2001} (RFs) apply the technique of bootstrap aggregation or bagging to decision trees.  The training
 data is independently drawn from the joint distribution of $(X,Y)$ and comprises $n$ $(p+1)$-tuples $(x_1,y_1),\ldots, (x_n,y_n)$.
@@
 
-The benefits of random forests models are:
+The benefits of random forest models are:
   \begin{itemize}
@@
 
-Since the original implementation of random forests there have been a number of developments leading to a better understanding of
+Since the original implementation of random forests, there have been a number of developments leading to a better understanding of
 the algorithm. 
 \begin{itemize}
-\item random forests are resistant against overfitting. However it is not true that they will not overfit,
+\item random forests are resistant to overfitting. However it is not true that they will not overfit,
   ~\cite{Segal.2004} gives examples of overfitting in the case where:
@@
   \end{itemize}
-However, keeping these difficult cases in mind, RF have a strong claim to being resistant to overfitting.
-
-\item \cite{Strobl.et.al.2007} demonstrates that the RF algorithm is subject to a bias in variable selection via two mechanisms:
+However, keeping these difficult cases in mind, random forests have a strong claim to being resistant to overfitting.
+
+\item \cite{Strobl.et.al.2007} demonstrates that the random forest algorithm is subject to a bias in variable selection via two mechanisms:
   \begin{enumerate}
@@
 
-\cite{Strobl.et.al.2007} introduce changes to the RF algorithm to fix 1) and 2).  However the level effect will be
+\cite{Strobl.et.al.2007} introduce changes to the random forest algorithm to fix 1) and 2).  However the level effect will be
 apparent in the CF algorithm as well.  However, there are many examples, particularly in genomics, where we have data
@@
 
-Subject to certain conditions, RF will not overfit even in the  $p > n$ case. 
+Subject to certain conditions, random forests will not overfit even in the $p > n$ case. 
 Here we demonstrate that this may hold true in extreme cases where $p \gg n$.
@@
 data.
-\marginpar{ \cite{Chen.and.Ishwaran.2012} cite many papers on extensions of RF to handle 
+\marginpar{ \cite{Chen.and.Ishwaran.2012} cite many papers on extensions of random forests to handle 
 GWAS data. What do we need to do in this area? What about linkage diseqelibrium?}
@@
 \subsection{Parameter settings}
-We consider the parameter setting for the RF algorithm. We use the R terms from the package RandomForests
+We consider the parameter setting for the random forest algorithm. We use the R terms from the {\sc randomForest} package
 \cite{Liaw.and.Weiner.2002} which incorporates the original Fortran code by Brieman and Cutler. We incorporate the 
@@
 \begin{itemize}
-\item \texttt{ntree} the number of trees.  The number of trees necessary for good performance grows with the number of predictors.
-  \marginpar{can we do proximity?}  \cite{Liaw.and.Weiner.2002} suggest that a high \texttt{ntree} is necessary to get stable
+\item \ntree\ the number of trees.  The number of trees necessary for good performance grows with the number of predictors.
+  \marginpar{can we do proximity?} \cite{Liaw.and.Weiner.2002} suggest that a high \ntree\ is necessary to get stable
   estimates of variable importance and proximity; however, even though the variable importance measures may vary from run to run,
-  the ranking of the importances is quite stable.  In addition, we note that it is possible for a RF model to have a very poor
+  the ranking of the importances is quite stable.  In addition, we note that it is possible for a random forest model to have a very poor
   fit, and be useless for prediction, and still have a useful ranking of variable importance.
-\item \texttt{mtry} --  the number of variables considered at each split (if \texttt{mtry}=$p$, we have a boosted decision
-  tree model).  If one has a very large number of variables but expects only very few to be ``important'', using larger \texttt{mtry} may give
+\item \mtry\ -- the number of variables considered at each split (if \mtry=$p$, we have a boosted decision
+  tree model).  If one has a very large number of variables but expects only very few to be ``important'', using larger \mtry\ may give
   better performance.
@@
   rule \marginpar{can we do this}
-\item the size and complexity of the individual trees is controlled in RandomForests by setting \texttt{nodesize}, the
+\item the size and complexity of the individual trees is controlled in {\sc randomForest} by setting \texttt{nodesize}, the
   minimum size of terminal nodes. It is controlled in Spark ML by setting \texttt{maxDepth}, the maximum depth of each
-  tree in the forest. The SparkML documentation somewhat unhelpfully sets \texttt{maxDepth=4} in their classification
+  tree in the forest. The Spark ML documentation somewhat unhelpfully sets \texttt{maxDepth=4} in their classification
   example\footnote{The documentation is available at \url{http://spark.apache.org/docs/latest/mllib-ensembles.html}.}.
 
-  In the ATCG example the RandomForests code produced trees with depths of 30+. We note that Random forests are a way of averaging
+  In the ATCG example the RandomForests code produced trees with depths of 30+. We note that random forests are a way of averaging
   multiple deep decision trees, trained on different parts of the same training set, with the goal of reducing the
@@
 \subsection{\cursedforest\ can evaluate variable importance for millions of features}\label{synthetic}
-%In this section we demonstrate that \cursedforest's ability to rapidly grow large number of trees, \texttt{ntree}, in
+%In this section we demonstrate that \cursedforest's ability to rapidly grow large number of trees, \ntree, in
 %parallel results in a stable feature importance ranking.  This concept was originally proposed by Liaw and
 %Weiner~\cite{Liaw.and.Weiner.2002}, who suggested that though the variable importance measures may vary from run to
-%run, the ranking of the variables will stabilise with growing \texttt{ntree}.
+%run, the ranking of the variables will stabilise with growing \ntree.
 In this section we demonstrate \cursedforest\ two main features, namely variable importance analysis to select the most
@@
 Figure~\ref{figure:synth}a illustrates that increasing the number of predictors sampled for splitting at each node
-(mtry) increases the accuracy with which features are ranked.  From $mtry=0.5$ the sort order of the features reflects
-the weight originally assigned to them (dotted line). Further increasing $mtry$ improves the predicted importance
+(\mtry) increases the accuracy with which features are ranked.  From $mtry=0.5$ the sort order of the features reflects
+the weight originally assigned to them (dotted line). Further increasing \mtry\ improves the predicted importance
 measure, which starts to converge to the weight originally assigned to them (coefficients).  It also increases the
@@
 
-Figure~\ref{figure:synth}b illustrates that increasing $mtry$ also improves the OOB (out of bag) classification error.
+Figure~\ref{figure:synth}b illustrates that increasing \mtry\ also improves the OOB (out of bag) classification error.
 %Piotr can you please explain the remaining error on the synthetic data
@@
 \captionof{figure}{{\bf Variable importance (Mean Decrease Gini) and OOB (out of bag) classification error estimate for
-    random forest with 100 trees trained on the synthetic dataset of 2.5M features and 5K samples.}  {\bf a)} Variable
+  random forests with 100 trees trained on the synthetic dataset of 2.5M features and 5K samples.}  {\bf a)} Variable
   importance of top 20 variabels normalized to the sum of the linear model coefficients. {\bf Others} represents
   aggregated importance of variables not present in the linear model.  {\bf b)} OOB (out of bag) classification error
-  estimate as funtion of mtry fraction.}
+  estimate as funtion of \mtry\ fraction.}
 \label{figure:synth}
@@
 
-In Table~\ref{table3}, we display the top 10 features, ranked by their Gini index in the Random Forest model.
+In Table~\ref{table3}, we display the top 10 features, ranked by their Gini index in the random forest model.
 
@@
 
-In order to asses these characteristics we ran \cursedforest\ classification on synthetics datasets with varying numbers of variables (features) and samples, similar to the dataset used in to evaluate ranger [REF: ranger], allocating varying number of CPU cores to the \cursedforest\ and also varying computational complexity of random forest by using range of mtry values.
+In order to asses these characteristics we ran \cursedforest\ classification on synthetics datasets with varying numbers of variables (features) and samples, similar to the dataset used in to evaluate ranger [REF: ranger], allocating varying number of CPU cores to the \cursedforest\ and also varying computational complexity of random forests by using range of \mtry\ values.
 
 We investigate the different synthetic datasets generated for section 5.1.
-and measured the time of building a random forest of 100 trees and the results reported below are averages of 5 runs.
+and measured the time of building a random forest model of 100 trees and the results reported below are averages of 5 runs.
 To improve consistency of measurements all the cases were executed with the same random seed.
 
-Firstly we look at CurstedForest horizontal scalability for a medium size dataset of 2.5 million variables and 5000 samples, by varying mtry fraction and the number of CPU cores allocated to the execution.
+Firstly we look at CurstedForest horizontal scalability for a medium size dataset of 2.5 million variables and 5000 samples, by varying \mtry\ fraction and the number of CPU cores allocated to the execution.
 The raw results are presented in the table [REF table1] and visualized in figure [REF: Fig 1] below
-Regardless of the number of cores used, \cursedforest\ displays approximately  linear dependency between the execution time and the mtry parameter (presented here as a fraction of the number of variables) [Fig 1 a)
-
-\cursedforest\ scales almost linearly with the number of CPU cores for medium values of mtry fraction but for both lower and higher values the performance degrades slightly (Fig 1 b)  In the latter case the likely cause is communication overhead (with lower mtry values the proportion of time for parallelizable computation to the time for internode communication is lower) while in the latter case its most likely caused by reaching the clusters computational capacity.
-
-Next we investigate \cursedforest\ scalability with regards to the size of data, by varying the number of variables and sample for a fixed mtry fraction of 0.25 and execution of 128 CPU cores.
+Regardless of the number of cores used, \cursedforest\ displays approximately linear dependency between the execution time and the \mtry\ parameter (presented here as a fraction of the number of variables) [Fig 1 a)
+
+\cursedforest\ scales almost linearly with the number of CPU cores for medium values of \mtry\ fraction but for both lower and higher values the performance degrades slightly (Fig 1 b).
+In the latter case the likely cause is communication overhead (with lower \mtry\ values the proportion of time for parallelizable computation to the time for internode communication is lower) while in the latter case its most likely caused by reaching the clusters computational capacity.
+
+Next we investigate \cursedforest\ scalability with regards to the size of data, by varying the number of variables and sample for a fixed \mtry\ fraction of 0.25 and execution of 128 CPU cores.
 The raw results are presented in the table [REF] and visualized in figure [ REF: Fig 2] below (please note log scale on the axes and the values on y axes are expresses as trees per hour).
@@
 Some irregularities in the graph can be attributed to computation vs communication tradeoff.
-It is also worth noting that keeping mtry fraction constants results with higher mtry values with growing number of variables, and this is what drives the performance down rather that the increase of dataset size itself.
+It is also worth noting that keeping \mtry\ fraction constants results with higher \mtry\ values with growing number of variables, and this is what drives the performance down rather that the increase of dataset size itself.
 
@@
 \caption{
-{\bf Time in seconds to build random forest of 100
-  trees with specified mtry fraction using given number of CPU cores.}}
+{\bf Time in seconds to build a random forest of 100
+  trees with specified \mtry\ fraction using given number of CPU cores.}}
 \begin{tabular}{|l|l||l|l|l|l|}
@@
 \hline
-\multirow{4}{*}{mtry (fraction)}        & \bf{0.10} & 708.3   & 340.0   & 186.0   & 134.5 \\
+\multirow{4}{*}{\mtry\ (fraction)}        & \bf{0.10} & 708.3   & 340.0   & 186.0   & 134.5 \\
                                         & \bf{0.25} & 1134.9  & 535.0   & 270.8   & 179.6 \\
@@
 \end{tabular}
-\captionof{figure}{{\bf Scalablity of the Wide Random Forest on the synthetic dataset of 2.5M features and 5K samples.} 
-   {\bf a)} Time in seconds to build 100 trees for differnet mtry fractions. 
+\captionof{figure}{{\bf Scalablity of the wide random forest on the synthetic dataset of 2.5M features and 5K samples.} 
+  {\bf a)} Time in seconds to build 100 trees for different \mtry\ fractions. 
   {\bf b)} Number of trees build per hour when using a growing number of CPU cores.}
@@
 \caption{
-{\bf Time in seconds to build random forest of 100
+{\bf Time in seconds to build a random forest of 100
   trees on 128 CPU cores with varying numbers of variables and samples.}}
@@
 %\multicolumn{4}{|l|}{\bf Heading1} & \multicolumn{4}{|l|}{\bf Heading2}\\ \hline
-\bf{Mtry}  & \bf{Mtree} & \bf{depth} & \bf{Accuracy} & \bf{Runtime} & \bf{Memory} \\
+\bf{ \mtry\ }  & \bf{Mtree} & \bf{depth} & \bf{Accuracy} & \bf{Runtime} & \bf{Memory} \\
 \hline
@@
  %http://stats.stackexchange.com/questions/173390/gradient-boosting-tree-vs-random-forest
-%overcomes this issue and is hence ideally suited to this application.  Furthermore, RF are able to capture interactions
+%overcomes this issue and is hence ideally suited to this application.  Furthermore, random forests are able to capture interactions
 %between features, which is of importance for modelling complex polygenic diseases.
 
-% We also investigated classifying the samples using Random Forests.  We chose random forests due to its ability to
+% We also investigated classifying the samples using random forests.  We chose random forests due to its ability to
 % capture complex interactions between features. Also, due to it being an ensemble method where trees can be formed
-% independently, which, is ideal for parallelisation. Furthermore, the ensemble of weak learners can help to minimise
+% independently, which, is ideal for parallelization. Furthermore, the ensemble of weak learners can help to minimise
 % overfitting and variance.

### 2016-10-26 doc/publications/plos16/plos16.tex
samlle changes to plot16.tex

@@
 more information about these events.  The challenge of big and ``wide'' data is especially pronounced in the health
-space where whole genome sequence (WGS) technology enables researchers to interrogate all 3 billion base pairs of the
+space where, for example, whole genome sequence (WGS) technology enables researchers to interrogate all 3 billion base pairs of the
 human genome.
@@
 datasets where multi-threading generally performs well. 
-\marginpar{``performant'' -- a person who performs, an actor or celebrant }
  \marginpar{Piotr this is your 20 trees has the important

### 2016-10-26 doc/publications/plos16/plos16.tex
Fix typos and things

--- 2016-10-25 doc/publications/plos16/plos16.tex
@@
   \item check Piotr's simulation  (done)
-  \item run sparkML or CursedForest on the  simulation data 
+  \item run spark ML or CursedForest on the  simulation data 
   \item we extend the simulations to many more variables 
@@
 Random forests~\cite{Breiman2001} are able to capture interactions between features while still allowing to
-prioritise SNPs for their contribution to disease.  The joint interaction is achieved by appling the technique of
+prioritise SNPs for their contribution to disease.  The joint interaction is achieved by applying the technique of
 bootstrap aggregation or bagging to decision trees, while a measure of variable importance can be provided, for example,
@@
 increasing feature size~\cite{Wright.and.Ziegle.2016}.  This is despite R now supporting long vectors (length greater
-than $2^{31}$), and the {\sc randomForest} implementation in R lending itself to paralelization to grow multiple trees
+than $2^{31}$), and the {\sc randomForest} implementation in R lending itself to parallelization to grow multiple trees
 simultaneously~\cite{Liaw.and.Weiner.2002} and them combine them.  More suitable implementations for ``wide'' genomic data have been
@@
   \item a single tree is the most appropriate model; and
-  \item the variables are hightly correlated so that the boostrap resampling does not give rise to uncorrelated trees.
+  \item the variables are highly correlated so that the bootstrap resampling does not give rise to uncorrelated trees.
   \end{itemize}
@@
   \item bagging introduces a bias, but sampling $(1- 1/e) \approx 0.632$ of the data without replacement seems to be an
-    effective stragey for removing the problem.
+    effective strategy for removing the problem.
   \end{enumerate}
@@
 \item very wide;
-\item all variables have the same number of levels (for example, measuments at different points along the genome);
+\item all variables have the same number of levels (for example, measurements at different points along the genome);
 \item the hypothesized functional relationship is of a form amenable to a decision tree.
@@
 are reduced to return the feature which gives the greatest information gain.  This process is then repeated until
-\cursedforest\ has calculated the entire decicion tree.
+\cursedforest\ has created the entire decision tree.
 
@@
 \begin{itemize}
-\item an ``information gained'' criteria for splitting. Let $f_q$ be the fraction of items labeled with value $q$ where $q= 1,
+\item an ``information gained'' criteria for splitting. Let $f_q$ be the fraction of items labelled with value $q$ where $q = 1,
   \ldots, Q$ at a node. The entropy of the node is $H = - \sum^{m}_{i=1} f_i \log^{}_2 f_i.$
@@
 \subsection{Datasets}
-%DENIS: Aidan can you please fill this in of where we got it from and how we process it for all the different tools to use
 \subsubsection{1000 Genomes Project}
-
-We obtained the 1000 Genomes Project data as VCF files from their FTP site.
+We obtained the 1000 Genomes Project data as VCF files from their FTP site.  Each VCF file contains the variants for every individual accross one chromosome.  
+We use the phase 3 dataset which contains 2,504 individuals with approximately 81 million variant positions (i.e. a matrix of $2,504 \times 81$ million).  
+For compatibility with the majority of the algorithms we demonstrate, we convert the original variant representations to double representations.  
+For example, we store the absence of a variant (0\textbar0) as 0.0, a heterozygous variant (1\textbar0 or 0\textbar1)as 1.0. and a homozygous variant (1\textbar1) as 2.0.
 
@@
 \subsubsection{Synthetic data} 
-
 Each dataset consists of $n$ samples and $m$ variables where $n << m$, and values for each variable are ordinal factors with
@@
   minimum size of terminal nodes. It is controlled in Spark ML by setting \texttt{maxDepth}, the maximum depth of each
-  tree in the forest. The Spark ML documentation somewhat unhelpfully sets \texttt{maxDepth=4} in their classification
+  tree in the forest. The Spark ML documentation sets \texttt{maxDepth} to 4 in their classification
   example\footnote{The documentation is available at \url{http://spark.apache.org/docs/latest/mllib-ensembles.html}.}.
 
-  In the ATCG example the RandomForests code produced trees with depths of 30+. We note that random forests are a way of averaging
+  In the TCGA example the RandomForests code produced trees with depths of 30+. We note that random forests are a way of averaging
   multiple deep decision trees, trained on different parts of the same training set, with the goal of reducing the
@@
 
-\begin{figure}[tbhp]
+\begin{figure}[tbph]
+\begin{adjustwidth}{-1.00in}{0in}
 \begin{tabular}{ll}
@@
 \end{tabular}
-\captionof{figure}{{\bf Variable importance (Mean Decrease Gini) and OOB (out of bag) classification error estimate for
+\caption{{\bf Variable importance (Mean Decrease Gini) and OOB (out of bag) classification error estimate for
   random forests with 100 trees trained on the synthetic dataset of 2.5M features and 5K samples.}  {\bf a)} Variable
@@
 \label{figure:synth}
+\end{adjustwidth}
 \end{figure}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{\cursedforest\ selects previously demonstrated cancer variants }
+\subsection{\cursedforest\ selects previously demonstrated cancer variants}
 %Feature selection can be useful for determining variables the contribute to the label or for perhaps selecting the n most important
@@
   trees with specified \mtry\ fraction using given number of CPU cores.}}
-\begin{tabular}{|l|l||l|l|l|l|}
+\begin{tabular}{|l|l|l|l|l|l|}
 \hline
@@
 \begin{figure}[tbhp]
+\begin{adjustwidth}{-1.00in}{0in}
 \begin{tabular}{ll}
@@
 \end{tabular}
-\captionof{figure}{{\bf Scalablity of the wide random forest on the synthetic dataset of 2.5M features and 5K samples.} 
+\caption{{\bf Scalablity of the wide random forest on the synthetic dataset of 2.5M features and 5K samples.} 
   {\bf a)} Time in seconds to build 100 trees for different \mtry\ fractions. 
@@
 \begin{table}[!ht]
+\begin{adjustwidth}{-1.00in}{0in}
 \centering
@@
 \label{table10}
+\end{adjustwidth}
 \end{table}
@@
 \end{tabular}
-\captionof{figure}{{\bf Scalablity of the Wide Random Forest on synthetic datasets with varying number of samples and variables.} 
+\caption{{\bf Scalablity of the Wide Random Forest on synthetic datasets with varying number of samples and variables.} 
   {\bf a)} Number of trees build per hour when growing number of variables. 
@@
 \begin{table}[!ht]
-%\begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
 \caption{
@@
 \label{table2}
-%\end{adjustwidth}
 \end{table}

### 2016-10-26 doc/publications/plos16/plos16.tex
add TCGA subsubsection

@@
 For example, we store the absence of a variant (0\textbar0) as 0.0, a heterozygous variant (1\textbar0 or 0\textbar1)as 1.0. and a homozygous variant (1\textbar1) as 2.0.
+
+\subsubsection{The Cancer Genome Atlas}
+We obtained The Cancer Genome Atlas (TCGA) BRCA data as VCF files from the GDC data portal.
+This data is in the category ``Simple Nucleotide Variation''.
+After downloading, we use a custom script to name the samples in the files and filter for only the "Normal" tissue column.
+We then use GATK to merge the files into a single VCF file.
 

### 2016-10-26 doc/publications/plos16/plos16.tex
changes to plos16.tex

@@
 Identifying the relevant and disease specific base-pair difference between individuals is the focus of genome wide
-association studies (GWAs).  These analysis are typically performed on only the most frequently differing base-pairs
-(SNPs) by applying linear or logistic regression analysis to each SNP separately~\cite{CCC2007}.  It has since been
+association studies (GWAs).  These analysis are typically performed on only the most frequently differing base-pairs, termed
+single nucleotide polymorphisms (SNPs), by applying linear or logistic regression analysis to each SNP separately~\cite{CCC2007}.  It has since been
 demonstrated that not taking joint interactions between SNPs into account when filtering for potential disease mutations
@@
 
-\marginpar{decison trees capture interaction -- there is no main effect term. Random forests inherit this property. I dont think it
-  has anything to do with boostrapping, the point of which is to ensure that successive trees are not correlated.}
-Random forests~\cite{Breiman2001} are able to capture interactions between features while still allowing to
-prioritise SNPs for their contribution to disease.  The joint interaction is achieved by applying the technique of
-bootstrap aggregation or bagging to decision trees, while a measure of variable importance can be provided, for example,
-by summing the number of times a variable is used to split a node, over all nodes and all trees.  Random forests are well suited for
+%\marginpar{decison trees capture interaction -- there is no main effect term. Random forests inherit this property. I dont think it
+%  has anything to do with boostrapping, the point of which is to ensure that successive trees are not correlated.}
+%Random forests~\cite{Breiman2001} are able to capture interactions between features while still allowing to
+%prioritise SNPs for their contribution to disease.  The joint interaction is achieved by applying the technique of
+%bootstrap aggregation or bagging to decision trees, while a measure of variable importance can be provided, for example,
+%by summing the number of times a variable is used to split a node, over all nodes and all trees.  
+Random forests are well suited for
 processing ``wide'' genomic data for two reasons.  Firstly, while other machine learning applications have the
@@
 Secondly, random forests are also very easy to parallelise. As the forest is a sum of decision trees, it it possible to grow
-separate trees on different processors and combine the trees.
+separate trees on different processors and combine the trees. See the discussion of the properties of random forests
+Section \ref{section:methods}.
 
@@
 simultaneously~\cite{Liaw.and.Weiner.2002} and them combine them.  More suitable implementations for ``wide'' genomic data have been
-developed in other programming languages.  {\sc CloudForest}~\cite{Bressler2015} written in Go, achieves fast running
+developed in other programming languages.  {\sc CloudForest}~\cite{Bressler2015} written in Go, achieves fast run
 times by effective use of the CPU cache, optimising for different classes of features and efficiently multi-threading.
@@
 
-\cursedforest\ is included in our previously developed framework for Spark-based genomic data analysis,
-VariantSpark~\cite{OBrien2015}, which therefore now enables supervised and unsupervised machine learning. This offers a
-comprehensive analysis toolkit that can scale for the future data demands, where the richness from WGS data can be
-utilised to ``fill in'' or impute the information at previously unobserved genomic positions in older array
-technology~\cite{Howie2012}.  This provides the potential to impute samples in the GWAs catalogue and generate datasets
-of hundreds of thousands of individuals with millions of variants, highlighting the need for incorporating modern
-compute paradigms to deal with these challenges.
-
 A Spark application runs on a ``driver'' node and distributes tasks to the many ``worker'' nodes, or ``executors''.
@@
 
+\cursedforest\ is included in our previously developed framework for Spark-based genomic data analysis,
+VariantSpark~\cite{OBrien2015}, which therefore now enables supervised and unsupervised machine learning. This offers a
+comprehensive analysis toolkit that can scale for the future data demands, where the richness from WGS data can be
+utilised to ``fill in'' or impute the information at previously unobserved genomic positions in older array
+technology~\cite{Howie2012}.  This provides the potential to impute samples in the GWAs catalogue and generate datasets
+of hundreds of thousands of individuals with millions of variants, highlighting the need for incorporating modern
+compute paradigms to deal with these challenges.
+
+In the first section we explore some of the
 \marginpar{Needs to be rewritten once we actually have the sections} 
-In the first section we explore some of the
 properties of random forests, demonstrating that any potential flaws in \cursedforest\ are an inherent property of random
@@
 \section{Methods}
+\label{section:methods}
 
@@
   \item can handle missing values;
-  \item are insensitive to useless predictors;
+  \item are insensitive to uninforamtive predictors;
   \item are able to capture interactions between features (which is of importance for modelling complex polygenic diseases).
@@
 
-Random forests~\cite{Breiman2001} (RFs) apply the technique of bootstrap aggregation or bagging to decision trees.  The training
-data is independently drawn from the joint distribution of $(X,Y)$ and comprises $n$ $(p+1)$-tuples $(x_1,y_1),\ldots, (x_n,y_n)$.
-$X$ and $\theta_b$ for $b=1,\ldots,B$ are i.i.d. random vectors and a tree is grown on each of these $\theta_b$ samples 
-($X$ is randomly sampled $B$ times with replacement and a tree is grown on each of these $B$ samples). 
+Random forests  apply the technique of bootstrap aggregation or bagging to decision trees.  The training
+data sets are independently drawn. with replacement,  from the joint distribution of $(X,Y)$ so that each set comprises $n$ $(p+1)$-tuples
+ $(x_1,y_1),\ldots, (x_n,y_n)$. This si done $B$ times and  a tree is grown on each of these $B$ samples.
 The results are combined in an appropriate way. In a classification problem we take the majority vote for the trees over
@@
 The average prediction error for an individual tree $h(X; \theta)$ is
+\marginpar{is this formulation correct in the case of classification? See Friedman on 0-1 losss}
 \begin{equation}
@@
 \item random forests are resistant to overfitting. However it is not true that they will not overfit,
-  ~\cite{Segal.2004} gives examples of overfitting in the case where:
+  ~\cite{Segal.2004} gives examples of overfitting in cases where:
   \begin{itemize}
@@
 
-\subsubsection{convergence}
 Biau {\it et al.}~\cite{Biau.2012} provide a proof that a random forest model will converge at a rate that depends on
@@
 rate that depends on the size of $S$ and not the number of noise variables in the data set. If the size of $S$ is small
-compared to the total number of variables, then the rate of convergence will be quicker.
-\marginpar{say something about the rate of convergence}
-
-Subject to certain conditions, random forests will not overfit even in the $p > n$ case. 
-Here we demonstrate that this may hold true in extreme cases where $p \gg n$.
-
-\marginpar{why is this relevant, i.e. if this is special to VariantSpark it should be explained better.}
-\subsubsection{variable selection}
+compared to the total number of variables, then the rate of convergence will be quicker. It is not clear waht the size
+of teh set $S$ will be in a GWAs application. There are cases of small groups of SNPs with large effectsd and also of
+contitions casued by the interaction of large groups of SNPs. 
+\marginpar{check this}
+
 \cite{Genuer.et.al.2010} and \cite{Diaz.and.Alvarez.2006} describe iterative schemes for doing variable selection using
@@
 \subsection{\cursedforest.}
-% DENIS: Aidan and Piotr can you please explain here how CursedForest is implemented
-Traditionally, VariantSpark made use of Spark's machine learning algorithms ``Spark ML''. These algorithms require data
+VariantSpark made use of Spark's machine learning algorithms ``Spark ML''. These algorithms require data
 in a ``DataFrame'', similar to DataFrames in R.
@@
 In this case, the information gain is IG = Entropy(parent node) - Weighted Sum of Entropy(Children)
-\item variable importance is the sum of IG over all notes and all trees for that variable
+\item variable importance is the sum of IG over all notes and all trees for that variable.
 \end{itemize}
@@
 
-%DENIS: Rob how was the synthetic data generated
 \subsubsection{Synthetic data} 
+\marginpar{these are not actually encoded as factors}
 Each dataset consists of $n$ samples and $m$ variables where $n << m$, and values for each variable are ordinal factors with
 three levels represented as numbers $\{0, 1, 2\}$ (which correspond to an additive effect encoding of genomics variation) randomly
-generated from uniform distribution with equal probabilities.  The dichotomous response is generated from a two level
-Bernoulli trail with $p$ (probability of success) calculated by applying inverse logit transformation to the continuous
+generated from uniform distribution with equal probabilities.  The dichotomous response is generated from a
+Bernoulli trail with $p$ (probability of success) calculated by applying an inverse logit transformation to the continuous
 linear response ($y$) of five variables:

### 2016-10-26 doc/publications/plos16/plos16.tex
changes to plos16.tex

@@
 \clearpage
-\section{Notes}
-\begin{itemize}
-\item I think the structure of the first section should be
-  \begin{itemize}
-  \item convergence results from Biau 2012  (done)
-  \item consider  \cite{Segal.2004} examples
-  \item simulations reproduced from Genuer et al 2010 (no -- we need classification examples)
-  \item check Piotr's simulation  (done)
-  \item run spark ML or CursedForest on the  simulation data 
-  \item we extend the simulations to many more variables 
-  \item I dont think I can get R to handle chromosme 1 etc. May have to run CF on chromosme 22, see table
-    \ref{table:smaller_data} -- On it, Aidan
-  \item variable  selection may succeed where accurate prediction does not
-  \item discussion of the effect of the \ntree\ and \mtry\ parameters for variable selection with many noise variables
-  \item   perhaps we argue that in the case where the relationship of $X$ and $y$ takes a particular form (step, ramp) we can recover
-    the significant variables in the presence of a very large number of noise variables
-  \item in the case of genomic data, our functions of interest are more likely to be stop or ramp functions rather than harmonic
-    functions. So we have some hope.
-  \end{itemize}
-\item check out 
-  \begin{itemize}
-  \item D\'iaz-Uriarte, R., \& Alvarez de Andr\'es, S. (2006). (done)
-  \item \cite{Goldstein.et.al.2011}
-notes that the 0, 1, 2 encoding of alleles is an additive model i.e. each additional minor allele increases the effect 
-\begin{verbatim}
-Different genetic effects
-Type	Mechanism	Partition
-Additive	Each additional minor allele increases variation	0, 1, 2
-Dominant	Presence of at least 1 minor allele increases variation	0, 1/2
-Recessive	Two minor alleles needed for variation	0/1, 2
-Heterosis	Heterozygote leads to variation	0/2, 1
-\end{verbatim}
-
-  \item Chen (2012)  (done)
-  \item VSURF and  varSelRF, (done) recursive feature (forwards backwards) using variable importance
-  \item \cite{Wright.and.Ziegle.2016} ranger package (doing it!)
-  \item  r2VIM: A new variable selection method for random forests in
-    genome-wide association studies
-  \item CloudForest: A Scalable and Efficient Random Forest
-    Implementation for Biological Data
-   Ensembles of decision trees in go/golang   --  do we need to look
-   at this? 
-  \item \cite{Tuv.et.al.2009} feature selection via adding permuted variables. Do we want to consider this? (no, leave it for now)
-  \end{itemize}
-\item other things to do
-  \begin{itemize}
-  \item what is the problem with bagging? Do we need to redo the thousand genome stuff?
-  \end{itemize}
-\end{itemize}
 
@@
   minimum size of terminal nodes. It is controlled in Spark ML by setting \texttt{maxDepth}, the maximum depth of each
-  tree in the forest. The Spark ML documentation sets \texttt{maxDepth} to 4 in their classification
-  example\footnote{The documentation is available at \url{http://spark.apache.org/docs/latest/mllib-ensembles.html}.}.
+  tree in the forest\footnote{The Spark ML documentation \cite[]{Spark.2016} sets \texttt{maxDepth}
+    to 4 in their classification example. This is wrong. The
+    individual trees should be grown to a large size to ensure that
+    they give an unbiased estimate.}.
 
@@
 
+\marginpar{wrong figure cited}
 Figure~\ref{figure:synth}b illustrates that increasing \mtry\ also improves the OOB (out of bag) classification error.

### 2016-10-27 doc/publications/plos16/plos16.tex
changes to ploos16.tex

@@
 
-Identifying the relevant and disease specific base-pair difference between individuals is the focus of genome wide
+Identifying the relevant or disease specific base-pair difference between individuals is the focus of genome wide
 association studies (GWAs).  These analysis are typically performed on only the most frequently differing base-pairs, termed
@@
 
-%\marginpar{decison trees capture interaction -- there is no main effect term. Random forests inherit this property. I dont think it
-%  has anything to do with boostrapping, the point of which is to ensure that successive trees are not correlated.}
-%Random forests~\cite{Breiman2001} are able to capture interactions between features while still allowing to
-%prioritise SNPs for their contribution to disease.  The joint interaction is achieved by applying the technique of
-%bootstrap aggregation or bagging to decision trees, while a measure of variable importance can be provided, for example,
-%by summing the number of times a variable is used to split a node, over all nodes and all trees.  
-Random forests are well suited for
+A tree based model is capable of taking this interaction of variables
+into account. In addition, random forests are well suited for
 processing ``wide'' genomic data for two reasons.  Firstly, while other machine learning applications have the
@@
 Secondly, random forests are also very easy to parallelise. As the forest is a sum of decision trees, it it possible to grow
-separate trees on different processors and combine the trees. See the discussion of the properties of random forests
-Section \ref{section:methods}.
+separate trees on different processors and combine the trees. See the
+discussion of the properties of random forests in Section \ref{section:methods}.
 
@@
 increasing feature size~\cite{Wright.and.Ziegle.2016}.  This is despite R now supporting long vectors (length greater
-than $2^{31}$), and the {\sc randomForest} implementation in R lending itself to parallelization to grow multiple trees
+than $2^{31}-1$), and the {\sc randomForest} implementation in R lending itself to parallelization to grow multiple trees
 simultaneously~\cite{Liaw.and.Weiner.2002} and them combine them.  More suitable implementations for ``wide'' genomic data have been
@@
 Using VariantSpark, we successfully built a k-means model on the full $2500 \times 80$ million matrix to cluster
-individuals by their ethnicity achieving an Adjusted Rand Index of 0.84 (with 1 being perfect clustering and -1 random
+individuals by their ethnicity achieving an Adjusted Rand Index (ARI) of 0.84 (with 1 being perfect clustering and -1 random
 clustering).
@@
 discussion.
-
-%% We therefore developed \cursedforest\, a Spark-based random forest algorithm able to process big and ``wide'' data.
-%% % Unlike the standard Apache implementation, \cursedforest\ is able to easily process data with millions of dimensions without
-%% % excessive resource requirements on the driver or worker nodes.  DENIS Piotr this is your 20 trees has the important feature in
-%% % the top 5 and 100 trees has it always as the top element, do we need to spell this out better, it should be a result section.
-%% \marginpar{Piotr this is your 20 trees has the important feature in the top 5 and 100 trees has it always as the top
-%%   element, do we need to spell this out better, it should be a result section.}  Harnessing the virtually unlimited
-%% capability to parallelise tasks, \cursedforest\ can build a large number of trees, which boosts the accuracy for
-%% variable selection as demonstrated by {\sc r2VIM}~\cite{Szymczak2016} on smaller datasets.
-
-%Hadoop/Spark overcomes this limitation by enabling programs to scale beyond compute-node boundaries and virtually utilise infinite number of CPUs.   
-%Therefore, these historically successful solutions are not catering for the big data demands that especially due to WGS are becoming more prevalent.
-Here, classification or feature importance analysis needs to be performed on datasets with a much larger number of features~\cite{Yano2016}.
-%For example, the 1000 Genome data consists of approximately 2500 samples with up to 80 million variants~\cite{1KG2012},
-%This dataset consumes close to a terabyte of disk space.  
-%We previously demonstrated the versatility and scalability of Spark by developing VariantSpark~\cite{OBrien2015}, a
-%framework allowing users to easily analyse Variant Call Format (VCF) files using ML algorithms on the Spark framework.
-%Using VariantSpark, we successfully built a k-means model on the full
-%$2500 \times 80$ million matrix to cluster individuals by
-%their ethnicity achieving an Adjusted Rand Index of 0.84 (with 1 being perfect clustering and -1 random clustering).
-
-%Spark ML's random forest implementation is not able to handle the extremely ``wide'' genomic data as it was developed for large number of samples with only modest dimensionality.  
-%Although Spark ML can build a random forest model on
-%a subset of the data (chromosome 1~and~2), the time taken is excessive due to the huge amount of data being aggregated
-%and processed by the single driver node during intermediate stages of building the model (see Result~\ref{comp}).  This
-%unbalanced work load where the driver node becomes the bottleneck and worker nodes being idle prevents a seamless
-%scaling to larger datasets. The memory requirements per executor also increases with dimensionality due to the data types
-%Spark ML uses, which we elaborate on in the discussion.
-%
-%We therefore developed \cursedforest\, a Spark-based random forest algorithm able to process big and ``wide'' data. 
-%Unlike
-%the standard Apache implementation, \cursedforest\ is able to easily process data with millions of dimensions without
-%excessive resource requirements on the driver or worker nodes.
-%DENIS Piotr this is your 20 trees has the important feature in the top 5 and 100 trees has it always as the top element, do we need to spell this out better, it should be a result section.
 
@@
 we report on the accuracy (in terms of the ARI)
-\marginpar{what is ARI?}
  and the error.  Finally, given the role different parameter values can
@@
 \item \ntree\ the number of trees.  The number of trees necessary for good performance grows with the number of predictors.
-  \marginpar{can we do proximity?} \cite{Liaw.and.Weiner.2002} suggest that a high \ntree\ is necessary to get stable
+\cite{Liaw.and.Weiner.2002} suggest that a high \ntree\ is necessary to get stable
   estimates of variable importance and proximity; however, even though the variable importance measures may vary from run to run,
@@
   better performance.
-\item For classification problems where the class frequencies are extremely unbalanced it may be necessary to change the prediction
-  rule \marginpar{can we do this}
 \item the size and complexity of the individual trees is controlled in {\sc randomForest} by setting \texttt{nodesize}, the
@@
     they give an unbiased estimate.}.
-
-  In the TCGA example the RandomForests code produced trees with depths of 30+. We note that random forests are a way of averaging
-  multiple deep decision trees, trained on different parts of the same training set, with the goal of reducing the
-  variance. \url{https://en.wikipedia.org/wiki/Random_forest} and refs therein.  
-\marginpar{are we including this example?  Supplementary info?}
+  In the TCGA example the RandomForests code produced trees with depths of 30+. 
 \end{itemize}

### 2016-10-28 doc/publications/plos16/plos16.tex
small changes to plos16.tex and added  supplementary_information.tex

@@
 method section.  Briefly we generate features by randomly sampling from a three value vector (1,2,3) with equal
+\marginpar{repicated methods section}
 probabilities, which mimics genomic mutation data and corresponds to homozygote reference, heterozygote or homozygote
@@
 5 feature in order of the weight that was assigned to them.  We are running Apache Spark 1.6.1 on a YARN cluster with 12
-worker nodes each with 16 Intel Xeon E5-2660@2.20GHz CPU cores and 128 GB or RAM.  We captured the OOB (out of bag)
+worker nodes each with 16 Intel Xeon E5-2660@2.20GHz CPU cores and 128 GB of RAM.  We captured the OOB (out of bag)
 classification error and the 20 most important variables.
 
+\marginpar{the accuracy hardly changes}
 Figure~\ref{figure:synth}a illustrates that increasing the number of predictors sampled for splitting at each node
@@
 {\bf Performance comparison between the different machine learning algorithms.}}
-\begin{tabular}{|l+l|l|l|l|l|l|p{1cm}|}
+\begin{tabular}{|l|l|l|l|l|l|p{1cm}|}
 \hline
@@
 \multirow{3}{*}{phase1\_chr22: 1,092 x 490,036} & \cursedforest\ & 0.077 (XX) &                  &                                                                   \\
-                                                & RF - Spark ML  &            &                  &                                                                   \\
+                                                & Spark ML  &            &                  &                                                                   \\
                                                 & RF - R         & 0.05       & 24 min           & 18GB\footnote{using the \texttt{parallel} packages with 10 nodes} \\
@@
 \multirow{3}{*}{phase3\_chr1: 2,504 x 6,450,364}    & \cursedforest\ & 0.063 (XX) & 1hr 58min        & 4GB \\
-                                                    & RF - Spark ML  & 0.94       & 8hr 29min        & 8GB \\
-                                                    & LR - Spark ML  &            & 2hr 6min         & 8GB \\
+                                                    & Spark ML  & 0.94       & 8hr 29min        & 8GB \\
                                                     & ranger - R      & 0.95       & 28 mins          & num.threads=10\footnote{don't
@@
                                                     & RF - Spark ML  &            &                                                      & \\
-                                                    & LR - Spark ML  &            & 14hr 48min                                           & 8GB \\
                                                     & ranger - R      &  0.97      & 58 mins\footnote{ but took 6 hours to load the
@@
 \multirow{4}{*}{phase3\_chr1-22: 2,504 x 81,047,467} & \cursedforest\ & XX (0.96) & 7hr 14min & 8GB \\
-& RF - Spark ML & - & - & - \\
-& LR - Spark ML & - & - & - \\ 
-& kmeans - Spark ML  & - (0.82)      & 30hr 44min    & 24GB \\ 
+& Spark ML & - & - & - \\
 & ranger (C++)       &        NA     &        NA     &            80 CPU -- runs out of memory (12.8GB per CPU) \\
@@
 \hline
-% \begin{tabular}{|p{1.2cm}|p{1cm}|p{1.5cm}+l|l|p{1.8cm}|p{1.5cm}|}
-% \hline
-% %\multicolumn{4}{|l|}{\bf Heading1} & \multicolumn{4}{|l|}{\bf Heading2}\\ \hline
-% \bf{Data- set} & \bf{Samp- les} & \bf{Feat- ures}  & \bf{Method} & \bf{Error (ARI)} & \bf{Runtime} & \bf{Memory/ Node} \\
-% \hline
-% %\multirow{3}{*}{phase1\_chr22: 1,092 x 490,036} & CursedForest & 0.077 (XX)  &  &  \\
-% phase1 chr22 &1,092 & 490,036 & CursedForest & 0.077 (XX)  & 9min 16sec & 1GB\footnote{\label{note10}10 nodes} \\
-% &&& RF - Spark ML & 0.069 (0.89) & 14min 13sec & 1GB\footref{note10}  \\
-% &&& RF - R &  0.05  & 24min  & 18GB\footref{note1}\footnote{using the \texttt{parallel} packages}\\
-% &&& ranger -R &         0.05 &        1min 10sec  &          16GB  (one processor) \\
-%  \hline
-% phase3 chr22 & 2,504 & 1,103,548 &  &  &  &  \\
-% &&& CursedForest & 0.053 & 9 min 8 sec & 2GB\footnote{\label{note20}20 nodes}\\
-% &&& RF - Spark ML & 0.058 (0.92) & 1hr 1min & 2GB\footref{note20} \\
-% &&& RF - R & NA  & NA & NA\footnote{long vectors ($> 2^31-1$)  are not supported in the .Fortran interface}\\
-% &&&  H2O   &        0.05 &     16 hours       &   600GB \footnote{java -Xmx600g -jar h2o.jar,   h2o.init(nthreads=20)} \\
-% \hline
-% phase3 chr1 & 2,504 & 6,450,364 & CursedForest & 0.063 (XX) & 1hr 58min & 4GB \\
-% &&& RF - Spark ML & (0.94) & 8hr 29min & 8GB \\
-% &&& LR - Spark ML&  & 2hr 6min & 8GB\\ 
-% &&& ranger  R   &  0.06   &     28 mins  &            num.threads = 10\footnote{don't know how to get the memory used, This
-%   was done on dpas-03} \\ 
-% \hline
-% phase3 chr1-3 & 2,504 & 19,328,051  & CursedForest & 0.046 (XX) & 4hr 0min & 4GB\\
-% &&& RF - Spark ML & - & -  & -\\
-% &&& LR - Spark ML & & 14hr 48min & 8GB \\ 
-% &&& ranger  R   &  0.05   &     58 mins\footnote{ but took 6 hours to load the data}  &      909GB\footnote{I think this is total memory. The job was run with
-%   ntasks-per-node=160 cores-per-socket=10 on ruby }  \\ 
-% \hline
-% phase3 chr1-22 & 2,504 & 81,047,467 & CursedForest & XX (0.96) & 7hr 14min & 8GB \\ 
-% &&& RF - Spark ML & - & - & - \\
-% &&& LR - Spark ML & - & - & - \\ 
-% %&&& kmeans - Spark ML & - (0.82) & 30hr 44min & 24GB \\ 
-% \hline
 \end{tabular}

### 2016-10-28 doc/publications/plos16/supplementary_information.tex
small changes to plos16.tex and added  supplementary_information.tex

\documentclass[11pt]{article}

\usepackage{geometry}
 \geometry{
 a4paper,
% total={210mm,297mm}
% }
 left=17mm,
 right=23mm,
 top=20mm, 
 bottom=20mm,
 }
\usepackage{url}
\usepackage{rotating}
\usepackage{natbib}
\usepackage{placeins}
\usepackage{longtable}
\usepackage[latin1]{inputenc}
\usepackage{multirow}
\usepackage{hyperref}
%\usepackage{showframe}
\usepackage{pdflscape}
%\usepackage{pbox}
\usepackage{xcolor}
\usepackage{amsmath}
\usepackage{rotating}
\usepackage{multicol}
\usepackage{caption}

\usepackage{pdflscape}

\usepackage{listings}

\lstset{
  basicstyle=\ttfamily, 
  basewidth=0.5em,                 %the default setting of listings with "fixed columns" has a space 0.6em wide, 
                                   %while the characters in Computer Modern Typewriter are 0.5em wide.
                                   %http://tex.stackexchange.com/questions/179071/spacing-looks-wrong-in-listings-when-using-fixed-columns
  backgroundcolor=\color{gray!10},
  keywordstyle=\color{green!40!black},
  columns=fixed,
  language=R,                     % the language of the code
  basicstyle=\footnotesize,       % the size of the fonts that are used for the code
  numbers=left,                   % where to put the line-numbers
  numberstyle=\tiny\color{gray},  % the style that is used for the line-numbers
  stepnumber=1,                   % the step between two line-numbers. If it's 1, each line
                                  % will be numbered
  numbersep=5pt,                  % how far the line-numbers are from the code
  showspaces=false,               % show spaces adding particular underscores
  showstringspaces=false,         % underline spaces within strings
  showtabs=false,                 % show tabs within strings adding particular underscores
  frame=single,                   % adds a frame around the code
  rulecolor=\color{black},        % if not set, the frame-color may be changed on line-breaks within not-black text (e.g. commens (green here))
  tabsize=2,                      % sets default tabsize to 2 spaces
  captionpos=b,                   % sets the caption-position to bottom
  breaklines=true,                % sets automatic line breaking
  breakatwhitespace=false,        % sets if automatic breaks should only happen at whitespace
  title=\lstname,                 % show the filename of files included with \lstinputlisting;
                                  % also try caption instead of title
  keywordstyle=\color{blue},      % keyword style
  commentstyle=\color{green},   % comment style
  stringstyle=\color{red},      % string literal style
  escapeinside={\%*}{*)},         % if you want to add a comment within your code
  morekeywords={*,...}            % if you want to add more keywords to the set
}

\usepackage{graphicx}
\usepackage{gensymb}
\usepackage{nag}   %It warns the user about the usage of old packages or commands (for example, using \it, \tt, etc.)
\usepackage{fixltx2e}
%fixltx2e package. It fixes some 'mistakes' in Latex. From the description:
%        ensure one-column floats don't get ahead of two-column floats;
%        correct page headers in twocolumn documents;
%        stop spaces disappearing in moving arguments;
%        allowing \fnysmbol to use text symbols;
%        allow the first word after a float to hyphenate;
%        \emph can produce caps/small caps text;
\usepackage{booktabs}
% \centering instead of \begin{center} \end{center} to center things inside tables/figures etc. \centering doesn't add any additional vertical space.
\usepackage{microtype}  %for small-scale typographic enhancements (character protrusion, font expansion, letter-spacing).
\usepackage{fancyvrb}   %get precise control in verbatim listings.
%\usepackage{siunitx} To typeset units
\usepackage{numprint} %format numbers nicely 
%~, the non-breakable space.

\parindent=0pt
\parskip=8pt
\setlength\itemsep{0em}

%\sloppy
%\SweaveOpts{echo=FALSE,prefix.string=script18/plot}
\renewcommand{\textfraction}{0.0}

\let\oldmarginpar\marginpar
\renewcommand\marginpar[1]{\-\oldmarginpar[\raggedleft\footnotesize #1]%
{\raggedright\footnotesize #1}}

\newcommand{\supp}{\mathop{\mathrm{supp}}}

\newcommand{\cursedforest}{{\sc CursedForest}}
\newcommand{\mtry}{{\texttt mtry}}
\newcommand{\ntree}{{\texttt ntree}}

\newsavebox\ltmcbox

\title{\cursedforest\ - A Random Forest Implementation for ``Big'' and ``Wide'' Data \\
Supplementary Information}
\author{}
\date{}

\begin{document}
%\setkeys{Gin}{width=8cm}
\setkeys{Gin}{height=8cm,width=0.9\columnwidth}
\maketitle
%%\tableofcontents
%% ## \begin{landscape}
%% ##   \small
%% ## \end{landscape}

%% \begin{lstlisting}
%% \end{lstlisting}

<<echo=FALSE,eval=FALSE>>=

@

\section{supplementary Information for Section 4.1}

\begin{table}[!ht]
\begin{minipage}{\textwidth}
%\begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
\centering
\caption{
{\bf Performance comparison between the different machine learning algorithms.}}
\begin{tabular}{|l|l|l|l|l|l|p{1cm}|}
\hline
\bf{Genome \%}                      & \bf{Method} & \bf{Error (ARI)} & \bf{Runtime} & \bf{Memory/Node} \\
\hline

\multirow{3}{*}{phase1\_chr22: 1,092 x 490,036} & \cursedforest\ & 0.077 (XX) &                  &                                                                   \\
                                                & RF - Spark ML  &            &                  &                                                                   \\
                                                & RF - R         & 0.05       & 24 min           & 18GB\footnote{using the \texttt{parallel} packages with 10 nodes} \\
                                                & ranger - R      & 0.94       & 1 min 10 seconds & 16GB  (one processor)                                             \\
                                                & ranger - c++     & 0.94       & 7 mins           & --nthreads 20                                                     \\

 \hline
\multirow{3}{*}{phase3\_chr22: 2,504 x 1,103,548}   &               &            &                  &                                                                \\
                                                    & RF - R        & NA         & NA               & NA\footnote{long vectors ($> 2^31-1$)  are not supported in the .Fortran interface} \\
                                                    & H2O           & 0.05       & 16 hours         & 600GB \footnote{java -Xmx600g -jar h2o.jar,   h2o.init(nthreads=20)} \\
                                                    & ranger - R    & 0.94       & 4 1/2 mins       & 21GB (one processor) \\
\hline
\multirow{3}{*}{phase3\_chr1: 2,504 x 6,450,364}    & \cursedforest\ & 0.063 (XX) & 1hr 58min        & 4GB \\
                                                    & RF - Spark ML  & 0.94       & 8hr 29min        & 8GB \\
                                                    & LR - Spark ML  &            & 2hr 6min         & 8GB \\
                                                    & ranger - R      & 0.95       & 28 mins          & num.threads=10\footnote{don't
                                                                                                      know how to get the memory
                                                                                                      used,   This was done on dpas-03} \\
\hline
\multirow{3}{*}{phase3\_chr1-3: 2,504 x 19,328,051} & \cursedforest\ & 0.046 (XX) & 4hr 0min                                             & 4GB \\
                                                    & RF - Spark ML  &            &                                                      & \\
                                                    & LR - Spark ML  &            & 14hr 48min                                           & 8GB \\
                                                    & ranger - R      &  0.97      & 58 mins\footnote{ but took 6 hours to load the
                                                                                   data}
                                                                                                                                         &
                                                                                                                                           909GB\footnote{I
                                                                                                                                           think
                                                                                                                                           this
                                                                                                                                           is
                                                                                                                                           total
                                                                                                                                           memory. The
                                                                                                                                           job
                                                                                                                                           was
                                                                                                                                           run
                                                                                                                                           with
                                                                                                                                                   ntasks-per-node=160
                                                                                                                                           cores-per-socket=10
                                                                                                                                           on
                                                                                                                                           ruby
                                                                                                                                           }  \\
& ranger - R        &             0.973     &      03:58           &       pearcey --cpus-per-task=15 --mem 1600gb   906GB \\
\hline

\multirow{4}{*}{phase3\_chr1-22: 2,504 x 81,047,467} & \cursedforest\ & XX (0.96) & 7hr 14min & 8GB \\
& RF - Spark ML & - & - & - \\
& LR - Spark ML & - & - & - \\ 
& kmeans - Spark ML  & - (0.82)      & 30hr 44min    & 24GB \\ 
& ranger (C++)       &        NA     &        NA     &            80 CPU -- runs out of memory (12.8GB per CPU) \\
& ranger (C++)       &        NA     &        NA     &            160
                                                        CPU -- runs
                                                        out of time at
                                                        20 hours \\
\hline
% \begin{tabular}{|p{1.2cm}|p{1cm}|p{1.5cm}+l|l|p{1.8cm}|p{1.5cm}|}
% \hline
% %\multicolumn{4}{|l|}{\bf Heading1} & \multicolumn{4}{|l|}{\bf Heading2}\\ \hline
% \bf{Data- set} & \bf{Samp- les} & \bf{Feat- ures}  & \bf{Method} & \bf{Error (ARI)} & \bf{Runtime} & \bf{Memory/ Node} \\
% \hline
% %\multirow{3}{*}{phase1\_chr22: 1,092 x 490,036} & CursedForest & 0.077 (XX)  &  &  \\
% phase1 chr22 &1,092 & 490,036 & CursedForest & 0.077 (XX)  & 9min 16sec & 1GB\footnote{\label{note10}10 nodes} \\
% &&& RF - Spark ML & 0.069 (0.89) & 14min 13sec & 1GB\footref{note10}  \\
% &&& RF - R &  0.05  & 24min  & 18GB\footref{note1}\footnote{using the \texttt{parallel} packages}\\
% &&& ranger -R &         0.05 &        1min 10sec  &          16GB  (one processor) \\
%  \hline
% phase3 chr22 & 2,504 & 1,103,548 &  &  &  &  \\
% &&& CursedForest & 0.053 & 9 min 8 sec & 2GB\footnote{\label{note20}20 nodes}\\
% &&& RF - Spark ML & 0.058 (0.92) & 1hr 1min & 2GB\footref{note20} \\
% &&& RF - R & NA  & NA & NA\footnote{long vectors ($> 2^31-1$)  are not supported in the .Fortran interface}\\
% &&&  H2O   &        0.05 &     16 hours       &   600GB \footnote{java -Xmx600g -jar h2o.jar,   h2o.init(nthreads=20)} \\
% \hline
% phase3 chr1 & 2,504 & 6,450,364 & CursedForest & 0.063 (XX) & 1hr 58min & 4GB \\
% &&& RF - Spark ML & (0.94) & 8hr 29min & 8GB \\
% &&& LR - Spark ML&  & 2hr 6min & 8GB\\ 
% &&& ranger  R   &  0.06   &     28 mins  &            num.threads = 10\footnote{don't know how to get the memory used, This
%   was done on dpas-03} \\ 
% \hline
% phase3 chr1-3 & 2,504 & 19,328,051  & CursedForest & 0.046 (XX) & 4hr 0min & 4GB\\
% &&& RF - Spark ML & - & -  & -\\
% &&& LR - Spark ML & & 14hr 48min & 8GB \\ 
% &&& ranger  R   &  0.05   &     58 mins\footnote{ but took 6 hours to load the data}  &      909GB\footnote{I think this is total memory. The job was run with
%   ntasks-per-node=160 cores-per-socket=10 on ruby }  \\ 
% \hline
% phase3 chr1-22 & 2,504 & 81,047,467 & CursedForest & XX (0.96) & 7hr 14min & 8GB \\ 
% &&& RF - Spark ML & - & - & - \\
% &&& LR - Spark ML & - & - & - \\ 
% %&&& kmeans - Spark ML & - (0.82) & 30hr 44min & 24GB \\ 
% \hline
\end{tabular}
\begin{flushleft} 
Comparison of different various machine learning libraries on different subsets of variant data 
from the 1000 Genomes Project.
Each random forest model consists of 50 trees.
\end{flushleft}
\label{table1}
%\end{adjustwidth}
\end{minipage}
\end{table}

\bibliographystyle{apalike}

%\bibliography{/home/dun280/Dropbox/TeX/bibliography/screening_compliance.bib}
\bibliography{/home/dun280/Dropbox/TeX/bibliography/ripley.bib}
\end{document}

### 2016-10-31 doc/publications/plos16/plos16.tex
changes to plos16.tex

@@
 \subsection{\cursedforest\ selects previously demonstrated cancer variants}
-%Feature selection can be useful for determining variables the contribute to the label or for perhaps selecting the n most important
-%variables for further analysis. 
-As established in the pervious section, \cursedforest\ allows the user to retrieve a subsection of features, ranked by
-their importance to identify the variables predictive of the label or selecting the $n$ most important variables for
-further analysis.  We demonstrate this feature on The Cancer Genome Atlas (TCGA) BRCA dataset.  We build this model on
-VCF files retrieved from the TCGA data portal.  For the label, we use the cell type (i.e. tumor or normal).
+We consider the question of feature discovery on the Cancer Genome Atlas (TCGA) BRCA dataset.  We build this model on
+VCF files retrieved from the TCGA data portal.  For the label, we use the cell type, tumor or normal.
 
@@
 \subsection{\cursedforest\ outperforms existing methods for classification problems}
-As established in Section~\ref{synthetic}, \cursedforest\ 
 In this section we compare \cursedforest 's performance against other published methods for predicting ethnicity on the
@@
                                                 & RF - R         & 0.05       & 24 min           & 18GB\footnote{using the \texttt{parallel} packages with 10 nodes} \\
-                                                & ranger - R      & 0.94       & 1 min 10 seconds & 16GB  (one processor)                                             \\
-                                                & ranger - c++     & 0.94       & 7 mins           & --nthreads 20                                                     \\
+                                                & ranger - R      & 0.06       & 1 min 10 seconds & 16GB  (one processor)                                             \\
+                                                & ranger - c++     & 0.06       & 7 mins           & --nthreads 20                                                     \\
 
@@
 
-In this section we explore the performance of \cursedforest\ by testing its ability to scale to different sizes of data and computational resources.
+In this section we explore the performance of \cursedforest\ by testing its ability to scale to different sizes of data
+and computational resources.
 
-In order to asses these characteristics we ran \cursedforest\ classification on synthetics datasets with varying numbers of variables (features) and samples, similar to the dataset used in to evaluate ranger [REF: ranger], allocating varying number of CPU cores to the \cursedforest\ and also varying computational complexity of random forests by using range of \mtry\ values.
+In order to asses these characteristics we ran \cursedforest\ classification on synthetics datasets with varying numbers
+of variables (features) and samples, similar to the dataset used in to evaluate ranger \cite{Wright.and.Ziegle.2016},
+allocating varying number of CPU cores to the \cursedforest\ and also varying computational complexity of random forests
+by using a range of \mtry\ values.
 
-We investigate the different synthetic datasets generated for section 5.1.
-and measured the time of building a random forest model of 100 trees and the results reported below are averages of 5 runs.
-To improve consistency of measurements all the cases were executed with the same random seed.
+We investigate the different synthetic datasets generated for section 5.1. and measured the time taken to build a random
+forest model of 100 trees. The results reported below are averages of 5 runs, and all the cases were executed with the
+same random seed, to improve consistency of measurements.
 
-Firstly we look at CurstedForest horizontal scalability for a medium size dataset of 2.5 million variables and 5000 samples, by varying \mtry\ fraction and the number of CPU cores allocated to the execution.
-The raw results are presented in the table [REF table1] and visualized in figure [REF: Fig 1] below
-Regardless of the number of cores used, \cursedforest\ displays approximately linear dependency between the execution time and the \mtry\ parameter (presented here as a fraction of the number of variables) [Fig 1 a)
+Firstly we look at CurstedForest horizontal scalability for a medium size dataset of 2.5 million variables and 5000
+samples, by varying \mtry\ fraction and the number of CPU cores allocated to the execution. The raw results are
+presented in table \ref{table10} and visualized in figure \ref{figure:synth} below. Regardless of the number of cores used,
+\cursedforest\ displays approximately linear dependency between the execution time and the \mtry\ parameter (presented
+here as a fraction of the number of variables) [Fig 1 a).
 
-\cursedforest\ scales almost linearly with the number of CPU cores for medium values of \mtry\ fraction but for both lower and higher values the performance degrades slightly (Fig 1 b).
-In the latter case the likely cause is communication overhead (with lower \mtry\ values the proportion of time for parallelizable computation to the time for internode communication is lower) while in the latter case its most likely caused by reaching the clusters computational capacity.
+\cursedforest\ scales almost linearly with the number of CPU cores for medium values of \mtry\ fraction but for both
+lower and higher values the performance degrades slightly (Fig 1 b). In the latter case the likely cause is
+communication overhead (with lower \mtry\ values the proportion of time for parallelizable computation to the time for
+internode communication is lower) while in the latter case its most likely caused by reaching the clusters computational
+capacity.
 
-Next we investigate \cursedforest\ scalability with regards to the size of data, by varying the number of variables and sample for a fixed \mtry\ fraction of 0.25 and execution of 128 CPU cores.
-The raw results are presented in the table [REF] and visualized in figure [ REF: Fig 2] below (please note log scale on the axes and the values on y axes are expresses as trees per hour).
+Next we investigate \cursedforest\ scalability with regards to the size of data, by varying the number of variables and
+sample for a fixed \mtry\ fraction of 0.25 and execution of 128 CPU cores. The raw results are presented in the table
+[REF] and visualized in figure [ REF: Fig 2] below (please note log scale on the axes and the values on y axes are
+expresses as trees per hour).
 
-Generally, the number of trees per hour decreases with increased number of variables and samples sizes.
-Some irregularities in the graph can be attributed to computation vs communication tradeoff.
-It is also worth noting that keeping \mtry\ fraction constants results with higher \mtry\ values with growing number of variables, and this is what drives the performance down rather that the increase of dataset size itself.
+Generally, the number of trees per hour decreases with increased number of variables and samples sizes. Some
+irregularities in the graph can be attributed to computation vs communication tradeoff. It is also worth noting that
+keeping \mtry\ fraction constants results with higher \mtry\ values with growing number of variables, and this is what
+drives the performance down rather that the increase of dataset size itself.
 
-To conclude \cursedforest\ is capable of processing datasets with 50M variables and 10K samples, which is the order of magnitude of the entire genome variants expected in such dataset, for which it can build approximately 60 trees per hour.
+To conclude \cursedforest\ is capable of processing datasets with 50M variables and 10K samples, which is the order of
+magnitude of the entire genome variants expected in such dataset, for which it can build approximately 60 trees per
+hour.
 
+\marginpar{can we remove this table and just rely on the figure?}
 \begin{table}[!ht]

### 2016-11-01 doc/publications/plos16/plos16.tex
changes to table in plos16.tex

@@
 \section{Introduction}
+\marginpar{
+we have parallelize and parallelise}
 
@@
 
-Here we introduce \cursedforest, a Hadoop/Spark based implementation of random forests specifically designed to cater for
-``big'' (many samples) and ``wide'' (many features) data sets. \cursedforest\ is capable of parallelising the split for
-each node in a tree thereby handling millions of features, such as in the 1000 Genomes Project data, which consists
-of approximately 2500 samples with up to 80 million variants~\cite{1KG2012}.  By also utilising Spark to read in and
-manipulate the standard genomic variant format (VCF) directly, \cursedforest\ outperforms existing tools even on small
-datasets where multi-threading generally performs well. 
- \marginpar{Piotr this is your 20 trees has the important
-  feature in the top 5 and 100 trees has it always as the top element, do we need to spell this out better, it should be
-  a result section.}
-Harnessing the virtually unlimited capability to parallelise tasks, \cursedforest\ can build a large number of trees,
-which boosts the accuracy for variable importance selection~\cite{Szymczak2016}.
-
-A Spark application runs on a ``driver'' node and distributes tasks to the many ``worker'' nodes, or ``executors''.
+Here we introduce \cursedforest, a Hadoop/Spark based implementation of random forests specifically designed to cater
+for ``big'' (many samples) and ``wide'' (many features) data sets. A Spark application runs on a ``driver'' node and
+distributes tasks to the many ``worker'' nodes, or ``executors''. Using this facility, \cursedforest\ is capable of
+parallelising the split for each node in a tree thereby handling millions of features, such as in the 1000 Genomes
+Project data, which consists of approximately 2500 samples with up to 80 million variants~\cite{1KG2012}.  By also
+utilising Spark to read in and manipulate the standard genomic variant format (VCF) directly, \cursedforest\ outperforms
+existing tools even on small datasets where multi-threading generally performs well. Harnessing the virtually unlimited
+capability to parallelise tasks, \cursedforest\ can build a large number of trees, which boosts the accuracy for
+variable importance selection~\cite{Szymczak2016}.
 
@@
 
-Here we extend this toolkit to encompass supervised learning using random forests.  Spark ML's random forest implementation is not
-able to handle the extremely ``wide'' genomic data as it was developed for large number of samples with only modest
-dimensionality.  Although Spark ML can build a random forest model on a subset of the data (chromosome 1~and~2), the time taken is
-excessive due to the huge amount of data being aggregated and processed by the single driver node during intermediate
-stages of building the model (see Result~\ref{comp}).  This unbalanced work load where the driver node becomes the
-bottleneck and worker nodes being idle prevents a seamless scaling to larger datasets. The memory requirements per
-executor also increases with dimensionality due to the data types Spark ML uses, which we elaborate on in the
-discussion.
+Here we extend this toolkit to encompass supervised learning using random forests.  Spark ML's random forest
+implementation is not able to handle the extremely ``wide'' genomic data as it was developed for large number of samples
+with only modest dimensionality.  Although Spark ML can build a random forest model on a subset of the data (chromosome
+1~and~2), 
+\margin[ar{something wrong here. We do not report on chromosome 1 and 2} 
+the time taken is excessive due to
+the large amount of data being aggregated and processed by the single driver node during intermediate stages of building
+the model (see Result~\ref{comp}).  This unbalanced work load where the driver node becomes the bottleneck and worker
+nodes being idle prevents a seamless scaling to larger datasets. The memory requirements per executor also increases
+with dimensionality due to the data types Spark ML uses, which we elaborate on in the discussion.
 
@@
 
-In the first section we explore some of the
-\marginpar{Needs to be rewritten once we actually have the sections} 
-properties of random forests, demonstrating that any potential flaws in \cursedforest\ are an inherent property of random
-forests. Here we provide demonstrations using synthetic data.  We then demonstrate the scalability of \cursedforest\ in
-respect to the dimensionality of the data, building a random forest model on whole-genome data from the 1000 Genomes Project. Here
-we report on the accuracy (in terms of the ARI)
- and the error.  Finally, given the role different parameter values can
-play in model construction, we explore the effect tuning these parameters can have on the prediction accuracy of the
-model.
+In the first section we explore some of the properties of random forests, demonstrating that any potential flaws in
+\cursedforest\ are an inherent property of random forests. Here we provide demonstrations using synthetic data.  We then
+demonstrate the scalability of \cursedforest\ in respect to the dimensionality of the data, building a random forest
+model on whole-genome data from the 1000 Genomes Project. Here we report on the accuracy (in terms of the ARI) 
+\marginpar{why not just error rate}
+and the error.  Finally, given the role different parameter values can play in model construction, we explore the effect tuning
+these parameters can have on the prediction accuracy of the model.
 
@@
   \item can handle missing values;
-  \item are insensitive to uninforamtive predictors;
+  \item are insensitive to uninformative predictors;
   \item are able to capture interactions between features (which is of importance for modelling complex polygenic diseases).
@@
 Random forests  apply the technique of bootstrap aggregation or bagging to decision trees.  The training
-data sets are independently drawn. with replacement,  from the joint distribution of $(X,Y)$ so that each set comprises $n$ $(p+1)$-tuples
- $(x_1,y_1),\ldots, (x_n,y_n)$. This si done $B$ times and  a tree is grown on each of these $B$ samples.
+data sets are independently drawn, with replacement,  from the joint distribution of $(X,Y)$ so that each set comprises $n$ $(p+1)$-tuples
+ $(x_1,y_1),\ldots, (x_n,y_n)$. This is done $B$ times and  a tree is grown on each of these $B$ samples.
 The results are combined in an appropriate way. In a classification problem we take the majority vote for the trees over
@@
 \begin{equation*}
-{{h_f}}= \arg \max_q \left(\sum_b I(h(x;\theta_b)=q)\right).
+{{h_f}}= \arg \max_q \left(\sum_b I(h(x;B_b)=q)\right).
 \end{equation*}
@@
 
-The average prediction error for an individual tree $h(X; \theta)$ is
+The average prediction error for an individual tree $h(X; B)$ is
 \marginpar{is this formulation correct in the case of classification? See Friedman on 0-1 losss}
 \begin{equation}
-PE_t = E_\theta E_{X,Y} (Y-h(X; \theta))^2.
+PE_t = E_B E_{X,Y} (Y-h(X; B))^2.
 \end{equation}
-Assume that, for all $\theta$, the tree is unbiased, i.e., $EY= E_X h(X; \theta)$. Then
+Assume that, for all $B$, the tree is unbiased, i.e., $EY= E_X h(X; B)$. Then
 \begin{equation}
@@
 \end{equation}
-where $\rho$ is the weighted correlation between residuals $(Y-h(X;\theta))$ and $(Y-h(X;\theta^\prime))$ for independent $\theta,
-\theta^\prime$.  
+where $\rho$ is the weighted correlation between residuals $(Y-h(X;B))$ and $(Y-h(X;B^\prime))$ for independent $B,B^\prime$.  
 So what is required for accurate random forest regression is (i) low correlation between residuals of differing tree in
@@
   \begin{itemize}
-  \item they give an out-of-bag estimate of model accuracy, by using the samples not selected for the bootstrapped sample
+  \item they give an out-of-bag (OOB) estimate of model accuracy, by using the samples not selected for the bootstrapped sample
     as an independent test set for each tree;
@@
 
-\cite{Strobl.et.al.2007} introduce changes to the random forest algorithm to fix 1) and 2).  However the level effect will be
-apparent in the CF algorithm as well.  However, there are many examples, particularly in genomics, where we have data
-sets that are
+\cite{Strobl.et.al.2007} introduce changes to the random forest algorithm to fix 1) and 2).  Effect 1) will be
+apparent in the \cursedforest algorithm as well.  However, there are many examples, particularly in genomics, where we have data
+sets that are:
 \begin{itemize}
@@
 \end{itemize}
+GWAs data sets are an obvious application areas for this algorithm.
 
@@
 rate that depends on the size of $S$ and not the number of noise variables in the data set. If the size of $S$ is small
-compared to the total number of variables, then the rate of convergence will be quicker. It is not clear waht the size
+compared to the total number of variables, then the rate of convergence will be quicker. It is not clear what the size
 of teh set $S$ will be in a GWAs application. There are cases of small groups of SNPs with large effectsd and also of
@@
 \subsection{\cursedforest.}
-VariantSpark made use of Spark's machine learning algorithms ``Spark ML''. These algorithms require data
-in a ``DataFrame'', similar to DataFrames in R.
-
-Traditionally, VariantSpark used the random forest algorithm from Spark ML. While Spark ML algorithms demonstrate
-scalability when dealing with a large number of samples, this scalability quickly breaks down as we include more
-features.
+VariantSpark made use of Spark's machine learning algorithms ``Spark ML''. These algorithms require data in a
+``DataFrame'', similar to DataFrames in R.  While Spark ML algorithms demonstrate some scalability when dealing with a large
+number of samples, this scalability eventuall break down as we include more features.
 
@@
 
-On the other hand, \cursedforest\ (CF) is specifically designed to handle wide ``cursed'' data. It avoids the relation
+On the other hand, \cursedforest\  is specifically designed to handle wide ``cursed'' data. It avoids the relation
 between memory and dimensionality by avoiding calculations that rely on entire feature vectors.
@@
 \subsubsection{1000 Genomes Project}
-We obtained the 1000 Genomes Project data as VCF files from their FTP site.  Each VCF file contains the variants for every individual accross one chromosome.  
-We use the phase 3 dataset which contains 2,504 individuals with approximately 81 million variant positions (i.e. a matrix of $2,504 \times 81$ million).  
-For compatibility with the majority of the algorithms we demonstrate, we convert the original variant representations to double representations.  
-For example, we store the absence of a variant (0\textbar0) as 0.0, a heterozygous variant (1\textbar0 or 0\textbar1)as 1.0. and a homozygous variant (1\textbar1) as 2.0.
+We obtained the 1000 Genomes Project data as VCF files from their FTP site.  Each VCF file contains the variants for
+every individual across one chromosome.  We use the phase 3 dataset which contains 2,504 individuals with approximately
+81 million variant positions (i.e. a matrix of $2,504 \times 81$ million).  For compatibility with the majority of the
+algorithms we demonstrate, we convert the original variant representations to double representations.  For example, we
+store the absence of a variant (0\textbar0) as 0, a heterozygous variant (1\textbar0 or 0\textbar1)as 1,  and a
+homozygous variant (1\textbar1) as 2.
 

### 2016-11-01 doc/publications/plos16/plos16.tex
a few changes

@@
 %
-% For tables that exceed the width of the text column, use the adjustwidth environment as illustrated in the example table in text below.
+% For tables that exceed the width of the text column, use the adjustwidth environment as illustrated in the example
+% table in text below.
 %
@@
 % IMPORTANT
-% Below are a few tips to help format your equations and other special characters according to our specifications.
-For more tips to help reduce the possibility of formatting errors during conversion, please see our LaTeX guidelines at http://journals.plos.org/plosone/s/latex
 %
-% For inline equations, please be sure to include all portions of an equation in the math environment.
-For example, x$^2$ is incorrect; this should be formatted as $x^2$ (or $\mathrm{x}^2$ if the romanized font is desired).
+% Below are a few tips to help format your equations and other special characters according to our specifications. For
+% more tips to help reduce the possibility of formatting errors during conversion, please see our LaTeX guidelines at
+% http://journals.plos.org/plosone/s/latex
 %
-% Do not include text that is not math in the math environment. For example, CO2 should be written as CO\textsubscript{2} instead of CO$_2$.
+% For inline equations, please be sure to include all portions of an equation in the math environment.  For example,
+% x$^2$ is incorrect; this should be formatted as $x^2$ (or $\mathrm{x}^2$ if the romanized font is desired).
+%
+% Do not include text that is not math in the math environment. For example, CO2 should be written as
+% CO\textsubscript{2} instead of CO$_2$.
 %
@@
 %
-% For inline equations, please do not include punctuation (commas, etc) within the math environment unless this is part of the equation.
+% For inline equations, please do not include punctuation (commas, etc) within the math environment unless this is part
+% of the equation.
 %
-% When adding superscript or subscripts outside of brackets/braces, please group using {}.  For example, change "[U(D,E,\gamma)]^2" to "{[U(D,E,\gamma)]}^2". 
+% When adding superscript or subscripts outside of brackets/braces, please group using {}.  For example, change
+% "[U(D,E,\gamma)]^2" to "{[U(D,E,\gamma)]}^2".
 %
@@
 \setlength{\marginparwidth}{5cm}
+\usepackage{caption}
+\usepackage{subcaption}
 
@@
 1~and~2), 
-\margin[ar{something wrong here. We do not report on chromosome 1 and 2} 
+\marginpar{something wrong here. We do not report on chromosome 1 and 2} 
 the time taken is excessive due to
@@
 of teh set $S$ will be in a GWAs application. There are cases of small groups of SNPs with large effectsd and also of
-contitions casued by the interaction of large groups of SNPs. 
+conditions caused by the interaction of large groups of SNPs. 
 \marginpar{check this}
@@
 \marginpar{ \cite{Chen.and.Ishwaran.2012} cite many papers on extensions of random forests to handle 
-GWAS data. What do we need to do in this area? What about linkage diseqelibrium?}
+GWAS data. What do we need to do in this area? What about linkage disequilibrium?}
 
@@
 ``DataFrame'', similar to DataFrames in R.  While Spark ML algorithms demonstrate some scalability when dealing with a large
-number of samples, this scalability eventuall break down as we include more features.
+number of samples, this scalability eventually break down as we include more features.
 
@@
 store the absence of a variant (0\textbar0) as 0, a heterozygous variant (1\textbar0 or 0\textbar1)as 1,  and a
-homozygous variant (1\textbar1) as 2.
+homozygous variant (1\textbar1) as 2.  This is not the only possible encoding \cite{Goldstein.et.al.2011} note that of
+the encoding listed in table \ref{table:encodings}, the additive encoding will present the most difficulty for a random
+forest model.
+
+\begin{table}[!ht]
+%\begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
+\begin{flushleft} 
+\centering
+\caption{\bf Possible encodings of GWAS data \cite[]{Goldstein.et.al.2011}}
+\begin{tabular}{|l|l|l|l||}
+\hline
+{\bf Type} & {\bf Mechanism}                                         & {\bf Partition} \\ 
+\thickhline
+Additive   & Each additional minor allele increases variation        & 0, 1, 2         \\
+Dominant   & Presence of at least 1 minor allele increases variation & 0, 1/2          \\
+Recessive  & Two minor alleles needed for variation                  & 0/1, 2          \\
+Heterosis  & Heterozygote leads to variation                         & 0/2, 1          \\ \hline
+%\end{adjustwidth}
+\end{tabular}
+\label{table:encodings}
+\end{flushleft}
+\end{table}
 
@@
 generated from uniform distribution with equal probabilities.  The dichotomous response is generated from a
-Bernoulli trail with $p$ (probability of success) calculated by applying an inverse logit transformation to the continuous
+Bernoulli trial with $p$ (probability of success) calculated by applying an inverse logit transformation to the continuous
 linear response ($y$) of five variables:
@@
 \cite{Liaw.and.Weiner.2002} which incorporates the original Fortran code by Brieman and Cutler.  We incorporate the
-advice of \cite{Liaw.and.Weiner.2002}, which we have found mirrors our own experience.
-
+advice of \cite{Liaw.and.Weiner.2002}, which we have found mirrors our own experience:
 \begin{itemize}
@@
   estimates of variable importance and proximity; however, even though the variable importance measures may vary from run to run,
-  the ranking of the importances is quite stable.  In addition, we note that it is possible for a random forest model to have a very poor
-  fit, and be useless for prediction, and still have a useful ranking of variable importance.
+  the ranking of the importances is quite stable.  In addition, we note that it is possible for a random forest model to
+  have a very poor
+\marginpar{is this true}
+  fit, and be useless for prediction, and still have a useful ranking of variable importance;
 \item \mtry\ -- the number of variables considered at each split (if \mtry=$p$, we have a boosted decision
   tree model).  If one has a very large number of variables but expects only very few to be ``important'', using larger \mtry\ may give
-  better performance.
+  better performance;
 \item the size and complexity of the individual trees is controlled in {\sc randomForest} by setting \texttt{nodesize}, the
@@
 method section.  Briefly we generate features by randomly sampling from a three value vector (1,2,3) with equal
-\marginpar{repicated methods section}
+\marginpar{replicated in methods section}
 probabilities, which mimics genomic mutation data and corresponds to homozygote reference, heterozygote or homozygote
@@
 complex trait of multiple interacting features.
+\marginpar{in the terminology of statistics, these are all main effects. There are no interaction terms}
 
@@
 5 feature in order of the weight that was assigned to them.  We are running Apache Spark 1.6.1 on a YARN cluster with 12
-worker nodes each with 16 Intel Xeon E5-2660@2.20GHz CPU cores and 128 GB of RAM.  We captured the OOB (out of bag)
+worker nodes each with 16 Intel Xeon E5-2660@2.20GHz CPU cores and 128 GB of RAM.  We captured the OOB 
 classification error and the 20 most important variables.
@@
 \marginpar{the accuracy hardly changes}
-Figure~\ref{figure:synth}a illustrates that increasing the number of predictors sampled for splitting at each node
+Fig.~\ref{figure:synth}a illustrates that increasing the number of predictors sampled for splitting at each node
 (\mtry) increases the accuracy with which features are ranked.  From $mtry=0.5$ the sort order of the features reflects
@@
 \marginpar{wrong figure cited}
-Figure~\ref{figure:synth}b illustrates that increasing \mtry\ also improves the OOB (out of bag) classification error.
+Fig.~\ref{figure:synth}b illustrates that increasing \mtry\ also improves the OOB (out of bag) classification error.
 %Piotr can you please explain the remaining error on the synthetic data
@@
 
-\begin{figure}[tbph]
-\begin{adjustwidth}{-1.00in}{0in}
-\begin{tabular}{ll}
-a)& b)\\
-\includegraphics[totalheight=6cm]{./figs/var_oder.png} & 
-\includegraphics[totalheight=6cm]{./figs/var_oob.png} \\
-\end{tabular}
-\caption{{\bf Variable importance (Mean Decrease Gini) and OOB (out of bag) classification error estimate for
-  random forests with 100 trees trained on the synthetic dataset of 2.5M features and 5K samples.}  {\bf a)} Variable
-  importance of top 20 variabels normalized to the sum of the linear model coefficients. {\bf Others} represents
-  aggregated importance of variables not present in the linear model.  {\bf b)} OOB (out of bag) classification error
-  estimate as funtion of \mtry\ fraction.}
+\marginpar{figure needs bigger text}
+\begin{figure}[tbhp] 
+  \begin{subfigure}[b]{0.5\linewidth}
+    \centering
+    \includegraphics[totalheight=6cm]{./figs/var_oder.png} 
+    \caption{Variable importance of top 20 variables.} 
+    \label{fig7:a} 
+    \vspace{4ex}
+  \end{subfigure}%% 
+  \begin{subfigure}[b]{0.5\linewidth}
+    \centering
+    \includegraphics[totalheight=6cm]{./figs/var_oob.png}
+    \caption{OOB classification error} 
+    \label{fig7:b} 
+    \vspace{4ex}
+  \end{subfigure} 
+\caption{{\bf Variable importance (Mean Decrease Gini) and OOB (out of bag) classification error estimate for random
+    forests with 100 trees trained on the synthetic dataset of 2.5M features and 5K samples.}  {\bf a)} Variable
+  importance of top 20 variables normalized to the sum of the linear model coefficients. {\bf Others} represents
+  aggregated importance of variables not present in the linear model. The dotted line represents the original linear
+  model coefficients.
+ {\bf b)} OOB (out of bag) classification error
+  estimate as function of \mtry\ fraction.}
   \label{figure:synth}
-\end{adjustwidth}
 \end{figure}
@@
 
-In Table~\ref{table3}, we display the top 10 features, ranked by their Gini index in the random forest model.
+In Table~\ref{table3}, we display the top 10 features, ranked by their Gini impurity measure in the random forest model.
+\marginpar{This is the Gini impurity measure, The Gini index or coefficient is an economic measure of income inequality}
 
@@
 \end{tabular}
-\begin{flushleft} Here the importance is the gini coefficient when building a model using the cell type as the label.
+\begin{flushleft} Here the importance is the Gini impurity measure when building a model using the cell type as the label.
 \end{flushleft}
@@
                                                     & H2O           & 0.05       & 16 hours         & 600GB \footnote{java -Xmx600g -jar h2o.jar,   h2o.init(nthreads=20)} \\
-                                                    & ranger - R    & 0.94       & 4 1/2 mins       & 21GB (one processor) \\
+                                                    & ranger - R    & 0.06       & 4 1/2 mins       & 21GB (one processor) \\
 \hline
@@
                                                     & Spark ML  & 0.94       & 8hr 29min        & 8GB \\
-                                                    & ranger - R      & 0.95       & 28 mins          & num.threads=10\footnote{don't
+                                                    & ranger - R      &  0.05       & 28 mins          & num.threads=10\footnote{don't
                                                                                                       know how to get the memory
@@
                                                     & RF - Spark ML  &            &                                                      & \\
-                                                    & ranger - R      &  0.97      & 58 mins\footnote{ but took 6 hours to load the
+                                                    & ranger - R      & 0.03      & 58 mins\footnote{ but took 6 hours to load the
                                                                                    data}
@@
 forest model of 100 trees. The results reported below are averages of 5 runs, and all the cases were executed with the
-same random seed, to improve consistency of measurements.
+same random seed, to improve the consistency of measurements.
 
@@
 communication overhead (with lower \mtry\ values the proportion of time for parallelizable computation to the time for
-internode communication is lower) while in the latter case its most likely caused by reaching the clusters computational
+internode communication is lower) while in the latter case it is most likely caused by reaching the clusters computational
 capacity.
@@
 
-\begin{table}[!ht]
-\caption{
-{\bf Sensitivity analysis on parameter choice.}}
-\begin{tabular}{|l|l|l|l|l|l|l|l|}
-\hline
-%\multicolumn{4}{|l|}{\bf Heading1} & \multicolumn{4}{|l|}{\bf Heading2}\\ \hline
-\bf{ \mtry\ }  & \bf{Mtree} & \bf{depth} & \bf{Accuracy} & \bf{Runtime} & \bf{Memory} \\
-\hline
-&&&&&\\ \hline
-\end{tabular}
-\begin{flushleft} 
-  Table notes here.
-\end{flushleft}
-\label{table2}
-\end{table}
+% \begin{table}[!ht]
+% \caption{
+% {\bf Sensitivity analysis on parameter choice.}}
+% \begin{tabular}{|l|l|l|l|l|l|l|l|}
+% \hline
+% %\multicolumn{4}{|l|}{\bf Heading1} & \multicolumn{4}{|l|}{\bf Heading2}\\ \hline
+% \bf{ \mtry\ }  & \bf{Mtree} & \bf{depth} & \bf{Accuracy} & \bf{Runtime} & \bf{Memory} \\
+% \hline
+% &&&&&\\ \hline
+% \end{tabular}
+% \begin{flushleft} 
+%   Table notes here.
+% \end{flushleft}
+% \label{table2}
+% \end{table}
 
@@
 %\end{table}
+
+% \begin{figure}[tbph]
+% \begin{adjustwidth}{-1.00in}{0in}
+% \begin{tabular}{ll}
+% a)& b)\\
+% \includegraphics[totalheight=6cm]{./figs/var_oder.png} & 
+% \includegraphics[totalheight=6cm]{./figs/var_oob.png} \\
+% \end{tabular}
+% \caption{{\bf Variable importance (Mean Decrease Gini) and OOB (out of bag) classification error estimate for random
+%     forests with 100 trees trained on the synthetic dataset of 2.5M features and 5K samples.}  {\bf a)} Variable
+%   importance of top 20 variables normalized to the sum of the linear model coefficients. {\bf Others} represents
+%   aggregated importance of variables not present in the linear model. The dotted line represents the original linear
+%   model coefficients.
+%  {\bf b)} OOB classification error
+%   estimate as function of \mtry\ fraction.}
+% \label{figure:synth}
+% \end{adjustwidth}
+% \end{figure}

### 2016-11-01 doc/publications/plos16/plos16.tex
a few changes

@@
 \marginpar{
-we have parallelize and parallelise}
+We have parallelize and parallelise.
+US spelling. Rob
+}
 
@@
 1~and~2), 
-\marginpar{something wrong here. We do not report on chromosome 1 and 2} 
+\marginpar{something wrong here. We do not report on chromosome 1 and 2.
+Aidan} 
 the time taken is excessive due to
@@
 model on whole-genome data from the 1000 Genomes Project. Here we report on the accuracy (in terms of the ARI) 
-\marginpar{why not just error rate}
 and the error.  Finally, given the role different parameter values can play in model construction, we explore the effect tuning
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\marginpar{Rob can you condense this down into what is relevant for this paper and not explained elsewhere in the literature}
 \subsection{Overview of Random Forests}
@@
 The average prediction error for an individual tree $h(X; B)$ is
-\marginpar{is this formulation correct in the case of classification? See Friedman on 0-1 losss}
+\marginpar{is this formulation correct in the case of classification? See Friedman on 0-1 losss, Rob}
 \begin{equation}
@@
 compared to the total number of variables, then the rate of convergence will be quicker. It is not clear what the size
-of teh set $S$ will be in a GWAs application. There are cases of small groups of SNPs with large effectsd and also of
+of the set $S$ will be in a GWAs application. There are cases of small groups of SNPs with large effectsd and also of
 conditions caused by the interaction of large groups of SNPs. 
-\marginpar{check this}
+\marginpar{check this. Denis}
 
@@
 \marginpar{ \cite{Chen.and.Ishwaran.2012} cite many papers on extensions of random forests to handle 
-GWAS data. What do we need to do in this area? What about linkage disequilibrium?}
+GWAS data. What do we need to do in this area? What about linkage disequilibrium? Denis}
 
@@
 \subsubsection{Synthetic data} 
-\marginpar{these are not actually encoded as factors}
-Each dataset consists of $n$ samples and $m$ variables where $n << m$, and values for each variable are ordinal factors with
+\marginpar{check details of simulation. Piotr}
+Each dataset consists of $n$ samples and $m$ variables where $n << m$, and values for each variable are ordinal variables with
 three levels represented as numbers $\{0, 1, 2\}$ (which correspond to an additive effect encoding of genomics variation) randomly
@@
   have a very poor
-\marginpar{is this true}
+\marginpar{is this true. If unsupported, remove it. Rob}
   fit, and be useless for prediction, and still have a useful ranking of variable importance;
@@
 alternative.  We then generate the response variable as a linear combination of 5 weighted features, modulating a
-complex trait of multiple interacting features.
-\marginpar{in the terminology of statistics, these are all main effects. There are no interaction terms}
+complex trait of multiple additive features.
 
@@
 
-\marginpar{the accuracy hardly changes}
 Fig.~\ref{figure:synth}a illustrates that increasing the number of predictors sampled for splitting at each node
@@
 
-\marginpar{wrong figure cited}
+\marginpar{wrong figures cited. Rob}
 Fig.~\ref{figure:synth}b illustrates that increasing \mtry\ also improves the OOB (out of bag) classification error.
@@
 By $mtry=0.8$ the error is converging to X, which is due to Y.
+\marginpar{what are ``X'' and ``Y''. Piotr}
 
@@
 \begin{figure}[tbhp] 
+ \begin{adjustwidth}{-1.00in}{0in}
   \begin{subfigure}[b]{0.5\linewidth}
@@
   \label{figure:synth}
+\end{adjustwidth}
 \end{figure}
@@
 \hline
-\bf{Genome \%}                      & \bf{Method} & \bf{Error (ARI)} & \bf{Runtime} & \bf{Memory/Node} \\
+\bf{Genome \%}                      & \bf{Method} & \bf{Error  & \bf{Runtime} & \bf{Memory/Node} \\
 \hline
 
-\multirow{3}{*}{phase1\_chr22: 1,092 x 490,036} & \cursedforest\ & 0.077 (XX) &                  &                                                                   \\
+\multirow{3}{*}{phase1\_chr22: 1,092 x 490,036} & \cursedforest\ & 0.077  &                  &                                                                   \\
                                                 & Spark ML  &            &                  &                                                                   \\
@@
 \hline
-\multirow{3}{*}{phase3\_chr1: 2,504 x 6,450,364}    & \cursedforest\ & 0.063 (XX) & 1hr 58min        & 4GB \\
+\multirow{3}{*}{phase3\_chr1: 2,504 x 6,450,364}    & \cursedforest\ & 0.063  & 1hr 58min        & 4GB \\
                                                     & Spark ML  & 0.94       & 8hr 29min        & 8GB \\
@@
 \hline
-\multirow{3}{*}{phase3\_chr1-3: 2,504 x 19,328,051} & \cursedforest\ & 0.046 (XX) & 4hr 0min                                             & 4GB \\
+\multirow{3}{*}{phase3\_chr1-3: 2,504 x 19,328,051} & \cursedforest\ & 0.046  & 4hr 0min                                             & 4GB \\
                                                     & RF - Spark ML  &            &                                                      & \\
@@
 hour.
-
-\marginpar{can we remove this table and just rely on the figure?}
-\begin{table}[!ht]
-\centering
-\caption{
-{\bf Time in seconds to build a random forest of 100
-  trees with specified \mtry\ fraction using given number of CPU cores.}}
-\begin{tabular}{|l|l|l|l|l|l|}
-\hline
-\multicolumn{2}{|l|}{\multirow{2}{*}{}}             & \multicolumn{4}{c|}{Number of CPU cores} \\
-\cline{3-6}
-\multicolumn{2}{|l|}{}                              & \bf{16} & \bf{32} & \bf{64} & \bf{128} \\
-\hline
-\multirow{4}{*}{\mtry\ (fraction)}        & \bf{0.10} & 708.3   & 340.0   & 186.0   & 134.5 \\
-                                        & \bf{0.25} & 1134.9  & 535.0   & 270.8   & 179.6 \\
-                                        & \bf{0.50} & 1724.1  & 856.6   & 428.7   & 243.8 \\
-                                        & \bf{0.80} & 2313.5  & 1125.7  & 569.1   & 364.6 \\
-\hline
-\end{tabular}
-\begin{flushleft} 
-\end{flushleft}
-\label{table10}
-\end{table}
 
@@
 % \end{figure}
+
+% \marginpar{can we remove this table and just rely on the figure?}
+% \begin{table}[!ht]
+% \centering
+% \caption{
+% {\bf Time in seconds to build a random forest of 100
+%   trees with specified \mtry\ fraction using given number of CPU cores.}}
+% \begin{tabular}{|l|l|l|l|l|l|}
+% \hline
+% \multicolumn{2}{|l|}{\multirow{2}{*}{}}             & \multicolumn{4}{c|}{Number of CPU cores} \\
+% \cline{3-6}
+% \multicolumn{2}{|l|}{}                              & \bf{16} & \bf{32} & \bf{64} & \bf{128} \\
+% \hline
+% \multirow{4}{*}{\mtry\ (fraction)}        & \bf{0.10} & 708.3   & 340.0   & 186.0   & 134.5 \\
+%                                         & \bf{0.25} & 1134.9  & 535.0   & 270.8   & 179.6 \\
+%                                         & \bf{0.50} & 1724.1  & 856.6   & 428.7   & 243.8 \\
+%                                         & \bf{0.80} & 2313.5  & 1125.7  & 569.1   & 364.6 \\
+% \hline
+% \end{tabular}
+% \begin{flushleft} 
+% \end{flushleft}
+% \label{table10}
+% \end{table}

### 2016-11-09 doc/publications/plos16/plos16.tex
fixed Gini coefficient
add new figures

@@
 \marginpar{figure needs bigger text}
+% \begin{figure}[tbhp] 
+%  \begin{adjustwidth}{-1.00in}{0in}
+%   \begin{subfigure}[b]{0.5\linewidth}
+%     \centering
+%     \includegraphics[totalheight=6cm]{./figs/var_oder.png} 
+%     \caption{Variable importance of top 20 variables.} 
+%     \label{fig7:a} 
+%     \vspace{4ex}
+%   \end{subfigure}%% 
+%   \begin{subfigure}[b]{0.5\linewidth}
+%     \centering
+%     \includegraphics[totalheight=6cm]{./figs/var_oob.png}
+%     \caption{OOB classification error} 
+%     \label{fig7:b} 
+%     \vspace{4ex}
+%   \end{subfigure} 
+% \caption{{\bf Variable importance (Mean Decrease Gini) and OOB (out of bag) classification error estimate for random
+%     forests with 100 trees trained on the synthetic dataset of 2.5M features and 5K samples.}  {\bf a)} Variable
+%   importance of top 20 variables normalized to the sum of the linear model coefficients. {\bf Others} represents
+%   aggregated importance of variables not present in the linear model. The dotted line represents the original linear
+%   model coefficients.
+%  {\bf b)} OOB (out of bag) classification error
+%   estimate as function of \mtry\ fraction.}
+%   \label{figure:synth}
+% \end{adjustwidth}
+% \end{figure}
+
 \begin{figure}[tbhp] 
@@
     \centering
-    \includegraphics[totalheight=6cm]{./figs/var_oder.png} 
-    \caption{Variable importance of top 20 variables.} 
+    \includegraphics[totalheight=8cm]{./figs/variable_importance.png} 
+    \caption{The sum of the variable importance of the 5 significant variables.} 
     \label{fig7:a} 
@@
     \centering
-    \includegraphics[totalheight=6cm]{./figs/var_oob.png}
+    \includegraphics[totalheight=8cm]{./figs/out-of-bag-prediction-error.png}
     \caption{OOB classification error} 
@@
   \end{subfigure} 
+  \begin{subfigure}[b]{0.5\linewidth}
+    \centering
+    \includegraphics[totalheight=8cm]{./figs/numnber_of_variables_found.png} 
+    \caption{The number of the 5 significant variables found in the top 5.} 
+    \label{fig7:a} 
+    \vspace{4ex}
+  \end{subfigure}%% 
+  \begin{subfigure}[b]{0.5\linewidth}
+    \centering
+    \includegraphics[totalheight=8cm]{./figs/rbo.png}
+    \caption{The  rank-biased-overlap measure.} 
+    \label{fig7:b} 
+    \vspace{4ex}
+  \end{subfigure} 
   \caption{{\bf Variable importance (Mean Decrease Gini) and OOB (out of bag) classification error estimate for random
-    forests with 100 trees trained on the synthetic dataset of 2.5M features and 5K samples.}  {\bf a)} Variable
-  importance of top 20 variables normalized to the sum of the linear model coefficients. {\bf Others} represents
-  aggregated importance of variables not present in the linear model. The dotted line represents the original linear
-  model coefficients.
- {\bf b)} OOB (out of bag) classification error
-  estimate as function of \mtry\ fraction.}
+      forests with 100 trees trained on the synthetic dataset of 2.5M features and 5K samples.}  Each measure is shown
+    as a function of \mtry and ntree. {\bf a)} The sum of the varible importances for the 5 significant variables.  {\bf
+      b)} OOB (out of bag) classification error. {\bf c)} the number of signifivant variables found in the top 5. 
+     {\bf d)} The rank-biased-overlap \cite[]{Webber.et.al.2010}. }
   \label{figure:synth}
@@
 In Table~\ref{table3}, we display the top 10 features, ranked by their Gini impurity measure in the random forest model.
-\marginpar{This is the Gini impurity measure, The Gini index or coefficient is an economic measure of income inequality}
 
@@
 \hline
-\bf{Genome \%}                      & \bf{Method} & \bf{Error  & \bf{Runtime} & \bf{Memory/Node} \\
+\bf{Genome \%}                      & \bf{Method} & \bf{Error}  & \bf{Runtime} & \bf{Memory/Node} \\
 \hline

### 2016-11-09 doc/publications/plos16/plos16.tex
changed to US spelling

@@
 \section{Introduction}
-\marginpar{
-We have parallelize and parallelise.
-US spelling. Rob
-}
-
 The digital revolution is seeing a dramatic increase in data collected about every aspect of life~\cite{Loebbecke2015}.
@@
 dimensionality''~\cite{Bauer2014, bellman1961adaptive}), decision trees are resistant to overfitting. 
-Secondly, random forests are also very easy to parallelise. As the forest is a sum of decision trees, it it possible to grow
+Secondly, random forests are also very easy to parallelize. As the forest is a sum of decision trees, it it possible to grow
 separate trees on different processors and combine the trees. See the
@@
 developed in other programming languages. {\sc CloudForest}~\cite{Bressler2015} written in Go, achieves fast run
-times by effective use of the CPU cache, optimising for different classes of features and efficiently multi-threading.
-Similarly, {\sc Ranger}~\cite{Wright.and.Ziegle.2016} written in C++ with an R front-end is specifically optimised for
+times by effective use of the CPU cache, optimizing for different classes of features and efficiently multi-threading.
+Similarly, {\sc Ranger}~\cite{Wright.and.Ziegle.2016} written in C++ with an R front-end is specifically optimized for
 dealing with large data sets.
@@
 However, the use of traditional compute infrastructure limits the parallelization strategies that can be employed by
-these methods.  The programs are limited to utilising only CPUs that are on the same computer node (multi-threading) or
+these methods.  The programs are limited to utilizing only CPUs that are on the same computer node (multi-threading) or
 else they run on CPUs distributed across nodes by virtue of the fact that there is no communication between the
@@
 distributes tasks to the many ``worker'' nodes, or ``executors''. Using this facility, \cursedforest\ is capable of
-parallelising the split for each node in a tree thereby handling millions of features, such as in the 1000 Genomes
+parallelization the split for each node in a tree thereby handling millions of features, such as in the 1000 Genomes
 Project data, which consists of approximately 2500 samples with up to 80 million variants~\cite{1KG2012}.  By also
-utilising Spark to read in and manipulate the standard genomic variant format (VCF) directly, \cursedforest\ outperforms
+utilizing Spark to read in and manipulate the standard genomic variant format (VCF) directly, \cursedforest\ outperforms
 existing tools even on small datasets where multi-threading generally performs well. Harnessing the virtually unlimited
-capability to parallelise tasks, \cursedforest\ can build a large number of trees, which boosts the accuracy for
+capability to parallelize tasks, \cursedforest\ can build a large number of trees, which boosts the accuracy for
 variable importance selection~\cite{Szymczak2016}.
@@
 We previously demonstrated the versatility and scalability of Spark by developing VariantSpark~\cite{OBrien2015}, a
-framework allowing users to easily analyse Variant Call Format (VCF) files using ML algorithms on the Spark framework.
+framework allowing users to easily analyze Variant Call Format (VCF) files using ML algorithms on the Spark framework.
 Using VariantSpark, we successfully built a k-means model on the full $2500 \times 80$ million matrix to cluster
@@
 comprehensive analysis toolkit that can scale for the future data demands, where the richness from WGS data can be
-utilised to ``fill in'' or impute the information at previously unobserved genomic positions in older array
+utilized to ``fill in'' or impute the information at previously unobserved genomic positions in older array
 technology~\cite{Howie2012}.  This provides the potential to impute samples in the GWAs catalogue and generate datasets
@@
 compared to the total number of variables, then the rate of convergence will be quicker. It is not clear what the size
-of the set $S$ will be in a GWAs application. There are cases of small groups of SNPs with large effectsd and also of
+of the set $S$ will be in a GWAs application. There are cases of small groups of SNPs with large effects and also of
 conditions caused by the interaction of large groups of SNPs. 
@@
       forests with 100 trees trained on the synthetic dataset of 2.5M features and 5K samples.}  Each measure is shown
-    as a function of \mtry and ntree. {\bf a)} The sum of the varible importances for the 5 significant variables.  {\bf
-      b)} OOB (out of bag) classification error. {\bf c)} the number of signifivant variables found in the top 5. 
+    as a function of \mtry and ntree. {\bf a)} The sum of the variable importances for the 5 significant variables.  {\bf
+      b)} OOB (out of bag) classification error. {\bf c)} the number of significant variables found in the top 5. 
      {\bf d)} The rank-biased-overlap \cite[]{Webber.et.al.2010}. }
@@
 
-Firstly we look at CurstedForest horizontal scalability for a medium size dataset of 2.5 million variables and 5000
+Firstly we look at CursedForest horizontal scalability for a medium size dataset of 2.5 million variables and 5000
 samples, by varying \mtry\ fraction and the number of CPU cores allocated to the execution. The raw results are

### 2016-11-09 doc/publications/plos16/plos16.tex
added conclusion

@@
 \begin{itemize}
-\item \ntree\ the number of trees.  The number of trees necessary for good performance grows with the number of predictors.
-\cite{Liaw.and.Weiner.2002} suggest that a high \ntree\ is necessary to get stable
-  estimates of variable importance and proximity; however, even though the variable importance measures may vary from run to run,
-  the ranking of the importances is quite stable.  In addition, we note that it is possible for a random forest model to
-  have a very poor
-\marginpar{is this true. If unsupported, remove it. Rob}
-  fit, and be useless for prediction, and still have a useful ranking of variable importance;
+\item \ntree\ the number of trees.  The number of trees necessary for good performance grows with the number of
+  predictors.  \cite{Liaw.and.Weiner.2002} suggest that a high \ntree\ is necessary to get stable estimates of variable
+  importance and proximity; however, even though the variable importance measures may vary from run to run, we note that
+  it is possible for a random forest model to have a very poor fit and still have a useful ranking of variable
+  importance;
 \item \mtry\ -- the number of variables considered at each split (if \mtry=$p$, we have a boosted decision
@@
 % \end{figure}
+\marginpar{this was done with n=100000. Needs to redone with n=2.5M}
 
@@
 \begin{minipage}{\textwidth}
-\begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
+%\begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
 \centering
@@
 {\bf Performance comparison between the different machine learning algorithms.}}
-\begin{tabular}{|l|l|l|l|l|l|p{1cm}|}
+\begin{tabular}{|l|l|l|l|l|l||}
 \hline
-\bf{Genome \%}                      & \bf{Method} & \bf{Error}  & \bf{Runtime} & \bf{Memory/Node} \\
+\bf{Genome \%}                      & \bf{Method} & \bf{Error}  & \bf{Runtime}  \\
 \hline
 
-\multirow{3}{*}{phase1\_chr22: 1,092 x 490,036} & \cursedforest\ & 0.077  &                  &                                                                   \\
-                                                & Spark ML  &            &                  &                                                                   \\
-                                                & RF - R         & 0.05       & 24 min           & 18GB\footnote{using the \texttt{parallel} packages with 10 nodes} \\
-                                                & ranger - R      & 0.06       & 1 min 10 seconds & 16GB  (one processor)                                             \\
-                                                & ranger - c++     & 0.06       & 7 mins           & --nthreads 20                                                     \\
+\multirow{3}{*}{phase1\_chr22: 1,092 x 490,036} & \cursedforest\ & 0.077  &                         \\
+                                                & Spark ML  &  Aidan           &           Aidan         \\
+                                                & RF - R         & 0.05       & 24 min           \\
+                                                & ranger - R      & 0.06       & 1 min 10 seconds       \\
+                                                & ranger - c++     & 0.06       & 7 mins            \\
 
  \hline
-\multirow{3}{*}{phase3\_chr22: 2,504 x 1,103,548}   &               &            &                  &                                                                \\
-                                                    & RF - R        & NA         & NA               & NA\footnote{long vectors ($> 2^31-1$)  are not supported in the .Fortran interface} \\
-                                                    & H2O           & 0.05       & 16 hours         & 600GB \footnote{java -Xmx600g -jar h2o.jar,   h2o.init(nthreads=20)} \\
-                                                    & ranger - R    & 0.06       & 4 1/2 mins       & 21GB (one processor) \\
+\multirow{3}{*}{phase3\_chr22: 2,504 x 1,103,548}   &               &            &                  \\
+                                                    & RF - R        & NA         & NA                \\
+                                                    & H2O           & 0.05       & 16 hours         \\
+                                                    & ranger - R    & 0.06       & 4 1/2 mins     \\
 \hline
-\multirow{3}{*}{phase3\_chr1: 2,504 x 6,450,364}    & \cursedforest\ & 0.063  & 1hr 58min        & 4GB \\
-                                                    & Spark ML  & 0.94       & 8hr 29min        & 8GB \\
-                                                    & ranger - R      &  0.05       & 28 mins          & num.threads=10\footnote{don't
-                                                                                                      know how to get the memory
-                                                                                                      used,   This was done on dpas-03} \\
+\multirow{3}{*}{phase3\_chr1: 2,504 x 6,450,364}    & \cursedforest\ & 0.063  & 1hr 58min         \\
+                                                    & Spark ML  & 0.94       & 8hr 29min        \\
+                                                    & ranger - R      &  0.05       & 28 mins         \\
 \hline
-\multirow{3}{*}{phase3\_chr1-3: 2,504 x 19,328,051} & \cursedforest\ & 0.046  & 4hr 0min                                             & 4GB \\
-                                                    & RF - Spark ML  &            &                                                      & \\
-                                                    & ranger - R      & 0.03      & 58 mins\footnote{ but took 6 hours to load the
-                                                                                   data}
-                                                                                                                                         &
-                                                                                                                                           909GB\footnote{I
-                                                                                                                                           think
-                                                                                                                                           this
-                                                                                                                                           is
-                                                                                                                                           total
-                                                                                                                                           memory. The
-                                                                                                                                           job
-                                                                                                                                           was
-                                                                                                                                           run
-                                                                                                                                           with
-                                                                                                                                                   ntasks-per-node=160
-                                                                                                                                           cores-per-socket=10
-                                                                                                                                           on
-                                                                                                                                           ruby
-                                                                                                                                           }  \\
-& ranger - R        &             0.973     &      03:58           &       pearcey --cpus-per-task=15 --mem 1600gb   906GB \\
+\multirow{3}{*}{phase3\_chr1-3: 2,504 x 19,328,051} & \cursedforest\ & 0.046  & 4hr 0min               \\
+                                                    & RF - Spark ML  &    Aidan         &   Aidan               \\
+                                                    & ranger - R      & 0.03      &         \\
+& ranger - R        &             0.973     &      03:58           \\
 \hline
 
-\multirow{4}{*}{phase3\_chr1-22: 2,504 x 81,047,467} & \cursedforest\ & XX (0.96) & 7hr 14min & 8GB \\
-& Spark ML & - & - & - \\
-& ranger (C++)       &        NA     &        NA     &            80 CPU -- runs out of memory (12.8GB per CPU) \\
-& ranger (C++)       &        NA     &        NA     &            160
-                                                        CPU -- runs
-                                                        out of time at
-                                                        20 hours \\
+\multirow{4}{*}{phase3\_chr1-22: 2,504 x 81,047,467} & \cursedforest\ & XX  & 7hr 14min \\
+& Spark ML & Aidan & Aidan  \\
+& ranger (C++)       &        NA     &        NA      \\
+& ranger (C++)       &        NA     &        NA    \\
 \hline
@@
 \label{table1}
-\end{adjustwidth}
+%\end{adjustwidth}
 \end{minipage}
@@
 
+We have demonstrated that using a different model of paralleization can extend the random forest to delaing with the
+case of an extremely large number of variables. By comparing this implementation to other implentations (including those
+optimized for large data sets) we have demonstrated the utility of this implentation.
+
+We have treated the case of variable selection in a $p >> n$ model where most of the variables are uninformative. We
+have also demostrated the utility of the model for large GWAs data sets.
+
+\marginpar{signal rather than ``significant variables''}
+
+Donoho and Tanner \cite{Donoho.and.Tanner.2009} give a ``universal phase change'' result based on arguments from
+combinatorial geometry. This has applications in a large number of areas including variable selection in high
+diensions. Consider figure \ref{figure:phase-diagram-equivalence.png} which shows the region where a model can recover
+the important variables, plotted as a function of $\delta = n/p$ and $\rho =k/n$ (where $k$ is the number of significant
+variables). There is a distinct doundary, shown empirically and also based on arguments from combinatorial geometry, to the
+region where we can reliably recover significant variables.
+
+We note that the examples we are considering lie on the extreme left of the plot, close to the origin. So there are
+limits to what can be recovered in the presence of noise. However in the case of data that is both big and wide, \cursedforest and
+other VariantSpark methods may provide a useful tool.
+
+\marginpar{redo plot}
+
+\begin{figure}[tbhp]
+\includegraphics[totalheight=6cm]{./figs/phase-diagram-equivalence.png} 
+\caption{{\bf The phase chage diagram.}}
+\label{figure:phase-diagram-equivalence.png} 
+\end{figure}
+
 \section{Supporting Information}
@@
 \end{document}
-
-%  \item D\iaz-Uriarte, R., \& Alvarez de Andr\es, S. (2006). Gene selection and classification of microarray data using random
-%    forest. BMC Bioinformatics, 7, 3. http://doi.org/10.1186/1471-2105-7-3
-%  \item Goldstein, B. A., Polley, E. C., & Briggs, F. B. S. (2011). Random Forests for Genetic Association Studies. Statistical
-%    Applications in Genetics and Molecular Biology, 10(1), 32. http://doi.org/10.2202/1544-6115.1691
-
-%http://stats.stackexchange.com/questions/77018/random-forest-is-it-a-boosting-algorithm
- %http://stats.stackexchange.com/questions/173390/gradient-boosting-tree-vs-random-forest
-%overcomes this issue and is hence ideally suited to this application.  Furthermore, random forests are able to capture interactions
-%between features, which is of importance for modelling complex polygenic diseases.
-
-% We also investigated classifying the samples using random forests.  We chose random forests due to its ability to
-% capture complex interactions between features. Also, due to it being an ensemble method where trees can be formed
-% independently, which, is ideal for parallelization. Furthermore, the ensemble of weak learners can help to minimise
-% overfitting and variance.
-
-% The reason for the failure can be attributed to the "curse of dimensionality". Where Spark ML is designed to process a
-% huge number of samples with a modest dimensionality, the 1000 Genomes data instead contains a huge number of
-% dimensions (features). As a result, the default random forests implementation is unable to partition the data into
-% small-enough tasks for nodes on the computer cluster to handle.
-
-% For figure citations, please use "Fig." instead of "Figure".
-%Nulla mi mi, Fig.~\ref{fig1}  Fusce fringilla erat porttitor lectus cursus, \nameref{S1_Video} vel sagittis arcu lobortis.
-Aliquam in enim semper, aliquam massa id, cursus neque.
-Praesent faucibus semper libero.
-%\begin{figure}[h]
-%\caption{{\bf Figure Title first bold sentence Nulla mi mi, venenatis sed ipsum varius, volutpat euismod diam.}
-%Figure Caption Proin rutrum vel massa non gravida. Quisque tempor sem et dignissim rutrum. A: Lorem ipsum dolor sit amet. B: Consectetur adipiscing elit.}
-%\label{fig1}
-%\end{figure}
-%\begin{enumerate}
-%\item{react}
-%\item{diffuse free particles}
-%\item{increment time by dt and go to 1}
-%\end{enumerate}
-
-% Results and Discussion can be combined.
-
-%\begin{table}[!ht]
-%\begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
-%\caption{
-%{\bf Table caption Nulla mi mi, venenatis sed ipsum varius, volutpat euismod diam.}}
-%\begin{tabular}{|l|l|l|l|l|l|l|l|}
-%\hline
-%\multicolumn{4}{|l|}{\bf Heading1} & \multicolumn{4}{|l|}{\bf Heading2}\\ \hline
-%$cell1 row1$ & cell2 row 1 & cell3 row 1 & cell4 row 1 & cell5 row 1 & cell6 row 1 & cell7 row 1 & cell8 row 1\\ \hline
-%$cell1 row2$ & cell2 row 2 & cell3 row 2 & cell4 row 2 & cell5 row 2 & cell6 row 2 & cell7 row 2 & cell8 row 2\\ \hline
-%$cell1 row3$ & cell2 row 3 & cell3 row 3 & cell4 row 3 & cell5 row 3 & cell6 row 3 & cell7 row 3 & cell8 row 3\\ \hline
-%\end{tabular}
-%\begin{flushleft} Table notes Phasellus venenatis, tortor nec vestibulum mattis, massa tortor interdum felis, nec pellentesque metus tortor nec nisl.
-Ut ornare mauris tellus, vel dapibus arcu suscipit sed.
-%\end{flushleft}
-%\label{table1}
-%\end{adjustwidth}
-%\end{table}
-
-% \begin{figure}[tbph]
-% \begin{adjustwidth}{-1.00in}{0in}
-% \begin{tabular}{ll}
-% a)& b)\\
-% \includegraphics[totalheight=6cm]{./figs/var_oder.png} & 
-% \includegraphics[totalheight=6cm]{./figs/var_oob.png} \\
-% \end{tabular}
-% \caption{{\bf Variable importance (Mean Decrease Gini) and OOB (out of bag) classification error estimate for random
-%     forests with 100 trees trained on the synthetic dataset of 2.5M features and 5K samples.}  {\bf a)} Variable
-%   importance of top 20 variables normalized to the sum of the linear model coefficients. {\bf Others} represents
-%   aggregated importance of variables not present in the linear model. The dotted line represents the original linear
-%   model coefficients.
-%  {\bf b)} OOB classification error
-%   estimate as function of \mtry\ fraction.}
-% \label{figure:synth}
-% \end{adjustwidth}
-% \end{figure}
-
-% \marginpar{can we remove this table and just rely on the figure?}
-% \begin{table}[!ht]
-% \centering
-% \caption{
-% {\bf Time in seconds to build a random forest of 100
-%   trees with specified \mtry\ fraction using given number of CPU cores.}}
-% \begin{tabular}{|l|l|l|l|l|l|}
-% \hline
-% \multicolumn{2}{|l|}{\multirow{2}{*}{}}             & \multicolumn{4}{c|}{Number of CPU cores} \\
-% \cline{3-6}
-% \multicolumn{2}{|l|}{}                              & \bf{16} & \bf{32} & \bf{64} & \bf{128} \\
-% \hline
-% \multirow{4}{*}{\mtry\ (fraction)}        & \bf{0.10} & 708.3   & 340.0   & 186.0   & 134.5 \\
-%                                         & \bf{0.25} & 1134.9  & 535.0   & 270.8   & 179.6 \\
-%                                         & \bf{0.50} & 1724.1  & 856.6   & 428.7   & 243.8 \\
-%                                         & \bf{0.80} & 2313.5  & 1125.7  & 569.1   & 364.6 \\
-% \hline
-% \end{tabular}
-% \begin{flushleft} 
-% \end{flushleft}
-% \label{table10}
-% \end{table}

### 2016-11-09 doc/publications/plos16/supplementary_information.tex
added conclusion

@@
 
+\begin{table}[!ht]
+\begin{minipage}{\textwidth}
+\begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
+\centering
+\caption{
+{\bf Performance comparison between the different machine learning algorithms.}}
+\begin{tabular}{|l|l|l|l|l|l|p{1cm}|}
+\hline
+\bf{Genome \%}                      & \bf{Method} & \bf{Error}  & \bf{Runtime} & \bf{Memory/Node} \\
+\hline
+
+\multirow{3}{*}{phase1\_chr22: 1,092 x 490,036} & \cursedforest\ & 0.077  &                  &                                                                   \\
+                                                & Spark ML  &            &                  &                                                                   \\
+                                                & RF - R         & 0.05       & 24 min           & 18GB\footnote{using the \texttt{parallel} packages with 10 nodes} \\
+                                                & ranger - R      & 0.06       & 1 min 10 seconds & 16GB  (one processor)                                             \\
+                                                & ranger - c++     & 0.06       & 7 mins           & --nthreads 20                                                     \\
+
+ \hline
+\multirow{3}{*}{phase3\_chr22: 2,504 x 1,103,548}   &               &            &                  &                                                                \\
+                                                    & RF - R        & NA         & NA               & NA\footnote{long vectors ($> 2^31-1$)  are not supported in the .Fortran interface} \\
+                                                    & H2O           & 0.05       & 16 hours         & 600GB \footnote{java -Xmx600g -jar h2o.jar,   h2o.init(nthreads=20)} \\
+                                                    & ranger - R    & 0.06       & 4 1/2 mins       & 21GB (one processor) \\
+\hline
+\multirow{3}{*}{phase3\_chr1: 2,504 x 6,450,364}    & \cursedforest\ & 0.063  & 1hr 58min        & 4GB \\
+                                                    & Spark ML  & 0.94       & 8hr 29min        & 8GB \\
+                                                    & ranger - R      &  0.05       & 28 mins          & num.threads=10\footnote{don't
+                                                                                                      know how to get the memory
+                                                                                                      used,   This was done on dpas-03} \\
+\hline
+\multirow{3}{*}{phase3\_chr1-3: 2,504 x 19,328,051} & \cursedforest\ & 0.046  & 4hr 0min                                             & 4GB \\
+                                                    & RF - Spark ML  &            &                                                      & \\
+                                                    & ranger - R      & 0.03      & 58 mins\footnote{ but took 6 hours to load the
+                                                                                   data}
+                                                                                                                                         &
+                                                                                                                                           909GB\footnote{I
+                                                                                                                                           think
+                                                                                                                                           this
+                                                                                                                                           is
+                                                                                                                                           total
+                                                                                                                                           memory. The
+                                                                                                                                           job
+                                                                                                                                           was
+                                                                                                                                           run
+                                                                                                                                           with
+                                                                                                                                                   ntasks-per-node=160
+                                                                                                                                           cores-per-socket=10
+                                                                                                                                           on
+                                                                                                                                           ruby
+                                                                                                                                           }  \\
+& ranger - R        &             0.973     &      03:58           &       pearcey --cpus-per-task=15 --mem 1600gb   906GB \\
+\hline
+
+\multirow{4}{*}{phase3\_chr1-22: 2,504 x 81,047,467} & \cursedforest\ & XX (0.96) & 7hr 14min & 8GB \\
+& Spark ML & - & - & - \\
+& ranger (C++)       &        NA     &        NA     &            80 CPU -- runs out of memory (12.8GB per CPU) \\
+& ranger (C++)       &        NA     &        NA     &            160
+                                                        CPU -- runs
+                                                        out of time at
+                                                        20 hours \\
+\hline
+\end{tabular}
+\begin{flushleft} 
+Comparison of different various machine learning libraries on different subsets of variant data 
+from the 1000 Genomes Project.
+Each random forest model consists of 50 trees.
+\end{flushleft}
+\label{table1}
+\end{adjustwidth}
+\end{minipage}
+\end{table}
+
 \bibliographystyle{apalike}

### 2016-11-13 doc/publications/plos16/plos16.tex
round of changes to plos16.tex

@@
 \section{Introduction}
-The digital revolution is seeing a dramatic increase in data collected about every aspect of life~\cite{Loebbecke2015}.
+The digital revolution is seeing a dramatic increase in data collected about almost every aspect of life~\cite{Loebbecke2015}.
 These data sets are not only growing vertically, by capturing more events, but also horizontally by capturing
@@
 demonstrated that not taking joint interactions between SNPs into account when filtering for potential disease mutations
-is inadequate~\cite{Manolio2009}.  
+is inadequate~\cite{Manolio2009}.  \marginpar{do  these two sentences say teh same thing?}
 Complex trait analysis has demonstrated that in order to explain the overall
@@
 distributes tasks to the many ``worker'' nodes, or ``executors''. Using this facility, \cursedforest\ is capable of
-parallelization the split for each node in a tree thereby handling millions of features, such as in the 1000 Genomes
+parallelization the split for each node in a tree thereby handling millions of features. It can, for example, fit a
+model to the 1000 Genomes
 Project data, which consists of approximately 2500 samples with up to 80 million variants~\cite{1KG2012}.  By also
@@
 the large amount of data being aggregated and processed by the single driver node during intermediate stages of building
-the model (see Result~\ref{comp}).  This unbalanced work load where the driver node becomes the bottleneck and worker
+the model (see Result~\ref{comp}). 
+\marginpar{missing ref}
+ This unbalanced work load where the driver node becomes the bottleneck and worker
 nodes being idle prevents a seamless scaling to larger datasets. The memory requirements per executor also increases
@@
 
-In the first section we explore some of the properties of random forests, demonstrating that any potential flaws in
-\cursedforest\ are an inherent property of random forests. Here we provide demonstrations using synthetic data.  We then
-demonstrate the scalability of \cursedforest\ in respect to the dimensionality of the data, building a random forest
-model on whole-genome data from the 1000 Genomes Project. Here we report on the accuracy (in terms of the ARI) 
-and the error.  Finally, given the role different parameter values can play in model construction, we explore the effect tuning
-these parameters can have on the prediction accuracy of the model.
+In the first section we explore some of the properties of random forests as these will also be apparent in the \cursedforest\
+implementation. Here we provide demonstrations using synthetic data.  We then demonstrate the scalability of
+\cursedforest\ in respect to the dimensionality of the data, building a random forest model on whole-genome data from
+the 1000 Genomes Project. Here we report on the accuracy (in terms of the ARI) and the error.  Finally, given the role
+different parameter values can play in model construction, we explore the effect tuning these parameters can have on the
+prediction accuracy of the model.
 
@@
  $(x_1,y_1),\ldots, (x_n,y_n)$. This is done $B$ times and  a tree is grown on each of these $B$ samples.
-The results are combined in an appropriate way. In a classification problem we take the majority vote for the trees over
+The results are then combined in an appropriate way. In a classification problem we take the majority vote for the trees over
 the $q=1,..,Q$ classes,
@@
   \item they inherit some of the qualities of tree models (but not, for example, the handling of missing values via surrogate
-    variables. This would be difficult as the trees as split on random subsets of variables).
+    variables. This would be difficult as the trees are split on random subsets of variables).
   \end{itemize}
@@
 increasingly large vectors must be loaded into memory.
+\marginpar{not clear. Are RDDs the problem or the solutuion}
 
@@
 
-The current implementation of \cursedforest\ uses:
-\begin{itemize}
-\item an ``information gained'' criteria for splitting. Let $f_q$ be the fraction of items labelled with value $q$ where $q = 1,
-  \ldots, Q$ at a node. The entropy of the node is $H = - \sum^{m}_{i=1} f_i \log^{}_2 f_i.$
-The expected information gain is the change in entropy $H$ from a prior state $T$ to a state that takes some
-information $a$ as given: 
+The current implementation of \cursedforest\ uses an ``information gained'' criteria for splitting. Let $f_q$ be the
+fraction of items labelled with value $q$ where $q = 1, \ldots, Q$ at a node. The entropy of the node is $H = -
+\sum^{m}_{i=1} f_i \log^{}_2 f_i.$ The expected information gain is the change in entropy $H$ from a prior state $T$ to
+a state that takes some information $a$ as given:
 \[ IG(T,a) = H(T) - H(T|a). \]
-In this case, the information gain is IG = Entropy(parent node) - Weighted Sum of Entropy(Children)
-\item variable importance is the sum of IG over all notes and all trees for that variable.
-\end{itemize}
+In this case, the information gain is IG = Entropy(parent node) - Weighted Sum of Entropy(Children). 
+Variable importance is the sum of IG over all notes and all trees for that variable.
 
@@
 81 million variant positions (i.e. a matrix of $2,504 \times 81$ million).  For compatibility with the majority of the
-algorithms we demonstrate, we convert the original variant representations to double representations.  For example, we
+algorithms we demonstrate, we convert the original variant representations to double representations. 
+\marginpar{are they stored a double or integer?}
+ For example, we
 store the absence of a variant (0\textbar0) as 0, a heterozygous variant (1\textbar0 or 0\textbar1)as 1,  and a
-homozygous variant (1\textbar1) as 2.  This is not the only possible encoding \cite{Goldstein.et.al.2011} note that of
+homozygous variant (1\textbar1) as 2.  This is not the only possible encoding.  \cite{Goldstein.et.al.2011} note that of
 the encoding listed in table \ref{table:encodings}, the additive encoding will present the most difficulty for a random
@@
 \marginpar{check details of simulation. Piotr}
-Each dataset consists of $n$ samples and $m$ variables where $n << m$, and values for each variable are ordinal variables with
+Each dataset consists of $n$ samples and $p$ variables where $n << p$, and values for each variable are ordinal variables with
 three levels represented as numbers $\{0, 1, 2\}$ (which correspond to an additive effect encoding of genomics variation) randomly
@@
 $$y = a_{10}* v_{10} + a_{100}*v_{100} + a_{1000}*v_{1000} + a_{10000}*v_{10000} + a_{10000}*v_{100000}$$
-where $a_{10} = 0.8, a_{100} = 0.9, a_{1000} = 1.1, a_{10000} = 1.0, a_{10000}=1.15$.
+where $a_{10} = 1.6, a_{100} = 1.8, a_{1000} = 2.2, a_{10000} = 2.0, a_{10000}=2.3$.
 
 \subsection{Parameter settings}
-We consider the parameter setting for the random forest algorithm. We use the R terms from the {\sc randomForest} package
+We consider the parameter setting for the random forest algorithm. We use the R notation from the {\sc randomForest} package
 \cite{Liaw.and.Weiner.2002} which incorporates the original Fortran code by Brieman and Cutler.  We incorporate the
@@
   importance and proximity; however, even though the variable importance measures may vary from run to run, we note that
-  it is possible for a random forest model to have a very poor fit and still have a useful ranking of variable
+  it is possible for a random forest model to have a very poor fit and still have an accurate ranking of variable
   importance;
@@
 We therefore generate a synthetic dataset with 2.5 million variables (features) and 5000 samples, as discussed in the
-method section.  Briefly we generate features by randomly sampling from a three value vector (1,2,3) with equal
-\marginpar{replicated in methods section}
-probabilities, which mimics genomic mutation data and corresponds to homozygote reference, heterozygote or homozygote
-alternative.  We then generate the response variable as a linear combination of 5 weighted features, modulating a
-complex trait of multiple additive features.
+method section.
 
@@
     \centering
-    \includegraphics[totalheight=8cm]{./figs/variable_importance.png} 
-    \caption{The sum of the variable importance of the 5 significant variables.} 
-    \label{fig7:a} 
-    \vspace{4ex}
-  \end{subfigure}%% 
-  \begin{subfigure}[b]{0.5\linewidth}
-    \centering
     \includegraphics[totalheight=8cm]{./figs/out-of-bag-prediction-error.png}
@@
   \end{subfigure} 
-  \begin{subfigure}[b]{0.5\linewidth}
-    \centering
-    \includegraphics[totalheight=8cm]{./figs/numnber_of_variables_found.png} 
-    \caption{The number of the 5 significant variables found in the top 5.} 
-    \label{fig7:a} 
-    \vspace{4ex}
-  \end{subfigure}%% 
   \begin{subfigure}[b]{0.5\linewidth}
@@
       forests with 100 trees trained on the synthetic dataset of 2.5M features and 5K samples.}  Each measure is shown
-    as a function of \mtry and ntree. {\bf a)} The sum of the variable importances for the 5 significant variables.  {\bf
-      b)} OOB (out of bag) classification error. {\bf c)} the number of significant variables found in the top 5. 
-     {\bf d)} The rank-biased-overlap \cite[]{Webber.et.al.2010}. }
+    as a function of \mtry and \ntree. {\bf a0} OOB (out of bag) classification error.  {\bf b)} The rank-biased-overlap \cite[]{Webber.et.al.2010}. }
   \label{figure:synth}

### 2016-11-17 doc/publications/plos16/plos16.tex
a few table numbers

@@
 
-\multirow{3}{*}{phase1\_chr22: 1,092 x 490,036} & \cursedforest\ & 0.077  &                         \\
-                                                & Spark ML  &  Aidan           &           Aidan         \\
+\multirow{3}{*}{phase1\_chr22: 1,092 x 490,036} & \cursedforest\ & 0.04  & 1 min 50 seconds                        \\
+                                                & RF - Spark ML  &  Aidan           &           Aidan         \\
                                                 & RF - R         & 0.05       & 24 min           \\
@@
                                                     & H2O           & 0.05       & 16 hours         \\
-                                                    & ranger - R    & 0.06       & 4 1/2 mins     \\
+                                                    & ranger - R    & 0.06       & 4 min 30 seconds     \\
 \hline
 \multirow{3}{*}{phase3\_chr1: 2,504 x 6,450,364}    & \cursedforest\ & 0.063  & 1hr 58min         \\
-                                                    & Spark ML  & 0.94       & 8hr 29min        \\
+                                                    & RF - Spark ML  & 0.04       & 8hr 29min        \\
                                                     & ranger - R      &  0.05       & 28 mins         \\
@@
 \multirow{3}{*}{phase3\_chr1-3: 2,504 x 19,328,051} & \cursedforest\ & 0.046  & 4hr 0min               \\
-                                                    & RF - Spark ML  &    Aidan         &   Aidan               \\
+                                                    & RF - Spark ML  &    NA         &   NA               \\
                                                     & ranger - R      & 0.03      &         \\
@@
 \multirow{4}{*}{phase3\_chr1-22: 2,504 x 81,047,467} & \cursedforest\ & XX  & 7hr 14min \\
-& Spark ML & Aidan & Aidan  \\
+& Spark ML & NA & NA  \\
 & ranger (C++)       &        NA     &        NA      \\

### 2016-11-24 doc/publications/plos16/plos16.tex
Add missing time

@@
 \multirow{3}{*}{phase1\_chr22: 1,092 x 490,036} & \cursedforest\ & 0.04  & 1 min 50 seconds                        \\
-                                                & RF - Spark ML  &  Aidan           &           Aidan         \\
+                                                & RF - Spark ML  &  0.05           &           7 min 44 seconds         \\
                                                 & RF - R         & 0.05       & 24 min           \\
                                                 & ranger - R      & 0.06       & 1 min 10 seconds       \\
-                                                & ranger - c++     & 0.06       & 7 mins            \\
+                                                & ranger - c++     & 0.06       & 7 min            \\
 

### 2016-11-24 doc/publications/plos16/plos16.tex
Fix ref to chr1,2 and verbosify table names

@@
 with only modest dimensionality.  Although Spark ML can build a random forest model on a subset of the data (chromosome
-1~and~2), 
-\marginpar{something wrong here. We do not report on chromosome 1 and 2.
-Aidan} 
-the time taken is excessive due to
-the large amount of data being aggregated and processed by the single driver node during intermediate stages of building
-the model (see Result~\ref{comp}). 
+1), the time taken is excessive due to the large amount of data being aggregated and processed by the driver node
+during intermediate stages of building the model (see Table~\ref{timingtable}). 
 \marginpar{missing ref}
@@
 
-In Table~\ref{table3}, we display the top 10 features, ranked by their Gini impurity measure in the random forest model.
+In Table~\ref{featuretable}, we display the top 10 features, ranked by their Gini impurity measure in the random forest model.
 
@@
 \end{flushleft}
-\label{table3}
+\label{featuretable}
 %\end{adjustwidth}
@@
 \end{flushleft}
-\label{table1}
+\label{timingtable}
 %\end{adjustwidth}
@@
 samples, by varying \mtry\ fraction and the number of CPU cores allocated to the execution. The raw results are
-presented in table \ref{table10} and visualized in figure \ref{figure:synth} below. Regardless of the number of cores used,
+presented in table \ref{synthetictimingtable} and visualized in figure \ref{figure:synth} below. Regardless of the number of cores used,
 \cursedforest\ displays approximately linear dependency between the execution time and the \mtry\ parameter (presented
@@
 \end{flushleft}
-\label{table10}
+\label{synthetictimingtable}
 \end{adjustwidth}

### 2016-11-26 doc/publications/plos16/plos16.tex
changes

@@
 association studies (GWAs).  These analysis are typically performed on only the most frequently differing base-pairs, termed
-single nucleotide polymorphisms (SNPs), by applying linear or logistic regression analysis to each SNP separately~\cite{CCC2007}.  It has since been
+single nucleotide polymorphisms (SNPs), by applying linear or logistic regression analysis to each SNP
+separately~\cite{CCC2007}.
+
+It has since been
 demonstrated that not taking joint interactions between SNPs into account when filtering for potential disease mutations
-is inadequate~\cite{Manolio2009}.  \marginpar{do  these two sentences say teh same thing?}
-Complex trait analysis has demonstrated that in order to explain the overall
-heritability a model needs to take the joint contribution of all variants into account~\cite{Yang2011}.
+is inadequate~\cite{Manolio2009,Yang2011}.
 
@@
 during intermediate stages of building the model (see Table~\ref{timingtable}). 
-\marginpar{missing ref}
 This unbalanced work load where the driver node becomes the bottleneck and worker
@@
 The average prediction error for an individual tree $h(X; B)$ is
-\marginpar{is this formulation correct in the case of classification? See Friedman on 0-1 losss, Rob}
+
+\cite{Friedman.1997}
+
 \begin{equation}

### 2016-11-27 doc/publications/plos16/plos16.tex
changes to plos16.tex

@@
 Random forests are a variance reduction technique. They take unpruned tree models, which have a low bias but a high
-variance, and by combining them reduce the variance. As the prediction error is the sum of the variance and the bias
-squared, this strategy should lead to a low prediction error.
+variance, and by combining them reduce the variance.  For regression problems, as the prediction error is the sum of the
+variance and the bias squared, this strategy should lead to a low prediction error.
 
-The average prediction error for an individual tree $h(X; B)$ is
+For classifiction problems there is no obvious decomposition of classification error into an equivalent bias plus
+variance form. A number of different additive decompositions of classification error have been considered
+in an effort to understand the success of bagging. They all less transparent than the regression
+case (see \cite{Friedman.1997} and citations therein).
 
-\cite{Friedman.1997}
-
+In the case of a regression tree, the average prediction error for an individual tree $h(X; B)$ is
 \begin{equation}
@@
 conditions caused by the interaction of large groups of SNPs.
-\marginpar{check this. Denis}
 
@@
 data.
-\marginpar{ \cite{Chen.and.Ishwaran.2012} cite many papers on extensions of random forests to handle 
-GWAS data. What do we need to do in this area? What about linkage disequilibrium? Denis}
 
@@
 81 million variant positions (i.e. a matrix of $2,504 \times 81$ million).  For compatibility with the majority of the
-algorithms we demonstrate, we convert the original variant representations to double representations. 
-\marginpar{are they stored a double or integer?}
- For example, we
+algorithms we demonstrate, we convert the original variant representations to double representations.  For example, we
 store the absence of a variant (0\textbar0) as 0, a heterozygous variant (1\textbar0 or 0\textbar1)as 1, and a

### 2016-11-28 doc/publications/plos16/plos16.tex
changes to plos16.tex, added reference

@@
 the important variables, plotted as a function of $\delta = n/p$ and $\rho =k/n$ (where $k$ is the number of significant
-variables). There is a distinct doundary, shown empirically and also based on arguments from combinatorial geometry, to the
-region where we can reliably recover significant variables.
+variables). There is a distinct doundary, shown empirically and also based on arguments from combinatorial geometry, to
+the region where we can reliably recover significant variables.
 
 We note that the examples we are considering lie on the extreme left of the plot, close to the origin. So there are
-limits to what can be recovered in the presence of noise. However in the case of data that is both big and wide, \cursedforest and
-other VariantSpark methods may provide a useful tool.
+limits to what can be recovered in the presence of noise. However in the case of data that is both big and wide,
+\cursedforest and other VariantSpark methods may provide a useful tool.
 
-\marginpar{redo plot}
+Notes:
+\begin{itemize}
+\item \cite{Donoho.and.Stodden.2006} evalute variable selection by 
+$$\frac{\Vert\hat{\beta}-\beta\Vert_2}{\Vert\beta\Vert_2}$$
+However there is no $\hat{\beta}$ in a tree model. Piotr was using vaiable importace as
+a measure of $\hat{\beta}$ (i.e. we expect varible importance to be proportional to $\hat{\beta}$. Is this correct?
+
+\end{itemize}
 
@@
 \section*{Acknowledgments}
-Cras egestas velit mauris, eu mollis turpis pellentesque sit amet.
-Interdum et malesuada fames ac ante ipsum primis in faucibus.
-Nam id pretium nisi.
-Sed ac quam id nisi malesuada congue.
-Sed interdum aliquet augue, at pellentesque quam rhoncus vitae.
+
+Cras egestas velit mauris, eu mollis turpis pellentesque sit
+amet. Interdum et malesuada fames ac ante ipsum primis in
+faucibus. Nam id pretium nisi. Sed ac quam id nisi malesuada
+congue. Sed interdum aliquet augue, at pellentesque quam rhoncus
+vitae.
 

### 2016-12-02 doc/publications/plos16/plos16.tex
added reference to plos16.tex

@@
 \section{Introduction}
-The digital revolution is seeing a dramatic increase in data collected about almost every aspect of life~\cite{Loebbecke2015}.
-These data sets are not only growing vertically, by capturing more events, but also horizontally by capturing
-more information about these events.  The challenge of big and ``wide'' data is especially pronounced in the health
-space where, for example, whole genome sequence (WGS) technology enables researchers to interrogate all 3 billion base pairs of the
-human genome.
+The digital revolution is seeing a dramatic increase in data collected about almost every aspect of
+life~\cite{Loebbecke2015}.  These data sets are not only growing vertically, by capturing more events, but also
+horizontally by capturing more information about these events.  The challenge of big and ``wide'' data is especially
+pronounced in the health space where, for example, whole genome sequence (WGS) technology enables researchers to
+interrogate all 3 billion base pairs of the human genome.
 
 Identifying the relevant or disease specific base-pair difference between individuals is the focus of genome wide
-association studies (GWAs).  These analysis are typically performed on only the most frequently differing base-pairs, termed
-single nucleotide polymorphisms (SNPs), by applying linear or logistic regression analysis to each SNP
-separately~\cite{CCC2007}.
+association studies (GWAs).  These analysis are typically performed on only the most frequently differing base-pairs,
+termed single nucleotide polymorphisms (SNPs), by applying linear or logistic regression analysis to each SNP
+separately~\cite{CCC2007}.  It has since been demonstrated that not taking joint interactions between SNPs into account
+when filtering for potential disease mutations is inadequate~\cite{Manolio2009,Yang2011}.
 
-It has since been
-demonstrated that not taking joint interactions between SNPs into account when filtering for potential disease mutations
-is inadequate~\cite{Manolio2009,Yang2011}.
-
-A tree based model is capable of taking this interaction of variables
-into account. In addition, random forests are well suited for
-processing ``wide'' genomic data for two reasons.  Firstly, while other machine learning applications have the
-propensity to overfit datasets with more features $p$ than samples $n$ (a consequence of the ``curse of
-dimensionality''~\cite{Bauer2014, bellman1961adaptive}), decision trees are resistant to overfitting. 
-Secondly, random forests are also very easy to parallelize. As the forest is a sum of decision trees, it it possible to grow
-separate trees on different processors and combine the trees. See the
-discussion of the properties of random forests in Section \ref{section:methods}.
+A tree based model is capable of taking this interaction of variables into account \cite{Wright.et.al.2016}. In addition, random forests are well
+suited for processing ``wide'' genomic data for two reasons.  Firstly, while other machine learning applications have
+the propensity to overfit datasets with more features $p$ than samples $n$ (a consequence of the ``curse of
+dimensionality''~\cite{Bauer2014, bellman1961adaptive}), decision trees are resistant to overfitting.  Secondly, random
+forests are also very easy to parallelize. As the forest is a sum of decision trees, it it possible to grow separate
+trees on different processors and combine the trees. See the discussion of the properties of random forests in Section
+\ref{section:methods}.
 
@@
 increasing feature size~\cite{Wright.and.Ziegle.2016}.  This is despite R now supporting long vectors (length greater
-than $2^{31}-1$), and the {\sc randomForest} implementation in R lending itself to parallelization to grow multiple trees
-simultaneously~\cite{Liaw.and.Weiner.2002} and them combine them.  More suitable implementations for ``wide'' genomic data have been
-developed in other programming languages. {\sc CloudForest}~\cite{Bressler2015} written in Go, achieves fast run
-times by effective use of the CPU cache, optimizing for different classes of features and efficiently multi-threading.
-Similarly, {\sc Ranger}~\cite{Wright.and.Ziegle.2016} written in C++ with an R front-end is specifically optimized for
-dealing with large data sets.
+than $2^{31}-1$), and the {\sc randomForest} implementation in R lending itself to parallelization to grow multiple
+trees simultaneously~\cite{Liaw.and.Weiner.2002} and them combine them.  More suitable implementations for ``wide''
+genomic data have been developed in other programming languages. {\sc CloudForest}~\cite{Bressler2015} written in Go,
+achieves fast run times by effective use of the CPU cache, optimizing for different classes of features and efficiently
+multi-threading.  Similarly, {\sc Ranger}~\cite{Wright.and.Ziegle.2016} written in C++ with an R front-end is
+specifically optimized for dealing with large data sets.
 
@@
 parallelization the split for each node in a tree thereby handling millions of features. It can, for example, fit a
-model to the 1000 Genomes
-Project data, which consists of approximately 2500 samples with up to 80 million variants~\cite{1KG2012}.  By also
-utilizing Spark to read in and manipulate the standard genomic variant format (VCF) directly, \cursedforest\ outperforms
-existing tools even on small datasets where multi-threading generally performs well. Harnessing the virtually unlimited
-capability to parallelize tasks, \cursedforest\ can build a large number of trees, which boosts the accuracy for
-variable importance selection~\cite{Szymczak2016}.
+model to the 1000 Genomes Project data, which consists of approximately 2500 samples with up to 80 million
+variants~\cite{1KG2012}.  By also utilizing Spark to read in and manipulate the standard genomic variant format (VCF)
+directly, \cursedforest\ outperforms existing tools even on small datasets where multi-threading generally performs
+well. Harnessing the virtually unlimited capability to parallelize tasks, \cursedforest\ can build a large number of
+trees, which boosts the accuracy for variable importance selection~\cite{Szymczak2016}.
 
@@
 Using VariantSpark, we successfully built a k-means model on the full $2500 \times 80$ million matrix to cluster
-individuals by their ethnicity achieving an Adjusted Rand Index (ARI) of 0.84 (with 1 being perfect clustering and -1 random
-clustering).
+individuals by their ethnicity achieving an Adjusted Rand Index (ARI) of 0.84 (with 1 being perfect clustering and -1
+random clustering).
 
@@
 with only modest dimensionality.  Although Spark ML can build a random forest model on a subset of the data (chromosome
-1), the time taken is excessive due to the large amount of data being aggregated and processed by the driver node
-during intermediate stages of building the model (see Table~\ref{timingtable}). 
-This unbalanced work load where the driver node becomes the bottleneck and worker
-nodes being idle prevents a seamless scaling to larger datasets. The memory requirements per executor also increases
-with dimensionality due to the data types Spark ML uses, which we elaborate on in the discussion.
+1), the time taken is excessive due to the large amount of data being aggregated and processed by the driver node during
+intermediate stages of building the model (see Table~\ref{timingtable}).  This unbalanced work load where the driver
+node becomes the bottleneck and worker nodes being idle prevents a seamless scaling to larger datasets. The memory
+requirements per executor also increases with dimensionality due to the data types Spark ML uses, which we elaborate on
+in the discussion.
 
@@
 
-In the first section we explore some of the properties of random forests as these will also be apparent in the \cursedforest\
-implementation. Here we provide demonstrations using synthetic data.  We then demonstrate the scalability of
-\cursedforest\ in respect to the dimensionality of the data, building a random forest model on whole-genome data from
+In the first section we explore some of the properties of random forests as these will also be apparent in the
+\cursedforest\ implementation. Here we provide demonstrations using synthetic data.  We then demonstrate the scalability
+of \cursedforest\ in respect to the dimensionality of the data, building a random forest model on whole-genome data from
 the 1000 Genomes Project. Here we report on the accuracy (in terms of the ARI) and the error.  Finally, given the role
@@
 
-Random forests  apply the technique of bootstrap aggregation or bagging to decision trees.  The training
-data sets are independently drawn, with replacement,  from the joint distribution of $(X,Y)$ so that each set comprises $n$ $(p+1)$-tuples
- $(x_1,y_1),\ldots, (x_n,y_n)$. This is done $B$ times and  a tree is grown on each of these $B$ samples.
-The results are then combined in an appropriate way. In a classification problem we take the majority vote for the trees over
-the $q=1,..,Q$ classes,
+Random forests apply the technique of bootstrap aggregation or bagging to decision trees.  The training data sets are
+independently drawn, with replacement, from the joint distribution of $(X,Y)$ so that each set comprises $n$
+$(p+1)$-tuples $(x_1,y_1),\ldots, (x_n,y_n)$. This is done $B$ times and a tree is grown on each of these $B$ samples.
+The results are then combined in an appropriate way. In a classification problem we take the majority vote for the trees
+over the $q=1,..,Q$ classes,
 \begin{equation*}
@@
 For classifiction problems there is no obvious decomposition of classification error into an equivalent bias plus
-variance form. A number of different additive decompositions of classification error have been considered
-in an effort to understand the success of bagging. They all less transparent than the regression
-case (see \cite{Friedman.1997} and citations therein).
+variance form. A number of different additive decompositions of classification error have been considered in an effort
+to understand the success of bagging. They all less transparent than the regression case (see \cite{Friedman.1997} and
+citations therein).
 
@@
 
-Since the original implementation of random forests, there have been a number of developments leading to a better understanding of
-the algorithm. 
+Since the original implementation of random forests, there have been a number of developments leading to a better
+understanding of the algorithm.
 \begin{itemize}
@@
 However, keeping these difficult cases in mind, random forests have a strong claim to being resistant to overfitting.
-
-\item \cite{Strobl.et.al.2007} demonstrates that the random forest algorithm is subject to a bias in variable selection via two mechanisms:
+\item \cite{Strobl.et.al.2007} demonstrates that the random forest algorithm is subject to a bias in variable selection
+  via two mechanisms:
   \begin{enumerate}
@@
 
-\cite{Strobl.et.al.2007} introduce changes to the random forest algorithm to fix 1) and 2).  Effect 1) will be
-apparent in the \cursedforest algorithm as well.  However, there are many examples, particularly in genomics, where we have data
+\cite{Strobl.et.al.2007} introduce changes to the random forest algorithm to fix 1) and 2).  Effect 1) will be apparent
+in the \cursedforest algorithm as well.  However, there are many examples, particularly in genomics, where we have data
 sets that are:
@@
 VariantSpark made use of Spark's machine learning algorithms ``Spark ML''. These algorithms require data in a
-``DataFrame'', similar to DataFrames in R.  While Spark ML algorithms demonstrate some scalability when dealing with a large
-number of samples, this scalability eventually break down as we include more features.
+``DataFrame'', similar to DataFrames in R.  While Spark ML algorithms demonstrate some scalability when dealing with a
+large number of samples, this scalability eventually break down as we include more features.
 
@@
 dimensions (variants) increases, the size of the vectors increases. This results in higher memory-requirements as
-increasingly large vectors must be loaded into memory.
-\marginpar{not clear. Are RDDs the problem or the solutuion}
+increasingly large vectors must be loaded into memory.  \marginpar{not clear. Are RDDs the problem or the solutuion}
 
@@
 
-Compared to Spark ML, \cursedforest\ further parallelizes work down to individual features.  For example, for each node of
-a tree, \cursedforest\ will distribute tasks that consist of single features (variants), for every individual.  Each of
-these tasks will calculate the information gain for that specific feature.  Once these tasks have completed, the results
-are reduced to return the feature which gives the greatest information gain.  This process is then repeated until
-\cursedforest\ has created the entire decision tree.
+Compared to Spark ML, \cursedforest\ further parallelizes work down to individual features.  For example, for each node
+of a tree, \cursedforest\ will distribute tasks that consist of single features (variants), for every individual.  Each
+of these tasks will calculate the information gain for that specific feature.  Once these tasks have completed, the
+results are reduced to return the feature which gives the greatest information gain.  This process is then repeated
+until \cursedforest\ has created the entire decision tree.
 
 The current implementation of \cursedforest\ uses an ``information gained'' criteria for splitting. Let $f_q$ be the
-fraction of items labelled with value $q$ where $q = 1, \ldots, Q$ at a node. The entropy of the node is $H = -
-\sum^{m}_{i=1} f_i \log^{}_2 f_i.$ The expected information gain is the change in entropy $H$ from a prior state $T$ to
-a state that takes some information $a$ as given:
+fraction of items labelled with value $q$ where $q = 1, \ldots, Q$ at a node. The entropy of the node is
+$H = - \sum^{m}_{i=1} f_i \log^{}_2 f_i.$ The expected information gain is the change in entropy $H$ from a prior state
+$T$ to a state that takes some information $a$ as given:
 \[ IG(T,a) = H(T) - H(T|a). \]
@@
 \subsubsection{The Cancer Genome Atlas}
-We obtained The Cancer Genome Atlas (TCGA) BRCA data as VCF files from the GDC data portal.
-This data is in the category ``Simple Nucleotide Variation''.
-After downloading, we use a custom script to name the samples in the files and filter for only the "Normal" tissue column.
-We then use GATK to merge the files into a single VCF file.
+We obtained The Cancer Genome Atlas (TCGA) BRCA data as VCF files from the GDC data portal.  This data is in the
+category ``Simple Nucleotide Variation''.  After downloading, we use a custom script to name the samples in the files
+and filter for only the "Normal" tissue column.  We then use GATK to merge the files into a single VCF file.
 
 \subsubsection{Synthetic data} 
-\marginpar{check details of simulation. Piotr}
-Each dataset consists of $n$ samples and $p$ variables where $n << p$, and values for each variable are ordinal variables with
-three levels represented as numbers $\{0, 1, 2\}$ (which correspond to an additive effect encoding of genomics variation) randomly
-generated from uniform distribution with equal probabilities.  The dichotomous response is generated from a
-Bernoulli trial with $p$ (probability of success) calculated by applying an inverse logit transformation to the continuous
-linear response ($y$) of five variables:
+\marginpar{check details of simulation. Piotr} Each dataset consists of $n$ samples and $p$ variables where $n << p$,
+and values for each variable are ordinal variables with three levels represented as numbers $\{0, 1, 2\}$ (which
+correspond to an additive effect encoding of genomics variation) randomly generated from uniform distribution with equal
+probabilities.  The dichotomous response is generated from a Bernoulli trial with $p$ (probability of success)
+calculated by applying an inverse logit transformation to the continuous linear response ($y$) of five variables:
 $$y = a_{10}* v_{10} + a_{100}*v_{100} + a_{1000}*v_{1000} + a_{10000}*v_{10000} + a_{10000}*v_{100000}$$
@@
 \subsection{Parameter settings}
-We consider the parameter setting for the random forest algorithm. We use the R notation from the {\sc randomForest} package
-\cite{Liaw.and.Weiner.2002} which incorporates the original Fortran code by Brieman and Cutler.  We incorporate the
-advice of \cite{Liaw.and.Weiner.2002}, which we have found mirrors our own experience:
+We consider the parameter setting for the random forest algorithm. We use the R notation from the {\sc randomForest}
+package \cite{Liaw.and.Weiner.2002} which incorporates the original Fortran code by Brieman and Cutler.  We incorporate
+the advice of \cite{Liaw.and.Weiner.2002}, which we have found mirrors our own experience:
 \begin{itemize}
@@
 \subsection{\cursedforest\ can evaluate variable importance for millions of features}\label{synthetic}
-%In this section we demonstrate that \cursedforest's ability to rapidly grow large number of trees, \ntree, in
-%parallel results in a stable feature importance ranking.  This concept was originally proposed by Liaw and
-%Weiner~\cite{Liaw.and.Weiner.2002}, who suggested that though the variable importance measures may vary from run to
-%run, the ranking of the variables will stabilise with growing \ntree.
+% In this section we demonstrate that \cursedforest's ability to rapidly grow large number of trees, \ntree, in parallel
+% results in a stable feature importance ranking.  This concept was originally proposed by Liaw and
+% Weiner~\cite{Liaw.and.Weiner.2002}, who suggested that though the variable importance measures may vary from run to
+% run, the ranking of the variables will stabilise with growing \ntree.
 In this section we demonstrate \cursedforest\ two main features, namely variable importance analysis to select the most
@@
 
-\marginpar{wrong figures cited. Rob}
-Fig.~\ref{figure:synth}b illustrates that increasing \mtry\ also improves the OOB (out of bag) classification error.
+\marginpar{wrong figures cited. Rob} Fig.~\ref{figure:synth}b illustrates that increasing \mtry\ also improves the OOB
+(out of bag) classification error.
 %Piotr can you please explain the remaining error on the synthetic data
@@
 \centering
-\caption{
-{\bf Performance comparison between the different machine learning algorithms.}}
+\caption{ {\bf Performance comparison between the different machine learning algorithms.}}
 \begin{tabular}{|l|l|l|l|l|l||}
@@
 \subsection{Scalability analysis}
-
 In this section we explore the performance of \cursedforest\ by testing its ability to scale to different sizes of data
@@
 samples, by varying \mtry\ fraction and the number of CPU cores allocated to the execution. The raw results are
-presented in table \ref{synthetictimingtable} and visualized in figure \ref{figure:synth} below. Regardless of the number of cores used,
-\cursedforest\ displays approximately linear dependency between the execution time and the \mtry\ parameter (presented
-here as a fraction of the number of variables) [Fig 1 a).
+presented in table \ref{synthetictimingtable} and visualized in figure \ref{figure:synth} below. Regardless of the
+number of cores used, \cursedforest\ displays approximately linear dependency between the execution time and the \mtry\
+parameter (presented here as a fraction of the number of variables) [Fig 1 a).
 
@@
 communication overhead (with lower \mtry\ values the proportion of time for parallelizable computation to the time for
-internode communication is lower) while in the latter case it is most likely caused by reaching the clusters computational
-capacity.
+internode communication is lower) while in the latter case it is most likely caused by reaching the clusters
+computational capacity.
 
@@
 \section{Conclusion}
-
 We have demonstrated that using a different model of paralleization can extend the random forest to delaing with the

### 2016-12-02 doc/publications/plos16/plos16.tex
changes to  plos16.tex

@@
 \section{Conclusion}
-We have demonstrated that using a different model of paralleization can extend the random forest to delaing with the
-case of an extremely large number of variables. By comparing this implementation to other implentations (including those
-optimized for large data sets) we have demonstrated the utility of this implentation.
-
-We have treated the case of variable selection in a $p >> n$ model where most of the variables are uninformative. We
-have also demostrated the utility of the model for large GWAs data sets.
+We have demonstrated that using a different paralleization model can extend random forests to the case of an
+extremely large number of variables. We have treated the case of variable selection in a $p >> n$ model where most of
+the variables are uninformative and have demonstrated the utility of the model for large GWAs data sets. By comparing
+this implementation to other implementations (including those optimized for large data sets) we have demonstrated the
+utility of this approach.
 
@@
 
-Donoho and Tanner \cite{Donoho.and.Tanner.2009} give a ``universal phase change'' result based on arguments from
-combinatorial geometry. This has applications in a large number of areas including variable selection in high
-diensions. Consider figure \ref{figure:phase-diagram-equivalence.png} which shows the region where a model can recover
-the important variables, plotted as a function of $\delta = n/p$ and $\rho =k/n$ (where $k$ is the number of significant
-variables). There is a distinct doundary, shown empirically and also based on arguments from combinatorial geometry, to
-the region where we can reliably recover significant variables.
+Donoho and Tanner \cite{Donoho.and.Tanner.2009} give a ``universal phase change'' result that has applications in a
+large number of areas including variable selection in high dimensions. Consider figure
+\ref{figure:phase-diagram-equivalence.png} which shows the region where a model can recover the important variables,
+plotted as a function of $\delta = n/p$ and $\rho =k/n$ (where $k$ is the number of significant variables). There is a
+distinct boundary, shown empirically and also based on arguments from combinatorial geometry, to the region where we can
+reliably recover significant variables.  \cite{Donoho.and.Stodden.2006} investigate the behaviour of a number of
+regression approaches for variable selection (lars, lasso and forward stepwise) and make the point that above the
+phase-transition line variable recovery is still possible by a combinatorial approach.
+
+It is not surprising that that it is more difficult to recover the signal variable in the upper-left area of the figure,
+as the problem is both under-determined and sparse. What is surprising is the connection with arguments from
+combinatorial geometry. This suggests that we are seeing a universal rule rather than an implementation issue. As
+\cursedforest is designed for extremely large numbers of variables it is likely to be operating in difficult regions of
+the figure.
 
@@
 \begin{itemize}
-\item \cite{Donoho.and.Stodden.2006} evalute variable selection by 
+\item \cite{Donoho.and.Stodden.2006} evaluate variable selection by 
 $$\frac{\Vert\hat{\beta}-\beta\Vert_2}{\Vert\beta\Vert_2}$$
-However there is no $\hat{\beta}$ in a tree model. Piotr was using vaiable importace as
-a measure of $\hat{\beta}$ (i.e. we expect varible importance to be proportional to $\hat{\beta}$. Is this correct?
-
+However there is no $\hat{\beta}$ in a tree model. We are using variable importance as
+a measure of the ordering of $\hat{\beta}$ (i.e. we expect variable importance to be proportional to $\hat{\beta}$). Is this correct?
+\item as RF is sort of greedy approximation to combinatorial approach, it may have some success in the dead zone
+\item  \cite{Donoho.and.Stodden.2006} are considering problems with $L_2$ errors (with additional  $L_1$ norms on the
+  $\beta$. $L_1$).
+How will their results carry over to 0-1 loss?
 \end{itemize}

### 2016-12-11 doc/publications/plos16/plos16.tex
changes figures 1 and 2 in plos16.tex
changes to plos16.tex

@@
 \usepackage{subcaption}
-
+\usepackage{placeins}
 \begin{document}
@@
 \subsection{\cursedforest\ can evaluate variable importance for millions of features}\label{synthetic}
-% In this section we demonstrate that \cursedforest's ability to rapidly grow large number of trees, \ntree, in parallel
-% results in a stable feature importance ranking.  This concept was originally proposed by Liaw and
-% Weiner~\cite{Liaw.and.Weiner.2002}, who suggested that though the variable importance measures may vary from run to
-% run, the ranking of the variables will stabilise with growing \ntree.
 In this section we demonstrate \cursedforest\ two main features, namely variable importance analysis to select the most
@@
 high-dimensional samples.
-%We therefore generate a synthetic dataset with varying numbers of variables (features) and samples, as discussed in the method section. 
 We therefore generate a synthetic dataset with 2.5 million variables (features) and 5000 samples, as discussed in the
@@
 
-\marginpar{wrong figures cited. Rob} Fig.~\ref{figure:synth}b illustrates that increasing \mtry\ also improves the OOB
+Fig.~\ref{figure:synth}b illustrates that increasing \mtry\ also improves the OOB
 (out of bag) classification error.
-%Piotr can you please explain the remaining error on the synthetic data
 By $mtry=0.8$ the error is converging to X, which is due to Y.
@@
 
-\marginpar{figure needs bigger text}
-% \begin{figure}[tbhp] 
-%  \begin{adjustwidth}{-1.00in}{0in}
-%   \begin{subfigure}[b]{0.5\linewidth}
-%     \centering
-%     \includegraphics[totalheight=6cm]{./figs/var_oder.png} 
-%     \caption{Variable importance of top 20 variables.} 
-%     \label{fig7:a} 
-%     \vspace{4ex}
-%   \end{subfigure}%% 
-%   \begin{subfigure}[b]{0.5\linewidth}
-%     \centering
-%     \includegraphics[totalheight=6cm]{./figs/var_oob.png}
-%     \caption{OOB classification error} 
-%     \label{fig7:b} 
-%     \vspace{4ex}
-%   \end{subfigure} 
-% \caption{{\bf Variable importance (Mean Decrease Gini) and OOB (out of bag) classification error estimate for random
-%     forests with 100 trees trained on the synthetic dataset of 2.5M features and 5K samples.}  {\bf a)} Variable
-%   importance of top 20 variables normalized to the sum of the linear model coefficients. {\bf Others} represents
-%   aggregated importance of variables not present in the linear model. The dotted line represents the original linear
-%   model coefficients.
-%  {\bf b)} OOB (out of bag) classification error
-%   estimate as function of \mtry\ fraction.}
-%   \label{figure:synth}
-% \end{adjustwidth}
-% \end{figure}
 \marginpar{this was done with n=100000. Needs to redone with n=2.5M}
+
+\subsubsection{simulation 1}
+
+Piotr's simulation with $n=1000$ and $p= 10000$
 
@@
     \includegraphics[totalheight=8cm]{./figs/rbo.png}
+    \caption{The  rank-biased-overlap measure.} 
+    \label{fig7:b} 
+    \vspace{4ex}
+  \end{subfigure} 
+  \caption{{\bf Variable importance (Mean Decrease Gini) and OOB (out of bag) classification error estimate for random
+      forests with 100 trees trained on the synthetic dataset of 2.5M features and 5K samples.}  Each measure is shown
+    as a function of \mtry and \ntree. {\bf a0} OOB (out of bag) classification error.  {\bf b)} The rank-biased-overlap \cite[]{Webber.et.al.2010}. }
+  \label{figure:synth}
+\end{adjustwidth}
+\end{figure}
+
+\FloatBarrier
+\subsubsection{simulation 2}
+Similar to Piotr's simulation with  except $y= w_1x_1 w_2x_2 w_3x_3 w_4x_4 w_5x_5$. $n=1000$ and $p= 10000$
+The significant variables can no longer be recovered by single variable tests for correlation with $y$
+
+\begin{figure}[tbhp] 
+ \begin{adjustwidth}{-1.00in}{0in}
+  \begin{subfigure}[b]{0.5\linewidth}
+    \centering
+    \includegraphics[totalheight=8cm]{./figs/out-of-bag-prediction-error-prod.png}
+    \caption{OOB classification error} 
+    \label{fig7:b} 
+    \vspace{4ex}
+  \end{subfigure} 
+  \begin{subfigure}[b]{0.5\linewidth}
+    \centering
+    \includegraphics[totalheight=8cm]{./figs/rbo-prod.png}
     \caption{The  rank-biased-overlap measure.} 
@@
 \begin{itemize}
+\item we try a small simulation, a regression problem with $\beta(1:k) \sim U(1,100)$ and $y= X\beta + \epsilon$, $\epsilon \sim  N(0,\sqrt{16})$
 \item \cite{Donoho.and.Stodden.2006} evaluate variable selection by 
 $$\frac{\Vert\hat{\beta}-\beta\Vert_2}{\Vert\beta\Vert_2}$$
+See figure \ref{figure:stepwise_lars_cv.png} 
+\item 
 However there is no $\hat{\beta}$ in a tree model. We are using variable importance as
@@
 \begin{figure}[tbhp] 
+ \begin{adjustwidth}{-1.00in}{0in}
+  \begin{subfigure}[b]{0.4\linewidth}
+    \centering
    \includegraphics[totalheight=6cm]{./figs/phase-diagram-equivalence.png} 
@@
 \label{figure:phase-diagram-equivalence.png} 
+    \vspace{4ex}
+  \end{subfigure} 
+  \begin{subfigure}[b]{0.6\linewidth}
+    \centering
+  \includegraphics[totalheight=6cm]{./figs/stepwise_lars_cv.png}
+\caption{Stepwise lars. $p=200$, $n$ and $k$ variable. Model selected by cross-validation
+}
+\label{figure:stepwise_lars_cv.png}
+  \end{subfigure} 
+  \caption{}
+\end{adjustwidth}
 \end{figure}
 
+\begin{figure}[tbhp]
+\includegraphics[totalheight=6cm]{/home/dun280/Dropbox/random_forests/exploration/figs/ranger.png}
+\caption{ ranger}
+\label{figure:ranger.png}
+\end{figure}
+
+\clearpage
 \section{Supporting Information}

### 2016-12-18 doc/publications/plos16/plos16.tex
added some figures
changes to plos16.tex

@@
 dimensionality''~\cite{Bauer2014, bellman1961adaptive}), decision trees are resistant to overfitting.  Secondly, random
-forests are also very easy to parallelize. As the forest is a sum of decision trees, it it possible to grow separate
+forests are also very easy to parallelize. As the forest is a sum of decision trees, it is possible to grow separate
 trees on different processors and combine the trees. See the discussion of the properties of random forests in Section
@@
 trees simultaneously~\cite{Liaw.and.Weiner.2002} and them combine them.  More suitable implementations for ``wide''
-genomic data have been developed in other programming languages. {\sc CloudForest}~\cite{Bressler2015} written in Go,
+genomic data have been developed in other programming languages. \textsc{CloudForest} \cite{Bressler2015} written in Go,
 achieves fast run times by effective use of the CPU cache, optimizing for different classes of features and efficiently
-multi-threading.  Similarly, {\sc Ranger}~\cite{Wright.and.Ziegle.2016} written in C++ with an R front-end is
+multi-threading.  Similarly, \textsc{Ranger}~\cite{Wright.and.Ziegle.2016} written in C++ with an R front-end is
 specifically optimized for dealing with large data sets.
@@
 of \cursedforest\ in respect to the dimensionality of the data, building a random forest model on whole-genome data from
-the 1000 Genomes Project. Here we report on the accuracy (in terms of the ARI) and the error.  Finally, given the role
+the 1000 Genomes Project. We report the error in predicting ethnicity.  Finally, given the role
 different parameter values can play in model construction, we explore the effect tuning these parameters can have on the
@@
 \subsubsection{Synthetic data} 
-\marginpar{check details of simulation. Piotr} Each dataset consists of $n$ samples and $p$ variables where $n << p$,
+Each dataset consists of $n$ samples and $p$ variables where $n << p$,
 and values for each variable are ordinal variables with three levels represented as numbers $\{0, 1, 2\}$ (which
 correspond to an additive effect encoding of genomics variation) randomly generated from uniform distribution with equal
-probabilities.  The dichotomous response is generated from a Bernoulli trial with $p$ (probability of success)
-calculated by applying an inverse logit transformation to the continuous linear response ($y$) of five variables:
-$$y = a_{10}* v_{10} + a_{100}*v_{100} + a_{1000}*v_{1000} + a_{10000}*v_{10000} + a_{10000}*v_{100000}$$
-where $a_{10} = 1.6, a_{100} = 1.8, a_{1000} = 2.2, a_{10000} = 2.0, a_{10000}=2.3$.
+probabilities.
+
+The model paramters are $w_i = 1/\sqrt{2^{i-1}}$ for $i =1,\ldots 5$ then 
+\begin{align}
+y=& \prod_{i=1}^{i=5} {w_i x_i} + \epsilon \textrm{ for $x_i \neq 0$} 
+\intertext{where}
+\epsilon \sim & N(0, \sigma_\epsilon^2= 0.12)
+\end{align}
+The dichotomous response is generated by thresholding $y$ at the $0.8$ quantile
 
@@
 \subsection{\cursedforest\ can evaluate variable importance for millions of features}\label{synthetic}
-In this section we demonstrate \cursedforest\ two main features, namely variable importance analysis to select the most
+In this section we demonstrate two features of \cursedforest\, variable importance analysis to select the most
 associated set of features amongst millions of dimensions as well as its ability to accurately classify the labels of
-high-dimensional samples.
-We therefore generate a synthetic dataset with 2.5 million variables (features) and 5000 samples, as discussed in the
-method section.
+high-dimensional samples.  We therefore generate a synthetic dataset with 2.5 million variables (features) and 5000
+samples, as discussed in the method section.
 
@@
 \marginpar{this was done with n=100000. Needs to redone with n=2.5M}
-
-\subsubsection{simulation 1}
-
-Piotr's simulation with $n=1000$ and $p= 10000$
-
-\begin{figure}[tbhp] 
- \begin{adjustwidth}{-1.00in}{0in}
-  \begin{subfigure}[b]{0.5\linewidth}
-    \centering
-    \includegraphics[totalheight=8cm]{./figs/out-of-bag-prediction-error.png}
-    \caption{OOB classification error} 
-    \label{fig7:b} 
-    \vspace{4ex}
-  \end{subfigure} 
-  \begin{subfigure}[b]{0.5\linewidth}
-    \centering
-    \includegraphics[totalheight=8cm]{./figs/rbo.png}
-    \caption{The  rank-biased-overlap measure.} 
-    \label{fig7:b} 
-    \vspace{4ex}
-  \end{subfigure} 
-  \caption{{\bf Variable importance (Mean Decrease Gini) and OOB (out of bag) classification error estimate for random
-      forests with 100 trees trained on the synthetic dataset of 2.5M features and 5K samples.}  Each measure is shown
-    as a function of \mtry and \ntree. {\bf a0} OOB (out of bag) classification error.  {\bf b)} The rank-biased-overlap \cite[]{Webber.et.al.2010}. }
-  \label{figure:synth}
-\end{adjustwidth}
-\end{figure}
-
-\FloatBarrier
-\subsubsection{simulation 2}
-Similar to Piotr's simulation with  except $y= w_1x_1 w_2x_2 w_3x_3 w_4x_4 w_5x_5$. $n=1000$ and $p= 10000$
-The significant variables can no longer be recovered by single variable tests for correlation with $y$
 
@@
 \begin{figure}[tbhp]
-\includegraphics[totalheight=6cm]{/home/dun280/Dropbox/random_forests/exploration/figs/ranger.png}
+\includegraphics[totalheight=6cm]{./figs/ranger.png}
 \caption{ ranger}

### 2016-12-18 doc/publications/plos16/plos16.tex
added reference and changes to plos16.tex

@@
   better performance;
-\item the size and complexity of the individual trees is controlled in {\sc randomForest} by setting \texttt{nodesize}, the
-  minimum size of terminal nodes. It is controlled in Spark ML by setting \texttt{maxDepth}, the maximum depth of each
-  tree in the forest\footnote{The Spark ML documentation \cite[]{Spark.2016} sets \texttt{maxDepth}
-    to 4 in their classification example. This is wrong. The
-    individual trees should be grown to a large size to ensure that
-    they give an unbiased estimate.}.
-  In the TCGA example the RandomForests code produced trees with depths of 30+. 
+\item the size and complexity of the individual trees is controlled in
+  {\sc randomForest} by setting \texttt{nodesize}, the minimum size of
+  terminal nodes. It is controlled in Spark ML by setting
+  \texttt{maxDepth}, the maximum depth of each tree in the
+  forest\footnote{The Spark ML documentation \cite[]{Spark.2016} sets
+    \texttt{maxDepth} to 4 in their classification example. This may
+    give an estimate with a higher bias and, as such, may be a poor
+    choice. See the discussion in \cite{Dietterich.2002} where it is
+    shown that bagging may reduce bias as well as variance but gives a
+    better result for low bias learners. \cite{Dietterich.2002} notes
+    that if the boostrap replicate approximation were correct, then
+    bagging would reduce variance without changing bias.} In the TCGA
+  example the RandomForests code produced trees with depths of 30+.
 \end{itemize}
@@
 \subsection{\cursedforest\ can evaluate variable importance for millions of features}\label{synthetic}
-In this section we demonstrate two features of \cursedforest\, variable importance analysis to select the most
+In this section we demonstrate two features of \cursedforest\, a variable importance analysis to select the most
 associated set of features amongst millions of dimensions as well as its ability to accurately classify the labels of
 high-dimensional samples.  We therefore generate a synthetic dataset with 2.5 million variables (features) and 5000
-samples, as discussed in the method section.
+samples, as discussed in the methods section.
 
@@
 worker nodes each with 16 Intel Xeon E5-2660@2.20GHz CPU cores and 128 GB of RAM.  We captured the OOB 
-classification error and the 20 most important variables.
+classification error and the rank-biased-overlap
+\cite[]{Webber.et.al.2010} measure.
 
-Fig.~\ref{figure:synth}a illustrates that increasing the number of predictors sampled for splitting at each node
-(\mtry) increases the accuracy with which features are ranked.  From $mtry=0.5$ the sort order of the features reflects
-the weight originally assigned to them (dotted line). Further increasing \mtry\ improves the predicted importance
-measure, which starts to converge to the weight originally assigned to them (coefficients).  It also increases the
-difference between the associated 5 variable and the remaining variables in the top 20 list.  By $mtry=0.8$ the variable
-with the smallest weight has a predicted importance measure that is 5 times larger than the aggregated importance of the
-15 remaining variables from the list.
+% Fig.~\ref{figure:out-of-bag-prediction-error-prod.png} illustrates that increasing the number of predictors sampled for splitting at each node
+% (\mtry) increases the accuracy with which features are ranked.  From $mtry=0.5$ the sort order of the features reflects
+% the weight originally assigned to them (dotted line). Further increasing \mtry\ improves the predicted importance
+% measure, which starts to converge to the weight originally assigned to them (coefficients).  It also increases the
+% difference between the associated 5 variable and the remaining variables in the top 20 list.  By $mtry=0.8$ the variable
+% with the smallest weight has a predicted importance measure that is 5 times larger than the aggregated importance of the
+% 15 remaining variables from the list.
 
-Fig.~\ref{figure:synth}b illustrates that increasing \mtry\ also improves the OOB
-(out of bag) classification error.
-By $mtry=0.8$ the error is converging to X, which is due to Y.
-\marginpar{what are ``X'' and ``Y''. Piotr}
+% Fig.~\ref{figure:rbo-prod.png} illustrates that increasing \mtry\ also improves the OOB
+% (out of bag) classification error.
+% By $mtry=0.8$ the error is converging to X, which is due to Y.
+% \marginpar{what are ``X'' and ``Y''. Piotr}
+
+% From the synthetic data we can conclude that \cursedforest\ is able to perform accurate feature selection as well as
+% accurate classification for large numbers of samples with large feature vectors.
+
+Fig.~\ref{figure:synth1} illustrates the effect of varying  \mtry\ and \ntree\ on these two measures. 
+Fig.~\ref{figure:out-of-bag-prediction-error-prod.png} shows that depends strongly on \ntree
+whereas Fig.~\ref{figure:rbo-prod.png} illustrates that the rbo measure is more dependent on \mtry.
 
@@
     \caption{OOB classification error} 
-    \label{fig7:b} 
+    \label{figure:out-of-bag-prediction-error-prod.png} 
     \vspace{4ex}
@@
     \caption{The  rank-biased-overlap measure.} 
-    \label{fig7:b} 
+    \label{figure:rbo-prod.png} 
     \vspace{4ex}
@@
       forests with 100 trees trained on the synthetic dataset of 2.5M features and 5K samples.}  Each measure is shown
-    as a function of \mtry and \ntree. {\bf a0} OOB (out of bag) classification error.  {\bf b)} The rank-biased-overlap \cite[]{Webber.et.al.2010}. }
-  \label{figure:synth}
+    as a function of \mtry\ and \ntree. Note that the plot colors have been chose so that the lighter color indicates
+    the better result, that is, a lower OOB and a higher RBO.}
+  \label{figure:synth1}
 \end{adjustwidth}
@@
                                                 & RF - Spark ML  &  0.05           &           7 min 44 seconds         \\
-                                                & RF - R         & 0.05       & 24 min           \\
+                                                & randomForest - R         & 0.05       & 24 min           \\
                                                 & ranger - R      & 0.06       & 1 min 10 seconds       \\
@@
 \multirow{3}{*}{phase3\_chr22: 2,504 x 1,103,548}   &               &            &                  \\
-                                                    & RF - R        & NA         & NA                \\
+                                                    & randomForest - R      & NA         & NA                \\
                                                     & H2O           & 0.05       & 16 hours         \\
@@
 & Spark ML & NA & NA  \\
-& ranger (C++)       &        NA     &        NA      \\
 & ranger (C++)       &        NA     &        NA      \\

### 2016-12-18 doc/publications/plos16/plos16.tex
new times in table

--- 2016-12-11 doc/publications/plos16/plos16.tex
@@
 \centering
-\caption{ {\bf Performance comparison between the different machine learning algorithms.}}
+\caption{ {\bf Performance comparison between the different random forest implementations.}}
 \begin{tabular}{|l|l|l|l|l|l||}
@@
 
-\multirow{3}{*}{phase1\_chr22: 1,092 x 490,036} & \cursedforest\ & 0.04  & 1 min 50 seconds                        \\
-                                                & RF - Spark ML  &  0.05           &           7 min 44 seconds         \\
-                                                & RF - R         & 0.05       & 24 min           \\
-                                                & ranger - R      & 0.06       & 1 min 10 seconds       \\
-                                                & ranger - c++     & 0.06       & 7 min            \\
+\multirow{5}{*}{phase1\_chr22: 1,092 x 490,036} & \cursedforest\ & 0.04  & 1 min 50 seconds                        \\
+                                                & Spark ML  &  0.05           &           7 min 44 seconds         \\
+                                                & randomForest (R)         & 0.05       & 24 min           \\
+                                                & ranger (R)      & 0.06       & 1 min 10 seconds       \\
+                                                & ranger (C++)     & 0.06       & 7 min            \\
 
  \hline
-\multirow{3}{*}{phase3\_chr22: 2,504 x 1,103,548}   &               &            &                  \\
-                                                    & RF - R        & NA         & NA                \\
+\multirow{5}{*}{phase3\_chr22: 2,504 x 1,103,548}   & \cursedforest\ & 0.06 & 4 min 2 seconds \\
+                                                    & Spark ML  & 0.06 & 37 min 19 seconds \\
+                                                    & randomForest (R)        & NA         & NA                \\
+                                                    & ranger (R)    & 0.06       & 4 min 30 seconds     \\
                                                     & H2O           & 0.05       & 16 hours         \\
-                                                    & ranger - R    & 0.06       & 4 min 30 seconds     \\
 \hline
-\multirow{3}{*}{phase3\_chr1: 2,504 x 6,450,364}    & \cursedforest\ & 0.063  & 1hr 58min         \\
-                                                    & RF - Spark ML  & 0.04       & 8hr 29min        \\
-                                                    & ranger - R      &  0.05       & 28 mins         \\
+\multirow{3}{*}{phase3\_chr1: 2,504 x 6,450,364}    & \cursedforest\ & 0.04  & 19 min 19 seconds         \\
+                                                    & Spark ML  & 0.04       & 8 hour 29 min        \\
+                                                    & ranger (R)      &  0.05       & 28 mins         \\
 \hline
-\multirow{3}{*}{phase3\_chr1-3: 2,504 x 19,328,051} & \cursedforest\ & 0.046  & 4hr 0min               \\
-                                                    & RF - Spark ML  &    NA         &   NA               \\
-                                                    & ranger - R      & 0.03      &         \\
-& ranger - R        &             0.973     &      03:58           \\
+\multirow{3}{*}{phase3\_chr1-3: 2,504 x 19,328,051} & \cursedforest\ & 0.02  & 56 min 43 seconds             \\
+                                                    & Spark ML &    NA         &   NA               \\
+                                                    & ranger (R)      & 0.03      &    3 hour 58 min     \\
 \hline
@@
                                                     & Spark ML & NA & NA  \\
-& ranger (C++)       &        NA     &        NA      \\
+                                                    & ranger (R)       &        NA     &        NA    \\
                                                     & ranger (C++)       &        NA     &        NA    \\
 \hline
+
 \end{tabular}

### 2016-12-18 doc/publications/plos16/plos16.tex

--- 2016-12-18 doc/publications/plos16/plos16.tex
@@
 \bf{Genome \%}                      & \bf{Method} & \bf{Error}  & \bf{Runtime}  \\
+
 \hline
 
-\multirow{3}{*}{phase1\_chr22: 1,092 x 490,036} & \cursedforest\ & 0.04  & 1 min 50 seconds                        \\
-                                                & RF - Spark ML  &  0.05           &           7 min 44 seconds         \\
-                                                & randomForest - R         & 0.05       & 24 min           \\
-                                                & ranger - R      & 0.06       & 1 min 10 seconds       \\
-                                                & ranger - c++     & 0.06       & 7 min            \\
+\multirow{5}{*}{phase1\_chr22: 1,092 x 490,036} & \cursedforest\ & 0.04  & 1 min 50 seconds                        \\
+                                                & Spark ML  &  0.05           &           7 min 44 seconds         \\
+                                                & randomForest (R)         & 0.05       & 24 min           \\
+                                                & ranger (R)      & 0.06       & 1 min 10 seconds       \\
+                                                & ranger (C++)     & 0.06       & 7 min            \\
 
 \hline
-\multirow{3}{*}{phase3\_chr22: 2,504 x 1,103,548}   &               &            &                  \\
-                                                    & randomForest - R      & NA         & NA                \\
+
+\multirow{5}{*}{phase3\_chr22: 2,504 x 1,103,548}   & \cursedforest\ & 0.06 & 4 min 2 seconds \\
+                                                    & Spark ML  & 0.06 & 37 min 19 seconds \\
+                                                    & randomForest (R)        & NA         & NA                \\
+                                                    & ranger (R)    & 0.06       & 4 min 30 seconds     \\
                                                     & H2O           & 0.05       & 16 hours         \\
-                                                    & ranger - R    & 0.06       & 4 min 30 seconds     \\
-\hline
-\multirow{3}{*}{phase3\_chr1: 2,504 x 6,450,364}    & \cursedforest\ & 0.063  & 1hr 58min         \\
-                                                    & RF - Spark ML  & 0.04       & 8hr 29min        \\
-                                                    & ranger - R      &  0.05       & 28 mins         \\
-\hline
-\multirow{3}{*}{phase3\_chr1-3: 2,504 x 19,328,051} & \cursedforest\ & 0.046  & 4hr 0min               \\
-                                                    & RF - Spark ML  &    NA         &   NA               \\
-                                                    & ranger - R      & 0.03      &         \\
-                                                    & ranger - R        &             0.973     &      03:58           \\
+
 \hline
 
-\multirow{4}{*}{phase3\_chr1-22: 2,504 x 81,047,467} & \cursedforest\ & XX  & 7hr 14min \\
+\multirow{3}{*}{phase3\_chr1: 2,504 x 6,450,364}    & \cursedforest\ & 0.04  & 19 min 19 seconds         \\
+                                                    & Spark ML  & 0.04       & 8 hour 29 min        \\
+                                                    & ranger (R)      &  0.05       & 28 mins         \\
+\hline
+
+\multirow{3}{*}{phase3\_chr1-3: 2,504 x 19,328,051} & \cursedforest\ & 0.02  & 56 min 43 seconds             \\
                                                     & Spark ML &    NA         &   NA               \\
+                                                    & ranger (R)      & 0.03      &    3 hour 58 min     \\
+\hline
+
+\multirow{4}{*}{phase3\_chr1-22: 2,504 x 81,047,467} & \cursedforest\ & XX  & 7hour 14min \\
+                                                    & Spark ML & NA & NA  \\
+                                                    & ranger (R)       &        NA     &        NA    \\
                                                     & ranger (C++)       &        NA     &        NA    \\
 \hline
+
 \end{tabular}

### 2016-12-19 doc/publications/plos16/plos16.tex
changes to plos16.tex

@@
 
-\newcommand{\cursedforest}{{\sc CursedForest}}
+\newcommand{\cursedforest}{{\textsc{CursedForest}}
 \newcommand{\mtry}{{\texttt mtry}}
@@
 increasing feature size~\cite{Wright.and.Ziegle.2016}.  This is despite R now supporting long vectors (length greater
-than $2^{31}-1$), and the {\sc randomForest} implementation in R lending itself to parallelization to grow multiple
+than $2^{31}-1$), and the \textsc{randomForest} implementation in R lending itself to parallelization to grow multiple
 trees simultaneously~\cite{Liaw.and.Weiner.2002} and them combine them.  More suitable implementations for ``wide''
@@
 \end{equation*}
+In addition to growing the trees
+on bootstrapped samples, each node in the tree is split on a randomly
+selected subset of the variables.
 
@@
 So what is required for accurate random forest regression is (i) low correlation between residuals of differing tree in
-the forest, and (ii) low prediction error for the individual trees \cite{Segal.2004}. In addition to growing the trees
-on bootstrapped samples, each node in the tree is split on a randomly selected subset of the variables.
+the forest, and (ii) low prediction error for the individual trees \cite{Segal.2004}.
 
@@
 algorithms we demonstrate, we convert the original variant representations to double representations.  For example, we
-store the absence of a variant (0\textbar0) as 0, a heterozygous variant (1\textbar0 or 0\textbar1)as 1, and a
+store the absence of a variant (0\textbar0) as 0, a heterozygous
+variant (1\textbar0 or 0\textbar1) as 1, and a
 homozygous variant (1\textbar1) as 2.  This is not the only possible encoding.  \cite{Goldstein.et.al.2011} note that of
@@
 \subsection{Parameter settings}
-We consider the parameter setting for the random forest algorithm. We use the R notation from the {\sc randomForest}
+We consider the parameter setting for the random forest algorithm. We use the R notation from the \textsc{randomForest}
 package \cite{Liaw.and.Weiner.2002} which incorporates the original Fortran code by Brieman and Cutler.  We incorporate
@@
 \item the size and complexity of the individual trees is controlled in
-  {\sc randomForest} by setting \texttt{nodesize}, the minimum size of
+  \textsc{randomForest} by setting \texttt{nodesize}, the minimum size of
   terminal nodes. It is controlled in Spark ML by setting
@@
 \end{figure}
+
+\marginpar{Does \cloudforest use Gini impurity or information gain? I
+  thought it used infomation gain but Piotr mentions Gini}
 
@@
 
-Firstly we look at CursedForest horizontal scalability for a medium size dataset of 2.5 million variables and 5000
+Firstly we look at \cursedforest horizontal scalability for a medium size dataset of 2.5 million variables and 5000
 samples, by varying \mtry\ fraction and the number of CPU cores allocated to the execution. The raw results are
@@
 
+\begin{figure}[tbhp] 
+ \begin{adjustwidth}{-1.00in}{0in}
+    \centering
+    \includegraphics[totalheight=6cm]{./figs/phase-diagram-equivalence.png} 
+    \caption{{\bf The phase chage diagram.}}
+    \label{figure:phase-diagram-equivalence.png} 
+    \vspace{4ex}
+  \end{adjustwidth}
+\end{figure}
+
+\section{not for the paper}
 Notes:
@@
   \begin{adjustwidth}{-1.00in}{0in}
-  \begin{subfigure}[b]{0.4\linewidth}
+    \begin{subfigure}[b]{0.5\linewidth}
       \centering
-   \includegraphics[totalheight=6cm]{./figs/phase-diagram-equivalence.png} 
-\caption{{\bf The phase chage diagram.}}
-\label{figure:phase-diagram-equivalence.png} 
+      \includegraphics[totalheight=6cm]{./figs/k.png}
+      \caption{k}
+      \label{figure:k.png}
       \vspace{4ex}
     \end{subfigure} 
-  \begin{subfigure}[b]{0.6\linewidth}
+    \begin{subfigure}[b]{0.5\linewidth}
       \centering
-  \includegraphics[totalheight=6cm]{./figs/stepwise_lars_cv.png}
-\caption{Stepwise lars. $p=200$, $n$ and $k$ variable. Model selected by cross-validation
-}
-\label{figure:stepwise_lars_cv.png}
+      \includegraphics[totalheight=6cm]{./figs/n.png}
+      \caption{n}
+      \label{figure:n.png}
     \end{subfigure} 
@@
 \begin{figure}[tbhp] 
+  \begin{adjustwidth}{-1.00in}{0in}
+    \begin{subfigure}[t]{0.5\linewidth}
+      \centering
+      \includegraphics[totalheight=6cm]{./figs/stepwise_lars_cv.png}
+      \caption{Stepwise lars. $p=200$, $n$ and $k$ variable. Model selected by cross-validation}
+      \label{figure:stepwise_lars_cv.png}
+      \vspace{4ex}
+    \end{subfigure} 
+    \begin{subfigure}[t]{0.5\linewidth}
+      \centering
       \includegraphics[totalheight=6cm]{./figs/ranger.png}
-\caption{ ranger}
+      \caption{ranger, RBO}
       \label{figure:ranger.png}
+    \end{subfigure} 
+    \caption{}
+  \end{adjustwidth}
 \end{figure}

### 2016-12-20 doc/publications/plos16/plos16.tex
changes to plos16.tex
add new table and add values to existing

@@
 
-\newcommand{\cursedforest}{{\textsc{CursedForest}}
+\newcommand{\cursedforest}{\textsc{CursedForest}}
 \newcommand{\mtry}{{\texttt mtry}}
@@
 
-\marginpar{Does \cloudforest use Gini impurity or information gain? I
+\marginpar{Does \cursedforest use Gini impurity or information gain? I
   thought it used infomation gain but Piotr mentions Gini}
@@
 \begin{minipage}{\textwidth}
+\centering
+\caption{ {\bf Performance comparison with \cursedforest\ on different subsets of the 1000 Genomes Project data.}}
+\begin{tabular}{| l | r | l |}
+\hline
+{\bf Genome \%} & {\bf Error} & {\bf Runtime} \\ \thickhline
+phase1\_chr22: 1,092 x 490,036 & 0.04 & 1 min 50 seconds \\ \hline
+phase3\_chr22: 2,504 x 1,103,548 & 0.06 & 4 min 2 seconds \\ \hline
+phase3\_chr1: 2,504 x 6,450,364 & 0.04 & 19 min 19 seconds \\ \hline
+phase3\_chr1-3: 2,504 x 19,328,051 & 0.02 & 56 min 43 seconds \\ \hline
+phase3\_chr1-22: 2,504 x 81,047,467 & 0.01 & 36 min 54 seconds \\ \hline
+\end{tabular}
+\begin{flushleft} 
+Comparison of runtime and error for models built using \cursedforest\ on different subsets of variant data 
+from the 1000 Genomes Project.
+Each random forest model consists of 50 trees and was built using 130 worker nodes---each with 8GB RAM.
+\end{flushleft}
+\label{cursedforesttable}
+\end{minipage}
+\end{table}
+
+\begin{table}[!ht]
+\begin{minipage}{\textwidth}
 %\begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
@@
 \caption{ {\bf Performance comparison between the different random forest implementations.}}
-\begin{tabular}{|l|l|l|l|l|l||}
+\begin{tabular}{| l | l | r | l |}
 \hline
@@
                                                 & ranger (C++)     & 0.06       & 7 min            \\
+                                                & H2O           & ??       & ??         \\
 
@@
                                                     & ranger (R)    & 0.06       & 4 min 30 seconds     \\
+                                                    & ranger (C++)     &     ??   & ??        \\
                                                     & H2O           & 0.05       & 16 hours         \\
@@
                                                     & Spark ML  & 0.04       & 8 hour 29 min        \\
+                                                    & randomForest (R)        & NA         & NA                \\
                                                     & ranger (R)      &  0.05       & 28 mins         \\
+                                                    & ranger (C++)     & ??       & ??            \\
+                                                    & H2O           & NA       & NA         \\
 \hline
@@
                                                     & Spark ML &    NA         &   NA               \\
+                                                    & randomForest (R)        & NA         & NA                \\
                                                     & ranger (R)      & 0.03      &    3 hour 58 min     \\
+                                                    & ranger (C++)     & ??       & ??            \\
+                                                    & H2O           & NA       & NA         \\
 \hline
 
-\multirow{4}{*}{phase3\_chr1-22: 2,504 x 81,047,467} & \cursedforest\ & XX  & 7hour 14min \\
+\multirow{4}{*}{phase3\_chr1-22: 2,504 x 81,047,467} & \cursedforest\ & 0.01  & 3hour 16min \\
                                                     & Spark ML & NA & NA  \\
+                                                    & randomForest (R)        & NA         & NA                \\
                                                     & ranger (R)       &        NA     &        NA    \\
                                                     & ranger (C++)       &        NA     &        NA    \\
+                                                    & H2O           & NA       & NA         \\
 \hline

### 2016-12-20 doc/publications/plos16/plos16.tex
new times

@@
 {\bf Genome \%} & {\bf Error} & {\bf Runtime} \\ \thickhline
-phase1\_chr22: 1,092 x 490,036 & 0.04 & 1 min 50 seconds \\ \hline
-phase3\_chr22: 2,504 x 1,103,548 & 0.06 & 4 min 2 seconds \\ \hline
-phase3\_chr1: 2,504 x 6,450,364 & 0.04 & 19 min 19 seconds \\ \hline
-phase3\_chr1-3: 2,504 x 19,328,051 & 0.02 & 56 min 43 seconds \\ \hline
+phase1\_chr22: 1,092 x 490,036 & 0.04 & 1 min 5 seconds \\ \hline
+phase3\_chr22: 2,504 x 1,103,548 & 0.06 & 1 min 37 seconds \\ \hline
+phase3\_chr1: 2,504 x 6,450,364 & 0.04 & 4 min 10 seconds \\ \hline
+phase3\_chr1-3: 2,504 x 19,328,051 & 0.02 & 10 min 49 seconds \\ \hline
 phase3\_chr1-22: 2,504 x 81,047,467 & 0.01 & 36 min 54 seconds \\ \hline

### 2017-01-10 doc/publications/plos16/plos16.tex
changed figures
changes to plos16.tex

@@
 
-\newcommand{\cursedforest}{\textsc{CursedForest}}
-\newcommand{\mtry}{{\texttt mtry}}
-\newcommand{\ntree}{{\texttt ntree}}
+\newcommand{\cursedforest}{\textsc{CursedForest}\ }
+\newcommand{\mtry}{{\texttt mtry\ }}
+\newcommand{\ntree}{{\texttt ntree\ }}
 
@@
 termed single nucleotide polymorphisms (SNPs), by applying linear or logistic regression analysis to each SNP
-separately~\cite{CCC2007}.  It has since been demonstrated that not taking joint interactions between SNPs into account
+separately~\cite{CCC2007}.  It has since been demonstrated that not taking interactions between SNPs into account
 when filtering for potential disease mutations is inadequate~\cite{Manolio2009,Yang2011}.
 
-A tree based model is capable of taking this interaction of variables into account \cite{Wright.et.al.2016}. In addition, random forests are well
+A tree based model is capable of taking the interaction of variables into account \cite{Wright.et.al.2016}. In addition, random forests are well
 suited for processing ``wide'' genomic data for two reasons.  Firstly, while other machine learning applications have
@@
 
-However, the use of traditional compute infrastructure limits the parallelization strategies that can be employed by  
+The use of traditional compute infrastructure limits the parallelization strategies that can be employed by  
 these methods.  The programs are limited to utilizing only CPUs that are on the same computer node (multi-threading) or
@@
 for ``big'' (many samples) and ``wide'' (many features) data sets. A Spark application runs on a ``driver'' node and
-distributes tasks to the many ``worker'' nodes, or ``executors''. Using this facility, \cursedforest\ is capable of
+distributes tasks to many ``worker'' nodes, or ``executors''. Using this facility, \cursedforest\ is capable of
 parallelization the split for each node in a tree thereby handling millions of features. It can, for example, fit a
@@
 trees, which boosts the accuracy for variable importance selection~\cite{Szymczak2016}.
+\marginpar{check this. It also allows us to set mtry to a large number, important when p is large}
 
@@
 \end{equation}
-where $\rho$ is the weighted correlation between residuals $(Y-h(X;B))$ and $(Y-h(X;B^\prime))$ for independent $B,B^\prime$.  
+where $\rho$ is the weighted correlation between residuals $(Y-h(X;B))$ and $(Y-h(X;B^\prime))$ for independent
+$B,B^\prime$.
+\marginpar{why is this true}  
 So what is required for accurate random forest regression is (i) low correlation between residuals of differing tree in
@@
 The current implementation of \cursedforest\ uses an ``information gained'' criteria for splitting. Let $f_q$ be the
+\marginpar{this is wrong -- it used Gini}
 fraction of items labelled with value $q$ where $q = 1, \ldots, Q$ at a node. The entropy of the node is
@@
 \subsubsection{Synthetic data} 
+\marginpar{this is wrong now}
 Each dataset consists of $n$ samples and $p$ variables where $n << p$,
@@
 Fig.~\ref{figure:synth1} illustrates the effect of varying  \mtry\ and \ntree\ on these two measures. 
-Fig.~\ref{figure:out-of-bag-prediction-error-prod.png} shows that depends strongly on \ntree
+Fig.~\ref{figure:out-of-bag-prediction-error-prod.png} shows that depends strongly on \ntree\ 
 whereas Fig.~\ref{figure:rbo-prod.png} illustrates that the rbo measure is more dependent on \mtry.
+\marginpar{this is not true anymore}
 
@@
   \begin{adjustwidth}{-1.00in}{0in}
+    \caption{\textbf{ OOB (out of bag) classification error and Rank-biased overlap estimates for random
+        forests with 100 trees trained on the synthetic dataset of 2.5M features and 5K samples.}}
+    \label{figure:synth1}
     \begin{subfigure}[b]{0.5\linewidth}
       \centering
-    \includegraphics[totalheight=8cm]{./figs/out-of-bag-prediction-error-prod.png}
+      \includegraphics[totalheight=8cm]{./figs/out-of-bag-prediction-error.png}
       \caption{OOB classification error} 
@@
       \centering
-    \includegraphics[totalheight=8cm]{./figs/rbo-prod.png}
+      \includegraphics[totalheight=8cm]{./figs/rbo.png}
       \caption{The  rank-biased-overlap measure.} 
@@
     \end{subfigure} 
-  \caption{{\bf Variable importance (Mean Decrease Gini) and OOB (out of bag) classification error estimate for random
-      forests with 100 trees trained on the synthetic dataset of 2.5M features and 5K samples.}  Each measure is shown
+    \begin{flushleft} 
+      Each measure is shown
       as a function of \mtry\ and \ntree. Note that the plot colors have been chose so that the lighter color indicates
-    the better result, that is, a lower OOB and a higher RBO.}
-  \label{figure:synth1}
+      the better result, that is, a lower OOB and a higher RBO.
+    \end{flushleft}
   \end{adjustwidth}
 \end{figure}
-
-\marginpar{Does \cursedforest use Gini impurity or information gain? I
-  thought it used infomation gain but Piotr mentions Gini}
 
@@
 \end{tabular}
-\begin{flushleft} Here the importance is the Gini impurity measure when building a model using the cell type as the label.
+\begin{flushleft} 
+Here the importance is the Gini impurity measure when building a model using the cell type as the label.
 \end{flushleft}
@@
 In order to asses these characteristics we ran \cursedforest\ classification on synthetics datasets with varying numbers
-of variables (features) and samples, similar to the dataset used in to evaluate ranger \cite{Wright.and.Ziegle.2016},
-allocating varying number of CPU cores to the \cursedforest\ and also varying computational complexity of random forests
-by using a range of \mtry\ values.
+of variables (features) and samples, similar to the dataset used in \cite{Wright.and.Ziegle.2016} to evaluate
+\textsc{Ranger}, allocating varying number of CPU cores to the \cursedforest\ and also varying computational complexity
+of random forests by using a range of \mtry\ values.
 
@@
 
-Firstly we look at \cursedforest horizontal scalability for a medium size dataset of 2.5 million variables and 5000
-samples, by varying \mtry\ fraction and the number of CPU cores allocated to the execution. The raw results are
-presented in table \ref{synthetictimingtable} and visualized in figure \ref{figure:synth} below. Regardless of the
-number of cores used, \cursedforest\ displays approximately linear dependency between the execution time and the \mtry\
-parameter (presented here as a fraction of the number of variables) [Fig 1 a).
+First we look at \cursedforest horizontal scalability for a medium size dataset of 2.5 million variables and 5000
+samples, by varying the \mtry  fraction and the number of CPU cores allocated to the execution. The raw results are
+presented in table \ref{synthetictimingtable} and visualized in figure \ref{figure:synthetictiming} below. Regardless of the
+number of cores used, \cursedforest displays approximately linear dependency between the execution time and the \mtry
+parameter (presented here as a fraction of the number of variables, Fig. \ref{figure:synthetictiming.a}
 
 \cursedforest\ scales almost linearly with the number of CPU cores for medium values of \mtry\ fraction but for both
-lower and higher values the performance degrades slightly (Fig 1 b). In the latter case the likely cause is
+lower and higher values the performance degrades slightly (Fig \ref{figure:synthetictiming.b}). In the latter case the likely cause is
 communication overhead (with lower \mtry\ values the proportion of time for parallelizable computation to the time for
@@
   \begin{adjustwidth}{-1.00in}{0in}
-\begin{tabular}{ll}
-a)& b)\\
-\includegraphics[totalheight=6cm]{./figs/mtry_cpu.png} & 
-\includegraphics[totalheight=6cm]{./figs/cpu_mtry_trees_per_hour.png} \\
-\end{tabular}
-\caption{{\bf Scalablity of the wide random forest on the synthetic dataset of 2.5M features and 5K samples.} 
-  {\bf a)} Time in seconds to build 100 trees for different \mtry\ fractions. 
-  {\bf b)} Number of trees build per hour when using a growing number of CPU cores.}
-\label{figure:synth}
+    \caption{\textbf{Scalablity of the wide random forest on the synthetic dataset of 2.5M features and 5K samples.}}
+    \label{figure:synthetictiming}
+    \begin{subfigure}[b]{0.5\linewidth}
+      \centering
+      \includegraphics[totalheight=6cm]{./figs/mtry_cpu.png} 
+      \caption{Time in seconds to build 100 trees for different \mtry\ fractions. } 
+      \label{figure:synthetictiming.a} 
+      \vspace{4ex}
+    \end{subfigure} 
+    \begin{subfigure}[b]{0.5\linewidth}
+      \centering
+      \includegraphics[totalheight=6cm]{./figs/cpu_mtry_trees_per_hour.png}
+      \caption{Number of trees build per hour when using a growing number of CPU cores.}
+    \label{figure:synthetictiming.b}
+    \vspace{4ex}
+  \end{subfigure} 
 \end{adjustwidth}
@@
 \centering
-\caption{
-{\bf Time in seconds to build a random forest of 100
+\caption{\textbf{
+ Time in seconds to build a random forest of 100
   trees on 128 CPU cores with varying numbers of variables and samples.}}
+\label{synthetictimingtable}
 \begin{tabular}{|l|l|l|l|l|l|l|}
@@
 \end{tabular}
-\begin{flushleft} 
-\end{flushleft}
-\label{synthetictimingtable}
 \end{adjustwidth}
@@
   \begin{adjustwidth}{-1.00in}{0in}
-\begin{tabular}{ll}
-a)& b)\\
-\includegraphics[totalheight=6cm]{./figs/nvars_nsamples.png} & 
-\includegraphics[totalheight=6cm]{./figs/nsamples_nvars.png} \\
-\end{tabular}
-\caption{{\bf Scalablity of the Wide Random Forest on synthetic datasets with varying number of samples and variables.} 
-  {\bf a)} Number of trees build per hour when growing number of variables. 
-  {\bf b)} Number of trees build per hour when growing number of samples.}
-\label{figure:synth}
+    \caption{\textbf{Scalablity of the Wide Random Forest on synthetic datasets with varying number of samples and variables.}}
+    \label{figure:synth2}
+    \begin{subfigure}[b]{0.5\linewidth}
+      \centering
+      \includegraphics[totalheight=6cm]{./figs/nvars_nsamples.png} 
+      \caption{Number of trees build per hour when growing number of variables.} 
+      \label{figure:synth2.a} 
+      \vspace{4ex}
+    \end{subfigure} 
+    \begin{subfigure}[b]{0.5\linewidth}
+      \centering
+      \includegraphics[totalheight=6cm]{./figs/nsamples_nvars.png}
+      \caption{Number of trees build per hour when growing number of samples.}
+    \label{figure:synth2.b}
+    \vspace{4ex}
+  \end{subfigure} 
 \end{adjustwidth}
 \end{figure}
+
+% \begin{figure}[tbhp]
+% \begin{adjustwidth}{-1.00in}{0in}
+% \begin{tabular}{ll}
+% a)& b)\\
+% \includegraphics[totalheight=6cm]{./figs/nvars_nsamples.png} & 
+% \includegraphics[totalheight=6cm]{./figs/nsamples_nvars.png} \\
+% \end{tabular}
+% \caption{{\bf Scalablity of the Wide Random Forest on synthetic datasets with varying number of samples and variables.} 
+%   {\bf a)} Number of trees build per hour when growing number of variables. 
+%   {\bf b)} Number of trees build per hour when growing number of samples.}
+% \label{figure:synth}
+% \end{adjustwidth}
+% \end{figure}
 

### 2017-01-11 doc/publications/plos16/plos16.tex
changes to plos16.tex

@@
 
-\newcommand{\cursedforest}{\textsc{CursedForest}\ }
-\newcommand{\mtry}{{\texttt mtry\ }}
-\newcommand{\ntree}{{\texttt ntree\ }}
+\newcommand{\cursedforest}{\textsc{CursedForest}\xspace}
+\newcommand{\mtry}{\texttt{mtry}\xspace}
+\newcommand{\ntree}{\texttt{ntree}\xspace}
 
@@
 \usepackage{placeins}
+\usepackage{xspace} % see http://tex.stackexchange.com/questions/31091/space-after-latex-commands
+
 \begin{document}
@@
 
+\marginpar{make consistent capitalization for ranger, randomForest}
+
 The use of traditional compute infrastructure limits the parallelization strategies that can be employed by  
@@
 for ``big'' (many samples) and ``wide'' (many features) data sets. A Spark application runs on a ``driver'' node and
-distributes tasks to many ``worker'' nodes, or ``executors''. Using this facility, \cursedforest\ is capable of
+distributes tasks to many ``worker'' nodes, or ``executors''. Using this facility, \cursedforest is capable of
 parallelization the split for each node in a tree thereby handling millions of features. It can, for example, fit a
@@
 directly, \cursedforest\ outperforms existing tools even on small datasets where multi-threading generally performs
-well. Harnessing the virtually unlimited capability to parallelize tasks, \cursedforest\ can build a large number of
-trees, which boosts the accuracy for variable importance selection~\cite{Szymczak2016}.
-\marginpar{check this. It also allows us to set mtry to a large number, important when p is large}
+well. Harnessing the virtually unlimited capability to parallelize tasks, \cursedforest can build a large number of
+trees and use a high \mtry value which leads to greater accurace (see section \ref{section:synthetic}). 
+\marginpar{does \cite{Szymczak2016} say that high \ntree gives greater  accuracy for variable importance selection?
+We do not see that}
 
@@
 $B,B^\prime$.
-\marginpar{why is this true}  
 So what is required for accurate random forest regression is (i) low correlation between residuals of differing tree in
@@
 
-The current implementation of \cursedforest\ uses an ``information gained'' criteria for splitting. Let $f_q$ be the
-\marginpar{this is wrong -- it used Gini}
-fraction of items labelled with value $q$ where $q = 1, \ldots, Q$ at a node. The entropy of the node is
-$H = - \sum^{m}_{i=1} f_i \log^{}_2 f_i.$ The expected information gain is the change in entropy $H$ from a prior state
-$T$ to a state that takes some information $a$ as given:
-\[ IG(T,a) = H(T) - H(T|a). \]
-In this case, the information gain is IG = Entropy(parent node) - Weighted Sum of Entropy(Children). 
-Variable importance is the sum of IG over all notes and all trees for that variable.
+The current implementation of \cursedforest\ uses an ``Gini impurity''
+criteria for splitting. Let $f_q$ be the fraction of items labelled
+with value $q$ where $q = 1, \ldots, Q$ at a node. The Gini impurity is
+\[
+I_G(f) = \Sigma_{q = 1}^Q f_q ( 1 − f_q ), 
+\]
+which is a minimum when all observatkions at the node are in the same class.
 
@@
 \subsubsection{Synthetic data} 
-\marginpar{this is wrong now}
  Each dataset consists of $n$ samples and $p$ variables where $n << p$,
@@
 The model paramters are $w_i = 1/\sqrt{2^{i-1}}$ for $i =1,\ldots 5$ then 
-\begin{align}
-y=& \prod_{i=1}^{i=5} {w_i x_i} + \epsilon \textrm{ for $x_i \neq 0$} 
-\intertext{where}
-\epsilon \sim & N(0, \sigma_\epsilon^2= 0.12)
-\end{align}
-The dichotomous response is generated by thresholding $y$ at the $0.8$ quantile
+\begin{align*}
+z=& \Sigma_{i=1}^{i=5} {w_i x_i} \\
+\intertext{and}
+\sigma_\epsilon^2 =& Var(z)(1- 0.125)/0.125 \\
+\intertext{so}
+y =&  z+\epsilon \textrm{ where } \epsilon \sim  N(0, \sigma_\epsilon^2).
+\end{align*}
+The dichotomous response is generated by thresholding $y$ at the $0.8$ quantile.
 
@@
 \begin{itemize}
-\item \ntree\ the number of trees.  The number of trees necessary for good performance grows with the number of
+\item \ntree{}  -- the number of trees.  The number of trees necessary for good performance grows with the number of
   predictors.  \cite{Liaw.and.Weiner.2002} suggest that a high \ntree\ is necessary to get stable estimates of variable
   importance and proximity; however, even though the variable importance measures may vary from run to run, we note that
-  it is possible for a random forest model to have a very poor fit and still have an accurate ranking of variable
+  it is possible for a random forest model to have a poorer fit and still have an accurate ranking of variable
   importance;
-\item \mtry\ -- the number of variables considered at each split (if \mtry=$p$, we have a boosted decision
+\item \mtry{}  -- the number of variables considered at each split (if \mtry=$p$, we have a boosted decision
   tree model).  If one has a very large number of variables but expects only very few to be ``important'', using larger \mtry\ may give
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{\cursedforest\ can evaluate variable importance for millions of features}\label{synthetic}
+\subsection{\cursedforest\ can evaluate variable importance for millions of features}
+\label{section:synthetic}
 In this section we demonstrate two features of \cursedforest\, a variable importance analysis to select the most
@@
 classification error and the rank-biased-overlap
+\marginpar{what is wrong with the formatting of this citation?}
 \cite[]{Webber.et.al.2010} measure.
@@
 
-Fig.~\ref{figure:synth1} illustrates the effect of varying  \mtry\ and \ntree\ on these two measures. 
-Fig.~\ref{figure:out-of-bag-prediction-error-prod.png} shows that depends strongly on \ntree\ 
-whereas Fig.~\ref{figure:rbo-prod.png} illustrates that the rbo measure is more dependent on \mtry.
-\marginpar{this is not true anymore}
+\marginpar{this was done with n=10000. Needs to redone with n=2.5M and some things re-written}
+Fig.~\ref{figure:synth1} illustrates the effect of varying \mtry and \ntree on these two measures. The default \mtry
+value for a classification problem is $\sqrt{p}=100$ so in this case we are considering values considerably larger than
+the default. Fig.~\ref{figure:out-of-bag-prediction-error-prod.png} shows the OOB error is minimized at the largerst
+values of \mtry and \ntree considered  whereas Fig.~\ref{figure:rbo-prod.png} illustrates that the rbo measure is less
+sensitive once we over a certain threshold.
 
@@
 
-\marginpar{this was done with n=100000. Needs to redone with n=2.5M}
-
 \begin{figure}[tbhp] 
   \begin{adjustwidth}{-1.00in}{0in}
-    \caption{\textbf{ OOB (out of bag) classification error and Rank-biased overlap estimates for random
-        forests with 100 trees trained on the synthetic dataset of 2.5M features and 5K samples.}}
+    \caption{\textbf{ out of bag classification error and Rank-biased overlap estimates}} 
     \label{figure:synth1}
@@
     \begin{flushleft} 
-      Each measure is shown
-      as a function of \mtry\ and \ntree. Note that the plot colors have been chose so that the lighter color indicates
-      the better result, that is, a lower OOB and a higher RBO.
+      Random forests with 100 trees were trained on the synthetic dataset of 2.5M features and 5K samples.
+      Each measure is shown as a function of \mtry and \ntree. Note that the plot colors have been chosen so that the lighter color indicates
+      the better result, that is, a lower OOB or a higher RBO.
     \end{flushleft}
@@
 1000 Genomes Project dataset.
+
+\marginpar{this section is not one sentence? has it been lost?}
 
@@
 \textsc{Ranger}, allocating varying number of CPU cores to the \cursedforest\ and also varying computational complexity
-of random forests by using a range of \mtry\ values.
+of random forests by using a range of \mtry values.
 
@@
 presented in table \ref{synthetictimingtable} and visualized in figure \ref{figure:synthetictiming} below. Regardless of the
-number of cores used, \cursedforest displays approximately linear dependency between the execution time and the \mtry
-parameter (presented here as a fraction of the number of variables, Fig. \ref{figure:synthetictiming.a}
-
-\cursedforest\ scales almost linearly with the number of CPU cores for medium values of \mtry\ fraction but for both
+number of cores used, \cursedforest displays approximately linear dependency between the execution time and \mtry
+(Fig. \ref{figure:synthetictiming.a}).
+
+\cursedforest scales almost linearly with the number of CPU cores for medium values of \mtry fraction but for both
 lower and higher values the performance degrades slightly (Fig \ref{figure:synthetictiming.b}). In the latter case the likely cause is
-communication overhead (with lower \mtry\ values the proportion of time for parallelizable computation to the time for
+communication overhead (with lower \mtry values the proportion of time for parallelizable computation to the time for
 internode communication is lower) while in the latter case it is most likely caused by reaching the clusters
@@
 
-Next we investigate \cursedforest\ scalability with regards to the size of data, by varying the number of variables and
-sample for a fixed \mtry\ fraction of 0.25 and execution of 128 CPU cores. The raw results are presented in the table
-[REF] and visualized in figure [ REF: Fig 2] below (please note log scale on the axes and the values on y axes are
+Next we investigate \cursedforest scalability with regards to the size of data, by varying the number of variables and
+sample for a fixed \mtry fraction of 0.25 and execution of 128 CPU cores. The raw results are presented in the table
+[REF] 
+\marginpar{have we lost this table?}
+and visualized in figure \ref{figure:synth2} below (please note log scale on the axes and the values on y axes are
 expresses as trees per hour).
@@
 irregularities in the graph can be attributed to computation vs communication tradeoff. It is also worth noting that
-keeping \mtry\ fraction constants results with higher \mtry\ values with growing number of variables, and this is what
+keeping the \mtry fraction constant results in higher \mtry values with the growing number of variables, and this is what
 drives the performance down rather that the increase of dataset size itself.
@@
   \begin{adjustwidth}{-1.00in}{0in}
-    \caption{\textbf{Scalablity of the wide random forest on the synthetic dataset of 2.5M features and 5K samples.}}
+    \caption{\textbf{Scalability of the wide random forest on the synthetic dataset of 2.5M features and 5K samples.}}
     \label{figure:synthetictiming}
-    \begin{subfigure}[b]{0.5\linewidth}
+    \begin{subfigure}[b]{0.45\linewidth}
       \centering
@@
     \end{subfigure} 
-    \begin{subfigure}[b]{0.5\linewidth}
+\hfill
+    \begin{subfigure}[b]{0.45\linewidth}
       \centering
@@
 \section{Conclusion}
-We have demonstrated that using a different paralleization model can extend random forests to the case of an
+We have demonstrated that using a different parallelization model can extend random forests to the case of an
 extremely large number of variables. We have treated the case of variable selection in a $p >> n$ model where most of
@@
 the figure.
+
+We note several things here:
+\begin{itemize}
+\item 
+\end{itemize}
 

### 2017-01-11 doc/publications/plos16/plos16.tex
Rewrite RDD section

--- 2017-01-10 doc/publications/plos16/plos16.tex
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{\cursedforest.}
-VariantSpark made use of Spark's machine learning algorithms ``Spark ML''. These algorithms require data in a
-``DataFrame'', similar to DataFrames in R.  While Spark ML algorithms demonstrate some scalability when dealing with a
-large number of samples, this scalability eventually break down as we include more features.
-
-One of the restrictions preventing Spark ML algorithms from scaling to high-dimensionality data is that the feature-set
-for each sample is stored as a vector data type. Spark vectors are stored in resilient distributed dataset (RDDs), and
-while RDDs can be split, vectors cannot. So as we add more samples (in our case individuals), Spark ML can simply split
-the RDD into more partitions and resource requirements will remain relatively stable. However, as the number of
-dimensions (variants) increases, the size of the vectors increases. This results in higher memory-requirements as
-increasingly large vectors must be loaded into memory.  \marginpar{not clear. Are RDDs the problem or the solutuion}
+\subsection{\cursedforest}
+VariantSpark made use of Spark's machine learning algorithms ``Spark ML''. While Spark ML algorithms demonstrate
+scalability when dealing with a large number of samples, this scalability eventually breaks down as we include more
+features.
+
+As is standard with Spark applications, we store our data in a Resilient Distributed Dataset (RDD), where an RDD is
+essentially a collection of elements. In the case of Spark ML, each element in our RDD is a sample. RDDs contribute to
+the scalability of Spark as they can distributed across multiple nodes and operated on in parallel. Even as we add more
+samples to a dataset, Spark can still simply schedule extra tasks to handle the additional items in the RDD.
+
+However, within an RDD, spark ML stores each sample as a vector. Unlike RDDs, which can be partitioned and distributed
+across multiple nodes, each vector must be present in its entirity on any node accessing it. This is no problem with
+typical datasets, however as dimensionality increases, the vectors eventually reach a size where they can no longer fit
+into a single node's memory.
+
+So in the case of adding more samples, Spark ML can simply create more tasks, keeping memory consumption within the
+cluster's bounds. However, as the dimensionality of each sample grows, the memory requirements of the job increases to
+enable these increasingly large vectors to be loaded into memory.
+\marginpar{not clear. Are RDDs the problem or the solutuion Aidan: Better?}
 

### 2017-01-11 doc/publications/plos16/plos16.tex
changes to plos16.tex

@@
 As is standard with Spark applications, we store our data in a Resilient Distributed Dataset (RDD), where an RDD is
-essentially a collection of elements. In the case of Spark ML, each element in our RDD is a sample. RDDs contribute to
+essentially a collection of elements. In the case of Spark ML, each element in the RDD is a sample. RDDs contribute to
 the scalability of Spark as they can distributed across multiple nodes and operated on in parallel. Even as we add more
@@
 enable these increasingly large vectors to be loaded into memory.
-\marginpar{not clear. Are RDDs the problem or the solutuion Aidan: Better?}
-
-On the other hand, \cursedforest\ is specifically designed to handle wide ``cursed'' data. It avoids the relation
-between memory and dimensionality by avoiding calculations that rely on entire feature vectors.
-
-Compared to Spark ML, \cursedforest\ further parallelizes work down to individual features.  For example, for each node
-of a tree, \cursedforest\ will distribute tasks that consist of single features (variants), for every individual.  Each
-of these tasks will calculate the information gain for that specific feature.  Once these tasks have completed, the
-results are reduced to return the feature which gives the greatest information gain.  This process is then repeated
-until \cursedforest\ has created the entire decision tree.
-
-The current implementation of \cursedforest\ uses an ``Gini impurity''
-criteria for splitting. Let $f_q$ be the fraction of items labelled
-with value $q$ where $q = 1, \ldots, Q$ at a node. The Gini impurity is
+
+On the other hand, \cursedforest is specifically designed to handle wide ``cursed'' data. It avoids the relation between
+memory and dimensionality by avoiding calculations that rely on entire feature vectors and taking the parallelization
+work down to the level of the individual features.  For example, for each node of a tree, \cursedforest will distribute
+tasks that consist of single features (variants), for every individual.  Each of these tasks will calculate the
+information gain for that specific feature.  Once these tasks have completed, the results are reduced to return the
+feature which gives the greatest information gain.  This process is then repeated until \cursedforest has created the
+entire decision tree.
+
+The current implementation of \cursedforest uses an ``Gini impurity'' criteria for splitting. Let $f_q$ be the fraction
+of items labelled with value $q$ where $q = 1, \ldots, Q$ at a node. The Gini impurity is
 \[
@@
 \item \ntree{}  -- the number of trees.  The number of trees necessary for good performance grows with the number of
-  predictors.  \cite{Liaw.and.Weiner.2002} suggest that a high \ntree\ is necessary to get stable estimates of variable
+  predictors.  \cite{Liaw.and.Weiner.2002} suggest that a high \ntree is necessary to get stable estimates of variable
   importance and proximity; however, even though the variable importance measures may vary from run to run, we note that
@@
 \item \mtry{}  -- the number of variables considered at each split (if \mtry=$p$, we have a boosted decision
-  tree model).  If one has a very large number of variables but expects only very few to be ``important'', using larger \mtry\ may give
+  tree model).  If one has a very large number of variables but expects only very few to be ``important'', using larger \mtry may give
   better performance;

### 2017-01-12 doc/publications/plos16/plos16.tex
added abstract to plos16.tex

@@
 \newcommand{\cursedforest}{\textsc{CursedForest}\xspace}
+\newcommand{\ranger}{\textsc{Ranger}\xspace}
 \newcommand{\mtry}{\texttt{mtry}\xspace}
@@
 \section{Abstract}
-The abstract.
+We introduce \cursedforest, an implementation of random forests designed to handle data that may have an extremely large
+number of variables. We discuss the utility of random forests for machine learning problems. We show on simulated and
+real data that the novel implementation can achieve reasonable prediction accuracy and variable importance estimates
+even in setting where other implementations, including some designed for large data, fail.
 
@@
 achieves fast run times by effective use of the CPU cache, optimizing for different classes of features and efficiently
-multi-threading.  Similarly, \textsc{Ranger}~\cite{Wright.and.Ziegle.2016} written in C++ with an R front-end is
+multi-threading.  Similarly, \ranger~\cite{Wright.and.Ziegle.2016} written in C++ with an R front-end is
 specifically optimized for dealing with large data sets.

### 2017-01-12 doc/publications/plos16/plos16.tex
consistent capitalization for package names plos16.tex

@@
 \newcommand{\ranger}{\textsc{Ranger}\xspace}
+\newcommand{\randomforest}{\textsc{randomForest}\xspace}
 \newcommand{\mtry}{\texttt{mtry}\xspace}
@@
 increasing feature size~\cite{Wright.and.Ziegle.2016}.  This is despite R now supporting long vectors (length greater
-than $2^{31}-1$), and the \textsc{randomForest} implementation in R lending itself to parallelization to grow multiple
+than $2^{31}-1$), and the \randomforest implementation in R lending itself to parallelization to grow multiple
 trees simultaneously~\cite{Liaw.and.Weiner.2002} and them combine them.  More suitable implementations for ``wide''
@@
 \subsection{Parameter settings}
-We consider the parameter setting for the random forest algorithm. We use the R notation from the \textsc{randomForest}
+We consider the parameter setting for the random forest algorithm. We use the R notation from the \randomforest
 package \cite{Liaw.and.Weiner.2002} which incorporates the original Fortran code by Brieman and Cutler.  We incorporate
@@
 \item the size and complexity of the individual trees is controlled in
-  \textsc{randomForest} by setting \texttt{nodesize}, the minimum size of
+  \randomforest by setting \texttt{nodesize}, the minimum size of
   terminal nodes. It is controlled in Spark ML by setting
@@
                                                 & Spark ML  &  0.05           &           7 min 44 seconds         \\
-                                                & randomForest (R)         & 0.05       & 24 min           \\
-                                                & ranger (R)      & 0.06       & 1 min 10 seconds       \\
-                                                & ranger (C++)     & 0.06       & 7 min            \\
+                                                & \randomforest (R)         & 0.05       & 24 min           \\
+                                                & \ranger (R)      & 0.06       & 1 min 10 seconds       \\
+                                                & \ranger (C++)     & 0.06       & 7 min            \\
                                                 & H2O           & ??       & ??         \\
@@
                                                     & Spark ML  & 0.06 & 37 min 19 seconds \\
-                                                    & randomForest (R)        & NA         & NA                \\
-                                                    & ranger (R)    & 0.06       & 4 min 30 seconds     \\
-                                                    & ranger (C++)     &     ??   & ??        \\
+                                                    & \randomforest (R)        & NA         & NA                \\
+                                                    & \ranger (R)    & 0.06       & 4 min 30 seconds     \\
+                                                    & \ranger (C++)     &     ??   & ??        \\
                                                     & H2O           & 0.05       & 16 hours         \\
@@
                                                     & Spark ML  & 0.04       & 8 hour 29 min        \\
-                                                    & randomForest (R)        & NA         & NA                \\
-                                                    & ranger (R)      &  0.05       & 28 mins         \\
-                                                    & ranger (C++)     & ??       & ??            \\
+                                                    & \randomforest (R)        & NA         & NA                \\
+                                                    & \ranger (R)      &  0.05       & 28 mins         \\
+                                                    & \ranger (C++)     & ??       & ??            \\
                                                     & H2O           & NA       & NA         \\
@@
                                                     & Spark ML &    NA         &   NA               \\
-                                                    & randomForest (R)        & NA         & NA                \\
-                                                    & ranger (R)      & 0.03      &    3 hour 58 min     \\
-                                                    & ranger (C++)     & ??       & ??            \\
+                                                    & \randomforest (R)        & NA         & NA                \\
+                                                    & \ranger (R)      & 0.03      &    3 hour 58 min     \\
+                                                    & \ranger (C++)     & ??       & ??            \\
                                                     & H2O           & NA       & NA         \\
@@
                                                     & Spark ML & NA & NA  \\
-                                                    & randomForest (R)        & NA         & NA                \\
-                                                    & ranger (R)       &        NA     &        NA    \\
-                                                    & ranger (C++)       &        NA     &        NA    \\
+                                                    & \randomforest (R)        & NA         & NA                \\
+                                                    & \ranger (R)       &        NA     &        NA    \\
+                                                    & \ranger (C++)       &        NA     &        NA    \\
                                                     & H2O           & NA       & NA         \\
@@
 of variables (features) and samples, similar to the dataset used in \cite{Wright.and.Ziegle.2016} to evaluate
-\textsc{Ranger}, allocating varying number of CPU cores to the \cursedforest\ and also varying computational complexity
+\ranger, allocating varying number of CPU cores to the \cursedforest\ and also varying computational complexity
 of random forests by using a range of \mtry values.

### 2017-01-12 doc/publications/plos16/plos16.tex
changes to  plos16.tex

@@
 {\Large
-\textbf\newline{\cursedforest\ - A Random Forest Implementation for ``Big'' and ``Wide'' Data} % Please use "title case" (capitalize all terms in the title except conjunctions, prepositions, and articles).
+\textbf\newline{\cursedforest - A Random Forest Implementation for ``Big'' and ``Wide'' Data} % Please use "title case" (capitalize all terms in the title except conjunctions, prepositions, and articles).
 }
@@
 
-\marginpar{make consistent capitalization for ranger, randomForest}
-
 The use of traditional compute infrastructure limits the parallelization strategies that can be employed by  
@@
 variants~\cite{1KG2012}.  By also utilizing Spark to read in and manipulate the standard genomic variant format (VCF)
-directly, \cursedforest\ outperforms existing tools even on small datasets where multi-threading generally performs
+directly, \cursedforest outperforms existing tools even on small datasets where multi-threading generally performs
 well. Harnessing the virtually unlimited capability to parallelize tasks, \cursedforest can build a large number of
-trees and use a high \mtry value which leads to greater accurace (see section \ref{section:synthetic}). 
+trees and use a high \mtry value which leads to greater accuracy (see the example in section \ref{section:synthetic}). 
 \marginpar{does \cite{Szymczak2016} say that high \ntree gives greater  accuracy for variable importance selection?
-We do not see that}
+We do not see that. }
 
@@
 
-\cursedforest\ is included in our previously developed framework for Spark-based genomic data analysis,
+\cursedforest is included in our previously developed framework for Spark-based genomic data analysis,
 VariantSpark~\cite{OBrien2015}, which therefore now enables supervised and unsupervised machine learning. This offers a
@@
 In the first section we explore some of the properties of random forests as these will also be apparent in the
-\cursedforest\ implementation. Here we provide demonstrations using synthetic data.  We then demonstrate the scalability
-of \cursedforest\ in respect to the dimensionality of the data, building a random forest model on whole-genome data from
+\cursedforest implementation. Here we provide demonstrations using synthetic data.  We then demonstrate the scalability
+of \cursedforest in respect to the dimensionality of the data, building a random forest model on whole-genome data from
 the 1000 Genomes Project. We report the error in predicting ethnicity.  Finally, given the role
@@
 the scalability of Spark as they can distributed across multiple nodes and operated on in parallel. Even as we add more
-samples to a dataset, Spark can still simply schedule extra tasks to handle the additional items in the RDD.
+samples to a dataset, Spark can simply schedule extra tasks to handle the additional items in the RDD.
 
@@
 memory and dimensionality by avoiding calculations that rely on entire feature vectors and taking the parallelization
-work down to the level of the individual features.  For example, for each node of a tree, \cursedforest will distribute
+work down to the level of the individual features.  For each node of a tree, \cursedforest will distribute
 tasks that consist of single features (variants), for every individual.  Each of these tasks will calculate the
@@
 of items labelled with value $q$ where $q = 1, \ldots, Q$ at a node. The Gini impurity is
-\[
-I_G(f) = \Sigma_{q = 1}^Q f_q ( 1 − f_q ), 
-\]
+$$
+I_G(f) = \sum_{q = 1}^Q f_q ( 1 − f_q ), 
+$$
 which is a minimum when all observatkions at the node are in the same class.
@@
 \subsubsection{Synthetic data} 
+\label{section:synthetic_data}
  Each dataset consists of $n$ samples and $p$ variables where $n << p$,
@@
 
-The model paramters are $w_i = 1/\sqrt{2^{i-1}}$ for $i =1,\ldots 5$ then 
-\begin{align*}
-z=& \Sigma_{i=1}^{i=5} {w_i x_i} \\
-\intertext{and}
-\sigma_\epsilon^2 =& Var(z)(1- 0.125)/0.125 \\
-\intertext{so}
-y =&  z+\epsilon \textrm{ where } \epsilon \sim  N(0, \sigma_\epsilon^2).
-\end{align*}
+The model paramters are $w_i = 1/\sqrt{2^{i-1}}$ for $i =1,\ldots 5$ and we set 
+$$
+z =  \sum_{i=1}^{i=5} {w_i x_i}.
+$$
+We let $\sigma_\epsilon^2 = Var(z)(1- 0.125)/0.125 $ and then 
+$y =   z + \epsilon$ where $\epsilon \sim  N(0, \sigma_\epsilon^2).$
 The dichotomous response is generated by thresholding $y$ at the $0.8$ quantile.
@@
   better performance;
-\item the size and complexity of the individual trees is controlled in
-  \randomforest by setting \texttt{nodesize}, the minimum size of
-  terminal nodes. It is controlled in Spark ML by setting
-  \texttt{maxDepth}, the maximum depth of each tree in the
-  forest\footnote{The Spark ML documentation \cite[]{Spark.2016} sets
-    \texttt{maxDepth} to 4 in their classification example. This may
-    give an estimate with a higher bias and, as such, may be a poor
-    choice. See the discussion in \cite{Dietterich.2002} where it is
-    shown that bagging may reduce bias as well as variance but gives a
-    better result for low bias learners. \cite{Dietterich.2002} notes
-    that if the boostrap replicate approximation were correct, then
-    bagging would reduce variance without changing bias.} In the TCGA
-  example the RandomForests code produced trees with depths of 30+.
+\item the size and complexity of the individual trees is controlled in \randomforest by setting \texttt{nodesize}, the
+  minimum size of terminal nodes. It is controlled in Spark ML by setting \texttt{maxDepth}, the maximum depth of each
+  tree in the forest\footnote{The Spark ML documentation \cite[]{Spark.2016} sets \texttt{maxDepth} to 4 in their
+    classification example. This may give an estimate with a higher bias and, as such, may be a poor choice. See the
+    discussion in \cite{Dietterich.2002} where it is shown that bagging may reduce bias as well as variance but gives a
+    better result for low bias learners. \cite{Dietterich.2002} notes that if the boostrap replicate approximation were
+    correct (i.e.~if the bootstrap sample came from an identical distribution to the data), then bagging would reduce
+    variance without changing bias.} In the TCGA example the RandomForests code produced trees with depths of 30+.
 \end{itemize}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{\cursedforest\ can evaluate variable importance for millions of features}
+\subsection{\cursedforest can evaluate variable importance for millions of features}
 \label{section:synthetic}
-In this section we demonstrate two features of \cursedforest\, a variable importance analysis to select the most
+In this section we demonstrate two features of \cursedforest, a variable importance analysis to select the most
 associated set of features amongst millions of dimensions as well as its ability to accurately classify the labels of
@@
 worker nodes each with 16 Intel Xeon E5-2660@2.20GHz CPU cores and 128 GB of RAM.  We captured the OOB 
-classification error and the rank-biased-overlap
+classification error and the rank-biased-overlap (RBO)
 \marginpar{what is wrong with the formatting of this citation?}
@@
 
-% From the synthetic data we can conclude that \cursedforest\ is able to perform accurate feature selection as well as
+% From the synthetic data we can conclude that \cursedforest is able to perform accurate feature selection as well as
 % accurate classification for large numbers of samples with large feature vectors.
@@
 the default. Fig.~\ref{figure:out-of-bag-prediction-error-prod.png} shows the OOB error is minimized at the largerst
-values of \mtry and \ntree considered  whereas Fig.~\ref{figure:rbo-prod.png} illustrates that the rbo measure is less
+values of \mtry and \ntree considered  whereas Fig.~\ref{figure:rbo-prod.png} illustrates that the RBO measure is less
 sensitive once we over a certain threshold.
 
-From the synthetic data we can conclude that \cursedforest\ is able to perform accurate feature selection as well as
+From the synthetic data we can conclude that \cursedforest is able to perform accurate feature selection as well as
 accurate classification for large numbers of samples with large feature vectors.
@@
   \begin{adjustwidth}{-1.00in}{0in}
-    \caption{\textbf{ out of bag classification error and Rank-biased overlap estimates}} 
+    \caption{\textbf{Out of bag classification error and rank-biased-overlap estimates}} 
     \label{figure:synth1}
@@
       \includegraphics[totalheight=8cm]{./figs/out-of-bag-prediction-error.png}
-      \caption{OOB classification error} 
+      \caption{OOB classification error.} 
       \label{figure:out-of-bag-prediction-error-prod.png} 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{\cursedforest\ selects previously demonstrated cancer variants}
+\subsection{\cursedforest selects previously demonstrated cancer variants}
 We consider the question of feature discovery on the Cancer Genome Atlas (TCGA) BRCA dataset.  We build this model on
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{\cursedforest\ outperforms existing methods for classification problems}
+\subsection{\cursedforest outperforms existing methods for classification problems}
 In this section we compare \cursedforest 's performance against other published methods for predicting ethnicity on the
@@
 \centering
-\caption{ {\bf Performance comparison with \cursedforest\ on different subsets of the 1000 Genomes Project data.}}
+\caption{ {\bf Performance comparison with \cursedforest on different subsets of the 1000 Genomes Project data.}}
 \begin{tabular}{| l | r | l |}
@@
 \begin{flushleft} 
-Comparison of runtime and error for models built using \cursedforest\ on different subsets of variant data 
+Comparison of runtime and error for models built using \cursedforest on different subsets of variant data 
 from the 1000 Genomes Project.
@@
 
-\multirow{5}{*}{phase1\_chr22: 1,092 x 490,036} & \cursedforest\ & 0.04  & 1 min 50 seconds                        \\
+\multirow{5}{*}{phase1\_chr22: 1,092 x 490,036} & \cursedforest & 0.04  & 1 min 50 seconds                        \\
                                                 & Spark ML  &  0.05           &           7 min 44 seconds         \\
@@
 
-\multirow{5}{*}{phase3\_chr22: 2,504 x 1,103,548}   & \cursedforest\ & 0.06 & 4 min 2 seconds \\
+\multirow{5}{*}{phase3\_chr22: 2,504 x 1,103,548}   & \cursedforest & 0.06 & 4 min 2 seconds \\
                                                     & Spark ML  & 0.06 & 37 min 19 seconds \\
@@
 
-\multirow{3}{*}{phase3\_chr1: 2,504 x 6,450,364}    & \cursedforest\ & 0.04  & 19 min 19 seconds         \\
+\multirow{3}{*}{phase3\_chr1: 2,504 x 6,450,364}    & \cursedforest & 0.04  & 19 min 19 seconds         \\
                                                     & Spark ML  & 0.04       & 8 hour 29 min        \\
@@
 
-\multirow{3}{*}{phase3\_chr1-3: 2,504 x 19,328,051} & \cursedforest\ & 0.02  & 56 min 43 seconds             \\
+\multirow{3}{*}{phase3\_chr1-3: 2,504 x 19,328,051} & \cursedforest & 0.02  & 56 min 43 seconds             \\
                                                     & Spark ML &    NA         &   NA               \\
@@
 
-\multirow{4}{*}{phase3\_chr1-22: 2,504 x 81,047,467} & \cursedforest\ & 0.01  & 3hour 16min \\
+\multirow{4}{*}{phase3\_chr1-22: 2,504 x 81,047,467} & \cursedforest & 0.01  & 3hour 16min \\
                                                     & Spark ML & NA & NA  \\
@@
 \subsection{Scalability analysis}
-In this section we explore the performance of \cursedforest\ by testing its ability to scale to different sizes of data
+In this section we explore the performance of \cursedforest by testing its ability to scale to different sizes of data
 and computational resources.
 
-In order to asses these characteristics we ran \cursedforest\ classification on synthetics datasets with varying numbers
+In order to asses these characteristics we ran \cursedforest classification on synthetic datasets with varying numbers
 of variables (features) and samples, similar to the dataset used in \cite{Wright.and.Ziegle.2016} to evaluate
-\ranger, allocating varying number of CPU cores to the \cursedforest\ and also varying computational complexity
-of random forests by using a range of \mtry values.
-
-We investigate the different synthetic datasets generated for section 5.1. and measured the time taken to build a random
-forest model of 100 trees. The results reported below are averages of 5 runs, and all the cases were executed with the
-same random seed, to improve the consistency of measurements.
+\ranger, allocating varying number of CPU cores to the \cursedforest and also varying the computational complexity
+of the random forests by using a range of \mtry values.
+
+We investigate the different synthetic datasets generated for section \ref{ection:synthetic_data} and measured the time
+taken to build a random forest model of 100 trees. The results reported below are averages of 5 runs, and all the cases
+were executed with the same random seed, to improve the consistency of measurements.
 
 First we look at \cursedforest horizontal scalability for a medium size dataset of 2.5 million variables and 5000
-samples, by varying the \mtry  fraction and the number of CPU cores allocated to the execution. The raw results are
-presented in table \ref{synthetictimingtable} and visualized in figure \ref{figure:synthetictiming} below. Regardless of the
-number of cores used, \cursedforest displays approximately linear dependency between the execution time and \mtry
-(Fig. \ref{figure:synthetictiming.a}).
+samples, by varying the \mtry  fraction and the number of CPU cores allocated to the execution. 
+The results are  visualized in figure \ref{figure:synthetictiming} below. 
+%The raw results are
+%presented in table \ref{synthetictimingtable} and visualized in figure \ref{figure:synthetictiming} below. 
+Regardless of the number of cores used, \cursedforest displays approximately linear dependency between the execution
+time and \mtry (Fig. \ref{figure:synthetictiming.a}).
 
@@
 
-Next we investigate \cursedforest scalability with regards to the size of data, by varying the number of variables and
-sample for a fixed \mtry fraction of 0.25 and execution of 128 CPU cores. The raw results are presented in the table
-[REF] 
-\marginpar{have we lost this table?}
-and visualized in figure \ref{figure:synth2} below (please note log scale on the axes and the values on y axes are
-expresses as trees per hour).
+\marginpar{tables 2 has been commented out -- there is no data in it. Piotr could add the data or we could remove both
+  tables just leave the figures. I have done this for the time being} 
+Next we investigate \cursedforest scalability with
+regards to the size of data, by varying the number of variables and sample for a fixed \mtry fraction of 0.25 and
+execution of 128 CPU cores. The results are visualized in figure \ref{figure:synth2} below (please note log scale on the
+axes and the values on y axes are expresses as trees per hour).
 
@@
 
-To conclude \cursedforest\ is capable of processing datasets with 50M variables and 10K samples, which is the order of
+To conclude \cursedforest is capable of processing datasets with 50M variables and 10K samples, which is the order of
 magnitude of the entire genome variants expected in such dataset, for which it can build approximately 60 trees per
@@
 
-%nvars 150000  500000  2500000 10000000  50000000
-%nsamples          
-%1000  21.6  26.7  59.8  133.5 668.8
-%5000  45.7  71.9  186.2 729.3 2887.4
-%10000 70.2  120.2 397.3 2237.8  5747.8
-
-\begin{table}[!ht]
-\begin{adjustwidth}{-1.00in}{0in}
-\centering
-\caption{\textbf{
- Time in seconds to build a random forest of 100
-  trees on 128 CPU cores with varying numbers of variables and samples.}}
-\label{synthetictimingtable}
-\begin{tabular}{|l|l|l|l|l|l|l|}
-\hline
-\multicolumn{2}{|l|}{\multirow{2}{*}{}}           & \multicolumn{5}{c|}{Number of variables} \\
-\cline{3-7}
-\multicolumn{2}{|l|}{}                               & \bf{150,000} & \bf{500,000} & \bf{2,500,000}  & \bf{10,000,000} & \bf{50,000,000} \\
-\hline                                                            
-\multirow{4}{*}{Number of samples}      & \bf{1000}  & 21.6  & 26.7  & 59.8  & 133.5  & 668.8 \\
-                                        & \bf{5000}  & 45.7  & 71.9  & 186.2 & 729.3  & 2887.4 \\
-                                        & \bf{10000} & 70.2  & 120.2 & 397.3 & 2237.8 & 5747.8 \\
-\hline
-\end{tabular}
-\end{adjustwidth}
-\end{table}
+% commented this table out -- see above
+% \begin{table}[!ht]
+%   \begin{adjustwidth}{-1.00in}{0in}
+%     \centering
+%     \caption{\textbf{
+%  Time in seconds to build a random forest of 100
+%   trees on 128 CPU cores with varying numbers of variables and samples.}}
+% \label{synthetictimingtable}
+% \begin{tabular}{|l|l|l|l|l|l|l|}
+% \hline
+% \multicolumn{2}{|l|}{\multirow{2}{*}{}}           & \multicolumn{5}{c|}{Number of variables} \\
+% \cline{3-7}
+% \multicolumn{2}{|l|}{}                               & \bf{150,000} & \bf{500,000} & \bf{2,500,000}  & \bf{10,000,000} & \bf{50,000,000} \\
+% \hline                                                            
+% \multirow{4}{*}{Number of samples}      & \bf{1000}  & 21.6  & 26.7  & 59.8  & 133.5  & 668.8 \\
+%                                         & \bf{5000}  & 45.7  & 71.9  & 186.2 & 729.3  & 2887.4 \\
+%                                         & \bf{10000} & 70.2  & 120.2 & 397.3 & 2237.8 & 5747.8 \\
+% \hline
+% \end{tabular}
+% \end{adjustwidth}
+% \end{table}
 
@@
 
-% \begin{figure}[tbhp]
-% \begin{adjustwidth}{-1.00in}{0in}
-% \begin{tabular}{ll}
-% a)& b)\\
-% \includegraphics[totalheight=6cm]{./figs/nvars_nsamples.png} & 
-% \includegraphics[totalheight=6cm]{./figs/nsamples_nvars.png} \\
-% \end{tabular}
-% \caption{{\bf Scalablity of the Wide Random Forest on synthetic datasets with varying number of samples and variables.} 
-%   {\bf a)} Number of trees build per hour when growing number of variables. 
-%   {\bf b)} Number of trees build per hour when growing number of samples.}
-% \label{figure:synth}
-% \end{adjustwidth}
-% \end{figure}
-
+% no data in this table 
 % \begin{table}[!ht]

### 2017-01-12 doc/publications/plos16/plos16.tex
Update my section a bit

--- 2017-01-12 doc/publications/plos16/plos16.tex
@@
 % Use Unicode characters when possible
-\usepackage[utf8x]{inputenc}
+\usepackage[utf8]{inputenc}
 
@@
 \newcommand{\cursedforest}{\textsc{CursedForest}\xspace}
-\newcommand{\ranger}{\textsc{Ranger}\xspace}
 \newcommand{\mtry}{\texttt{mtry}\xspace}
@@
 \usepackage{multirow}
+\usepackage[T1]{fontenc}
 
@@
 \section{Abstract}
-We introduce \cursedforest, an implementation of random forests designed to handle data that may have an extremely large
-number of variables. We discuss the utility of random forests for machine learning problems. We show on simulated and
-real data that the novel implementation can achieve reasonable prediction accuracy and variable importance estimates
-even in setting where other implementations, including some designed for large data, fail.
+The abstract.
 
@@
 achieves fast run times by effective use of the CPU cache, optimizing for different classes of features and efficiently
-multi-threading.  Similarly, \ranger~\cite{Wright.and.Ziegle.2016} written in C++ with an R front-end is
+multi-threading.  Similarly, \textsc{Ranger}~\cite{Wright.and.Ziegle.2016} written in C++ with an R front-end is
 specifically optimized for dealing with large data sets.
@@
 As is standard with Spark applications, we store our data in a Resilient Distributed Dataset (RDD), where an RDD is
-essentially a collection of elements. In the case of Spark ML, each element in the RDD is a sample. RDDs contribute to
+essentially a collection of elements. In the case of Spark ML, each element in our RDD is a sample. RDDs contribute to
 the scalability of Spark as they can distributed across multiple nodes and operated on in parallel. Even as we add more
@@
 enable these increasingly large vectors to be loaded into memory.
-
-On the other hand, \cursedforest is specifically designed to handle wide ``cursed'' data. It avoids the relation between
-memory and dimensionality by avoiding calculations that rely on entire feature vectors and taking the parallelization
-work down to the level of the individual features.  For example, for each node of a tree, \cursedforest will distribute
-tasks that consist of single features (variants), for every individual.  Each of these tasks will calculate the
-information gain for that specific feature.  Once these tasks have completed, the results are reduced to return the
-feature which gives the greatest information gain.  This process is then repeated until \cursedforest has created the
-entire decision tree.
-
-The current implementation of \cursedforest uses an ``Gini impurity'' criteria for splitting. Let $f_q$ be the fraction
-of items labelled with value $q$ where $q = 1, \ldots, Q$ at a node. The Gini impurity is
+\marginpar{not clear. Are RDDs the problem or the solutuion Aidan: Better?}
+
+On the other hand, \cursedforest\ is specifically designed to handle wide ``cursed'' data. It avoids the relation
+between memory and dimensionality by avoiding calculations that rely on entire feature vectors.
+
+Compared to Spark ML, \cursedforest\ further parallelizes work down to individual features.  For example, for each node
+of a tree, \cursedforest\ will distribute tasks that consist of single features (variants), for every individual.  Each
+of these tasks will calculate the information gain for that specific feature.  Once these tasks have completed, the
+results are reduced to return the feature which gives the greatest information gain.  This process is then repeated
+until \cursedforest\ has created the entire decision tree.
+
+The current implementation of \cursedforest\ uses an ``Gini impurity''
+criteria for splitting. Let $f_q$ be the fraction of items labelled
+with value $q$ where $q = 1, \ldots, Q$ at a node. The Gini impurity is
 \[
-I_G(f) = \Sigma_{q = 1}^Q f_q ( 1 − f_q ), 
+I_G(f) = \Sigma_{q = 1}^Q f_q ( 1 - f_q ), 
 \]
@@
 
-\subsubsection{The Cancer Genome Atlas}
-We obtained The Cancer Genome Atlas (TCGA) BRCA data as VCF files from the GDC data portal.  This data is in the
-category ``Simple Nucleotide Variation''.  After downloading, we use a custom script to name the samples in the files
-and filter for only the "Normal" tissue column.  We then use GATK to merge the files into a single VCF file.
+\subsubsection{Bone mineral density dataset}
+We obtained a VCF file from the authors of \cite{Duncan2011}, which contains variant data from 1,936 postmenopausal women.
+This dataset covers approximately 7 million variant positions. We have information about the bone mineral density (BMD)
+for each individual, where a low BMD is a major predisposing factor to fracture \cite{Duncan2011}.
 
@@
 \item \ntree{}  -- the number of trees.  The number of trees necessary for good performance grows with the number of
-  predictors.  \cite{Liaw.and.Weiner.2002} suggest that a high \ntree is necessary to get stable estimates of variable
+  predictors.  \cite{Liaw.and.Weiner.2002} suggest that a high \ntree\ is necessary to get stable estimates of variable
   importance and proximity; however, even though the variable importance measures may vary from run to run, we note that
@@
 \item \mtry{}  -- the number of variables considered at each split (if \mtry=$p$, we have a boosted decision
-  tree model).  If one has a very large number of variables but expects only very few to be ``important'', using larger \mtry may give
+  tree model).  If one has a very large number of variables but expects only very few to be ``important'', using larger \mtry\ may give
   better performance;
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{\cursedforest\ selects previously demonstrated cancer variants}
-We consider the question of feature discovery on the Cancer Genome Atlas (TCGA) BRCA dataset.  We build this model on
-VCF files retrieved from the TCGA data portal.  For the label, we use the cell type, tumor or normal.
+\subsection{\cursedforest\ selects previously demonstrated bone-density variants}
+We consider the problem of feature discovery on a genomic dataset for a specific phenotype.
+In this case we investigate the ability of  ability to locate variants in the same loci as genes previously demonstrated to
+be associated with a high or low bone mass density (BMD) and rank these variants as being "important".
+From a random forest model built on this data, we produce a list of the to 5,000 features. 
+Applying the Mann–Whitney U test to these fetures we observe a p-value of x, supporting our
+hypothesis that features in close proximity to the known BMD genes are Contributing more information to our classification model.
+
+One observation we make is that of the most-important features---for example, the top ten variants ranked by Gini impurity---is
+that not all of these are variants are in close proximity of known BMD genes.
+One of these is COLEC10. Featuring a transcription factor binding site for FOXL1, COLEC10 is also reported as containing a SNP
+associated with BMD \cite{Liu2008}.
 

### 2017-01-12 doc/publications/plos16/plos16.tex

--- 2017-01-12 doc/publications/plos16/plos16.tex
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{\cursedforest selects previously demonstrated cancer variants}
-We consider the question of feature discovery on the Cancer Genome Atlas (TCGA) BRCA dataset.  We build this model on
-VCF files retrieved from the TCGA data portal.  For the label, we use the cell type, tumor or normal.
+\subsection{\cursedforest\ selects previously demonstrated bone-density variants}
+We consider the problem of feature discovery on a genomic dataset for a specific phenotype.
+In this case we investigate the ability of  ability to locate variants in the same loci as genes previously demonstrated to
+be associated with a high or low bone mass density (BMD) and rank these variants as being "important".
+From a random forest model built on this data, we produce a list of the to 5,000 features. 
+Applying the Mann–Whitney U test to these fetures we observe a p-value of x, supporting our
+hypothesis that features in close proximity to the known BMD genes are Contributing more information to our classification model.
+
+One observation we make is that of the most-important features---for example, the top ten variants ranked by Gini impurity---is
+that not all of these are variants are in close proximity of known BMD genes.
+One of these is COLEC10. Featuring a transcription factor binding site for FOXL1, COLEC10 is also reported as containing a SNP
+associated with BMD \cite{Liu2008}.
 

### 2017-01-12 doc/publications/plos16/plos16.tex
Change to better hyphen
changes to plos16.tex

--- 2017-01-12 doc/publications/plos16/plos16.tex
@@
 
-\marginpar{signal rather than ``significant variables''}
-
 Donoho and Tanner \cite{Donoho.and.Tanner.2009} give a ``universal phase change'' result that has applications in a
@@
 \begin{itemize}
-\item 
+\item the Donoho-Tanner phase transition (in this setting) refers to estimating the $\beta$ in data generated by a
+  linear model;
+\item in a decision tree (random forest) there is no notion of estimating the $\beta$
+\item also a decision tree (random forest) is a heruistic search. It may recover a relationshipin the space
+  of the combinatorial search.
 \end{itemize}
@@
     \centering
-    \includegraphics[totalheight=6cm]{./figs/phase-diagram-equivalence.png} 
+    \includegraphics[totalheight=6cm]{./figs/phase.png} 
     \caption{{\bf The phase chage diagram.}}
@@
     \vspace{4ex}
-  \end{adjustwidth}
-\end{figure}
-
-\section{not for the paper}
-Notes:
-\begin{itemize}
-\item we try a small simulation, a regression problem with $\beta(1:k) \sim U(1,100)$ and $y= X\beta + \epsilon$, $\epsilon \sim  N(0,\sqrt{16})$
-\item \cite{Donoho.and.Stodden.2006} evaluate variable selection by 
-$$\frac{\Vert\hat{\beta}-\beta\Vert_2}{\Vert\beta\Vert_2}$$
-See figure \ref{figure:stepwise_lars_cv.png} 
-\item 
-However there is no $\hat{\beta}$ in a tree model. We are using variable importance as
-a measure of the ordering of $\hat{\beta}$ (i.e. we expect variable importance to be proportional to $\hat{\beta}$). Is this correct?
-\item as RF is sort of greedy approximation to combinatorial approach, it may have some success in the dead zone
-\item  \cite{Donoho.and.Stodden.2006} are considering problems with $L_2$ errors (with additional  $L_1$ norms on the
-  $\beta$. $L_1$).
-How will their results carry over to 0-1 loss?
-\end{itemize}
-
-\begin{figure}[tbhp] 
-  \begin{adjustwidth}{-1.00in}{0in}
-    \begin{subfigure}[b]{0.5\linewidth}
-      \centering
-      \includegraphics[totalheight=6cm]{./figs/k.png}
-      \caption{k}
-      \label{figure:k.png}
-      \vspace{4ex}
-    \end{subfigure} 
-    \begin{subfigure}[b]{0.5\linewidth}
-      \centering
-      \includegraphics[totalheight=6cm]{./figs/n.png}
-      \caption{n}
-      \label{figure:n.png}
-    \end{subfigure} 
-    \caption{}
-  \end{adjustwidth}
-\end{figure}
-
-\begin{figure}[tbhp] 
-  \begin{adjustwidth}{-1.00in}{0in}
-    \begin{subfigure}[t]{0.5\linewidth}
-      \centering
-      \includegraphics[totalheight=6cm]{./figs/stepwise_lars_cv.png}
-      \caption{Stepwise lars. $p=200$, $n$ and $k$ variable. Model selected by cross-validation}
-      \label{figure:stepwise_lars_cv.png}
-      \vspace{4ex}
-    \end{subfigure} 
-    \begin{subfigure}[t]{0.5\linewidth}
-      \centering
-      \includegraphics[totalheight=6cm]{./figs/ranger.png}
-      \caption{ranger, RBO}
-      \label{figure:ranger.png}
-    \end{subfigure} 
-    \caption{}
   \end{adjustwidth}

### 2017-01-12 doc/publications/plos16/plos16.tex

--- 2017-01-12 doc/publications/plos16/plos16.tex
@@
 $$
-I_G(f) = \sum_{q = 1}^Q f_q ( 1 − f_q ), 
+I_G(f) = \sum_{q = 1}^Q f_q ( 1 - f_q ), 
 $$

### 2017-01-13 doc/publications/plos16/plos16.tex
changes to plos16.tex

@@
 * Denis.Bauer@CSIRO.au
+\end{flushleft}
 
-\end{flushleft}
 \tableofcontents
-\clearpage
-
 \clearpage
@@
 reliably recover significant variables.  \cite{Donoho.and.Stodden.2006} investigate the behaviour of a number of
-regression approaches for variable selection (lars, lasso and forward stepwise) and make the point that above the
+regression approaches for variable selection (LARS, Lasso and forward stepwise) and make the point that above the
 phase-transition line variable recovery is still possible by a combinatorial approach.
@@
 \cursedforest is designed for extremely large numbers of variables it is likely to be operating in difficult regions of
-the figure.
+the figure where the ratio $\delta = n/p$ is small.
 
@@
 \begin{itemize}
-\item the Donoho-Tanner phase transition (in this setting) refers to estimating the $\beta$ in data generated by a
-  linear model;
-\item in a decision tree (random forest) there is no notion of estimating the $\beta$
-\item also a decision tree (random forest) is a heruistic search. It may recover a relationshipin the space
-  of the combinatorial search.
+\item the Donoho-Tanner phase transition arises in recovering the $\beta$ in data generated by a
+  linear model. However, in a decision tree (random forest) there is no notion of estimating the $\beta$
+\item A decision tree (random forest) is a heruistic search. It may recover a relationship in the space of the
+  combinatorial search.
 \end{itemize}
 
-We note that the examples we are considering lie on the extreme left of the plot, close to the origin. So there are
-limits to what can be recovered in the presence of noise. However in the case of data that is both big and wide,
+The existence of the  Donoho-Tanner phase transition is a salutory warning. There are likley to be limits, both
+computionaal and logical, to the recovery of signals from noisy data. \cursedforest  is a contribution to addressing the
+practivcal limits but the logical limits will still apply. However in the case of data that is both big and wide,
 \cursedforest and other VariantSpark methods may provide a useful tool.

### 2017-01-13 doc/publications/plos16/plos16.tex
some more changes

--- 2017-01-12 doc/publications/plos16/plos16.tex
@@
 \usepackage{multirow}
-\usepackage[T1]{fontenc}
+%\usepackage[T1]{fontenc}
 
@@
 We consider the problem of feature discovery on a genomic dataset for a specific phenotype.
-In this case we investigate the ability of  ability to locate variants in the same loci as genes previously demonstrated to
+In this case we investigate the ability to retrieve variants in the same loci as genes previously demonstrated to
 be associated with a high or low bone mass density (BMD) and rank these variants as being "important".
-From a random forest model built on this data, we produce a list of the to 5,000 features. 
-Applying the Mann–Whitney U test to these fetures we observe a p-value of x, supporting our
-hypothesis that features in close proximity to the known BMD genes are Contributing more information to our classification model.
+We build a random forest model on this data, and produce a list of the top 5,000 contributing features, ranked by their
+Gini impurity. We apply the Mann–Whitney U test to these fetures and observe a p-value of x, supporting our hypothesis
+that features in close proximity to the known BMD genes are contributing more information to our classification model.
 
-One observation we make is that of the most-important features---for example, the top ten variants ranked by Gini impurity---is
-that not all of these are variants are in close proximity of known BMD genes.
-One of these is COLEC10. Featuring a transcription factor binding site for FOXL1, COLEC10 is also reported as containing a SNP
+However, we also observe that from the set of highest-ranked features---for example, the top ten---not all of them exist
+within close proximity to known BMD genes.
+
+One example is COLEC10. Featuring a transcription factor binding site for FOXL1, COLEC10 is also reported as containing a SNP
 associated with BMD \cite{Liu2008}.
+
+We also frequently see highly-ranked variants in chromosome 22q11
 
@@
 
+\marginpar{signal rather than ``significant variables''}
+
 Donoho and Tanner \cite{Donoho.and.Tanner.2009} give a ``universal phase change'' result that has applications in a
@@
 \begin{itemize}
-\item the Donoho-Tanner phase transition (in this setting) refers to estimating the $\beta$ in data generated by a
-  linear model;
-\item in a decision tree (random forest) there is no notion of estimating the $\beta$
-\item also a decision tree (random forest) is a heruistic search. It may recover a relationshipin the space
-  of the combinatorial search.
+\item 
 \end{itemize}
@@
     \centering
-    \includegraphics[totalheight=6cm]{./figs/phase.png} 
+    \includegraphics[totalheight=6cm]{./figs/phase-diagram-equivalence.png} 
     \caption{{\bf The phase chage diagram.}}
@@
     \vspace{4ex}
+  \end{adjustwidth}
+\end{figure}
+
+\section{not for the paper}
+Notes:
+\begin{itemize}
+\item we try a small simulation, a regression problem with $\beta(1:k) \sim U(1,100)$ and $y= X\beta + \epsilon$, $\epsilon \sim  N(0,\sqrt{16})$
+\item \cite{Donoho.and.Stodden.2006} evaluate variable selection by 
+$$\frac{\Vert\hat{\beta}-\beta\Vert_2}{\Vert\beta\Vert_2}$$
+See figure \ref{figure:stepwise_lars_cv.png} 
+\item 
+However there is no $\hat{\beta}$ in a tree model. We are using variable importance as
+a measure of the ordering of $\hat{\beta}$ (i.e. we expect variable importance to be proportional to $\hat{\beta}$). Is this correct?
+\item as RF is sort of greedy approximation to combinatorial approach, it may have some success in the dead zone
+\item  \cite{Donoho.and.Stodden.2006} are considering problems with $L_2$ errors (with additional  $L_1$ norms on the
+  $\beta$. $L_1$).
+How will their results carry over to 0-1 loss?
+\end{itemize}
+
+\begin{figure}[tbhp] 
+  \begin{adjustwidth}{-1.00in}{0in}
+    \begin{subfigure}[b]{0.5\linewidth}
+      \centering
+      \includegraphics[totalheight=6cm]{./figs/k.png}
+      \caption{k}
+      \label{figure:k.png}
+      \vspace{4ex}
+    \end{subfigure} 
+    \begin{subfigure}[b]{0.5\linewidth}
+      \centering
+      \includegraphics[totalheight=6cm]{./figs/n.png}
+      \caption{n}
+      \label{figure:n.png}
+    \end{subfigure} 
+    \caption{}
+  \end{adjustwidth}
+\end{figure}
+
+\begin{figure}[tbhp] 
+  \begin{adjustwidth}{-1.00in}{0in}
+    \begin{subfigure}[t]{0.5\linewidth}
+      \centering
+      \includegraphics[totalheight=6cm]{./figs/stepwise_lars_cv.png}
+      \caption{Stepwise lars. $p=200$, $n$ and $k$ variable. Model selected by cross-validation}
+      \label{figure:stepwise_lars_cv.png}
+      \vspace{4ex}
+    \end{subfigure} 
+    \begin{subfigure}[t]{0.5\linewidth}
+      \centering
+      \includegraphics[totalheight=6cm]{./figs/ranger.png}
+      \caption{ranger, RBO}
+      \label{figure:ranger.png}
+    \end{subfigure} 
+    \caption{}
   \end{adjustwidth}

### 2017-01-13 doc/publications/plos16/plos16.tex
put T1 fontenc back in
little change
some more changes

@@
 
-We also frequently see highly-ranked variants in chromosome 22q11
+We also frequently see highly-ranked variants in chromosome 22q11. Stagi2010 observe that individuals with a microdeletion
+in this region have a significant in bone mass.\ref{Stagi2010} In this same region, we also observe variants present within the
+PRODH gene---a gene which is overexpressed in bone.
 

### 2017-01-17 doc/publications/plos16/plos16.tex
add to BMD section

--- 2017-01-13 doc/publications/plos16/plos16.tex
@@
 \usepackage{multirow}
-%\usepackage[T1]{fontenc}
+\usepackage[T1]{fontenc}
 
@@
 \subsection{\cursedforest\ selects previously demonstrated bone-density variants}
-We consider the problem of feature discovery on a genomic dataset for a specific phenotype.
-In this case we investigate the ability to retrieve variants in the same loci as genes previously demonstrated to
-be associated with a high or low bone mass density (BMD) and rank these variants as being "important".
-We build a random forest model on this data, and produce a list of the top 5,000 contributing features, ranked by their
-Gini impurity. We apply the Mann–Whitney U test to these fetures and observe a p-value of x, supporting our hypothesis
-that features in close proximity to the known BMD genes are contributing more information to our classification model.
+We investigate the ability of using a random forest model we build on genomic data to identify variants related to a
+specific phenotype.
+In this case we aim to identify variants associated with Bone Minderal Density (BMD). We build a model on the BMD dataset
+(see datasets) using a binary high or low BMD value as the label.
+From this model, we output a list of the top 10,000 contributing variants, ranked by their Gini impurity. The variant with
+the highest Gini impurity---and hence, importance---is ranked 1, and the variant with the lowest Gini impurity is ranked 10,000.
+Then, using the 26 known and 6 novel BMD genes from Duncan et al., we compile a list of variants that are present in or near one of these
+32 genes.
+
+We then apply the Mann–Whitney U test to this list to verify that from our list, variants from known BMD loci are ranked
+higher than variants from elsewhere in the genome.
+We observe a p-value of XXXX, supporting our hypothesis that variants in close proximity to the known BMD genes are
+contributing more information to our classification model.
 
 However, we also observe that from the set of highest-ranked features---for example, the top ten---not all of them exist
-within close proximity to known BMD genes.
+within close proximity to known BMD genes. We investigate the literature to determine the plausibility of these variants
+contributing to BMD.
 
-One example is COLEC10. Featuring a transcription factor binding site for FOXL1, COLEC10 is also reported as containing a SNP
-associated with BMD \cite{Liu2008}.
+One gene is COLEC10 (collectin 10). Featuring a transcription factor binding site for FOXL1, COLEC10 is also reported as
+containing variants associated with BMD \cite{Kemp2014,Liu2008}. However there currently exists no information on how this gene
+may contribute to BMD.
 
-We also frequently see highly-ranked variants in chromosome 22q11
+We also frequently see highly-ranked variants in chromosome 22q11. Stagi2010 observe that individuals with a microdeletion
+in this region have a significant in bone mass\cite{Stagi2010}. In this same region, we also observe variants present in
+PRODH (proline dehydrogenase). PRODH catalyzes the first step in proline degredation, and as proposed by Phang et al., proline
+is made available by collagen degradation \cite{Phang2008}.
+
+DCDC5 (Doublecortin Domain Containing 5) is one of the 26 known genes, however it is one of the genes that Duncan et al.
+failed to replicate in their study. DCDC5 does appear in our list of genes, however the first occurance of a variant in
+or near this gene has a rank of 9,667 out of 10,000. Also, this gene is not highly expressed in bone, and any mechanisms
+by which this gene might regulate BMD are unclear \cite{Thakker2012}. Although variants near DCDC5 are associated with
+BMD \cite{Rivadeneira2009}, it is questionable as to whether the gene itself contributes to BMD.
 
@@
  \begin{adjustwidth}{-1.00in}{0in}
+    \caption{\textbf{The Donoho-Tanner phase chage diagram.}}
     \centering
     \includegraphics[totalheight=6cm]{./figs/phase.png} 
-    \caption{{\bf The phase chage diagram.}}
     \label{figure:phase-diagram-equivalence.png} 
@@
 
-\bibliography{./CursedForest.bib}
+\bibliography{CursedForest}
 

### 2017-01-17 doc/publications/plos16/plos16.tex
standardize and hopefully fix a few citations

@@
 worker nodes each with 16 Intel Xeon E5-2660@2.20GHz CPU cores and 128 GB of RAM.  We captured the OOB 
-classification error and the rank-biased-overlap (RBO)
-\marginpar{what is wrong with the formatting of this citation?}
-\cite[]{Webber.et.al.2010} measure.
+classification error and the rank-biased-overlap (RBO) \cite{Webber2010} measure.
 
@@
 higher than variants from elsewhere in the genome.
-We observe a p-value of XXXX, supporting our hypothesis that variants in close proximity to the known BMD genes are
+We observe $P=1.3\times10^{-07}$, supporting our hypothesis that variants in close proximity to the known BMD genes are
 contributing more information to our classification model.
@@
 
-We also frequently see highly-ranked variants in chromosome 22q11. Stagi2010 observe that individuals with a microdeletion
+We also frequently see highly-ranked variants in chromosome 22q11. Stagi et al. observe that individuals with a microdeletion
 in this region have a significant in bone mass\cite{Stagi2010}. In this same region, we also observe variants present in

### 2017-01-18 doc/publications/plos16/plos16.tex
1000 genomes section

@@
 
-\marginpar{this section is not one sentence? has it been lost?}
+In a previous paper, we clustered individuals from the 1000 Genomes Project using k-means clustering\cite{Obrien2016}.
+Although we were able to cluster individuals on all 80 million variants, it was relatively slow, with the entire
+process (pre-processing and clustering) taking over 30 hours.
+We investigate building random forest models on this data---however we observed this process to not only be slower
+than building a k-means model, but also unable to scale to as many dimensions (see Table~\ref{timingtable}). To keep
+the following comparisons fair, we limit the number of executors available to Spark. On the relatively low-dimensional
+phase1\_chr22 dataset (1,092 x 490,036), we observed a runtime of under 8 minutes. Although slower than Ranger---
+presumably due to the overhead of Spark's distributed platform on this relatively small dataset---we observe Spark ML to
+be about 16 minutes faster than R. Moving to the larger dataset, phase3\_chr1 (2,504 x 6,450,364), we observe an increase
+in Spark ML's runtime to over 8 hours. Moving to our next largest dataset, phase3\_chr1\-3 (2,504 x 19,328,051), and the
+job fails to complete within our arbitrary cutoff time of 20 hours.
+
+From our trials, it was apparent that Spark ML---although designed to handle many samples---is unable to efficiently cope
+with a huge number of features. With the decreasing costs of genome sequencing, this case will surely become increasingly
+common.
+
+\cursedforest solves this problem of high-dimensionality genomic data. Not only is it faster on the above smaller datasets,
+but it is also able to scale to the entire genome. In table~\ref{cursedforesttable} we present our observed times building
+models with CursedForest using all available resources on our Spark cluster.
+
+As we previously mentioned, building a random forest model on the phase3\_chr1 dataset with Spark ML takes over 8 hours.
+Building a model with \cursedforest, however, takes only 20 minutes using the same cluster setup and this time drops to
+nearly 4 minutes using the entire compute cluster. Subsequently, phase3\_chr1\-3 takes about 11 minutes, and building a
+model on the entire genome is just shy of 37 minutes. Not only does the job complete successfully, but building a random
+forest model on all 22 autosomes (phase3\_chr1\-22) using \cursedforest is more than twice as fast as building a random
+forest on the chromosome 1 (phase3\_chr1) using Spark ML.
 

### 2017-01-19 doc/publications/plos16/plos16.tex
added figure for run comparison and commented out table

@@
 Piotr Szul\textsuperscript{2},
-Stephanie Li\textsuperscript{3},
-James Doecke\textsuperscript{3},
-Nick Ellis\textsuperscript{4},
-Robert Dunne\textsuperscript{5}, and
+Robert Dunne\textsuperscript{3}, 
+Paul Leo \textsuperscript{4},
+Emma L. Duncan \textsuperscript{4,5,6},
+Stephanie Li\textsuperscript{7},
+James Doecke\textsuperscript{7},
+Nick Ellis\textsuperscript{8}, and
 Denis C. Bauer\textsuperscript{1,*}
@@
 \\
-\textbf{3} Health \& Biosecurity, CSIRO, Brisbane, QLD, Australia
+\textbf{3} Data61, CSIRO, Sydney, NSW, Australia
 \\
-\textbf{4} Oceans \& Atmosphere, CSIRO, Brisbane, QLD, Australia
+\textbf{4} Institute of Health and Biomedical Innovation, Queensland University of Technology, Brisbane, QLD, Australia
 \\
-\textbf{5} Data61, CSIRO, Sydney, NSW, Australia
+\textbf{5} Department of Endocrinology, Royal Brisbane and Women's Hospital, Brisbane, QLD, Australia
 \\
+\textbf{6} Faculty of Medicine and Biomedical Sciences, University of Queensland, Brisbane, QLD, Australia
+\\
+\textbf{7} Health \& Biosecurity, CSIRO, Brisbane, QLD, Australia
+\\
+\textbf{8} Oceans \& Atmosphere, CSIRO, Brisbane, QLD, Australia
 \bigskip
@@
 
+\begin{figure}[tbhp] 
+    \caption{\textbf{Runtime comparison on the 1000 genomes project}} 
+    \label{figure:synth1}
+      \includegraphics[totalheight=8cm]{./figs/1000genomesRuntime.pdf}
+    \begin{flushleft} 
+      The figure shows the runtime for the 5 subsets of the 1000 genomes project and the different methods investigated.
+$\bigotimes$ marks the last dataset the different method completed before running out of memory or running more than 24 hours.
+The dashed lines mark the trajectory for how ranger and SparkML had performed given larger memory machines.
+    \end{flushleft}
+\end{figure}
+
+\iffalse
+
 \begin{table}[!ht]
@@
 
-\multirow{5}{*}{phase1\_chr22: 1,092 x 490,036} & \cursedforest & 0.04  & 1 min 50 seconds                        \\
-                                                & Spark ML  &  0.05           &           7 min 44 seconds         \\
-                                                & \randomforest (R)         & 0.05       & 24 min           \\
-                                                & \ranger (R)      & 0.06       & 1 min 10 seconds       \\
-                                                & \ranger (C++)     & 0.06       & 7 min            \\
+\multirow{5}{*}{phase1\_chr22: 1,092 x 490,036} & \cursedforest & 0.04  & 1 min 50 seconds    \\ %110
+                                                & Spark ML  &  0.05           &           7 min 44 seconds         \\  %464
+                                                & \randomforest (R)         & 0.05       & 24 min           \\ %1440
+                                                & \ranger (R)      & 0.06       & 1 min 10 seconds       \\ %70
+                                                & \ranger (C++)     & 0.06       & 7 min            \\ %420
                                                 & H2O           & ??       & ??         \\
@@
 
-\multirow{5}{*}{phase3\_chr22: 2,504 x 1,103,548}   & \cursedforest & 0.06 & 4 min 2 seconds \\
-                                                    & Spark ML  & 0.06 & 37 min 19 seconds \\
+\multirow{5}{*}{phase3\_chr22: 2,504 x 1,103,548}   & \cursedforest & 0.06 & 4 min 2 seconds \\  %240
+                                                    & Spark ML  & 0.06 & 37 min 19 seconds \\  %2239
                                                     & \randomforest (R)        & NA         & NA                \\
-                                                    & \ranger (R)    & 0.06       & 4 min 30 seconds     \\
+                                                    & \ranger (R)    & 0.06       & 4 min 30 seconds     \\  %270
                                                     & \ranger (C++)     &     ??   & ??        \\
-                                                    & H2O           & 0.05       & 16 hours         \\
+                                                    & H2O           & 0.05       & 16 hours         \\ %57600
 
@@
 
-\multirow{3}{*}{phase3\_chr1: 2,504 x 6,450,364}    & \cursedforest & 0.04  & 19 min 19 seconds         \\
-                                                    & Spark ML  & 0.04       & 8 hour 29 min        \\
+\multirow{3}{*}{phase3\_chr1: 2,504 x 6,450,364}    & \cursedforest & 0.04  & 19 min 19 seconds         \\ %1159
+                                                    & Spark ML  & 0.04       & 8 hour 29 min        \\ % 3403
                                                     & \randomforest (R)        & NA         & NA                \\
-                                                    & \ranger (R)      &  0.05       & 28 mins         \\
+                                                    & \ranger (R)      &  0.05       & 28 mins         \\ %1680
                                                     & \ranger (C++)     & ??       & ??            \\
@@
 
-\multirow{3}{*}{phase3\_chr1-3: 2,504 x 19,328,051} & \cursedforest & 0.02  & 56 min 43 seconds             \\
-                                                    & Spark ML &    NA         &   NA               \\
+\multirow{3}{*}{phase3\_chr1-3: 2,504 x 19,328,051} & \cursedforest & 0.02  & 56 min 43 seconds             \\ %3403
+                                                    & Spark ML &    NA         &   NA               \\ %30540
                                                     & \randomforest (R)        & NA         & NA                \\
-                                                    & \ranger (R)      & 0.03      &    3 hour 58 min     \\
+                                                    & \ranger (R)      & 0.03      &    3 hour 58 min     \\ %14280
                                                     & \ranger (C++)     & ??       & ??            \\
@@
 
-\multirow{4}{*}{phase3\_chr1-22: 2,504 x 81,047,467} & \cursedforest & 0.01  & 3hour 16min \\
+\multirow{4}{*}{phase3\_chr1-22: 2,504 x 81,047,467} & \cursedforest & 0.01  & 3hour 16min \\ %11760
                                                     & Spark ML & NA & NA  \\
@@
 \end{table}
+
+\fi
 

### 2017-01-24 doc/publications/plos16/plos16.tex
Fix figure and table refs

@@
 homozygous variant (1\textbar1) as 2.  This is not the only possible encoding.  \cite{Goldstein.et.al.2011} note that of
-the encoding listed in table \ref{table:encodings}, the additive encoding will present the most difficulty for a random
+the encoding listed in Table~\ref{table:encodings}, the additive encoding will present the most difficulty for a random
 forest model.
@@
 \centering
-\caption{\bf Possible encodings of GWAS data \cite[]{Goldstein.et.al.2011}}
+\caption{\bf Possible encodings of GWAS data \cite{Goldstein.et.al.2011}}
 \begin{tabular}{|l|l|l|l||}
@@
 
-% Fig.~\ref{figure:out-of-bag-prediction-error-prod.png} illustrates that increasing the number of predictors sampled for splitting at each node
+% Fig~\ref{figure:out-of-bag-prediction-error-prod.png} illustrates that increasing the number of predictors sampled for splitting at each node
 % (\mtry) increases the accuracy with which features are ranked.  From $mtry=0.5$ the sort order of the features reflects
@@
 
-% Fig.~\ref{figure:rbo-prod.png} illustrates that increasing \mtry\ also improves the OOB
+% Fig~\ref{figure:rbo-prod.png} illustrates that increasing \mtry\ also improves the OOB
 % (out of bag) classification error.
@@
 \marginpar{this was done with n=10000. Needs to redone with n=2.5M and some things re-written}
-Fig.~\ref{figure:synth1} illustrates the effect of varying \mtry and \ntree on these two measures. The default \mtry
+Fig~\ref{figure:synth1} illustrates the effect of varying \mtry and \ntree on these two measures. The default \mtry
 value for a classification problem is $\sqrt{p}=100$ so in this case we are considering values considerably larger than
-the default. Fig.~\ref{figure:out-of-bag-prediction-error-prod.png} shows the OOB error is minimized at the largerst
-values of \mtry and \ntree considered  whereas Fig.~\ref{figure:rbo-prod.png} illustrates that the RBO measure is less
+the default. Fig~\ref{figure:out-of-bag-prediction-error-prod.png} shows the OOB error is minimized at the largerst
+values of \mtry and \ntree considered  whereas Fig~\ref{figure:rbo-prod.png} illustrates that the RBO measure is less
 sensitive once we over a certain threshold.
@@
 \cursedforest solves this problem of high-dimensionality genomic data. Not only is it faster on the above smaller datasets,
-but it is also able to scale to the entire genome. In table~\ref{cursedforesttable} we present our observed times building
+but it is also able to scale to the entire genome. In Table~\ref{cursedforesttable} we present our observed times building
 models with CursedForest using all available resources on our Spark cluster.
@@
 samples, by varying the \mtry  fraction and the number of CPU cores allocated to the execution. 
-The results are  visualized in figure \ref{figure:synthetictiming} below. 
+The results are  visualized in Fig~\ref{figure:synthetictiming} below. 
 %The raw results are
-%presented in table \ref{synthetictimingtable} and visualized in figure \ref{figure:synthetictiming} below. 
+%presented in Table~\ref{synthetictimingtable} and visualized in figure \ref{figure:synthetictiming} below. 
 Regardless of the number of cores used, \cursedforest displays approximately linear dependency between the execution
-time and \mtry (Fig. \ref{figure:synthetictiming.a}).
+time and \mtry (Fig~\ref{figure:synthetictiming.a}).
 
 \cursedforest scales almost linearly with the number of CPU cores for medium values of \mtry fraction but for both
-lower and higher values the performance degrades slightly (Fig \ref{figure:synthetictiming.b}). In the latter case the likely cause is
+lower and higher values the performance degrades slightly (Fig~\ref{figure:synthetictiming.b}). In the latter case the likely cause is
 communication overhead (with lower \mtry values the proportion of time for parallelizable computation to the time for
@@
 regards to the size of data, by varying the number of variables and sample for a fixed \mtry fraction of 0.25 and
-execution of 128 CPU cores. The results are visualized in figure \ref{figure:synth2} below (please note log scale on the
+execution of 128 CPU cores. The results are visualized in Fig~\ref{figure:synth2} below (please note log scale on the
 axes and the values on y axes are expresses as trees per hour).
@@
 Donoho and Tanner \cite{Donoho.and.Tanner.2009} give a ``universal phase change'' result that has applications in a
-large number of areas including variable selection in high dimensions. Consider figure
-\ref{figure:phase-diagram-equivalence.png} which shows the region where a model can recover the important variables,
+large number of areas including variable selection in high dimensions. Consider
+Fig~\ref{figure:phase-diagram-equivalence.png} which shows the region where a model can recover the important variables,
 plotted as a function of $\delta = n/p$ and $\rho =k/n$ (where $k$ is the number of significant variables). There is a

### 2017-01-24 doc/publications/plos16/plos16.tex
Mostly typo corrections

@@
 horizontally by capturing more information about these events.  The challenge of big and ``wide'' data is especially
-pronounced in the health space where, for example, whole genome sequence (WGS) technology enables researchers to
+pronounced in the health space where, for example, whole genome sequencing (WGS) technology enables researchers to
 interrogate all 3 billion base pairs of the human genome.
@@
 variance form. A number of different additive decompositions of classification error have been considered in an effort
-to understand the success of bagging. They all less transparent than the regression case (see \cite{Friedman.1997} and
+to understand the success of bagging. They are all less transparent than the regression case (see \cite{Friedman.1997} and
 citations therein).
@@
   \item differences in variable scale (for $X_i$ continuous) and number of levels (for $X_i$ discrete) introduces a
-    bias. A uninformative variable with a large number of levels may be selected over a more informative variable with
+    bias. An uninformative variable with a large number of levels may be selected over a more informative variable with
     fewer levels;
@@
 
-Biau {\it et al.}~\cite{Biau.2012} provide a proof that a random forest model will converge at a rate that depends on
+Biau~{\it et al.}~\cite{Biau.2012} provide a proof that a random forest model will converge at a rate that depends on
 the cardinality of the set $S$ of ``strong predictors'' rather than on the number of variables $p$. That is, given a
@@
 However, within an RDD, spark ML stores each sample as a vector. Unlike RDDs, which can be partitioned and distributed
-across multiple nodes, each vector must be present in its entirity on any node accessing it. This is no problem with
+across multiple nodes, each vector must be present in its entirety on any node accessing it. This is no problem with
 typical datasets, however as dimensionality increases, the vectors eventually reach a size where they can no longer fit
@@
 
-The current implementation of \cursedforest uses an ``Gini impurity'' criteria for splitting. Let $f_q$ be the fraction
+The current implementation of \cursedforest uses a ``Gini impurity'' criteria for splitting. Let $f_q$ be the fraction
 of items labelled with value $q$ where $q = 1, \ldots, Q$ at a node. The Gini impurity is
@@
 $$
-which is a minimum when all observatkions at the node are in the same class.
+which is a minimum when all observations at the node are in the same class.
 
@@
 variant (1\textbar0 or 0\textbar1) as 1, and a
-homozygous variant (1\textbar1) as 2.  This is not the only possible encoding.  \cite{Goldstein.et.al.2011} note that of
-the encoding listed in Table~\ref{table:encodings}, the additive encoding will present the most difficulty for a random
+homozygous variant (1\textbar1) as 2.  However, this is not the only possible encoding \cite{Goldstein.et.al.2011}. Note that from
+the encodings listed in Table~\ref{table:encodings}, the additive encoding will present the most difficulty for a random
 forest model.
@@
   minimum size of terminal nodes. It is controlled in Spark ML by setting \texttt{maxDepth}, the maximum depth of each
-  tree in the forest\footnote{The Spark ML documentation \cite[]{Spark.2016} sets \texttt{maxDepth} to 4 in their
+  tree in the forest\footnote{The Spark ML documentation \cite{Spark.2016} sets \texttt{maxDepth} to 4 in their
     classification example. This may give an estimate with a higher bias and, as such, may be a poor choice. See the
     discussion in \cite{Dietterich.2002} where it is shown that bagging may reduce bias as well as variance but gives a
-    better result for low bias learners. \cite{Dietterich.2002} notes that if the boostrap replicate approximation were
+    better result for low bias learners. \cite{Dietterich.2002} notes that if the bootstrap replicate approximation were
     correct (i.e.~if the bootstrap sample came from an identical distribution to the data), then bagging would reduce
@@
 specific phenotype.
-In this case we aim to identify variants associated with Bone Minderal Density (BMD). We build a model on the BMD dataset
+In this case we aim to identify variants associated with Bone Mineral Density (BMD). We build a model on the BMD dataset
 (see datasets) using a binary high or low BMD value as the label.
@@
 the highest Gini impurity---and hence, importance---is ranked 1, and the variant with the lowest Gini impurity is ranked 10,000.
-Then, using the 26 known and 6 novel BMD genes from Duncan et al., we compile a list of variants that are present in or near one of these
+Then, using the 26 known and 6 novel BMD genes from Duncan~{\it et al.}, we compile a list of variants that are present in or near one of these
 32 genes.
@@
 
-We also frequently see highly-ranked variants in chromosome 22q11. Stagi et al. observe that individuals with a microdeletion
+We also frequently see highly-ranked variants in chromosome 22q11. Stagi~{\it et al.} observe that individuals with a microdeletion
 in this region have a significant in bone mass\cite{Stagi2010}. In this same region, we also observe variants present in
-PRODH (proline dehydrogenase). PRODH catalyzes the first step in proline degredation, and as proposed by Phang et al., proline
+PRODH (proline dehydrogenase). PRODH catalyzes the first step in proline degradation, and as proposed by Phang~{\it et al.}, proline
 is made available by collagen degradation \cite{Phang2008}.
 
-DCDC5 (Doublecortin Domain Containing 5) is one of the 26 known genes, however it is one of the genes that Duncan et al.
-failed to replicate in their study. DCDC5 does appear in our list of genes, however the first occurance of a variant in
+DCDC5 (Doublecortin Domain Containing 5) is one of the 26 known genes, however it is one of the genes that Duncan~{\it et al.}.
+failed to replicate in their study. DCDC5 does appear in our list of genes, however the first occurrence of a variant in
 or near this gene has a rank of 9,667 out of 10,000. Also, this gene is not highly expressed in bone, and any mechanisms
@@
 \subsection{\cursedforest outperforms existing methods for classification problems}
-In this section we compare \cursedforest 's performance against other published methods for predicting ethnicity on the
+In this section we compare \cursedforest's performance against other published methods for predicting ethnicity on the
 1000 Genomes Project dataset.
@@
 $\bigotimes$ marks the last dataset the different method completed before running out of memory or running more than 24 hours.
-The dashed lines mark the trajectory for how ranger and SparkML had performed given larger memory machines.
+The dashed lines mark the trajectory for how ranger and Spark ML had performed given larger memory machines.
     \end{flushleft}
@@
   linear model. However, in a decision tree (random forest) there is no notion of estimating the $\beta$
-\item A decision tree (random forest) is a heruistic search. It may recover a relationship in the space of the
+\item A decision tree (random forest) is a heuristic search. It may recover a relationship in the space of the
   combinatorial search.
@@
 
-The existence of the  Donoho-Tanner phase transition is a salutory warning. There are likley to be limits, both
-computionaal and logical, to the recovery of signals from noisy data. \cursedforest  is a contribution to addressing the
-practivcal limits but the logical limits will still apply. However in the case of data that is both big and wide,
+The existence of the  Donoho-Tanner phase transition is a salutary warning. There are likely to be limits, both
+computational and logical, to the recovery of signals from noisy data. \cursedforest  is a contribution to addressing the
+practical limits but the logical limits will still apply. However in the case of data that is both big and wide,
 \cursedforest and other VariantSpark methods may provide a useful tool.

### 2017-01-26 doc/publications/plos16/plos16.tex
chaged figuers, fixed references, etc

@@
 \usepackage[T1]{fontenc}
+%% required the installation of cm-super 55MB - TeX font package (full version) with CM (EC) in Type1 in T1, T2*, TS1, X2 enc
 
@@
 well. Harnessing the virtually unlimited capability to parallelize tasks, \cursedforest can build a large number of
-trees and use a high \mtry value which leads to greater accuracy (see the example in section \ref{section:synthetic}). 
+trees and use a high \mtry value which may lead to greater accuracy (see the example in section \ref{section:synthetic}).
+
 \marginpar{does \cite{Szymczak2016} say that high \ntree gives greater  accuracy for variable importance selection?
-We do not see that. }
+Check this. }
 
@@
 
-Biau~{\it et al.}~\cite{Biau.2012} provide a proof that a random forest model will converge at a rate that depends on
+\cite{Bau.2012} provides a proof that a random forest model will converge at a rate that depends on
 the cardinality of the set $S$ of ``strong predictors'' rather than on the number of variables $p$. That is, given a
@@
 \subsubsection{Bone mineral density dataset}
-We obtained a VCF file from the authors of \cite{Duncan2011}, which contains variant data from 1,936 postmenopausal women.
+We obtained a VCF file from the authors of \cite{Duncan.2011}, which contains variant data from 1,936 postmenopausal women.
 This dataset covers approximately 7 million variant positions. We have information about the bone mineral density (BMD)
-for each individual, where a low BMD is a major predisposing factor to fracture \cite{Duncan2011}.
+for each individual, where a low BMD is a major predisposing factor to fracture \cite{Duncan.2011}.
 
@@
 
-\marginpar{this was done with n=10000. Needs to redone with n=2.5M and some things re-written}
 Fig~\ref{figure:synth1} illustrates the effect of varying \mtry and \ntree on these two measures. The default \mtry
-value for a classification problem is $\sqrt{p}=100$ so in this case we are considering values considerably larger than
-the default. Fig~\ref{figure:out-of-bag-prediction-error-prod.png} shows the OOB error is minimized at the largerst
-values of \mtry and \ntree considered  whereas Fig~\ref{figure:rbo-prod.png} illustrates that the RBO measure is less
-sensitive once we over a certain threshold.
+value for a classification problem is $\sqrt{p} \approx 1581$ so in this case we are considering values considerably larger than
+the default. Fig~\ref{figure:out-of-bag-prediction-error-prod.png} shows the OOB error is minimized at the largest
+values of \mtry and \ntree considered  whereas Fig~\ref{figure:rbo-prod.png} illustrates that the RBO measure depends
+heavily on \ntree but gives a better valuye with lower values of \mtry.
 
@@
       \centering
-      \includegraphics[totalheight=8cm]{./figs/out-of-bag-prediction-error.png}
+      \includegraphics[totalheight=8cm]{./figs/oob_levleplot.png}
       \caption{OOB classification error.} 
@@
       \centering
-      \includegraphics[totalheight=8cm]{./figs/rbo.png}
+      \includegraphics[totalheight=8cm]{./figs/rbo_levleplot.png}
       \caption{The rank-biased-overlap measure.} 
@@
 We investigate the ability of using a random forest model we build on genomic data to identify variants related to a
-specific phenotype.
-In this case we aim to identify variants associated with Bone Mineral Density (BMD). We build a model on the BMD dataset
-(see datasets) using a binary high or low BMD value as the label.
-From this model, we output a list of the top 10,000 contributing variants, ranked by their Gini impurity. The variant with
-the highest Gini impurity---and hence, importance---is ranked 1, and the variant with the lowest Gini impurity is ranked 10,000.
-Then, using the 26 known and 6 novel BMD genes from Duncan~{\it et al.}, we compile a list of variants that are present in or near one of these
-32 genes.
+specific phenotype.  In this case we aim to identify variants associated with Bone Mineral Density (BMD). We build a
+model on the BMD dataset (see datasets) using a binary high or low BMD value as the label.  From this model, we output a
+list of the top 10,000 contributing variants, ranked by their Gini impurity. The variant with the highest Gini impurity,
+and hence importance, is ranked 1, and the variant with the lowest Gini impurity is ranked 10,000.  Then, using the 26
+known and 6 novel BMD genes from \cite{Duncan.2011}, we compile a list of variants that are present in or near one of
+these 32 genes.
 
 We then apply the Mann–Whitney U test to this list to verify that from our list, variants from known BMD loci are ranked
-higher than variants from elsewhere in the genome.
-We observe $P=1.3\times10^{-07}$, supporting our hypothesis that variants in close proximity to the known BMD genes are
-contributing more information to our classification model.
-
-However, we also observe that from the set of highest-ranked features---for example, the top ten---not all of them exist
+higher than variants from elsewhere in the genome.  We observe $P=1.3\times10^{-07}$, supporting our hypothesis that
+variants in close proximity to the known BMD genes are contributing more information to our classification model.
+
+However, we also observe that from the set of highest-ranked features, for example, the top ten, not all of them exist
 within close proximity to known BMD genes. We investigate the literature to determine the plausibility of these variants
@@
 
-We also frequently see highly-ranked variants in chromosome 22q11. Stagi~{\it et al.} observe that individuals with a microdeletion
-in this region have a significant in bone mass\cite{Stagi2010}. In this same region, we also observe variants present in
-PRODH (proline dehydrogenase). PRODH catalyzes the first step in proline degradation, and as proposed by Phang~{\it et al.}, proline
-is made available by collagen degradation \cite{Phang2008}.
-
-DCDC5 (Doublecortin Domain Containing 5) is one of the 26 known genes, however it is one of the genes that Duncan~{\it et al.}.
+We also frequently see highly-ranked variants in chromosome 22q11. \cite{Stagi.2010}, observe that individuals with a
+microdeletion in this region have a significant in bone mass\cite{Stagi2010}. In this same region, we also observe
+variants present in PRODH (proline dehydrogenase). PRODH catalyzes the first step in proline degradation, and as
+proposed by \cite{Phang.2008}, proline is made available by collagen degradation \cite{Phang2008}.
+
+DCDC5 (Doublecortin Domain Containing 5) is one of the 26 known genes, however it is one of the genes that \cite{Duncan.2011}.
 failed to replicate in their study. DCDC5 does appear in our list of genes, however the first occurrence of a variant in
@@
 
-In a previous paper, we clustered individuals from the 1000 Genomes Project using k-means clustering\cite{Obrien2016}.
+In a previous paper, we clustered individuals from the 1000 Genomes Project using k-means clustering \cite{Obrien2016}.
+\marginpar{reference is missing}
 Although we were able to cluster individuals on all 80 million variants, it was relatively slow, with the entire
 process (pre-processing and clustering) taking over 30 hours.
-We investigate building random forest models on this data---however we observed this process to not only be slower
-than building a k-means model, but also unable to scale to as many dimensions (see Table~\ref{timingtable}). To keep
+We investigate building random forest models on this data. However, we observed this process to not only be slower
+than building a k-means model, but does not scale to as many dimensions (see Table~\ref{timingtable}). To keep
 the following comparisons fair, we limit the number of executors available to Spark. On the relatively low-dimensional
-phase1\_chr22 dataset (1,092 x 490,036), we observed a runtime of under 8 minutes. Although slower than Ranger---
-presumably due to the overhead of Spark's distributed platform on this relatively small dataset---we observe Spark ML to
+phase1\_chr22 dataset (1,092 x 490,036), we observed a runtime of under 8 minutes. Although slower than Ranger, 
+presumably due to the overhead of Spark's distributed platform on this relatively small dataset, we observe Spark ML to
 be about 16 minutes faster than R. Moving to the larger dataset, phase3\_chr1 (2,504 x 6,450,364), we observe an increase
@@
 
-From our trials, it was apparent that Spark ML---although designed to handle many samples---is unable to efficiently cope
+From our trials, it was apparent that Spark ML, although designed to handle many samples, is unable to efficiently cope
 with a huge number of features. With the decreasing costs of genome sequencing, this case will surely become increasingly
@@
 from the 1000 Genomes Project.
-Each random forest model consists of 50 trees and was built using 130 worker nodes---each with 8GB RAM.
+Each random forest model consists of 50 trees and was built using 130 worker nodes, each with 8GB RAM.
 \end{flushleft}
@@
 
-We investigate the different synthetic datasets generated for section \ref{ection:synthetic_data} and measured the time
+We investigate the different synthetic datasets generated for section \ref{section:synthetic_data} and measured the time
 taken to build a random forest model of 100 trees. The results reported below are averages of 5 runs, and all the cases

### 2017-01-30 doc/publications/plos16/plos16.tex
fixed a couple of references

@@
 
-\cite{Bau.2012} provides a proof that a random forest model will converge at a rate that depends on
+\cite{Biau.2012} provides a proof that a random forest model will converge at a rate that depends on
 the cardinality of the set $S$ of ``strong predictors'' rather than on the number of variables $p$. That is, given a
@@
 We also frequently see highly-ranked variants in chromosome 22q11. \cite{Stagi.2010}, observe that individuals with a
-microdeletion in this region have a significant in bone mass\cite{Stagi2010}. In this same region, we also observe
+microdeletion in this region have a significant in bone mass \cite{Stagi.2010}. In this same region, we also observe
 variants present in PRODH (proline dehydrogenase). PRODH catalyzes the first step in proline degradation, and as
-proposed by \cite{Phang.2008}, proline is made available by collagen degradation \cite{Phang2008}.
+proposed by \cite{Phang.2008}, proline is made available by collagen degradation \cite{Phang.2008}.
 

### 2017-01-31 doc/publications/plos16/plos16.tex
updated result section 2

--- 2017-01-26 doc/publications/plos16/plos16.tex
@@
 \subsection{\cursedforest\ selects previously demonstrated bone-density variants}
-We investigate the ability of using a random forest model we build on genomic data to identify variants related to a
-specific phenotype.  In this case we aim to identify variants associated with Bone Mineral Density (BMD). We build a
+In this section we use \cursedforest\ to identify genomic locations that are associated with a specific trait. To do this, we perform feature selection over XX genomic variants to 
+identify the locations associated with Bone Mineral Density (BMD)~\cite{Duncan.2011}. We build a
 model on the BMD dataset (see datasets) using a binary high or low BMD value as the label.  From this model, we output a
 list of the top 10,000 contributing variants, ranked by their Gini impurity. The variant with the highest Gini impurity,
-and hence importance, is ranked 1, and the variant with the lowest Gini impurity is ranked 10,000.  Then, using the 26
-known and 6 novel BMD genes from \cite{Duncan.2011}, we compile a list of variants that are present in or near one of
-these 32 genes.
-
-We then apply the Mann–Whitney U test to this list to verify that from our list, variants from known BMD loci are ranked
-higher than variants from elsewhere in the genome.  We observe $P=1.3\times10^{-07}$, supporting our hypothesis that
-variants in close proximity to the known BMD genes are contributing more information to our classification model.
-
-However, we also observe that from the set of highest-ranked features, for example, the top ten, not all of them exist
-within close proximity to known BMD genes. We investigate the literature to determine the plausibility of these variants
-contributing to BMD.
-
-One gene is COLEC10 (collectin 10). Featuring a transcription factor binding site for FOXL1, COLEC10 is also reported as
-containing variants associated with BMD \cite{Kemp2014,Liu2008}. However there currently exists no information on how this gene
-may contribute to BMD.
-
-We also frequently see highly-ranked variants in chromosome 22q11. \cite{Stagi.2010}, observe that individuals with a
-microdeletion in this region have a significant in bone mass\cite{Stagi2010}. In this same region, we also observe
+and hence importance, is ranked 1, and the variant with the lowest Gini impurity is ranked 10,000.
+
+%DENIS: number
+To test whether the associated features have biological meaning, we investigate whether the known BMD genes investigated by Duncan {\it et al.}~\cite{Duncan.2011} are enriched amongst the 
+highly ranked genes. We therefore label all variants that fall within XXbp of these 32 BMD genes and observe that they are
+ranked significantly higher than other locations in the genome (Mann-Whitney U, $p=1.3\times10^{-07}$), supporting our hypothesis that biologically relevant features are contributing more information to our classification model.
+
+We therefore investigate the highest-ranked features not amongst the associated gene list by Duncan {\it et al.}~\cite{Duncan.2011}. 
+%DENIS is this correct about the transcription factor binding site?
+COLEC10 (collectin 10) contains variants more recently associated with BDM~\cite{Kemp2014,Liu2008}, with a transcription factor binding site for the known BMD gene FOXL1 overlapping the highly ranked variant.
+%However, we also observe that from the set of highest-ranked features, for example, the top ten, not all of them exist
+%within close proximity to known BMD genes. We investigate the literature to determine the plausibility of these variants
+%contributing to BMD.
+%One gene is COLEC10 (collectin 10). Featuring a transcription factor binding site for FOXL1, COLEC10 is also reported as
+%containing variants associated with BMD \cite{Kemp2014,Liu2008}. However there currently exists no information on how this gene
+%may contribute to BMD. 
+Also highly-ranked are variants in chromosome 22q11. Stagi {\it et al.}~\cite{Stagi.2010}, observe that individuals with a
+microdeletion in this region have a significant reduction in bone mass. In this same region, we also observe
 variants present in PRODH (proline dehydrogenase). PRODH catalyzes the first step in proline degradation, and as
-proposed by \cite{Phang.2008}, proline is made available by collagen degradation \cite{Phang2008}.
-
-DCDC5 (Doublecortin Domain Containing 5) is one of the 26 known genes, however it is one of the genes that \cite{Duncan.2011}.
-failed to replicate in their study. DCDC5 does appear in our list of genes, however the first occurrence of a variant in
-or near this gene has a rank of 9,667 out of 10,000. Also, this gene is not highly expressed in bone, and any mechanisms
-by which this gene might regulate BMD are unclear \cite{Thakker2012}. Although variants near DCDC5 are associated with
-BMD \cite{Rivadeneira2009}, it is questionable as to whether the gene itself contributes to BMD.
+proposed by Phang {\it et al.}~\cite{Phang.2008}, proline is made available by collagen degradation.
+
+Investigating the lowly-ranked genes, we observe DCDC5 (Doublecortin Domain Containing 5) ranked 9,667 out of 10,000. This gene is 
+one of the BMD-associated genes postulated by literature that Duncan {\it et al.}~\cite{Duncan.2011} inspect.
+Similar to their findings, our results also indicate that this gene does not have a strong association with BMD. 
+In fact, DCDC5 is not highly expressed in bone, and any mechanisms by which this gene might regulate BMD are unclear \cite{Thakker2012}. 
+Although variants near DCDC5 are associated with BMD \cite{Rivadeneira2009}, it is questionable as to whether the gene itself contributes to BMD.
 

### 2017-01-31 doc/publications/plos16/plos16.tex
fix typos

--- 2017-01-30 doc/publications/plos16/plos16.tex
@@
 \usepackage[T1]{fontenc}
-%% required the installation of cm-super 55MB - TeX font package (full version) with CM (EC) in Type1 in T1, T2*, TS1, X2 enc
 
@@
 well. Harnessing the virtually unlimited capability to parallelize tasks, \cursedforest can build a large number of
-trees and use a high \mtry value which may lead to greater accuracy (see the example in section \ref{section:synthetic}).
-
+trees and use a high \mtry value which leads to greater accuracy (see the example in section \ref{section:synthetic}). 
 \marginpar{does \cite{Szymczak2016} say that high \ntree gives greater  accuracy for variable importance selection?
-Check this. }
+We do not see that. }
 
@@
 
-\cite{Biau.2012} provides a proof that a random forest model will converge at a rate that depends on
+Biau~{\it et al.}~\cite{Biau.2012} provide a proof that a random forest model will converge at a rate that depends on
 the cardinality of the set $S$ of ``strong predictors'' rather than on the number of variables $p$. That is, given a
@@
 \subsubsection{Bone mineral density dataset}
-We obtained a VCF file from the authors of \cite{Duncan.2011}, which contains variant data from 1,936 postmenopausal women.
+We obtained a VCF file from the authors of \cite{Duncan2011}, which contains variant data from 1,936 postmenopausal women.
 This dataset covers approximately 7 million variant positions. We have information about the bone mineral density (BMD)
-for each individual, where a low BMD is a major predisposing factor to fracture \cite{Duncan.2011}.
+for each individual, where a low BMD is a major predisposing factor to fracture \cite{Duncan2011}.
 
@@
 
+\marginpar{this was done with n=10000. Needs to redone with n=2.5M and some things re-written}
 Fig~\ref{figure:synth1} illustrates the effect of varying \mtry and \ntree on these two measures. The default \mtry
-value for a classification problem is $\sqrt{p} \approx 1581$ so in this case we are considering values considerably larger than
-the default. Fig~\ref{figure:out-of-bag-prediction-error-prod.png} shows the OOB error is minimized at the largest
-values of \mtry and \ntree considered  whereas Fig~\ref{figure:rbo-prod.png} illustrates that the RBO measure depends
-heavily on \ntree but gives a better valuye with lower values of \mtry.
+value for a classification problem is $\sqrt{p}=100$ so in this case we are considering values considerably larger than
+the default. Fig~\ref{figure:out-of-bag-prediction-error-prod.png} shows the OOB error is minimized at the largerst
+values of \mtry and \ntree considered  whereas Fig~\ref{figure:rbo-prod.png} illustrates that the RBO measure is less
+sensitive once we over a certain threshold.
 
@@
       \centering
-      \includegraphics[totalheight=8cm]{./figs/oob_levleplot.png}
+      \includegraphics[totalheight=8cm]{./figs/out-of-bag-prediction-error.png}
       \caption{OOB classification error.} 
@@
       \centering
-      \includegraphics[totalheight=8cm]{./figs/rbo_levleplot.png}
+      \includegraphics[totalheight=8cm]{./figs/rbo.png}
       \caption{The rank-biased-overlap measure.} 
@@
 We investigate the ability of using a random forest model we build on genomic data to identify variants related to a
-specific phenotype.  In this case we aim to identify variants associated with Bone Mineral Density (BMD). We build a
-model on the BMD dataset (see datasets) using a binary high or low BMD value as the label.  From this model, we output a
-list of the top 10,000 contributing variants, ranked by their Gini impurity. The variant with the highest Gini impurity,
-and hence importance, is ranked 1, and the variant with the lowest Gini impurity is ranked 10,000.  Then, using the 26
-known and 6 novel BMD genes from \cite{Duncan.2011}, we compile a list of variants that are present in or near one of
-these 32 genes.
+specific phenotype.
+In this case we aim to identify variants associated with Bone Mineral Density (BMD). We build a model on the BMD dataset
+(see datasets) using a binary high or low BMD value as the label.
+From this model, we output a list of the top 10,000 contributing variants, ranked by their Gini impurity. The variant with
+the highest Gini impurity---and hence, importance---is ranked 1, and the variant with the lowest Gini impurity is ranked 10,000.
+Then, using the 26 known and 6 novel BMD genes from Duncan~{\it et al.}, we compile a list of variants that are present in or near one of these
+32 genes.
 
 We then apply the Mann–Whitney U test to this list to verify that from our list, variants from known BMD loci are ranked
-higher than variants from elsewhere in the genome.  We observe $P=1.3\times10^{-07}$, supporting our hypothesis that
-variants in close proximity to the known BMD genes are contributing more information to our classification model.
-
-However, we also observe that from the set of highest-ranked features, for example, the top ten, not all of them exist
+higher than variants from elsewhere in the genome.
+We observe $P=1.3\times10^{-07}$, supporting our hypothesis that variants in close proximity to the known BMD genes are
+contributing more information to our classification model.
+
+However, we also observe that from the set of highest-ranked features---for example, the top ten---not all of them exist
 within close proximity to known BMD genes. We investigate the literature to determine the plausibility of these variants
@@
 
-We also frequently see highly-ranked variants in chromosome 22q11. \cite{Stagi.2010}, observe that individuals with a
-microdeletion in this region have a significant in bone mass \cite{Stagi.2010}. In this same region, we also observe
-variants present in PRODH (proline dehydrogenase). PRODH catalyzes the first step in proline degradation, and as
-proposed by \cite{Phang.2008}, proline is made available by collagen degradation \cite{Phang.2008}.
-
-DCDC5 (Doublecortin Domain Containing 5) is one of the 26 known genes, however it is one of the genes that \cite{Duncan.2011}.
+We also frequently see highly-ranked variants in chromosome 22q11. Stagi~{\it et al.} observe that individuals with a microdeletion
+in this region have a significant in bone mass\cite{Stagi2010}. In this same region, we also observe variants present in
+PRODH (proline dehydrogenase). PRODH catalyzes the first step in proline degradation, and as proposed by Phang~{\it et al.}, proline
+is made available by collagen degradation \cite{Phang2008}.
+
+DCDC5 (Doublecortin Domain Containing 5) is one of the 26 known genes, however it is one of the genes that Duncan~{\it et al.}.
 failed to replicate in their study. DCDC5 does appear in our list of genes, however the first occurrence of a variant in
@@
 
-In a previous paper, we clustered individuals from the 1000 Genomes Project using k-means clustering \cite{Obrien2016}.
-\marginpar{reference is missing}
+In a previous paper, we clustered individuals from the 1000 Genomes Project using k-means clustering\cite{Obrien2016}.
 Although we were able to cluster individuals on all 80 million variants, it was relatively slow, with the entire
 process (pre-processing and clustering) taking over 30 hours.
-We investigate building random forest models on this data. However, we observed this process to not only be slower
-than building a k-means model, but does not scale to as many dimensions (see Table~\ref{timingtable}). To keep
+We investigate building random forest models on this data---however we observed this process to not only be slower
+than building a k-means model, but also unable to scale to as many dimensions (see Table~\ref{timingtable}). To keep
 the following comparisons fair, we limit the number of executors available to Spark. On the relatively low-dimensional
-phase1\_chr22 dataset (1,092 x 490,036), we observed a runtime of under 8 minutes. Although slower than Ranger, 
-presumably due to the overhead of Spark's distributed platform on this relatively small dataset, we observe Spark ML to
+phase1\_chr22 dataset (1,092 x 490,036), we observed a runtime of under 8 minutes. Although slower than Ranger---
+presumably due to the overhead of Spark's distributed platform on this relatively small dataset---we observe Spark ML to
 be about 16 minutes faster than R. Moving to the larger dataset, phase3\_chr1 (2,504 x 6,450,364), we observe an increase
@@
 
-From our trials, it was apparent that Spark ML, although designed to handle many samples, is unable to efficiently cope
+From our trials, it was apparent that Spark ML---although designed to handle many samples---is unable to efficiently cope
 with a huge number of features. With the decreasing costs of genome sequencing, this case will surely become increasingly
@@
 from the 1000 Genomes Project.
-Each random forest model consists of 50 trees and was built using 130 worker nodes, each with 8GB RAM.
+Each random forest model consists of 50 trees and was built using 130 worker nodes---each with 8GB RAM.
 \end{flushleft}
@@
 
-We investigate the different synthetic datasets generated for section \ref{section:synthetic_data} and measured the time
+We investigate the different synthetic datasets generated for section \ref{ection:synthetic_data} and measured the time
 taken to build a random forest model of 100 trees. The results reported below are averages of 5 runs, and all the cases
@@
 execution of 128 CPU cores. The results are visualized in Fig~\ref{figure:synth2} below (please note log scale on the
-axes and the values on y axes are expresses as trees per hour).
-
-Generally, the number of trees per hour decreases with increased number of variables and samples sizes. Some
+axes and the values on y axes are expressed as trees per hour).
+
+Generally, the number of trees per hour decreases with an increased number of variables and samples sizes. Some
 irregularities in the graph can be attributed to computation vs communication tradeoff. It is also worth noting that

### 2017-01-31 doc/publications/plos16/plos16.tex
Reword a sentence in part 2

@@
 %DENIS is this correct about the transcription factor binding site?
-COLEC10 (collectin 10) contains variants more recently associated with BDM~\cite{Kemp2014,Liu2008}, with a transcription factor binding site for the known BMD gene FOXL1 overlapping the highly ranked variant.
+COLEC10 (collectin 10) contains variants more recently associated with BDM~\cite{Kemp2014,Liu2008}, with a verified transcription factor binding site for the known BMD gene FOXL1.
 %However, we also observe that from the set of highest-ranked features, for example, the top ten, not all of them exist

### 2017-01-31 doc/publications/plos16/plos16.tex
updated result section 1

--- 2017-01-31 doc/publications/plos16/plos16.tex
@@
 \label{section:synthetic}
-In this section we demonstrate two features of \cursedforest, a variable importance analysis to select the most
+In this section we demonstrate the two features of \cursedforest, namely variable importance analysis to select the most
 associated set of features amongst millions of dimensions as well as its ability to accurately classify the labels of
-high-dimensional samples.  We therefore generate a synthetic dataset with 2.5 million variables (features) and 5000
-samples, as discussed in the methods section.
-
-After fitting the random forest model and performing the variable importance analysis we hence expect to retrieve these
-5 feature in order of the weight that was assigned to them.  We are running Apache Spark 1.6.1 on a YARN cluster with 12
-worker nodes each with 16 Intel Xeon E5-2660@2.20GHz CPU cores and 128 GB of RAM.  We captured the OOB 
-classification error and the rank-biased-overlap (RBO) \cite{Webber2010} measure.
-
-% Fig~\ref{figure:out-of-bag-prediction-error-prod.png} illustrates that increasing the number of predictors sampled for splitting at each node
-% (\mtry) increases the accuracy with which features are ranked.  From $mtry=0.5$ the sort order of the features reflects
-% the weight originally assigned to them (dotted line). Further increasing \mtry\ improves the predicted importance
-% measure, which starts to converge to the weight originally assigned to them (coefficients).  It also increases the
-% difference between the associated 5 variable and the remaining variables in the top 20 list.  By $mtry=0.8$ the variable
-% with the smallest weight has a predicted importance measure that is 5 times larger than the aggregated importance of the
-% 15 remaining variables from the list.
-
-% Fig~\ref{figure:rbo-prod.png} illustrates that increasing \mtry\ also improves the OOB
-% (out of bag) classification error.
-% By $mtry=0.8$ the error is converging to X, which is due to Y.
-% \marginpar{what are ``X'' and ``Y''. Piotr}
-
-% From the synthetic data we can conclude that \cursedforest is able to perform accurate feature selection as well as
-% accurate classification for large numbers of samples with large feature vectors.
-
-Fig~\ref{figure:synth1} illustrates the effect of varying \mtry and \ntree on these two measures. The default \mtry
-value for a classification problem is $\sqrt{p} \approx 1581$ so in this case we are considering values considerably larger than
-the default. Fig~\ref{figure:out-of-bag-prediction-error-prod.png} shows the OOB error is minimized at the largest
-values of \mtry and \ntree considered  whereas Fig~\ref{figure:rbo-prod.png} illustrates that the RBO measure depends
-heavily on \ntree but gives a better valuye with lower values of \mtry.
+high-dimensional samples.  We therefore generate a synthetic dataset with 2.5 million variables (features, p) of which 5 
+are the response variable and 5000 samples (n), as discussed in the methods section.
+
+We fit the random forest model to obtain the classification accuracy by capturing the out-of-bag error (OOB) and perform the variable importance analysis where we expect to retrieve the
+5 features in order of the weight that was assigned to them, which is captured by the rank-biased-overlap (RBO) \cite{Webber2010} measure.
+We are running Apache Spark 1.6.1 on a YARN cluster with 12 worker nodes each with 16 Intel Xeon E5-2660@2.20GHz CPU cores and 128 GB of RAM.
+
+%mtry = number of variable seclection at each node - proportion of p 
+%ntree= number of trees
+
+% mtry large makes trees correlated because the same important features have a higher chance of being selected
+% overfitting because it picks random that happen to be
+
+As shown in Fig~\ref{figure:out-of-bag-prediction-error-prod.png} the default value for \mtry (number of variables selected at each node) does not result in a 
+good classification performance for this large feature dataset. Note, that the plot shows the proportion of $n/p$ for \mtry\ with the default being $\sqrt{p} \approx 1581$. 
+The OOB  for this value of \mtry does not exceed 0.5 on a scale from 0 (perfect) to 1 (random), even when the number of trees are increased (\ntree).
+
+Increasing \mtry\ in combination with \ntree\ yields best performance with the OOB error essentially constant around 0.4 across a large range of 
+of \mtry and \ntree values. 
+This is in contrast to the feature-selection performance, where the RBO measure heavily depends on \ntree\ and gives better results with lower values of \mtry, where 
+the best RBO is around 0.85 on a scale from 0 (no feature recovered) to 1 (fully recovered).
+
+This may be because a large \mtry\ generates more correlated trees as the same important features have a higher chance of being selected in all, not yielding good performance outcomes. 
+This issue is less pronounced for classification tasks where random features can mimic the response variable, hence resulting more in a performance plateau. 
+Increasing the number of trees on the other hand improves performance especially when 
+the trees are kept diverse (small \mtry) but appropriate for large feature datasets (\mtry larger than default).
+\cursedforest's feature of parallelising trees at the node level hence cater perfectly for the
+requirement of large feature datasets as more trees can be built given a fixed time budget.
+
+%Fig~\ref{figure:synth1} illustrates the effect of varying \mtry and \ntree on these two measures. The default \mtry
+%value for a classification problem is $\sqrt{p} \approx 1581$ so in this case we are considering values considerably larger than
+%the default. Fig~\ref{figure:out-of-bag-prediction-error-prod.png} shows the OOB error is constant across a large range of 
+%of \mtry and \ntree values whereas Fig~\ref{figure:rbo-prod.png} illustrates that the RBO measure depends
+%heavily on \ntree but gives a better value with lower values of \mtry.
 
@@
 accurate classification for large numbers of samples with large feature vectors.
+We can also conclude that fine tuning the model to accurately classify the dataset does not ensure good feature selection performance as there seems to be little correspondence between the OOB and RBO measure.
 

### 2017-02-01 doc/publications/plos16/plos16.tex
Updated equasions in methods section

@@
 $$
-We let $\sigma_\epsilon^2 = Var(z)(1- 0.125)/0.125 $ and then 
+We let $\sigma_\epsilon^2 = Var(z)(1-\theta)/\theta $  where $\theta$ is a parameter controling the fraction of variance explained by the infomative variables and in our study we chose $\theta = 0.125$ .
+Then
 $y =   z + \epsilon$ where $\epsilon \sim  N(0, \sigma_\epsilon^2).$
-The dichotomous response is generated by thresholding $y$ at the $0.8$ quantile.
+The dichotomous response is generated by thresholding $y$ at the $0.5$ quantile.
+
+$$
+\ddot{y} =    \left\{ \begin{array}{rcl}
+			  0 & \mbox{for} & y \leq Q2(y) \\
+			  1 & \mbox{for} & y > Q2(y)
+		 \end{array}\right.
+$$
 

### 2017-02-01 doc/publications/plos16/plos16.tex
some small changes to results section

@@
 \label{section:synthetic}
-In this section we demonstrate the two features of \cursedforest, namely variable importance analysis to select the most
-associated set of features amongst millions of dimensions as well as its ability to accurately classify the labels of
-high-dimensional samples.  We therefore generate a synthetic dataset with 2.5 million variables (features, p) of which 5 
-are the response variable and 5000 samples (n), as discussed in the methods section.
-
-We fit the random forest model to obtain the classification accuracy by capturing the out-of-bag error (OOB) and perform the variable importance analysis where we expect to retrieve the
-5 features in order of the weight that was assigned to them, which is captured by the rank-biased-overlap (RBO) \cite{Webber2010} measure.
-We are running Apache Spark 1.6.1 on a YARN cluster with 12 worker nodes each with 16 Intel Xeon E5-2660@2.20GHz CPU cores and 128 GB of RAM.
+In this section we demonstrate the two features of \cursedforest, namely: variable selection analysis to select the most
+associated set of features amongst millions of dimensions; and its ability to accurately classify the labels of
+high-dimensional samples.  We therefore generate a synthetic dataset with 2.5 million variables ($p=2500000$) of which 5 
+are related to the response variable and 5000 samples ($n=5000$), as discussed in the methods section.
+
+We fit the random forest model and estimate the classification accuracy by capturing the out-of-bag error (OOB) and the
+variable selection performance by capturing the rank-biased-overlap (RBO) \cite{Webber2010} measure. We expect to
+retrieve the 5 features in order of the weight that was assigned to them and with a higher variable importance measure
+than the noise variables. RBO provides a quantiative measure of this.  We are running Apache Spark 1.6.1 on a YARN
+cluster with 12 worker nodes each with 16 Intel Xeon E5-2660@2.20GHz CPU cores and 128 GB of RAM.
 
@@
 
-As shown in Fig~\ref{figure:out-of-bag-prediction-error-prod.png} the default value for \mtry (number of variables selected at each node) does not result in a 
-good classification performance for this large feature dataset. Note, that the plot shows the proportion of $n/p$ for \mtry\ with the default being $\sqrt{p} \approx 1581$. 
-The OOB  for this value of \mtry does not exceed 0.5 on a scale from 0 (perfect) to 1 (random), even when the number of trees are increased (\ntree).
-
-Increasing \mtry\ in combination with \ntree\ yields best performance with the OOB error essentially constant around 0.4 across a large range of 
-of \mtry and \ntree values. 
-This is in contrast to the feature-selection performance, where the RBO measure heavily depends on \ntree\ and gives better results with lower values of \mtry, where 
-the best RBO is around 0.85 on a scale from 0 (no feature recovered) to 1 (fully recovered).
-
-This may be because a large \mtry\ generates more correlated trees as the same important features have a higher chance of being selected in all, not yielding good performance outcomes. 
-This issue is less pronounced for classification tasks where random features can mimic the response variable, hence resulting more in a performance plateau. 
-Increasing the number of trees on the other hand improves performance especially when 
-the trees are kept diverse (small \mtry) but appropriate for large feature datasets (\mtry larger than default).
-\cursedforest's feature of parallelising trees at the node level hence cater perfectly for the
-requirement of large feature datasets as more trees can be built given a fixed time budget.
+As shown in Fig~\ref{figure:out-of-bag-prediction-error-prod.png} the default value for \mtry (number of variables
+selected at each node) does not result in a good classification performance for this large feature dataset. Note, that
+the plot shows the proportion of $n/p$ for \mtry\ with the default being $\sqrt{p} \approx 1581$ ($n/p=0.0006$).  The
+OOB for this value of \mtry does not drop below 0.5, even when the number of trees are increased (\ntree).
+
+Increasing \mtry\ in combination with \ntree\ yields best performance with the OOB error essentially constant around 0.4
+across a large range of of \mtry and \ntree values.  This is in contrast to the feature-selection performance, where the
+RBO measure heavily depends on \ntree\ and gives better results with lower values of \mtry, where the best RBO is around
+0.85 on a scale from 0 (no feature recovered) to 1 (fully recovered).
+
+This may be because a large \mtry\ generates more correlated trees as the same important features have a higher chance
+of being selected in all, not yielding good performance outcomes.  This issue is less pronounced for classification
+tasks where random features can mimic the response variable, hence resulting more in a performance plateau.  Increasing
+the number of trees on the other hand improves performance especially when the trees are kept diverse (small \mtry) but
+appropriate for large feature datasets (\mtry larger than default). \cursedforest's feature of parallelising trees at
+the node level hence cater perfectly for the requirement of large feature datasets as more trees can be built given a
+fixed time budget.
 
@@
 From the synthetic data we can conclude that \cursedforest is able to perform accurate feature selection as well as
-accurate classification for large numbers of samples with large feature vectors.
-We can also conclude that fine tuning the model to accurately classify the dataset does not ensure good feature selection performance as there seems to be little correspondence between the OOB and RBO measure.
+accurate classification for large numbers of samples with large feature vectors.  We can also conclude that fine tuning
+the model to accurately classify the dataset does not ensure good feature selection performance as there seems to be
+little correspondence between the OOB and RBO measure.
 
@@
       \label{figure:out-of-bag-prediction-error-prod.png} 
-      \vspace{4ex}
+%      \vspace{4ex}
     \end{subfigure} 
@@
       \label{figure:rbo-prod.png} 
-      \vspace{4ex}
+%      \vspace{4ex}
     \end{subfigure} 
@@
 
-\marginpar{tables 2 has been commented out -- there is no data in it. Piotr could add the data or we could remove both
-  tables just leave the figures. I have done this for the time being} 
 Next we investigate \cursedforest scalability with

### 2017-02-01 doc/publications/plos16/plos16.tex
Updated part ii

@@
 \subsection{\cursedforest\ selects previously demonstrated bone-density variants}
-In this section we use \cursedforest\ to identify genomic locations that are associated with a specific trait. To do this, we perform feature selection over XX genomic variants to 
-identify the locations associated with Bone Mineral Density (BMD)~\cite{Duncan.2011}. We build a
-model on the BMD dataset (see datasets) using a binary high or low BMD value as the label.  From this model, we output a
-list of the top 10,000 contributing variants, ranked by their Gini impurity. The variant with the highest Gini impurity,
-and hence importance, is ranked 1, and the variant with the lowest Gini impurity is ranked 10,000.
-
-%DENIS: number
-To test whether the associated features have biological meaning, we investigate whether the known BMD genes investigated by Duncan {\it et al.}~\cite{Duncan.2011} are enriched amongst the 
-highly ranked genes. We therefore label all variants that fall within XXbp of these 32 BMD genes and observe that they are
-ranked significantly higher than other locations in the genome (Mann-Whitney U, $p=1.3\times10^{-07}$), supporting our hypothesis that biologically relevant features are contributing more information to our classification model.
-
-We therefore investigate the highest-ranked features not amongst the associated gene list by Duncan {\it et al.}~\cite{Duncan.2011}. 
-%DENIS is this correct about the transcription factor binding site?
-COLEC10 (collectin 10) contains variants more recently associated with BDM~\cite{Kemp2014,Liu2008}, with a verified transcription factor binding site for the known BMD gene FOXL1.
+In this section we use \cursedforest\ to identify genomic locations that are associated with a specific trait. To do
+this, we perform feature selection over 7,225,008 genomic variants to identify the locations associated with Bone
+Mineral Density (BMD)~\cite{Duncan.2011}. We build a model on the BMD dataset (see datasets) using a binary high or low
+BMD value as the label.  From this model, we output a list of the top 10,000 contributing variants, ranked by their Gini
+impurity. The variant with the highest Gini impurity, and hence importance, is ranked 1, and the variant with the lowest
+Gini impurity is ranked 10,000.
+
+To test whether the associated features have biological meaning, we investigate whether the known BMD genes investigated
+by Duncan {\it et al.}~\cite{Duncan.2011} are enriched amongst the highly ranked genes. We therefore label all variants
+that fall within 100 kb of these 32 BMD genes and observe that they are ranked significantly higher than other locations
+in the genome (Mann-Whitney U, $p=1.3\times10^{-07}$), supporting our hypothesis that biologically relevant features are
+contributing more information to our classification model.
+
+We therefore investigate the highest-ranked features not amongst the associated gene list by Duncan
+{\it et al.}~\cite{Duncan.2011}. COLEC10 (collectin 10) contains variants more recently associated with
+BDM~\cite{Kemp2014,Liu2008}, with a verified transcription factor binding site for the known BMD gene FOXL1~\cite{DECODE}.
 %However, we also observe that from the set of highest-ranked features, for example, the top ten, not all of them exist

### 2017-02-02 doc/publications/plos16/plos16.tex
changes to results section

@@
 The model paramters are $w_i = 1/\sqrt{2^{i-1}}$ for $i =1,\ldots 5$ and we set 
-$$
+\begin{equation}
+\label{equation:synthetic.data}
 z =  \sum_{i=1}^{i=5} {w_i x_i}.
-$$
+\end{equation}
 We let $\sigma_\epsilon^2 = Var(z)(1-\theta)/\theta $  where $\theta$ is a parameter controling the fraction of variance explained by the infomative variables and in our study we chose $\theta = 0.125$ .
@@
 We fit the random forest model and estimate the classification accuracy by capturing the out-of-bag error (OOB) and the
-variable selection performance by capturing the rank-biased-overlap (RBO) \cite{Webber2010} measure. We expect to
-retrieve the 5 features in order of the weight that was assigned to them and with a higher variable importance measure
-than the noise variables. RBO provides a quantiative measure of this.  We are running Apache Spark 1.6.1 on a YARN
+variable selection performance by capturing the rank-biased-overlap (RBO) \cite{Webber2010} measure. RBO compares two
+lists (in this case, the paramters $w_i, i=1,\ldots , 5$ (equation \ref{equation:synthetic.data}), and the top 1000
+ranked variable importances from the \cursedforest model). It is a mesaure of similarity between rankings that handles
+list of different length, weighs high ranks more heavily than low, and is monotonic with increasing depth of
+evaluation. We expect to retrieve the 5 features in order of the weight that was assigned to them and with a higher
+variable importance measure than the noise variables. RBO provides a quantiative measure of how close we come to this on
+a scale from 0 (no feature recovered) to 1 (fully recovered). See supplementary information (S*) for more details.
+
+We are running Apache Spark 1.6.1 on a YARN
 cluster with 12 worker nodes each with 16 Intel Xeon E5-2660@2.20GHz CPU cores and 128 GB of RAM.
@@
 As shown in Fig~\ref{figure:out-of-bag-prediction-error-prod.png} the default value for \mtry (number of variables
-selected at each node) does not result in a good classification performance for this large feature dataset. Note, that
-the plot shows the proportion of $n/p$ for \mtry\ with the default being $\sqrt{p} \approx 1581$ ($n/p=0.0006$).  The
+selected at each node) does not result in a good classification performance for this large feature dataset. Note that
+the plot shows the proportion, $n/p,$ for \mtry\ with the default being $\sqrt{p} \approx 1581$ ($n/p=0.0006$).  The
 OOB for this value of \mtry does not drop below 0.5, even when the number of trees are increased (\ntree).
@@
 across a large range of of \mtry and \ntree values.  This is in contrast to the feature-selection performance, where the
-RBO measure heavily depends on \ntree\ and gives better results with lower values of \mtry, where the best RBO is around
-0.85 on a scale from 0 (no feature recovered) to 1 (fully recovered).
+RBO measure heavily depends on \ntree\ and gives better results with lower values of \mtry.
 
@@
 of being selected in all, not yielding good performance outcomes.  This issue is less pronounced for classification
-tasks where random features can mimic the response variable, hence resulting more in a performance plateau.  Increasing
+tasks where random features can mimic the response variable, hence resulting more in a performance plateau.  
+\marginpar{unclear what this means}
+Increasing
 the number of trees on the other hand improves performance especially when the trees are kept diverse (small \mtry) but

### 2017-02-08 doc/publications/plos16/plos16.tex
Many general changes and new reference

@@
 The digital revolution is seeing a dramatic increase in data collected about almost every aspect of
-life~\cite{Loebbecke2015}.  These data sets are not only growing vertically, by capturing more events, but also
+life~\cite{Loebbecke2015}.  These datasets are not only growing vertically, by capturing more events, but also
 horizontally by capturing more information about these events.  The challenge of big and ``wide'' data is especially
@@
 
-Identifying the relevant or disease specific base-pair difference between individuals is the focus of genome wide
-association studies (GWAs).  These analysis are typically performed on only the most frequently differing base-pairs,
+Identifying the relevant or disease specific base pair differences between individuals is the focus of genome wide
+association studies (GWAS).  These analyses are typically performed on only the most frequently differing base pairs,
 termed single nucleotide polymorphisms (SNPs), by applying linear or logistic regression analysis to each SNP
@@
 
-A tree based model is capable of taking the interaction of variables into account \cite{Wright.et.al.2016}. In addition, random forests are well
+A tree-based model is capable of taking the interaction of variables into account \cite{Wright.et.al.2016}. In addition, random forests are well
 suited for processing ``wide'' genomic data for two reasons.  Firstly, while other machine learning applications have
@@
 than $2^{31}-1$), and the \randomforest implementation in R lending itself to parallelization to grow multiple
-trees simultaneously~\cite{Liaw.and.Weiner.2002} and them combine them.  More suitable implementations for ``wide''
-genomic data have been developed in other programming languages. \textsc{CloudForest} \cite{Bressler2015} written in Go,
+trees simultaneously~\cite{Liaw.and.Weiner.2002} and then combine them.  More suitable implementations for ``wide''
+genomic data have been developed in other programming languages. \textsc{CloudForest} \cite{Bressler2015}, written in Go,
 achieves fast run times by effective use of the CPU cache, optimizing for different classes of features and efficiently
-multi-threading.  Similarly, \ranger~\cite{Wright.and.Ziegle.2016} written in C++ with an R front-end is
-specifically optimized for dealing with large data sets.
+multithreading.  Similarly, \ranger~\cite{Wright.and.Ziegle.2016}, written in C++ with an R front-end, is
+specifically optimized for dealing with large datasets.
 
 The use of traditional compute infrastructure limits the parallelization strategies that can be employed by  
-these methods.  The programs are limited to utilizing only CPUs that are on the same computer node (multi-threading) or
-else they run on CPUs distributed across nodes by virtue of the fact that there is no communication between the
-processes (a separate tree is grown on each node).  Hadoop/Spark overcomes these limitation by enabling programs to
+these methods.  The programs are limited to utilizing only CPUs that are on the same computer node (multithreading) or
+else they can run on CPUs distributed across nodes by virtue of the fact that there is no communication between the
+processes (a separate tree is grown on each node).  Hadoop/Spark overcomes these limitations by enabling programs to
 scale beyond compute-node boundaries and hence enable more sophisticated parallelization strategies.  In the case of
@@
 
-Here we introduce \cursedforest, a Hadoop/Spark based implementation of random forests specifically designed to cater
-for ``big'' (many samples) and ``wide'' (many features) data sets. A Spark application runs on a ``driver'' node and
+Here we introduce \cursedforest, a Hadoop/Spark-based implementation of random forests specifically designed to cater
+for ``big'' (many samples) and ``wide'' (many features) datasets. A Spark application runs on a ``driver'' node and
 distributes tasks to many ``worker'' nodes, or ``executors''. Using this facility, \cursedforest is capable of
-parallelization the split for each node in a tree thereby handling millions of features. It can, for example, fit a
+parallelizing the split for each node in a tree thereby handling millions of features. It can, for example, fit a
 model to the 1000 Genomes Project data, which consists of approximately 2500 samples with up to 80 million
 variants~\cite{1KG2012}.  By also utilizing Spark to read in and manipulate the standard genomic variant format (VCF)
-directly, \cursedforest outperforms existing tools even on small datasets where multi-threading generally performs
+directly, \cursedforest outperforms existing tools even on small datasets where multithreading generally performs
 well. Harnessing the virtually unlimited capability to parallelize tasks, \cursedforest can build a large number of
@@
 Here we extend this toolkit to encompass supervised learning using random forests.  Spark ML's random forest
-implementation is not able to handle the extremely ``wide'' genomic data as it was developed for large number of samples
+implementation is not able to handle the extremely ``wide'' genomic data as it was developed for a large number of samples
 with only modest dimensionality.  Although Spark ML can build a random forest model on a subset of the data (chromosome
@@
 VariantSpark~\cite{OBrien2015}, which therefore now enables supervised and unsupervised machine learning. This offers a
-comprehensive analysis toolkit that can scale for the future data demands, where the richness from WGS data can be
+comprehensive analysis toolkit that can scale for future data demands, where the richness from WGS data can be
 utilized to ``fill in'' or impute the information at previously unobserved genomic positions in older array
-technology~\cite{Howie2012}.  This provides the potential to impute samples in the GWAs catalogue and generate datasets
+technology~\cite{Howie2012}.  This provides the potential to impute samples in the GWAS catalogue and generate datasets
 of hundreds of thousands of individuals with millions of variants, highlighting the need for incorporating modern
@@
 In the first section we explore some of the properties of random forests as these will also be apparent in the
-\cursedforest implementation. Here we provide demonstrations using synthetic data.  We then demonstrate the scalability
-of \cursedforest in respect to the dimensionality of the data, building a random forest model on whole-genome data from
-the 1000 Genomes Project. We report the error in predicting ethnicity.  Finally, given the role
-different parameter values can play in model construction, we explore the effect tuning these parameters can have on the
-prediction accuracy of the model.
+\cursedforest implementation. Here we provide demonstrations using synthetic data. Secondly, we demonstrate the ability of
+\cursedforest to select previously demonstrated features from a bone mineral density (BMD) dataset. 
+Thirdly, we demonstrate the scalability of \cursedforest in respect to the dimensionality of the data by building a random
+forest model on whole-genome data from the 1000 Genomes Project. Here we report the error in predicting ethnicity.
+Finally, given the role different parameter values can play in model construction, we explore the effect that tuning these
+parameters can have on the prediction accuracy of the model.
 
@@
   \item are insensitive to uninformative predictors;
-  \item are able to capture interactions between features (which is of importance for modelling complex polygenic diseases).
+  \item are able to capture interactions between features (which is important for modelling complex polygenic diseases).
   \end{itemize}
@@
 
-Random forests apply the technique of bootstrap aggregation or bagging to decision trees.  The training data sets are
+Random forests apply the technique of bootstrap aggregation or bagging to decision trees.  The training datasets are
 independently drawn, with replacement, from the joint distribution of $(X,Y)$ so that each set comprises $n$
@@
 \end{itemize}
-GWAs data sets are an obvious application areas for this algorithm.
+GWAS datasets are an obvious application area for this algorithm.
 
@@
 function $y=f(S)$ depending on a set of variables $S$, a random forest will converge to the true function $y$ with a
-rate that depends on the size of $S$ and not the number of noise variables in the data set. If the size of $S$ is small
+rate that depends on the size of $S$ and not the number of noise variables in the dataset. If the size of $S$ is small
 compared to the total number of variables, then the rate of convergence will be quicker. It is not clear what the size
-of the set $S$ will be in a GWAs application. There are cases of small groups of SNPs with large effects and also of
-conditions caused by the interaction of large groups of SNPs.
+of the set $S$ will be in a GWAS application. There are cases of small groups of SNPs with large effects and also of
+conditions caused by the interactions of large groups of SNPs.
 
@@
 essentially a collection of elements. In the case of Spark ML, each element in the RDD is a sample. RDDs contribute to
-the scalability of Spark as they can distributed across multiple nodes and operated on in parallel. Even as we add more
+the scalability of Spark as they can be distributed across multiple nodes and operated on in parallel. Even as we add more
 samples to a dataset, Spark can simply schedule extra tasks to handle the additional items in the RDD.
 
-However, within an RDD, spark ML stores each sample as a vector. Unlike RDDs, which can be partitioned and distributed
+However, within an RDD, Spark ML stores each sample as a vector. Unlike RDDs, which can be partitioned and distributed
 across multiple nodes, each vector must be present in its entirety on any node accessing it. This is no problem with
@@
 The current implementation of \cursedforest uses a ``Gini impurity'' criteria for splitting. Let $f_q$ be the fraction
-of items labelled with value $q$ where $q = 1, \ldots, Q$ at a node. The Gini impurity is
+of items labelled with value $q$, where $q = 1, \ldots, Q$ at a node. The Gini impurity is
 $$
@@
 $$
-which is a minimum when all observations at the node are in the same class.
+which is at a minimum when all observations at the node are in the same class.
 
@@
 and values for each variable are ordinal variables with three levels represented as numbers $\{0, 1, 2\}$ (which
-correspond to an additive effect encoding of genomics variation) randomly generated from uniform distribution with equal
+correspond to an additive effect encoding of genomics variation) randomly generated from a uniform distribution with equal
 probabilities.
 
-The model paramters are $w_i = 1/\sqrt{2^{i-1}}$ for $i =1,\ldots 5$ and we set 
+The model parameters are $w_i = 1/\sqrt{2^{i-1}}$ for $i =1,\ldots 5$ and we set 
 \begin{equation}
@@
 \end{equation}
-We let $\sigma_\epsilon^2 = Var(z)(1-\theta)/\theta $  where $\theta$ is a parameter controling the fraction of variance explained by the infomative variables and in our study we chose $\theta = 0.125$ .
+We let $\sigma_\epsilon^2 = Var(z)(1-\theta)/\theta $  where $\theta$ is a parameter controlling the fraction of variance explained by the informative variables and in our study we chose $\theta = 0.125$ .
 Then
@@
 \subsection{Parameter settings}
-We consider the parameter setting for the random forest algorithm. We use the R notation from the \randomforest
+We consider the parameter settings for the random forest algorithm. We use the R notation from the \randomforest
 package \cite{Liaw.and.Weiner.2002} which incorporates the original Fortran code by Brieman and Cutler.  We incorporate
@@
   minimum size of terminal nodes. It is controlled in Spark ML by setting \texttt{maxDepth}, the maximum depth of each
-  tree in the forest\footnote{The Spark ML documentation \cite{Spark.2016} sets \texttt{maxDepth} to 4 in their
+  tree in the forest.\footnote{The Spark ML documentation \cite{Spark.2016} sets \texttt{maxDepth} to 4 in their
     classification example. This may give an estimate with a higher bias and, as such, may be a poor choice. See the
-    discussion in \cite{Dietterich.2002} where it is shown that bagging may reduce bias as well as variance but gives a
+    discussion in \cite{Dietterich.2002} where it is shown that bagging may reduce bias as well as variance but give a
     better result for low bias learners. \cite{Dietterich.2002} notes that if the bootstrap replicate approximation were
     correct (i.e.~if the bootstrap sample came from an identical distribution to the data), then bagging would reduce
-    variance without changing bias.} In the TCGA example the RandomForests code produced trees with depths of 30+.
+    variance without changing bias.}
 \end{itemize}
@@
 associated set of features amongst millions of dimensions; and its ability to accurately classify the labels of
-high-dimensional samples.  We therefore generate a synthetic dataset with 2.5 million variables ($p=2500000$) of which 5 
-are related to the response variable and 5000 samples ($n=5000$), as discussed in the methods section.
+high-dimensional samples.  We therefore generate a synthetic dataset with 2.5 million variables ($p=2500000$), of which 5 
+are related to the response variable, and 5000 samples ($n=5000$), as discussed in the methods section.
 
 We fit the random forest model and estimate the classification accuracy by capturing the out-of-bag error (OOB) and the
-variable selection performance by capturing the rank-biased-overlap (RBO) \cite{Webber2010} measure. RBO compares two
-lists (in this case, the paramters $w_i, i=1,\ldots , 5$ (equation \ref{equation:synthetic.data}), and the top 1000
+variable selection performance by capturing the rank-biased overlap (RBO) \cite{Webber2010} measure. RBO compares two
+lists (in this case, the parameters $w_i, i=1,\ldots , 5$ (Equation~\ref{equation:synthetic.data}), and the top 1000
 ranked variable importances from the \cursedforest model). It is a mesaure of similarity between rankings that handles
-list of different length, weighs high ranks more heavily than low, and is monotonic with increasing depth of
+lists of different lengths, weighs high ranks more heavily than low, and is monotonic with increasing depth of
 evaluation. We expect to retrieve the 5 features in order of the weight that was assigned to them and with a higher
@@
 
-Increasing \mtry\ in combination with \ntree\ yields best performance with the OOB error essentially constant around 0.4
+Increasing \mtry\ in combination with \ntree\ yields the best performance with the OOB error essentially constant around 0.4
 across a large range of of \mtry and \ntree values.  This is in contrast to the feature-selection performance, where the
@@
 appropriate for large feature datasets (\mtry larger than default). \cursedforest's feature of parallelising trees at
-the node level hence cater perfectly for the requirement of large feature datasets as more trees can be built given a
+the node level hence caters perfectly to the requirement of large feature datasets as more trees can be built given a
 fixed time budget.
@@
 In this section we use \cursedforest\ to identify genomic locations that are associated with a specific trait. To do
-this, we perform feature selection over 7,225,008 genomic variants to identify the locations associated with Bone
+this, we perform feature selection over 7.2 million genomic variants to identify the locations associated with Bone
 Mineral Density (BMD)~\cite{Duncan.2011}. We build a model on the BMD dataset (see datasets) using a binary high or low
@@
 extremely large number of variables. We have treated the case of variable selection in a $p >> n$ model where most of
-the variables are uninformative and have demonstrated the utility of the model for large GWAs data sets. By comparing
-this implementation to other implementations (including those optimized for large data sets) we have demonstrated the
+the variables are uninformative and have demonstrated the utility of the model for large GWAS datasets. By comparing
+this implementation to other implementations (including those optimized for large datasets) we have demonstrated the
 utility of this approach.

### 2017-02-08 doc/publications/plos16/plos16.tex
add BMD table

@@
 {\bf Variant} & {\bf Gene} & {\bf Importance}\\ \thickhline
-x:xxxx & cell2 row 1 & cell3 row 1\\ \hline
-x:xxxx & cell2 row 2 & cell3 row 2\\ \hline
-x:xxxx & cell2 row 3 & cell3 row 3\\ \hline
+8:120003727 & COLEC10 & 1.6393260970E-6\\ \hline
+1:22702231  & ZBTB40  & 1.3901355155E-6\\ \hline
+22:18923745 & PRODH   & 1.3769982791E-6\\ \hline
+1:22702187  & ZBTB40  & 1.3350582926E-6\\ \hline
+3:41111035  & CTNNB1  & 1.2707227479E-6\\ \hline
+1:693731    &         & 1.1786957578E-6\\ \hline
+1:22700351  & ZBTB40  & 1.1547011076E-6\\ \hline
+8:120040181 & COLEC10 & 1.1511986773E-6\\ \hline
+1:22701761  & ZBTB40  & 1.1495539020E-6\\ \hline
+3:41120417  & CTNNB1  & 1.1325242379E-6\\ \hline
 \end{tabular}
 \begin{flushleft} 
-Here the importance is the Gini impurity measure when building a model using the cell type as the label.
+The ten genes proximal to the top variants as ranked by their Gini impurity from a random forest model built on
+individuals using high/low bone mineral density as the label.
 \end{flushleft}

### 2017-02-08 doc/publications/plos16/plos16.tex
fix my ref

@@
 
-In a previous paper, we clustered individuals from the 1000 Genomes Project using k-means clustering \cite{Obrien2016}.
-\marginpar{reference is missing}
+In a previous paper, we clustered individuals from the 1000 Genomes Project using k-means clustering \cite{OBrien2015}.
 Although we were able to cluster individuals on all 80 million variants, it was relatively slow, with the entire

### 2017-02-09 doc/publications/plos16/plos16.tex
Move figure to the left
fix typos and number in BMD section

@@
 utilized to ``fill in'' or impute the information at previously unobserved genomic positions in older array
-technology~\cite{Howie2012}.  This provides the potential to impute samples in the GWAS catalogue and generate datasets
+technology~\cite{Howie2012}.  This provides the potential to impute samples in the GWAS catalog and generate datasets
 of hundreds of thousands of individuals with millions of variants, highlighting the need for incorporating modern
@@
 \cursedforest to select previously demonstrated features from a bone mineral density (BMD) dataset. 
-Thirdly, we demonstrate the scalability of \cursedforest in respect to the dimensionality of the data by building a random
+Thirdly, we demonstrate the scalability of \cursedforest in respect to the dimensionality of data by building a random
 forest model on whole-genome data from the 1000 Genomes Project. Here we report the error in predicting ethnicity.
@@
   \item are insensitive to uninformative predictors;
-  \item are able to capture interactions between features (which is important for modelling complex polygenic diseases).
+  \item are able to capture interactions between features (which is important for modeling complex polygenic diseases).
   \end{itemize}
@@
 The current implementation of \cursedforest uses a ``Gini impurity'' criteria for splitting. Let $f_q$ be the fraction
-of items labelled with value $q$, where $q = 1, \ldots, Q$ at a node. The Gini impurity is
+of items labeled with value $q$, where $q = 1, \ldots, Q$ at a node. The Gini impurity is
 $$
@@
 and values for each variable are ordinal variables with three levels represented as numbers $\{0, 1, 2\}$ (which
-correspond to an additive effect encoding of genomics variation) randomly generated from a uniform distribution with equal
+correspond to an additive effect encoding of genomic variation) randomly generated from a uniform distribution with equal
 probabilities.
@@
 the number of trees on the other hand improves performance especially when the trees are kept diverse (small \mtry) but
-appropriate for large feature datasets (\mtry larger than default). \cursedforest's feature of parallelising trees at
+appropriate for large feature datasets (\mtry larger than default). \cursedforest's feature of parallelizing trees at
 the node level hence caters perfectly to the requirement of large feature datasets as more trees can be built given a
@@
 
-Investigating the lowly-ranked genes, we observe DCDC5 (Doublecortin Domain Containing 5) ranked 9,667 out of 10,000. This gene is 
+Investigating the lowly-ranked genes, we observe DCDC5 (Doublecortin Domain Containing 5) ranked 9,830 out of 10,000. This gene is 
 one of the BMD-associated genes postulated by literature that Duncan {\it et al.}~\cite{Duncan.2011} inspect.
@@
 \begin{figure}[tbhp]
+  \begin{adjustwidth}{-1.00in}{0in}
     \caption{\textbf{Runtime comparison on the 1000 genomes project}}
@@
       The figure shows the runtime for the 5 subsets of the 1000 genomes project and the different methods investigated.
-$\bigotimes$ marks the last dataset the different method completed before running out of memory or running more than 24 hours.
-The dashed lines mark the trajectory for how ranger and Spark ML had performed given larger memory machines.
+      $\bigotimes$ marks the last dataset the different method completed before running out of memory or running more than
+      24 hours. The dashed lines mark the trajectory for how ranger and Spark ML had performed given larger memory machines.
     \end{flushleft}
+  \end{adjustwidth}
 \end{figure}
@@
 distinct boundary, shown empirically and also based on arguments from combinatorial geometry, to the region where we can
-reliably recover significant variables.  \cite{Donoho.and.Stodden.2006} investigate the behaviour of a number of
+reliably recover significant variables.  \cite{Donoho.and.Stodden.2006} investigate the behavior of a number of
 regression approaches for variable selection (LARS, Lasso and forward stepwise) and make the point that above the

### 2017-02-09 doc/publications/plos16/plos16.tex
Reword and explain new genes

@@
 
-To test whether the associated features have biological meaning, we investigate whether the known BMD genes investigated
-by Duncan {\it et al.}~\cite{Duncan.2011} are enriched amongst the highly ranked genes. We therefore label all variants
-that fall within 100 kb of these 32 BMD genes and observe that they are ranked significantly higher than other locations
-in the genome (Mann-Whitney U, $p=1.3\times10^{-07}$), supporting our hypothesis that biologically relevant features are
-contributing more information to our classification model.
+To test whether the associated features have biological meaning, we investigate whether variants in or near
+BMD-associated genes are enriched are enriched toward the top of our list of our ranked variant list. To begin with,
+we start with 26 known BMD genes~\cite{Duncan.2011}. We also include the 6 genes that Duncan {\it et al.} find with
+a new BMD association. We then add to this list the genes that Estrada {\it et al.} found with a new BMD association.
+In total, this gives us a list of 100 genes.
+
+We label our contributing variants that fall within 100 kb of these 100 genes and observe that they are ranked
+significantly higher than variants from other locations in the genome (Mann-Whitney U, $p=1.3\times10^{-07}$),
+supporting our hypothesis that biologically relevant features are contributing more information to our classification model.
 

### 2017-02-09 doc/publications/plos16/plos16.tex
Few fix ups of previous commit

--- 2017-02-09 doc/publications/plos16/plos16.tex
@@
 In this section we use \cursedforest\ to identify genomic locations that are associated with a specific trait. To do
-this, we perform feature selection over 7.2 million genomic variants to identify the locations associated with Bone
+this, we perform feature selection on over 7.2 million genomic variants to identify the locations associated with Bone
 Mineral Density (BMD)~\cite{Duncan.2011}. We build a model on the BMD dataset (see datasets) using a binary high or low
@@
 
-To test whether the associated features have biological meaning, we investigate whether the known BMD genes investigated
-by Duncan {\it et al.}~\cite{Duncan.2011} are enriched amongst the highly ranked genes. We therefore label all variants
-that fall within 100 kb of these 32 BMD genes and observe that they are ranked significantly higher than other locations
-in the genome (Mann-Whitney U, $p=1.3\times10^{-07}$), supporting our hypothesis that biologically relevant features are
-contributing more information to our classification model.
+To test whether the associated features have biological meaning, we investigate whether variants in or near
+BMD-associated genes are enriched toward the top of our ranked variant list. To form this list of "known genes", we start
+with the 26 known BMD genes listed by Duncan {\it et al.}~\cite{Duncan.2011}. We also include the 6 genes that they find
+with a new BMD association. We then add to this list the genes that Estrada {\it et al.} found with a new BMD association.
+In total, this gives us a list of 100 genes.
+
+We label variants from our ranked list that fall within 100 kb of these 100 genes and observe that they are ranked
+significantly higher than variants from other locations in the genome (Mann-Whitney U, $p=1.3\times10^{-07}$),
+supporting our hypothesis that biologically relevant features are contributing more information to our classification model.
 

### 2017-02-09 doc/publications/plos16/plos16.tex
Few more changes to second para of that section

@@
 with the 26 known BMD genes listed by Duncan {\it et al.}~\cite{Duncan.2011}. We also include the 6 genes that they find
-with a new BMD association. We then add to this list the genes that Estrada {\it et al.} found with a new BMD association.
-In total, this gives us a list of 100 genes.
+with a new BMD association. We then add to this list the genes that Estrada {\it et al.} found with a new BMD
+association~\cite{Estrada2012}. In total, this gives us a list of 100 different genes.
 
@@
 
-We therefore investigate the highest-ranked features not amongst the associated gene list by Duncan
-{\it et al.}~\cite{Duncan.2011}. COLEC10 (collectin 10) contains variants more recently associated with
+We therefore investigate the highest-ranked features that are not located near genes from our known-gene list. One gene,
+COLEC10	 (collectin 10) contains variants more recently associated with
 BDM~\cite{Kemp2014,Liu2008}, with a verified transcription factor binding site for the known BMD gene FOXL1~\cite{DECODE}.

### 2017-02-09 doc/publications/plos16/plos16.tex
BMD section added notes

--- 2017-02-09 doc/publications/plos16/plos16.tex
@@
 utilized to ``fill in'' or impute the information at previously unobserved genomic positions in older array
-technology~\cite{Howie2012}.  This provides the potential to impute samples in the GWAS catalog and generate datasets
+technology~\cite{Howie2012}.  This provides the potential to impute samples in the GWAS catalogue and generate datasets
 of hundreds of thousands of individuals with millions of variants, highlighting the need for incorporating modern
@@
 \cursedforest to select previously demonstrated features from a bone mineral density (BMD) dataset. 
-Thirdly, we demonstrate the scalability of \cursedforest in respect to the dimensionality of data by building a random
+Thirdly, we demonstrate the scalability of \cursedforest in respect to the dimensionality of the data by building a random
 forest model on whole-genome data from the 1000 Genomes Project. Here we report the error in predicting ethnicity.
@@
   \item are insensitive to uninformative predictors;
-  \item are able to capture interactions between features (which is important for modeling complex polygenic diseases).
+  \item are able to capture interactions between features (which is important for modelling complex polygenic diseases).
   \end{itemize}
@@
 The current implementation of \cursedforest uses a ``Gini impurity'' criteria for splitting. Let $f_q$ be the fraction
-of items labeled with value $q$, where $q = 1, \ldots, Q$ at a node. The Gini impurity is
+of items labelled with value $q$, where $q = 1, \ldots, Q$ at a node. The Gini impurity is
 $$
@@
 and values for each variable are ordinal variables with three levels represented as numbers $\{0, 1, 2\}$ (which
-correspond to an additive effect encoding of genomic variation) randomly generated from a uniform distribution with equal
+correspond to an additive effect encoding of genomics variation) randomly generated from a uniform distribution with equal
 probabilities.
@@
 the number of trees on the other hand improves performance especially when the trees are kept diverse (small \mtry) but
-appropriate for large feature datasets (\mtry larger than default). \cursedforest's feature of parallelizing trees at
+appropriate for large feature datasets (\mtry larger than default). \cursedforest's feature of parallelising trees at
 the node level hence caters perfectly to the requirement of large feature datasets as more trees can be built given a
@@
 
-To test whether the associated features have biological meaning, we investigate whether variants in or near
-BMD-associated genes are enriched are enriched toward the top of our list of our ranked variant list. To begin with,
-we start with 26 known BMD genes~\cite{Duncan.2011}. We also include the 6 genes that Duncan {\it et al.} find with
-a new BMD association. We then add to this list the genes that Estrada {\it et al.} found with a new BMD association.
-In total, this gives us a list of 100 genes.
-
-We label our contributing variants that fall within 100 kb of these 100 genes and observe that they are ranked
-significantly higher than variants from other locations in the genome (Mann-Whitney U, $p=1.3\times10^{-07}$),
-supporting our hypothesis that biologically relevant features are contributing more information to our classification model.
+To test whether the associated features have biological meaning, we investigate whether the known BMD genes investigated
+by Duncan {\it et al.}~\cite{Duncan.2011} are enriched amongst the highly ranked genes. We therefore label all variants
+that fall within 100 kb of these 32 BMD genes and observe that they are ranked significantly higher than other locations
+in the genome (Mann-Whitney U, $p=1.3\times10^{-07}$), supporting our hypothesis that biologically relevant features are
+contributing more information to our classification model.
 
@@
 
-Investigating the lowly-ranked genes, we observe DCDC5 (Doublecortin Domain Containing 5) ranked 9,830 out of 10,000. This gene is 
+Investigating the lowly-ranked genes, we observe DCDC5 (Doublecortin Domain Containing 5) ranked 9,667 out of 10,000. This gene is 
 one of the BMD-associated genes postulated by literature that Duncan {\it et al.}~\cite{Duncan.2011} inspect.
@@
 
+%1:693731 rs12238997
+% rs1035665 A genome-wide association study of global gene expression http://www.nature.com/ng/journal/v39/n10/full/ng2109.html rs1035665 as master regulator of ATP Binding Cassette Subfamily F Member 2 ABCF1  with ABCF2 634 known
+% rs6702460 1	91536	91536	14	1.0296212827525244E-6
+% rs2058553 2	20573002	20573002	15	9.896033178354981E-7
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 distinct boundary, shown empirically and also based on arguments from combinatorial geometry, to the region where we can
-reliably recover significant variables.  \cite{Donoho.and.Stodden.2006} investigate the behavior of a number of
+reliably recover significant variables.  \cite{Donoho.and.Stodden.2006} investigate the behaviour of a number of
 regression approaches for variable selection (LARS, Lasso and forward stepwise) and make the point that above the

### 2017-02-10 doc/publications/plos16/plos16.tex
updated result section 2

--- 2017-02-09 doc/publications/plos16/plos16.tex
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{\cursedforest\ selects previously demonstrated bone-density variants}
-In this section we use \cursedforest\ to identify genomic locations that are associated with a specific trait. To do
-this, we perform feature selection on over 7.2 million genomic variants to identify the locations associated with Bone
-Mineral Density (BMD)~\cite{Duncan.2011}. We build a model on the BMD dataset (see datasets) using a binary high or low
-BMD value as the label.  From this model, we output a list of the top 10,000 contributing variants, ranked by their Gini
-impurity. The variant with the highest Gini impurity, and hence importance, is ranked 1, and the variant with the lowest
-Gini impurity is ranked 10,000.
+\subsection{GWAS-style analysis using \cursedforest\ identifies novel variants associated with Bone Mineral Density}
+In this section we use \cursedforest\ to perform feature selection on over 7.2 million genomic variants to identify the locations associated with Bone
+Mineral Density (BMD)~\cite{Duncan.2011}. We build a model on the BMD dataset as described in Methods using a binary high or low
+BMD value as the label.  From this model, we output the top 10,000 contributing variants, ranked by their Gini
+impurity. The variant with the highest Gini impurity, and hence importance, is ranked 1, the lowest 10,000.
 
 To test whether the associated features have biological meaning, we investigate whether variants in or near
-BMD-associated genes are enriched toward the top of our ranked variant list. To form this list of "known genes", we start
-with the 26 known BMD genes listed by Duncan {\it et al.}~\cite{Duncan.2011}. We also include the 6 genes that they find
-with a new BMD association. We then add to this list the genes that Estrada {\it et al.} found with a new BMD association.
-In total, this gives us a list of 100 genes.
-
-We label variants from our ranked list that fall within 100 kb of these 100 genes and observe that they are ranked
-significantly higher than variants from other locations in the genome (Mann-Whitney U, $p=1.3\times10^{-07}$),
-supporting our hypothesis that biologically relevant features are contributing more information to our classification model.
-
-We therefore investigate the highest-ranked features not amongst the associated gene list by Duncan
-{\it et al.}~\cite{Duncan.2011}. COLEC10 (collectin 10) contains variants more recently associated with
-BDM~\cite{Kemp2014,Liu2008}, with a verified transcription factor binding site for the known BMD gene FOXL1~\cite{DECODE}.
-%However, we also observe that from the set of highest-ranked features, for example, the top ten, not all of them exist
-%within close proximity to known BMD genes. We investigate the literature to determine the plausibility of these variants
-%contributing to BMD.
-%One gene is COLEC10 (collectin 10). Featuring a transcription factor binding site for FOXL1, COLEC10 is also reported as
-%containing variants associated with BMD \cite{Kemp2014,Liu2008}. However there currently exists no information on how this gene
-%may contribute to BMD. 
-Also highly-ranked are variants in chromosome 22q11. Stagi {\it et al.}~\cite{Stagi.2010}, observe that individuals with a
-microdeletion in this region have a significant reduction in bone mass. In this same region, we also observe
-variants present in PRODH (proline dehydrogenase). PRODH catalyzes the first step in proline degradation, and as
+BMD-associated genes are enriched toward the top of our ranked variant list.
+To form the list of "known genes", we combine the 26 known and 6 novel BMD genes listed by Duncan {\it et al.}~\cite{Duncan.2011} with genes identified by Estrada {\it et al.}~\cite{Estrada2012, Kemp2014,Phang.2008}, resulting in 102 genes.
+All genes are covered in the 10,000 extracted variants and the gene name is assigned to a variant if it falls within 100 kb of the gene.
+
+As shown in Figure~\ref{figure:ranksumtest}, the known genes are ranked significantly higher than variants from other locations in the genome (Mann-Whitney U, $p=1.3\times10^{-07}$), supporting our hypothesis that biologically relevant features are contributing more information to our classification model.
+
+The gene with the highest-ranked features is COLEC10 (collectin 10) as shown in Figure~\ref{figure:knowngeneranks}, a well established contributor to BDM~\cite{Kemp2014,Liu2008} with verified transcription factor binding site for FOXL1, also a known BDM gene~\cite{DECODE}.
+The second highest ranked gene is ZBTB40 on 22q11. Stagi {\it et al.}~\cite{Stagi.2010}, observe that individuals with a
+micro-deletion in this region have a significant reduction in bone mass. In this same region, we also observe the next highest ranked gene, PRODH (proline dehydrogenase). 
+PRODH catalyses the first step in proline degradation, and as
 proposed by Phang {\it et al.}~\cite{Phang.2008}, proline is made available by collagen degradation.
 
-Investigating the lowly-ranked genes, we observe DCDC5 (Doublecortin Domain Containing 5) ranked 9,830 out of 10,000. This gene is 
-one of the BMD-associated genes postulated by literature that Duncan {\it et al.}~\cite{Duncan.2011} inspect.
-Similar to their findings, our results also indicate that this gene does not have a strong association with BMD. 
+The lowest ranked gene with literature implied BMD association is DCDC5 (Doublecortin Domain Containing 5) ranked 9,830 out of 10,000 (see Figure~\ref{figure:knowngeneranks}). 
+Similar to the conclusions drawn by Duncan {\it et al.}~\cite{Duncan.2011}, our results also indicate that this gene does not have a strong association with BMD. 
 In fact, DCDC5 is not highly expressed in bone, and any mechanisms by which this gene might regulate BMD are unclear \cite{Thakker2012}. 
@@
 
-In Table~\ref{featuretable}, we display the top 10 features, ranked by their Gini impurity measure in the random forest model.
-
-\begin{table}[!ht]
-%\begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
-\centering
-\caption{
-{\bf Features from the TCGA BRCA dataset ranked by their importance.}}
-\begin{tabular}{|l|l|l|l|l|l|l|}
-\hline
-{\bf Variant} & {\bf Gene} & {\bf Importance}\\ \thickhline
-8:120003727 & COLEC10 & 1.6393260970E-6\\ \hline
-1:22702231  & ZBTB40  & 1.3901355155E-6\\ \hline
-22:18923745 & PRODH   & 1.3769982791E-6\\ \hline
-1:22702187  & ZBTB40  & 1.3350582926E-6\\ \hline
-3:41111035  & CTNNB1  & 1.2707227479E-6\\ \hline
-1:693731    &         & 1.1786957578E-6\\ \hline
-1:22700351  & ZBTB40  & 1.1547011076E-6\\ \hline
-8:120040181 & COLEC10 & 1.1511986773E-6\\ \hline
-1:22701761  & ZBTB40  & 1.1495539020E-6\\ \hline
-3:41120417  & CTNNB1  & 1.1325242379E-6\\ \hline
-\end{tabular}
+Hence focusing on the top 20 most associated variants, the majority are from five known BMD genes (left side of the vertical line in Figure~\ref{figure:knowngeneranks}), however there are also five novel locations as annotated in Figure~\ref{figure:ranksumtest}.
+Most notably the SNP in the intronic region of Unc13c (rs12916554) is linked to quality of sleep~\cite{Nicod2016}, which in turn is reported to be the best predictor of osteoporosis risk~\cite{Magali2016}, a condition diagnosed by measuring BMD.
+Also listed is rs1035665, which was identified in a recent GWAS study to regulate ABCF1 (ATP Binding Cassette Subfamily F Member 1)~\cite{Dixon2007} with member 2 of this subfamily being a known BMD gene.
+The other three novel loci are on or near ncRNA.
+
+\cursedforest\ can successfully be used in GWAS-type analysis replicating previous findings and uncovering novel associations due to the ability of random forests to pick up interacting features.
+
+%In Table~\ref{featuretable}, we display the top 10 features, ranked by their Gini impurity measure in the random forest model.
+
+%\begin{table}[!ht]
+%%\begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
+%\centering
+%\caption{
+%{\bf Features from the TCGA BRCA dataset ranked by their importance.}}
+%\begin{tabular}{|l|l|l|l|l|l|l|}
+%\hline
+%{\bf Variant} & {\bf Gene} & {\bf Importance}\\ \thickhline
+%8:120003727 & COLEC10 & 1.6393260970E-6\\ \hline
+%1:22702231  & ZBTB40  & 1.3901355155E-6\\ \hline
+%22:18923745 & PRODH   & 1.3769982791E-6\\ \hline
+%1:22702187  & ZBTB40  & 1.3350582926E-6\\ \hline
+%3:41111035  & CTNNB1  & 1.2707227479E-6\\ \hline
+%1:693731    &         & 1.1786957578E-6\\ \hline  
+%1:22700351  & ZBTB40  & 1.1547011076E-6\\ \hline
+%8:120040181 & COLEC10 & 1.1511986773E-6\\ \hline
+%1:22701761  & ZBTB40  & 1.1495539020E-6\\ \hline
+%3:41120417  & CTNNB1  & 1.1325242379E-6\\ \hline
+%\end{tabular}
+%\begin{flushleft} 
+%The ten genes proximal to the top variants as ranked by their Gini impurity from a random forest model built on
+%individuals using high/low bone mineral density as the label.
+%\end{flushleft}
+%\label{featuretable}
+%%\end{adjustwidth}
+%\end{table}
+
+\begin{figure}[tbhp]
+  \begin{adjustwidth}{-1.00in}{0in}
+    \caption{\textbf{Distribution of known BMD genes amongst the top 10,000 features}}
+    \label{figure:ranksumtest}
+    \includegraphics[totalheight=8cm]{./figs/BMDTop10K.pdf}
     \begin{flushleft}
-The ten genes proximal to the top variants as ranked by their Gini impurity from a random forest model built on
-individuals using high/low bone mineral density as the label.
+      The figure shows the top 10,000 features by their rank and Gini impurity score (Importance).
+The line denotes the top 20 features which contains five novel loci annotated with text.
+The $p$-value is the Mann-Whitney U test.
      \end{flushleft}
-\label{featuretable}
-%\end{adjustwidth}
-\end{table}
+  \end{adjustwidth}
+\end{figure}
+
+\begin{figure}[tbhp]
+  \begin{adjustwidth}{-2.00in}{0in}
+    \caption{\textbf{Rank of literature assigned BMD genes}}
+    \label{figure:knowngeneranks}
+    \includegraphics[totalheight=6.5cm]{./figs/BMDgenes_landscape.pdf}
+    \begin{flushleft}
+      The figure shows the rank of the 100 literature assigned BMD genes.
+Each ball represents a SNP assigned to the closest gene with colour representing the distance (max 100 kb).
+The size shows the Gini impurity score with larger balls representing more important SNPs.
+The horizontal line separates the genes with SNPs amongst the Top 20 associated features.
+     \end{flushleft}
+  \end{adjustwidth}
+\end{figure}
+
+%1:693731 rs12238997 lncRNA
+% rs1035665 A genome-wide association study of global gene expression http://www.nature.com/ng/journal/v39/n10/full/ng2109.html rs1035665 as master regulator of ATP Binding Cassette Subfamily F Member 2 ABCF1  with ABCF2 634 known
+% rs6702460 1	91536	91536	14	1.0296212827525244E-6 ncRNA
+% rs2058553 2	20573002	20573002	15	9.896033178354981E-7 ncRNA
+% rs12916554 15	 54607255	54607255	17	9.687391928109767E-7  Unc13c quality of sleep http://www.nature.com/ng/journal/v48/n8/full/ng.3595.html   https://www.ncbi.nlm.nih.gov/pubmed/27655463
 
@@
 \subsection{\cursedforest outperforms existing methods for classification problems}
-In this section we compare \cursedforest's performance against other published methods for predicting ethnicity on the
+In this section we demonstrate \cursedforest's classification accuracy on biological data and compare its performance against other published methods by predicting ethnicity on the
 1000 Genomes Project dataset.
 
-In a previous paper, we clustered individuals from the 1000 Genomes Project using k-means clustering \cite{OBrien2015}.
-Although we were able to cluster individuals on all 80 million variants, it was relatively slow, with the entire
-process (pre-processing and clustering) taking over 30 hours.
-We investigate building random forest models on this data. However, we observed this process to not only be slower
-than building a k-means model, but does not scale to as many dimensions (see Table~\ref{timingtable}). To keep
-the following comparisons fair, we limit the number of executors available to Spark. On the relatively low-dimensional
+%In a previous paper, we clustered individuals from the 1000 Genomes Project using k-means clustering \cite{OBrien2015}.
+%Although we were able to cluster individuals on all 80 million variants, it was relatively slow, with the entire
+%process (pre-processing and clustering) taking over 30 hours.
+%We investigate building random forest models on this data. However, we observed this process to not only be slower
+%than building a k-means model, but does not scale to as many dimensions (see Table~\ref{timingtable}). 
+To keep
+the following comparisons fair, we limit the number of executors available to Spark.
+
+On the relatively low-dimensional
 phase1\_chr22 dataset (1,092 x 490,036), we observed a runtime of under 8 minutes. Although slower than Ranger, 

### 2017-02-11 doc/publications/plos16/plos16.tex
upgraded to new version of plot template

@@
 % Template for PLoS
-% Version 3.3 June 2016
+% Version 3.4 January 2017
 %
@@
 %
-% For tables that exceed the width of the text column, use the adjustwidth environment as illustrated in the example
-% table in text below.
+% For tables that exceed the width of the text column, use the adjustwidth environment as illustrated in the example table in text below.
 %
@@
 % IMPORTANT
-%
-% Below are a few tips to help format your equations and other special characters according to our specifications. For
-% more tips to help reduce the possibility of formatting errors during conversion, please see our LaTeX guidelines at
-% http://journals.plos.org/plosone/s/latex
-%
-% For inline equations, please be sure to include all portions of an equation in the math environment.  For example,
-% x$^2$ is incorrect; this should be formatted as $x^2$ (or $\mathrm{x}^2$ if the romanized font is desired).
-%
-% Do not include text that is not math in the math environment. For example, CO2 should be written as
-% CO\textsubscript{2} instead of CO$_2$.
+% Below are a few tips to help format your equations and other special characters according to our specifications.
+For more tips to help reduce the possibility of formatting errors during conversion, please see our LaTeX guidelines at http://journals.plos.org/plosone/s/latex
+%
+% For inline equations, please be sure to include all portions of an equation in the math environment.
+For example, x$^2$ is incorrect; this should be formatted as $x^2$ (or $\mathrm{x}^2$ if the romanized font is desired).
+%
+% Do not include text that is not math in the math environment. For example, CO2 should be written as CO\textsubscript{2} instead of CO$_2$.
 %
@@
 %
-% For inline equations, please do not include punctuation (commas, etc) within the math environment unless this is part
-% of the equation.
-%
-% When adding superscript or subscripts outside of brackets/braces, please group using {}.  For example, change
-% "[U(D,E,\gamma)]^2" to "{[U(D,E,\gamma)]}^2".
+% For inline equations, please do not include punctuation (commas, etc) within the math environment unless this is part of the equation.
+%
+% When adding superscript or subscripts outside of brackets/braces, please group using {}.  For example, change "[U(D,E,\gamma)]^2" to "{[U(D,E,\gamma)]}^2". 
 %
@@
 \documentclass[10pt,letterpaper]{article}
-\usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}
+\usepackage[top=0.85in,left=2.75in,footskip=0.75in,showframe]{geometry}
 
@@
 
+\newcommand{\lorem}{{\bf LOREM}}
+\newcommand{\ipsum}{{\bf IPSUM}}
+
+%% END MACROS SECTION
+
+%% Additional packages and local edits
+\usepackage{footmisc}
+\usepackage{multirow}
+\usepackage[T1]{fontenc}
+%% required the installation of cm-super 55MB - TeX font package (full version) with CM (EC) in Type1 in T1, T2*, TS1, X2 enc
+
+\let\oldmarginpar\marginpar
+\renewcommand\marginpar[1]{\-\oldmarginpar[\raggedleft\footnotesize #1]%hv
+{\raggedright\footnotesize #1}}
+\reversemarginpar
+
+\setlength{\marginparwidth}{5cm}
+\usepackage{caption}
+\usepackage{subcaption}
+\usepackage{placeins}
+\usepackage{xspace} % see http:
+%\usepackage{showframe}
+
 \newcommand{\cursedforest}{\textsc{CursedForest}\xspace}
@@
 \newcommand{\ntree}{\texttt{ntree}\xspace}
-
-%% END MACROS SECTION
-
-%% Additional packages
-\usepackage{footmisc}
-\usepackage{multirow}
-\usepackage[T1]{fontenc}
-%% required the installation of cm-super 55MB - TeX font package (full version) with CM (EC) in Type1 in T1, T2*, TS1, X2 enc
-
-%%local edits (rob)
-\let\oldmarginpar\marginpar
-\renewcommand\marginpar[1]{\-\oldmarginpar[\raggedleft\footnotesize #1]%hv
-{\raggedright\footnotesize #1}}
-\reversemarginpar
-\setlength{\marginparwidth}{5cm}
-\usepackage{caption}
-\usepackage{subcaption}
-\usepackage{placeins}
-\usepackage{xspace} % see http://tex.stackexchange.com/questions/31091/space-after-latex-commands
 
@@
 \subsection{GWAS-style analysis using \cursedforest\ identifies novel variants associated with Bone Mineral Density}
-In this section we use \cursedforest\ to perform feature selection on over 7.2 million genomic variants to identify the locations associated with Bone
-Mineral Density (BMD)~\cite{Duncan.2011}. We build a model on the BMD dataset as described in Methods using a binary high or low
-BMD value as the label.  From this model, we output the top 10,000 contributing variants, ranked by their Gini
-impurity. The variant with the highest Gini impurity, and hence importance, is ranked 1, the lowest 10,000.
+In this section we use \cursedforest\ to perform feature selection on over 7.2 million genomic variants to identify the
+locations associated with Bone Mineral Density (BMD)~\cite{Duncan.2011}. We build a model on the BMD dataset as
+described in Methods using a binary high or low BMD value as the label.  From this model, we output the top 10,000
+contributing variants, ranked by their Gini impurity. The variant with the highest Gini impurity, and hence importance,
+is ranked 1, the lowest 10,000.
 
 To test whether the associated features have biological meaning, we investigate whether variants in or near
-BMD-associated genes are enriched toward the top of our ranked variant list.
-To form the list of "known genes", we combine the 26 known and 6 novel BMD genes listed by Duncan {\it et al.}~\cite{Duncan.2011} with genes identified by Estrada {\it et al.}~\cite{Estrada2012, Kemp2014,Phang.2008}, resulting in 102 genes.
-All genes are covered in the 10,000 extracted variants and the gene name is assigned to a variant if it falls within 100 kb of the gene.
-
-As shown in Figure~\ref{figure:ranksumtest}, the known genes are ranked significantly higher than variants from other locations in the genome (Mann-Whitney U, $p=1.3\times10^{-07}$), supporting our hypothesis that biologically relevant features are contributing more information to our classification model.
-
-The gene with the highest-ranked features is COLEC10 (collectin 10) as shown in Figure~\ref{figure:knowngeneranks}, a well established contributor to BDM~\cite{Kemp2014,Liu2008} with verified transcription factor binding site for FOXL1, also a known BDM gene~\cite{DECODE}.
-The second highest ranked gene is ZBTB40 on 22q11. Stagi {\it et al.}~\cite{Stagi.2010}, observe that individuals with a
-micro-deletion in this region have a significant reduction in bone mass. In this same region, we also observe the next highest ranked gene, PRODH (proline dehydrogenase). 
-PRODH catalyses the first step in proline degradation, and as
-proposed by Phang {\it et al.}~\cite{Phang.2008}, proline is made available by collagen degradation.
-
-The lowest ranked gene with literature implied BMD association is DCDC5 (Doublecortin Domain Containing 5) ranked 9,830 out of 10,000 (see Figure~\ref{figure:knowngeneranks}). 
-Similar to the conclusions drawn by Duncan {\it et al.}~\cite{Duncan.2011}, our results also indicate that this gene does not have a strong association with BMD. 
-In fact, DCDC5 is not highly expressed in bone, and any mechanisms by which this gene might regulate BMD are unclear \cite{Thakker2012}. 
-Although variants near DCDC5 are associated with BMD \cite{Rivadeneira2009}, it is questionable as to whether the gene itself contributes to BMD.
-
-Hence focusing on the top 20 most associated variants, the majority are from five known BMD genes (left side of the vertical line in Figure~\ref{figure:knowngeneranks}), however there are also five novel locations as annotated in Figure~\ref{figure:ranksumtest}.
-Most notably the SNP in the intronic region of Unc13c (rs12916554) is linked to quality of sleep~\cite{Nicod2016}, which in turn is reported to be the best predictor of osteoporosis risk~\cite{Magali2016}, a condition diagnosed by measuring BMD.
-Also listed is rs1035665, which was identified in a recent GWAS study to regulate ABCF1 (ATP Binding Cassette Subfamily F Member 1)~\cite{Dixon2007} with member 2 of this subfamily being a known BMD gene.
-The other three novel loci are on or near ncRNA.
+BMD-associated genes are enriched toward the top of our ranked variant list. To form the list of "known genes", we
+combine the 26 known and 6 novel BMD genes listed by Duncan {\it et al.}~\cite{Duncan.2011} with genes identified by
+Estrada {\it et al.}~\cite{Estrada2012, Kemp2014,Phang.2008}, resulting in 102 genes.  All genes are covered in the
+10,000 extracted variants and the gene name is assigned to a variant if it falls within 100 kb of the gene.
+
+As shown in Fig~\ref{figure:ranksumtest}, the known genes are ranked significantly higher than variants from other
+locations in the genome (Mann-Whitney U, $p=1.3\times10^{-07}$), supporting our hypothesis that biologically relevant
+features are contributing more information to our classification model.
+
+The gene with the highest-ranked features is COLEC10 (collectin 10) as shown in Fig~\ref{figure:knowngeneranks}, a
+well established contributor to BDM~\cite{Kemp2014,Liu2008} with verified transcription factor binding site for FOXL1,
+also a known BDM gene~\cite{DECODE}.  The second highest ranked gene is ZBTB40 on 22q11. Stagi {\it et
+  al.}~\cite{Stagi.2010}, observe that individuals with a micro-deletion in this region have a significant reduction in
+bone mass. In this same region, we also observe the next highest ranked gene, PRODH (proline dehydrogenase).  PRODH
+catalyses the first step in proline degradation, and as proposed by Phang {\it et al.}~\cite{Phang.2008}, proline is
+made available by collagen degradation.
+
+The lowest ranked gene with literature implied BMD association is DCDC5 (Doublecortin Domain Containing 5) ranked 9,830
+out of 10,000 (see Fig~\ref{figure:knowngeneranks}).  Similar to the conclusions drawn by Duncan {\it et
+  al.}~\cite{Duncan.2011}, our results also indicate that this gene does not have a strong association with BMD.  In
+fact, DCDC5 is not highly expressed in bone, and any mechanisms by which this gene might regulate BMD are unclear
+\cite{Thakker2012}.  Although variants near DCDC5 are associated with BMD \cite{Rivadeneira2009}, it is questionable as
+to whether the gene itself contributes to BMD.
+
+Hence focusing on the top 20 most associated variants, the majority are from five known BMD genes (left side of the
+vertical line in Fig~\ref{figure:knowngeneranks}), however there are also five novel locations as annotated in
+Fig~\ref{figure:ranksumtest}. Most notably the SNP in the intronic region of Unc13c (rs12916554) is linked to quality
+of sleep~\cite{Nicod2016}, which in turn is reported to be the best predictor of osteoporosis risk~\cite{Magali2016}, a
+condition diagnosed by measuring BMD.  Also listed is rs1035665, which was identified in a recent GWAS study to regulate
+ABCF1 (ATP Binding Cassette Subfamily F Member 1)~\cite{Dixon2007} with member 2 of this subfamily being a known BMD
+gene. The other three novel loci are on or near ncRNA.
 

### 2017-02-13 doc/publications/plos16/plos16.tex
changed figures to higher reolution
first pass through all result sections and abstract

@@
 \documentclass[10pt,letterpaper]{article}
-\usepackage[top=0.85in,left=2.75in,footskip=0.75in,showframe]{geometry}
+\usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}
 
@@
 \section{Abstract}
-We introduce \cursedforest, an implementation of random forests designed to handle data that may have an extremely large
-number of variables. We discuss the utility of random forests for machine learning problems. We show on simulated and
-real data that the novel implementation can achieve reasonable prediction accuracy and variable importance estimates
-even in setting where other implementations, including some designed for large data, fail.
+
+Analyses are performed on ever larger datasets with unprecedentedly detailed information per data point.
+Specifically in genomics, cohort sizes are in the thousands, each comprising of Millions of variants that make up a persons genetic differences.
+Traditional approaches are unable to cater for this task and even standard implementations using novel compute paradigms, such as Hadoop/Spark, are limited by the finite resource of computer memory on such data.
+Specifically, Spark ML?s supervised machine learning methods suffer from "the curse of dimensionality".
+That is, they scale well with samples (n) but not with features (p, variants).
+
+We introduce \cursedforest, a tailored implementation of random forests designed to handle data with extremely large number of variables per sample.
+We showcase \cursedforest\ to accurately predict ethnicity from whole genome sequencing profiles of 2,500 individuals and 80 million genomic variants each (1000 Genomes Project) in 30 minutes.
+\cursedforest\ was the only implementation scaling to the full dataset, outperforming other methods on smaller subsets by 80\% or more.
+We also demonstrate a GWAS-style analysis using  \cursedforest's feature selection approach on an imputed SNP array dataset (2000 x 7 Million) on Bone mineral density (BMD), successfully predicting known BMD-genes as well as identifying novel candidates.
+
+\cursedforest is included in our earlier variant interpretation package, VariantSpark, allowing it to perform near real-time classification of population-scale patient cohorts in "patients like mine" scenarios, as well as performing GWAS analysis on large unfiltered whole genome sequencing cohorts.
 
@@
 \subsection{\cursedforest outperforms existing methods for classification problems}
-In this section we demonstrate \cursedforest's classification accuracy on biological data and compare its performance against other published methods by predicting ethnicity on the
-1000 Genomes Project dataset.
-
+In this section we demonstrate \cursedforest's classification accuracy on biological data and compare its performance against other published methods. 
+We do this by predicting the ethnicity of the individuals in the 1000 Genomes Project from their genomic profiles and evaluate against the annotated label. 
+We record the prediction error (OOB) as well as the runtime for each method. 
 %In a previous paper, we clustered individuals from the 1000 Genomes Project using k-means clustering \cite{OBrien2015}.
@@
 %than building a k-means model, but does not scale to as many dimensions (see Table~\ref{timingtable}). 
-To keep
-the following comparisons fair, we limit the number of executors available to Spark.
-
-On the relatively low-dimensional
-phase1\_chr22 dataset (1,092 x 490,036), we observed a runtime of under 8 minutes. Although slower than Ranger, 
-presumably due to the overhead of Spark's distributed platform on this relatively small dataset, we observe Spark ML to
-be about 16 minutes faster than R. Moving to the larger dataset, phase3\_chr1 (2,504 x 6,450,364), we observe an increase
-in Spark ML's runtime to over 8 hours. Moving to our next largest dataset, phase3\_chr1\-3 (2,504 x 19,328,051), and the
-job fails to complete within our arbitrary cutoff time of 20 hours.
-
-From our trials, it was apparent that Spark ML, although designed to handle many samples, is unable to efficiently cope
-with a huge number of features. With the decreasing costs of genome sequencing, this case will surely become increasingly
-common.
-
-\cursedforest solves this problem of high-dimensionality genomic data. Not only is it faster on the above smaller datasets,
-but it is also able to scale to the entire genome. In Table~\ref{cursedforesttable} we present our observed times building
-models with CursedForest using all available resources on our Spark cluster.
-
-As we previously mentioned, building a random forest model on the phase3\_chr1 dataset with Spark ML takes over 8 hours.
-Building a model with \cursedforest, however, takes only 20 minutes using the same cluster setup and this time drops to
-nearly 4 minutes using the entire compute cluster. Subsequently, phase3\_chr1\-3 takes about 11 minutes, and building a
-model on the entire genome is just shy of 37 minutes. Not only does the job complete successfully, but building a random
-forest model on all 22 autosomes (phase3\_chr1\-22) using \cursedforest is more than twice as fast as building a random
-forest on the chromosome 1 (phase3\_chr1) using Spark ML.
+%@Aidan 
+We limit the number of executors to X that are available to \cursedforest\ to keep the comparison fair for all methods.
+
+As shown in Fig.~\ref{figure:1000genomes}, only \cursedforest\ is able to scale to the full phase3\_chr1-22 dataset (2,504 samples x 19,328,051 variants), with
+the standard random forest implementation in R and ranger's C++ version running out of resources on the relatively low-dimensional
+phase1\_chr22 dataset (1,092 x 490,036).
+Ranger's R implementation is able to process the second largest dataset (phase3\_chr1-3 with 2,504 x 19,328,051), however, for larger datasets the runtime increases beyond a reasonable compute time for this task (>20h).
+The standard implementation of random forest in SparkML already runs out of memory after processing the third largest dataset, phase3\_chr1 (2,504 x 6,450,364). 
+This demonstrates that Spark ML, although designed to handle many samples, is unable to efficiently cope with large numbers of features.
+
+%With the decreasing costs of genome sequencing, this case will surely become increasingly
+%common.
+All tested methods have a worse runtime than \cursedforest\ for the datasets they processed successfully, except for the smallest dataset where the overhead of Spark's distributed platform made Ranger's R implementation faster with 1 min 10 seconds compared to 1 min 50 seconds.
+For example, on phase3\_chr1-3 \cursedforest\ finishes in 1h, with Ranger(R) requiring 4h and SparkML already needing 8 hours on a smaller dataset.
+Extrapolating from this, all comparable implementations hence would require more than 100 days to finish, assuming the use of larger memory nodes.
+
+%with the ranger implementation in R and SparkML 
+%On the relatively low-dimensional
+%phase1\_chr22 dataset (1,092 samples x 490,036 variants), we observed a runtime of under 8 minutes. Although slower than Ranger, 
+%presumably due to the overhead of Spark's distributed platform on this relatively small dataset, we observe Spark ML to
+%be about 16 minutes faster than R. Moving to the larger dataset, phase3\_chr1 (2,504 x 6,450,364), we observe an increase
+%in Spark ML's runtime to over 8 hours. Moving to our next largest dataset, phase3\_chr1\-3 (2,504 x 19,328,051), and the
+%job fails to complete within our arbitrary cutoff time of 20 hours.
+
+%From our trials, it was apparent that Spark ML, although designed to handle many samples, is unable to efficiently cope
+%with a huge number of features. With the decreasing costs of genome sequencing, this case will surely become increasingly
+%common.
+
+%\cursedforest solves this problem of high-dimensionality genomic data. Not only is it faster on the above smaller datasets,
+%but it is also able to scale to the entire genome.
+
+As our implementation can efficiently utilise large numbers of commodity nodes, the performance of \cursedforest\ can be substantially improved further.  
+Table~\ref{cursedforesttable} shows the runtime when utilising the full Spark cluster (12 worker nodes).
+
+%As we previously mentioned, building a random forest model on the phase3\_chr1 dataset with Spark ML takes over 8 hours.
+Building a model with \cursedforest\ on phase3\_chr1, which previously took 20 minutes, can be reduced to 4 minutes. 
+Subsequently, phase3\_chr1-3 takes about 11 minutes, and building a
+model on the entire genome is 37 minutes.
+
+Not only does the job complete successfully using \cursedforest\, but building a random
+forest model on all 22 autosomes (phase3\_chr1-22)  leads to a more accurate result. 
+The error can be reduced from OOB=0.04 when using only data from Chr1, which represents processing power limit for most tools, down to OOB=0.01 when using the full dataset.
+
+%than twice as fast as building a random
+%forest on the chromosome 1 (phase3\_chr1) using Spark ML.
 
@@
     \caption{\textbf{Runtime comparison on the 1000 genomes project}}
-    \label{figure:synth1}
+    \label{figure:1000genomes}
     \includegraphics[totalheight=8cm]{./figs/1000genomesRuntime.pdf}
@@
 \subsection{Scalability analysis}
-In this section we explore the performance of \cursedforest by testing its ability to scale to different sizes of data
+In this section we explore the performance of \cursedforest\ in more detail by testing its ability to scale to different sizes of data
 and computational resources.
 
-In order to asses these characteristics we ran \cursedforest classification on synthetic datasets with varying numbers
+In order to asses these characteristics, we run \cursedforest classification on synthetic datasets with varying numbers
 of variables (features) and samples, similar to the dataset used in \cite{Wright.and.Ziegle.2016} to evaluate
@@
 samples, by varying the \mtry  fraction and the number of CPU cores allocated to the execution. 
-The results are  visualized in Fig~\ref{figure:synthetictiming} below. 
+%The results are visualised in Fig~\ref{figure:synthetictiming}. 
 %The raw results are
@@
 
-To conclude \cursedforest is capable of processing datasets with 50M variables and 10K samples, which is the order of
-magnitude of the entire genome variants expected in such dataset, for which it can build approximately 60 trees per
-hour.
+To conclude \cursedforest is capable of processing 60 trees per hour on a datasets with 50 million variables and 10,000 samples, which 
+is the size range for whole genome sequencing experiments of clinically relevant cohort sizes.
 
@@
 \section*{Acknowledgments}
-
-Cras egestas velit mauris, eu mollis turpis pellentesque sit
-amet. Interdum et malesuada fames ac ante ipsum primis in
-faucibus. Nam id pretium nisi. Sed ac quam id nisi malesuada
-congue. Sed interdum aliquet augue, at pellentesque quam rhoncus
-vitae.
+AO and PS implemented the method, AO conducted the BMD and 1000genomes experiments while PS generated the scaling data, analysed by RD with input from NE.
+PL and ELD aided in the interpretation of the BMD data, while SL and JD contributed to the 1000genomes analysis.
+DCB visualised the BMD and 1000genomes results.
+All authors contributed to the manuscript.
+The authors would like to acknowledge CSIRO, specifically PS, for setting up and maintaining the in-house Spark cluster.
 

### 2017-02-14 doc/publications/plos16/plos16.tex
introduction updated

@@
 
-\tableofcontents
-\clearpage
+%\tableofcontents
+%\clearpage
 
@@
 Analyses are performed on ever larger datasets with unprecedentedly detailed information per data point.
-Specifically in genomics, cohort sizes are in the thousands, each comprising of Millions of variants that make up a persons genetic differences.
+Specifically in genomics, cohort sizes are in the thousands, each comprising of Millions of variants that make up a person's genetic differences.
 Traditional approaches are unable to cater for this task and even standard implementations using novel compute paradigms, such as Hadoop/Spark, are limited by the finite resource of computer memory on such data.
-Specifically, Spark ML?s supervised machine learning methods suffer from "the curse of dimensionality".
+Specifically, Spark ML's supervised machine learning methods suffer from "the curse of dimensionality".
 That is, they scale well with samples (n) but not with features (p, variants).
 
-We introduce \cursedforest, a tailored implementation of random forests designed to handle data with extremely large number of variables per sample.
+We introduce \cursedforest, a tailored implementation of random forests, designed to handle data with extremely large number of variables per sample.
 We showcase \cursedforest\ to accurately predict ethnicity from whole genome sequencing profiles of 2,500 individuals and 80 million genomic variants each (1000 Genomes Project) in 30 minutes.
@@
 
-A tree-based model is capable of taking the interaction of variables into account \cite{Wright.et.al.2016}. In addition, random forests are well
+Hence using more sophisticated machine learning (ML) approaches, in particular tree-based models, have been successful
+for taking the interaction of variables into account~\cite{Wright.et.al.2016}. In addition, random forests are well
 suited for processing ``wide'' genomic data for two reasons.  Firstly, while other machine learning applications have
@@
 dimensionality''~\cite{Bauer2014, bellman1961adaptive}), decision trees are resistant to overfitting.  Secondly, random
-forests are also very easy to parallelize. As the forest is a sum of decision trees, it is possible to grow separate
-trees on different processors and combine the trees. See the discussion of the properties of random forests in Section
-\ref{section:methods}.
-
-The most popular random forest software packages are R-based, although current implementations do not scale well with
-increasing feature size~\cite{Wright.and.Ziegle.2016}.  This is despite R now supporting long vectors (length greater
-than $2^{31}-1$), and the \randomforest implementation in R lending itself to parallelization to grow multiple
-trees simultaneously~\cite{Liaw.and.Weiner.2002} and then combine them.  More suitable implementations for ``wide''
+forests are also very easy to parallelise. As the forest is a sum of decision trees, it is possible to grow separate
+trees on different processors and combine the trees. 
+%See the discussion of the properties of random forests in Section \ref{section:methods}.
+
+The most popular random forest software packages to date are R-based, although it has been reported that current implementations 
+do not scale well with increasing feature size~\cite{Wright.and.Ziegle.2016}.  This is despite R now supporting long vectors (length greater
+than $2^{31}-1$), and the \randomforest implementation in R lending itself to parallelisation by growing multiple
+trees simultaneously and combining them for the final result~\cite{Liaw.and.Weiner.2002}.  More suitable implementations for ``wide''
 genomic data have been developed in other programming languages. \textsc{CloudForest} \cite{Bressler2015}, written in Go,
-achieves fast run times by effective use of the CPU cache, optimizing for different classes of features and efficiently
+achieves fast run times by effective use of the CPU cache, optimising for different classes of features and efficiently
 multithreading.  Similarly, \ranger~\cite{Wright.and.Ziegle.2016}, written in C++ with an R front-end, is
-specifically optimized for dealing with large datasets.
-
-The use of traditional compute infrastructure limits the parallelization strategies that can be employed by  
-these methods.  The programs are limited to utilizing only CPUs that are on the same computer node (multithreading) or
-else they can run on CPUs distributed across nodes by virtue of the fact that there is no communication between the
-processes (a separate tree is grown on each node).  Hadoop/Spark overcomes these limitations by enabling programs to
-scale beyond compute-node boundaries and hence enable more sophisticated parallelization strategies.  In the case of
-random forests this means that the computations for each node of a tree can be handed off to separate processors.
-
-Here we introduce \cursedforest, a Hadoop/Spark-based implementation of random forests specifically designed to cater
-for ``big'' (many samples) and ``wide'' (many features) datasets. A Spark application runs on a ``driver'' node and
-distributes tasks to many ``worker'' nodes, or ``executors''. Using this facility, \cursedforest is capable of
-parallelizing the split for each node in a tree thereby handling millions of features. It can, for example, fit a
-model to the 1000 Genomes Project data, which consists of approximately 2500 samples with up to 80 million
-variants~\cite{1KG2012}.  By also utilizing Spark to read in and manipulate the standard genomic variant format (VCF)
+specifically optimised for dealing with large datasets.
+
+However, the use of traditional compute infrastructure limits the parallelisation strategies that can be employed.  
+The programs are limited to utilising only CPUs that are on the same computer node (multithreading) or
+farm out independent tasks to CPUs distributed across nodes that do not require communication between the
+processes (a separate tree is grown on each node).
+%else they can run on CPUs distributed across nodes by virtue of the fact that there is no communication between the
+%processes (a separate tree is grown on each node).  
+Hadoop/Spark overcomes these limitations by enabling programs to
+scale beyond compute-node boundaries and hence enable more sophisticated parallelisation strategies.  In the case of
+random forests, the computations for each node of a tree can hence be handed off to separate processors.
+
+Despite overcoming the node-boundary limitation, the standard implementation of random forest in Spark ML
+is not able to handle the extremely ``wide'' genomic data as it was developed for a large number of samples
+with only modest dimensionality~\cite{NIPS2016_6366}.  Although Spark ML can build a random forest model on a subset of the data (chromosome
+1), we show that the time taken is excessive due to the large amount of data being aggregated and processed by the driver node during
+intermediate stages of building the model.  This unbalanced work load where the driver
+node becomes the bottleneck and worker nodes being idle prevents a seamless scaling to larger datasets. We also show that the memory
+requirements per executor increases with dimensionality due to the data types Spark ML uses.%, which we elaborate on in the discussion.
+
+Here we introduce \cursedforest, a tailored Hadoop/Spark-based implementation of random forests specifically designed to cater
+for ``big'' (many samples) and ``wide'' (many features) datasets. 
+\cursedforest extends our previously developed variant interpretation framework, VariantSpark~\cite{OBrien2015}, to now offer
+supervised as well as unsupervised ML algorithms in the Spark framework.
+In our implementation a Spark application runs on a ``driver'' node and
+distributes tasks to many ``worker'' nodes, or ``executors''. 
+By also utilising Spark to read in and manipulate the standard genomic variant format (VCF)
 directly, \cursedforest outperforms existing tools even on small datasets where multithreading generally performs
-well. Harnessing the virtually unlimited capability to parallelize tasks, \cursedforest can build a large number of
-trees and use a high \mtry value which may lead to greater accuracy (see the example in section \ref{section:synthetic}).
-
-\marginpar{does \cite{Szymczak2016} say that high \ntree gives greater  accuracy for variable importance selection?
-Check this. }
-
-We previously demonstrated the versatility and scalability of Spark by developing VariantSpark~\cite{OBrien2015}, a
-framework allowing users to easily analyze Variant Call Format (VCF) files using ML algorithms on the Spark framework.
-Using VariantSpark, we successfully built a k-means model on the full $2500 \times 80$ million matrix to cluster
-individuals by their ethnicity achieving an Adjusted Rand Index (ARI) of 0.84 (with 1 being perfect clustering and -1
-random clustering).
-
-Here we extend this toolkit to encompass supervised learning using random forests.  Spark ML's random forest
-implementation is not able to handle the extremely ``wide'' genomic data as it was developed for a large number of samples
-with only modest dimensionality.  Although Spark ML can build a random forest model on a subset of the data (chromosome
-1), the time taken is excessive due to the large amount of data being aggregated and processed by the driver node during
-intermediate stages of building the model (see Table~\ref{timingtable}).  This unbalanced work load where the driver
-node becomes the bottleneck and worker nodes being idle prevents a seamless scaling to larger datasets. The memory
-requirements per executor also increases with dimensionality due to the data types Spark ML uses, which we elaborate on
-in the discussion.
-
-\cursedforest is included in our previously developed framework for Spark-based genomic data analysis,
-VariantSpark~\cite{OBrien2015}, which therefore now enables supervised and unsupervised machine learning. This offers a
-comprehensive analysis toolkit that can scale for future data demands, where the richness from WGS data can be
-utilized to ``fill in'' or impute the information at previously unobserved genomic positions in older array
-technology~\cite{Howie2012}.  This provides the potential to impute samples in the GWAS catalog and generate datasets
-of hundreds of thousands of individuals with millions of variants, highlighting the need for incorporating modern
-compute paradigms to deal with these challenges.
-
-In the first section we explore some of the properties of random forests as these will also be apparent in the
-\cursedforest implementation. Here we provide demonstrations using synthetic data. Secondly, we demonstrate the ability of
-\cursedforest to select previously demonstrated features from a bone mineral density (BMD) dataset. 
+well. Harnessing the virtually unlimited capability to parallelise tasks, \cursedforest can hence explore the solution 
+space faster by building a larger number of diverse models to generate a consensus from.
+
+Using this facility, \cursedforest is capable of parallelising the split for each node in a tree thereby handling millions 
+of features, as required to process whole genome sequencing data or SNP array data with unobserved genotypes 
+imputed~\cite{Howie2012}.  This provides the potential to generate datasets of hundreds of thousands of individuals 
+with millions of variants (imputing the GWAS catalog), highlighting the need for modern
+compute paradigms in the genomics space.
+
+%It can, for example, fit a model to the 1000 Genomes Project data, which consists of approximately 2500 samples with up to 80 million
+%variants~\cite{1KG2012}.  
+%build a larger number of
+%trees hence building a larger number of diverse models to generate the consensus from.
+%and use a high \mtry value which may lead to greater accuracy (see the example in section \ref{section:synthetic}).
+%\marginpar{does \cite{Szymczak2016} say that high \ntree gives greater  accuracy for variable importance selection?
+%Check this. }
+%We previously demonstrated the versatility and scalability of Spark by developing VariantSpark~\cite{OBrien2015}, a
+%framework allowing users to easily analyze Variant Call Format (VCF) files using ML algorithms on the Spark framework.
+%Using VariantSpark, we successfully built a k-means model on the full $2500 \times 80$ million matrix to cluster
+%individuals by their ethnicity achieving an Adjusted Rand Index (ARI) of 0.84 (with 1 being perfect clustering and -1
+%random clustering).
+%Using VariantSpark, we successfully built a k-means model on the full $2500 \times 80$ million matrix to cluster
+%individuals by their ethnicity achieving an Adjusted Rand Index (ARI) of 0.84 (with 1 being perfect clustering and -1
+%random clustering).
+%Here we extend this toolkit to encompass supervised learning using random forests.
+
+VariantSpark~\cite{OBrien2015} with the \cursedforest extension therefore offers a
+comprehensive analysis toolkit that can scale to future data demands. To showcase the framework's ability we demonstrate
+a classification as well as feature-selection task on synthetic data in the first section. In the second section, we demonstrate the ability of
+\cursedforest to successfully replicate findings from a previous GWAS study as well as identify novel variants associated with bone 
+mineral density (BMD). 
 Thirdly, we demonstrate the scalability of \cursedforest in respect to the dimensionality of data by building a random
-forest model on whole-genome data from the 1000 Genomes Project. Here we report the error in predicting ethnicity.
+forest model on whole-genome data from the 1000 Genomes Project~\cite{1KG2012} to predict ethnicity.
 Finally, given the role different parameter values can play in model construction, we explore the effect that tuning these

### 2017-02-14 doc/publications/plos16/plos16.tex
new p-value

@@
 As shown in Fig~\ref{figure:ranksumtest}, the known genes are ranked significantly higher than variants from other
-locations in the genome (Mann-Whitney U, $p=1.3\times10^{-07}$), supporting our hypothesis that biologically relevant
+locations in the genome (Mann-Whitney U, $p=9.71\times10^{-15}$), supporting our hypothesis that biologically relevant
 features are contributing more information to our classification model.

### 2017-02-14 doc/publications/plos16/supplementary_information.tex
first pass through all result sections and abstract
changes to the supplementary_information.tex

@@
 
-<<echo=FALSE,eval=FALSE>>=
-
-@
-
 \section{supplementary Information for Section 4.1}
 
-\begin{table}[!ht]
+\begin{sidewaystable}[!ht]
 \begin{minipage}{\textwidth}
@@
                                                     & H2O           & 0.05       & 16 hours         & 600GB \footnote{java -Xmx600g -jar h2o.jar,   h2o.init(nthreads=20)} \\
-                                                    & ranger - R    & 0.94       & 4 1/2 mins       & 21GB (one processor) \\
+                                                    & ranger - R    & 0.06       & 4 1/2 mins       & 21GB (one processor) \\
 \hline
-\multirow{3}{*}{phase3\_chr1: 2,504 x 6,450,364}    & \cursedforest\ & 0.063 (XX) & 1hr 58min        & 4GB \\
-                                                    & RF - Spark ML  & 0.94       & 8hr 29min        & 8GB \\
+\multirow{3}{*}{phase3\_chr1: 2,504 x 6,450,364}    & \cursedforest\ & 0.06 & 1hr 58min        & 4GB \\
+                                                    & RF - Spark ML  & 0.06       & 8hr 29min        & 8GB \\
                                                     & LR - Spark ML  &            & 2hr 6min         & 8GB \\
-                                                    & ranger - R      & 0.95       & 28 mins          & num.threads=10\footnote{don't
+                                                    & ranger - R      & 0.05       & 28 mins          & num.threads=10\footnote{don't
                                                                                                       know how to get the memory
@@
 \hline
-\multirow{3}{*}{phase3\_chr1-3: 2,504 x 19,328,051} & \cursedforest\ & 0.046 (XX) & 4hr 0min                                             & 4GB \\
+\multirow{3}{*}{phase3\_chr1-3: 2,504 x 19,328,051} & \cursedforest\ & 0.04 & 4hr 0min                                             & 4GB \\
                                                     & RF - Spark ML  &            &                                                      & \\
                                                     & LR - Spark ML  &            & 14hr 48min                                           & 8GB \\
-                                                    & ranger - R      &  0.97      & 58 mins\footnote{ but took 6 hours to load the
+                                                    & ranger - R      &  0.03      & 58 mins\footnote{ but took 6 hours to load the
                                                                                    data}
@@
                                                                                                                                            }  \\
-& ranger - R        &             0.973     &      03:58           &       pearcey --cpus-per-task=15 --mem 1600gb   906GB \\
+& ranger - R        &             0.027     &      03:58           &       pearcey --cpus-per-task=15 --mem 1600gb   906GB \\
 \hline
 
-\multirow{4}{*}{phase3\_chr1-22: 2,504 x 81,047,467} & \cursedforest\ & XX (0.96) & 7hr 14min & 8GB \\
+\multirow{4}{*}{phase3\_chr1-22: 2,504 x 81,047,467} & \cursedforest\ & 0.04 & 7hr 14min & 8GB \\
 & RF - Spark ML & - & - & - \\
 & LR - Spark ML & - & - & - \\ 
-& kmeans - Spark ML  & - (0.82)      & 30hr 44min    & 24GB \\ 
+& kmeans - Spark ML  & 0.18      & 30hr 44min    & 24GB \\ 
 & ranger (C++)       &        NA     &        NA     &            80 CPU -- runs out of memory (12.8GB per CPU) \\
@@
 \end{minipage}
-\end{table}
+\end{sidewaystable}
 
-\begin{table}[!ht]
+\bibliographystyle{apalike}
+
+%\bibliography{/home/dun280/Dropbox/TeX/bibliography/screening_compliance.bib}
+\bibliography{/home/dun280/Dropbox/TeX/bibliography/ripley.bib}
+\end{document}
+
+\begin{sidewaystable}[!ht]
 \begin{minipage}{\textwidth}
-\begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
+%\begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
 \centering
@@
 \label{table1}
-\end{adjustwidth}
+%\end{adjustwidth}
 \end{minipage}
-\end{table}
+\end{sidewaystable}
 
-\bibliographystyle{apalike}
-
-%\bibliography{/home/dun280/Dropbox/TeX/bibliography/screening_compliance.bib}
-\bibliography{/home/dun280/Dropbox/TeX/bibliography/ripley.bib}
-\end{document}
+% \begin{tabular}{|p{1.2cm}|p{1cm}|p{1.5cm}+l|l|p{1.8cm}|p{1.5cm}|}
+% \hline
+% %\multicolumn{4}{|l|}{\bf Heading1} & \multicolumn{4}{|l|}{\bf Heading2}\\ \hline
+% \bf{Data- set} & \bf{Samp- les} & \bf{Feat- ures}  & \bf{Method} & \bf{Error (ARI)} & \bf{Runtime} & \bf{Memory/ Node} \\
+% \hline
+% %\multirow{3}{*}{phase1\_chr22: 1,092 x 490,036} & CursedForest & 0.077 (XX)  &  &  \\
+% phase1 chr22 &1,092 & 490,036 & CursedForest & 0.077 (XX)  & 9min 16sec & 1GB\footnote{\label{note10}10 nodes} \\
+% &&& RF - Spark ML & 0.069 (0.89) & 14min 13sec & 1GB\footref{note10}  \\
+% &&& RF - R &  0.05  & 24min  & 18GB\footref{note1}\footnote{using the \texttt{parallel} packages}\\
+% &&& ranger -R &         0.05 &        1min 10sec  &          16GB  (one processor) \\
+%  \hline
+% phase3 chr22 & 2,504 & 1,103,548 &  &  &  &  \\
+% &&& CursedForest & 0.053 & 9 min 8 sec & 2GB\footnote{\label{note20}20 nodes}\\
+% &&& RF - Spark ML & 0.058 (0.92) & 1hr 1min & 2GB\footref{note20} \\
+% &&& RF - R & NA  & NA & NA\footnote{long vectors ($> 2^31-1$)  are not supported in the .Fortran interface}\\
+% &&&  H2O   &        0.05 &     16 hours       &   600GB \footnote{java -Xmx600g -jar h2o.jar,   h2o.init(nthreads=20)} \\
+% \hline
+% phase3 chr1 & 2,504 & 6,450,364 & CursedForest & 0.063 (XX) & 1hr 58min & 4GB \\
+% &&& RF - Spark ML & (0.94) & 8hr 29min & 8GB \\
+% &&& LR - Spark ML&  & 2hr 6min & 8GB\\ 
+% &&& ranger  R   &  0.06   &     28 mins  &            num.threads = 10\footnote{don't know how to get the memory used, This
+%   was done on dpas-03} \\ 
+% \hline
+% phase3 chr1-3 & 2,504 & 19,328,051  & CursedForest & 0.046 (XX) & 4hr 0min & 4GB\\
+% &&& RF - Spark ML & - & -  & -\\
+% &&& LR - Spark ML & & 14hr 48min & 8GB \\ 
+% &&& ranger  R   &  0.05   &     58 mins\footnote{ but took 6 hours to load the data}  &      909GB\footnote{I think this is total memory. The job was run with
+%   ntasks-per-node=160 cores-per-socket=10 on ruby }  \\ 
+% \hline
+% phase3 chr1-22 & 2,504 & 81,047,467 & CursedForest & XX (0.96) & 7hr 14min & 8GB \\ 
+% &&& RF - Spark ML & - & - & - \\
+% &&& LR - Spark ML & - & - & - \\ 
+% %&&& kmeans - Spark ML & - (0.82) & 30hr 44min & 24GB \\ 
+% \hline

### 2017-02-15 doc/publications/plos16/plos16.tex
added conclusion

@@
 
-\cursedforest scales almost linearly with the number of CPU cores for medium values of \mtry fraction but for both
-lower and higher values the performance degrades slightly (Fig~\ref{figure:synthetictiming.b}). In the latter case the likely cause is
-communication overhead (with lower \mtry values the proportion of time for parallelizable computation to the time for
-internode communication is lower) while in the latter case it is most likely caused by reaching the clusters
+\cursedforest scales almost linearly with the number of CPU cores for medium values of \mtry fraction but for both lower
+and higher values the performance degrades slightly (Fig~\ref{figure:synthetictiming.b}). In the latter case the likely
+cause is communication overhead (with lower \mtry values the proportion of time for parallelizable computation to the
+time for internode communication is lower) while in the latter case it is most likely caused by reaching the clusters
 computational capacity.
 
-Next we investigate \cursedforest scalability with
-regards to the size of data, by varying the number of variables and sample for a fixed \mtry fraction of 0.25 and
-execution of 128 CPU cores. The results are visualized in Fig~\ref{figure:synth2} below (please note log scale on the
-axes and the values on y axes are expressed as trees per hour).
+Next we investigate \cursedforest scalability with regards to the size of data, by varying the number of variables and
+sample for a fixed \mtry fraction of 0.25 and execution of 128 CPU cores. The results are visualized in
+Fig~\ref{figure:synth2} below (please note log scale on the axes and the values on y axes are expressed as trees per
+hour).
 
@@
 irregularities in the graph can be attributed to computation vs communication tradeoff. It is also worth noting that
-keeping the \mtry fraction constant results in higher \mtry values with the growing number of variables, and this is what
-drives the performance down rather that the increase of dataset size itself.
-
-To conclude \cursedforest is capable of processing 60 trees per hour on a datasets with 50 million variables and 10,000 samples, which 
-is the size range for whole genome sequencing experiments of clinically relevant cohort sizes.
+keeping the \mtry fraction constant results in higher \mtry values with the growing number of variables, and this is
+what drives the performance down rather that the increase of dataset size itself.
+
+To conclude \cursedforest is capable of processing 60 trees per hour on a datasets with 50 million variables and 10,000
+samples, which is the size range for whole genome sequencing experiments of clinically relevant cohort sizes.
 
@@
 \section{Conclusion}
+
 We have demonstrated that using a different parallelization model can extend random forests to the case of an
@@
 Donoho and Tanner \cite{Donoho.and.Tanner.2009} give a ``universal phase change'' result that has applications in a
-large number of areas including variable selection in high dimensions. Consider
-Fig~\ref{figure:phase-diagram-equivalence.png} which shows the region where a model can recover the important variables,
-plotted as a function of $\delta = n/p$ and $\rho =k/n$ (where $k$ is the number of significant variables). There is a
-distinct boundary, shown empirically and also based on arguments from combinatorial geometry, to the region where we can
-reliably recover significant variables.  \cite{Donoho.and.Stodden.2006} investigate the behavior of a number of
-regression approaches for variable selection (LARS, Lasso and forward stepwise) and make the point that above the
-phase-transition line variable recovery is still possible by a combinatorial approach.
-
-It is not surprising that that it is more difficult to recover the signal variable in the upper-left area of the figure,
-as the problem is both under-determined and sparse. What is surprising is the connection with arguments from
-combinatorial geometry. This suggests that we are seeing a universal rule rather than an implementation issue. As
-\cursedforest is designed for extremely large numbers of variables it is likely to be operating in difficult regions of
-the figure where the ratio $\delta = n/p$ is small.
-
-We note several things here:
-\begin{itemize}
-\item the Donoho-Tanner phase transition arises in recovering the $\beta$ in data generated by a
-  linear model. However, in a decision tree (random forest) there is no notion of estimating the $\beta$
-\item A decision tree (random forest) is a heuristic search. It may recover a relationship in the space of the
+large number of areas including variable selection in high dimensions. They show that there is a sharp disjunction
+between the cases where significant variables may be recovered with a high accuracy by procedures like stepwise variable
+selection, and the cases where they can not be recovered. The boundary is shown in
+Fig~\ref{figure:phase-diagram-equivalence.png} in a space parameterized by the level of underdeterminedness,
+$\delta = n/p$, and by the sparsity, $\rho =k/n$ (where $k$ is the number of significant variables). The theoretical
+boundary is based on arguments from combinatorial geometry. Above the phase-transition line variable recovery is still
+possible by a combinatorial approach such as all-subsets variable selection.
+
+\begin{figure}[tbhp] 
+ \begin{adjustwidth}{-1.00in}{0in}
+    \centering
+    \includegraphics[totalheight=6cm]{./figs/phase-diagram-equivalence.png} 
+    \caption{{\bf The phase change diagram.}}
+    \label{figure:phase-diagram-equivalence.png} 
+    \vspace{4ex}
+  \end{adjustwidth}
+\end{figure}
+
+\cite{Donoho.and.Stodden.2006} investigate the behavior of a number of
+regression approaches for variable selection in a small simulation.  They consider a regression probem with 
+$\underset{n\times p}{X}\sim N(0,1)$ with $p$ fixed at 200 and $n$ variable.
+They  set $\beta(1:k) \sim U(1,100)$ and $\beta((k+1):p) =0$. Then  $y= X\beta + \epsilon$, $\epsilon \sim
+N(0,\sqrt{16})$. They evaluate variable selection by the error measure 
+$$\frac{\Vert\hat{\beta}-\beta\Vert_2}{\Vert\beta\Vert_2}$$
+
+They consider a number of variable selection methods, including a false discovery rate criteria. This involves adding
+the variable with the maximum $t$-value to the linear model if the $p$-value is less than 
+$25(\text{number of  terms  currently  in  the  model})/(\text{total  number  of  variables})$.  
+See Fig~\ref{figure:error_Stodden_FDR.png} for the error measure and
+figure \ref{figure:rbo_Stodden_FDR.png} for the RBO, comparing the ranking (i.e. values) of $\hat{\beta}$ and $\beta$.
+It apparent that there is a marked drop in the accuracy of the variable recovery in line with the pediction of the
+Donoho-Tanner phase transition.
+
+\begin{figure}[tbhp] 
+    \begin{subfigure}[b]{0.5\linewidth}
+      \centering
+      \includegraphics[totalheight=6cm]{./figs/k.png}
+      \caption{$k$}
+      \label{figure:k.png}
+    \end{subfigure} 
+    \begin{subfigure}[b]{0.5\linewidth}
+      \centering
+      \includegraphics[totalheight=6cm]{./figs/n.png}
+      \caption{$n$}
+      \label{figure:n.png}
+    \end{subfigure} 
+    \caption{The simulation in \protect{\cite{Donoho.and.Stodden.2006}} considers $\delta = n/p$ and $\rho =k/n$. Here we plot the
+      space of $\{ \delta, \rho$\}, colored  by the paramters $n$ and $k$}
+\end{figure}
+
+ \begin{figure}[tbhp] 
+%   \begin{adjustwidth}{-1.00in}{0in}
+     \begin{subfigure}[t]{0.5\linewidth}
+       \centering
+       \includegraphics[totalheight=6cm]{./figs/error_Stodden_FDR.png}
+       \caption{The error measure.}
+       \label{figure:error_Stodden_FDR.png}
+%       \vspace{4ex}
+     \end{subfigure} 
+     \begin{subfigure}[t]{0.5\linewidth}
+       \centering
+       \includegraphics[totalheight=6cm]{./figs/rbo_Stodden_FDR.png}
+       \caption{The RBO measure}
+       \label{figure:rbo_Stodden_FDR.png}
+     \end{subfigure} 
+     \caption{Forward stepwise variable selction with a false discovery rate stopping criteria.}
+       \label{figure:error_Stodden_FDR.png}
+%   \end{adjustwidth}
+ \end{figure}
+
+ \begin{figure}[tbhp] 
+%   \begin{adjustwidth}{-1.00in}{0in}
+     \begin{subfigure}[t]{0.5\linewidth}
+       \centering
+       \includegraphics[totalheight=6cm]{./figs/ranger_error_Stodden_simulation.png}
+       \caption{Stepwise with FDR stopping criteria. $p=200$, $n$ and $k$ variable. }
+       \label{figure:ranger_error_Stodden_simulation.png}
+       \vspace{4ex}
+     \end{subfigure} 
+     \begin{subfigure}[t]{0.5\linewidth}
+       \centering
+       \includegraphics[totalheight=6cm]{./figs/ranger_rbo_Stodden_simulation.png}
+       \caption{RBO}
+       \label{figure:ranger_rbo_Stodden_simulation.png}
+     \end{subfigure} 
+     \caption{}
+     \label{figure:ranger_rbo_Stodden_simulation.png}
+%   \end{adjustwidth}
+ \end{figure}
+
+We have seen no work on the question of such a phase transition in the case of classification problems. Still it amy be
+instructive to consider the implications of the pahse transition for our work. As \cursedforest is designed for
+extremely large numbers of variables it is likely to be operating in difficult regions of the figure where the ratio
+$\delta = n/p$ is small. In the simulated example in section \ref{section:synthetic_data} the underdeterminedness $\delta = 0.002$ and the
+sparsity $\rho =2\times 10^{-6}$
+
+We have used a random forest on the same simulation as figure \ref{figure:error_Stodden_FDR.png}.  The Donoho-Tanner phase transition arises in
+recovering the $\beta$ in data generated by a linear model. However, in a decision tree (random forest) we are fitting a
+non-linear model and there is no notion of estimating the $\beta$. Becsaue of this we have evaluted the performance of
+the random forest using the RBO measure on the variable importance.
+
+We note that while a random forest is a long way from an all-seubsets search, it is a limited combinatorial search of
+the feature space. As such it may perform outside of the bounds of the phase transition. 
   combinatorial search.
-\end{itemize}
+
+Fig~\ref{figure:ranger_rbo_Stodden_simulation.png} shosw the OOB prediction error and the RBO error for a random forest with 10000 trees and the \mtry value set
+to the default for a regression problem (depending on $n$). It shows no evidence of following the shape of the
+Donoho-Tanner phase transition.
+
+As \cursedforest is designed for extremely large numbers of variables it is likely to be operating in difficult regions of the figure.
+We note that the examples we are considering lie on the extreme left of the plot, close to the origin. So there are
+limits to what can be recovered in the presence of noise. However in the case of data that is both big and wide,
+\cursedforest and other VariantSpark methods may provide a useful tool.
 
@@
 \cursedforest and other VariantSpark methods may provide a useful tool.
-
-\begin{figure}[tbhp] 
- \begin{adjustwidth}{-1.00in}{0in}
-    \caption{\textbf{The Donoho-Tanner phase chage diagram.}}
-    \centering
-    \includegraphics[totalheight=6cm]{./figs/phase.png} 
-    \label{figure:phase-diagram-equivalence.png} 
-    \vspace{4ex}
-  \end{adjustwidth}
-\end{figure}
 
@@
 \section*{Acknowledgments}
-AO and PS implemented the method, AO conducted the BMD and 1000genomes experiments while PS generated the scaling data, analysed by RD with input from NE.
-PL and ELD aided in the interpretation of the BMD data, while SL and JD contributed to the 1000genomes analysis.
-DCB visualised the BMD and 1000genomes results.
-All authors contributed to the manuscript.
+AO and PS implemented the method, AO conducted the BMD and 1000genomes experiments while PS generated the scaling data,
+analysed by RD with input from NE. PL and ELD aided in the interpretatio nof the BMD data, while SL and JD contributed
+to the 1000genomes analysis. DCB visualised the BMD and 1000genomes results. All authors contributed to the manuscript.
 The authors would like to acknowledge CSIRO, specifically PS, for setting up and maintaining the in-house Spark cluster.
@@
 \end{document}
+
+% We have demonstrated that using a different parallelization model can extend random forests to the case of an
+% extremely large number of variables. We have treated the case of variable selection in a $p >> n$ model where most of
+% the variables are uninformative and have demonstrated the utility of the model for large GWAS datasets. By comparing
+% this implementation to other implementations (including those optimized for large datasets) we have demonstrated the
+% utility of this approach.
+
+% Donoho and Tanner \cite{Donoho.and.Tanner.2009} give a ``universal phase change'' result that has applications in a
+% large number of areas including variable selection in high dimensions. Consider
+% Fig~\ref{figure:phase-diagram-equivalence.png} which shows the region where a model can recover the important variables,
+% plotted as a function of $\delta = n/p$ and $\rho =k/n$ (where $k$ is the number of significant variables). There is a
+% distinct boundary, shown empirically and also based on arguments from combinatorial geometry, to the region where we can
+% reliably recover significant variables.  \cite{Donoho.and.Stodden.2006} investigate the behavior of a number of
+% regression approaches for variable selection (LARS, Lasso and forward stepwise) and make the point that above the
+% phase-transition line variable recovery is still possible by a combinatorial approach.
+
+% It is not surprising that that it is more difficult to recover the signal variable in the upper-left area of the figure,
+% as the problem is both under-determined and sparse. What is surprising is the connection with arguments from
+% combinatorial geometry. This suggests that we are seeing a universal rule rather than an implementation issue. As
+% \cursedforest is designed for extremely large numbers of variables it is likely to be operating in difficult regions of
+% the figure where the ratio $\delta = n/p$ is small.
+
+% We note several things here:
+% \begin{itemize}
+% \item the Donoho-Tanner phase transition arises in recovering the $\beta$ in data generated by a
+%   linear model. However, in a decision tree (random forest) there is no notion of estimating the $\beta$
+% \item A decision tree (random forest) is a heuristic search. It may recover a relationship in the space of the
+%   combinatorial search.
+% \end{itemize}
+
+% The existence of the  Donoho-Tanner phase transition is a salutary warning. There are likely to be limits, both
+% computational and logical, to the recovery of signals from noisy data. \cursedforest  is a contribution to addressing the
+% practical limits but the logical limits will still apply. However in the case of data that is both big and wide,
+% \cursedforest and other VariantSpark methods may provide a useful tool.
+
+% \begin{figure}[tbhp] 
+%  \begin{adjustwidth}{-1.00in}{0in}
+%     \caption{\textbf{The Donoho-Tanner phase chage diagram.}}
+%     \centering
+%     \includegraphics[totalheight=6cm]{./figs/phase.png} 
+%     \label{figure:phase-diagram-equivalence.png} 
+%     \vspace{4ex}
+%   \end{adjustwidth}
+% \end{figure}

### 2017-02-15 doc/publications/plos16/plos16.tex
changes to conclustion

@@
 selection, and the cases where they can not be recovered. The boundary is shown in
-Fig~\ref{figure:phase-diagram-equivalence.png} in a space parameterized by the level of underdeterminedness,
-$\delta = n/p$, and by the sparsity, $\rho =k/n$ (where $k$ is the number of significant variables). The theoretical
-boundary is based on arguments from combinatorial geometry. Above the phase-transition line variable recovery is still
-possible by a combinatorial approach such as all-subsets variable selection.
+Fig~\ref{figure:phase-diagram-equivalence.png} in a space parameterized by the level of underdetermination,
+$\delta = n/p$, and by the sparsity, $\rho =k/n$ (where $k$ is the number of significant variables).  Above the
+phase-transition line variable recovery is still possible by a combinatorial search approach such as all-subsets
+variable selection.  The theoretical boundary is based on arguments from combinatorial geometry.
 
@@
 \cite{Donoho.and.Stodden.2006} investigate the behavior of a number of
-regression approaches for variable selection in a small simulation.  They consider a regression probem with 
+regression approaches for variable selection in a small simulation.  They consider a regression problem with 
 $\underset{n\times p}{X}\sim N(0,1)$ with $p$ fixed at 200 and $n$ variable.
-They  set $\beta(1:k) \sim U(1,100)$ and $\beta((k+1):p) =0$. Then  $y= X\beta + \epsilon$, $\epsilon \sim
+They  set $\beta(1:k) \sim U(1,100)$ and $\beta((k+1):p) =0$, then  $y= X\beta + \epsilon$, $\epsilon \sim
 N(0,\sqrt{16})$. They evaluate variable selection by the error measure 
@@
 figure \ref{figure:rbo_Stodden_FDR.png} for the RBO, comparing the ranking (i.e. values) of $\hat{\beta}$ and $\beta$.
-It apparent that there is a marked drop in the accuracy of the variable recovery in line with the pediction of the
+It apparent that there is a marked drop in the accuracy of the variable recovery in line with the prediction of the
 Donoho-Tanner phase transition.
-
-\begin{figure}[tbhp] 
-    \begin{subfigure}[b]{0.5\linewidth}
-      \centering
-      \includegraphics[totalheight=6cm]{./figs/k.png}
-      \caption{$k$}
-      \label{figure:k.png}
-    \end{subfigure} 
-    \begin{subfigure}[b]{0.5\linewidth}
-      \centering
-      \includegraphics[totalheight=6cm]{./figs/n.png}
-      \caption{$n$}
-      \label{figure:n.png}
-    \end{subfigure} 
-    \caption{The simulation in \protect{\cite{Donoho.and.Stodden.2006}} considers $\delta = n/p$ and $\rho =k/n$. Here we plot the
-      space of $\{ \delta, \rho$\}, colored  by the paramters $n$ and $k$}
-\end{figure}
 
@@
      \end{subfigure} 
-     \caption{Forward stepwise variable selction with a false discovery rate stopping criteria.}
+     \caption{Forward stepwise variable selection with a false discovery rate stopping criteria.}
        \label{figure:error_Stodden_FDR.png}
@@
        \includegraphics[totalheight=6cm]{./figs/ranger_error_Stodden_simulation.png}
-       \caption{Stepwise with FDR stopping criteria. $p=200$, $n$ and $k$ variable. }
+       \caption{The OOB error rate.}
        \label{figure:ranger_error_Stodden_simulation.png}
@@
        \includegraphics[totalheight=6cm]{./figs/ranger_rbo_Stodden_simulation.png}
-       \caption{RBO}
+       \caption{The RBO measure}
        \label{figure:ranger_rbo_Stodden_simulation.png}
      \end{subfigure} 
-     \caption{}
+     \caption{The simulated data of \protect{\cite{Donoho.and.Stodden.2006}} with a random forest model.}
      \label{figure:ranger_rbo_Stodden_simulation.png}
@@
 
-We have seen no work on the question of such a phase transition in the case of classification problems. Still it amy be
-instructive to consider the implications of the pahse transition for our work. As \cursedforest is designed for
+We have seen no work on the question of such a phase transition in the case of classification problems. Still it may be
+instructive to consider the implications of the phase transition for our work. As \cursedforest is designed for
 extremely large numbers of variables it is likely to be operating in difficult regions of the figure where the ratio
-$\delta = n/p$ is small. In the simulated example in section \ref{section:synthetic_data} the underdeterminedness $\delta = 0.002$ and the
+$\delta = n/p$ is small. In the simulated example in section \ref{section:synthetic_data} the underdetermination $\delta = 0.002$ and the
 sparsity $\rho =2\times 10^{-6}$
@@
 recovering the $\beta$ in data generated by a linear model. However, in a decision tree (random forest) we are fitting a
-non-linear model and there is no notion of estimating the $\beta$. Becsaue of this we have evaluted the performance of
+non-linear model and there is no notion of estimating the $\beta$. Because of this we have evaluated the performance of
 the random forest using the RBO measure on the variable importance.
 
-We note that while a random forest is a long way from an all-seubsets search, it is a limited combinatorial search of
+We note that while a random forest is a long way from an all-subsets search, it is a limited combinatorial search of
 the feature space. As such it may perform outside of the bounds of the phase transition. 
@@
 
-Fig~\ref{figure:ranger_rbo_Stodden_simulation.png} shosw the OOB prediction error and the RBO error for a random forest with 10000 trees and the \mtry value set
+Fig~\ref{figure:ranger_rbo_Stodden_simulation.png} shows the OOB prediction error and the RBO error for a random forest with 10000 trees and the \mtry value set
 to the default for a regression problem (depending on $n$). It shows no evidence of following the shape of the
@@
 % \end{figure}
+
+\begin{figure}[tbhp] 
+    \begin{subfigure}[b]{0.5\linewidth}
+      \centering
+      \includegraphics[totalheight=6cm]{./figs/k.png}
+      \caption{$k$}
+      \label{figure:k.png}
+    \end{subfigure} 
+    \begin{subfigure}[b]{0.5\linewidth}
+      \centering
+      \includegraphics[totalheight=6cm]{./figs/n.png}
+      \caption{$n$}
+      \label{figure:n.png}
+    \end{subfigure} 
+    \caption{The simulation in \protect{\cite{Donoho.and.Stodden.2006}} considers $\delta = n/p$ and $\rho =k/n$. Here we plot the
+      space of $\{ \delta, \rho$\}, colored  by the parameters $n$ and $k$}
+\end{figure}

### 2017-02-15 doc/publications/plos16/plos16.tex
changes to the conclusion

@@
 
-\cite{Donoho.and.Stodden.2006} investigate the behavior of a number of
-regression approaches for variable selection in a small simulation.  They consider a regression problem with 
+\cite{Donoho.and.Stodden.2006} investigate the phase transition in a small simulation.  They consider a regression problem with 
 $\underset{n\times p}{X}\sim N(0,1)$ with $p$ fixed at 200 and $n$ variable.
@@
 N(0,\sqrt{16})$. They evaluate variable selection by the error measure 
-$$\frac{\Vert\hat{\beta}-\beta\Vert_2}{\Vert\beta\Vert_2}$$
+$$\frac{\Vert\hat{\beta}-\beta\Vert_2}{\Vert\beta\Vert_2}.$$
 
@@
 the variable with the maximum $t$-value to the linear model if the $p$-value is less than 
-$25(\text{number of  terms  currently  in  the  model})/(\text{total  number  of  variables})$.  
+$0.25(\text{number of  terms  currently  in  the  model})/(\text{total  number  of  variables})$.  
 See Fig~\ref{figure:error_Stodden_FDR.png} for the error measure and
@@
 We have seen no work on the question of such a phase transition in the case of classification problems. Still it may be
-instructive to consider the implications of the phase transition for our work. As \cursedforest is designed for
-extremely large numbers of variables it is likely to be operating in difficult regions of the figure where the ratio
-$\delta = n/p$ is small. In the simulated example in section \ref{section:synthetic_data} the underdetermination $\delta = 0.002$ and the
-sparsity $\rho =2\times 10^{-6}$
+instructive to consider the implications of the phase transition for our work.
 
@@
 
-We note that while a random forest is a long way from an all-subsets search, it is a limited combinatorial search of
+We note that while a random forest is a long way from an all-subsets search, it is a limited search of
 the feature space. As such it may perform outside of the bounds of the phase transition.
-  combinatorial search.
 
@@
 
-As \cursedforest is designed for extremely large numbers of variables it is likely to be operating in difficult regions of the figure.
-We note that the examples we are considering lie on the extreme left of the plot, close to the origin. So there are
-limits to what can be recovered in the presence of noise. However in the case of data that is both big and wide,
-\cursedforest and other VariantSpark methods may provide a useful tool.
-
-The existence of the  Donoho-Tanner phase transition is a salutary warning. There are likely to be limits, both
-computational and logical, to the recovery of signals from noisy data. \cursedforest  is a contribution to addressing the
-practical limits but the logical limits will still apply. However in the case of data that is both big and wide,
-\cursedforest and other VariantSpark methods may provide a useful tool.
+As \cursedforest is designed for extremely large numbers of variables it is likely to be operating in difficult regions
+of the figure where the ratio $\delta = n/p$ is small. In the simulated example in section \ref{section:synthetic_data}
+the underdetermination $\delta = 0.002$ and the sparsity $\rho =2\times 10^{-6}.$ The existence of the Donoho-Tanner
+phase transition is a salutary warning. There are likely to be limits, both computational and logical, to the recovery
+of signals from noisy data. \cursedforest is a contribution to addressing the practical limits but the logical limits
+will still apply. However in the case of data that is both big and wide, \cursedforest and other VariantSpark methods
+may provide a useful tool.
 

### 2017-02-17 doc/publications/plos16/supplementary_information.tex
added H2O times to sulementary info

@@
                                                 & ranger - c++     & 0.06       & 7 mins           & --nthreads 20                                                    \\
+                                                & H2O           & 0.06       & 5 hours         & 600GB \footnote{java -Xmx600g -jar h2o.jar,   h2o.init(nthreads=20)} \\
 

### 2017-02-17 doc/publications/plos16/plos16.tex
updated BMD section with PRODH as novel gene

@@
 \subsection{GWAS-style analysis using \cursedforest\ identifies novel variants associated with Bone Mineral Density}
-In this section we use \cursedforest\ to perform feature selection on over 7.2 million genomic variants to identify the
-locations associated with Bone Mineral Density (BMD)~\cite{Duncan.2011}. We build a model on the BMD dataset as
-described in Methods using a binary high or low BMD value as the label.  From this model, we output the top 10,000
-contributing variants, ranked by their Gini impurity. The variant with the highest Gini impurity, and hence importance,
-is ranked 1, the lowest 10,000.
-
-To test whether the associated features have biological meaning, we investigate whether variants in or near
-BMD-associated genes are enriched toward the top of our ranked variant list. To form the list of "known genes", we
-combine the 26 known and 6 novel BMD genes listed by Duncan {\it et al.}~\cite{Duncan.2011} with genes identified by
-Estrada {\it et al.}~\cite{Estrada2012, Kemp2014,Phang.2008}, resulting in 102 genes.  All genes are covered in the
+In this section we use \cursedforest\ to perform feature selection on over 7.2 million genomic variants and identify the
+locations associated with Bone Mineral Density (BMD)~\cite{Duncan.2011}. 
+%We build a model on the BMD dataset as described in Methods using a binary high or low BMD value as the label.  
+We rank variants by their Gini impurity, with the most important one ranked at position 1, and return the top 10,000 variants.
+
+To test whether the associated variants have biological meaning, we investigate whether variants in or near
+BMD-associated genes are enriched toward the top of our ranked list. To form the list of "known genes", we
+combine the 26 known and 6 novel BMD genes listed by Duncan {\it et al.}~\cite{Duncan.2011} with the genes identified by
+Estrada {\it et al.}~\cite{Estrada2012} and Lui {\it et al.}~\cite{Liu2008}, resulting in 102 genes.  All genes are covered in the
 10,000 extracted variants and the gene name is assigned to a variant if it falls within 100 kb of the gene.
@@
 As shown in Fig~\ref{figure:ranksumtest}, the known genes are ranked significantly higher than variants from other
-locations in the genome (Mann-Whitney U, $p=9.71\times10^{-15}$), supporting our hypothesis that biologically relevant
+locations in the genome (Mann-Whitney U, $p=1.83\times10^{-14}$), supporting our hypothesis that biologically relevant
 features are contributing more information to our classification model.
 
-The gene with the highest-ranked features is COLEC10 (collectin 10) as shown in Fig~\ref{figure:knowngeneranks}, a
-well established contributor to BDM~\cite{Kemp2014,Liu2008} with verified transcription factor binding site for FOXL1,
-also a known BDM gene~\cite{DECODE}.  The second highest ranked gene is ZBTB40 on 22q11. Stagi {\it et
-  al.}~\cite{Stagi.2010}, observe that individuals with a micro-deletion in this region have a significant reduction in
-bone mass. In this same region, we also observe the next highest ranked gene, PRODH (proline dehydrogenase).  PRODH
-catalyses the first step in proline degradation, and as proposed by Phang {\it et al.}~\cite{Phang.2008}, proline is
-made available by collagen degradation.
-
-The lowest ranked gene with literature implied BMD association is DCDC5 (Doublecortin Domain Containing 5) ranked 9,830
+The two gene with the highest-ranked features are the known BMD associated genes COLEC10 (collectin 10) and ZBTB40 (Zinc Finger And BTB Domain-Containing Protein 40) as shown in Fig~\ref{figure:knowngeneranks}.
+The association for both gene has been replicated since in independent studies (COLEC10~\cite{Kemp2014} and ZBTB40~\cite{Nielson2016}).
+%with verified transcription factor binding site for FOXL1,
+%also a known BDM gene~\cite{DECODE}.  
+%The second highest ranked gene is ZBTB40, also a well established BMD associated gene~\cite{27476799}.
+
+In contrast, the lowest ranked gene with literature implied BMD association is DCDC5 (Doublecortin Domain Containing 5) ranked 9,830
 out of 10,000 (see Fig~\ref{figure:knowngeneranks}).  Similar to the conclusions drawn by Duncan {\it et
@@
 
-Hence focusing on the top 20 most associated variants, the majority are from five known BMD genes (left side of the
-vertical line in Fig~\ref{figure:knowngeneranks}), however there are also five novel locations as annotated in
-Fig~\ref{figure:ranksumtest}. Most notably the SNP in the intronic region of Unc13c (rs12916554) is linked to quality
+Hence focusing on the top 20 most associated variants, the majority are from the four known BMD genes (left side of the
+vertical line in Fig~\ref{figure:knowngeneranks}), however there are also six novel locations as annotated in
+Fig~\ref{figure:ranksumtest}. Most notably PRODH (proline dehydrogenase) located on 22q11, a loci observed by Stagi {\it et
+  al.}~\cite{Stagi.2010} to significantly reduce bone mass in individuals with a micro-deletion in this region. The variant in PRODH is the only 
+  highly associated feature in this region and not in linkage disequilibrium (LD) with another gene (see Fig~\ref{zoomplot}). PRODH
+catalyses the first step in proline degradation, and as proposed by Phang {\it et al.}~\cite{Phang.2008}, proline is
+made available by collagen degradation. Also amongst the top 20 associated variants is a SNP in the intronic region of Unc13c (rs12916554), a gene that is linked to quality
 of sleep~\cite{Nicod2016}, which in turn is reported to be the best predictor of osteoporosis risk~\cite{Magali2016}, a
@@
 \begin{figure}[tbhp]
-  \begin{adjustwidth}{-1.00in}{0in}
-    \caption{\textbf{Distribution of known BMD genes amongst the top 10,000 features}}
+  \begin{adjustwidth}{-2.00in}{0in}
+    \caption{\textbf{Importance of 10,000 top ranked features and visualisation of the 22q.11 region}}
     \label{figure:ranksumtest}
-    \includegraphics[totalheight=8cm]{./figs/BMDTop10K.pdf}
+     \begin{subfigure}[b]{0.5\linewidth}
+    \includegraphics[totalheight=7cm]{./figs/BMDTop10K.pdf}
+     \caption{Distribution of known BMD genes amongst the top 10,000 features.} 
+      \label{figure:rbo-prod.png} 
+%      \vspace{4ex}
+    \end{subfigure} 
+ \begin{subfigure}[b]{0.5\linewidth}
+      \centering
+      \includegraphics[totalheight=7cm]{./figs/EURchr22_18-20mb.pdf}
+      \caption{Zoom plot of the 22q.11 region centred on PRODH.} 
+      \label{zoomplot} 
+%      \vspace{4ex}
+    \end{subfigure} 
     \begin{flushleft}
-      The figure shows the top 10,000 features by their rank and Gini impurity score (Importance).
+      (a) The figure shows the top 10,000 features by their rank and Gini impurity score (Importance).
 The line denotes the top 20 features which contains five novel loci annotated with text.
 The $p$-value is the Mann-Whitney U test.
+      (b) Each point represents a variant with the normalised Gini impurity score on the left y axis and LD in colour with the highest $r^2$ in red.  
      \end{flushleft}

### 2017-02-17 doc/publications/plos16/supplementary_information.tex
supplement updated

--- 2017-02-14 doc/publications/plos16/supplementary_information.tex
@@
 
-\multirow{3}{*}{phase1\_chr22: 1,092 x 490,036} & \cursedforest\ & 0.077 (XX) &                  &                                                                   \\
-                                                & RF - Spark ML  &            &                  &                                                                   \\
-                                                & RF - R         & 0.05       & 24 min           & 18GB\footnote{using the \texttt{parallel} packages with 10 nodes} \\
-                                                & ranger - R      & 0.06       & 1 min 10 seconds & 16GB  (one processor)                                             \\
-                                                & ranger - c++     & 0.06       & 7 mins           & --nthreads 20                                                     \\
+\multirow{3}{*}{phase1\_chr22: 1,092 x 490,036} & \cursedforest & 0.04  & 1 min 50 seconds  &  \\ %110 %\cursedforest\ & 0.077 (XX) &                  &                                                                   \\
+                                                & RF - Spark ML  &  0.05           &           7 min 44 seconds        &  \\  %464  %&            &                  &                                                                   \\
+                                                & RF - R         & 0.05       & 24 min           & 18GB\footnote{using the \texttt{parallel} packages with 10 nodes} \\ %1440
+                                                & ranger - R      & 0.06       & 1 min 10 seconds & 16GB  (one processor)                                             \\ %70
+                                                & ranger - c++     & 0.06       & 7 mins           & --nthreads 20                                                    \\  %420
+                                                & H2O           & 0.06       & 5 hours         & 600GB \footnote{java -Xmx600g -jar h2o.jar,   h2o.init(nthreads=20)} \\ %18000
 
  \hline
-\multirow{3}{*}{phase3\_chr22: 2,504 x 1,103,548}   &               &            &                  &                                                                \\
-                                                    & RF - R        & NA         & NA               & NA\footnote{long vectors ($> 2^31-1$)  are not supported in the .Fortran interface} \\
-                                                    & H2O           & 0.05       & 16 hours         & 600GB \footnote{java -Xmx600g -jar h2o.jar,   h2o.init(nthreads=20)} \\
-                                                    & ranger - R    & 0.06       & 4 1/2 mins       & 21GB (one processor) \\
+\multirow{3}{*}{phase3\_chr22: 2,504 x 1,103,548}   & \cursedforest & 0.06 & 4 min 2 seconds & \\  %240 &               &            &                  &                                                                \\
+						  & RF - Spark ML   & 0.06 & 37 min 19 seconds  & \\  %2239
+                                                    & RF - R        & NA         & NA               & NA\footnote{long vectors ($> 2^31-1$)  are not supported in the .Fortran interface} \\ %OK
+                                                    & ranger - R    & 0.06       & 4 30 seconds       & 21GB (one processor) \\ %270
+                                                    & ranger - c++ & ?? & ?? &   \\
+			                           & H2O           & 0.05       & 16 hours         & 600GB \footnote{java -Xmx600g -jar h2o.jar,   h2o.init(nthreads=20)} \\  %57600
 \hline
-\multirow{3}{*}{phase3\_chr1: 2,504 x 6,450,364}    & \cursedforest\ & 0.06 & 1hr 58min        & 4GB \\
-                                                    & RF - Spark ML  & 0.06       & 8hr 29min        & 8GB \\
-                                                    & LR - Spark ML  &            & 2hr 6min         & 8GB \\
-                                                    & ranger - R      & 0.05       & 28 mins          & num.threads=10\footnote{don't
-                                                                                                      know how to get the memory
-                                                                                                      used,   This was done on dpas-03} \\
+\multirow{3}{*}{phase3\_chr1: 2,504 x 6,450,364}    & \cursedforest & 0.04  & 19 min 19 seconds &        \\ %1159   \cursedforest\ & 0.06 & 1hr 58min        & 4GB \\
+                                                    & RF - Spark ML  &  0.04       & 8 hour 29 min   &     \\ % 3403 0.06       & 8hr 29min        & 8GB \\
+  %                                                  & LR - Spark ML  &            & 2hr 6min         & 8GB \\
+                                                   & ranger - R      & 0.05       & 28 mins          & num.threads=10 \\ %\footnote{don't know how to get the memory used, This was done on dpas-03} \\ %1680
+                                                   & ranger - c++ & ?? & ?? &  \\
+                                                    &H2O & NA & NA &   \\
 \hline
-\multirow{3}{*}{phase3\_chr1-3: 2,504 x 19,328,051} & \cursedforest\ & 0.04 & 4hr 0min                                             & 4GB \\
-                                                    & RF - Spark ML  &            &                                                      & \\
-                                                    & LR - Spark ML  &            & 14hr 48min                                           & 8GB \\
-                                                    & ranger - R      &  0.03      & 58 mins\footnote{ but took 6 hours to load the
-                                                                                   data}
-                                                                                                                                         &
-                                                                                                                                           909GB\footnote{I
-                                                                                                                                           think
-                                                                                                                                           this
-                                                                                                                                           is
-                                                                                                                                           total
-                                                                                                                                           memory. The
-                                                                                                                                           job
-                                                                                                                                           was
-                                                                                                                                           run
-                                                                                                                                           with
-                                                                                                                                                   ntasks-per-node=160
-                                                                                                                                           cores-per-socket=10
-                                                                                                                                           on
-                                                                                                                                           ruby
-                                                                                                                                           }  \\
-& ranger - R        &             0.027     &      03:58           &       pearcey --cpus-per-task=15 --mem 1600gb   906GB \\
+\multirow{3}{*}{phase3\_chr1-3: 2,504 x 19,328,051} & \cursedforest & 0.02  & 56 min 43 seconds     &        \\ %3403%& 0.04 & 4hr 0min                                             & 4GB \\
+                                                    & RF - Spark ML  &  NA           &  NA                                                     & \\ %OK
+  %                                                  & LR - Spark ML  &            & 14hr 48min                                           & 8GB \\
+  						 & ranger - R      & 0.03      &    3 hour 58 min  &   \\ %14280
+
+						 %DENIS: is it 6 h 58 m or the above quoted 3 h 58 m ? 
+ %                                                   & ranger - R      &  0.03      & 58 mins\footnote{ but took 6 hours to load the data}
+ %                                                                                                                                        & 909GB\footnote{I
+  %                                                                                                                                         think
+  %                                                                                                                                         this
+  %                                                                                                                                         is
+  %                                                                                                                                         total
+   %                                                                                                                                        memory. The
+   %                                                                                                                                        job
+   %                                                                                                                                        was
+   %                                                                                                                                        run
+   %                                                                                                                                        with
+    %                                                                                                                                               ntasks-per-node=160
+     %                                                                                                                                      cores-per-socket=10
+      %                                                                                                                                     on
+       %                                                                                                                                    ruby
+       %                                                                                                                                    }  \\
+						%DENIS: our best result is on this subset is 10 min 49 seconds using 130 worker nodes, each with 8GB RAM so are you sure about the ranger result?
+%						& ranger - R        &             0.027     &      03:58           &       pearcey --cpus-per-task=15 --mem 1600gb   906GB \\
+						& ranger - c++ & ?? & ?? &  \\
 \hline
-
-\multirow{4}{*}{phase3\_chr1-22: 2,504 x 81,047,467} & \cursedforest\ & 0.04 & 7hr 14min & 8GB \\
-& RF - Spark ML & - & - & - \\
-& LR - Spark ML & - & - & - \\ 
-& kmeans - Spark ML  & 0.18      & 30hr 44min    & 24GB \\ 
-& ranger (C++)       &        NA     &        NA     &            80 CPU -- runs out of memory (12.8GB per CPU) \\
-& ranger (C++)       &        NA     &        NA     &            160
-                                                        CPU -- runs
-                                                        out of time at
-                                                        20 hours \\
+\multirow{4}{*}{phase3\_chr1-22: 2,504 x 81,047,467} & \cursedforest\ & 0.01  & 3 hour 16min & \\% \\ %11760 %& 0.04 & 7hr 14min & 8GB \\
+%& kmeans - Spark ML  & 0.18      & 30hr 44min    & 24GB \\ 
+										& ranger - R       &        NA     &        NA   & \\
+										& ranger - c++       &        NA     &        NA     &            80 CPU -- runs out of memory (12.8GB per CPU) \\
+										& ranger - c++       &        NA     &        NA     &            160 CPU -- runs out of time at 20 hours \\
 \hline

### 2017-02-19 doc/publications/plos16/supplementary_information.tex
added to table in supp file

@@
                                                     & ranger - R    & 0.06       & 4 30 seconds       & 21GB (one processor) \\ %270
-                                                    & ranger - c++ & ?? & ?? &   \\
+                                                    & ranger - c++ & 0.05  & 63 min &   \\
 			                           & H2O           & 0.05       & 16 hours         & 600GB \footnote{java -Xmx600g -jar h2o.jar,   h2o.init(nthreads=20)} \\  %57600
@@
                                                    & ranger - R      & 0.05       & 28 mins          & num.threads=10 \\ %\footnote{don't know how to get the memory used, This was done on dpas-03} \\ %1680
-                                                   & ranger - c++ & ?? & ?? &  \\
+                                                   & ranger - c++ & 0.04 & 5 hours 27 min &  \\
                                                     &H2O & NA & NA &   \\

### 2017-02-19 doc/publications/plos16/supplementary_information.tex
changes to table in supp

--- 2017-02-17 doc/publications/plos16/supplementary_information.tex
@@
                                                     & RF - R        & NA         & NA               & NA\footnote{long vectors ($> 2^31-1$)  are not supported in the .Fortran interface} \\ %OK
-                                                    & ranger - R    & 0.06       & 4 30 seconds       & 21GB (one processor) \\ %270
-                                                    & ranger - c++ & ?? & ?? &   \\
+                                                    & ranger - R    & 0.06       & 18 min        & 21GB (one processor) \\ %270
+                                                    & ranger - c++ & 0.05  & 63 min &   \\
 			                           & H2O           & 0.05       & 16 hours         & 600GB \footnote{java -Xmx600g -jar h2o.jar,   h2o.init(nthreads=20)} \\  %57600
@@
   %                                                  & LR - Spark ML  &            & 2hr 6min         & 8GB \\
-                                                   & ranger - R      & 0.05       & 28 mins          & num.threads=10 \\ %\footnote{don't know how to get the memory used, This was done on dpas-03} \\ %1680
-                                                   & ranger - c++ & ?? & ?? &  \\
+                                                   & ranger - R      & 0.05       & 1 hour 42 min          & num.threads=10 \\ %\footnote{don't know how to get the memory used, This was done on dpas-03} \\ %1680
+                                                   & ranger - c++ & 0.04 & 5 hours 27 min &  \\
                                                     &H2O & NA & NA &   \\
@@
 %						& ranger - R        &             0.027     &      03:58           &       pearcey --cpus-per-task=15 --mem 1600gb   906GB \\
-						& ranger - c++ & ?? & ?? &  \\
+						& ranger - c++ & 0.04 & 8 hours 54 min &  \\
 \hline

### 2017-02-19 doc/publications/plos16/supplementary_information.tex
changes to table in supp

@@
                                                 & RF - Spark ML  &  0.05           &           7 min 44 seconds        &  \\  %464  %&            &                  &                                                                   \\
-                                                & RF - R         & 0.05       & 24 min           & 18GB\footnote{using the \texttt{parallel} packages with 10 nodes} \\ %1440
+                                                & randomForest - R         & 0.05       & 24 min           & 18GB\footnote{using the \texttt{parallel} packages with 10 nodes} \\ %1440
                                                 & ranger - R      & 0.06       & 1 min 10 seconds & 16GB  (one processor)                                             \\ %70
@@
 						  & RF - Spark ML   & 0.06 & 37 min 19 seconds  & \\  %2239
-                                                    & RF - R        & NA         & NA               & NA\footnote{long vectors ($> 2^31-1$)  are not supported in the .Fortran interface} \\ %OK
+                                                    & randomForest - R        & NA         & NA               & NA\footnote{long vectors ($> 2^31-1$)  are not supported in the .Fortran interface} \\ %OK
                                                     & ranger - R    & 0.06       & 18 min        & 21GB (one processor) \\ %270

### 2017-02-19 doc/publications/plos16/supplementary_information.tex
updated 1000 project section and updated figure with new runtimes, also cleaned out supp table

@@
                                                     & randomForest - R        & NA         & NA               & NA\footnote{long vectors ($> 2^31-1$)  are not supported in the .Fortran interface} \\ %OK
-                                                    & ranger - R    & 0.06       & 18 min        & 21GB (one processor) \\ %270
-                                                    & ranger - c++ & 0.05  & 63 min &   \\
+                                                    & ranger - R    & 0.06       & 18 min        & 21GB (one processor) \\ %1080
+                                                    & ranger - c++ & 0.05  & 1 hour 3 min &   \\  %3780
 			                           & H2O           & 0.05       & 16 hours         & 600GB \footnote{java -Xmx600g -jar h2o.jar,   h2o.init(nthreads=20)} \\  %57600
@@
   %                                                  & LR - Spark ML  &            & 2hr 6min         & 8GB \\
-                                                   & ranger - R      & 0.05       & 1 hour 42 min          & num.threads=10 \\ %\footnote{don't know how to get the memory used, This was done on dpas-03} \\ %1680
-                                                   & ranger - c++ & 0.04 & 5 hours 27 min &  \\
+                                                   & ranger - R      & 0.05       & 1 hour 42 min          & num.threads=10  \\ %6120
+                                                   & ranger - c++ & 0.04 & 5 hours 27 min &  \\ %19620
                                                     &H2O & NA & NA &   \\
@@
   						 & ranger - R      & 0.03      &    3 hour 58 min  &   \\ %14280
-
-						 %DENIS: is it 6 h 58 m or the above quoted 3 h 58 m ? 
- %                                                   & ranger - R      &  0.03      & 58 mins\footnote{ but took 6 hours to load the data}
- %                                                                                                                                        & 909GB\footnote{I
-  %                                                                                                                                         think
-  %                                                                                                                                         this
-  %                                                                                                                                         is
-  %                                                                                                                                         total
-   %                                                                                                                                        memory. The
-   %                                                                                                                                        job
-   %                                                                                                                                        was
-   %                                                                                                                                        run
-   %                                                                                                                                        with
-    %                                                                                                                                               ntasks-per-node=160
-     %                                                                                                                                      cores-per-socket=10
-      %                                                                                                                                     on
-       %                                                                                                                                    ruby
-       %                                                                                                                                    }  \\
-						%DENIS: our best result is on this subset is 10 min 49 seconds using 130 worker nodes, each with 8GB RAM so are you sure about the ranger result?
-%						& ranger - R        &             0.027     &      03:58           &       pearcey --cpus-per-task=15 --mem 1600gb   906GB \\
-						& ranger - c++ & 0.04 & 8 hours 54 min &  \\
+						& ranger - c++ & 0.04 & 8 hours 54 min &  \\ %32040
 \hline

### 2017-02-19 doc/publications/plos16/supplementary_information.tex
small change

--- 2017-02-19 doc/publications/plos16/supplementary_information.tex
@@
   %                                                  & LR - Spark ML  &            & 2hr 6min         & 8GB \\
-                                                   & ranger - R      & 0.05       & 1 hour 42 min          & num.threads=10 \\ %\footnote{don't know how to get the memory used, This was done on dpas-03} \\ %1680
+                                                   & ranger - R      & 0.05       & 1 hour 42 min          & num.threads=10 \\ 
                                                    & ranger - c++ & 0.04 & 5 hours 27 min &  \\

### 2017-02-19 doc/publications/plos16/plos16.tex

@@
 We record the prediction error (OOB) as well as the runtime for each method. 
-%In a previous paper, we clustered individuals from the 1000 Genomes Project using k-means clustering \cite{OBrien2015}.
-%Although we were able to cluster individuals on all 80 million variants, it was relatively slow, with the entire
-%process (pre-processing and clustering) taking over 30 hours.
-%We investigate building random forest models on this data. However, we observed this process to not only be slower
-%than building a k-means model, but does not scale to as many dimensions (see Table~\ref{timingtable}). 
-%@Aidan 
-We limit the number of executors to X that are available to \cursedforest\ to keep the comparison fair for all methods.
+%@Aidan is 12 correct?
+We limit the number of executors to 12 that are available to \cursedforest\ to keep the comparison fair for all methods.
 
 As shown in Fig.~\ref{figure:1000genomes}, only \cursedforest\ is able to scale to the full phase3\_chr1-22 dataset (2,504 samples x 19,328,051 variants), with
-the standard random forest implementation in R and ranger's C++ version running out of resources on the relatively low-dimensional
+the standard random forest implementation in R running out of resources on the relatively low-dimensional
 phase1\_chr22 dataset (1,092 x 490,036). 
-Ranger's R implementation is able to process the second largest dataset (phase3\_chr1-3 with 2,504 x 19,328,051), however, for larger datasets the runtime increases beyond a reasonable compute time for this task (>20h).
+Ranger's c++ implementation and the R front-end version of it are able to process the second largest dataset (phase3\_chr1-3 with 2,504 x 19,328,051), however, both are unable to process larger datasets successfully.
+While the c++ implementation runs out of memory the R front-end, which also covers the loading and pre-processing, needs an unreasonable compute time for this task (>20h), see Supplementary Table 1.
 The standard implementation of random forest in SparkML already runs out of memory after processing the third largest dataset, phase3\_chr1 (2,504 x 6,450,364). 
@@
 
-%With the decreasing costs of genome sequencing, this case will surely become increasingly
-%common.
-All tested methods have a worse runtime than \cursedforest\ for the datasets they processed successfully, except for the smallest dataset where the overhead of Spark's distributed platform made Ranger's R implementation faster with 1 min 10 seconds compared to 1 min 50 seconds.
-For example, on phase3\_chr1-3 \cursedforest\ finishes in 1h, with Ranger(R) requiring 4h and SparkML already needing 8 hours on a smaller dataset.
-Extrapolating from this, all comparable implementations hence would require more than 100 days to finish, assuming the use of larger memory nodes.
-
-%with the ranger implementation in R and SparkML 
-%On the relatively low-dimensional
-%phase1\_chr22 dataset (1,092 samples x 490,036 variants), we observed a runtime of under 8 minutes. Although slower than Ranger, 
-%presumably due to the overhead of Spark's distributed platform on this relatively small dataset, we observe Spark ML to
-%be about 16 minutes faster than R. Moving to the larger dataset, phase3\_chr1 (2,504 x 6,450,364), we observe an increase
-%in Spark ML's runtime to over 8 hours. Moving to our next largest dataset, phase3\_chr1\-3 (2,504 x 19,328,051), and the
-%job fails to complete within our arbitrary cutoff time of 20 hours.
-
-%From our trials, it was apparent that Spark ML, although designed to handle many samples, is unable to efficiently cope
-%with a huge number of features. With the decreasing costs of genome sequencing, this case will surely become increasingly
-%common.
-
-%\cursedforest solves this problem of high-dimensionality genomic data. Not only is it faster on the above smaller datasets,
-%but it is also able to scale to the entire genome.
+All tested methods have a worse runtime than \cursedforest\ for the datasets they processed successfully, except for the smallest dataset where the overhead of Spark's distributed platform made ranger's R implementation faster with 1 min 10 seconds compared to 1 min 50 seconds.
+For example, on phase3\_chr1-3 \cursedforest\ finishes in 1 hour, with ranger(R) requiring 4 hours and SparkML already needing 8 hours on a smaller dataset.
+Extrapolating from this, all comparable implementations hence would require more than 24 hours to finish, assuming the use of larger memory nodes.
 
 As our implementation can efficiently utilise large numbers of commodity nodes, the performance of \cursedforest\ can be substantially improved further. 
-Table~\ref{cursedforesttable} shows the runtime when utilising the full Spark cluster (12 worker nodes).
-
-%As we previously mentioned, building a random forest model on the phase3\_chr1 dataset with Spark ML takes over 8 hours.
+%@Aidan is 130 correct?
+Table~\ref{cursedforesttable} shows the runtime when utilising the full Spark cluster (130 worker nodes).
+
 Building a model with \cursedforest\ on phase3\_chr1, which previously took 20 minutes, can be reduced to 4 minutes. 
@@
 The error can be reduced from OOB=0.04 when using only data from Chr1, which represents processing power limit for most tools, down to OOB=0.01 when using the full dataset.
-
-%than twice as fast as building a random
-%forest on the chromosome 1 (phase3\_chr1) using Spark ML.
 
@@
 
-\iffalse
-
-\begin{table}[!ht]
-\begin{minipage}{\textwidth}
-%\begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
-\centering
-\caption{ {\bf Performance comparison between the different random forest implementations.}}
-\begin{tabular}{| l | l | r | l |}
-\hline
-\bf{Genome \%}                      & \bf{Method} & \bf{Error}  & \bf{Runtime}  \\
-
-\hline
-
-\multirow{5}{*}{phase1\_chr22: 1,092 x 490,036} & \cursedforest & 0.04  & 1 min 50 seconds    \\ %110
-                                                & Spark ML  &  0.05           &           7 min 44 seconds         \\  %464
-                                                & \randomforest (R)         & 0.05       & 24 min           \\ %1440
-                                                & \ranger (R)      & 0.06       & 1 min 10 seconds       \\ %70
-                                                & \ranger (C++)     & 0.06       & 7 min            \\ %420
-                                                & H2O           & ??       & ??         \\
-
-\hline
-
-\multirow{5}{*}{phase3\_chr22: 2,504 x 1,103,548}   & \cursedforest & 0.06 & 4 min 2 seconds \\  %240
-                                                    & Spark ML  & 0.06 & 37 min 19 seconds \\  %2239
-                                                    & \randomforest (R)        & NA         & NA                \\
-                                                    & \ranger (R)    & 0.06       & 4 min 30 seconds     \\  %270
-                                                    & \ranger (C++)     &     ??   & ??        \\
-                                                    & H2O           & 0.05       & 16 hours         \\ %57600
-
-\hline
-
-\multirow{3}{*}{phase3\_chr1: 2,504 x 6,450,364}    & \cursedforest & 0.04  & 19 min 19 seconds         \\ %1159
-                                                    & Spark ML  & 0.04       & 8 hour 29 min        \\ % 3403
-                                                    & \randomforest (R)        & NA         & NA                \\
-                                                    & \ranger (R)      &  0.05       & 28 mins         \\ %1680
-                                                    & \ranger (C++)     & ??       & ??            \\
-                                                    & H2O           & NA       & NA         \\
-\hline
-
-\multirow{3}{*}{phase3\_chr1-3: 2,504 x 19,328,051} & \cursedforest & 0.02  & 56 min 43 seconds             \\ %3403
-                                                    & Spark ML &    NA         &   NA               \\ %30540
-                                                    & \randomforest (R)        & NA         & NA                \\
-                                                    & \ranger (R)      & 0.03      &    3 hour 58 min     \\ %14280
-                                                    & \ranger (C++)     & ??       & ??            \\
-                                                    & H2O           & NA       & NA         \\
-\hline
-
-\multirow{4}{*}{phase3\_chr1-22: 2,504 x 81,047,467} & \cursedforest & 0.01  & 3hour 16min \\ %11760
-                                                    & Spark ML & NA & NA  \\
-                                                    & \randomforest (R)        & NA         & NA                \\
-                                                    & \ranger (R)       &        NA     &        NA    \\
-                                                    & \ranger (C++)       &        NA     &        NA    \\
-                                                    & H2O           & NA       & NA         \\
-\hline
-
-\end{tabular}
-\begin{flushleft} 
-Comparison of different various machine learning libraries on different subsets of variant data 
-from the 1000 Genomes Project.
-Each random forest model consists of 50 trees.
-\end{flushleft}
-\label{timingtable}
-%\end{adjustwidth}
-\end{minipage}
-\end{table}
-
-\fi
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Section 4: Sensitivity 
+% Section 4: Sensitivity to different parameters and scalability
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2017-02-19 doc/publications/plos16/supplementary_information.tex
small change to table in supp

--- 2017-02-19 doc/publications/plos16/supplementary_information.tex
@@
 \multirow{3}{*}{phase1\_chr22: 1,092 x 490,036} & \cursedforest & 0.04  & 1 min 50 seconds  &  \\ %110 %\cursedforest\ & 0.077 (XX) &                  &                                                                   \\
-                                                & RF - Spark ML  &  0.05           &           7 min 44 seconds        &  \\  %464  %&            &                  &                                                                   \\
-                                                & randomForest - R         & 0.05       & 24 min           & 18GB\footnote{using the \texttt{parallel} packages with 10 nodes} \\ %1440
+                                                & RF - Spark ML  &  0.05           &           7 mins 44 seconds        &  \\  %464  %&            &                  &                                                                   \\
+                                                & randomForest - R         & 0.05       & 24 mins           & 18GB\footnote{using the \texttt{parallel} packages with 10 nodes} \\ %1440
                                                 & ranger - R      & 0.06       & 1 min 10 seconds & 16GB  (one processor)                                             \\ %70
@@
  \hline
-\multirow{3}{*}{phase3\_chr22: 2,504 x 1,103,548}   & \cursedforest & 0.06 & 4 min 2 seconds & \\  %240 &               &            &                  &                                                                \\
-						  & RF - Spark ML   & 0.06 & 37 min 19 seconds  & \\  %2239
+\multirow{3}{*}{phase3\_chr22: 2,504 x 1,103,548}   & \cursedforest & 0.06 & 4 mins 2 seconds & \\  %240 &               &            &                  &                                                                \\
+						  & RF - Spark ML   & 0.06 & 37 mins 19 seconds  & \\  %2239
                                                     & randomForest - R        & NA         & NA               & NA\footnote{long vectors ($> 2^31-1$)  are not supported in the .Fortran interface} \\ %OK
-                                                    & ranger - R    & 0.06       & 18 min        & 21GB (one processor) \\ %1080
+                                                    & ranger - R    & 0.06       & 18 mins        & 21GB (one processor) \\ %1080
                                                     & ranger - c++ & 0.05  & 1 hour 3 min &   \\  %3780
@@
 \hline
-\multirow{3}{*}{phase3\_chr1: 2,504 x 6,450,364}    & \cursedforest & 0.04  & 19 min 19 seconds &        \\ %1159   \cursedforest\ & 0.06 & 1hr 58min        & 4GB \\
-                                                    & RF - Spark ML  &  0.04       & 8 hour 29 min   &     \\ % 3403 0.06       & 8hr 29min        & 8GB \\
+\multirow{3}{*}{phase3\_chr1: 2,504 x 6,450,364}    & \cursedforest & 0.04  & 19 mins 19 seconds &        \\ %1159   \cursedforest\ & 0.06 & 1hr 58min        & 4GB \\
+                                                    & RF - Spark ML  &  0.04       & 8 hours 29 mins   &     \\ % 3403 0.06       & 8hr 29min        & 8GB \\
   %                                                  & LR - Spark ML  &            & 2hr 6min         & 8GB \\
-                                                   & ranger - R      & 0.05       & 1 hour 42 min          & num.threads=10  \\ %6120
-                                                   & ranger - c++ & 0.04 & 5 hours 27 min &  \\ %19620
+                                                   & ranger - R      & 0.05       & 1 hour 42 mins          & num.threads=10  \\ %6120
+                                                   & ranger - c++ & 0.04 & 5 hours 27 mins &  \\ %19620
                                                     &H2O & NA & NA &   \\
 \hline
-\multirow{3}{*}{phase3\_chr1-3: 2,504 x 19,328,051} & \cursedforest & 0.02  & 56 min 43 seconds     &        \\ %3403%& 0.04 & 4hr 0min                                             & 4GB \\
+\multirow{3}{*}{phase3\_chr1-3: 2,504 x 19,328,051} & \cursedforest & 0.02  & 56 mins 43 seconds     &        \\ %3403%& 0.04 & 4hr 0min                                             & 4GB \\
                                                     & RF - Spark ML  &  NA           &  NA                                                     & \\ %OK
   %                                                  & LR - Spark ML  &            & 14hr 48min                                           & 8GB \\
-  						 & ranger - R      & 0.03      &    3 hour 58 min  &   \\ %14280
-						& ranger - c++ & 0.04 & 8 hours 54 min &  \\ %32040
+  						 & ranger - R      & 0.03      &    3 hours 58 mins  &   \\ %14280
+						& ranger - c++ & 0.04 & 8 hours 54 mins &  \\ %32040
 \hline
-\multirow{4}{*}{phase3\_chr1-22: 2,504 x 81,047,467} & \cursedforest\ & 0.01  & 3 hour 16min & \\% \\ %11760 %& 0.04 & 7hr 14min & 8GB \\
+\multirow{4}{*}{phase3\_chr1-22: 2,504 x 81,047,467} & \cursedforest\ & 0.01  & 3 hours 16min & \\% \\ %11760 %& 0.04 & 7hr 14min & 8GB \\
 %& kmeans - Spark ML  & 0.18      & 30hr 44min    & 24GB \\ 

### 2017-02-19 doc/publications/plos16/plos16.tex
small change cption in paper

@@
       $\bigotimes$ marks the last dataset the different method completed before running out of memory or running more than
-      24 hours. The dashed lines mark the trajectory for how ranger and Spark ML had performed given larger memory machines.
+      20 hours. The dashed lines mark the trajectory for how \ranger and Spark ML had performed given larger memory machines.
     \end{flushleft}

### 2017-02-21 doc/publications/plos16/plos16.tex
removed margin note

@@
 
-\setlength{\marginparwidth}{5cm}
+%\setlength{\marginparwidth}{5cm}
 \usepackage{caption}
@@
 
-This may be because a large \mtry\ generates more correlated trees as the same important features have a higher chance
-of being selected in all, not yielding good performance outcomes.  This issue is less pronounced for classification
-tasks where random features can mimic the response variable, hence resulting more in a performance plateau.  
-\marginpar{unclear what this means}
-Increasing
-the number of trees on the other hand improves performance especially when the trees are kept diverse (small \mtry) but
+This may be because a large \mtry leads to more correlated trees as the same important features have a higher chance of
+being selected in all, not yielding good performance outcomes.  This issue is less pronounced for classification error
+where random features can mimic the response variable, hence resulting more in a performance plateau.  Increasing the
+number of trees on the other hand improves performance especially when the trees are kept diverse (small \mtry) but
 appropriate for large feature datasets (\mtry larger than default). \cursedforest's feature of parallelizing trees at
@@
 
-The two gene with the highest-ranked features are the known BMD associated genes COLEC10 (collectin 10) and ZBTB40 (Zinc Finger And BTB Domain-Containing Protein 40) as shown in Fig~\ref{figure:knowngeneranks}.
-The association for both gene has been replicated since in independent studies (COLEC10~\cite{Kemp2014} and ZBTB40~\cite{Nielson2016}).
+The two gene with the highest-ranked features are the known BMD associated genes COLEC10 (collectin 10) and ZBTB40 (Zinc
+Finger And BTB Domain-Containing Protein 40) as shown in Fig~\ref{figure:knowngeneranks}.  The association for both gene
+has been replicated since in independent studies (COLEC10~\cite{Kemp2014} and ZBTB40~\cite{Nielson2016}).
 %with verified transcription factor binding site for FOXL1,
@@
 vertical line in Fig~\ref{figure:knowngeneranks}), however there are also six novel locations as annotated in
-Fig~\ref{figure:ranksumtest}. Most notably PRODH (proline dehydrogenase) located on 22q11, a loci observed by Stagi {\it et
-  al.}~\cite{Stagi.2010} to significantly reduce bone mass in individuals with a micro-deletion in this region. The variant in PRODH is the only 
-  highly associated feature in this region and not in linkage disequilibrium (LD) with another gene (see Fig~\ref{zoomplot}). PRODH
-catalyses the first step in proline degradation, and as proposed by Phang {\it et al.}~\cite{Phang.2008}, proline is
-made available by collagen degradation. Also amongst the top 20 associated variants is a SNP in the intronic region of Unc13c (rs12916554), a gene that is linked to quality
-of sleep~\cite{Nicod2016}, which in turn is reported to be the best predictor of osteoporosis risk~\cite{Magali2016}, a
+Fig~\ref{figure:ranksumtest}. Most notably PRODH (proline dehydrogenase) located on 22q11, a loci observed by Stagi {\it
+  et al.}~\cite{Stagi.2010} to significantly reduce bone mass in individuals with a micro-deletion in this region. The
+variant in PRODH is the only highly associated feature in this region and not in linkage disequilibrium (LD) with
+another gene (see Fig~\ref{zoomplot}). PRODH catalyses the first step in proline degradation, and as proposed by Phang
+{\it et al.}~\cite{Phang.2008}, proline is made available by collagen degradation. Also amongst the top 20 associated
+variants is a SNP in the intronic region of Unc13c (rs12916554), a gene that is linked to quality of
+sleep~\cite{Nicod2016}, which in turn is reported to be the best predictor of osteoporosis risk~\cite{Magali2016}, a
 condition diagnosed by measuring BMD.  Also listed is rs1035665, which was identified in a recent GWAS study to regulate
@@
 
-\cursedforest\ can successfully be used in GWAS-type analysis replicating previous findings and uncovering novel associations due to the ability of random forests to pick up interacting features.
+\cursedforest\ can successfully be used in GWAS-type analysis replicating previous findings and uncovering novel
+associations due to the ability of random forests to pick up interacting features.
 
@@
     \begin{flushleft}
-      (a) The figure shows the top 10,000 features by their rank and Gini impurity score (Importance).
-The line denotes the top 20 features which contains five novel loci annotated with text.
-The $p$-value is the Mann-Whitney U test.
-      (b) Each point represents a variant with the normalised Gini impurity score on the left y axis and LD in colour with the highest $r^2$ in red.  
+      (a) The figure shows the top 10,000 features by their rank and Gini impurity score (Importance). The line denotes
+      the top 20 features which contains five novel loci annotated with text. The $p$-value is the Mann-Whitney U test.
+      (b) Each point represents a variant with the normalised Gini impurity score on the left y axis and LD in colour
+      with the highest $r^2$ in red.
      \end{flushleft}

### 2017-02-21 doc/publications/plos16/plos16.tex
changes to conclusion

@@
 We have demonstrated that using a different parallelization model can extend random forests to the case of an
-extremely large number of variables. We have treated the case of variable selection in a $p >> n$ model where most of
-the variables are uninformative and have demonstrated the utility of the model for large GWAS datasets. By comparing
+extremely large number of variables. We have treated the case of variable selection in a $p >> n$ model, where most of
+the variables are uninformative, and have demonstrated the utility of the model for large GWAS datasets. By comparing
 this implementation to other implementations (including those optimized for large datasets) we have demonstrated the
@@
 Donoho and Tanner \cite{Donoho.and.Tanner.2009} give a ``universal phase change'' result that has applications in a
-large number of areas including variable selection in high dimensions. They show that there is a sharp disjunction
-between the cases where significant variables may be recovered with a high accuracy by procedures like stepwise variable
-selection, and the cases where they can not be recovered. The boundary is shown in
-Fig~\ref{figure:phase-diagram-equivalence.png} in a space parameterized by the level of underdetermination,
-$\delta = n/p$, and by the sparsity, $\rho =k/n$ (where $k$ is the number of significant variables).  Above the
+large number of areas, including variable selection in high dimensions. The theoretical phase change boundary is based on arguments
+from combinatorial geometry.
+
+They argue that there is a sharp disjunction between the cases where informative variables may be recovered with a high
+accuracy by procedures like stepwise variable selection, and the cases where they can not be recovered. The boundary is
+shown in Fig~\ref{figure:phase-diagram-equivalence.png} in a space parameterized by the level of underdetermination,
+$\delta = n/p$, and by the sparsity, $\rho =k/n$ (where $k$ is the number of informative variables).  Above the
 phase-transition line variable recovery is still possible by a combinatorial search approach such as all-subsets
-variable selection.  The theoretical boundary is based on arguments from combinatorial geometry.
+variable selection.
 
@@
 
-\cite{Donoho.and.Stodden.2006} investigate the phase transition in a small simulation.  They consider a regression problem with 
-$\underset{n\times p}{X}\sim N(0,1)$ with $p$ fixed at 200 and $n$ variable.
-They  set $\beta(1:k) \sim U(1,100)$ and $\beta((k+1):p) =0$, then  $y= X\beta + \epsilon$, $\epsilon \sim
-N(0,\sqrt{16})$. They evaluate variable selection by the error measure 
+\cite{Donoho.and.Stodden.2006} investigate the phase transition in a small simulation.  They consider a regression
+problem with $\underset{n\times p}{X}\sim N(0,1)$ with $p$ fixed at 200 and $n$ variable.  They set
+$\beta(1:k) \sim U(1,100)$ and $\beta((k+1):p) =0$, then $y= X\beta + \epsilon$ with $\epsilon \sim N(0,\sqrt{16})$. They
+evaluate variable selection by the error measure
 $$\frac{\Vert\hat{\beta}-\beta\Vert_2}{\Vert\beta\Vert_2}.$$
@@
      \caption{Forward stepwise variable selection with a false discovery rate stopping criteria.}
-       \label{figure:error_Stodden_FDR.png}
+       \label{figure:error_and_rbo_Stodden_FDR.png}
 %   \end{adjustwidth}
@@
      \caption{The simulated data of \protect{\cite{Donoho.and.Stodden.2006}} with a random forest model.}
-     \label{figure:ranger_rbo_Stodden_simulation.png}
+     \label{figure:ranger_error_and_rbo_Stodden_simulation.png}
 %   \end{adjustwidth}
@@
 
-We have used a random forest on the same simulation as figure \ref{figure:error_Stodden_FDR.png}.  The Donoho-Tanner phase transition arises in
-recovering the $\beta$ in data generated by a linear model. However, in a decision tree (random forest) we are fitting a
-non-linear model and there is no notion of estimating the $\beta$. Because of this we have evaluated the performance of
-the random forest using the RBO measure on the variable importance.
+We have used a random forest on the same simulation data used for figure \ref{figure:error_and_rbo_Stodden_FDR.png}.  The
+Donoho-Tanner phase transition arises in recovering the $\beta$ in data generated by a linear model. However, in a
+decision tree (random forest) we are fitting a non-linear model and there is no notion of estimating the
+$\beta$. Because of this we have evaluated the performance of the random forest using the RBO measure on the variable
+importance.
 
@@
 
-Fig~\ref{figure:ranger_rbo_Stodden_simulation.png} shows the OOB prediction error and the RBO error for a random forest with 10000 trees and the \mtry value set
-to the default for a regression problem (depending on $n$). It shows no evidence of following the shape of the
-Donoho-Tanner phase transition.
+Fig~\ref{figure:ranger_error_and_rbo_Stodden_simulation.png} shows the OOB prediction error and the RBO error for a
+random forest with 10000 trees and the \mtry value set to the default for a regression problem (depending on $n$). It
+shows no evidence of following the shape of the Donoho-Tanner phase transition.
 
@@
 of the figure where the ratio $\delta = n/p$ is small. In the simulated example in section \ref{section:synthetic_data}
-the underdetermination $\delta = 0.002$ and the sparsity $\rho =2\times 10^{-6}.$ The existence of the Donoho-Tanner
-phase transition is a salutary warning. There are likely to be limits, both computational and logical, to the recovery
-of signals from noisy data. \cursedforest is a contribution to addressing the practical limits but the logical limits
-will still apply. However in the case of data that is both big and wide, \cursedforest and other VariantSpark methods
-may provide a useful tool.
+the underdetermination parameter $\delta = 0.002$ and the sparsity parameter $\rho =2\times 10^{-6}.$
+
+The existence of the Donoho-Tanner phase transition is a salutary warning. There are likely to be limits, both
+computational and logical, to the recovery of signals from noisy data. \cursedforest is a contribution to addressing the
+practical limits but the logical limits will still apply. However in the case of data that is both big and wide,
+\cursedforest and other VariantSpark methods may provide a useful tool.
 

### 2017-03-09 doc/publications/plos16/plos16.tex
updated BMI section with new model

@@
 %We build a model on the BMD dataset as described in Methods using a binary high or low BMD value as the label.  
-We rank variants by their Gini impurity, with the most important one ranked at position 1, and return the top 10,000 variants.
-
-To test whether the associated variants have biological meaning, we investigate whether variants in or near
+We rank variants by their Gini impurity, with the most important one ranked at position 1, and return the top 100,000 variants.
+
+To test whether the associated variants have biological meaning, we investigate if variants in or near
 BMD-associated genes are enriched toward the top of our ranked list. To form the list of "known genes", we
 combine the 26 known and 6 novel BMD genes listed by Duncan {\it et al.}~\cite{Duncan.2011} with the genes identified by
-Estrada {\it et al.}~\cite{Estrada2012} and Lui {\it et al.}~\cite{Liu2008}, resulting in 102 genes.  All genes are covered in the
-10,000 extracted variants and the gene name is assigned to a variant if it falls within 100 kb of the gene.
-
-As shown in Fig~\ref{figure:ranksumtest}, the known genes are ranked significantly higher than variants from other
-locations in the genome (Mann-Whitney U, $p=1.83\times10^{-14}$), supporting our hypothesis that biologically relevant
-features are contributing more information to our classification model.
-
-The two gene with the highest-ranked features are the known BMD associated genes COLEC10 (collectin 10) and ZBTB40 (Zinc
-Finger And BTB Domain-Containing Protein 40) as shown in Fig~\ref{figure:knowngeneranks}.  The association for both gene
-has been replicated since in independent studies (COLEC10~\cite{Kemp2014} and ZBTB40~\cite{Nielson2016}).
-%with verified transcription factor binding site for FOXL1,
-%also a known BDM gene~\cite{DECODE}.  
-%The second highest ranked gene is ZBTB40, also a well established BMD associated gene~\cite{27476799}.
-
-In contrast, the lowest ranked gene with literature implied BMD association is DCDC5 (Doublecortin Domain Containing 5) ranked 9,830
-out of 10,000 (see Fig~\ref{figure:knowngeneranks}).  Similar to the conclusions drawn by Duncan {\it et
+Estrada {\it et al.}~\cite{Estrada2012} and Lui {\it et al.}~\cite{Liu2008}, resulting in 102 genes.  All genes are covered within the
+top 40,000 variants and a gene name is assigned to the variant if it falls within 100 kb of this gene.
+
+The known genes rank significantly higher than variants from other locations in the genome (Mann-Whitney U, $p=4.5\times10^{-14}$, Supplemental File 1), 
+supporting our hypothesis that biologically relevant features are contributing more information to our classification model.
+
+As shown in Fig~\ref{figure:knowngeneranks}, the top 30 features are dominated by variants from five well established BMD genes; TNFRSF11B (TNF Receptor Superfamily Member 11b or Osteoprotegerin), CTNNB (Catenin Beta 1), ZBTB40 (Zinc Finger And BTB Domain-Containing Protein 40), IBSP (Integrin Binding Sialoprotein) and MEF2C (Myocyte Enhancer Factor 2C).
+[MORE ABOUT THESE GENES].
+All SNPs in linkage disequilibrium (LD) with the top ranking one also show a high importance score (Fig 3b).
+%~\ref{manhattenzoom}).
+%has been replicated since in independent studies (COLEC10~\cite{Kemp2014} and ZBTB40~\cite{Nielson2016}).
+
+%MAPT microtubule associated protein tau \cite{Dengler-Crish2017}
+%"MAPT is primarily expressed in the nervous system and has not been identified in bone tissue, therefore, low BMD in a tauopathy-specific mouse model that lacks A?
+pathology would provide evidence that central mechanisms of bone homeostasis are dysfunctional in AD, rather than only direct peripheral effects on bone tissue itself."
+%"Reduced bone mineral density (BMD) and its clinical sequelae, osteoporosis, occur at a much greater rate in patients with Alzheimer?s disease (AD), often emerging early in the disease before significant cognitive decline is seen.
+"
+
+In contrast, one of the lowest ranked gene with literature implied BMD association is DCDC5 (Doublecortin Domain Containing 5) ranked 23,842
+(see Fig 3a).
+%~\ref{figure:knowngeneranks}).  
+Similar to the conclusions drawn by Duncan {\it et
   al.}~\cite{Duncan.2011}, our results also indicate that this gene does not have a strong association with BMD.  In
@@
 to whether the gene itself contributes to BMD.
-
-Hence focusing on the top 20 most associated variants, the majority are from the four known BMD genes (left side of the
-vertical line in Fig~\ref{figure:knowngeneranks}), however there are also six novel locations as annotated in
-Fig~\ref{figure:ranksumtest}. Most notably PRODH (proline dehydrogenase) located on 22q11, a loci observed by Stagi {\it
+Similarly, the lowest ranked gene MAPT (microtubule associated protein tau) at 36,262 is primarily expressed in the nervous system and has not been identified in bone tissue. 
+It is therefore likely that the association is a co-morbidity with Alzheimer's disease~\cite{Dengler-Crish2017} and not a major contributing factor in this cohort.
+
+%2 TEX2  EN rs112558362 SNORNA  ERN1 Gliablastoma in older people Alzheimer's
+%5 rs6961378  Woolfe2005
+%13 rs17195090
+
+The top 30 most associated variants are dominated by the known BMD genes, however there are also four novel locations as annotated in
+Fig 3a.
+%~\ref{figure:rbo-prod.png}. 
+Most notably PRODH (proline dehydrogenase) located on 22q11, a loci observed by Stagi {\it
   et al.}~\cite{Stagi.2010} to significantly reduce bone mass in individuals with a micro-deletion in this region. The
-variant in PRODH is the only highly associated feature in this region and not in linkage disequilibrium (LD) with
-another gene (see Fig~\ref{zoomplot}). PRODH catalyses the first step in proline degradation, and as proposed by Phang
-{\it et al.}~\cite{Phang.2008}, proline is made available by collagen degradation. Also amongst the top 20 associated
-variants is a SNP in the intronic region of Unc13c (rs12916554), a gene that is linked to quality of
-sleep~\cite{Nicod2016}, which in turn is reported to be the best predictor of osteoporosis risk~\cite{Magali2016}, a
-condition diagnosed by measuring BMD.  Also listed is rs1035665, which was identified in a recent GWAS study to regulate
-ABCF1 (ATP Binding Cassette Subfamily F Member 1)~\cite{Dixon2007} with member 2 of this subfamily being a known BMD
-gene. The other three novel loci are on or near ncRNA.
-
-\cursedforest\ can successfully be used in GWAS-type analysis replicating previous findings and uncovering novel
-associations due to the ability of random forests to pick up interacting features.
+variant in PRODH is the only highly associated feature in this region and all LD SNPs show high importance (see Fig 3c).%~\ref{zoomplot}). 
+PRODH catalyses the first step in proline degradation, and as proposed by Phang
+{\it et al.}~\cite{Phang.2008}, proline is made available by collagen degradation.
+Ranked at position 5 (rs6961378) and 18 (rs146034056) are located on chromosome 7 overlapping the lincRNA (LINC01006), which as been linked to vertebrate development~\cite{Woolfe2005}.
+In contrast rs17195090 and rs112558362 are single feature loci (Supplementary Figure 2 and 3) hence likely due to erroneous imputation or genotyping.
+
+% Also amongst the top 20 associated
+%variants is a SNP in the intronic region of Unc13c (rs12916554), a gene that is linked to quality of
+%sleep~\cite{Nicod2016}, which in turn is reported to be the best predictor of osteoporosis risk~\cite{Magali2016}, a
+%condition diagnosed by measuring BMD.  Also listed is rs1035665, which was identified in a recent GWAS study to regulate
+%ABCF1 (ATP Binding Cassette Subfamily F Member 1)~\cite{Dixon2007} with member 2 of this subfamily being a known BMD
+%gene. The other three novel loci are on or near ncRNA.
+
+\cursedforest\ can successfully perform GWAS-type analyses replicating previous findings and uncovering novel
+associations due to random forests also modelling interacting features.
 
@@
   \begin{adjustwidth}{-2.00in}{0in}
-    \caption{\textbf{Importance of 10,000 top ranked features and visualisation of the 22q.11 region}}
-    \label{figure:ranksumtest}
-     \begin{subfigure}[b]{0.5\linewidth}
-    \includegraphics[totalheight=7cm]{./figs/BMDTop10K.pdf}
-     \caption{Distribution of known BMD genes amongst the top 10,000 features.} 
-      \label{figure:rbo-prod.png} 
-%      \vspace{4ex}
-    \end{subfigure} 
- \begin{subfigure}[b]{0.5\linewidth}
-      \centering
-      \includegraphics[totalheight=7cm]{./figs/EURchr22_18-20mb.pdf}
-      \caption{Zoom plot of the 22q.11 region centred on PRODH.} 
-      \label{zoomplot} 
-%      \vspace{4ex}
-    \end{subfigure} 
-    \begin{flushleft}
-      (a) The figure shows the top 10,000 features by their rank and Gini impurity score (Importance). The line denotes
-      the top 20 features which contains five novel loci annotated with text. The $p$-value is the Mann-Whitney U test.
-      (b) Each point represents a variant with the normalised Gini impurity score on the left y axis and LD in colour
-      with the highest $r^2$ in red.
-     \end{flushleft}
-  \end{adjustwidth}
-\end{figure}
-
-\begin{figure}[tbhp]
-  \begin{adjustwidth}{-2.00in}{0in}
     \caption{\textbf{Rank of literature assigned BMD genes}}
@@
 \end{figure}
+
+\begin{figure}[tbhp]
+  \begin{adjustwidth}{-2.00in}{0in}
+    \caption{\textbf{Genome-wide importance of SNPs and visualisation of the 22q.11 region}}
+    \label{figure:ranksumtest}
+     \begin{subfigure}[b]{0.5\linewidth}
+     \includegraphics[totalheight=7cm]{./figs/BMDTopWholeGenome.png}
+      \label{figure:rbo-prod.png} 
+    \end{subfigure} 
+ \begin{subfigure}[b]{0.5\linewidth}
+      \centering
+      \includegraphics[totalheight=7cm]{./figs/new_10k_TNFRSF11B.pdf}
+      \label{zoomplotTNFRSF} 
+   \end{subfigure} 
+    \begin{subfigure}[b]{0.5\linewidth}
+     \includegraphics[totalheight=7cm]{./figs/new_10k_PRODH.pdf}
+      \label{zoomplotprodh} 
+    \end{subfigure} 
+        \begin{subfigure}[b]{0.5\linewidth}
+     \includegraphics[totalheight=7cm]{./figs/EUR_chr7_156_2-156_6mb.pdf}
+      \label{zoomplotlinc} 
+    \end{subfigure} 
+    \begin{flushleft}
+      (a) The figure shows genome-wide SNPs and their importance. Highlighted are the 100 literature assigned BMD genes (green) and four novel loci. The horizontal line indicates the top 30 features. 
+      Locus Zoomplot for (b) the top ranked gene TNFRSF11B and the two putative novel associations (c) PRODH and (e) LINC01006.
+Each point represents a variant with the normalised Gini impurity score on the left y axis and LD to the central SNP in colour with the highest $r^2$ in red.
+     \end{flushleft}
+  \end{adjustwidth}
+        \label{manhattenzoom} 
+\end{figure}
+
+%\begin{figure}[tbhp]
+%  \begin{adjustwidth}{-2.00in}{0in}
+%    \caption{\textbf{Importance of 10,000 top ranked features and visualisation of the 22q.11 region}}
+%    \label{figure:ranksumtest}
+%     \begin{subfigure}[b]{0.5\linewidth}
+%    \includegraphics[totalheight=7cm]{./figs/BMDTop10K.pdf}
+%     \caption{Distribution of known BMD genes amongst the top 10,000 features.} 
+%      \label{figure:rbo-prod.png} 
+%%      \vspace{4ex}
+%    \end{subfigure} 
+% \begin{subfigure}[b]{0.5\linewidth}
+%      \centering
+%      \includegraphics[totalheight=7cm]{./figs/EURchr22_18-20mb.pdf}
+%      \caption{Zoom plot of the 22q.11 region centred on PRODH.} 
+%      \label{zoomplot} 
+%%      \vspace{4ex}
+%    \end{subfigure} 
+%    \begin{flushleft}
+%      (a) The figure shows the top 10,000 features by their rank and Gini impurity score (Importance). The line denotes
+%      the top 20 features which contains five novel loci annotated with text. The $p$-value is the Mann-Whitney U test.
+%      (b) Each point represents a variant with the normalised Gini impurity score on the left y axis and LD in colour
+%      with the highest $r^2$ in red.
+%     \end{flushleft}
+%  \end{adjustwidth}
+%\end{figure}
+
+%\begin{figure}[tbhp]
+%  \begin{adjustwidth}{-2.00in}{0in}
+%    \caption{\textbf{Importance of genome-wide SNPs}}
+%    \label{figure:knowngeneranks}
+%\includegraphics[totalheight=6.5cm]{./figs/BMDTopWholeGenome.pdf}
+%    \begin{flushleft}
+%      The figure shows genome-wide SNPs and their importance. Highlighted are the 100 literature assigned BMD genes. The horizontal line indicates the top 20. 
+%     \end{flushleft}
+%  \end{adjustwidth}
+%\end{figure}
 

### 2017-03-10 doc/publications/plos16/plos16.tex
updated BMD section

@@
 \\
-Aidan O'Brien\textsuperscript{1},
-Piotr Szul\textsuperscript{2},
-Robert Dunne\textsuperscript{3}, 
-Paul Leo \textsuperscript{4},
-Emma L. Duncan \textsuperscript{4,5,6},
-Stephanie Li\textsuperscript{7},
-James Doecke\textsuperscript{7},
-Nick Ellis\textsuperscript{8}, and
+Aidan O'Brien\textsuperscript{1,2},
+Piotr Szul\textsuperscript{3},
+Robert Dunne\textsuperscript{4}, 
+Paul Leo \textsuperscript{5},
+Emma L. Duncan \textsuperscript{5,6,7},
+Natalie Twine\textsuperscript{1,8}
+Stephanie Li\textsuperscript{9},
+James Doecke\textsuperscript{9},
+Nick Ellis\textsuperscript{10}, and
 Denis C. Bauer\textsuperscript{1,*}
@@
 \\
-\textbf{2} Data61, CSIRO, Brisbane, QLD, Australia
+\textbf{2} ANU, Canberra, Australia
 \\
-\textbf{3} Data61, CSIRO, Sydney, NSW, Australia
+\textbf{3} Data61, CSIRO, Brisbane, QLD, Australia
 \\
-\textbf{4} Institute of Health and Biomedical Innovation, Queensland University of Technology, Brisbane, QLD, Australia
+\textbf{4} Data61, CSIRO, Sydney, NSW, Australia
 \\
-\textbf{5} Department of Endocrinology, Royal Brisbane and Women's Hospital, Brisbane, QLD, Australia
+\textbf{5} Institute of Health and Biomedical Innovation, Queensland University of Technology, Brisbane, QLD, Australia
 \\
-\textbf{6} Faculty of Medicine and Biomedical Sciences, University of Queensland, Brisbane, QLD, Australia
+\textbf{6} Department of Endocrinology, Royal Brisbane and Women's Hospital, Brisbane, QLD, Australia
 \\
-\textbf{7} Health \& Biosecurity, CSIRO, Brisbane, QLD, Australia
+\textbf{7} Faculty of Medicine and Biomedical Sciences, University of Queensland, Brisbane, QLD, Australia
 \\
-\textbf{8} Oceans \& Atmosphere, CSIRO, Brisbane, QLD, Australia
+\textbf{8} MQ, QLD, Australia
+\\
+\textbf{9} Health \& Biosecurity, CSIRO, Brisbane, QLD, Australia
+\\
+\textbf{10} Oceans \& Atmosphere, CSIRO, Brisbane, QLD, Australia
 \bigskip
@@
 
-As shown in Fig~\ref{figure:knowngeneranks}, the top 30 features are dominated by variants from five well established BMD genes; TNFRSF11B (TNF Receptor Superfamily Member 11b or Osteoprotegerin), CTNNB (Catenin Beta 1), ZBTB40 (Zinc Finger And BTB Domain-Containing Protein 40), IBSP (Integrin Binding Sialoprotein) and MEF2C (Myocyte Enhancer Factor 2C).
-[MORE ABOUT THESE GENES].
-All SNPs in linkage disequilibrium (LD) with the top ranking one also show a high importance score (Fig 3b).
+As shown in Figure~\ref{figure:knowngeneranks}, the top 30 features are dominated by variants from five well established BMD genes; TNFRSF11B (TNF Receptor Superfamily Member 11b or Osteoprotegerin), CTNNB (Catenin Beta 1), ZBTB40 (Zinc Finger And BTB Domain-Containing Protein 40), IBSP (Integrin Binding Sialoprotein) and MEF2C (Myocyte Enhancer Factor 2C).
+The BMD association has been independently replicated for four of the five genes: CTNNB1~\cite{Pei2016, Rivadeneira2009}, MEF2C~\cite{Pei2016a}, TNFRSF11B~\cite{Rivadeneira2009,Zhang2014,Kemp2014,Paternoster2013} and ZBTB40~\cite{Rivadeneira2009,  Zhang2014,Nielson2016}.
+The association of IBSP has not been replicated in other GWAS studies, but homozygous knockouts of this gene have a known skeletal phenotype in mice.
+Mice deficient in IBSP have impaired bone growth and mineralization, with dramatically reduced bone formation~\cite{Malaval2008}.
+CTNBB1, MEF2C and TNFRSF11B also have known skeletal phenotypes.
+CTNNB1 is a key downstream member of the Wnt signaling pathway, and deletion of CTNNB1 in mice results in severe osteopenia, while constitutive activation of CTNNB1 results in massively increased bone deposition~\cite{Holmen2005}.
+MEF2C is a transcription factor which controls bone mass by regulating osteoclastic bone resorption, where mice with a MEF2C deletion in osteocytes had increased bone mass compared with normal controls~\cite{Kramer2012}.
+TNFRSF11B (or OPG) is a secreted glycoprotein which regulates bone mass~\cite{Simonet1997}.
+ZBTB40 has no known biological or molecular involvement in skeletal phenotypes~\cite{Styrkarsdottir2008}, even though it has been associated with BMD in multiple GWAS and is expressed in bone.
+SNPs in linkage disequilibrium (LD) with each other at these loci also show a high importance score, as exemplified by TNFRSF11B in Figure 3b (Supplemental Figure 2 and 3 for the other genes).
 %~\ref{manhattenzoom}).
-%has been replicated since in independent studies (COLEC10~\cite{Kemp2014} and ZBTB40~\cite{Nielson2016}).
 
@@
 In contrast, one of the lowest ranked gene with literature implied BMD association is DCDC5 (Doublecortin Domain Containing 5) ranked 23,842
-(see Fig 3a).
+(see Figure 2).
 %~\ref{figure:knowngeneranks}).  
 Similar to the conclusions drawn by Duncan {\it et
-  al.}~\cite{Duncan.2011}, our results also indicate that this gene does not have a strong association with BMD.  In
-fact, DCDC5 is not highly expressed in bone, and any mechanisms by which this gene might regulate BMD are unclear
+  al.}~\cite{Duncan.2011}, our results also indicate that this gene does not have a strong association with BMD.
+DCDC5 is not highly expressed in bone, and any mechanisms by which this gene might regulate BMD are unclear
 \cite{Thakker2012}.  Although variants near DCDC5 are associated with BMD \cite{Rivadeneira2009}, it is questionable as
 to whether the gene itself contributes to BMD.
-Similarly, the lowest ranked gene MAPT (microtubule associated protein tau) at 36,262 is primarily expressed in the nervous system and has not been identified in bone tissue. 
-It is therefore likely that the association is a co-morbidity with Alzheimer's disease~\cite{Dengler-Crish2017} and not a major contributing factor in this cohort.
+Similarly, the lowest ranked gene with annotated BMD links, MAPT (microtubule associated protein tau) at 36,262, is primarily expressed in the nervous system and has not been identified in bone tissue.
+It is therefore likely that the association is through a co-morbidity with Alzheimer's disease~\cite{Dengler-Crish2017} and not a major contributing factor in this cohort.
 
@@
 
-The top 30 most associated variants are dominated by the known BMD genes, however there are also four novel locations as annotated in
-Fig 3a.
+There are also four novel locations amongst the top 30 most associated variants as annotated in Figure 3a.
 %~\ref{figure:rbo-prod.png}. 
-Most notably PRODH (proline dehydrogenase) located on 22q11, a loci observed by Stagi {\it
-  et al.}~\cite{Stagi.2010} to significantly reduce bone mass in individuals with a micro-deletion in this region. The
-variant in PRODH is the only highly associated feature in this region and all LD SNPs show high importance (see Fig 3c).%~\ref{zoomplot}). 
-PRODH catalyses the first step in proline degradation, and as proposed by Phang
-{\it et al.}~\cite{Phang.2008}, proline is made available by collagen degradation.
-Ranked at position 5 (rs6961378) and 18 (rs146034056) are located on chromosome 7 overlapping the lincRNA (LINC01006), which as been linked to vertebrate development~\cite{Woolfe2005}.
-In contrast rs17195090 and rs112558362 are single feature loci (Supplementary Figure 2 and 3) hence likely due to erroneous imputation or genotyping.
+Most notably a lincRNA (LINC01006) on chromosome 7, which has been associated with serum creatinine levels in two independent GWAS studies~\cite{Pattaro2010,Chambers2010}. 
+Subjects with low serum creatinine levels were at a higher risk for low BMD~\cite{Huh2015} and in a rat study high creatine monohydrate diet resulted in greater lumbar BMD~\cite{Antolic2007}, 
+with creatinine being the breakdown product of creatine.
+The variants overlapping the 5' end of the lincRNA are the highest ranked features in this region and all LD SNPs also show high importance (see Fig 3d).
+Also amonst the top 30 features is PRODH (proline dehydrogenase) located on 22q11, a loci observed by Stagi {\it
+  et al.}~\cite{Stagi.2010} to significantly reduce bone mass in individuals with a micro-deletion in this region. 
+PRODH catalyses the first step in proline degradation, and as proposed by Phang {\it et al.}~\cite{Phang.2008}, proline is made available by collagen degradation. 
+Again, all LD SNPs also show high importance (see Figure 3c). %~\ref{zoomplot}). 
+%Ranked at position 5 (rs6961378) and 18 (rs146034056) are located on chromosome 7 overlapping the lincRNA (LINC01006), which as been linked to vertebrate development~\cite{Woolfe2005}.
+In contrast rs17195090 and rs112558362 are single feature loci (Supplementary Figure 4) hence likely due to erroneous imputation or genotyping.
 

### 2017-03-16 doc/publications/plos16/plos16.tex
added figures
small changes to paper

@@
 
-Analyses are performed on ever larger datasets with unprecedentedly detailed information per data point.
-Specifically in genomics, cohort sizes are in the thousands, each comprising of Millions of variants that make up a person's genetic differences.
-Traditional approaches are unable to cater for this task and even standard implementations using novel compute paradigms, such as Hadoop/Spark, are limited by the finite resource of computer memory on such data.
-Specifically, Spark ML's supervised machine learning methods suffer from "the curse of dimensionality".
-That is, they scale well with samples (n) but not with features (p, variants).
-
-We introduce \cursedforest, a tailored implementation of random forests, designed to handle data with extremely large number of variables per sample.
-We showcase \cursedforest\ to accurately predict ethnicity from whole genome sequencing profiles of 2,500 individuals and 80 million genomic variants each (1000 Genomes Project) in 30 minutes.
-\cursedforest\ was the only implementation scaling to the full dataset, outperforming other methods on smaller subsets by 80\% or more.
-We also demonstrate a GWAS-style analysis using  \cursedforest's feature selection approach on an imputed SNP array dataset (2000 x 7 Million) on Bone mineral density (BMD), successfully predicting known BMD-genes as well as identifying novel candidates.
-
-\cursedforest is included in our earlier variant interpretation package, VariantSpark, allowing it to perform near real-time classification of population-scale patient cohorts in "patients like mine" scenarios, as well as performing GWAS analysis on large unfiltered whole genome sequencing cohorts.
+Analyses are performed on ever larger datasets with unprecedentedly detailed information per data point. Specifically in
+genomics, cohort sizes are in the thousands, each comprising of millions of variants that make up a person's genetic
+differences. Traditional approaches are unable to cater for this task and even standard implementations using novel
+compute paradigms, such as Hadoop/Spark, are limited by the finite resource of computer memory on such
+data. Specifically, Spark ML's supervised machine learning methods suffer from "the curse of dimensionality". That is,
+they scale well with samples ($n$) but not with features ($p$, variants).
+
+We introduce \cursedforest, a tailored implementation of random forests, designed to handle data with extremely large
+number of variables per sample. We showcase \cursedforest\ to accurately predict ethnicity from whole genome sequencing
+profiles of 2,500 individuals and 80 million genomic variants each (1000 Genomes Project) in 30 minutes. \cursedforest\
+was the only implementation scaling to the full dataset, outperforming other methods on smaller subsets by 80\% or
+more. We also demonstrate a GWAS-style analysis using \cursedforest's feature selection approach on an imputed SNP array
+dataset (2000 x 7 Million) on Bone mineral density (BMD), successfully predicting known BMD-genes as well as identifying
+novel candidates.
+
+\cursedforest is included in our earlier variant interpretation package, VariantSpark, allowing it to perform near
+real-time classification of population-scale patient cohorts in "patients like mine" scenarios, as well as performing
+GWAS analysis on large unfiltered whole genome sequencing cohorts.
 
@@
 intermediate stages of building the model.  This unbalanced work load where the driver
-node becomes the bottleneck and worker nodes being idle prevents a seamless scaling to larger datasets. We also show that the memory
+node becomes the bottleneck and worker nodes are idle prevents a seamless scaling to larger datasets. We also show that the memory
 requirements per executor increases with dimensionality due to the data types Spark ML uses.%, which we elaborate on in the discussion.
@@
 of the set $S$ will be in a GWAS application. There are cases of small groups of SNPs with large effects and also of
-conditions caused by the interactions of large groups of SNPs.
+conditions caused by the interactions of larger groups of SNPs.
 
@@
 across multiple nodes, each vector must be present in its entirety on any node accessing it. This is no problem with
-typical datasets, however as dimensionality increases, the vectors eventually reach a size where they can no longer fit
+typical datasets; however, as dimensionality increases, the vectors eventually reach a size where they can no longer fit
 into a single node's memory.
@@
 Additive   & Each additional minor allele increases variation        & 0, 1, 2         \\
-Dominant   & Presence of at least 1 minor allele increases variation & 0, 1/2          \\
-Recessive  & Two minor alleles needed for variation                  & 0/1, 2          \\
-Heterosis  & Heterozygote leads to variation                         & 0/2, 1          \\ \hline
+Dominant   & Presence of at least 1 minor allele increases variation & 0, $\{1,2\}$          \\
+Recessive  & Two minor alleles needed for variation                  & $\{0,1\}$, 2          \\
+Heterosis  & Heterozygote leads to variation                         & $\{0,2\}$, 1          \\ \hline
 %\end{adjustwidth}
@@
     classification example. This may give an estimate with a higher bias and, as such, may be a poor choice. See the
-    discussion in \cite{Dietterich.2002} where it is shown that bagging may reduce bias as well as variance but give a
+    discussion in \cite{Dietterich.2002} where it is shown that the question of reducing bias versus variance is more
+    complicated than first thought. They show that bagging may reduce bias as well as variance but give a
     better result for low bias learners. \cite{Dietterich.2002} notes that if the bootstrap replicate approximation were
@@
 associated set of features amongst millions of dimensions; and its ability to accurately classify the labels of
-high-dimensional samples.  We therefore generate a synthetic dataset with 2.5 million variables ($p=2500000$), of which 5 
+high-dimensional samples.  We therefore generate a synthetic dataset with 2.5 million variables ($p=2,500,000$), of which 5 
 are related to the response variable, and 5000 samples ($n=5000$), as discussed in the methods section.
@@
 accuracy by procedures like stepwise variable selection, and the cases where they can not be recovered. The boundary is
-shown in Fig~\ref{figure:phase-diagram-equivalence.png} in a space parameterized by the level of underdetermination,
+shown in Fig~\ref{figure:phase-diagram-equivalence.png} in a space parameterized by the level of underdeterminatio,n
 $\delta = n/p$, and by the sparsity, $\rho =k/n$ (where $k$ is the number of informative variables).  Above the
-phase-transition line variable recovery is still possible by a combinatorial search approach such as all-subsets
+phase-transition line variable recovery is still possible by a combinatorial approach such as all-subsets
 variable selection.
@@
     \centering
-    \includegraphics[totalheight=6cm]{./figs/phase-diagram-equivalence.png} 
+    \includegraphics[totalheight=6cm]{./figs/phase2.png} 
     \caption{{\bf The phase change diagram.}}

### 2017-04-24 doc/publications/plos16/nm.tex
rewrite

--- 2017-03-10 doc/publications/plos16/plos16.tex
@@
 \setlength{\headheight}{27.023pt}
-\lhead{\includegraphics[width=2.0in]{PLOS-submission.eps}}
+
+%\lhead{\includegraphics[width=2.0in]{PLOS-submission.eps}}
 \rfoot{\thepage/\pageref{LastPage}}
@@
 \fancyfootoffset[L]{2.25in}
-\lfoot{\sf PLOS}
+%\lfoot{\sf PLOS}
 
@@
 Emma L. Duncan \textsuperscript{5,6,7},
-Natalie Twine\textsuperscript{1,8}
+Natalie A. Twine\textsuperscript{1,8}
 Stephanie Li\textsuperscript{9},
@@
 separately~\cite{CCC2007}.  It has since been demonstrated that not taking interactions between SNPs into account
-when filtering for potential disease mutations is inadequate~\cite{Manolio2009,Yang2011}.
+when filtering for potential disease mutations is inadequate~\cite{Manolio2009,Yang2011} and the approach should 
+in-fact be extended to interrogating all genomic positions[CITATION WES vs WGS?].
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Section 1:  Feature Selection and classification on synthetic data
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{\cursedforest can evaluate variable importance for millions of features}
-\label{section:synthetic}
-In this section we demonstrate the two features of \cursedforest, namely: variable selection analysis to select the most
-associated set of features amongst millions of dimensions; and its ability to accurately classify the labels of
-high-dimensional samples.  We therefore generate a synthetic dataset with 2.5 million variables ($p=2500000$), of which 5 
-are related to the response variable, and 5000 samples ($n=5000$), as discussed in the methods section.
-
-We fit the random forest model and estimate the classification accuracy by capturing the out-of-bag error (OOB) and the
-variable selection performance by capturing the rank-biased overlap (RBO) \cite{Webber2010} measure. RBO compares two
-lists (in this case, the parameters $w_i, i=1,\ldots , 5$ (Equation~\ref{equation:synthetic.data}), and the top 1000
-ranked variable importances from the \cursedforest model). It is a mesaure of similarity between rankings that handles
-lists of different lengths, weighs high ranks more heavily than low, and is monotonic with increasing depth of
-evaluation. We expect to retrieve the 5 features in order of the weight that was assigned to them and with a higher
-variable importance measure than the noise variables. RBO provides a quantiative measure of how close we come to this on
-a scale from 0 (no feature recovered) to 1 (fully recovered). See supplementary information (S*) for more details.
-
-We are running Apache Spark 1.6.1 on a YARN
-cluster with 12 worker nodes each with 16 Intel Xeon E5-2660@2.20GHz CPU cores and 128 GB of RAM.
-
-%mtry = number of variable seclection at each node - proportion of p 
-%ntree= number of trees
-
-% mtry large makes trees correlated because the same important features have a higher chance of being selected
-% overfitting because it picks random that happen to be
-
-As shown in Fig~\ref{figure:out-of-bag-prediction-error-prod.png} the default value for \mtry (number of variables
-selected at each node) does not result in a good classification performance for this large feature dataset. Note that
-the plot shows the proportion, $n/p,$ for \mtry\ with the default being $\sqrt{p} \approx 1581$ ($n/p=0.0006$).  The
-OOB for this value of \mtry does not drop below 0.5, even when the number of trees are increased (\ntree).
-
-Increasing \mtry\ in combination with \ntree\ yields the best performance with the OOB error essentially constant around 0.4
-across a large range of of \mtry and \ntree values.  This is in contrast to the feature-selection performance, where the
-RBO measure heavily depends on \ntree\ and gives better results with lower values of \mtry.
-
-This may be because a large \mtry leads to more correlated trees as the same important features have a higher chance of
-being selected in all, not yielding good performance outcomes.  This issue is less pronounced for classification error
-where random features can mimic the response variable, hence resulting more in a performance plateau.  Increasing the
-number of trees on the other hand improves performance especially when the trees are kept diverse (small \mtry) but
-appropriate for large feature datasets (\mtry larger than default). \cursedforest's feature of parallelizing trees at
-the node level hence caters perfectly to the requirement of large feature datasets as more trees can be built given a
-fixed time budget.
-
-%Fig~\ref{figure:synth1} illustrates the effect of varying \mtry and \ntree on these two measures. The default \mtry
-%value for a classification problem is $\sqrt{p} \approx 1581$ so in this case we are considering values considerably larger than
-%the default. Fig~\ref{figure:out-of-bag-prediction-error-prod.png} shows the OOB error is constant across a large range of 
-%of \mtry and \ntree values whereas Fig~\ref{figure:rbo-prod.png} illustrates that the RBO measure depends
-%heavily on \ntree but gives a better value with lower values of \mtry.
-
-From the synthetic data we can conclude that \cursedforest is able to perform accurate feature selection as well as
-accurate classification for large numbers of samples with large feature vectors.  We can also conclude that fine tuning
-the model to accurately classify the dataset does not ensure good feature selection performance as there seems to be
-little correspondence between the OOB and RBO measure.
-
-\begin{figure}[tbhp] 
-  \begin{adjustwidth}{-1.00in}{0in}
-    \caption{\textbf{Out of bag classification error and rank-biased-overlap estimates}} 
-    \label{figure:synth1}
-    \begin{subfigure}[b]{0.5\linewidth}
-      \centering
-      \includegraphics[totalheight=8cm]{./figs/oob_levleplot.png}
-      \caption{OOB classification error.} 
-      \label{figure:out-of-bag-prediction-error-prod.png} 
-%      \vspace{4ex}
-    \end{subfigure} 
-    \begin{subfigure}[b]{0.5\linewidth}
-      \centering
-      \includegraphics[totalheight=8cm]{./figs/rbo_levleplot.png}
-      \caption{The rank-biased-overlap measure.} 
-      \label{figure:rbo-prod.png} 
-%      \vspace{4ex}
-    \end{subfigure} 
-    \begin{flushleft} 
-      Random forests with 100 trees were trained on the synthetic dataset of 2.5M features and 5K samples.
-      Each measure is shown as a function of \mtry and \ntree. Note that the plot colors have been chosen so that the lighter color indicates
-      the better result, that is, a lower OOB or a higher RBO.
-    \end{flushleft}
-  \end{adjustwidth}
-\end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 % Section 2: Feature Selection on real data 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{GWAS-style analysis using \cursedforest\ identifies novel variants associated with Bone Mineral Density}
+%\subsection{GWAS-style analysis using \cursedforest\ identifies novel variants associated with Bone Mineral Density}
+\subsection{GWAS-style analyses using \cursedforest\ faithfully identify known disease-accociated genes}
 In this section we use \cursedforest\ to perform feature selection on over 7.2 million genomic variants and identify the
-locations associated with Bone Mineral Density (BMD)~\cite{Duncan.2011}. 
+genomic locations associated with Bone Mineral Density (BMD)~\cite{Duncan.2011}. 
 %We build a model on the BMD dataset as described in Methods using a binary high or low BMD value as the label.  
@@
 
-As shown in Figure~\ref{figure:knowngeneranks}, the top 30 features are dominated by variants from five well established BMD genes; TNFRSF11B (TNF Receptor Superfamily Member 11b or Osteoprotegerin), CTNNB (Catenin Beta 1), ZBTB40 (Zinc Finger And BTB Domain-Containing Protein 40), IBSP (Integrin Binding Sialoprotein) and MEF2C (Myocyte Enhancer Factor 2C).
-The BMD association has been independently replicated for four of the five genes: CTNNB1~\cite{Pei2016, Rivadeneira2009}, MEF2C~\cite{Pei2016a}, TNFRSF11B~\cite{Rivadeneira2009,Zhang2014,Kemp2014,Paternoster2013} and ZBTB40~\cite{Rivadeneira2009,  Zhang2014,Nielson2016}.
+%~\ref{figure:BMDTopWholeGenome}, 
+As shown in Figure 1a, the top 30 features are dominated by variants from five well established BMD genes; TNFRSF11B (TNF Receptor Superfamily Member 11b or Osteoprotegerin), CTNNB (Catenin Beta 1), ZBTB40 (Zinc Finger And BTB Domain-Containing Protein 40), IBSP (Integrin Binding Sialoprotein) and MEF2C (Myocyte Enhancer Factor 2C).
+The BMD association has been independently replicated for four of the five genes: TNFRSF11B~\cite{Rivadeneira2009,Zhang2014,Kemp2014,Paternoster2013}, CTNNB1~\cite{Pei2016, Rivadeneira2009}, ZBTB40~\cite{Rivadeneira2009,  Zhang2014,Nielson2016}, and MEF2C~\cite{Pei2016a}.
 The association of IBSP has not been replicated in other GWAS studies, but homozygous knockouts of this gene have a known skeletal phenotype in mice.
@@
 In contrast, one of the lowest ranked gene with literature implied BMD association is DCDC5 (Doublecortin Domain Containing 5) ranked 23,842
-(see Figure 2).
-%~\ref{figure:knowngeneranks}).  
-Similar to the conclusions drawn by Duncan {\it et
+(see Supplemental Figure 1). Similar to the conclusions drawn by Duncan {\it et
   al.}~\cite{Duncan.2011}, our results also indicate that this gene does not have a strong association with BMD.
@@
 
+\subsection{\cursedforest\ amplifies signal over traditional methods to identify novel associations}
+
+Figure 1b shows a traditional GWAS analysis using logistic regression~\cite{Duncan.2011} where the resulting $p$-values do not clearly distinguish these known BMD genes from the noise. 
+In contrast, \cursedforest\ seems to amplify the signal to identify associated features (Figure 1a). 
+As such, this approach allows to employ a very stringent cut-off for identifying strongly associated feature. Here we chose a 99.98 False Discovery Rate (FDR) from 60 random experiments (see Methods).
+
 %2 TEX2  EN rs112558362 SNORNA  ERN1 Gliablastoma in older people Alzheimer's
@@
 
-There are also four novel locations amongst the top 30 most associated variants as annotated in Figure 3a.
+Above this FDR cut-off we identify four novel locations as annotated in Figure 1a.
 %~\ref{figure:rbo-prod.png}. 
@@
 
+To evaluate whether the here identified features are significantly associated with BMD, we train a traditional linear mixed model to return a $p$-value for each included feature.
+Interestingly, while the three known BMD genes with the highest Importance score are indeed significantly associated (TNFRSF11B $p$-value=XX, CTNNB1 XX, ZBTB40 XX), IBSP and MEF2C were not significant.
+All of our newly associated features, on the other hand, were deemed significant, and in-fact rs17195090 was the most significant feature evaluated with $p$-value=XX (rs112558362 XX, LINC01006 XX, PRODH XX).
+
 % Also amongst the top 20 associated
@@
 
-\cursedforest\ can successfully perform GWAS-type analyses replicating previous findings and uncovering novel
-associations due to random forests also modelling interacting features.
+%\cursedforest\ can successfully perform GWAS-type analyses replicating previous findings and uncovering novel associations due to random forests also modelling interacting features.
 
@@
 
-\begin{figure}[tbhp]
-  \begin{adjustwidth}{-2.00in}{0in}
-    \caption{\textbf{Rank of literature assigned BMD genes}}
-    \label{figure:knowngeneranks}
-    \includegraphics[totalheight=6.5cm]{./figs/BMDgenes_landscape.pdf}
-    \begin{flushleft}
-      The figure shows the rank of the 100 literature assigned BMD genes.
+%\begin{figure}[tbhp]
+%  \begin{adjustwidth}{-2.00in}{0in}
+%    \caption{\textbf{Rank of literature assigned BMD genes}}
+%    \label{figure:knowngeneranks}
+%    \includegraphics[totalheight=6.5cm]{./figs/BMDgenes_landscape.pdf}
+%    \begin{flushleft}
+%      The figure shows the rank of the 100 literature assigned BMD genes.
 Each ball represents a SNP assigned to the closest gene with colour representing the distance (max 100 kb).
@@
 The horizontal line separates the genes with SNPs amongst the Top 20 associated features.
-     \end{flushleft}
-  \end{adjustwidth}
-\end{figure}
+%     \end{flushleft}
+%  \end{adjustwidth}
+%\end{figure}
 
@@
   \begin{adjustwidth}{-2.00in}{0in}
-    \caption{\textbf{Genome-wide importance of SNPs and visualisation of the 22q.11 region}}
+    \caption{\textbf{Genome-wide importance of SNPs for Bone Mineral Density association}}
     \label{figure:ranksumtest}
      \begin{subfigure}[b]{0.5\linewidth}
-     \includegraphics[totalheight=7cm]{./figs/BMDTopWholeGenome.png}
-      \label{figure:rbo-prod.png} 
+     \includegraphics[totalheight=7cm]{./figs/BMDTopWholeGenome_annotated.png}
+      \label{figure:BMDTopWholeGenome} 
     \end{subfigure} 
@@
       \centering
-      \includegraphics[totalheight=7cm]{./figs/new_10k_TNFRSF11B.pdf}
+      \includegraphics[totalheight=7cm]{./figs/BMDTopWholeGenome_BW_random.png}
       \label{zoomplotTNFRSF} 
@@
     \begin{subfigure}[b]{0.5\linewidth}
-     \includegraphics[totalheight=7cm]{./figs/new_10k_PRODH.pdf}
+    \includegraphics[totalheight=7cm]{./figs/new_10k_TNFRSF11B.pdf}
+%     \includegraphics[totalheight=7cm]{./figs/new_10k_PRODH.pdf}
       \label{zoomplotprodh} 
@@
     \begin{flushleft}
-      (a) The figure shows genome-wide SNPs and their importance. Highlighted are the 100 literature assigned BMD genes (green) and four novel loci. The horizontal line indicates the top 30 features. 
-      Locus Zoomplot for (b) the top ranked gene TNFRSF11B and the two putative novel associations (c) PRODH and (e) LINC01006.
+      Genome-wide SNPs and their importance for the BMD dataset {\bf(a)}; highlighted are the 100 literature assigned BMD genes (green) and four novel loci.
+The horizontal line indicates the top 99.98 FDR.
+      {\bf(b)} shows a previous GWAS analysis using logistic regression. 
+      Locus Zoomplot for {\bf(c)} the top ranked gene TNFRSF11B and the two putative novel associations {\bf(d)} LINC01006.
 Each point represents a variant with the normalised Gini impurity score on the left y axis and LD to the central SNP in colour with the highest $r^2$ in red.
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% Section 1:  Feature Selection and classification on synthetic data
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{\cursedforest can evaluate variable importance and perform classification for millions of features}
+%\label{section:synthetic}
+In this section we test the limits of \cursedforest\'s association testing and classify ability on high-dimensional data. 
+%In this section we demonstrate the two features of \cursedforest, namely: variable selection analysis to select the most
+%associated set of features amongst millions of dimensions; and its ability to accurately classify the labels of
+%high-dimensional samples.  
+We therefore generate a synthetic dataset with 2.5 million variables ($p=2500000$), of which 5 
+are related to the response variable, and 5000 samples ($n=5000$), as discussed in the methods section.
+
+We fit the random forest model and estimate the classification accuracy by capturing the out-of-bag error (OOB) and the
+variable selection performance by capturing the rank-biased overlap (RBO) \cite{Webber2010} measure. RBO compares two
+lists (in this case, the parameters $w_i, i=1,\ldots , 5$ (Equation~\ref{equation:synthetic.data}), and the top 1000
+ranked variable importances from the \cursedforest model). It is a mesaure of similarity between rankings that handles
+lists of different lengths, weighs high ranks more heavily than low, and is monotonic with increasing depth of
+evaluation. We expect to retrieve the 5 features in order of the weight that was assigned to them and with a higher
+variable importance measure than the noise variables. RBO provides a quantiative measure of how close we come to this on
+a scale from 0 (no feature recovered) to 1 (fully recovered). See supplementary information (S*) for more details.
+
+We are running Apache Spark 1.6.1 on a YARN
+cluster with 12 worker nodes each with 16 Intel Xeon E5-2660@2.20GHz CPU cores and 128 GB of RAM.
+
+%mtry = number of variable seclection at each node - proportion of p 
+%ntree= number of trees
+
+% mtry large makes trees correlated because the same important features have a higher chance of being selected
+% overfitting because it picks random that happen to be
+
+As shown in Fig~\ref{figure:out-of-bag-prediction-error-prod.png} the default value for \mtry (number of variables
+selected at each node) does not result in a good classification performance for this large feature dataset. Note that
+the plot shows the proportion, $n/p,$ for \mtry\ with the default being $\sqrt{p} \approx 1581$ ($n/p=0.0006$).  The
+OOB for this value of \mtry does not drop below 0.5, even when the number of trees are increased (\ntree).
+
+Increasing \mtry\ in combination with \ntree\ yields the best performance with the OOB error essentially constant around 0.4
+across a large range of of \mtry and \ntree values.  This is in contrast to the feature-selection performance, where the
+RBO measure heavily depends on \ntree\ and gives better results with lower values of \mtry.
+
+This may be because a large \mtry leads to more correlated trees as the same important features have a higher chance of
+being selected in all, not yielding good performance outcomes.  This issue is less pronounced for classification error
+where random features can mimic the response variable, hence resulting more in a performance plateau.  Increasing the
+number of trees on the other hand improves performance especially when the trees are kept diverse (small \mtry) but
+appropriate for large feature datasets (\mtry larger than default). \cursedforest's feature of parallelizing trees at
+the node level hence caters perfectly to the requirement of large feature datasets as more trees can be built given a
+fixed time budget.
+
+%Fig~\ref{figure:synth1} illustrates the effect of varying \mtry and \ntree on these two measures. The default \mtry
+%value for a classification problem is $\sqrt{p} \approx 1581$ so in this case we are considering values considerably larger than
+%the default. Fig~\ref{figure:out-of-bag-prediction-error-prod.png} shows the OOB error is constant across a large range of 
+%of \mtry and \ntree values whereas Fig~\ref{figure:rbo-prod.png} illustrates that the RBO measure depends
+%heavily on \ntree but gives a better value with lower values of \mtry.
+
+From the synthetic data we can conclude that \cursedforest is able to perform accurate feature selection as well as
+accurate classification for large numbers of samples with large feature vectors.  We can also conclude that fine tuning
+the model to accurately classify the dataset does not ensure good feature selection performance as there seems to be
+little correspondence between the OOB and RBO measure.
+
+\begin{figure}[tbhp] 
+  \begin{adjustwidth}{-1.00in}{0in}
+    \caption{\textbf{Out of bag classification error and rank-biased-overlap estimates}} 
+    \label{figure:synth1}
+    \begin{subfigure}[b]{0.5\linewidth}
+      \centering
+      \includegraphics[totalheight=8cm]{./figs/oob_levleplot.png}
+      \caption{OOB classification error.} 
+      \label{figure:out-of-bag-prediction-error-prod.png} 
+%      \vspace{4ex}
+    \end{subfigure} 
+    \begin{subfigure}[b]{0.5\linewidth}
+      \centering
+      \includegraphics[totalheight=8cm]{./figs/rbo_levleplot.png}
+      \caption{The rank-biased-overlap measure.} 
+      \label{figure:rbo-prod.png} 
+%      \vspace{4ex}
+    \end{subfigure} 
+    \begin{flushleft} 
+      Random forests with 100 trees were trained on the synthetic dataset of 2.5M features and 5K samples.
+      Each measure is shown as a function of \mtry and \ntree. Note that the plot colors have been chosen so that the lighter color indicates
+      the better result, that is, a lower OOB or a higher RBO.
+    \end{flushleft}
+  \end{adjustwidth}
+\end{figure}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-We have demonstrated that using a different parallelization model can extend random forests to the case of an
-extremely large number of variables. We have treated the case of variable selection in a $p >> n$ model, where most of
-the variables are uninformative, and have demonstrated the utility of the model for large GWAS datasets. By comparing
-this implementation to other implementations (including those optimized for large datasets) we have demonstrated the
-utility of this approach.
-
-Donoho and Tanner \cite{Donoho.and.Tanner.2009} give a ``universal phase change'' result that has applications in a
-large number of areas, including variable selection in high dimensions. The theoretical phase change boundary is based on arguments
-from combinatorial geometry.
-
-They argue that there is a sharp disjunction between the cases where informative variables may be recovered with a high
-accuracy by procedures like stepwise variable selection, and the cases where they can not be recovered. The boundary is
-shown in Fig~\ref{figure:phase-diagram-equivalence.png} in a space parameterized by the level of underdetermination,
-$\delta = n/p$, and by the sparsity, $\rho =k/n$ (where $k$ is the number of informative variables).  Above the
-phase-transition line variable recovery is still possible by a combinatorial search approach such as all-subsets
-variable selection.
-
-\begin{figure}[tbhp] 
- \begin{adjustwidth}{-1.00in}{0in}
-    \centering
-    \includegraphics[totalheight=6cm]{./figs/phase-diagram-equivalence.png} 
-    \caption{{\bf The phase change diagram.}}
-    \label{figure:phase-diagram-equivalence.png} 
-    \vspace{4ex}
-  \end{adjustwidth}
-\end{figure}
-
-\cite{Donoho.and.Stodden.2006} investigate the phase transition in a small simulation.  They consider a regression
-problem with $\underset{n\times p}{X}\sim N(0,1)$ with $p$ fixed at 200 and $n$ variable.  They set
-$\beta(1:k) \sim U(1,100)$ and $\beta((k+1):p) =0$, then $y= X\beta + \epsilon$ with $\epsilon \sim N(0,\sqrt{16})$. They
-evaluate variable selection by the error measure
-$$\frac{\Vert\hat{\beta}-\beta\Vert_2}{\Vert\beta\Vert_2}.$$
-
-They consider a number of variable selection methods, including a false discovery rate criteria. This involves adding
-the variable with the maximum $t$-value to the linear model if the $p$-value is less than 
-$0.25(\text{number of  terms  currently  in  the  model})/(\text{total  number  of  variables})$.  
-See Fig~\ref{figure:error_Stodden_FDR.png} for the error measure and
-figure \ref{figure:rbo_Stodden_FDR.png} for the RBO, comparing the ranking (i.e. values) of $\hat{\beta}$ and $\beta$.
-It apparent that there is a marked drop in the accuracy of the variable recovery in line with the prediction of the
-Donoho-Tanner phase transition.
-
- \begin{figure}[tbhp] 
+Using a novel parallelization approach enabled by Spark, we extended random forests to cope with extremely high dimensional data (large number of variables). 
+We have demonstrated that for datasets with more variable ($p$) than samples ($n$), such as GWAS applications, \cursedforest can perform effective variable selection replicating previous findings and uncovering novel associations in a Bone Mineral Density study.
+We are able to elevate the signal compared to previous approaches using logistic regression due to random forests also modelling interacting features.
+
+In fact, \cursedforest's ability to almost exhaustively sample the combinatorial space by efficiently building large numbers of trees puts it closer than any other method to the theoretical boundary for accurate recovery of informative variables.
+Donoho and Tanner~\cite{Donoho.and.Tanner.2009} define this as the ``universal phase change'', which depends on underdetermination, $\delta = n/p$, and the sparsity, $\rho =k/n$ (where $k$ is the number of informative variables).
+By effectively operating in the difficult area of this space with $p >> n$ and small number of informative features, \cursedforest is pushing the practical limits of this paradigm.
+
+To showcase the breadth of application scope, we also demonstrate that \cursedforest's implementation aids in classification problems such as predicting the ethnicity of the 1000 genome project. 
+By comparing \cursedforest\ to other implementations (including those optimised for large datasets) we have demonstrated that runtime can be improved by 80\% allowing analysis in timeframes relevant for point-of-care diagnostics such as for "Patients like mine" scenarios thereby enabling evidence based self learning health care.
+
+In summary, \cursedforest\  offers the broadly needed capability of performing sophisticated machine learning tasks on big, complex life science data.
+Specifically, the here demonstrated ability to uncover novel biological information from previously analysed GWAS studies showcases the enormous potential of re-analysing the GWAS catalogue given this unprecedented computational capacity to perform joint-loci association testing and identify the much-anticipated cumulative multi-gene disease contribution.
+
+%Donoho and Tanner \cite{Donoho.and.Tanner.2009} give a ``universal phase change'' result that has applications in a
+%large number of areas, including variable selection in high dimensions. The theoretical phase change boundary is based on arguments
+%from combinatorial geometry.
+%
+%They argue that there is a sharp disjunction between the cases where informative variables may be recovered with a high
+%accuracy by procedures like stepwise variable selection, and the cases where they can not be recovered. The boundary is
+%shown in Fig~\ref{figure:phase-diagram-equivalence.png} in a space parameterized by the level of underdetermination,
+%$\delta = n/p$, and by the sparsity, $\rho =k/n$ (where $k$ is the number of informative variables).  Above the
+%phase-transition line variable recovery is still possible by a combinatorial search approach such as all-subsets
+%variable selection. 
+%
+%\begin{figure}[tbhp] 
 % \begin{adjustwidth}{-1.00in}{0in}
-     \begin{subfigure}[t]{0.5\linewidth}
-       \centering
-       \includegraphics[totalheight=6cm]{./figs/error_Stodden_FDR.png}
-       \caption{The error measure.}
-       \label{figure:error_Stodden_FDR.png}
+%    \centering
+%    \includegraphics[totalheight=6cm]{./figs/phase-diagram-equivalence.png} 
+%    \caption{{\bf The phase change diagram.}}
+%    \label{figure:phase-diagram-equivalence.png} 
 %    \vspace{4ex}
-     \end{subfigure} 
-     \begin{subfigure}[t]{0.5\linewidth}
-       \centering
-       \includegraphics[totalheight=6cm]{./figs/rbo_Stodden_FDR.png}
-       \caption{The RBO measure}
-       \label{figure:rbo_Stodden_FDR.png}
-     \end{subfigure} 
-     \caption{Forward stepwise variable selection with a false discovery rate stopping criteria.}
-       \label{figure:error_and_rbo_Stodden_FDR.png}
 %  \end{adjustwidth}
- \end{figure}
-
- \begin{figure}[tbhp] 
-%   \begin{adjustwidth}{-1.00in}{0in}
-     \begin{subfigure}[t]{0.5\linewidth}
-       \centering
-       \includegraphics[totalheight=6cm]{./figs/ranger_error_Stodden_simulation.png}
-       \caption{The OOB error rate.}
-       \label{figure:ranger_error_Stodden_simulation.png}
-       \vspace{4ex}
-     \end{subfigure} 
-     \begin{subfigure}[t]{0.5\linewidth}
-       \centering
-       \includegraphics[totalheight=6cm]{./figs/ranger_rbo_Stodden_simulation.png}
-       \caption{The RBO measure}
-       \label{figure:ranger_rbo_Stodden_simulation.png}
-     \end{subfigure} 
-     \caption{The simulated data of \protect{\cite{Donoho.and.Stodden.2006}} with a random forest model.}
-     \label{figure:ranger_error_and_rbo_Stodden_simulation.png}
-%   \end{adjustwidth}
- \end{figure}
-
-We have seen no work on the question of such a phase transition in the case of classification problems. Still it may be
-instructive to consider the implications of the phase transition for our work.
-
-We have used a random forest on the same simulation data used for figure \ref{figure:error_and_rbo_Stodden_FDR.png}.  The
-Donoho-Tanner phase transition arises in recovering the $\beta$ in data generated by a linear model. However, in a
-decision tree (random forest) we are fitting a non-linear model and there is no notion of estimating the
-$\beta$. Because of this we have evaluated the performance of the random forest using the RBO measure on the variable
-importance.
-
-We note that while a random forest is a long way from an all-subsets search, it is a limited search of
-the feature space. As such it may perform outside of the bounds of the phase transition.
-
-Fig~\ref{figure:ranger_error_and_rbo_Stodden_simulation.png} shows the OOB prediction error and the RBO error for a
-random forest with 10000 trees and the \mtry value set to the default for a regression problem (depending on $n$). It
-shows no evidence of following the shape of the Donoho-Tanner phase transition.
-
-As \cursedforest is designed for extremely large numbers of variables it is likely to be operating in difficult regions
-of the figure where the ratio $\delta = n/p$ is small. In the simulated example in section \ref{section:synthetic_data}
-the underdetermination parameter $\delta = 0.002$ and the sparsity parameter $\rho =2\times 10^{-6}.$
-
-The existence of the Donoho-Tanner phase transition is a salutary warning. There are likely to be limits, both
-computational and logical, to the recovery of signals from noisy data. \cursedforest is a contribution to addressing the
-practical limits but the logical limits will still apply. However in the case of data that is both big and wide,
-\cursedforest and other VariantSpark methods may provide a useful tool.
+%\end{figure}
+%
+%
+%\cite{Donoho.and.Stodden.2006} investigate the phase transition in a small simulation.  They consider a regression
+%problem with $\underset{n\times p}{X}\sim N(0,1)$ with $p$ fixed at 200 and $n$ variable.  They set
+%$\beta(1:k) \sim U(1,100)$ and $\beta((k+1):p) =0$, then $y= X\beta + \epsilon$ with $\epsilon \sim N(0,\sqrt{16})$. They
+%evaluate variable selection by the error measure
+%$$\frac{\Vert\hat{\beta}-\beta\Vert_2}{\Vert\beta\Vert_2}.$$
+%
+%They consider a number of variable selection methods, including a false discovery rate criteria. This involves adding
+%the variable with the maximum $t$-value to the linear model if the $p$-value is less than 
+%$0.25(\text{number of  terms  currently  in  the  model})/(\text{total  number  of  variables})$.  
+%See Fig~\ref{figure:error_Stodden_FDR.png} for the error measure and
+%figure \ref{figure:rbo_Stodden_FDR.png} for the RBO, comparing the ranking (i.e. values) of $\hat{\beta}$ and $\beta$.
+%It apparent that there is a marked drop in the accuracy of the variable recovery in line with the prediction of the
+%Donoho-Tanner phase transition.
+%
+%
+%
+% \begin{figure}[tbhp] 
+%%   \begin{adjustwidth}{-1.00in}{0in}
+%     \begin{subfigure}[t]{0.5\linewidth}
+%       \centering
+%       \includegraphics[totalheight=6cm]{./figs/error_Stodden_FDR.png}
+%       \caption{The error measure.}
+%       \label{figure:error_Stodden_FDR.png}
+%%       \vspace{4ex}
+%     \end{subfigure} 
+%     \begin{subfigure}[t]{0.5\linewidth}
+%       \centering
+%       \includegraphics[totalheight=6cm]{./figs/rbo_Stodden_FDR.png}
+%       \caption{The RBO measure}
+%       \label{figure:rbo_Stodden_FDR.png}
+%     \end{subfigure} 
+%     \caption{Forward stepwise variable selection with a false discovery rate stopping criteria.}
+%       \label{figure:error_and_rbo_Stodden_FDR.png}
+%%   \end{adjustwidth}
+% \end{figure}
+%
+% \begin{figure}[tbhp] 
+%%   \begin{adjustwidth}{-1.00in}{0in}
+%     \begin{subfigure}[t]{0.5\linewidth}
+%       \centering
+%       \includegraphics[totalheight=6cm]{./figs/ranger_error_Stodden_simulation.png}
+%       \caption{The OOB error rate.}
+%       \label{figure:ranger_error_Stodden_simulation.png}
+%       \vspace{4ex}
+%     \end{subfigure} 
+%     \begin{subfigure}[t]{0.5\linewidth}
+%       \centering
+%       \includegraphics[totalheight=6cm]{./figs/ranger_rbo_Stodden_simulation.png}
+%       \caption{The RBO measure}
+%       \label{figure:ranger_rbo_Stodden_simulation.png}
+%     \end{subfigure} 
+%     \caption{The simulated data of \protect{\cite{Donoho.and.Stodden.2006}} with a random forest model.}
+%     \label{figure:ranger_error_and_rbo_Stodden_simulation.png}
+%%   \end{adjustwidth}
+% \end{figure}
+%
+%
+%We have seen no work on the question of such a phase transition in the case of classification problems. Still it may be
+%instructive to consider the implications of the phase transition for our work. 
+%
+%We have used a random forest on the same simulation data used for figure \ref{figure:error_and_rbo_Stodden_FDR.png}.  The
+%Donoho-Tanner phase transition arises in recovering the $\beta$ in data generated by a linear model. However, in a
+%decision tree (random forest) we are fitting a non-linear model and there is no notion of estimating the
+%$\beta$. Because of this we have evaluated the performance of the random forest using the RBO measure on the variable
+%importance.
+%
+%We note that while a random forest is a long way from an all-subsets search, it is a limited search of
+%the feature space. As such it may perform outside of the bounds of the phase transition. 
+%
+%
+%Fig~\ref{figure:ranger_error_and_rbo_Stodden_simulation.png} shows the OOB prediction error and the RBO error for a
+%random forest with 10000 trees and the \mtry value set to the default for a regression problem (depending on $n$). It
+%shows no evidence of following the shape of the Donoho-Tanner phase transition.
+%
+%As \cursedforest is designed for extremely large numbers of variables it is likely to be operating in difficult regions
+%of the figure where the ratio $\delta = n/p$ is small. In the simulated example in section \ref{section:synthetic_data}
+%the underdetermination parameter $\delta = 0.002$ and the sparsity parameter $\rho =2\times 10^{-6}.$
+%
+%The existence of the Donoho-Tanner phase transition is a salutary warning. There are likely to be limits, both
+%computational and logical, to the recovery of signals from noisy data. \cursedforest is a contribution to addressing the
+%practical limits but the logical limits will still apply. However in the case of data that is both big and wide,
+%\cursedforest and other VariantSpark methods may provide a useful tool.
 

### 2017-04-24 doc/publications/plos16/supplementary_information.tex
updated BMI section with new model
updated BMD section
rewrite

@@
 \usepackage{caption}
+\usepackage{subfigure}
 
@@
 
-\section{supplementary Information for Section 4.1}
+\section{Supplementary Information for Section 4.1}
+
+\begin{figure}[tbhp]
+  \begin{adjustwidth}{-2.00in}{0in}
+    \caption{\textbf{Rank of literature assigned BMD genes}}
+    \label{figure:knowngeneranks}
+    \includegraphics[totalheight=6.5cm]{./figs/BMDgenes_landscape.pdf}
+    \begin{flushleft}
+      The figure shows the rank of the 100 literature assigned BMD genes.
+Each ball represents a SNP assigned to the closest gene with colour representing the distance (max 100 kb).
+The size shows the Gini impurity score with larger balls representing more important SNPs.
+The horizontal line separates the genes with SNPs amongst the Top 20 associated features.
+     \end{flushleft}
+  \end{adjustwidth}
+\end{figure}
+
+\newpage
+
+\section{supplementary Information for Section 4.2}
 
@@
 \end{sidewaystable}
+
+\clearpage
+\newpage
+
+\section{supplementary Information for Section 4.2}
+
+\begin{figure}[tbhp]
+    \caption{\textbf{Rank of literature assigned BMD genes}}
+    \label{figure:knowngeneranks}
+    \includegraphics{./figs/BMDTop10K.pdf}
+    \begin{flushleft}
+      The figure shows the top 10,000 features by their rank and Gini impurity score (Importance). The line denotes the top 30 features which contains six novel loci. 
+     \end{flushleft}
+\end{figure}
+
+\begin{figure}[tbhp]
+    \caption{\textbf{Zoomplot in order of importance}}
+    \label{figure:knowngeneranks}
+    \includegraphics{./figs/new_10k_TNFRSF11B.pdf}
+     \includegraphics{./figs/new_10k_CTNNB1.pdf}
+    \begin{flushleft}
+     \end{flushleft}
+\end{figure}
+
+\begin{figure}[tbhp]
+    \caption{\textbf{Zoomplot in order of importance}}
+    \label{figure:knowngeneranks}
+      \includegraphics{./figs/new_10k_ZBTB40.pdf}
+        \includegraphics{./figs/new_10k_IBSP.pdf}
+    \begin{flushleft}
+     \end{flushleft}
+\end{figure}
+
+\begin{figure}[tbhp]
+    \caption{\textbf{Zoomplot of likely artefacts}}
+    \label{figure:knowngeneranks}
+      \includegraphics{./figs/EUR_chr13_66_4-67mb.pdf}
+        \includegraphics{./figs/EUR_chr17_62_1-62_3mb.pdf}
+    \begin{flushleft}
+     \end{flushleft}
+\end{figure}
+
+\begin{figure}[tbhp]
+    \caption{\textbf{Manhattan plot of PRODH and LINC01006}}
+    \label{figure:knowngeneranks}
+        \includegraphics{./figs/BMDTopWholeGenome_22.pdf}
+      \includegraphics{./figs/BMDTopWholeGenome_7.pdf}
+    \begin{flushleft}
+     \end{flushleft}
+\end{figure}
 

### 2017-04-24 doc/publications/plos16/nm.tex
add robs changes from plos16.tex to nm.tex

@@
 across multiple nodes, each vector must be present in its entirety on any node accessing it. This is no problem with
-typical datasets, however as dimensionality increases, the vectors eventually reach a size where they can no longer fit
+typical datasets; however as dimensionality increases, the vectors eventually reach a size where they can no longer fit
 into a single node's memory.
@@
     classification example. This may give an estimate with a higher bias and, as such, may be a poor choice. See the
-    discussion in \cite{Dietterich.2002} where it is shown that bagging may reduce bias as well as variance but give a
+    discussion in \cite{Dietterich.2002} where it is shown that bagging may reduce bias as well as variance but is more
+    complicated than first thought. They show that bagging may reduce bias as well as variance but give a    
     better result for low bias learners. \cite{Dietterich.2002} notes that if the bootstrap replicate approximation were
@@
 %high-dimensional samples.  
-We therefore generate a synthetic dataset with 2.5 million variables ($p=2500000$), of which 5 
+We therefore generate a synthetic dataset with 2.5 million variables ($p=2,500,000$), of which 5 
 are related to the response variable, and 5000 samples ($n=5000$), as discussed in the methods section.

### 2017-04-25 doc/publications/plos16/nm.tex
changes to nm.tex
1) million not Million
2) removed the spacing symbold from  \cursedforest\  -- the definition ends in an \xspace which will space it properly
3) fixed double quotes so LaTeX  will draw left and right quotation marks
4) make n and p italic
5) we have inconsistent capitalization on Bone Mineral Density, Bone mineral density, bone mineral density
I think we get rid of all the capitals
6) datasets or data sets ?
7)  We have a number of instances where we start a sentence with a reference. This is OK in an author-year style but with a numeric style I have had editors complain abou

@@
 
-Analyses are performed on ever larger datasets with unprecedentedly detailed information per data point.
-Specifically in genomics, cohort sizes are in the thousands, each comprising of Millions of variants that make up a person's genetic differences.
-Traditional approaches are unable to cater for this task and even standard implementations using novel compute paradigms, such as Hadoop/Spark, are limited by the finite resource of computer memory on such data.
-Specifically, Spark ML's supervised machine learning methods suffer from "the curse of dimensionality".
-That is, they scale well with samples (n) but not with features (p, variants).
-
-We introduce \cursedforest, a tailored implementation of random forests, designed to handle data with extremely large number of variables per sample.
-We showcase \cursedforest\ to accurately predict ethnicity from whole genome sequencing profiles of 2,500 individuals and 80 million genomic variants each (1000 Genomes Project) in 30 minutes.
-\cursedforest\ was the only implementation scaling to the full dataset, outperforming other methods on smaller subsets by 80\% or more.
-We also demonstrate a GWAS-style analysis using  \cursedforest's feature selection approach on an imputed SNP array dataset (2000 x 7 Million) on Bone mineral density (BMD), successfully predicting known BMD-genes as well as identifying novel candidates.
-
-\cursedforest is included in our earlier variant interpretation package, VariantSpark, allowing it to perform near real-time classification of population-scale patient cohorts in "patients like mine" scenarios, as well as performing GWAS analysis on large unfiltered whole genome sequencing cohorts.
+Analyses are being performed on ever larger datasets with unprecedentedly detailed information per data point. Specifically in
+genomics, cohort sizes are in the thousands, each comprising of millions of variants that make up a person's genetic
+differences. Traditional approaches are unable to cater for this task and even standard implementations using novel
+compute paradigms, such as Hadoop/Spark, are limited by the finite resource of computer memory on such
+data. Specifically, Spark ML's supervised machine learning methods suffer from ``the curse of dimensionality''. That is,
+they scale well with samples ($n$) but not with features ($p$, variants).
+
+We introduce \cursedforest, a tailored implementation of random forests, designed to handle data with extremely large
+number of variables per sample. We showcase \cursedforest to accurately predict ethnicity from whole genome sequencing
+profiles of 2,500 individuals and 80 million genomic variants each (1000 Genomes Project) in 30 minutes. \cursedforest
+was the only implementation scaling to the full dataset, outperforming other methods on smaller subsets by 80\% or
+more. We also demonstrate a GWAS-style analysis using \cursedforest's feature selection approach on an imputed SNP array
+dataset (2000 x 7 Million) on Bone mineral density (BMD), successfully predicting known BMD-genes as well as identifying
+novel candidates.
+
+\cursedforest is included in our earlier variant interpretation package, VariantSpark, allowing it to perform near
+real-time classification of population-scale patient cohorts in ``patients like mine'' scenarios, as well as performing
+GWAS analysis on large unfiltered whole genome sequencing cohorts.
 
@@
 well. Harnessing the virtually unlimited capability to parallelise tasks, \cursedforest can hence explore the solution 
-space faster by building a larger number of diverse models to generate a consensus from.
+space faster by building a larger number of diverse models to generate a consensus classifier.
 
@@
 \item they provide a measure of variable importance, for example by summing the number of times a variable is used to
-  split a node, over all nodes and all trees. More complex measures of variable importance are also used;
+  split a node, over all nodes and all trees. Other, more complex, measures of variable importance are also used;
 \item they inherit some of the qualities of tree models (but not, for example, the handling of missing values via surrogate
@@
 variant (1\textbar0 or 0\textbar1) as 1, and a
-homozygous variant (1\textbar1) as 2.  However, this is not the only possible encoding \cite{Goldstein.et.al.2011}. Note that from
+homozygous variant (1\textbar1) as 2, so we have an ``additive'' model.  However, this is not the only possible encoding \cite{Goldstein.et.al.2011}. Note that from
 the encodings listed in Table~\ref{table:encodings}, the additive encoding will present the most difficulty for a random
@@
 $y =   z + \epsilon$ where $\epsilon \sim  N(0, \sigma_\epsilon^2).$
-The dichotomous response is generated by thresholding $y$ at the $0.5$ quantile.
+The dichotomous response is generated by thresholding $y$ at the $0.5$ quantile,
 
@@
 			  0 & \mbox{for} & y \leq Q2(y) \\
-			  1 & \mbox{for} & y > Q2(y)
-		 \end{array}\right.
+			  1 & \mbox{for} & y > Q2(y).
+		 \end{array}\right 
 $$
@@
     classification example. This may give an estimate with a higher bias and, as such, may be a poor choice. See the
-    discussion in \cite{Dietterich.2002} where it is shown that bagging may reduce bias as well as variance but is more
-    complicated than first thought. They show that bagging may reduce bias as well as variance but give a    
-    better result for low bias learners. \cite{Dietterich.2002} notes that if the bootstrap replicate approximation were
-    correct (i.e.~if the bootstrap sample came from an identical distribution to the data), then bagging would reduce
-    variance without changing bias.}
+    discussion in \cite{Dietterich.2002} where it is shown that the action of a RF is complicated than first
+    thought. They show that bagging may reduce bias as well as variance but give a better result for low bias
+    learners. \cite{Dietterich.2002} notes that if the bootstrap replicate approximation were correct (i.e.~if the
+    bootstrap sample came from an identical distribution to the data), then bagging would reduce variance without
+    changing bias.}
 \end{itemize}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%\subsection{GWAS-style analysis using \cursedforest\ identifies novel variants associated with Bone Mineral Density}
-\subsection{GWAS-style analyses using \cursedforest\ faithfully identify known disease-accociated genes}
-In this section we use \cursedforest\ to perform feature selection on over 7.2 million genomic variants and identify the
+%\subsection{GWAS-style analysis using \cursedforest identifies novel variants associated with Bone Mineral Density}
+\subsection{GWAS-style analyses using \cursedforest faithfully identify known disease-accociated genes}
+In this section we use \cursedforest to perform feature selection on over 7.2 million genomic variants and identify the
 genomic locations associated with Bone Mineral Density (BMD)~\cite{Duncan.2011}. 
@@
 To test whether the associated variants have biological meaning, we investigate if variants in or near
-BMD-associated genes are enriched toward the top of our ranked list. To form the list of "known genes", we
+BMD-associated genes are enriched toward the top of our ranked list. To form the list of ``known genes'', we
 combine the 26 known and 6 novel BMD genes listed by Duncan {\it et al.}~\cite{Duncan.2011} with the genes identified by
@@
 
-\subsection{\cursedforest\ amplifies signal over traditional methods to identify novel associations}
+\subsection{\cursedforest amplifies signal over traditional methods to identify novel associations}
 
 Figure 1b shows a traditional GWAS analysis using logistic regression~\cite{Duncan.2011} where the resulting $p$-values do not clearly distinguish these known BMD genes from the noise. 
-In contrast, \cursedforest\ seems to amplify the signal to identify associated features (Figure 1a). 
+In contrast, \cursedforest seems to amplify the signal to identify associated features (Figure 1a). 
 As such, this approach allows to employ a very stringent cut-off for identifying strongly associated feature. Here we chose a 99.98 False Discovery Rate (FDR) from 60 random experiments (see Methods).
@@
 The variants overlapping the 5' end of the lincRNA are the highest ranked features in this region and all LD SNPs also show high importance (see Fig 3d).
-Also amonst the top 30 features is PRODH (proline dehydrogenase) located on 22q11, a loci observed by Stagi {\it
+Also amongst the top 30 features is PRODH (proline dehydrogenase) located on 22q11, a loci observed by Stagi {\it
   et al.}~\cite{Stagi.2010} to significantly reduce bone mass in individuals with a micro-deletion in this region. 
@@
 
-%\cursedforest\ can successfully perform GWAS-type analyses replicating previous findings and uncovering novel associations due to random forests also modelling interacting features.
+%\cursedforest can successfully perform GWAS-type analyses replicating previous findings and uncovering novel associations due to random forests also modelling interacting features.
 
@@
 %@Aidan is 12 correct?
-We limit the number of executors to 12 that are available to \cursedforest\ to keep the comparison fair for all methods.
-
-As shown in Fig.~\ref{figure:1000genomes}, only \cursedforest\ is able to scale to the full phase3\_chr1-22 dataset (2,504 samples x 19,328,051 variants), with
+We limit the number of executors to 12 that are available to \cursedforest to keep the comparison fair for all methods.
+
+As shown in Fig.~\ref{figure:1000genomes}, only \cursedforest is able to scale to the full phase3\_chr1-22 dataset (2,504 samples x 19,328,051 variants), with
 the standard random forest implementation in R running out of resources on the relatively low-dimensional
@@
 
-All tested methods have a worse runtime than \cursedforest\ for the datasets they processed successfully, except for the smallest dataset where the overhead of Spark's distributed platform made ranger's R implementation faster with 1 min 10 seconds compared to 1 min 50 seconds.
-For example, on phase3\_chr1-3 \cursedforest\ finishes in 1 hour, with ranger(R) requiring 4 hours and SparkML already needing 8 hours on a smaller dataset.
+All tested methods have a worse runtime than \cursedforest for the datasets they processed successfully, except for the smallest dataset where the overhead of Spark's distributed platform made ranger's R implementation faster with 1 min 10 seconds compared to 1 min 50 seconds.
+For example, on phase3\_chr1-3 \cursedforest finishes in 1 hour, with ranger(R) requiring 4 hours and SparkML already needing 8 hours on a smaller dataset.
 Extrapolating from this, all comparable implementations hence would require more than 24 hours to finish, assuming the use of larger memory nodes.
 
-As our implementation can efficiently utilise large numbers of commodity nodes, the performance of \cursedforest\ can be substantially improved further. 
+As our implementation can efficiently utilise large numbers of commodity nodes, the performance of \cursedforest can be substantially improved further. 
 %@Aidan is 130 correct?
@@
 
-Building a model with \cursedforest\ on phase3\_chr1, which previously took 20 minutes, can be reduced to 4 minutes. 
+Building a model with \cursedforest on phase3\_chr1, which previously took 20 minutes, can be reduced to 4 minutes. 
 Subsequently, phase3\_chr1-3 takes about 11 minutes, and building a
@@
 
-Not only does the job complete successfully using \cursedforest\, but building a random
+Not only does the job complete successfully using \cursedforest, but building a random
 forest model on all 22 autosomes (phase3\_chr1-22)  leads to a more accurate result. 
@@
 %\label{section:synthetic}
-In this section we test the limits of \cursedforest\'s association testing and classify ability on high-dimensional data. 
+In this section we test the limits of \cursedforest's association testing and classify ability on high-dimensional data. 
 %In this section we demonstrate the two features of \cursedforest, namely: variable selection analysis to select the most
@@
 selected at each node) does not result in a good classification performance for this large feature dataset. Note that
-the plot shows the proportion, $n/p,$ for \mtry\ with the default being $\sqrt{p} \approx 1581$ ($n/p=0.0006$).  The
+the plot shows the proportion, $n/p,$ for \mtry with the default being $\sqrt{p} \approx 1581$ ($n/p=0.0006$).  The
 OOB for this value of \mtry does not drop below 0.5, even when the number of trees are increased (\ntree).
 
-Increasing \mtry\ in combination with \ntree\ yields the best performance with the OOB error essentially constant around 0.4
+Increasing \mtry in combination with \ntree yields the best performance with the OOB error essentially constant around 0.4
 across a large range of of \mtry and \ntree values.  This is in contrast to the feature-selection performance, where the
-RBO measure heavily depends on \ntree\ and gives better results with lower values of \mtry.
+RBO measure heavily depends on \ntree and gives better results with lower values of \mtry.
 
@@
 \subsection{Scalability analysis}
-In this section we explore the performance of \cursedforest\ in more detail by testing its ability to scale to different sizes of data
+In this section we explore the performance of \cursedforest in more detail by testing its ability to scale to different sizes of data
 and computational resources.
@@
       \includegraphics[totalheight=6cm]{./figs/mtry_cpu.png} 
-      \caption{Time in seconds to build 100 trees for different \mtry\ fractions. } 
+      \caption{Time in seconds to build 100 trees for different \mtry fractions. } 
       \label{figure:synthetictiming.a} 
@@
 % %\multicolumn{4}{|l|}{\bf Heading1} & \multicolumn{4}{|l|}{\bf Heading2}\\ \hline
-% \bf{ \mtry\ }  & \bf{Mtree} & \bf{depth} & \bf{Accuracy} & \bf{Runtime} & \bf{Memory} \\
+% \bf{ \mtry }  & \bf{Mtree} & \bf{depth} & \bf{Accuracy} & \bf{Runtime} & \bf{Memory} \\
 % \hline
@@
 To showcase the breadth of application scope, we also demonstrate that \cursedforest's implementation aids in classification problems such as predicting the ethnicity of the 1000 genome project. 
-By comparing \cursedforest\ to other implementations (including those optimised for large datasets) we have demonstrated that runtime can be improved by 80\% allowing analysis in timeframes relevant for point-of-care diagnostics such as for "Patients like mine" scenarios thereby enabling evidence based self learning health care.
-
-In summary, \cursedforest\  offers the broadly needed capability of performing sophisticated machine learning tasks on big, complex life science data.
+By comparing \cursedforest to other implementations (including those optimised for large datasets) we have demonstrated that runtime can be improved by 80\% allowing analysis in timeframes relevant for point-of-care diagnostics such as for "Patients like mine" scenarios thereby enabling evidence based self learning health care.
+
+In summary, \cursedforest  offers the broadly needed capability of performing sophisticated machine learning tasks on big, complex life science data.
 Specifically, the here demonstrated ability to uncover novel biological information from previously analysed GWAS studies showcases the enormous potential of re-analysing the GWAS catalogue given this unprecedented computational capacity to perform joint-loci association testing and identify the much-anticipated cumulative multi-gene disease contribution.

### 2017-04-26 doc/publications/plos16/nm.tex
small changes to nm.tex

@@
 
-Analyses are being performed on ever larger datasets with unprecedentedly detailed information per data point. Specifically in
+Analyses are being performed on ever larger data sets with unprecedentedly detailed information per data point. Specifically in
 genomics, cohort sizes are in the thousands, each comprising of millions of variants that make up a person's genetic
@@
 more. We also demonstrate a GWAS-style analysis using \cursedforest's feature selection approach on an imputed SNP array
-dataset (2000 x 7 Million) on Bone mineral density (BMD), successfully predicting known BMD-genes as well as identifying
+dataset (2000 x 7 Million) on bone mineral density (BMD), successfully predicting known BMD-genes as well as identifying
 novel candidates.
@@
 The digital revolution is seeing a dramatic increase in data collected about almost every aspect of
-life~\cite{Loebbecke2015}.  These datasets are not only growing vertically, by capturing more events, but also
+life~\cite{Loebbecke2015}.  These data sets are not only growing vertically, by capturing more events, but also
 horizontally by capturing more information about these events.  The challenge of big and ``wide'' data is especially
@@
 suited for processing ``wide'' genomic data for two reasons.  Firstly, while other machine learning applications have
-the propensity to overfit datasets with more features $p$ than samples $n$ (a consequence of the ``curse of
+the propensity to overfit data sets with more features $p$ than samples $n$ (a consequence of the ``curse of
 dimensionality''~\cite{Bauer2014, bellman1961adaptive}), decision trees are resistant to overfitting.  Secondly, random
@@
 multithreading.  Similarly, \ranger~\cite{Wright.and.Ziegle.2016}, written in C++ with an R front-end, is
-specifically optimised for dealing with large datasets.
+specifically optimised for dealing with large data sets.
 
@@
 intermediate stages of building the model.  This unbalanced work load where the driver
-node becomes the bottleneck and worker nodes being idle prevents a seamless scaling to larger datasets. We also show that the memory
+node becomes the bottleneck and worker nodes being idle prevents a seamless scaling to larger data sets. We also show that the memory
 requirements per executor increases with dimensionality due to the data types Spark ML uses.%, which we elaborate on in the discussion.
@@
 Here we introduce \cursedforest, a tailored Hadoop/Spark-based implementation of random forests specifically designed to cater
-for ``big'' (many samples) and ``wide'' (many features) datasets. 
+for ``big'' (many samples) and ``wide'' (many features) data sets. 
 \cursedforest extends our previously developed variant interpretation framework, VariantSpark~\cite{OBrien2015}, to now offer
@@
 By also utilising Spark to read in and manipulate the standard genomic variant format (VCF)
-directly, \cursedforest outperforms existing tools even on small datasets where multithreading generally performs
+directly, \cursedforest outperforms existing tools even on small data sets where multithreading generally performs
 well. Harnessing the virtually unlimited capability to parallelise tasks, \cursedforest can hence explore the solution 
@@
 of features, as required to process whole genome sequencing data or SNP array data with unobserved genotypes 
-imputed~\cite{Howie2012}.  This provides the potential to generate datasets of hundreds of thousands of individuals 
+imputed~\cite{Howie2012}.  This provides the potential to generate data sets of hundreds of thousands of individuals 
 with millions of variants (imputing the GWAS catalog), highlighting the need for modern
@@
 
-Random forests apply the technique of bootstrap aggregation or bagging to decision trees.  The training datasets are
+Random forests apply the technique of bootstrap aggregation or bagging to decision trees.  The training data sets are
 independently drawn, with replacement, from the joint distribution of $(X,Y)$ so that each set comprises $n$
@@
 \end{itemize}
-GWAS datasets are an obvious application area for this algorithm.
+GWAS data sets are an obvious application area for this algorithm.
 
@@
 across multiple nodes, each vector must be present in its entirety on any node accessing it. This is no problem with
-typical datasets; however as dimensionality increases, the vectors eventually reach a size where they can no longer fit
+typical data sets; however as dimensionality increases, the vectors eventually reach a size where they can no longer fit
 into a single node's memory.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% This section talks about the datasets
+% This section talks about the data sets
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Datasets}
+\subsection{Data Sets}
 \subsubsection{1000 Genomes Project}
@@
 In this section we use \cursedforest to perform feature selection on over 7.2 million genomic variants and identify the
-genomic locations associated with Bone Mineral Density (BMD)~\cite{Duncan.2011}. 
+genomic locations associated with bone mineral density (BMD)~\cite{Duncan.2011}. 
 %We build a model on the BMD dataset as described in Methods using a binary high or low BMD value as the label.  
@@
   \begin{adjustwidth}{-2.00in}{0in}
-    \caption{\textbf{Genome-wide importance of SNPs for Bone Mineral Density association}}
+    \caption{\textbf{Genome-wide importance of SNPs for bone mineral density association}}
     \label{figure:ranksumtest}
@@
 %@Aidan is 12 correct?
-We limit the number of executors to 12 that are available to \cursedforest to keep the comparison fair for all methods.
+We limit the number of executors available to \cursedforest to 12 to keep the comparison fair for all methods.
 
@@
 phase1\_chr22 dataset (1,092 x 490,036). 
-Ranger's c++ implementation and the R front-end version of it are able to process the second largest dataset (phase3\_chr1-3 with 2,504 x 19,328,051), however, both are unable to process larger datasets successfully.
+Ranger's c++ implementation and the R front-end version of it are able to process the second largest dataset (phase3\_chr1-3 with 2,504 x 19,328,051), however, both are unable to process larger data sets successfully.
 While the c++ implementation runs out of memory the R front-end, which also covers the loading and pre-processing, needs an unreasonable compute time for this task (>20h), see Supplementary Table 1.
@@
 
-All tested methods have a worse runtime than \cursedforest for the datasets they processed successfully, except for the smallest dataset where the overhead of Spark's distributed platform made ranger's R implementation faster with 1 min 10 seconds compared to 1 min 50 seconds.
+All tested methods have a worse runtime than \cursedforest for the data sets they processed successfully, except for the smallest dataset where the overhead of Spark's distributed platform made ranger's R implementation faster with 1 min 10 seconds compared to 1 min 50 seconds.
 For example, on phase3\_chr1-3 \cursedforest finishes in 1 hour, with ranger(R) requiring 4 hours and SparkML already needing 8 hours on a smaller dataset.
@@
 number of trees on the other hand improves performance especially when the trees are kept diverse (small \mtry) but
-appropriate for large feature datasets (\mtry larger than default). \cursedforest's feature of parallelizing trees at
-the node level hence caters perfectly to the requirement of large feature datasets as more trees can be built given a
+appropriate for large feature data sets (\mtry larger than default). \cursedforest's feature of parallelizing trees at
+the node level hence caters perfectly to the requirement of large feature data sets as more trees can be built given a
 fixed time budget.
@@
 
-In order to asses these characteristics, we run \cursedforest classification on synthetic datasets with varying numbers
+In order to asses these characteristics, we run \cursedforest classification on synthetic data sets with varying numbers
 of variables (features) and samples, similar to the dataset used in \cite{Wright.and.Ziegle.2016} to evaluate
@@
 
-We investigate the different synthetic datasets generated for section \ref{section:synthetic_data} and measured the time
+We investigate the different synthetic data sets generated for section \ref{section:synthetic_data} and measured the time
 taken to build a random forest model of 100 trees. The results reported below are averages of 5 runs, and all the cases
@@
 
-To conclude \cursedforest is capable of processing 60 trees per hour on a datasets with 50 million variables and 10,000
+To conclude \cursedforest is capable of processing 60 trees per hour on a data sets with 50 million variables and 10,000
 samples, which is the size range for whole genome sequencing experiments of clinically relevant cohort sizes.
@@
   \begin{adjustwidth}{-1.00in}{0in}
-    \caption{\textbf{Scalablity of the Wide Random Forest on synthetic datasets with varying number of samples and variables.}}
+    \caption{\textbf{Scalablity of the Wide Random Forest on synthetic data sets with varying number of samples and variables.}}
     \label{figure:synth2}
@@
 Using a novel parallelization approach enabled by Spark, we extended random forests to cope with extremely high dimensional data (large number of variables). 
-We have demonstrated that for datasets with more variable ($p$) than samples ($n$), such as GWAS applications, \cursedforest can perform effective variable selection replicating previous findings and uncovering novel associations in a Bone Mineral Density study.
+We have demonstrated that for data sets with more variable ($p$) than samples ($n$), such as GWAS applications, \cursedforest can perform effective variable selection replicating previous findings and uncovering novel associations in a bone mineral density study.
 We are able to elevate the signal compared to previous approaches using logistic regression due to random forests also modelling interacting features.
@@
 To showcase the breadth of application scope, we also demonstrate that \cursedforest's implementation aids in classification problems such as predicting the ethnicity of the 1000 genome project. 
-By comparing \cursedforest to other implementations (including those optimised for large datasets) we have demonstrated that runtime can be improved by 80\% allowing analysis in timeframes relevant for point-of-care diagnostics such as for "Patients like mine" scenarios thereby enabling evidence based self learning health care.
+By comparing \cursedforest to other implementations (including those optimised for large data sets) we have demonstrated that runtime can be improved by 80\% allowing analysis in timeframes relevant for point-of-care diagnostics such as for "Patients like mine" scenarios thereby enabling evidence based self learning health care.
 
@@
 % extremely large number of variables. We have treated the case of variable selection in a $p >> n$ model where most of
-% the variables are uninformative and have demonstrated the utility of the model for large GWAS datasets. By comparing
-% this implementation to other implementations (including those optimized for large datasets) we have demonstrated the
+% the variables are uninformative and have demonstrated the utility of the model for large GWAS data sets. By comparing
+% this implementation to other implementations (including those optimized for large data sets) we have demonstrated the
 % utility of this approach.

### 2017-04-27 doc/publications/plos16/nm.tex
changes to  nm.tex

@@
 \section{Abstract}
-
-Analyses are being performed on ever larger data sets with unprecedentedly detailed information per data point. Specifically in
+Modern data analyses are performed on ever larger data sets with unprecedentedly detailed information per data point. Specifically in
 genomics, cohort sizes are in the thousands, each comprising of millions of variants that make up a person's genetic
@@
 horizontally by capturing more information about these events.  The challenge of big and ``wide'' data is especially
-pronounced in the health space where, for example, whole genome sequencing (WGS) technology enables researchers to
+pronounced in the biomedical space where, for example, whole genome sequencing (WGS) technology enables researchers to
 interrogate all 3 billion base pairs of the human genome.
 
-Identifying the relevant or disease specific base pair differences between individuals is the focus of genome wide
+Identifying the relevant or disease specific genetic variations between individuals is the focus of genome wide
 association studies (GWAS).  These analyses are typically performed on only the most frequently differing base pairs,
@@
 Hence using more sophisticated machine learning (ML) approaches, in particular tree-based models, have been successful
-for taking the interaction of variables into account~\cite{Wright.et.al.2016}. In addition, random forests are well
+for taking the interaction of variables into account~\cite{Wright.et.al.2016}. In addition, random forests \cite{Breiman2001} are well
 suited for processing ``wide'' genomic data for two reasons.  Firstly, while other machine learning applications have
@@
 forests are also very easy to parallelise. As the forest is a sum of decision trees, it is possible to grow separate
-trees on different processors and combine the trees. 
+trees on different compute nodes and combine the trees. 
 %See the discussion of the properties of random forests in Section \ref{section:methods}.
@@
 			  1 & \mbox{for} & y > Q2(y).
-		 \end{array}\right 
+		 \end{array}\right. 
 $$

### 2017-04-27 doc/publications/plos16/supplementary_information.tex

@@
 \newpage
+
+SNP	Pvalue	Name	PaulPVALUE	RF
+rs9969672	0.0323	TNFRSF11B	1.30E-05	1
+rs62259232	0.0022	CTNNB1	4.12E-06	2
+rs112558362	0.0003	rs112558362	9.90E-07	3
+rs12563939	0.9795	ZBTB40	6.69E-07	4
+rs6961378	0.0179	LINC0106	8.92E-06	5
+rs1054629	0.5631	IBSP	8.62E-05	6
+rs6894139	0.5229	MEF2C	6.92E-08	7
+rs17195090	0.0643	rs17195090	1.80E-05	8
+rs17743906	??	PRODH	1.21E-06	9
 

### 2017-05-03 doc/publications/plos16/nm.tex
Short abstract

@@
 \section{Abstract}
-Modern data analyses are performed on ever larger data sets with unprecedentedly detailed information per data point. Specifically in
-genomics, cohort sizes are in the thousands, each comprising of millions of variants that make up a person's genetic
-differences. Traditional approaches are unable to cater for this task and even standard implementations using novel
-compute paradigms, such as Hadoop/Spark, are limited by the finite resource of computer memory on such
-data. Specifically, Spark ML's supervised machine learning methods suffer from ``the curse of dimensionality''. That is,
-they scale well with samples ($n$) but not with features ($p$, variants).
-
+%Modern data analyses are performed on ever larger data sets with unprecedentedly detailed information per data point. Specifically in
+%genomics, cohort sizes are in the thousands, each comprising of millions of variants that make up a person's genetic
+%differences. Traditional approaches are unable to cater for this task and even standard implementations using novel
+%compute paradigms, such as Hadoop/Spark, are limited by the finite resource of computer memory on such
+%data. Specifically, Spark ML's supervised machine learning methods suffer from ``the curse of dimensionality''. That is,
+%they scale well with samples ($n$) but not with features ($p$, variants).
+Modern genomic sequencing studies have seen increasingly large cohort in the past decade. This lead to genomic sequencing datasets with not only bigger sample sizes,
+but also, more crucially, exponentially growing variant space. Traditional data analysis approaches are unable to cater such kind of big and wide datasets. Even the machine
+learning library in the highly parallelised Spark computing engine suffer from this ``the curse of dimensionality'.
 We introduce \cursedforest, a tailored implementation of random forests, designed to handle data with extremely large
-number of variables per sample. We showcase \cursedforest to accurately predict ethnicity from whole genome sequencing
-profiles of 2,500 individuals and 80 million genomic variants each (1000 Genomes Project) in 30 minutes. \cursedforest
-was the only implementation scaling to the full dataset, outperforming other methods on smaller subsets by 80\% or
-more. We also demonstrate a GWAS-style analysis using \cursedforest's feature selection approach on an imputed SNP array
-dataset (2000 x 7 Million) on bone mineral density (BMD), successfully predicting known BMD-genes as well as identifying
-novel candidates.
-
+number of variables per sample. With public genome sequencing and SNP array datasets, we demonstrate the utilities of \cursedforest 
+for accurately predicting sample ethnicity and identifying known and novel variants associated with the phenotype of interest.
+%We showcase \cursedforest to accurately predict ethnicity from whole genome sequencing
+%profiles of 2,500 individuals and 80 million genomic variants each (1000 Genomes Project) in 30 minutes. \cursedforest
+%was the only implementation scaling to the full dataset, outperforming other methods on smaller subsets by 80\% or
+%more. We also demonstrate a GWAS-style analysis using \cursedforest's feature selection approach on an imputed SNP array
+%dataset (2000 x 7 Million) on bone mineral density (BMD), successfully predicting known BMD-genes as well as identifying
+%novel candidates. 
 \cursedforest is included in our earlier variant interpretation package, VariantSpark, allowing it to perform near
 real-time classification of population-scale patient cohorts in ``patients like mine'' scenarios, as well as performing
-GWAS analysis on large unfiltered whole genome sequencing cohorts.
+GWAS-style analysis on large unfiltered whole genome sequencing cohorts.
 

### 2017-05-05 doc/publications/plos16/nm.tex
added some of Oscars chnages

@@
 
-The most popular random forest software packages to date are R-based, although it has been reported that current implementations 
-do not scale well with increasing feature size~\cite{Wright.and.Ziegle.2016}.  This is despite R now supporting long vectors (length greater
-than $2^{31}-1$), and the \randomforest implementation in R lending itself to parallelisation by growing multiple
-trees simultaneously and combining them for the final result~\cite{Liaw.and.Weiner.2002}.  More suitable implementations for ``wide''
-genomic data have been developed in other programming languages. \textsc{CloudForest} \cite{Bressler2015}, written in Go,
-achieves fast run times by effective use of the CPU cache, optimising for different classes of features and efficiently
-multithreading.  Similarly, \ranger~\cite{Wright.and.Ziegle.2016}, written in C++ with an R front-end, is
-specifically optimised for dealing with large data sets.
+The most popular random forest software packages to date are implemented in R \cite[]{R-language}, although it has been
+reported that current implementations do not scale well with increasing feature size~\cite{Wright.and.Ziegle.2016}.
+This is despite R now supporting long vectors (length greater than $2^{31}-1$), and the \randomforest implementation in
+R lending itself to parallelisation by growing multiple trees simultaneously and combining them for the final
+result~\cite{Liaw.and.Weiner.2002}.  More suitable implementations for ``wide'' genomic data have been developed in
+other programming languages. \textsc{CloudForest} \cite{Bressler2015}, written in Go, achieves fast run times by
+effective use of the CPU cache, optimising for different classes of features and efficiently multithreading.  Similarly,
+\ranger~\cite{Wright.and.Ziegle.2016}, written in C++ with an R front-end, is specifically optimised for dealing with
+large data sets.
 
@@
 scale beyond compute-node boundaries and hence enable more sophisticated parallelisation strategies. In the case of
-random forests, the computations for each node of a tree can hence be handed off to separate processors.
+random forests, the computations for each node of a tree can hence be handed off to separate processors (cores).
 
@@
 \subsubsection{1000 Genomes Project}
-We obtained the 1000 Genomes Project data as VCF files from their FTP site.  Each VCF file contains the variants for
+We obtained the 1000 Genomes Project data as VCF files from their FTP site \cite[]{1KG2012}.  Each VCF file contains the variants for
 every individual across one chromosome.  We use the phase 3 dataset which contains 2,504 individuals with approximately

### 2017-05-19 doc/publications/IJCAI2017/IJCAI.tex
added conference paper

--- 2017-03-16 doc/publications/plos16/plos16.tex
@@
-% Template for PLoS
-% Version 3.4 January 2017
-%
-% % % % % % % % % % % % % % % % % % % % % %
-%
-% -- IMPORTANT NOTE
-%
-% This template contains comments intended 
-% to minimize problems and delays during our production 
-% process. Please follow the template instructions
-% whenever possible.
-%
-% % % % % % % % % % % % % % % % % % % % % % % 
-%
-% Once your paper is accepted for publication, 
-% PLEASE REMOVE ALL TRACKED CHANGES in this file 
-% and leave only the final text of your manuscript. 
-% PLOS recommends the use of latexdiff to track changes during review, as this will help to maintain a clean tex file.
-% Visit https://www.ctan.org/pkg/latexdiff?lang=en for info or contact us at latex@plos.org.
-%
-%
-% There are no restrictions on package use within the LaTeX files except that 
-% no packages listed in the template may be deleted.
-%
-% Please do not include colors or graphics in the text.
-%
-% The manuscript LaTeX source should be contained within a single file (do not use \input, \externaldocument, or similar commands).
-%
-% % % % % % % % % % % % % % % % % % % % % % %
-%
-% -- FIGURES AND TABLES
-%
-% Please include tables/figure captions directly after the paragraph where they are first cited in the text.
-%
-% DO NOT INCLUDE GRAPHICS IN YOUR MANUSCRIPT
-% - Figures should be uploaded separately from your manuscript file. 
-% - Figures generated using LaTeX should be extracted and removed from the PDF before submission. 
-% - Figures containing multiple panels/subfigures must be combined into one image file before submission.
-% For figure citations, please use "Fig" instead of "Figure".
-% See http://journals.plos.org/plosone/s/figures for PLOS figure guidelines.
-%
-% Tables should be cell-based and may not contain:
-% - spacing/line breaks within cells to alter layout or alignment
-% - do not nest tabular environments (no tabular environments within tabular environments)
-% - no graphics or colored text (cell background color/shading OK)
-% See http://journals.plos.org/plosone/s/tables for table guidelines.
-%
-% For tables that exceed the width of the text column, use the adjustwidth environment as illustrated in the example table in text below.
-%
-% % % % % % % % % % % % % % % % % % % % % % % %
-%
-% -- EQUATIONS, MATH SYMBOLS, SUBSCRIPTS, AND SUPERSCRIPTS
-%
-% IMPORTANT
-% Below are a few tips to help format your equations and other special characters according to our specifications.
-For more tips to help reduce the possibility of formatting errors during conversion, please see our LaTeX guidelines at http://journals.plos.org/plosone/s/latex
-%
-% For inline equations, please be sure to include all portions of an equation in the math environment.
-For example, x$^2$ is incorrect; this should be formatted as $x^2$ (or $\mathrm{x}^2$ if the romanized font is desired).
-%
-% Do not include text that is not math in the math environment. For example, CO2 should be written as CO\textsubscript{2} instead of CO$_2$.
-%
-% Please add line breaks to long display equations when possible in order to fit size of the column. 
-%
-% For inline equations, please do not include punctuation (commas, etc) within the math environment unless this is part of the equation.
-%
-% When adding superscript or subscripts outside of brackets/braces, please group using {}.  For example, change "[U(D,E,\gamma)]^2" to "{[U(D,E,\gamma)]}^2". 
-%
-% Do not use \cal for caligraphic font.  Instead, use \mathcal{}
-%
-% % % % % % % % % % % % % % % % % % % % % % % % 
-%
-% Please contact latex@plos.org with any questions.
-%
-% % % % % % % % % % % % % % % % % % % % % % % %
+\documentclass[a4paper]{amsart}
+\usepackage{geometry}                % See geometry.pdf to learn the layout options. There are lots.
+\geometry{letterpaper}                   % ... or a4paper or a5paper or ... 
+%\geometry{landscape}                % Activate for for rotated page geometry
+%\usepackage[parfill]{parskip}    % Activate to begin paragraphs with an empty line rather than an indent
+\usepackage{graphicx}
+\usepackage{amssymb}
+\usepackage{epstopdf}
+\usepackage{subcaption}
 
-\documentclass[10pt,letterpaper]{article}
-\usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}
-
-% amsmath and amssymb packages, useful for mathematical formulas and symbols
-\usepackage{amsmath,amssymb}
-
-% Use adjustwidth environment to exceed column width (see example table in text)
-\usepackage{changepage}
-
-% Use Unicode characters when possible
-\usepackage[utf8x]{inputenc}
-
-% textcomp package and marvosym package for additional characters
-\usepackage{textcomp,marvosym}
-
-% cite package, to clean up citations in the main text. Do not remove.
-\usepackage{cite}
-
-% Use nameref to cite supporting information files (see Supporting Information section for more info)
-\usepackage{nameref,hyperref}
-
-% line numbers
-\usepackage[right]{lineno}
-
-% ligatures disabled
-\usepackage{microtype}
-\DisableLigatures[f]{encoding = *, family = * }
-
-% color can be used to apply background shading to table cells only
-\usepackage[table]{xcolor}
-
-% array package and thick rules for tables
-\usepackage{array}
-
-% create "+" rule type for thick vertical lines
-\newcolumntype{+}{!{\vrule width 2pt}}
-
-% create \thickcline for thick horizontal lines of variable length
-\newlength\savedwidth
-\newcommand\thickcline[1]{%
-  \noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
-  \cline{#1}%
-  \noalign{\vskip\arrayrulewidth}%
-  \noalign{\global\arrayrulewidth\savedwidth}%
-}
-
-% \thickhline command for thick horizontal lines that span the table
-\newcommand\thickhline{\noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
-\hline
-\noalign{\global\arrayrulewidth\savedwidth}}
-
-% Remove comment for double spacing
-%\usepackage{setspace} 
-%\doublespacing
-
-% Text layout
-\raggedright
-\setlength{\parindent}{0.5cm}
-\textwidth 5.25in 
-\textheight 8.75in
-
-% Bold the 'Figure #' in the caption and separate it from the title/caption with a period
-% Captions will be left justified
-\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off]{caption}
-\renewcommand{\figurename}{Fig}
-
-% Use the PLoS provided BiBTeX style
-\bibliographystyle{plos2015}
-
-% Remove brackets from numbering in List of References
-\makeatletter
-\renewcommand{\@biblabel}[1]{\quad#1.}
-\makeatother
-
-% Leave date blank
-\date{}
-
-% Header and Footer with logo
-\usepackage{lastpage,fancyhdr,graphicx}
-\usepackage{epstopdf}
-\pagestyle{myheadings}
-\pagestyle{fancy}
-\fancyhf{}
-\setlength{\headheight}{27.023pt}
-\lhead{\includegraphics[width=2.0in]{PLOS-submission.eps}}
-\rfoot{\thepage/\pageref{LastPage}}
-\renewcommand{\footrule}{\hrule height 2pt \vspace{2mm}}
-\fancyheadoffset[L]{2.25in}
-\fancyfootoffset[L]{2.25in}
-\lfoot{\sf PLOS}
-
-%% Include all macros below
-
-\newcommand{\lorem}{{\bf LOREM}}
-\newcommand{\ipsum}{{\bf IPSUM}}
-
-%% END MACROS SECTION
-
-%% Additional packages and local edits
-\usepackage{footmisc}
-\usepackage{multirow}
-\usepackage[T1]{fontenc}
-%% required the installation of cm-super 55MB - TeX font package (full version) with CM (EC) in Type1 in T1, T2*, TS1, X2 enc
-
-\let\oldmarginpar\marginpar
-\renewcommand\marginpar[1]{\-\oldmarginpar[\raggedleft\footnotesize #1]%hv
-{\raggedright\footnotesize #1}}
-\reversemarginpar
-
-%\setlength{\marginparwidth}{5cm}
-\usepackage{caption}
-\usepackage{subcaption}
-\usepackage{placeins}
-\usepackage{xspace} % see http:
-%\usepackage{showframe}
+\DeclareGraphicsRule{.tif}{png}{.png}{`convert #1 `dirname #1`/`basename #1 .tif`.png}
 
@@
 
+\title{CursedForest - A Random Forest Implementation for ``Big'' and ``Wide'' Data}
+%\section{}
+%\subsection{}
 \begin{document}  
-\vspace*{0.2in}
+\maketitle
 
-% Title must be 250 characters or less.
-\begin{flushleft}
-{\Large
-\textbf\newline{\cursedforest - A Random Forest Implementation for ``Big'' and ``Wide'' Data} % Please use "title case" (capitalize all terms in the title except conjunctions, prepositions, and articles).
-}
-\newline
-% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
-\\
-Aidan O'Brien\textsuperscript{1,2},
-Piotr Szul\textsuperscript{3},
-Robert Dunne\textsuperscript{4}, 
-Paul Leo \textsuperscript{5},
-Emma L. Duncan \textsuperscript{5,6,7},
-Natalie Twine\textsuperscript{1,8}
-Stephanie Li\textsuperscript{9},
-James Doecke\textsuperscript{9},
-Nick Ellis\textsuperscript{10}, and
-Denis C. Bauer\textsuperscript{1,*}
-%, with the Lorem Ipsum Consortium\textsuperscript{\textpilcrow}
-\\
-\bigskip
-\textbf{1} Health \& Biosecurity, CSIRO, Sydney, NSW, Australia
-\\
-\textbf{2} ANU, Canberra, Australia
-\\
-\textbf{3} Data61, CSIRO, Brisbane, QLD, Australia
-\\
-\textbf{4} Data61, CSIRO, Sydney, NSW, Australia
-\\
-\textbf{5} Institute of Health and Biomedical Innovation, Queensland University of Technology, Brisbane, QLD, Australia
-\\
-\textbf{6} Department of Endocrinology, Royal Brisbane and Women's Hospital, Brisbane, QLD, Australia
-\\
-\textbf{7} Faculty of Medicine and Biomedical Sciences, University of Queensland, Brisbane, QLD, Australia
-\\
-\textbf{8} MQ, QLD, Australia
-\\
-\textbf{9} Health \& Biosecurity, CSIRO, Brisbane, QLD, Australia
-\\
-\textbf{10} Oceans \& Atmosphere, CSIRO, Brisbane, QLD, Australia
-\bigskip
+\begin{abstract}
+Genomic data analyses are performed on ever larger patient cohorts with unprecedentedly detailed genome-wide information for each individual.
+Machine learning (ML) is employed to detect the complex genomic interactions that can lead to diseases like diabetes or cancer.
+However, current ML approaches are unable to cope with these data volumes and even implementations using novel compute paradigms, such as Hadoop/Spark, are limited by the finite resource of computer memory.
+Specifically, Spark ML?s supervised machine learning methods suffer from "the curse of dimensionality".
+That is, they scale well with samples (n) but not with features (p, variants).
+We introduce CursedForest, a tailored implementation of random forests, designed to handle data with extremely large number of variables per sample.
+We showcase CursedForest to accurately predict ethnicity from whole genome sequencing profiles of 2,500 individuals and 80 million genomic variants each (1000 Genomes Project) in 30 minutes.
+CursedForest was the only implementation scaling to the full dataset, outperforming other methods on smaller subsets by 80\% or more.
+We also demonstrate a GWAS-style analysis using CursedForest?s feature selection approach on an imputed SNP array dataset (2000 x 7 Million) on Bone mineral density (BMD), successfully predicting known BMD-genes as well as identifying novel candidates.
+CursedForest is included in our earlier genome interpretation package, VariantSpark, allowing it to perform near real-time classification of population-scale patient cohorts in "patients like mine" scenarios, as well as performing GWAS analysis on large unfiltered whole genome sequencing cohorts.
+\end{abstract}
 
-% Insert additional author notes using the symbols described below. Insert symbol callouts after author names as necessary.
-% 
-% Remove or comment out the author notes below if they aren't used.
-%
-% Primary Equal Contribution Note
-%\Yinyang These authors contributed equally to this work.
-
-% Additional Equal Contribution Note
-% Also use this double-dagger symbol for special authorship notes, such as senior authorship.
-%\ddag These authors also contributed equally to this work.
-
-% Current address notes
-%\textcurrency a Insert current address of first author with an address update
-% \textcurrency b Insert current address of second author with an address update
-% \textcurrency c Insert current address of third author with an address update
-
-% Deceased author note
-%\dag Deceased
-
-% Group/Consortium Author Note
-%\textpilcrow Membership list can be found in the Acknowledgments section.
-
-% Use the asterisk to denote corresponding authorship and provide email address in note below.
-* Denis.Bauer@CSIRO.au
-\end{flushleft}
-
-%\tableofcontents
-%\clearpage
-
-% Please keep the abstract below 300 words
-\section{Abstract}
-
-Analyses are performed on ever larger datasets with unprecedentedly detailed information per data point. Specifically in
-genomics, cohort sizes are in the thousands, each comprising of millions of variants that make up a person's genetic
-differences. Traditional approaches are unable to cater for this task and even standard implementations using novel
-compute paradigms, such as Hadoop/Spark, are limited by the finite resource of computer memory on such
-data. Specifically, Spark ML's supervised machine learning methods suffer from "the curse of dimensionality". That is,
-they scale well with samples ($n$) but not with features ($p$, variants).
-
-We introduce \cursedforest, a tailored implementation of random forests, designed to handle data with extremely large
-number of variables per sample. We showcase \cursedforest\ to accurately predict ethnicity from whole genome sequencing
-profiles of 2,500 individuals and 80 million genomic variants each (1000 Genomes Project) in 30 minutes. \cursedforest\
-was the only implementation scaling to the full dataset, outperforming other methods on smaller subsets by 80\% or
-more. We also demonstrate a GWAS-style analysis using \cursedforest's feature selection approach on an imputed SNP array
-dataset (2000 x 7 Million) on Bone mineral density (BMD), successfully predicting known BMD-genes as well as identifying
-novel candidates.
-
-\cursedforest is included in our earlier variant interpretation package, VariantSpark, allowing it to perform near
-real-time classification of population-scale patient cohorts in "patients like mine" scenarios, as well as performing
-GWAS analysis on large unfiltered whole genome sequencing cohorts.
-
-\linenumbers
+%\twocolumn
 
@@
 do not scale well with increasing feature size~\cite{Wright.and.Ziegle.2016}.  This is despite R now supporting long vectors (length greater
-than $2^{31}-1$), and the \randomforest implementation in R lending itself to parallelisation by growing multiple
+than $2^{31}-1$), and the random forest implementation in R lending itself to parallelisation by growing multiple
 trees simultaneously and combining them for the final result~\cite{Liaw.and.Weiner.2002}.  More suitable implementations for ``wide''
@@
 achieves fast run times by effective use of the CPU cache, optimising for different classes of features and efficiently
-multithreading.  Similarly, \ranger~\cite{Wright.and.Ziegle.2016}, written in C++ with an R front-end, is
+multithreading.  Similarly, ranger~\cite{Wright.and.Ziegle.2016}, written in C++ with an R front-end, is
 specifically optimised for dealing with large datasets.
@@
 
-Here we introduce \cursedforest, a tailored Hadoop/Spark-based implementation of random forests specifically designed to cater
+Here we introduce CursedForest, a tailored Hadoop/Spark-based implementation of random forests specifically designed to cater
 for ``big'' (many samples) and ``wide'' (many features) datasets. 
-\cursedforest extends our previously developed variant interpretation framework, VariantSpark~\cite{OBrien2015}, to now offer
+CursedForest extends our previously developed variant interpretation framework, VariantSpark~\cite{OBrien2015}, to now offer
 supervised as well as unsupervised ML algorithms in the Spark framework.
@@
 By also utilising Spark to read in and manipulate the standard genomic variant format (VCF)
-directly, \cursedforest outperforms existing tools even on small datasets where multithreading generally performs
-well. Harnessing the virtually unlimited capability to parallelise tasks, \cursedforest can hence explore the solution 
+directly, CursedForest outperforms existing tools even on small datasets where multithreading generally performs
+well. Harnessing the virtually unlimited capability to parallelise tasks, CursedForest can hence explore the solution 
 space faster by building a larger number of diverse models to generate a consensus from.
 
-Using this facility, \cursedforest is capable of parallelising the split for each node in a tree thereby handling millions 
+Using this facility, CursedForest is capable of parallelising the split for each node in a tree thereby handling millions 
 of features, as required to process whole genome sequencing data or SNP array data with unobserved genotypes 
@@
 
-%It can, for example, fit a model to the 1000 Genomes Project data, which consists of approximately 2500 samples with up to 80 million
-%variants~\cite{1KG2012}.  
-%build a larger number of
-%trees hence building a larger number of diverse models to generate the consensus from.
-%and use a high \mtry value which may lead to greater accuracy (see the example in section \ref{section:synthetic}).
-%\marginpar{does \cite{Szymczak2016} say that high \ntree gives greater  accuracy for variable importance selection?
-%Check this. }
-%We previously demonstrated the versatility and scalability of Spark by developing VariantSpark~\cite{OBrien2015}, a
-%framework allowing users to easily analyze Variant Call Format (VCF) files using ML algorithms on the Spark framework.
-%Using VariantSpark, we successfully built a k-means model on the full $2500 \times 80$ million matrix to cluster
-%individuals by their ethnicity achieving an Adjusted Rand Index (ARI) of 0.84 (with 1 being perfect clustering and -1
-%random clustering).
-%Using VariantSpark, we successfully built a k-means model on the full $2500 \times 80$ million matrix to cluster
-%individuals by their ethnicity achieving an Adjusted Rand Index (ARI) of 0.84 (with 1 being perfect clustering and -1
-%random clustering).
-%Here we extend this toolkit to encompass supervised learning using random forests.
-
-VariantSpark~\cite{OBrien2015} with the \cursedforest extension therefore offers a
+VariantSpark~\cite{OBrien2015} with the CursedForest extension therefore offers a
 comprehensive analysis toolkit that can scale to future data demands. To showcase the framework's ability we demonstrate
 a classification as well as feature-selection task on synthetic data in the first section. In the second section, we demonstrate the ability of
-\cursedforest to successfully replicate findings from a previous GWAS study as well as identify novel variants associated with bone 
+CursedForest to successfully replicate findings from a previous GWAS study as well as identify novel variants associated with bone 
 mineral density (BMD). 
-Thirdly, we demonstrate the scalability of \cursedforest in respect to the dimensionality of data by building a random
+Thirdly, we demonstrate the scalability of CursedForest in respect to the dimensionality of data by building a random
 forest model on whole-genome data from the 1000 Genomes Project~\cite{1KG2012} to predict ethnicity.
@@
 \section{Methods}
-\label{section:methods}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Overview of Random Forests}
-
-Decision trees have a number of desirable features, they:
-\begin{itemize}
-  \item are invariant to monotonic scaling of the data;
-  \item can handle categorical and real valued inputs;
-  \item can handle missing values;
-  \item are insensitive to uninformative predictors;
-  \item are able to capture interactions between features (which is important for modeling complex polygenic diseases).
-  \end{itemize}
-However, the tree fitting algorithm is greedy and may generate an unstable model. That is, a small change in the data may
-lead to a very different model.
-
-Random forests apply the technique of bootstrap aggregation or bagging to decision trees.  The training datasets are
-independently drawn, with replacement, from the joint distribution of $(X,Y)$ so that each set comprises $n$
-$(p+1)$-tuples $(x_1,y_1),\ldots, (x_n,y_n)$. This is done $B$ times and a tree is grown on each of these $B$ samples.
-The results are then combined in an appropriate way. In a classification problem we take the majority vote for the trees
-over the $q=1,..,Q$ classes,
-\begin{equation*}
-{{h_f}}= \arg \max_q \left(\sum_b I(h(x;B_b)=q)\right).
-\end{equation*}
-In addition to growing the trees
-on bootstrapped samples, each node in the tree is split on a randomly
-selected subset of the variables.
-
-Random forests are a variance reduction technique. They take unpruned tree models, which have a low bias but a high
-variance, and by combining them reduce the variance.  For regression problems, as the prediction error is the sum of the
-variance and the bias squared, this strategy should lead to a low prediction error.
-
-For classifiction problems there is no obvious decomposition of classification error into an equivalent bias plus
-variance form. A number of different additive decompositions of classification error have been considered in an effort
-to understand the success of bagging. They are all less transparent than the regression case (see \cite{Friedman.1997} and
-citations therein).
-
-In the case of a regression tree, the average prediction error for an individual tree $h(X; B)$ is
-\begin{equation}
-PE_t = E_B E_{X,Y} (Y-h(X; B))^2.
-\end{equation}
-Assume that, for all $B$, the tree is unbiased, i.e., $EY= E_X h(X; B)$. Then
-\begin{equation}
-PE_f \leq \rho PE_t
-\end{equation}
-where $\rho$ is the weighted correlation between residuals $(Y-h(X;B))$ and $(Y-h(X;B^\prime))$ for independent
-$B,B^\prime$.
-So what is required for accurate random forest regression is (i) low correlation between residuals of differing tree in
-the forest, and (ii) low prediction error for the individual trees \cite{Segal.2004}.
-
-The benefits of random forest models are:
-\begin{itemize}
-\item they give an out-of-bag (OOB) estimate of model accuracy, by using the samples not selected for the bootstrapped sample
-  as an independent test set for each tree;
-\item they provide a measure of variable importance, for example by summing the number of times a variable is used to
-  split a node, over all nodes and all trees. More complex measures of variable importance are also used;
-\item they inherit some of the qualities of tree models (but not, for example, the handling of missing values via surrogate
-  variables. This would be difficult as the trees are split on random subsets of variables).
-\end{itemize}
-
-Since the original implementation of random forests, there have been a number of developments leading to a better
-understanding of the algorithm.
-\begin{itemize}
-\item random forests are resistant to overfitting. However it is not true that they will not overfit,
-  ~\cite{Segal.2004} gives examples of overfitting in cases where:
-  \begin{itemize}
-  \item a single tree is the most appropriate model; and
-  \item the variables are highly correlated so that the bootstrap resampling does not give rise to uncorrelated trees.
-  \end{itemize}
-However, keeping these difficult cases in mind, random forests have a strong claim to being resistant to overfitting.
-\item \cite{Strobl.et.al.2007} demonstrates that the random forest algorithm is subject to a bias in variable selection
-  via two mechanisms:
-  \begin{enumerate}
-  \item differences in variable scale (for $X_i$ continuous) and number of levels (for $X_i$ discrete) introduces a
-    bias. An uninformative variable with a large number of levels may be selected over a more informative variable with
-    fewer levels;
-  \item bagging introduces a bias, but sampling $(1- 1/e) \approx 0.632$ of the data without replacement seems to be an
-    effective strategy for removing the problem.
-  \end{enumerate}
-\end{itemize}
-
-\cite{Strobl.et.al.2007} introduce changes to the random forest algorithm to fix 1) and 2).  Effect 1) will be apparent
-in the \cursedforest algorithm as well.  However, there are many examples, particularly in genomics, where we have data
-sets that are:
-\begin{itemize}
-\item very wide;
-\item all variables have the same number of levels (for example, measurements at different points along the genome);
-\item the hypothesized functional relationship is of a form amenable to a decision tree.
-\end{itemize}
-GWAS datasets are an obvious application area for this algorithm.
-
-\cite{Biau.2012} provides a proof that a random forest model will converge at a rate that depends on
-the cardinality of the set $S$ of ``strong predictors'' rather than on the number of variables $p$. That is, given a
-function $y=f(S)$ depending on a set of variables $S$, a random forest will converge to the true function $y$ with a
-rate that depends on the size of $S$ and not the number of noise variables in the dataset. If the size of $S$ is small
-compared to the total number of variables, then the rate of convergence will be quicker. It is not clear what the size
-of the set $S$ will be in a GWAS application. There are cases of small groups of SNPs with large effects and also of
-conditions caused by the interactions of larger groups of SNPs.
-
-\cite{Genuer.et.al.2010} and \cite{Diaz.and.Alvarez.2006} describe iterative schemes for doing variable selection using
-the importance measure. See \cite{Chen.and.Ishwaran.2012} for a review of the area with particular reference to genomic
-data.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{\cursedforest}
-VariantSpark made use of Spark's machine learning algorithms ``Spark ML''. While Spark ML algorithms demonstrate
-scalability when dealing with a large number of samples, this scalability eventually breaks down as we include more
-features.
-
+\subsection{CursedForest}
 As is standard with Spark applications, we store our data in a Resilient Distributed Dataset (RDD), where an RDD is
@@
 
-On the other hand, \cursedforest is specifically designed to handle wide ``cursed'' data. It avoids the relation between
+On the other hand, CursedForest is specifically designed to handle wide ``cursed'' data. It avoids the relation between
 memory and dimensionality by avoiding calculations that rely on entire feature vectors and taking the parallelization
-work down to the level of the individual features.  For each node of a tree, \cursedforest will distribute
+work down to the level of the individual features.  For each node of a tree, CursedForest will distribute
 tasks that consist of single features (variants), for every individual.  Each of these tasks will calculate the
 information gain for that specific feature.  Once these tasks have completed, the results are reduced to return the
-feature which gives the greatest information gain.  This process is then repeated until \cursedforest has created the
+feature which gives the greatest information gain.  This process is then repeated until CursedForest has created the
 entire decision tree.
 
-The current implementation of \cursedforest uses a ``Gini impurity'' criteria for splitting. Let $f_q$ be the fraction
+The current implementation of CursedForest uses a ``Gini impurity'' criteria for splitting. Let $f_q$ be the fraction
 of items labeled with value $q$, where $q = 1, \ldots, Q$ at a node. The Gini impurity is
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% This section talks about the datasets
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Datasets}
-\subsubsection{1000 Genomes Project}
-We obtained the 1000 Genomes Project data as VCF files from their FTP site.  Each VCF file contains the variants for
-every individual across one chromosome.  We use the phase 3 dataset which contains 2,504 individuals with approximately
-81 million variant positions (i.e. a matrix of $2,504 \times 81$ million).  For compatibility with the majority of the
-algorithms we demonstrate, we convert the original variant representations to double representations.  For example, we
-store the absence of a variant (0\textbar0) as 0, a heterozygous
-variant (1\textbar0 or 0\textbar1) as 1, and a
-homozygous variant (1\textbar1) as 2.  However, this is not the only possible encoding \cite{Goldstein.et.al.2011}. Note that from
-the encodings listed in Table~\ref{table:encodings}, the additive encoding will present the most difficulty for a random
-forest model.
-
-\begin{table}[!ht]
-%\begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
-\begin{flushleft} 
-\centering
-\caption{\bf Possible encodings of GWAS data \cite{Goldstein.et.al.2011}}
-\begin{tabular}{|l|l|l|l||}
-\hline
-{\bf Type} & {\bf Mechanism}                                         & {\bf Partition} \\ 
-\thickhline
-Additive   & Each additional minor allele increases variation        & 0, 1, 2         \\
-Dominant   & Presence of at least 1 minor allele increases variation & 0, $\{1,2\}$          \\
-Recessive  & Two minor alleles needed for variation                  & $\{0,1\}$, 2          \\
-Heterosis  & Heterozygote leads to variation                         & $\{0,2\}$, 1          \\ \hline
-%\end{adjustwidth}
-\end{tabular}
-\label{table:encodings}
-\end{flushleft}
-\end{table}
-
-\subsubsection{Bone mineral density dataset}
-We obtained a VCF file from the authors of \cite{Duncan.2011}, which contains variant data from 1,936 postmenopausal women.
-This dataset covers approximately 7 million variant positions. We have information about the bone mineral density (BMD)
-for each individual, where a low BMD is a major predisposing factor to fracture \cite{Duncan.2011}.
-
-\subsubsection{Synthetic data} 
+\subsection{Synthetic data} 
 \label{section:synthetic_data}
@@
 \subsection{Parameter settings}
-We consider the parameter settings for the random forest algorithm. We use the R notation from the \randomforest
+We consider the parameter settings for the random forest algorithm. We use the R notation from the random forest
 package \cite{Liaw.and.Weiner.2002} which incorporates the original Fortran code by Brieman and Cutler.  We incorporate
@@
 \begin{itemize}
-\item \ntree{}  -- the number of trees.  The number of trees necessary for good performance grows with the number of
-  predictors.  \cite{Liaw.and.Weiner.2002} suggest that a high \ntree is necessary to get stable estimates of variable
+\item ntree  -- the number of trees.  The number of trees necessary for good performance grows with the number of
+  predictors.  \cite{Liaw.and.Weiner.2002} suggest that a high ntree is necessary to get stable estimates of variable
   importance and proximity; however, even though the variable importance measures may vary from run to run, we note that
@@
   importance;
-\item \mtry{}  -- the number of variables considered at each split (if \mtry=$p$, we have a boosted decision
-  tree model).  If one has a very large number of variables but expects only very few to be ``important'', using larger \mtry may give
+\item mtry  -- the number of variables considered at each split (if mtry=$p$, we have a boosted decision
+  tree model).  If one has a very large number of variables but expects only very few to be ``important'', using larger mtry may give
   better performance;
-\item the size and complexity of the individual trees is controlled in \randomforest by setting \texttt{nodesize}, the
+\item the size and complexity of the individual trees is controlled in random forest by setting \texttt{nodesize}, the
   minimum size of terminal nodes. It is controlled in Spark ML by setting \texttt{maxDepth}, the maximum depth of each
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%
-% Results
-%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Results and Discussion}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Section 1:  Feature Selection and classification on synthetic data
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{\cursedforest can evaluate variable importance for millions of features}
-\label{section:synthetic}
-In this section we demonstrate the two features of \cursedforest, namely: variable selection analysis to select the most
-associated set of features amongst millions of dimensions; and its ability to accurately classify the labels of
-high-dimensional samples.  We therefore generate a synthetic dataset with 2.5 million variables ($p=2,500,000$), of which 5 
-are related to the response variable, and 5000 samples ($n=5000$), as discussed in the methods section.
-
-We fit the random forest model and estimate the classification accuracy by capturing the out-of-bag error (OOB) and the
-variable selection performance by capturing the rank-biased overlap (RBO) \cite{Webber2010} measure. RBO compares two
-lists (in this case, the parameters $w_i, i=1,\ldots , 5$ (Equation~\ref{equation:synthetic.data}), and the top 1000
-ranked variable importances from the \cursedforest model). It is a mesaure of similarity between rankings that handles
-lists of different lengths, weighs high ranks more heavily than low, and is monotonic with increasing depth of
-evaluation. We expect to retrieve the 5 features in order of the weight that was assigned to them and with a higher
-variable importance measure than the noise variables. RBO provides a quantiative measure of how close we come to this on
-a scale from 0 (no feature recovered) to 1 (fully recovered). See supplementary information (S*) for more details.
-
-We are running Apache Spark 1.6.1 on a YARN
-cluster with 12 worker nodes each with 16 Intel Xeon E5-2660@2.20GHz CPU cores and 128 GB of RAM.
-
-%mtry = number of variable seclection at each node - proportion of p 
-%ntree= number of trees
-
-% mtry large makes trees correlated because the same important features have a higher chance of being selected
-% overfitting because it picks random that happen to be
-
-As shown in Fig~\ref{figure:out-of-bag-prediction-error-prod.png} the default value for \mtry (number of variables
-selected at each node) does not result in a good classification performance for this large feature dataset. Note that
-the plot shows the proportion, $n/p,$ for \mtry\ with the default being $\sqrt{p} \approx 1581$ ($n/p=0.0006$).  The
-OOB for this value of \mtry does not drop below 0.5, even when the number of trees are increased (\ntree).
-
-Increasing \mtry\ in combination with \ntree\ yields the best performance with the OOB error essentially constant around 0.4
-across a large range of of \mtry and \ntree values.  This is in contrast to the feature-selection performance, where the
-RBO measure heavily depends on \ntree\ and gives better results with lower values of \mtry.
-
-This may be because a large \mtry leads to more correlated trees as the same important features have a higher chance of
-being selected in all, not yielding good performance outcomes.  This issue is less pronounced for classification error
-where random features can mimic the response variable, hence resulting more in a performance plateau.  Increasing the
-number of trees on the other hand improves performance especially when the trees are kept diverse (small \mtry) but
-appropriate for large feature datasets (\mtry larger than default). \cursedforest's feature of parallelizing trees at
-the node level hence caters perfectly to the requirement of large feature datasets as more trees can be built given a
-fixed time budget.
-
-%Fig~\ref{figure:synth1} illustrates the effect of varying \mtry and \ntree on these two measures. The default \mtry
-%value for a classification problem is $\sqrt{p} \approx 1581$ so in this case we are considering values considerably larger than
-%the default. Fig~\ref{figure:out-of-bag-prediction-error-prod.png} shows the OOB error is constant across a large range of 
-%of \mtry and \ntree values whereas Fig~\ref{figure:rbo-prod.png} illustrates that the RBO measure depends
-%heavily on \ntree but gives a better value with lower values of \mtry.
-
-From the synthetic data we can conclude that \cursedforest is able to perform accurate feature selection as well as
-accurate classification for large numbers of samples with large feature vectors.  We can also conclude that fine tuning
-the model to accurately classify the dataset does not ensure good feature selection performance as there seems to be
-little correspondence between the OOB and RBO measure.
-
-\begin{figure}[tbhp] 
-  \begin{adjustwidth}{-1.00in}{0in}
-    \caption{\textbf{Out of bag classification error and rank-biased-overlap estimates}} 
-    \label{figure:synth1}
-    \begin{subfigure}[b]{0.5\linewidth}
-      \centering
-      \includegraphics[totalheight=8cm]{./figs/oob_levleplot.png}
-      \caption{OOB classification error.} 
-      \label{figure:out-of-bag-prediction-error-prod.png} 
-%      \vspace{4ex}
-    \end{subfigure} 
-    \begin{subfigure}[b]{0.5\linewidth}
-      \centering
-      \includegraphics[totalheight=8cm]{./figs/rbo_levleplot.png}
-      \caption{The rank-biased-overlap measure.} 
-      \label{figure:rbo-prod.png} 
-%      \vspace{4ex}
-    \end{subfigure} 
-    \begin{flushleft} 
-      Random forests with 100 trees were trained on the synthetic dataset of 2.5M features and 5K samples.
-      Each measure is shown as a function of \mtry and \ntree. Note that the plot colors have been chosen so that the lighter color indicates
-      the better result, that is, a lower OOB or a higher RBO.
-    \end{flushleft}
-  \end{adjustwidth}
-\end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Section 2: Feature Selection on real data 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{GWAS-style analysis using \cursedforest\ identifies novel variants associated with Bone Mineral Density}
-In this section we use \cursedforest\ to perform feature selection on over 7.2 million genomic variants and identify the
-locations associated with Bone Mineral Density (BMD)~\cite{Duncan.2011}. 
-%We build a model on the BMD dataset as described in Methods using a binary high or low BMD value as the label.  
-We rank variants by their Gini impurity, with the most important one ranked at position 1, and return the top 100,000 variants.
-
-To test whether the associated variants have biological meaning, we investigate if variants in or near
-BMD-associated genes are enriched toward the top of our ranked list. To form the list of "known genes", we
-combine the 26 known and 6 novel BMD genes listed by Duncan {\it et al.}~\cite{Duncan.2011} with the genes identified by
-Estrada {\it et al.}~\cite{Estrada2012} and Lui {\it et al.}~\cite{Liu2008}, resulting in 102 genes.  All genes are covered within the
-top 40,000 variants and a gene name is assigned to the variant if it falls within 100 kb of this gene.
-
-The known genes rank significantly higher than variants from other locations in the genome (Mann-Whitney U, $p=4.5\times10^{-14}$, Supplemental File 1), 
-supporting our hypothesis that biologically relevant features are contributing more information to our classification model.
-
-As shown in Figure~\ref{figure:knowngeneranks}, the top 30 features are dominated by variants from five well established BMD genes; TNFRSF11B (TNF Receptor Superfamily Member 11b or Osteoprotegerin), CTNNB (Catenin Beta 1), ZBTB40 (Zinc Finger And BTB Domain-Containing Protein 40), IBSP (Integrin Binding Sialoprotein) and MEF2C (Myocyte Enhancer Factor 2C).
-The BMD association has been independently replicated for four of the five genes: CTNNB1~\cite{Pei2016, Rivadeneira2009}, MEF2C~\cite{Pei2016a}, TNFRSF11B~\cite{Rivadeneira2009,Zhang2014,Kemp2014,Paternoster2013} and ZBTB40~\cite{Rivadeneira2009,  Zhang2014,Nielson2016}.
-The association of IBSP has not been replicated in other GWAS studies, but homozygous knockouts of this gene have a known skeletal phenotype in mice.
-Mice deficient in IBSP have impaired bone growth and mineralization, with dramatically reduced bone formation~\cite{Malaval2008}.
-CTNBB1, MEF2C and TNFRSF11B also have known skeletal phenotypes.
-CTNNB1 is a key downstream member of the Wnt signaling pathway, and deletion of CTNNB1 in mice results in severe osteopenia, while constitutive activation of CTNNB1 results in massively increased bone deposition~\cite{Holmen2005}.
-MEF2C is a transcription factor which controls bone mass by regulating osteoclastic bone resorption, where mice with a MEF2C deletion in osteocytes had increased bone mass compared with normal controls~\cite{Kramer2012}.
-TNFRSF11B (or OPG) is a secreted glycoprotein which regulates bone mass~\cite{Simonet1997}.
-ZBTB40 has no known biological or molecular involvement in skeletal phenotypes~\cite{Styrkarsdottir2008}, even though it has been associated with BMD in multiple GWAS and is expressed in bone.
-SNPs in linkage disequilibrium (LD) with each other at these loci also show a high importance score, as exemplified by TNFRSF11B in Figure 3b (Supplemental Figure 2 and 3 for the other genes).
-%~\ref{manhattenzoom}).
-
-%MAPT microtubule associated protein tau \cite{Dengler-Crish2017}
-%"MAPT is primarily expressed in the nervous system and has not been identified in bone tissue, therefore, low BMD in a tauopathy-specific mouse model that lacks A?
-pathology would provide evidence that central mechanisms of bone homeostasis are dysfunctional in AD, rather than only direct peripheral effects on bone tissue itself."
-%"Reduced bone mineral density (BMD) and its clinical sequelae, osteoporosis, occur at a much greater rate in patients with Alzheimer?s disease (AD), often emerging early in the disease before significant cognitive decline is seen.
-"
-
-In contrast, one of the lowest ranked gene with literature implied BMD association is DCDC5 (Doublecortin Domain Containing 5) ranked 23,842
-(see Figure 2).
-%~\ref{figure:knowngeneranks}).  
-Similar to the conclusions drawn by Duncan {\it et
-  al.}~\cite{Duncan.2011}, our results also indicate that this gene does not have a strong association with BMD.
-DCDC5 is not highly expressed in bone, and any mechanisms by which this gene might regulate BMD are unclear
-\cite{Thakker2012}.  Although variants near DCDC5 are associated with BMD \cite{Rivadeneira2009}, it is questionable as
-to whether the gene itself contributes to BMD.
-Similarly, the lowest ranked gene with annotated BMD links, MAPT (microtubule associated protein tau) at 36,262, is primarily expressed in the nervous system and has not been identified in bone tissue.
-It is therefore likely that the association is through a co-morbidity with Alzheimer's disease~\cite{Dengler-Crish2017} and not a major contributing factor in this cohort.
-
-%2 TEX2  EN rs112558362 SNORNA  ERN1 Gliablastoma in older people Alzheimer's
-%5 rs6961378  Woolfe2005
-%13 rs17195090
-
-There are also four novel locations amongst the top 30 most associated variants as annotated in Figure 3a.
-%~\ref{figure:rbo-prod.png}. 
-Most notably a lincRNA (LINC01006) on chromosome 7, which has been associated with serum creatinine levels in two independent GWAS studies~\cite{Pattaro2010,Chambers2010}. 
-Subjects with low serum creatinine levels were at a higher risk for low BMD~\cite{Huh2015} and in a rat study high creatine monohydrate diet resulted in greater lumbar BMD~\cite{Antolic2007}, 
-with creatinine being the breakdown product of creatine.
-The variants overlapping the 5' end of the lincRNA are the highest ranked features in this region and all LD SNPs also show high importance (see Fig 3d).
-Also amonst the top 30 features is PRODH (proline dehydrogenase) located on 22q11, a loci observed by Stagi {\it
-  et al.}~\cite{Stagi.2010} to significantly reduce bone mass in individuals with a micro-deletion in this region. 
-PRODH catalyses the first step in proline degradation, and as proposed by Phang {\it et al.}~\cite{Phang.2008}, proline is made available by collagen degradation. 
-Again, all LD SNPs also show high importance (see Figure 3c). %~\ref{zoomplot}). 
-%Ranked at position 5 (rs6961378) and 18 (rs146034056) are located on chromosome 7 overlapping the lincRNA (LINC01006), which as been linked to vertebrate development~\cite{Woolfe2005}.
-In contrast rs17195090 and rs112558362 are single feature loci (Supplementary Figure 4) hence likely due to erroneous imputation or genotyping.
-
-% Also amongst the top 20 associated
-%variants is a SNP in the intronic region of Unc13c (rs12916554), a gene that is linked to quality of
-%sleep~\cite{Nicod2016}, which in turn is reported to be the best predictor of osteoporosis risk~\cite{Magali2016}, a
-%condition diagnosed by measuring BMD.  Also listed is rs1035665, which was identified in a recent GWAS study to regulate
-%ABCF1 (ATP Binding Cassette Subfamily F Member 1)~\cite{Dixon2007} with member 2 of this subfamily being a known BMD
-%gene. The other three novel loci are on or near ncRNA.
-
-\cursedforest\ can successfully perform GWAS-type analyses replicating previous findings and uncovering novel
-associations due to random forests also modelling interacting features.
-
-%In Table~\ref{featuretable}, we display the top 10 features, ranked by their Gini impurity measure in the random forest model.
-
-%\begin{table}[!ht]
-%%\begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
-%\centering
-%\caption{
-%{\bf Features from the TCGA BRCA dataset ranked by their importance.}}
-%\begin{tabular}{|l|l|l|l|l|l|l|}
-%\hline
-%{\bf Variant} & {\bf Gene} & {\bf Importance}\\ \thickhline
-%8:120003727 & COLEC10 & 1.6393260970E-6\\ \hline
-%1:22702231  & ZBTB40  & 1.3901355155E-6\\ \hline
-%22:18923745 & PRODH   & 1.3769982791E-6\\ \hline
-%1:22702187  & ZBTB40  & 1.3350582926E-6\\ \hline
-%3:41111035  & CTNNB1  & 1.2707227479E-6\\ \hline
-%1:693731    &         & 1.1786957578E-6\\ \hline  
-%1:22700351  & ZBTB40  & 1.1547011076E-6\\ \hline
-%8:120040181 & COLEC10 & 1.1511986773E-6\\ \hline
-%1:22701761  & ZBTB40  & 1.1495539020E-6\\ \hline
-%3:41120417  & CTNNB1  & 1.1325242379E-6\\ \hline
-%\end{tabular}
-%\begin{flushleft} 
-%The ten genes proximal to the top variants as ranked by their Gini impurity from a random forest model built on
-%individuals using high/low bone mineral density as the label.
-%\end{flushleft}
-%\label{featuretable}
-%%\end{adjustwidth}
-%\end{table}
-
-\begin{figure}[tbhp]
-  \begin{adjustwidth}{-2.00in}{0in}
-    \caption{\textbf{Rank of literature assigned BMD genes}}
-    \label{figure:knowngeneranks}
-    \includegraphics[totalheight=6.5cm]{./figs/BMDgenes_landscape.pdf}
-    \begin{flushleft}
-      The figure shows the rank of the 100 literature assigned BMD genes.
-Each ball represents a SNP assigned to the closest gene with colour representing the distance (max 100 kb).
-The size shows the Gini impurity score with larger balls representing more important SNPs.
-The horizontal line separates the genes with SNPs amongst the Top 20 associated features.
-     \end{flushleft}
-  \end{adjustwidth}
-\end{figure}
-
-\begin{figure}[tbhp]
-  \begin{adjustwidth}{-2.00in}{0in}
-    \caption{\textbf{Genome-wide importance of SNPs and visualisation of the 22q.11 region}}
-    \label{figure:ranksumtest}
-     \begin{subfigure}[b]{0.5\linewidth}
-     \includegraphics[totalheight=7cm]{./figs/BMDTopWholeGenome.png}
-      \label{figure:rbo-prod.png} 
-    \end{subfigure} 
- \begin{subfigure}[b]{0.5\linewidth}
-      \centering
-      \includegraphics[totalheight=7cm]{./figs/new_10k_TNFRSF11B.pdf}
-      \label{zoomplotTNFRSF} 
-   \end{subfigure} 
-    \begin{subfigure}[b]{0.5\linewidth}
-     \includegraphics[totalheight=7cm]{./figs/new_10k_PRODH.pdf}
-      \label{zoomplotprodh} 
-    \end{subfigure} 
-        \begin{subfigure}[b]{0.5\linewidth}
-     \includegraphics[totalheight=7cm]{./figs/EUR_chr7_156_2-156_6mb.pdf}
-      \label{zoomplotlinc} 
-    \end{subfigure} 
-    \begin{flushleft}
-      (a) The figure shows genome-wide SNPs and their importance. Highlighted are the 100 literature assigned BMD genes (green) and four novel loci. The horizontal line indicates the top 30 features. 
-      Locus Zoomplot for (b) the top ranked gene TNFRSF11B and the two putative novel associations (c) PRODH and (e) LINC01006.
-Each point represents a variant with the normalised Gini impurity score on the left y axis and LD to the central SNP in colour with the highest $r^2$ in red.
-     \end{flushleft}
-  \end{adjustwidth}
-        \label{manhattenzoom} 
-\end{figure}
-
-%\begin{figure}[tbhp]
-%  \begin{adjustwidth}{-2.00in}{0in}
-%    \caption{\textbf{Importance of 10,000 top ranked features and visualisation of the 22q.11 region}}
-%    \label{figure:ranksumtest}
-%     \begin{subfigure}[b]{0.5\linewidth}
-%    \includegraphics[totalheight=7cm]{./figs/BMDTop10K.pdf}
-%     \caption{Distribution of known BMD genes amongst the top 10,000 features.} 
-%      \label{figure:rbo-prod.png} 
-%%      \vspace{4ex}
-%    \end{subfigure} 
-% \begin{subfigure}[b]{0.5\linewidth}
-%      \centering
-%      \includegraphics[totalheight=7cm]{./figs/EURchr22_18-20mb.pdf}
-%      \caption{Zoom plot of the 22q.11 region centred on PRODH.} 
-%      \label{zoomplot} 
-%%      \vspace{4ex}
-%    \end{subfigure} 
-%    \begin{flushleft}
-%      (a) The figure shows the top 10,000 features by their rank and Gini impurity score (Importance). The line denotes
-%      the top 20 features which contains five novel loci annotated with text. The $p$-value is the Mann-Whitney U test.
-%      (b) Each point represents a variant with the normalised Gini impurity score on the left y axis and LD in colour
-%      with the highest $r^2$ in red.
-%     \end{flushleft}
-%  \end{adjustwidth}
-%\end{figure}
-
-%\begin{figure}[tbhp]
-%  \begin{adjustwidth}{-2.00in}{0in}
-%    \caption{\textbf{Importance of genome-wide SNPs}}
-%    \label{figure:knowngeneranks}
-%\includegraphics[totalheight=6.5cm]{./figs/BMDTopWholeGenome.pdf}
-%    \begin{flushleft}
-%      The figure shows genome-wide SNPs and their importance. Highlighted are the 100 literature assigned BMD genes. The horizontal line indicates the top 20. 
-%     \end{flushleft}
-%  \end{adjustwidth}
-%\end{figure}
-
-%1:693731 rs12238997 lncRNA
-% rs1035665 A genome-wide association study of global gene expression http://www.nature.com/ng/journal/v39/n10/full/ng2109.html rs1035665 as master regulator of ATP Binding Cassette Subfamily F Member 2 ABCF1  with ABCF2 634 known
-% rs6702460 1	91536	91536	14	1.0296212827525244E-6 ncRNA
-% rs2058553 2	20573002	20573002	15	9.896033178354981E-7 ncRNA
-% rs12916554 15	 54607255	54607255	17	9.687391928109767E-7  Unc13c quality of sleep http://www.nature.com/ng/journal/v48/n8/full/ng.3595.html   https://www.ncbi.nlm.nih.gov/pubmed/27655463
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Section 3: Classification selection
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{\cursedforest outperforms existing methods for classification problems}
-In this section we demonstrate \cursedforest's classification accuracy on biological data and compare its performance against other published methods. 
-We do this by predicting the ethnicity of the individuals in the 1000 Genomes Project from their genomic profiles and evaluate against the annotated label. 
-We record the prediction error (OOB) as well as the runtime for each method. 
-%@Aidan is 12 correct?
-We limit the number of executors to 12 that are available to \cursedforest\ to keep the comparison fair for all methods.
-
-As shown in Fig.~\ref{figure:1000genomes}, only \cursedforest\ is able to scale to the full phase3\_chr1-22 dataset (2,504 samples x 19,328,051 variants), with
-the standard random forest implementation in R running out of resources on the relatively low-dimensional
-phase1\_chr22 dataset (1,092 x 490,036). 
-Ranger's c++ implementation and the R front-end version of it are able to process the second largest dataset (phase3\_chr1-3 with 2,504 x 19,328,051), however, both are unable to process larger datasets successfully.
-While the c++ implementation runs out of memory the R front-end, which also covers the loading and pre-processing, needs an unreasonable compute time for this task (>20h), see Supplementary Table 1.
-The standard implementation of random forest in SparkML already runs out of memory after processing the third largest dataset, phase3\_chr1 (2,504 x 6,450,364). 
-This demonstrates that Spark ML, although designed to handle many samples, is unable to efficiently cope with large numbers of features.
-
-All tested methods have a worse runtime than \cursedforest\ for the datasets they processed successfully, except for the smallest dataset where the overhead of Spark's distributed platform made ranger's R implementation faster with 1 min 10 seconds compared to 1 min 50 seconds.
-For example, on phase3\_chr1-3 \cursedforest\ finishes in 1 hour, with ranger(R) requiring 4 hours and SparkML already needing 8 hours on a smaller dataset.
-Extrapolating from this, all comparable implementations hence would require more than 24 hours to finish, assuming the use of larger memory nodes.
-
-As our implementation can efficiently utilise large numbers of commodity nodes, the performance of \cursedforest\ can be substantially improved further. 
-%@Aidan is 130 correct?
-Table~\ref{cursedforesttable} shows the runtime when utilising the full Spark cluster (130 worker nodes).
-
-Building a model with \cursedforest\ on phase3\_chr1, which previously took 20 minutes, can be reduced to 4 minutes. 
-Subsequently, phase3\_chr1-3 takes about 11 minutes, and building a
-model on the entire genome is 37 minutes.
-
-Not only does the job complete successfully using \cursedforest\, but building a random
-forest model on all 22 autosomes (phase3\_chr1-22)  leads to a more accurate result. 
-The error can be reduced from OOB=0.04 when using only data from Chr1, which represents processing power limit for most tools, down to OOB=0.01 when using the full dataset.
-
-\begin{table}[!ht]
-\begin{minipage}{\textwidth}
-\centering
-\caption{ {\bf Performance comparison with \cursedforest on different subsets of the 1000 Genomes Project data.}}
-\begin{tabular}{| l | r | l |}
-\hline
-{\bf Genome \%} & {\bf Error} & {\bf Runtime} \\ \thickhline
-phase1\_chr22: 1,092 x 490,036 & 0.04 & 1 min 5 seconds \\ \hline
-phase3\_chr22: 2,504 x 1,103,548 & 0.06 & 1 min 37 seconds \\ \hline
-phase3\_chr1: 2,504 x 6,450,364 & 0.04 & 4 min 10 seconds \\ \hline
-phase3\_chr1-3: 2,504 x 19,328,051 & 0.02 & 10 min 49 seconds \\ \hline
-phase3\_chr1-22: 2,504 x 81,047,467 & 0.01 & 36 min 54 seconds \\ \hline
-\end{tabular}
-\begin{flushleft} 
-Comparison of runtime and error for models built using \cursedforest on different subsets of variant data 
-from the 1000 Genomes Project.
-Each random forest model consists of 50 trees and was built using 130 worker nodes, each with 8GB RAM.
-\end{flushleft}
-\label{cursedforesttable}
-\end{minipage}
-\end{table}
-
-\begin{figure}[tbhp]
-  \begin{adjustwidth}{-1.00in}{0in}
-    \caption{\textbf{Runtime comparison on the 1000 genomes project}}
-    \label{figure:1000genomes}
-    \includegraphics[totalheight=8cm]{./figs/1000genomesRuntime.pdf}
-    \begin{flushleft}
-      The figure shows the runtime for the 5 subsets of the 1000 genomes project and the different methods investigated.
-      $\bigotimes$ marks the last dataset the different method completed before running out of memory or running more than
-      20 hours. The dashed lines mark the trajectory for how \ranger and Spark ML had performed given larger memory machines.
-    \end{flushleft}
-  \end{adjustwidth}
-\end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Section 4: Sensitivity to different parameters and scalability
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Scalability analysis}
-In this section we explore the performance of \cursedforest\ in more detail by testing its ability to scale to different sizes of data
+\section{Results}
+\subsection{Scalability}
+In this section we explore the performance of CursedForest in more detail by testing its ability to scale to different sizes of data
 and computational resources.
 
-In order to asses these characteristics, we run \cursedforest classification on synthetic datasets with varying numbers
+In order to asses these characteristics, we run CursedForest classification on synthetic datasets with varying numbers
 of variables (features) and samples, similar to the dataset used in \cite{Wright.and.Ziegle.2016} to evaluate
-\ranger, allocating varying number of CPU cores to the \cursedforest and also varying the computational complexity
-of the random forests by using a range of \mtry values.
+ranger, allocating varying number of CPU cores to the CursedForest and also varying the computational complexity
+of the random forests by using a range of mtry values.
 
@@
 
-First we look at \cursedforest horizontal scalability for a medium size dataset of 2.5 million variables and 5000
-samples, by varying the \mtry  fraction and the number of CPU cores allocated to the execution. 
+First we look at CursedForest horizontal scalability for a medium size dataset of 2.5 million variables and 5000
+samples, by varying the mtry  fraction and the number of CPU cores allocated to the execution. 
 %The results are visualised in Fig~\ref{figure:synthetictiming}. 
@@
 %presented in Table~\ref{synthetictimingtable} and visualized in figure \ref{figure:synthetictiming} below. 
-Regardless of the number of cores used, \cursedforest displays approximately linear dependency between the execution
-time and \mtry (Fig~\ref{figure:synthetictiming.a}).
+Regardless of the number of cores used, CursedForest displays approximately linear dependency between the execution
+time and mtry (Fig~\ref{figure:synthetictiming.a}).
 
-\cursedforest scales almost linearly with the number of CPU cores for medium values of \mtry fraction but for both lower
+CursedForest scales almost linearly with the number of CPU cores for medium values of mtry fraction but for both lower
 and higher values the performance degrades slightly (Fig~\ref{figure:synthetictiming.b}). In the latter case the likely
-cause is communication overhead (with lower \mtry values the proportion of time for parallelizable computation to the
+cause is communication overhead (with lower mtry values the proportion of time for parallelizable computation to the
 time for internode communication is lower) while in the latter case it is most likely caused by reaching the clusters
@@
 
-Next we investigate \cursedforest scalability with regards to the size of data, by varying the number of variables and
-sample for a fixed \mtry fraction of 0.25 and execution of 128 CPU cores. The results are visualized in
+Next we investigate CursedForest scalability with regards to the size of data, by varying the number of variables and
+sample for a fixed mtry fraction of 0.25 and execution of 128 CPU cores. The results are visualized in
 Fig~\ref{figure:synth2} below (please note log scale on the axes and the values on y axes are expressed as trees per
@@
 irregularities in the graph can be attributed to computation vs communication tradeoff. It is also worth noting that
-keeping the \mtry fraction constant results in higher \mtry values with the growing number of variables, and this is
+keeping the mtry fraction constant results in higher mtry values with the growing number of variables, and this is
 what drives the performance down rather that the increase of dataset size itself.
 
-To conclude \cursedforest is capable of processing 60 trees per hour on a datasets with 50 million variables and 10,000
+To conclude CursedForest is capable of processing 60 trees per hour on a datasets with 50 million variables and 10,000
 samples, which is the size range for whole genome sequencing experiments of clinically relevant cohort sizes.
@@
 \begin{figure}[tbhp]
-  \begin{adjustwidth}{-1.00in}{0in}
     \caption{\textbf{Scalability of the wide random forest on the synthetic dataset of 2.5M features and 5K samples.}}
@@
       \centering
-      \includegraphics[totalheight=6cm]{./figs/mtry_cpu.png} 
-      \caption{Time in seconds to build 100 trees for different \mtry\ fractions. } 
+      \includegraphics[totalheight=5cm]{../plos16/figs/mtry_cpu.png} 
+      \caption{Time in seconds to build 100 trees for different mtry fractions. } 
       \label{figure:synthetictiming.a} 
@@
       \centering
-      \includegraphics[totalheight=6cm]{./figs/cpu_mtry_trees_per_hour.png}
+      \includegraphics[totalheight=5cm]{../plos16/figs/cpu_mtry_trees_per_hour.png}
       \caption{Number of trees build per hour when using a growing number of CPU cores.}
@@
     \end{subfigure} 
-  \end{adjustwidth}
 \end{figure}
 
-% commented this table out -- see above
-% \begin{table}[!ht]
-%   \begin{adjustwidth}{-1.00in}{0in}
-%     \centering
-%     \caption{\textbf{
-%  Time in seconds to build a random forest of 100
-%   trees on 128 CPU cores with varying numbers of variables and samples.}}
-% \label{synthetictimingtable}
-% \begin{tabular}{|l|l|l|l|l|l|l|}
-% \hline
-% \multicolumn{2}{|l|}{\multirow{2}{*}{}}           & \multicolumn{5}{c|}{Number of variables} \\
-% \cline{3-7}
-% \multicolumn{2}{|l|}{}                               & \bf{150,000} & \bf{500,000} & \bf{2,500,000}  & \bf{10,000,000} & \bf{50,000,000} \\
-% \hline                                                            
-% \multirow{4}{*}{Number of samples}      & \bf{1000}  & 21.6  & 26.7  & 59.8  & 133.5  & 668.8 \\
-%                                         & \bf{5000}  & 45.7  & 71.9  & 186.2 & 729.3  & 2887.4 \\
-%                                         & \bf{10000} & 70.2  & 120.2 & 397.3 & 2237.8 & 5747.8 \\
-% \hline
-% \end{tabular}
-% \end{adjustwidth}
-% \end{table}
-
 \begin{figure}[tbhp]
-  \begin{adjustwidth}{-1.00in}{0in}
     \caption{\textbf{Scalablity of the Wide Random Forest on synthetic datasets with varying number of samples and variables.}}
     \label{figure:synth2}
-    \begin{subfigure}[b]{0.5\linewidth}
+    \begin{subfigure}[b]{0.45\linewidth}
       \centering
-      \includegraphics[totalheight=6cm]{./figs/nvars_nsamples.png} 
+      \includegraphics[totalheight=5cm]{../plos16/figs/nvars_nsamples.png} 
       \caption{Number of trees build per hour when growing number of variables.} 
@@
     \end{subfigure} 
-    \begin{subfigure}[b]{0.5\linewidth}
+    \begin{subfigure}[b]{0.45\linewidth}
       \centering
-      \includegraphics[totalheight=6cm]{./figs/nsamples_nvars.png}
+      \includegraphics[totalheight=5cm]{../plos16/figs/nsamples_nvars.png}
       \caption{Number of trees build per hour when growing number of samples.}
@@
   \end{subfigure}
-\end{adjustwidth}
 \end{figure}
 
-% no data in this table 
-% \begin{table}[!ht]
-% \caption{
-% {\bf Sensitivity analysis on parameter choice.}}
-% \begin{tabular}{|l|l|l|l|l|l|l|l|}
-% \hline
-% %\multicolumn{4}{|l|}{\bf Heading1} & \multicolumn{4}{|l|}{\bf Heading2}\\ \hline
-% \bf{ \mtry\ }  & \bf{Mtree} & \bf{depth} & \bf{Accuracy} & \bf{Runtime} & \bf{Memory} \\
-% \hline
-% &&&&&\\ \hline
-% \end{tabular}
-% \begin{flushleft} 
-%   Table notes here.
-% \end{flushleft}
-% \label{table2}
-% \end{table}
+\subsection{Exploring theoretical recovery rate of wide data}
+
+ Donoho and Tanner \cite{Donoho.and.Tanner.2009} give a ``universal phase change'' result that has applications in a
+ large number of areas including variable selection in high dimensions. Consider
+ Fig~\ref{figure:phase-diagram-equivalence.png} which shows the region where a model can recover the important variables,
+ plotted as a function of $\delta = n/p$ and $\rho =k/n$ (where $k$ is the number of significant variables). There is a
+ distinct boundary, shown empirically and also based on arguments from combinatorial geometry, to the region where we can
+ reliably recover significant variables.  \cite{Donoho.and.Stodden.2006} investigate the behavior of a number of
+ regression approaches for variable selection (LARS, Lasso and forward stepwise) and make the point that above the
+ phase-transition line variable recovery is still possible by a combinatorial approach.
+
+ It is not surprising that that it is more difficult to recover the signal variable in the upper-left area of the figure,
+ as the problem is both under-determined and sparse. What is surprising is the connection with arguments from
+ combinatorial geometry. This suggests that we are seeing a universal rule rather than an implementation issue. As
+ CursedForest is designed for extremely large numbers of variables it is likely to be operating in difficult regions of
+ the figure where the ratio $\delta = n/p$ is small.
+
+ We note several things here:
+ \begin{itemize}
+ \item the Donoho-Tanner phase transition arises in recovering the $\beta$ in data generated by a
+   linear model. However, in a decision tree (random forest) there is no notion of estimating the $\beta$
+ \item A decision tree (random forest) is a heuristic search. It may recover a relationship in the space of the
+   combinatorial search.
+ \end{itemize}
+
+ The existence of the  Donoho-Tanner phase transition is a salutary warning. There are likely to be limits, both
+ computational and logical, to the recovery of signals from noisy data. CursedForest  is a contribution to addressing the
+ practical limits but the logical limits will still apply. However in the case of data that is both big and wide,
+ CursedForest and other VariantSpark methods may provide a useful tool.
+
+ \begin{figure}[tbhp] 
+     \caption{\textbf{The Donoho-Tanner phase chage diagram.}}
+     \centering
+     \includegraphics[totalheight=6cm]{../plos16/figs/phase.png} 
+     \label{figure:phase-diagram-equivalence.png} 
+     \vspace{4ex}
+ \end{figure}
+
+\begin{figure}[tbhp] 
+    \begin{subfigure}[b]{0.45\linewidth}
+      \centering
+      \includegraphics[totalheight=6cm]{../plos16/figs/k.png}
+      \caption{$k$}
+      \label{figure:k.png}
+    \end{subfigure} 
+    \begin{subfigure}[b]{0.45\linewidth}
+      \centering
+      \includegraphics[totalheight=6cm]{../plos16/figs/n.png}
+      \caption{$n$}
+      \label{figure:n.png}
+    \end{subfigure} 
+    \caption{The simulation in \protect{\cite{Donoho.and.Stodden.2006}} considers $\delta = n/p$ and $\rho =k/n$. Here we plot the
+      space of $\{ \delta, \rho$\}, colored  by the parameters $n$ and $k$}
+\end{figure}
+
+\subsection{Biological data}
+We apply CursedForest to two biological datasets.
+Firstly, the 1000 genomes project to test its classification accuracy and secondly, to a bone mineral density dataset to demonstrate a GWAS-style analysis.
+We training CursedForest on the 1000 genomes dataset, which consists of 2,504 samples with 81,047,467 features each to predict the ethnicity from genomic profiles.
+CursedForest achieves an out of bag error of OOB=0.01 and completes in 36 min 54 seconds, demonstrating its capability to run on population-scale cohorts of real world applications.
+Next we perform feature selection on over 7.2 million genomic variants and identify the
+locations associated with Bone Mineral Density (BMD) in a previously published GWAS dataset~\cite{Duncan.2011}.
+We faithfully recover 5 known BMD genes that were previously identified in GWAS studies, however also find two probable new associations that were previously only suggestive.
+This demonstrates the utility of our approach as well as the ability to amplify signal by taking SNP interactions into account rather than limiting the analysis to individual strong responders.
 
 \section{Conclusion}
-
 We have demonstrated that using a different parallelization model can extend random forests to the case of an
@@
 
-Donoho and Tanner \cite{Donoho.and.Tanner.2009} give a ``universal phase change'' result that has applications in a
-large number of areas, including variable selection in high dimensions. The theoretical phase change boundary is based on arguments
-from combinatorial geometry.
-
-They argue that there is a sharp disjunction between the cases where informative variables may be recovered with a high
-accuracy by procedures like stepwise variable selection, and the cases where they can not be recovered. The boundary is
-shown in Fig~\ref{figure:phase-diagram-equivalence.png} in a space parameterized by the level of underdeterminatio,n
-$\delta = n/p$, and by the sparsity, $\rho =k/n$ (where $k$ is the number of informative variables).  Above the
-phase-transition line variable recovery is still possible by a combinatorial approach such as all-subsets
-variable selection.
-
-\begin{figure}[tbhp] 
- \begin{adjustwidth}{-1.00in}{0in}
-    \centering
-    \includegraphics[totalheight=6cm]{./figs/phase2.png} 
-    \caption{{\bf The phase change diagram.}}
-    \label{figure:phase-diagram-equivalence.png} 
-    \vspace{4ex}
-  \end{adjustwidth}
-\end{figure}
-
-\cite{Donoho.and.Stodden.2006} investigate the phase transition in a small simulation.  They consider a regression
-problem with $\underset{n\times p}{X}\sim N(0,1)$ with $p$ fixed at 200 and $n$ variable.  They set
-$\beta(1:k) \sim U(1,100)$ and $\beta((k+1):p) =0$, then $y= X\beta + \epsilon$ with $\epsilon \sim N(0,\sqrt{16})$. They
-evaluate variable selection by the error measure
-$$\frac{\Vert\hat{\beta}-\beta\Vert_2}{\Vert\beta\Vert_2}.$$
-
-They consider a number of variable selection methods, including a false discovery rate criteria. This involves adding
-the variable with the maximum $t$-value to the linear model if the $p$-value is less than 
-$0.25(\text{number of  terms  currently  in  the  model})/(\text{total  number  of  variables})$.  
-See Fig~\ref{figure:error_Stodden_FDR.png} for the error measure and
-figure \ref{figure:rbo_Stodden_FDR.png} for the RBO, comparing the ranking (i.e. values) of $\hat{\beta}$ and $\beta$.
-It apparent that there is a marked drop in the accuracy of the variable recovery in line with the prediction of the
-Donoho-Tanner phase transition.
-
- \begin{figure}[tbhp] 
-%   \begin{adjustwidth}{-1.00in}{0in}
-     \begin{subfigure}[t]{0.5\linewidth}
-       \centering
-       \includegraphics[totalheight=6cm]{./figs/error_Stodden_FDR.png}
-       \caption{The error measure.}
-       \label{figure:error_Stodden_FDR.png}
-%       \vspace{4ex}
-     \end{subfigure} 
-     \begin{subfigure}[t]{0.5\linewidth}
-       \centering
-       \includegraphics[totalheight=6cm]{./figs/rbo_Stodden_FDR.png}
-       \caption{The RBO measure}
-       \label{figure:rbo_Stodden_FDR.png}
-     \end{subfigure} 
-     \caption{Forward stepwise variable selection with a false discovery rate stopping criteria.}
-       \label{figure:error_and_rbo_Stodden_FDR.png}
-%   \end{adjustwidth}
- \end{figure}
-
- \begin{figure}[tbhp] 
-%   \begin{adjustwidth}{-1.00in}{0in}
-     \begin{subfigure}[t]{0.5\linewidth}
-       \centering
-       \includegraphics[totalheight=6cm]{./figs/ranger_error_Stodden_simulation.png}
-       \caption{The OOB error rate.}
-       \label{figure:ranger_error_Stodden_simulation.png}
-       \vspace{4ex}
-     \end{subfigure} 
-     \begin{subfigure}[t]{0.5\linewidth}
-       \centering
-       \includegraphics[totalheight=6cm]{./figs/ranger_rbo_Stodden_simulation.png}
-       \caption{The RBO measure}
-       \label{figure:ranger_rbo_Stodden_simulation.png}
-     \end{subfigure} 
-     \caption{The simulated data of \protect{\cite{Donoho.and.Stodden.2006}} with a random forest model.}
-     \label{figure:ranger_error_and_rbo_Stodden_simulation.png}
-%   \end{adjustwidth}
- \end{figure}
-
-We have seen no work on the question of such a phase transition in the case of classification problems. Still it may be
-instructive to consider the implications of the phase transition for our work.
-
-We have used a random forest on the same simulation data used for figure \ref{figure:error_and_rbo_Stodden_FDR.png}.  The
-Donoho-Tanner phase transition arises in recovering the $\beta$ in data generated by a linear model. However, in a
-decision tree (random forest) we are fitting a non-linear model and there is no notion of estimating the
-$\beta$. Because of this we have evaluated the performance of the random forest using the RBO measure on the variable
-importance.
-
-We note that while a random forest is a long way from an all-subsets search, it is a limited search of
-the feature space. As such it may perform outside of the bounds of the phase transition.
-
-Fig~\ref{figure:ranger_error_and_rbo_Stodden_simulation.png} shows the OOB prediction error and the RBO error for a
-random forest with 10000 trees and the \mtry value set to the default for a regression problem (depending on $n$). It
-shows no evidence of following the shape of the Donoho-Tanner phase transition.
-
-As \cursedforest is designed for extremely large numbers of variables it is likely to be operating in difficult regions
-of the figure where the ratio $\delta = n/p$ is small. In the simulated example in section \ref{section:synthetic_data}
-the underdetermination parameter $\delta = 0.002$ and the sparsity parameter $\rho =2\times 10^{-6}.$
-
-The existence of the Donoho-Tanner phase transition is a salutary warning. There are likely to be limits, both
-computational and logical, to the recovery of signals from noisy data. \cursedforest is a contribution to addressing the
-practical limits but the logical limits will still apply. However in the case of data that is both big and wide,
-\cursedforest and other VariantSpark methods may provide a useful tool.
-
-\clearpage
-\section{Supporting Information}
-
-% Include only the SI item label in the subsection heading. Use the \nameref{label} command to cite SI items in the text.
-%\subsection*{S1 Video}
-%\label{S1_Video}
-%{\bf Bold the first sentence.}  Maecenas convallis mauris sit amet sem ultrices gravida.
-Etiam eget sapien nibh.
-Sed ac ipsum eget %enim egestas ullamcorper nec euismod ligula.
-Curabitur fringilla pulvinar lectus consectetur pellentesque.
-
-\section*{Acknowledgments}
-AO and PS implemented the method, AO conducted the BMD and 1000genomes experiments while PS generated the scaling data,
-analysed by RD with input from NE. PL and ELD aided in the interpretatio nof the BMD data, while SL and JD contributed
-to the 1000genomes analysis. DCB visualised the BMD and 1000genomes results. All authors contributed to the manuscript.
-The authors would like to acknowledge CSIRO, specifically PS, for setting up and maintaining the in-house Spark cluster.
-
-\nolinenumbers
-
-\bibliography{CursedForest}
+\bibliographystyle{plain}
+\bibliography{../plos16/CursedForest}
 
 \end{document}
-
-% We have demonstrated that using a different parallelization model can extend random forests to the case of an
-% extremely large number of variables. We have treated the case of variable selection in a $p >> n$ model where most of
-% the variables are uninformative and have demonstrated the utility of the model for large GWAS datasets. By comparing
-% this implementation to other implementations (including those optimized for large datasets) we have demonstrated the
-% utility of this approach.
-
-% Donoho and Tanner \cite{Donoho.and.Tanner.2009} give a ``universal phase change'' result that has applications in a
-% large number of areas including variable selection in high dimensions. Consider
-% Fig~\ref{figure:phase-diagram-equivalence.png} which shows the region where a model can recover the important variables,
-% plotted as a function of $\delta = n/p$ and $\rho =k/n$ (where $k$ is the number of significant variables). There is a
-% distinct boundary, shown empirically and also based on arguments from combinatorial geometry, to the region where we can
-% reliably recover significant variables.  \cite{Donoho.and.Stodden.2006} investigate the behavior of a number of
-% regression approaches for variable selection (LARS, Lasso and forward stepwise) and make the point that above the
-% phase-transition line variable recovery is still possible by a combinatorial approach.
-
-% It is not surprising that that it is more difficult to recover the signal variable in the upper-left area of the figure,
-% as the problem is both under-determined and sparse. What is surprising is the connection with arguments from
-% combinatorial geometry. This suggests that we are seeing a universal rule rather than an implementation issue. As
-% \cursedforest is designed for extremely large numbers of variables it is likely to be operating in difficult regions of
-% the figure where the ratio $\delta = n/p$ is small.
-
-% We note several things here:
-% \begin{itemize}
-% \item the Donoho-Tanner phase transition arises in recovering the $\beta$ in data generated by a
-%   linear model. However, in a decision tree (random forest) there is no notion of estimating the $\beta$
-% \item A decision tree (random forest) is a heuristic search. It may recover a relationship in the space of the
-%   combinatorial search.
-% \end{itemize}
-
-% The existence of the  Donoho-Tanner phase transition is a salutary warning. There are likely to be limits, both
-% computational and logical, to the recovery of signals from noisy data. \cursedforest  is a contribution to addressing the
-% practical limits but the logical limits will still apply. However in the case of data that is both big and wide,
-% \cursedforest and other VariantSpark methods may provide a useful tool.
-
-% \begin{figure}[tbhp] 
-%  \begin{adjustwidth}{-1.00in}{0in}
-%     \caption{\textbf{The Donoho-Tanner phase chage diagram.}}
-%     \centering
-%     \includegraphics[totalheight=6cm]{./figs/phase.png} 
-%     \label{figure:phase-diagram-equivalence.png} 
-%     \vspace{4ex}
-%   \end{adjustwidth}
-% \end{figure}
-
-\begin{figure}[tbhp] 
-    \begin{subfigure}[b]{0.5\linewidth}
-      \centering
-      \includegraphics[totalheight=6cm]{./figs/k.png}
-      \caption{$k$}
-      \label{figure:k.png}
-    \end{subfigure} 
-    \begin{subfigure}[b]{0.5\linewidth}
-      \centering
-      \includegraphics[totalheight=6cm]{./figs/n.png}
-      \caption{$n$}
-      \label{figure:n.png}
-    \end{subfigure} 
-    \caption{The simulation in \protect{\cite{Donoho.and.Stodden.2006}} considers $\delta = n/p$ and $\rho =k/n$. Here we plot the
-      space of $\{ \delta, \rho$\}, colored  by the parameters $n$ and $k$}
-\end{figure}

### 2017-05-19 doc/publications/plos16/nm_alt.tex

--- 2017-05-03 doc/publications/plos16/nm.tex
@@
+% This is a separate version of the CursedForest manuscript following Nature Method formatting guidelines.
+A Word document has been created based on this.
+See the Word document for the latest version of the paper.
+%
 % Template for PLoS
@@
 {\Large
-\textbf\newline{\cursedforest - A Random Forest Implementation for ``Big'' and ``Wide'' Data} % Please use "title case" (capitalize all terms in the title except conjunctions, prepositions, and articles).
+\textbf\newline{\cursedforest - A Random Forest Machine Learning Implementation for ``Big'' and ``Wide'' Data} % Please use "title case" (capitalize all terms in the title except conjunctions, prepositions, and articles).
 }
@@
 life~\cite{Loebbecke2015}.  These data sets are not only growing vertically, by capturing more events, but also
-horizontally by capturing more information about these events.  The challenge of big and ``wide'' data is especially
+horizontally by capturing more information about these events.  The challenge of ``big" and ``wide'' data is especially
 pronounced in the biomedical space where, for example, whole genome sequencing (WGS) technology enables researchers to
@@
 for taking the interaction of variables into account~\cite{Wright.et.al.2016}. In addition, random forests \cite{Breiman2001} are well
-suited for processing ``wide'' genomic data for two reasons.  Firstly, while other machine learning applications have
+suited for processing ``wide'' genomic data for two reasons: firstly, while other machine learning applications have
 the propensity to overfit data sets with more features $p$ than samples $n$ (a consequence of the ``curse of
-dimensionality''~\cite{Bauer2014, bellman1961adaptive}), decision trees are resistant to overfitting.  Secondly, random
-forests are also very easy to parallelise. As the forest is a sum of decision trees, it is possible to grow separate
+dimensionality''~\cite{Bauer2014, bellman1961adaptive}), decision trees are resistant to overfitting; secondly, random
+forests are easy to parallelise, as the forest is a sum of decision trees, it is possible to grow separate
 trees on different compute nodes and combine the trees. 
@@
 
-\section{Methods}
-\label{section:methods}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Overview of Random Forests}
-
-Decision trees have a number of desirable features, they:
-\begin{itemize}
-  \item are invariant to monotonic scaling of the data;
-  \item can handle categorical and real valued inputs;
-  \item can handle missing values;
-  \item are insensitive to uninformative predictors;
-  \item are able to capture interactions between features (which is important for modeling complex polygenic diseases).
-  \end{itemize}
-However, the tree fitting algorithm is greedy and may generate an unstable model. That is, a small change in the data may
-lead to a very different model.
-
-Random forests apply the technique of bootstrap aggregation or bagging to decision trees.  The training data sets are
-independently drawn, with replacement, from the joint distribution of $(X,Y)$ so that each set comprises $n$
-$(p+1)$-tuples $(x_1,y_1),\ldots, (x_n,y_n)$. This is done $B$ times and a tree is grown on each of these $B$ samples.
-The results are then combined in an appropriate way. In a classification problem we take the majority vote for the trees
-over the $q=1,..,Q$ classes,
-\begin{equation*}
-{{h_f}}= \arg \max_q \left(\sum_b I(h(x;B_b)=q)\right).
-\end{equation*}
-In addition to growing the trees
-on bootstrapped samples, each node in the tree is split on a randomly
-selected subset of the variables.
-
-Random forests are a variance reduction technique. They take unpruned tree models, which have a low bias but a high
-variance, and by combining them reduce the variance.  For regression problems, as the prediction error is the sum of the
-variance and the bias squared, this strategy should lead to a low prediction error.
-
-For classifiction problems there is no obvious decomposition of classification error into an equivalent bias plus
-variance form. A number of different additive decompositions of classification error have been considered in an effort
-to understand the success of bagging. They are all less transparent than the regression case (see \cite{Friedman.1997} and
-citations therein).
-
-In the case of a regression tree, the average prediction error for an individual tree $h(X; B)$ is
-\begin{equation}
-PE_t = E_B E_{X,Y} (Y-h(X; B))^2.
-\end{equation}
-Assume that, for all $B$, the tree is unbiased, i.e., $EY= E_X h(X; B)$. Then
-\begin{equation}
-PE_f \leq \rho PE_t
-\end{equation}
-where $\rho$ is the weighted correlation between residuals $(Y-h(X;B))$ and $(Y-h(X;B^\prime))$ for independent
-$B,B^\prime$.
-So what is required for accurate random forest regression is (i) low correlation between residuals of differing tree in
-the forest, and (ii) low prediction error for the individual trees \cite{Segal.2004}.
-
-The benefits of random forest models are:
-\begin{itemize}
-\item they give an out-of-bag (OOB) estimate of model accuracy, by using the samples not selected for the bootstrapped sample
-  as an independent test set for each tree;
-\item they provide a measure of variable importance, for example by summing the number of times a variable is used to
-  split a node, over all nodes and all trees. Other, more complex, measures of variable importance are also used;
-\item they inherit some of the qualities of tree models (but not, for example, the handling of missing values via surrogate
-  variables. This would be difficult as the trees are split on random subsets of variables).
-\end{itemize}
-
-Since the original implementation of random forests, there have been a number of developments leading to a better
-understanding of the algorithm.
-\begin{itemize}
-\item random forests are resistant to overfitting. However it is not true that they will not overfit,
-  ~\cite{Segal.2004} gives examples of overfitting in cases where:
-  \begin{itemize}
-  \item a single tree is the most appropriate model; and
-  \item the variables are highly correlated so that the bootstrap resampling does not give rise to uncorrelated trees.
-  \end{itemize}
-However, keeping these difficult cases in mind, random forests have a strong claim to being resistant to overfitting.
-\item \cite{Strobl.et.al.2007} demonstrates that the random forest algorithm is subject to a bias in variable selection
-  via two mechanisms:
-  \begin{enumerate}
-  \item differences in variable scale (for $X_i$ continuous) and number of levels (for $X_i$ discrete) introduces a
-    bias. An uninformative variable with a large number of levels may be selected over a more informative variable with
-    fewer levels;
-  \item bagging introduces a bias, but sampling $(1- 1/e) \approx 0.632$ of the data without replacement seems to be an
-    effective strategy for removing the problem.
-  \end{enumerate}
-\end{itemize}
-
-\cite{Strobl.et.al.2007} introduce changes to the random forest algorithm to fix 1) and 2).  Effect 1) will be apparent
-in the \cursedforest algorithm as well.  However, there are many examples, particularly in genomics, where we have data
-sets that are:
-\begin{itemize}
-\item very wide;
-\item all variables have the same number of levels (for example, measurements at different points along the genome);
-\item the hypothesized functional relationship is of a form amenable to a decision tree.
-\end{itemize}
-GWAS data sets are an obvious application area for this algorithm.
-
-\cite{Biau.2012} provides a proof that a random forest model will converge at a rate that depends on
-the cardinality of the set $S$ of ``strong predictors'' rather than on the number of variables $p$. That is, given a
-function $y=f(S)$ depending on a set of variables $S$, a random forest will converge to the true function $y$ with a
-rate that depends on the size of $S$ and not the number of noise variables in the dataset. If the size of $S$ is small
-compared to the total number of variables, then the rate of convergence will be quicker. It is not clear what the size
-of the set $S$ will be in a GWAS application. There are cases of small groups of SNPs with large effects and also of
-conditions caused by the interactions of large groups of SNPs.
-
-\cite{Genuer.et.al.2010} and \cite{Diaz.and.Alvarez.2006} describe iterative schemes for doing variable selection using
-the importance measure. See \cite{Chen.and.Ishwaran.2012} for a review of the area with particular reference to genomic
-data.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{\cursedforest}
-VariantSpark made use of Spark's machine learning algorithms ``Spark ML''. While Spark ML algorithms demonstrate
-scalability when dealing with a large number of samples, this scalability eventually breaks down as we include more
-features.
-
-As is standard with Spark applications, we store our data in a Resilient Distributed Dataset (RDD), where an RDD is
-essentially a collection of elements. In the case of Spark ML, each element in the RDD is a sample. RDDs contribute to
-the scalability of Spark as they can be distributed across multiple nodes and operated on in parallel. Even as we add more
-samples to a dataset, Spark can simply schedule extra tasks to handle the additional items in the RDD.
-
-However, within an RDD, Spark ML stores each sample as a vector. Unlike RDDs, which can be partitioned and distributed
-across multiple nodes, each vector must be present in its entirety on any node accessing it. This is no problem with
-typical data sets; however as dimensionality increases, the vectors eventually reach a size where they can no longer fit
-into a single node's memory.
-
-So in the case of adding more samples, Spark ML can simply create more tasks, keeping memory consumption within the
-cluster's bounds. However, as the dimensionality of each sample grows, the memory requirements of the job increases to
-enable these increasingly large vectors to be loaded into memory.
-
-On the other hand, \cursedforest is specifically designed to handle wide ``cursed'' data. It avoids the relation between
-memory and dimensionality by avoiding calculations that rely on entire feature vectors and taking the parallelization
-work down to the level of the individual features.  For each node of a tree, \cursedforest will distribute
-tasks that consist of single features (variants), for every individual.  Each of these tasks will calculate the
-information gain for that specific feature.  Once these tasks have completed, the results are reduced to return the
-feature which gives the greatest information gain.  This process is then repeated until \cursedforest has created the
-entire decision tree.
-
-The current implementation of \cursedforest uses a ``Gini impurity'' criteria for splitting. Let $f_q$ be the fraction
-of items labeled with value $q$, where $q = 1, \ldots, Q$ at a node. The Gini impurity is
-$$
-I_G(f) = \sum_{q = 1}^Q f_q ( 1 - f_q ), 
-$$
-which is at a minimum when all observations at the node are in the same class.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% This section talks about the data sets
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Data Sets}
-\subsubsection{1000 Genomes Project}
-We obtained the 1000 Genomes Project data as VCF files from their FTP site.  Each VCF file contains the variants for
-every individual across one chromosome.  We use the phase 3 dataset which contains 2,504 individuals with approximately
-81 million variant positions (i.e. a matrix of $2,504 \times 81$ million).  For compatibility with the majority of the
-algorithms we demonstrate, we convert the original variant representations to double representations.  For example, we
-store the absence of a variant (0\textbar0) as 0, a heterozygous
-variant (1\textbar0 or 0\textbar1) as 1, and a
-homozygous variant (1\textbar1) as 2, so we have an ``additive'' model.  However, this is not the only possible encoding \cite{Goldstein.et.al.2011}. Note that from
-the encodings listed in Table~\ref{table:encodings}, the additive encoding will present the most difficulty for a random
-forest model.
-
-\begin{table}[!ht]
-%\begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
-\begin{flushleft} 
-\centering
-\caption{\bf Possible encodings of GWAS data \cite{Goldstein.et.al.2011}}
-\begin{tabular}{|l|l|l|l||}
-\hline
-{\bf Type} & {\bf Mechanism}                                         & {\bf Partition} \\ 
-\thickhline
-Additive   & Each additional minor allele increases variation        & 0, 1, 2         \\
-Dominant   & Presence of at least 1 minor allele increases variation & 0, $\{1,2\}$          \\
-Recessive  & Two minor alleles needed for variation                  & $\{0,1\}$, 2          \\
-Heterosis  & Heterozygote leads to variation                         & $\{0,2\}$, 1          \\ \hline
-%\end{adjustwidth}
-\end{tabular}
-\label{table:encodings}
-\end{flushleft}
-\end{table}
-
-\subsubsection{Bone mineral density dataset}
-We obtained a VCF file from the authors of \cite{Duncan.2011}, which contains variant data from 1,936 postmenopausal women.
-This dataset covers approximately 7 million variant positions. We have information about the bone mineral density (BMD)
-for each individual, where a low BMD is a major predisposing factor to fracture \cite{Duncan.2011}.
-
-\subsubsection{Synthetic data} 
-\label{section:synthetic_data}
- Each dataset consists of $n$ samples and $p$ variables where $n << p$,
-and values for each variable are ordinal variables with three levels represented as numbers $\{0, 1, 2\}$ (which
-correspond to an additive effect encoding of genomic variation) randomly generated from a uniform distribution with equal
-probabilities.
-
-The model parameters are $w_i = 1/\sqrt{2^{i-1}}$ for $i =1,\ldots 5$ and we set 
-\begin{equation}
-\label{equation:synthetic.data}
-z =  \sum_{i=1}^{i=5} {w_i x_i}.
-\end{equation}
-We let $\sigma_\epsilon^2 = Var(z)(1-\theta)/\theta $  where $\theta$ is a parameter controlling the fraction of variance explained by the informative variables and in our study we chose $\theta = 0.125$ .
-Then
-$y =   z + \epsilon$ where $\epsilon \sim  N(0, \sigma_\epsilon^2).$
-The dichotomous response is generated by thresholding $y$ at the $0.5$ quantile,
-
-$$
-\ddot{y} =    \left\{ \begin{array}{rcl}
-			  0 & \mbox{for} & y \leq Q2(y) \\
-			  1 & \mbox{for} & y > Q2(y).
-		 \end{array}\right. 
-$$
-
-\subsection{Parameter settings}
-We consider the parameter settings for the random forest algorithm. We use the R notation from the \randomforest
-package \cite{Liaw.and.Weiner.2002} which incorporates the original Fortran code by Brieman and Cutler.  We incorporate
-the advice of \cite{Liaw.and.Weiner.2002}, which we have found mirrors our own experience:
-\begin{itemize}
-\item \ntree{}  -- the number of trees.  The number of trees necessary for good performance grows with the number of
-  predictors.  \cite{Liaw.and.Weiner.2002} suggest that a high \ntree is necessary to get stable estimates of variable
-  importance and proximity; however, even though the variable importance measures may vary from run to run, we note that
-  it is possible for a random forest model to have a poorer fit and still have an accurate ranking of variable
-  importance;
-\item \mtry{}  -- the number of variables considered at each split (if \mtry=$p$, we have a boosted decision
-  tree model).  If one has a very large number of variables but expects only very few to be ``important'', using larger \mtry may give
-  better performance;
-\item the size and complexity of the individual trees is controlled in \randomforest by setting \texttt{nodesize}, the
-  minimum size of terminal nodes. It is controlled in Spark ML by setting \texttt{maxDepth}, the maximum depth of each
-  tree in the forest.\footnote{The Spark ML documentation \cite{Spark.2016} sets \texttt{maxDepth} to 4 in their
-    classification example. This may give an estimate with a higher bias and, as such, may be a poor choice. See the
-    discussion in \cite{Dietterich.2002} where it is shown that the action of a RF is complicated than first
-    thought. They show that bagging may reduce bias as well as variance but give a better result for low bias
-    learners. \cite{Dietterich.2002} notes that if the bootstrap replicate approximation were correct (i.e.~if the
-    bootstrap sample came from an identical distribution to the data), then bagging would reduce variance without
-    changing bias.}
-\end{itemize}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%
-% Results
-%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Results and Discussion}
-
+\section{Results}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-\section{Conclusion}
-
+\section{Discussion}
 Using a novel parallelization approach enabled by Spark, we extended random forests to cope with extremely high dimensional data (large number of variables). 
@@
 Specifically, the here demonstrated ability to uncover novel biological information from previously analysed GWAS studies showcases the enormous potential of re-analysing the GWAS catalogue given this unprecedented computational capacity to perform joint-loci association testing and identify the much-anticipated cumulative multi-gene disease contribution.
+
+\section{Methods}
+\label{section:methods}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Overview of Random Forests (This should go to Supplemental Information)}
+
+Decision trees have a number of desirable features, they:
+\begin{itemize}
+  \item are invariant to monotonic scaling of the data;
+  \item can handle categorical and real valued inputs;
+  \item can handle missing values;
+  \item are insensitive to uninformative predictors;
+  \item are able to capture interactions between features (which is important for modeling complex polygenic diseases).
+  \end{itemize}
+However, the tree fitting algorithm is greedy and may generate an unstable model. That is, a small change in the data may
+lead to a very different model.
+
+Random forests apply the technique of bootstrap aggregation or bagging to decision trees.  The training data sets are
+independently drawn, with replacement, from the joint distribution of $(X,Y)$ so that each set comprises $n$
+$(p+1)$-tuples $(x_1,y_1),\ldots, (x_n,y_n)$. This is done $B$ times and a tree is grown on each of these $B$ samples.
+The results are then combined in an appropriate way. In a classification problem we take the majority vote for the trees
+over the $q=1,..,Q$ classes,
+\begin{equation*}
+{{h_f}}= \arg \max_q \left(\sum_b I(h(x;B_b)=q)\right).
+\end{equation*}
+In addition to growing the trees
+on bootstrapped samples, each node in the tree is split on a randomly
+selected subset of the variables.
+
+Random forests are a variance reduction technique. They take unpruned tree models, which have a low bias but a high
+variance, and by combining them reduce the variance.  For regression problems, as the prediction error is the sum of the
+variance and the bias squared, this strategy should lead to a low prediction error.
+
+For classifiction problems there is no obvious decomposition of classification error into an equivalent bias plus
+variance form. A number of different additive decompositions of classification error have been considered in an effort
+to understand the success of bagging. They are all less transparent than the regression case (see \cite{Friedman.1997} and
+citations therein).
+
+In the case of a regression tree, the average prediction error for an individual tree $h(X; B)$ is
+\begin{equation}
+PE_t = E_B E_{X,Y} (Y-h(X; B))^2.
+\end{equation}
+Assume that, for all $B$, the tree is unbiased, i.e., $EY= E_X h(X; B)$. Then
+\begin{equation}
+PE_f \leq \rho PE_t
+\end{equation}
+where $\rho$ is the weighted correlation between residuals $(Y-h(X;B))$ and $(Y-h(X;B^\prime))$ for independent
+$B,B^\prime$.
+So what is required for accurate random forest regression is (i) low correlation between residuals of differing tree in
+the forest, and (ii) low prediction error for the individual trees \cite{Segal.2004}.
+
+The benefits of random forest models are:
+\begin{itemize}
+\item they give an out-of-bag (OOB) estimate of model accuracy, by using the samples not selected for the bootstrapped sample
+  as an independent test set for each tree;
+\item they provide a measure of variable importance, for example by summing the number of times a variable is used to
+  split a node, over all nodes and all trees. Other, more complex, measures of variable importance are also used;
+\item they inherit some of the qualities of tree models (but not, for example, the handling of missing values via surrogate
+  variables. This would be difficult as the trees are split on random subsets of variables).
+\end{itemize}
+
+Since the original implementation of random forests, there have been a number of developments leading to a better
+understanding of the algorithm.
+\begin{itemize}
+\item random forests are resistant to overfitting. However it is not true that they will not overfit,
+  ~\cite{Segal.2004} gives examples of overfitting in cases where:
+  \begin{itemize}
+  \item a single tree is the most appropriate model; and
+  \item the variables are highly correlated so that the bootstrap resampling does not give rise to uncorrelated trees.
+  \end{itemize}
+However, keeping these difficult cases in mind, random forests have a strong claim to being resistant to overfitting.
+\item \cite{Strobl.et.al.2007} demonstrates that the random forest algorithm is subject to a bias in variable selection
+  via two mechanisms:
+  \begin{enumerate}
+  \item differences in variable scale (for $X_i$ continuous) and number of levels (for $X_i$ discrete) introduces a
+    bias. An uninformative variable with a large number of levels may be selected over a more informative variable with
+    fewer levels;
+  \item bagging introduces a bias, but sampling $(1- 1/e) \approx 0.632$ of the data without replacement seems to be an
+    effective strategy for removing the problem.
+  \end{enumerate}
+\end{itemize}
+
+\cite{Strobl.et.al.2007} introduce changes to the random forest algorithm to fix 1) and 2).  Effect 1) will be apparent
+in the \cursedforest algorithm as well.  However, there are many examples, particularly in genomics, where we have data
+sets that are:
+\begin{itemize}
+\item very wide;
+\item all variables have the same number of levels (for example, measurements at different points along the genome);
+\item the hypothesized functional relationship is of a form amenable to a decision tree.
+\end{itemize}
+GWAS data sets are an obvious application area for this algorithm.
+
+\cite{Biau.2012} provides a proof that a random forest model will converge at a rate that depends on
+the cardinality of the set $S$ of ``strong predictors'' rather than on the number of variables $p$. That is, given a
+function $y=f(S)$ depending on a set of variables $S$, a random forest will converge to the true function $y$ with a
+rate that depends on the size of $S$ and not the number of noise variables in the dataset. If the size of $S$ is small
+compared to the total number of variables, then the rate of convergence will be quicker. It is not clear what the size
+of the set $S$ will be in a GWAS application. There are cases of small groups of SNPs with large effects and also of
+conditions caused by the interactions of large groups of SNPs.
+
+\cite{Genuer.et.al.2010} and \cite{Diaz.and.Alvarez.2006} describe iterative schemes for doing variable selection using
+the importance measure. See \cite{Chen.and.Ishwaran.2012} for a review of the area with particular reference to genomic
+data.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{\cursedforest (This should go to Results section)}
+VariantSpark made use of Spark's machine learning algorithms ``Spark ML''. While Spark ML algorithms demonstrate
+scalability when dealing with a large number of samples, this scalability eventually breaks down as we include more
+features.
+
+As is standard with Spark applications, we store our data in a Resilient Distributed Dataset (RDD), where an RDD is
+essentially a collection of elements. In the case of Spark ML, each element in the RDD is a sample. RDDs contribute to
+the scalability of Spark as they can be distributed across multiple nodes and operated on in parallel. Even as we add more
+samples to a dataset, Spark can simply schedule extra tasks to handle the additional items in the RDD.
+
+However, within an RDD, Spark ML stores each sample as a vector. Unlike RDDs, which can be partitioned and distributed
+across multiple nodes, each vector must be present in its entirety on any node accessing it. This is no problem with
+typical data sets; however as dimensionality increases, the vectors eventually reach a size where they can no longer fit
+into a single node's memory.
+
+So in the case of adding more samples, Spark ML can simply create more tasks, keeping memory consumption within the
+cluster's bounds. However, as the dimensionality of each sample grows, the memory requirements of the job increases to
+enable these increasingly large vectors to be loaded into memory.
+
+On the other hand, \cursedforest is specifically designed to handle wide ``cursed'' data. It avoids the relation between
+memory and dimensionality by avoiding calculations that rely on entire feature vectors and taking the parallelization
+work down to the level of the individual features.  For each node of a tree, \cursedforest will distribute
+tasks that consist of single features (variants), for every individual.  Each of these tasks will calculate the
+information gain for that specific feature.  Once these tasks have completed, the results are reduced to return the
+feature which gives the greatest information gain.  This process is then repeated until \cursedforest has created the
+entire decision tree.
+
+The current implementation of \cursedforest uses a ``Gini impurity'' criteria for splitting. Let $f_q$ be the fraction
+of items labeled with value $q$, where $q = 1, \ldots, Q$ at a node. The Gini impurity is
+$$
+I_G(f) = \sum_{q = 1}^Q f_q ( 1 - f_q ), 
+$$
+which is at a minimum when all observations at the node are in the same class.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% This section talks about the data sets
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Data Sets}
+\subsubsection{1000 Genomes Project}
+We obtained the 1000 Genomes Project data as VCF files from their FTP site.  Each VCF file contains the variants for
+every individual across one chromosome.  We use the phase 3 dataset which contains 2,504 individuals with approximately
+81 million variant positions (i.e. a matrix of $2,504 \times 81$ million).  For compatibility with the majority of the
+algorithms we demonstrate, we convert the original variant representations to double representations.  For example, we
+store the absence of a variant (0\textbar0) as 0, a heterozygous
+variant (1\textbar0 or 0\textbar1) as 1, and a
+homozygous variant (1\textbar1) as 2, so we have an ``additive'' model.  However, this is not the only possible encoding \cite{Goldstein.et.al.2011}. Note that from
+the encodings listed in Table~\ref{table:encodings}, the additive encoding will present the most difficulty for a random
+forest model.
+
+\begin{table}[!ht]
+%\begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
+\begin{flushleft} 
+\centering
+\caption{\bf Possible encodings of GWAS data \cite{Goldstein.et.al.2011}}
+\begin{tabular}{|l|l|l|l||}
+\hline
+{\bf Type} & {\bf Mechanism}                                         & {\bf Partition} \\ 
+\thickhline
+Additive   & Each additional minor allele increases variation        & 0, 1, 2         \\
+Dominant   & Presence of at least 1 minor allele increases variation & 0, $\{1,2\}$          \\
+Recessive  & Two minor alleles needed for variation                  & $\{0,1\}$, 2          \\
+Heterosis  & Heterozygote leads to variation                         & $\{0,2\}$, 1          \\ \hline
+%\end{adjustwidth}
+\end{tabular}
+\label{table:encodings}
+\end{flushleft}
+\end{table}
+
+\subsubsection{Bone mineral density dataset}
+We obtained a VCF file from the authors of \cite{Duncan.2011}, which contains variant data from 1,936 postmenopausal women.
+This dataset covers approximately 7 million variant positions. We have information about the bone mineral density (BMD)
+for each individual, where a low BMD is a major predisposing factor to fracture \cite{Duncan.2011}.
+
+\subsubsection{Synthetic data} 
+\label{section:synthetic_data}
+ Each dataset consists of $n$ samples and $p$ variables where $n << p$,
+and values for each variable are ordinal variables with three levels represented as numbers $\{0, 1, 2\}$ (which
+correspond to an additive effect encoding of genomic variation) randomly generated from a uniform distribution with equal
+probabilities.
+
+The model parameters are $w_i = 1/\sqrt{2^{i-1}}$ for $i =1,\ldots 5$ and we set 
+\begin{equation}
+\label{equation:synthetic.data}
+z =  \sum_{i=1}^{i=5} {w_i x_i}.
+\end{equation}
+We let $\sigma_\epsilon^2 = Var(z)(1-\theta)/\theta $  where $\theta$ is a parameter controlling the fraction of variance explained by the informative variables and in our study we chose $\theta = 0.125$ .
+Then
+$y =   z + \epsilon$ where $\epsilon \sim  N(0, \sigma_\epsilon^2).$
+The dichotomous response is generated by thresholding $y$ at the $0.5$ quantile,
+
+$$
+\ddot{y} =    \left\{ \begin{array}{rcl}
+			  0 & \mbox{for} & y \leq Q2(y) \\
+			  1 & \mbox{for} & y > Q2(y).
+		 \end{array}\right. 
+$$
+
+\subsection{Parameter settings}
+We consider the parameter settings for the random forest algorithm. We use the R notation from the \randomforest
+package \cite{Liaw.and.Weiner.2002} which incorporates the original Fortran code by Brieman and Cutler.  We incorporate
+the advice of \cite{Liaw.and.Weiner.2002}, which we have found mirrors our own experience:
+\begin{itemize}
+\item \ntree{}  -- the number of trees.  The number of trees necessary for good performance grows with the number of
+  predictors.  \cite{Liaw.and.Weiner.2002} suggest that a high \ntree is necessary to get stable estimates of variable
+  importance and proximity; however, even though the variable importance measures may vary from run to run, we note that
+  it is possible for a random forest model to have a poorer fit and still have an accurate ranking of variable
+  importance;
+\item \mtry{}  -- the number of variables considered at each split (if \mtry=$p$, we have a boosted decision
+  tree model).  If one has a very large number of variables but expects only very few to be ``important'', using larger \mtry may give
+  better performance;
+\item the size and complexity of the individual trees is controlled in \randomforest by setting \texttt{nodesize}, the
+  minimum size of terminal nodes. It is controlled in Spark ML by setting \texttt{maxDepth}, the maximum depth of each
+  tree in the forest.\footnote{The Spark ML documentation \cite{Spark.2016} sets \texttt{maxDepth} to 4 in their
+    classification example. This may give an estimate with a higher bias and, as such, may be a poor choice. See the
+    discussion in \cite{Dietterich.2002} where it is shown that the action of a RF is complicated than first
+    thought. They show that bagging may reduce bias as well as variance but give a better result for low bias
+    learners. \cite{Dietterich.2002} notes that if the bootstrap replicate approximation were correct (i.e.~if the
+    bootstrap sample came from an identical distribution to the data), then bagging would reduce variance without
+    changing bias.}
+\end{itemize}
 

### 2017-08-09 doc/publications/IJCAI2017/IJCAI.tex
tex update

@@
 \usepackage{epstopdf}
-\usepackage{subcaption}
+\usepackage{subcaption}q
 
@@
 
-\title{CursedForest - A Random Forest Implementation for ``Big'' and ``Wide'' Data}
+\title{Breaking the curse of dimensionality for machine learning on genomic data}
 %\section{}
@@
 However, current ML approaches are unable to cope with these data volumes and even implementations using novel compute paradigms, such as Hadoop/Spark, are limited by the finite resource of computer memory.
-Specifically, Spark ML?s supervised machine learning methods suffer from "the curse of dimensionality".
+Specifically, Spark ML's supervised machine learning methods suffer from "the curse of dimensionality".
 That is, they scale well with samples (n) but not with features (p, variants).
@@
 CursedForest was the only implementation scaling to the full dataset, outperforming other methods on smaller subsets by 80\% or more.
-We also demonstrate a GWAS-style analysis using CursedForest?s feature selection approach on an imputed SNP array dataset (2000 x 7 Million) on Bone mineral density (BMD), successfully predicting known BMD-genes as well as identifying novel candidates.
+We also demonstrate a GWAS-style analysis using CursedForest's feature selection approach on an imputed SNP array dataset (2000 x 7 Million) on Bone mineral density (BMD), successfully predicting known BMD-genes as well as identifying novel candidates.
 CursedForest is included in our earlier genome interpretation package, VariantSpark, allowing it to perform near real-time classification of population-scale patient cohorts in "patients like mine" scenarios, as well as performing GWAS analysis on large unfiltered whole genome sequencing cohorts.
